### 2018-06-23 media/fibonacci/proof.tex
added fibonacci post

\documentclass{article}

\usepackage{amsthm}
\usepackage{amsmath}

\title{\huge{\textbf{A Fibonacci Proof}}}
\author{\large{\textit{Marcel Goh}}}
\date{23 June 2018}

\newtheorem*{theorem}{Theorem}
\DeclareMathOperator{\fib}{fib}

\begin{document}

\maketitle

\begin{theorem}
    For all $n\in\mathcal{N}$, the $n$-th number in the Fibonacci sequence $(1,1,2,3,5\ldots\fib(n))$ is the closest integer to $\phi^n/\sqrt{5}$ where $\phi = (1+\sqrt{5})/2$
\end{theorem}
\begin{proof}
    For natural numbers $n$, the Fibonacci sequence is defined as follows:
    $$\fib(n)=\begin{cases}
                  1 & n=1,2\\
                  \fib(n-1)+\fib(n-2) & n\geq3\\
              \end{cases}$$
    The proof consists of two parts. First we will show that $\fib(n)$ differs from $\phi^n/\sqrt{5}$ by $\psi^n/\sqrt{5}$ where $\psi = (1-\sqrt{5})/2$. Then we will prove that this difference is strictly less than 0.5 for all natural numbers $n$. Observe the following two cases:\\ \\
    For n=1:
    $$\frac{(\frac{1+\sqrt{5}}{2})^1}{\sqrt{5}} - \frac{(\frac{1-\sqrt{5}}{2})^1}{\sqrt{5}} = \frac{(1+\sqrt{5})-(1-\sqrt{5})}{2\sqrt{5}} = \frac{\sqrt{5} + \sqrt{5}}{2\sqrt{5}} = 1$$
    For n=2:
    $$\frac{(\frac{1+\sqrt{5}}{2})^2}{\sqrt{5}} - \frac{(\frac{1-\sqrt{5}}{2})^2}{\sqrt{5}} = \frac{(1+2\sqrt{5}+5)-(1-2\sqrt{5}+5)}{4\sqrt{5}} = \frac{2\sqrt{5} + 2\sqrt{5}}{4\sqrt{5}} = 1$$\\
    So if we set $n=3$, we know the following two identities hold:
    $$\fib(n-1) = \frac{\phi^{n-1}-\psi^{n-1}}{\sqrt{5}} \quad \text{and} \quad \fib(n-2) = \frac{\phi^{n-2}-\psi^{n-2}}{\sqrt{5}}$$
    We want to show that for all $n\in\mathcal{N},n\geq3$, $\fib(n) = fib(n-1)+fib(n-2)$.
    \begin{align} \label{eq:1}
        \frac{\phi^n-\psi^n}{\sqrt{5}} &= \frac{\phi^{n-1}-\psi^{n-1}}{\sqrt{5}} + \frac{\phi^{n-2}-\psi^{n-2}}{\sqrt{5}} \nonumber \\
        \phi^2\cdot\phi^{n-2}-\psi^2\cdot\psi^{n-2} &= \phi\cdot\phi^{n-2}-\psi\cdot\psi^{n-2} + \phi^{n-2} - \psi^{n-2} \nonumber \\
        \phi^2\cdot\phi^{n-2}-\psi^2\cdot\psi^{n-2} &= (\phi+1)\phi^{n-2}-(\psi+1)\psi^{n-2}
    \end{align}
    To prove that identity \eqref{eq:1} holds, we must equate the coefficients and prove that $\phi^2 = \phi+1$ and $\psi^2 = \psi + 1$:\\\\
    For $\phi$:
    \begin{align*}
        (\frac{1+\sqrt{5}}{2})^2 &= \frac{1+\sqrt{5}}{2} + 1\\
        \frac{1+2\sqrt{5}+5}{4} &= \frac{1+\sqrt{5}}{2} + \frac{2}{2}\\
        \frac{6+2\sqrt{5}}{4} &= \frac{1+\sqrt{5}+2}{2}\\
        \frac{3+\sqrt{5}}{2} &= \frac{3+\sqrt{5}}{2}
    \end{align*}
    For $\psi$:
    \begin{align*}
        (\frac{1-\sqrt{5}}{2})^2 &= \frac{1-\sqrt{5}}{2} + 1\\
        \frac{1-2\sqrt{5}+5}{4} &= \frac{1-\sqrt{5}}{2} + \frac{2}{2}\\
        \frac{6-2\sqrt{5}}{4} &= \frac{1-\sqrt{5}+2}{2}\\
        \frac{3-\sqrt{5}}{2} &= \frac{3-\sqrt{5}}{2}
    \end{align*}
    So identity \eqref{eq:1} holds and we have proven that $\fib(n)$ differs from $\phi^n/\sqrt{5}$ by $\psi^n/\sqrt{5}$.\\ \\
    Now we will show that this difference is less than $0.5$. Concretely, this means that for all $n\in\mathcal{N}$, $\left|\phi^n/\sqrt{5}\right| < 0.5$. $\psi \approx -0.618$ so if $n=1$, we have $\psi/\sqrt{5} \approx -0.276\ldots$, the absolute value of which is less than $0.5$. And because $\left|\psi\right| < 1$, $\left|\psi^{n+1}\right| < \left|\psi^n\right|$ and for all $n\in\mathcal{N}$, $\left|\psi^n/\sqrt{5}\right| < 0.5$.\\ \\
    Therefore, for all natural numbers $n$, $\fib(n)$ differs from $\phi^n/\sqrt{5}$ by less than 0.5 and the theorem is proved.
\end{proof}

\end{document}

### 2019-06-09 misc/study-notes/nopt048notes.tex
Added linear programming notes

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amsthm,amssymb,amsfonts,tikz,tikzsymbols}
\usepackage[shortlabels]{enumitem}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}

\newcommand{\M}{\mathbb{M}}

\newcommand{\F}{\mathcal{F}}
\newcommand{\pp}{\mathcal{P}}

\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\conv}{conv}
\DeclareMathOperator{\cone}{cone}
\DeclareMathOperator{\aff}{aff}

\renewcommand\qedsymbol{$\Smiley$}

\newenvironment{problem}[2][Problem]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
\newenvironment{definition}[2][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem*{theorem*}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem*{lemma*}{Lemma}

\begin{document}

\title{NOPT048 Notes}
\author{Marcel Goh}
\maketitle

\section{Definitions and theorems}
\subsection{Linear and integer programs}
The following is a \textbf{linear program}:
\begin{alignat*}{2}
    & \text{maximise} & \quad & c^Tx \\
    & \text{subject to} & & Ax \leq b
\end{alignat*}
Here $x, b, c\in \R^n$ and $A$ is an $m\times n$ matrix over $\R$. The function $c^Tx$ is called the \textbf{objective function} and it is what the linear program is trying to maximise (or minimise). Any vector $x$ that satisfies the inequality $Ax \leq b$ is called a \textbf{feasible solution}. A vector $x^*$ is called an \textbf{optimum solution} if $c^Tx^* \geq c^Tx$ for all feasible solutions $x$. A linear program is called \textbf{infeasible} if there are no feasible solutions. Even if a linear program is feasible, it might be \textbf{unbounded}. This happens when its objective function can be made arbitrarily "good".\\
\indent In a certain sense, linear programs are easy to solve. Adding an integrality constraint to a linear program gives an \textbf{integer program}:
\begin{alignat*}{2}
    & \text{maximise} & \quad & c^Tx \\
    & \text{subject to} & & Ax \leq b\\
    &                   & & x \in \Z^n
\end{alignat*}
Integer programs are much harder to solve in the general case. Suppose we want to solve the following integer program:
\begin{alignat*}{2}
    & \text{maximise} & \quad & c^Tx \\
    & \text{subject to} & & Ax \leq b\\
    &                   & & x \in \{0,1\}
\end{alignat*}
It is often a good strategy to solve the \textbf{linear program relaxation} of the integer program, which gives an approxmiate answer:
\begin{alignat*}{2}
    & \text{maximise} & \quad & c^Tx \\
    & \text{subject to} & & Ax \leq b\\
    &                   & & 0\leq x\leq 1
\end{alignat*}
Then we can take the solution of the linear program and round each element to the nearest integer value to get a solution of the original integer program. In some cases, the linear program relaxation gives quite a good answer. For example, if $G$ is a graph with minimum vertex cover of size $n$, then formulating an integer program and solving its linear program relaxation will spit out a vertex cover $S$ with $|S|\leq 2n$.

\subsection{Geometric interpretation}
The inequality $Ax\leq b$ defines a \textbf{polyhedron}. More precisely, we say that a polyhedron is the intersection of finitely many \textbf{halfspaces}, where a halfspace is the set $\{x : a^Tx\leq b\}$ for some $a\in \R^n$ and $b\in \R$. If a polyhedron is bounded, we call it a \textbf{polytope}. A polyhedron is closed and convex. The finite intersection of closed sets is closed and the intersection of convex sets is convex.\\
\indent Let $S\subset \R^n$ be a a set of $m$ points. Then the \textbf{convex hull} of $S$ is the set
$$\conv S = \{x : x = \sum_{i=1}^m\lambda_i s_i, \sum_{i=1}^m\lambda_i = 1, \lambda_i \geq 0\}.$$
The \textbf{conic hull} of $S$ is the set
$$\cone S = \{x : x = \sum_{i=1}^m\lambda_i s_i, \lambda_i\geq 0\}.$$
The \textbf{affine hull} of $S$ is the set
$$\aff S = \{x : x = \sum_{i=1}^m\lambda_i s_i, \sum_{i=1}^m\lambda_i = 1\}.$$
The convex hull can be obtained from the conic hull by intersecting it with a hyperplane. The affine hull is the smallest-dimensional subspace that contains all the points in $S$.\\
\indent Let $P = \{x: Ax\leq b\}$ be a polyhedron. An inequality $\alpha^Tx \leq \beta$ is said to be a \textbf{valid inequality} if and only if for all $y\in P$, $\alpha^T y\leq \beta$. Then a subset $F\subset P$ is a \textbf{face} of $P$ if and only if there exists a valid inequality $\alpha^T x\leq \beta$ such that $F = P\cap \{x : \alpha^T x\leq \beta\}$. The trivial faces of any polyhedron $P$ are $\emptyset$ and $P$ itself (formed by intersecting $P$ with the valid inequalities $0^Tx \leq 1$ and $0^Tx\leq 0$ respectively). Any face that is not trivial is called \textbf{proper}.\\
\indent The \textbf{dimension} of a polyhedron $P$, denoted $\dim P$, is the largest dimension of a ball inside $P$. A \textbf{vertex} is a face of dimension 0, an \textbf{edge} is a face of dimension 1, and a \textbf{facet} is a face of dimension $\dim P - 1$. A point $x\in P$ is called an \textbf{extreme point} if it is not a convex combination of points of $P$. Note that a point $x$ is an extreme point if and only if it is a vertex.
\begin{theorem*}{\textup{(Minkowski-Weyl)}}
    A set of points $P$ is a polyhedron if and only if there exist finite sets $V$ and $Y$ such that $P = \conv V + \cone Y$ where the sum $A + B$ denotes the set $\{a + b : a\in A, b\in B\}$.
\end{theorem*}

\subsection{Simplex method}
A polyhedron is \textbf{pointed} if it does not contain a line.
\begin{lemma*}
    Every non-empty pointed polyhedron $P$ has at least one vertex.
\end{lemma*}
\begin{theorem*}
    Consider the linear program
    \begin{alignat*}{2}
        & \textup{maximise} & \quad & c^Tx \\
        & \textup{subject to} & & Ax \leq b.
    \end{alignat*}
    If the polyhedron $P$ defined by $Ax\leq b$ is pointed and if an optimum exists, then there exists a vertex of $P$ that attains the maximum.
\end{theorem*}
The result of the above statements is an algorithm for solving linear programs, called the \textbf{simplex method}. Here is the general idea:
\begin{enumerate}
    \item Find an initial vertex. It is a 0-dimensional face with some 1-dimensional faces incident on it.
    \item At any vertex, check for optimality. If optimal, we are done.
    \item If the vertex is not optimal, increase the value of the objective function by following an edge to a new vertex. Return to step 2.
\end{enumerate}
\indent Let $A$ be an $m\times n$ matrix, $b\in \R^m$, $x \in\R^n$, with $n \geq m$. Let $B\subset [n] =\{1,\ldots,n\}$ of cardinality $m$. Then $A_B$ denotes the square matrix formed by selecting the columns of $A$ indexed by the elements of $B$ and likewise $x_B$ denotes the vector of length $m$ formed by selecting the elements indexed by the elements of $B$. We define a vector $x$ to be a \textbf{basic feasible solution} of the linear program
\begin{alignat*}{2}
    & \textup{maximise} & \quad & c^Tx \\
    & \textup{subject to} & & Ax \leq b
\end{alignat*}
if and only if there exists $B\subset [n]$ with $|B|= m$ such that:
\begin{enumerate}[i)]
    \item $A_B$ is non-singular
    \item $x_j = 0$ for all $j\notin B$
    \item $x_B = A_B^{-1}b$
\end{enumerate}
A set $B\subset [n]$ is called a \textbf{basis} if and only if $A_B$ is non-singular, and $B$ is a \textbf{feasible basis} if the unique solution of $A_Bx_B=b$ is non-negative. Note that a basic feasible solution is uniquely determined by the feasible basis $B$, but a vertex does \textit{not} have to be determined by a unique basis.\\
\indent A matrix $A$ is \textbf{totally unimodular} if and only if every square submatrix of $A$ has determinant 0, 1, or $-1$.

\subsection{Duality}
As always, we start with the linear program
\begin{alignat*}{2}
    & \textup{maximise} & \quad & c^Tx \\
    & \textup{subject to} & & Ax \leq b\\
    &                     & & x \geq 0.
\end{alignat*}
Inequalities of a linear program can be added together without changing the feasible region. Likewise, one can also multiply any inequality by a non-negative scalar. So we can manipulate inequalities to get an upper bound on the optimum value of the objective function. We want to pick coefficients to get the \textit{least} upper bound possible. Let $y$ denote a vector of coefficients. We want to
\begin{alignat*}{2}
    & \textup{minimise} & \quad & b^Ty \\
    & \textup{subject to} & & A^Ty \geq c\\
    &                     & & y \geq 0
\end{alignat*}
The original linear program is called the \textbf{primal} and this new one is called the \textbf{dual}. Any feasible solution of the dual linear program gives an upper bound on the primal and the optimum of the dual is equal to the optimum of the original program.

\begin{theorem*}{\textup{(Weak Duality Theorem)}}
    Let $P$ and $D$ be primal and dual linear programs of the above form and let $x$ and $y$ be respective feasible solutions. Then $c^Tx \leq b^Ty$.
\end{theorem*}

\begin{theorem*}{\textup{(Strong Duality Theorem)}}
    Let $P$ and $D$ be primal and dual linear programs. Then exactly one of the following holds:
    \begin{enumerate}[i)]
        \item P infeasible, D infeasible
        \item P infeasible, D unbounded
        \item P unbounded, D infeasible
        \item P and D both feasible and bounded and furthermore, if $x^*$ and $y^*$ are primal and dual optimal solutions, then $c^Tx^* = b^Ty^*$.
    \end{enumerate}
\end{theorem*}

\begin{lemma*}{\textup{(Farkas lemma)}}
    For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
    \begin{enumerate}[i)]
        \item There exists a vector $x$ such that $Ax = b$ and $x \geq 0$.
        \item There exists a vector $y$ such that $A^Ty \geq 0$ and $b^Ty< 0$.
    \end{enumerate}
\end{lemma*}

\begin{theorem*}{\textup{(Farkas lemma 2.0)}}
    For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
    \begin{enumerate}[i)]
        \item There exists a vector $x$ such that $Ax \leq b$ and $x \geq 0$.
        \item There exists a vector $y$ such that $A^Ty \geq 0$, $y\geq0$ and $b^Ty< 0$.
    \end{enumerate}
\end{theorem*}

\subsection{Applications}
\subsubsection{Matching and vertex cover in bipartite graphs}
Given a graph $G=(V,E)$, a \textbf{matching} is a set $M\subset E$ such that for all $v\in V$ there is at most one edge $e\in M$ with $v\in e$. The linear program that finds the maximal matching in a bipartite graph is
\begin{alignat*}{2}
    & \textup{maximise} & \quad & \sum_{e\in E} x_e \\
    & \textup{subject to} & & \sum_{v\in e}x_e \leq 1,\ \text{for all}\ v\in V\\
    &                     & & x_e \geq 0,\ \text{for all}\ e\in E.
\end{alignat*}
The dual of this linear program is the following:
\begin{alignat*}{3}
    & \textup{minimise} & \quad & \sum_{v\in V} y_v &\\
    & \textup{subject to} & & y_u + y_v \geq 1,\  \text{for all}\ \{u,v\}\in E\\
    &                     & & y_v \geq 0, \  \text{for all}\ v\in V.
\end{alignat*}
    Given a graph $G=(V,E)$, a \textbf{vertex cover} is a set $S \subset V$ such that for all $\{u, v\}\in E$, $\{u,v\}\cap S\geq 1$. The dual linear program above finds the minimum vertex cover in a bipartite graph. The duality of these two programs results in K\H{o}nig's theorem.
\begin{theorem*}{\textup{(K\H{o}nig's theorem)}}
    Suppose $G=(V,E)$ is a bipartite graph. Let $M\subset E$ denote the maximum matching and let $S\subset V$ denote the minimum vertex cover. Then $|M|=|S|$.
\end{theorem*}

\subsubsection{Matching in weighted graphs}
Let $G=(V,E)$ be a graph (not necessarily bipartite) with a weight function $w:E\rightarrow \R^+$. Suppose we want to find the maximum weight matching in $G$. For any subset $E'\subset E$, the \textbf{characteristic vector} of $E'$ is the vector $\chi^{E'}$ given by
$$ \chi_e^{E'} = \begin{cases} 1 & e\in E' \\ 0 & e\notin E'\end{cases} $$
for all $e\in E$. Then the \textbf{matching polytope} of the graph $G$ is the set
$$ M(G) = \conv\{ x \in \{0,1\}^{|E|} : x\ \text{is the characteristic vector of some matching in}\ G\}.$$
To get the maximum weight matching, we need only solve the linear program
\begin{alignat*}{2}
    & \textup{maximise} & \quad & w^Tx \\
    & \textup{subject to} & & x\in M(G).
\end{alignat*}
What does the polytope $M(G)$ look like in terms of inequalities? It turns out that $M(G)$ is the set of all vectors $x$ that satisfy
\begin{enumerate}
    \item $\displaystyle\sum_{v\in e}x_e \leq 1$ for all $v\in V$ and
    \item $\displaystyle\sum_{e\in E[U]}x_e \leq \frac{|U|-1}{2}$ for every set $U\subset V$ of odd cardinality.
\end{enumerate}
In a graph $G=(V,E)$, a matching $M\subset E$ is \textbf{perfect} if every vertex in $V$ is incident to some edge in $M$. The perfect matching polytope is the set
$$ PM(G) = \conv\{ x \in \{0,1\}^{|E|} : x\ \text{is the characteristic vector of some perfect matching in}\ G\}.$$
For a set $U\subset V$ of vertices, let $\delta(U)$ denote the number of edges that have one endpoint in $U$ and one endpoint outside $U$. Then we find that $PM(G)$ is the set of all vectors $x$ satisfying
\begin{enumerate}
    \item $\displaystyle\sum_{v\in e}x_e = 1$ for all $v\in V$,
    \item $\displaystyle\sum_{e\in \delta(U)}x_e\geq 1$ for every set $U\subset V$ of odd cardinality, and
    \item $x_e \geq 0$ for every $e\in E$.
\end{enumerate}
Note that the number of inequalities in these polytopes may be very large. For this reason, a polynomial-time algorithm may not exist.

\subsubsection{The travelling salesman problem}
Given a graph $G=(V,E)$, a \textbf{closed tour} is a walk in $G$ with no repeated edges and no repeated vertices except for the first and last vertex, which are the same. In a graph with edge costs $(c_e : e\in E)$, \textbf{travelling salesman problem} asks us to find the minimum-weight closed tour that contains each vertex exactly once. To solve the travelling salesman problem, we start with the following linear program:
\begin{alignat*}{2}
    & \textup{minimise} & \quad & \sum_{e\in E} c_ex_e \\
    & \textup{subject to} & & \sum_{v\in e} = 2\ \text{for all}\ v\in V,\\
    &  & & 0\leq x_e \leq 1\ \text{for all}\ e\in E.
\end{alignat*}
What we really want to do is add the inequalities $\displaystyle\sum_{e\in \delta(S)} x_e \geq 2$ for every $\emptyset \neq S \neq V$. This is called \textbf{subtour elimination} because it forbids any tours smaller than the entire closed tour we're trying to build. Adding this inequality, we get the linear program relaxation of the integer program for the travelling salesman problem. The problem is that eliminating subtours adds an exponential number of ineqalities to the program.\\
\indent Instead of adding all of these inequalities, we use the cutting-plane approach. Given a polytope $P$, a \textbf{cutting plane} is an inequality that is valid for all integer points in $P$. It is called a cutting plane because it may cut the original polytope. We can apply this idea as follows:
\begin{enumerate}
    \item We start with the linear program without the subtour constraints.
    \item Find the optimal solution to the linear program.
    \item If we find that some subtour constraints are violated by the optimal solution, we add these cutting planes to our program and return to step 2. If no subtour constraints are violated, we have a solution and we may stop.
\end{enumerate}
Step 3 is our main concern. How do we figure out if there are any violated subtour constraints? Suppose we are at a certain step of the algorithm and the linear program has just given us an optimal solution $x^*$. Let $G(x^*)$ denote the graph with vertices $V$ and edges $\{e : x_e^* > 0\}$. Without loss of generality, assume $G(x^*)$. Set capacities $u_e = x_e^*$ for every edge $e$. Now for any $S\subset V$, the weight of the cut $(S, V\setminus S)$ equals the value of $\displaystyle\sum_{e\in \delta(S)} x_e$. So we can solve the min-cut problem over $G(x^*)$ and if we get a value less than 2, we know that some subtour constraint is violated. Of course, we need some heuristics to decide which constraints to add, so there is still no guarantee of polynomial-time termination.

\subsection{Ellipsoid algorithm}
So far, we do not know if linear programs can be solved in polynomial time. It turns out that optimisation is polynomially equivalent to separating over a polytope. Given $x\in \R^n$ and a polyhedron $P$, we separate as follows: Output "yes" if $x\in P$; otherwise, output "no" and a pair $(\alpha, \beta)$ such that $\alpha^Ty\leq\beta$ is a valid inequality for $P$ and $\alpha^T x > \beta$. The \textbf{ellipsoid algorithm} uses an oracle for separation to answer the question of feasibility: Given a polytope $P$, find an $x\in P$ or conclude that $P = \emptyset$.\\
\indent Before we get into the actual method, let's convince ourselves that checking for feasibility is equivalent to optimising. That optimisation implies feasibility is obvious. So let's suppose we have a method to check if a linear program is feasible. Can we use that to optimise the linear program
\begin{alignat*}{2}
    & \textup{minimise} & \quad & c^Tx\\
    & \textup{subject to} & & Ax \leq b\\
    &  & & x\geq 0?
\end{alignat*}
Well, from duality we know that if an optimum exists, it will also be the optimum value of the dual linear program. So we can construct a polyhedron that contains only this value and then check if it is feasible. So we check if the polyhedron
$$\{\begin{bmatrix} x\\y\end{bmatrix} : c^Tx = b^Ty, Ax \leq b, A^Ty \geq c, x \geq 0, y\geq 0\}$$
is feasible and that will give us the optimum value of $x$, if it exists.\\
\indent On to ellipsoids. An \textbf{ellipsoid} $E(A,a)$ is the set $\{x \in \R^n : (x - a)^TA^{-1}(x-a) \leq 1\}$ where $A$ is a \textbf{positive definite} matrix. This means that $(x-a)^TA^{-1}(x-a)$ is positive and moreover, $A = M\cdot M$ for some square matrix $M$. If we let $B$ denote the unit ball in $\R^n$, then $E(A,a) = \{Mx + a : x \in B\}$.\\
\indent For simplicity, assume that $P$ is bounded and assume that $P$ is either empty or full-dimensional. Assume also that we have an oracle that outputs a separating hyperplane. Then the \textbf{ellipsoid algorithm} determines if $P = \{x : Ax \leq b\}$ is feasible.
\begin{enumerate}
    \item Start with an ellipsoid large enough that it definitely contains a point in the polytope, if one exists.
    \item Let $c$ be the centre of the ellipsoid. Invoke the separation oracle. If $c\in P$, output $c$ and halt.
    \item If we have reached the maximum amount of iterations, output $P = \emptyset$.
    \item Otherwise, the oracle gives us a separating hyperplane that cuts the ellipsoid. Determine which half of the ellipsoid $P$ lies, then draw a new, smaller ellipsoid that contains this half. Return to step 2.
\end{enumerate}
The maximum number of iterations is calculated using a function polynomial in the size of the linear program, so this algorithm runs in polynomial time.

\subsection{Matroids and greedy algorithms}
Given a graph $G=(V,E)$, a \textbf{spanning tree} is a set $T\subset E$ such that every vertex is incident to some edge in $T$ and $G=(V,T)$ is a contains no cycles. Kruskal's algorithm obtains the minimum weight spanning tree by maintaining a forest $F$. If $F$ spans every vertex, then it is a spanning tree. If not, add the minimum weight edge $e$ such that $F\cup \{e\}$ does not contain a cycle and repeat.\\
\indent Let $G=(V,E)$ be a graph and let $F$ denote the set of all forests in $G$. Then $(E,F)$ is called the \textbf{graphic matroid} of $G$. In fact, there are lots of different types of matroids. A tuple $M=(E,I)$ where $I\subset \mathcal{P}(E)$ is a \textbf{matroid} if
\begin{enumerate}[i)]
    \item $\emptyset\in I$,
    \item If $X\subset Y$ and $Y\in I$, then $X\in I$,
    \item If there exist $X,Y \in I$ such that $|Y| > |X|$, then there exists some $y\in Y\setminus X$ such that $X\cup\{y\}\in I$.
\end{enumerate}
A system that satisfies i) and ii) is called an \textbf{independence system}. Associated with any independence system $M = (E,I)$ is its \textbf{rank function} $r : \mathcal{P}(E)\rightarrow \R$ defined by
$$r(X) = \max\{ |Y| : Y \subset X, Y\in I\}.$$
We call $r(E)$ the \textbf{rank} of $M$. A set $S\in I$ such that $|S| = r(E)$ is called a \textbf{base} of $M$. If $M=(E,I)$ is a matroid, then some properties of its rank function $r$ include:
\begin{enumerate}
    \item $0\leq r(X) \leq |X|$ for all $x\subset E$.
    \item If $X\subset Y$ then $r(X)\leq r(Y)$ for all $X, Y\subset E$.
    \item $r(X\cup Y)+r(X\cap Y) \leq r(X) + r(Y)$ for all $X,Y\subset E$.
\end{enumerate}
Property 3 above is called \textbf{submodularity}. If $r$ is the rank function of a general independence system, it need only satisfy properties 1 and 2, as well as the weaker property $r(X\cup Y) \leq r(X) + r(Y)$.\\
\indent Matroids can be directly characterised by the rank function.
\begin{theorem*}{\textup{(Rank characterisation of matroids)}}
    Let $E$ be a finite set and suppose that $r : \mathcal{P}(E)\rightarrow \R$ satisfies the three peoperties of the rank function. Then $M=(E,I)$, with
    $$I = \{Y \subset E : |Y| = r(Y)\},$$
    is a matroid with $r$ as its rank function.
\end{theorem*}
To further understand the link between spanning trees and matroids, note that Kruskal's algorithm is an instance of the \textbf{weighted greedy algorithm}. Let $M=(E,I)$ be a matroid with a weight function $c$ defined on edges. The algorithm proceeds as follows:
\begin{enumerate}
    \item Set $S_0\gets\emptyset$, and $E$ is the element set of $M$. Set $i\gets 0$.
    \item Select $e\in E$ with maximum $c(e)$. Set $E\gets E\setminus\{e\}$.
    \item If $S_i\cup\{e\}\in I$, set $S_{i+1}\gets S_i\cup\{e\}$, $i\gets i + 1$.
    \item If $E = \emptyset$, output $S_i$. Otherwise, return to step 2.
\end{enumerate}
Then the following two theorems establishes a certain equivalence between matroids and problems that can be solved with the greedy algorithm.
\begin{theorem*}
    Let $M=(E,I)$ be a matroid with rank function $r$. The weighted greedy algorithm finds the maximum weighted independent set $S_k$ of cardinality $k$ for each $1\leq k\leq r(E)$.
\end{theorem*}
\begin{theorem*}{\textup{(Greedy equivalence)}}
    Let $M=(E,I)$ be an independence system. Then if the weighted greedy algorithm works for all (non-negative) weights, then $M$ is a matroid.
\end{theorem*}
\indent It is worth it to note that spanning trees and matroids have corresponding polytopes. For a graph $G=(V,E)$, the spanning tree polytope $ST(G)$ is the set of all vectors $x\in \R^{|E|}$ that satisfy
\begin{enumerate}
    \item $x_e \geq 0$,
    \item $\displaystyle\sum_{e\in E} x_e = |V| - 1$, and
    \item $\displaystyle\sum_{e\in E[S]} x_e \leq |S| - 1$, for all $S\subsetneq V, |S| \geq 2$.
\end{enumerate}
For a matroid $M=(E,I)$ with rank function $r$, the matroid polytope $P_M$ is the set of all vectors $x\in \R^{|E|}$ satisfying $\displaystyle\sum_{e\in T} x_e \leq r(T)$ for all $T\subset E$, where $x_e\geq 0$.

\section{Problems and proofs}
\begin{problem}{1}
    Describe how to find an initial basic feasible solution for an arbitrary LP in standard (equational) form.
\end{problem}
\begin{proof}[Solution]
    Suppose the constraints are $Ax = b, x\geq 0$. Examine the matrix $A$. We want to create an identity submatrix. If one already exists, we use that as an initial basis. If not, for any row that doesn't have a unique variable with coefficient 1, introduce a new variable. So we introduce new variables $y_1,\ldots,y_k$. Then all these new variables form an initial basis of an auxiliary linear program. We minimise the sum over all $y_i$. Optimising over this minimisation linear program gives us a basis for our original linear program.
\end{proof}

\begin{problem}{2}
    Write an integer program formulation for the matching/perfect matching problem.
\end{problem}
\begin{proof}[Solution]
    The integer program corresponding to the matching problem in a graph $G=(V,E)$ with a cost vector $c$ corresponding to edges.
    \begin{alignat*}{2}
        & \textup{maximise} & \quad & c^Tx\\
        & \textup{subject to} & & \sum_{v\in e} x_e \leq 1\ \text{for all}\ v\in V\\
        &  & & x_e\in\{0,1\}.
    \end{alignat*}
    For a perfect matching, change the inequality for each vertex to equality.
\end{proof}

\begin{problem}{3}
    Prove that the linear program relaxation of the standard integer program for matching/perfect matching is integral for bipartite graphs.
\end{problem}
\begin{proof}
    Let $LP(G,c)$ denote the linear program relaxation of the perfect matching integer program over a graph $G=(V,E)$ with weights $c$. If $LP(G,c)$ is infeasible, then $G$ has no perfect matching. For a vector $x$, let $I(x)$ denote the set $\{i:x_i\in\{0,1\}\}$. If $LP(G,c)$ is feasible, there exists at least one optimal solution. Let $x^*$ an optimal solution that maximises the number of integer elements, i.e.\@ $|I(x^*)| \geq |I(x)|$ for all optimal solutions $x$. We want to show that $|I(x^*)|=|E|$.\\
    \indent Suppose, towards a contradiction, that $|I(x^*)| < |E|$. Then there exists an $i$ such that $0<x_i<1$. It must correspond to some edge $e_1 = \{a_1, b_1\}$ in the graph $G$. Since the sum at each vertex is 1, there must be another edge $e_2$ leaving $b_1$ with $x^*_{e_2}$ non-integral. Continue adding edges in this way. Because the graph has finitely many edges, we must eventually hit a vertex we've seen before and we have a cycle $C\subset E$ in which $0< x^*_e <1$ for all edges. Since the graph $G$ is bipartite, the length $t$ of the cycle $C$ is even. To simplify notation let's rename the edges of $C$ to $e_1,e_2,\ldots,e_t$. Now for a small real number $\epsilon$ define a vector $x'$ by
    \[
        x'_e =
        \begin{cases}
            x^*_e - \epsilon & \text{for}\ e\in\{e_1,e_3,\ldots,e_{t-1}\}\\
            x^*_e + \epsilon & \text{for}\ e\in\{e_2,e_4,\ldots,e_t\}\\
            x^*_e & \text{otherwise}
        \end{cases}
    \]
    It's easy to see that this new $x'$ still maintains the sum of 1 at each vertex. Now we just have to determine how to pick $\epsilon\in[0,1]$. Let $\epsilon$ be the smallest number that causes one of the $x^*_e$ to hit 0 or 1. Then $x'$ has one more element that is integral compared to $x^*$, contradicting maximality of $I(x^*)|$. So $I(x^*)|=|E|$.
\end{proof}

\begin{problem}{4}
    The following is the integer program for vertex cover in a graph $G=(V,E)$:
    \begin{alignat*}{2}
        & \textup{minimise} & \quad & \sum_{v\in V}x_v\\
        & \textup{subject to} & & x_u + x_v \geq 1\ \text{for all}\ \{u,v\}\in E\\
        &  & & x_v\in\{0,1\}\ \text{for all}\ v\in V.
    \end{alignat*}
    Prove that changing $x_v\in \{0,1\}$ to $0\leq x_v\leq 1$ and then solving the resulting linear program results in a 2-approximation.
\end{problem}
\begin{proof}
    Suppose that $x$ is the optimum feasible solution of the linear program relaxation. Define $S=\{v\in V: x_v^* \geq \frac{1}{2}\}$. $S$ is a vertex cover. We want to show that $|S| \leq 2\cdot|S^*|$. Let $S^*$ be the actual optimal vertex cover and $x^*$ its characteristic vector. Then
    $$ |S| = \sum_{v\in S} 1 \leq 2\cdot\sum_{v\in V} x_v \leq 2\cdot\sum_{v\in V}x^*_v = 2\cdot|S^*|,$$
    which is what we wanted.
\end{proof}

\begin{problem}{5}
    Prove the max-flow min-cut theorem via duality.
\end{problem}
\begin{proof}
    Given a directed graph $G=(V,E)$ with capacities $c$ for each edge, source vertex $s$ and sink vertex $t$. Let $P$ be the set of all paths from $s$ to $t$. Then we can formulate the maximum flow problem as the following linear program:
    \begin{alignat*}{2}
        & \textup{maximise} & \quad & \sum_{p\in P} x_p\\
        & \textup{subject to} & & \sum_{e\in p} x_p \leq c(e)\ \text{for all}\ e\in E\\
        &  & & x_p \geq 0\ \text{for all}\ p\in P.
    \end{alignat*}
    Now we construct the dual. We have one dual variable $y_e$ for every edge $e\in E$.
    \begin{alignat*}{2}
        & \textup{minimise} &\quad & \sum_{e\in E} c(e)y_e\\
        & \textup{subject to} & & \sum_{e\in p} y_e\geq 1\ \text{for all}\ p\in P\\
        &  & & y_e\geq 0\ \text{for all}\ e\in E.
    \end{alignat*}
    This is the linear program relaxation of the minimum cut integer program. By the Strong Duality Theorem, the optimal value of the objective functions of these two linear programs must be equal, so the maximum flow equals the value of the minimum cut.
\end{proof}

\begin{problem}{6}
    Prove K\H{o}nig's theorem via duality.
\end{problem}
\begin{proof}
    Let $G=(V,E)$ be a bipartite graph with $m$ edges and $n$ vertices. Consider the integer program
    \begin{alignat*}{2}
        & \textup{maximise} & \quad & \sum_{j=1}^m x_j\\
        & \textup{subject to} & & Ax \leq 1\\
        &  & & x\geq 0\\
        &  & & x\in \Z^m.
    \end{alignat*}
    where $A$ is the incidence matrix of $G$. Because each row can sum to at most 1 and the incidence matrix of a graph contains only 0s and 1s, we discover that $x_j\in \{0,1\}$ for all $j$ and that the edges $e_j$ with $x^*_j = 1$ in an optimal solution $x^*$ form a maximum matching in $G$.\\
    \indent Now consider the integer program
    \begin{alignat*}{2}
        & \textup{maximise} & \quad & \sum_{i=1}^n y_i\\
        & \textup{subject to} & & A^Ty \geq 1\\
        &  & & y\geq 0\\
        &  & & y\in \Z^n.
    \end{alignat*}
    where again, $A$ is the incidence matrix of $G$. In this integer program, every row of $A^T$ corresponding to edge $e_j$ must be at least 1. So in any optimal solution $y^*$ we have $y^*_i\in \{0,1\}$, since any larger value could be decreased to 1. Then the vertices $v_i$ with $y^*_i = 1$ form a minimum vertex cover of $G$.\\
    \indent In both integer programs we may drop the integrality constraints, without affecting the optima, because $A$ is unimodular and by Problem 8, $b$ being integral implies that any optimum vector is as well. The resulting linear programs are dual to one another, so their optimal values are equal. Hence the size of a maximum matching equals the size of a minimum vertex cover in $G$.
\end{proof}

\begin{problem}{7}
    Let $G=(V,E)$ be a graph and with $n$ vertices $v_1,\ldots,v_n$ and $m$ edges $e_1,\ldots,e_m$. The incidence matrix of $G$ is an $n\times m$ matrix $A$ such that $A_{ij} = 1$ if $v_i\in e_j$ and 0 otherwise, for $1\leq i\leq m$ and $1\leq j\leq n$. Prove that if $G$ is bipartite, its incidence matrix is totally unimodular.
\end{problem}
\begin{proof}
    Let $G=(X\cup Y, E)$ be a bipartite graph ($X\cap Y = \emptyset$) and let $A$ denote its incidence matrix. We want to show that every $k\times k$ submatrix $Q$ of $A$ has determinant 0, 1, or $-1$. We proceed by induction on $k$. The base case $k=1$ is easy, because all the entries of an incidence matrix are 0 or 1.\\
    \indent Now consider $k>1$ and a $k\times k$ submatrix $Q$. Since the columns of $Q$ correspond to edges, each column of $Q$ has at most two nonzero entries (which are 1). If there is a column with only zero entries, we get $\det Q = 0$, and if there is a column with only one nonzero entry, we can expand the determinant on this column and get that up to sign, $\det Q$ equals the determinant of a $(k-1)\times(k-1)$ submatrix.\\
    \indent The last case is if every column of $Q$ has exactly two 1s. Here we claim that $\det Q = 0$. To see this, observe that the sum of all rows of $Q$ corresponding to vertices in $X$ is the row vector $(1,\ldots,1)$, since for each column of $Q$, exactly one of its two 1s comes from a vertex in $X$. For the same reason, we get $(1,\ldots,1)$ by summing up the rows for vertices in $Y$, and it follows that the rows of $Q$ are linearly dependent.
\end{proof}

\begin{problem}{8}
    Prove that a polyhedron described by a totally unimodular matrix and integral right-hand side is integral.
\end{problem}
\begin{proof}
    Suppose that $P = \{x\in R^n : Ax\leq b\}$ for a totally unimodular matrix $A$ and vector $b$ with integer elements. Let $x$ be a vertex of $P$. Then there exists some $B\subset [n]$ such that $A_B$ is non-singular, $x_i = 0$ for all $i\notin B$, and $x_B = A_B^{-1}b$. By Cramer's rule, all the entries of $A_B^{-1}$ can be written as rational numbers with common denominator of $\det(A_B)$. But since $A$ is a non-singular unimodular matrix, $\det A_B^{-1} \in\{-1,1\}$, so every element of $A_B^{-1}$ is integral. So $x_B$ has integral elements and $x$ has integral non-zero elements, which is what we wanted.
\end{proof}

\begin{problem}{9}
    Prove that an independence system is a matroid if and only if the weighted greedy algorithm produces a correct result for all non-negative objective functions.
\end{problem}
\begin{proof}
    First we prove the "if" direction. Let $M=(E,I)$ be an independence system. We proceed by contraposition. We assume $M$ is not a matroid and show that there exists some non-negative objective function for which the greedy algorithm fails. So suppose that there exist $X, Y\in I$ such that $|Y|>|X|$ and for all $e\in Y\setminus X$, $X\cup \{e\}\notin I$. Let $c$ be a weight function given by
    \[
        c =
        \begin{cases}
            1+\epsilon & \text{if}\ e\in X\\
            1 & \text{if}\ e\in Y\setminus X\\
            0 & \text{otherwise}
        \end{cases}
    \]
    with $\epsilon$ to be determined. The Greedy Algorithm should choose a maximum-weight independent set with cardinality $|Y|$. But instead, the algorithm chooses all of $X$, then it completes $X$ to an independent set $X'$ of cardinality $|Y|$ by choosing 0-weight elements, for a total weight of $|X|(1+\epsilon)$. Now if we set $\epsilon<\frac{1}{|E|}$, then $c(X') < |X| + 1 \leq |Y| \leq c(Y)$. So the Greedy Algorithm chose the wrong set.\\
    \indent Now we prove the "only if" direction. Let $M=(E,I)$ denote a matroid this time and we want to show that if $c$ is a positive objective function, the Greedy Algorithm finds maximum-weight independent sets of cardinality $k$ for every $k$ satisfying $1\leq k\leq r(E)$.\\
    \indent Suppose, towards a contradiction, that the algorithm fails for a given $k$. Concretely, suppose the algorithm choose $S_k = \{s_1, s_2, \ldots, s_k\}$ with $c(s_1)\geq c(s_2)\geq \ldots \geq c(s_k)$ and there exists some set $T_k = \{t_1, t_2, \ldots, t_k\}$ with $c(t_1)\geq c(t_2)\geq \ldots \geq c(t_k)$ and $c(T_k) > c(S_k)$. This means there exists some $p$ with $1\leq p\leq k$ such that $c(t_p) > c(s_p)$. Look at the smallest such $p$. Consider $X = \{s_1, s_2, \ldots, s_{p-1}\}$ and $Y = \{t_1, t_2,\ldots, t_p\}$. Note that $|Y|>|X|$, so by property 3 of matroids, there exists a $t_j\in Y\setminus X$ such that $X\cup\{t_j\}\in I$. By observation, $c(t_j)\geq c(t_p)>c(s_p)$, so the Greedy Algorithm should have chosen $t_j$, not $s_p$, a contradiction.
\end{proof}

\begin{problem}{10}
    Prove that the rank function of a matroid is submodular.
\end{problem}
\begin{proof}
    Let $M=(E,I)$ be a matroid and let $r$ denote its rank function. We want to show that for all $X,Y\subset E$, $r(X\cup Y) + r(X\cap Y) \leq r(X)+r(Y)$. Let $J$ be a maximal independent subset of $X\cap Y$. Extend $J$ to $J_X$, a maximal independent subset of $X$, and likewise to $J_Y$, a maximal independent subset of $Y$. We have $r(X \cap Y)=|J| = |J_X\cap J_Y|$. If we can show that $r(X\cup Y)\leq |J_X\cup J_Y|$, then submodularity follows, because $|J_X\cup J_Y|+|J_X\cap J_Y|=|J_X|+|J_Y|$. Extend $J_X$ to a maximal independent subset $K$ of $X\cup Y$.\\
    \indent Suppose, towards a contradiction, that $|K|>|J_X\cup J_Y|$. Because $J_X\setminus J$ is contained in both $K$ and $J_X\cup J_Y$, we have $|K\setminus (J_X\setminus J)|>|J_Y|$. Now, by the choice of $J_X$, we have that $K\setminus(J_X\setminus J)$ is an independent subset of $Y$. This contradicts the choice of $J_Y$.
\end{proof}

\end{document}

### 2019-06-15 misc/study-notes/ndmi011notes.tex
Added graph theory notes and TeX macros

% Notes for NDMI011 Combinatorics and Graph Theory
% Charles University in Prague, Summer 2019
% Instructor: Prof. Andreas Feldmann
% Notes by Marcel Goh

\input fontmac
\input mathmac

\leftrighttop{NDMI011 COMBINATORICS AND GRAPH THEORY}{MARCEL GOH}

\maketitle{NDMI011 Combinatorics and Graph Theory}{Notes by}{Marcel K. Goh {\rm (Prague, Czech Rep.)}}{10 June 2019}

\section 1. PRELIMINARIES

We say that a function $f(n)$ is $O\big(g(n)\big)$ (read ``big-oh of $g(n)$'') if there exist constants $n_0$ and $C$ such that for all $n\geq n_0$, $f(n)\leq C\cdot g(n)$. If the ratio ${f(n)\over g(n)}$ approaches 0 as n approaches infinity, then we say $f(n)$ is $o\big(g(n)\big)$ (read ``little-oh of $g(n)$''). If $f(n)$ is both $O\big(g(n)\big)$ and $o\big(g(n)\big)$, then $f(n)$ is $\Theta\big(g(n)\big)$ (read ``big-theta of $g\big(n)\big)$''). Finally, if the ratio ${f(n)\over g(n)}$ approaches 1 as $n$ approaches infinity, then we write $f(n) \sim g(n)$.

Let $G=(V,E)$ be a graph. The graph $G'=(V',E')$ is a {\it subgraph} of $G$ if $V'\subseteq V$ and $E'\subseteq E$. A {it cycle} is a graph on $v_1, \ldots, v_n$ with edges $\{v_1, v_2\}, \ldots, \{v_{n-1}, v_n\}$. A {\it tree} is a connected graph that contains no cycle (i.e\  does not have a cycle as a subgraph).

\proclaim Theorem. In any graph $G=(V,E)$,
$$\displaystyle\sum_{v\in V} \deg v = 2\cdot|E|$$

\proof Every edge $\{u, v\}\in E$ adds 1 to the degree of $u$ and adds 1 to the degree of $v$. Hence if you sum over all vertices, each edge gets counted twice.\slug

\proclaim Theorem.
The number of vertices with odd degree is even.

\proof Divide $V$ into $V_1$ and $V_2$ where $V_1$ is the set of vertices with odd degree and $V_2$ is the set of vertices with even degree. Then we have $\sum_{v\in V} \deg v = \sum_{v\in V_1} \deg v + \sum_{v\in V_2} \deg v$. Observe that the sum over all odd vertices $\sum_{v\in V_1}\deg v$ equals $2\cdot|E| - \sum_{v\in V_2}\deg v = 2(|E| - x)$ for some $x$, meaning it is even. Since the sum is even but each vertex has an odd degree, the number of vertices must be even.\slug

\proclaim Theorem.
A tree with $n$ vertices has $n-1$ edges.

\proof Let $G = (V,E)$ be a tree and let $n$ denote $|V|$, the number of vertices. Using induction on $n$, we show that it has $n-1$ vertices. The base case $n = 1$ is easy, since a tree with only one vertex has no edges. Now assume that a tree with $n$ vertices has $n-1$ edges and we consider the case where $G$ has $n+1$ vertices. Let us create $G'$ by removing a vertex from $G$. We cannot remove an internal vertex, since then $G'$ would not be connected. So we must remove a leaf, along with the edge that connected it to $G$. Now $G'$ has $n$ vertices, so by the induction hypothesis it has $n-1$ edges. This implies that $G$ had $n$ edges.
\slug

\section 2. GENERATING FUNCTIONS

\vskip -\medskipamount

\section 2.1. Power series and generating functions

Given an infinite sequence $a_0, a_1, \ldots = (a_i)_{i=0}^\infty$, the power series
$$\sum_{i=0}^\infty a_ix^i$$
is called the {\it generating function} of $(a_i)_{i=0}^\infty$ and is denoted $a(x)$. For example, the sequence $1, 1, 1, \ldots$ has generating function $\sum_{i=0}^\infty x^i$, which converges to $1/(1-x)$ for $x\in (-1, 1)$. Since we only really care about the coefficients of a generating function, we assume that $x$ is within the power series' interval of convergence.

\section 2.2. Binomial coefficients

For $r\in \R$ and $k\in \N$, the {\it binomial coefficient} is given by
$${r\choose k} = \left(\prod_{i=0}^{k-1} (r-i)\right) /k!$$
where ${r\choose 0} = 1$. If $r\in \N$, then we can think of ${r\choose k}$ as the number of ways to choose $k$ objects out of a pool of $r$ objects.

\thmparens{}{Generalised binomial theorem} For any $r\in \R$,
$$\sum_{i\geq 0} {r\choose i} x^i = (1 + x)^r.$$

\proof In this proof we will gloss over technicalities relating to convergence. Let $a(x)$ denote $(1+x)^r$. Then the derivative of $a(x)$ is $a'(x)=r(1+x)^{r-1}$, the second derivative $a''(x)$ is $r(r-1)(1+x)^{r-2}$, and so on. Generally,
$$a^{(i)}(x) = \left(\prod_{j=0}^{i-1} (r-j)\right)(1+x)^{r-i}.$$
If we evaluate the function at $x=0$, we get
$$a^{(i)}(0) = \prod_{j=0}^{i-1} (r-j)$$
so we can use Taylor's theorem to get
$$\eqalign{
    a(x) &= \sum_{i\geq 0} {a^{(i)}(0)\over i!} x^i\cr
    &= \sum_{i\geq 0}{r\choose i} x^i,
}$$
which is what we wanted.\slug

\section 2.3. Existence of a closed form

It is clear from the theorem above that the generating function for the sequence ${r\choose 1}, {r\choose 2}, \ldots$ is $(1+x)^r$. We may wonder what kinds of sequences have a {\it closed form} generating function. The following theorem, presented without proof, answers this question.

\proclaim Theorem. Let $(a_i)_{i=0}^\infty$ be a sequence of reals such that there exists some $k>0$ for which $|a_i|\leq k^i$ for every $i\geq 1$. Then for all $x$ in the interval $(-1/k, 1/k)$, the power series
$$a(x) = \sum_{i=0}^\infty a_ix^i$$
converges, so the generating function $a(x)$ has a closed form. Moreover, for any arbitrarily small $\epsilon >0$, the values of $a(x)$ for $x$ in the interval $(-\epsilon, \epsilon)$ uniquely determine the sequence $(a_i)_{i=0}^\infty$ where, by Taylor's theorem, $a_i = a^{(i)}(0)/i!$\slug

\section 2.4. Application to counting

Generating functions give us a method for counting the number of elements with some parameter $i$. For example, we could count the number of trees with $i$ vertices or the number of steps performed by an algorithm with input size $i$. Often, $a_i$ is defined inductively or recursively.

For example, suppose that for some number $c$ our sequence $a_i$ is given by $a_0 = c$, and $a_i = a_{i-1} + c$ for $i\geq 1$. Then we get that
$$\eqalign{
    a(x) &= \sum_{i\geq 0} a_ix^i\cr
    &= cx^0 + \sum_{i\geq 1} (a_{i-1} + c) x^i\cr
    &=  cx^0 + \sum_{i\geq 1}cx^i + \sum_{i\geq 1} a_{i-1}\cr
    &= {c\over 1-x} + x\cdot a(x).
}$$
This implies that $a(x) = c/(1-x)$, and by Taylor's theorem, $a_i = a^{(0)}(0)/i!$. Calculating this directly can be very tedious, so very often, we will express $a(x)$ in terms of generating functions that we know. In the case of this example, we note that
$$\eqalign{
    {c\over (1-x)^2} &= c\cdot {1\over 1-x}\cdot {1\over 1-x} \cr
    &= c\left(\sum_{i\geq 0} x^i\right) \left(\sum_{i\geq 0} x^i\right)\cr
    &= \sum_{i\geq 0} c(i+1) x^i,
}$$
which gives us that $a_i = c(i+1)$ (which we could have easily gotten by observing the recurrence).

\section 2.5. Operations on generating functions

Some useful operations on generating functions include:
\medskip
{\bf Addition.} $a(x) + b(x) = \displaystyle\sum_{i\geq 0} (a_i + b_i) x^i$.
\smallskip
{\bf Multiplication.} $\beta a(x) = \displaystyle\sum_{i\geq 0} (\beta a_i) x^i$.
\smallskip
{\bf Scaling.} $a(\beta x) = \displaystyle\sum_{i\geq 0} (\beta^i a_i) x^i$.
\smallskip
{\bf Gaps.} $a(x^k) = \displaystyle\sum_{i\geq 0} a_i x^{ki}$.
\smallskip
{\bf Shift right.} $xa(x) = \displaystyle\sum_{i\geq 0} (a_{i-1}) x^{i}$.
\smallskip
{\bf Shift left.} $\left(a(x) - a_0\right)/x = \displaystyle\sum_{i\geq 0} (a_{i+1}) x^{i}$.
\smallskip
{\bf Derivative.} $a'(x) = \displaystyle\sum_{i\geq 0} \big((a+1) a_{i+1}\big) x^{i}$.
\smallskip
{\bf Integral.} $\displaystyle\int_0^x a(x)dx = \displaystyle\sum_{i\geq 1} (a_{i-1}/i) x^{i}$.
\smallskip
{\bf Convolution.} $a(x)b(x) = \displaystyle\sum_{i\geq 0} \left(\displaystyle\sum_{k=0}^i a_k b_{i-k}\right) x^{i}$.
\smallskip
{\bf Partial sums.} $a(x)/(1-x) = \displaystyle\sum_{i\geq 0} \left(\displaystyle\sum_{k\geq 0} a_k \right) x^{i}$.
\medskip

\section 2.6. Some applications

Imagine a situation at a grocery store. A customer has six \$1 coins, five \$2 coins, and four \$5 bills. Suppose she has to pay \$21. How many ways can she do this? To find the answer, we need only compute the coefficient of $x^{21}$ in the polynomial
$$ (1 + x + x^2 + \cdots + x^6)\cdot(1+x^2+x^4+\cdots+x^{10})\cdot(1+x^5+x^{10}+\cdots+x^{20}).$$
Now suppose that a box contains 30 red balls, 40 blue balls, and 50 white balls. In how many ways can one draw 70 balls? As you may have suspected, we need to get the coefficient of $x^{70}$ in
$$ (1+x+\cdots+x^{30})\cdot(1+x+\cdots+x^{40})\cdot(1+x+\cdots+x^{50}).$$
It may be instructive to work through a full example. We will use generating functions to count binary trees. A {\it binary tree} $T$ is either empty ($T=\emptyset$) or consists of a root adjacent to a left subtree $T_l$ and a right subtree $T_r$. Let $b_n$ denote the number of binary trees with $n$ vertices. We will try to find a closed formula for $b_n$. There is exactly one binary tree with 0 vertices (the empty one), and one binary tree with 1 vertex. From the definition, $n=0$ or $n = 1 + n_l + n_r$ where $n_l = k$ and $n_r = n-1-k$ for some $k\in \{0,\ldots, n-1\}$. Hence we obtain the recurrence $b_n = b_0\cdot b_{n-1} + b_1\cdot b_{n-2} + \cdots b_{n-1}\cdot b_0$ for $n\geq 1$, i.e.
$$ b_n = \sum_{k=0}^{n-1} b_k\cdot b_{n-1-k}. $$
Then the generating function $b(x)$ can be derived:
$$\eqalign{
    b(x) &= \sum_{n\geq 0} b_nx^n\cr
    &= 1x^0 + \sum_{n\geq 1} \left(\sum_{k\geq 0}^{n-1} b_k\cdot b_{n-1-k}\right) x^n\cr
    &= 1 + x\sum_{n\geq 0} \left(\sum_{k\geq 0}^{n-1} b_k\cdot b_{n-1-k}\right) x^n\cr
    &= 1 + xb(x)b(x).
}$$
This implies that $xb(x)^2 - b(x) + 1 = 0$, which means that
$$b(x) = {1\pm\sqrt{1-4x}\over 2x}$$
Which one is correct? Well, note that $b(0) = \sum_{n\geq 0} b_n0^n = b_0 = 1$, whereas
$$\lim_{x\rightarrow 0} {1 + \sqrt{1-4x}\over 2x} = +\infty.$$
We try the other root, starting with
$$\lim_{x\rightarrow 0} {1 - \sqrt{1-4x}\over 2x},$$
upon which we can apply l'Hospital's rule to get a limit of 1. So this is the generating function we want.

Applying Newton's formula and the scaling operation, we observe that
$$\sqrt{1-4x}=\big(1+(-4x)\big)^{1/2} = \sum_{n\geq 0} (-4)^n {{1/2}\choose n}x^n.$$
We can substitute this into our generating function to get that
$$b(x) = {1\over 2x}\left( 1 - \sum_{n\geq 0} (-4)^n {{1/2}\choose n}x^n\right).$$
Since $1 = \sum_{n\geq 0} (-4)^n {{1/2}\choose n}x^0$, we can further simplify, obtaining
$$\eqalign{
    b(x) &= {1\over x}\sum_{n\geq 1} -{1\over 2}(-4)^n{{1/2}\choose n}x^n\cr
    &= \sum_{n\geq 0} -{1\over 2} (-4)^{n+1}{{1/2}\choose n+1}x^n\cr
    &= \sum_{n\geq 0} {1\over n+1}{2n\choose n}x^n.
}$$
So the number of binary trees with $n$ vertices is ${2n\choose n}/(n+1)$. These are the {\it Catalan numbers}.

\section 3. GRAPHS

\vskip -\medskipamount

\section 3.1. Network flow

\vskip -\medskipamount

\section 3.1.1. Edge capacities

Suppose we are given a directed graph $G=(V,E)$ ($E\subseteq V\times V$) with designated source $s\in V$ and target $t\in V$. To each edge $e$ is assigned a capacity $c(e)$, which may be a non-negative real number or $+\infty$. Then an {\it s-t-flow} is a function $f:E\rightarrow \R$ such that for every edge $e\in E$, $0\leq f(e)\leq c(e)$. These are called the {\it capacity constraints}. Our flow must also obey the law of {\it flow conservation}, also known as {\it Kirchhoff's First Law} (which was originally observed in electrical circuits): For every vertex $v\in V\setminus\{s, t\}$, $\sum_{uv\in E} f(uv) = \sum_{vu\in E} f(vu)$. We want to maximise the flow that leaves $s$, subject to these constraints. The {\it value} of an s-t-flow is
$$|f| = \sum_{su\in E} f(su) - f(us),$$
and a {\it max-flow} is an s-t-flow with maximum value.

The intuitive way to check that a flow is maximal is to study ``bottlenecks''. Concretely, if we let $S\subseteq V$ be such that $s\in S$ and $t\notin S$, then the set $\delta^+(S) = \{uv\in E : u\in S, v\notin S\}$ is an {\it s-t-cut} or {\it s-t-edge-cut}. After removing $\delta^+(S)$ from $G$, there is no longer a path from $s$ to $t$, i.e.\  $\delta^+(S)$ {\it separates} $s$ from $t$. (Since we only removed edges {\it leaving} $S$, there may still be a path from $t$ to $s$.) The {\it capacity} of an s-t-cut is given by
$$ c\big(\delta^+(S)\big) = \sum_{uv\in \delta^+(S)} c(uv).$$
Then we can define a {\it min-cut} to be an s-t-cut with minimum capacity.

\proclaim Lemma. For any s-t-flow $f$ and s-t-cut $\delta^+(S)$,
$$|f| = \sum_{e\in \delta^+(S)} f(e) - \sum_{e\in \delta^-(S)} f(e),$$
where $\delta^-(S)$ is the set $\{uv : u\notin S, v\in S\}$.

\proof By flow conservation, we know that for every $v \neq s,t$,
$$\sum_{vu\in E} f(vu) - \sum_{uv\in E} f(uv) = 0.$$
Then the value of the s-t-flow is the sum of all equations for the cut $S$:
$$|f| = \sum_{v\in S} \left( \sum_{vu\in E} f(vu) - \sum_{uv\in E} f(uv) \right).$$
We need to consider three possible cases for an edge $e = uv$:
\medskip
\item {1.} ($u, v\in S$) If both $u$ and $v$ are in $S$, then we will add $f(uv)$ and then subtract $f(uv)$, so this case does not contribute to $|f|$.
\smallskip
\item {2.} ($u\in S$, $v\notin S$) In this case, $uv\in \delta^+(S)$, so we will have a contribution of $+f(uv)$ to $|f|$.
\smallskip
\item {3.} ($u\notin S$, $v\in S$) In this case, $uv\in \delta^-(S)$, so we will have a contribution of $-f(uv)$ to $|f|$.
\medskip
This implies that
$$ |f| = \sum_{e\in \delta^+(S)} f(e) - \sum_{e\in \delta^-(S)} f(e) ,$$
which is exactly what we wanted to prove.\slug

We can use this to derive an easy corollary:

\proclaim Corollary. If $f$ is a max-flow and $\delta^+(S)$ is a min-cut, then $|f| \leq c\big(\delta^+(S)\big)$.

\proof By the preceding lemma, $|f| \leq \sum_{e\in \delta^+(S)} f(e)$, which by capacity constraints is no greater than $\sum_{e\in \delta^+(S)} c(e)$. But by the definition of the capacity of a min-cut, this is just $c\big(\delta^+(S)\big)$. \slug

In fact, the value of a max-flow {\it equals} the capacity of a min-cut, but the other inequality requires a more involved proof.

\proclaim Theorem. If $f$ is a max-flow and $\delta^+(S)$ is a min-cut, then $|f| = c\big(\delta^+(S)\big)$.

\proof By the preceding corollary, it suffices to show that $|f| \geq c\big(\delta^+(S)\big)$. For any max-flow $f$, the idea is to construct some s-t-cut $\delta^+(S)$ with $|f| = \delta^+(S)$. This would imply that any {\it min-cut} would have capacity no greater than $|f|$.

Consider the following procedure. We start with $S\gets \{s\}$. While there exists an edge $uv\in \delta^+(S)$ such that $c(uv) > f(uv)$ {\it or} there exists an edge $vu\in \delta^-(S)$ such that $f(uv) > 0$, we add $v$ to $S$. The claim is that if $f$ is a max-flow, then this algorithm will produce an $S$ such that $t\notin S$, i.e.\  $S$ is an s-t-cut. If this claim is true, then we are done; since $\sum_{e\in \delta^-(S)} = 0$ after the algorithm terminates, we will have $|f| = c\big(\delta^+(S)\big)$.

Suppose, towards a contradiction, that $t\in S$. Then there exists a sequence of vertices $v_0,\ldots, v_k\in S$ such that $v_0 = s$, $v_k = t$, and for all $i\in \{0,\ldots, k-1\}$, either
\medskip
\item {a)} $v_iv_{i+1}\in E$ and $c(v_iv_{i+1}) > f(v_iv_{i+1})$, or
\smallskip
\item {b)} $v_{i+1}v_i\in E$ and $f(v_{i+1}v_i) > 0$.
\medskip
For each $i$, define a small real number $\epsilon_i$ given by
$$\epsilon_i = \cases{
    c(v_iv_{i+1}) - f(v_iv_{i+1}) & \hbox{if case (a) holds}\cr
    f(v_{i+1}) & \hbox{if case (b) holds}
}$$
Then we can let $\epsilon = \min\{e_i : 0\leq i\leq k-1\}$. Note that, by construction, $\epsilon > 0$.

Now we create a new function $f':E\rightarrow \R$ given by
$$ f'(e) = \cases{
    f(e) + \epsilon & \hbox{if}\ e\ \hbox{satisfies case (a)}\cr
    f(e) - \epsilon & \hbox{if}\ e\ \hbox{satisfies case (b)}\cr
    f(e) & \hbox{otherwise}
}$$
The claim is that $f'$ is an s-t-flow. The values of $f'(e)$ are non negative, because $f'(e)\geq f(e)\geq 0$ in case a), and in case b), $f'(e) = f(e) - \epsilon \geq 0$. The capacity constraints are satisfied, since in case a), $\epsilon\leq c(e) - f(e)$ implies that $f'(e) = f(e) + \epsilon \leq c(e)$; and in case b), $f'(e)\leq f(e)\leq c(e)$. And we can see that the law of flow conservation is obeyed as well, since for every $v\in V\setminus\{s, t\}$, either flow is unchanged at $v$ or exactly two edges are changed (one adds $\epsilon$ and the other one subtracts $\epsilon$ from $f$).

Computing the value of the flow $f'$, we find that
$$ |f'| = \sum_{sv\in E} f'(sv) - \sum_{vs\in E} f'(vs) = |f| + \epsilon.$$
This implies that $f$ is not a max-flow, a contradiction.\slug

So to determine if a flow is maximal, we can figure out what $S$ is and see if $t\in S$. If not, then $f$ is a max-flow and $\delta^+(S)$ is a min-cut. The proof of the theorem also suggests an algorithm to compute max-flow and min-cut:

\algbegin Algorithm M (Find max-flow/min-cut). Given a directed graph $G=(V,E)$, this algorithm finds a max-flow $f$ and a set $S\subset V$ such that $\delta^+(S)$ is a min-cut.

\algstep M1. [Initialise.] Set $i\gets 0$ and $f_0(e) \gets 0$ for every $e\in E$.
\algstep M2. [Find $S_i$] Construct the set $S_i$ for the flow $f_i$ using the procedure outlined in the preceding proof.
\algstep M3. [Is $t$ in $S_i$?] If $t\notin S_i$, the flow $f_i$ is maximal. Output $f_i$ and $S_i$.
\algstep M4. [Increase flow.] Find the ``path''$P$ from $s$ to $t$ in $S$, and compute $\epsilon$. Let $f_{i+1}$ be the result of augmenting $f_i$ by $\epsilon$ along $P$. Set $i\gets i+1$.
\algstep M5. [Repeat.] Go to step M2.\slug

It is clear that if the algorithm terminates, it will output the correct answer. So it is natural to wonder when the algorithm is sure to terminate. In the case of integer capacities, termination is easy to prove:

\proclaim Theorem. If for all $e\in E$, $c(e)\in \N_0$, then the capacity $n$ of the min-cut is integral and Algorithm M terminates in at most $n$ iterations.

\proof We prove, by induction on $i$, that at every step of the algorithm, $|f_i|\in \N_0$. The base case is simple because the algorithm starts with $|f_0|=0$. Now assume that $f_i\in \N_0$. In the step that increases $i$ to $i+1$, we choose $\epsilon$ to be the minimum value computed over a path $P$ of edges. So for some $e\in P$, either $\epsilon = c(e) - f_i(e)$ or $\epsilon = f_i(e)$. Since both $c(e), f_i(e)\in \N_0$, we have that $\epsilon\in \N_0$ and the value of the flow at each iteration is integral.

Then because $\epsilon > 0$, we have from integrality that $\epsilon \geq 1$ meaning that for every iteration of the algorithm, the value $|f_{i+1}| = |f_i| + \epsilon \geq |f_i| + 1$. Since the value of the max-flow is exactly $n$, the algorithm terminates in at most $n$ iterations.\slug

\section 3.1.2. Vertex capacites

Instead of constraining flow at the edges, we can instead assign capacities $d(v)$ to vertices $v\in V\setminus\{s,t\}$. Then for all $v\in V\setminus\{s,t\}$, an s-t-flow $f:E\rightarrow \R$ must satisfy $\sum_{uv\in E} f(uv)\leq d(v)$ and $\sum_{uv\in E} f(uv)= \sum_{vu\in E} f(vu)$. Note that by this definition, if $st\in E$, then the value of the max-flow will be infinite, so assume that $st\notin E$.

Now we define an {\it s-t-vertex-cut} to be a set $T\subseteq V\setminus\{s,t\}$ such that the graph $G\setminus T$ has no path from $s$ to $t$. For ease of notation, let $d(T) = \sum_{v\in T} d(v)$.

\proclaim Theorem. Let $f$ be a max-flow in a network with vertex capacities. If $T$ is a min-cut, then $|f| = d(T)$, and furthermore, if all capacities are integers, then there is an integer max-flow.

\proof We simply convert our vertex-constrained network to an edge-constrained one. Replace each constrained vertex $v_i$ with two new vertices $v_{i1}$ and $v_{i2}$. Connect these vertices with a single edge $e_i$ with $c(e_i) = d(v_i)$. Set the capacties of all other edges in the new graph to $+\infty$. Then the theorem follows from our previous theorems regarding edge-constrained networks.\slug

\section 3.2. Bipartite matchings

A graph $G=(V,E)$ is {\it bipartite} if $V$ can be split into disjoint subsets $A$ and $B$ such that every edge in $E$ has one endpoint in $A$ and the other in $B$. A {\it matching} is a set $M\subseteq E$ such that no two edges of $M$ share a vertex. A {\it maximum matching} is matching of maximum size.

Consider the so-called {\it marriage problem}. Given a set $A$ of boys and $B$ of girls, can we marry off all boys to girls that they know? We can model this problem with a bipartite graph $(A\cup B, E)$, where an edge exists between a boy and a girl if they know each other. First we define the set of {\it neigbours} of a vertex set $S\subseteq A$ to be $N(S) = \{v\in B : uv\in E\ \hbox{and}\ u\in S\}$. Then the following theorem tells us exactly when the marriage problem is solvable.

\thmparens{}{Hall} A bipartite graph $G=(A\cup B, E)$ (with $A$ and $B$ disjoint) has a matching $M\subseteq E$ of size $|M| = |A|$ if and only if $|N(S)|\geq |S|$ for all $S\subseteq A$.

\proof The ``only if'' direction is obvious. We prove the ``if'' direction by contrapositive, i.e.\  if there is no matching of size $|A|$ then there exists some $S\subseteq A$ such that $|N(S) < |S|$.

The idea is to use the vertex version of the max-flow min-cut theorem. We construct a directed graph $H=(V',E')$ from G. Introduce two new vertices $s$ and $t$ and set $V' = A\cup B \cup \{s,t\}$. We want to connect $s$ to every vertex in $A$ and connect every vertex in $B$ to $t$. Then for every edge already in the graph, ensure it is directed from $A$ to $B$. So
$$ E' = \{ su : u\in A\}\cup\{vt : v\in B\}\cup\{uv : uv\in E, u\in A, v\in B\}.$$
Now we set all the capacities on every vertex $v\in A\cup B$ to 1. We know from the max-flow min-cut theorem that there exists an integer max-flow, call it $f$, which is a {\it 0-1 flow}, meaning that $f(e)\in \{0,1\}$ for all $e\in E'$. Note that if $f\geq |A|$, then there exists a matching of size $|A|$, since the 0-1 flow uses at most one incoming and one outgoing edge of any vertex in $V\setminus \{s,t\}$.

But we assumed that no matching of size $|A|$ exists, so $|f| < |A|$. Let $T$ be a min-cut. From the max-flow min-cut theorem, we know that $d(T) = |f| < |A|$. Consider the sets $X = A\cap T$ and $Y = B\cap T$. Since $|X| + |Y| = |T| = \sum_{v\in T} 1 = d(T)$, we have that $|Y| < |A| - |X|$. There is no edge from $A\setminus X$ to $B\setminus Y$, as $T$ is an s-t-vertex-cut. So $N(A\setminus X)\subseteq Y$ in $G$ and thus $|N(A\setminus X)| \leq |Y| < |A| - |X| = |A\setminus X|$. The set $A\setminus X$ is exactly the $S$ we were looking for.\slug

In a graph $G = (V,E)$, a {\it vertex cover} is a set $C\subseteq V$ such that every edge is incident to at least one vertex in $C$. A {\it minimum vertex cover} is a vertex cover of minimum cardinality. Notice that every vertex cover of $G$ is an s-t-vertex cut in $H$, otherwise there would be a path from $s$ to $t$. So we can derive the following theorem as a corollary.

\thmparens{}{K\H onig's theorem} Let $G$ be a bipartite graph with maximum matching $M$ and minimum vertex cover $C$. Then $|M| = |C|$.\slug

\section 3.3. Graph connectivity

\vskip -\medskipamount

\section 3.3.1. Definitions and inequalities

A graph $G=(V,E)$ is {\it connected} if there is a path between any two vertices. Otherwise, we say $G$ is {\it disconnected}. A {\it component} of $G$ is a maximal connected subgraph. If $G$ is connected and removing a set $W$ of vertices or edges causes it to become disconnected, then we say $W$ {\it separates} $G$. We have special names for $W$ if $|W| = 1$. If such a $W\subseteq V$, we call it a {\it cut vertex}. If $W\subseteq E, |W| = 1$, we call $W$ a {\it bridge}.

A graph $G=(V,E)$ is {\it k-connected} if $|V|>k$ and no vertex set $W\subseteq V$ of size $|W|<k$ separates $G$. Note that the complete graph $K_n$ is $(n-1)$-connected and that if $G$ is $k$-connected then it is also $k'$-connected for any $k'\leq k$. If $|V|\geq 2$ and no edge set $W\subseteq E$ of size $|W|<k$ separates $G$, then $G$ is {\it k-edge-connected}.

The {\it connectivity} of a graph $G$, denoted $\kappa(G)$, is the maximum $k\in \N$ such that $G$ is $k$-connected. Likewise, the {\it edge-connectivity} $\lambda(G)$ of a graph $G$ is the maximum $k$ such that $G$ is $k$-edge-connected. Note that if $G$ is not a complete graph, then both $\kappa(G)$ and $\lambda(G)$ are at most $|V|-2$.

\proclaim Lemma. Let $G=(V,E)$ be a graph. Then for all edges $uv\in E$, $\kappa(G\setminus\{uv\})\geq \kappa(G) - 1$.

\proof Remove an edge $uv$ from $G$. Now we find a set of vertices $W$, with $|W| = \kappa(G\setminus\{uv\})$, that disconnects $G$. So $V$ has been split into three disjoint subsets $L\cup W\cup R$. If $u$ and $v$ both lie in $L$ or both lie in $R$, then $\kappa(G) \leq |W| = \kappa(G\setminus\{uv\})$. If $u\in L$ and $v\in R$ (or vice-versa), then if there exists a vertex $w\neq u\in L$, then $W\cup \{u\}$ separates $G$, implying that $\kappa(G)\leq |W\cup \{u\}|\leq \kappa(G\setminus\{u,v\}) + 1$. The last case is if $L = \{u\}$ and $R=\{v\}$. Then $|V| = |W\cup\{u,v\}|\leq \kappa(G\setminus\{u,v\} + 2$, meaning that $\kappa(G)\leq |V|-1 \leq \kappa(G\setminus\{u,v\})+1$. \slug

\proclaim Lemma. Let $G=(V,E)$ be a graph with $|V|=n$. Then the following both hold:
\medskip
\item {1.} $\kappa(G) - 1 \leq \kappa(G-v)$ for all $v\in V$.
\smallskip
\item {2.}$\lambda(G) - 1 \leq \lambda(G-e)\leq \lambda(G)$ for all $e\in E$.
\medskip

\proof We prove each part separately.
\medskip
\item {1.} If $G = K_n$, then $G-v=K_{n-1}$, so $\kappa(G) - 1 = n-2 = \kappa(G-v)$ and we are done. So assume $G\neq K_n$. If $G\neq K_n$ and $G-v = K_{n-1}$, then there must exist some vertex $u\in V$ such that $uv\notin E$. Then removing the set $W = V\setminus\{u,v\}$ separates $G$, so $\kappa(G) \leq |W| = n-2 = \kappa(G-v)$. The last case is that $G\neq K_n$ and $G-v \neq K_{n-1}$. Then there exists a $W$ with $|W|=\kappa(G-v)$ such that $W\cup \{v\}$ separates $G$, which means that $\kappa(G) \leq |W\cup\{v\}| = \kappa(G-v)+1$.
\smallskip
\item {2.} There exists a set $W\subseteq E$ with $|W| = \lambda(G-e)$. So $e\notin W$ but $W\cup\{e\}$ separates $G$. So $\lambda(G) \leq |W\cup\{e\}| = \lambda(G-e) + 1$. For the second inequality, let $W$ be the set that separates $G$, so $|W|=\lambda(G)$. Then $W\setminus\{e\}$ separates $G-e$. So $\lambda(G-e) \leq |W\setminus\{e\}|\leq |W|=\lambda(G)$.\slug

To summarise, removing an edge from $G$ causes both $\kappa(G)$, $\lambda(G)$ to decrease by {\it at most} 1. Removing a vertex from $G$ can cause $\kappa(G)$ to decrease by at most 1 as well, but may cause $\lambda(G)$ to decrease a lot (possibly down to 0). The next lemma describes the relationship between vertex- and edge-connectivity.

\proclaim Lemma. Let $G=(V,E)$ be a graph and let $\delta(G)$ denote the minimum degree over all vertices in $G$. Then
$$ \kappa(G) \leq \lambda(G) \leq \delta(G).$$

\proof The inequality $\lambda(G) \leq \delta(G)$ is easy, because if $v$ is the vertex of minimum degree in $G$, then removing the $\delta(G)$ edges adjacent to $v$ separates $v$ from the rest of the graph.

To show that $\kappa(G) \leq \lambda(G)$, we use that $\kappa(G-e) \geq \kappa(G)-1$ for all $e\in E$. Suppose $\lambda(G) = n$. So removing a set of edges $\{e_1,\ldots,e_n\}$ causes $G$ to be disconnected. This means that $0=\kappa(G-\{e_1,\ldots,e_n\}) \geq \kappa(G) - n$, which means that $\kappa(G) \leq n = \lambda(G)$. \slug

\section 3.3.2. Connectivity and paths

For the following important theorem, we will need a new definition. We say two s-t-paths $P$ and $Q$ are {\it independent} if $V(P)\cap V(Q) = \{s,t\}$. The paths are {\it edge-disjoint} if $E(P)\cap E(Q) = \emptyset$.

\thmparens{M}{Menger}
Let $G$ be a graph. The following both hold:
\medskip
\item {1.} $G$ is $k$-connected if and only if there are $k$ independent paths between any two vertices $s$ and $t$.
\smallskip
\item {2.} $G$ is $k$-edge-connected if and only if there are $k$ edge-disjoint paths between any two vertices $s$ and $t$.
\medskip

\proof The ``if'' direction is left as an exercise for the student. To prove the ``only if'' direction, we use the max-flow min-cut theorem. We construct a directed graph $G'$ from $G$:
$$\eqalign{
    V(G') &= V(G)\cr
    E(G') &= \{uv, vu : uv\in E(G)\}
}$$
To prove the vertex version of the theorem, for every vertex $v\in V'\setminus\{s,t\}$, we set a vertex capacity $d(v) = 1$. To prove the edge version of the theorem, we set $c(e) = 1$ for all edges in $E(G')$. Now we consider the vertex and edge cases separately.
\medskip
\item {1.} Assume that $st\notin E(G')$. Any s-t-vertex cut of $G'$ separates $G$; since any s-t-path in $G'$ is also a t-s-path, the s-t-vertex cut also separates $t$ from $s$. Let $T$ denote a minimum-cardinality s-t-vertex-cut in $G'$ and we get that $d(T) \geq \kappa(G)$. There exists a 0-1 flow with $|f|\geq \kappa(G)$. So at each vertex, the number of incoming edges is at most 1 and the number of outgoing edges is at most 1 as well. This implies that there are $|f|$ independent pats from $s$ to $t$. But by the max-flow min-cut theorem, $|f| = d(T) \geq \kappa(G) = k$. If it so happens that $st\in E(G)$, then $(s, st, t)$ is an additional independent s-t-path to those in $G-st$ and we know that $\kappa(G-st)\geq \kappa(G) - 1$. Hence $G$ being $k$-connected implies that there are at least $k$ independent paths between any $s, t$.
\smallskip
\item {2.} In the case of edge capacities, any s-t-edge-cut separates $G$ as any t-s-path is also an s-t-path in G'. Let $S$ denote a minimum s-t-edge cut. Then $c(E)\geq \lambda(G)$. Again, there exists a 0-1 flow f with $|f|\geq \lambda(G)$, but this time it may be possible that a vertex has many incoming and outgoing edges. However, we know that the number of incoming edges that carry flow must equal the number of outgoing edges that carry flow at any vertex. Let $H=(V,F)$ be a subgraph of $G$, where $F = \{e\in E(G) : f(e) = 1\}$. While there exists an s-t-path $P$ in $H$, we can remove any all edges of $P$ from $H$ while maintaining flow conservation at every vertex $v\in V\setminus\{s,t\}$. We can repeat the algorithm as many times as there are edges coming out of $S$, so at least $|f|$ times. Hence $|f|\geq \lambda(G)\geq k$.\slug

\section 3.3.3. 2-connectivity and ear decompositions

An {\it ear decomposition} of a graph $G$ is a sequence $G_0, G_1, \ldots, G_k$ of subgraphs of $G$ such that $G_0$ is a cycle and for any $i\in \{1,\ldots,k\}$, the graph $G_i$ is obtained from $G_{i-1}$ by adding a path $P_i$ to $G_{i-1}$ such that $P_i$ shares exactly its endpoints with $G_{i-1}$. Each of these paths $P_i$ is called an {\it ear}.

\proclaim Theorem. A graph is 2-connected if and only if it has an ear decomposition.

\proof The only way for a graph to not be 2-connected is for it to have a cut vertex. So the ``if'' direction is easy, because if there is an ear decomposition, there is no cut vertex.

To prove the ``only if'' direction, suppose that $G$ is a 2-connected graph, and let $G_0$ be an arbitrary cycle in $G$. To get from any subgraph $G_{i-1}$ to $G_i$, we need to find an ear. Since $G$ is connected and $G_{i-1}\neq G$, there exists an edge $uv\in E(G)\setminus E(G_{i-1})$ such that $u\in V(G_{i-1})$. If $v\in V(G_{i-1})$, then $uv$ is an ear and we can set $P_i = uv$. Otherwise, there exists an edge $uw\in E(G_{i-1})$. This we know because $G_0\subseteq G_{i-1}$ so $|V(G_{i-1})| \geq 3$ and $G_{i-1}$ is connected.

By Menger's theorem, there exist at least 2 independent paths between $v$ and $w$ in $G$. So there exists some v-w-path $P$ such that $u\notin V(P)$. Let $u'$ be the first vertex of $P$ that is in $G_{i-1}$. Then we can append $u$ to the front of $P$ to get a path $(u, uv, v, \ldots, u')$, which is an ear.\slug

For a graph $G$, we define an {\it edge addition} is the operation of adding an edge between two vertices that were not connected in $G$. An {\it edge subdivision} is the operation of splitting an edge into two edges, with a new vertex in between them.

\proclaim Corollary. A graph is 2-connected if it can be obtained from $K_3$ by a sequence of edge additions and subdivisions.

This corollary is immediate from the observation that any starting cycle $G_0$ is a subdivision of $K_3$, and any ear $P_i$ is a subdivision of an edge addition.

\section 3.4. Counting spanning trees

A {\it spanning tree} $T$ of a graph $G$ is a subgraph that contains all vertices in $G$ and is a tree with edges from $G$. Let $T(G)$ denote the number of spanning trees of a graph $G$. For example, $T(K_3) = 3$. We want to find out the value of $T(G)$ for different classes of graphs.

In our investigation of $T(K_n)$, we will need to define a certain type of finite sequence. A Pr{\"u}fer code is a sequence $(p_1, \ldots, p_{n-2})\in \{1,\ldots,n\}^{n-2}$. It turns out that there is a bijection between Pr{\"u}fer codes and trees of length $n-2$ and trees on $n$ vertices. First we consider the algorithm for getting the Pr{\"u}fer code from a tree on $n$ vertices:

\algbegin Algorithm P (Build Pr{\"u}fer code). Given a tree with $n$ vertices $T$, this algorithm constructs a Pr{\"u}fer code.

\algstep P1. [Initialise.] Set $i\gets 1$, $T_0\gets T$.
\algstep P2. [Get next element.] Let $l_i$ denote the leaf of $T_{i-1}$ with smallest index. Set $p_i$ equal to the (single) neighbour of $l_i$ in $T_{i-1}$.
\algstep P3. [Remove a leaf.] Set $T_i\gets T_{i-1}-l_i$.
\algstep P4. [Done?] If $i=n-2$, we output the sequence $(p_1,\ldots, p_{n-2})$. 
\algstep P5. [Repeat.] Otherwise, set $i\gets i+1$ and return to step P2. \slug

There is a reverse algorithm that builds a tree from a Prufer code:

\algbegin Algorithm T (Build a tree). Given a Pr{\"u}fer code $(p_1,\ldots, p_{n-2})$, this algorithm outputs a tree with $n$ labelled vertices.

\algstep T1. [Calculate $l_i$.] For every $i\in \{1, \ldots, n\}$, let $l_i$ denote the number of times $i$ appears in the Pr{\"u}fer code, plus 1.
\algstep T2. [Initialise.] Let $T_0 \gets(\emptyset, \emptyset)$ to begin with. Set $i\gets 1$.
\algstep T3. [Add an edge.] Let $j$ be the smallest index with $l_j = 1$. Set $T_i \gets T_{i-1} + \{p_i, j\}$, where the addition operation denotes adding the edge as well as the vertices $p_i, j$, if one (or both) of them is not yet in the tree.
\algstep T4. [Update $l$.] Set $l_{p_i} \gets l_{p_i} - 1$ and $l_j \gets l_j - 1$.
\algstep T5. [Done?] If $i = n-2$, we have worked through the whole Pr{\"u}fer code. There are two remaining non-zero elements in $l$, call them $l_x$ and $l_y$. Output $T_i + \{l_x, l_y\}$.
\algstep T6. [Repeat.] Otherwise, set $i\gets i+1$ and return to step T3. \slug

Since Algorithms P and T are both deterministic and they are inverse to one another, this implies a bijection between trees on $n$ vertices and Pr{\"u}fer codes of length $n-2$. Hence we have the following formula.

\thmparens{C}{Cayley's formula} The number of spanning trees of a complete graph with $n$ vertices is
$$ T(K_n) = n^{n-2}.$$

\proof Since every possible edge between two vertices exists in $K_n$, the number of spanning trees of $K_n$ is exactly the number of trees with $n$ vertices. This is the same value as the number of Pr{\"u}fer codes of length $n-2$, which is the value $n^{n-2}$, since there are $n$ choices for each of $n-2$ elements.\slug

The following theorem gives a formula for the number of spanning trees of $K_n$, after a single edge is removed.

\proclaim Theorem. Let $e$ be an edge in $K_n$. Then
$$ T(K_n-e) = (n-2)n^{n-3}.$$

\proof The idea is to double-count the number of edges in all spanning trees of $K_n$. Each spanning tree has $n-1$ edges so the total number of edges is $(n-1)n^{n-2}$. Now let $k_e$ denote the number of spanning trees of $K_n$ containing $e$. By symmetry, this is the same for any choice of $e$ in $K_n$. There are ${n\choose 2}$ edges in $K_n$, each contained in $k_e$ trees. This implies that
    $$ {n\choose 2}k_e = (n-1)n^{n-2}.$$
    Solving for $k_e$, we obtain the value $k_e = 2n^{n-3}$. Therefore $T(K_n - e) = T(K_n)-k_e = n^{n-2} - 2n^{n-3} = (n-2)n^{n-3}$, which is what we wanted.\slug

\section 4. EXTREMAL THEORY

\vskip -\medskipamount

\section 4.1. Extremal graphs

This section deals with the general question: ``What is the extremal (maximum/minimum) number of objects, subject to restriction $R$?'' Some simple examples from graph theory include
\medskip
\item {$\bullet$} Question: What is the maximum number of edges in a graph with $n$ vertices? Answer: ${n\choose 2}$.
\smallskip
\item {$\bullet$} Question: What is the maximum number of edges in a graph with no cycles? Answer: $n-1$.
\medskip

A harder question is ``What is the maximum number of edges in a graph not containing $C_3$ as a subgraph?'' First let us consider a class of graphs that definitely do not contain $C_3$ as a subgraph: bipartite graphs (they do not contain any odd cycles). What is the maximum number of edges in a bipartite graph? Well, the complete bipartite graph $K_{a,b}$ has $ab$ edges, so we want to maximise the product $ab$ when $a+b =n$. Intuitively this is when $a$ and $b$ are as close to $n/2$ as possible. So we get that the number of edges in $K_{a,b}$ is at least $\lfloor n^2/4 \rfloor$, thus we have a lower bound for the number of edges in a graph not containing $C_3$. The following theorem proves that the bound is tight.

\proclaim Theorem. The maximum number of edges of any graph $G$ not containing $C_3$ as a subgraph is at most $\lfloor n^2/4 \rfloor$.

\proof We want to show that there exist $a,b\in \N$ such that $|E(G)|\leq |E(K_{a,b})| \leq \lfloor n^2/4 \rfloor$. To this end, we find disjoint subsets $A,B\subseteq V(G)$ such that $A\cup B = V(G)$. and let $H$ = $K_{a,b}$ on $A$ and $B$.

Then it suffices to show that for any $v\in V(G)$, $\deg_G(v) \leq \deg_H(v)$, since the number of edges in any graph equals the sum over degrees of vertices, divided by 2. So let $v_0$ be a vertex of maximum degree in $G$. Then we can set $B = N_G(v_0)$ and $A = V\setminus B$ Now for any $v\in A$, $\deg_H(v) = |B| = \deg_G(v_0) \geq \deg_G(v)$. Note that in $G$, no two vertices of $B$ are adjacent, since then there would be a cycle of length 3. Then for any $v\in B$, $N_G(v)\subseteq A$, hence $\deg_G(v)\leq |A| = \deg_H(v)$.\slug

Compared to ${n\choose 2}$, $\lfloor n^2/4 \rfloor$ is still quite a lot, but remember that disallowing any cycles at all, we have a much smaller bound of $n-1$. A natural question is ``How long do forbidden cycles have to be so that the maximum number of edges is much smaller than ${n\choose 2}$?'' As we will see from the following theorem, disallowing $C_4$ lowers this bound quite a bit.

\proclaim Theorem. The maximum number of edges of any graph $G$ not containing $C_4$ as a subgraph is at most $(n^{3/2} + n) / 2$.

\proof The idea is to double-count the size of the set $M$ of pairs $(\{u, u'\}, v)$, where $u\neq\ u'\neq v$ and $uv, u'v\in E$. Note that for any set $\{u, u'\}$, there is at most one $v\in V$ such that $(\{u, u'\}, v)\in M$, because otherwise we would have a cycle of length 4, which is forbidden. This implies that $|M|\leq {n\choose 2}$, which is the number of sets $\{u, u'\}$.

Another way to count $|M|$ is to consider the number of sets $\{u, u'\}$ that are contributed to by each $v\in V$. For every set $\{u, u'\}\subseteq N(v)$ we get a pair $(\{u, u'\}, v)\in M$, so $v$ contributes ${\deg(v)\choose 2}$ elements to $M$. For ease of notation, let us number the vertices $V = \{1, \ldots, n\}$ and let $d_i$ denote $\deg(i)$. Then
$$ |M| = \sum_{i=1}^n {d_i\choose 2}\leq {n\choose 2} .$$
To get a bound on the number of edges, we will relate $(\sum_{i=1}^n d_i)/2$ to $|M|$. We know that ${n\choose 2}\leq n^2/2$. Without loss of generality, we may assume that $d_i\geq 1$ for all $i\in V$, since adding an edge between a vertex $i$ with $d_i = 0$ to any other vertex $j$ increases the number of edges without introducing a $C_4$. Therefore,
$$ {d_i\choose 2} = {d_i(d_i -1)\over 2} \geq {1\over 2}(d_i - 1)^2 $$
for every $i\in V$. This implies that
$$ \sum_{i=1}^n (d_i - 1)^2 \leq n^2. $$
Now we use the famous {\it Cauchy-Schwarz inequality}, which applies to vectors $(x_1, \ldots, x_n)$, $(y_1, \ldots, y_n)\in \R^n$:
$$\sum_{i=1}^n x_iy_i \leq \sqrt{ \sum_{i=1}^n x_i^2 }\sqrt{ \sum_{i=1}^n y_i^2 }.$$
We apply the formula with $x_i = d_i - 1$ and $y_i = 1$ to obtain
$$\sum_{i=1}^n (d_i - 1) \cdot 1 \leq \sqrt{\sum_{i=1}^n (d_i-1)^2} \sqrt{\sum_{i=1}^n 1} \leq \sqrt{n^2}\sqrt{n} = n^{3/2}.$$
Because $\sum_{i=1}^n (d_i-1) = \sum_{i=1}^n d_i - n$, we get $\sum_{i=1}^n d_i \leq n^{3/2} + n$. So we can put a bound on the number of edges:
$$ |E| = {1\over 2}\sum_{i=1}^n d_i \leq {n^{3/2} + n\over 2} .\noskipslug$$

\section 4.2. Partially-ordered sets

A {\it partially-ordered set} or {\it poset} is a pair $({\cal L}, \subseteq)$ where $\subseteq$ is a partial order on the set ${\cal L} \subseteq 2^X$ over $X = \{1,\ldots , n\}$. A {\it chain} is a subset $\{A_1, \ldots, A_k\}$ of ${\cal L}$ such that $A_1\subseteq A_2\subseteq\cdots\subseteq A_k$. An {\it antichain} or {\it independent set system} is a subset $\{A_1, \ldots, A_k\}$ of ${\cal L}$ such that $A_i\not\subset A_j$ for all $i,j\in\{1,\ldots, k\}, i\neq j$. A chain is {\it maximal} if adding any set breaks the chain property.

The maximum length of a chain in $({\cal L}, \subseteq)$ is $|X| + 1$, where $X$ denotes the motherset. This is because $\emptyset$ can be part of the chain as well. Now consider the maximum length of an {\it antichain}. One way to get an antichain is to take all sets $A\in {\cal L}$ such that $|A| = i$ for some $i$. This number is at least ${n\choose i} \leq {n\choose \lfloor n/2 \rfloor}$ (in the case that ${\cal L} = 2^X$). The following theorem gives an upper bound:

\thmparens{S}{Sperner} Any antichain of a poset of $X = \{1, \ldots, n\}$ has size at most $n\choose \lfloor n/2 \rfloor$

\proof We will work with ${\cal L} = 2^X$. Let an antichain $M\subset 2^X$ be given. Our goal is to bound the cardinality of $M$. The idea is to double count the number of pairs $(R,A)$ where $A\in M$ and $R$ is a maximal chain containing $A$. Notice that a maximal chain contains exactly one set of size $i$ for $i\in \{0,\ldots, n\}$:
$$\emptyset \subseteq \{x_1\} \subseteq \{x_1, x_2\} \subseteq \cdots \subseteq \{x_1, x_2, \ldots, x_n\}.$$
where $x_1\ldots x_n$ are elements of $X$ written in some order. This implies that the number of maximal chains is $n!$ as every possible ordering defines a chain. By observation, any $R$ contains at most one $A\in M$, so the number of pairs $(R,A)$ is at most $n!$.

The other way we will count the pairs is to ask how many maximal chains actually contain a set $A\in M$. If $R$ is made up of elements $x_1,\ldots,x_n$ as shown above, then $A\in R$ if and only if $A = \{x_1,\ldots,x_k\}$ for some $k$. Hence to form $R$ we first introduce $x_1,\ldots,x_k$ in $k!$ ways, then $x_{k+1},\ldots,x_n$ in $(n-k)!$ ways, so the number of pairs $(R,A)$ is
$$ \sum_{A\in M} |A|! (n-|A|)! \leq n!$$
Dividing by $n!$ on both sides of the equation we get that
$$\sum_{A\in M} {|A|! (n-|A|)!\over n!} = \sum_{A\in M} {n\choose |A|}^{-1} \leq 1.$$
We know that ${n\choose |A|}\leq {n\choose \lfloor n/2 \rfloor}$, hence we may replace the reciprocal of the former with the reciprocal of the latter to get
$$1\geq \sum_{A\in M} {n\choose \lfloor n/2 \rfloor}^{-1} = {|M|\over {n\choose \lfloor n/2\rfloor}},$$
which means that $|M| \leq {n\choose \lfloor n/2 \rfloor}$.\slug

\section 5. FINITE PROJECTIVE PLANES

Let $X$ be a finite point set and ${\cal L}\subseteq 2^X$ be a set of subsets of $X$. $(X, {\cal L})$ is called a {\it finite projective plane} if the following properties hold:
\medskip
\item {P0)} There exists a set $F\subseteq X$ such that $|F| = 4$ and for all $L\in {\cal L}$, $|F\cap L|\leq 2$.
\smallskip
\item {P1)} For all $L_1,L_2\in {\cal L}$ with $L_1\neq L_2$, we have $|L_1\cap L_2| = 1$. 
\smallskip
\item {P2)} For all $x_1, x_2\in X$ with $x_1\neq x_2$, there is {\it exactly one} $L\in {\cal L}$ such that $x_1,x_2\in L$ and we may denote it $L = \overline{x_1x_2}$.
\medskip
We call any $x\in X$ a {\it point} and any $L\in {\cal L}$ a {\it line}. Finite projective planes have very interesting ``symmetrical'' properties.

\proclaim Lemma. Let $(X,{\cal L})$ be a finite projective plane. Then for all $L, L'\in {\cal L}$, there exists an $z\in L$ such that $x\notin L\cup L'$

\proof By (P0), there exists a set $F\subseteq X$ with $|F| = 4$ and $|F\cap L|\leq 2$ for a $|F\cap L'|\leq 2$. If $F$ is a proper subset of $L\cup L'$, then we're done, because $z\in F$. So assume $F\subset L\cup L'$: say $F = \{a, b,c,d\}$, $F\cap L = \{a, b\}$, and $F\cap L' = \{c, d\}$. Then by (P2), there exist lines $L_1 = \overline{ac}$ and $L_2 = \overline{bd}$; and by (P1), there exists a point $z\in L_1\cap L_2$. We claim that $z\notin L\cup L'$.

Suppose, towards a contradiction, that $z\in L$. Then since, by (P1), $|L\cap L_1| = 1$ and $z\in L_1$, we have that $z = a$. This implies that $L_2$ contains $a,b,d$. So $|F\cap L_2| \geq 3$, a contradiction to (P0). An analogous argument can be made for $L'$ so we are done. \slug

\proclaim Lemma. Let $(X,{\cal L})$ be a finite projective plane. Then for all $L,L'\in {\cal L}$, $|L| = |L'|$.

\proof The idea is to find a bijection $\phi : L \rightarrow L'$. By the preceding lemma, there exists some $z\notin L\cup L'$, so for any $x\in L$, we let $\phi(x)$ be the point in $L'\cap \overline{zx}$. The function $\phi$ is well-defined, since $\overline{zx}$ exists by (P2), and $|L' \cap \overline{zx}| = 1$ by (P1).

Now we prove that $\phi$ is a bijection. Let $y\in L'$ be given. It suffices to show that $|\phi^{-1}(y)| = 1$. Let $x\in L\cap \overline{yz}$. Then the set $\{x,z\}\subseteq \overline{yz}\cap\overline{xz}$, i.e.\  $|\overline{yz}\cap\overline{xz}|\geq 2$, so by (P1), $\overline{yz} = \overline{xz}$. This means that $\phi(x) = y$ and $x$ is unique by (P1), so $|\phi^{-1}(y)| = 1$.\slug

The {\it order} of a finite projective plane $(X, {\cal L})$ is $|L| - 1$ for any $L\in X$.

\proclaim Theorem. Let $(X,{\cal L})$ be a finite projective plane of order $n$. Then the following all hold:
\medskip
\item {1.} Exactly $n+1$ lines pass through any point $x\in X$.
\smallskip
\item {2.} $|X| = n^2 + n + 1$.
\smallskip
\item {3.} $|{\cal L}| = n^2 + n + 1$.
\medskip

\proof First we show that for any $x\in X$, there exists some line $L\in {\cal L}$ such that $x\notin L$. By (P0), there exists some $F = \{a,b,c,d\}$ and without loss of generality we will assume that $x\notin\{a,b,c\}$. So $\overline{ab} \cap F = \{a,b\}$ and $\overline{ac}\cap F = \{a, c\}$. Now if $x = d$, then $x\notin \overline{ab}$ and we're done. If $x\in \overline{ab}$, then $x\notin\overline{ac}$ and if $x\in \overline{ac}$, then $x\notin\overline{ab}$, since $|\overline{ab} \cap\overline{ac}| = 1$.

Let $x\in X$ be given and let $L$ be a line in ${\cal L}$ such that $x\notin L$. Now we may prove each part of the theorem.
\medskip
\item {1.} For all $y\in L$, there exists a line $\overline{xy}$ passing through $x$. There are $n+1$ points $y\in L$, so there are at least $n+1$ lines passing through $x$. Now for all $L'\in{\cal L}$ such that $x\in L'$, $|L\cap L'| = 1$ by (P1), so $L'$ was already counted above. So there are exactly $n+1$ lines through $x$.
\smallskip
\item {2.} Let $L_1 = \overline{xx_i}$ where $x_i\in L, i\in\{1,\ldots, n+1\}$. By (P1), $L_i\cap L_j = \{x\}$ for all $i\neq j$. As $|L_i\setminus\{x\}| = n$ we have
    $$|X|\geq |\bigcup_{i=1}^{n+1} L_i| = (n+1)n + 1 = n^2 + n + 1,$$
    so $|X|\geq n^2 + n + 1$.
    Now we show that for all $p\in X$, there exists an $i\in \{1,\ldots,n+1\}$ such that $p\in L_i$. If $p=x$. this is clearly true, so assume $p\neq x$. Then $\overline{px}\cap L = \{x_i\}$ for some $i$ by (P1). So $\overline{px} = L_i$ for some $i$ and we already counted $p$ above. So $|X| = n^2 + n + 1$.
\smallskip
\item {3.} This follows from item 2 and duality, which we will introduce next.\slug

The {\it incidence graph} of a finite projective plane $(X,{\cal L})$, is a bipartite graph $G = (V, E)$ with $V = X\cup {\cal L}$ and $E = \{xL : if x\in L\}$. The concept of duality involves switching the roles of $X$ and ${\cal L}$ in this bipartite graph, interpreting $X$ as lines and ${\cal L}$ as points.

Formally, the {\it dual} of a finite projective plane $(X,{\cal L})$ is the set system $({\cal L}, \Lambda)$ where $\Lambda\in 2^{\cal L}$ contains an element $\{L\in {\cal L} : x\in L\}$ for every $x\in X$.

\proclaim Lemma. The dual $({\cal L}, \Lambda)$ of a finite projective plane $(X,{\cal L})$ is also a finite projective plane.

\proof We show that $({\cal L}, \Lambda)$ satisfies each of (P0), (P1), and (P2).
\medskip
\item {(P1)} We need to find lines $L_1, L_2, L_3, L_4$ such that any for any $\lambda\in \Lambda$, with $\lambda = \{L\in {\cal L} : x\in L\}$ for some $x\in X$, the point $x$ is contained in at most two of $L_1, L_2, L_3, L_4$. Let $F = \{a,b,c,d\}$ be the set given by applying (P0) to $(X,{\cal L})$. Let $L_1 = \overline{ab}, L_2 = \overline{cd}, L_3 = \overline{ad}, L_4 = \overline{bc}$. Suppose, towards a contradiction, that $x$ is in three of these lines. Without loss of generality, suppose $x\in L_1\cap L_2\cap L_3$. By (P1) for $(X,{\cal L})$, $|L_i\cap L_j|$ for all $i\neq j$. Since $x\in L_1\cap L_3$, we have that $x = a$. But since $x\in L_2\cap L_3$, we have that $x = d$. This is a contradiction, since $a\neq d$. The same is true for any other triple of these four lines.
\smallskip
\item {(P2)} Let $\lambda_1,\lambda_2\in \Lambda, \lambda_1\neq\lambda_2$ be given, i.e.\  $\lambda_1 = \{ L\in {\cal L} : x_1\in L\}$ and $\lambda_2 = \{L\in {\cal L} : x_2\in L\}$ for some points $x_1\neq x_2$. We need that $|\lambda_1\cap\lambda_2| = 1$, but this follows from (P2) for $(X,{\cal L})$, since $x_1$ and $x_2$ intersect at exactly one line.
\smallskip
\item {(P3)} Let $L_1, L_2\in {\cal L}$ with $L_1\neq L_2$. It suffices to show that there exists a unique $\lambda \in \Lambda$ such that $L_1\in \lambda$ and $L_2\in \lambda$. But since $\lambda = \{L\in {\cal L} : x\in L\}$ for some $x$, this is exactly (P1) for $(X,{\cal L})$.\slug

\proclaim Corollary. The order of the dual finite projective plane is the same as the order of the original.

\proof For any $\lambda \in \Lambda$ of the dual, $|\{L\in {\cal L} : x\in L\}| = n+1$. This implies that the order or $({\cal L}, \Lambda)$ is $n$.\slug

One may wonder whether there is a finite projective plane of any order. The answer is no. There are no finite projective planes of order 1, 6, or 10. There are finite projective planes of order 3, 4, 5, 7, 8, 9, and 11. Whether there is a finite projective plane of order 12 is an open problem. The existence of finite projective planes of some higer orders is known though, as stated by the following theorem.

\proclaim Theorem. A finite projective plane exists of order $n$ if $n$ is a prime power.

We will not present the proof, which follows from ${\bf F}_n$ being a field.\slug

A corollary of this theorem is that there are infinitely many finite projective planes. Recall that a graph on $n$ vertices without $C_4$ as its subgraph has at most $(n^{3/2} + n)/2$ edges. Then a corollary of the infinity of finite projective planes is given below.

\proclaim Corollary. For infinitely many values of $n$ there is a graph on $n$ vertices without $C_4$ as a subgraph that contains at least $(n/2)^{3/2}$ edges.

\proof Consider the incidence graph of the finite projective plane $(X,{\cal L})$. The number of vertices $n = |X| + |{\cal L}| = 2(n^2 + n + 1)$. Then for all $L\in {\cal L}$, since $|L|=m+1$, the degree of $L$ in $G$ is $m+1$ as well. This means that $|E| = (m+1)|{\cal L}| = (m+1)(m^2+m+1) \geq (m^2+m+1)^{3/2} = (n/2)^{3/2}$. So the incidence graph has at least $(n/2)^{3/2}$ edges, so we just need to show that it does not have $C_4$ as a subgraph. This is easy because a $C_4$ in the bipartite incidence graph implies that the intersection of two lines $L\cap L'$ contains two points $x,x'$, which contradicts (P1).\slug

\section 6. LATIN SQUARES

A {\it Latin square} of order $n$ is a matrix $A \in \{1,\ldots, n\}^{n\times n}$ such that for any row $r$, $A_{ri} \neq A_{rj}$ if $i\neq j$ and for any column $c$, $A_{ic}\neq A_{jc}$ if $i\neq j$. Two Latin squares $A$ and $B$ are {\it orthogonal} if $(A_{ij}, B_{ij}) \neq (A_{xy}, B_{xy})$ whenever $i\neq x$ or $j\neq y$. Note that the number of ordered pairs of $\{1,\ldots,n\}$ is $n^2$ and there are only $n^2$ cells, so if all the pairs are different, each pair appears exactly once.

\proclaim Theorem. Let $M$ be a set of pairwise orthogonal Latin squares of order $n$. Then $|M|\leq n-1$.

\proof Let $A,B$ be orthogonal Latin squares of order $n$ and let $\pi$ be some permutation of $\{1,\ldots,n\}$. Consider $A'$ where $A'_{ij} = \pi(A_{ij})$. Note that $(A'_{ij}, B_{ij}) = (A'_{xy}, B_{xy})$ if and only if $(A'_{ij}, B_{ij}) = (A'_{xy}, B_{xy})$. So $A'$ and $B$ are also orthogonal.

So let $M = \{A_1, \ldots, A_t\}$ and for each $A_i\in M$, permute the entries such that the first row of the resulting Latin square $A'_i$ is $(1,2,3,\ldots,n)$. By our earlier observation, $A'_1,\ldots,A'_k$ are pairwise orthogonal. Now we zoom in on the second row, first column if $A'_i$ for some $i$. Call this cell $k$. First we notice that $k\neq 1$, since the first element of the first row is 1. Then since comparing any two $A'_i$ pairwise, we get all the pairs $(1,1), (2,2), \ldots, (n,n)$ in the first row, each of $2,3,\ldots,n$ can appear in the second row, first column of at most one $A'_i$. So $t = |M|\leq n-1$.\slug

The following theorem relates orthogonal Latin squares to finite projective planes.

\proclaim Theorem.
    For $n\geq 2$, a finite projective plane of order $n$ exists if and only if there exists a set of $n-1$ pairwise orthogonal Latin squares of order $n$.

\section 7. RAMSEY THEORY

Ramsey theory deals with the general question: ``How big must a mathematical structure have to be such that some property holds?"

\section 7.1. Independent sets and cliques

In a graph $G=(V,E)$, an {\it independent set} is a set $I\subseteq V$ such that $uv\notin E$ for all $u,v\in I$. A {\it clique} is a set $K\subseteq V$ such that $uv\in E$ for all $u,v\in K, u\neq v$. The {\it independent set number} $\alpha(G)$ is the maximum size of any independent set and the {\it clique number} $\omega(G)$ is the maximum size of any clique in $G$.

\proclaim Theorem. Let $G=(V,E)$ be a graph. If $|V|\geq {k-l-2\choose k-1} = {k+l-2\choose l-1}$, then $\alpha(G)\geq l$ or $\omega(G)\geq k$.

\proof By induction on $k+l$. In the base case $k = 1$ or $l=1$, all the theorem says is that there is an independent set or clique of size 4. For the inductive step, let $k,l\geq 2$ and assume that the claim holds for $k,l-1$ or $k-1,l$. By Pascal's formula,
$${k+l-2\choose k-1} = {k+l-3\choose k-1} + {k+l-3\choose k-2}.$$
Call the left-hand side $n$ and from the right-hand side, call the first summand $n_1$ and the second summand $n_2$. Now pick any vertex $u\in V$. Let $B = N(u)$ and $A = V\setminus(B\cup \{u\})$. First note that it is impossible that both $|A| < n_1$ and $|B|<n_2$ since that would imply that
$$n = 1 + |A| + |B| \leq 1 + (n_1-1) + (n_2=1) = n-1,$$
a contradiction. So either $|A|\geq n_1$ or $|B|\geq n_2$.

If $|A|\geq n_1$, by the induction hypothesis $\omega(G[A])\geq k$ or $\alpha(G[A])\geq l-1$. Then a clique in $G[A]$ is also a clique in $G$, for the first case; and for the second case, adding the vertex $u$ to an independent set in $G[A]$ gives an independent set in $G$, so $\alpha(G)\geq l$.

If it happens instead that $|B|\geq n_2$, then by induction we have either $\omega(G[B]) \geq k-1$ or $\alpha(G)\geq l$. In the first of these cases, an independent set in $G[B]$ is also an independent set in $G$, so $\alpha(G)\geq l$; and in the second case, adding $u$ to a clique in $G[B]$ gives a clique in $G$, so $\omega(G) \geq K$. \slug

For each $k,l\in \N$, the {\it Ramsey number} $r(k,l)$ is the minimum $n$ such that for any graph $G$ with $n$ vertices, $\omega(G)\geq k$ or $\alpha(G)\geq l$.

\proclaim Corollary. $$r(k,l) \leq {k+l-1\choose k-1} = {k+l-2\choose l-1}.$$

Very little is known about Ramsey numbers. By observation, we have that $r(k,1) = r(1,l) = 1$; $r(2,l) = l$ and $r(k,2) = k$. For $k=l$, we only know that $r(3,3) = 6$ and $r(4,4) = 18$. The exact values of $r(n,n)$ for $n\geq 5$ are unknown. The above corollary gave an upper bound and the next theorem gives a lower bound.

\proclaim Theorem. For $k\geq 3$, $r(k,k)>2^{k/2}$.

\proof First we introduce some basic notions from probability that we will need. Let $\Omega$ be a probability space and let $X$ be an event. Then $\Pr[X] = |X|/|\Omega| < 1$ if and only if $|X| < |\Omega|$, which would imply that $\neg X \neq \emptyset$.

Now we create a random graph $G$ by starting with a set of $n$ vertices and adding each edge with probablity $1/2$ uniformly at random. We're trying to prove that if $n\leq 2^{k/2}$ there is a $G$ with $|V| = n$ such that $\alpha(G) < k$ and $\omega(G) < k$, so it suffices to show that
$$\Pr[\alpha(G) \geq k\ \hbox{or}\ \omega(G)\geq k : n\leq k^{k/2}] < 1.$$
Let $A$ be a subset of vertices with $|A|=k$. Let $K_A$ denote the proposition ``A forms a clique'' and let $I_A$ denote the proposition ``A forms an independent set". Then let $X_A = K_A \cup I_A$. $\Pr[K_A] = (1/2)^{k\choose 2} = \Pr[I_A]$ and since $K_A \cap I_A = \emptyset$,
$$X_A = 2\cdot\left({1\over 2}\right)^{k\choose 2} = 2^{1-{k\choose 2}}.$$
Let $\Pr(Y)$ denote the probability that $\alpha(G) \geq k$ or $\omega(G) \geq k$. This is equal to the probablilty that $X_A$ holds for some $A\subseteq V$ with $|A|\geq k$. So
$$\Pr[Y] \leq \sum_{A\subseteq V, |A|\geq k} 2^{1-{k\choose 2}} = {n\choose 2}2^{1-{k\choose 2}} \leq {n^k\over k!}2^{1-{k\choose 2}},$$
and we can further derive the strict inequality
$${n^k\over k!}2^{1-{k\choose 2}} < {n^k\over {2^{k/2+1}}}2^{1-(k^2/2-k/2)} = \left({n\over 2^{k/2}}\right)^k,$$
which is less than $1$, since $n\leq 2^{k/2}$. \slug

\section 7.2. Increasing/decreasing subsequences

Given a finite sequence $S = (x_1, x_2, \ldots, x_n)$ of numbers, an {\it increasing} ({\it decreasing}) {\it subsequence} of length $k$ is a sequence $(x_{i_1}, x_{i_2}, \ldots, x_{i_n})$ such that for all $j,l$ with $1\leq j < l\leq k$, $i_j < i_l$ and $x_{i_j} \leq x_{i_l}$ ($x_{i_j} \geq x_{i_l}$).

\thmparens{E}{Erd\H os-Szekeres} For any finite sequence $S$ of at least $(r+1)(s+1) + 1$ numbers, there is an increasing subsequence of length $r$ or a decreasing subsequence of length $s$.

\proof Suppose, towards a contradiction, that every increasing/decreasing subsequence has length at most $r-1$/$s-1$ respectively. Then we label each $x_i$ of $S$ with a pair $(a_i, b_i)$ where $a_i$/$b_i$ is the length of the longest increasing/decreasing subsequence ending in $x_i$. (Note that each of the numbers $x_i, x_j$ have different labels.) Without loss of generality, assume $i<j$. If $x_i\leq x_j$, then $a_i < a_j$, as any increasing subsequence ending in $x_i$ can be extended by $x_j$. Likewise, if $x_j\leq x_i$, then $a_j < a_i$, as any increasing subsequence ending in $x_j$ can be extended by $x_i$.

By our assumption, $a_i\leq r-1$ and $b_i\leq s-1$ for all $i$. So the number of labels $(a_i, b_i)\leq (r-1)(s-1)$ which implies that the length of the sequence $S$ is at most $(r-1)(s-1)$, a contradiction.\slug

The given bound is tight; there are sequences with length $(r-1)(s-1)$ numbers that neither have an increasing subsequence of length $r$, nor have a decreasing subsequence of length $s$. To construct one, take a grid of $(r-1)(s-1)$ points in the plane and rotate it {\it very} slightly counterclockwise such that no two points are on the same horizontal or vertical line. Then we have a set of points $(x_i, y_i),\ldots(x_n,y_n)$ where $n=(r-1)(s-1)$. Now order the $y$-coordinates {\it by the value of the $x$-coordinates}. Any increasing subsequence can take at most one element from each column, so at most $r-1$, and any decreasing subsequence can take at most one element from each row, so at most $s-1$.

\section 8. ERROR-CORRECTING CODES

Suppose we have two people who need to communicate over a channel that is ``noisy'' in the sense that errors may be introduced into the message. To detect and solve this issue, checksums may be used. Suppose we have an alphabet $\Sigma = \{0,\ldots,q-1\}$ where $q$ is prime. Let $x$ denote a message $x\in \Sigma^k$ for some positive integer $k$. Then the {\it checksum} $c$ of $x$ is $\sum_{i=1}^k x_i \pmod{q}$. Then the message becomes $y\in {\bf F}_q^{k+1}$ where
$$ y_i = \cases{
    x_i & for $i\leq k$ \cr
    c   & for $i = k+1$
}$$
Since $q$ is prime, for any $x,y\in {\bf F}_q^kk$, if $x$ and $y$ differ at {\it exactly one} place, then their checksums will differ. But what happens if more errors occur? Also, is it possible for the receiver to determine where the error occured without asking the sender to resend the message?

To formalise the problem further, we introduce some more definitions. The {\it block length} $n$ is the length of the encoded message. Suppose the original (non-encoded) message has length $k$. Then the {\it encoding function} $E : \Sigma^k \rightarrow \Sigma^n $ is applied by the sender before transmission and the {\it decoding function} $D : \Sigma^n \rightarrow \Sigma^k$ is applied by the receiver upon the message's arrival. The {\it code} $C\subseteq \Sigma^n$ is the image of $E$. A code is {\it binary} if $\Sigma = \{0,1\}$. Checksums work by increasing the distance between two messages $x,y\in \Sigma^k$ if $x\neq y$.

The {\it Hamming distance} between two messages $x,y\in \Sigma^m$ is given by
$$ \Delta(x,y) = \sum_{i=1}^m [x_i\neq y_i].$$
For a code $C\subseteq \Sigma^n$, the {\it minimum distance} of $C$ is
$$ \Delta(C) = \min_{x,y\in C} \{\Delta(x,y)\}.$$
An {\it $(n,k,d)_q$-code} is a code $C$ such that
\medskip
\item {i)} $C\subseteq \Sigma^n$;
\smallskip
\item {ii)} $k = \log_q |C|$;
\smallskip
\item {iii)} and $d = \Delta(C)$.
\medskip
Note that $|C|$ is the size of the code and $k$ need not be an integer. A checksum is a $(n, n-1, 2)_q$-code: Suppose that $x\in C$. Let $y$ be such that $\Delta(x,y) = 1$. Is it possible that $y\in C$? No, because for all $x,y\in C$, if $x\neq y$ then $\Delta(x,y)\geq 2$.

If, for a code $C$, we can be sure that at most $r$ errors occur during transmission, then:
\medskip
\item {i)} If $\Delta(C) \geq r+1$, then the error can be {\it detected} because for any $x\in C$, if we have $y$ that differs from $x$ at not less than $r+1$ places, i.e.\  $\Delta(x,y)\leq r$, then $y\notin C$.
\smallskip
\item {ii)} If $r\leq \lfloor (\Delta(C) - 1)/2 \rfloor$, then the error can be {\it corrected}. Construct a graph with $V = \Sigma^n$ with an edge between $x,y\in \Sigma^n$ if $\Delta(x,y) = 1$. Now if we have $y\in \Sigma^n$ which we know is a garbled message, we can find the closest valid message in the graph. If the error is not too large, then the closest valid message will have been the intended one.
\medskip

The general aim is to construct codes with small $n$ and large $\Delta(C)$. For any finite projective plane $(X,{\cal L})$ of order $m$, we can construct a $(m^2+m+1, \log_2(m^2+m+1), 2m)_2$-code by associating every letter in the alphabet with a characteristic vector of $L\in {\cal L}$. For example, the letter associated with the line $\{1,5,6\}$ in the Fano plane will be coded as $1000110$. Since for any $L_1, L_2\in {\cal L}, L_1\neq L_2$, $|L_1\cap L_2| = 1$ and each of $L_1, L_2$ have $m+1$ points, $\Delta(y_{L_1}, y_{L_2}) = 2m$, where $y_{L_i}$ is the letter associated with the line $L_i$.

A {\it linear} code $C\subseteq {\bf F}_q^n$ is a linear subspace of ${\bf F}_q$, i.e.\  for all $x,y\in C$, $\alpha\in {\bf F}_q$, $\alpha x+y\in C$. There exist $k$ independent vectors $x_1,\ldots, x_k\in {\bf F}_q^n$ such that
$$C = \{\sum_{i=1}^k \alpha_ix_i : \alpha_i,\ldots \alpha_n\in {\bf F}_q\}.$$
More succinctly, we can express $C$ using the generator matrix $G\in {\bf F}_q^{n\times k}$. The matrix $G = \matrix{ x_1 & \cdots & x_k }$ and $C = \{G_\alpha : \alpha \in {\bf F}_q^k\}$. (Multiplying any message by $G$ on the left encodes it.)

Alternatively, represent $C$ by the {\it null space}. Recall that for any $k$-dimensional linear subspace $C$, there is an $(n-k)$-dimensional linear subspace $C^\perp$ such that for all $x\in C, y\in C^\perp$, $x^Ty = 0$. Using the generator matrix $H\in {\bf F}_k^{n\times (n-k)}$ of $C^\perp$, express $C = \{ x : x^TH = 0\}$. Then $H$ is called the {\it parity check} matrix of $C$. In a sense, linear codes are simple because we can detect errors simply by computing $x^TH$.

The {\it support} of a vector $x$ is the set $\{i : x_i \neq 0\}$ and the {\it Hamming weight} of $x$ is $\hbox{wt} x = |\{i : x_i \neq 0\}|$.

\proclaim Theorem. For a linear code $C$,
$$ \Delta(C) = \min\{\hbox{wt} x : x\in C\ \hbox{and}\ x\neq 0\}.$$

\proof For ease of notation, let $d$ denote $\min\{\hbox{wt} x : x\in C\ \hbox{and}\ x\neq 0\}$. First we show that $d\leq \Delta(C)$. Let $y,z\in C$ be such that $\Delta(y,z)$ = $\Delta(C)$. Since $C$ is linear, $x = y-z \in C$. Note that $\Delta(y,z) = |\{i : x_i \neq 0\}|$, so $d \leq \hbox{wt} x = \Delta(C)$.

Now we show that $d\geq \Delta(C)$. Note that $0\in C$. Let $x\neq 0$ be such that $\hbox{wt} x = d$. Then $\Delta(C)\leq \Delta(0,x) = |\{i : x_i \neq 0\}| = \hbox{wt} x = d$.\slug

We can use this theorem to construct a linear code with $\Delta(C) \geq 3$. We find a parity check matrix $H$ such that $x^TH \neq 0$ for any non-zero $x$ with $\hbox{wt} x\in \{1,2\}$. For simplicity, we will work with $q=2$. If $\hbox{wt} x = 1$, then $X$ has exactly one 1-entry, say it is at position $i$. So $x^TH$ is the $i$-th row $H_i$ of $H$. We need that $H_i$ is non-zero for every $i$. If $\hbox{wt} x = 2$, then $x^T$ has exactly two non-zero elements, say they are at positions $i$ and $j$. Then $x^TH$ = $(H_i + H_j)^T$ hence we need $H_i\neq H_j$ for all $i\neq j$.

If all rows of $H$ are distinct and non-zero, then we get $\Delta(C) \geq 3$. The largest number of such rows if $H\in {\bf F}_2^{n\times l}$ is $2^l-1$ so $n = 2^l-1$ for any $l = n-k$. Hence we obtain an $(n, n-\log_2(n+1), 3)_2$-code called the {\it Hamming code}.

Let $x\in {\bf F}_q^n$. The {\it ball} around $x$ of radius $r$ is given by
$$B(x,r) = \{y\in {\bf F}_q^n : \Delta(x,y) \leq r\}.$$
The {\it volume} of such a ball is
$$\hbox{vol}(r,n) = |B(x,r)| = \sum_{i=0}^r {n\choose i} (q-1)^i.$$

\thmparens{S}{Hamming bound} If an $(n,k,d)_q$-code exists, then
$$q^k\cdot\hbox{vol}\left(\lfloor {d-1\over 2} \rfloor, n\right) \leq q^n.$$

\proof Let $r = \lfloor (d-1)/2 \rfloor$ and consider $\bigcup_{x\in C} B(x,r)\subset \{1,\ldots,q\}^n$. By observation,
$$|\bigcup_{x\in C} B(x,r)| = \sum_{x\in C} \hbox{vol}(r,n) = q^k\cdot \hbox{vol}(r,n) \leq q_n.\noskipslug$$

Note that if $d = 3$ and $q = 2$ then the Hamming bound is $2^k(n+1)\leq 2^n$ and the Hamming code matches this bound with equality as $k=n-log_2(n+1)$. An $(n,k,d)_2$-code is called {\it perfect} if $q^k\cdot\hbox{vol}(\lfloor (d-1)/2 \rfloor, n) = q^n$. There are no codes with larger size $k$ than a $(n, n-d+1, d)_2$-code (even if we let $q>2$), as the following theorem shows.

\thmparens{B}{Singleton bound} If $C$ is an $(n,k,d)_q$-code, then $k\leq n-d+1$.

\proof Define a function $f:\Sigma^n\rightarrow\Sigma^{k-1}$ such that $f(x_1,\ldots, x_n) = (x_1,\dots, x_{k-1})$. As $|\Sigma^{k-1}| = q^{k-1}$ and $|C| = q^k > q^{k-1}$, there are $x,y\in C$, with $x\neq y$ and $f(x) = f(y)$. So $x$ and $y$ can only differ in the last $n-k+1$ entries. Since $d = \Delta(C) \leq \Delta(x,y) \leq n-k+1$, $k\leq n-d+1$.\slug

An $(n,k,d)_q$ code is called {\it maximum-distance separable} or MDS if $k = n-d+1$.

\end

### 2019-06-15 misc/study-notes/nopt048notes.tex
Added graph theory notes and TeX macros

@@
-\documentclass[12pt]{article}
-\usepackage[margin=1in]{geometry}
-\usepackage{amsmath,amsthm,amssymb,amsfonts,tikz,tikzsymbols}
-\usepackage[shortlabels]{enumitem}
+% Notes for NOPT048 Linear Programming and Combinatorial Optimization
+% Charles University in Prague, Summer 2019
+% Instructor: Prof. Hans Raj Tiwary
+% Notes by Marcel Goh
 
-\newcommand{\N}{\mathbb{N}}
-\newcommand{\Z}{\mathbb{Z}}
-\newcommand{\Q}{\mathbb{Q}}
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\C}{\mathbb{C}}
+\input fontmac
+\input mathmac
 
-\newcommand{\M}{\mathbb{M}}
+\def\conv{\hbox{conv}\,}
+\def\cone{\hbox{cone}\,}
+\def\aff{\hbox{aff}\,}
 
-\newcommand{\F}{\mathcal{F}}
-\newcommand{\pp}{\mathcal{P}}
+\def\maximise{\hbox{maximise}\hfill}
+\def\minimise{\hbox{minimise}\hfill}
+\def\subject{\hbox{subject to}\hfill}
+\def\hcr{\hfill\cr}
 
-\DeclareMathOperator{\dom}{dom}
-\DeclareMathOperator{\conv}{conv}
-\DeclareMathOperator{\cone}{cone}
-\DeclareMathOperator{\aff}{aff}
+\leftrighttop{NOPT048 LINEAR PROGRAMMING AND COMBINATORIAL OPTIMIZATION}{MARCEL GOH}
 
-\renewcommand\qedsymbol{$\Smiley$}
+\maketitle{NOPT048 Linear Programming and Combinatorial Optimization}{Notes by}{Marcel K. Goh {\rm (Prague, Czech Rep.)}}{9 June 2019}
 
-\newenvironment{problem}[2][Problem]{\begin{trivlist}
-\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
-\newenvironment{definition}[2][Definition]{\begin{trivlist}
-\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
+\section 1. DEFINITIONS AND THEOREMS
 
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}
-\newtheorem*{theorem*}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem*{lemma*}{Lemma}
+\vskip -\medskipamount
 
-\begin{document}
+\section 1.1. Linear and integer programs
 
-\title{NOPT048 Notes}
-\author{Marcel Goh}
-\maketitle
+The following is a {\it linear program}:
+$$\matrix{
+    & \maximise & \quad & c^Tx\hcr
+    & \subject & & Ax \leq b.\hfill
+}$$
 
-\section{Definitions and theorems}
-\subsection{Linear and integer programs}
-The following is a \textbf{linear program}:
-\begin{alignat*}{2}
-    & \text{maximise} & \quad & c^Tx \\
-    & \text{subject to} & & Ax \leq b
-\end{alignat*}
 Here $x, b, c\in \R^n$ and $A$ is an $m\times n$ matrix over $\R$.
-The function $c^Tx$ is called the \textbf{objective function} and it is what the linear program is trying to maximise (or minimise).
-Any vector $x$ that satisfies the inequality $Ax \leq b$ is called a \textbf{feasible solution}.
-A vector $x^*$ is called an \textbf{optimum solution} if $c^Tx^* \geq c^Tx$ for all feasible solutions $x$.
-A linear program is called \textbf{infeasible} if there are no feasible solutions.
-Even if a linear program is feasible, it might be \textbf{unbounded}.
-This happens when its objective function can be made arbitrarily "good".\\
-\indent In a certain sense, linear programs are easy to solve. Adding an integrality constraint to a linear program gives an \textbf{integer program}:
-\begin{alignat*}{2}
-    & \text{maximise} & \quad & c^Tx \\
-    & \text{subject to} & & Ax \leq b\\
-    &                   & & x \in \Z^n
-\end{alignat*}
+The function $c^Tx$ is called the {\it objective function} and it is what the linear program is trying to maximise (or minimise).
+Any vector $x$ that satisfies the inequality $Ax \leq b$ is called a {\it feasible solution}.
+A vector $x^*$ is called an {\it optimum solution} if $c^Tx^* \geq c^Tx$ for all feasible solutions $x$.
+A linear program is called {\it infeasible} if there are no feasible solutions.
+Even if a linear program is feasible, it might be {\it unbounded}.
+This happens when its objective function can be made arbitrarily ``good''.
+
+In a certain sense, linear programs are easy to solve. Adding an integrality constraint to a linear program gives an {\it integer program}:
+$$\matrix{
+    & \maximise & \quad & c^Tx \hcr
+    & \subject & & Ax \leq b,\hcr
+    &                   & & x \in \Z^n.\hfill
+}$$
 Integer programs are much harder to solve in the general case. Suppose we want to solve the following integer program:
-\begin{alignat*}{2}
-    & \text{maximise} & \quad & c^Tx \\
-    & \text{subject to} & & Ax \leq b\\
-    &                   & & x \in \{0,1\}
-\end{alignat*}
-It is often a good strategy to solve the \textbf{linear program relaxation} of the integer program, which gives an approxmiate answer:
-\begin{alignat*}{2}
-    & \text{maximise} & \quad & c^Tx \\
-    & \text{subject to} & & Ax \leq b\\
-    &                   & & 0\leq x\leq 1
-\end{alignat*}
+$$\matrix{
+    & \maximise & \quad & c^Tx \hcr
+    & \subject & & Ax \leq b,\hcr
+    &                   & & x \in \{0,1\}.\hfill
+}$$
+It is often a good strategy to solve the {\it linear program relaxation} of the integer program, which gives an approxmiate answer:
+$$\matrix{
+    & \maximise & \quad & c^Tx \hcr
+    & \subject & & Ax \leq b,\hcr
+    &                   & & 0\leq x\leq 1.\hfill
+}$$
 Then we can take the solution of the linear program and round each element to the nearest integer value to get a solution of the original integer program.
@@
 
-\subsection{Geometric interpretation}
-The inequality $Ax\leq b$ defines a \textbf{polyhedron}.
-More precisely, we say that a polyhedron is the intersection of finitely many \textbf{halfspaces}, where a halfspace is the set $\{x : a^Tx\leq b\}$ for some $a\in \R^n$ and $b\in \R$.
-If a polyhedron is bounded, we call it a \textbf{polytope}.
+\section 1.2. Geometric interpretation
+
+The inequality $Ax\leq b$ defines a {\it polyhedron}.
+More precisely, we say that a polyhedron is the intersection of finitely many {\it halfspaces}, where a halfspace is the set $\{x : a^Tx\leq b\}$ for some $a\in \R^n$ and $b\in \R$.
+If a polyhedron is bounded, we call it a {\it polytope}.
 A polyhedron is closed and convex.
-The finite intersection of closed sets is closed and the intersection of convex sets is convex.\\
-\indent Let $S\subset \R^n$ be a a set of $m$ points. Then the \textbf{convex hull} of $S$ is the set
-$$\conv S = \{x : x = \sum_{i=1}^m\lambda_i s_i, \sum_{i=1}^m\lambda_i = 1, \lambda_i \geq 0\}.$$
-The \textbf{conic hull} of $S$ is the set
-$$\cone S = \{x : x = \sum_{i=1}^m\lambda_i s_i, \lambda_i\geq 0\}.$$
-The \textbf{affine hull} of $S$ is the set
-$$\aff S = \{x : x = \sum_{i=1}^m\lambda_i s_i, \sum_{i=1}^m\lambda_i = 1\}.$$
-The convex hull can be obtained from the conic hull by intersecting it with a hyperplane. The affine hull is the smallest-dimensional subspace that contains all the points in $S$.\\
-\indent Let $P = \{x: Ax\leq b\}$ be a polyhedron.
-An inequality $\alpha^Tx \leq \beta$ is said to be a \textbf{valid inequality} if and only if for all $y\in P$, $\alpha^T y\leq \beta$.
-Then a subset $F\subset P$ is a \textbf{face} of $P$ if and only if there exists a valid inequality $\alpha^T x\leq \beta$ such that $F = P\cap \{x : \alpha^T x\leq \beta\}$.
+The finite intersection of closed sets is closed and the intersection of convex sets is convex.
+
+Let $S\subset \R^n$ be a a set of $m$ points. Then the {\it convex hull} of $S$ is the set
+$$\conv S = \{x : x = \displaystyle\sum_{i=1}^m\lambda_i s_i, \displaystyle\sum_{i=1}^m\lambda_i = 1, \lambda_i \geq 0\}.$$
+The {\it conic hull} of $S$ is the set
+$$\cone S = \{x : x = \displaystyle\sum_{i=1}^m\lambda_i s_i, \lambda_i\geq 0\}.$$
+The {\it affine hull} of $S$ is the set
+$$\aff S = \{x : x = \displaystyle\sum_{i=1}^m\lambda_i s_i, \displaystyle\sum_{i=1}^m\lambda_i = 1\}.$$
+The convex hull can be obtained from the conic hull by intersecting it with a hyperplane. The affine hull is the smallest-dimensional subspace that contains all the points in $S$.
+
+Let $P = \{x: Ax\leq b\}$ be a polyhedron.
+An inequality $\alpha^Tx \leq \beta$ is said to be a {\it valid inequality} if and only if for all $y\in P$, $\alpha^T y\leq \beta$.
+Then a subset $F\subset P$ is a {\it face} of $P$ if and only if there exists a valid inequality $\alpha^T x\leq \beta$ such that $F = P\cap \{x : \alpha^T x\leq \beta\}$.
 The trivial faces of any polyhedron $P$ are $\emptyset$ and $P$ itself (formed by intersecting $P$ with the valid inequalities $0^Tx \leq 1$ and $0^Tx\leq 0$ respectively).
-Any face that is not trivial is called \textbf{proper}.\\
-\indent The \textbf{dimension} of a polyhedron $P$, denoted $\dim P$, is the largest dimension of a ball inside $P$.
-A \textbf{vertex} is a face of dimension 0, an \textbf{edge} is a face of dimension 1, and a \textbf{facet} is a face of dimension $\dim P - 1$.
-A point $x\in P$ is called an \textbf{extreme point} if it is not a convex combination of points of $P$.
+Any face that is not trivial is called {\it proper}.
+
+The {\it dimension} of a polyhedron $P$, denoted $\dim P$, is the largest dimension of a ball inside $P$.
+A {\it vertex} is a face of dimension 0, an {\it edge} is a face of dimension 1, and a {\it facet} is a face of dimension $\dim P - 1$.
+A point $x\in P$ is called an {\it extreme point} if it is not a convex combination of points of $P$.
 Note that a point $x$ is an extreme point if and only if it is a vertex.
-\begin{theorem*}{\textup{(Minkowski-Weyl)}}
-    A set of points $P$ is a polyhedron if and only if there exist finite sets $V$ and $Y$ such that $P = \conv V + \cone Y$ where the sum $A + B$ denotes the set $\{a + b : a\in A, b\in B\}$.
-\end{theorem*}
 
-\subsection{Simplex method}
-A polyhedron is \textbf{pointed} if it does not contain a line.
-\begin{lemma*}
-    Every non-empty pointed polyhedron $P$ has at least one vertex.
-\end{lemma*}
-\begin{theorem*}
-    Consider the linear program
-    \begin{alignat*}{2}
-        & \textup{maximise} & \quad & c^Tx \\
-        & \textup{subject to} & & Ax \leq b.
-    \end{alignat*}
+\thmparens{M}{Minkowski-Weyl} A set of points $P$ is a polyhedron if and only if there exist finite sets $V$ and $Y$ such that $P = \conv V + \cone Y$ where the sum $A + B$ denotes the set $\{a + b : a\in A, b\in B\}$.
+
+\section 1.3. Simplex method
+
+A polyhedron is {\it pointed} if it does not contain a line.
+\proclaim Lemma. Every non-empty pointed polyhedron $P$ has at least one vertex.
+
+\proclaim Theorem. Consider the linear program
+$$\matrix{
+    & \hbox{\rm maximise}\hfill & \quad & c^Tx \hcr
+    & \hbox{\rm subject to}\hfill & & Ax \leq b.\hfill
+}$$
 If the polyhedron $P$ defined by $Ax\leq b$ is pointed and if an optimum exists, then there exists a vertex of $P$ that attains the maximum.
-\end{theorem*}
-The result of the above statements is an algorithm for solving linear programs, called the \textbf{simplex method}. Here is the general idea:
-\begin{enumerate}
-    \item Find an initial vertex. It is a 0-dimensional face with some 1-dimensional faces incident on it.
-    \item At any vertex, check for optimality. If optimal, we are done.
-    \item If the vertex is not optimal, increase the value of the objective function by following an edge to a new vertex. Return to step 2.
-\end{enumerate}
-\indent Let $A$ be an $m\times n$ matrix, $b\in \R^m$, $x \in\R^n$, with $n \geq m$.
+
+The result of the above statements is an algorithm for solving linear programs, called the {\it simplex method}. Here is the general idea:
+
+\algbegin Algorithm X (Simplex method). Find an optimum solution of a linear program.
+
+\algstep X1. Find an initial vertex. It is a 0-dimensional face with some 1-dimensional faces incident on it.
+\algstep X2. At any vertex, check for optimality. If optimal, we are done.
+\algstep X3. If the vertex is not optimal, increase the value of the objective function by following an edge to a new vertex. Return to step X2.\slug
+
+Let $A$ be an $m\times n$ matrix, $b\in \R^m$, $x \in\R^n$, with $n \geq m$.
 Let $B\subset [n] =\{1,\ldots,n\}$ of cardinality $m$.
 Then $A_B$ denotes the square matrix formed by selecting the columns of $A$ indexed by the elements of $B$ and likewise $x_B$ denotes the vector of length $m$ formed by selecting the elements indexed by the elements of $B$.
-We define a vector $x$ to be a \textbf{basic feasible solution} of the linear program
-\begin{alignat*}{2}
-    & \textup{maximise} & \quad & c^Tx \\
-    & \textup{subject to} & & Ax \leq b
-\end{alignat*}
+We define a vector $x$ to be a {\it basic feasible solution} of the linear program
+$$\matrix{
+    & \maximise & \quad & c^Tx \hcr
+    & \subject & & Ax \leq b\hfill
+}$$
 if and only if there exists $B\subset [n]$ with $|B|= m$ such that:
-\begin{enumerate}[i)]
-    \item $A_B$ is non-singular
-    \item $x_j = 0$ for all $j\notin B$
-    \item $x_B = A_B^{-1}b$
-\end{enumerate}
-A set $B\subset [n]$ is called a \textbf{basis} if and only if $A_B$ is non-singular, and $B$ is a \textbf{feasible basis} if the unique solution of $A_Bx_B=b$ is non-negative.
-Note that a basic feasible solution is uniquely determined by the feasible basis $B$, but a vertex does \textit{not} have to be determined by a unique basis.\\
-\indent A matrix $A$ is \textbf{totally unimodular} if and only if every square submatrix of $A$ has determinant 0, 1, or $-1$.
+\medskip
+\item {i)} $A_B$ is non-singular;
+\smallskip
+\item {ii)} $x_j = 0$ for all $j\notin B$;
+\smallskip
+\item {iii)} $x_B = A_B^{-1}b$.
+\medskip
+A set $B\subset [n]$ is called a {\it basis} if and only if $A_B$ is non-singular, and $B$ is a {\it feasible basis} if the unique solution of $A_Bx_B=b$ is non-negative.
+Note that a basic feasible solution is uniquely determined by the feasible basis $B$, but a vertex does {\it not} have to be determined by a unique basis.
 
-\subsection{Duality}
+A matrix $A$ is {\it totally unimodular} if and only if every square submatrix of $A$ has determinant 0, 1, or $-1$.
+
+\section 1.4. Duality
+
 As always, we start with the linear program
-\begin{alignat*}{2}
-    & \textup{maximise} & \quad & c^Tx \\
-    & \textup{subject to} & & Ax \leq b\\
-    &                     & & x \geq 0.
-\end{alignat*}
+$$\matrix{
+    & \maximise & \quad & c^Tx \hcr
+    & \subject & & Ax \leq b,\hcr
+    &                     & & x \geq 0.\hfill\hfill
+}$$
 Inequalities of a linear program can be added together without changing the feasible region.
@@
 So we can manipulate inequalities to get an upper bound on the optimum value of the objective function.
-We want to pick coefficients to get the \textit{least} upper bound possible.
+We want to pick coefficients to get the {\it least} upper bound possible.
 Let $y$ denote a vector of coefficients.
 We want to
-\begin{alignat*}{2}
-    & \textup{minimise} & \quad & b^Ty \\
-    & \textup{subject to} & & A^Ty \geq c\\
-    &                     & & y \geq 0
-\end{alignat*}
-The original linear program is called the \textbf{primal} and this new one is called the \textbf{dual}.
+$$\matrix{
+    & \minimise & \quad & b^Ty \hcr
+    & \subject & & A^Ty \geq c,\hcr
+    &                     & & y \geq 0.\hfill
+}$$
+The original linear program is called the {\it primal} and this new one is called the {\it dual}.
 Any feasible solution of the dual linear program gives an upper bound on the primal and the optimum of the dual is equal to the optimum of the original program.
 
-\begin{theorem*}{\textup{(Weak Duality Theorem)}}
-    Let $P$ and $D$ be primal and dual linear programs of the above form and let $x$ and $y$ be respective feasible solutions. Then $c^Tx \leq b^Ty$.
-\end{theorem*}
+\thmparens{W}{Weak Duality Theorem} Let $P$ and $D$ be primal and dual linear programs of the above form and let $x$ and $y$ be respective feasible solutions. Then $c^Tx \leq b^Ty$.
 
-\begin{theorem*}{\textup{(Strong Duality Theorem)}}
-    Let $P$ and $D$ be primal and dual linear programs. Then exactly one of the following holds:
-    \begin{enumerate}[i)]
-        \item P infeasible, D infeasible
-        \item P infeasible, D unbounded
-        \item P unbounded, D infeasible
-        \item P and D both feasible and bounded and furthermore, if $x^*$ and $y^*$ are primal and dual optimal solutions, then $c^Tx^* = b^Ty^*$.
-    \end{enumerate}
-\end{theorem*}
+\thmparens{S}{Strong Duality Theorem} Let $P$ and $D$ be primal and dual linear programs. Then exactly one of the following holds:
+\medskip
+\item {i)} P infeasible, D infeasible
+\smallskip
+\item {ii)} P infeasible, D unbounded
+\smallskip
+\item {iii)} P unbounded, D infeasible
+\smallskip
+\item {iv)} P and D both feasible and bounded and furthermore, if $x^*$ and $y^*$ are primal and dual optimal solutions, then $c^Tx^* = b^Ty^*$.
+\medskip
 
-\begin{lemma*}{\textup{(Farkas lemma)}}
+\lemparens{F}{Farkas lemma}
 For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
-    \begin{enumerate}[i)]
-        \item There exists a vector $x$ such that $Ax = b$ and $x \geq 0$.
-        \item There exists a vector $y$ such that $A^Ty \geq 0$ and $b^Ty< 0$.
-    \end{enumerate}
-\end{lemma*}
+\medskip
+\item {i)} There exists a vector $x$ such that $Ax = b$ and $x \geq 0$.
+\smallskip
+\item {ii)} There exists a vector $y$ such that $A^Ty \geq 0$ and $b^Ty< 0$.
+\medskip
 
-\begin{theorem*}{\textup{(Farkas lemma 2.0)}}
+\lemparens{G}{Farkas lemma 2.0}
 For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
-    \begin{enumerate}[i)]
-        \item There exists a vector $x$ such that $Ax \leq b$ and $x \geq 0$.
-        \item There exists a vector $y$ such that $A^Ty \geq 0$, $y\geq0$ and $b^Ty< 0$.
-    \end{enumerate}
-\end{theorem*}
+\medskip
+\item {i)} There exists a vector $x$ such that $Ax \leq b$ and $x \geq 0$.
+\smallskip
+\item {ii)} There exists a vector $y$ such that $A^Ty \geq 0$, $y\geq0$ and $b^Ty< 0$.
+\medskip
 
-\subsection{Applications}
-\subsubsection{Matching and vertex cover in bipartite graphs}
-Given a graph $G=(V,E)$, a \textbf{matching} is a set $M\subset E$ such that for all $v\in V$ there is at most one edge $e\in M$ with $v\in e$.
+\section 1.5. Applications
+
+\vskip -\medskipamount
+
+\section 1.5.1. Matching and vertex cover in bipartite graphs
+
+Given a graph $G=(V,E)$, a {\it matching} is a set $M\subset E$ such that for all $v\in V$ there is at most one edge $e\in M$ with $v\in e$.
 The linear program that finds the maximal matching in a bipartite graph is
-\begin{alignat*}{2}
-    & \textup{maximise} & \quad & \sum_{e\in E} x_e \\
-    & \textup{subject to} & & \sum_{v\in e}x_e \leq 1,\ \text{for all}\ v\in V\\
-    &                     & & x_e \geq 0,\ \text{for all}\ e\in E.
-\end{alignat*}
+$$\matrix{
+    & \maximise   & \quad & \displaystyle\sum_{e\in E} x_e  \hfill    & \hcr
+    & \subject &       & \displaystyle\sum_{v\in e}x_e \leq 1  \hfill &\hbox{for all}\ v\in V,\hcr
+    &                   &       & x_e \geq 0      \hfill          &\hbox{for all}\ e\in E.\hfill
+}$$
 The dual of this linear program is the following:
-\begin{alignat*}{3}
-    & \textup{minimise} & \quad & \sum_{v\in V} y_v &\\
-    & \textup{subject to} & & y_u + y_v \geq 1,\  \text{for all}\ \{u,v\}\in E\\
-    &                     & & y_v \geq 0, \  \text{for all}\ v\in V.
-\end{alignat*}
-    Given a graph $G=(V,E)$, a \textbf{vertex cover} is a set $S \subset V$ such that for all $\{u, v\}\in E$, $\{u,v\}\cap S\geq 1$.
+$$\matrix{
+    & \minimise   & \quad & \displaystyle\sum_{v\in V} y_v  \hfill & \hcr
+    & \subject &       & y_u + y_v \geq 1   \hfill &\hbox{for all}\ \{u,v\}\in E,\hcr
+    &                   &       & y_v \geq 0        \hfill  &\hbox{for all}\ v\in V.\hfill
+}$$
+    Given a graph $G=(V,E)$, a {\it vertex cover} is a set $S \subset V$ such that for all $\{u, v\}\in E$, $\{u,v\}\cap S\geq 1$.
 The dual linear program above finds the minimum vertex cover in a bipartite graph.
 The duality of these two programs results in K\H{o}nig's theorem.
-\begin{theorem*}{\textup{(K\H{o}nig's theorem)}}
-    Suppose $G=(V,E)$ is a bipartite graph. Let $M\subset E$ denote the maximum matching and let $S\subset V$ denote the minimum vertex cover. Then $|M|=|S|$.
-\end{theorem*}
+\thmparens{K}{K\H{o}nig's theorem}
+    Suppose $G=(V,E)$ is a bipartite graph. Let $M\subset E$ denote the maximum matching and let $S\subset V$ denote the minimum vertex cover. Then $|M|=|S|$.\slug
 
-\subsubsection{Matching in weighted graphs}
+\section 1.5.2. Matching in weighted graphs
+
 Let $G=(V,E)$ be a graph (not necessarily bipartite) with a weight function $w:E\rightarrow \R^+$.
 Suppose we want to find the maximum weight matching in $G$.
-For any subset $E'\subset E$, the \textbf{characteristic vector} of $E'$ is the vector $\chi^{E'}$ given by
-$$ \chi_e^{E'} = \begin{cases} 1 & e\in E' \\ 0 & e\notin E'\end{cases} $$
-for all $e\in E$. Then the \textbf{matching polytope} of the graph $G$ is the set
-$$ M(G) = \conv\{ x \in \{0,1\}^{|E|} : x\ \text{is the characteristic vector of some matching in}\ G\}.$$
+For any subset $E'\subset E$, the {\it characteristic vector} of $E'$ is the vector $\chi^{E'}$ given by
+$$\chi_e^{E'} = \cases{
+    1 & if $e\in E'$ \hcr
+    0 & if $e\notin E'$\hfill
+}$$
+for all $e\in E$. Then the {\it matching polytope} of the graph $G$ is the set
+$$ M(G) = \conv\{ x \in \{0,1\}^{|E|} : x\ \hbox{is the characteristic vector of some matching in}\ G\}.$$
 To get the maximum weight matching, we need only solve the linear program
-\begin{alignat*}{2}
-    & \textup{maximise} & \quad & w^Tx \\
-    & \textup{subject to} & & x\in M(G).
-\end{alignat*}
+$$\matrix{
+    & \maximise & \quad & w^Tx \hcr
+    & \subject & & x\in M(G).\hfill
+}$$
 What does the polytope $M(G)$ look like in terms of inequalities? It turns out that $M(G)$ is the set of all vectors $x$ that satisfy
-\begin{enumerate}
-    \item $\displaystyle\sum_{v\in e}x_e \leq 1$ for all $v\in V$ and
-    \item $\displaystyle\sum_{e\in E[U]}x_e \leq \frac{|U|-1}{2}$ for every set $U\subset V$ of odd cardinality.
-\end{enumerate}
-In a graph $G=(V,E)$, a matching $M\subset E$ is \textbf{perfect} if every vertex in $V$ is incident to some edge in $M$. The perfect matching polytope is the set
-$$ PM(G) = \conv\{ x \in \{0,1\}^{|E|} : x\ \text{is the characteristic vector of some perfect matching in}\ G\}.$$
+\medskip
+\item {1.} $\displaystyle\sum_{v\in e}x_e \leq 1$ for all $v\in V$ and
+\smallskip
+\item {2.} $\displaystyle\sum_{e\in E[U]}x_e \leq {|U|-1\over 2}$ for every set $U\subset V$ of odd cardinality.
+\medskip
+In a graph $G=(V,E)$, a matching $M\subset E$ is {\it perfect} if every vertex in $V$ is incident to some edge in $M$. The perfect matching polytope is the set
+$$ PM(G) = \conv\{ x \in \{0,1\}^{|E|} : x\ \hbox{is the characteristic vector of some perfect matching in}\ G\}.$$
 For a set $U\subset V$ of vertices, let $\delta(U)$ denote the number of edges that have one endpoint in $U$ and one endpoint outside $U$.
 Then we find that $PM(G)$ is the set of all vectors $x$ satisfying
-\begin{enumerate}
-    \item $\displaystyle\sum_{v\in e}x_e = 1$ for all $v\in V$,
-    \item $\displaystyle\sum_{e\in \delta(U)}x_e\geq 1$ for every set $U\subset V$ of odd cardinality, and
-    \item $x_e \geq 0$ for every $e\in E$.
-\end{enumerate}
+\medskip
+\item {1.} $\displaystyle\sum_{v\in e}x_e = 1$ for all $v\in V$,
+\smallskip
+\item {2.} $\displaystyle\sum_{e\in \delta(U)}x_e\geq 1$ for every set $U\subset V$ of odd cardinality, and
+\smallskip
+\item {3.} $x_e \geq 0$ for every $e\in E$.
+\medskip
 Note that the number of inequalities in these polytopes may be very large. For this reason, a polynomial-time algorithm may not exist.
 
-\subsubsection{The travelling salesman problem}
-Given a graph $G=(V,E)$, a \textbf{closed tour} is a walk in $G$ with no repeated edges and no repeated vertices except for the first and last vertex, which are the same.
-In a graph with edge costs $(c_e : e\in E)$, \textbf{travelling salesman problem} asks us to find the minimum-weight closed tour that contains each vertex exactly once.
+\section 1.5.3. The travelling salesman problem
+
+Given a graph $G=(V,E)$, a {\it closed tour} is a walk in $G$ with no repeated edges and no repeated vertices except for the first and last vertex, which are the same.
+In a graph with edge costs $(c_e : e\in E)$, {\it travelling salesman problem} asks us to find the minimum-weight closed tour that contains each vertex exactly once.
 To solve the travelling salesman problem, we start with the following linear program:
-\begin{alignat*}{2}
-    & \textup{minimise} & \quad & \sum_{e\in E} c_ex_e \\
-    & \textup{subject to} & & \sum_{v\in e} = 2\ \text{for all}\ v\in V,\\
-    &  & & 0\leq x_e \leq 1\ \text{for all}\ e\in E.
-\end{alignat*}
-What we really want to do is add the inequalities $\displaystyle\sum_{e\in \delta(S)} x_e \geq 2$ for every $\emptyset \neq S \neq V$.
-This is called \textbf{subtour elimination} because it forbids any tours smaller than the entire closed tour we're trying to build.
+$$\matrix{
+    & \minimise & \quad & \displaystyle\sum_{e\in E} c_ex_e \hcr
+    & \subject & & \displaystyle\sum_{v\in e} = 2\ \hbox{for all}\ v\in V,\hcr
+    &  & & 0\leq x_e \leq 1\ \hbox{for all}\ e\in E.\hfill
+}$$
+What we really want to do is add the inequalities $\sum_{e\in \delta(S)} x_e \geq 2$ for every $\emptyset \neq S \neq V$.
+This is called {\it subtour elimination} because it forbids any tours smaller than the entire closed tour we're trying to build.
 Adding this inequality, we get the linear program relaxation of the integer program for the travelling salesman problem.
-The problem is that eliminating subtours adds an exponential number of ineqalities to the program.\\
-\indent Instead of adding all of these inequalities, we use the cutting-plane approach.
-Given a polytope $P$, a \textbf{cutting plane} is an inequality that is valid for all integer points in $P$.
-It is called a cutting plane because it may cut the original polytope.
-We can apply this idea as follows:
-\begin{enumerate}
-    \item We start with the linear program without the subtour constraints.
-    \item Find the optimal solution to the linear program.
-    \item If we find that some subtour constraints are violated by the optimal solution, we add these cutting planes to our program and return to step 2.
-If no subtour constraints are violated, we have a solution and we may stop.
-\end{enumerate}
+The problem is that eliminating subtours adds an exponential number of ineqalities to the program.
+Instead of adding all of these inequalities, we approach the problem as follows:
+
+\algbegin Algorithm T (Subtour elimination). This algorithm solves the travelling salesman problem by adding violated subtour constraints one at a time.
+
+\algstep T1. We start with the linear program without the subtour constraints.
+\algstep T2. Find the optimal solution to the linear program.
+\algstep T3.
+If we find that some subtour constraints are violated by the optimal solution, we add these inequalities to our program and return to step T2.
+If no subtour constraints are violated, we have a solution.\slug
+
 Step 3 is our main concern.
@@
 Set capacities $u_e = x_e^*$ for every edge $e$.
-Now for any $S\subset V$, the weight of the cut $(S, V\setminus S)$ equals the value of $\displaystyle\sum_{e\in \delta(S)} x_e$.
+Now for any $S\subset V$, the weight of the cut $(S, V\setminus S)$ equals the value of $\sum_{e\in \delta(S)} x_e$.
 So we can solve the min-cut problem over $G(x^*)$ and if we get a value less than 2, we know that some subtour constraint is violated.
@@
 
-\subsection{Ellipsoid algorithm}
+\section 1.6. Ellipsoid algorithm
+
 So far, we do not know if linear programs can be solved in polynomial time.
 It turns out that optimisation is polynomially equivalent to separating over a polytope.
-Given $x\in \R^n$ and a polyhedron $P$, we separate as follows: Output "yes" if $x\in P$; otherwise, output "no" and a pair $(\alpha, \beta)$ such that $\alpha^Ty\leq\beta$ is a valid inequality for $P$ and $\alpha^T x > \beta$.
-The \textbf{ellipsoid algorithm} uses an oracle for separation to answer the question of feasibility: Given a polytope $P$, find an $x\in P$ or conclude that $P = \emptyset$.\\
-\indent Before we get into the actual method, let's convince ourselves that checking for feasibility is equivalent to optimising.
+Given $x\in \R^n$ and a polyhedron $P$, we separate as follows: Output ``yes'' if $x\in P$; otherwise, output ``no'' and a pair $(\alpha, \beta)$ such that $\alpha^Ty\leq\beta$ is a valid inequality for $P$ and $\alpha^T x > \beta$.
+The {\it ellipsoid algorithm} uses an oracle for separation to answer the question of feasibility: Given a polytope $P$, find an $x\in P$ or conclude that $P = \emptyset$.
+
+Before we get into the actual method, let's convince ourselves that checking for feasibility is equivalent to optimising.
 That optimisation implies feasibility is obvious.
@@
 Can we use that to optimise the linear program
-\begin{alignat*}{2}
-    & \textup{minimise} & \quad & c^Tx\\
-    & \textup{subject to} & & Ax \leq b\\
-    &  & & x\geq 0?
-\end{alignat*}
+$$\matrix{
+    & \minimise & \quad & c^Tx\hcr
+    & \subject & & Ax \leq b,\hcr
+    &  & & x\geq 0?\hfill
+}$$
 Well, from duality we know that if an optimum exists, it will also be the optimum value of the dual linear program.
@@
 So we check if the polyhedron
-$$\{\begin{bmatrix} x\\y\end{bmatrix} : c^Tx = b^Ty, Ax \leq b, A^Ty \geq c, x \geq 0, y\geq 0\}$$
-is feasible and that will give us the optimum value of $x$, if it exists.\\
-\indent On to ellipsoids.
-An \textbf{ellipsoid} $E(A,a)$ is the set $\{x \in \R^n : (x - a)^TA^{-1}(x-a) \leq 1\}$ where $A$ is a \textbf{positive definite} matrix.
+$$\{\pmatrix{ x\hcr y } : c^Tx = b^Ty, Ax \leq b, A^Ty \geq c, x \geq 0, y\geq 0\}$$
+is feasible and that will give us the optimum value of $x$, if it exists.
+
+On to ellipsoids.
+An {\it ellipsoid} $E(A,a)$ is the set $\{x \in \R^n : (x - a)^TA^{-1}(x-a) \leq 1\}$ where $A$ is a {\it positive definite} matrix.
 This means that $(x-a)^TA^{-1}(x-a)$ is positive and moreover, $A = M\cdot M$ for some square matrix $M$.
-If we let $B$ denote the unit ball in $\R^n$, then $E(A,a) = \{Mx + a : x \in B\}$.\\
-\indent For simplicity, assume that $P$ is bounded and assume that $P$ is either empty or full-dimensional.
+If we let $B$ denote the unit ball in $\R^n$, then $E(A,a) = \{Mx + a : x \in B\}$.
+
+For simplicity, assume that $P$ is bounded and assume that $P$ is either empty or full-dimensional.
 Assume also that we have an oracle that outputs a separating hyperplane.
-Then the \textbf{ellipsoid algorithm} determines if $P = \{x : Ax \leq b\}$ is feasible.
-\begin{enumerate}
-    \item Start with an ellipsoid large enough that it definitely contains a point in the polytope, if one exists.
-    \item Let $c$ be the centre of the ellipsoid. Invoke the separation oracle. If $c\in P$, output $c$ and halt.
-    \item If we have reached the maximum amount of iterations, output $P = \emptyset$.
-    \item Otherwise, the oracle gives us a separating hyperplane that cuts the ellipsoid.
+Then the {\it ellipsoid algorithm} determines if $P = \{x : Ax \leq b\}$ is feasible.
+
+\algbegin Algorithm E (Ellipsoid algorithm).
+This algorithm presupposes that we have a separation oracle.
+Given a bounded polytope $P$ and assuming, for the sake of simplicity, that $P$ is either empty or full-dimensional, the algorithm determines either outputs a point $c$ in $P$ or concludes that $P$ is empty.
+
+\algstep E1. Start with an ellipsoid large enough that it definitely contains a point in the polytope, if one exists.
+\algstep E2. Let $c$ be the centre of the ellipsoid. Invoke the separation oracle. If $c\in P$, output $c$ and halt.
+\algstep E3. If we have reached the maximum amount of iterations, output $P = \emptyset$.
+\algstep E4.
+Otherwise, the oracle gives us a separating hyperplane that cuts the ellipsoid.
 Determine which half of the ellipsoid $P$ lies, then draw a new, smaller ellipsoid that contains this half.
-Return to step 2.
-\end{enumerate}
+Return to step E2.\slug
+
 The maximum number of iterations is calculated using a function polynomial in the size of the linear program, so this algorithm runs in polynomial time.
 
-\subsection{Matroids and greedy algorithms}
-Given a graph $G=(V,E)$, a \textbf{spanning tree} is a set $T\subset E$ such that every vertex is incident to some edge in $T$ and $G=(V,T)$ is a contains no cycles.
+\section 1.7. Matroids and greedy algorithms
+
+Given a graph $G=(V,E)$, a {\it spanning tree} is a set $T\subset E$ such that every vertex is incident to some edge in $T$ and $G=(V,T)$ is a contains no cycles.
 Kruskal's algorithm obtains the minimum weight spanning tree by maintaining a forest $F$.
 If $F$ spans every vertex, then it is a spanning tree.
-If not, add the minimum weight edge $e$ such that $F\cup \{e\}$ does not contain a cycle and repeat.\\
-\indent Let $G=(V,E)$ be a graph and let $F$ denote the set of all forests in $G$.
-Then $(E,F)$ is called the \textbf{graphic matroid} of $G$.
+If not, add the minimum weight edge $e$ such that $F\cup \{e\}$ does not contain a cycle and repeat.
+
+Let $G=(V,E)$ be a graph and let $F$ denote the set of all forests in $G$.
+Then $(E,F)$ is called the {\it graphic matroid} of $G$.
 In fact, there are lots of different types of matroids.
-A tuple $M=(E,I)$ where $I\subset \mathcal{P}(E)$ is a \textbf{matroid} if
-\begin{enumerate}[i)]
-    \item $\emptyset\in I$,
-    \item If $X\subset Y$ and $Y\in I$, then $X\in I$,
-    \item If there exist $X,Y \in I$ such that $|Y| > |X|$, then there exists some $y\in Y\setminus X$ such that $X\cup\{y\}\in I$.
-\end{enumerate}
-A system that satisfies i) and ii) is called an \textbf{independence system}.
-Associated with any independence system $M = (E,I)$ is its \textbf{rank function} $r : \mathcal{P}(E)\rightarrow \R$ defined by
+A tuple $M=(E,I)$ where $I\subset {\cal P}(E)$ is a {\it matroid} if
+\medskip
+\item {i)} $\emptyset\in I$,
+\smallskip
+\item {ii)} If $X\subset Y$ and $Y\in I$, then $X\in I$,
+\smallskip
+\item {iii)} If there exist $X,Y \in I$ such that $|Y| > |X|$, then there exists some $y\in Y\setminus X$ such that $X\cup\{y\}\in I$.
+\medskip
+A system that satisfies i) and ii) is called an {\it independence system}. Associated with any independence system $M = (E,I)$ is its {\it rank function} $r : {\cal P}(E)\rightarrow \R$ defined by
 $$r(X) = \max\{ |Y| : Y \subset X, Y\in I\}.$$
-We call $r(E)$ the \textbf{rank} of $M$. A set $S\in I$ such that $|S| = r(E)$ is called a \textbf{base} of $M$. If $M=(E,I)$ is a matroid, then some properties of its rank function $r$ include:
-\begin{enumerate}
-    \item $0\leq r(X) \leq |X|$ for all $x\subset E$.
-    \item If $X\subset Y$ then $r(X)\leq r(Y)$ for all $X, Y\subset E$.
-    \item $r(X\cup Y)+r(X\cap Y) \leq r(X) + r(Y)$ for all $X,Y\subset E$.
-\end{enumerate}
-Property 3 above is called \textbf{submodularity}.
-If $r$ is the rank function of a general independence system, it need only satisfy properties 1 and 2, as well as the weaker property $r(X\cup Y) \leq r(X) + r(Y)$.\\
-\indent Matroids can be directly characterised by the rank function.
-\begin{theorem*}{\textup{(Rank characterisation of matroids)}}
-    Let $E$ be a finite set and suppose that $r : \mathcal{P}(E)\rightarrow \R$ satisfies the three peoperties of the rank function. Then $M=(E,I)$, with
+We call $r(E)$ the {\it rank} of $M$. A set $S\in I$ such that $|S| = r(E)$ is called a {\it base} of $M$. If $M=(E,I)$ is a matroid, then the properties of its rank function $r$ are:
+\medskip
+\item {1.} $0\leq r(X) \leq |X|$ for all $x\subset E$.
+\smallskip
+\item {2.} If $X\subset Y$ then $r(X)\leq r(Y)$ for all $X, Y\subset E$.
+\smallskip
+\item {3.} $r(X\cup Y)+r(X\cap Y) \leq r(X) + r(Y)$ for all $X,Y\subset E$.
+\medskip
+Property 3 above is called {\it submodularity}.
+If $r$ is the rank function of a general independence system, it need only satisfy properties 1 and 2, as well as the weaker property $r(X\cup Y) \leq r(X) + r(Y)$.
+
+Matroids can be directly characterised by the rank function.
+
+\thmparens{R}{Rank characterisation of matroids} Let $E$ be a finite set and suppose that $r : {\cal P}(E)\rightarrow \R$ satisfies the three peoperties of the rank function. Then $M=(E,I)$, with
 $$I = \{Y \subset E : |Y| = r(Y)\},$$
 is a matroid with $r$ as its rank function.
-\end{theorem*}
-To further understand the link between spanning trees and matroids, note that Kruskal's algorithm is an instance of the \textbf{weighted greedy algorithm}.
+
+To further understand the link between spanning trees and matroids, note that Kruskal's algorithm is an instance of the {\it weighted greedy algorithm}.
 Let $M=(E,I)$ be a matroid with a weight function $c$ defined on edges.
 The algorithm proceeds as follows:
-\begin{enumerate}
-    \item Set $S_0\gets\emptyset$, and $E$ is the element set of $M$. Set $i\gets 0$.
-    \item Select $e\in E$ with maximum $c(e)$. Set $E\gets E\setminus\{e\}$.
-    \item If $S_i\cup\{e\}\in I$, set $S_{i+1}\gets S_i\cup\{e\}$, $i\gets i + 1$.
-    \item If $E = \emptyset$, output $S_i$. Otherwise, return to step 2.
-\end{enumerate}
-Then the following two theorems establishes a certain equivalence between matroids and problems that can be solved with the greedy algorithm.
-\begin{theorem*}
-    Let $M=(E,I)$ be a matroid with rank function $r$. The weighted greedy algorithm finds the maximum weighted independent set $S_k$ of cardinality $k$ for each $1\leq k\leq r(E)$.
-\end{theorem*}
-\begin{theorem*}{\textup{(Greedy equivalence)}}
-    Let $M=(E,I)$ be an independence system. Then if the weighted greedy algorithm works for all (non-negative) weights, then $M$ is a matroid.
-\end{theorem*}
-\indent It is worth it to note that spanning trees and matroids have corresponding polytopes.
-For a graph $G=(V,E)$, the spanning tree polytope $ST(G)$ is the set of all vectors $x\in \R^{|E|}$ that satisfy
-\begin{enumerate}
-    \item $x_e \geq 0$,
-    \item $\displaystyle\sum_{e\in E} x_e = |V| - 1$, and
-    \item $\displaystyle\sum_{e\in E[S]} x_e \leq |S| - 1$, for all $S\subsetneq V, |S| \geq 2$.
-\end{enumerate}
-For a matroid $M=(E,I)$ with rank function $r$, the matroid polytope $P_M$ is the set of all vectors $x\in \R^{|E|}$ satisfying $\displaystyle\sum_{e\in T} x_e \leq r(T)$ for all $T\subset E$, where $x_e\geq 0$.
 
-\section{Problems and proofs}
-\begin{problem}{1}
-    Describe how to find an initial basic feasible solution for an arbitrary LP in standard (equational) form.
-\end{problem}
-\begin{proof}[Solution]
-    Suppose the constraints are $Ax = b, x\geq 0$.
+\algbegin Algorithm G (Greedy algorithm). Given a matroid $M=(E,I)$ with a weight function $c$ defined on edges, this algorithm finds the set $S\in I$ with maximum weight.
+
+\algstep G1. Set $S_0\gets\emptyset$, and $E$ is the element set of $M$. Set $i\gets 0$.
+\algstep G2. Select $e\in E$ with maximum $c(e)$. Set $E\gets E\setminus\{e\}$.
+\algstep G3. If $S_i\cup\{e\}\in I$, set $S_{i+1}\gets S_i\cup\{e\}$, $i\gets i + 1$.
+\algstep G4. If $E = \emptyset$, output $S = S_i$. Otherwise, return to step 2.\slug
+
+The following two theorems establishes a certain equivalence between matroids and problems that can be solved with the greedy algorithm.
+
+\proclaim Theorem.
+Let $M=(E,I)$ be a matroid with rank function $r$.
+The weighted greedy algorithm finds the maximum weighted independent set $S_k$ of cardinality $k$ for each $1\leq k\leq r(E)$.\slug
+
+\thmparens{G}{Greedy equivalence} Let $M=(E,I)$ be an independence system. Then if the weighted greedy algorithm works for all (non-negative) weights, then $M$ is a matroid.\slug
+
+It is worth it to note that spanning trees and matroids have corresponding polytopes. For a graph $G=(V,E)$, the spanning tree polytope $ST(G)$ is the set of all vectors $x\in \R^{|E|}$ that satisfy
+\medskip
+\item {1.} $x_e \geq 0$,
+\smallskip
+\item {2.} $\displaystyle\sum_{e\in E} x_e = |V| - 1$, and
+\smallskip
+\item {3.} $\displaystyle\sum_{e\in E[S]} x_e \leq |S| - 1$, for every proper subset $S$ of $V$ with $|S| \geq 2$.
+\medskip
+For a matroid $M=(E,I)$ with rank function $r$, the matroid polytope $P_M$ is the set of all vectors $x\in \R^{|E|}$ satisfying $\sum_{e\in T} x_e \leq r(T)$ for all $T\subset E$, where $x_e\geq 0$.
+
+\section 2. PROBLEMS AND PROOFS
+
+\proclaim Problem 1. Describe how to find an initial basic feasible solution for an arbitrary LP in standard (equational) form.
+
+\solution Suppose the constraints are $Ax = b, x\geq 0$.
 Examine the matrix $A$.
@@
 We minimise the sum over all $y_i$.
-Optimising over this minimisation linear program gives us a basis for our original linear program.
-\end{proof}
+Optimising over this minimisation linear program gives us a basis for our original linear program.\slug
 
-\begin{problem}{2}
-    Write an integer program formulation for the matching/perfect matching problem.
-\end{problem}
-\begin{proof}[Solution]
-    The integer program corresponding to the matching problem in a graph $G=(V,E)$ with a cost vector $c$ corresponding to edges.
-    \begin{alignat*}{2}
-        & \textup{maximise} & \quad & c^Tx\\
-        & \textup{subject to} & & \sum_{v\in e} x_e \leq 1\ \text{for all}\ v\in V\\
-        &  & & x_e\in\{0,1\}.
-    \end{alignat*}
-    For a perfect matching, change the inequality for each vertex to equality.
-\end{proof}
+\proclaim Problem 2. Write an integer program formulation for the matching/perfect matching problem.
 
-\begin{problem}{3}
-    Prove that the linear program relaxation of the standard integer program for matching/perfect matching is integral for bipartite graphs.
-\end{problem}
-\begin{proof}
-    Let $LP(G,c)$ denote the linear program relaxation of the perfect matching integer program over a graph $G=(V,E)$ with weights $c$.
+\solution The integer program corresponding to the matching problem in a graph $G=(V,E)$ with a cost vector $c$ corresponding to edges.
+$$\matrix{
+    & \maximise & \quad & \displaystyle\sum_{e\in E} c_ex_e \hfill & \hcr
+    & \subject & & \displaystyle\sum_{v\in e} x_e \leq 1 \hfill & \hbox{for all}\ v\in V,\hcr
+    &  & & x_e\in\{0,1\}.\hfill & 
+}$$
+For a perfect matching, change the inequality for each vertex to equality.\slug
+
+\proclaim Problem 3. Prove that the linear program relaxation of the standard integer program for matching/perfect matching is integral for bipartite graphs.
+
+\proof Let $LP(G,c)$ denote the linear program relaxation of the perfect matching integer program over a graph $G=(V,E)$ with weights $c$.
 If $LP(G,c)$ is infeasible, then $G$ has no perfect matching.
@@
 If $LP(G,c)$ is feasible, there exists at least one optimal solution.
-Let $x^*$ an optimal solution that maximises the number of integer elements, i.e.\@ $|I(x^*)| \geq |I(x)|$ for all optimal solutions $x$.
-We want to show that $|I(x^*)|=|E|$.\\
-    \indent Suppose, towards a contradiction, that $|I(x^*)| < |E|$.
+Let $x^*$ an optimal solution that maximises the number of integer elements, i.e.\  $|I(x^*)| \geq |I(x)|$ for all optimal solutions $x$.
+We want to show that $|I(x^*)|=|E|$.
+
+Suppose, towards a contradiction, that $|I(x^*)| < |E|$.
 Then there exists an $i$ such that $0<x_i<1$.
@@
 Now for a small real number $\epsilon$ define a vector $x'$ by
-    \[
-        x'_e =
-        \begin{cases}
-            x^*_e - \epsilon & \text{for}\ e\in\{e_1,e_3,\ldots,e_{t-1}\}\\
-            x^*_e + \epsilon & \text{for}\ e\in\{e_2,e_4,\ldots,e_t\}\\
-            x^*_e & \text{otherwise}
-        \end{cases}
-    \]
+$$x'_e = \cases{
+    x^*_e - \epsilon & for $e\in\{e_1,e_3,\ldots,e_{t-1}\}$\hcr
+    x^*_e + \epsilon & for $e\in\{e_2,e_4,\ldots,e_t\}$\hcr
+    x^*_e & otherwise
+}$$
 It's easy to see that this new $x'$ still maintains the sum of 1 at each vertex.
@@
 Then $x'$ has one more element that is integral compared to $x^*$, contradicting maximality of $I(x^*)|$.
-So $I(x^*)|=|E|$.
-\end{proof}
+So $I(x^*)|=|E|$.\slug
 
-\begin{problem}{4}
-    The following is the integer program for vertex cover in a graph $G=(V,E)$:
-    \begin{alignat*}{2}
-        & \textup{minimise} & \quad & \sum_{v\in V}x_v\\
-        & \textup{subject to} & & x_u + x_v \geq 1\ \text{for all}\ \{u,v\}\in E\\
-        &  & & x_v\in\{0,1\}\ \text{for all}\ v\in V.
-    \end{alignat*}
+\proclaim Problem 4. The following is the integer program for vertex cover in a graph $G=(V,E)$:
+$$\matrix{
+    & \hbox{\rm minimise}\hfill & \quad & \displaystyle\sum_{v\in V}x_v\hcr
+    & \hbox{\rm subject to}\hfill & & x_u + x_v \geq 1\ \hfill & \hbox{\rm for all}\ \{u,v\}\in E,\hcr
+    &  & & x_v\in\{0,1\}\ \hfill& \hbox{\rm for all}\ v\in V.\hfill
+}$$
 Prove that changing $x_v\in \{0,1\}$ to $0\leq x_v\leq 1$ and then solving the resulting linear program results in a 2-approximation.
-\end{problem}
-\begin{proof}
-    Suppose that $x$ is the optimum feasible solution of the linear program relaxation.
-Define $S=\{v\in V: x_v^* \geq \frac{1}{2}\}$.
+
+\proof Suppose that $x$ is the optimum feasible solution of the linear program relaxation.
+Define $S=\{v\in V: x_v^* \geq {1\over 2}\}$.
 $S$ is a vertex cover.
@@
     which is what we wanted.
-\end{proof}
+\slug
 
-\begin{problem}{5}
-    Prove the max-flow min-cut theorem via duality.
-\end{problem}
-\begin{proof}
-    Given a directed graph $G=(V,E)$ with capacities $c$ for each edge, source vertex $s$ and sink vertex $t$.
+\proclaim Problem 5. Prove the max-flow min-cut theorem via duality.
+
+\proof Given a directed graph $G=(V,E)$ with capacities $c$ for each edge, source vertex $s$ and sink vertex $t$.
 Let $P$ be the set of all paths from $s$ to $t$.
 Then we can formulate the maximum flow problem as the following linear program:
-    \begin{alignat*}{2}
-        & \textup{maximise} & \quad & \sum_{p\in P} x_p\\
-        & \textup{subject to} & & \sum_{e\in p} x_p \leq c(e)\ \text{for all}\ e\in E\\
-        &  & & x_p \geq 0\ \text{for all}\ p\in P.
-    \end{alignat*}
+$$\matrix{
+    & \maximise & \quad & \displaystyle\sum_{p\in P} x_p\hcr
+    & \subject & & \displaystyle\sum_{e\in p} x_p \leq c(e)\ \hfill & \hbox{for all}\ e\in E,\hcr
+    &  & & x_p \geq 0\ \hfill & \hbox{for all}\ p\in P.\hfill
+}$$
 Now we construct the dual. We have one dual variable $y_e$ for every edge $e\in E$.
-    \begin{alignat*}{2}
-        & \textup{minimise} &\quad & \sum_{e\in E} c(e)y_e\\
-        & \textup{subject to} & & \sum_{e\in p} y_e\geq 1\ \text{for all}\ p\in P\\
-        &  & & y_e\geq 0\ \text{for all}\ e\in E.
-    \end{alignat*}
+$$\matrix{
+    & \minimise &\quad & \displaystyle\sum_{e\in E} c(e)y_e\hcr
+    & \subject & & \displaystyle\sum_{e\in p} y_e\geq 1\  \hfill & \hbox{for all}\ p\in P,\hcr
+    &  & & y_e\geq 0\  \hfill & \hbox{for all}\ e\in E.\hfill
+}$$
 This is the linear program relaxation of the minimum cut integer program.
-By the Strong Duality Theorem, the optimal value of the objective functions of these two linear programs must be equal, so the maximum flow equals the value of the minimum cut.
-\end{proof}
+By the Strong Duality Theorem, the optimal value of the objective functions of these two linear programs must be equal, so the maximum flow equals the value of the minimum cut.\slug
 
-\begin{problem}{6}
-    Prove K\H{o}nig's theorem via duality.
-\end{problem}
-\begin{proof}
-    Let $G=(V,E)$ be a bipartite graph with $m$ edges and $n$ vertices. Consider the integer program
-    \begin{alignat*}{2}
-        & \textup{maximise} & \quad & \sum_{j=1}^m x_j\\
-        & \textup{subject to} & & Ax \leq 1\\
-        &  & & x\geq 0\\
-        &  & & x\in \Z^m.
-    \end{alignat*}
+\proclaim Problem 6. Prove K\H onig's theorem via duality.
+
+\proof Let $G=(V,E)$ be a bipartite graph with $m$ edges and $n$ vertices. Consider the integer program
+$$\matrix{
+    & \maximise & \quad & \displaystyle\sum_{j=1}^m x_j\hcr
+    & \subject & & Ax \leq 1,\hcr
+    &  & & x\geq 0,\hcr
+    &  & & x\in \Z^m.\hfill
+}$$
 where $A$ is the incidence matrix of $G$.
-Because each row can sum to at most 1 and the incidence matrix of a graph contains only 0s and 1s, we discover that $x_j\in \{0,1\}$ for all $j$ and that the edges $e_j$ with $x^*_j = 1$ in an optimal solution $x^*$ form a maximum matching in $G$.\\
-    \indent Now consider the integer program
-    \begin{alignat*}{2}
-        & \textup{maximise} & \quad & \sum_{i=1}^n y_i\\
-        & \textup{subject to} & & A^Ty \geq 1\\
-        &  & & y\geq 0\\
-        &  & & y\in \Z^n.
-    \end{alignat*}
+Because each row can sum to at most 1 and the incidence matrix of a graph contains only 0s and 1s, we discover that $x_j\in \{0,1\}$ for all $j$ and that the edges $e_j$ with $x^*_j = 1$ in an optimal solution $x^*$ form a maximum matching in $G$.
+
+Now consider the integer program
+$$\matrix{
+    & \maximise & \quad & \displaystyle\sum_{i=1}^n y_i\hcr
+    & \subject & & A^Ty \geq 1,\hcr
+    &  & & y\geq 0,\hcr
+    &  & & y\in \Z^n.\hfill
+}$$
 where again, $A$ is the incidence matrix of $G$.
@@
 So in any optimal solution $y^*$ we have $y^*_i\in \{0,1\}$, since any larger value could be decreased to 1.
-Then the vertices $v_i$ with $y^*_i = 1$ form a minimum vertex cover of $G$.\\
-    \indent In both integer programs we may drop the integrality constraints, without affecting the optima, because $A$ is unimodular and by Problem 8, $b$ being integral implies that any optimum vector is as well.
+Then the vertices $v_i$ with $y^*_i = 1$ form a minimum vertex cover of $G$.
+
+In both integer programs we may drop the integrality constraints, without affecting the optima, because $A$ is unimodular and by Problem 8, $b$ being integral implies that any optimum vector is as well.
 The resulting linear programs are dual to one another, so their optimal values are equal.
-Hence the size of a maximum matching equals the size of a minimum vertex cover in $G$.
-\end{proof}
+Hence the size of a maximum matching equals the size of a minimum vertex cover in $G$.\slug
 
-\begin{problem}{7}
+\proclaim Problem 7.
 Let $G=(V,E)$ be a graph and with $n$ vertices $v_1,\ldots,v_n$ and $m$ edges $e_1,\ldots,e_m$.
@@
 Prove that if $G$ is bipartite, its incidence matrix is totally unimodular.
-\end{problem}
-\begin{proof}
-    Let $G=(X\cup Y, E)$ be a bipartite graph ($X\cap Y = \emptyset$) and let $A$ denote its incidence matrix.
+
+\proof Let $G=(X\cup Y, E)$ be a bipartite graph ($X\cap Y = \emptyset$) and let $A$ denote its incidence matrix.
 We want to show that every $k\times k$ submatrix $Q$ of $A$ has determinant 0, 1, or $-1$.
 We proceed by induction on $k$.
-The base case $k=1$ is easy, because all the entries of an incidence matrix are 0 or 1.\\
-    \indent Now consider $k>1$ and a $k\times k$ submatrix $Q$.
+The base case $k=1$ is easy, because all the entries of an incidence matrix are 0 or 1.
+
+Now consider $k>1$ and a $k\times k$ submatrix $Q$.
 Since the columns of $Q$ correspond to edges, each column of $Q$ has at most two nonzero entries (which are 1).
-If there is a column with only zero entries, we get $\det Q = 0$, and if there is a column with only one nonzero entry, we can expand the determinant on this column and get that up to sign, $\det Q$ equals the determinant of a $(k-1)\times(k-1)$ submatrix.\\
-    \indent The last case is if every column of $Q$ has exactly two 1s.
+If there is a column with only zero entries, we get $\det Q = 0$, and if there is a column with only one nonzero entry, we can expand the determinant on this column and get that up to sign, $\det Q$ equals the determinant of a $(k-1)\times(k-1)$ submatrix.
+
+The last case is if every column of $Q$ has exactly two 1s.
 Here we claim that $\det Q = 0$.
 To see this, observe that the sum of all rows of $Q$ corresponding to vertices in $X$ is the row vector $(1,\ldots,1)$, since for each column of $Q$, exactly one of its two 1s comes from a vertex in $X$.
-For the same reason, we get $(1,\ldots,1)$ by summing up the rows for vertices in $Y$, and it follows that the rows of $Q$ are linearly dependent.
-\end{proof}
+For the same reason, we get $(1,\ldots,1)$ by summing up the rows for vertices in $Y$, and it follows that the rows of $Q$ are linearly dependent.\slug
 
-\begin{problem}{8}
-    Prove that a polyhedron described by a totally unimodular matrix and integral right-hand side is integral.
-\end{problem}
-\begin{proof}
-    Suppose that $P = \{x\in R^n : Ax\leq b\}$ for a totally unimodular matrix $A$ and vector $b$ with integer elements.
+\proclaim Problem 8. Prove that a polyhedron described by a totally unimodular matrix and integral right-hand side is integral.
+
+\proof Suppose that $P = \{x\in R^n : Ax\leq b\}$ for a totally unimodular matrix $A$ and vector $b$ with integer elements.
 Let $x$ be a vertex of $P$.
@@
 But since $A$ is a non-singular unimodular matrix, $\det A_B^{-1} \in\{-1,1\}$, so every element of $A_B^{-1}$ is integral.
-So $x_B$ has integral elements and $x$ has integral non-zero elements, which is what we wanted.
-\end{proof}
+So $x_B$ has integral elements and $x$ has integral non-zero elements, which is what we wanted.\slug
 
-\begin{problem}{9}
-    Prove that an independence system is a matroid if and only if the weighted greedy algorithm produces a correct result for all non-negative objective functions.
-\end{problem}
-\begin{proof}
-    First we prove the "if" direction.
+\proclaim Problem 9. Prove that an independence system is a matroid if and only if the weighted greedy algorithm produces a correct result for all non-negative objective functions.
+
+\proof First we prove the ``if'' direction.
 Let $M=(E,I)$ be an independence system.
@@
 Let $c$ be a weight function given by
-    \[
-        c =
-        \begin{cases}
-            1+\epsilon & \text{if}\ e\in X\\
-            1 & \text{if}\ e\in Y\setminus X\\
-            0 & \text{otherwise}
-        \end{cases}
-    \]
+$$c = \cases{
+    1+\epsilon & if $e\in X$\hcr
+    1 & if $e\in Y\setminus X$\hcr
+    0 & otherwise
+}$$
 with $\epsilon$ to be determined.
@@
 But instead, the algorithm chooses all of $X$, then it completes $X$ to an independent set $X'$ of cardinality $|Y|$ by choosing 0-weight elements, for a total weight of $|X|(1+\epsilon)$.
-Now if we set $\epsilon<\frac{1}{|E|}$, then $c(X') < |X| + 1 \leq |Y| \leq c(Y)$.
-So the Greedy Algorithm chose the wrong set.\\
-    \indent Now we prove the "only if" direction.
-Let $M=(E,I)$ denote a matroid this time and we want to show that if $c$ is a positive objective function, the Greedy Algorithm finds maximum-weight independent sets of cardinality $k$ for every $k$ satisfying $1\leq k\leq r(E)$.\\
-    \indent Suppose, towards a contradiction, that the algorithm fails for a given $k$.
+Now if we set $\epsilon<{1\over |E|}$, then $c(X') < |X| + 1 \leq |Y| \leq c(Y)$.
+So the Greedy Algorithm chose the wrong set.
+
+Now we prove the ``only if'' direction.
+Let $M=(E,I)$ denote a matroid this time and we want to show that if $c$ is a positive objective function, the Greedy Algorithm finds maximum-weight independent sets of cardinality $k$ for every $k$ satisfying $1\leq k\leq r(E)$.
+
+Suppose, towards a contradiction, that the algorithm fails for a given $k$.
 Concretely, suppose the algorithm choose $S_k = \{s_1, s_2, \ldots, s_k\}$ with $c(s_1)\geq c(s_2)\geq \ldots \geq c(s_k)$ and there exists some set $T_k = \{t_1, t_2, \ldots, t_k\}$ with $c(t_1)\geq c(t_2)\geq \ldots \geq c(t_k)$ and $c(T_k) > c(S_k)$.
@@
 By observation, $c(t_j)\geq c(t_p)>c(s_p)$, so the Greedy Algorithm should have chosen $t_j$, not $s_p$, a contradiction.
-\end{proof}
+\slug
 
-\begin{problem}{10}
-    Prove that the rank function of a matroid is submodular.
-\end{problem}
-\begin{proof}
-    Let $M=(E,I)$ be a matroid and let $r$ denote its rank function.
+\proclaim Problem 10. Prove that the rank function of a matroid is submodular.
+
+\proof Let $M=(E,I)$ be a matroid and let $r$ denote its rank function.
 We want to show that for all $X,Y\subset E$, $r(X\cup Y) + r(X\cap Y) \leq r(X)+r(Y)$.
@@
 If we can show that $r(X\cup Y)\leq |J_X\cup J_Y|$, then submodularity follows, because $|J_X\cup J_Y|+|J_X\cap J_Y|=|J_X|+|J_Y|$.
-Extend $J_X$ to a maximal independent subset $K$ of $X\cup Y$.\\
-    \indent Suppose, towards a contradiction, that $|K|>|J_X\cup J_Y|$.
+Extend $J_X$ to a maximal independent subset $K$ of $X\cup Y$.
+
+Suppose, towards a contradiction, that $|K|>|J_X\cup J_Y|$.
 Because $J_X\setminus J$ is contained in both $K$ and $J_X\cup J_Y$, we have $|K\setminus (J_X\setminus J)|>|J_Y|$.
 Now, by the choice of $J_X$, we have that $K\setminus(J_X\setminus J)$ is an independent subset of $Y$.
-This contradicts the choice of $J_Y$.
-\end{proof}
+This contradicts the choice of $J_Y$.\slug
 

### 2019-06-16 misc/study-notes/ndmi011notes.tex
Fixed many typos in graph theory notes

@@
 So at each vertex, the number of incoming edges is at most 1 and the number of outgoing edges is at most 1 as well.
-This implies that there are $|f|$ independent pats from $s$ to $t$.
+This implies that there are $|f|$ independent paths from $s$ to $t$.
 But by the max-flow min-cut theorem, $|f| = d(T) \geq \kappa(G) = k$.
@@
 
-\proclaim Lemma. Let $(X,{\cal L})$ be a finite projective plane. Then for all $L, L'\in {\cal L}$, there exists an $z\in L$ such that $x\notin L\cup L'$
+\proclaim Lemma. Let $(X,{\cal L})$ be a finite projective plane. Then for all $L, L'\in {\cal L}$, there exists an $z\in L$ such that $z\notin L\cup L'$
 
@@
 
-\proclaim Theorem. Let $(X,{\cal L})$ be a finite projective plane of order $n$. Then the following all hold:
+\proclaim Theorem. Let $(X,{\cal L})$ be a finite projective plane of order $n$. Then the following statements all hold:
 \medskip
@@
 \smallskip
-\item {2.} Let $L_1 = \overline{xx_i}$ where $x_i\in L, i\in\{1,\ldots, n+1\}$. By (P1), $L_i\cap L_j = \{x\}$ for all $i\neq j$. As $|L_i\setminus\{x\}| = n$ we have
+\item {2.} Let $L_i = \overline{xx_i}$ where $x_i\in L, i\in\{1,\ldots, n+1\}$. By (P1), $L_i\cap L_j = \{x\}$ for all $i\neq j$. As $|L_i\setminus\{x\}| = n$ we have
     $$|X|\geq |\bigcup_{i=1}^{n+1} L_i| = (n+1)n + 1 = n^2 + n + 1,$$
@@
     Now we show that for all $p\in X$, there exists an $i\in \{1,\ldots,n+1\}$ such that $p\in L_i$.
-If $p=x$.
-this is clearly true, so assume $p\neq x$.
+If $p=x$, this is clearly true, so assume $p\neq x$.
 Then $\overline{px}\cap L = \{x_i\}$ for some $i$ by (P1).
@@
 
-The {\it incidence graph} of a finite projective plane $(X,{\cal L})$, is a bipartite graph $G = (V, E)$ with $V = X\cup {\cal L}$ and $E = \{xL : if x\in L\}$.
+The {\it incidence graph} of a finite projective plane $(X,{\cal L})$, is a bipartite graph $G = (V, E)$ with $V = X\cup {\cal L}$ and $E = \{xL : \hbox{for all pairs}\ x,L\ \hbox{such that}\ x\in L\}$.
 The concept of duality involves switching the roles of $X$ and ${\cal L}$ in this bipartite graph, interpreting $X$ as lines and ${\cal L}$ as points.
@@
 \medskip
-\item {(P1)} We need to find lines $L_1, L_2, L_3, L_4$ such that any for any $\lambda\in \Lambda$, with $\lambda = \{L\in {\cal L} : x\in L\}$ for some $x\in X$, the point $x$ is contained in at most two of $L_1, L_2, L_3, L_4$.
+\item {P1)} We need to find lines $L_1, L_2, L_3, L_4$ such that any for any $\lambda\in \Lambda$, with $\lambda = \{L\in {\cal L} : x\in L\}$ for some $x\in X$, the point $x$ is contained in at most two of $L_1, L_2, L_3, L_4$.
 Let $F = \{a,b,c,d\}$ be the set given by applying (P0) to $(X,{\cal L})$.
@@
 \smallskip
-\item {(P2)} Let $\lambda_1,\lambda_2\in \Lambda, \lambda_1\neq\lambda_2$ be given, i.e.\  $\lambda_1 = \{ L\in {\cal L} : x_1\in L\}$ and $\lambda_2 = \{L\in {\cal L} : x_2\in L\}$ for some points $x_1\neq x_2$.
+\item {P2)} Let $\lambda_1,\lambda_2\in \Lambda, \lambda_1\neq\lambda_2$ be given, i.e.\  $\lambda_1 = \{ L\in {\cal L} : x_1\in L\}$ and $\lambda_2 = \{L\in {\cal L} : x_2\in L\}$ for some points $x_1\neq x_2$.
 We need that $|\lambda_1\cap\lambda_2| = 1$, but this follows from (P2) for $(X,{\cal L})$, since $x_1$ and $x_2$ intersect at exactly one line.
 \smallskip
-\item {(P3)} Let $L_1, L_2\in {\cal L}$ with $L_1\neq L_2$.
+\item {P3)} Let $L_1, L_2\in {\cal L}$ with $L_1\neq L_2$.
 It suffices to show that there exists a unique $\lambda \in \Lambda$ such that $L_1\in \lambda$ and $L_2\in \lambda$.
@@
 
-So let $M = \{A_1, \ldots, A_t\}$ and for each $A_i\in M$, permute the entries such that the first row of the resulting Latin square $A'_i$ is $(1,2,3,\ldots,n)$.
+So let $M = \{A_1, \ldots, A_t\}$ and for each $A_i\in M$, permute the $n$ columns such that the first row of the resulting Latin square $A'_i$ is $(1,2,3,\ldots,n)$.
 By our earlier observation, $A'_1,\ldots,A'_k$ are pairwise orthogonal.
@@
 First note that it is impossible that both $|A| < n_1$ and $|B|<n_2$ since that would imply that
-$$n = 1 + |A| + |B| \leq 1 + (n_1-1) + (n_2=1) = n-1,$$
+$$n = 1 + |A| + |B| \leq 1 + (n_1-1) + (n_2-1) = n-1,$$
 a contradiction. So either $|A|\geq n_1$ or $|B|\geq n_2$.
@@
 Let $\Pr(Y)$ denote the probability that $\alpha(G) \geq k$ or $\omega(G) \geq k$. This is equal to the probablilty that $X_A$ holds for some $A\subseteq V$ with $|A|\geq k$. So
-$$\Pr[Y] \leq \sum_{A\subseteq V, |A|\geq k} 2^{1-{k\choose 2}} = {n\choose 2}2^{1-{k\choose 2}} \leq {n^k\over k!}2^{1-{k\choose 2}},$$
+$$\Pr[Y] \leq \sum_{A\subseteq V\atop |A|\geq k} 2^{1-{k\choose 2}} = {n\choose 2}2^{1-{k\choose 2}} \leq {n^k\over k!}2^{1-{k\choose 2}},$$
 and we can further derive the strict inequality
@@
 Then we label each $x_i$ of $S$ with a pair $(a_i, b_i)$ where $a_i$/$b_i$ is the length of the longest increasing/decreasing subsequence ending in $x_i$.
-(Note that each of the numbers $x_i, x_j$ have different labels.) Without loss of generality, assume $i<j$.
+
+We claim that each of the numbers $x_i, x_j$ have different labels.
+Let $i\neq j$ and without loss of generality, assume $i<j$.
 If $x_i\leq x_j$, then $a_i < a_j$, as any increasing subsequence ending in $x_i$ can be extended by $x_j$.
-Likewise, if $x_j\leq x_i$, then $a_j < a_i$, as any increasing subsequence ending in $x_j$ can be extended by $x_i$.
+Likewise, if $x_i\geq x_j$, then $b_j > b_i$, as any decreasing subsequence ending in $x_i$ can be extended by $x_j$.
 
@@
 Let $x$ denote a message $x\in \Sigma^k$ for some positive integer $k$.
-Then the {\it checksum} $c$ of $x$ is $\sum_{i=1}^k x_i \pmod{q}$.
-Then the message becomes $y\in {\bf F}_q^{k+1}$ where
+The {\it checksum} $c$ of $x$ is $\sum_{i=1}^k x_i \pmod{q}$ and we can encode te message as $y\in {\bf F}_q^{k+1}$ where
 $$ y_i = \cases{
@@
 
+\smallskip
+
 A {\it linear} code $C\subseteq {\bf F}_q^n$ is a linear subspace of ${\bf F}_q$, i.e.\  for all $x,y\in C$, $\alpha\in {\bf F}_q$, $\alpha x+y\in C$.
 There exist $k$ independent vectors $x_1,\ldots, x_k\in {\bf F}_q^n$ such that
-$$C = \{\sum_{i=1}^k \alpha_ix_i : \alpha_i,\ldots \alpha_n\in {\bf F}_q\}.$$
+$$C = \bigg\{\sum_{i=1}^k \alpha_ix_i : \alpha_i,\ldots \alpha_n\in {\bf F}_q\bigg\}.$$
 More succinctly, we can express $C$ using the generator matrix $G\in {\bf F}_q^{n\times k}$.
@@
 
-Alternatively, represent $C$ by the {\it null space}.
+Alternatively, we may represent $C$ as the {\it null space} of a matrix.
 Recall that for any $k$-dimensional linear subspace $C$, there is an $(n-k)$-dimensional linear subspace $C^\perp$ such that for all $x\in C, y\in C^\perp$, $x^Ty = 0$.
@@
 
-The {\it support} of a vector $x$ is the set $\{i : x_i \neq 0\}$ and the {\it Hamming weight} of $x$ is $\hbox{wt} x = |\{i : x_i \neq 0\}|$.
+The {\it support} of a vector $x$ is the set $\{i : x_i \neq 0\}$ and the {\it Hamming weight} of $x$ is $\hbox{wt}\,x = |\{i : x_i \neq 0\}|$.
 
 \proclaim Theorem. For a linear code $C$,
-$$ \Delta(C) = \min\{\hbox{wt} x : x\in C\ \hbox{and}\ x\neq 0\}.$$
-
-\proof For ease of notation, let $d$ denote $\min\{\hbox{wt} x : x\in C\ \hbox{and}\ x\neq 0\}$.
+$$ \Delta(C) = \min\{\hbox{wt}\,x : x\in C\ \hbox{and}\ x\neq 0\}.$$
+
+\proof For ease of notation, let $d$ denote $\min\{\hbox{wt}\,x : x\in C\ \hbox{and}\ x\neq 0\}$.
 First we show that $d\leq \Delta(C)$.
@@
 Since $C$ is linear, $x = y-z \in C$.
-Note that $\Delta(y,z) = |\{i : x_i \neq 0\}|$, so $d \leq \hbox{wt} x = \Delta(C)$.
-
-Now we show that $d\geq \Delta(C)$. Note that $0\in C$. Let $x\neq 0$ be such that $\hbox{wt} x = d$. Then $\Delta(C)\leq \Delta(0,x) = |\{i : x_i \neq 0\}| = \hbox{wt} x = d$.\slug
+Note that $\Delta(y,z) = |\{i : x_i \neq 0\}|$, so $d \leq \hbox{wt}\,x = \Delta(C)$.
+
+Now we show that $d\geq \Delta(C)$. Note that $0\in C$. Let $x\neq 0$ be such that $\hbox{wt}\,x = d$. Then $\Delta(C)\leq \Delta(0,x) = |\{i : x_i \neq 0\}| = \hbox{wt}\,x = d$.\slug
 
@@
 For simplicity, we will work with $q=2$.
-If $\hbox{wt} x = 1$, then $X$ has exactly one 1-entry, say it is at position $i$.
+If $\hbox{wt}\,x = 1$, then $X$ has exactly one 1-entry, say it is at position $i$.
 So $x^TH$ is the $i$-th row $H_i$ of $H$.
 We need that $H_i$ is non-zero for every $i$.
-If $\hbox{wt} x = 2$, then $x^T$ has exactly two non-zero elements, say they are at positions $i$ and $j$.
+If $\hbox{wt}\,x = 2$, then $x^T$ has exactly two non-zero elements, say they are at positions $i$ and $j$.
 Then $x^TH$ = $(H_i + H_j)^T$ hence we need $H_i\neq H_j$ for all $i\neq j$.
@@
 \thmparens{S}{Hamming bound} If an $(n,k,d)_q$-code exists, then
-$$q^k\cdot\hbox{vol}\left(\lfloor {d-1\over 2} \rfloor, n\right) \leq q^n.$$
+$$q^k\cdot\hbox{\rm vol}\left(\left\lfloor {d-1\over 2} \right\rfloor, n\right) \leq q^n.$$
 
 \proof Let $r = \lfloor (d-1)/2 \rfloor$ and consider $\bigcup_{x\in C} B(x,r)\subset \{1,\ldots,q\}^n$. By observation,
-$$|\bigcup_{x\in C} B(x,r)| = \sum_{x\in C} \hbox{vol}(r,n) = q^k\cdot \hbox{vol}(r,n) \leq q_n.\noskipslug$$
+$$\left|\bigcup_{x\in C} B(x,r)\right| = \sum_{x\in C} \hbox{vol}\,(r,n) = q^k\cdot \hbox{vol}\,(r,n) \leq q^n.\noskipslug$$
 
 Note that if $d = 3$ and $q = 2$ then the Hamming bound is $2^k(n+1)\leq 2^n$ and the Hamming code matches this bound with equality as $k=n-log_2(n+1)$.
-An $(n,k,d)_2$-code is called {\it perfect} if $q^k\cdot\hbox{vol}(\lfloor (d-1)/2 \rfloor, n) = q^n$.
+An $(n,k,d)_2$-code is called {\it perfect} if $q^k\cdot\hbox{vol}\,(\lfloor (d-1)/2 \rfloor, n) = q^n$.
 There are no codes with larger size $k$ than a $(n, n-d+1, d)_2$-code (even if we let $q>2$), as the following theorem shows.

### 2019-06-18 misc/study-notes/ndmi011notes.tex
Fixed errors in both notes

@@
 
-\maketitle{NDMI011 Combinatorics and Graph Theory}{Notes by}{Marcel K. Goh {\rm (Prague, Czech Rep.)}}{10 June 2019}
+\maketitle{NDMI011 Combinatorics and Graph Theory\footnote{$^*$}{\eightpt Course given by Prof.
+Andreas Feldmann at Charles University in Prague}}{Notes by}{Marcel K.
+Goh {\rm (Prague, Czech Rep.)}}{10 June 2019}
 
@@
     c(v_iv_{i+1}) - f(v_iv_{i+1}) & \hbox{if case (a) holds}\cr
-    f(v_{i+1}) & \hbox{if case (b) holds}
+    f(v_{i+1}v_i) & \hbox{if case (b) holds}
 }$$
@@
 
-So let $M = \{A_1, \ldots, A_t\}$ and for each $A_i\in M$, permute the $n$ columns such that the first row of the resulting Latin square $A'_i$ is $(1,2,3,\ldots,n)$.
+So let $M = \{A_1, \ldots, A_t\}$ and for each $A_i\in M$, permute the $n$ elements such that the first row of the resulting Latin square $A'_i$ is $(1,2,3,\ldots,n)$.
 By our earlier observation, $A'_1,\ldots,A'_k$ are pairwise orthogonal.
@@
 Then we have a set of points $(x_i, y_i),\ldots(x_n,y_n)$ where $n=(r-1)(s-1)$.
-Now order the $y$-coordinates {\it by the value of the $x$-coordinates}.
+Now order the $y$-coordinates {\sl by the value of the $x$-coordinates}.
 Any increasing subsequence can take at most one element from each column, so at most $r-1$, and any decreasing subsequence can take at most one element from each row, so at most $s-1$.
@@
 Let $x$ denote a message $x\in \Sigma^k$ for some positive integer $k$.
-The {\it checksum} $c$ of $x$ is $\sum_{i=1}^k x_i \pmod{q}$ and we can encode te message as $y\in {\bf F}_q^{k+1}$ where
+The {\it checksum} $c$ of $x$ is $\sum_{i=1}^k x_i \pmod{q}$ and we can encode the message as $y\in {\bf F}_q^{k+1}$ where
 $$ y_i = \cases{
@@
 $$C = \bigg\{\sum_{i=1}^k \alpha_ix_i : \alpha_i,\ldots \alpha_n\in {\bf F}_q\bigg\}.$$
-More succinctly, we can express $C$ using the generator matrix $G\in {\bf F}_q^{n\times k}$.
-The matrix $G = \matrix{ x_1 & \cdots & x_k }$ and $C = \{G_\alpha : \alpha \in {\bf F}_q^k\}$.
-(Multiplying any message by $G$ on the left encodes it.)
+More succinctly, we can express $C$ using the generator matrix $G = \left[\matrix{ x_1 & \cdots & x_k }\right]\in {\bf F}_q^{n\times k}$.
+Multiplying any message by $G$ on the left encodes it and $C = \{G_\alpha : \alpha \in {\bf F}_q^k\}$.
 
@@
 So $x$ and $y$ can only differ in the last $n-k+1$ entries.
-Since $d = \Delta(C) \leq \Delta(x,y) \leq n-k+1$, $k\leq n-d+1$.\slug
+Since $d = \Delta(C) \leq \Delta(x,y) \leq n-k+1$, we get $k\leq n-d+1$.\slug
 

### 2019-07-18 misc/study-notes/ndmi011notes.tex
Added letters to theorems and lemmas

@@
 
-\proclaim Theorem. In any graph $G=(V,E)$,
+\proclaim Theorem D. In any graph $G=(V,E)$,
 $$\displaystyle\sum_{v\in V} \deg v = 2\cdot|E|$$
@@
 
-\proclaim Theorem.
-The number of vertices with odd degree is even.
+\proclaim Theorem E. The number of vertices with odd degree is even.
 
@@
 
-\proclaim Theorem.
+\proclaim Theorem T.
 A tree with $n$ vertices has $n-1$ edges.
@@
 
-\thmparens{}{Generalised binomial theorem} For any $r\in \R$,
+\thmparens{G}{Generalised binomial theorem} For any $r\in \R$,
 $$\sum_{i\geq 0} {r\choose i} x^i = (1 + x)^r.$$
@@
 
-\proclaim Theorem.
+\proclaim Theorem S.
 Let $(a_i)_{i=0}^\infty$ be a sequence of reals such that there exists some $k>0$ for which $|a_i|\leq k^i$ for every $i\geq 1$.
@@
 
-\proclaim Lemma. For any s-t-flow $f$ and s-t-cut $\delta^+(S)$,
+\proclaim Lemma D. For any s-t-flow $f$ and s-t-cut $\delta^+(S)$,
 $$|f| = \sum_{e\in \delta^+(S)} f(e) - \sum_{e\in \delta^-(S)} f(e),$$
@@
 
-\proclaim Theorem. If $f$ is a max-flow and $\delta^+(S)$ is a min-cut, then $|f| = c\big(\delta^+(S)\big)$.
+\proclaim Theorem C. If $f$ is a max-flow and $\delta^+(S)$ is a min-cut, then $|f| = c\big(\delta^+(S)\big)$.
 
@@
 
-\proclaim Theorem. If for all $e\in E$, $c(e)\in \N_0$, then the capacity $n$ of the min-cut is integral and Algorithm M terminates in at most $n$ iterations.
+\proclaim Theorem I. If for all $e\in E$, $c(e)\in \N_0$, then the capacity $n$ of the min-cut is integral and Algorithm M terminates in at most $n$ iterations.
 
@@
 
-\proclaim Theorem.
+\proclaim Theorem J.
 Let $f$ be a max-flow in a network with vertex capacities.
@@
 
-\thmparens{}{Hall} A bipartite graph $G=(A\cup B, E)$ (with $A$ and $B$ disjoint) has a matching $M\subseteq E$ of size $|M| = |A|$ if and only if $|N(S)|\geq |S|$ for all $S\subseteq A$.
+\thmparens{H}{Hall} A bipartite graph $G=(A\cup B, E)$ (with $A$ and $B$ disjoint) has a matching $M\subseteq E$ of size $|M| = |A|$ if and only if $|N(S)|\geq |S|$ for all $S\subseteq A$.
 
@@
 
-\thmparens{}{K\H onig's theorem} Let $G$ be a bipartite graph with maximum matching $M$ and minimum vertex cover $C$. Then $|M| = |C|$.\slug
+\thmparens{K}{K\H onig's theorem} Let $G$ be a bipartite graph with maximum matching $M$ and minimum vertex cover $C$. Then $|M| = |C|$.\slug
 
@@
 
-\proclaim Lemma. Let $G=(V,E)$ be a graph. Then for all edges $uv\in E$, $\kappa(G\setminus\{uv\})\geq \kappa(G) - 1$.
+\proclaim Lemma K. Let $G=(V,E)$ be a graph. Then for all edges $uv\in E$, $\kappa(G\setminus\{uv\})\geq \kappa(G) - 1$.
 
@@
 
-\proclaim Lemma. Let $G=(V,E)$ be a graph with $|V|=n$. Then the following both hold:
+\proclaim Lemma R. Let $G=(V,E)$ be a graph with $|V|=n$. Then the following both hold:
 \medskip
@@
 
-\proclaim Lemma. Let $G=(V,E)$ be a graph and let $\delta(G)$ denote the minimum degree over all vertices in $G$. Then
+\proclaim Lemma C. Let $G=(V,E)$ be a graph and let $\delta(G)$ denote the minimum degree over all vertices in $G$. Then
 $$ \kappa(G) \leq \lambda(G) \leq \delta(G).$$
@@
 
-\proclaim Theorem. A graph is 2-connected if and only if it has an ear decomposition.
+\proclaim Theorem E. A graph is 2-connected if and only if it has an ear decomposition.
 
@@
 
-\proclaim Theorem. Let $e$ be an edge in $K_n$. Then
+\proclaim Theorem L. Let $e$ be an edge in $K_n$. Then
 $$ T(K_n-e) = (n-2)n^{n-3}.$$
@@
 
-\proclaim Theorem. The maximum number of edges of any graph $G$ not containing $C_3$ as a subgraph is at most $\lfloor n^2/4 \rfloor$.
+\proclaim Theorem T. The maximum number of edges of any graph $G$ not containing $C_3$ as a subgraph is at most $\lfloor n^2/4 \rfloor$.
 
@@
 
-\proclaim Theorem. The maximum number of edges of any graph $G$ not containing $C_4$ as a subgraph is at most $(n^{3/2} + n) / 2$.
+\proclaim Theorem F. The maximum number of edges of any graph $G$ not containing $C_4$ as a subgraph is at most $(n^{3/2} + n) / 2$.
 
@@
 
-\proclaim Lemma. Let $(X,{\cal L})$ be a finite projective plane. Then for all $L, L'\in {\cal L}$, there exists an $z\in L$ such that $z\notin L\cup L'$
+\proclaim Lemma Z. Let $(X,{\cal L})$ be a finite projective plane. Then for all $L, L'\in {\cal L}$, there exists an $z\in L$ such that $z\notin L\cup L'$
 
@@
 
-\proclaim Lemma. Let $(X,{\cal L})$ be a finite projective plane. Then for all $L,L'\in {\cal L}$, $|L| = |L'|$.
+\proclaim Lemma C. Let $(X,{\cal L})$ be a finite projective plane. Then for all $L,L'\in {\cal L}$, $|L| = |L'|$.
 
@@
 
-\proclaim Theorem. Let $(X,{\cal L})$ be a finite projective plane of order $n$. Then the following statements all hold:
+\proclaim Theorem F. Let $(X,{\cal L})$ be a finite projective plane of order $n$. Then the following statements all hold:
 \medskip
@@
 
-\proclaim Lemma. The dual $({\cal L}, \Lambda)$ of a finite projective plane $(X,{\cal L})$ is also a finite projective plane.
+\proclaim Lemma D. The dual $({\cal L}, \Lambda)$ of a finite projective plane $(X,{\cal L})$ is also a finite projective plane.
 
@@
 
-\proclaim Theorem. A finite projective plane exists of order $n$ if $n$ is a prime power.
+\proclaim Theorem P. A finite projective plane exists of order $n$ if $n$ is a prime power.
 
@@
 
-\proclaim Theorem. Let $M$ be a set of pairwise orthogonal Latin squares of order $n$. Then $|M|\leq n-1$.
+\proclaim Theorem L. Let $M$ be a set of pairwise orthogonal Latin squares of order $n$. Then $|M|\leq n-1$.
 
@@
 
-\proclaim Theorem.
+\proclaim Theorem O.
     For $n\geq 2$, a finite projective plane of order $n$ exists if and only if there exists a set of $n-1$ pairwise orthogonal Latin squares of order $n$.
@@
 
-\proclaim Theorem. Let $G=(V,E)$ be a graph. If $|V|\geq {k-l-2\choose k-1} = {k+l-2\choose l-1}$, then $\alpha(G)\geq l$ or $\omega(G)\geq k$.
+\proclaim Theorem R. Let $G=(V,E)$ be a graph. If $|V|\geq {k-l-2\choose k-1} = {k+l-2\choose l-1}$, then $\alpha(G)\geq l$ or $\omega(G)\geq k$.
 
@@
 
-\proclaim Theorem. For $k\geq 3$, $r(k,k)>2^{k/2}$.
+\proclaim Theorem L. For $k\geq 3$, $r(k,k)>2^{k/2}$.
 
@@
 
-\proclaim Theorem. For a linear code $C$,
+\proclaim Theorem L. For a linear code $C$,
 $$ \Delta(C) = \min\{\hbox{wt}\,x : x\in C\ \hbox{and}\ x\neq 0\}.$$
@@
 
-\thmparens{S}{Hamming bound} If an $(n,k,d)_q$-code exists, then
+\thmparens{H}{Hamming bound} If an $(n,k,d)_q$-code exists, then
 $$q^k\cdot\hbox{\rm vol}\left(\left\lfloor {d-1\over 2} \right\rfloor, n\right) \leq q^n.$$
@@
 
-\thmparens{B}{Singleton bound} If $C$ is an $(n,k,d)_q$-code, then $k\leq n-d+1$.
+\thmparens{S}{Singleton bound} If $C$ is an $(n,k,d)_q$-code, then $k\leq n-d+1$.
 

### 2019-07-18 misc/study-notes/nopt048notes.tex
Fixed errors in both notes
Added letters to theorems and lemmas

@@
 
-\maketitle{NOPT048 Linear Programming and Combinatorial Optimization}{Notes by}{Marcel K. Goh {\rm (Prague, Czech Rep.)}}{9 June 2019}
+\maketitle{NOPT048 Linear Programming and Combinatorial Optimization\footnote{$^*$}{\eightpt Course given by Prof.
+Hans Raj Tiwary at Charles University in Prague}}{Notes by}{Marcel K.
+Goh {\rm (Prague, Czech Rep.)}}{9 June 2019}
 
@@
 A polyhedron is {\it pointed} if it does not contain a line.
-\proclaim Lemma. Every non-empty pointed polyhedron $P$ has at least one vertex.
+\proclaim Lemma P. Every non-empty pointed polyhedron $P$ has at least one vertex.
 
-\proclaim Theorem. Consider the linear program
+\proclaim Theorem E. Consider the linear program
 $$\matrix{
@@
 
-\proclaim Theorem.
+\proclaim Theorem G.
 Let $M=(E,I)$ be a matroid with rank function $r$.
@@
 
-\thmparens{G}{Greedy equivalence} Let $M=(E,I)$ be an independence system. Then if the weighted greedy algorithm works for all (non-negative) weights, then $M$ is a matroid.\slug
+\thmparens{E}{Greedy equivalence} Let $M=(E,I)$ be an independence system. Then if the weighted greedy algorithm works for all (non-negative) weights, then $M$ is a matroid.\slug
 

### 2019-07-21 misc/arithmeticae/disquisitiones-arithmeticae.tex
Added Arithmeticae

\input fontmac
\input mathmac

\centerline {\bigbold Disquitiones Arithmeticae}
\medskip
\centerline {\eightpt auctore}
\medskip
\centerline {\tensc D. Carolo Friderico Gauss}

\bigskip

\def\ssection #1\par {
    \section #1\par
    \vskip -\medskipamount
}

\def\numsection #1. {
    \medskip
    \centerline {#1.}
    \smallskip
}

% INTRODUCTION
\ssection I. DE NUMERORUM CONGRUENTIA IN GENERE

\ssection Numeri congrui, moduli, residua et nonresidua

\numsection 1. Si numerus $a$ numerorum $b,c$ differentiam metitur, $b$ et $c$ {\it secundum $a$ congrui} dicuntur, sin minus, {\it incongrui}: ipsum {\it a modulum} appellamus. Uterque numerorum $b,c$ priori in casu alterius {\it residuum}, in posteriori vero {\it nonresiduum} vocatur.

Hae notiones de omnibus numeris integris tam positivis quam negativis valent\footnote{$^1$}{Modulus manifesto semper {\it absolute} i. e. sine omni signo est sumendus.}, neque vero ad fractos sunt extendendae. E. g.\ $-9$ et $16$ secundum modulum $5$ sunt congrui; $-7$ ipsius $15$ secundum modulum $11$ residuum, secondum modulum $3$ vero nonresiduum. Ceterum quoniam cifram numerus quisque metitur, omnis numerus tamquam sibi ipsi congruus secundum modulum quemcunque est spectandus.

\numsection 2. Omnia numeri dati $a$ residua secundum modulum $m$ sub formula $a+km$ comprehenduntur, designante $k$ numerum integrum indeterminatum. Propositionum quas post trademus faciliores nullo negotio hinc demonstrari possunt: sed istarum quidem veritatem aeque facile quivis intuendo poterit perspicere.

Numerorum congruentiam hoc signo, $\equiv$, in posterum denotabimus, modulum ubi opus erit in clausulis adiungentes: $-16\equiv 9 \pmod{5}$, $-7\equiv 15 \pmod{11}$\footnote{$^2$}{Hoc signum propter magnam analogiam quae inter aequalitatem atque congruentiam invenitur adoptavimus. Ob eandem caussam ill.\ Le Gendre in comment.\ infra saepius laudanda ipsum aequalitatis signum pro congruentia retinuit, quod nos ne ambiguitas oriatur imitari dubitavimus.}.

\numsection 3.

\proclaim Theorema. Propositis $m$ numeris integris successivis
$$ a, a+1, a+2, \ldots, a+m-1$$
alioque $A$, illorum aliquis huic secundum modulum $m$ congruus erit, et quidem unicus tantum.

Si enim ${a-A\over m}$ integer, erit $a\equiv A$, sin fractus, sit integer proxime maior (aut quando est negativus, proxime {\it minor}, si ad signum non respiciatur) aequalis $k$, cadetque $A+km$ inter $a$ et $a+m$, quare erit numerus quaesitus. Et manifestum est omnes quotientes ${a-A \over m}$, ${a+1-A\over m}$, ${a+2-A\over m}$ etc.\ inter $k-1$ et $k+1$ sitos esse; quare plures quam unus integri esse nequeunt.

\ssection Residua minima

\numsection 4. Quisque igitur numerus residuum habebit tum in hac serie, $0,1,2,\ldots,m-1$, tum in hac, $0,-1,-2,\ldots,\allowbreak -(m-1)$, quae {\it residua minima} dicemus, patetque, nisi $0$ fuerit residuum, bina semper dari, {\it positivum} alterum, alterum {\it negativum}. Quae si magnitudine sunt inaequalia, alterum erit minus quam ${m\over 2}$, sin secus utrumque ${m\over 2}$ aequalis, signi respectu non habito. Unde patet, quemvis numerum residuum habere moduli semissem non superans quod {\it absolute minimum} vocabitur.

E. g.\ $-13$ secundum modulum $5$, habet residuum minimum positivum $2$, quod simul est absolute minimum, $-3$ vero residuum minimum negativum; $5$ secundum modulum $7$ sui ipsius est residuum minimum positivum, $-2$ negativum, simulque absolute minimum.

\ssection Propositiones elementares de congruentiis

\numsection 5. His notionibus stabilitis eas numerorum congruorum propietates quae prima fronte se offerunt colligamus.

{\sl Qui numeri secundum modulum compositum sunt congrui, etiam secundum quemvis eius divisorem congrui. Si plures numeri eidem numero secundum eundem modulum sunt congrui, inter se erunt congrui (secundum eundem modulum).}

Haec modulorum indentitas etiam in sequentibus est subintelligenda.

{\sl Numeri congrui residua minima habent eadem, incongrui diversa.}

\numsection 6. {\sl Si habentur quotcunque numeri $A,B,C$ etc.\ totidemque alii $a,b,c$ etc.\ illus secundum modulum quemque congrui
$$A\equiv a, B\equiv b,\hbox{etc., erit}\ A+B+C+\hbox{etc.}\equiv a+b+c+\hbox{etc.}$$}
Si $A\equiv a, B\equiv b$, erit $A-B \equiv a-b$.

\numsection 7. {\sl Si $A\equiv a$, erit quoque $kA\equiv ka$.}

Si $k$ numerus positivus, hoc est tantummodo casus particularis propos.\ art.\ praec.\ ponendo ibi $A=B=C=\hbox{etc.}$, $a=b=c\ \hbox{etc.}$ Si $k$ negativus, erit $-k$ positivus, adeoque $-kA\equiv -ka$, unde $kA \equiv ka$.

{\sl Si $A\equiv a$, $B\equiv b$, erit $AB\equiv ab$. Namque $AB\equiv Ab\equiv ba$.}

\numsection 8. {\sl Si habentur quotcunque numeri $A,B,C,\hbox{etc.}$ totidemque alii $a,b,c,\hbox{etc.}$ his congrui, $A\equiv a$, $B\equiv b,\hbox{etc.}$, producta ex utrisque erunt congrua, $ABC\ \hbox{etc.}\equiv abc\ \hbox{etc.}$}

Ex artic.\ praec.\ $AB\equiv ab$, et ob eandem rationem $ABC\equiv abc$; eodemque modo quotcunque alii factores accedere possunt.

Si omnes numeri $A,B,C,\ldots$ aequales assumuntur, nec non respondentes $a,b,c,\ldots$ habetur hoc theorema: {\sl Si $A\equiv a$ et $k$ integer positivus, erit $A^k\equiv a^k$.}

\end

### 2019-07-21 misc/arithmeticae/disquisitiones-arithmeticae.tex
Added Erdos-Szekeres

@@
 \centerline {\tensc D. Carolo Friderico Gauss}
+\medskip
+\centerline {\eightpt transcriptae a}
+\medskip
+\centerline {\tensc Marcello K. Goh}
 
@@
 
-Hae notiones de omnibus numeris integris tam positivis quam negativis valent\footnote{$^1$}{Modulus manifesto semper {\it absolute} i.
+Hae notiones de omnibus numeris integris tam positivis quam negativis valent\footnote{$^{*}$}{Modulus manifesto semper {\it absolute} i.
 e.
@@
 
-Numerorum congruentiam hoc signo, $\equiv$, in posterum denotabimus, modulum ubi opus erit in clausulis adiungentes: $-16\equiv 9 \pmod{5}$, $-7\equiv 15 \pmod{11}$\footnote{$^2$}{Hoc signum propter magnam analogiam quae inter aequalitatem atque congruentiam invenitur adoptavimus.
+Numerorum congruentiam hoc signo, $\equiv$, in posterum denotabimus, modulum ubi opus erit in clausulis adiungentes: $-16\equiv 9 \pmod{5}$, $-7\equiv 15 \pmod{11}$\footnote{$^{**}$}{Hoc signum propter magnam analogiam quae inter aequalitatem atque congruentiam invenitur adoptavimus.
 Ob eandem caussam ill.\ Le Gendre in comment.\ infra saepius laudanda ipsum aequalitatis signum pro congruentia retinuit, quod nos ne ambiguitas oriatur imitari dubitavimus.}.
@@
 Quisque igitur numerus residuum habebit tum in hac serie, $0,1,2,\ldots,m-1$, tum in hac, $0,-1,-2,\ldots,\allowbreak -(m-1)$, quae {\it residua minima} dicemus, patetque, nisi $0$ fuerit residuum, bina semper dari, {\it positivum} alterum, alterum {\it negativum}.
-Quae si magnitudine sunt inaequalia, alterum erit minus quam ${m\over 2}$, sin secus utrumque ${m\over 2}$ aequalis, signi respectu non habito.
+Quae si magnitudine sunt inaequalia, alterum erit $<{m\over 2}$, sin secus utrumque ${m\over 2}$ aequalis, signi respectu non habito.
 Unde patet, quemvis numerum residuum habere moduli semissem non superans quod {\it absolute minimum} vocabitur.

### 2019-07-21 misc/study-notes/ndmi011notes.tex
Faulty italic
Added Erdos-Szekeres

@@
 The graph $G'=(V',E')$ is a {\it subgraph} of $G$ if $V'\subseteq V$ and $E'\subseteq E$.
-A {it cycle} is a graph on $v_1, \ldots, v_n$ with edges $\{v_1, v_2\}, \ldots, \{v_{n-1}, v_n\}$.
+A {\it cycle} is a graph on $v_1, \ldots, v_n$ with edges $\{v_1, v_2\}, \ldots, \{v_{n-1}, v_n\}$.
 A {\it tree} is a connected graph that contains no cycle (i.e\  does not have a cycle as a subgraph).
@@
 
-\thmparens{E}{Erd\H os-Szekeres} For any finite sequence $S$ of at least $(r+1)(s+1) + 1$ numbers, there is an increasing subsequence of length $r$ or a decreasing subsequence of length $s$.
+\thmparens{E}{Erd\H os-Szekeres} For any finite sequence $S$ of at least $(r-1)(s-1) + 1$ numbers, there is an increasing subsequence of length $r$ or a decreasing subsequence of length $s$.
 

### 2019-07-25 misc/proofs/coprime-probability.tex
Added coprime prob

% Probablity that random u and v are relatively prime

\input fontmac
\input mathmac

\font\eightsl=cmsl8

\maketitle{On the probability that two integers are relatively prime\footnote{$^*$}{\eightpt This document functions as a complete solution to Exercise 4.5.2--10 of {\eightsl The Art of Computer Programming} by D.\ E.\ Knuth.}}{by}{Marcel K. Goh (\rm Saigon, Vietnam)}{26 July 2019}

\medskip

By the fundamental theorem of arithmetic, any integer $n>1$ may be expressed as a product of primes
$$n = p_1^{e_1}p_2^{e_2}\cdots p_k^{e_k},$$
and this representation is unique up to the order of the factors. If none of the primes are repeated, i.e.\ each exponent $e_i$ is at most $1$, then we say that $n$ is {\it square-free}. The {\it M\"obius function} $\mu : {\bf N} \rightarrow \{0, \pm 1\}$ is given by
$$\mu(n)=
\cases{
    1 & if $n=1$ or $n$ is square-free and has an even number of prime factors;\cr
    -1 & if $n$ is square-free and has an odd number of prime factors; and\cr
    0 & otherwise.
}$$

If two integers $u,v$ have no prime factors in common (i.e.\ $\gcd(u,v) = 1$), we say they are {\it relatively prime} and write $u\perp v$. For any positive integer $n$, let $q_n$ denote the number of ordered pairs of integers $(u,v)$ such that $1\leq u,v\leq n$and $u\perp v$. Then the probability that any pair of integers $u,v$ lying in the range $[1\ ..\ n]$ are relatively prime is $q_n/n^2$. Our goal is to find out what happens when $u$ and $v$ may be {\it any} positive integers. In other words, we will investigate the behaviour of $q_n/n^2$ as $n\rightarrow \infty$. We begin with the following lemma:

\proclaim Lemma A. For any positive integer $n$, we have
$$q_n = \sum_{k\geq 1} \mu(k)\left\lfloor {n\over k}\right\rfloor^2,$$
where the floor function $\lfloor x\rfloor$ returns the largest integer that is no larger than $x$.

\proof Let $X$ denote the set of all ordered pairs $(u,v)$ in the range $1\geq u,v\geq n$. Note that $X$ contains $n^2$ elements. For each prime number $p_i$, let $S_{p_i}\subset X$ denote the set of pairs $(u,v)$ such that $p_i$ divides both $u$ and $v$. Since $u\perp v$ if and only if no prime divides both $u$ and $v$, the number of pairs $(u,v)$ that lie in {\it none} of the sets $S_{p_i}$ is exactly $q_n$. Thus
$$ q_n = |X \setminus \bigcup_{p_i} S_{p_i}|.$$
By the inclusion-exclusion principle, we can expand this to get
$$\eqalign{
    q_n &= |X| - \sum_{p_i} |S_{p_i}| + \sum_{p_i < p_j} |S_{p_i} \cap S_{p_j}| - \sum_{p_i < p_j < p_k} |S_{p_i}\cap S_{p_j}\cap S_{p_k}| + \cdots \cr
    q_n &= n^2 - \sum_{p_i} \left\lfloor {n\over p_i} \right\rfloor^2 + \sum_{p_i < p_j} \left\lfloor {n\over p_ip_j} \right\rfloor^2 - \sum_{p_i < p_j < p_k} \left\lfloor {n\over p_ip_jp_k} \right\rfloor^2 + \cdots
}$$
where the sums are taken over all prime numbers.

First, notice that although each sum is taken over all prime numbers, each sum is actually finite due to the floor function. From the uniqueness of prime decompositions, each positive integer $k\geq 1$ appears in the denominator of at most one term in the overall alternating sum. If $k$ is not square-free, it does not appear at all, since no term contains a repeated prime in its denominator. If the prime decomposition of $k$ consists of an odd number of distinct primes, then we see from the sum above that it contributes $-\lfloor n/k\rfloor^2$ to the sum. Finally, if $k = 1$ or $k$ is the product of an even number of distinct primes, then it contributes $+\lfloor n/k\rfloor^2$ to the sum. This corresponds to the definition of the M\"obius function. So we have $q_n = \sum_{k\geq 1} \mu(k) \lfloor n/k\rfloor ^2$, as we had hoped.\slug

In the proof of the next result, we will employ notation that is useful when describing the asymptotic behaviour of a function. Let $f$ and $g$ be functions such that $g$ is strictly positive for large enough values of $n$. We write $f(n) = O\bigl(g(n)\bigr)$ if and only if there exist positive real numbers $c$ and $n_0$ such that $|f(n)| \leq c\cdot g(n)$ for $n\geq n_0$; and we write $f(n) = o\bigl(g(n)\bigr)$ if and only if for all choices of positive real $\epsilon$, there exists a real $n_0>0$ such that $|f(n)| \leq \epsilon\cdot g(n)$ for any $n\geq n_0$. Intuitively, $f(n)$ being $O\bigl(g(n)\bigr)$ means that as $n$ gets very large and omitting constant factors, $f(n)$ grows no faster than $g(n)$. The notation $f(n) = o\bigl(g(n)\bigr)$ makes a stronger statement, implying that as $n$ gets very large, $f(n)$ becomes insignificant relative to $g(n)$.

\proclaim Lemma B.
$$\lim_{n\rightarrow\infty} {q_n\over n^2} = \sum_{k\geq 1} {\mu(k)\over k^2}.$$

\proof To prove this limit, we will investigate the difference
$$q_n - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 = \sum_{k\geq 1}\mu(k)\left\lfloor {n\over k}\right\rfloor^2 - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2.$$
Because $\lfloor n/k \rfloor = 0$ for $k>n$, the first summation on the right-hand side need only range up to $k=n$, and we can reorganise the summations as follows:
$$\eqalignno{
    \sum_{k\geq 1}\mu(k)\left\lfloor {n\over k}\right\rfloor^2 - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 &= \sum_{k=1}^n \mu(k)\left\lfloor {n\over k}\right\rfloor^2 - \sum_{k=1}^n \mu(k)\left({n\over k}\right)^2 - \sum_{k>n}\mu(k)\left({n\over k}\right)^2\cr
    q_n - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 &= \sum_{k=1}^n \mu(k) \left(\left\lfloor {n\over k} \right\rfloor ^2 - \left({n \over k}\right)^2\right) - \sum_{k>n}\mu(k)\left({n\over k}\right)^2 & ({\oldstyle 1})
}$$
We will want to put asymptotic bounds on these rearranged sums. First, note that for all positive real $x$,
$$x^2 - \lfloor x\rfloor^2 = O(x).\oldno 2$$
For $x^2 - \lfloor x\rfloor ^2$ can be rewritten $(x + \lfloor x\rfloor)(x-\lfloor x\rfloor)$, which is at most $2x\cdot x = 3x$. Next, we use standard formulas to analyse the asymptotic behaviour of the series $\sum_{k>n}(n/k)^2$:
$$\eqalign{
     \sum_{k>n} \left({n\over k}\right)^2 &= \sum_{k\geq 1} \left({n\over k}\right)^2 - \sum_{k=1}^n \left({n\over k}\right)^2 \cr
    &= {n^2\pi^2\over 6} - n^2\left({\pi^2\over 6} -  {1\over n} + O( n^{-2} )\right) \cr
    &= n - O(1) \cr
    &= O(n)
}\oldno 3$$
Since $\mu(k)\leq 1$, we can discard this factor when applying Eqs.\ ({\oldstyle 2}) and ({\oldstyle 3}) to get the asymptotic behaviour of Eq.\ ({\oldstyle 1}); this yields
$$\eqalign{
    q_n - \sum_{k\geq 1}\mu(k)\left({n\over k}\right)^2 &= \sum_{k=1}^n O(n/k) - O(n)\cr
    &= O(nH_n) - O(n),
}$$
where $H_n = \sum_{k=1}^n 1/k$ denotes the $n$-th harmonic number. Because both $O(nH_n)$ and $O(n)$ are $o(n^2)$, we can replace these terms by $o(n^2)$ and divide by $n^2$ to arrive at
$${q_n\over n^2} - \sum_{k\geq 1}{\mu(k)\over k^2} = o(1).$$
This means that for any real $\epsilon > 0$, we can find a number $n_0$ such that for all $n \geq n_0$ the difference $|q_n - \sum_{k\geq 1} \mu(k)/k^2| \leq \epsilon$; the sequence $q_n/n^2$ converges exactly as stated by the lemma.\slug

We now know that $q_n/n^2$ tends to the mysterious summation $\sum_{k\geq 1}\mu(k)/k^2$ as $n$ approaches infinity. Before we may compute the value of this summation, we need the following theorem, whose proof is simple enough that it may be included as well.

\proclaim Theorem M. The sum $\sum_{d\backslash n} \mu(d)$, taken over all integers $d$ that divide $n$, equals 1 when $n=1$ and equals 0 for all integers $n>1$.

\proof The case $n=1$ is obviously true, so let $n>1$. By the fundamental theorem of arithmetic, $n = p_1^{e_1}p_2^{e_2}\cdots p_k^{e_k}$. Only square-free divisors contribute to the sum $\sum_{d\backslash n}\mu(d)$, so we have
$$\sum_{d\backslash n}\mu(d) = \mu(1) + \sum_{1\leq i\leq k} \mu(p_i) + \sum_{1\leq i<j\leq k} \mu(p_ip_j) + \cdots + \mu(p_1\cdots p_k).$$
Since the magnitude of $\mu(d)$ is 1 (whenever $d$ is square-free) and the sign of $\mu(d)$ depends on the number of primes that make up $d$, this can be re-expressed as the alternating sum of binomial coefficients
$$\sum_{d\backslash n}\mu(d) = {k \choose 0} - {k\choose 1} + {k\choose 2} - \cdots + (-1)^k{k\choose k}.$$
which equals 0. (We will not prove this claim, but one may be convinced of this fact by trying it out on a couple of rows of Pascal's Triangle.)\slug

We are now ready for the final lemma.

\proclaim Lemma C.
$$\biggl(\sum_{k\geq 1} {\mu(k) \over k^2}\biggr) \biggl( \sum_{m \geq 1} {1\over m^2}\biggr) = 1$$.

\proof Both series in the product are absolutely convergent, so we may make use of the identity
$$\biggl(\sum_{k\geq 1} {a_k\over k^z}\biggr) \biggl(\sum_{m\geq 1} {b_m\over m^z}\biggr) = \sum_{n\geq 1}\biggl( \sum_{d\backslash n} a_d b_{n/d}\biggr) \bigg/ n^z,$$
setting $a_k = \mu(k)$, $b_m = 1$, and $z = 2$. This yields
$$\biggl(\sum_{k\geq 1} {\mu(k)\over k^2}\biggr) \biggl(\sum_{m\geq 1} {1\over m^2}\biggr) = \sum_{n\geq 1}\biggl( \sum_{d\backslash n} \mu(d)\cdot 1\biggr) \bigg/ n^2.$$
As a result of Theorem M, the only non-zero term in the sum on the right-hand side occurs when $n=1$, so the product simplifies to $\mu(1)/1^2 = 1$.\slug

Since $\sum_{m\geq 1} 1/m^2 = \pi^2/6$, we must have
$$\sum_{k\geq 1}{\mu(k)\over k^2} = {6 \over \pi^2} \approx 60.8\%,$$
and this is the probability that two random integers are relatively prime.

\end

### 2019-07-25 misc/proofs/coprime-probability.tex
Too many mays
Changed at most 3x to less than 2x

@@
 $$x^2 - \lfloor x\rfloor^2 = O(x).\oldno 2$$
-For $x^2 - \lfloor x\rfloor ^2$ can be rewritten $(x + \lfloor x\rfloor)(x-\lfloor x\rfloor)$, which is at most $2x\cdot x = 3x$.
+For $x^2 - \lfloor x\rfloor ^2$ can be rewritten $(x + \lfloor x\rfloor)(x-\lfloor x\rfloor)$, which is less than $2x$.
 Next, we use standard formulas to analyse the asymptotic behaviour of the series $\sum_{k>n}(n/k)^2$:
@@
 We now know that $q_n/n^2$ tends to the mysterious summation $\sum_{k\geq 1}\mu(k)/k^2$ as $n$ approaches infinity.
-Before we may compute the value of this summation, we need the following theorem, whose proof is simple enough that it may be included as well.
+Before we are able to compute the value of this summation, we need the following theorem, whose proof is simple enough that it may be included as well.
 

### 2019-07-25 misc/proofs/coprime-probability.tex
Less than 2x+1

@@
 $$x^2 - \lfloor x\rfloor^2 = O(x).\oldno 2$$
-For $x^2 - \lfloor x\rfloor ^2$ can be rewritten $(x + \lfloor x\rfloor)(x-\lfloor x\rfloor)$, which is less than $2x$.
+For $x^2 - \lfloor x\rfloor ^2$ can be rewritten $(x + \lfloor x\rfloor)(x-\lfloor x\rfloor)$, which is less than $2x+1$.
 Next, we use standard formulas to analyse the asymptotic behaviour of the series $\sum_{k>n}(n/k)^2$:

### 2019-08-05 misc/study-notes/abstract-algebra.tex
Fixed typo in graph theory and added algebra

% Abstract Algebra Notes
% Based on lectures given in Fall 2003 by Prof. Benedict Gross
% Notes by Marcel Goh

\input fontmac
\input mathmac

% Useful functions

\def\Sym{\mathop{\rm Sym}\nolimits}

\leftrighttop{NOTES ON ABSTRACT ALGEBRA}{MARCEL GOH}

\maketitlenodate{Notes on Abstract Algebra\footnote{$^1$}{\eightpt Based on a series of lectures given at Harvard University in Fall 2003 by Prof. Benedict Gross.}}{by}{Marcel K.\ Goh}

\section 1. PRELIMINARY NOTIONS FROM LINEAR ALGEBRA

This set of notes assumes that the reader has had some exposure to linear algebra; the most crucial notions will be outlined in this section.

\section 1.1. Basic properties of square matrices

We shall mainly concern ourselves with matrices of the form
$$\pmatrix{
    a_{11} & \cdots & a_{1n} \cr
    \vdots & \ddots & \vdots \cr
    a_{n1} & \cdots & a_{nn}
}$$
where $n$ is some positive integer and each entry $a_{ij}\in \R$. We call the set of all $n\times n$ matrices with real entries $M_n(\R)$.

For two matrices $A$ and $B$, we may form their {\it sum} $A+B = (a_{ij} + b_{ij})$. (This notation means that at the $i$th row and $j$th column, the entry is the sum of $a_{ij}$ and $b_{ij}.)$. Given a real number $\alpha$, we obtain the scalar product $\alpha A  = (\alpha\cdot a_{ij})$ by multiplying every entry in $A$ by $\alpha$.

We can also {\it multiply} $n\times n$ matrices $A$ and $B$ with the following formula:
$$A\cdot B = (c_{ij})\quad\hbox{where } c_{ij} = \sum_{k=1}^n a_{ik}\cdot b_{kj}.\oldno 1$$
Since a matrix represents a linear transformation, multiplying matrices is like composing functions. If $S$ and $T$ are the transformations represented by the matrices $B$ and $A$, respectively, then the matrix product $A\cdot B$ can be thought of as the following composition of transformations:
$$ \matrix{ \R^n & \buildrel S\over \longrightarrow & \R^n & \buildrel T\over \longrightarrow & \R^n } $$
For matrices $A$ and $B$ the commutativity of addition
$$A + B = B + A \oldno 2$$
is valid, and for three matrices $A$, $B$, and $C$, the distributive law
$$A\cdot (B + C) = A\cdot B + A\cdot C\oldno 3$$
and the associativity of multiplication
$$A\cdot (B\cdot C) = (A\cdot B)\cdot C \oldno 4$$
may be proven to hold as well. In particular, ({\oldstyle 4}) is laborious to prove from the definition given in ({\oldstyle 1}), but easy to derive when reasoning about matrices as transformations.

Unlike addition, multiplication is not commutative: $A\cdot B$ does not equal $B\cdot A$ in general. To prove this, we simply note that
$$\pmatrix{ 0 & 1 \cr 0 & 0} \cdot\pmatrix{ 0 & 0 \cr 0 & 1 } = \pmatrix{ 0 & 1 \cr 0 & 0 } \hbox{ but } \pmatrix{ 0 & 0 \cr 0 & 1} \cdot\pmatrix{ 0 & 1 \cr 0 & 0 } = \pmatrix{ 0 & 0 \cr 0 & 0 }.$$
The matrix
$$I =
\pmatrix{
    1 & 0 & \cdots & 0 \cr
    0 & 1 & \cdots & 0 \cr
    \vdots & \vdots \ddots & 0 \cr
    0 0 0 & \cdots & 1
}$$
is called the {\it identity matrix} and has the property that $AI = IA = A$ for any matrix $A$. [When no ambiguity can arise, we often omit the $\cdot$ symbol when denoting a product.]

We say that a matrix $A$ is {\it invertible} if and only if there exists a matrix $B$ such that $AB = BA = I$; otherwise, it is called {\it singular.} Not all matrices are invertible. For example, the matrix ${\bf 0}$, all of whose entries are $0$, is not invertible in any dimension. The identity matrix is easily seen to be invertible (take $B = I$). Note that if an inverse matrix exists for a given matrix $A$, then it is unique, for if $AB = AB' = I$ and $B'A = BA = I$, then, multiplying the first identity by $B$ on the left, we arrive at $BAB = BAB'$, i.e.\ $B = B'$.

Since a $1\times 1$ matrix $(a)$ contains only one real number, it is invertible if and only if $a \neq 0$. A $2\times 2$ matrix $a\,b \choose c\,d$ is invertible if and only if $ad-bc = 0$, since for any such matrix,
$$\pmatrix { a & b \cr c & d } \pmatrix { d & -c \cr -c & a} = \pmatrix { ad-bc & 0 \cr 0 & ad-bc },$$
and we can multiply both sides by $1/(ad-bc)$ to obtain the identity on the right, provided that $ad-bc$ is nonzero.

In general, there exists a function $\det : M_n(\R) \rightarrow \R$ such that a matrix $A\in M_n(\R)$ is invertible if and only if $\det A\neq 0$. This determinant can be calculated as a sum over $n!$ terms; this formula will not be useful for our purposes.

\section 1.2. The general linear group

Let us now restrict our attention to a certain subset of square matrices, namely those whose determinant is nonzero. This set is called the {\it general linear group of degree $n$} and is denoted $GL_n(\R)$ when all matrix entries are real numbers. Remark that, since $(-1) + (1) = (0)$, this set is not closed under addition; scalar multiplication is also no longer a safe operation, since multiplying any matrix by $0$ results in a singular matrix.

In return for these two forfeited closure properties, we get closure under matrix multiplication.

\proclaim Proposition I. Suppose that two matrices $A$ and $B$ are invertible. Then their product $AB$ is also invertible.

\proof Consider $B^{-1}A^{-1}$ and the product $(B^{-1}A^{-1})(AB)$. By associativity of multiplication, this becomes $B^{-1}(A^{-1}A)B = B^{-1}IB = IB^{-1}B = II = I$. Alternatively, use the fact that $\det(AB) = \det(A)\det(B)$, which is nonzero because both $\det(A)$ and $\det(B)$ are nonzero.\slug

Thus the set $GL_n(\R)$, under the operation of matrix multiplication, has a multiplicative inverse $A^{_1}$ for every matrix $A$. It also contains an identity element $I$ and the multiplication operation is associative. These are the properties of a group.

\section 2. GROUPS

\vskip -\medskipamount

\section 2.1. Properties and basic examples

A {\it group} $G$ is a set on which is defined a rule of combination such that the product of two elements $g,h\in G$, denoted $g\cdot h$ or $gh$, is also in $G$. Furthermore, the following three properties must hold:
\medskip
\item {a)} Multiplication must be associative: for all $g,h,k\in G$, $(gh)k = g(hk)$.
\smallskip
\item {b)} There exists an identity element $e\in G$ such that $ge = eg = g$ for all $g\in G$. This element is also often denoted $1$.
\smallskip
\item {c)} For every element $g\in G$, there exists an inverse element $g^{-1}$ such that $gg^{-1} = g^{-1}g = e$.
\medskip
If $gh = hg$ for all $g,h\in G$, then we say the group $G$ is {\it commutative} or {\it abelian}.

The most familiar abelian group is the set of all integers, denoted $\Z$, under the operation of addition. It has $1$ as its identity, inverses $-a$ for every whole number $a$, and the commutative property; likewise, any vector space $V$ is also an abelian group under vector addition.

Group theory is intimately connected to the study of symmetries of an object. Let $T$ be a set and let $\Sym(T)$ denote the set of all bijections from $T$ to itself. This is called the {\it symmetric group} on $T$ as, under composition of functions, it obeys all the group axioms: It contains the identity transformation $e$ and every bijection $f$ has an inverse bijection $f^{-1}$. In some sense, the symmetric group is the most general group, because all other groups arise from adding restrictions to these bijections. For instance, $GL_n(\R) \subset \Sym(\R^n)$. If $T = \{1,\ldots,n\}$, then $\Sym(T)$ is called the symmetric group on $n$ letters and is denoted $S_n$. This group has $n!$ elements and is non-abelian when $n\geq 3$.

\section 2.2 Subgroups

\end

### 2019-08-06 misc/study-notes/abstract-algebra.tex
Added flairs

@@
 \def\Sym{\mathop{\rm Sym}\nolimits}
+\def\Aut{\mathop{\rm Aut}\nolimits}
 
@@
     0 & 1 & \cdots & 0 \cr
-    \vdots & \vdots \ddots & 0 \cr
-    0 0 0 & \cdots & 1
+    \vdots & \vdots & \ddots & \vdots \cr
+    0 & 0 & \cdots & 1
 }$$
@@
 If $gh = hg$ for all $g,h\in G$, then we say the group $G$ is {\it commutative} or {\it abelian}.
+The number of elements in $G$ is called its {\it order} and this value, also denoted $|G|$, need not be finite.
 
@@
 If $T = \{1,\ldots,n\}$, then $\Sym(T)$ is called the symmetric group on $n$ letters and is denoted $S_n$.
-This group has $n!$ elements and is non-abelian when $n\geq 3$.
 
@@
 
+Let $G$ be a group. We call a nonempty subset $H\subset G$ a {\it subgroup} and write $H\leq G$ provided that
+\medskip
+\item {a)} The set $H$ is closed under the multiplication operation of $G$.
+\smallskip
+\item {b)} Whenever $H$ contains an element $a\in G$, $H$ contains its inverse $a^{-1}$ as well.
+\medskip
+It is immediate from these requirements that a subgroup $H$ contains the identity element.
+Since $H$ is nonempty, it contains an element $h$ as well as its inverse $h^{-1}$.
+Then from closure of multiplication we conclude that $hh^{-1} = e\in H$.
+
+We turn our attention to $S_3$, the permutation group on three letters. Notice that this group is not commutative, since the elements
+$$\sigma = \pmatrix{1 & 2 & 3\cr 2 & 3 & 1} \quad \hbox{and} \quad
+  \tau = \pmatrix{1 & 2 & 3\cr 2 & 1 & 3}$$
+do not commute:
+$$\sigma\tau = \pmatrix{1 & 2 & 3 \cr 3 & 2 & 1}, \quad \hbox{but} \quad
+  \tau\sigma = \pmatrix{1 & 2 & 3 \cr 1 & 3 & 2}.$$
+Note that for $k\leq n$, $S_k\leq S_n$ because we can simply fix the letters $k+1,k+2,\ldots,n$.
+An easy corollary, then, is that $S_n$ does not commute for $n\geq 3$.
+This is because $S_3\leq S_n$ and we can simply take $\sigma$ and $\tau$ as elements of the larger group $S_n$ that do not commute.
+
+Another example of a subgroup is the set of $2\times 2$ matrices that stabilise the line $y=0$ (vectors lying on this line remain on this line after transformation).
+In terms of matrices, this set looks like
+$$S = \biggl\{ \pmatrix{a & c \cr 0 & d} : ad\neq 0\biggr\}.$$
+Showing that this set is closed is a simple matter of computing
+$$\pmatrix{a & c \cr 0 & d}\pmatrix{a' & c' \cr 0 & d'}
+  = \pmatrix{aa' & ac' + cd' \cr 0 & dd'},$$
+and observing that the determinant of this matrix is nonzero since all of $a,a',d,d'$ were assumed to be nonzero. (Closure under inversion is also easily derived.)
+
+It is not always easy to characterise all subgroups of a given group. For the additive group of integers, however, the following proposition does so nicely.
+
+\proclaim Proposition S. The subgroups of $\Z$ under addition are precisely given by $b\Z$, where $b$ is a fixed integer.
+
+\proof First, we fix an integer $b$ and show that $b\Z$ is a subgroup.
+Adding two integers $bm + bn$ gives a third integer $b(m+n)$, which is also in $b\Z$, so the set is closed under the operation of addition; likewise $-(bm) = b(-m)$, so the set is closed under additive inverse.
+
+Now we must show that these $b\Z$ are all the possible subgroups.
+Let $H\leq \Z$.
+It is possible that $H$ contains only the identity $0$, in which case $H = 0\Z$.
+If not, let $b$ be the smallest positive integer contained in $H$.
+We know from closure that every multiple of $b$ is in $H$, so $b\Z\subset H$.
+Now let $h$ be any element in $H$.
+By the Euclidean division algorithm, we can divide $h$ by $b$ to get $h = mb + r$, where $mb$ is some multiple of $b$ and $r$ is a remainder lying in the range $0\leq r < b$.
+Since $r = (-mb) + h$, we have $r\in H$.
+But then necessarily we have $r=0$, since $b$ is the smallest positive integer in $H$.
+So $h$ is an integer multiple of $b$ and we have shown that $H \subset b\Z$.
+\slug
+
+Finally, we introduce a specific class of subgroup. If $G$ is a group with an element $g$, the {\it cyclic subgroup} generated by $g$ is the set
+$$\langle g\rangle = \{ g^m : m\in \Z\}.$$
+This is a subgroup because $g^mg^n = g^{m+n}$ and $(g^m)^{-1} = g^{-m}$.
+Note that not all powers are distinct!
+For example, in the group $S_3$, the cyclic subgroup generated by $\tau$ contains only the identity element and $\tau$ itself.
+If $g^m = e$ and $m$ is the smallest positive integer for which this holds, we say that the {\it order} of $g$ is $m$ and write $|g| = m$.
+If no such $m$ exists, then we say $g$ has infinite order.
+
+If $G$ is a group containing an element $g$ such that $\langle g\rangle = G$, then $G$ is called a {\it cyclic group}.
+
+\section 2.3 Isomorphisms and homomorphisms
+
+Consider the group $G_1 = \{\pm 1, \pm i\}$ under complex multiplication alongside the group $G_2 = \langle\rho\rangle\leq S_4$, where $\rho$ is the permutation that takes $1\mapsto 2$, $2\mapsto 3$, $3\mapsto 4$, and $4\mapsto 1$.
+The groups have the following multiplication tables:
+% Each cell is right-aligned in the first table, centred in the second.
+$$G_1{\rm :}\quad\vcenter{\vbox{\offinterlineskip
+    \halign{ 
+        \hfil $#$ & \hfil\vrule# \hfil \enskip &
+        \hfil $#$ & \hfil $#$ & \hfil $#$ & \hfil $#$ \cr
+        %\hfil $#$ \hfil & \hfil\vrule# \hfil \enskip &
+        %\hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil \cr
+        && 1 & i & -1 & -i \cr
+        & height4pt &&& \cr
+        \noalign{\hrule}
+        & height4pt &&&\cr
+        1 && 1 & i & -1 & -i \cr
+        & height4pt &&&\cr
+        i && i & -1 & -i & 1 \cr
+        & height4pt &&&\cr
+        -1 && -1 & -i & 1 & i \cr
+        & height4pt &&&\cr
+        -i && -i & 1 & i & -1 \cr
+    }
+}} \qquad
+G_2{\rm :}\quad\vcenter{\vbox{\offinterlineskip
+    \halign{ 
+        \hfil $#$ \hfil & \hfil\vrule# \hfil \enskip &
+        \hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil \cr
+        && e & \rho & \rho^2 & \rho^3 \cr
+        & height2pt &&& \cr
+        \noalign{\hrule}
+        & height2pt &&&\cr
+        e && e & \rho & \rho^2 & \rho^3 \cr
+        & height2pt &&&\cr
+        \rho && \rho & \rho^2 & \rho^3 & e \cr
+        & height2pt &&&\cr
+        \rho^2 && \rho^2 & \rho^3 & e & \rho \cr
+        & height2pt &&&\cr
+        \rho^3 && \rho^3 & e & \rho & \rho^2 \cr
+    }
+}}
+$$
+It doesn't take long to realise that these two multiplication tables are the same, up to relabeling $1 \equiv e$, $i\equiv \rho$, $-1 \equiv \rho^2$, and $-i \equiv \rho^3$.
+This is an example of the concept of isomorphisms between groups.
+
+Formally, an {\it isomorphism} is a bijection $f:G\rightarrow G'$ from a group to another such that
+$$f(x\cdot y) = f(x) \cdot f(y).$$
+The multiplication on the left-hand side is taking place in $G$ while the right-hand multiplication takes place in $G'$.
+In the above example, the function $f : G_1 \rightarrow G_2$ that takes $i^k \mapsto \rho^k$ for $k=0,1,2,3$ gives an explicit isomorphism.
+If there exists an isomorphism between two groups, we say that they are {\it isomorphic}.
+
+Any two cyclic groups of order $n$ are isomorphic.
+We will not formally prove this, but it is clear that if $G_1$ and $G_2$ are cyclic groups of the same order generated by $g_1$ and $g_2$ respectively, then the function $f : G_1 \rightarrow G_2$ given by $f({g_1}^k) = {g_2}^k$ for any integer $k$ will be a well-defined isomorphism.
+
+As a perhaps surprising example, the group of real numbers under addition and the group of positive real numbers under multiplication are isomorphic to one another.
+The function $f(x) = e^x$ is a bijection from $\R$ to $\R^+$ and $e^{x+y} = e^xe^y$ for real numbers $x$ and $y$.
+
+If two groups $G_1$ and $G_2$ are isomorphic, then
+\medskip
+\item {a)} The groups have the same order, i.e.\ $|G_1| = |G_2|$.
+\smallskip
+\item {b)} Either $G_1$ and $G_2$ are both abelian or they are both non-abelian.
+\smallskip
+\item {c)} Both groups have the same number of elements of every order.
+\medskip
+These properties are useful in showing that two groups are not isomorphic.
+For example, $S_3$ has no element of order $6$, so it is not isomorphic to the cyclic group of order $6$, which has two such elements: 1 and 5.
+
+An isomorphism from a set to itself is called an {\it automorphism}.
+Given a group $G$, we can construct a set $\Aut(G)$ of all automorphisms of $G$ and in fact, it is easily verifiable that $\Aut(G)$ is a group under function composition.
+
+We may obtain a generalisation of an isomorphism by relaxing the requirement that the function be bijective.
+Any map $f : G_1 \rightarrow G_2$ between groups that satisfies $f(x\cdot y) = f(x)\cdot f(y)$ for all $x,y\in G_1$ is called a {\it homomorphism}, and an isomorphism is simply a bijective homomorphism.
+It follows from the definition that any homomorphism maps the identity of the first group to the identity of the second, and that inverses are mapped to inverses.
+The simplest example of a homomorphism is the trivial homomorphism that maps every element of $G_1$ to the identity of $G_2$.
+Another homomorphism that is not an isomorphism is the map from $\Z$ to $S_2$ that takes all even integers to the identity and all odd integers to the permutation that switches 1 and 2.
+
+The {\it kernel} of a homomorphism is a the set of all elements in the domain that map to the identity in the target group.
+The {\it image} of a homomorphism $f: G_1\rightarrow G_2$ is the set of all elements in $G_2$ that equal $f(x)$ for some $x\in G_1$.
+
 \end

### 2019-08-13 misc/study-notes/abstract-algebra.tex
Changed LaTeX to TeX
Added proofmac

@@
 
-\section 1.1. Basic properties of square matrices
+\section 1.1. Basic Properties of Square Matrices
 
@@
 For two matrices $A$ and $B$, we may form their {\it sum} $A+B = (a_{ij} + b_{ij})$.
-(This notation means that at the $i$th row and $j$th column, the entry is the sum of $a_{ij}$ and $b_{ij}.)$.
-Given a real number $\alpha$, we obtain the scalar product $\alpha A  = (\alpha\cdot a_{ij})$ by multiplying every entry in $A$ by $\alpha$.
+(This notation means that at the $i$th row and $j$th column, the entry is the sum of $a_{ij}$ and $b_{ij}.)$ Given a real number $\alpha$, we obtain the scalar product $\alpha A  = (\alpha\cdot a_{ij})$ by multiplying every entry in $A$ by $\alpha$.
 
@@
 
-\section 1.2. The general linear group
+\section 1.2. The General Linear Group
 
@@
 
-\section 2.1. Properties and basic examples
+\section 2.1. Properties and Basic Examples
 
@@
 
-\section 2.3 Isomorphisms and homomorphisms
+\section 2.3 Isomorphisms and Homomorphisms
 
@@
 
+\section ?.?. Group Actions on Sets
+
+Let $G$ be a group and $S$ a set of elements.
+We say that $G$ {\it acts on} $S$ if there exists a map from $G\times S\rightarrow S$ (whose pairs $(g,s)$ are written $g(s)$) such that, for all $s\in S$, $e(s) = s$ and $(gh)(s) = g\big(h(s)\big)$ for every $g,h\in G$.
+For any element $s\in S$, the {\it orbit} is set of elements in $S$ to which $s$ may be taken by $G$:
+$$O_s = \{g(s)\in S : g\in G\}$$
+If there exists an $s$ such that $O_s = S$, then we say that $G$ acts {\it transitively} on $S$.
+In fact, because the orbits of a group action partition the set $S$ into equivalence classes, transitivity of a group action implies there is only one orbit, and for any $s,s'\in S$, there exists some $g\in G$ such that $g(s) = s'$.
+
+The {\it stabiliser} of $s$ is the set of all elements in $G$ that leave $s$ fixed:
+$$G_s = \{g\in G : g(s) = s\}$$
+
+Consider a model example of a transitive action.
+Let $G$ be a group with a subgroup $H\leq G$; let $S$ be the set $\{aH : a\in G\}$ of left cosets of $H$.
+The group $G$ acts on $S$ by taking a coset $aH$ to $gaH$.
+This action is transitive: For any $a, a'$, there exists some $g$ such that $g(a) = a'$, so we can take any $aH$ to $a'H$ via $g$ as well.
+The closure property of subgroups gives us $G_H = H$, and the stabiliser of an arbitrary coset $aH$ is the set $\{g\in G : gaH = aH\}$.
+We find that this is exactly the conjugate subgroup $aHa^{-1}$.
+More generally, if $G$ acts transitively on a set $S$ and $g(s) = s'$, then $G_{s'} = gG_sg^{-1}\subset G$.
+
+\section ?.?. The Sylow Theorems
+
+Suppose that a group $G$ acts on itself by conjugation:
+$$ s\mapsto gsg^{-1}.$$
+Then the orbit of an element $s$ is its conjugacy class and its stabiliser is the set $\{g\in G : gsg^{-1}\}$, also called the centraliser.
+Let $\overline H$ denote the set of all subgroups of $G$.
+Then $G$ also acts on $\overline H$ by conjugation: $g(H) = gHg^{-1}$.
+In this action, the stabiliser of a subgroup $H$ is the set
+$$ G_H = \{ g \in G : gHg^{-1} = H\}.$$
+Note that this is not the centraliser of $H$; elements may not be taken to themselves pointwise, so long as the sets $gHg^{-1}$ and $H$ are equal.
+This subset of $G$ is called the {\it normaliser} of $H$ and is itself a subgroup, denoted $N(H)$.
+
 \end

### 2019-09-15 misc/study-notes/groups.tex
Edited group theory notes

--- 2019-08-13 misc/study-notes/abstract-algebra.tex
@@
-% Abstract Algebra Notes
-% Based on lectures given in Fall 2003 by Prof. Benedict Gross
+% Group Theory Notes
 % Notes by Marcel Goh
@@
 \def\Aut{\mathop{\rm Aut}\nolimits}
+\def\Im{\mathop{\rm Im}\nolimits}
+\def\Id{\mathop{\rm Id}\nolimits}
 
-\leftrighttop{NOTES ON ABSTRACT ALGEBRA}{MARCEL GOH}
+\leftrighttop{NOTES ON THE THEORY OF GROUPS}{MARCEL GOH}
 
-\maketitlenodate{Notes on Abstract Algebra\footnote{$^1$}{\eightpt Based on a series of lectures given at Harvard University in Fall 2003 by Prof. Benedict Gross.}}{by}{Marcel K.\ Goh}
+\maketitlenodate{Notes on the Theory of Groups}{by}{Marcel K.\ Goh}
 
@@
 
-Thus the set $GL_n(\R)$, under the operation of matrix multiplication, has a multiplicative inverse $A^{_1}$ for every matrix $A$.
+Thus the set $GL_n(\R)$, under the operation of matrix multiplication, has a multiplicative inverse $A^{-1}$ for every matrix $A$.
 It also contains an identity element $I$ and the multiplication operation is associative.
@@
 \medskip
-If $gh = hg$ for all $g,h\in G$, then we say the group $G$ is {\it commutative} or {\it abelian}.
-The number of elements in $G$ is called its {\it order} and this value, also denoted $|G|$, need not be finite.
+An immediate consequence of the axioms is that the identity $e$ is unique. For if both $e$ and $e'$ are the identity of a group, then $e = ee' = e'$.
 
-The most familiar abelian group is the set of all integers, denoted $\Z$, under the operation of addition.
-It has $1$ as its identity, inverses $-a$ for every whole number $a$, and the commutative property; likewise, any vector space $V$ is also an abelian group under vector addition.
+The term {\it order} serves a somewhat dual purpose in group theory.
+The order of a group $G$ is the number of elements it contains and and this value, also denoted $|G|$, need not be finite.
+On the other hand, we define the order of a group {\it element} $g$ to be the smallest $k>0$ such that $g^k=e$.
+If no such $k$ exists, then $g$ is said to have infinite order.
+
+Probably the most familiar group is the set of all integers, denoted $\Z$, under the operation of addition.
+It has $1$ as its identity, inverses $-a$ for every whole number $a$, and the commutative property; likewise, any vector space $V$ is also a group under vector addition.
+The set of nonzero real numbers forms a group under multiplication.
+In these examples, the binary operation has the property that for any $g,h\in G$, the products $gh$ and $hg$ are equal.
+A group where this holds is called an {\it abelian} or {\it commutative} group.
+An example of a non-abelian group is the group $Q_8$ of quaternions with identity $1$, governed by the identities $(-1)^2 = 1$ and $i^2 = j^2 = k^2 = ijk = -1$.
+It has the following multiplication table:
+% Right-aligned table with some extra spacing
+$$\vcenter{\vbox{\offinterlineskip
+    \halign{
+        \hfil $#$ & \hfil\vrule# \hfil \enskip &
+        \hfil $\;#\;$ & \hfil $\;#\;$ & \hfil $\;#\;$ & \hfil $\;#\;$ &
+        \hfil $\;#\;$ & \hfil $\;#\;$ & \hfil $\;#\;$ & \hfil $\;#\;$ \cr
+        && 1 & -1 & i & -i & j & -j & k & - k \cr
+        & height4pt &&&&&&& \cr
+        \noalign{\hrule}
+        & height4pt &&&&&&& \cr
+        1 && 1 & -1 & i & -i & j & -j & k & - k \cr
+        & height4pt &&&&&&& \cr
+        -1 && -1 & 1 & -i & i & -j & j & -k & k \cr
+        & height4pt &&&&&&& \cr
+        i && i & -i & -1 & 1 & k & -k & -j & j \cr
+        & height4pt &&&&&&& \cr
+        -i && -i & i & 1 & -1 & -k & k & j & -j \cr
+        & height4pt &&&&&&& \cr
+        j && j & -j & -k & k & -1 & 1 & i & -i \cr
+        & height4pt &&&&&&& \cr
+        -j && -j & j & k & -k & 1 & -1 & -i & i \cr
+        & height4pt &&&&&&& \cr
+        k && k & -k & j & -j & -i & i & -1 & 1 \cr
+        & height4pt &&&&&&& \cr
+        -k && -k & k & -j & j & i & -i & 1 & -1 \cr
+    }
+}}$$
+The general linear group $GL_n(\R)$ of invertible $n\times n$ matrices is another example of a non-abelian group, whenever $n>1$.
 
@@
 Let $T$ be a set and let $\Sym(T)$ denote the set of all bijections from $T$ to itself.
-This is called the {\it symmetric group} on $T$ as, under composition of functions, it obeys all the group axioms: It contains the identity transformation $e$ and every bijection $f$ has an inverse bijection $f^{-1}$.
+This is called the {\it symmetric group} on $T$ as, under composition of functions, it obeys all the group axioms: It contains the identity transformation $\Id_T$ and every bijection $f$ has an inverse bijection $f^{-1}$.
 In some sense, the symmetric group is the most general group, because all other groups arise from adding restrictions to these bijections.
 For instance, $GL_n(\R) \subset \Sym(\R^n)$.
-If $T = \{1,\ldots,n\}$, then $\Sym(T)$ is called the symmetric group on $n$ letters and is denoted $S_n$.
 
-\section 2.2 Subgroups
+\section 2.2. Permutation Groups
+
+A bijection $\sigma$ from a set $T$ to itself is also called a {\it permutation}.
+We will focus on the case where $T$ is finite and we may simply number the elements of $T=\{1,\ldots,n\}$.
+Then the set of permutations of $T$ is denoted $S_n$ and called the {\it symmetric group on $n$ letters} or, alternatively, the {\it permutation group on $n$ letters}.
+An element $\sigma$ of this group may be explicitly presented in such a way that we see where each element is taken to by $\sigma$:
+$$\sigma = \pmatrix{1 & 2 & \cdots & n \cr \sigma(1) & \sigma(2) & \cdots & \sigma(n)}$$
+It is easy to see that there are $n!
+= n(n-1)(n-2)\cdots 1$ different permutations of $n$ letters.
+There are $n$ choices for $\sigma(1)$; subsequently there remain $n-1$ choices for $\sigma(2)$, $n-2$ choices for $\sigma(3)$ and so on until our hand is forced for $\sigma(n)$.
+So $|S_n| = n!$.
+Let us now consider the concrete example of the permutation
+$$\sigma = \pmatrix{1 & 2 & 3 & 4 & 5 & 6 & 7 \cr 1 & 7 & 6 & 4 & 5 & 2 & 3},$$
+which is an element of $S_7$.
+Notice that the elements 1, 4, and 5 are untouched by the permutation and the remaining four items are permuted in the cycle $2\mapsto 7$, $7\mapsto 3$, $3\mapsto 6$, and $6\mapsto 2$.
+This suggests a more concise notation for $\sigma$, since this cycle $(2\,7\,3\,6)$ completely determines the permutation.
+Notation-wise, $(2\,7\,3\,6)$ denotes exactly the same cycle as $(3\,6\,2\,7)$; to avoid confusion, we usually select the one that starts with the smallest number.
+Not every permutation is a cycle; for example the permutation
+$$\tau = \pmatrix{1 & 2 & 3 & 4 \cr 4 & 3 & 2 & 1}$$
+cannot be represented as a single cycle.
+However, it contains two cycles, $(1\,4)$ and $(2\,3)$, so we can simply represent $\tau$ as $(1\,4)(2\,3)$.
+When two cycles permute disjoint sets of elements, the cycles commute, so $(1\,4)(2\,3) = (2\,3)(1\,4)$.
+Every permutation can be represented as a product of disjoint cycles.
+In particular, the identity permutation can be represented as $(1)$, but we will often simply denote it $e$ or $()$.
+
+\section 2.3. Subgroups
 
@@
 
-We turn our attention to $S_3$, the permutation group on three letters. Notice that this group is not commutative, since the elements
-$$\sigma = \pmatrix{1 & 2 & 3\cr 2 & 3 & 1} \quad \hbox{and} \quad
-  \tau = \pmatrix{1 & 2 & 3\cr 2 & 1 & 3}$$
-do not commute:
-$$\sigma\tau = \pmatrix{1 & 2 & 3 \cr 3 & 2 & 1}, \quad \hbox{but} \quad
-  \tau\sigma = \pmatrix{1 & 2 & 3 \cr 1 & 3 & 2}.$$
+We turn our attention to $S_3$.
+This group is not commutative, since the elements $\sigma = (1\,2\,3)$ and $\tau=(1\,2)$ do not commute.
+Multiplying $\sigma\tau$ we get $(1\,3)$ whereas the product $\tau\sigma = (2\,3)$.
+
 Note that for $k\leq n$, $S_k\leq S_n$ because we can simply fix the letters $k+1,k+2,\ldots,n$.
@@
 
-\section 2.3 Isomorphisms and Homomorphisms
+\section 2.4. Cosets and Normal Subgroups
+
+Let $H$ be a subgroup of a group $G$. For any $g\in G$, we can form a {\it left coset} of $H$ by multiplying every element of $H$ by $g$ on the left:
+$$gH = \{ gh : h\in H\}$$
+Symmetrically, we could form a {\it right coset} $Hg$ of $H$. Now we prove that the cosets partition the group $G$.
+
+\proclaim Lemma P. Let $H$ be a subgroup of a group $G$. Then every element of $G$ is in exactly one left coset of $H$.
+
+\proof Let $g\in G$ be given.
+We commence by noting that since $H$ contains the identity element $e$, the element $g$ is in at least one coset, namely $gH$.
+Now we show that any two cosets are either disjoint or equal.
+Suppose $aH$ and $bH$ are two cosets that are not disjoint.
+This implies that there exist $h_1, h_2\in H$ such that $ah_1 = bh_2$.
+We can manipulate this identity to obtain $a^{-1}b = h_1h_2^{-1}$.
+So $a^{-1}b\in H$.
+Then from closure properties of subgroups, the cosets must be equal, since
+$$aH = a(a^{-1}bH) = (aa^{-1})bH = bH.$$
+Thus $g$ is in exactly one coset of $H$.\slug
+
+Any two cosets have the same size, namely the size of $H$, since the for any coset $aH$, the function $f:H\rightarrow aH$ given by $a\mapsto aH$ establishes a bijection.
+This gives us some information about $H$, since it induces a partition of $G$ into equally-sized disjoint cosets.
+In fact, we have just proved the following famous theorem:
+
+\thmparens L {Lagrange} Let $H$ be a subgroup of a group $G$. Then the order of $H$ must divide the order of $G$.\slug
+
+Thus the value $|G|/|H|$ is an integer; it is called the {\it index} of $H$ in $G$ and denoted $[G:H]$.
+Lagrange's theorem also tells us that, for any element $g$ of a group $G$ with order $n$, the order of the cyclic subgroup $\langle g\rangle$ must divide $n$, so the order of the element $g$ divides $n$.
+Consequently, we know $g^n=e$ for any element $g\in G$.
+
+We now know that a subgroup $H$ can partition a group $G$ in two ways, namely into left cosets of the form $aH$ or into right cosets of the form $Ha$.
+Remark that in general, the left cosets {\it do not equal} the right cosets.
+
+\section 2.5. Isomorphisms and Homomorphisms
 
@@
 
-The {\it kernel} of a homomorphism is a the set of all elements in the domain that map to the identity in the target group.
-The {\it image} of a homomorphism $f: G_1\rightarrow G_2$ is the set of all elements in $G_2$ that equal $f(x)$ for some $x\in G_1$.
+Let $f:G\rightarrow G'$.
+The {\it kernel} of $f$, denoted $\ker(f)$, is a the set of all elements in the $G$ that map to the identity $e'$ of the target group $G'$.
+The {\it image} of $f$, denoted $\Im(f)$, is the set of all elements in $G'$ that equal $f(g)$ for some $g\in G$.
+Then the homomorphism $f$ is an isomorphism if and only if $\ker(f) = \{e\}$ and $\Im(f) = G'$.
+It is an easy exercise to verify that $\ker(f)$ and $\Im(f)$ are subgroups of $G$ and $G'$ respectively.
+In fact, the kernel of a homomorphism is a normal subgroup.
 
@@
 
+\section REFERENCES
+
+The contents of this document are heavily based on the following three sets of lectures: Math 122 given by Benedict Gross at Harvard University, Fall 2003; MATH 235 given by Dani Wise at McGill University, Fall 2018; and MATH 456 given by Mika\"el Pichot at McGill University, Fall 2019.
+
 \end

### 2019-09-19 misc/study-notes/groups.tex
Added more to groups doc

@@
 \medskip
-An immediate consequence of the axioms is that the identity $e$ is unique. For if both $e$ and $e'$ are the identity of a group, then $e = ee' = e'$.
+An immediate consequence of the axioms is that the identity $e$ is unique.
+For if both $e$ and $e'$ are the identity of a group, then $e = ee' = e'$.
+Likewise, any element $g$ has a unique inverse, in the sense that if two elements $h$ and $h'$ are both inverses of $g$, then
+$$h = he = h(gh') = (hg)h' = eh' = h',$$
+and they were the same to begin with.
 
@@
 Every permutation can be represented as a product of disjoint cycles.
-In particular, the identity permutation can be represented as $(1)$, but we will often simply denote it $e$ or $()$.
+The identity permutation can be represented as a singleton cycle, e.g\ $(1)$, but we will often simply denote it $e$ or $()$.
+A cycle that moves $k$ elements is called a $k$-cycle; a 2-cycle is also called a {\it transposition}.
+
+Every cycle can be expressed as a product of transpositions in the following manner:
+$$(x_1\,x_2\,x_3\,\ldots\,x_{n-1}\,x_n) = (x_1\,x_2)(x_2\,x_3)\cdots(x_{n-1}\,x_n)$$
+A consequence of this is that every permutation $\sigma\in S_n$ can be expressed as a product $\sigma = \tau_1\tau_2\cdots\tau_k$ of transpositions.
+We can classify permutations as {\it even} or {\it odd} depending on the parity of $k$.
+In particular, the identity permutation $()$ is a product of 0 cycles, so it is even.
+Thus the set of even permutations of $n$ elements has a group structure.
+It is called the {\it alternating group on $n$ letters} and is denoted $A_n$.
+Because the set $A_n$ is a a subset of $S_n$ and $A_n$ is a group under the same binary operation that defines $S_n$, $A_n$ is an example of a subgroup.
 
@@
 
-Any two cosets have the same size, namely the size of $H$, since the for any coset $aH$, the function $f:H\rightarrow aH$ given by $a\mapsto aH$ establishes a bijection.
+Any two cosets have the same size, namely the size of $H$, since the for any coset $aH$, the function $f:H\rightarrow aH$ given by $h\mapsto ah$ establishes a bijection.
 This gives us some information about $H$, since it induces a partition of $G$ into equally-sized disjoint cosets.
@@
 Remark that in general, the left cosets {\it do not equal} the right cosets.
+As an example, let $H = \{(), (1\,2)\} \leq S_3$.
+Then we have
+$$(1\,3)H = \{(1\,3), (1\,2\,3)\} \neq \{(1\,3),(1\,3\,2)\} = H(1\,3).$$
+However, there do exist subgroups whose left cosets equal their right cosets.
+If $N$ is a subgroup of a group $G$ with the property that $aN = Na$ for all $a\in G$, then we call $N$ a {\it normal} subgroup.
+Normal subgroups have the property that the product of two cosets is itself a coset:
+$$aH\cdot bH = a\cdot Hb\cdot H = a \cdot bH \cdot H = (ab)\cdot HH = (ab)H$$
 

### 2019-09-30 misc/study-notes/groups.tex
Fixed typo in groups and fixed teletype
Possibly fixed issues regarding mathmac and fontmac clashing

@@
 The term {\it order} serves a somewhat dual purpose in group theory.
-The order of a group $G$ is the number of elements it contains and and this value, also denoted $|G|$, need not be finite.
+The order of a group $G$ is the number of elements it contains and this value, also denoted $|G|$, need not be finite.
 On the other hand, we define the order of a group {\it element} $g$ to be the smallest $k>0$ such that $g^k=e$.
@@
 
-\thmparens L {Lagrange} Let $H$ be a subgroup of a group $G$. Then the order of $H$ must divide the order of $G$.\slug
+\parenproclaim Theorem L (Lagrange). Let $H$ be a subgroup of a group $G$. Then the order of $H$ must divide the order of $G$.\slug
 
@@
 $$aH\cdot bH = a\cdot Hb\cdot H = a \cdot bH \cdot H = (ab)\cdot HH = (ab)H$$
+
+Now we introduce the concept of conjugacy, which is closely intertwined with normal subgroups.
+For elements $a$ and $b$ of a group $G$, we say $a$ is {\it conjugate} to $b$ if there exists an element $g\in G$ such that $gag^{-1} = b$.
+It is easy to see that this relation is reflexive (with $g=e$) and symmetric (replace any $g$ with its inverse), so it makes sense to say that $a$ and $b$ are conjugate elements of a group.
+We can also prove transitivity: If there exists $g\in G$ such that $gag^{-1} = b$ and $h\in G$ such that $hbh^{-1} = c$, then $hgag^{-1}h^{-1}= (hg)a(hg)^{-1} = c$.
+Thus conjugacy is an equivalence relation that partitions a group into {\it conjugacy classes}.
+
+The following lemma describes the link between conjugacy and normal subgroups.
+
+\proclaim Lemma C. Normal subgroups are exactly the subgroups that are stable under conjugation.
+
+\proof Let $H$ be a normal subgroup of a group $G$.
+Let $g\in G$ and $h\in H$ be arbitrarily chosen.
+Since $gH=Hg$, there is an element $h'\in H$ such that $gh = h'g$.
+Then we see that $ghg^{-1} = h'$, an element in $H$.
+So we conclude that normal subgroups are stable under conjugation.
+
+On the other hand, suppose $H$ is a subgroup such that, for any $g\in G$ and $h\in H$, $ghg^{-1}\in H$.
+Then $gHg^{-1}\subset H$ and we can conclude that $gH\subset Hg$.
+But $g^{-1}$ is also an element of $G$, so we can apply our hypothesis to get that $g^{-1}Hg\subset H$ and $Hg \subset gH$.
+So $gH=Hg$ for all $g\in G$ and $H$ is normal.\slug
 
@@
 It is an easy exercise to verify that $\ker(f)$ and $\Im(f)$ are subgroups of $G$ and $G'$ respectively.
-In fact, the kernel of a homomorphism is a normal subgroup.
+In fact, the kernel of a homomorphism is a normal subgroup, as we shall now prove.
+
+\proclaim Theorem K. Let $f:G\rightarrow G'$ be a group homomorphism. Then $K = \ker(f)$ is a normal subgroup of $G$.
+
+\proof By Lemma 2.4.C, we need only show that $K$ is stable under conjugation.
+Let $k\in K$ be given and let us conjugate it by an element $a\in G$ to get $aka^{-1}$.
+Applying the homomorphism $f$ to $aka^{-1}$, we find that
+$$f(aka^{-1}) = f(a)f(k)f(a^{-1}) = f(a)e'f(a)^{-1},$$
+where $e'$ is the identity of $G'$. But $f(a)e'f(a)^{-1} = e'$, so $aka^{-1}\in \ker(f) = K$.\slug
 

### 2019-09-30 misc/study-notes/ndmi011notes.tex
Fixed typo in graph theory and added algebra
Added flairs
Possibly fixed issues regarding mathmac and fontmac clashing

@@
 
-\thmparens{G}{Generalised binomial theorem} For any $r\in \R$,
+\parenproclaim Theorem G (Generalised binomial theorem). For any $r\in \R$,
 $$\sum_{i\geq 0} {r\choose i} x^i = (1 + x)^r.$$
@@
 
-\thmparens{H}{Hall} A bipartite graph $G=(A\cup B, E)$ (with $A$ and $B$ disjoint) has a matching $M\subseteq E$ of size $|M| = |A|$ if and only if $|N(S)|\geq |S|$ for all $S\subseteq A$.
+\parenproclaim Theorem H (Hall).
+A bipartite graph $G=(A\cup B, E)$ (with $A$ and $B$ disjoint) has a matching $M\subseteq E$ of size $|M| = |A|$ if and only if $|N(S)|\geq |S|$ for all $S\subseteq A$.
 
@@
 
-\thmparens{K}{K\H onig's theorem} Let $G$ be a bipartite graph with maximum matching $M$ and minimum vertex cover $C$. Then $|M| = |C|$.\slug
+\parenproclaim Theorem K (K\H onig's theorem). Let $G$ be a bipartite graph with maximum matching $M$ and minimum vertex cover $C$. Then $|M| = |C|$.\slug
 
@@
 
-\thmparens{M}{Menger}
-Let $G$ be a graph. The following both hold:
+\parenproclaim Theorem M (Menger). Let $G$ be a graph. The following both hold:
 \medskip
@@
 
-\thmparens{C}{Cayley's formula} The number of spanning trees of a complete graph with $n$ vertices is
+\parenproclaim Theorem C (Cayley's formula). The number of spanning trees of a complete graph with $n$ vertices is
 $$ T(K_n) = n^{n-2}.$$
@@
 
-\thmparens{S}{Sperner} Any antichain of a poset of $X = \{1, \ldots, n\}$ has size at most $n\choose \lfloor n/2 \rfloor$
+\parenproclaim Theorem S (Sperner). Any antichain of a poset of $X = \{1, \ldots, n\}$ has size at most $n\choose \lfloor n/2 \rfloor$
 
@@
 
-\proclaim Theorem R. Let $G=(V,E)$ be a graph. If $|V|\geq {k-l-2\choose k-1} = {k+l-2\choose l-1}$, then $\alpha(G)\geq l$ or $\omega(G)\geq k$.
+\proclaim Theorem R. Let $G=(V,E)$ be a graph. If $|V|\geq {k+l-2\choose k-1} = {k+l-2\choose l-1}$, then $\alpha(G)\geq l$ or $\omega(G)\geq k$.
 
 \proof By induction on $k+l$.
-In the base case $k = 1$ or $l=1$, all the theorem says is that there is an independent set or clique of size 4.
+In the base case $k = 1$ or $l=1$, all the theorem says is that there is an independent set or clique of size 1.
 For the inductive step, let $k,l\geq 2$ and assume that the claim holds for $k,l-1$ or $k-1,l$.
@@
 
-\thmparens{E}{Erd\H os-Szekeres} For any finite sequence $S$ of at least $(r-1)(s-1) + 1$ numbers, there is an increasing subsequence of length $r$ or a decreasing subsequence of length $s$.
+\parenproclaim Theorem E (Erd\H os-Szekeres).
+For any finite sequence $S$ of at least $(r-1)(s-1) + 1$ numbers, there is an increasing subsequence of length $r$ or a decreasing subsequence of length $s$.
 
@@
 
-\thmparens{H}{Hamming bound} If an $(n,k,d)_q$-code exists, then
+\parenproclaim Theorem H (Hamming bound). If an $(n,k,d)_q$-code exists, then
 $$q^k\cdot\hbox{\rm vol}\left(\left\lfloor {d-1\over 2} \right\rfloor, n\right) \leq q^n.$$
@@
 
-\thmparens{S}{Singleton bound} If $C$ is an $(n,k,d)_q$-code, then $k\leq n-d+1$.
+\parenproclaim Theorem S (Singleton bound). If $C$ is an $(n,k,d)_q$-code, then $k\leq n-d+1$.
 

### 2019-09-30 misc/study-notes/nopt048notes.tex
Possibly fixed issues regarding mathmac and fontmac clashing

@@
 
-\thmparens{M}{Minkowski-Weyl} A set of points $P$ is a polyhedron if and only if there exist finite sets $V$ and $Y$ such that $P = \conv V + \cone Y$ where the sum $A + B$ denotes the set $\{a + b : a\in A, b\in B\}$.
+\parenproclaim Theorem M (Minkowski-Weyl).
+A set of points $P$ is a polyhedron if and only if there exist finite sets $V$ and $Y$ such that $P = \conv V + \cone Y$ where the sum $A + B$ denotes the set $\{a + b : a\in A, b\in B\}$.
 
@@
 
-\thmparens{W}{Weak Duality Theorem} Let $P$ and $D$ be primal and dual linear programs of the above form and let $x$ and $y$ be respective feasible solutions. Then $c^Tx \leq b^Ty$.
+\parenproclaim Theorem W (Weak Duality Theorem). Let $P$ and $D$ be primal and dual linear programs of the above form and let $x$ and $y$ be respective feasible solutions. Then $c^Tx \leq b^Ty$.
 
-\thmparens{S}{Strong Duality Theorem} Let $P$ and $D$ be primal and dual linear programs. Then exactly one of the following holds:
+\parenproclaim Theorem S (Strong Duality Theorem). Let $P$ and $D$ be primal and dual linear programs. Then exactly one of the following holds:
 \medskip
@@
 
-\lemparens{F}{Farkas lemma}
-For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
+\parenproclaim Lemma F (Farkas lemma). For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
 \medskip
@@
 
-\lemparens{G}{Farkas lemma 2.0}
-For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
+\parenproclaim Lemma G (Farkas lemma 2.0). For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
 \medskip
@@
 The duality of these two programs results in K\H{o}nig's theorem.
-\thmparens{K}{K\H{o}nig's theorem}
-    Suppose $G=(V,E)$ is a bipartite graph. Let $M\subset E$ denote the maximum matching and let $S\subset V$ denote the minimum vertex cover. Then $|M|=|S|$.\slug
+
+\parenproclaim Theorem K (K\H{o}nig's theorem).
+Suppose $G=(V,E)$ is a bipartite graph.
+Let $M\subset E$ denote the maximum matching and let $S\subset V$ denote the minimum vertex cover.
+Then $|M|=|S|$.\slug
 
@@
 
-\thmparens{R}{Rank characterisation of matroids} Let $E$ be a finite set and suppose that $r : {\cal P}(E)\rightarrow \R$ satisfies the three peoperties of the rank function. Then $M=(E,I)$, with
+\parenproclaim Theorem R (Rank characterisation of matroids).
+Let $E$ be a finite set and suppose that $r : {\cal P}(E)\rightarrow \R$ satisfies the three peoperties of the rank function.
+Then $M=(E,I)$, with
 $$I = \{Y \subset E : |Y| = r(Y)\},$$
@@
 
-\thmparens{E}{Greedy equivalence} Let $M=(E,I)$ be an independence system. Then if the weighted greedy algorithm works for all (non-negative) weights, then $M$ is a matroid.\slug
+\parenproclaim Theorem E (Greedy equivalence). Let $M=(E,I)$ be an independence system. Then if the weighted greedy algorithm works for all (non-negative) weights, then $M$ is a matroid.\slug
 

### 2019-10-28 misc/study-notes/groups.tex
Studying for groups midterm

@@
 \def\Aut{\mathop{\rm Aut}\nolimits}
+\def\Out{\mathop{\rm Out}\nolimits}
+\def\Inn{\mathop{\rm Inn}\nolimits}
 \def\Im{\mathop{\rm Im}\nolimits}
@@
 
-If $G$ is a group containing an element $g$ such that $\langle g\rangle = G$, then $G$ is called a {\it cyclic group}.
+If $G$ is a group containing an element $g$ such that $\langle g\rangle = G$, then $G$ is called {\it cyclic} or {\it singly-generated}.
+More generally, if $S\subset G$ and we let $S^{-1}$ denote the set $\{s^{-1} : s\in S\}$, then the set
+$$\langle S\rangle = \{s_1s_2\cdots s_n : n\in\N, s_i\in S\cup S^{-1}\}$$
+is a subgroup of $G$, called the subgroup generated by $S$. If it so happens that $\langle S\rangle = G$, then we way that $S$ is a {\it generating set} of $G$.
 
@@
 In the above example, the function $f : G_1 \rightarrow G_2$ that takes $i^k \mapsto \rho^k$ for $k=0,1,2,3$ gives an explicit isomorphism.
-If there exists an isomorphism between two groups, we say that they are {\it isomorphic}.
+If there exists an isomorphism between two groups $G$ and $H$, we say that they are {\it isomorphic} and write $G\simeq H$.
 
@@
 
-An isomorphism from a set to itself is called an {\it automorphism}.
-Given a group $G$, we can construct a set $\Aut(G)$ of all automorphisms of $G$ and in fact, it is easily verifiable that $\Aut(G)$ is a group under function composition.
-
 We may obtain a generalisation of an isomorphism by relaxing the requirement that the function be bijective.
@@
 
-Let $f:G\rightarrow G'$.
-The {\it kernel} of $f$, denoted $\ker(f)$, is a the set of all elements in the $G$ that map to the identity $e'$ of the target group $G'$.
-The {\it image} of $f$, denoted $\Im(f)$, is the set of all elements in $G'$ that equal $f(g)$ for some $g\in G$.
-Then the homomorphism $f$ is an isomorphism if and only if $\ker(f) = \{e\}$ and $\Im(f) = G'$.
-It is an easy exercise to verify that $\ker(f)$ and $\Im(f)$ are subgroups of $G$ and $G'$ respectively.
-In fact, the kernel of a homomorphism is a normal subgroup, as we shall now prove.
+\section 2.6. The First Isomorphism Theorem
 
-\proclaim Theorem K. Let $f:G\rightarrow G'$ be a group homomorphism. Then $K = \ker(f)$ is a normal subgroup of $G$.
+Let $f:G\rightarrow H$ be a homomorphism.
+The {\it kernel} of $f$, denoted $\ker(f)$, is a the set of all elements in the $G$ that map to the identity $e_H$ of the target group $H$.
+The {\it image} of $f$, denoted $\Im(f)$, is the set of all elements in $H$ that equal $f(g)$ for some $g\in G$.
+In particular, the homomorphism $f$ is an isomorphism if and only if $\ker(f) = \{e\}$ and $\Im(f) = G$.
+It is an easy exercise to verify that $\ker(f)$ and $\Im(f)$ are always subgroups of $G$ and $H$ respectively.
 
-\proof By Lemma 2.4.C, we need only show that $K$ is stable under conjugation.
-Let $k\in K$ be given and let us conjugate it by an element $a\in G$ to get $aka^{-1}$.
-Applying the homomorphism $f$ to $aka^{-1}$, we find that
-$$f(aka^{-1}) = f(a)f(k)f(a^{-1}) = f(a)e'f(a)^{-1},$$
-where $e'$ is the identity of $G'$. But $f(a)e'f(a)^{-1} = e'$, so $aka^{-1}\in \ker(f) = K$.\slug
+Now we want to describe the preimages of $\Im(f)$.
+For each element $h\in Im(f)\leq H$, the subset $\{g : f(g) = h\}$ is its {\it fibre}.
+The following lemma establishes that the kernel of a homomorphism is normal and relates the fibres of the image to cosets of the kernel.
+Since the image of a homomorphism is a subgroup, we can assume that the homomorphism is surjective without losing generality.
+
+\parenproclaim Lemma F (Fibre lemma).
+Let $f:G\rightarrow H$ be a surjective group homomorphism with kernel $K$.
+Then for any $h\in H$ and $g\in G$ that satisfies $f(g) = h$, we have $f^{-1}(h) = gK = Kg$, implying that $K$ is a normal subgroup of $G$.
+
+\proof Let $h\in H$ and let $X$ denote $f^{-1}(h)$, the fibre of $h$.
+Now take $g\in G$ such that $f(g) = h$.
+Since $f(k) = e_H$ for all $k\in K$, we have $f(sk) = f(s)f(k) = h = f(k)f(s) = f(ks)$.
+So $sK\subset X$ and $Ks\subset H$.
+To prove the reverse inclusions, we let $x\in X$ (so $f(x) = h$).
+Then since $f$ is a homomorphism, we have $f(g^{-1}) = h^{-1}$ and
+$$f(xg^{-1}) = f(x)f(g^{-1}) = hh^{-1} = e_H = h^{-1}h = f(g^{-1})f(x) = f(g^{-1}x).$$
+This implies that $xg^{-1}\in K$, so multiplying on the right by $g$, we find that $x\in Kg$. Similarly, $g^{-1}x\in K$ so $x\in gK$.\slug
+
+Now for any surjective map $f$ from sets $G$ to $H$, we can define an equivalence relation on elements $a,b\in G$:
+$$a\sim b\quad \hbox{if and only if}\quad f(a) = f(b)$$
+The set of equivalence classes is denoted $G/{\sim}$ and there exists a bijection $f':G/{\sim} \rightarrow H$.
+Now if $f$ is a group homomorphism, then we can pull back the group structure of $H$ to $G/{\sim}$ using the bijection.
+Then the set
+$$G/K = \{gK : g\in G\}.$$
+has a group structure and is called the {\it quotient group}.
+Multiplication in the quotient group is given by $gK\cdot hK = (g\cdot h) K$, which is well-defined because $K$ is normal.
+So the group structure of $G/K$ depends only on the kernel $K$ and not the homomorphism that induced it.
+The results of this section are summed up in the following theorem.
+
+\parenproclaim Theorem I (First isomorphism theorem).
+Let $f:G\rightarrow H$ be a surjective group homomorphism.
+Then $K = \ker(f)$ is a normal subgroup of $G$ and there exists a group homomorphism $f':G/K \rightarrow H$.\slug
+
+\section 2.7. Inner and Outer Automorphisms
+
+An isomorphism from a group to itself is called an {\it automorphism}.
+Given a group $G$, we can construct a set $\Aut(G)$ of all automorphisms of $G$ and in fact, it is easily verifiable that $\Aut(G)$ is a group under function composition.
+
+For any element $a$ of a group $G$, the map $f_a$ defined by
+$$f_a(s) = asa^{-1}$$
+is an automorphism.
+Such automorphisms, which conjugate the group by a fixed element, are called {\it inner} automorphisms.
+An automorphism that is not inner is called an {\it outer} automorphism.
+There is a homomorphism $\phi:G\rightarrow \Aut(G)$ given by $a\mapsto f_a$; that is, $f_af_b = f_{ab}$.
+This follows from associativity:
+$$f_af_b(s) = a(bsb^{-1})a^{-1} = (ab)s(ab)^{-1} = f_{ab}(s)$$
+The image of the homomorphism $\phi$ is denoted $\Inn(G)$.
+
+The {\it centre} of a group $G$ is the subgroup
+$$Z(G) = \{z\in G : zg = gz\ \hbox{for all}\ g\in G\}.$$
+
+\proclaim Lemma Z. The centre of $G$ is the kernel of the map from $G$ to $\Aut(G)$ given by $a\mapsto f_a$.
+
+\proof Let $a\in Z(G)$ and note that for an arbitrary element $s\in G$,
+$$f_a(s) = asa^{-1} = aa^{-1}s = s$$
+so $f_a$ is the identity automorphism.
+Likewise, if some $f_a\in \Aut(G)$ is the identity automorphism, then $f_a(s) = asa^{-1} = s$ for all $s\in G$.
+So $as = sa$ for all $s\in G$ and $a\in Z(G)$.\slug
+
+From Lemma 2.4.C, we know that normal subgroups are exactly the subgroups that are stable under conjugacy, i.e.\ stable under inner automorphism.
+We can extend this into a more general definition.
+If $H\leq G$ and $f(H) = H$ for all $f\in \Aut(G)$, then we say that $H$ is a {\it characteristic} subgroup.
+(Of course, every characteristic subgroup is also normal.)
+
+\proclaim Lemma C. The centre $Z(G)$ of a group $G$ is characteristic.
+
+\proof It must be shown that if $g\in Z(G)$, then for all $f\in\Aut(G)$, $f(g)\in Z(G)$.
+We prove this by contraposition.
+Let $g\in G$ and suppose there exists $f\in\Aut(G)$ such that $f(g)\notin Z(G)$.
+Then there exists $s\in G$ such that $f(g)\cdot s \neq s\cdot f(g)$.
+Now we apply the inverse automorphism $f^{-1}$ to both sides to get that $g\cdot f^{-1}(s) \neq f^{-1}(s)\cdot g$.
+Hence there exists an element of $G$, namely $f^{-1}(s)$, that $g$ does not commute with, so $g\notin Z(G)$.\slug
+
+By the first isomorphism theorem, $\Inn(G)\simeq G/Z(G)$; in particular, if $G$ is abelian, we have $Z(G) = G$ and $\Inn(G)$ is trivial.
+Because it arose as the image of a group homomorphism $\phi$, we know that $\Inn(G)\leq \Aut(G)$.
+In fact, we can prove that $\Inn(G)$ is normal in $\Aut(G)$.
+
+\proclaim Lemma N. Let $G$ be a group. Then $\Inn(G)$ is a normal subgroup of $\Aut(G)$.
+
+\proof Let $f_a\in Inn(G)$ be the map that conjugates by an element $a\in G$. Let $g\in \Aut(G)$ be arbitrary. Then
+$$g\circ f_a\circ g^{-1} = g(a\cdot g^{-1}(s)\cdot a^{-1}) = g(a)\cdot s\cdot g(a)^{-1},$$
+and we see that $g\circ f_a \circ g^{-1} = f_{g(a)}\in Inn(G)$. So $\Inn(G)$ is stable under conjugation in $Aut(G)$.\slug
+
+Now by the first isomorphism theorem, we have a group $\Out(G) = \Aut(G)/\Inn(G)$, called the {\it outer automorphism group}.
+Note that this is not the set of outer automorphisms!
+(The set of outer automorphisms do not form a group because the identity automorphism is inner.)
+
+\section ?.?. Simplicity of the Alternating Group
+
+A group $G$ is {\it simple} if its only normal subgroups are the trivial group $\{e\}$ and the whole group $G$.
+In this section we show that for $n\geq 5$, the alternating group $A_n$ is simple.
+First we establish that the set of 3-cycles generate $A_n$.
+
+\proclaim Lemma G. Every permutation in $A_n$ can be written as a product of 3-cycles.
+
+\proof Since every permutation in $A_n$ can be written as a product of an even number of transpositions, it suffices to show that any product $(a\,b)(c\,d)$ can be rewritten as a product of 3-cycles.
+There are three cases to consider.
+Firstly, if the transpositions have two numbers in common, then $(a\,b) = (c\,d)$ and the product is simply the identity permutation.
+If the transpositions have a single point in common, then the product is a 3-cycle, since $(a\,b)(b\,c) = (a\,b\,c)$.
+Lastly, in the case that all four points are distinct, we have
+$$(a\,b)(c\,d) = (a\,b\,c)(b\,c\,d),$$
+a product of 3-cycles.
+Hence the set of 3-cycles generates $A_n$.\slug
+
+In fact, we can shrink the generating set to 3-cycles of a specific form.
+
+\proclaim Lemma H. The $n-2$ permutations $(1\,2\,3), (1\,2\,4),\ldots,(1\,2\,n)$ generate $A_n$.
+
+\proof By Lemma G, we need only show that any 3-cycle can be rewritten as a product of cycles $(1\,2\,k)$.
+Since the square of $(1\,2\,k)$ is $(2\,1\,k)$, any cycle that contains both 1 and 2 can be generated by the permutations above.
+Then, note that any cycle $(a\,b\,c)$ that fixes 1 can be rewritten $(1\,a\,b)(1\,b\,c)$.
+Now we can rewrite each of these cycles as follows:
+$$(1\,j\,k) = (1\,2\,k)(1\,2\,k)(1\,2\,j)(1\,2\,k)$$
+So every 3-cycle can be expressed as a product of cycles of the form $(1\,2\,k)$.\slug
+
+This result can be used to show that any normal subgroup of $A_n$ that contains a 3-cycles is necessarily the whole group $A_n$.
+
+\proclaim Lemma K. Let $n\geq 4$ and let $N$ be a normal subgroup of the group $A_n$. If $N$ contains a 3-cycle, then $N = A_n$.
+
+\proof Without loss of generality, suppose that $N$ contains the cycle $(1\,2\,3)$. Then it also contains its inverse $(2\,1\,3)$ and for any $\sigma\in A_n$,
+$$\sigma (2\,1\,3) \sigma^{-1}\in N.$$
+So let $\sigma = (1\,2)(3\,k)$ for $k\geq 4$ and we have
+$$\sigma (2\,1\,3) \sigma^{-1} = (1\,2\,k).$$
+By Lemma H, these generate $A_n$, so $N$ is the entire group.\slug
+
+We may finally prove that $A_n$ is simple for $n\geq 5$. The following theorem is due to E. Galois.
+
+\proclaim Theorem S. Let $n\geq 5$ and let $N\geq A_n$ be non-trivial. Then $N=A_n$.
+
+\proof Let $\tau\in N$ be a non-identity permutation that fixes as many symbols as possible. We will show, by contradiction, that $\tau$ must be a 3-cycle, i.e.\ it fixes all symbols except three.
+
+Suppose that $\tau$ moves more than three symbols.
+Then either $\tau$ has a cycle with more than three symbols, or $\tau$ consists of disjoint transpositions.
+Without losing generality, we use the symbols 1, 2, 3, 4, and 5.
+In the first case, we have
+$$\tau = (1\,2\,3\,\ldots)\ldots$$
+and we can conjugate by $(3\,4\,5)$ to get
+$$\tau' = (3\,4\,5)\tau(3\,4\,5)^{-1} = (1\,2\,4\,\ldots)\ldots.$$
+Then $\tau^{-1}\tau'$ fixes 1 where $\tau$ did not, a contradiction.
+
+In the case that $\tau$ only contains transpositions, we have
+$$\tau = (1\,2)(3\,4)\ldots,$$
+and again we conjugate by $(3\,4\,5)$ to get
+$$\tau' = (3\,4\,5)\tau(3\,4\,5)^{-1} = (1\,2)(4\,5)\ldots)\ldots.$$
+Then $\tau^{-1}\tau'$ fixes 1 and 2 and $\tau$ does not, leading to a contradiction.
+
+The contradiction in both cases implies that $\tau$ moves three symbols or less.
+In the alternating group, this implies that $\tau$ is a 3-cycle.
+So $N$ contains a 3-cycle and by Lemma K, $N=A_n$.\slug
 
@@
 Let $G$ be a group and $S$ a set of elements.
-We say that $G$ {\it acts on} $S$ if there exists a map from $G\times S\rightarrow S$ (whose pairs $(g,s)$ are written $g(s)$) such that, for all $s\in S$, $e(s) = s$ and $(gh)(s) = g\big(h(s)\big)$ for every $g,h\in G$.
+We say that $G$ {\it acts on} $S$ if there exists a map from $G\times S\rightarrow S$ (whose pairs $(g,s)$ are written $g(s)$ or sometimes $gs$) such that, for all $s\in S$, $e(s) = s$ and $(gh)(s) = g\big(h(s)\big)$ for every $g,h\in G$.
 For any element $s\in S$, the {\it orbit} is set of elements in $S$ to which $s$ may be taken by $G$:

### 2019-11-26 misc/study-notes/groups.tex
Edited group theory notes

@@
 
+Let $G$ be a group, $H$ a subgroup of $G$, and let $S$ be a non-empty set of elements of $G$.
+The set of $h\in H$ such that $hSh^{-1} = H$ is called the {\it normaliser of $S$ in $H$}, denoted $N_H(S)$.
+The set of $h\in H$ such that $hsh^{-1} = s$ for all $s\in S$ is called the {\it centraliser of $S$ in $H$}, denoted $Z_H(S)$.
+It is easy to check that these are both subgroups of $H$.
+Note that if $S$ consists only of one element, then the centraliser and normaliser are equal and the inclusion $Z_H(S) \subset N_H(S)$ always holds.
+If $H = G$, then we may omit the subscript and simply call the subgroups the normaliser and centraliser of $S$.
+
 \section 2.5. Isomorphisms and Homomorphisms
@@
 
-The {\it centre} of a group $G$ is the subgroup
+The {\it centre} of a group $G$ is defined as the centraliser of $G$ in itself:
 $$Z(G) = \{z\in G : zg = gz\ \hbox{for all}\ g\in G\}.$$
@@
 Let $G$ be a group and $S$ a set of elements.
-We say that $G$ {\it acts on} $S$ if there exists a map from $G\times S\rightarrow S$ (whose pairs $(g,s)$ are written $g(s)$ or sometimes $gs$) such that, for all $s\in S$, $e(s) = s$ and $(gh)(s) = g\big(h(s)\big)$ for every $g,h\in G$.
+We say that $G$ {\it acts on} $S$ if there exists a map from $G\times S\rightarrow S$ (whose pairs $(g,s)$ are written $g(s)$ or $gs$) such that, for all $s\in S$, $e(s) = s$ and $(gh)(s) = g\big(h(s)\big)$ for every $g,h\in G$.
 For any element $s\in S$, the {\it orbit} is set of elements in $S$ to which $s$ may be taken by $G$:
@@
 
+We can equivalently describe a group action of $G$ on a set $S$ as a homomorphism from $G$ to $\Sym(S)$.
+If this map is injective, we say the action is {\it faithful}.
+Most group actions that we will study are faithful, and a very simple one is given by the following theorem.
+
+\parenproclaim Theorem C (Cayley). Let $G$ be a finite group. Then $G$ is isomorphic to a subgroup of $\Sym(G)$.
+
+\proof For every element $g$ of $G$, let $\sigma_g : G\leftarrow \Sym(G)$ be given by $x\mapsto gx$.
+The map $\sigma_g$ is injective, since in $G$, $g_1x = g_2x$ implies that $g_1 = g_2$.
+For any given $x\in G$, we have $\sigma_g(g^{-1}x) = x$, so the map is also surjective.
+Then we also have $\sigma_{gh}(x) = (gh)x = g(hx) = \sigma_g(\sigma_h(x)) = (\sigma_g\circ\sigma_h)(x)$, establishing the isomorphism.\slug
+
+The injection from $G$ to $\Sym(G)$ is faithful, since if $\sigma_{g_1} = \sigma_{g_2}$, then $\sigma_{g_1}(e) = g_1e = g_2e = \sigma_{g_2}(e)$, and $g_1 = g_2$.
+We also observe that $\sigma_e$ is the identity permutation.
+This action of a group $G$ on itself is called the {\it left Cayley action} or {\it left regular action}.
+We could also have described an isomorphism given by $g \mapsto (x \mapsto g)$, but this does not give a homomorphism since now, $\sigma_{gh} = \sigma_h\circ\sigma_g$.
+This can be mended by letting $\sigma_g$ be given by $\sigma_g(x) = xg^{-1}$.
+This is called the {\it right regular action}.
+
 \section ?.?. The Sylow Theorems
@@
 $$ s\mapsto gsg^{-1}.$$
-Then the orbit of an element $s$ is its conjugacy class and its stabiliser is the set $\{g\in G : gsg^{-1}\}$, also called the centraliser.
+Then the orbit of an element $s$ is its conjugacy class and its stabiliser is its centraliser.
 Let $\overline H$ denote the set of all subgroups of $G$.
 Then $G$ also acts on $\overline H$ by conjugation: $g(H) = gHg^{-1}$.
-In this action, the stabiliser of a subgroup $H$ is the set
-$$ G_H = \{ g \in G : gHg^{-1} = H\}.$$
-Note that this is not the centraliser of $H$; elements may not be taken to themselves pointwise, so long as the sets $gHg^{-1}$ and $H$ are equal.
-This subset of $G$ is called the {\it normaliser} of $H$ and is itself a subgroup, denoted $N(H)$.
+In this action, the stabiliser of a subgroup $H$ is $N(H)$, its normaliser.
 

### 2019-12-04 misc/study-notes/groups.tex
Added LING 360 and more group theory

@@
 In some sense, the symmetric group is the most general group, because all other groups arise from adding restrictions to these bijections.
-For instance, $GL_n(\R) \subset \Sym(\R^n)$.
+For instance, $GL_n(\R) \subseteq \Sym(\R^n)$.
 
@@
 
-Let $G$ be a group. We call a nonempty subset $H\subset G$ a {\it subgroup} and write $H\leq G$ provided that
+Let $G$ be a group. We call a nonempty subset $H\subseteq G$ a {\it subgroup} and write $H\leq G$ provided that
 \medskip
@@
 If not, let $b$ be the smallest positive integer contained in $H$.
-We know from closure that every multiple of $b$ is in $H$, so $b\Z\subset H$.
+We know from closure that every multiple of $b$ is in $H$, so $b\Z\subseteq H$.
 Now let $h$ be any element in $H$.
@@
 But then necessarily we have $r=0$, since $b$ is the smallest positive integer in $H$.
-So $h$ is an integer multiple of $b$ and we have shown that $H \subset b\Z$.
+So $h$ is an integer multiple of $b$ and we have shown that $H \subseteq b\Z$.
 \slug
@@
 If $G$ is a group containing an element $g$ such that $\langle g\rangle = G$, then $G$ is called {\it cyclic} or {\it singly-generated}.
-More generally, if $S\subset G$ and we let $S^{-1}$ denote the set $\{s^{-1} : s\in S\}$, then the set
+More generally, if $S\subseteq G$ and we let $S^{-1}$ denote the set $\{s^{-1} : s\in S\}$, then the set
 $$\langle S\rangle = \{s_1s_2\cdots s_n : n\in\N, s_i\in S\cup S^{-1}\}$$
@@
 
-The following lemma describes the link between conjugacy and normal subgroups.
+Given a group $G$ and any set $S$ of elements from $G$, we can conjugate the entire set by an element:
+$$gSg^{-1} = \{gsg^{-1} : s\in S$$
+When the set $S$ is a subgroup, the conjugate set will also be a subgroup, but not necessarily the same subgroup $S$.
+It turns out that we can obtain an equivalent definition of normal subgroups in terms of conjugacy.
 
@@
 On the other hand, suppose $H$ is a subgroup such that, for any $g\in G$ and $h\in H$, $ghg^{-1}\in H$.
-Then $gHg^{-1}\subset H$ and we can conclude that $gH\subset Hg$.
-But $g^{-1}$ is also an element of $G$, so we can apply our hypothesis to get that $g^{-1}Hg\subset H$ and $Hg \subset gH$.
+Then $gHg^{-1}\subseteq H$ and we can conclude that $gH\subseteq Hg$.
+But $g^{-1}$ is also an element of $G$, so we can apply our hypothesis to get that $g^{-1}Hg\subseteq H$ and $Hg \subseteq gH$.
 So $gH=Hg$ for all $g\in G$ and $H$ is normal.\slug
@@
 It is easy to check that these are both subgroups of $H$.
-Note that if $S$ consists only of one element, then the centraliser and normaliser are equal and the inclusion $Z_H(S) \subset N_H(S)$ always holds.
-If $H = G$, then we may omit the subscript and simply call the subgroups the normaliser and centraliser of $S$.
+Note that if $S$ consists only of one element, then the centraliser and normaliser are equal and the inclusion $Z_H(S) \subseteq N_H(S)$ always holds.
+If $H = G$, we may omit the subscript and simply call the subgroups the normaliser and centraliser of $S$.
+It is clear that a subgroup $H$ of $G$ is normal if and only if $N(H) = G$.
+
+\proclaim Theorem N. Let $S$ be a set of elements of a group $G$. If $H\leq G$, the index in $H$ of the normaliser of $S$ in $H$, $[H:N_H(S)]$, is exactly the number of conjugates of $S$ under $H$.
+
+\proof Write $N = N_H(S)$ for short and suppose that $H$ is the disjoint union of cosets
+$$H = N \cup h_1N \cup \cdots \cup h_kN,$$
+where $k = [H:N_H(S)]$.
+Now suppose that $h_i$ and $h_j$ are elements of $H$ such that $h_iSh_i^{-1} = h_jSh_j^{-1}$.
+This is true if and only if $S = (h_j^{-1}h_i)S(h_j^{-1}h_i)^{-1}$.
+So $h_j^{-1}h_i$ is in $N_H(S) = N$ and $h_i\in h_jN$.
+So two conjugates of $S$ are the same if and only if the conjugating elements belong to the same left coset of $N$.
+Thus $[H:N_H(S)]$ is exactly the number of conjugates of $S$ under $H$.\slug
 
@@
 
-\section 2.6. The First Isomorphism Theorem
+\section 2.6. The Isomorphism Theorems
 
@@
 Since $f(k) = e_H$ for all $k\in K$, we have $f(sk) = f(s)f(k) = h = f(k)f(s) = f(ks)$.
-So $sK\subset X$ and $Ks\subset H$.
+So $sK\subseteq X$ and $Ks\subseteq H$.
 To prove the reverse inclusions, we let $x\in X$ (so $f(x) = h$).
@@
 
-\parenproclaim Theorem I (First isomorphism theorem).
+\parenproclaim Theorem I (First Isomorphism Theorem).
 Let $f:G\rightarrow H$ be a surjective group homomorphism.
 Then $K = \ker(f)$ is a normal subgroup of $G$ and there exists a group homomorphism $f':G/K \rightarrow H$.\slug
+
+The second and third isomorphism theorems give further results on the relationships between specific types of quotient groups.
 
@@
 
+\section ?.? Direct Products and Abelian Groups
+
+[Under construction.]
+
 \section ?.?. Group Actions on Sets
@@
 We find that this is exactly the conjugate subgroup $aHa^{-1}$.
-More generally, if $G$ acts transitively on a set $S$ and $g(s) = s'$, then $G_{s'} = gG_sg^{-1}\subset G$.
+More generally, if $G$ acts transitively on a set $S$ and $g(s) = s'$, then $G_{s'} = gG_sg^{-1}\subseteq G$.
 
@@
 
-\section ?.?. The Sylow Theorems
-
 Suppose that a group $G$ acts on itself by conjugation:
@@
 
+\section ?.?. The Sylow Theorems
+
+Lagrange's Theorem, presented in Section 2.4, states that if $G$ is a group with $|G|=n$, then the order of any element of $G$ divides $n$.
+The converse of this theorem does not hold: There does not necessarily exist a subgroup with size $m$ for all $m$ dividing $n$.
+For example, $A_4$ has no subgroup of order 6.
+However, the results of this section provide a partial converse to Lagrange's Theorem.
+To this end, we first introduce the concept of double cosets.
+
+Let $H$ and $K$ be subgroups, not necessarily distinct, of a group $G$. For a fixed element $g$ of $G$, the set
+$$HgK = \{hgk : h\in H\ \hbox{and}\ k\in K\}$$
+is called a {\it double coset}.
+
+First we show that, just like ordinary cosets, double cosets partition a group.
+First, every element $g$ of a group belongs to some double coset of $H$ and $K$, namely $HgK$.
+Then we have the following lemma:
+
+\proclaim Lemma D. Let $H$ and $K$ be subgroups of a group $G$. For elements $x,y\in G$, the cosets $HxK$ and $HyK$ are either equal or disjoint.
+
+\proof If $HxK$ and $HyK$ are not disjoint, then we can find $g\in HxK \cap HyK$.
+Suppose that $g = h_1xk_1 = h_2yk_2$; so $x = h_1^{-1}h_2yk_2k_1^{-1}$ and $y = h_2^{-1}h_1xk_1k_2^{-1}$.
+Then for any $hxk\in HxK$,
+$$hxk = hh_1^{-1}h_2yk_2k_1^{-1}k,$$
+meaning that $HxK\subseteq HyK$. Likewise,
+$$hyk = hh_2^{-1}h_1xk_1k_2^{-1}k$$
+and $HyK\subseteq HxK$.\slug
+
+A double coset $HgK$ contains left cosets of $K$ that are of the form $(hg)K$ as well as right cosets of $H$ of the form $H(gk)$.
+In fact, we can calculate the exact quantity of such left and right cosets.
+
+\proclaim Lemma Q. The number of right cosets of $H$ in $HgK$ is $[gKg^{-1} : H\cap gKg^{-1}]$ and the number of left cosets of $K$ in $HgK$ is $[H : H\cap gKg^{-1}]$.
+
+\proof We form a bijection $f:HgK\rightarrow HgKg^{-1}$ that takes an element $hgk$ to $hgkg^{-1}$.
+This creates a correspondence between left cosets $h(gKg^{-1})$ of $gKg^{-1}$ and left cosets $(hg)K$ of $K$ as well as a correspondence between right cosets $H(gkg^{-1})$ of $H$ and right cosets $H(gk)$ of $H$ in $HgK$.
+
+Let $D = (H\cap gKg^{-1}) \leq gKg^{-1}$. Then $gKg^{-1}$ can be written as a union of disjoint right cosets
+$$gKg^{-1} = D \cup Dx_2 \cup \cdots \cup Dx_n,$$
+where $n = [gKg^{-1} : D]$.
+Each $x_i$ belongs to $gKg^{-1}$ and the claim is that $H, Hx_2, \ldots, Hx_n$ are exactly the right cosets of $H$ that are in $HgK$.
+These cosets are distinct since if $Hx_i = Hx_j$, then $x_ix_j^{-1}\in H$, but since both $x_i$ and $x_j$ are in $gKg^{-1}$, this means that $x_ix_j^{-1}\in D$ and $Dx_i = Dx_j$, contradicting our choice of coset representatives.
+Now every right coset of $H$ in $HgKg^{-1}$ is of the form $Hx$, where $x = dx_i$ for some $d\in D$.
+But since $D\subseteq H$, $Hx = Hdx_i = Hx_i$.
+So the number of right cosets of $H$ in $HgKg^{-1}$ is $[gKg^{-1} :  D] = [gKg^{-1} : H\cap gKg^{-1}]$ and by the bijection above, this is the number of right cosets of $H$ in $HgK$.
+
+In a similar manner, it can be shown that the number of left cosets of $gKg^{-1}$ in $HgKg^{-1}$ is $[H:D] = [H:H\cap gKg^{-1}]$ and this is exactly the number of left cosets of $K$ in $HgK$.\slug
+
+So much for double cosets.
+We saw earlier that if $n$ is the order of a group $G$ and $d$ divides $n$, there does not necessarily exist a subgroup of order $d$.
+However, if $d$ is prime or a power of a prime, then there is such a subgroup.
+We begin with a useful theorem, due to Cauchy.
+
+\parenproclaim Theorem C (Cauchy). If $G$ is a group that has order dividing a prime $p$, then $G$ contains an element of order $p$.
+
+\proof Let $G$ be a group with order $mp$, where $p$ is prime. The proof is by strong induction on $m$. If $m=1$, then $G$ is cyclic, generated by an element of order $p$.
+
+Now suppose that the theorem holds for all smaller multiples of $p$.
+Then if $G$ contains a proper subgroup $H$ whose index $[G:H]$ is not divisible by $p$, then the order of $H$ is divisible by $P$ and, by induction, $H$ contains an element of order $p$.
+In the case that all proper subgroups of $G$ have indices divisible by $p$, we split $G$ into a union of disjoint conjugacy classes:
+$$G = \{e\} \cup C_2 \cup \cdots \cup C_k$$
+Suppose that each conjugacy class $C_i$ has order $n_i\geq 1$.
+Let $x_i$ be a representative from each $C_i$.
+Then $n = n_1 + \cdots + n_k$, where, by Theorem 1.4.N, $n_i = [G:Z(\{s_i\})]$.
+If $n_i \neq 1$, then $n_i$ is the index of a proper subgroup of $G$, and by hypothesis is divisible by $p$.
+Surely $n_1 = 1$, so the number of distinct $i$ for which $n_i = 1$ divides $p$.
+But $n_i = 1$ if and only if $s_i\in Z(G)$, so the centre of $G$ has order dividing $p$.
+By the Fundamental Theorem of Finite Abelian Groups, this implies that it contains an element of order $p$.\slug
+
+Cauchy's Theorem tells us that if $p$ divides the order of $G$, at least one subgroup of $G$ has order $p$. The first of Sylow's theorems proves a stronger result.
+
+[Sylow's Theorems.]
+
 \section REFERENCES

### 2019-12-04 misc/study-notes/ling360notes.tex
Added LING 360 and more group theory

% LING 360 Notes
% Notes by Marcel Goh

\input fontmac
\input mathmac

% Useful functions

\def\Sym{\mathop{\rm Sym}\nolimits}
\def\Aut{\mathop{\rm Aut}\nolimits}
\def\Out{\mathop{\rm Out}\nolimits}
\def\Inn{\mathop{\rm Inn}\nolimits}
\def\Im{\mathop{\rm Im}\nolimits}
\def\Id{\mathop{\rm Id}\nolimits}
\def\lexent#1#2{{\it #1}|{\rm #2}}

\leftrighttop{LING 360 INTRODUCTION TO SEMANTICS}{MARCEL GOH}

\maketitle{LING 360 Introduction to Semantics\footnote{$^*$}{\eightpt Course given by Prof.\ Brendan S.\ Gillon at McGill University. These notes are heavily based on the textbook.}}{Notes by}{Marcel K. Goh {\rm (Montr\'eal, Qu\'ebec)}}{1 December 2019}

\section 1. INTRODUCTION

This set of notes assumes that the reader has knowledge of basic set theory, logic, and English grammar. When those topics come up, only course-specific terms and concepts will be included.

\section 1.1. Linguistics and Psychology

The study of language has existed in various forms since prehistory. Early on, the study of grammar was mostly {\it prescriptivist}, establishing and defending an idealised ``correct'' version of language. It was only in the last two centuries a modern view of linguistic study began to emerge. At the beginning of the twentieth century, Ferdinand de Saussure noted that the study of language can be undertaken in two distinct ways. One can study language at a certain point in time ({\it synchronic}), or over a period of time ({\it diachronic}).

The connection between language and psychology has been evident since Wilhelm Wundt established psychology as an empirical science. (Previously, psychology was a subfield of philosophy.) It was Leonard Bloomfield who first saw linguistics as an independent branch of psychology and even today, most linguists view their field as a domain of psychology. In the early twentieth century, psychology was under the influence of {\it behaviourism}, initiated by the psychologist John B. Watson. Behaviourism rejects all appeal to unobservable entities (instinct in animals, the mind in humans) and focuses on measurable data. Behaviourists study an organism's stimuli and responses.

By the end of World War Two, behaviourism was falling out of fashion and it was once again acceptable to scientifically study instinct. This led to the rise of {\it ethology}, the study of animal behaviour that grew out of the work of zoologists like Nikolaas Tinbergen. It attempts to determine what behaviours are innate (instictive) and which are acquired (learned). Ethologists are especially interested in what is known as a {\it fixed action pattern}, in which a determinate sequence of actions, once initiated, goes to completion. One way to find out if this sequence of actions is innate or acquired is to perform a {\it deprivation expermient}. A young animal is raised in isolation from other animals of its species and one studies its behaviour. Those that are present are deemed to be innate. One can also figure out what experience is required to cause learned behaviours to emerge.

Human language is different from animal communicative behaviour in that it is both {\it discrete} and {\it infinite}. Both types of communication are present in the animal kingdom. An example of discrete communication is the vervet monkey, which has three distinct vocal signals: one to signal a leopard, one for a python, and a third for an eagle. An example of infinite, but continuous communicaton, is the honeybee, which can do a waggle dance. The axis of the dance indicates the direction of nectar and the rate indicates the distance. Human communication is also able to express an infinite amount of ideas, but it is made up of discrete components. Infinity is achieved via recursion.

\section 1.2. Chomsky and the Poverty of the Stimulus Argument

Unlike his predecessor Bloomfield, Noam Chomsky was a critic of behaviourism. His criticism of B. F. Skinner in the late 1950s was an influential text in linguistics. Chomsky hypothesised that human language relies on an innate {\it language acquisition device} that is unique to our species. To support his theory, he exposes the poverty of the stimulus argument with a number of observations:
\medskip
\item{i)} The structure of language is complex and abstract from its actual acoustic signal. While the sounds may be linear, the actual structure of sentences is tree-like.
\smallskip
\item{ii)} A child acquires grammatical competence in a very short span of time. This means that some of the capacity to speak and understand language must be innate.
\smallskip
\item{iii)} A child can learn grammar perfectly even if he/she has little exposure to every single relevant structure and even when the utterances are defective.
\smallskip
\item{iv)} Even though the utterances that each individual child hears can be very different, children of the same language converge upon the same grammatical competence.
\smallskip
\item{v)} The grammatical rules that people actually use to speak are not taught. For example, speakers of English learn how to form plurals long before the rule is taught (imperfectly) in school. Children will append a [z] to the word ``dog'', [s] to the word ``cat'', and [iz] to the word ``horse'', even though no parent would actually explicitly teach their child that these are the rules.
\smallskip
\item{vi)} A child's acquisition of language is independent of most other developmental factors, including intelligence, motivation, and emotional disposition.
\smallskip
\item{vii)} Children are not genetically predisposed to learn any one language over another. A child born to speakers of Mandarin will learn English perfectly if raised by fluent English speakers.
\medskip
In addition to these observations, it appears there is a {\it critical period} during which a child can acquire language. It is unethical to perform a deprivation experiment on a human infant, but there have been a number of cases throughout history that support this idea. Children raised without being exposed to language end up never learning to speak properly, even with extensive training as an adult.

\section 1.3. Grammar and Structure

Ovee 2500 years ago, unknown thinkers in Indo-Aryan tribes in what is now Pakistan and northwest India had a deep interest in their language, Sanskrit. They formulated what is now known as a {\it generative grammar}. Either written or compiled by someone named P\=a\d nini, the {\sl A\d s\d t\=adhy\=ay\=\i} comprised a finite set of rules and a finite set of minimal expressions from which each and every expression of Sanskrit could be derived in a finite number of steps. It was not until the middle of the twentieth century that generative grammars of other languages were formulated by linguists.

The key insight in the development of a grammar is that the set of grammatical expressions is infinite and thus impossible to learn one-by-one. Structuralist linguists such as Leonard Bloomfield generalised the work of P\=a\d nini, observing that complex expressions are made out of smaller expressions and so on, and the meaning of these smaller expressions contributes to the meaning of the larger expression. Central to structuralism is the notion of {\it immediate constituency analysis}:
\medskip
\item{i)} Each complex expression can be analysed into immediate subexpressions, typically two, which can themselves be analysed in the same manner until the minimal constituents are reached.
\smallskip
\item{ii)} Each expression can be put into a set of expressions that can be substituted one for another without affecting acceptability.
\smallskip
\item{iii)} Each of these sets of expressions can be assigned a syntactic category.
\medskip
Recursion, the property whereby a constituent of a certain type can contain another constituent of the same type is a property of every known human language. It is the means by which one can generate an infinite number of expressions from a finite set of rules. Two types of recursive specification are usually employed by linguists: {\it constituency grammar} and {\it categorial grammar}.

Constituency grammar arose out of the study of language itself, having its origins in P\=a\d nini's {\sl A\d s\d t\=adhy\=ay\=\i}. Bloomfield studied P\=a\d nini's grammar and applied his techniques not only to phonology and morphology, but also to syntax. His work was extended by others, including Zellig Harris, Charles Hockett, and Rulon Wells.

Categorial grammar, on the other hand, arose out of the study of logic. It was devised by Kazimierz Ajdukiewicz to mathematically characterise the notation used in classical quantificational logic. He anecdotally pointed out some of its connections to natural language structure, but his main concern was with logic. Yehoshua Bar-Hillel was the first person to apply notions of categorial grammar to natural language. Later on, this work was generalised by the mathematical Joachim Lambek in his {\it Lambek calculus}. The study of how semantic values are assigned to syntactic structure is based on a subdiscipline of logic known as {\it model theory}. Its founder, Alfred Tarski, noted how model theory might be pertinent to studying how the meaning of an expression can described in terms of the meanings of constituent expressions, though he did not formalise such an application himself.

When studying grammar at an abstract level, we deal with idealisations of natural language. In particular, it is important to distinguish an individual's {\it competence} from his/her {\it performance}. Indeed, a person may have perfect grammatical competence despite mispronunciations, hesitation, stuttering, etc.

\section 1.4. Hypothesis Testing

When dealing with empirical evidence, such as linguistic data, a key method we use is Mill's method of agreement and difference, formulated by John Stuart Mill in 1843. The {\it method of agreement} states that if multiple instances of a phenomenon have only one circumstance in common, then that circumstance must be the effect or cause of the phenomenon. Similarly, the {\it method of difference} states that if an instance in which a phenomenon occurs and one in which it does not occur have every circumstance in common except one, then that circumstance must be the effect or cause of the phenomenon. An example of the method of difference is the {\it minimal pair}. For example, consider the sentences
$$\hbox{``Whales eat plankton.''}\quad\hbox{and}\quad \hbox{``Asteroids eat plankton.''}$$
They differ only in the first word and have different meanings, so we conclude that the words ``whales'' and ``asteroids'' have different semantic value.

\section 2. BASIC SET THEORY

Most of the concepts in this chapter will be familiar to the mathematically-inclined reader. Only the more obscure or hard-to-remember definitions are listed here.

Let $R = \langle X,Y,G\rangle$ be a binary relation. It is {\it left total} if for each $x\in X$, there exists $y\in Y$ such that $xRy$. The relation is {\it left monogamous} if for every $x\in X$, if $y,z\in Y$ are such that $xRy$ and $xRz$, then $y=z$. The definitions of right totality and right monogamy are analogous. A function is a relation that is left total and left monogamous.

Let $X$ be a set with a subset $Y\subseteq X$. The {\it characteristic function} of $Y$ is the map $f:X\rightarrow\{0,1\}$ given by
$$f(x) = \cases{1 & if  $x\in Y$;\cr 0 & if $x \notin Y$.}$$

A function $f$ is a {\it near-variant} of a function $g$ if their domains and codomains are equal and if the domain contains at most one element $x$ such that $f(x) \neq g(x)$. A consequence of the definition is that every function is a near-variant of itself.

\section 3. BASIC ENGLISH GRAMMAR

We gloss over preliminaries regarding English grammar and parts of speech and discuss constituency grammars.

Let $L$ be a language. Then $G = \langle {\rm BX}, {\rm CT}, {\rm LX}, {\rm FR}\rangle$ is a (synthesis) {\it constituency grammar} of $L$ if and only if
\medskip
\item{i)} ${\rm BX}$ is a nonempty, finite set of the basic expressions of $L$.
\smallskip
\item{ii)} ${\rm CT}$ is a nonempty, finite set of the categories of $L$.
\smallskip
\item{iii)} ${\rm LX}$ is a nonempty, finite set of ordered pairs $\lexent{v}{{\it C}}$, where $v\in {\rm BX}$ and $C\in {\rm CT}$. These are the lexical entries of $L$.
\smallskip
\item{iv)} ${\rm FR}$ is a nonempty, finite set of constituency formation rules of $L$, where each rule has the form $C_1\ldots C_n \rightarrow C$ and $C,C_1,\ldots,C_n$ are in ${\rm CT}$.
\medskip
When an expression accomodates more than one constituency analysis, it is said to be {\it structurally ambiguous} or {\it amphibolous}. For example, the sentence ``Alice met a friend of Bob's mother'' is amphibolous.

\section 4. LANGUAGE AND CONTEXT

An expression whose full understanding requires knowledge of the setting is called an {\it exophor}. Examples include the pronouns ``I'' and ``you''. One needs to know who is speaking and who is the addressee to fully understand the meaning that the pronouns convey. English exophoric expressions can often require knowledge of person, temporal order, or spatial location.

An expression whose understanding relies on other expressions uttered before or after it (the cotext) is called an {\it endophor} or a {\it proform}. Third-person pronouns are a good example of this, but demonstrative pronouns like ``this'' and ``that'' are also endophors.

\section REFERENCES
\frenchspacing

\item{} B. S. Gillon, {\sl Natural Language Semantics: Formation and Valuation}, 2019, MIT Press.

\bye

### 2019-12-11 misc/study-notes/groups.tex
Added more isomorphism theorems

@@
 \def\Sym{\mathop{\rm Sym}\nolimits}
+\def\Stab{\mathop{\rm Stab}\nolimits}
 \def\Aut{\mathop{\rm Aut}\nolimits}
@@
 
-\section 1. PRELIMINARY NOTIONS FROM LINEAR ALGEBRA
+\section 1. INTRODUCTION
 
-This set of notes assumes that the reader has had some exposure to linear algebra; the most crucial notions will be outlined in this section.
+This set of notes assumes that the reader has had some exposure to linear algebra; the most crucial notions are briefly outlined here.
+Next, the group axioms are introduced and some basic properties of groups are given.
 
@@
 
-\section 2. GROUPS
-
-\vskip -\medskipamount
-
-\section 2.1. Properties and Basic Examples
+\section 1.3. Groups
 
@@
 
-\section 2.2. Permutation Groups
+\section 1.4. Permutation Groups
 
@@
 
-\section 2.3. Subgroups
+\section 1.5. Subgroups
 
@@
 
-\section 2.4. Cosets and Normal Subgroups
+\section 2. NORMAL SUBGROUPS AND HOMOMORPHISMS
+
+\vskip -\medskipamount
+
+\section 2.1. Cosets and Normal Subgroups
 
@@
 
-\section 2.5. Isomorphisms and Homomorphisms
+\section 2.2. Simplicity of the Alternating Group
+
+We take a brief detour to prove a famous result in the classification of groups, using only the basic techniques we have learned so far.
+A group $G$ is {\it simple} if its only normal subgroups are the trivial group $\{e\}$ and the whole group $G$.
+In this section we show that for $n\geq 5$, the alternating group $A_n$ is simple.
+First we establish that the set of 3-cycles generate $A_n$.
+
+\proclaim Lemma G. Every permutation in $A_n$ can be written as a product of 3-cycles.
+
+\proof Since every permutation in $A_n$ can be written as a product of an even number of transpositions, it suffices to show that any product $(a\,b)(c\,d)$ can be rewritten as a product of 3-cycles.
+There are three cases to consider.
+Firstly, if the transpositions have two numbers in common, then $(a\,b) = (c\,d)$ and the product is simply the identity permutation.
+If the transpositions have a single point in common, then the product is a 3-cycle, since $(a\,b)(b\,c) = (a\,b\,c)$.
+Lastly, in the case that all four points are distinct, we have
+$$(a\,b)(c\,d) = (a\,b\,c)(b\,c\,d),$$
+a product of 3-cycles.
+Hence the set of 3-cycles generates $A_n$.\slug
+
+In fact, we can shrink the generating set to 3-cycles of a specific form.
+
+\proclaim Lemma H. The $n-2$ permutations $(1\,2\,3), (1\,2\,4),\ldots,(1\,2\,n)$ generate $A_n$.
+
+\proof By Lemma G, we need only show that any 3-cycle can be rewritten as a product of cycles $(1\,2\,k)$.
+Since the square of $(1\,2\,k)$ is $(2\,1\,k)$, any cycle that contains both 1 and 2 can be generated by the permutations above.
+Then, note that any cycle $(a\,b\,c)$ that fixes 1 can be rewritten $(1\,a\,b)(1\,b\,c)$.
+Now we can rewrite each of these cycles as follows:
+$$(1\,j\,k) = (1\,2\,k)(1\,2\,k)(1\,2\,j)(1\,2\,k)$$
+So every 3-cycle can be expressed as a product of cycles of the form $(1\,2\,k)$.\slug
+
+This result can be used to show that any normal subgroup of $A_n$ that contains a 3-cycles is necessarily the whole group $A_n$.
+
+\proclaim Lemma K. Let $n\geq 4$ and let $N$ be a normal subgroup of the group $A_n$. If $N$ contains a 3-cycle, then $N = A_n$.
+
+\proof Without loss of generality, suppose that $N$ contains the cycle $(1\,2\,3)$. Then it also contains its inverse $(2\,1\,3)$ and for any $\sigma\in A_n$,
+$$\sigma (2\,1\,3) \sigma^{-1}\in N.$$
+So let $\sigma = (1\,2)(3\,k)$ for $k\geq 4$ and we have
+$$\sigma (2\,1\,3) \sigma^{-1} = (1\,2\,k).$$
+By Lemma H, these generate $A_n$, so $N$ is the entire group.\slug
+
+We may finally prove that $A_n$ is simple for $n\geq 5$. The following theorem is due to E. Galois.
+
+\proclaim Theorem S. Let $n\geq 5$ and let $N\geq A_n$ be non-trivial. Then $N=A_n$.
+
+\proof Let $\tau\in N$ be a non-identity permutation that fixes as many symbols as possible. We will show, by contradiction, that $\tau$ must be a 3-cycle, i.e.\ it fixes all symbols except three.
+
+Suppose that $\tau$ moves more than three symbols.
+Then either $\tau$ has a cycle with more than three symbols, or $\tau$ consists of disjoint transpositions.
+Without losing generality, we use the symbols 1, 2, 3, 4, and 5.
+In the first case, we have
+$$\tau = (1\,2\,3\,\ldots)\ldots$$
+and we can conjugate by $(3\,4\,5)$ to get
+$$\tau' = (3\,4\,5)\tau(3\,4\,5)^{-1} = (1\,2\,4\,\ldots)\ldots.$$
+Then $\tau^{-1}\tau'$ fixes 1 where $\tau$ did not, a contradiction.
+
+In the case that $\tau$ only contains transpositions, we have
+$$\tau = (1\,2)(3\,4)\ldots,$$
+and again we conjugate by $(3\,4\,5)$ to get
+$$\tau' = (3\,4\,5)\tau(3\,4\,5)^{-1} = (1\,2)(4\,5)\ldots)\ldots.$$
+Then $\tau^{-1}\tau'$ fixes 1 and 2 and $\tau$ does not, leading to a contradiction.
+
+The contradiction in both cases implies that $\tau$ moves three symbols or less.
+In the alternating group, this implies that $\tau$ is a 3-cycle.
+So $N$ contains a 3-cycle and by Lemma K, $N=A_n$.\slug
+
+With this theorem in hand, it becomes easy to check that $A_4$ is the only alternating group that possesses a normal subgroup.
+
+\section 2.3. Isomorphisms and Homomorphisms
 
@@
 
-\section 2.6. The Isomorphism Theorems
+\section 2.4. The Isomorphism Theorems
 
@@
 So the group structure of $G/K$ depends only on the kernel $K$ and not the homomorphism that induced it.
-The results of this section are summed up in the following theorem.
+The following theorem sums up what we have discovered so far.
 
@@
 Let $f:G\rightarrow H$ be a surjective group homomorphism.
-Then $K = \ker(f)$ is a normal subgroup of $G$ and there exists a group homomorphism $f':G/K \rightarrow H$.\slug
+Then $K = \ker(f)$ is a normal subgroup of $G$ and there exists a group isomorphism $f':G/K \rightarrow H$.\slug
 
-The second and third isomorphism theorems give further results on the relationships between specific types of quotient groups.
+This result is fundamental and is often used in conjunction with other laws of homomorphisms to establish isomorphisms.
+The other two isomorphism theorems provide further insight into the properties of group homomorphisms.
 
-\section 2.7. Inner and Outer Automorphisms
+\parenproclaim Theorem J (Second Isomorphism Theorem).
+Let $G$ be a group with subgroups $A$ and $B$.
+Suppose that $A\leq N_G(B)$.
+Then $AB$ is a group and $B$ is a normal subgroup of $AB$.
+Then there exists a surjective homomorphism $f:A\rightarrow AB/B$ with kernel $A\cap B$, given by $f(a) = aB$.
+
+\proof First we show that $AB$ is a group.
+Since $A\leq N_G(B)$, for any product $a_1b_1a_2b_2$, we can find $b'\in B$ such that $a_2b' = b_1a_2$.
+Then $a_1b_1a_2b_2 = a_1a_2b'b_2\in AB$.
+Similarly, $AB$ is stable under inverses, since for any product $ab$, we can find $b'\in B$ such that $(ab)^{-1} = b^{-1}a^{-1} = a^{-1}b'$.
+Because both $A$ and $B$ are subgroups of $N_G(B)$, so is $AB$ and in particular, $B$ is a normal subgroup of $AB$.
+Then $f:A\rightarrow AB/B$ is the restriction of the quotient map $AB\rightarrow AB/B$ to $A$, so it is surjective.
+
+Now let $a\in \ker(f)$.
+By definition, $f(a) = aB = B$, so $a\in B$.
+Thus $a\in A\cap B$.
+Now let $a\in A\cap B$.
+Then $aB = B$ so $a\in \ker(f)$.
+Thus $\ker(f) = A\cap B$.
+In fact, we can now apply the first isomorphism theorem to deduce that $A/\ker(f) \cong AB/B$.
+\slug
+
+\parenproclaim Theorem K (Third Isomorphism Theorem).
+Suppose that $A\leq B\leq G$ is a chain of subgroups and furthermore, both $A$ and $B$ are normal subgroups of $G$.
+Then the kernel of the surjective homomorphism $f:G/A\rightarrow G/B$ is $B/A$.
+
+\proof The homomorphism $f$ is well defined, since if $sA = tA$, then $t^{-1}s\in A$, implying that $t^{-1}s\in B$ and $sB = tB$.
+It is also surjective, since for any $sB\in G/B$, we have $sA\in G/A$ and $f(sA) = sB$.
+
+Now let $sA\in \ker(f)$.
+We have $f(sA) = sB = B$, so $s\in B$.
+This means that $sA\in B/A$.
+Conversely, if $s\in B$ then $f(sA) = sB = B$.
+So $\ker(f) = B/A$ and by the first isomorphism theorem, we can express this as the isomorphism $(G/A)/(B/A)\cong G/B$.\slug
+
+[Include correspondence theorem.]
+
+\section 3. AUTOMORPHISMS
 
@@
 Given a group $G$, we can construct a set $\Aut(G)$ of all automorphisms of $G$ and in fact, it is easily verifiable that $\Aut(G)$ is a group under function composition.
+
+\section 3.1. Inner and Outer Automorphisms
 
@@
 
-\proof Let $f_a\in Inn(G)$ be the map that conjugates by an element $a\in G$. Let $g\in \Aut(G)$ be arbitrary. Then
+\proof Let $f_a\in \Inn(G)$ be the map that conjugates by an element $a\in G$. Let $g\in \Aut(G)$ be arbitrary. Then
 $$g\circ f_a\circ g^{-1} = g(a\cdot g^{-1}(s)\cdot a^{-1}) = g(a)\cdot s\cdot g(a)^{-1},$$
-and we see that $g\circ f_a \circ g^{-1} = f_{g(a)}\in Inn(G)$. So $\Inn(G)$ is stable under conjugation in $Aut(G)$.\slug
+and we see that $g\circ f_a \circ g^{-1} = f_{g(a)}\in \Inn(G)$. So $\Inn(G)$ is stable under conjugation in $\Aut(G)$.\slug
 
@@
 
-\section ?.?. Simplicity of the Alternating Group
-
-A group $G$ is {\it simple} if its only normal subgroups are the trivial group $\{e\}$ and the whole group $G$.
-In this section we show that for $n\geq 5$, the alternating group $A_n$ is simple.
-First we establish that the set of 3-cycles generate $A_n$.
-
-\proclaim Lemma G. Every permutation in $A_n$ can be written as a product of 3-cycles.
-
-\proof Since every permutation in $A_n$ can be written as a product of an even number of transpositions, it suffices to show that any product $(a\,b)(c\,d)$ can be rewritten as a product of 3-cycles.
-There are three cases to consider.
-Firstly, if the transpositions have two numbers in common, then $(a\,b) = (c\,d)$ and the product is simply the identity permutation.
-If the transpositions have a single point in common, then the product is a 3-cycle, since $(a\,b)(b\,c) = (a\,b\,c)$.
-Lastly, in the case that all four points are distinct, we have
-$$(a\,b)(c\,d) = (a\,b\,c)(b\,c\,d),$$
-a product of 3-cycles.
-Hence the set of 3-cycles generates $A_n$.\slug
-
-In fact, we can shrink the generating set to 3-cycles of a specific form.
-
-\proclaim Lemma H. The $n-2$ permutations $(1\,2\,3), (1\,2\,4),\ldots,(1\,2\,n)$ generate $A_n$.
-
-\proof By Lemma G, we need only show that any 3-cycle can be rewritten as a product of cycles $(1\,2\,k)$.
-Since the square of $(1\,2\,k)$ is $(2\,1\,k)$, any cycle that contains both 1 and 2 can be generated by the permutations above.
-Then, note that any cycle $(a\,b\,c)$ that fixes 1 can be rewritten $(1\,a\,b)(1\,b\,c)$.
-Now we can rewrite each of these cycles as follows:
-$$(1\,j\,k) = (1\,2\,k)(1\,2\,k)(1\,2\,j)(1\,2\,k)$$
-So every 3-cycle can be expressed as a product of cycles of the form $(1\,2\,k)$.\slug
-
-This result can be used to show that any normal subgroup of $A_n$ that contains a 3-cycles is necessarily the whole group $A_n$.
-
-\proclaim Lemma K. Let $n\geq 4$ and let $N$ be a normal subgroup of the group $A_n$. If $N$ contains a 3-cycle, then $N = A_n$.
-
-\proof Without loss of generality, suppose that $N$ contains the cycle $(1\,2\,3)$. Then it also contains its inverse $(2\,1\,3)$ and for any $\sigma\in A_n$,
-$$\sigma (2\,1\,3) \sigma^{-1}\in N.$$
-So let $\sigma = (1\,2)(3\,k)$ for $k\geq 4$ and we have
-$$\sigma (2\,1\,3) \sigma^{-1} = (1\,2\,k).$$
-By Lemma H, these generate $A_n$, so $N$ is the entire group.\slug
-
-We may finally prove that $A_n$ is simple for $n\geq 5$. The following theorem is due to E. Galois.
-
-\proclaim Theorem S. Let $n\geq 5$ and let $N\geq A_n$ be non-trivial. Then $N=A_n$.
-
-\proof Let $\tau\in N$ be a non-identity permutation that fixes as many symbols as possible. We will show, by contradiction, that $\tau$ must be a 3-cycle, i.e.\ it fixes all symbols except three.
-
-Suppose that $\tau$ moves more than three symbols.
-Then either $\tau$ has a cycle with more than three symbols, or $\tau$ consists of disjoint transpositions.
-Without losing generality, we use the symbols 1, 2, 3, 4, and 5.
-In the first case, we have
-$$\tau = (1\,2\,3\,\ldots)\ldots$$
-and we can conjugate by $(3\,4\,5)$ to get
-$$\tau' = (3\,4\,5)\tau(3\,4\,5)^{-1} = (1\,2\,4\,\ldots)\ldots.$$
-Then $\tau^{-1}\tau'$ fixes 1 where $\tau$ did not, a contradiction.
-
-In the case that $\tau$ only contains transpositions, we have
-$$\tau = (1\,2)(3\,4)\ldots,$$
-and again we conjugate by $(3\,4\,5)$ to get
-$$\tau' = (3\,4\,5)\tau(3\,4\,5)^{-1} = (1\,2)(4\,5)\ldots)\ldots.$$
-Then $\tau^{-1}\tau'$ fixes 1 and 2 and $\tau$ does not, leading to a contradiction.
-
-The contradiction in both cases implies that $\tau$ moves three symbols or less.
-In the alternating group, this implies that $\tau$ is a 3-cycle.
-So $N$ contains a 3-cycle and by Lemma K, $N=A_n$.\slug
-
-\section ?.? Direct Products and Abelian Groups
+\section ?.? Structure of Abelian Groups
 
@@
 Let $G$ be a group and $S$ a set of elements.
-We say that $G$ {\it acts on} $S$ if there exists a map from $G\times S\rightarrow S$ (whose pairs $(g,s)$ are written $g(s)$ or $gs$) such that, for all $s\in S$, $e(s) = s$ and $(gh)(s) = g\big(h(s)\big)$ for every $g,h\in G$.
+We say that $G$ {\it acts on} $S$ if there exists a map from $G\times S\rightarrow S$ (whose pairs $(g,s)$ are written $g(s)$ or $gs$) such that, for all $s\in S$, $es = s$ and $(gh)s = g\big(hs\big)$ for every $g,h\in G$.
 For any element $s\in S$, the {\it orbit} is set of elements in $S$ to which $s$ may be taken by $G$:
-$$O_s = \{g(s)\in S : g\in G\}$$
-If there exists an $s$ such that $O_s = S$, then we say that $G$ acts {\it transitively} on $S$.
-In fact, because the orbits of a group action partition the set $S$ into equivalence classes, transitivity of a group action implies there is only one orbit, and for any $s,s'\in S$, there exists some $g\in G$ such that $g(s) = s'$.
+$$Gs = \{gs\in S : g\in G\}$$
+If there exists an $s$ such that $Gs = S$, then we say that $G$ acts {\it transitively} on $S$.
+In fact, because the orbits of a group action partition the set $S$ into equivalence classes, transitivity of a group action implies there is only one orbit, and for any $s,s'\in S$, there exists some $g\in G$ such that $gs = s'$.
 
 The {\it stabiliser} of $s$ is the set of all elements in $G$ that leave $s$ fixed:
-$$G_s = \{g\in G : g(s) = s\}$$
+$$\Stab(s) = \{g\in G : gs = s\}$$
 
@@
 We find that this is exactly the conjugate subgroup $aHa^{-1}$.
-More generally, if $G$ acts transitively on a set $S$ and $g(s) = s'$, then $G_{s'} = gG_sg^{-1}\subseteq G$.
+More generally, if $G$ acts transitively on a set $S$ and $gs = s'$, then $\Stab(s') = g\Stab(s)g^{-1}\subseteq G$.
 
@@
 
-[Sylow's Theorems.]
+\parenproclaim Theorem S (First Sylow Theorem).
 

### 2019-12-29 misc/solutions/baby-sols.tex
Added textbook solutions

\input fontmac
\input mathmac

\font\bigboldsl=cmbxsl12
\font\eightsl=cmsl8

\maketitlenodate{Answers to Selected Exercises in {\bigboldsl Principles of Mathematical Analysis}\footnote{$^*$}{\frenchspacing \eightpt Walter Rudin. 1986. {\eightsl Principles of Mathematical Analysis, 3rd ed}. McGraw-Hill, Inc., USA.}}{Solutions by}{Marcel K.\ Goh}

\section CHAPTER 2. BASIC TOPOLOGY

\proclaim 1. Prove that the empty set is a subset of every set.

\proof Let $S$ be an arbitrary set. Then every element of $\emptyset$ is an element of $S$. So $\emptyset \subseteq S$.\slug

\proclaim 2. A complex number $z$ is said to be algebraic if there are integers $a_0,\ldots,a_n$, not all zero, such that
$$a_0z^n + a_1z^{n-1} + \cdots + a_{n-1}z + a_n = 0.$$
Prove that the set of all algebraic numbers is countable. {\it Hint:} For every positive integer $N$, there are only finitely many equations with
$$n + |a_0| + |a_1| + \cdots + |a_n| = N$$

\proof Let $A$ denote the set of all algebraic numbers and partition $A$ as follows: For each $z\in A$, calculate the positive integer $N$ that corresponds to its equation and place it in a set $E_N$. So
$$A = \bigcup_{N\in \N} E_N,$$
where each $E_N$ is finite. Then apply the corollary of Theorem 2.12 to find that $A$ is countable.\slug

\proclaim 3. Prove that there exist real numbers which are not algebraic.

\proof Let $A$ denote the set of all algebraic numbers and suppose, towards a contradiction, that all real numbers are algebraic. Then ${\bf R}\subset A$. But the set of real numbers is uncountable and we know from Problem 2 that $A$ is countable. The contradiction completes the proof.\slug

\proclaim 4. Is the set of all irrational real numbers countable?

The answer is no.

\proof If $\R \setminus \Q$ is countable, then $\R = \Q \cup (\R\setminus\Q)$, is countable, which we know to be false.\slug

\proclaim 5. Construct a bounded set of real numbers with exactly three limit points.

$\{{1\over n} : n\in \N\} \cup\{{1\over n} + 2: n\in \N\} \cup \{{1\over n} + 4: n\in \N\}$ has limit points 0, 2, and 4. It is bounded above by 6 and below by 0.

\bye

### 2019-12-29 misc/solutions/papa-sols.tex
Added textbook solutions

\input fontmac
\input mathmac

\font\bigboldsl=cmbxsl12
\font\eightsl=cmsl8

\maketitlenodate{Answers to Selected Exercises in {\bigboldsl Real and Complex Analysis}\footnote{$^*$}{\frenchspacing \eightpt Walter Rudin. 1987. {\eightsl Real and complex analysis, 3rd ed}. McGraw-Hill, Inc., USA.}}{Solutions by}{Marcel K.\ Goh}

\section CHAPTER 1. ABSTRACT INTEGRATION

\proclaim 1. Does there exist an infinite $\sigma$-algebra which has only countably many members?

The answer is no.

\proof Let $X$ be a ground set and let ${\cal F}$ be a $\sigma$-algebra (over $X$) with infinitely many members. First we claim that we can always find a set $E\neq \emptyset$ such that the set $\{ F \cap E^c : F\in {\cal F}\}$ is infinite. If this did not hold, then take any $E\neq \emptyset$ in ${\cal F}$. Then by assumption, ${\cal S}_1 = \{ F \cap E^c : F\in {\cal F}\}$ is finite and because $E^c\in {\cal F}$, ${\cal S}_2 = \{ F \cap E : F\in {\cal F}\}$ is finite as well. Since any member of ${\cal F}$ can be expressed as a union of an element of ${\cal S}_1$ with an element of ${\cal S}_2$, the implies that ${\cal F}$ is finite, a contradiction.

Now we may use the claim to construct a countable sequence of pairwise disjoint elements of ${\cal F}$. Let $G_1$ be the set $E$ given by the claim. Now the infinite set ${\cal S}_1$ we constructed before is also a $\sigma$-algebra, so repeat the argument to get a set $G_2$, disjoint from $G_1$. Continuing in this manner, we obtain a sequence $G_1, G_2,\ldots$ where the $G_i$ are pairwise disjoint. Now we see that the map from the power set of the natural numbers to ${\cal F}$ given by
$$ A \mapsto \bigcup_{i\in A} G_i $$
is injective. So the uncountability of ${\cal F}$ follows from the uncountability of ${\cal P}({\bf N})$.\slug

\bye

### 2019-12-29 misc/solutions/waerden-sols.tex
Added textbook solutions

% Solutions to exercises in Modern Algebra by
% B. L. van der Waerden, translated by Fred Blum (1949)

\input fontmac
\input mathmac

\font\bigboldsl=cmbxsl12
\font\eightsl=cmsl8

\maketitlenodate{Answers to Selected Exercises in {\bigboldsl Modern Algebra}\footnote{$^*$}{\eightpt B.\ L.\ van der Waerden, {\eightsl Modern Algebra}, translated by Fred Blum, New York: Ungar, 1949.}}{Solutions by}{Marcel K.\ Goh}

\section CHAPTER I: NUMBERS AND SETS

\vskip -\medskipamount

\section 2. Mappings. Cardinality

\proclaim 1. For an arbitrary set $A$, prove that $A\sim A$.

\proof For each element $a\in A$, let $\phi(a) = a$. It is easy to see that this is a one-to-one correspondence.\slug

\proclaim 2. Given sets $A$ and $B$, prove that $A\sim B$ implies $B\sim A$.

\proof Since $A\sim B$, there exists a one-to-one correspondence $\phi$ from $A$ onto $B$. Then $\phi^{-1}$ is a one-to-one correspondence from $B$ onto $A$.\slug

\proclaim 3. For sets $A$, $B$, and $C$, prove that if $A\sim B$ and $B\sim C$, then $A\sim C$.

\proof We have the existence of biunique mappings $\phi : A \rightarrow B$ and $\psi : B\rightarrow C$. Then $\psi\phi$ is a one-to-one correspondence from $A$ to $C$ (with $\psi^{-1}\phi^{-1}$ as its inverse).\slug

\section 3. The Number Sequence

\proclaim 1. Let a property $E$ be true, first for $n=3$, and second for $n+1$ whenever it is true for $n\geq 3$. Prove that $E$ is true for all numbers $\geq 3$.

\proof For a number $k$, let $F$ denote the property ``$E$ is true for $k+2$''. Then $E$ is true for all numbers $n\geq 3$ if and only if $F$ is true for all natural numbers $k$. We find that $F$ is true for $k=1$, since $E$ is true for $n = 3$. Then from the second statement about the property $E$, we can derive that also $F$ is true for $k+1$ whenever it is true for $k\geq 1$. So by the principle of complete induction, we have $F$ true for all natural numbers.\slug

\proclaim 3. The same as Ex.\ 1 with the number 3 replaced by 0.

\proof For a natural number $k$, let $F$ denote the property ``$E$ is true for $k-1$''. Then proceed as in the solution to Exercise 1.\slug

\section CHAPTER II: GROUPS

\vskip -\medskipamount

\section 6. The Group Concept

\proclaim 1. The Euclidean motions of space combined with reflections (i.e.\ those transformations that preserve all distances between pairs of points) form an infinite non-abelian group.

\proof We work in 2-dimensional space, but most of the work generalises to higher dimensions. Any Euclidean motion $M$ can be described as one of the following:
\medskip
\item{a)} a translation $t_{PQ}$ that takes a point $P$ to a point $Q$;
\smallskip
\item{b)} a rotation $s_P(\theta)$ of $\theta$ radians about the point $P$;
\smallskip
\item{c)} a reflection $r_H$ where $H$ is a hyperplane (line).
\medskip

(Note that the same motion may be described in multiple ways. For example, if $PQ$ and $ST$ are parallel line-segments with the same length, then $t_{PQ} = t_{ST}$.) The motion $t_{PP}$ that fixes every point satisfies the requirements of an identity element. Associativity follows from the fact that, for any point $X$ in Euclidean space, multiplication equates to composing motions. Thus $(M_1M_2)M_3(X) = M_1\big(M_2(M_3(X))\big) = M_1(M_2M_3)(X)$; since the two motions act identically on every point in the space, they are the same transformation. To see that every transformation has an inverse, we need only note that $(t_{PQ})^{-1} = t_{QP}$ for all points $P$ and $Q$; $(s_P(\theta))^{-1} = s_P(-\theta)$ for all points $P$ and choices of $\theta$; and $(r_H)^{-1} = r_H$ for every hyperplane $H$. The group is not abelian because rotations do not commute with reflections in general.\slug

\proclaim 2. Prove that the elements $e,a$ form a group (abelian) if the group operation is defined by
$$ee = e, \quad ea = a, \quad ae = a,\quad aa = e.$$

\proof By inspection, we see that $e$ is the identity element, and both elements are their own inverses. Associativity may be checked by hand, examining all eight possible triples. And finally, the group is abelian because $ea = ae = a$.\slug

\proclaim 3. Construct a multiplication table for the group of all permutations on three digits.

\solution For brevity, cycle notation is employed:
$$\vbox{\offinterlineskip
    \halign{ 
        \hfil $#$ \hfil & \hfil\vrule# \quad\hfil &
        \hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil &
        \hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil \cr
        \cdot && () & (1\,2) & (1\,3) & (2\,3) & (1\,2\,3) & (1\,3\,2) \cr
        & height2pt &&&&&\cr
        \noalign{\hrule}
        & height2pt &&&&&\cr
        () && () & (1\,2) & (1\,3) & (2\,3) & (1\,2\,3) & (1\,3\,2) \cr
        & height2pt &&&&&\cr
        (1\,2) && (1\,2) & () & (1\,3\,2) & (1\,2\,3) & (2\,3) & (1\,3)\cr
        & height2pt &&&&&\cr
        (1\,3) && (1\,3) & (1\,2\,3) & () & (1\,3\,2) & (1\,2) & (2\,3)\cr
        & height2pt &&&&&\cr
        (2\,3) && (2\,3) & (1\,3\,2) & (1\,2\,3) & () & (1\,3) & (1\,2)\cr
        & height2pt &&&&&\cr
        (1\,2\,3) && (1\,2\,3) & (1\,3) & (2\,3) & (1\,2) & (1\,3\,2) & ()\cr
        & height2pt &&&&&\cr
        (1\,3\,2) && (1\,3\,2) & (2\,3) & (1\,2) & (1\,3) & () & (1\,2\,3)\cr
    }
}\noskipslug
$$

\end

### 2019-12-31 misc/solutions/papa-sols.tex
Added some integrals

@@
 
+\proclaim 3. Prove that if $f$ is a real function on a measurable space $X$ such that $\{x:f(x)\geq r\}$ is measurable for every rational $r$, then $f$ is measurable.
+
+\proof We immediately infer that for rational $r$ and $s$, $f^{-1}\big([r,s)\big)$ is measurable, since
+$$f^{-1}\big([r,s)\big) = \{ x : f(x)\geq r\} \cap \{ x : f(x) \geq s\}^c.$$
+Next, let $I\subseteq \R$ be an open interval. Setting $J = I\cap \Q$, we can show that $f^{-1}(I)$ is measurable, since
+$$f^{-1}(I) = f^{-1}\bigg(\bigcup_{r\in J}\bigcup_{s\in J} [r,s)\bigg) = \bigcup_{r\in J}\bigcup_{s\in J} f^{-1}\big([r,s)\big)$$
+is a countable union of measurable sets.
+
+Now let $V\subseteq \R$ be open. By Lindel\"of's lemma, we can express $V$ as a countable union of open intervals $V=\bigcup_{n=1}^\infty I_n$. Then
+$$f^{-1}(V) = f^{-1}\bigg(\bigcup_{n=1}^\infty I_n\bigg) = \bigcup_{n=1}^\infty f^{-1}(I_n)$$
+is measurable.\slug
+
+\proclaim 5. Prove the following statements:
+\medskip
+\item{a)} If $f:X\rightarrow[-\infty,\infty]$ and $g:x\rightarrow[-\infty,\infty]$ are measurable, then the sets
+$$\{x : f(x) < g(x)\}\ \hbox{and}\ \{x : f(x) = g(x)\}$$
+are measurable.
+\item{b)} The set of points at which a sequence of measurable real-valued functions converges (to a finite limit) is measurable.
+
+\proof Observe that $f(x) < g(x)$ if and only if there exists a rational number $r$ such that $f(x) < r < g(x)$. So
+$$\{x : f(x) < g(x)\} = \bigcup_{r\in \Q} \big(\{x : f(x) < r\} \cap \{x : g(x) > r\}\big) = \bigcup_{r\in \Q} \Big(f^{-1}\big([-\infty,r)\big) \cap g^{-1}\big((r,\infty]\big)\Big)$$
+is a countable union of measurable sets (since $f$ and $g$ are measurable functions). Then
+$$\{x : f(x) = g(x)\} = \{x : f(x) < g(x)\}^c \cap \{x : g(x) < f(x)\}^c$$
+is measurable as well, proving (a).
+
+Let $(f_n)$ be a sequence of real-valued measurable functions.
+Then $(f_n)$ converges at a point $x$ if and only if it is Cauchy at $x$, i.e.\ for any $\epsilon > 0$ there exists $N\in \N$ such that for all $m,n\geq N$, $|f_m(x) - f_n(x)| < \epsilon$.
+By the Archimedean property, we can replace $\epsilon$ with $1/k$ for some $k\in \N$.
+So the set of all points at which $(f_n)$ converges can be written thus:
+$$\bigcap_{k=1}^\infty \bigcup_{N=1}^\infty \bigcap_{m=1}^\infty \bigcap_{n=1}^\infty \big\{x : |f_m(x) - f_n(x)| < 1/k\big\}$$
+Set $g_{m,n} = f_m - f_n$; this is a measurable function. Hence the set
+$$\big\{x : |f_m(x) - f_n(x)| < 1/k\big\} = \big\{x : |g_{m,n}(x)| < 1/k\big\} = {g_{m,n}}^{-1}\big((-1/k, 1/k)\big)$$
+is measurable for every $m$, $n$, and $k$. We have thus proved that the set of all points at which $(f_n)$ converges is a countable union of measurable sets.\slug
+
+\proclaim 6.
+Let $X$ be an uncountable set, let ${\cal M}$ be the collection of all sets $E\subseteq X$ such that either $E$ or $E^c$ is at most countable, and define $\mu(E) = 0$ in the first case, $\mu(E) = 1$ in the second.
+Prove that ${\cal M}$ is a $\sigma$-algebra in $X$ and that $\mu$ is a measure on ${\cal M}$.
+Describe the corresponding measurable functions and their integrals.
+
+First we prove that ${\cal M}$ is a $\sigma$-algebra and that $\mu$ is a measure.
+
+\proof Since $\emptyset$ is countable, $X\in {\cal M}$.
+Then for any set $E\in {\cal M}$, either $E$ or $E^c$ is at most countable; in any case $E^c\in {\cal M}$.
+Lastly, let $\{E_n\}$ be a collection of sets in ${\cal M}$.
+If every set $E_n$ is at most countable, then $\bigcup_{n=1}^\infty E_n$ is countable and thus in ${\cal M}$.
+Otherwise, there is some $k$ for which $E_k$ is uncountable.
+Then $E_k^c\in {\cal M}$ is at most countable and $\left(\bigcup_{n=1}^\infty E_n\right)^c = \bigcap_{n=1}^\infty E_n^c \subseteq E_k^c$ is countable, so $\bigcup_{n=1}^\infty E_n$ is in ${\cal M}$.
+This shows that ${\cal M}$ is a $\sigma$-algebra.
+
+It is clear that the range of $\mu$ is in $[0,\infty]$.
+To show that $\mu$ is countably additive, let $\{E_n\}$ be a disjoint collection of sets in ${\cal M}$.
+If every $E_n$ is at most countable, then $\bigcup_{n=1}^\infty E_n$ is at most countable and $\mu(\bigcup_{n=1}^\infty E_n) = 0 = \sum_{n=1}^\infty \mu(E_n)$.
+Otherwise, there exists $E_k$ such that $E_k^c$ is countable and thus we have $\mu(E_k) = 1$ and $\mu(E_k^c) = 0$.
+Since the $E_n$ are pairwise disjoint, $E_n\subseteq E_k^c$ for all $n\neq k$.
+So $\mu(E_n) = 0$ for all $n\neq k$.
+So $\mu(\bigcup_{n=1}^\infty E_n) = 1 = \sum_{n=1}^\infty \mu(E_n)$ and $\mu$ is a measure.\slug
+
+Now we claim that the measurable functions are those that are constant at all but countably many points.
+
+\proof It suffices to prove this for a real-valued function.
+Suppose $f:X\rightarrow[-\infty,\infty]$ is measurable.
+For any $a\in \R$, let $E_a = f^{-1}\big([-\infty,a)\big)$.
+Note that for any $a$, either $E_a$ is countable or $E_a^c$ is countable.
+Note also that if $a\leq b$, then $E_a \subseteq E_b$.
+So let there is a constant $c$ such that
+$$k = \sup\{a : E_a\ \hbox{is countable}\}.$$
+This supremum is not $-\infty$, since if it were, then $E_a^c$ would be countable for all $a\in \R$, and since $\bigcap_{n=0}^\infty E_{-n} = \emptyset$, $X = \bigcap_{n=0}^\infty E_{-n}^c$ is countable, a contradiction.
+By a similar argument, we know that the supremum is not $\infty$.
+So $k\in \R$ and there exists a sequence $(a_n)$ all of whose terms are less than $k$ and whose limit is $k$.
+Hence $E_k = \bigcup_{n=1}^\infty E_{a_n}$ is countable.
+Now let $(b_n)$ be a sequence that converges to $k$ such that $b_n > k$ for all $n$.
+Then note that if $f(x) > k$, then $f(x) \in [b_n, \infty) = E_{b_n}^c$ for some $n$.
+So
+$$\{x : f(x) \neq k\} = E_k \cup \{x : f(x) > k\} \subseteq E_k \cup \Big(\bigcup_{n=0}^\infty E_{b_n}^c\Big)$$
+is countable and $f$ equals $k$ at all but countably many points.\slug
+
+Lastly, we show that if $f$ is measurable and takes on the value $k$ at all but countably many points, then $\int_E f\,d\mu = k$ for all uncountable $E\in {\cal M}$.
+(If $E$ is countable then $\int_E f\,d\mu = 0$.)
+
+\proof Let $E\in {\cal M}$ be uncountable; so $\mu(E) = 1$.
+Let $s$ be a simple measurable function such that $0\leq s\leq f$.
+Suppose that $s$ takes values $\alpha_i$ on the $n$ disjoint sets $A_i$ that cover $X$.
+We know that one of the $\alpha_i$, call it $\alpha_j$, is equal to a constant $k_s$ and that $A_i$ is at most countable for all $i \neq j$.
+So for all $i \neq j$, $A_i \cap E$ is countable and $A_j \cap E$ must be uncountable.
+Then we have
+$$\int_E s\,d\mu = \sum_{i=1}^n \alpha_i \mu(A_i\cap E) = \alpha_j\mu(A_j\cap E) = k_s\cdot 1 = k_s,$$
+where $0\leq k_s \leq k$. So we have
+$$\int_E f\,d\mu = \sup \Big\{ \int_E s\,d\mu : 0 \leq s\leq f\Big\} = \sup \{k_s : 0\leq s \leq f\} = k,$$
+which is what we had to show.\slug
 \bye

### 2020-01-03 misc/solutions/baby-sols.tex
Lots of changes
Added more solutions

@@
 \font\eightsl=cmsl8
+
+\def\d{\,d}
 
@@
 \proof Let $A$ denote the set of all algebraic numbers and suppose, towards a contradiction, that all real numbers are algebraic.
-Then ${\bf R}\subset A$.
+Then ${\bf R}\subseteq A$.
 But the set of real numbers is uncountable and we know from Problem 2 that $A$ is countable.
@@
 
+\section CHAPTER 11. THE LEBESGUE THEORY
+
+\proclaim 1.
+If $f\geq 0$ and $\int_E f\d\mu = 0$, prove that $f(x) = 0$ almost everywhere on $E$.
+{\it Hint:} Let $E_n$ be the subset of $E$ on which $f(x)> 1/n$.
+Write $A = \bigcup_{n=1}^\infty E_n$.
+Then $\mu(A) = 0$ if and only if $\mu(E_n) = 0$ for every $n$.
+
+\proof First we claim that $\mu(E_n) = 0$ for every $n$. Suppose, towards a contradiction, that $\mu(E_n) > 0$ for some $n$. Then since $f(x) > 1/n$ for all $x\in E_n$, we find that
+$$\int_E f\d\mu \geq \int_{E_n} f \d\mu > {\mu(E_n)\over n} > 0,$$
+contradicting our hypothesis. Now since $\mu(E_n) = 0$ for all $n$, $\mu(A) \leq \sum_{n=1}^\infty \mu(E_n) = 0$. So $\mu(A) = 0$. Now we write
+$$\int_E f\d\mu = \int_{E\setminus A} f\d\mu + \int_{A} f\d\mu,$$
+and $f$ is equal to 0 on $E\setminus A$ by our construction of $A$.\slug
+
 \bye

### 2020-01-03 misc/solutions/papa-sols.tex
Added more solutions

@@
 
-Lastly, we show that if $f$ is measurable and takes on the value $k$ at all but countably many points, then $\int_E f\,d\mu = k$ for all uncountable $E\in {\cal M}$.
-(If $E$ is countable then $\int_E f\,d\mu = 0$.)
+Lastly, we show that if $f$ is measurable and takes on the value $k$ at all but countably many points, then $\int_E f\d\mu = k$ for all uncountable $E\in {\cal M}$.
+(If $E$ is countable then $\int_E f\d\mu = 0$.)
 
@@
 Then we have
-$$\int_E s\,d\mu = \sum_{i=1}^n \alpha_i \mu(A_i\cap E) = \alpha_j\mu(A_j\cap E) = k_s\cdot 1 = k_s,$$
+$$\int_E s\d\mu = \sum_{i=1}^n \alpha_i \mu(A_i\cap E) = \alpha_j\mu(A_j\cap E) = k_s\cdot 1 = k_s,$$
 where $0\leq k_s \leq k$. So we have
-$$\int_E f\,d\mu = \sup \Big\{ \int_E s\,d\mu : 0 \leq s\leq f\Big\} = \sup \{k_s : 0\leq s \leq f\} = k,$$
+$$\int_E f\d\mu = \sup \Big\{ \int_E s\d\mu : 0 \leq s\leq f\Big\} = \sup \{k_s : 0\leq s \leq f\} = k,$$
 which is what we had to show.\slug
+
+\proclaim 7.
+Suppose $f_n : X\rightarrow[0,\infty]$ is measurable for $n=1,2,3,\ldots$, $f_1\geq f_2\geq f_3\geq\cdots\geq 0$, $f_n(x)\rightarrow f(x)$ as $n\rightarrow\infty$ for every $x\in X$, and $f_1\in L^1(\mu)$.
+Prove that then
+$$\lim_{n\rightarrow\infty} \int_X f_n \d\mu = \int_X f\d\mu$$
+and show that this conclusion does {\it not} follow if the condition ``$f_1\in L^1(\mu)$'' is omitted.
+
+\proof Consider the sequence of functions $(f_1 - f_n)$.
+This sequence is nonnegative, nondecreasing and for any $x\in X$, $\lim_{n\rightarrow\infty} (f_1 - f_n)(x) = f_1(x) - f(x)$.
+So by the monotone convergence theorem,
+$$= \lim_{n\rightarrow \infty} \int_X f_1 - f_n \d\mu = \int_X f_1 - f\d\mu.$$
+Since the integrals of $f_n$ are finite, we can apply Theorem 1.27 to get
+$$\int_X f_1\d\mu - \lim_{n\rightarrow \infty}\int_X f_n \d\mu = \int_X f_1\d\mu -  \int_X f \d\mu,$$
+which gives us what we want after subtracting $\int_X f_1 \d\mu$ (which is finite) and multiplying by $-1$.\slug
+
+The condition $f_1\in L^1(\mu)$ is necessary.
+Let $I = [0,1]$ and $f_n = {1\over nx^2}$, which converges to the constant function 0.
+Then $\lim_{n\rightarrow \infty} \int_I f_n \d\mu = +\infty$ but $\int_I 0\d\mu = 0$.
+
 \bye

### 2020-01-05 misc/solutions/papa-sols.tex
Fatou's lemma question

@@
 
+\proclaim 8. Put $f_n = \chi_E$ if $n$ is odd, $f_n = 1 - \chi_E$ if $n$ is even. What is the relevance of this example to Fatou's lemma?
+
+This gives an example of strict inequality arising. Let $X$ be a measure space such that $\mu(X) = 1$ and let $E\subseteq X$ be such that $\mu(E) = 2/3$. Then
+$$\int_X \liminf_{n\rightarrow\infty} f_n \d\mu = \int_X 0 \d\mu = 0,$$
+while on the other hand
+$$\liminf_{n\rightarrow\infty} \int_X f_n\d\mu = \int_X 1 - \chi_E \d\mu = \int_X 1 \d\mu - \int_X \chi_E \d\mu = \mu(X) - \mu(E) = {1\over 3},$$
+and we have $0<1/3$.
+
 \bye

### 2020-01-10 misc/study-notes/ergodic.tex
Added ergodic notes

% Ergodic Group Theory Notes
% Class given by Prof. Mikael Pichot. Notes by Marcel Goh

\input fontmac
\input mathmac

% Useful functions

\def\Sym{\mathop{\rm Sym}\nolimits}
\def\Stab{\mathop{\rm Stab}\nolimits}
\def\Aut{\mathop{\rm Aut}\nolimits}
\def\Out{\mathop{\rm Out}\nolimits}
\def\Inn{\mathop{\rm Inn}\nolimits}
\def\Im{\mathop{\rm Im}\nolimits}
\def\Id{\mathop{\rm Id}\nolimits}
\def\graph{\mathop{\rm graph}\nolimits}

\leftrighttop{NOTES ON ERGODIC GROUP THEORY}{MARCEL GOH}

\maketitlenodate{Notes on Ergodic Group Theory}{by}{Marcel K.\ Goh}

\floattext6 Disclaimer. These notes are based on a course given by Prof. Mika\"el Pichot at McGill University in Winter 2020. Where I found it necessary, I have supplied extra material (e.g.\ definitions). My notes are not officially endorsed in any way and I may have introduced errors. Please contact me if you find any such error.

\section 1. PRELIMINARY NOTIONS

Let $X$ be a topological space. The smallest $\sigma$-algebra containing all the open sets in $X$ is called the {\it Borel $\sigma$-algebra} and members of this collection are called {\it Borel sets}. A map $f$ between topological spaces $X$ and $Y$ is called {\it Borel} provided that for every open set $V$ in $Y$, $f^{-1}(V)$ is a Borel set in $X$. The {\it graph} of a Borel map $f$ is the set
$$\graph(f) = \{(x, f(x)) : x\in X\}.$$
[Borelness and other things.]
A topological space $X$ is called a {\it Polish space} if it is homeomorphic to a separable complete metric space. In other words, it is a separable completely metrisable topological space. Examples of Polish spaces are $[0,1]$, $(0,1)$, manifolds, and Hilbert spaces.

Let $G$ be a group and let $X$ be a Polish space. We define a {\it Borel action} to be a Borel map from $G\times X\rightarrow X$ (where pairs $(s,x)$ are usually denoted $sx$) such that $ex = x$ for all $x\in X$ and $s(tx) = (st)x$ for all $s,t\in G, x\in X$. The set
$$Gx = \{ y \in X : y = sx\ \hbox{for some}\ s\in G\}$$
is called the {\it orbit} of $x\in X$. The partition of $X$ into orbits is called the {\it orbit equivalence relation} of the action.

\section REFERENCES

The contents of this document are heavily based on MATH 596 lectures given by Mika\"el Pichot at McGill University in Winter 2020.

\end

### 2020-04-16 misc/proofs/burnside.tex
Added Burnside problem

\input fontmac
\input mathmac

\def\d{\,d}
\def\bar{\overline}
\def\graph{\mathop{\rm graph}\nolimits}
\def\ran{\mathop{\rm ran}\nolimits}
\def\Stab{\mathop{\rm Stab}\nolimits}
\def\Out{\mathop{\rm Out}\nolimits}
\def\Aut{\mathop{\rm Aut}\nolimits}
\def\Inn{\mathop{\rm Inn}\nolimits}
\def\matr#1#2#3#4{
    \big({#1\atop #3}{#2\atop #4}\big)
}
\def\tto{\longrightarrow}
\def\m{\frak m}
\def\inj{\hookrightarrow}
\def\surj{\rightarrow\!\!\!\!\!\rightarrow}

\input cyracc.def
    \font\tencyr=wncyr10
    \font\tencyri=wncyi10
    \def\cyr{\tencyr\cyracc}
    \def\cyri{\tencyri\cyracc}

\maketitle{The Burnside Problem}{by}{Marcel K. Goh}{15 April 2020}

\section 1. Introduction

\jankscsp ONE\ OF\ THE oldest and most studied problems in group theory began life in a 1902 paper by William Burnside [3]. He wrote: ``A still undecided point in the theory of discontinuous groups is whether the order of a group may not be finite while the order of every operation it contains is finite.'' This question is now known as the {\it general} Burnside problem and it was solved in 1964 by E. S. Golod and I. R. Shafarevich [4], who found an example of an infinite finitely-generated group all of whose elements have finite order.

A more specialised version of the problem, which Burnside himself tackled in his original paper, requires that the order of every element of the group be bounded by a {\it fixed} integer $n$. The problem of whether such groups are necessarily finite is called the Burnside problem and to solve it, we consider a specific class of groups. The {\it free Burnside group with $r$ generators of order $n$}, denoted $B(r,n)$, is the group with $r$ generators $x_1, \ldots, x_r$ such that for every $s\in B(r,n)$, $s^n = e$. It can be regarded as the quotient of the free group $F_r$ by the normal subgroup $\{s^n : s\in F_r\}$. Any group with $r$ generators and all orders dividing $n$ is the image of a homomorphism from $B(r,n)$, so the Burnside problem boils down to determining if $B(r,n)$ is finite for all $r$ and $n$.

\section 2. Small Cases

The simplest case occurs when $n=2$.

\parenproclaim Theorem A (Burnside, {\rm 1902}). The order of $B(r,2)$ is $2^r$.

\proof For every $s,t\in B(r,2)$, $s^2 = t^2 = (st)^2 = e$, since every element is an involution. Then from $stst = e$ we can multiply by $s$ on the left and $t$ on the right to obtain $ts = st$. Thus $B(r,2)$ is abelian and, having $r$ generators as a basis, is isomorphic to $(\Z/2\Z)^r$, with order $2^r$.\slug

Burnside also proved that the order of $B(2,3)$ is $27$, that $B(2,4)$ has order at most $2^{12}$, and that $B(r,3)$ is finite for all $r$. Using a convoluted method that spans two and a half pages, Burnside proved that $|B(r,3)| \leq 3^{2^r-1}$, but finiteness of $|B(r,3)|$ can be shown more easily if we relax the bound. The following argument was presented by Marshall Hall, Jr.\ in his 1959 textbook {\sl The Theory of Groups} [6].

\proclaim Theorem B. For $r\geq 1$, the order of $B(r,3)$ is of the form $3^{m(r)}$ for some integer $m(r) \leq 3^{r-1}$.

\proof Because every non-identity element in $B(r,3)$ has order $3$, the group is a $3$-group and has order a power of $3$. We will perform induction on $r$. When $r=1$, the group is cyclic and its order is 3, so setting $m(1) = 0$ satisfies the inequality with equality. Now suppose that the theorem holds for some integer $k$; i.e.\ $|B(k,3)| = 3^{m(k)}$ and $m(k) \leq 3^{k-1}$. Note that for any Burnside group of exponent 3, the relation $(st)^3 = e$ implies that
$$tst = s^{-1}t^{-1}s^{-1} \oldno 1$$
for any $s,t$ in the group.

We form $B(k+1,3)$ by adding a new generator $g$ to the generating set of $B(k,3)$. Let $s\in B(k+1,3)$ It can be expressed as a finite product of the form
$$s = s_1g^{\pm 1}s_2g^{\pm 1} \cdots g^{\pm 1}s_n, \oldno 2$$
where $n$ is some integer and the $s_i$ belong to $B(k,3)$. If, in this product, any two consecutive $g$'s have the same sign in the exponent, we can use the identity ({\oldstyle 1}) with $g = t$ to perform the following replacements, each reducing the total number of $g^{\pm 1}$ terms by one:
$$gs_ig = {s_i}^{-1}g^{-1}{s_i}^{-1} \qquad \hbox{and}\qquad g^{-1}s_ig^{-1} = {s_i}^{-1}g{s_i}^{-1} \oldno 3$$
In this way, $s$ can be reexpressed in the form ({\oldstyle 2}) but with the exponents of $g$ alternating in sign.

But we can make further reductions. Since $g^3 = e$, we have $g^{-1} = g^2$ and we can perform surgery on $s$ as follows:
$$\eqalign{
s &= s_1\cdots s_i g s_{i+1} g^{-1} s_{i+2} g \cdots s_n\cr
&= s_1\cdots s_i g s_{i+1} g\cdot g s_{i+2} g \cdots s_n\cr
&= s_1\cdots (s_i {s_{i+1}}^{-1}) g^{-1} ({s_{i+1}}^{-1}{s_{i+2}}^{-1}) g^{-1} {s_{i+2}}^{-1} \cdots s_n,\cr
}$$
reducing the total instances of $g^{\pm 1}$ by one \big(the third line is due to ({\oldstyle 3})\big). An analogous reduction can be done when $s$ contains a $s_i g^{-1} s_{i+1} g s_{i+2} g^{-1}$, and in fact, in both cases we now have two consecutive matching exponents so we can immediately apply one of the identities in ({\oldstyle 3}) again. After doing this repeatedly, we find that any $s\in B(k+1,3)$ is of the form ({\oldstyle 2}) with at most two $g^{\pm 1}$s. As a final refinement, using $g = g^{-1}g^{-1}$ we can write
$$s_1g^{-1}s_2gs_3 = s_1g^{-1}s_2g^{-1}g^{-1}s_3 = (s_1{s_2}^{-1})g{s_2}^{-1}g^{-1}s_3,$$
so any $s\in B(k+1,3)$ can be written in one of the forms
$$s_1,\quad s_1gs_2, \quad s_1g^{-1}s_2, \quad s_1gs_2g^{-1}s_3,$$
where $s_1$, $s_2$, and $s_3$ are in $B(k,3)$. There are $3^{m(k)}$ possibilities for the first form, $3^{2m(k)}$ possibilites for each of the second and third forms, and $3^{3m(k)}$ for the fourth. Thus
$$|B(k+1,3)| = 3^{m(k)} + 2\cdot3^{2m(k)} + 3^{3m(k)} < 3^{3m(k) + 1},$$
so $m(k+1) \leq 3m(k)$ and $|B(k+1,3)| \leq 3^{3^{r-1}}$.\slug

In 1933, F. W. Levi and B. L. van der Waerden [8] determined the exact value of the exponent to be
$$m(r) = {r\choose 3} + {r\choose 2} + r.$$

\section 3. Burnside Groups of Exponent Four

For $n$ even just slightly larger than 2 or 3, much less is known about $B(r,n)$. For example, the exact orders of the groups $B(r,4)$ are only known for $r\leq 5$. It is known, however, that $B(r,4)$ is finite for all $r$. This result was proved by I. N. Sanov in 1940 [9], but since his paper is written in Russian, we present a treatment due to M. Hall [6]. The proof has been reorganised for the sake of clarity. We begin with a lemma.

\proclaim Lemma C. Let $G$ be a finite group of order $M$, all of whose elements have orders dividing 4. Then if we adjoin to $G$ a new generator $g\notin G$ such that $g^2\in G$, the resulting group $G' = G \cup \langle g\rangle$ is finite, when divided by the normal subgroup $\{s^n : s\in G'\}$.

\proof Since $g^{-1} = g^3$ and $(gs)^4 = e$ for any $s\in G'$, we have
$$\eqalign{
gsg &= s^{-1}g^{-1}s^{-1}g^{-1}s^{-1} \cr
&= s^{-1}g (g^2 s^{-1} g^2) gs^{-1};\cr
}$$
since $g^2\in G$, we have, for any $s\in G'$,
$$gsg = s^{-1}gs'gs^{-1} \oldno 4$$
for some $s'\in G$.

Armed with this identity, we proceed to fix some $s\in G'$, which, since $g = g^{-1}$, can be expressed as
$$s = s_1gs_2g\cdots gs_n,$$
where each $s_i\in G$. We will say that this word ``has length $n$'' because it consists of $n$ elements of $G$. Applying ({\oldstyle 4}) to some $s_i$ we get
$$s = s_1g \cdots g (s_{i-1}{s_i}^{-1})g {s_i}' g ({s_i}^{-1}s_{i+1}) g \cdots gs_n,$$
and this word has not increased in length. Note that $s_{i-1}$ has been replaced by $s_{i-1}{s_i}^{-1}$. If some $s_i = e$, then its two neighbouring $g$ terms combine to form an element of $G$, so the length of the word decreases by one.

Repeatedly applying ({\oldstyle 4}), we can replace $s_{i-1}$ with $s_{i-1}{s_i}^{-1}$, then replace $s_{i-2}$ with $s_{i-2}(s_{i-1}{s_i}^{-1})^{-1} = s_{i-2}s_i{s_{i-1}}^{-1}$, etc. So, in particular, we can replace $s_2$ with any one of
$$s_2, \quad s_2{s_3}^{-1}, \quad s_2s_4{s_3}^{-1},\quad s_2s_4{s_5}^{-1}{s_3}^{-1},\quad  \ldots\ \oldno 5$$
(we repeat this $n-2$ times). Observe that the there is a pattern to the way new $s_i$'s appear. When $i$ is even, $s_i$ appears to the left of $s_{i-1}$ and when $i$ is odd, it appears to the right of $s_{i-1}$ and it is inverted. If any one of these products is the identity, we can reduce the length of the word.

Suppose that $n\geq M+2$. Then in the list ({\oldstyle 5}) of $M$ possible replacements for $s_2$, either one of them is the identity or there is a repeated element, say
$$s_2\cdots s_{2j}{s_{2j+1}}^{-1}\cdots {s_3}^{-1} = s_2 \cdots s_{2k}{s_{2k+1}}^{-1} \cdots {s_3}^{-1},$$
where $j<k$. Then we can deduce that
$$s_{2j+2}\cdots  s_{2k}{s_{2k+1}}^{-1} \cdots {s_{2j+3}}^{-1} = e.$$
But it is possible, by repeated application of ({\oldstyle 4}), to replace $s_{2j+2}$ with this value, so we can shorten the representation of $s$. This handles the case where we have $s_{2j}{s_{2j+1}}^{-1}$ in the middle of the repeated term. It is also possible to have $s_{2j}{s_{2j-1}}^{-1}$ (here the odd term has smaller index), in which case we find that we can replace $s_{2j+1}$ with a word whose product is the identity.

So any word of length $n\geq M+2$ can be reduced. Thus any $s\in G'$ can be represented by a word that contains at most $M+1$ elements of $G$. This shows that there are only $M^{M+1}$ possibilities for $s$, so $G'$ is finite.\slug

This lemma performs most of the hard work in the following finiteness proof.

\proclaim Theorem D. For any positive integer $r$, $B(r,4)$ is finite.

\proof For a group $G$ we let
$$G^* = G/\{s^4 : s\in G\}.$$
The proof is by induction on $r$. When $r=1$, the group is cyclic of order 4. Now assume that the theorem holds for $r=k$, i.e.\ $B(k,4)$ is finite. We want to add a new generator $g$ of order 4 to this group and show that, after setting $s^4 = e$ for every $s$, the new group $B(k+1,4) = \big(B(k,4) \cup \langle g\rangle\big)^*$ is finite.

Well, since $g^4 = {(g^2)}^2 = e\in B(k,n)$, we can apply Lemma C to find that the group $G' = \big(B(k,4)\cup \langle g^2\rangle\big)^*$ is finite. Now the square of $g$ is in $G'$, so applying Lemma C once more, we find that $B(k+1,4) = \big(B(k+1,4) \cup \langle g\rangle\big)^* = \big(G' \cup \langle g\rangle\big)^*$ is finite.\slug

\section 4. The Restricted Burnside Problem

After over a century of investigation, we still know surprisingly little information about the structure of Burnside groups. Towards the middle of the 20th century, various mathematicians investigated a different, weaker version of the Burnside problem: {\sl Fix some integer $n$. For every integer $r$, does there exist an integer $b_{r,n}$ such that every finite group of exponent $n$ generated by $r$ elements has order at most $b_{r,n}$?}

If the statement is true for some $n$, then there would have to exist some largest finite group $R(r,n)$ of exponent $n$ generated by $r$ elements. Note that if $B(r,n)$ is finite, then $B(r,n) \cong R(r,n)$. In 1956, Philip Hall and Graham Higman [7] proved that $R(r,6)$ exists and that its order is
$$2^a3^{b+{b\choose 2} + {b\choose 3}},\oldno 6$$
where
$$a = 1 + (r-1)3^{r + {r\choose 2} + {r\choose 3}}\qquad\hbox{and}\qquad b=1 + (r-1)2^r.$$
Though this result does not imply the finiteness of $B(r,6)$, P. Hall and Higman showed in the same paper that $B(r,6)$ has a normal series
$$B(r,6), \supseteq M \supseteq M' \supseteq \{e\},$$
where $M'$ is a maximal normal subgroup of order coprime to 2, $M/M'$ is a $2$-group, and $G/M$ has order coprime to $2$. Marshall Hall, Jr., motivated by these results, showed in 1958 [5] that $G/M$ is finite of exponent 3, $M/M'$ is finite of exponent 2, and $M'$ is of exponent 3 and thus finite by the work of Levi and van der Waerden [8]. This proved that $B(r,6)$ is finite, of order given by ({\oldstyle 6}).

The work of P. Hall and Higman would also eventually lead to a positive solution of the restricted Burnside problem for all exponents. In their paper they proved the following theorem.

\parenproclaim Theorem E (Hall-Higman). Let $n$ be an integer whose prime decomposition is ${p_1}^{k_1}\cdots {p_n}^{k_n}$. Then the restricted Burnside problem holds for exponent $n$ provided that all of the following statements hold:
\medskip
\item{i)} The restricted Burnside problem holds for exponent ${p_i}^{k_i}$ for all $i$.
\smallskip
\item{ii)} There are finitely many simple groups of exponent $n$.
\smallskip
\item{iii)} For any finite simple group $G$ of exponent $n$, the outer automorphism group $\Out(G) = \Aut(G)/\Inn(G)$ is soluble.\noskipslug

The statements (ii) and (iii) were proven in the 1980s during the effort to classify all finite simple groups, so the last hurdle was to prove the restricted Burnside problem for exponents equal to powers of primes. The final proof was published by E. I. Zelmanov, first for odd exponents in 1990 [10], then for powers of two in 1991 [11]. He received the Fields Medal for this work in 1994.

\section 5. Infinite Burnside Groups

For $n$ not equal to 2, 3, 4, or 6, it is not known whether $B(r,n)$ is finite for all $r$. And as it turns out, the answer to the original problem of whether $B(r,n)$ is finite for all $r$ and $n$ is no. In 1968, S. I. Adian and P. S. Novikov showed that $B(r,n)$ is infinite for $n$ odd and $\geq 4381$ by means of a long combinatorial proof [1]. The result was improved to $n$ odd, $n\geq 665$ by Adian in 1975 [2].

\section REFERENCES

\frenchspacing
\item{[1]} Sergei Ivanovich Adian and Pyotr Sergeyevich Novikov, ``{\cyr O beckonechnykh periodicheskikh gruppakh} I, II, III = O beskonechnykh periodicheskikh gruppakh I, II, III,'' {\sl Izv. Akad. Nauk SSSR Ser. Mat.} {\bf 32} (1968), 212--244; 251--524; 709--731.
\smallskip
\smallskip
\item{[2]} Sergei Ivanovich Adian, {\cyri Problema Bernsa\u ida i tozhdestva v gruppakh} = {\sl Problema Bernsa\u\i da i tozhdestva v gruppakh,} (Moscow: Izdat. ``Nauka'', 1975), 335 pages.
\item{[3]} William Burnside, ``On an unsettled question in the theory of discontinuous groups.'' {\sl Quart J. Pure and Applied Math.} {\bf 33} (1902), 230--238.
\smallskip
\item{[4]} Evgenii Solomonovich Golod and Igor Rostislavovich Shafarevich, ``{\cyr O vaxne pole\u i klassov} = O vashne pole\u\i\ klassov,'' {\sl Izv.\ Akad.\ Nauk SSSR Ser.\ Mat.} {\bf 28},2 (1964), 261--272.
\smallskip
\item{[5]} Marshall Hall, Jr., ``Solution of the Burnside Problem for Exponent Six,'' {\sl Illinois J.\ of Math.} {{\bf 2} (1958), 764--786.
\smallskip
\item{[6]} Marshall Hall, Jr., {\sl The Theory of Groups}, (New York: MacMillan and Co., 1959), $\hbox{xiii} + 431$ pages.
\smallskip
\item{[7]} Philip Hall and Graham Higman, ``On the $p$-length of $p$-soluble groups and reduction theorems for Burnside's Problem,'' {\sl Proc.\ London Math.\ Soc. (3)} {\bf 6} (1956), 1--42.
\smallskip
\item{[8]} Friedrich Wilhelm Levi and Bartel Leendert van der Waerden, ``\"Uber eine besondere Klasse von Gruppen,'' {\sl Abh. Math. Sem. Hamburg\/} {\bf 9} (1933), 154--158.
\smallskip
\item{[9]} Ivan Nikolaevich Sanov, ``{\cyr Rexenie problemy B\"ernsa\u ida dlya pokazatelya 4} = Reshenie problemy B\"ernsa\u\i da dl\t\i a pokazatel\t\i a 4,'' {\sl Leningrad State Univ.\ Ann\ 10} (1940), 166--170.
\smallskip
\item{[10]} Efim Isaakovich Zelmanov, ``{\cyr Rexenie oslablenno\u i problemy Bernsa\u ida dlya grupp nechetnogo pokazatelya} = Reshenie oslablenno\u\i\ problemy Bernsa\u\i da dl\t\i a grupp nechetnogo pokazatel\t\i a,'' {\sl Izv.\ Akad. Nauk SSSR Ser.\ Mat.} {\bf 54},1 (1990) 42--59.
\smallskip
\item{[11]} Efim Isaakovich Zelmanov, ``{\cyr Rexenie oslablenno\u i problemy Bernsa\u ida dlya 2-grupp} = Reshenie oslablenno\u\i\ problemy Bernsa\u\i da dl\t\i a 2-grupp,'' {\sl Mat.\ Sb.}, {\bf 182},4 (1991), 568--592.
\bye

### 2020-04-19 misc/study-notes/math457.tex
Algebra 4 notes

\input fontmac
\input mathmac

\font\ninebf=cmbx9

\def\lcm{\mathop {\rm lcm}}
\def\Im{\mathop {\rm Im}}
\def\Sym{\mathop {\rm Sym}}
\def\End{\mathop {\rm End}}
\def\Ann{\mathop {\rm Ann}}
\def\Tor{\mathop {\rm Tor}}
\def\Spec{\mathop {\rm Spec}}
\def\Specmax{\Spec_{\max}}
\def\Nil{\mathop {\rm Nil}}
\def\Jac{\mathop {\rm Jac}}
\def\inj{\hookrightarrow}
\def\surj{\rightarrow\!\!\!\!\!\rightarrow}
\def\defn{\noindent {\bf Definition.\enspace}}
\def\bar{\overline}
\def\mod{\!\!\!\!\pmod}

\maketitle{MATH 457 Review}{by}{Marcel K. Goh}{21 April 2020}

\floattext5 \noindent {\ninebf Note.} \ninept These notes are quite rough and skip over a lot of details. \ninept Most proofs are either omitted or distilled to their main ideas.

\section 1. Rings

A {\it ring} $R$ is a set with operations $+$ and $\cdot$ such that
\medskip
\item{i)} $(R,+)$ is an abelian group;
\smallskip
\item{ii)} $(R,\cdot)$ is a semigroup;
\smallskip
\item{iii)} $\cdot$ distributes over $+$ on both sides:
$$a\cdot (b+c) = a\cdot b + a\cdot c \quad\hbox{and}\quad (a+b)\cdot c = a\cdot c + b\cdot c$$
A {\it semiring is the same as a ring} except that condition (i) above becomes
\medskip
\item{i')} $(R,+)$ is a monoid with absorbing unit 0.
\medskip
A ring is {\it unital} if $(R,\cdot)$ has a unit $1$. We always assume that $1\neq 0$, since if $1=0$ then $R = \{0\}$. Observe that in a unital ring, $(R,+)$ is necessarily abelian. A ring is said to be {\it commutative} if $(R,\cdot)$ is.

Even for commutative rings, there are many possible ring structures for $(R,+) = \Z^2$. For example we can take the {\it Gaussian integers} $\Z[i] = \{a+bi : a,b\in\Z\}$ or the {\it Eisenstein integers} $\Z[\omega] = \{a + b\omega : a,b\in \Z\}$ where
$$\omega = -{1 + i\sqrt{3} \over 2}.$$
In both cases the second binary operation is complex multiplication. Since $i$ and $\omega$ are both solutions to equations of the form $x^2 + Bx + C = 0$, they are called {\it quadratic integers} and $\Z[i]$ and $\Z[\omega]$ are called {\it quadratic rings}.

The definition of a ring is meant to describe a class of $\Z$-like objects, but many rings have properties different from the integers. For example, the ring $\Z[\sqrt{-5}]$ does not have Euclidean division. There are also many non-commutative rings such as the {\it Lipschitz quaternions}
$$\{a + bi + cj + dk : a,b,c,d\in \Z\}$$
or the {\it Hurwitz quaternions}
$$\bigg\{a + bi + cj + dk : a,b,c,d\in \Z\ \hbox{or}\ a,b,c,d\in \Z + {1\over 2}\bigg\}.$$

If $R$ is a ring, then a subgroup of $(R,+)$ that is closed under multiplication is called a {\it subring}. If a ring is unital, then any unital subring will have the same unit. A {\it homomorphism} between two rings $R$ and $S$ is a map $f:R\to S$ that preserves both operations:
$$f(a+b) = f(a) + f(b) \qquad\hbox{and}\qquad f(a\cdot b) = f(a)\cdot f(b)$$
A homomorphism that preserves the units is called {\it unital}.

An {\it ideal} in a ring $R$ is a subgroup $(I,+)$ such that
\medskip
\item{i)} $ab\in I$ for all $a\in R$, $b\in I$;
\smallskip
\item{ii)} $ab\in I$ for all $b\in I$, $a\in R$.
\medskip
If (i) holds, $I$ is called a {\it left ideal} and if (ii) holds, $I$ is called a right ideal. Let $I \subseteq R$ be an ideal. One defines the {\it quotient ring} $R/I$ as follows. Since $(I,+)$ is a normal subgroup of $(R,+)$, $R/I$ is an abelian group. We associate $r\sim r'$ if $r-r'\in I$. Then we can define multiplication in $R/I$ as $(a+I)(b+I) = ab+I$. This is well-defined because $I$ is an ideal and distributivity holds.

The isomorphism theorems for groups extend to rings as well.

\parenproclaim Theorem A (First isomorphism theorem). Let $f:R\surj S$ be a surjective ring homomorphism. Then $f$ descends to a a ring homomorphism $f' : R/I \to S$ that takes $a+I$ to $f(a)$, where $I$ is the kernel of $f$.\slug

\parenproclaim Theorem B (Second isomorphism theorem). Let $S$ be a subring and $I$ and ideal in a ring $R$. Then $S+I$ is a subring of $R$, $I$ is an ideal in $S+I$, and the map $S \surj (S+I)/I$ is a surjective ring homomorphism with kernel $S\cap I$.\slug

\parenproclaim Theorem C (Third isomorphism theorem). Let $R$ be a ring and $I\subseteq J\subseteq R$ be ideals. Then $R/I \surj R/J$ is a surjective ring homomorphism with kernel $J/I$.\slug

\parenproclaim Theorem D (Fourth isomorphism theorem). Let $f:R\surj S$ be a surjective ring homomorphism. There is a bijection between the ideals in $R$ containing $\ker f$ and the set of all ideals in $S$.\slug

Note that the correspondence in Theorem D works with subrings as well, not just ideals.

An element $r$ in a unital ring $R$ is said to be {\it invertible} if there exists $s\in R$ such that $rs = sr = 1$. The set of invertible elements is denoted $R^\times$ and this is a group under $\times$, called the {\it group of units}. A {\it field} is a ring in which every nonzero element is a unit. Non-commutative fields are called {\it division rings} or {\it skew fields} (the quaternions are an example of a skew field).

Let $K$ be a field. The set $K[x]$ of polynomials with coefficients in $K$ is a ring. Then the set
$$K(x) = \{ f / g : f,g\in K[x], g\neq 0\}$$
is a field, called the {\it field of rational functions}. The set $K[[x]]$ is called the {\it ring of formal series}: possibly infinite sums $\sum_{n\geq 0} a_nx^n$. Addition is done pointwise and multiplication is convolution of power series. The map $K[x] \to K[[x]]$ is a homomorphism and some elements become invertible. For example, $1-x$ becomes invertible, since $1/(1-x) = \sum_{n\geq 0} x^n$. Not every element in $K[[x]]$ is invertible, but one can invert the elements to get a new field $K((x))$: the set of sequences $K^\Z$ that are eventually zero when going to the left.

A {\it zero-divisor} is an element $r\in R$, $r\neq 0$ for which there exists $s\in R$ such that $rs = 0$. A ring is {\it cancellative} if $rs = rs'$ implies that $s = s'$. Then we define an {\it integral domain} to be a unital, commutative, cancellative ring. Every integral domain embeds into a field, called the {\it field of fractions}. The construction is analogous to building the rational numbers from the integers.

\proclaim Proposition Z. If $R$ is a ring with unity, there exists a unique unital homomorphism $f:\Z\to R$.\slug

\proof Since $f(1) = 1$, we have $f(n) = 1 + 1 + \cdots + 1 \in R$.\slug

The nonnegative integer $n$ which generates $\ker f$ is called the {\it characteristic} of $R$. The image of $f$ is called the {\it characteristic subring}. For example $\Z/n\Z$ has characteristic $n$.

\proclaim Proposition P. The characteristic of an integral domain $R$ is either $0$ or a prime number.\slug

An {\it algebra} over a commutative ring $R$ is a ring $A$ with a homomorphism $\eta : \R\to A$ whose image lies in the {\it centre} of $A$. Examples of algebras include rings of functions and matrices $M_n(R)$.

For a group $G$ and a ring $R$, we can define the {\it group ring} $G[R]$ as the set of all finitely supported functions from $G$ to $R$. This forms a ring with addition $(f+g)(s) = f(s) = g(s)$ and multiplication $(fg)(s) = \sum_{uv=s} f(u)g(u)$.

\section 2. Ideals

Every element $r$ in a unital ring $R$ generates a {\it principal} ideal $(r)$. More generally any subset $S\subseteq R$ does. The ideal $(S)$ is the intersection of all ideals that contain $S$. If $R$ is commutative, then $(r) = rR = Rr$. In $\Z$, the ideals are the of the form $(n) = n\Z$. Then $(n)\subseteq (m)$ if and only if $m\setminus n$ (this is true in any commutative ring). A ring $R$ in which every ideal is principal is called a {\it principal ring} and if $R$ is also an integral domain, we call it a {\it principal ideal domain} or PID.

Principal ideals determine their generators up to unit. If $(r) = (s)$, then $s = ar$ and $r = bs$ together imply that both $a$ and $b$ are units. Elements $r$ and $s$ of a ring $R$ are called {\it associate} if there exists a unit $a$ such that $r=as$.

We can define three operations on ideals. Let $I,J\subseteq R$ be ideals.
\medskip
\item{i)} $I\cap J$ is an ideal.
\smallskip
\item{ii)} $I+J = \{a + b : a\in I, b\in J\} = (I\cup J)$ is an ideal.
\smallskip
\item{iii)} $IJ = \{ab : a\in I, b\in J\}$ is an ideal.
\medskip

\proclaim Lemma P. Let $R$ be a commutative ring. Let $I = (S)$ and $J = (T)$ be two ideals. Then $IJ = (ST)$. \slug

In the ring of integers $\Z$, we have $(m)(n) = (mn)$, $(m)\cap(n) = \big(\lcm(m,n)\big)$, and $(m)+(n) = \big(\gcd(m,n)\big)$. When $I\subseteq J$ is an inclusion of ideals, one may think of it as a kind of divisibility $J\setminus I$. For example, $\gcd(m,n)\setminus \lcm(m,n)\setminus mn$.

\proclaim Lemma D. If $I,J\subseteq R$ are ideals, then
$$IJ \subseteq I\cap J \subseteq I+J.\noskipslug$$

The set of ideals forms a semiring where the two operations are $I+J$ and $IJ$. The semiring in $\Z$ is $\N$ with the addition $m + n = \gcd(m,n)$ and ordinary multiplication.

For an ideal $I\subseteq R$, we define the {\it radical} of $I$ to be the set
$$\sqrt{I} = \{a \in R : a^n \in I\ \hbox{for some}\ n\in \N\}.$$
This is an ideal and it has the property that $\sqrt{\sqrt{I}} = \sqrt{I}$. Furthermore, if $I\subseteq J$, then $\sqrt{I} \subseteq \sqrt{J}$.

An ideal $I\subseteq R$ is called {\it maximal} if it is proper and whenever $I\subseteq J\subseteq R$, then either $J = I$ or $J=R$.

\proclaim Lemma M. Let $R$ be a unital ring. Then every proper ideal is included in a maximal ideal.

\proof This is an application of Zorn's Lemma. Let $I$ be a proper ideal and let $X$ be the set of all proper ideals containing $I$, ordered by inclusion. Then this set is inductive (increasing union of ideals is an ideal) so there is a maximal element $M$.\slug

\proclaim Lemma F. Let $R$ be unital and commutative. Then an ideal $I\subseteq R$ is maximal if and only if $R/I$ is a field.

\proof This follows from the fourth isomorphism theorem.\slug

Let $R$ be a unital ring. An ideal $I$ of $R$ is {\it prime} if it is proper and for any ideals $A,B$ of $R$, $AB\subseteq I$ implies that $A\subseteq I$ or $B\subseteq I$. The {\it spectrum} of $R$ is the set of all prime ideals and it is denoted $\Spec(R)$. The {\it maximal spectrum} of $R$, denoted $\Specmax(R)$, is the set of all maximal ideals of $R$.

Maximal ideals are always prime (so $\Specmax(R) \subseteq \Spec(R)$), but not all prime ideals are maximal. For example, $(0)$ is prime in $\Z$ but certainly not maximal. A ring is called {\it local} if it has a unique maximal ideal. A ring $R$ is local if and only if $R\setminus R^\times$ is an ideal.

\proclaim Lemma C. Let $R$ be a unital commutative ring. Let $I\subseteq R$ be a proper ideal. Then $I$ is prime if and only if $ab\in I$ implies that $a\in I$ or $b\in $I.\slug

\proclaim Lemma I. Let $R$ be a unital commutative ring. Then $I\subseteq R$ is a prime ideal if and only if $R/I$ is an integral domain.\slug

Since all fields are integral domains, this proves that all maximal ideals are prime. We also have that a commutative ring $R$ is an integral domain if and only if $(0)$ is a prime ideal in $R$ (if $R$ is not commutative, then we say it is a {\it prime ring}). If $R$ is a PID, then every nonzero prime ideal is maximal.

We can view elements in a commutative unital ring $R$ as ``functions'' on the set $\Spec(R)$ of prime ideals. To $r\in R$ we identify a function $f_r$ such that $f_r(P) = r \bmod P\in R/P$. We have a bundle at every $P\in \Spec(R)$ and a fibre $R/P$ which is an integral domain. The {\it total space} $B(R)$ is the union of $R/P$ over all prime ideals $P$. A {\it section} is a map $s: \Spec(R)\to B(R)$ such that $s(P)\in R/P$. $\Gamma(R)$ is the set of all sections and $\Gamma_{\max}(R)$ is its restriction to $\Specmax(R)$. Let $\pi : R\to \Gamma(R)$ map $r\mapsto f_r$ and $\pi_{\max} : R\to \Gamma_{\max}(R)$ take $r$ to $f_r$, restricted to $\Specmax(R)$. We want to know when $\pi$ and $\pi_{\max}$ are faithful.

\proclaim Proposition K. The kernel of $\pi$ is the intersection of all prime ideals and the kernel of $\pi_{\max}$ is the intersection of all maximal ideals.\slug

For a unital commutative ring $R$, we define the {\it nilradical} of $R$ to be the intersection $\Nil(R) = \bigcap P$ of all prime ideals $P$. The {\it Jacobean radical} is the intersection $\Jac(R) = \bigcap M$ of all maximal ideals $M$. Since $\Specmax(R) \subseteq \Spec(R)$, $\Jac(R) \supseteq \Nil(R)$. An element $r\neq 0$ in a ring $R$ is called {\it nilpotent} if $r^n = 0$ for some $n$. It turns out that there is a connection between nilpotency and prime ideals.

\proclaim Proposition N. Let $R$ be unital and commutative. Then $\Nil(R)$ is the set of all nilpotent elements, i.e.
$$\sqrt{(0)} = \{r\in R : r^n = 0\ \hbox{for some}\ n\in \N\} = \bigcap_{P\in \Spec(R)} P.$$

\proof To show that a nilpotent element $r$ belongs to every prime ideal $P$, note that $r^n \in P$, so $r\cdot r^{n-1}\in P$ and we can iterate this until we get that $r\in P$. Conversely, if $r$ is not nilpotent, we can let $X$ be the set of ideals $I$ such that $r^n$ is not in $I$ for any $n$. $X$ is nonempty and inductive, so by Zorn's Lemma there is a maximal element and it can be shown that this ideal is prime.\slug

Let $R$ be a commutative ring and let $p\in R$ be a nonzero non-unit. Then $p$ is said to be
\medskip
\item{i)} {\it prime} if $p\setminus ab$ implies that $p\setminus a$ or $p\setminus b$;
\smallskip
\item{ii)} {\it irreducible} if $p=ab$ implies $a$ is a unit or $b$ is a unit.
\medskip

To find irreducible elements in a ring, may attempt the ``bisection process''. Let $r\in R$. If $r$ is irreducible, we stop. If $r$ is not irreducible, then $r = r_1 r_2$. If neither is irreducible, we continue by splitting $r_1$ and $r_2$ in the same way. This process may not terminate.

\proclaim Proposition I. Let $R$ be an integral domain. If an element $p\in R$ is prime, then it is irreducible.

\proof. Let $p\in R$ be a prime element. Assume that $p=ab$. This implies that $p\setminus a$ or $p\setminus b$. Say $a = pc$ for some $c\in R$. Then $p=ab=pcb$ and $cb = 1$. So $b$ is a unit.\slug

Note that the converse does not hold. For example, in the ring $\Z[\sqrt{-3}]$, we have $4=(1+\sqrt {-3})(1-\sqrt {-3})$. The element 2 is irreducible, but it is not prime because $2$ divides $4$ but does not divide either of $(1+\sqrt{-3})$ and $(1-\sqrt{-3})$.

\proclaim Proposition A. Let $R$ be an integral domain. Let $p$ be a nonzero element in $R$. Then $p$ is prime if and only if $(p)$ is prime and $p$ is irreducible if and only if $(p)$ is maximal among principal ideals.\slug

This proposition implies that in a PID, irreducible elements are prime.

A ring $R$ is a {\it unique factorisation domain} if every $r\in R$ can be expressed as a product $r = p_1\cdots p_n$ of irreducible elements, which is unique up to the order of the $p_i$. The rings $\Z$, $K[x]$, and $K[x,y]$ are all examples of UFDs. Every PID is a UFD and in a UFD, all irreducible elements are prime.

\proclaim Lemma S. In a PID, every chain of ideals stabilises.

\proof $I = \bigcup_{n\geq 1} I_n$ is an ideal. Since $R$ is a PID, $I = (x)$ for some $x$ and $x\in I_n$ for some $n$. This implies that $I = I_n$.\slug

\proclaim Lemma N. Let $R$ be a unital ring. Then every increasing chain of ideals stabilises if and only if every ideal is finitely generated.

\proof If $I = (x_1,x_2,\ldots)$ is not finitely generated, then $I_n = (x_1,\ldots,x_n)$ is an increasing chain of ideals that does not stabilise. Conversely, if every ideal is finitely generated, then let $I_1\subseteq I_2\subseteq \cdots$ be a chain of ideals and let $I = \bigcup_{n\geq 1} I_n$. There exist $(x_1,\ldots,x_n)$ that generate $I$, so there exists a $k$ such $x_i\in I_k$ for all $i$ and we find that $I = I_k$.\slug

A ring is called {\it Noetherian} if it the equivalent conditions from Lemma N hold.

For elements $r$ and $s$ of a ring, a {\it greatest common divisor} or gcd is an element $d$ dividing both $r$ and $s$ such that if any $d'$ divides both $r$ and $s$, then $d'$ divides $d$. An integral domain $R$ is called a {\it B\'ezout domain} if $(r)+(s)$ is principal for every $r,s\in R$ (of course, every PID is a B\'ezout domain) and it is called a {\it GCD domain} if any two $r,s\in R$ have a gcd. Every UFD is a GCD domain.

\proclaim Lemma B. The following statements regarding B\'ezout domains are true.
\medskip
\item{i)} A ring $R$ is B\'ezout if and only if every finitely generated ideal is principal.
\smallskip
\item{ii)} A B\'ezout domain is a GCD.
\smallskip
\item{iii)} If a ring is both Noetherian and a B\'ezout domain, then it is a PID.\noskipslug
\medskip

\section 3. Gaussian Integers

Recall from Section 1 that the Gaussian integers are the ring
$$\Z[i] = \{a + bi : a,b\in \Z\}.$$
We write $N$ for the complex modulus, squared. So $N(z) = z\bar z = a^2 + b^2$. This is called the {\it norm} and it is a group homomorphism $\C^\times\to \R^\times$, since $N(zz') = N(z)N(z')$. $N(z) = 0$ implies that $z = 0$. The norm $n$ takes $\Z[i]$ to $\N$. The kernel of $N$ on $\C^\times$ is the unit circle $\{z\in \C : |z| = 1\}$. Let $\ker N$ denote the kernel of $N$ restricted to $\Z[i]$, i.e.\ $\{\pm 1, \pm i\}$. These are the units of $\Z[i]$.

The image of $N$ is
$$\Im(N) = \{n\in \N : n = a^2 + b^2\ \hbox{for some}\ a,b\in \Z\}.$$
This set is stable under product, since if $n = N(z)$ and $n' = N(z')$, then $nn' = N(zz')$. Gauss was interested in studying the number of integer numbers less than a given $n$ that can be expressed as a sum of two squares. We will return to this point later.

We say that a prime number {\it splits} if it is no longer prime in $\Z[i]$ and we say that it is {\it inert} otherwise.

\proclaim Lemma S. Let $p$ be a prime. Then $p$ is a sum of two squares if and only if it splits in $\Z[i]$.

\proof If $p = a^2 + b^2$ then $p = (a + ib)(a-ib)$ and $N(a+ib)N(a-ib) = p^2$ implies that neither of these factors are units. So $p$ is not prime in $\Z[i]$. Conversely, if $p = \alpha\beta$ in $\Z[i]$, then $N(\alpha) = N(\beta) = p$ means that $p$ is the sum of two squares.\slug

\proclaim Lemma I. A prime $p$ splits if and only if $p\equiv 1 \mod 4$.

\proof If $p$ splits, then by the previous lemma, $p = a^2 + b^2$ and the sum of two squares is never $3$ modulo 4. So if $p$ is an odd prime it is congruent to 1 modulo 4. Conversely, assume that $p\equiv 1\mod 4$. Then $p = 1 + 4n$ for some $n$ and there exists $x\in \Z$ such that $x^2 \equiv 1\mod p$. (In fact, $x = (2n)!$ works.) Then $p$ divides $x^2 + 1 = (x+i)(x-i)$. So $p$ divides $(x+1)$ or $(x-i)$, so $p$ divides $i$ and is not inert.\slug

All this talk of divisibility leads nicely into a discussion of Euclidean division. In $\Z$, the goal of Euclidean division for integers $a$ and $b$ is to find a $q\in \Z$ such that $a-bq$ is small, in some sense. The following proves a similar result in $\Z[i]$.

\proclaim Proposition E. There is a Euclidean division in $\Z[i]$.

\proof. Let $a,b\in \Z[i]$, $b\neq 0$. We can divide them in $\C$ to get $z = a/b$. Then there is a (not necessarily unique) $q\in \Z[i]$ that is of minimal distance to $z$. We have $|z-q|<1$; in fact $|z-q| \leq \sqrt{2}/2 < 1$. So $|a-bq| < |b|$.\slug

Let us now define this generally. An integral domain $R$ is a {\it Euclidean domain} if there exists a function $N : R\to \N$ called the {\it norm} such that $N(0) = 0$ and for all $a,b\in R$, $b\neq 0$, either $b$ divides $a$ or there exists $q\in R$ such that $N(a-bq) < N(b)$. Proposition E showed that $Z[i]$ is a Euclidean domain with the complex norm, and other familiar examples include $\Z$ with the absolute value function and $K[x]$ with the degree of a polynomial as its norm. In general, the Euclidean division algorithm does not give a unique answer. Even in $\Z$, we can end up with $q$ or $-q$ as a quotient.

\proclaim Proposition T. $\Z[\sqrt{-2}]$ is a Euclidean domain.

\proof We repeat the same proof as for $\Z[i]$ except for the computation of $|z-q|$, which is now $\leq \sqrt{3}/2$.\slug

Recall that $\Z[\sqrt{-3}]$ is not a Euclidean domain. It is not even a UFD, since $4 = 2\cdot 2 = (1+\sqrt{-3})(1-\sqrt{-3})$. But $\Z[\sqrt{-3}]\subseteq \Z[\omega]$ and this is a Euclidean domain, with norm $N(a+b\omega) = a^2 - ab + b^2$. The units in $\Z[\omega]$ are the elements of norm 1: $\{\pm 1, \pm \omega, \pm \omega^2\}$ and we have unique factorisation up to units.

\proclaim Proposition P. Every Euclidean domain is a PID (and consequently a UFD).

\proof Let $R$ be a Euclidean domain and $I\subseteq R$ an ideal. Let $b\neq 0$ be an element of $I$ of minimal norm. If $a\in I$ then $b$ divides $a$. Otherwise, there exists $q\in R$ such that $N(a-bq) < N(b)$, contradicting the minimality of $b$'s norm. So $I$ is principal.\slug

A corollary of this fact is that every ideal in $\Z[i]$ is principal.

\section 4. Modules

For any set $X$, the set of symmetries $\Sym(X)$ is a group and an action of a group $G$ on $X$ is a group homomorphism $G\to \Sym(X)$. If $X$ is a group, we can define the {\it ring of endomorphisms} $\End(X)$ as the set of group homomorphisms from $X$ to $X$.

\proclaim Lemma M. Let $M$ be an abelian group. Then $\End(M)$ is a ring.

\proof Addition is pointwise addition from $M$ and multiplication is composition of maps.\slug

Let $R$ be a a unital commutative ring. A {\it module} $M$ over $R$ is a ring homomorphism $R\to \End(M)$. Explicitly, the list of axioms of a module are very similar to those of a vector space (in fact, if $R$ is a field, then a module is a vector space). For $r,s\in R$ and $m,n\in M$, we have
\medskip
\item{i)} $r(m+n) = rm + rn$;
\smallskip
\item{ii)} $(r+s)m) = rm + sm$;
\smallskip
\item{iii)} $(rs)m = r(sm)$;
\smallskip
\item{iv)} $1m = m$.
\medskip
These axioms also work if $R$ is not commutative; in this case, we call $M$ a {\it left $R$-module}. The kernel of $R\to \End(M)$ is called the {\it annihilator} of $M$:
$$\Ann(M) = \{r\in R : rm = 0\ \hbox{for all}\ m\in M\}$$
A module is said to be {\it faithful} if $\Ann(M)$ is trivial. If $M$ is an $R$-module, then $M$ is a faithful $S$-module where $S = R/\Ann(M)$.

\proclaim Proposition I. Any ideal $I$ in a ring $R$ is a module over $R$.

\proof For $a\in I$ and $r\in R$, we have $ra\in I$. The rest of the axioms follow.\slug

Quotients $R/I$ are also modules. When $R = \Z$, the action is determined by the group structure in $M$. For example,
$$2m = (1 + 1)m = 1m + 1m = m + m.$$
When $R = K[x]$ for some field $K$, we have the following interesting lemma.

\proclaim Lemma V. $K[x]$-modules are operators on vector spaces and vice versa.

\proof Let $M$ be a $K[x]$-module. The restriction of the $K[x]$ action to $K$ gives a $K$-module structure on $M$. This is a vector space. Furthermore, the indeterminate $x$ also acts on $M$ by taking $m\mapsto xm$. This gives a map $x: M \to M$ such that $x(m + n) = x(m)$ and $x(rm) = (xr)m = (rx)m = r\cdot x(m)$. So $x$ is a linear map.

Conversely, if $V$ is a $K$-vector space, and $T:V\to V$ is a linear map, then $V$ is a $K[x]$-module, because for any $p\in K[x]$, $p(T)$ is a linear map on $V$.\slug

Note that the module is not faithful, because $K[x]$ has infinite dimension but $\End(V)$ has finite dimension when $V$ has finite dimension. If $G$ is a group, $K[G]$ is the group ring and a $K[G]$-module is a linear representation of $G$.

A {\it submodule} $M'$ of $M$ is a subgroup that is stable under the action of the ring, i.e.\ for all $m, n\in M'$ and $r\in R$, $m+rn\in M'$. For example, ideals are submodules of $R$ and if $M$' is a submodule, we can define the {\it quotient module} $M/M'$ with the action of $R$:
$$ r(m+M') = rm + M'$$
If $M$ and $M'$ are modules, then $M\times M'$ is a module. If a module has no proper nontrivial submodules, then it is called {\it simple}.

An $R$-module map is a group homomorphism $f:M\to M'$ such that $f(rm) = rf(m)$ for all $r\in R$ and $m\in M$. The kernel $\ker f$ is a submodule of $M$ and the image of $f$ is a submodule of $M'$. The isomorphism theorems for modules are exactly analogous to the ones given for rings in Section 1.

\parenproclaim Lemma S (Schur's lemma). Let $M$ be a simple module. Then $\End_R(M)$ is a skew field.

\proof Let $f:M\to M'$ be a module map that is not identically zero. The kernel of $f$ is a submodule of $M$, so since $f\neq 0$, $\ker f = \{0\}$. Then the image of $f$ is a submodule of $M'$ since $f\neq 0$, $\Im f = M'$. Hence $f$ is an isomorphism.\slug

If $M$ is an $R$-module and $I\subseteq R$ is an ideal, then
$$IM = \Big\{\sum r_im_i : r_i\in I, m_i\in M\Big\} \subseteq M$$
is a submodule.

\parenproclaim Theorem C (Chinese remainder theorem). Let $I,J$ be ideals in a ring $R$. Let $M$ be an $R$-module. Then the map
$$M\to M/IM \times M/JM$$
has kernel $IM\cap JM$.\slug

If $I+J = R$ then the map is surjective and $(I\cap J)M = IJM$. With $n$ ideals such that $I_k + I_l = R$ for $k\neq l$, we have
$$M/(I_1\cdots I_n)M \cong M/I_1M \times \cdots \times M/I_nM.$$

Let $M$ be an $R$-module. If $A\subseteq M$, then
$$(A) = \Big\{\sum r_ia_i : r_i\in R, a_i \in A\Big\}$$
is the submodule of $M$ {\it generated} by $A$. A module is {\it finitely generated} if it admits a finite generating set and {\it cyclic} (or {\it singly generated}) if it is generated by one element. If $M = (a)$ is cyclic, then the map $R\to M$ that sends $r\mapsto ra$ is surjective with kernel $\Ann(M)$.

\proclaim Lemma P. Let $R$ be an integral domain. Then the nonzero principal ideals are isomorphic to $R$.

\proof Let $I = (a)$ be an ideal (so it is an $R$-module). If $r\in \Ann(I)$ then $r$ is a zero divisor. So $R\to I$ is an isomorphism.\slug

A finitely generated $R$-module $M$ is called {\it free} if it is isomorphic to $R^n$ for some $n$. For example, if $R$ is a field, every module (finite-dimensional vector space) is free. Equivalently, an $R$-module is free if there exists a basis, that is, a generating set $A$ such that any $m\in M$ can be written in a unique way as a finite sum
$$m = \sum_{a\in A} r_a a.$$
The set $A$ is called a {\it free generating set} and the cardinality of $A$ is called the {\it rank} of $M$.

In a PID, every ideal is a free module (isomorphic to the ring itself). For any set $A$ and ring $R$, we can let $F_A$ be the set of all functions from $A$ to $R$ with finite support. This is a group under pointwise addition and $r$ acts on $F_A$: $(rf)(a) = r\cdot f(a)$. A basis for $F_A$ is the set $(\delta_a)_{a\in A}$ of delta functions, where $\delta_a(b) = 1$ if $b=a$ and $0$ otherwise.

\parenproclaim Proposition U (Universal property of free modules). Let $\phi$ be a map from a set $A$ to an $R$-module $M$. Then there is a unique extension of $\phi$ to a module map $\bar\phi : F_A\to M$.

\proof Take any element $f\in F_A$ and express it as
$$f = \sum_{a\in A} r_a\delta_a$$
for some $r_a\in R$. Then let $\bar\phi$ be given by
$$\bar\phi(f) = \sum_{a\in A} r_a\phi(a).\noskipslug$$

\proclaim Proposition S. Let $N\inj M \surj F$ be a short exact sequence of modules (so $F \cong N/M$), where $F$ is a free module. Then the sequence splits, i.e.\ $M\cong N\oplus F$.

\proof We need to construct the section $s$ of $\pi : M\surj F$.

\bye

### 2020-04-19 misc/study-notes/math457.tex
Finished modules

@@
 \def\Im{\mathop {\rm Im}}
+\def\Id{\mathop {\rm Id}}
 \def\Sym{\mathop {\rm Sym}}
@@
 
-\maketitle{MATH 457 Review}{by}{Marcel K. Goh}{21 April 2020}
+\maketitle{MATH 457 Honours Algebra 4\footnote{$^*$}{\eightpt Course given by Prof. Mika\"el Pichot at McGill University}}{Notes by}{Marcel K. Goh}{23 April 2020}
 
@@
 \medskip
-\item{i')} $(R,+)$ is a monoid with absorbing unit 0.
+\item{i')} $(R,+)$ is a monoid with absorbing identity 0.
 \medskip
@@
 \proof We need to construct the section $s$ of $\pi : M\surj F$.
+Let $A$ be a basis of $F$.
+Since $\pi$ is surjective, for any $a\in A$ we can find $m_a\in M$ such that $\pi(m_a) = m$.
+This gives a map $s_* : A \to M$ and by the universal property there is a unique extension $s : F_A\to M$.
+We have $\pi \circ s = \Id$ on the basis and therefore everywhere on $F$.
+So $s$ is a section of $\pi$.
+Let $F' = \Im(s) \subseteq M$.
+So $F'\cong F$ as $R$-modules.
+We claim that $M = N\oplus F'$ (viewing $N$ as a submodule of $M$).
+
+Firstly, $N\cap F' = \{0\}$, since if $m\in N\cap F'$, then $\pi(m) = 0$ and there exists $f\in F$ such that $s(f) = m$.
+But this implies that $f = \pi\big(s(f)\big) = \pi(m) = 0$, so $m=0$.
+And $M = N + F'$ because any $m\in M$ can be expressed as the sum of $\big(m-s\circ \pi(m)\big) + s\circ\pi (m)$.\slug
+
+The following theorem shows that the rank of a free module is well-defined.
+
+\proclaim Theorem R. If $R^n\cong R^m$ as $R$-modules, then $n = m$.
+
+\proof Since $R$ is unital and commutative, it contains a maximal ideal $M$.
+Let $K = R/M$ and consider the submodule $MR^n = \big\{s(x_1,\cdots,x_n) \in R^n: s\in M, x_i\in R\big\}$.
+The quotient module is $K^n = (R/M)^n$ and a module isomorphism $R^n \cong R^m$ descends to a $R$-module isomorphism $K^n \cong K^m$.
+This map has kernel $M$ and is a $K$-vector space isomorphism.
+So the dimension of the two vector spaces are the same and thus $n=m$.\slug
+
+Let $M$ be a module over an integral domain $R$. The set of {\it torsion elements}
+$$\Tor(M) = \{m\in M : rm = 0\ \hbox{for some}\ r\neq 0\}$$
+is a submodule of $M$.
+A module is called {\it torsion} if $\Tor(M) = M$ and {\it torsion-free} if $\Tor(M) = \{0\}$.
+Note that $R^n$ is torsion-free, since it has a basis $\{e_n\}$ and if $am = ra_1e_1 + \cdots + ra_ne_n = 0$, then $ra_i = 0$ for all $i$ and $m = 0$.
+
+\proclaim Proposition T. For any module $M$ over an integral domain $R$, $M/Tor(M)$ is torsion-free.
+
+\proof Let $N = M/\Tor(M)$ and let $\bar m\in N$.
+Suppose there exists $r\neq 0$ such that $r\bar m = 0$.
+So $\bar{rm} = 0$ and $rm \in \Tor(M)$.
+Thus there exists $s\neq 0$ such that $(rs)\bar m = 0$.
+But $\bar{rs}\neq 0$ so $m$ must be 0.
+Hence $\Tor(N) = \{0\}$.\slug
+
+\proclaim Lemma G. A module $M$ over an integral domain $R$ is torsion if and only if it is generated by torsion elements.
+
+\proof The forward direction is clear.
+Conversely, suppose $M = (A)$ and every element in $A$ is torsion.
+Let $m = s_1a_1 + \cdots + s_na_n \in M$ for some $s_i\in R$ and $a_i\in A$.
+Each $a_i$ is a torsion element so there is $r_i$ such that $r_ia_i = 0$.
+Let $r = r_1\cdots r_n$.
+Then $rm = 0$.\slug
+
+\proclaim Proposition F. Let $M$ be a finitely generated module over an integral domain $R$. There exists a free module $F\subseteq M$ such that $M/F$ is torsion.
+
+\proof Let $M = (A)$ where $A$ is finite.
+Let $B\subseteq A$ be a maximal basis which generates a free module $F$ of rank $n = |B|$.
+Let $N$ be the quotient $M/F$.
+For every $a\in A\setminus B$, the module $(B\cup \{a\})$ is not free.
+So there exists $r\in R, r_b\in R$, not all zero, such that
+$$ra + \sum_{b\in B} r_b b = 0.$$
+Note that $r\neq 0$, otherwise $B$ would not be a basis. But $ra = 0 \bmod F$, so $N$ is generated by torsion elements and by the previous lemma, $N$ is torsion.\slug
+
+\section 5. Modules over PIDs
+
+An $R$-module has properties very much like a vector space when $R$ is a PID.
+
+\proclaim Proposition F. Let $R$ be a PID. Then every submodule of a free module $R^n$ is free of rank $k\leq n$.
+
+\proof We proceed by induction.
+When $n=1$, every ideal $I\subseteq R$ is free, isomorphic to $R$.
+Now assume the proposition is true for $R^n$.
+Let $M$ be a submodule of $R^{n+1}$.
+Let $\pi : R^{n+1} \surj R^n$ be the projection map on to the first $n$ coordinates.
+So we have a short exact sequence
+$$\ker(\pi_{|M}) \inj M \surj \pi(M).$$
+But $\pi(M)$ is a submodule of $R^n$ so, by the induction hypothesis, it is free and the sequence splits. Thus $M\cong \ker(\pi_{|M}) \oplus \pi(M)$ is free.\slug
+
+A module $M$ over a unital ring $R$ is called a {\it Noetherian module} if every submodule is finitely generated.
+
+\proclaim Propostion N. Let $M$ be a left $R$-module. The following are equivalent:
+\medskip
+\item{i)} $M$ is Noetherian.
+\smallskip
+\item{ii)} $M$ satisfies the ascending chain condition on left modules.
+\smallskip
+\item{iii)} If $\frak F$ is a nonempty family of submodules, there exists a maximal element in $\frak F$ with respect to inclusion.
+\medskip
+
+\proof To show that (i) implies (ii), we let $N_1\subseteq N_2 \subseteq N_3 \subseteq \cdots$ be an increasing sequence of modules.
+We need to know there is an upper bound.
+Let $N = \bigcup_{i\geq 1} N_i$.
+Since $N$ is finitely generated, there will be a first index $i$ such that all the generators of $N$ belong to $N_i$.
+Thus $N = N_i$ for some $i$.
+
+We show that (ii) implies (iii) by contraposition.
+If (iii) fails, then there exists a family $\frak F$ of submodules for which there is no maximal element.
+Pick $N_1\in \frak F$.
+We can find $N_2$ such that $N_1$ is properly contained in $N_2$.
+Continuing in this way, we are left with an increasing chain that does not stabilise.
+
+Lastly, assume that (i) does not hold; i.e.\ there is a submodule $N$ that is not finitely generated.
+Let $\{n_1,n_2,\ldots\}$ be an infinite countable subset of $N$ such that, for every $k$, $N_k = (n_1,\ldots,n_k)$ is properly contained in $N_{k+1} = (n_1,\ldots,n_{k+1})$.
+Now $\frak F = \{N_k\}$ is a family of submodules without a maximal element, so (iii) fails.\slug
+
+\proclaim Proposition S. Let $N \inj P\surj Q$ be an exact sequence of modules. Then $N$ and $Q$ are Noetherian if and only if $P$ is Noetherian.
+
+\proof Clearly $N$ is Noetherian if $P$ is.
+To show that $Q$ is Noetherian, let $M\subseteq Q$ be a submodule.
+Then, by the fourth isomorphism theorem, $M$ is the image of a submodule $M'$ of $P$.
+Since $M'$ is finitely generated, $M$ is as well.
+
+Conversely, assume that $N$ and $Q$ are Noetherian and let $M\subseteq P$ be a submodule.
+Let $\pi:P\surj Q$ be the quotient map and consider the exact sequence $\ker(\pi_{|M})\inj M\surj \pi(M)$.
+Let $X$ be a finite generating set for $\ker(\pi_{|M})$ and $Y$ be a finite set in $M$ such that $\bar Y = \pi(Y)$ is a finite generating set of $\pi(M)$.
+Then for any $m\in M$, then there exist some $r_x$ and $r_y$ in $R$ such that
+$$m = \sum_{x\in X} r_x x + \sum_{y\in Y} r_y y$$
+and $X\cup Y$ generates $M$.\slug
+
+\proclaim Theorem R. The following are equivalent:
+\medskip
+\item{i)} $R$ is a Noetherian ring.
+\smallskip
+\item{ii)} The free module $R^n$ is Noetherian for every $n$.
+\smallskip
+\item{iii)} Every finitely generated $R$-module is Noetherian.\noskipslug
+\medskip
+
+A corollary of Theorem R is that every finitely generated module over a PID is Noetherian. For example, $R = K[x_1,\ldots,x_n]$ is Noetherian.
+
+\proclaim Lemma N. If $R$ is a PID and $M$ a torsion-free $R$-module, then $M$ is free.
+
+\proof In general, we showed that there exists $F$ free such that $F\inj M \surj T$, where $T$ is torsion.
+Since $M$ is Noetherian, we can choose a maximal $F$ satisfying this property.
+We claim that $M=F$.
+Let $\pi:M\surj T$ denote the quotient map and let $m\in M$.
+Since $\pi(m)\in T$ is a torsion element, there exists $r\in R$ such that $r\pi(m) = 0$.
+So $rm\in \ker\pi$ and $rm\in F$.
+Let $f_r : M \to M$ be the map that sends $m\mapsto rm$.
+This map is injective because $M$ is torsion-free.
+Since $f_r(F) \subseteq F$ and $f_r(M) \subseteq F$, so the submodule $f_r(F,M)$ is contained in $F$.
+By Proposition F, $f_r(F,M)$ is free, so $M$ is free.\slug
+
+\proclaim Theorem T. Let $M$ be a finitely generated module over a PID $R$. Then $M\cong R^n \oplus \Tor(M)$.
+
+\proof Consider the exact sequence $\Tor(M)\inj M\surj N$ where $N$ is torsion-free. Since $R$ is a PID, $N$ is free and the sequence splits, giving us the desired direct sum decomposition.\slug
+
+The integer $n$ given by Theorem T is called the {\it free rank} of a module.
+If two modules $M$ and $N$ are isomorphic, then their free ranks are equal and $\Tor(M)\cong \Tor(N)$.
+The following theorem is called the structure theorem for finitely generated modules over a PID.
+
+\proclaim Theorem S.
+Let $R$ be a PID and let $F \cong R^n$ be a finitely generated free module.
+Let $M$ be a finitely generated submodule of $F$.
+Then there exists a basis $(e_1,\ldots, e_n)$ of $F$ and elements $r_1,\ldots,r_m\in R$ such that $(r_1e_1,\ldots,r_me_m)$ forms a basis of $M$:
+$$F/M \cong R^{n-m} \oplus R/(r_1) \oplus \cdots \oplus R/(r_m)$$
+The elements $r_i$ are unique up to multiplication by a unit if we assume that $r_i$ divides $r_{i+1}$.\slug
+
+The elements $r_i$ in Theorem S are called the {\it invariant factors} of the module.
 

### 2020-04-19 misc/study-notes/math457.tex
Added new sections

@@
 \def\Id{\mathop {\rm Id}}
+\def\Ev{\mathop {\rm Ev}}
 \def\Sym{\mathop {\rm Sym}}
+\def\Aut{\mathop {\rm Aut}}
 \def\End{\mathop {\rm End}}
@@
 \def\mod{\!\!\!\!\pmod}
+\def\F{{\bf F}}
 
@@
 
-\floattext5 \noindent {\ninebf Note.} \ninept These notes are quite rough and skip over a lot of details. \ninept Most proofs are either omitted or distilled to their main ideas.
+\floattext5 \noindent {\ninebf Note.} \ninept These notes are rough and may skip over some details. \ninept Some proofs are either omitted or distilled to their main ideas.
 
@@
 
-\section 5. Modules over PIDs
+\section 5. Modules Over PIDs
 
@@
 
+\section 6. Fields and Polynomials
+
+Because the kernel of a homomorphism is an ideal, then any nontrivial homomorphism $f:K\to R$ is injective when $K$ is a field.
+If $R = L$ is a field, then $K\subseteq L$ is a subfield and we call $L$ an {\it extension} of $K$.
+We will  often denote this by $L/K$.
+If $L/K$ is a field extension then $L$ is a vector space over $K$ and the dimension $[L:k] = \dim_K L$ is called the {\it degree} of the extension.
+Every there is a basis $(\alpha_i)$ of $L$ such that any element $l\in L$ can be expressed as $\lambda_1\alpha_1 + \cdots \lambda_n\alpha_n$ where $\lambda_i\in K$.
+If $K\subseteq L\subseteq M$ is a chain of extensions and $(\beta_j)$ is a basis of $M$ over $L$, then it can shown that $(\alpha_i\beta_j)$ is a basis of $M$ over $K$.
+So $[M:K] = [M:L][L:K]$.
+
+A field is {\it prime} if it contains no proper nontrivial subfields.
+Any field $K$ is an extension of a prime field that contains $1$, $1+1$, etc.
+as well as their inverses.
+If the characteristic of $K$ is 0, then the prime field is $\Q$, and if the characteristic is a prime $p$, then the prime field is $\F_p$.
+
+\proclaim Lemma F. Let $K$ be a field of characteristic $p$. The Frobenius map $x\mapsto x^p$ is a field homomorphism (so it is injective).
+
+\proof For any $x,y\in K$, we have $(x+y)^p = x^p + y^p$ (by the binomial theorem) and $(xy)^p = x^py^p$.\slug
+
+Let $L/K$ be an extension and $S\subseteq L$ be a set.
+Then $K(S)$ is the subfield of $L$ generated by $S$.
+The extension field is finitely generated if $S$ is finite.
+If $S$ consists of a single element $\alpha$, then $L = K(\alpha)$ is called a {\it simple} extension and $\alpha$ is a {\it primitive element}.
+For $\alpha_1,\ldots,\alpha_n$, the extensions $K(\alpha_1,\ldots,\alpha_n)$ and $K(\alpha_1)\cdots(\alpha_n)$ are the same and the order in which the elements are adjoined does not matter.
+
+If $K$ is a field then $K[x]$ is a PID. So for any irreducible polynomial $f\in K[x]$, $(f)$ is maximal and $L = K[x]/(f)$ is a field extension. This is called the {\it Kronecker construction}.
+
+\proclaim Lemma R. Every $f\in K[x]$ admits a root in a finite-degree extension.
+
+\proof We may assume that $f$ is irreducible. It is of finite degree so $L = K[x]/(f)$ is a finite degree extension and if $\alpha = x \bmod f$, then $f(\alpha) = 0$ in $L$.\slug
+
+Kronecker's construction is universal in the following sense.
+Let $L/K$ be an arbitrary extension and let $\alpha\in L$.
+Consider the map $\Ev_\alpha : K[\alpha] \to L$ that takes a polynomial $f$ to $f(\alpha)$.
+Since $K[x]$ is a PID, the $\ker(\Ev_\alpha)$ is principal and equals $(f_\alpha)$ for some polynomial $f_a\in K[x]$.
+Because $L$ is an integral domain, one of two things may happen.
+The first is that $f_\alpha$ is an irreducible polynomial, in which case we say that $\alpha$ is {\it algebraic}.
+The second is that the kernel is trivial and in this case we call $\alpha$ {\it transcendental}.
+
+When $\alpha$ is algebraic, the unique monic irreducible polynomial $f_\alpha$ such that $f_\alpha(\alpha) = 0$ is called the {\it minimal polynomial} of $\alpha$ and $K(\alpha) \subseteq L$ is obtained by the Kronecker construction
+$$K(\alpha)\cong K[x]/(f_\alpha).$$
+If $\alpha$ is transcendental, $\Ev_\alpha : K[x] \inj L$ is injective and it extends to the fraction field $K(x) \inj L$ by taking $f/g \mapsto f(\alpha)/g(\alpha)$ (since $g(\alpha) \neq 0$ whenever $g\neq 0)$.
+Liouville established the existence of transcendental numbers in 1844 by proving that
+$$L = \sum_{n\geq 0} {1\over 10^{n!}}$$
+is transcendental. We have $\Q(L)\cong \Q(x) \subseteq \R$. Other famous transcendental numbers are $\pi$ and $e$.
+
+An extension $L/K$ is {\it algebraic} if every $\alpha\in L$ is algebraic over $K$ and the following lemma proves some properties of algebraic extensions.
+
+\proclaim Lemma A. Assume that an extension $L$ is generated by $\alpha_1,\ldots,\alpha_n$ over $K$. The following are equivalent:
+\medskip
+\item{i)} The elements $\alpha_1,\ldots,\alpha_n$ are algebraic over $K$.
+\smallskip
+\item{ii)} The degree $[L:K]$ is finite.
+\smallskip
+\item{iii)} Every $\alpha\in L$ is algebraic over $K$.
+\medskip
+
+\proof Suppose (i) holds. We have
+$$K \subseteq K(\alpha_1)\subseteq K(\alpha_1, \alpha_2) \subseteq \cdots \subseteq L,$$
+and since each $\alpha_i$ is algebraic over $K$, it is algebraic over $K(\alpha_1,\ldots,\alpha_{i-1}$, whence
+$$\big[ K(\alpha_1,\ldots,\alpha_i) : K(\alpha,\ldots,\alpha_{i-1})\big] < \infty.$$
+By the multiplicativity of the degree, $[L:K] \leq \infty$.
+
+Suppose (iii) fails, i.e.\ some $\alpha\in L$ is not algebraic. Then $K(x) \cong K(\alpha) \inj L$ is an injection into $L$, contradicting the fact that $[K(x) : K] < \infty$. Thus (ii) implies (iii).
+
+That (iii) implies (i) is obvious, so we are done.\slug
+
+To construct extensions of a field, we need to find irreducible polynomials.
+Over $\Q$, we can consider $x^n - p$ where $p$ is a prime.
+Then $\Q(\root n\of p)$ is an extension of degree $n$ over $Q$.
+A general check for irreducibility is given by the following criterion.
+
+\parenproclaim Theorem E (Eisenstein's criterion).
+Let $R$ be an integral domain and let $f\in R[x]$ be a monic polynomial of degree $n$.
+If there exists a prime ideal $\frak p$ such that $f = x^n \bmod \frak p$ and $f(0) \notin \frak p^2$, then $f$ is irreducible.
+
+\proof Suppose, towards a contradiction, that $f = ab$ is reducible.
+Then, we have $x^n = \bar a \bar b$, where the bar indicates the polynomials modulo $\frak p$.
+In particular, $\bar a\bar b$ has zero constant term.
+The ideal $\frak p$ is prime, so $R/\frak p$ is an integral domain, so both $\bar a$ and $\bar b$ have zero constant term modulo $\frak p$, meaning that the constant terms of $a$ and $b$ both belong to $\frak p$.
+This is a contradiction, since it is clear that the constant term of $f$ belongs to $\frak p^2$.\slug
+
+We can use Eisenstein's criterion to show that cyclotomic polynomials of the form
+$$\Phi_p(x) = {x^p -1\over x-1} = x^{p-1} + x^{p-2} + \cdots + x + 1$$
+for $p$ prime are irreducible. The criterion does not immediately apply, but if we consider
+$$\Phi_p(x+1) = x^{p-1} + px^{p-2} + {p(p-1)\over 2}x + p,$$
+we find that Eisenstein's criterion applies, so $\Phi_p(x+1)$ is irreducible and this implies that $\Phi_p(x)$ is irreducible, since any factorisation for $\Phi_p(x)$ would give a factorisation for $\Phi_p(x+1)$ (replacing $x$ with $x+1$).
+
+There are other many other criteria for reducibility/irreducibility; we give two more famous ones.
+
+\parenproclaim Theorem C (Cohn's criterion). Suppose that $p$ is prime and in some base $b$,
+$$p = a_nb^n + \cdots + a_1b + a_0.$$
+Then $f = a_nx^n + \cdots + a_1x + a_0$ in $\Z[x]$.\slug
+
+\parenproclaim Lemma G (Gauss' lemma).
+Let $R$ be a UFD with fraction field $K$ and let $f\in R[x]$ be a polynomial of degree $n$ such that $\gcd(a_n, \ldots, a_0) = 1$.
+If $f$ is reducible in $K[x]$, then $f$ is reducible in $R[x]$.\slug
+
+Gauss' lemma is often applied with $R = \Z$ and $K = \Q$.
+
+\section 7. Splitting Fields
+
+We begin with a lemma regarding the interchangeability of the roots of a polynomial.
+
+\proclaim Lemma I.
+Let $f\in K[x]$ be a monic, irreducible polynomial and let $L/K$ be an extension of the field $K$.
+If $\alpha,\beta$ are two roots of $f$ in $L$, then there is a field isomorphism $K(\alpha) \cong K(\beta)$.
+
+\proof This follows from the universality of the Kronecker construction: $K(\alpha) \cong K[x]/(f) \cong K(\beta)$.\slug
+
+More generally, any field isomorphism $\phi : K\to L$ extends uniquely to a ring isomorphism $\bar\phi : K[x] \to L[x]$ defined by applying $\phi$ on the coefficients.
+Then $f\in K[x]$ is irreducible if and only if $\bar\phi(f)$ is irreducible.
+Let $\alpha$ be an arbitrary root of an irreducible polynomial $f\in K[x]$ and let $\beta$ be an arbitrary root of $\bar\phi(f)$.
+Then there exists a unique field isomorphism $\phi^* : K(\alpha) \to L(\beta)$ that takes $\alpha$ to $\beta$ and whose restriction to $K$ is $\phi$.
+A corollary of this fact is that if $f\in K[x]$ is irreducible, then all roots of $f$ have the same multiplicity in an algebraic closure (this will be expanded on later).
+
+A {\it splitting field} for a polynomial $f\in K[x]$ is an extension $L/K$ such that
+$$f = \prod_{i=1}^n (x - \alpha_i)$$
+for $\alpha_i\in L$ and $L = K(\alpha_1,\ldots,\alpha_n)$.
+
+\proclaim Proposition S. Every polynomial $f\in K[x]$ of degree $n$ admits a splitting field of degree at most $n!$.
+
+\proof Let $\alpha_1$ be an abstract root of $f$ in $K(\alpha_1)$ obtained by the Kronecker construction.
+Then $f = (x-\alpha_1)f_1$ for some $f_1\in K(\alpha_1)$.
+Let $\alpha_2$ be a root of an irreducible factor of $f_1$ and extend the field to $K(\alpha_1,\alpha_2)$.
+This process happens at most $n$ times, by which time we will have found a splitting field $L$ of $f$.
+We have
+$$K \subseteq K(\alpha_1) \subseteq K(\alpha_1,\alpha_2) \subseteq \cdots \subseteq L,$$
+and the degree of each $f_i$ is $n-i$. So by multiplicity of the degrees we have $[L:K] \leq n!$.\slug
+
+For example, the polynomial $f = x^4 - 1$ has roots $\pm 1, \pm i$.
+When $K = \Q$, the abstract bound for the degree of the splitting field is $4!
+= 24$, but in fact $\Q(i)$ is a splitting field for $f$, of degree 2.
+More generally, when $f = x^n - 1$, the roots are the $n$th roots of unity $1, \omega, \ldots, \omega^{n-1} \in\C$, where $\omega = e^{2ki\pi/n}$ for $i = 0,\ldots,n-1$.
+The roots form a group isomorphic to $\Z/n\Z = \langle\omega\rangle$ and $K(\omega)$ is the splitting field of $x^n - 1$, since if one root is added, all of its powers come along for the ride.
+If $n$ is prime, then the degree of this splitting field is $n-1$; in general, the degree is equal to $\varphi(n)$ where $\varphi$ denotes Euler's totient function.
+
+The following theorem shows that splitting fields are unique up to $K$-isomorphism.
+
+\parenproclaim Theorem K (Kronecker, {\rm 1887}).
+Let $f\in K[x]$ be an irreducible polynomial.
+Let $\alpha$ and $\alpha'$ be two roots of $f$ in two splitting fields $L/K$ and $L'/K$ respectively.
+Assume the existence of a map $\theta_0 \in \Aut(K)$ that fixes coefficients of $f$.
+Then there exists an isomorphism $\theta : L \to L'$ such that $\theta_{|K} = \theta_0$ and $\theta(\alpha) = \alpha'$.
+
+\proof Extend $\theta_0$ to $\bar{\theta_0} : K[x]\to K[x]$.
+Then since $\theta_0$ fixes the coefficients of $f$, $\bar{\theta_0}(f) = f$.
+By the uniqueness of the Kronecker extension, there is an extension $\theta_1 : K(\alpha) \to K(\alpha)$ such that $\theta_1(\alpha) = \alpha'$ and $\theta_1_{|K} = \theta_0$.
+Write $f = (x - \alpha)f_1
+
 \bye

### 2020-04-20 misc/study-notes/math457.tex
Steinitz

@@
 \def\Aut{\mathop {\rm Aut}}
+\def\Gal{\mathop {\rm Gal}}
 \def\End{\mathop {\rm End}}
@@
 
-\maketitle{MATH 457 Honours Algebra 4\footnote{$^*$}{\eightpt Course given by Prof. Mika\"el Pichot at McGill University}}{Notes by}{Marcel K. Goh}{23 April 2020}
+\maketitle{MATH 457 Honours Algebra 4\footnote{$^*$}{\eightpt Course given by Prof.\ Mika\"el Pichot at McGill University}}{Notes by}{Marcel K. Goh}{23 April 2020}
 
@@
 
-\proclaim Propostion N. Let $M$ be a left $R$-module. The following are equivalent:
+\proclaim Proposition N. Let $M$ be a left $R$-module. The following are equivalent:
 \medskip
@@
 
-\parenproclaim Theorem K (Kronecker, {\rm 1887}).
+\parenproclaim Theorem K (Kronecker{\rm, 1887}).
 Let $f\in K[x]$ be an irreducible polynomial.
@@
 
-\proof Extend $\theta_0$ to $\bar{\theta_0} : K[x]\to K[x]$.
-Then since $\theta_0$ fixes the coefficients of $f$, $\bar{\theta_0}(f) = f$.
-By the uniqueness of the Kronecker extension, there is an extension $\theta_1 : K(\alpha) \to K(\alpha)$ such that $\theta_1(\alpha) = \alpha'$ and $\theta_1_{|K} = \theta_0$.
-Write $f = (x - \alpha)f_1
+\proof The proof is by strong induction on $n = \deg f$.
+If $f$ splits in $K[x]$, then $L = K = L'$.
+We can take $\theta = \theta_0$, finishing the case when $n=1$ and when every irreducible factor of $f$ has degree 1.
+
+Now we assume that the theorem is proved for any field $K$, automorphism $\theta_0$ and polynomial $f$ of degree less than $n$.
+Now let $p\in K[x]$ be an irreducible factor of $f$ of degree at least 2.
+If $\alpha\in L$ and $\alpha'\in L'$ are roots of $p$, then we can extend $\theta_0$ to an isomorphism $\theta : K(\alpha) \to K(\alpha')$.
+Let $K_1 = K(\alpha)$ and ${K_1}' = K(\alpha')$ for short.
+We have $f = (x-\alpha)f_1$ in $K_1$ and $f = (x-\alpha'){f_1}'$ in ${K_1}'$ where $f_1$ and ${f_1}'$ have degree $n-1$.
+Now $L$ is a splitting field for $f_1$ over $K_1$ and $L'$ is a splitting field for ${f_1}'$ over ${K_1}'$.
+Since the degrees of $f_1$ and ${f_1}'$ are less than $n$, by the induction hypothesis there is an isomorphism $\theta^* : L \to L$ that extends the isomorphism $\theta : K_1 \to {K_1}'$.
+The restriction of $\theta^*$ to $K_1$ is $\theta$, and the restriction of that onto $K$ is $\theta_0$.\slug
+
+Let $L/K$ and $L'/K$ be field extensions.
+A {\it $K$-embedding} $L \inj L$ is an injective homomorphism that fixes $K$.
+If $\theta : L\to L'$ is a bijection, we call it a {\it $K$-automorphism}.
+The {\it Galois group} of a polynomial $f\in K[x]$ is the group of $K$-automorphisms of a splitting field of $K$.
+If $L$ is a splitting field, we denote this group by $\Gal(L/K)$ or $\Aut(L/K)$.
+For short, we may use the notation $\Gal(f)$ for a polynomial $f$, but this is only well-defined up to conjugacy.
+If $L/K$ and $L'/K$ are two splitting fields, then by Theorem K there exists a $K$-isomorphism $\theta:L\to L'$ and $\Aut(L/K) \cong \Aut(L'/K)$.
+
+\proclaim Lemma A. Let $R$ be the roots of a polynomial $f$. Then $\Gal(f)$ acts on $R$.
+
+\proof If $\theta\in \Gal(f)$ and $f(\alpha) = 0$, we have
+$$\alpha^n + \lambda_{n-1}\alpha^{n-1} + \cdots + \lambda_1\alpha + \lambda_0 = 0$$
+and
+$$\theta(\alpha)^n + \lambda_{n-1}\theta(\alpha)^{n-1} + \cdots + \lambda_1\theta(\alpha) + \lambda_0 = 0$$
+so $\theta(\alpha)\in R$. This defines an action $\Gal(f) \to \Sym(R)$.\slug
+
+\proclaim Proposition F. The action of $\Gal(f)$ on the set of roots $R$ is faithful.
+
+\proof Let $L = K(R) = K(\alpha_1,\ldots,\alpha_n)$ be a splitting field.
+Suppose that $\theta\in \Gal(f)$ acts trivially on $R$, i.e.\ $\theta(\alpha_i) = \alpha_i$ for all $i$.
+Since $\theta_{|K} = \Id$, $\theta = \Id$ as an automorphism of $L$.\slug
+
+We have established that $\Gal(f) \inj \Sym(R)$ is a group of permutations of the roots of $f$.
+
+\proclaim Proposition O.
+Let $\Gal(f)$ act on the set $R$ of roots of $f$.
+Then the orbit $\Gal(f)\alpha$ of $\alpha\in R$ is the set $R_1$ of roots of $f_1$ where $f_1$ is the irreducible factor of $f$ such that $f_1(\alpha) = 0$.
+
+\proof If $f_1(\alpha) = 0$, since $f_1\in K[x]$.
+we have $f_1\big(\theta(\alpha)\big) = 0$ so $\theta(\alpha)\in R_1$ (the set of roots of $f_1$).
+Thus $\theta(R) \subseteq R_1$.
+Furthermore, since $f_1$ is irreducible, Kronecker's uniqueness theorem shows that for any two roots $\alpha$ and $\beta$ of $f_1$, there exists some $\theta_i : L\to L$ such that $\theta_i(\alpha) = \beta$.
+So $\Gal(f)\alpha = R_1$.\slug
+
+As a corollary, if $f$ is already irreducible, then $\Gal(f)$ is transitive.
+In general, it is not tractable to classify all the transitive subgroups (up to conjugacy) of $S_n$.
+This has been done for small $n$, however; for example, when $n = 6$ there are 16 different possible subgroups.
+
+Let us look at an example.
+Consider $f = (x^2 - 2)(x^2 - 3)$ over the field $K = \Q$.
+Then $R = \{\pm \alpha = \sqrt 2, \pm \beta = \sqrt 3\}$ and $L = \Q(\alpha, \beta)$.
+The Galois group cannot take $\sqrt 2$ to $\sqrt 3$ because such a permutation does not preserve the relations between the roots: If $\theta(\alpha) = \beta$, then $2 = \theta(\alpha^2) = \theta(\alpha)^2 = 3$, a contradiction.
+In this case, $\Gal(f)$ is the Klein four-group $V$.
+Let $\theta_2$ be the permutation that switches $\pm \sqrt 2$ and let $\theta_3$ transpose $\pm \sqrt 3$.
+Then the two commute and generate a group isomorphic to $V$.
+
+As another example, consider $f = x^4 - 2$ over $K = \Q$.
+Eisenstein's criterion with $p=2$ tells us that $f$ is irreducible, and the set of roots turns out to be $R = \{\pm \alpha = \root 4 \of 2, \pm \beta = i \root 4 \of 2\}$.
+The splitting field is $L = \Q(\alpha, \beta)$.
+We can employ a useful trick, namely that {\sl if the coefficients of $f$ are real, then complex conjugation permutes the roots.} This gives us an element in $\Gal(f)$ of order 2: the permutation that fixes $\alpha$ and takes $\beta\mapsto -\beta$.
+In any case, we will employ a more general method to compute $\Gal(f)$.
+Let $\theta\in \Gal(f)$.
+Since $\alpha^2 + \beta^2 = 0$, we have
+$$\theta(\alpha)^2 + \theta(\beta)^2 = 0.$$
+Suppose that $\theta(\alpha) = \beta$. Then $\theta(\beta)^2 = -\beta^2$ and $\theta(\beta)$ is $\pm i\beta$. If $\theta(\beta) = -i\beta = \alpha$, then we have the order 2 automorphism
+$$s = (\alpha \leftrightarrow \beta, -\alpha\leftrightarrow-\beta).$$
+If, instead, we have $\theta(\beta) = i\beta = -\alpha$, then we have the order 4 automorphism
+$$r = (\alpha \mapsto \beta \mapsto -\alpha \mapsto -\beta \mapsto \alpha).$$
+We conclude that $\Gal(f) = \langle r,s \rangle = D_4$.
+
+A field is {\it algebraically closed} if every $f\in K[x]$ admits a root in $K$.
+For example, $\C$ is algebraically closed.
+Every algebraically closed field is infinite.
+The Kronecker construction tells us that for any finite set $F\subseteq K[x]$, there is a finite extension $L$ of $K$ such that every polynomial $f\in F$ splits in $L$.
+We can extend our definition of a splitting field to (not necessarily finite) subsets of $K[x]$.
+Then an {\it algebraic closure} of $F$ is a splitting field for $K[x]$.
+
+The following theorem gives the existence and uniqueness (up to $K$-isomorphism) of the algebraic closure $\bar K$ for every field $K$.
+The construction is ``universal'' in the sense that if $L/K$ is an algebraic extension, then there exists a $K$-embedding $L\inj \bar K$.
+
+\parenproclaim Theorem S (Steinitz{\rm, 1910}). Let $K$ be a field. There exists an algebraic closure $\bar K$ of $K$ and this extension is unique up to $K$-isomorphism.
+
+\proof First we show existence.
+Let $\frak A$ be the set of all algebraic extensions of $K$, ordered by set inclusion.
+Observe that $\frak A$ is inductive; indeed, if $L_1\subseteq L_2\subseteq \cdots$ is a chain, then $\bigcup_{i=1}^\infty L_i$ is in $\frak A$.
+By Zorn's Lemma, there exists a maximal element of $\frak A$, call it $L$.
+We claim that $L$ is algebraically closed.
+Let $f\in L[x]$ be a polynomial.
+By the Kronecker construction, there is an extension $L(\alpha)$ of $L$.
+Then $L(\alpha)$ is algebraic and $L\subseteq L(\alpha)$.
+Since $L$ is a maximal element in $\frak A$, $L = L(\alpha)$ and $f$ admits a root in $L$.
+
+Next we show uniqueness of the algebraic closure.
+It is enough to show that if $L$ is algebraic over $K$, then $L\inj \bar K$ (if there were two algebraic closures, this would make them isomorphic).
+So fix an algebraic extension $L/K$.
+Consider the set of all intermediate extensions $K\subseteq L_\phi\subseteq L$ and for each $L_\phi$ there exists an embedding $\phi : L_\phi \to \bar K$.
+Let $\frak B$ be the set of all such $\phi$, partially ordered in the following manner: $\phi \leq \phi'$ if $L_\phi \subseteq L_{\phi'}$, i.e.\ $\phi'$ extends $\phi$.
+
+The poset $\frak B$ is inductive, since if $\phi_1 \leq \phi_2 \leq \cdots$, then we can let
+$$L_\phi = \bigcup_{i=1}^\infty L_{\phi_i}.$$
+For any $\alpha\in L_\phi$, we have $\alpha\in L_{\phi_i}$ for some $i$ and we can simply set $\phi(\alpha) = \phi_i(\alpha)$.
+(This does not depend on the choice of $i$ since the functions extend one another.) By Zorn's Lemma, $\frak B$ admits a maximal element $\phi : L_\phi \to \bar K$.
+We want to show that $L_\phi = L$.
+If not, then there exists $\alpha\in L\setminus L_\phi$.
+By the uniqueness of Kronecker, $\phi: L_\phi \to \bar K$ admits an extension $\bar\phi : L_\phi(\alpha) \to \bar K$.
+But since $\bar \phi$ is an extension of $\phi$, by the maximality of $\phi$ we have $\phi= \bar\phi$ so $\alpha\in L_\phi$.\slug
+
+As corollaries of Theorem S we have $\bar K = \bar{\bar K}$ and if $L/\bar K$ is algebraic, then $L = \bar K$.
 

### 2020-04-21 misc/study-notes/math457.tex
Separable elements and fixed fields

@@
 \def\Tor{\mathop {\rm Tor}}
+\def\sep{{\rm sep}}
 \def\Spec{\mathop {\rm Spec}}
@@
 \def\F{{\bf F}}
+\def\bsep#1{[#1]_{\sep}}
 
@@
 $$p = a_nb^n + \cdots + a_1b + a_0.$$
-Then $f = a_nx^n + \cdots + a_1x + a_0$ in $\Z[x]$.\slug
+Then $f = a_nx^n + \cdots + a_1x + a_0$ is irreducible in $\Z[x]$.\slug
 
@@
 
-Gauss' lemma is often applied with $R = \Z$ and $K = \Q$.
+Gauss' lemma is often applied with $R = \Z$ and $K = \Q$. It says that it $f$ is irreducible in $\Z[x]$, then it is irreducible in $\Q[x]$.
 
@@
 
+\proclaim Proposition R.
+Let $K$ be a field and $f\in K[x]$ monic and irreducible.
+Let $\alpha,\beta\in \bar K$ be roots of $f$.
+Then there exists $\theta\in \Aut_K(\bar K)$ such that $\theta(\alpha) = \beta$.
+Conversely, if $\theta(\alpha) = \beta$ then $\alpha$ and $\beta$ have the same minimal polynomial.
+
+\proof Theorem K provides an isomorphism $\theta : K(\alpha) \to K(\beta)$ such that $\theta(\alpha) = \beta$.
+Then Theorem S extends $\theta$ (not uniquely) to a map in $\Aut_K(\bar K)$.
+Conversely, if $\theta$ fixes $K$ then it fixes the coefficients of the minimal polynomial of $\alpha$ and $\beta$.\slug
+
+The group $\Aut_K(\bar K)$ is called the {\it absolute Galois group} over $K$ and it acts on $\bar K$ with finite orbits.
+The orbits are precisely the set of roots of irreducible polynomials in $K[x]$.
+
+\proclaim Theorem N. Let $K \subseteq L\subseteq \bar K$ be an intermediate extension of $K$. The following are equivalent:
+\medskip
+\item{i)} $L/K$ is a splitting field for a subset of polynomials over $K$.
+\smallskip
+\item{ii)} Every irreducible polynomial $f\in K[x]$ which admits a root in $L$ splits in $L$.
+\smallskip
+\item{iii)} For all $\theta\in \Aut_K(\bar K)$, $\theta(L) = L$.
+\medskip
+
+\proof Conditions (i) and (ii) are clearly equivalent.
+To show that (ii) implies (iii), let $\theta\in \Aut_K(\bar K)$ and $\alpha\in L$.
+There exists $f\in K[x]$ irreducible such that $f(\alpha) = 0$.
+So $\theta(\alpha)$ is a root of $f$ and $\theta(\alpha)\in L$.
+Lastly, suppose (iii) holds.
+Let $f\in K[x]$ be an irreducible polynomial and let $\alpha\in L$ be a root of $f$.
+If $\beta$ is another root of $f$, by Proposition R there is a $\theta\in \Aut_K(\bar K)$ that takes $\alpha$ to $\beta$, and by hypothesis, $\theta(L) = L$, so $\beta\in L$.\slug
+
+Any intermediate extension satisfying the equivalent conditions of Theorem N is called a {\it normal} extension.
+For any normal extension $L$, and any $\theta\in \Aut_K(\bar K)$, we can restrict $\theta$ to $L$.
+This gives a surjective group homomorphism from $\Aut_K(\bar K)$ to $\Aut_K(L)$.
+The latter is denoted $\Gal(L/K)$ and is called the {\it Galois group} of $L/K$.
+
+\section 8. Separable Extensions
+
+Besides normality, there is another important concept in the Galois theory called separability.
+Let $f\in K[x]$ be irreducible and $L = K[x]/(f) = K(\alpha)$ be the Kronecker extension, where $\alpha$ is the abstract root, and let $\bar K$ be an algebraic closure of $K$.
+How many $K$-embeddings $L\inj \bar K$ are there?
+This number is called the {\it separable degree} of $L/K$ and is denoted $\bsep{L:K}$.
+This is the degree of an intermediate subfield $L_\sep\subseteq L$, called the {\it separable closure} of $K$ in $L$.
+Then
+$$\bsep{L:K} = [L_\sep : K].$$
+We define the {\it separable degree} $\deg_\sep(f)$ to be the number fo distinct roots of $f$ in a splitting field.
+Thus if $\alpha$ is a root in an extension $K(\alpha)$ of $K$ and $f_\alpha$ is its minimal polynomial, then
+$$\bsep{K(\alpha):K} = \deg_\sep(f_\alpha).$$
+A polynomial $f$ is called {\it separable} if its roots are all distinct, i.e.\ $\deg(f) = \deg_\sep(f)$.
+We can make similar definitions for an element $\alpha$ of an extension.
+Its {\it separable degree} is the separable degree of its minimal polynomial and it is {\it separable} if its minimal polynomial is separable.
+This is equivalent to $\bsep{K(\alpha):K} = [K(\alpha):K]$.
+An algebraic extension is {\it separable} if it contains only separable elements. The ``universal property'' in Theorem 7S gives us the multiplicativity formula
+$$\bsep{M:K} = \bsep{M:L} \bsep{L:K}$$
+whenever $K\subseteq L\subseteq M$ is a chain of algebraic extensions.
+
+\proclaim Proposition M. Let $L = K(\alpha_1,\alpha_2,\ldots,\alpha_n)$ be an algebraic extension. Then $\bsep{L:K} \leq [L:K]$ and the following are equivalent:
+\medskip
+\item{i)} The elements $\alpha_1,\alpha_2,\ldots,\alpha_n$ are separable over $K$.
+\smallskip
+\item{ii)} $\bsep{L:K} = [L:K]$.
+\smallskip
+\item{iii)} Every element of $L$ is separable over $K$.
+\medskip
+
+\proof By invoking the multiplicativity of degrees, it is enough to prove the proposition for a primitive extension $K(\alpha)$.
+The equivalence of (i) and (ii) is discussed above and (iii) trivially implies (i).
+So we prove that (ii) implies (iii).
+If $\beta\in K(\alpha)$, then $K\subseteq K(\beta) \subseteq K(\alpha) = L$ and
+$$[L:K] = [K(\alpha) : K(\beta)][K(\beta) : K].$$
+If $\alpha$ is separable over $K$, then it is also separable over $K(\beta)$ so
+$$[K(\alpha):K(\beta)] = \bsep{K(\alpha):K(\beta)}.$$
+Thus if we assume that $\bsep{L:K} = [L:K]$, then we also have $\bsep{K(\beta) : K} = [K(\beta) : K]$ so $\beta$ is separable over $K$.\slug
+
+A {\it Galois extension} is an extension that is algebraic, normal, and separable.
+
+\proclaim Proposition G. An extension is Galois if and only if it is the splitting field of a family of separable polynomials.
+
+\proof If an extension is the splitting field of a family of separable polynomials then it is clearly normal and separablility comes from the fact that $L/K$ is generated by the roots of separable polynomials.
+Conversely, if $\alpha\in L$, then the minimal polynomial of $\alpha$ splits into linear factors in $L$ with no multiple roots, so the extension is both normal and separable.\slug
+
+If $K\subseteq L\subseteq M$ is a tower of extensions, then $M/K$ may not be Galois even though $M/L$ and $L/K$ are. In particular, the normality condition may fail. The standard example is
+$$\Q \subseteq \Q[\sqrt 2] \subseteq \Q[\root 4\of 2].$$
+The extension $\Q[\root 4\of 2]$ is not normal, but the intermediate extensions are quadratic extensions and therefore Galois.
+However, it is true that $M/K$ is separable if and only if $M/L$ and $L/K$ are separable.
+
+Earlier, for an extension $L/K$ we defined an intermediate subfield $L_\sep\subseteq L$ called the separable closure of $K$ in $L$.
+This is the set of elements in $L$ which are separable over $K$.
+Since $L_\sep/K$ is separable, $[L_\sep:K]$ is exactly the number of $K$-embeddings from $L_\sep\inj \bar K$.
+Since $\bar K$ contains the roots of every polynomial in $K[x]$, we can define the {\it separable closure} of $K$ to be $K^\sep = \bar K_\sep$.
+A field is called {\it perfect} if every irreducible polynomial over it is separable.
+We will prove that $\bar K_\sep = \bar K$ if and only if $K$ is perfect.
+
+For a polynomial $f\in K[x]$, we define its {\it derivative} $f'\in K[x]$ in the the usual way. If $f(x) = a_nx^n + \cdots + a_1x + a_0$, then $f'(x) = na_nx^{n-1} + \cdots + a_1$.
+
+\proclaim Lemma D. Let $K$ be a field. Then $f$ is separable if and only if $f$ and $f'$ have no common root in $\bar K$, which is true if and only if $\gcd(f,f') = 1$.
+
+\proof Observe that $\gcd(f,f')\neq 1$ if and only if $f$ and $f'$ have a common irreducible factor and this is equivalent to them having a common root in $\bar K$.
+If $\alpha\in \bar K$ is a root of $f$, then $f = (x-\alpha)g$ for some $g$ and by the product rule, $f' = g + (x-\alpha)g'$ so if $\alpha$ is also a root of $f'$, then $g(\alpha) = 0$ and $f = (x-\alpha)^2h$ for some $h$.
+Conversely, if $f = (x-\alpha)^2h$ for some $h$, then $\alpha$ is a root of both $f$ and $f'$.\slug
+
+If $K$ is a field and $f$ is irreducible, then $f$ is separable if and only if $f' \neq 0$.
+If the characteristic is zero, then $\deg(f') = \deg(f) - 1$, so $f' \neq 0$ is equivalent to $f$ being a nonconstant irreducible.
+Thus fields of characteristic zero are perfect.
+A field of characteristic $p$ is perfect if the Frobenius map that takes $x\mapsto x^p$ is surjective.
+
+\section 9. Fixed Fields
+
+Let $G$ be a semigroup and $L$ be a field. A {\it multiplicative character} is a map $\chi : G\to L$ which is multiplicative:
+$$\chi(st) = \chi(s)\chi(t)$$
+
+\parenproclaim Lemma D (Dedekind's lemma). Multiplicative characters on a semigroup are linearly independent, i.e.\ if $\chi_1,\ldots,\chi_n$ are distinct and there exist $\beta_i\in L$ such that
+$$\sum_{i=1}^n \beta_i\chi_i = 0,$$
+then all $\beta_i$ are 0.
+
+\proof The proof is by induction.
+If $n=1$, then $\beta_1\chi_1(s) = 0$ for all $s$ implies that $\beta = 0$.
+Now assume the lemma holds for $n$ and assume that $\sum_{i=1}^{n+1} \beta_i\chi_i = 0$.
+Then for any $s,t\in G$, we infer from multiplicativity that
+$$\sum_{i=1}^{n+1} \beta_i\chi_i(s)\chi_i(t) = 0.$$
+On the other hand,
+$$\Big(\sum_{i=1}^{n+1} \beta_i\chi_i\Big)\chi_{n+1}(t) = 0,$$
+so by subtracting the two equations we get
+$$\sum_{i=1}^n \Big(\beta_i\big(x_i(t) - \chi_{n+1}(t)\big)\Big) \chi_i(s) = 0$$
+for all $s\in G$. By the induction hypothesis, $\chi_1,\ldots,\chi_n$ are independent over $L$, so for $i = 1,\ldots,n$
+$$\beta_i\big(\chi_{i}(t) - \chi_{n+1}(t)\big) = 0$$
+for all $t\in G$.
+Now, $\chi_i \neq \chi_{n+1}$ if and only if there exists $t\in G$, $\chi_i(t) \neq \chi_{n+1}(t)$, so $\beta_i = 0$ for all $i \leq n$.
+Then $\beta_{n+1}\chi_{n+1} = 0$ so $\beta_{n+1} = 0$.\slug
+
+Dedekind's lemma tells us that if $L/K$ is a field extension, then the set of homomorphisms from $K$ to $L$ is linearly independent.
+This provides an upper bound on the number of homomorphisms if the dimension of the extension is finite.
+
+Let $L$ be a field and $G\subseteq \Aut(L)$ be a subgroup. The {\it fixed field} of $G$ is the field
+$$L^G = \{\alpha\in L : \theta(\alpha) = \alpha\ \hbox{for all}\ \theta\in G\}.$$
+To give a simple proof of the main result of the section, we will require the primitive element theorem. The proof given is the treatment given by van der Waerden.
+
+\parenproclaim Theorem P (Primitive element theorem).
+Let $L = K(\alpha_1,\ldots,\alpha_n)$ be a finite algebraic extension of $K$ and suppose that $\alpha_2,\ldots,\alpha_n$ are separable (it is not required that $\alpha_1$ be separable).
+Then there exists an element $\gamma$ such that $L = K(\gamma)$.
+
+\proof We can suppose that $L$ is infinite, since otherwise $K$ is finite and we can let $\gamma$ be a primitive root of unity that generates $K^\times$.
+Moreover, it suffices to show the theorem for two elements $\alpha$ and $\beta$, with $\beta$ separable, since a simple induction will extend the result to arbitrary $n$.
+Let $f$ and $g$ be irreducible polynomials for which $\alpha$ and $\beta$ are roots, respectively.
+Let $\alpha_1,\ldots,\alpha_r$ be the distinct roots of $f$ and $\beta_1,\ldots,\beta_s$ be roots of $g$; let $\alpha =\alpha_1$ and $\beta = \beta_1$.
+
+For $k\neq 1$ we have $\beta_k\neq \beta_1$ so the equation
+$$\alpha_i + \beta_k x = \alpha_1 + \beta_1x$$
+has at most one solution in $K$ for every $i$ and every $k\neq 1$. If we take $c\in K$ to be different from any of these solutions, we have
+$$\alpha_i + c\beta_k \neq \alpha_1 + c\beta_1$$
+for every $i$ and $k\neq 1$. Now we let
+$$\gamma = \alpha_1 + c\beta_1 = \alpha + c\beta;$$
+this is an element of $K(\alpha,\beta)$.
+
+The element $\beta$ satisfies
+$$g(\beta) = 0\qquad\hbox{and}\qquad f(\gamma - c\beta) = f(\alpha) = 0,$$
+with coefficients in $K(\gamma)$.
+The polynomials $g(x)$ and $f(\gamma - cx)$ only have the root $\beta$ in common, since for $k\neq 1$, $i = 1,\ldots,r$, $a_i \neq \theta - c\beta_k$ and $f(\gamma - c\beta_k) \neq 0$.
+Since $\beta$ is separable, the polynomials $g(x)$ and $f(\gamma - cx)$ only have the factor $x-\beta$ in common and the coefficients of this factor must lie in $K(\gamma)$, so $\beta\in K(\gamma)$.
+The same thing can be shown for $\alpha$ from the identity $\alpha = \gamma - c\beta$.
+So $K(\gamma) = K(\alpha,\beta)$.\slug
+
+The main theorem of this section links fixed fields to Galois extensions.
+
+\parenproclaim Theorem A (Artin's fixed field theorem). Let $L$ be a field and let $G$ be a subgroup of $\Aut(L)$.
+\medskip
+\item{i)} If $G$ acts with finite orbits, then $L$ is a Galois extension of $L^G$.
+\smallskip
+\item{ii)} If $|G|$ is finite, then $[L:L^G] = |G|$ and $G$ is the Galois group $\Gal(L/L^G)$.
+\medskip
+
+\proof Let $\alpha \in L\setminus L^G$ and let $\{\alpha_1,\ldots,\alpha_n\}$ be the orbit under the $G$-action. Then
+$$p(x) = \prod_{i=1}^n (x-\alpha_i)$$
+is $G$-invariant and $p\in L^G[x]$ is separable. $L$ is the splitting field of $p$ so $L$ is a Galois extension of $L^G$.
+
+To prove (ii), suppose that $G$ is finite and let $n=|G|$.
+Take $\alpha\in L\setminus L^G$ and since $|G\alpha|\leq n$ for any $\alpha\in L$, $[L^G(\alpha) : L^G] \leq n$.
+We use this in our proof that $[L:L^G] \leq n$.
+Take $\alpha\in L$ such that $[L^G(\alpha):L^G]$ is maximal and let $\beta\in L$.
+Then $L^G(x,y)$ is a finite extension.
+By the primitive element theorem, $L^G(\alpha,\beta) = L^G(\gamma)$ for some $\gamma\in L$.
+But by the maximality of $\alpha$, we have $[L^G(\alpha):L^G]\geq L^G(\gamma):L^G]$ so $L^G(\alpha) = L^G(\gamma)$.
+This means that $\beta\in L^G(\alpha)$ and since our choice of $\beta$ was arbitrary, $L = L^G(\alpha)$.
+In particular, $[L:L^G]\leq n$.
+
+Now if $[L:L^G] < n$, then $L$ cannot have $L$ automorphisms over $L^G$. But $G$ is a subgroup of $\Aut(L/L^G)$ with $n$ elements. So $[L:L^G] = n$ and $G = \Aut(L/L^G)$.\slug
+
 \bye

### 2020-04-21 misc/study-notes/math457.tex
Finished Algebra 4 notes

@@
 
-\floattext5 \noindent {\ninebf Note.} \ninept These notes are rough and may skip over some details. \ninept Some proofs are either omitted or distilled to their main ideas.
+\floattext5 \noindent {\ninebf Note.} \ninept These notes form a large subset of the content covered by the lectures/notes given over the semester.
+\ninept Some proofs are either skipped or distilled to their main ideas.
+\ninept Note that because of the coronavirus outbreak, some concepts that are normally covered in Algebra 4 were omitted this year, e.g.\ solvability by radicals.
+\ninept I take responsibility for any errors that may be present.
 
@@
 
-The isomorphism theorems for groups extend to rings as well.
+The isomorphism theorems for groups extend to rings as well. Their proofs are analogous to the ones in group theory.
 
@@
 
+This theorem tells us that finite groups give field extensions, in the following sense.
+
+\proclaim Corollary G. If $G$ is a finite group, then there exists an extension $L/K$ with $G = \Aut(L/L^G)$.
+
+\proof By the fixed field theorem, it suffices to represent $G$ as a group of automorphisms over some field $L$.
+Let $\{x_s\}_{s\in G}$ be a set of $|G|$ indeterminate variables over some field $F$.
+Then $G = \{e,s,t,\ldots\}$ acts on the polynomial ring $F[x_e,x_s,x_t,\ldots]$ by acting on the variables:
+$$sx_t = x_{st}$$
+The action extends to monomials as well; for a subset $T\subseteq G$,
+$$s\Big(\prod_{t\in T} x_t\Big) = \prod_{t\in T} x_{st}$$
+and the action of $G$ on $F[x_s : s\in G]$ is by ring automorphisms.
+We can extend the action to the fraction field $L = F(x_s : s\in G)$ and the result is a faithful action by field automorphisms.
+Thus $G\subseteq \Aut(L)$ and $L$ has degree $|G|$ over $L^G$.\slug
+
+A well-known open problem is the following: {\sl Given a finite group $G$, is it possible to to find an extension $L/\Q$ such that $G = \Aut(L/\Q)$ and $\Q = L^G$?
+Equivalently, is $G$ the Galois group of a polynomial with rational coefficients?} For example, the group of quaternions $Q_8$ is the Galois group of the polynomial
+$$f = x^8 - 72x^6 + 180x^4 - 144x^2 + 36$$
+and the group ${\rm PSL}_2(7)$ of order 168 (the second non-abelian simple group) is the Galois group of
+$$x^7 - 56x^6 + 609x^5 + 1190x^4 + 6536x^3 + 4536x^2 - 6804x - 5832 - tx^3(x + 1),$$
+where the parameter $t$ can be any integer $t\equiv 1\mod{31}$.
+
+\proclaim Proposition S. If $f\in \Q[x]$ is irreducible of prime degree $p$ with exactly two roots in $\C\setminus \R$, then the Galois group of $f$ is $S_p$.
+
+\proof Complex conjugation is a non-trivial Galois automorphism, which acts as a transposition since there are two non-real roots.
+We just need to show that there exists a $p$-cycle.
+But the number $p$ of roots of $f$ divides the order of the Galois group, so by Cauchy's theorem we have an element of order $p$.
+This must be a $p$-cycle and we have generated $S_p$.\slug
+
+\section 10. The Galois Correspondence
+
+If $L/K$ is a Galois extension, then $\alpha\in L\setminus K$ must be displaced by some automorphism. The converse is also true, giving us the following theorem.
+
+\parenproclaim Theorem F (Galois' fixed field theorem). Let $L/K$ be an algebraic extension with automorphism group $G = \Aut_K(L)$. Then $L/K$ is a Galois extension if and only if $K = L^G$.
+
+\proof For the forward direction, let $\alpha\in L^G$ have minimal polynomial $f$ and splitting field $M$.
+Because $M$ is normal, we assume that $M\subseteq L$.
+If $\beta\in M$ is another root of $f$, then we can find $\theta\in \Aut_K(M)$ such that $\theta(\alpha) = \beta$.
+Extend $\theta$ to an element in $\Aut_K(L)$.
+Since $\alpha\in L^G$, we have $\beta = \theta(\alpha)  = \alpha$, i.e.\ $f$ has only one root.
+Since $L$ is separable, $f$ has degree 1, so $\alpha\in K$.
+
+For the converse, assume that $K = L^G$. Let $\alpha\in L$ have minimal polynomial $f$ with set of roots $R\subseteq \bar L$, and let
+$$g = \prod_{\beta \in R\cap L} (x - \beta).$$
+Note that $g$ is separable and invariant under every $\theta\in \Aut_K(L)$, since $\theta(R\cap L) = R\cap L$.
+This implies that the coefficients of $g$ are fixed by $G = \Aut_K(L)$.
+Since $L^G = K$, $g\in K[x]$ and since $g(\alpha) = 0$, it divides the minimal polynomial.
+Thus $f =g$ and all the roots of $f$ are in $L$, proving normality.
+Lastly, $f$ is separable because $g$ is.\slug
+
+The Galois correspondence was used by Galois to compute Galois groups of polynomial equations. It comprises the following results:
+\medskip
+\item{i)} Equations that are solvable by radicals have solvable Galois groups.
+\smallskip
+\item{ii)} Conversely, if $\Gal(f)$ is solvable, then the roots of $f$ can be expressed as radicals.
+\smallskip
+\item{iii)} $A_5$ is the smallest group that is not solvable (so every equation of degree $\leq 4$ can be solved by radicals).
+\smallskip
+\item{iv)} $A_5$ is the Galois group of a polynomial of degree 5 with rational coefficients, which is therefore not solvable by radicals.
+\medskip
+If $L/K$ is a Galois extension, then $L/M$ is Galois for any intermediate extension $K\subseteq M\subseteq L$.
+
+\parenproclaim Theorem C (Classical Galois correspondence).
+Let $L/K$ be a finite Galois extension.
+Then there is a bijection $\phi$ from subgroups of $\Gal(L/K)$ and intermediate fields in $L/K$.
+This is explicitly given by $\phi(H) = L^H$ for $H\subseteq \Gal(L/K)$ and $\phi^{-1}(M) = \Gal(L/M)$ for an intermediate extension $K\subseteq M\subseteq L$.
+
+\proof From Artin's fixed field theorem and Theorem F we know that $|G| = [L:K]$ and that the map $\phi$ is injective.
+Let $M$ be an intermediate extension and let $H = \Gal(L/M)$.
+Observe that since $L/M$ is Galois, Theorem F asserts that $M = L^H$.
+Thus $\phi^{-1}$ is injective and we have a bijection.\slug
+
+The correspondence is order-reversing in that the bigger the subgroup $H$, the smaller the fixed field $L^H$.
+Another corollary is that a Galois extension can only have finitely many intermediate subfields and that the number of subfields is at most the number of subgroups in the Galois group.
+
+\proclaim Proposition N. Let $L/K$ be a Galois extension and $M$ be an intermediate field. The following are equivalent:
+\medskip
+\item{i)} $\Gal(L/M)$ is a normal subgroup of $\Gal(L/K)$.
+\smallskip
+\item{ii)} If $s\in \Gal(L/K)$, then $s(M) = M$.
+\smallskip
+\item{iii)} $M/K$ is a normal extension.
+\smallskip
+\item{iv)} $K = M^H$ where $H = \Aut_K(M)$.
+\smallskip
+\item{v)} If $M/K$ is a finite extension, then $[M:K] = |\Aut_K(M)|$.
+\smallskip
+\item{vi)} The map $\Gal(L/K)\to \Gal(M/K)$ that maps $s\mapsto s_{|M}$ is surjective with kernel $\Gal(L/M)$.\noskipslug
+\medskip
+
+Let $L/K$ be a Galois extension.
+Let $K\subseteq M,N\subseteq L$ be intermediate subfields.
+The {\it compositum} $MN$ of $M$ and $N$ is the subfield of $L$ that they generate, i.e.\ the smallest subfield of $L$ containing them.
+It can be shown that
+$$MN = \bigg\{\sum_{i=1}^m \sum_{j=1}^n \alpha_i\beta_j : \alpha_1,\ldots,\alpha_m \in M, \beta_1,\ldots,\beta_n\in N\bigg\}.$$
+
+\proclaim Proposition I. If $M/K$ and $N/K$ are Galois, then so is the compositum $MN$.
+
+\proof It suffices to show normality.
+Suppose that $M$ and $N$ are finite normal extensions.
+Write $M = K(R_1)$ and $N=K(R_2)$ as splitting fields of two polynomials $f_1$ and $f_2$ with respective root sets $R_1$ and $R_2$.
+Then the splitting field of the product $f = f_1f_2$ is $K(R_1\cup R_2)$, which is the compositum $MN = K(R_1)K(R_2)$.
+If $M$ and $N$ are infinite, we can reuse the argument by writing $M$ and $N$ as splitting fields of families $F_1$ and $F_2$ of polynomials.\slug
+
+The Galois correspondence interchanges intersection and compositum, as shown by the following proposition.
+
+\proclaim Proposition S. Let $L/K$ be a finite Galois extension and let $G$ and $H$ be subgroups of $\Gal(L/K)$. Then
+\medskip
+\item{i)} $L^G \cap L^H = L^{\langle G,H \rangle}$;
+\smallskip
+\item{ii)} $L^GL^H = L^{G\cap H}$.
+\medskip
+
+\proof This is a corollary of Theorem C.
+Because $\langle G,H\rangle$ is the smallest subgroup containing both $G$ and $H$, it corresponds to the largest subfield $L^G\cap L^H$ of both $L^G$ and $L^H$.
+Likewise, the compositum $L^GL^H$ is the smallest subfield containing both $L^G$ and $L^H$, so it corresponds to the largest subgroup $G\cap H$ of both $G$ and $H$.\slug
+
+If $M$ and $N$ are Galois then $MN$ is Galois and we have a homomorphism
+$$f : \Gal(MN/K) \to \Gal(M/K) \times \Gal(N/K)$$
+that maps $\sigma\mapsto (\sigma_{|M},\sigma_{|N})$.
+It can be shown that $\ker f = \{e\}$ and $\Im f$ consists of pairs $(s,t)$ for which $s_{|M\cap N} = t_{|M\cap N}$.
+In terms of polynomials, if $f = f_1f_2$, the Galois group of $f$ is the subgroup of $\Gal(f_1)\times \Gal_(f_2)$ consisting of pairs of permutations that coincide on $K(R_1)\cap K(R_2)$, where $R_1$ and $R_2$ are the root sets of $f_1$ and $f_2$.
+
+Thus, if $L/K$ is a Galois extension such that $L = MN$ for normal extensions $M/K$ and $N/K$, and furthermore $M\cap N = K$, then $\Gal(L/K)$ can be expresed as the direct product
+$$\Gal(L/K)\cong \Gal(M/K) \times \Gal(N/K).$$
+
 \bye

### 2020-04-23 misc/study-notes/math457.tex
K infinite
Some mistakes in Algebra 4

@@
 For a unital commutative ring $R$, we define the {\it nilradical} of $R$ to be the intersection $\Nil(R) = \bigcap P$ of all prime ideals $P$.
-The {\it Jacobean radical} is the intersection $\Jac(R) = \bigcap M$ of all maximal ideals $M$.
+The {\it Jacobson radical} is the intersection $\Jac(R) = \bigcap M$ of all maximal ideals $M$.
 Since $\Specmax(R) \subseteq \Spec(R)$, $\Jac(R) \supseteq \Nil(R)$.
@@
 
-\proof. Let $p\in R$ be a prime element. Assume that $p=ab$. This implies that $p\setminus a$ or $p\setminus b$. Say $a = pc$ for some $c\in R$. Then $p=ab=pcb$ and $cb = 1$. So $b$ is a unit.\slug
+\proof Let $p\in R$ be a prime element. Assume that $p=ab$. This implies that $p\setminus a$ or $p\setminus b$. Say $a = pc$ for some $c\in R$. Then $p=ab=pcb$ and $cb = 1$. So $b$ is a unit.\slug
 
@@
 
-\proof.
-Let $a,b\in \Z[i]$, $b\neq 0$.
+\proof Let $a,b\in \Z[i]$, $b\neq 0$.
 We can divide them in $\C$ to get $z = a/b$.
@@
 \smallskip
-\item{ii)} $(r+s)m) = rm + sm$;
+\item{ii)} $(r+s)m = rm + sm$;
 \smallskip
@@
 
-\proof We can suppose that $L$ is infinite, since otherwise $K$ is finite and we can let $\gamma$ be a primitive root of unity that generates $K^\times$.
+\proof We can suppose that $K$ is infinite, since otherwise we can let $\gamma$ be a primitive root of unity that generates $K^\times$.
 Moreover, it suffices to show the theorem for two elements $\alpha$ and $\beta$, with $\beta$ separable, since a simple induction will extend the result to arbitrary $n$.

### 2020-05-30 about/vita.tex
Added CV

% Marcel Goh CV

\font\elevensc=cmcsc11
\font\bigbold=cmbx12

\def\sectheader#1{{\bigskip\noindent\elevensc #1}\smallskip\hrule\medskip}
\def\leftright#1#2{\noindent{#1}\hfill{#2}\par}

\leftright{\bigbold Marcel Goh}{3660 Rue Hutchison, Apt.\ 4}
\leftright{Email: marcel\_goh@yahoo.ca}{Montr\'eal, QC H2X 2H3, Canada}
\leftright{URL: marcelgoh.github.io}{+1$\;$(438)$\;$886-2338}

\sectheader{Education}

\leftright{\bf McGill University}{\bf September 2017 -- May 2021}
\leftright{\sl BSc, Honours Mathematics and Computer Science, minor Linguistics}{\sl Montr\'eal, QC, Canada}
\noindent CGPA: 3.79/4.
\medskip
\leftright{\bf Charles University}{\bf February -- June 2019}
\leftright{\sl Exchange semester, Faculty of Mathematics and Physics}{\sl Prague, Czech Republic}
\noindent Grade: 1 on all courses (highest score attainable).

\sectheader{Awards}

\leftright{\bf NSERC Undergraduate Student Research Award}{\bf May 2020}
\smallskip
\leftright{\bf Governor General's Academic Medal -- Bronze}{\bf June 2015}

\sectheader{Research Experience}

\leftright{\bf Sorting algorithms on manifolds}{\bf May 2020 -- present}
\leftright{\sl Summer research project}{\sl McGill University}
\smallskip
\noindent Summer research project under the supervision of Michael Lipnowski.
Studied algorithms that investigate the topology of group actions on locally symmetric spaces.
Work involved writing of code in OCaml and PostScript.
\medskip

\leftright{\bf Interactive proofs}{\bf September -- December 2019}
\leftright{\sl Honours research project}{\sl McGill University}
\smallskip
\noindent Semester-long research project in the Computation and Logic Group, supervised
by Brigitte Pientka.
Proved a theorem in constructive logic concerning the formal verification
of an interactive proof assistant and wrote OCaml code as part of
ongoing work on the functional programming language Beluga.
\medskip

\leftright{\bf Bytecode compiler and interpreter}{\bf February -- June 2019}
\leftright{\sl Individual software project}{\sl Charles University}
\smallskip
\noindent Semester-long individual software project under the supervision of Adam Dingle.
Created OPythn, a bytecode compiler and virtual machine for a subset of Python.
OPythn includes support for lists, dictionaries, named and anonymous functions, objects and classes.
It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing.
\medskip

\sectheader{Work and Volunteer Experience}

\leftright{\bf Department of Mathematics and Statistics, McGill University}{\bf September 2019 -- present}
\leftright{\sl Student Grader}{\sl Montr\'eal, QC, Canada}
\smallskip
\noindent Graded MATH 235, an introductory course on the theory of groups, rings, and fields, in Fall 2019
and MATH 240, a discrete mathematics course, in Winter 2020.
\medskip

\leftright{\bf Computer Science Undergraduate Society Helpdesk}{\bf September 2018 -- present}
\leftright{\sl Tutor}{\sl Montr\'eal, QC, Canada}
\smallskip
\noindent Hold twice-weekly office hours to tutor students in a variety of undergraduate courses.
Topics covered include elementary data structures and algorithms, command-line scripting,
and functional programming. Recipient of the Tomlinson Engagement Award for Mentoring.
\medskip

\filbreak % Break page here

\leftright{\bf McGill Student Emergency Response Team}{\bf October 2017 -- present}
\leftright{\sl First Responder}{\sl Montr\'eal, QC, Canada}
\smallskip
\noindent On call on a weekly basis to provide emergency medical care at campus residences overnight as well as at
university events such as frosh, sports games, and formals. Attend team training sessions twice a month to keep
first-aid skills up-to-date.
\medskip

\leftright{\bf Society of Undergraduate Mathematics Students}{\bf May 2019 -- January 2020}
\leftright{\sl Vice President, Academic}{\sl Montr\'eal, QC, Canada}
\smallskip
\noindent Oversaw academic affairs within SUMS council and acted as liaison between the undergraduate community and
mathematics faculty. Duties included representing the student body at department meetings and helping students with
academic concerns.
\medskip

\leftright{\bf Singapore Armed Forces}{\bf August 2015 -- August 2017}
\leftright{\sl Infantryman}{\sl Singapore}
\smallskip
\noindent Held appointment of machine-gun team commander in the 3rd Battalion, Singapore Infantry Regiment.
Led a six-person team consisting of a medic, signaller, sensor, and two-machine gunners within a rifle platoon.
\medskip

\sectheader{Skills}

\noindent{\bf Programming Languages}\par
\noindent OCaml, C, Haskell, Standard ML, Scheme, PostScript, Assembly, Python
\medskip

\noindent{\bf Technologies}\par
\noindent UNIX, Vim, \TeX, Git.
\medskip

\noindent{\bf Languages}\par
\noindent Fluent: English, French. Proficient: Mandarin, Italian.
\medskip

\sectheader{Interests}

\noindent Analysis and design of algorithms, analysis of random discrete structures, combinatorics, group theory,
programming language theory and design, compilers.

\sectheader{Other}

\noindent Recipient of 0x\$0.40 in Knuth reward cheques.

\bye

### 2020-05-31 about/vita.tex
Cleaned up CV tex

@@
 
-\font\elevensc=cmcsc11
-\font\bigbold=cmbx12
+\font\elevensc=cmcsc11  % For section headers
+\font\bigbold=cmbx12  % Name
 
-\def\sectheader#1{{\bigskip\noindent\elevensc #1}\smallskip\hrule\medskip}
-\def\leftright#1#2{\noindent{#1}\hfill{#2}\par}
+\parindent=0pt  % No indentation
 
-\leftright{\bigbold Marcel Goh}{3660 Rue Hutchison, Apt.\ 4}
+\def\sectheader#1{{\bigskip\elevensc #1}\smallskip\hrule\medskip}
+\def\leftright#1#2{{#1}\hfill{#2}\par}
+
+\leftright{\bigbold Marcel K.\ Goh}{3660 Rue Hutchison, Apt.\ 4}
 \leftright{Email: marcel\_goh@yahoo.ca}{Montr\'eal, QC H2X 2H3, Canada}
@@
 \leftright{\sl BSc, Honours Mathematics and Computer Science, minor Linguistics}{\sl Montr\'eal, QC, Canada}
-\noindent CGPA: 3.79/4.
+CGPA: 3.79/4.
 \medskip
@@
 \leftright{\sl Exchange semester, Faculty of Mathematics and Physics}{\sl Prague, Czech Republic}
-\noindent Grade: 1 on all courses (highest score attainable).
+Grade: 1 on all courses (highest score attainable).
 
@@
 \smallskip
-\noindent Summer research project under the supervision of Michael Lipnowski.
+Summer research project under the supervision of Michael Lipnowski.
 Studied algorithms that investigate the topology of group actions on locally symmetric spaces.
@@
 \smallskip
-\noindent Semester-long research project in the Computation and Logic Group, supervised
+Semester-long research project in the Computation and Logic Group, supervised
 by Brigitte Pientka.
@@
 \smallskip
-\noindent Semester-long individual software project under the supervision of Adam Dingle.
+Semester-long individual software project under the supervision of Adam Dingle.
 Created OPythn, a bytecode compiler and virtual machine for a subset of Python.
@@
 \smallskip
-\noindent Graded MATH 235, an introductory course on the theory of groups, rings, and fields, in Fall 2019
+Graded MATH 235, an introductory course on the theory of groups, rings, and fields, in Fall 2019
 and MATH 240, a discrete mathematics course, in Winter 2020.
@@
 \smallskip
-\noindent Hold twice-weekly office hours to tutor students in a variety of undergraduate courses.
+Hold twice-weekly office hours to tutor students in a variety of undergraduate courses.
 Topics covered include elementary data structures and algorithms, command-line scripting,
@@
 \smallskip
-\noindent On call on a weekly basis to provide emergency medical care at campus residences overnight as well as at
+On call on a weekly basis to provide emergency medical care at campus residences overnight as well as at
 university events such as frosh, sports games, and formals. Attend team training sessions twice a month to keep
@@
 \smallskip
-\noindent Oversaw academic affairs within SUMS council and acted as liaison between the undergraduate community and
+Oversaw academic affairs within SUMS council and acted as liaison between the undergraduate community and
 mathematics faculty. Duties included representing the student body at department meetings and helping students with
@@
 \smallskip
-\noindent Held appointment of machine-gun team commander in the 3rd Battalion, Singapore Infantry Regiment.
+Held appointment of machine-gun team commander in the 3rd Battalion, Singapore Infantry Regiment.
 Led a six-person team consisting of a medic, signaller, sensor, and two-machine gunners within a rifle platoon.
@@
 
-\noindent{\bf Programming Languages}\par
-\noindent OCaml, C, Haskell, Standard ML, Scheme, PostScript, Assembly, Python
+{\bf Programming Languages}\par
+OCaml, C, Haskell, Standard ML, Scheme, PostScript, Assembly, Python
 \medskip
 
-\noindent{\bf Technologies}\par
-\noindent UNIX, Vim, \TeX, Git.
+{\bf Technologies}\par
+UNIX, Vim, \TeX, Git.
 \medskip
 
-\noindent{\bf Languages}\par
-\noindent Fluent: English, French. Proficient: Mandarin, Italian.
+{\bf Languages}\par
+Fluent: English, French. Proficient: Mandarin, Italian.
 \medskip
@@
 
-\noindent Analysis and design of algorithms, analysis of random discrete structures, combinatorics, group theory,
+Analysis and design of algorithms, analysis of random discrete structures, combinatorics, group theory,
 programming language theory and design, compilers.
@@
 
-\noindent Recipient of 0x\$0.40 in Knuth reward cheques.
+Recipient of 0x\$0.40 in Knuth reward cheques.
 

### 2020-07-11 misc/expo/genfcts.tex
Added generating functions crash course

\input fontmac
\input mathmac

\def\Seq{\op{\tensc Seq}}
\def\Set{\op{\tensc Set}}
\def\Cyc{\op{\tensc Cyc}}
\def\A{{\cal A}}
\def\B{{\cal B}}
\def\C{{\cal C}}
\def\D{{\cal D}}
\def\E{{\cal E}}
\def\F{{\cal F}}
\def\I{{\cal I}}
\def\M{{\cal M}}
\def\P{{\cal P}}
\def\T{{\cal T}}
\def\Z{{\cal Z}}
\def\bul{\bullet}

\def\eval{\big|}

\def\beginref{\noindent}
\def\endref{\smallskip}
\input cyracc.def
    \font\tencyr=wncyr10
    \font\tencyri=wncyi10
    \def\cyr{\tencyr\cyracc}
    \def\cyri{\tencyri\cyracc}

\maketitle{A crash course on generating functions}{by}{Marcel K. Goh}{10 July 2020}

\section\the\sectcount. Introduction

When counting discrete structures, we often run into sequences that are recursively defined.
For example, suppose we have the recurrence
$$R_n = \cases{1,&if $n=0$; \cr 2R_{n-1}+3, & if $n\geq 1$.}$$
We might want to know how the sequence $(R_n)$ grows as $n$ gets large.
Now, this particular recurrence
is easy enough to solve by the substitution method. But often, it is easier to work with the
{\it generating function} of a sequence. For a sequence $(A_n)$, this is the formal power series
$$A(z) = \sum_{n\geq 0} A_nz^n.$$
In his textbook, titled {\sl generatingfunctionology},
H. S. Wilf famously wrote: ``A generating function is a clothesline on which we hang up a sequence
of numbers for display.'' That is exactly how we should treat the function $A(z)$ for now;
in later sections, we will discuss how this series can be treated as an analytic object.

Let us try to find the generating function for the sequence $(R_n)$ defined above. Right off the bat, we have
$$R(z) = \sum_{n\geq 0} R_nz^n = R_0 + \sum_{n\geq 1} R_nz^n.$$
Plugging in the recursive definition of $R_n$, we get
\newcount\cgenfcteq
\cgenfcteq=\eqcount
$$\eqalign{
R(z) &= 1 + \sum_{n\geq 1} (2R_{n-1}+3)z^n \cr
&= 1 + 2\sum_{n\geq 1} R_{n-1}z^n + 3\sum_{n\geq 1} z^n \cr
&= 1 + 2z\sum_{n\geq 0} R_nz^n + 3\Big(\sum_{n\geq 0} z^n - 1\Big) \cr
}\adveq$$
The first summation on the right-hand side is $R(z)$ and the second one is well-known,
so we derive the generating function
$$\eqalign{
R(z) &= 1+ 2zR(z) + 3\bigg({1\over 1-z} - 1\bigg) \cr
(1-2z)R(z) &= {1+2z\over 1-z} \cr
R(z) &= {1+2z \over (1-z)(1-2z)}. \cr
}$$
Here we can use the partial fraction expansion to get
$$\eqalign{
R(z) &= {1+2z\over (1-z)(1-2z)} = {4\over 1-2z} - {3\over 1-z} \cr
&= 4\sum_{n\geq 0} (2z)^n - 3\sum_{n\geq 0} z^n \cr
&= \sum_{n\geq 0}(2^{n+2} - 3)z^n, \cr
}\adveq$$
implying that $R_n = 2^{n+2} -3$. This could have been found by easier methods, but this example
demonstrates how the generating function of a sequence provides a bridge from the recurrence to a closed
form. In later sections, we will see that if we only want asymptotic estimates, then we can actually
stop at the generating function equation found in \refeq{\the\cgenfcteq}.

\advance\sectcount by 1
\section\the\sectcount. The Symbolic Method and Ordinary Generating Functions

Deriving generating functions is a rather drawn-out process, and it would not be very enjoyable
to have to work from scratch every time. Luckily, generating functions are often made up of simpler
ones, so we can use the {\it symbolic method} to arrive directly at a generating function equation.
This is the subject of the monolithic textbook {\sl Analytic Combinatorics}, by P. Flajolet and R. Sedgewick.
In these notes, we will often skip some details and proofs; anything we gloss over here can certainly
be found somewhere in {\sl Analytic Combinatorics}. We will refer to this book from time to time and abbreviate
it by `AC'.

We define a {\it combinatorial class} $\A$ to be a countable set with a {\it size function}
$|\cdot|:\A\to \N_0$ for which the number of elements of any given size is finite. Here are examples
of combinatorial classes:
\medskip
\item{i)} The set of all strings of 0 and 1, with size given by string length.
\smallskip
\item{ii)} The set of all permutations. A permutation of size $n$ is a one-to-one correspondence
\smallskip
\item{iii)} The set of all binary trees, with size given by the number of nodes.
from $[1\twodots n]$ to itself.
\medskip

\boldlabel Ordinary generating functions. The {\it ordinary generating function}, or OGF, of a combinatorial
class $\A$ is defined to be the formal power series
$$A(z) = \sum_{a\in \A} z^{|a|} = \sum_{n\geq 0} A_nz^n,$$
where $A_n$ is the number of elements in $\A$ of size $n$. It is easy to see that the two summations
are equivalent characterisations of $A(z)$, and $A_n = [z^n]A(z)$ is the coefficient of $z^n$ in $A(z)$.
We now define the {\it neutral class} $\E$ containing only one element, of size 0.
Thus the generating function of $\E$ is 1.
Likewise, we define the {\it atomic class} $\Z$, which contains a single element of size 1.
The generating function of $\Z$ is $z$.

When classes $\A$ and $\B$ are disjoint, we will denote their {\it union} by
$$\A + \B = \A\cup\B.$$
It is easy to see that the generating function of $\A + \B$ is $A(z) + B(z)$.
In a similar vein, let the {\it product} of two combinatorial classes be given by
$$\A\times\B = \{(a,b) : a\in \A,\; b\in B\};$$
the generating function of $\A\times\B$ is $A(z)\cdot B(z)$. We will denote the product of
$n$ copies of $\A$ by $\A^n$ (this is $\E$ if $n=0$), which has the generating function $A(z)^n$.

A more complex construction is the {\it sequence class}. For a class $\A$, this is
$$\Seq(\A) = \E + \A + (\A\times\A) + (\A\times\A\times\A) + \cdots$$
It follows that the generating function of $\Seq(\A)$ is
$$S(z) = 1 + A(z) + A(z)^2 + A(z)^3 + \cdots = {1\over 1-A(z)}.$$
More generally, if $\Omega$ is a subset of $\N_0$, we define $\Seq_\Omega(\A)$ as the class
$\sum_{n\in \Omega} A^n$, with generating function $\sum_{n\in\Omega} A(z)^n$.
A list of further constructions can be found in Fig.~I.18 of AC.
\medskip

\boldlabel Integers and compositions. The set of all positive integers is a combinatorial class.
To get its generating function, we note that the positive integers can be associated with nonempty sequences
of unlabelled atoms:
$$\{1,2,3,\ldots\} \cong \{\bul,\bul\;\bul,\bul\bul\bul,\ldots\}$$
This gives the construction ${\cal I} = \Seq_{\geq 1}(\Z)$, which immediately tells us that
$I(z) = z/(1-z)$ and $I_n = 1$ for all $n\geq 1$.

A {\it composition} of an integer $n$ is expression of $n$ as a sum of a sequence of positive integers
(so the order of the terms matters). For example, there are four compositions of the number 3:
\newcount\threecomp
\threecomp=\eqcount
$$1 + 1 + 1\qquad1 + 2\qquad2 + 1\qquad3\adveq$$
Let $\C$ be the set of all integer compositions. We can describe a composition as a sequence of integers, so
$$\C = \Seq(\I).$$
From this, we find that
$$C(z) = {1\over 1 - z/(1-z)} = {1-z\over 1-2z}.\adveq$$
Reexpressing this as the series
$$C(z) = {1\over 1-2z}-{z\over 1-2z}
       =\sum_{n\geq 0} (2z)^n - \sum_{n\geq 0} 2^nz^{n+1}= 1+\sum_{n\geq 1} 2^{n-1}z^n,$$
we find that $C_0 = 1$ and $C_n = 2^{n-1}$ for $n\geq 1$. Ignoring $C_0$,
this makes sense because there is a one-to-one
correspondence between compositions and a subset of $n-1$ possible bars between $n$ balls:
the compositions in \refeq{\the\threecomp} correspond to
$$\bul\mid\bul\mid\bul\qquad \bul\mid\bul\bul\qquad \bul\;\bul\mid\bul\qquad \bul\bul\bul.$$
\medskip

\boldlabel Fibonacci numbers. Let $\I^*$ denote the cute combinatorial class $\{1,2\}$,
which clearly has the generating function $I^*(z) = z + z^2$. Let $\T$ be the set of partitions
using only the numbers 1 and 2 (for simplicity's sake, we also include the neutral object, i.e.\ $T_0=1$).
We have
$$\eqalign{
1 &= 1, \cr
2 &= 1+1 = 2,\cr
3 &= 1 + 1 + 1 = 1 + 2 = 2+1, \cr
4 &= 1 + 1 + 1 + 1 = 1 + 1 + 2 = 1 + 2 + 1 = 2 + 1 + 1 = 2 + 2,\cr
}$$
and so on. After computing a few more terms, we notice the pattern
$$(T_n)_{n\geq 0} = (1,1,2,3,5,8,13,\ldots).$$
This is almost the infamous Fibonacci sequence; we have $T_n = F_{n+1}$.
Hence $T(z)$ must be the generating function
of the Fibonacci numbers, divided by $z$.
We can easily derive this function with the symbolic method, since a composition
using only 1s and 2s is simply a sequence of elements of $\I^*$. We conclude that
$$\T = \Seq(\I^*)\qquad\hbox{and}\qquad T(z) = {1\over 1-I^*(z)} = {1\over 1-z-z^2}.\adveq,$$
so we know that the generating function of the Fibonacci numbers is
\newcount\fibeq
\fibeq=\eqcount
$$F(z) = {z\over 1-z-z^2}.\adveq$$
\medskip

\boldlabel Catalan numbers. Let $\B$ be the class of all binary trees. This combinatorial class
can be described recursively as follows: ``A binary tree is either empty, or else it consists of
a root node adjoined to two more binary trees.''. An empty binary tree has size 0, so it is represented
by $\E$ and a node has size $1$, so we represent it by $\Z$. We arrive at the symbolic description
$$\B = \E + \Z\times\B\times\B,$$
whence we derive a functional equation
$$B(z) = 1 + zB(z)^2$$
that the generating function must satisfy. Treating $B(z)$ as a variable and employing
the quadratic formula, we find that
\newcount\catalaneq
\catalaneq=\eqcount
$$B(z) = {1 - \sqrt{1-4z}\over 2z}.\adveq$$
(Since we know that $B(0) = B_0 = 1$, the positive branch of the square root is invalid.
On the other hand, l'Hospital's rule can be applied to the negative version to get a limit of 1.)
From here, one could use Newton's generalisation of the binomial theorem to get that
\newcount\catalanformula
\catalanformula=\eqcount
$$B_n = {1\over n+1}{2n\choose n},\adveq$$
and indeed, $B(z)$ is the generating function of the Catalan numbers. From here, Stirling's approximation
can be applied to give asymptotics, but the techniques of later sections will allow us to characterise
the asymptotic growth directly from the generating function \refeq{\the\catalaneq}.
\medskip

\boldlabel Making change. Suppose a cashier has exactly
five loonies, three toonies, and four \$5 bills in her till.
How many ways can she give \$23 in change? To solve this, we assign each coin/bill
a size proportional to its value;
so a loonie is described by $\Z$, a toonie by $\Z^2$, and a \$5 bill by $\Z^5$.
Now we can describe the set of all possiblities with the combinatorial class
$$\M = \Seq_{\leq 5}(\Z) \times \Seq_{\leq 3}(\Z^2) \times \Seq_{\leq 4}(\Z^5).$$
This immediately gives the generating function
$$M(z) = (1 + z + z^2 + z^3 + z^4 + z^5)(1 + z^2 + z^4 + z^6)(1 + z^5 + z^{10} + z^{15} + z^{20}),$$
and the number of ways to make \$23 is exactly $[z^{23}]M(z)$. Note that if the cashier had
an unlimited supply of coins and bills, we would have
$$\M^* = \Seq(\Z) \times \Seq(\Z^2) \times \Seq(\Z^5).$$
and the number of ways to make $n$ dollars out of these denominations is
$$[z^n]\bigg({1\over 1-z}\bigg)\bigg({1\over 1-z^2}\bigg)\bigg({1\over 1-z^5}\bigg).$$
The change-making problem was used as a primary example in G. P\'olya's 1956 paper on the applications
of generating functions.

\advance\sectcount by 1
\section\the\sectcount. Labelled Structures and Exponential Generating Functions

Ordinary generating functions deal with atoms that are unlabelled, meaning these atoms are indistinguishable
from one another.
When we want to count labelled objects, we will use an {\it exponential generating function}, or EGF.
For a combinatorial class $\A$, this is the formal power series
$$A(z) = \sum_{a\in \A} {z^{|a|} \over |a|!} = \sum_{n\geq 0} A_n {z^n\over n!}.$$
In particular, $A_n = [z^n/n!]A(z)$ for any EGF $A(z)$.

The symbolic method applies to EGFs as well. As before, $\E$ denotes the neutral class, with a single object
of size 0 and $\Z$ denotes the atomic class, with a single labelled object of size 1.
If $\A$ and $\B$ are disjoint combinatorial classes with
EGFs $A(z)$ and $B(z)$, then the EGF of their union $\A + \B$ is $A(z) + B(z)$. Instead of an ordinary
Cartesian product, we have the notion of a {\it labelled product}, in which atoms are relabelled in all
consistent ways. For example, a valid relabelling of the labelled triple $(1,3,2)$ is $(2,7,5)$,
because the relative ordering of the atoms is unchanged. The set of all labelled pairs of elements from
$\A$ and $\B$ is denoted $\A\star \B$ and its EGF is $A(z)\cdot B(z)$.

The labelled sequence class $\Seq(\A)$, consists of sequences of elements of $\A$. It has the EGF
$1/(1-A(z))$, just as with the OGF, but notice that the EGF counts a lot more things.
The easiest way to see this is to consider
the generating function of $\Seq(\Z)$, which is $S(z) = 1/(1-z)$. When considered as an OGF, there is only one
element of each given size, since $[z^n]S(z) = 1$ for all $n$. But when taken as an EGF, we find that there are
$n!$ objects of size $n$, because we have $[z^n/n!]S(z) = n!$ for all $n$.
Of course, this makes sense since permuting the atoms
in a sequence of length $n$ creates a new labelled object, but does not change the unlabelled one.

There are two more constructions that will be useful to define on labelled classes. First, we let
$\Set_k(\A)$ denote the class of all sets of size $k$ with elements in $\A$. This is like a sequence of
length $k$, except that the order is not important, meaning we can divide by a factor of $k!$.
The generating function of $\Set_k(\A)$ is therefore $A(z)^k/k!$. Now we let
$$\Set(\A) = \E + \A + \Set_2(\A) + \Set_3(\A) + \cdots$$
and derive the generating function for $\Set(\A)$:
\newcount\setgenfct
\setgenfct=\eqcount
$$T(z) = \sum_{k\geq 0} {A(z)^k \over k!} = e^{A(z)}\adveq$$
The last construction we will handle is the class of $k$-cycles with elements in $\A$, denoted $\Cyc_k(\A)$.
There are $k$ ways to represent a cycle as a sequence of $k$ objects, so the generating function of $\Cyc_k(\A)$
is $A(z)/k$. Now we define the class of unrestricted cycles
$$\Cyc(\A) = \Cyc_1(\A) + \Cyc_2(\A) + \Cyc_3(\A) + \cdots$$
and find that the generating function of $\Cyc(\A)$ is
$$U(z) = \sum_{k\geq 1} {A(z)^k \over k} = \log {1\over 1-A(z)}.\adveq$$
There are many other interesting EGF constructions; see, for example, Fig.\ II.18 in AC.

Note that a permutation is simply a set of cycles:
$$\Seq(\Z) = \Set(\Cyc(\Z))$$
Of course, the cycle decomposition of a permutation is well-known, but generating functions give us
another way of seeing that this is true, since
$${1\over 1-z} = \exp\bigg(\log{1\over 1-z}\bigg).$$

Now is as good a time as any to state, without proof, an important classic theorem that will be used in the next
example.

\parenproclaim Theorem L (Lagrange inversion theorem). If a generating function $g(z)$ satisfies the equation
$z = f(g(z))$, with $f(0) = 0$ and $f'(0) \neq 0$, then for an arbitrary function $H(u)$, we have
$$[z^n]H(g(z)) = {1\over n}[u^{n-1}]H'(u)\bigg({u\over f(u)}\bigg)^n.$$
In particular, for $H(u) = u$ we have
$$[z^n]g(z) = {1\over n}[u^{n-1}]\bigg({u\over f(u)}\bigg)^n.$$
(This particular form of of Lagrange's inversion theorem is often called the B\"urmann form.)\slug
\medskip

\boldlabel Cayley trees. We are now equipped to count rooted non-plane labelled trees, sometimes called
Cayley trees. Children of a given node are not ordered and nodes in these trees have unrestricted degree,
so we have the specification
$$\T = \Z \star \Set(\T).$$
This immediately gives the functional equation
$$T(z) = ze^{T(z)}$$
and by Lagrange's inversion theorem, with $f(u) = u/e^u$ and $H(u) = u$, we obtain
$$n![z^n]T(z) = n!\bigg({1\over n}[u^{n-1}]e^{un}\bigg)
              = n!\cdot{1\over n}\cdot{n^{n-1}\over(n-1)!} = n^{n-1}.$$
Note that there are $n^{n-2}$ {\it unrooted} Cayley trees, since there are $n$ choices for the root.
This is called Cayley's formula and the standard combinatorial proof involves establishing
a bijection with Pr\"ufer sequences; see Pr\"ufer (1918).
\medskip

\boldlabel Derangements. A derangement is a permutation in which no element is fixed. This can equally
be characterised as a permutation with no cycle of length 1. We saw earlier that
permutations is simply a set of cycles, so disallowing cycles of length 1 gives the specification
$$\D = \Set(\Cyc_{\geq 1}(\Z)).$$
We remove singleton cycles from the generating function by simply subtracting the generating function of
$\Cyc_1(\Z)$ (which is simply $z$, because a singleton cycle is just an atom) from $\Cyc_(\Z)$, so
$$D(z) = \exp\bigg(\log {1\over 1-z}- z\bigg) = {e^{-z}\over 1-z}.\adveq$$
We will revisit this generating function in the context of a word problem once we have the tools
to analyse its asymptotics.

\advance\sectcount by 1
\section\the\sectcount. Counting Parameters with Bivariate Generating Functions

Some situations may require us to do more than just count the number of objects of a certain size.
In addition, we might like to know how many of a certain substructure is embedded in an object
of a combinatorial class. We can use a {\it bivariate generating function} or BGF to accomplish
this. (BGFs are part of a larger class of {\it multivariate generating functions}, but we will
keep things simple in this crash course. See Chapter III of AC for many more constructions.)

Consider a two-dimensional array of numbers $(a_{n,k})$, where $n$ counts the number of objects
of a certain size and $k$ counts some other combinatorial parameter. For example, we might have
\medskip
\item{i)} The number of $n$-bit strings of 0s and 1s that have exactly $k$ 0s.
\smallskip
\item{ii)} The number of $n$-letter permutations that have exactly $k$ cycles.
\smallskip
\item{iii)} The number of binary trees of size $n$ that have $k$ leaves.
\medskip
Symbolically, suppose we have a combinatorial class $\A$ with an extra scalar parameter $\chi : \A \to \N_0$.
The ordinary and exponential BGFs are defined, respectively, to be the power series
$$\sum_{a\in\A} z^{|a|} u^{\chi(a)} \qquad\hbox{and}\qquad
           \sum_{a\in\A} {z^{|a|}\over n!} u^{\chi(a)},$$
which can be rewritten as
$$\sum_{n,k} a_{n,k} z^n u^k \qquad\hbox{and}\qquad \sum_{n,k} a_{n,k} {z^n\over n!} u^k,$$
where $a_{n,k}$ is the number of objects of size $a$ with $\chi(a) = k$. Note that substituting $u=1$ into
a BGF returns the ordinary counting sequence for the class $\A$.
\medskip

\boldlabel Bitstrings. To warm up, we will consider a very simple example. Let $\B$ be the class of all
binary strings, where for $b\in \B$, its size $n$ is given by its length and $\chi(b)$ is the number
of zeroes in the string. There are two distinguishable atoms, $\Z_0$ and $\Z_1$, but we introduce a $u$
only when a zero is present, so we have the specification
$$\B = \Seq(u\Z_0 + \Z_1)$$
with generating function
$$B(z,u) = {1\over 1-(u+1)z}.\adveq$$
When we substitute $u=1$, we get
$$B(z,1) = {1\over 1-2z},$$
which is the generating function of $2^n$, the number of binary strings with $n$ bits.
The number of $n$-bit binary strings with $k$ zeroes can now be extracted. It is equal to
$$[u^k][z^n]B(z,u) = [u^k](u+1)^n = {n\choose k},$$
by an application of the binomial theorem.
Hence the probability that an $n$-bit binary string has $k$ zeroes is
$${[u^k][z^n] B(z,u) \over [z^n] B(z,1)} = {{n\choose k}\over 2^n}.\adveq$$
\medskip

\boldlabel Probabilities and moments. The above example has an elementary result,
but it illustrates how BGFs allow us to extract probabilities.
For all $a$ in a class $\A$, we have
$$\pr\{\chi(a) = k\mid |a| = n\} = {[u^k][z^n] A(z,u) \over [z^n] A(z,1)}.\adveq$$
We can also calculate higher moments with the BGF. Notice that if we take the partial derivative
with respect to $u$, the $\chi$ values of each term pop down into the coefficient of each term:
$${\partial \over\partial u} A(z,u) = {\partial \over\partial u}\sum_{a\in \A} z^{|a|} u^{\chi(a)}
  = \sum_{a\in \A} \chi(a) z^{|a|} u^{\chi(a) -1}$$
Of course, the exponents of the variable $u$ have changed, but
letting $A_u(z,u)$ denote the partial derivative with respect to $u$ and evaluating at $u=1$, we get
the cumulated cost
$$A_u(z,u)|_{u=1} = \sum_{n\geq 0} \Big(\sum_{a\in \A_n} \chi(a) \Big)z^n,\adveq$$
whence we can calculate the mean over $\A_n$, the class of all $a\in \A$ of size $n$, to be
\newcount\firstmoment
\firstmoment=\eqcount
$$\ex_{\A_n}\{\chi\} = {[z^n] A_u(z,u)\eval_{u=1} \over [z^n]A(z,1)}.\adveq$$
These formulas are valid even when the generating functions are exponential, because the $n!$ factors cancel out.
More generally, by taking repeated derivatives, we can calculate factorial moments of $\chi$.
Let $A_u^r(z,u)$ denote the $r$th partial derivative of $A$ with respect to $u$. We have
$$\ex_{\A_n}\{\chi(\chi-1)\cdots(\chi-r+1)\} = {[z^n] A_u^r(z,u)\eval_{u=1} \over [z^n]A(z,1)}.\adveq$$
For example, by linearity of expectation, the second moment is easily seen to be
\newcount\secondmoment
\secondmoment=\eqcount
$$\eqalign{
\ex_{\A_n}\{\chi^2\} &= \ex_{\A_n}\{\chi + \chi(\chi-1)\} \cr
&= \ex_{\A_n}\{\chi\} + \ex_{\A_n}\{\chi(\chi-1)\} \cr
&= {[z^n] A_u(z,u)\eval_{u=1} \over [z^n]A(z,1)} + {[z^n] A_u^2(z,u)\eval_{u=1} \over [z^n]A(z,1)} \cr
}.\adveq$$
This can be easily used to calculate the variance and standard deviation of $\chi$, from the usual formula
$$\var_{\A_n}\{\chi\} = \sigma(\chi)^2 = \ex_{\A_n}\{\chi^2\} - \ex_{\A_n}\{\chi\}^2.\adveq$$
\medskip

\boldlabel Cycles in a permutation. Recall that a permutation is a set of cycles. If we mark every cycle
with a $u$, we get the specification
$$\P = \Set(u\Cyc(\Z)),$$
which gives the bivariate exponential generating function
$$P(z,u) = \exp\bigg(u\log{1\over 1-z}\bigg) = \exp\bigg(\log{1\over (1-z)^u}\bigg) = \bigg({1\over 1-z}\bigg)^u\adveq$$
for the number of $n$-letter permutations with exactly $k$ cycles. We can calculate the partial derivatives
$$P_u(z,u) = \bigg({1\over 1-z}\bigg)^u\log{1\over 1-z}$$
Here the random variable $\chi$ counts the number of cycles, and we have
$$\ex_{\P_n}\{\chi\} = {[z^n] P_u(z,u)\eval_{u=1} \over [z^n]P(z,1)} = 1 + {1\over 2} + \cdots + {1\over n} = H_n.$$
The harmonic numbers $H_n$ can be expanded as $\log n + \gamma + O(1/n)$, so we find that there are roughly
$\log n$ cycles in a random permutation on $n$ letters.
A slightly harder analysis of Taylor coefficients retrives the second moment.
From this, it can be shown that $\var_{\P_n}\{\chi\}\sim \log n$ an well.
\medskip

\boldlabel External nodes in a binary tree. Hidden inside the specification for binary trees,
we can see that we are actually only counting internal nodes, since we have given $\E$, the external node,
the generating function of 1 (meaning it has size 0). If we mark it with a $u$,
we arrive at
$$\B = u\E + \Z\times\B\times\B$$
and the bivariate ordinary generating function satisfies
$$B(z,u) = u + zB(z)^2.$$
Using the quadratic formula once again, we find that the generating function is
$$B(z,u) = {1-\sqrt{1-4uz} \over 2z},$$
which checks out, since setting $u=1$ gives exactly \refeq{\the\catalaneq}. We have the partial derivatives
$$P_u(z,u) = {1\over \sqrt{1-4uz}} \qquad\hbox{and}\qquad P_u^2(z,u) = {2z \over (1-4uz)^{3/2}}.$$
With the substitution $u=1$,
these are both relatively well-known generating functions. The first is the generating function
for the central binomial coefficients ${2n\choose n}$ and
the second is $z {\partial\over \partial z} P_u(z,1) $, so it has coefficients $n{2n\choose n}$
(the powers of $n$ drop down into the coefficient when differentiating, and multiplying by $z$ restores
the correct power).
We calculated in \refeq{\the\catalanformula} that the
total number of binary trees with $n$ internal nodes is ${2n\choose n}/(n+1)$, so applying our method gives us
$$\ex_{\B_n}\{\chi\} = {{2n\choose n}\over {1\over n+1}{2n\choose n}} = n+1$$
and
$$\ex_{\B_n}\{\chi^2\} = {n{2n\choose n} \over {1\over n+1}{2n\choose n}} + n + 1 = n(n+1)+n+1 = (n+1)^2.$$
So the average number of external nodes in a tree with $n$ internal nodes is $n+1$ and since
$$\var_{\B_n}\{\chi\} = \ex_{\B_n}\{\chi^2\} - \ex_{\B_n}\{\chi\}^2 = (n+1)^2 - (n+1)^2 = 0,$$
we discover that, in fact, every tree with $n$ internal nodes has exactly $n+1$ external nodes.
(We have performed what is probably the most convoluted proof of this simple fact, but it is a good exercise
in calculating moments using bivariate generating functions.)

\section References
\frenchspacing

\parskip=0pt
\hyphenpenalty=-1000 \pretolerance=-1 \tolerance=1000
\doublehyphendemerits=-100000 \finalhyphendemerits=-100000
\frenchspacing
\def\beginref{
\par\begingroup\nobreak\smallskip\parindent=0pt\kern1pt\nobreak
\everypar{\strut}
}
\def\endref{\kern1pt\endgroup\smallbreak\noindent}

\beginref Philippe Flajolet
and Robert Sedgewick,
{\sl Analytic Combinatorics}
(New York:
Cambridge University Press,
2009).
\endref

\beginref George P\'olya,
``On picture-writing,''
{\sl American Mathematical Monthly}\/
{\bf 63}
(1956),
689--697.
\endref

\beginref Heinz Pr\"ufer,
``Neuer Beweis eines Satzes \"uber Permutationen,''
{\sl Archiv der Mathematik und Physik}\/
{\bf 27}
(1918),
142--144.
\endref

\beginref Herbert S.~Wilf,
{\sl generatingfunctionology}
(Boston:
Academic Press,
1994).
\endref

\bye

### 2020-07-20 misc/expo/genfcts.tex
Added Anna to homies

@@
 \def\Cyc{\op{\tensc Cyc}}
+\def\Res{\op{\rm Res}}
 \def\A{{\cal A}}
@@
 
+\def\d{\,d}
+
+\def\CC{{\bf C}}
+\def\RR{{\bf R}}
+\def\eps{\epsilon}
+
 \def\eval{\big|}
@@
 
-\maketitle{A crash course on generating functions}{by}{Marcel K. Goh}{10 July 2020}
+\maketitle{A crash course on generating functions}{by}{Marcel K. Goh}{17 July 2020}
 
+\bigskip
+{\obeylines\eightssi
+\hfill A generating function is a clothesline
+\hfill on which we hang up a sequence for display.
+\eightss
+\smallskip
+\hfill --- HERBERT S. WILF, {\eightssi generatingfunctionology} (1989)}
 \section\the\sectcount. Introduction
@@
 $$A(z) = \sum_{n\geq 0} A_nz^n.$$
-In his textbook, titled {\sl generatingfunctionology},
-H. S. Wilf famously wrote: ``A generating function is a clothesline on which we hang up a sequence
-of numbers for display.'' That is exactly how we should treat the function $A(z)$ for now;
+H. S. Wilf famously described a generating function as a clothesline, and
+that is exactly how we should treat the function $A(z)$ for now;
 in later sections, we will discuss how this series can be treated as an analytic object.
@@
 using only 1s and 2s is simply a sequence of elements of $\I^*$. We conclude that
-$$\T = \Seq(\I^*)\qquad\hbox{and}\qquad T(z) = {1\over 1-I^*(z)} = {1\over 1-z-z^2}.\adveq,$$
-so we know that the generating function of the Fibonacci numbers is
+$$\T = \Seq(\I^*)\qquad\hbox{and}\qquad T(z) = {1\over 1-I^*(z)} = {1\over 1-z-z^2},\adveq$$
+and the generating function of the Fibonacci numbers is
 \newcount\fibeq
@@
 $$F(z) = {z\over 1-z-z^2}.\adveq$$
-\medskip
+\bigskip
+{\obeylines\eightssi
+\hfill ``What's one and one and one and one and one
+\hfill and one and one and one and one and one?''
+\hfill ``I don't know,'' said Alice. ``I lost count.''
+\hfill ``She can't do Addition,'' the Red Queen interrupted.
+\eightss
+\smallskip
+\hfill --- LEWIS CARROLL, {\eightssi Through the Looking Glass} (1871)}
+\bigskip
 
@@
 and by Lagrange's inversion theorem, with $f(u) = u/e^u$ and $H(u) = u$, we obtain
+\newcount\cayleyexact
+\cayleyexact=\eqcount
 $$n![z^n]T(z) = n!\bigg({1\over n}[u^{n-1}]e^{un}\bigg)
-              = n!\cdot{1\over n}\cdot{n^{n-1}\over(n-1)!} = n^{n-1}.$$
+              = n!\cdot{1\over n}\cdot{n^{n-1}\over(n-1)!} = n^{n-1}.\adveq$$
 Note that there are $n^{n-2}$ {\it unrooted} Cayley trees, since there are $n$ choices for the root.
@@
 which gives the bivariate exponential generating function
-$$P(z,u) = \exp\bigg(u\log{1\over 1-z}\bigg) = \exp\bigg(\log{1\over (1-z)^u}\bigg) = \bigg({1\over 1-z}\bigg)^u\adveq$$
+$$P(z,u) = \exp\bigg(u\log{1\over 1-z}\bigg) = \exp\bigg(\log{1\over (1-z)^u}\bigg)
+    = \bigg({1\over 1-z}\bigg)^u\adveq$$
 for the number of $n$-letter permutations with exactly $k$ cycles. We can calculate the partial derivatives
@@
 Here the random variable $\chi$ counts the number of cycles, and we have
-$$\ex_{\P_n}\{\chi\} = {[z^n] P_u(z,u)\eval_{u=1} \over [z^n]P(z,1)} = 1 + {1\over 2} + \cdots + {1\over n} = H_n.$$
+$$\ex_{\P_n}\{\chi\} = {[z^n] P_u(z,u)\eval_{u=1} \over [z^n]P(z,1)}
+     = 1 + {1\over 2} + \cdots + {1\over n} = H_n.\adveq$$
 The harmonic numbers $H_n$ can be expanded as $\log n + \gamma + O(1/n)$, so we find that there are roughly
@@
 we arrive at
-$$\B = u\E + \Z\times\B\times\B$$
+$$\B = u\E + \Z\times\B\times\B\adveq$$
 and the bivariate ordinary generating function satisfies
@@
 Using the quadratic formula once again, we find that the generating function is
-$$B(z,u) = {1-\sqrt{1-4uz} \over 2z},$$
+$$B(z,u) = {1-\sqrt{1-4uz} \over 2z},\adveq$$
 which checks out, since setting $u=1$ gives exactly \refeq{\the\catalaneq}. We have the partial derivatives
-$$P_u(z,u) = {1\over \sqrt{1-4uz}} \qquad\hbox{and}\qquad P_u^2(z,u) = {2z \over (1-4uz)^{3/2}}.$$
+$$P_u(z,u) = {1\over \sqrt{1-4uz}} \qquad\hbox{and}\qquad P_u^2(z,u) = {2z \over (1-4uz)^{3/2}}.\adveq$$
 With the substitution $u=1$,
@@
 total number of binary trees with $n$ internal nodes is ${2n\choose n}/(n+1)$, so applying our method gives us
-$$\ex_{\B_n}\{\chi\} = {{2n\choose n}\over {1\over n+1}{2n\choose n}} = n+1$$
+$$\ex_{\B_n}\{\chi\} = {{2n\choose n}\over {1\over n+1}{2n\choose n}} = n+1\adveq$$
 and
@@
 So the average number of external nodes in a tree with $n$ internal nodes is $n+1$ and since
-$$\var_{\B_n}\{\chi\} = \ex_{\B_n}\{\chi^2\} - \ex_{\B_n}\{\chi\}^2 = (n+1)^2 - (n+1)^2 = 0,$$
+$$\var_{\B_n}\{\chi\} = \ex_{\B_n}\{\chi^2\} - \ex_{\B_n}\{\chi\}^2 = (n+1)^2 - (n+1)^2 = 0,\adveq$$
 we discover that, in fact, every tree with $n$ internal nodes has exactly $n+1$ external nodes.
@@
 in calculating moments using bivariate generating functions.)
+\bigskip
+{\obeylines\eightssi
+\hfill And if I should live to be
+\hfill The last leaf upon the tree
+\eightss
+\smallskip
+\hfill --- OLIVER WENDELL HOLMES, {\eightssi The Last Leaf} (1831)}
+\bigskip
 
+\advance\sectcount by 1
+\section\llap{*}\the\sectcount. Rudiments of Complex Analysis
+
+In this section, we will collect definitions and theorems that will be useful in studying the coefficient
+asymptotics of generating functions. We will state results without proof; the details can be found any
+introductory textbook on complex analysis (e.g., Flanigan (1983)). Readers who are familiar with
+complex numbers may choose to skim or skip this section.
+
+The complex plane $\CC$ is the set of all numbers $x + iy$ were $x$ and $y$ are real numbers and $i$
+is the square root of $-1$. If $z = x + iy$, then the {\it real part} of $z$ is $\Re z = x$ and the
+{\it imaginary part} of $z$ is $\Im z = y$. The complex conjugate of $z = x + iy$ is $x-iy$, and it is
+easily verified, from the identity $i^2 = -1$, that
+$$z\cdot \bar z = x^2 + y^2.$$
+Letting $|z| = \sqrt{x^2 + y^2}$ denote the {\it modulus} of $z$, we find that
+$${z\cdot \bar z\over |z|^2} = 1,$$
+making it clear that the inverse $1/z$ of a complex number $z$ is equal to $\bar z / |z|^2$.
+
+For real $\theta$, we have the well-known formula
+$$e^{i\theta} = \cos \theta + i\sin \theta,\adveq$$
+which gives another way to represent a complex number. The complex number $z = x+iy$ can also be written
+as $|z|e^{i\theta}$, where $\theta$ is the angle that $z$ makes with the positive real axis.
+Note that $e^{i\theta} = e^{i(\theta + 2\pi k)}$ for any integer $k$.
+\bigskip
+{\obeylines\eightssi
+\hfill The imaginary relish is so sweet
+\hfill That it enchants my sense.
+\eightss
+\smallskip
+\hfill --- WILLIAM SHAKESPEARE, {\eightssi Twelfth Night} (1602)}
+\bigskip
+
+\boldlabel Complex differentation. A complex function defined on a domain $D$
+is said to be {\it analytic} at a point $z_0\in D$ if for some open ball $\Omega$ centred at
+$z_0$, $f$ is representable as a power series
+$$f(z) = \sum_{n\geq 0} f_n(z-z_0)^n.$$
+For example, $1/(1-z)$ is analytic for at any $z_0$ in the ball of radius 1 about the origin.
+A complex function whose domain is D
+is called {\it differentiable} or {\it holomorphic} at $z_0\in D$ if the limit
+$$f'(z_0) = \lim_{z\to z_0} {f(z) - f(z_0) \over z- z_0} $$
+exists, in the sense that $f'(z_0)$ has the same value regardless of the manner in which $z$ approaches
+$z_0$ in the complex plane. The following theorem says that these two notions are equivalent.
+
+\parenproclaim Theorem E (Basic Equivalence Theorem). A complex function $f$ defined on a domain
+$D$ is analytic at a point $z_0\in D$ if and only if it is differentiable at $z_0$.\slug
+
+It is a useful fact that if a function is differentiable at $z_0$,
+then its derivative is also differentiable at $z_0$, meaning
+that it admits derivatives of all orders, and we have the expansion
+$$f(z) = \sum_{n\geq 0} {f^{(n)} \over n!} (z-z_0)^n.\adveq$$
+Of course, if a function is analytic in a disc around $z_0$, then we can
+differentiate it by taking derivatives of the terms in its power series.
+\medskip
+
+\boldlabel Complex integration. The Fundamental Theorem of Calculus states that
+for a real-valued function $f$ with antiderivative $F$,
+$$\int_a^b f(x)\d x = F(b) - F(a).\adveq$$
+We can think of this as summing the values of $f$ over a path from $a$ to $b$. Since $\RR$ is
+one-dimensional,
+there is only one way to get from $a$ to $b$, but on the complex plane, we can integrate over
+any path $\gamma$ from $z_0\in \CC$ to $z_1\in \CC$. We have the following complex analogue of
+the Fundamental Theorem of Calculus.
+
+\parenproclaim Theorem P (Path integration). Let $f$ a complex function,
+continuous in a domain $D$, and let $F$
+be a complex function, analytic in $D$, that satisfies $F'(z) = f(z)$ for all $z\in D$.
+If $\gamma$ is any piecewise smooth path from $z_0\in D$ to $z_1\in D$, then
+$$\int_\gamma f(z)\d z = F(z_1) - F(z_0).\noskipslug$$
+
+We can evaluate complex integrals over paths by parametrising the curve $\gamma$ as a function
+of a real variable $t$ and then performing a change of variables. If $\gamma(t)$ is a parametrised curve
+and $t$ ranges from $a$ to $b$, then
+$$\int_\gamma f(z)\d z = \int_a^b f\big(\gamma(t)\big)\gamma'(t)\d t.\adveq$$
+
+As an example, if $f(z) = z^2$ and we want to find the integral of $f$ from $z_0 = 0$ to $z_1 = 1+2i$. We can
+walk from 0 to $1+2i$ via a path $\gamma_1$ that goes from $0$ to $2i$ and then another path $\gamma_2$ that goes
+from $2i$ to $1+2i$. So $\gamma_1(t) = it$ for $0\leq t\leq 2$ and $\gamma_2(t) = 2i + t$ for $0\leq t\leq 1$ and
+we have the derivatives ${\gamma_1}'(t) = i$ and ${\gamma_2}'(t) = 1$.
+Putting this all together, the integral over the
+whole path is
+$$\eqalign{
+\int_\gamma z^2 \dz &= \int_0^2 (it)^2\cdot i\d t + \int_0^1 (2i + t)^2\d t \cr
+&= \int_0^2 -it^2\d t + \int_0^1 t^2 + 4it - 4 \d t \cr
+&= \bigg[-{it^3\over 3}\bigg]_{t=0}^2 + \bigg[{t^3\over 3} + 2it^2 - 4t\bigg]_{t=0}^1 \cr
+&= -{8i\over 3} + {1\over 3} + 2i - 4 \cr
+&= -{11\over 3} - {2i\over 3} \cr
+}\adveq$$
+Of course, in this example, we could have just noticed that the antiderivative
+of $f(z) = z^2$ is $F(z) = z^3/3$ and applied
+Theorem P to get
+$$\int_\gamma f(z)\d z = F(1+2i) - F(0) = {(1+2i)^3\over 3} = -{11\over 3} - {2i\over 3}.\adveq$$
+
+When $\gamma$ is a {\it closed Jordan curve}, i.e.\ it starts and ends at the same point and does not
+cross itself, we often write the
+integral sign with a circle around it.
+The rules of integration are the same; the circle is just there to remind
+us that our path goes around in a loop. Integrals over closed curves are important, because of the following
+theorem.
+
+\parenproclaim Theorem I (Cauchy Integral Theorem). Let $\gamma$ be a closed Jordan curve in a domain
+$D$. Suppose $f$ is a function that is analytic both on $\gamma$ and the interior of $\gamma$. Then
+$$\oint_\gamma f(z)\d z = 0.\noskipslug$$
+
+A corollary of Theorem I is that if $\gamma_1$ is a closed curve that is fully contained in the interior
+of another closed curve $\gamma_2$ and $f$ is analytic on both curves as well as the region between
+the two curves, then
+$$\oint_{\gamma_1} f(z)\d z = \oint_{\gamma_2} f(z)\d z.\adveq$$
+\medskip
+
+\boldlabel Singularities. If functions tended to be analytic everywhere on $\CC$, then Theorem I tells
+us that integration wouldn't be very interesting. Luckily for us, most of the functions we study
+will fail to be analytic at certain points. Let $D$ be a domain and let $z_0\in D$ be an interior point.
+If $f$ is not analytic at $z_0$ but it is analytic on $D\setminus \{z_0\}$, then the point $z_0$ is said
+to be a {\it singularity} of $f$. There are three types of singularities:
+\medskip
+\item{i)} If $\lim_{z\to z_0} f(z) = w_0$ for some finite value $w_0$, then
+it is possible to continuously extend $f$ to an analytic function $g$ on all of $D$
+(just set $g(z_0) = w_0$). In this case, $z_0$ is called a {\it removable singularity}.
+\smallskip
+\item{ii)} If $\lim_{z\to z_0} |f(z)| = \infty$, then $z_0$ is called a {\it pole} of $f$.
+\smallskip
+\item{iii)} If neither (i) nor (ii) hold, then $z_0$ is called an {\it essential singularity} of $f$.
+\medskip
+We will now show that each of these types of singularities actually pop up in the wild with some examples.
+
+First, we can create a function with a removable singularity by taking a {\it bona fide} analytic function $f$
+and consider its restriction $f'$ on the domain $D\setminus \{z_0\}$. Now $z_0$ is a removable singularity
+of the function $f'$, since $f'$ can be continuously extended to an analytic function on all of $D$.
+
+An easy example of a function with a pole is $f(z) = 1/z$. This function is analytic everywhere on $\CC$ except at
+$z_0 = 0$, and $\lim_{z\to z_0} |1/z| = \infty$. This tells us that $f(z)$ may have an interesting integral
+around the origin. It does not matter which closed curve we take around 0, so the unit circle $C$ is a tempting
+choice. We can parametrise $C$ by $\gamma(\theta) = e^{i\theta}$ for $0\leq \theta< 2\pi$, which has the
+derivative $\gamma'(\theta) = ie^{i\theta}$. We now find that
+\newcount\twopii
+\twopii=\eqcount
+$$\oint_C {1\over z}\d z = \int_0^{2\pi} e^{-i\theta} ie^{i\theta} \d\theta= i\int_0^{2\pi} 1\d\theta = 2\pi i.\adveq$$
+This integral crops up a lot in complex analysis; commit it to memory!
+
+A function with an essential singularity is the square root function $\sqrt z$. This is because, given
+$z = re^{i\theta}$, we can rewrite
+$z = re^{i\theta + 2\pi ik}$ for any integer $k$, so for any integer $k$, $w = \sqrt re^{i\theta/2 + \pi i k}$
+satisfies the property that $w^2 = z$. To get a well-defined square root function, we set
+$$\sqrt z = \sqrt re^{i\theta/2},\adveq$$
+where $\theta\in (-\pi, \pi]$. This spells big trouble for complex numbers on the negative real axis.
+For example, let $z = -1 = e^{i\pi}$. Note that $\sqrt{e^{i(\pi - \eps)}} = e^{i(\pi/2 - \eps/2)}$ and
+$\sqrt{e^{-i(\pi -\eps)}} = e^{-i(\pi/2 - \eps/2)}$ (so both angles are in the range $(-\pi, \pi]$.
+As $\eps\to 0$, the arguments to the function both
+get arbitrarily close to $-1$ but
+the outputs have different limits ($i$ and $-i$). So the square root function has an infinite
+number of essential singularities (any negative real number); these are often called {\it algebraic
+singularities}.
+
+Another function with essential singularities is the natural logarithm $\log z$. Note that, given
+$z = re^{i\theta}$, the number $w = \log r + 2k\pi i\theta$ satisfies $e^w = z$ for any integer $k$.
+Once again, to get a well-defined function we must restrict $\theta$ to a single loop around the unit circle:
+$$\log z = \log r + i\theta,\adveq$$
+where $\theta\in (-\pi, \pi]$. Now, just as with the square root, we have a discontinuity in the argument
+for all real numbers $>1$. These are the {\it logarithmic singularities}.
+\medskip
+
+\boldlabel Residues. If we allow a power series expansion around a point $z_0$
+to have a finite number of negative powers, then it is
+called a {\it Laurent series}; this has the form
+$$f(z) = \sum_{n\geq -M} f_n (z-z_0)^n = {f_{-M} \over (z-z_0)^M} + \cdots + {f_{-1}\over (z-z_0)} + f_0
+    + f_1(z-z_0) + f_2(z-z_0)^2 + \cdots,$$
+for some nonnegative integer $M$. Note that, integrating on a circle of radius 1 around $z_0$, we
+can make the substitution $z - z_0 = e^{i\theta}$ to get $dz = ie^{i\theta}\d\theta$ and
+$$\oint (z-z_0)^M \d z = i\int_0^{2\pi} e^{i(M+1)\theta}\d\theta
+    = \cases{2\pi i & if $M=-1$; \cr 0 & otherwise.}\adveq$$
+So the coefficient $f_{-1}$ of $(z-z_0)^{-1}$ is special; it is the only term that survives the integration
+of a Laurent series. We call this the {\it residue of $f$ at $z_0$}
+and denote it by $\Res(f,z_0)$; we have
+$$\Res(f,z_0) = {1\over 2\pi i} \oint f(z)\d z.\adveq$$
+This is the crucial fact that will allows the following theorem to extract coefficients of generating functions.
+
+\parenproclaim Theorem C (Cauchy coefficient formula). Let $f(z)$ be analytic. The coefficient
+of $z^n$ is given by the formula
+$$[z^n]f(z) = {1\over 2\pi i} \oint {f(z) \over z^{n+1}}\d z,\adveq$$
+where the integral is taken on any closed curve about the origin.\slug
+
+\advance\sectcount by 1
+\section\the\sectcount. Generating Functions as Analytic Functions
+
+This section represents a turning point in our study of generating functions. Up until now, we have treated
+generating functions as purely formal objects. However, we can get a lot of information about the asymptotic
+behaviour of a power series' coefficients by treating it as a function of a complex argument. This section
+will give a general feel for how to do so, without getting too formal. A key theorem that allows us to work
+with generating functions as analytic objects is the following.
+
+\parenproclaim Theorem P (Pringsheim). If $f(z)$ has a series expansion whose coefficients are all
+nonnegative and the radius of convergence of $f(z)$ is $R$, then the real number $R$ is a singularity
+of $f(z)$.\slug
+
+Since counting objects usually induces a generating function with nonnegative coefficients, this theorem
+applies to most of the interesting cases we study.
+Note that we will only deal with the case where there is a {\it unique}
+root of smallest modulus, meaning that $R$ is the only singularity on the circle $|z| = R$. This assumption
+eliminates complications that involve periodicities in the coefficients.
+\medskip
+
+\boldlabel Rational functions. Treating a generating function as a complex function pays off immediately
+when the function is a ratio $f(z)/g(z)$ of two polynomials. Since a polynomial of degree $n$ always has
+$n$ complex roots, we can factorise $g(z)$ into a product of linear factors and then use the partial fraction
+expansion to extract the coefficients. For example, if $f(z) = z$ and $g(z) = 18z^3 - 3z^2 - 4z + 1$, then
+$$\eqalign{
+{f(z)\over g(z)} &= {z\over 18z^3 - 3z^2 - 4z + 1} = {z\over 18(z-1/3)^2(z+1/2)} \cr
+&= {1\over 45(z-1/3)^2} + {1\over 25(z-1/3)} - {1\over 25(z+1/2)}\cr
+&= {1\over 5(1-3z)^2} - {3\over 25(1-3z)} - {2\over 25(1+2z)}.\cr
+}\adveq$$
+This immediately tells us that
+\newcount\exactrat
+\exactrat=\eqcount
+$$\eqalign{
+[z^n]{f(z)\over g(z)} &= {1\over 5}(n+1)3^{n} - {3\over 25} 3^n - {2\over 25}(-2)^n \cr
+&=  {1\over 5}n3^n + {2\over 25} 3^n - {2\over 25}(-2)^n. \cr
+}\adveq$$
+Notice that only the first term is important, because $3^n$ is exponentially larger than $2^n$. We got
+an exponential factor of $\sim 3^n$ because the root of $g(z)$ that is closest to the origin is $1/3$. The fact
+that this root has multiplicity $2$ gives a factor of $\sim n$ in the first term. It is safe to ignore
+all the other terms because they are much smaller, as formalised in the following theorem.
+
+\parenproclaim Theorem G (Rational asymptotics). Let $f(z)$ and $g(z)$ be polynomials that are relatively prime,
+with $g(z) \neq 0$. If $g(z)$ has a a unique real root $\alpha$ of smallest modulus whose multiplicity is $r$, then
+$$[z^n]{f(z)\over g(z)} \sim C\alpha^{-n} n^{r-1} \qquad\hbox{where}\qquad
+   C = {r(-1/\alpha)^r f(\alpha) \over g^{(r)}(\alpha)}.\noskipslug$$
+
+This will become clear later on when we discuss a more general result.
+Applying this theorem to our most recent example, we have $\alpha = 1/3$, $r = 2$, and
+$g''(z) = 108z-6$. So we can derive
+$$[z^n]{z\over 18z^3 - 3z^2 - 4z + 1} = {2(-3)^2 (1/3) \over 108(1/3)-6} 3^n n^{2-1} = {1\over 5} n3^n,\adveq$$
+which is exactly the first term in \refeq{\the\exactrat}.
+\medskip
+
+\boldlabel Fibonacci asymptotics. As an example, recall the generating
+function for the Fibonacci numbers, $z/(1-z-z^2)$. This function has poles at
+$$-\phi = {-1-\sqrt 5\over 2}\approx -1.618 \qquad\hbox{and}\qquad
+   \bar\phi = {-1+\sqrt 5\over 2}\approx0.618,\adveq$$
+where $\phi$ is the golden ratio.
+The root of smaller modulus is $\alpha = \bar\phi = 1/\phi$, and the multiplicity $r = 1$. This gives
+$$F_n = [z^n]{z\over 1-z-z^2} \sim {(-\phi)(1/\phi)\over 1-2/\phi} \phi^n
+    = {\phi \over \phi + 2}\phi^n = {1 + \sqrt 5 \over 5 + \sqrt 5}\phi^n = {\phi ^n\over \sqrt 5},\adveq$$
+which says that the ratio between successive Fibonacci numbers tends towards $\phi$.
+\medskip
+
+\boldlabel Meromorphic functions. Rational functions are special cases of a broader class of functions
+called meromorphic functions. A complex function is {\it meromorphic} if it can be expressed as the ratio
+of two analytic functions. A function is meromorphic in a region if and only if it is analytic except for
+at a finite set of poles.
+We will use the fact that any function that is meromorphic at the point $z_0$
+admits a Laurent expansion
+$$h(z) = {h_{-M} \over (z-z_0)^M} + \cdots + {h_{-1}\over (z-z_0)} + h_0
+    + h_1(z-z_0) + h_2(z-z_0)^2 + \cdots,$$
+where $M$ is a nonnegative integer. We say that $h$ has a {\it pole of order $M$} at the point $z_0$.
+In this case, the function $(z-z_0)^M h(z)$ is analytic at $z_0$, because the Laurent series becomes
+a common-or-garden Taylor series. We have the following generalisation of Cauchy's coefficient formula
+for meromorphic functions.
+
+\parenproclaim Theorem R (Residue theorem). Let $h(z)$ is meromorphic in a region $D$ and let
+$\gamma$ be a closed loop in $D$. Suppose that $\gamma$ encloses $k$ poles of $h$; call them
+$\alpha_1,\ldots,\alpha_k$. Then
+$$[z^n]h(z) = {1\over 2\pi i} \oint_\gamma h(z)\d z = \sum_{i=1}^k \Res(h, \alpha_i).\adveq$$
+In particular, if $h$ is meromorphic in closed ball $|z|\leq R$, analytic at 0,
+and the poles have orders $m_1,\ldots,m_k$, then
+$$[z^n]h(z) = {p_1(n)\over {\alpha_1}^n} + \cdots + {p_k(n)\over {\alpha_k}^n} + O(R^{-N}),\adveq$$
+where each polynomial $p_i$ has degree $m_i - 1$.\slug
+
+As before, we can get a very good approximation by looking only at the pole $\alpha$ of smallest modulus.
+The residue around that pole contributes exponentially more to the coefficient asymptotics than the others,
+so we should analyse the Laurent expansion of $h(z)$ around $\alpha$. If $\alpha$ has order $M$, then
+this looks like
+$$h(z) = {h_{-M} \over (z-\alpha)^M} + \cdots + {h_{-1}\over (z-\alpha)} + h_0
+    + h_1(z-\alpha) + h_2(z-\alpha)^2 + \cdots,$$
+and when $z$ is very close to $\alpha$, only the first term matters. In other words,
+$$h(z) \sim {h_{-M} \over (z-\alpha)^M} = {h_{-M}\over (-\alpha)^M (1-z/\alpha)^M}
+   = (-1)^M{h_{-M}\over\alpha^M} \sum_{n\geq 0} {n+M-1\over \alpha^n}z^n.\adveq$$
+The actual residue $h_{-M}$ can be computed as a limit, by repeated application of l'Hospital's Rule.
+If $h(z)$ is equal to the ratio $f(z)/g(z)$ of analytic functions, then
+$$h_{-M} = \lim_{z\to \alpha} {(z-\alpha)^M f(z) \over g(z)} = {Mf(\alpha) \over g^{(M)}(\alpha)}.\adveq$$
+Putting this all together, we have the following theorem.
+
+\parenproclaim Theorem M (Meromorphic asymptotics). If $h(z) = f(z)/g(z)$ is a meromorphic function with
+a unique pole of smallest modulus $\alpha$ whose order is $M$, then
+$$[z^n]h(z) \sim (-1)^M {Mf(\alpha) \over \alpha^M g^{(M)}(\alpha)} \alpha^{-n} n^{M-1}.\noskipslug$$
+
+We now see that Theorem G is a special case of this, since a multiplicity-$r$
+root of a polynomial in the denominator induces a pole of order $r$.
+\medskip
+
+\boldlabel Asymptotics of derangements. Consider the following word problem: It is a Tuesday night and
+$n$ froshies leave their Residence to go to Caf\'e Campus. After attaining a sufficient state of inebriation,
+the $n$ froshies all make it safely back to Residence, but each of them has found a random room to sleep in;
+that is, they return to their rooms in a random permutation. What is the likelihood that no froshie returns
+to his or her own room?
+
+We can regard the $n$ froshies as $n$ labelled atoms and since a permutation as a set of cycles,
+this problem asks for the proportion of all $n$-letter permutations
+that have no singleton cycle. These are exactly the derangements, which have the EGF
+$D(z) = e^{-z}/(1-z)$.
+We can use Theorem M to get an estimate of the number of $n$-letter derangements. We have $f(z) = e^{-z}$,
+$g(z) = 1-z$, and a pole of order $M=1$ at $\alpha=1$. The derivative of $g(z)$ is $-1$ so after a rumble and
+a tumble and a great deal of cancelling out 1s, we have
+$$n![z^n]D(z) \sim n! (-1)^1 {1\cdot e^{-1} \over 1^1 (-1)} 1^{-n} n^0 = {1\over e}n!.\adveq$$
+There are $n!$ total permutations so the proportion of permutations that are derangements is $1/e$;
+we have a $1/e \approx 36.8\%$ chance that the froshies all end up in rooms not belonging to them.
+
+\advance\sectcount by 1
+\section\the\sectcount. Singularity Analysis
+
+We have now dealt with generating functions that have poles. It remains to tackle functions that have
+essential singularities, such as those caused by square roots and logarithms. First we need a couple of facts
+about the gamma function. For positive integers $n$, we have $\Gamma(n) = (n-1)!$ and
+for all $z\in \CC$ with $\Re z > 0$, $\Gamma$ is defined by
+$$\Gamma(z) = \int_0^\infty x^{z-1} e^{-x}\d x.\adveq$$
+Now the full gamma function is defined to be the analytic continuation to a meromorphic function
+that has analytic everywhere except the nonpositive integers. The nonpositive integers are poles of $\Gamma$.
+Just like with factorials, we have $\Gamma(z+1) = z\Gamma(z)$. The important identity
+$$\Gamma(1/2) = \int_0^\infty x^{-1/2} e^{-x}\d x = 2\int_0^\infty e^{-x^2}\d x = \sqrt\pi\adveq$$
+should also be memorised. The {\it Hankel contour} $H$ is the path that
+starts at $\infty$, hovers $+\delta$ above the real line until reaching the unit circle, travels around the unit
+circle counterclockwise until reaching a point a distance of $\delta$ below the real line, then increases
+off to $\infty$ once again. The distance $\delta$ is called the {\it slit width}. We have the identity
+\newcount\hankel
+\hankel=\eqcount
+$${1\over \Gamma(z)} = {1\over 2\pi i} \int_H (-x)^{-z} e^{-x} \d x,\adveq$$
+which will come in handy later.
+\bigskip
+{\obeylines\eightssi
+\hfill Singularity is almost invariably a clue.
+\eightss
+\smallskip
+\hfill --- SHERLOCK HOLMES, {\eightssi The Boscombe Valley Mystery} (1891)}
+\bigskip
+
+\boldlabel The standard function scale. We saw that the function $\sqrt z$ is singular when $z$ is negative
+and real. More generally, the function $(1-z)^{-\alpha}$ is singular when $z$ is real and $>1$. This leads
+to the fundamental result that makes singularity analysis possible.
+
+\parenproclaim Theorem S (Standard function scale). If $\alpha\in \CC$ is not a nonpositive real number,
+then
+$$[z^n](1-z)^{-\alpha} \sim {n^{\alpha - 1}\over \Gamma(\alpha)}.\adveq$$
+
+\proof (Sketch.) Let $H$ be a slight variant of the Hankel contour of slit width $1/n$
+that circles around $z=1$ instead
+of the origin. We start with the coefficient formula
+$$[z^n](1-z)^{-\alpha} = {1\over 2\pi i} \oint {(1-z)^{-\alpha} \over z^{n+1}}\d z,$$
+where the path of integration goes in a circle around the origin. We deform this circle
+to a ball of infinite radius that avoids the essential singularities; this is exactly the contour $H$.
+The change of variable
+$z = 1+t/n$ and $dz = dt/n$ gives
+$$[z^n](1-z)^{-\alpha} = {1\over 2\pi i} \int_H {(-t)^{-\alpha} \over n^{-\alpha} z^{n+1}}{1\over n}\d t
+= {n^{\alpha -1} \over 2\pi i} \int_H (-t)^{-\alpha}\bigg(1+{t\over n}\bigg)^{-n-1} \d t
+\sim {n^{\alpha - 1}\over \Gamma(\alpha)},\adveq$$
+where the last $\sim$ comes from the approximation
+$$\bigg(1+{t\over n}\bigg)^{-n-1} \sim \bigg(1+{t\over n}\bigg)^{-n} \to e^{-t}$$
+combined with $\refeq{\the\hankel}$.\slug
+
+A similar but equally involved process proves the more general version of the standard function scale,
+which also handles logarithms.
+
+\parenproclaim Theorem S$'$ (Standard function scale, logarithms).
+If $\alpha\in \CC$ is not a nonpositive real number, then
+$$[z^n](1-z)^{-\alpha}\bigg({1\over z}\log {1\over 1-z}\bigg)^\beta
+    \sim {N^{\alpha - 1}\over \Gamma(\alpha)}(\log n)^\beta.\noskipslug\adveq$$
+
+If $f(z)$ is singular at $z=\rho$, then $g(z) = f(\rho z)$ is singular at $1$ and
+$$[z^n] f(z) = \rho^{-n} [z^n] f(\rho z) = \rho^{-n} [z^n] g(z).\adveq$$
+When a generating function is not meromorphic, the approach to getting its coefficient asymptotics is
+as follows:
+\medskip
+\item{i)} First, we locate the dominant singularity $\rho$ (the one closest to the origin).
+Check that it is the only singularity on the circle of convergence.
+\smallskip
+\item{ii)} Next, we find functions $\sigma$ and $\tau$ on the standard scale such that,
+for $z$ near $\rho$,
+$$f(z) = \sigma(z/\rho) + O\big(\tau(z/\rho)\big).$$
+We should have $\tau(z) = o\big(\sigma(z)\big)$.
+\smallskip
+\item{iii)} Lastly, we apply Theorem S$'$ to $\sigma$ and $\tau$ to get an approximation for $f$.
+\medskip
+
+\boldlabel Catalan asymptotics. We are now ready to derive an asymptotic estimate for the growth
+of the Catalan numbers. Earlier, we derived the OGF
+$$C(z) = {1 - \sqrt{1-4z} \over 2z},$$
+which has a radius of convergence of $1/4$. By Pringsheim's theorem, we have an algebraic
+singularity at $\rho = 1/4$. Now we rewrite
+$$C(z) = {1\over 2z} - {(1-4z)^{1/2} \over 2z};$$
+when $z$ is close to $1/4$ this is
+$$C(z) \sim 2 - 2(1-4z)^{1/2}.\adveq$$
+Now we use the standard scale with $\alpha = -1/2$. From $z\Gamma(z) = \Gamma(z+1)$ we infer that
+$$\eqalign{
+-{1\over 2}\Gamma(-1/2) &= \Gamma(1/2) \cr
+\Gamma(-1/2) &= -2\sqrt \pi \cr
+},\adveq$$
+whence the famous approximation
+$$[z^n]C(z) \sim [z^n]\big(-2(1-4z)^{1/2}\big)
+  \sim {-2n^{-3/2}4^n \over \Gamma(-1/2)} = {4^n \over \sqrt \pi n^{3/2}}\adveq$$
+can be derived.
+\medskip
+
+\boldlabel Simple varieties of trees. Many generating functions associated with tree classes
+have a generating function that satisfies $f(z) = z\phi\big(f(z)\big)$. We have the following
+theorem that generalises singularity analysis of simple varieties of trees.
+
+\parenproclaim Theorem V (Simple varieties of trees). Suppose $f(z) = z\phi\big(f(z)\big)$ and furthermore,
+the following conditions hold:
+\medskip
+\item{i)} The function $\phi(u)$ is nonlinear, satisfies $\phi(0)\neq 0$, is analytic at 0 and has
+nonnegative coefficients.
+\smallskip
+\item{ii)} Within the open ball of convergence $|z| < R$ of $\phi$, there exists a
+positive real solution $\tau$ to the equation
+$$\phi(\tau) = \tau\phi'(\tau).\adveq$$
+\medskip
+\noindent Then $\rho = \tau/\phi(\tau)$ is the radius of convergence of $f(z)$ and
+$$[z^n]f(z) \sim \sqrt{{\phi(\tau) \over 2\pi \phi''(\tau)}} {\rho^{-n} \over n^{3/2}}.\adveq$$
+
+\proof The proof is beyond the scope of the notes, but the general gist is to first show that
+$$f(z) \sim \tau - \sqrt{2\phi(\tau) / \phi''(\tau)} \sqrt{(1-z\phi'(\tau))}\adveq$$
+and then use Theorem S.\slug
+\medskip
+
+\boldlabel Cayley trees and Stirling's formula.
+We will apply Theorem V to the family of Cayley trees, whose EGF satisfies
+$$T(z) = ze^{T(z)}.$$
+So $\phi(u) = \phi'(u) = \phi''(u) = e^u$, $\tau = 1$, $\rho = 1/e$ and
+$$n![z^n]T(z) \sim n!\sqrt{{e \over 2\pi e}} {e^n \over n^{3/2}} = {n!e^n\over \sqrt{2\pi} n^{3/2}}.\adveq$$
+We saw earlier that $n![z^n]T(z)$ is exactly $n^{n-1}$. This can be substituted to obtain
+$$\eqalign{
+{n! e^n \over \sqrt{2\pi} n^{3/2}} &\sim n^{n-1} \cr
+n! &\sim {\sqrt{2\pi} n^{3/2} \cdot n^{n-1} \over e^n} = \sqrt{2\pi n}\bigg({n\over e}\bigg)^n,\cr
+}\adveq$$
+and we have, as a byproduct of counting trees, derived Stirling's formula!
 \section References
@@
 2009).
+\endref
+
+\beginref Francis J.~Flanigan,
+{\sl Complex Variables: Harmonic and Analytic Functions}
+(New York:
+Dover,
+1983).
 \endref

### 2020-07-30 misc/literate/lll.c
Small error in LLL and revamped font macros

@@ -161,7 +161,7 @@ mu[k][i]= mu[k][i]-rounded*mu[l][i];
 #line 184 "./lll.w"
 ;
 if(/*8:*/
-#line 261 "./lll.w"
+#line 264 "./lll.w"
 
 B[k]>=(0.75-mu[k][k-1]*mu[k][k-1])*B[k-1]
 
@@ -191,7 +191,7 @@ mu[k][i]= mu[k][i]-rounded*mu[l][i];
 ++k;
 }else{
 /*9:*/
-#line 267 "./lll.w"
+#line 270 "./lll.w"
 
 set(temp,bb[k]);
 set(bb[k],bb[k-1]);
@@ -241,7 +241,7 @@ return H;
 int main(){
 n= DIM;
 /*11:*/
-#line 302 "./lll.w"
+#line 305 "./lll.w"
 
 double input_lattice[DIM][DIM]= INPUT_BASIS;
 double input_quad[DIM][DIM]= INPUT_QUAD;
@@ -253,7 +253,7 @@ Q[i+1][j+1]= input_quad[i][j];
 }
 printf("Input lattice basis:\n");
 /*13:*/
-#line 325 "./lll.w"
+#line 328 "./lll.w"
 
 for(int i= 1;i<=n;++i){
 printf("(");
@@ -265,7 +265,7 @@ printf(")\n");
 }
 
 /*:13*/
-#line 312 "./lll.w"
+#line 315 "./lll.w"
 ;
 printf("Input Q matrix:\n");
 for(int j= 1;j<=n;++j){
@@ -282,11 +282,11 @@ int**H;
 H= lll(bb);
 if(H!=NULL){
 /*12:*/
-#line 321 "./lll.w"
+#line 324 "./lll.w"
 
 printf("Reduced basis:\n");
 /*13:*/
-#line 325 "./lll.w"
+#line 328 "./lll.w"
 
 for(int i= 1;i<=n;++i){
 printf("(");
@@ -298,14 +298,14 @@ printf(")\n");
 }
 
 /*:13*/
-#line 323 "./lll.w"
+#line 326 "./lll.w"
 ;
 
 /*:12*/
 #line 70 "./lll.w"
 ;
 /*14:*/
-#line 337 "./lll.w"
+#line 340 "./lll.w"
 
 printf("H matrix:\n");
 for(int j= 1;j<=n;++j){

### 2020-07-31 about/vita.tex
Added note to 400
Added papers to research page
Updated vita

@@
 \font\elevensc=cmcsc11  % For section headers
-\font\bigbold=cmbx12  % Name
+\font\bigbold=cmb12  % Name
+\font\tenb=cmb10
+\font\mc=cmr9
+
+\let\tenbf=\tenb
+
+% For publication lists (adapted from Knuth's algorithms)
+\newdimen\itemindent
+\newif\ifitempar \itempartrue
+\def\itemindentset#1{\setbox0\hbox{{\bf #1.\kern.25em}}\itemindent=\wd0\relax}
+\def\pubbegin #1{\itemindentset{#11}} % when all have 1 digit
+\def\ppubbegin #1{\itemindentset{#111}} % when 10 or more publications
+\def\pubitem#1.{\ifitempar\smallskip\noindent\else\itempartrue
+  \hskip-\parindent\fi
+  \hbox to\itemindent{\bf\hfil #1.\kern.25em}%
+  \hangindent=\itemindent\hangafter=1\ignorespaces}
+
+\def\up#1{\leavevmode \raise.16ex\hbox{#1}}
+\def\sectheader#1{{\bigskip\elevensc #1}\smallskip\hrule\medskip}
+\def\leftright#1#2{{#1}\hfill{#2}\par}
+\def\myname{{\tensc Marcel K. Goh}}
+\def\CEE{{\mc C}}
+\def\UNIX{{\mc UNIX}}
+\def\thing{\item{$\bullet$}}
 
@@
 
-\def\sectheader#1{{\bigskip\elevensc #1}\smallskip\hrule\medskip}
-\def\leftright#1#2{{#1}\hfill{#2}\par}
+\newcount\submittedcount
+\submittedcount=1
+\def\subitem{\pubitem S\the\submittedcount. \global\advance\submittedcount by 1}
+\newcount\reportcount
+\reportcount=1
+\def\repitem{\pubitem R\the\reportcount. \global\advance\reportcount by 1}
+
+% =========== START ============ %
 
@@
 \leftright{Email: marcel\_goh@yahoo.ca}{Montr\'eal, QC H2X 2H3, Canada}
-\leftright{URL: marcelgoh.github.io}{+1$\;$(438)$\;$886-2338}
+\leftright{URL: marcelgoh.github.io}{+1\thinspace\up(438\up)\thinspace 886-2338}
 
@@
 CGPA: 3.79/4.
-\medskip
+\medbreak
 \leftright{\bf Charles University}{\bf February -- June 2019}
@@
 Work involved writing of code in OCaml and PostScript.
-\medskip
+\medbreak
 
@@
 ongoing work on the functional programming language Beluga.
-\medskip
+\medbreak
 
@@
 Created OPythn, a bytecode compiler and virtual machine for a subset of Python.
-OPythn includes support for lists, dictionaries, named and anonymous functions, objects and classes.
+OPythn includes support for lists, dictionaries, named and anonymous functions, objects, and classes.
 It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing.
-\medskip
+
+\sectheader{Submitted Papers}
+
+\pubbegin S
+\subitem (with Anna M. Brandenberger and Luc Devroye)
+Root estimation in Galton-Watson trees. {\sl arXiv preprint 2007.05681}.
+Submitted July 2020.
+
+\sectheader{Reports}
+
+\pubbegin R
+\repitem Typechecking proof scripts: making interactive proof assistants robust.
+Honours project report, McGill University (Montr\'eal, Qu\'ebec, December 2019), 10 pp.
 
@@
 and MATH 240, a discrete mathematics course, in Winter 2020.
-\medskip
+\medbreak
 
@@
 and functional programming. Recipient of the Tomlinson Engagement Award for Mentoring.
-\medskip
-
-\filbreak % Break page here
+\medbreak
 
@@
 first-aid skills up-to-date.
-\medskip
+\medbreak
 
@@
 Oversaw academic affairs within SUMS council and acted as liaison between the undergraduate community and
-mathematics faculty. Duties included representing the student body at department meetings and helping students with
+mathematics faculty. Duties included representing the student body at department meetings,
+organising midterm and final review sessions,
+and helping students with
 academic concerns.
-\medskip
+\medbreak
 
@@
 Led a six-person team consisting of a medic, signaller, sensor, and two-machine gunners within a rifle platoon.
-\medskip
+\medbreak
 
@@
 {\bf Programming Languages}\par
-OCaml, C, Haskell, Standard ML, Scheme, PostScript, Assembly, Python
-\medskip
+\CEE, OCaml, Haskell, Standard ML, Scheme Lisp, PostScript, {\tt CWEB}, {\tt MIXAL}, Python
+\medbreak
 
 {\bf Technologies}\par
-UNIX, Vim, \TeX, Git.
-\medskip
+\UNIX, Vim, \TeX, Git.
+\medbreak
 
@@
 Fluent: English, French. Proficient: Mandarin, Italian.
-\medskip
+\medbreak
 
@@
 
-Recipient of 0x\$0.40 in Knuth reward cheques.
+\parindent=10pt
+\thing Scored 10/120 on the William Lowell Putnam Mathematical Competition in 2019
+(the median score that year was 2/120).
+\smallskip
+\thing Recipient of 0x\$0.40 in Knuth reward cheques.
 

### 2020-09-12 misc/proofs/random-divisibility.tex
Games I've finished

% Probablity that random variable divisibility

\input fontmac
\input mathmac

\classicmode

\def\Res{\op{Res}}

\font\eightsl=cmsl8

\maketitle{On the divisibility of a random variable\footnote{$^*$}{\eightpt This is a generalisation of an assignment
question given by Prof. Luc Devroye in his COMP 690 class, Fall 2020.}}
{by}{Marcel K. Goh (\rm Montr\'eal, Qu\'ebec)}{11 September 2020}

\medskip

\noindent If $X$ is an integer-valued random variable and $n$ is a positive integer, we might want to know
the probability that $n$ divides $X$. To this end, we will make use of the probability generating
function
$$p(z) = \sum_{j=0}^\infty p_jz^j.$$
We have the following result:

\proclaim Theorem A. Let $X$ be a nonnegative integer-valued random variable whose probability generating function
$p(z)$ has radius of convergence $R>1$. Let $n$ be a positive integer and let $\zeta_1,\ldots,\zeta_n$ denote the
$n$th roots of unity. The probability that $n$ divides $X$ is given by two equivalent formulas:
$$\pr\big\{X \equiv 0 \!\!\!\!\pmod{n}\big\}
   = {1\over n}\sum_{k=1}^n p(\zeta_k) = {1\over n}\sum_{k=1}^n \Re\,p(\zeta_k)$$

\proof
Let $p_j = \pr\{X = j\}$ for all positive integers $j$. We are trying to compute the sum
$$p^* = p_0 + p_n + p_{2n} + \cdots.$$
Consider the generating function
$$f(z) = \bigg(1 + {1\over z^n} + {1\over z^{2n}} + \cdots\bigg) p(z).$$
For any multiple of $kn$ of $n$, there is some term of the infinite sum that will pull $p_{kn}$ into the constant
term of $f(z)$. So it is clear that $[z^0] f(z) = p_0 + p_n + p_{2n} +\cdots = p^*$.
We can rewrite $f(z)$ as
$$f(z) = {p(z)\over 1 - z^{-n}} = {z^np(z) \over z^n - 1}.$$
Letting $g(z) = f(z)/z$ and applying Cauchy's Integral Formula, the constant coefficient is given by
$$[z^0]f(z) = [z^{-1}] g(z) = {1\over 2\pi i}\oint g(z)\,dz,$$
where the path of integration is taken in the annulus of convergence.
The function
$$g(z) = {z^{n-1}p(z)\over z^n - 1}$$
has only $n$ singularities on the unit circle: a pole of order 1 at each of the $n$ roots of unity.
So we may take
our path of integration to be any positively-oriented loop around the origin that stays outside the closed unit disk
and inside the disk $|z| < R$.
By the Residue Theorem, this is the sum of the residues at each of the $n$ poles, so we have
$$p^* = {1\over 2\pi i}\oint g(z) \,dz =
    {1\over 2\pi i}\cdot 2\pi i\big(\Res(g;\zeta_1) + \cdots + \Res(g; \zeta_n)\big).$$
We let $g(z) = a(z)/b(z)$, and since $a(z) = z^{n-1}p(z)$ and $b(z) = z^n - 1$ are both holomorphic in neighbourhoods
around each of the poles, for any pole $\zeta_i$ of $g$ we have
$$\Res(g;\zeta_k) = {a(\zeta_k)\over b'(\zeta_k)} = {{\zeta_k}^{n-1}p(\zeta_k)\over n{\zeta_k}^{n-1}}
            = {p(\zeta_k)\over n}.$$
Plugging the $n$th roots of unity into the formula, we have
$$p^* = \sum_{k=1}^n \Res(g;\zeta_k) = \sum_{k=1}^n {p(\zeta_k)\over n}$$
Note if $\zeta_k$ is not real, then $\overline{\zeta_k}$ is also an $n$th root of unity. Using the identity
$\Re z + \Re(\overline z) = (z + \overline z)$, we find that
$$p^* = {1\over n}\sum_{k=1}^n p(\zeta_k) = {1\over n}\sum_{k=1}^n \Re\,p(\zeta_k),$$
assuring us that $p^*$ is real and proving the theorem statement.\slug

\end

### 2020-10-09 about/vita.tex
Lots of changes

@@
 
+% From cwebmac: Date and time
+\def\today{\ifcase\month\or
+  January\or February\or March\or April\or May\or June\or
+  July\or August\or September\or October\or November\or December\fi
+  \space\number\day, \number\year}
+\newcount\twodigits
+\def\hours{\twodigits=\time \divide\twodigits by 60 \printtwodigits
+  \multiply\twodigits by-60 \advance\twodigits by\time :\printtwodigits}
+\def\gobbleone1{}
+\def\printtwodigits{\advance\twodigits100
+  \expandafter\gobbleone\number\twodigits
+  \advance\twodigits-100 }
+
+\font\eightrm=cmr8
 \font\elevensc=cmcsc11  % For section headers
@@
 
+\def\datethis{\eightrm Last updated \today\ at\ \hours.}
+
 % =========== START ============ %
+
+\footline{\phantom{\datethis}\hss\folio\hss\datethis}  % \phantom for centering
 
@@
 \leftright{Email: marcel\_goh@yahoo.ca}{Montr\'eal, QC H2X 2H3, Canada}
-\leftright{URL: marcelgoh.github.io}{+1\thinspace\up(438\up)\thinspace 886-2338}
+\leftright{URL: marcelgoh.github.io}{+1\thinspace\up(825\up)\thinspace 440-0681}
 
@@
 
-\leftright{\bf McGill University}{\bf September 2017 -- May 2021}
+\leftright{\bf McGill University}{\bf September 2017 -- April 2021}
 \leftright{\sl BSc, Honours Mathematics and Computer Science, minor Linguistics}{\sl Montr\'eal, QC, Canada}
@@
 
-\leftright{\bf Sorting algorithms on manifolds}{\bf May 2020 -- present}
+\leftright{\bf Probabilistic analysis of branching processes}{\bf May 2020 -- present}
+\leftright{\sl Research group}{\sl McGill University}
+\smallskip
+Ongoing research on branching processes, headed by Luc Devroye. Studied estimation problems
+on Galton-Watson trees. Attended informal weekly seminars over the summer, discussing
+various topics related to branching
+processes. (Gave a presentation at three of these: one on root estimation in Galton-Watson
+trees, and two on generating functions and elementary analytic combinatorics.)
+
+\medbreak
+\leftright{\bf Sorting algorithms on manifolds}{\bf May -- August 2020}
 \leftright{\sl Summer research project}{\sl McGill University}
 \smallskip
-Summer research project under the supervision of Michael Lipnowski.
+Research project under the supervision of Michael Lipnowski.
 Studied algorithms that investigate the topology of group actions on locally symmetric spaces.
-Work involved writing of code in OCaml and PostScript.
+Work involved writing of code in OCaml and PostScript that generates illustrations of certain quotient
+spaces, in various models of hyperbolic geometry.
 \medbreak
@@
 Proved a theorem in constructive logic concerning the formal verification
-of an interactive proof assistant and wrote OCaml code as part of
+of the interactive proof assistant Harpoon and wrote OCaml code as part of
 ongoing work on the functional programming language Beluga.
@@
 
+\filbreak
+
 \sectheader{Work and Volunteer Experience}
@@
 \smallskip
-Graded MATH 235, an introductory course on the theory of groups, rings, and fields, in Fall 2019
-and MATH 240, a discrete mathematics course, in Winter 2020.
+Grading of assignments in the following courses:
+\begingroup\parindent=10pt
+\smallskip
+\thing Fall 2020: MATH 323 Probability, MATH 456 Honours Algebra 3
+\smallskip
+\thing Winter 2020: MATH 240 Discrete Structures
+\smallskip
+\thing Fall 2019: MATH 235 Algebra 1
+\endgroup
 \medbreak
@@
 
+\leftright{\bf Bakir Contracting Corp.}{\bf May -- August 2018}
+\leftright{\sl Painter}{\sl Edmonton, AB, Canada}
+\smallskip
+Exterior painting (siding, decks, fences, trim, etc.) for residential clients.
+\medbreak
+
 \leftright{\bf Singapore Armed Forces}{\bf August 2015 -- August 2017}
@@
 
+\leftright{\bf YoWoChAs Outdoor Education Centre}{\bf June -- August 2014}
+\leftright{\sl Summer Camp Counsellor}{\sl Fallis, AB, Canada}
+\smallskip
+Led children aged 4--15 through various activities (e.g. archery, canoeing, zipline) at a sleepaway camp.
+\medbreak
+
 \sectheader{Skills}
@@
 {\bf Programming Languages}\par
-\CEE, OCaml, Haskell, Standard ML, Scheme Lisp, PostScript, {\tt CWEB}, {\tt MIXAL}, Python
+\CEE, OCaml, Haskell, Standard ML, Scheme Lisp, PostScript, {\tt CWEB}, {\tt MIXAL}, Python.
 \medbreak
@@
 \thing Recipient of 0x\$0.40 in Knuth reward cheques.
+\medbreak
+
+\filbreak
 

### 2020-10-25 misc/tex-macros/gohlatex-demo.tex
GohLaTeX

% Test document GohLaTeX

\documentclass[10pt]{article}

\usepackage{gohlatex}

\begin{document}

\maketitle{Title}{a demo by}{Marcel K. Goh}{32 February 2143}

\gohsection 1. Introduction

This is a demo of all the features of \GohLaTeX. Shame on you for not learning Plain \TeX. This
is not a perfect imitation; the keen Plain \TeX\ user will notice that some things (e.g., spacing
between paragraphs) still isn't as beautiful as in Plain.

\proclaim Theorem 1. This is how to make a theorem.

\proof Prove the theorem afterwards.\slug

\parenproclaim Lemma 2 (Parentheses). Sometimes you want to name your theorems/lemmas.

\solution Sometimes you want to put the slug in display math mode. We have shown that
$$2+2 = 4.\noskipslug$$

Sometimes you want to typeset an algorithm:

\algbegin Algorithm A (Name). Description of algorithm.

\algstep A1. [Initialise.] Use \GohLaTeX.

\algstep A2. [Fall in love.] The formatting is so exquisite that you want to do things to it (write
your own \TeX\ macros).

\algstep A3. [Convert.] Switch to Plain \TeX.

\algstep A4. [Enlightenment.] Your soul transcends.\slug

I'm not sure why you have to leave a blank space between each algorithm step for it to work. You don't have
to do that in Plain \TeX. When the algorithm has $\geq 10$ steps, you'll want {\tt \char`\\aalgbegin} instead.

\aalgbegin Algorithm B (Math). These are some math macros I added. There are not too many of them and you
should probably use your own macros for other things you like.

\algstep B1. [Sets.] We have the inclusion $\NN\subseteq \ZZ \subseteq \QQ\subseteq  \RR\subseteq  \CC$.

\algstep B2. [Probability.] We find that $\pr\{A\} = 1$, $\ex\{X\} = 2$, and $\var\{X\} = \sigma^2$.

\algstep B3. [Indicators.] The indicator of an event $\one_A$ equals 1 if $A$ is true and 0 if $A$ is false.
You can also spell out the event; for example if $A$ is the event that $u\edge v$, then you can
write $\indic{u\edge v}$.

\algstep B4. [Dots.] We sometimes want to define $[1\twodots n] = \{1,2,\ldots,n\}$.

\algstep B5. [Equation numbers.] You can number your equations with old-style numerals:
$$[z^n]f(z) = {1\over 2\pi i}\oint {f(z)\over z^{n+1}}\;dz.\oldno 1$$

\algstep B6. [Reference.] You can reference an equation using \refeq{1234567890}.

\algstep B7. [Credit where it's due.] Some of these macros are lifted right out of {\tt plain.tex}, which
was written by Knuth himself.

\algstep B8. [Operators.] You can make your own operators and functions $\op{myfunc}(x)$ and they can
even have limits, like
$$\limitop{mylim}_{n\to\infty} \op{myfunc}(n).$$

\algstep B9. [Stalling.] Can't you tell I'm just trying to get to ten steps?

\algstep B10. [Slug.] Don't forget to end your algorithm with a slug!\slug

This is the end of a subsection

\medskip
\boldlabel Big bold label. Use this when you don't want to start a whole new section, but you still want
to break up your text.

\end{document}

### 2020-10-25 misc/tex-macros/gohlatex-demo.tex
Some typos in gohlatex

@@
 
-\algstep A2. [Fall in love.] The formatting is so exquisite that you want to do things to it (write
-your own \TeX\ macros).
+\algstep A2. [Fall in love.] The formatting is so exquisite that you want to do unspeakable things to it (like
+write your own \TeX\ macros).
 
@@
 
-This is the end of a subsection
+This is the end of a subsection.
 

### 2020-10-25 misc/tex-macros/gohlatex-demo.tex
More typos

@@
 \algstep B3. [Indicators.] The indicator of an event $\one_A$ equals 1 if $A$ is true and 0 if $A$ is false.
-You can also spell out the event; for example if $A$ is the event that $u\edge v$, then you can
+You can also spell out the event; for example, if $A$ is the event that $u\edge v$, then you can
 write $\indic{u\edge v}$.
@@
 
-\algstep B8. [Operators.] You can make your own operators and functions $\op{myfunc}(x)$ and they can
+\algstep B8. [Operators.] You can make your own operators and functions and they can
 even have limits, like
-$$\limitop{mylim}_{n\to\infty} \op{myfunc}(n).$$
+$$\limitop{mylim}_{n\to\infty} \op{myfunc}_n(x).$$
 

### 2020-11-13 about/vita.tex
Updated vita

@@
 
-\footline{\phantom{\datethis}\hss\folio\hss\datethis}  % \phantom for centering
+% \footline{\phantom{\datethis}\hss\folio\hss\datethis}  % \phantom for centering
 
@@
 
+\leftright{\bf \'Ecole Secondaire Beaumont Composite High School}{\bf September 2014 -- June 2015}
+\leftright{\sl Tutor}{\sl Beaumont, AB, Canada}
+\smallskip
+Tutored various students in grades 4 through 11 in chemistry, physics, math, and French.
+\medbreak
+
 \leftright{\bf YoWoChAs Outdoor Education Centre}{\bf June -- August 2014}
@@
 
-Analysis and design of algorithms, analysis of random discrete structures, combinatorics, group theory,
+Analysis and design of algorithms, analysis of random discrete structures, enumerative combinatorics,
+additive combinatorics, group theory,
 programming language theory and design, compilers.
@@
 \smallskip
-\thing Recipient of 0x\$0.40 in Knuth reward cheques.
+\thing Recipient of 0x\$3.40 in Knuth reward cheques.
 \medbreak
+
+\rightline{\eightrm \datethis}
 

### 2020-11-16 misc/solutions/papa-sols.tex
Fixed numbering in Papa solutions

@@
 \maketitlenodate{Answers to Selected Exercises in {\bigboldsl Real and Complex Analysis}\footnote{$^*$}{\frenchspacing \eightpt Walter Rudin.
-1987.
-{\eightsl Real and complex analysis, 3rd ed}.
+1966.
+{\eightsl Real and complex analysis}.
 McGraw-Hill, Inc., USA.}}{Solutions by}{Marcel K.\ Goh}
@@
 
-\proclaim 1. Does there exist an infinite $\sigma$-algebra which has only countably many members?
+\proclaim 2. Put $f_n = \chi_E$ if $n$ is odd, $f_n = 1 - \chi_E$ if $n$ is even. What is the relevance of this example to Fatou's lemma?
 
-The answer is no.
+This gives an example of strict inequality arising. Let $X$ be a measure space such that $\mu(X) = 1$ and let $E\subseteq X$ be such that $\mu(E) = 2/3$. Then
+$$\int_X \liminf_{n\rightarrow\infty} f_n \d\mu = \int_X 0 \d\mu = 0,$$
+while on the other hand
+$$\liminf_{n\rightarrow\infty} \int_X f_n\d\mu = \int_X 1 - \chi_E \d\mu = \int_X 1 \d\mu - \int_X \chi_E \d\mu = \mu(X) - \mu(E) = {1\over 3},$$
+and we have $0<1/3$.
 
-\proof Let $X$ be a ground set and let ${\cal F}$ be a $\sigma$-algebra (over $X$) with infinitely many members.
-First we claim that we can always find a set $E\neq \emptyset$ such that the set $\{ F \cap E^c : F\in {\cal F}\}$ is infinite.
-If this did not hold, then take any $E\neq \emptyset$ in ${\cal F}$.
-Then by assumption, ${\cal S}_1 = \{ F \cap E^c : F\in {\cal F}\}$ is finite and because $E^c\in {\cal F}$, ${\cal S}_2 = \{ F \cap E : F\in {\cal F}\}$ is finite as well.
-Since any member of ${\cal F}$ can be expressed as a union of an element of ${\cal S}_1$ with an element of ${\cal S}_2$, the implies that ${\cal F}$ is finite, a contradiction.
+\proclaim 3.
+Suppose $f_n : X\rightarrow[0,\infty]$ is measurable for $n=1,2,3,\ldots$, $f_1\geq f_2\geq f_3\geq\cdots\geq 0$, $f_n(x)\rightarrow f(x)$ as $n\rightarrow\infty$ for every $x\in X$, and $f_1\in L^1(\mu)$.
+Prove that then
+$$\lim_{n\rightarrow\infty} \int_X f_n \d\mu = \int_X f\d\mu$$
+and show that this conclusion does {\it not} follow if the condition ``$f_1\in L^1(\mu)$'' is omitted.
 
-Now we may use the claim to construct a countable sequence of pairwise disjoint elements of ${\cal F}$.
-Let $G_1$ be the set $E$ given by the claim.
-Now the infinite set ${\cal S}_1$ we constructed before is also a $\sigma$-algebra, so repeat the argument to get a set $G_2$, disjoint from $G_1$.
-Continuing in this manner, we obtain a sequence $G_1, G_2,\ldots$ where the $G_i$ are pairwise disjoint.
-Now we see that the map from the power set of the natural numbers to ${\cal F}$ given by
-$$ A \mapsto \bigcup_{i\in A} G_i $$
-is injective. So the uncountability of ${\cal F}$ follows from the uncountability of ${\cal P}({\bf N})$.\slug
+\proof Consider the sequence of functions $(f_1 - f_n)$.
+This sequence is nonnegative, nondecreasing and for any $x\in X$, $\lim_{n\rightarrow\infty} (f_1 - f_n)(x) = f_1(x) - f(x)$.
+So by the monotone convergence theorem,
+$$= \lim_{n\rightarrow \infty} \int_X f_1 - f_n \d\mu = \int_X f_1 - f\d\mu.$$
+Since the integrals of $f_n$ are finite, we can apply Theorem 1.27 to get
+$$\int_X f_1\d\mu - \lim_{n\rightarrow \infty}\int_X f_n \d\mu = \int_X f_1\d\mu -  \int_X f \d\mu,$$
+which gives us what we want after subtracting $\int_X f_1 \d\mu$ (which is finite) and multiplying by $-1$.\slug
 
-\proclaim 3. Prove that if $f$ is a real function on a measurable space $X$ such that $\{x:f(x)\geq r\}$ is measurable for every rational $r$, then $f$ is measurable.
+The condition $f_1\in L^1(\mu)$ is necessary.
+Let $I = [0,1]$ and $f_n = {1\over nx^2}$, which converges to the constant function 0.
+Then $\lim_{n\rightarrow \infty} \int_I f_n \d\mu = +\infty$ but $\int_I 0\d\mu = 0$.
+
+\proclaim 4. Prove that if $f$ is a real function on a measurable space $X$ such that $\{x:f(x)\geq r\}$ is measurable for every rational $r$, then $f$ is measurable.
 
@@
 
-\proclaim 5. Prove the following statements:
-\medskip
-\item{a)} If $f:X\rightarrow[-\infty,\infty]$ and $g:x\rightarrow[-\infty,\infty]$ are measurable, then the sets
-$$\{x : f(x) < g(x)\}\ \hbox{and}\ \{x : f(x) = g(x)\}$$
-are measurable.
-\item{b)} The set of points at which a sequence of measurable real-valued functions converges (to a finite limit) is measurable.
+\proclaim 5. Prove that the set of points at which a sequence of measurable real-valued functions converges (to a finite limit) is measurable.
 
-\proof Observe that $f(x) < g(x)$ if and only if there exists a rational number $r$ such that $f(x) < r < g(x)$. So
-$$\{x : f(x) < g(x)\} = \bigcup_{r\in \Q} \big(\{x : f(x) < r\} \cap \{x : g(x) > r\}\big) = \bigcup_{r\in \Q} \Big(f^{-1}\big([-\infty,r)\big) \cap g^{-1}\big((r,\infty]\big)\Big)$$
-is a countable union of measurable sets (since $f$ and $g$ are measurable functions). Then
-$$\{x : f(x) = g(x)\} = \{x : f(x) < g(x)\}^c \cap \{x : g(x) < f(x)\}^c$$
-is measurable as well, proving (a).
-
-Let $(f_n)$ be a sequence of real-valued measurable functions.
+\proof Let $(f_n)$ be a sequence of real-valued measurable functions.
 Then $(f_n)$ converges at a point $x$ if and only if it is Cauchy at $x$, i.e.\ for any $\epsilon > 0$ there exists $N\in \N$ such that for all $m,n\geq N$, $|f_m(x) - f_n(x)| < \epsilon$.
@@
 
-First we prove that ${\cal M}$ is a $\sigma$-algebra and that $\mu$ is a measure.
+\noindent First we prove that ${\cal M}$ is a $\sigma$-algebra and that $\mu$ is a measure.
 
@@
 
-Now we claim that the measurable functions are those that are constant at all but countably many points.
+\noindent Now we claim that the measurable functions are those that are constant at all but countably many points.
 
@@
 
-Lastly, we show that if $f$ is measurable and takes on the value $k$ at all but countably many points, then $\int_E f\d\mu = k$ for all uncountable $E\in {\cal M}$.
+\noindent Lastly, we show that if $f$ is measurable and takes on the value $k$ at all but countably many points, then $\int_E f\d\mu = k$ for all uncountable $E\in {\cal M}$.
 (If $E$ is countable then $\int_E f\d\mu = 0$.)
@@
 
-\proclaim 7.
-Suppose $f_n : X\rightarrow[0,\infty]$ is measurable for $n=1,2,3,\ldots$, $f_1\geq f_2\geq f_3\geq\cdots\geq 0$, $f_n(x)\rightarrow f(x)$ as $n\rightarrow\infty$ for every $x\in X$, and $f_1\in L^1(\mu)$.
-Prove that then
-$$\lim_{n\rightarrow\infty} \int_X f_n \d\mu = \int_X f\d\mu$$
-and show that this conclusion does {\it not} follow if the condition ``$f_1\in L^1(\mu)$'' is omitted.
+\proclaim 7. Does there exist an infinite $\sigma$-algebra which has only countably many members?
 
-\proof Consider the sequence of functions $(f_1 - f_n)$.
-This sequence is nonnegative, nondecreasing and for any $x\in X$, $\lim_{n\rightarrow\infty} (f_1 - f_n)(x) = f_1(x) - f(x)$.
-So by the monotone convergence theorem,
-$$= \lim_{n\rightarrow \infty} \int_X f_1 - f_n \d\mu = \int_X f_1 - f\d\mu.$$
-Since the integrals of $f_n$ are finite, we can apply Theorem 1.27 to get
-$$\int_X f_1\d\mu - \lim_{n\rightarrow \infty}\int_X f_n \d\mu = \int_X f_1\d\mu -  \int_X f \d\mu,$$
-which gives us what we want after subtracting $\int_X f_1 \d\mu$ (which is finite) and multiplying by $-1$.\slug
+\noindent The answer is no.
 
-The condition $f_1\in L^1(\mu)$ is necessary.
-Let $I = [0,1]$ and $f_n = {1\over nx^2}$, which converges to the constant function 0.
-Then $\lim_{n\rightarrow \infty} \int_I f_n \d\mu = +\infty$ but $\int_I 0\d\mu = 0$.
+\proof Let $X$ be a ground set and let ${\cal F}$ be a $\sigma$-algebra (over $X$) with infinitely many members.
+First we claim that we can always find a set $E\neq \emptyset$ such that the set $\{ F \cap E^c : F\in {\cal F}\}$ is infinite.
+If this did not hold, then take any $E\neq \emptyset$ in ${\cal F}$.
+Then by assumption, ${\cal S}_1 = \{ F \cap E^c : F\in {\cal F}\}$ is finite and because $E^c\in {\cal F}$, ${\cal S}_2 = \{ F \cap E : F\in {\cal F}\}$ is finite as well.
+Since any member of ${\cal F}$ can be expressed as a union of an element of ${\cal S}_1$ with an element of ${\cal S}_2$, the implies that ${\cal F}$ is finite, a contradiction.
 
-\proclaim 8. Put $f_n = \chi_E$ if $n$ is odd, $f_n = 1 - \chi_E$ if $n$ is even. What is the relevance of this example to Fatou's lemma?
-
-This gives an example of strict inequality arising. Let $X$ be a measure space such that $\mu(X) = 1$ and let $E\subseteq X$ be such that $\mu(E) = 2/3$. Then
-$$\int_X \liminf_{n\rightarrow\infty} f_n \d\mu = \int_X 0 \d\mu = 0,$$
-while on the other hand
-$$\liminf_{n\rightarrow\infty} \int_X f_n\d\mu = \int_X 1 - \chi_E \d\mu = \int_X 1 \d\mu - \int_X \chi_E \d\mu = \mu(X) - \mu(E) = {1\over 3},$$
-and we have $0<1/3$.
+Now we may use the claim to construct a countable sequence of pairwise disjoint elements of ${\cal F}$.
+Let $G_1$ be the set $E$ given by the claim.
+Now the infinite set ${\cal S}_1$ we constructed before is also a $\sigma$-algebra, so repeat the argument to get a set $G_2$, disjoint from $G_1$.
+Continuing in this manner, we obtain a sequence $G_1, G_2,\ldots$ where the $G_i$ are pairwise disjoint.
+Now we see that the map from the power set of the natural numbers to ${\cal F}$ given by
+$$ A \mapsto \bigcup_{i\in A} G_i $$
+is injective. So the uncountability of ${\cal F}$ follows from the uncountability of ${\cal P}({\bf N})$.\slug
 

### 2020-11-21 misc/expo/littlewood-offord.tex
littlewood-offord

\input fontmac
\input mathmac

\classicmode

\def\norm#1{|\!|#1|\!|}
\def\C{{\cal C}}
\def\F{{\cal F}}
\def\eps{\epsilon}

\maketitle{The Littlewood-Offord problem}{notes by}{Marcel K. Goh}{20 November 2020}

\medskip
Let $x_1,x_2,\ldots,x_n$ be vectors in some normed space $B$ of norm at least 1.
There are $2^n$ ways to form sums out of these
vectors (where the empty sum is taken to be 0) and the Littlewood-Offord problem asks how many of these sums
differ by less than 1. In its original 1943 formulation by J. E. Littlewood and A. C. Offord,
the normed space in question was the complex plane $\CC$.

Before we get started, let us recall the properties of a {\it normed space}. It is a vector space $B$ over
a field $F$ along with
a function $\norm{\cdot} : B \to \RR$ such that
\medskip
\item{i)} $\norm x\geq 0$ for all $x\in B$;
\smallskip
\item{ii)} $\norm x = 0$ if and only if $x = 0$;
\smallskip
\item{iii)} for every $\lambda \in F$, $\norm{\lambda x} = |\lambda|\norm x$;
\smallskip
\item{iv)} the triangle inequality $\norm{x+y} \leq \norm x + \norm y$ for all $x,y\in B$.
\medskip
Note that in (iii), we have assumed some kind of norm on the field $F$. In the case that $F = \RR$, this
is the ordinary absolute value function, and in the case that $F=\CC$, we can take the modulus.
The {\it distance} between two vectors $x$ and $y$ is the norm of the difference $x-y$.

On a completely different note, we will also need some notions regarding families of subsets of a finite set.
If $X = \{1,2,\ldots,n\}$,
then the power set $2^X$ has a natural partial order given by inclusion. A {\it chain} is a set of subsets
$\{A_1,A_2, \ldots, A_k\}$ of $X$ such that $A_1\subseteq A_2\subseteq\cdots\subseteq A_k$, while an
{\it antichain} or {\it Sperner family} is a family of subsets $\{A_1,A_2\ldots,A_k\}$ such that if $i\neq j$
then $A_i\not\subseteq A_j$, for any $1\leq i,j \leq k$. A chain is {\it maximal} if one cannot add any set
to the family without violating the chain property. A chain can be as large as $|X|+1$, since for example,
we have the maximal chain
$$\emptyset\subseteq \{1\}\subseteq \{1,2\} \subseteq \cdots \subseteq X.$$
How big can an antichain be? Well, we could take every set to be of the same size $i$, and this would certainly
form an antichain. Since ${n\choose i}$ is maximised, when $i =\lfloor n/2\rfloor$, we see that the maximum
size of an antichain must be at least ${n\choose \lfloor n/2\rfloor}$. The following famous theorem shows that
we can do no better.

\parenproclaim Theorem S (Sperner, {\rm 1928}). Let $X = \{1,2,\ldots, n\}$ and let $\F\subseteq 2^X$ be an
antichain. Then $|\F|\leq {n\choose \lfloor n/2\rfloor}$.

\proof We will double count the number of pairs $(\C, A)$ where $\C$ is a maximal chain and
$A\in \F\cap\C$. Observe that any permutation $\sigma_1, \sigma_2, \ldots, \sigma_n$
of $\{1,2,\ldots,n\}$ defines a unique maximal chain
$$\emptyset \subseteq \{\sigma_1\} \subseteq \{\sigma_1,\sigma_2\}\subseteq \cdots\subseteq\{\sigma_1,\sigma_2,
\ldots,\sigma_n\}.$$
Since the intersection of a chain with the antichain $\F$ consists of at most one set $A$, the number of
pairs $(\C,A)$ is at most $n!$.

On the other hand, for a fixed set $A = \{x_1, x_2,\ldots, x_k\}\in \F$,
we must ask how many chains actually contain $A$. Note that a maximal chain $\C$ contains $A$ if and only if
the first $k$ nonempty sets in the chain introduce the elements $x_1,x_2,\ldots x_k$ in some order. After
that, the elements of $X\setminus A$ can be introduced in any order. So for a fixed $A$, there are $|A|!(n-|A|)!$
pairs $(\C,A)$. We have deduced that
$$\sum_{A\in \F} |A|!(n-|A|)! \leq n!,$$
and dividing both sides by $n!$, we have
$$\sum_{A\in \F} {n\choose |A|}^{-1} \leq 1.$$
Note that ${n\choose |A|}\leq {n\choose \lfloor n/2\rfloor}$ for all $A$, so we find that
$${|\F|\over {n\choose \lfloor n/2\rfloor}} \leq \sum_{a\in \F}{n\choose |A|}^{-1}\leq 1,$$
which allows us to conclude that $|\F|\leq {n\choose \lfloor n/2\rfloor}$.\slug

\parenproclaim Lemma R (Erd\H os, {\rm 1945}). Let $x_1, x_2, \ldots, x_n$ be real numbers with absolute
value at least $1$. At most ${n\choose \lfloor n/2\rfloor}$ of their sums differ by less than $1$ from
each other.

\proof For $A\subseteq \{1,2,\ldots,n\}$, let
$$x_A = \sum_{i\in A} x_i.$$
Without loss of generality, we may assume that all the $x_i$ are positive. This is because replacing
$x_i$ with $-x_i$ and $A$ with $A\triangle \{i\}$ does not change the relative differences between
the sums; this operation causes the $x_A$ to permute amongst themselves and increase by $x_i$.

Let $\F$ be a family of subsets of $\{1,2,\ldots,n\}$ such that $|x_A - x_B|<1$ for every pair of distinct
$A,B\in \F$. It is an antichain because if $A\subseteq B$ then $x_B>x_A$ and in particular,
$$|x_B - x_A| = \sum_{x\in B}x_i - \sum_{x\in A} x_i = \sum_{x\in B\setminus A} x_i = x_{B\setminus A}\geq 1.$$
So at most one of $A$ and $B$ can be in $\F$. Applying Theorem S, $|\F|\leq {n\choose \lfloor n/2\rfloor}$.\slug

Note that Lemma R gives the best possible bound, because if $x_1 = x_2 = \cdots = x_n = c$ for some positive
constant $c$, then ${n\choose \lfloor n/2\rfloor}$ of the sums are equal to $c\lfloor n/2\rfloor$. With a bit
of extra work, we can derive the following statement (and in fact, it is more than a corollary, because it is
equivalent to Lemma R).

\proclaim Corollary T.
Let $x_1,x_2,\ldots x_n$ be real numbers, all with absolute value at least $1$ and consider
the $2^n$ different sums $\sum_{i=1}^n \eps_ix_i$, where $\eps_i$ is either $-1$ or $1$. For any $x\in \RR$ with
$|x|\geq 1$, at most ${n\choose \lfloor n/2\rfloor}$ of them are at distance less than $1$ from $x$.\slug

\section References

\bye

### 2020-11-30 misc/expo/littlewood-offord.tex
nullstellensatz

@@
 \input mathmac
-
-\classicmode
 
@@
 \def\eps{\epsilon}
+
+\classicmode
 
@@
 
-Before we get started, let us recall the properties of a {\it normed space}. It is a vector space $B$ over
-a field $F$ along with
-a function $\norm{\cdot} : B \to \RR$ such that
-\medskip
-\item{i)} $\norm x\geq 0$ for all $x\in B$;
-\smallskip
-\item{ii)} $\norm x = 0$ if and only if $x = 0$;
-\smallskip
-\item{iii)} for every $\lambda \in F$, $\norm{\lambda x} = |\lambda|\norm x$;
-\smallskip
-\item{iv)} the triangle inequality $\norm{x+y} \leq \norm x + \norm y$ for all $x,y\in B$.
-\medskip
-Note that in (iii), we have assumed some kind of norm on the field $F$. In the case that $F = \RR$, this
-is the ordinary absolute value function, and in the case that $F=\CC$, we can take the modulus.
-The {\it distance} between two vectors $x$ and $y$ is the norm of the difference $x-y$.
-
-On a completely different note, we will also need some notions regarding families of subsets of a finite set.
+Before we get started, we will also need some notions regarding families of subsets of a finite set.
 If $X = \{1,2,\ldots,n\}$,
@@
 
+P. Erd\H os used Sperner's theorem to give the following neat solution to the Littlewood-Offord problem
+in the case $B=\RR$:
+
 \parenproclaim Lemma R (Erd\H os, {\rm 1945}). Let $x_1, x_2, \ldots, x_n$ be real numbers with absolute

### 2020-12-04 misc/expo/furstenberg.tex
Furstenberg

\input fontmac
\input mathmac

\def\F{{\cal F}}

\classicmode

\maketitle{Ergodic theory and arithmetic progressions}{by}{Marcel K. Goh}{4 December 2020}

\advsect Introduction

Let $(Z,+)$ be an abelian group. For any integer $n$ and $z\in Z$ we let $nz$ denote the iterated sum; for
instance, $2z = z+z$ and $-3z = -z-z-z$.
Letting $[a,b]$ denote the discrete interval $\{n\in \ZZ:a\leq n\leq b\}$
for $a,b\in \ZZ$, we define an {\it arithmetic progression} to be a set of the form
$$a+[0,k-1]\cdot r = \{a,a+r,a+2r,\ldots,a+(k-1)r\},\adveq$$
where $k>1$ is the {\it length}, $a\in Z$ is the {\it base point}, and $r\in Z$ is the {\it common difference}
or {\it step}. If $r\neq 0$, the progression is {\it nontrivial}, and this condition is assumed unless otherwise
stated. In these notes, we will study the case $Z = \ZZ$.

The quest to find arithmetic progressions in subsets of $\ZZ$ began with the following celebrated theorem
of B. L. van der Waerden, which answered a conjecture of P. J. H. Baudet.

\parenproclaim Theorem V (van der Waerden, {\rm 1927}). For any integers $r$ and $k$, there exists $N\in \NN$
such that for any partition of $[1,N]$ into disjoint sets $C_1, C_2, \ldots, C_r$, some $C_i$ contains
a $k$-term arithmetic progression.\slug

\noindent
An infinitary formulation of this theorem states that if the positive integers are coloured with $r$ colours,
then some colour contains arbitrarily long arithmetic progressions. P. Erd\H os and P. Tur\'an suspected that
this had to do with the density of the partition. A subset $A$ of the nonnegative integers is said to have
{\it positive (upper) density} if
$$\limsup_{N\to \infty} {\big|A\cap [0,N]\big|\over N+1} > 0.\adveq$$
In the more general setting where $A\subseteq \ZZ^d$ for $d\geq 1$, we get a similar definition by intersecting
with $[-N,N]^d$ and dividing by $(2N+1)^d$. (It does not really matter exactly which interval we intersect
with, since we only care to know whether the density is positive or zero.)
Erd\H os and Tur\'an's conjecture was that van der Waerden's
theorem holds because at least one of the $C_i$ must have density at least $1/r$. This was proved for the case $k=3$
by K. F. Roth (1953) using Fourier-analytic methods. A proof for $k=4$ did not arrive until 1969;
it was due to E. Szemer\'edi, who also proved the general case six years later:

\parenproclaim Theorem S (Szemer\'edi, {\rm 1975}). Any subset of the nonnegative integers with positive upper
density contains arithmetic progressions of arbitrary length. Equivalently, given any real $\delta > 0$ and
$k\in \NN$, there exists $N\in \NN$ such that for any $A\subseteq [0,N]$ with $|A|\geq \delta(N+1)$, there exists
an arithmetic progression of length $k$ contained in $A$.

\noindent
The proof that Szemer\'edi gave was long and complicated, relying on an auxiliary ``graph-regularity lemma''
that has had far-reaching consequences in combinatorics. We will only supply a proof of the fact
that the infinitary version of the theorem is in fact equivalent to the finitary one. (This treatment
can be found in Furstenberg, Katznelson, and Ornstein (1982).)

\proof It is clear that the finitary statement implies the infinitary one. Now suppose that the finitary
statement does not hold; that is, there exists $\delta>0$ and $k\in \NN$ such that for all $N\in \NN$,
there exists $A_N\subseteq [0,N]$ such that $|A_N|\geq \delta(N+1)$ but there is no arithmetic progression
of length $k$ contained in $A$. Note that the presence of arithmetic progressions in a set $A$ does not
change if we shift all the elements of the set by a fixed value: for any $s\in \NN$,
$s+A = \{s + a : a\in A\}$ has the
same number of arithmetic progressions as $A$. So we can translate the sets $A_N$ far enough apart that no
arithmetic progression can belong to two of the $A_N$ at once. Letting $A = \bigcup_{N\in \NN} A_N$,
we find that $A$ has upper density $\geq \delta$, but does not contain arithmetic progressions of length $k$.\slug

Szemer\'edi's quantitative bounds on the size of $N$ relative to $\delta$ and $k$ were not as tight
as Roth's bounds for the $k=3$ case. In fact, Roth's Fourier-analytic approach was generalisable,
but this was not discovered until 1998 for the case
$k=4$ and finally for the general case in 2001 (both of these results, which established much better bounds,
are due to W. T. Gowers). On the other
hand, if we are only interested in the infinitary statement and allow non-constructive proofs, then ergodic
methods can be applied to give much shorter arguments than were employed in Szemer\'edi's original paper.
This approach, devised by H. Furstenberg in 1977, led to many generalisations of Szemer\'edi's result.

\advsect The ergodic formulation

We will need some terminology from the realm of dynamical systems.
Let $(X,\F,\mu)$ be a probability space and let $T:X\to X$ be measure-preserving, i.e.,
$\mu(T^{-n} A) = \mu(A)$ for all $A\in \F$. Note that $T^{-n}A$ is the set of all $x\in X$ such that $Tx\in A$;
one should not be tricked into thinking that $T$ is necessarily invertible!
The quadruple $(X,\F,\mu,T)$ is called a {\it measure-preserving
system}. For a bounded measurable function $f:X\to\RR$, $T$ acts as a shift operator by sending $f$ to $T^n f$,
given by $T^nf(x) = f(T^nx)$. We have the following lemma.

\proclaim Lemma D. Let $(X,\F,\mu)$ be a probability space and let $f$ be a measurable function on $X$
such that $\int_Xf\d\mu>0$ and such that for all $x\in X$, $0\leq f(x)\leq C$ for some constant $C$.
Let $\delta>0$ be such that
$$G = \{x\in X : f(x)\geq \delta\}$$
has positive measure. For any $T:X\to X$ that is measure-preserving,
there exists $E_T$ with $\mu(E_T)\geq \mu(G)/2$ such that the set
$$R_x = \{n\in \NN : T^nf(x) \geq \delta\}$$
has upper density $\geq \mu(G)/2$ for every $x\in E_T$.

\noindent This is a variant of Lemma 1 from a blog post by Terry Tao, dated 10 February 2008. The
proof we supply follows the same outline (with some details added).

\proof First, note that $\int_Xf\d\mu\leq C$, since $X$ is a probability space. Next, notice that
the for any $x\in X$ and any $n$, $T^nf(x)\geq \delta$ if and only if $f(T^nx)\geq \delta$,
which is equivalent to
$x\in T^{-n}G$. Because $\mu$ is shift-invariant, we have
\newcount\eqA
\eqA=\eqcount
$$\mu(G) = \int_X {1\over N} \sum_{n=1}^N \chi_{G}\d\mu = \int_X{1\over N}\sum_{n=1}^N \chi_{T^{-n}G}\d\mu.\adveq$$
The integrand on the right-hand side no more than $1$, so the set
$$A_N = \bigg\{ x\in X : {1\over N} \sum_{n=1}^N \chi_{T^{-n}G}(x)\geq {\mu(G)\over 2} \bigg\}$$
has measure at least $\mu(G)/2$. For otherwise, we would have
$$\int_X {1\over N}\sum_{n=1}^N \chi_{T^{-n}G}\d\mu \leq \int_{A_N} 1
  + \int_{{A_N}^c} {1\over N}\sum_{n=1}^N \chi_{T^{-n}G}\d\mu < {\mu(G)\over 2} + {\mu({A_N}^c)\mu(G)\over 2}
  \leq\mu(G),\adveq$$
contradicting \refeq{\the\eqA}. Note that
$$\limsup_{N\to\infty} A_N = \bigcap_{N=1}^\infty \bigcup_{n=N}^\infty A_N$$
is the set of all $x\in X$ that are in $A_N$ for infinitely many $N$. Put another way, this is the set
of all $x\in X$ for which
$$\limsup_{N\to\infty} {R_x \cap [1,N]\over N} \geq {\mu(G)\over 2}.$$
Since $\mu(A_N)\geq \mu(G)/2$ for every $N$, we have
$$\mu\Big(\limsup_{N\to\infty} A_N\Big) = \mu\Big(\bigcap_{n=1}^\infty \bigcup_{n=N}^\infty A_n\Big)\geq
\mu(G)/2\adveq$$
and $E_T = \limsup_{N\to\infty} A_N$ is a set of positive measure such that for every point $x\in E_T$,
$R_x$ has upper density $\geq \mu(G)/2$.\slug

\noindent We can now formulate Furstenberg's multiple recurrence theorem:

\parenproclaim Theorem F (Furstenberg, {\rm 1977}). Let $(X,\F,\mu,T)$ be a measure-preserving
system and let $f$ be a bounded measurable function such that $\int_X f\d\mu >0$. For any $k\in \NN$, we have
$$\liminf_{N\to\infty} {1\over N} \sum_{s=1}^N \int_X f (T^sf)\cdots (T^{(k-1)s}f)\d\mu>0.\noskipslug\adveq$$

In the particular case that $k=2$, this is a form of Poincar\'e's recurrence theorem. We will not be able
to prove Theorem F directly in these notes, but we will prove the important correspondence between Furstenberg's
multiple recurrence and arithmetic progressions in $\NN_0$.

\proclaim Lemma E. Theorems S and F are equivalent.

\proof First, suppose that Theorem S holds. Let $(X,\F,\mu,T)$ be a measure-preserving system, $f$ a bounded
measurable function with $\int_X f\d\mu>0$, and let $k\in \NN$ be arbitrary. Let $\delta>0$ be small enough that
$G = \{x\in X : f(x)\geq \delta\}$ has positive measure; by Lemma D, there exists a set $E_T$ of measure at least
$\mu(G)/2$ such that $R_x = \{n\in \NN : T^nf(x)\ge\delta\}$ has upper density $\ge\mu(G)/2$ for every $x\in E_T$.
By Theorem S, there exists an $N_0$ such that for any $x\in E_T$, there is a $k$-term arithmetic progression
in $R_x\cap [0,N_0]$. But the set of $k$-term arithmetic progressions
in $[0,N_0]$ is finite, with some cardinality $K$;
as a crude upper bound we have $K\leq {N_0\choose k}$. So there exists
some fixed progression $P_T = \{a, a+r, \ldots, a+(k-1)r\}\subseteq [0,N_0]$
such that the set $E'_T = \{x\in E_T : P_T\subseteq R_x\}$ has $\mu(E'_T)\ge\mu(G)/(2K)$. We have
\newcount\eqC
\eqC=\eqcount
$$\eqalign{
\int_X f(T^rf)\cdots(T^{(k-1)r}f)\d\mu &\geq \int_{T^{-a}E'_T} f(T^rf)\cdots(T^{(k-1)r}f)\d\mu \cr
&= \int_{E'_T} T^af(T^{a+r}f)\cdots(T^{a+(k-1)r}f)\d\mu\cr
&\geq {\delta^k\mu(G)\over 2K}.\cr
}\adveq$$
The progression $P_T$ is particular to the transformation $T$, but the bound depends only on $f$ and $k$,
so in the argument above we could have applied Lemma D to $T^m$ for $m\in \NN$ and obtained \refeq{\the\eqC}
with $T^m$ in place of $T$. Since $r$ cannot exceed $N_0$ no matter what $T^m$ is, we find that
$${1\over N_0} \sum_{s'=1}^{N_0} \int_X f(T^{ms'}f)\cdots(T^{(k-1)ms'})\d\mu \geq {\delta^k\mu(G)\over 2K},\adveq$$
where $m\in \NN$ is arbitrary. For simplicity's sake, suppose that $N\geq {N_0}^2$ and averaging over all
$1\leq m\leq N/N_0$, we have
$${N_0\over N} \sum_{m=1}^{N/N_0}
{1\over N_0} \sum_{s'=1}^{N_0} \int_X f(T^{ms'}f)\cdots(T^{(k-1)ms'})\d\mu \geq {\delta^k\mu(G)\over 2K},\adveq$$
and since every $1\leq s\leq N$ has at most $\min(N_0, N/N_0) = N_0$ representations of the form $ms'$, where
$1\leq m\leq N/N_0$ and $1\leq s'\leq N_0$, this gives
$${1\over N} \sum_{s=1}^{N} \int_X f(T^{s}f)\cdots(T^{(k-1)s})\d\mu \geq {\delta^k\mu(G)\over 2KN_0}\adveq$$
for all large enough $N$, yielding Theorem F\footnote{$^*$}{\eightpoint
Thank you to Terry Tao, whose answer to my question on his blog
helped me get the correct constants in the final bound.}.

For the other direction, fix a subset $A\subseteq \NN_0$ with positive upper density and let $k\in \NN$.
We can find a sequence of natural numbers
$N_1,N_2,\ldots$ such that
$$\liminf_{j\to\infty} {A \cap [0,N_j] \over N_j+1} >0.$$
Consider the linear space $\ell^\infty(\NN)$ of all bounded real-valued sequences under the supremum norm. By
the Hahn-Banach theorem applied to the functional $\lambda(x) = \lim_{j\to\infty} x_j$, which is defined
on the subspace of convergent sequences, we can construct a linear functional $\Lambda : \ell^\infty(\NN)\to\RR$
such that
$$\liminf_{j\to\infty} x_j \leq \Lambda(x) \leq \limsup_{j\to\infty} x_j.\adveq$$
Now let $X = \{0,1\}^{\NN_0}$ be the space of all binary sequences indexed on the nonnegative integers
with the product topology and Borel $\sigma$-algebra $\F$. Let $T$ be the left-shift operator given by
$Tx = (x_{n+1})_{n\in \NN_0}$ and let $a\in X$ be given by
$$a_n = \cases{1,&if $n\in A$;\cr0,&otherwise.\cr}\adveq$$
For any sequence of indices $s_1,\ldots,s_m \in \NN_0$, define a measure $\mu_{s_1,\ldots,s_m}$ on $\{0,1\}^m$
given by
$$\mu_{s_1,\ldots,s_m}(E_1\times\ldots\times E_m) = \Lambda\Bigg(\!\!\bigg(
{1\over N_j+1} \sum_{i=0}^{N_j} \chi_{E_1}(T^{s_1}a_i)\cdots\chi_{E_m}(T^{s_m}a_i)\bigg)_{j=1}^\infty\Bigg),\adveq$$
where $E_1,\ldots,E_m\subseteq \{0,1\}$.
Permuting the indices $s_1, \ldots,s_m$ does not change the value on the right-hand
side (multiplication is commutative), and for any $s_{m+1}\in \NN_0$,
$$\eqalign{
\mu_{s_1,\ldots,s_m,s_{m+1}}(E_1\times\ldots\times E_m\times \{0,1\})
&= \Lambda\Bigg(\!\!\bigg(
{1\over N_j+1} \sum_{i=0}^{N_j} \chi_{E_1}(T^{s_1}a_i)\cdots\chi_{E_m}(T^{s_m}a_i)\chi_{\{0,1\}}(T^{s_{m+1}}a_i)
\bigg)_{j=1}^\infty\Bigg)\cr
&= \Lambda\Bigg(\!\!\bigg(
{1\over N_j+1} \sum_{i=0}^{N_j} \chi_{E_1}(T^{s_1}a_i)\cdots\chi_{E_m}(T^{s_m}a_i)\bigg)_{j=1}^\infty\Bigg)\cr
&= \mu_{s_1,\ldots,s_m}(E_1\times\ldots\times E_m).
}\adveq$$
So by the Kolmogorov consistency theorem, there exists a measure $\mu$ on $X$ such that for all $s_1,\ldots s_m\in
\NN_0$, and $E_1,\ldots,E_m\subseteq \{0,1\}$,
$$\mu\big(\{x \in X : x_{s_1}\in E_1,\,\ldots,\,x_{s_m}\in E_m\}\big)
= \mu_{s_1,\ldots,s_m}(E_1\times\ldots\times E_m).$$
This is ``almost'' invariant under the shift $T$, since
$$\eqalign{
\mu\big(T^{-1}\{x \in X : x_{s_1}\in E_1,\,\ldots,\,x_{s_m}\in E_m\}\big)
&= \mu\big(\{x \in X : x_{s_1+1}\in E_1,\,\ldots,\,x_{s_m+1}\in E_m\}\big)\cr
&= \mu_{s_1+1,\ldots,s_m+1}(E_1\times\ldots\times E_m)\cr
&= \Lambda\Bigg(\!\!\bigg(
{1\over N_j+1} \sum_{i=0}^{N_j} \chi_{E_1}(T^{s_1+1}a_i)\cdots\chi_{E_m}(T^{s_m+1}a_i)\bigg)_{j=1}^\infty\Bigg) \cr
&= \Lambda\Bigg(\!\!\bigg(
{1\over N_j+1} \sum_{i=1}^{N_j+1} \chi_{E_1}(T^{s_1}a_i)\cdots\chi_{E_m}(T^{s_m}a_i)\bigg)_{j=1}^\infty\Bigg).\cr
}\adveq$$
But as $j\to\infty$, shifting the terms over which we take the average does not change the $\limsup$ or the
$\liminf$, so this is the same as
$\mu\big(\{x \in X : x_{s_1}\in E_1, \ldots, x_{s_m}\in E_m\}\big)$, provided that the sequence
$$\Bigg(\!\!\bigg(
{1\over N_j+1} \sum_{i=0}^{N_j} \chi_{E_1}(T^{s_1}a_i)\cdots\chi_{E_m}(T^{s_m}a_i)\bigg)_{j=1}^\infty\Bigg)$$
is convergent.
This will be enough for our purposes (though there should be some way to fix it in general).
Let $B$ be the cylinder of all $x\in X$
with $x_0 = 1$. Note that
$$\eqalign{
\mu(B) &= \mu\big(\{x \in X : x_0 = 1\}\big) \cr
&= \Lambda\Bigg(\!\!\bigg({1\over N_j+1} \sum_{i=0}^{N_j} \chi_{\{1\}}(a_i)\bigg)_{j=1}^\infty\Bigg) \cr
&= \Lambda\Bigg(\!\!\bigg({A\cap [0,N_j]\over N_j+1}\bigg)_{j=1}^\infty\Bigg) \cr
&= \liminf_{j\to\infty}{A\cap [0,N_j]\over N_j+1},\cr
}\adveq$$
and by our choice of the sequence $N_j$, this is positive and equals the $\limsup$.
By Theorem F applied to $k$ and the function $\chi_B$,
there exists some $r$ such that
$$\mu(B\cap T^{-r}B \cap \cdots\cap T^{-(k-1)r}B) > 0.$$
Letting $A-s$ denote the set $\{a-s : a\in A\}$, and working backwards, we discover that
$$\eqalign{
\limsup_{j\to\infty}&{A\cap (A-r)\cap \cdots \cap (A-(k-1)r)\cap [0,N_j]\over N_j+1}\cr
&\geq\Lambda\Bigg(\!\!\bigg({A\cap (A-r)\cap \cdots \cap (A-(k-1)r)\cap [0,N_j]\over N_j+1}\bigg)_{j=1}^\infty\Bigg)\cr
&=\Lambda\Bigg(\!\!\bigg(
{1\over N_j+1}\sum_{i=0}^{N_j}\chi_{\{1\}}(a_i)\chi_{\{1\}}(T^ra_i) \cdots \chi_{\{1\}}(T^{(k-1)r}a_i)
\bigg)_{j=1}^\infty\Bigg)\cr
&= \mu\big(\{x\in X : x_0 = 1,\,x_r = 1,\,\ldots,\,x_{(k-1)r} = 1\}\big)\cr
&= \mu(B\cap T^{-r}B \cap \cdots\cap T^{-(k-1)r}B),\cr
}\adveq$$
and since this is positive, there exists an arithmetic progression of length $k$ in $A$.\slug

The first part of the presented proof draws from the aforementioned blog post of Terry Tao;
the second part is a fleshed-out version of the proof sketch found in Tao and Vu (2006).

\advsect Generalisations

The techniques that Furstenberg used to link arithmetic progressions to recurrence phenomena came to be known
as the Furstenberg correspondence principle and, as mentioned above, it led to various generalisations of
Szemer\'edi's theorem. We list a few of them here, along with their equivalent statements in the domain
of dynamical systems.

\parenproclaim Theorem M (Furstenberg, Katznelson, {\rm 1979}). The following are equivalent.
\medskip
\item{i)} Let $d\geq 1$ and let $A\subseteq \ZZ^d$ have positive upper density, that is,
$$\limsup_{N\to\infty} {A\cap [-N,N]^d\over (2N+1)^d} \geq 0.$$
For any $v_1,\ldots,v_k\in \ZZ^d$, there exist infinitely many pairs $(a,r)\in \ZZ^d$ such that
$\{a+rv_1,\ldots,a+rv_k\} \subseteq A$.
\smallskip
\item{ii)} Let $(X,\F,\mu)$ be a probability space and $k\in \NN$. If $T_1,T_2,\ldots,T_k:X\to X$ are
measure-preserving maps that commute wieh each other and $E$ is a set of positive measure, then there exists
$r>0$ such that ${T_1}^rE\cap {T_2}^rE\cap\cdots\cap{T_k}^rE$ is nonempty.\slug

\noindent This multidimensional version of Szemer\'edi's theorem is often called the constellation theorem because,
roughly speaking, it asserts that a fixed finite constellation (up to translation and scaling)
can be found in any set of points with positive density. A polynomial version of the theorem is also known:

\parenproclaim Theorem P (Bergelson, Leibman, {\rm 1996}). The following are equivalent.
\medskip
\item{i)} Let $P_1, \ldots, P_k : \ZZ \to \ZZ$ be polynomials such that $P_1(0) = \cdots = P_k(0) = 0$.
Let $A\subseteq \ZZ^d$ have positive upper density. Then there exist infinitely many pairs
$(a,r)\in \ZZ^d\times\NN_0$ such that $a+P_1(r), \ldots, a+P_k(r)\in A$.
\smallskip
\item{ii)} Let $(X,\F,\mu)$ be a probability space, let $k\in \NN$, and let $T_1,T_2,\ldots,T_k:X\to X$ be
commuting measure-preserving maps. Let
$P_1, \ldots, P_k : \ZZ \to \ZZ$ be polynomials such that $P_1(0) = \cdots = P_k(0) = 0$. There exists $r>0$
such that
$$T^{-P_1(r)}E + T^{-P_2(r)}E\cap \cdots\cap T^{-P_k(r)}E$$
is nonempty, where $T^{(a_1,\ldots,a_k)}$ is defined to be ${T_1}^{a_1}\cdots {T_k}^{a_k}$.\slug
\medskip

\noindent Lastly, we have the density Hales-Jewett theorem, which tells us that subsets of large enough density must
contain a combinatorial line.

\parenproclaim Theorem D (Furstenberg, Katznelson, {\rm 1991}).
Let $n\geq 1$ and $0<\delta\leq 1$. There exists an integer $d\geq 1$ such that if $A$ is any
subset of $[0,n-1]^d$ of cardinality $|A|\geq \delta n^d$, then $A$ contains a proper arithmetic progression
$\{a, a+v, \ldots, a+(k-1)v\}$ for some $a\in [0,n-1]^d$ and $v\in [0,1]^d$.\slug

\noindent Theorem D also has a an ergodic counterpart, but it is rather
complicated. For many years, the ergodic proof of the density Hales-Jewett theorem was the only one known.
In 2009, in an online initiative called the Polymath Project spearheaded by W. T. Gowers,
a group of over 40 mathematicians found a purely combinatorial proof of Theorem D. The results were published
under the pseudonym D. H. J. Polymath and the success of this experiment
led to over a dozen other online Polymath Projects.

\section References

\bye

### 2020-12-24 about/vita.tex
Arithmetic subsequences

@@
 \leftright{\sl BSc, Honours Mathematics and Computer Science, minor Linguistics}{\sl Montr\'eal, QC, Canada}
-CGPA: 3.79/4.
+CGPA: 3.81/4.
 \medbreak
@@
 Submitted July 2020.
+\subitem (with Rosie Y. Zhao)
+Arithmetic subsequences in a random ordering of an additive set. {\sl arXiv preprint 2012.12339}.
+Submitted December 2020.
 

### 2020-12-24 misc/expo/adventures.tex
Arithmetic subsequences

\input fontmac
\input mathmac
\classicmode

\def\A{{\cal A}}
\def\C{{\cal C}}
\def\F{{\cal F}}

\def\edge{\mathrel-\!\!\!\mathrel-}

\maketitle{Adventures in additive combinatorics}{by}{Marcel K. Goh}{18 December 2020}

\advsect Introduction

This set of notes will contain neat theorems and problems that are (at least a little bit)
related to additive combinatorics. The individual sections are intended to be largely self-contained, and
will not delve too much into theory, but focus on famous problems and their solutions, especially if
the proofs are rather elegant (particular preference has been given to probabilistic proofs). None of these
proofs are new in any way; they all appear in one of the references listed at the end of the
document.

\medskip
\boldlabel Notation and terminology. The notation $[a,b]$ always denotes the {\it discrete interval}
$\{n\in \ZZ : a\leq n\leq b\}$. For a set $X$, we let $2^X$ denote the set of all subsets of $X$ (the {\it
power set}) and the notation $X^{(r)}$ indicates the collection of all subsets
of $X$ with cardinality $r$.

A {\it graph} $G$ is a pair $(V,E)$ where $E\subseteq V^{(2)}$. If, for $u,v\in
V$, we have $\{u,v\}\in E$, then we write $u \edge v$ and say that $u$ and $v$ are {\it adjacent}. The
{\it neighbourhood} of a node $u\in V$ is the set $N(u)=\{v\in V: u\edge v\}$, that is, the set of
nodes adjacent to $u$, and the {\it degree} of $u$, denoted $d(u)$, is the cardinality of this set.

\advsect The Littlewood-Offord problem

We begin with a simple problem, which J. E. Littlewood and A. C. Offord came across in 1943 while studying the
zeroes of random polynomials. Let $z_1, z_2, \ldots, z_n$ be complex numbers of modulus at least $1$ and form
all $2^n$ possible sums. How many of these sums can differ by less than $1$? To attack this problem, we will
need a couple of definitions and inequalities. Given a ground set $X$,
an {\it antichain} or {\it Sperner system} is a collection
of subsets $\F$ such that for any $A\subseteq B\in 2^X$, either $A\notin \F$ or $B\notin F$. In other words,
if all the elements of $2^X$ are ordered by inclusion, then no two elements of $\F$ are comparable. On the
other hand, a {\it chain} is a collection $\C = \{C_1, C_2, \ldots, C_k\}$ of subsets of $X$ such that
$C_1\subset C_2 \subset \cdots \subset C_k$.

The following inequality was proved independently by D.~Lubell, K.~Yamamoto, and L.~D.~Meshalkin,
and has thus become known as the LYM inequality. However, it also follows from an earlier result of B.~Bollob\'as,
and in fact, the probabilistic proof we present is due to Bollob\'as.

\parenproclaim Lemma B (LYM inequality). Let $X$ be a finite set of size $n$;
for simplicity, we assume that $X = [1,n]$. Let $\F\subseteq 2^X$ be an antichain; then
$$\sum_{A\in \F} {n\choose |A|}^{-1} \leq 1.$$

\proof Let $\sigma : X\to X$ be a permutation chosen uniformly from all $n!$ such permutations. For
$A\subseteq X$, $\sigma(A)$ denote the set $\{\sigma(x) : x\in A\}$ and consider the probability that
$\sigma(A) = [1,|A|]$. Note that $\sigma(A)$ is uniformly distributed over the ${n\choose |A|}$ members of
$X^{(|A|)}$, so the probability that it equals $[1,|A|]$ is ${n\choose |A|}^{-1}$. Note that for two sets
$A,B\in 2^X$, if $\sigma(A) = [1,|A|]$ and $\sigma(B) = [1,|B|]$ for the same fixed $\sigma$,
then either $A\subseteq B$ or $B\subseteq
A$. Since $\F$ is an antichain, the events $\sigma(A) = [1,|A|]$ must then be disjoint and the sum of their
probabilities is at most $1$. This proves the inequality.\slug

Because ${n \choose k}$ is maximal when $k = \lfloor n/2\rfloor$, we can derive the following theorem of
E. Sperner as a corollary.

\parenproclaim Corollary S (Sperner, {\rm 1928}).
Let $X$ be a finite set and let $\F\subseteq 2^X$ be an antichain. Then $\F$ has size at most ${n\choose \lfloor n/2\rfloor}$.

\proof We have, by Lemma B,
$${|A|\over {n\choose \lfloor n/2\rfloor}} \leq \sum_{A\in \F} {n\choose |A|}^{-1} \leq 1.$$
Multiplying on both sides by ${n\choose \lfloor n/2\rfloor}$ yields the result.\slug

With this theorem in hand, we can already solve the Littlewood-Offord problem in the special case
when all the $z_i$ are real. The
following construction is due to P. Erd\H os (1945). Let $x_1, x_2, \ldots, x_n$ be real numbers with absolute
value at least 1. For $A\subseteq [1,n]$, let $x_A = \sum_{i\in A}x_i$. We may assume all of the $x_i$ are
positive, because replacing $x_i$ with $-x_i$ and $A$ with $A \cup \{i\} \setminus \big(A\cap \{i\}\big)$ does not
change the relative differences between the $2^n$ sums.
This operation simply causes all the $x_A$ to permute amongst themselves and shift by $-x_i$. Now let $\F$
be a family of subsets of $[1,n]$ such that $|x_A-x_B|<1$ for every distinct pair of $A,B\in \F$. It is an
antichain because if $A$ is a proper subset of $B$ then $x_B>x_A$ and we have
$$|x_B - x_A| = \sum_{x\in B}x_i - \sum_{x\in A}x_i = \sum_{x\in B\setminus A}x_i = x_{B\setminus A},$$
and this is at least 1 since $B\setminus A$ is nonempty.
Hence at most one of $A$ and $B$ belongs to $\F$. By Corollary~S, $|\F|\leq {n\choose n/2}$.

\advsect Independent sets

For a graph $G=(V,E)$, an {\it independent set} is a subset $S\subseteq V$ of vertices such that no two
elements of $S$ are adjacent. The largest independent set in $G$ is denoted $\alpha(G)$.
In this section, we present two beautiful probabilistic proofs that give lower bounds for $\alpha(G)$.
The first is due to Tur\'an; sometimes this theorem is called Tur\'an's theorem, but more often, that name is
reserved for a somewhat dual statement about the maximum number of edges in a graph that has no
complete graph on $r+1$ vertices as a subgraph.

\proclaim Theorem T. Let $G=(V,E)$ be a graph. The size $\alpha(G)$ of the largest independent set in $G$
satisfies the inequality
$$\alpha(G) \geq \sum_{u\in V} {1\over d(u) + 1}.$$

\proof Suppose $V$ has size $n$ and let $\sigma:V\to [1,n]$ be a bijection, chosen uniformly at random from
all $n!$ such bijections. Regarding this as an ordering of the vertices, consider the set
$$S = \{ u \in V : \sigma(u) < \sigma(v)\ \hbox{for all}\ v\in N(u)\},$$
which consist of vertices that come before all of their neighbours in the ordering. For any ordering,
the set $S$ that it induces must be an independent set, for if two vertices $u,v\in V$ are adjacent
then the vertex which was assigned a larger value in the ordering cannot be in $S$. Fixing a node $u\in V$,
the probability that it is the first among all of its neighbours is $1/\big(d(u) + 1\big)$, and by linearity
of expectation, we have
$$\ex\{|S|\} = \ex\Big\{\sum_{u\in V} \indic{u\in S}\Big\} = \sum_{u\in V}{1\over d(u) + 1}.$$
So there exists an independent set $S$ with size at least the value of this sum, which was what we
wanted.\slug

In particular, if maximal degree over all vertices in $G$ is $d$, then there is an independent set of size
at least $n/(d+1)$.

\section References

\bye

### 2020-12-24 misc/expo/regularity.tex
Arithmetic subsequences

\input fontmac
\input mathmac

\def\P{{\cal P}}

\classicmode
\maketitle{Szemer\'edi's regularity lemma}{notes by}{Marcel K. Goh}{22 December 2020}

\advsect Introduction

[Put some kind of introduction here.]

\advsect The regularity lemma

In this section we will prove Szemer\'edi's regularity lemma via a sequence of auxiliary ones. We will
first require some definitions. Fix a graph $G=(V,E)$ and let $X$ and $Y$ be subsets of $V$ (not necessarily
disjoint). Let $e(X,Y) = \{xy \in E : x\in X,\,y\in Y\}$ denote the number of edges that have an endpoint
in each of $X$ and $Y$. We define the {\it edge density} between $X$ and $Y$ to be the ratio
$$d(X,Y) = {e(X,Y)\over |X||Y|}.$$
If $X$ and $Y$ are disjoint, then this is the fraction of all possible edges between $X$ and $Y$ that are
actually present in the graph (and in the case that they are not disjoint, it isn't awfully far off anyway).

We say that a pair of vertex subsets $(X,Y)$ is {\it $\eps$-regular} if for all subsets $A\subseteq X$ and
$B\subseteq Y$ with $|A|\geq \eps|X|$ and $|B|\geq \eps|Y|$, we have $|d(A,B)- d(X,Y)|\leq \eps$. This means that
if we zoom in to look at the edges between a subset of $X$ and a subset of $Y$, we find that the picture is sort of
a ``scale-model'' of the whole of $X$ and the whole of $Y$ in the sense that the number of edges that
we see is proportional to the sizes of the subsets, unless the subsets are taken to be very small. If the pair
$(X,Y)$ is {\it not} $\eps$-regular, then there must be some $A\subseteq X$ and $B\subseteq Y$, with
$|A|\geq \eps|X|$ and $|B| \geq \eps|Y|$, such that $|d(A,B)-d(X,Y)|>\eps$. The pair $(A,B)$ is said to
{\it witness} the irregularity.

A {\it partition} $\P$ is a collection $V_1,\ldots,V_k$ of disjoint subsets of $V$ whose union is all of $V$.
We will say that a partition is {\it equitable} if the sizes of any two parts do not differ by more than $1$.
A partition is said to be {\it $\eps$-regular} if the sum of $|V_i||V_j|$, taken over all pairs $(V_i,V_j)$
that are not $\eps$-regular, is less than $\eps |V|^2$. If the partition is equitable, then this is equivalent
to saying that at most $\eps k^2$ of the pairs $(V_i,V_j)$ are not $\eps$-regular. Szemer\'edi's regularity
lemma says that every graph admits an $\eps$-regular equitable partition into a number of parts depending
only on $\eps$ (and not the size of the graph).

\medskip
\boldlabel Energy increment. :q

\bye

### 2021-01-13 misc/expo/regularity.tex
Winter 2021 update

@@
 \def\P{{\cal P}}
+\def\Q{{\cal Q}}
+\def\R{{\cal R}}
 
 \classicmode
-\maketitle{Szemer\'edi's regularity lemma}{notes by}{Marcel K. Goh}{22 December 2020}
+\maketitle{Szemer\'edi's regularity lemma}{notes by}{Marcel K. Goh}{12 January 2020}
 
@@
 
-\advsect The regularity lemma
-
-In this section we will prove Szemer\'edi's regularity lemma via a sequence of auxiliary ones. We will
-first require some definitions. Fix a graph $G=(V,E)$ and let $X$ and $Y$ be subsets of $V$ (not necessarily
+Fix a graph $G=(V,E)$ and let $X$ and $Y$ be subsets of $V$ (not necessarily
 disjoint). Let $e(X,Y) = \{xy \in E : x\in X,\,y\in Y\}$ denote the number of edges that have an endpoint
 in each of $X$ and $Y$. We define the {\it edge density} between $X$ and $Y$ to be the ratio
-$$d(X,Y) = {e(X,Y)\over |X||Y|}.$$
+$$d(X,Y) = {e(X,Y)\over |X||Y|}.\adveq$$
 If $X$ and $Y$ are disjoint, then this is the fraction of all possible edges between $X$ and $Y$ that are
@@
 
-A {\it partition} $\P$ is a collection $V_1,\ldots,V_k$ of disjoint subsets of $V$ whose union is all of $V$.
+A {\it partition} $\P$ is a collection $\{V_1,\ldots,V_k\}$ of disjoint subsets of $V$ whose union is all of $V$.
 We will say that a partition is {\it equitable} if the sizes of any two parts do not differ by more than $1$.
@@
 
+References for the original instances of the proofs are listed at the bottom of this document; a large part of
+these notes are based on lectures given by Yufei Zhao in 2019.
+
+\advsect The regularity lemma
+
+In this section we will prove Szemer\'edi's regularity lemma via a sequence of auxiliary ones. The idea runs
+as follows. We begin with a partition of $G=(V,E)$ that is given to us
+(e.g., the trivial partition $\P = \{V\}$) and while the partition
+is not $\eps$-regular, we iteratively refine
+it by subdividing each element of the partition into further parts.
+Using an ``energy increment argument'', we show that this
+process terminates after a bounded number of steps, and therefore the number of parts in the final partition
+is bounded. For vertex sets $U$ and $W$, we define the {\it energy} of $(U,W)$
+to be the quantity
+$$q(U,W) = {|U||W|\over n^2}d(U,W)^2,\adveq$$
+where $n=|V|$, and for partitions $\P_U = \{U_1,U_2, \ldots,U_k\}$ and $\P_W\{W_1,W_2,\ldots, W_l\}$
+of $U$ and $W$ respectively, we will define the {\it energy} of the two partitions to be the sum
+$$q(\P_U, \P_W) = \sum_{i=1}^k\sum_{j=1}^l q(U_i, W_j).\adveq$$
+We will write $q(\P) = q(\P,\P)$ when the two partitions are equal. Note that if $\P$ is a partition of
+$V$ into $k$ parts, we have
+$$q(\P) = \sum_{i=k}^k\sum_{j=1}^k {|V_i||V_j|\over n^2} d(V_i,V_j)^2 \leq 1,\adveq$$
+since edge density is at most $1$.
+The first lemma states that the energy of a pair of refined partitions is at least the energy of the original
+pair.
+
+\proclaim Lemma A. Let $G=(V,E)$ be a graph, let $U,W\subseteq V$ and suppose that
+$\P_U = \{U_1,U_2, \ldots,U_k\}$ and $\P_W\{W_1,W_2,\ldots, W_l\}$ are partitions of $U$ and $W$ respectively.
+Then $q(\P_U,\P_W)\geq q(U,W)$.
+
+\proof We will define a random variable $Z$ as follows. We select vertices $u\in U$ and $w\in W$ uniformly
+at random; suppose that $U_i\in \P_U$ contains $u$ and $W_j\in \P_W$ contains $w$. We let $Z = d(U_i,W_j)$.
+We compute the first moment
+$$\ex\{Z\} = \sum_{i=1}^k{|U_i|\over |U|}\sum_{j=1}^l {|W_j|\over|W|} d(U_i,W_j) = {e(U,W)\over |U||W|} = d(U,W)
+\adveq$$
+and the second moment
+$$\ex\{Z^2\}=\sum_{i=1}^k{|U_i|\over |U|}\sum_{j=1}^l {|W_j|\over|W|} d(U_i,W_j)^2={n^2\over |U||W|}q(\P_U,\P_W)
+\adveq$$
+of $Z$, where $n$ denotes the size of $V$. By Jensen's inequality, we have $\ex\{Z^2\}\geq\ex\{Z\}^2$ and therefore
+$$q(\P_U,\P_W)\geq {|U||W|\over n^2}d(U,W) = q(U,W).\noskipslug\adveq$$
+
+In particular,
+if $\P$ is a partition of $V$ and $\P'$ refines $\P$, then we can apply Lemma A to every pair $(V_i,V_j)$ of
+sets in $\P$ to conclude that $q(\P')\geq q(\P)$. The next lemma shows that the inequality in Lemma A is sometimes
+strict, a fact we will need for the energy increment argument.
+
+\proclaim Lemma B. With the same definitions as in Lemma A, suppose furthermore that for some $\eps>0$,
+the pair $(U,W)$ is not $\eps$-regular and the irregularity is witnessed by $U_1\subseteq U$ and $W_1\subseteq W$.
+Then
+$$q(\{U_1,U\setminus U_1\}, \{W_1,W\setminus W_1\}) \geq q(U,W)+\eps^4{|U||W|\over n^2},\adveq$$
+where $n = |V|$.
+
+\proof Define the random variable $Z$ as in the proof of Lemma $A$. Note that the variance of $Z$ is
+$$\eqalign{
+\var\{Z\}&=\ex\{Z^2\}-\ex\{Z\}^2 \cr
+&={n^2\over |U||W|}q(\{U_1,U\setminus U_1\},\{W_1,W\setminus W_1\}) - d(U,W)^2\cr
+&= {n^2\over |U||W|}\big(q(\{U_1,U\setminus U_1\}, \{W_1,W\setminus W_1\}) - q(U,W)\big).\cr
+}\adveq$$
+But we also have the formula
+$$\eqalign{
+\var\{Z\}&=\ex\big\{(Z-\ex\{Z\})^2\big\}\cr
+&= {|U_1||W_1\over |U||W|}\big(d(U_1,W_1)-d(U,W)\big)^2
+ + {|U_1||W\setminus W_1|\over |U||W|} \big(d(U_1,W\setminus W_1)-d(U,W)\big)^2\cr
+&\quad+ {|U\setminus U_1||W_1|\over |U||W|} \big(d(U\setminus U_1,W_1)-d(U,W)\big)^2
+ + {|U\setminus U_1||W\setminus W_1|\over|U||W|} \big(d(U\setminus U_1,W\setminus W_1)-d(U,W)\big)^2\cr
+&\geq {|U_1|\over |U|}\cdot{|W_1|\over |W|}\cdot \big(d(U_1,W_1)-d(U,W)\big)^2\cr
+&\geq \eps\cdot\eps\cdot\eps^2,\cr
+}\adveq$$
+where the final inequality follows from the fact that $(U_1,W_1)$ was the witness for the
+non-$\eps$-regularity of $(U,W)$. Combining both calculations for the variance proves the inequality we need.\slug
+
+We are now able to formulate the step in the inner loop of our regularisation procedure.
+
+\proclaim Lemma C. Let $G=(V,E)$ be a graph, let $\P = \{V_1,V_2\ldots,V_k\}$ be a partition of $V$, and let
+$\eps>0$. If the partition $\P$ is not $\eps$-regular, then there exists a refinement $\Q$ of $\P$ in which every
+$V_i$ is partitioned into at most $2^k$ parts and such that
+$$q(\Q) \geq q(\P) + \eps^5.\adveq$$
+
+\proof If $\P_1$ and $\P_2$ are refinements of $\P$ that subdivide $V_i$ into $V_{i1}\cup V_{i1}'$ and
+$V_{i2}\cup V_{i2}'$ respectively, then the {\it common refinement} of $\P_1$ and $\P_2$ divides $V_i$
+into the union
+$$V_i = (V_{i1}\cap V_{i2}) \cup (V_{i1}'\cap V_{i2}) \cup (V_{i1}\cap V_{i2}') \cup (V_{i1}'\cap V_{i2}');\adveq$$
+by induction, we can similarly define the common refinement of any finite number of partitions that refine $\P$.
+For every pair $(i,j)$ for which $(V_i,V_j)$ is not $\eps$-regular, we can find $A_{ij}\subseteq V_i$ and
+$A_{ji}\subseteq V_j$ that witnesses the irregularity. Lemma B will produce a refinement of $\P$ for each $(i,j)$
+that divides $V_i$ and $V_j$ each into two new parts,
+and we can let $\Q$ be the common refinement of these partitions, as defined above. Note that we have constructed
+$\Q$ such that it does not have more than $2^k$ parts.
+
+Let $\R$ be the set of all $(i,j)\in [1,k]^2$ such that $(V_i,V_j)$ is $\eps$-regular. For each $i$,
+let $\Q_{V_i}$ denote the subdivision of $V_i$ given by $\Q$. By Lemma A, we have
+$$\eqalign{
+q(\Q) &= \sum_{i=1}^k\sum_{j=1}^k q(\Q_{V_i},\Q_{V_j}) \cr
+&\geq \sum_{(i,j)\in \R} q(V_i,V_j) + \sum_{(i,j)\notin \R} q(\{A_{ij},V_i\setminus A_{ij}\},\{A_{ji},V_j\setminus
+A_{ji}\}).\cr
+}\adveq$$
+Applying Lemma B, we find that
+$$\eqalign{
+q(\Q) &\geq \sum_{(i,j)\in \R} q(V_i,V_j) + \sum_{(i,j)\notin \R} q(V_i,V_j) + \eps^4{|V_i||V_j|\over n^2}\cr
+&\geq q(\P) + \eps^5,\cr
+}\adveq$$
+as desired.\slug
+
+With Lemma C in hand, we can state and prove Szemer\'edi's regularity lemma without too much further effort.
+
+\parenproclaim Theorem R (Szemer\'edi, {\rm 1978}).
+For all $\eps>0$ there exists an $M$ with such that every graph admits an
+$\eps$-regular partition of its vertices into no more than $M$ parts. The constant $M$ depends only on
+$\eps$ (not on the size of the graph) and we have the upper bound
+\font\sevenit=cmmi7
+$$M\leq 4^{4^{\lower3pt\hbox{\sevenit $\sevenit\cdot$}
+\lower1.75pt\hbox{\sevenit$\cdot$}
+\lower0.5pt\hbox{\sevenit$\cdot$}^4}}\adveq$$
+where the tower of $4$s consists of $\eps^{-5}$ repeated exponents.
+
+\proof We start with the trivial partition $\P = \{V\}$ and apply Lemma C while the current partition is not
+$\eps$-regular. Since $0\leq q(\P)\leq 1$, and the energy of the partition increases by at least $\eps^5$ with
+each iteration, the algorithm terminates after at most $\eps^{-5}$ steps. At any particular step, if $\P$ has
+$k$ parts, Lemma C outputs a partition with $k2^k\leq 4^k$ parts, and this observation yields the upper bound
+in the theorem statement.\slug
+
+The bound we stated above is not tight, but it turns out that the tower of exponents is inescapable. It was shown
+by Gowers (1997) that there exists a $c>0$ such that for all $\eps>0$ small enough, one can construct a graph
+whose $\eps$-regular partition requires more than $M$ parts, where $M$ is a exponential tower of $2$s that is
+$\eps^{-c}$ high.
+
 \medskip
-\boldlabel Energy increment. :q
+\boldlabel Equitable partitions. In Lemma C, one can require
+that the resulting partition $\Q$ be equitable (no two parts differ by more than one element), and the inequality
+would still hold, although with an increment that is possibly less than $\eps^5$. The difference is negligible,
+in the sense that the final bound obtained
+for $M$ will still be of the same order. It is important to note that it is
+{\it not} possible to obtain an $\eps$-regular partition by proving the regularity
+lemma with Lemma C as stated above, and then further subdividing and merging the partitions afterwards, because
+\section References
 

### 2021-01-13 about/vita.tex
Updated vita

@@
 \font\elevensc=cmcsc11  % For section headers
-\font\bigbold=cmb12  % Name
+\font\bigbold=cmb10 at 12pt  % Name
 \font\tenb=cmb10
@@
 \def\leftright#1#2{{#1}\hfill{#2}\par}
-\def\myname{{\tensc Marcel K. Goh}}
 \def\CEE{{\mc C}}
@@
 
-\leftright{\bigbold Marcel K.\ Goh}{3660 Rue Hutchison, Apt.\ 4}
+\leftright{\bigbold Marcel K.\ Goh
+\ifx\Umathchar\undefined\else% Chinese characters do not appear when using pdfTeX
+\font\chinese="STKaiti" at 12pt%
+\font\twelverm=cmr12%
+\twelverm\up({\chinese 吳麒仁}\up)%
+\fi}{3660 Rue Hutchison, Apt.\ 4}
 \leftright{Email: marcel\_goh@yahoo.ca}{Montr\'eal, QC H2X 2H3, Canada}
@@
 \leftright{\sl BSc, Honours Mathematics and Computer Science, minor Linguistics}{\sl Montr\'eal, QC, Canada}
-CGPA: 3.81/4.
+CGPA: 3.82/4.
 \medbreak
@@
 \begingroup\parindent=10pt
+\smallskip
+\thing Winter 2021: MATH 457 Honours Algebra 4
 \smallskip
@@
 \thing Recipient of 0x\$3.40 in Knuth reward cheques.
+\smallskip
+\thing Contributed sequences A335562, A338550, and A338993 to the On-line Encyclopedia of Integer Sequences.
 \medbreak

### 2021-01-23 misc/expo/regularity.tex
Added more reports

@@
 \def\R{{\cal R}}
+\def\given{\,|\;}
+\def\zed#1{{\bf Z}/#1{\bf Z}}
 
@@
 
-\advsect Introduction
+\advsect Definitions and notation.
 
-[Put some kind of introduction here.]
-
+In these notes, we consider undirected graphs $G = (V,E)$, where $V$ is an arbitrary finite set and $E$ consists
+of $2$-element subsets of $V$. We sometimes write $V(G)$ for $V$ and $E(G)$ for $E$, and for convenience, we will
+write $uv$ instead of $\{u,v\}$ when the vertices $u$ and $v$ are adjacent.
 Fix a graph $G=(V,E)$ and let $X$ and $Y$ be subsets of $V$ (not necessarily
@@
 
-References for the original instances of the proofs are listed at the bottom of this document; a large part of
-these notes are based on lectures given by Yufei Zhao in 2019.
+A function $f(n)$ is said to be $O\big(g(n)\big)$ if there exists a constant $C$ such that
+$f(n)\leq Cg(n)$ for all integers $n\geq 1$. A function $f(n)$ is $o\big(g(n)\big)$ if $lim_{n\to\infty}
+f(n)/g(n) = 0$.
+The notation $[a,b]$ will always denote the {\it discrete interval} $\{n\in \ZZ : a\leq n\leq b\}$.
+
+References for the original instances of the proofs are listed at the bottom of this document; the particular
+presentation of these notes is heavily informed by lectures given by Yufei Zhao in 2019.
 
@@
 lemma with Lemma C as stated above, and then further subdividing and merging the partitions afterwards, because
+
+\advsect Graph counting and removal
+
+A {\it graph homomorphism} $f: H\to G$ is a function from $V(H)$ to $V(G)$ such that for any edge $uv\in E(H)$,
+$f(u)f(v)$ is an edge in $E(G)$.
+We begin by counting the number of homomorphic copies of a fixed graph $H$ in a larger graph
+$G$, provided some regularity conditions on the vertex set of $G$ hold.
+
+\parenproclaim Lemma C (Graph counting lemma). Let $H$ be a graph with vertex set $[1,k]$.
+Let $\eps>0$ and let $G$ be another
+graph with vertex subsets $V_1,V_2,\ldots,V_k\subseteq V(G)$ such that $(V_i,V_j)$ are $\eps$-regular for all
+$ij\in E(H)$. If $m = \big|E(H)\big|$ and $n$ is the number of $k$-tuples
+$(v_1,v_2,\ldots,v_k)\in V_1\times V_2\times\cdots\times V_k$ such that
+$v_iv_j\in E(G)$ whenever $ij\in E(H)$, then we have
+$$\Big|n - \prod_{ij\in E(H)} d(V_i,V_j) \prod_{i=1}^k |V_i|\Big|\leq m\eps \prod_{i=1}^k |V_i|.\adveq$$
+
+\proof To avoid trivialities, we assume $k\geq 3$.
+If we sample $\xi_1$ uniformly from $V_1$, $\xi_2$ from $V_2$, and so on, and
+let $B$ be the event that $\xi_i\xi_j\in E(G)$ for all $ij\in E(H)$,
+then the above inequality is equivalent to the statement
+$$\Big|\pr\{B\} - \prod_{ij\in E(H)} d(V_i,V_j)\Big| \leq m\eps.\adveq$$
+We will prove this statement by induction on $m$. If $m = 0$, both sides equal $0$, since the empty product equals 1.
+Now assume that $m\geq 1$ and by relabelling the vertices of $H$, we can assume that $12\in E(H)$. Let $B'$ be the
+event that $\xi_i\xi_j\in E(G)$ for all $ij\in E(H)\setminus\{12\}$. By the induction hypothesis, we have
+$$\bigg|\pr\{B'\} - {1\over d(V_1,V_2)}\prod_{ij\in E(H)} d(V_i,V_j)\bigg| \leq (m-1)\eps,\adveq$$
+and since edge density is bounded above by $1$, we can rewrite this as
+\newcount\ih
+\ih=\eqcount
+$$\Big|d(V_1,V_2)\pr\{B'\} - \prod_{ij\in E(H)} d(V_i,V_j)\Big| \leq (m-1)\eps.\adveq$$
+
+Now let $(v_3,v_4,\ldots,v_k)$ in $V_3\times V_4\times\cdots\times V_k$ be arbitrary and let $A_1$ be the set
+of all $v_1\in V_1$ such that $v_1v_i\in E(G)$ for all $1i\in E(H)$, $i\neq 2$. Likewise, let $A_2$ be the set of all
+$v_2\in V_1$ such that $v_2v_i \in E(G)$ for all $2i\in E(H)$, $i\neq 1$. If $|A|\geq\eps|V_1|$ and $|A_2|\geq\eps
+|V_2|$, then since $(V_1,V_2)$ is $\eps$-regular, the inequality $\big|d(A_1,A_2)-d(V_1,V_2)\big|\leq \eps$ implies
+that
+\newcount\diff
+\diff=\eqcount
+$$\bigg| {e(A_1,A_2)\over |V_1|\cdot|V_2|} - d(V_1,V_2){|A_1|\cdot|A_2|\over |V_1|\cdot|V_2|}\bigg|\leq
+\eps{|A_1|\cdot|A_2|\over |V_1|\cdot|V_2|} \leq \eps;\adveq$$
+this inequality also holds if $|A_1|<\eps|V_1$ or $|A_2|<\eps|V_2|$, since in this case both terms are less than
+$\eps$.
+The first term is the probability that the choices for $\xi_1$ and $\xi_2$ have the correct neighbours in the set
+$\{v_3,v_4,\ldots,v_k\}$ {\it and}
+are connected, while the second term does not require that $\xi_1$ and $\xi_2$ be connected. In other words,
+$${e(A_1,A_2)\over |V_1|\cdot|V_2|} = \pr\{B\given \xi_3 = v_3,\,\xi_4 = v_4,\,\ldots,\,\xi_k=v_k\}\quad\hbox{and}\quad
+{|A_1|\cdot|A_2|\over |V_1|\cdot|V_2|} = \{B'\given\xi_3 = v_3,\,\xi_4 = v_4,\,\ldots,\,\xi_k=v_k\}.\adveq$$
+Substituting this back into~\refeq{\the\diff}, we have
+$$\big|\pr\{B\given \xi_3 = v_3,\,\xi_4 = v_4,\,\ldots,\,\xi_k=v_k\} - d(V_1,V_2)\pr\{B'\given
+\xi_3 = v_3,\,\xi_4 = v_4,\,\ldots,\,\xi_k=v_k\}\big|\leq \eps.\adveq$$
+Since this holds for {\it any} choice of $(v_3,v_4,\ldots,v_k)$, we have {\it a fortiori}
+$$\big|\pr\{B\} - d(V_1,V_2)\pr\{B'\}\big|\leq \eps,\adveq$$
+which results by letting $\xi_3,\xi_4,\ldots,\xi_k$ once again be chosen uniformly from their respective sets.
+Adding the induction hypothesis~\refeq{\the\ih} and applying the triangle inequality, we have
+$$\Big|\pr\{B\} - \prod_{ij\in E(H)} d(V_i,V_j)\Big| \le
+\big|\pr\{B\} - d(V_1,V_2)\pr\{B'\}\big| +
+\Big|d(V_1,V_2)\pr\{B'\} - \prod_{ij\in E(H)} d(V_i,V_j)\Big| \leq
+m\eps,\adveq$$
+which was what we had to show.\slug
+
+We say that a graph $G$ is {\it $H$-free} if it does not admit a graph homomorphism $f: H\to G$.
+We can now formulate the graph removal lemma, which states that if a graph $G$ does not contain too many homomorphic
+copies of a graph $H$, then one does not need to remove that many edges to make it $H$-free. The following
+theorem was first proven in 1978 by I.~Ruzsa and E.~Szemer\'edi for the case $H = K_3$, and extended to arbitrary
+$H$ in 1986 By P.~Erd\H os, P.~Frankl, and V.~R\"odl.
+
+\parenproclaim Lemma R (Graph removal lemma). Fix a graph $H$ with $k\geq 3$ vertices and $m$ edges. For any $\eps>0$
+there exists
+$\delta>0$ depending on $H$ and $\eps$
+such that every $n$-vertex graph $G$ with at most $\delta n^k$ homomorphic copies of $H$ can be made
+$H$-free by removing $\leq \eps n^2$ edges.
+
+\proof Let $G$ be a graph on $n$ vertices and let
+$\P = \{V_i\}$ be an $(\eps/4)$-regular partition
+of $V(G)$ given by the regularity lemma (thus it has no more than $M$ parts, where $M$ depends only on $\eps$).
+We will remove all edges between parts $V_i$ and $V_j$ if
+\medskip
+\item{i)} $(V_i,V_j)$ is not $(\eps/4)$-regular;
+\smallskip
+\item{ii)} $d(V_i,V_j)\leq \eps/2$; or
+\smallskip
+\item{iii)} either $V_i$ or $V_j$ has $\leq (\eps/4M)n$ vertices.
+\medskip
+The total number of edges deleted in case (i) is $\eps n^2/4$, by the definition of an $(\eps/4)$-regular partition.
+In case (ii), we remove no more than $\eps n^2/2$ edges and case (iii) happens no more than $M\cdot n$ times,
+causing no more than $\eps n^2/4$ edges to be deleted. Thus we have removed at most $\eps n^2$ edges in total.
+If there exists a copy of $H$ in this post-surgery graph, the vertices $1,2,\ldots,k$ of $H$ must lie in
+parts (not necessarily distinct) $V_1,V_2,\ldots,V_k\in \P$, each with greater than $(\eps/4M)n$ vertices,
+such that for all $(i,j)\in [1,k]^2$, $(V_i,V_j)$ is $(\eps/4)$-regular and $d(V_i,V_j) > \eps/2$.
+By the graph counting lemma, the number $C$ of homomorphic copies of $H$ in $G$ satisfies
+$$\Big|C - \prod_{ij\in E(H)} d(V_i,V_j) \prod_{i=1}^k |V_i|\Big|\leq m{\eps\over 4} \prod_{i=1}^k |V_i|.\adveq$$
+This means that there are at least
+$$ {1\over k!}
+\bigg({\eps\over 2}\bigg)^m \bigg({\eps\over 4M}\bigg)^k n^k - m{\eps\over 4}\bigg({\eps\over 4M}\bigg)^k n^k\adveq$$
+homomorphic copies of $H$ in $G$, where we have divided by $k!$ to handle the overcounting that occurs if some
+(or indeed all) of the $V_i$ are actually equal. So we may set
+$$\delta < \bigg({\eps\over 2}\bigg)^m \bigg(1- m{\eps\over 4}\bigg)\bigg({\eps\over 4M}\bigg)^k\adveq$$
+and by contraposition, if $G$ did not contain more than $\delta n^k$ copies of $H$ to begin with, then the procedure
+above removed all copies of $H$ in $G$.\slug
+
+The triangle counting lemma can be stated in a more succinct way that avoids $\eps$-$\delta$ terminology:
+{\sl If $H$ has $k$ vertices, then any graph with $o(n^k)$ copies of $H$
+can be made $H$-free by deleting $o(n^2)$ edges.}
+Of course, given a single graph on $n$ vertices, one cannot tell whether it has $o(n^k)$ copies of $H$, so
+more precisely, we're saying that given any $f(n) = o(n^k)$, there exists $g(n) = o(n^2)$ such that any graph
+on $n$ vertices with $\leq f(n)$ copies of $H$ can be made $H$-free by deleting at most $g(n)$ edges. Finally,
+although we have counted homomorphisms that are not necessarily injective, as the size of $G$ gets large relative
+to a fixed graph $H$, the proportion of homomorphisms that are non-injective quickly becomes negligible, so the number
+of subgraphs is roughly the number of homomorphisms.
+
+\advsect Roth's theorem
+
+By letting $H$ be the triangle graph $K_3$, we can prove a special case of Szemer\'edi's theorem, namely Roth's
+theorem on arithmetic progressions of length $3$, which was first proved in 1953.
+Since any homomorphism from $K_3$ to a graph $G$ must be injective, the graph counting
+lemma counts {\it bona fide} subgraphs of $K_3$ in larger graphs $G$. We begin with the following corollary of
+the removal lemma.
+
+\proclaim Lemma T. If $G$ is an $n$-vertex graph and every edge lies in exactly one triangle, then the number
+of edges of $G$ is $o(n^2)$.
+
+\proof Suppose there are $m$ edges in $G$. Then the number of triangles is $m'=m/3$, but since $m\leq {n\choose 2}$,
+this is $o(n^3)$. Applying the triangle removal lemma, we only need to remove $o(n^2)$ edges to make
+the graph triangle-free. But then again, we needed to remove at least $1/3$ of the edges to make $G$ triangle-free,
+so the total number of edges is $o(n^2)$ as well.\slug
+
+Roth's theorem states that for any $\delta>0$, there exists $N$ such that for any $n\geq N$, every subset of $[1,n]$
+with size at least $\delta n$ has an arithmetic progression of length $3$. We restate this slightly differently
+using the asymptotic notation described above and then prove it using the triangle removal lemma.
+
+\parenproclaim Theorem R (Roth's theorem).
+If $A\subseteq [1,n]$ does not have any $3$-term arithmetic progressions, then $|A| = o(n)$.
+
+\proof Let $A\subseteq [1,n]$ be free of $3$-term arithmetic progressions. Let $m=2n+1$ and note that if
+$A$ is taken to be a subset of the cyclic group $\zed m$, it also does not have any $3$-term arithmetic
+progressions in this group. Construct $G$ such that $V(G) = X \cup Y \cup Z$, where $X = Y = Z = \zed m$.
+We define the edge set of $G$ as follows:
+\medskip
+\item{i)} Connect $x\in X$ and $y\in Y$ whenever $y-x\in A$.
+\smallskip
+\item{ii)} Connect $x\in X$ and $z\in Z$ whenever $(z-x)/2\in A$. We are allowed to divide by $2$ because the
+multiplicative inverse of $2$ in $\zed m$ is $n+1$.
+\smallskip
+\item{iii)} Connect $y\in Y$ and $z\in Z$ whenever $z-y\in A$.
+\medskip
+Since we are doing addition in $\zed m$, each element of $A$ contributes $m$ edges to $E(G)$ in each case.
+If $x\in X$, $y\in Y$, and $z\in Z$ are the vertices of a triangle in $G$, then $y-x$, $(z-x)/2$, and $z-y$ all
+belong to $A$. But $A$ does not contain any proper arithmetic progressions of length $3$, so these elements
+must all be the same, that is,
+$$y-x = {z-x\over 2} = z-y.\adveq$$
+Given any two of $x,y,z$, we can determine the third from this equation, so every edge belongs to exactly one
+triangle. Thus the number of edges is $o(m^2)$. On the other hand, the number of edges is exactly $3m|A|$, so
+$3m|A| = o(m^2)$ and $|A| = o(m) = o(n)$.\slug
+
+Szemer\'edi's theorem has the same statement as Roth's theorem, with $3$ replaced by an arbitrary integer $k$.
+Note that we cannot extend this result to longer arithmetic progressions with the graph regularity lemma alone.
+However, it was first shown in a 2007 paper by W.~T.~Gowers,
+that one can extend the regularity lemma to hypergraphs, and by encoding
+the faces of a $(k-1)$-simplex as a $(k-1)$-uniform hypergraph as above, one can prove Szemer\'edi's theorem directly
+from the hypergraph removal lemma.
+
+\advsect Property testing
+
+Fix a graph $H$ on $k$ vertices.
+Suppose we have a graph $G$ with a large number $n$ of vertices and we know, for some $\eps>0$,
+that it is either $H$-free or ``$\eps$-far''
+from $H$-free, in that it cannot be made $H$-free by removing less than $\eps n^2$ edges. Consider the following
+algorithm to test whether $G$ is $H$-free. We sample $k$ random vertices $v_1,v_2,\ldots,v_k\in V(G)$
+(not necessarily distinct) and check if $v_iv_j\in E(G)$ whenever $ij\in E(H)$. If a copy of $H$ is found,
+we report that $G$ is $\eps$-far from $H$-free. If $C$ samples are performed and no copy of $H$ is found,
+we report that $G$ is $H$-free. Of course, there is a probability that this algorithm outputs an incorrect answer.
+The following theorem bounds this probability.
+
+\parenproclaim Theorem H (Find homomorphism). There exists a constant $C$ depending only on $\eps$ such that
+the algorithm above succeeds with probability at most $1-1/e$.
+
+\proof If $G$ is $H$-free, the algorithm always succeeds. So we are only concerned with the case where $G$ is
+$\eps$-far from $H$-free. The graph removal lemma says that in this case, there must exist at least $\delta n^k$
+copies of $H$ in $G$ for some $\delta$ depending on $\eps$. Setting $C = 1/\delta$ and letting $F$ be the
+event that the algorithm fails, we have
+$$\pr\{F\} \leq \bigg(1-{\delta n^k\over n^k}\bigg)^C \leq (1-\delta)^{1/\delta} \leq {1\over e}.$$
+Thus the algorithm succeeds with probability at most $1-1/e$.\slug
+
+So one can test whether graphs $G$ are $H$-free with arbitrary precision in $O(1)$ time, i.e.,
+not depending on the size of $G$.
+For example, if we want a probability of error $<1\%$, we only need to perform $5C$ tests no matter how large $G$
+is. Of course,
+since $\delta$ was given by the regularity lemma, the constant $C$ is astronomically large depending on $\eps$.
+
 \section References

### 2021-02-02 about/vita.tex
Added more reports
Tonogenesis

@@
 
+\goodbreak
+
 \sectheader{Reports}
@@
 \pubbegin R
+\repitem The OPythn programming language. Software project report, Charles University (Prague,
+Czech Republic, June 2019), 10 pp.
 \repitem Typechecking proof scripts: making interactive proof assistants robust.
 Honours project report, McGill University (Montr\'eal, Qu\'ebec, December 2019), 10 pp.
-
-\filbreak
+\repitem Grid-building algorithms on manifolds. Summer research report, McGill University (Montr\'eal, Qu\'ebec,
+August 2020), 10 pp.
 
@@
 \smallskip
-\thing Contributed sequences A335562, A338550, and A338993 to the On-line Encyclopedia of Integer Sequences.
+\thing Contributed sequences A335562, A338550, A338993, A339941, and A339942
+to the On-line Encyclopedia of Integer Sequences.
 \medbreak

### 2021-04-20 misc/expo/furstenberg.tex
Furstenberg video

@@
 \noindent
-The proof that Szemer\'edi gave was long and complicated, relying on an auxiliary ``graph-regularity lemma''
-that has had far-reaching consequences in combinatorics. We will only supply a proof of the fact
+The proof that Szemer\'edi gave was long and complicated, relying on an intricate combinatorial argument.
+We will only supply a proof of the fact
 that the infinitary version of the theorem is in fact equivalent to the finitary one. (This treatment
@@
 Since $\mu(A_N)\geq \mu(G)/2$ for every $N$, we have
-$$\mu\Big(\limsup_{N\to\infty} A_N\Big) = \mu\Big(\bigcap_{n=1}^\infty \bigcup_{n=N}^\infty A_n\Big)\geq
+$$\mu\Big(\limsup_{N\to\infty} A_N\Big) = \mu\Big(\bigcap_{N=1}^\infty \bigcup_{n=N}^\infty A_n\Big)\geq
 \mu(G)/2\adveq$$
@@
 $${1\over N} \sum_{s=1}^{N} \int_X f(T^{s}f)\cdots(T^{(k-1)s})\d\mu \geq {\delta^k\mu(G)\over 2KN_0}\adveq$$
-for all large enough $N$, yielding Theorem F\footnote{$^*$}{\eightpoint
-Thank you to Terry Tao, whose answer to my question on his blog
-helped me get the correct constants in the final bound.}.
+for all large enough $N$, yielding Theorem F.
 
@@
 = \mu_{s_1,\ldots,s_m}(E_1\times\ldots\times E_m).$$
-This is ``almost'' invariant under the shift $T$, since
+This is invariant under the shift $T$, since
 $$\eqalign{
@@
 &= \Lambda\Bigg(\!\!\bigg(
-{1\over N_j+1} \sum_{i=1}^{N_j+1} \chi_{E_1}(T^{s_1}a_i)\cdots\chi_{E_m}(T^{s_m}a_i)\bigg)_{j=1}^\infty\Bigg).\cr
+{1\over N_j+1} \sum_{i=1}^{N_j+1} \chi_{E_1}(T^{s_1}a_i)\cdots\chi_{E_m}(T^{s_m}a_i)\bigg)_{j=1}^\infty\Bigg),\cr
 }\adveq$$
-But as $j\to\infty$, shifting the terms over which we take the average does not change the $\limsup$ or the
-$\liminf$, so this is the same as
-$\mu\big(\{x \in X : x_{s_1}\in E_1, \ldots, x_{s_m}\in E_m\}\big)$, provided that the sequence
-$$\Bigg(\!\!\bigg(
-{1\over N_j+1} \sum_{i=0}^{N_j} \chi_{E_1}(T^{s_1}a_i)\cdots\chi_{E_m}(T^{s_m}a_i)\bigg)_{j=1}^\infty\Bigg)$$
-is convergent.
-This will be enough for our purposes (though there should be some way to fix it in general).
+and the $j$th term in the inner sequence differs from the original sequence by no more than $1/(N_j+1)$. So the
+differences between these two sequences goes to zero and their values under $\Lambda$ are the same.
+
 Let $B$ be the cylinder of all $x\in X$
@@
 }\adveq$$
-and since this is positive, there exists an arithmetic progression of length $k$ in $A$.\slug
+and since this is positive, there exists an arithmetic progression of length $k$ in $A$\footnote{$^*$}{\eightpoint
+Thank you to Terry Tao, whose answers to my questions on his blog helped me work out some of the details in this
+proof.}.\slug
 

### 2021-05-12 about/vita.tex
New sequence
Update

@@
 
+\font\ninesc=cmcsc9
 \font\eightrm=cmr8
-\font\elevensc=cmcsc11  % For section headers
+\font\elevensc=cmcsc10 at 11pt  % For section headers
 \font\bigbold=cmb10 at 12pt  % Name
@@
 
-\def\datethis{\eightrm Last updated \today\ at\ \hours.}
+\def\datethis{\eightrm \today}
 
@@
 
-\leftright{\bigbold Marcel K.\ Goh
+\leftright{\bigbold Marcel Kieren Goh
 \ifx\Umathchar\undefined\else% Chinese characters do not appear when using pdfTeX
@@
 \twelverm\up({\chinese 吳麒仁}\up)%
-\fi}{3660 Rue Hutchison, Apt.\ 4}
+\fi}{\datethis}
+\smallskip
+
+\sectheader{Personal and Contact Information}
+
+\leftright{Citizen of Canada and Singapore}{3660 Rue Hutchison, Apt.\ 4}
 \leftright{Email: marcel\_goh@yahoo.ca}{Montr\'eal, QC H2X 2H3, Canada}
-\leftright{URL: marcelgoh.github.io}{+1\thinspace\up(825\up)\thinspace 440-0681}
+\leftright{Webpage: marcelgoh.github.io}{+1\thinspace\up(825\up)\thinspace 440-0681}
 
@@
 
-\leftright{\bf McGill University}{\bf September 2017 -- April 2021}
-\leftright{\sl BSc, Honours Mathematics and Computer Science, minor Linguistics}{\sl Montr\'eal, QC, Canada}
-CGPA: 3.82/4.
+\leftright{\bf M.Sc.\ Mathematics}{\bf May 2021 -- present}
+\leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
+\smallskip
+Supervisor: Luc Devroye.
 \medbreak
-\leftright{\bf Charles University}{\bf February -- June 2019}
-\leftright{\sl Exchange semester, Faculty of Mathematics and Physics}{\sl Prague, Czech Republic}
+\leftright{\bf B.Sc.\ Joint Honours Mathematics and Computer Science}{\bf September 2017 -- April 2021}
+\leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
+\smallskip
+Minor: Linguistics. CGPA: 3.84/4.
+\medbreak
+\leftright{\bf Exchange semester}{\bf February -- June 2019}
+\leftright{\sl Faculty of Mathematics and Physics, Charles University}{\sl Prague, Czech Republic}
+\smallskip
 Grade: 1 on all courses (highest score attainable).
 
-\sectheader{Awards}
+\sectheader{Awards and Funding}
 
-\leftright{\bf NSERC Undergraduate Student Research Award}{\bf May 2020}
+\leftright{\bf NSERC Canada Graduate Scholarship -- Master's \rm (\$17,500)}{\bf May 2021 -- April 2022}
+\smallskip
+\leftright{\bf NSERC Undergraduate Student Research Award \rm (\$7,125)}{\bf May -- August 2020}
 \smallskip
@@
 
-\sectheader{Research Experience}
+\sectheader{Research Interests}
 
-\leftright{\bf Probabilistic analysis of branching processes}{\bf May 2020 -- present}
-\leftright{\sl Research group}{\sl McGill University}
-\smallskip
-Ongoing research on branching processes, headed by Luc Devroye. Studied estimation problems
-on Galton-Watson trees. Attended informal weekly seminars over the summer, discussing
-various topics related to branching
-processes. (Gave a presentation at three of these: one on root estimation in Galton-Watson
-trees, and two on generating functions and elementary analytic combinatorics.)
-
-\medbreak
-\leftright{\bf Sorting algorithms on manifolds}{\bf May -- August 2020}
-\leftright{\sl Summer research project}{\sl McGill University}
-\smallskip
-Research project under the supervision of Michael Lipnowski.
-Studied algorithms that investigate the topology of group actions on locally symmetric spaces.
-Work involved writing of code in OCaml and PostScript that generates illustrations of certain quotient
-spaces, in various models of hyperbolic geometry.
-\medbreak
-
-\leftright{\bf Interactive proofs}{\bf September -- December 2019}
-\leftright{\sl Honours research project}{\sl McGill University}
-\smallskip
-Semester-long research project in the Computation and Logic Group, supervised
-by Brigitte Pientka.
-Proved a theorem in constructive logic concerning the formal verification
-of the interactive proof assistant Harpoon and wrote OCaml code as part of
-ongoing work on the functional programming language Beluga.
-\medbreak
-
-\leftright{\bf Bytecode compiler and interpreter}{\bf February -- June 2019}
-\leftright{\sl Individual software project}{\sl Charles University}
-\smallskip
-Semester-long individual software project under the supervision of Adam Dingle.
-Created OPythn, a bytecode compiler and virtual machine for a subset of Python.
-OPythn includes support for lists, dictionaries, named and anonymous functions, objects, and classes.
-It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing.
+Analysis of random discrete structures, analysis and design of algorithms, enumerative combinatorics,
+additive combinatorics, algebra.
 
@@
 August 2020), 10 pp.
+\repitem Finding regularity in Tlingit verb prefixes. Semester project report, McGill University
+(Montr\'eal, Qu\'ebec, April 2021), 7 pp.
+
+\sectheader{Research Experience}
+
+\leftright{\bf Probabilistic analysis of branching processes}{\bf May 2020 -- present}
+\leftright{\sl Research group}{\sl McGill University}
+\smallskip
+Ongoing research on branching processes, headed by Luc Devroye. Studied estimation problems
+on Galton-Watson trees. Attended informal seminars over the year, discussing
+various topics related to branching
+processes. (Gave a presentation at five of these: one on root estimation in Galton-Watson
+trees, two on generating functions and elementary analytic combinatorics, and two on graph regularity.)
+[S1]
+
+\medbreak
+\leftright{\bf Sorting algorithms on manifolds}{\bf May -- August 2020}
+\leftright{\sl Summer research project}{\sl McGill University}
+\smallskip
+Research project under the supervision of Michael Lipnowski.
+Studied algorithms that investigate the topology of group actions on locally symmetric spaces.
+Work involved writing of code in OCaml and PostScript that generates illustrations of certain quotient
+spaces, in various models of hyperbolic geometry. [R3]
+\medbreak
+
+\leftright{\bf Interactive proofs}{\bf September -- December 2019}
+\leftright{\sl Honours research project}{\sl McGill University}
+\smallskip
+Semester-long research project in the Computation and Logic Group, supervised
+by Brigitte Pientka.
+Proved a theorem in constructive logic concerning the formal verification
+of the interactive proof assistant Harpoon and wrote OCaml code as part of
+ongoing work on the functional programming language Beluga. [R2]
+\medbreak
+
+\leftright{\bf Bytecode compiler and interpreter}{\bf February -- June 2019}
+\leftright{\sl Individual software project}{\sl Charles University}
+\smallskip
+Semester-long individual software project under the supervision of Adam Dingle.
+Created OPythn, a bytecode compiler and virtual machine for a subset of Python.
+OPythn includes support for lists, dictionaries, named and anonymous functions, objects, and classes.
+It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing. [R1]
 
@@
 
-\leftright{\bf Department of Mathematics and Statistics, McGill University}{\bf September 2019 -- present}
-\leftright{\sl Student Grader}{\sl Montr\'eal, QC, Canada}
+\leftright{\bf First responder}{\bf October 2017 -- present}
+\leftright{\sl McGill Student Emergency Response Team}{\sl Montr\'eal, QC, Canada}
+\smallskip
+On call on a weekly basis to provide emergency medical care at campus residences overnight as well as at
+university events such as frosh, sports games, and formals. Attend team training sessions twice a month to keep
+first-aid skills up-to-date. Most recent first responder certification: September 2021.
+\medbreak
+
+\leftright{\bf Grader}{\bf September 2019 -- April 2021}
+\leftright{\sl Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
 \smallskip
@@
 
-\leftright{\bf Computer Science Undergraduate Society Helpdesk}{\bf September 2018 -- present}
-\leftright{\sl Tutor}{\sl Montr\'eal, QC, Canada}
+\leftright{\bf Helpdesk tutor}{\bf September 2018 -- April 2021}
+\leftright{\sl Computer Science Undergraduate Society Helpdesk}{\sl Montr\'eal, QC, Canada}
 \smallskip
@@
 
-\leftright{\bf McGill Student Emergency Response Team}{\bf October 2017 -- present}
-\leftright{\sl First Responder}{\sl Montr\'eal, QC, Canada}
-\smallskip
-On call on a weekly basis to provide emergency medical care at campus residences overnight as well as at
-university events such as frosh, sports games, and formals. Attend team training sessions twice a month to keep
-first-aid skills up-to-date.
-\medbreak
-
-\leftright{\bf Society of Undergraduate Mathematics Students}{\bf May 2019 -- January 2020}
-\leftright{\sl Vice President, Academic}{\sl Montr\'eal, QC, Canada}
+\leftright{\bf Vice President, Academic}{\bf May 2019 -- January 2020}
+\leftright{\sl Society of Undergraduate Mathematics Students}{\sl Montr\'eal, QC, Canada}
 \smallskip
@@
 
-\leftright{\bf Bakir Contracting Corp.}{\bf May -- August 2018}
-\leftright{\sl Painter}{\sl Edmonton, AB, Canada}
+\leftright{\bf Painter}{\bf May -- August 2018}
+\leftright{\sl Bakir Contracting Corp.}{\sl Edmonton, AB, Canada}
 \smallskip
@@
 
-\leftright{\bf Singapore Armed Forces}{\bf August 2015 -- August 2017}
-\leftright{\sl Infantryman}{\sl Singapore}
+\leftright{\bf Infantryman}{\bf August 2015 -- August 2017}
+\leftright{\sl Singapore Armed Forces}{\sl Singapore}
 \smallskip
@@
 
-\leftright{\bf \'Ecole Secondaire Beaumont Composite High School}{\bf September 2014 -- June 2015}
-\leftright{\sl Tutor}{\sl Beaumont, AB, Canada}
+\leftright{\bf Tutor}{\bf September 2014 -- June 2015}
+\leftright{\sl \'Ecole Secondaire Beaumont Composite High School}{\sl Beaumont, AB, Canada}
 \smallskip
@@
 
-\leftright{\bf YoWoChAs Outdoor Education Centre}{\bf June -- August 2014}
-\leftright{\sl Summer Camp Counsellor}{\sl Fallis, AB, Canada}
+\leftright{\bf Summer camp counsellor}{\bf June -- August 2014}
+\leftright{\sl YoWoChAs Outdoor Education Centre}{\sl Fallis, AB, Canada}
 \smallskip
@@
 {\bf Programming Languages}\par
-\CEE, OCaml, Haskell, Standard ML, Scheme Lisp, PostScript, {\tt CWEB}, {\tt MIXAL}, Python.
+\CEE, OCaml, Python, Java, PostScript, Haskell, Scheme Lisp,  Standard ML, {\tt CWEB}, {\tt MIXAL}, MIPS Assembly.
 \medbreak
@@
 
-\sectheader{Interests}
-
-Analysis and design of algorithms, analysis of random discrete structures, enumerative combinatorics,
-additive combinatorics, group theory,
-programming language theory and design, compilers.
-
 \sectheader{Other}
@@
 \smallskip
-\thing Contributed sequences A335562, A338550, A338993, A339941, and A339942
+\thing Contributed sequences A335562, A338550, A338993, A339941, A339942, and A341822
 to the On-line Encyclopedia of Integer Sequences.
 \medbreak
-
-\rightline{\eightrm \datethis}
 

### 2021-05-26 about/vita.tex
New paper

@@
 \leftright{Email: marcel\_goh@yahoo.ca}{Montr\'eal, QC H2X 2H3, Canada}
-\leftright{Webpage: marcelgoh.github.io}{+1\thinspace\up(825\up)\thinspace 440-0681}
+\leftright{Webpage: marcelgoh.ca}{+1\thinspace\up(825\up)\thinspace 440-0681}
 
@@
 \subitem (with Anna M. Brandenberger and Luc Devroye)
-Root estimation in Galton-Watson trees. {\sl arXiv preprint 2007.05681}.
+Root estimation in Galton--Watson trees. {\sl arXiv preprint 2007.05681}.
 Submitted July 2020.
@@
 Submitted December 2020.
+\subitem (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
+Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl arXiv preprint 2105.12046}.
+Submitted May 2021.
 
@@
 trees, two on generating functions and elementary analytic combinatorics, and two on graph regularity.)
-[S1]
+[S1, S3]
 

### 2021-05-26 misc/expo/incidence.tex
New paper

\input fontmac
\input mathmac

\def\down{\downarrow\!}
\def\up{\uparrow\!}
\maketitle{Posets and incidence algebras}{by}{Marcel K. Goh}{23 May 2021}

\advsect Preliminaries

A {\it partially ordered set} or {\it poset}
is a pair $(P,\leq)$ where $P$ is a ground set and $\leq$ is a binary relation on $P$
that satisfies the following three properties.
\medskip\axiombegin P
\axiom P1. [Reflexivity.] For all $x\in P$, $x\leq x$.
\smallskip
\axiom P2. [Antisymmetry.] For all $x,y\in P$, $x\leq y$ and $y\leq x$ implies that $x=y$.
\smallskip
\axiom P3. [Transitivity.] For all $x,y,z\in P$, if $x\leq y$ and $y\leq z$, then $x\leq z$.
\medskip
\noindent
When $x\leq y$, we write $y\geq x$; when $x\leq y$ and $x\neq y$, we may write $x<y$ and $y>x$.
Since it is cumbersome to always to refer to a poset as a pair,
will also allow ourselves to denote the entire poset by its ground set $P$ when no confusion can arise.

For our purposes, a {\it subposet} of a poset $(P,\leq)$ is obtained by taking a subset $S\subseteq P$
and declaring that $x\leq y$ in $(S,\leq)$ if and only if $x\leq y$ in $(P,\leq)$ (this is sometimes
called an {\it induced} subposet). An {\it order ideal} is a subset $I$ of $P$ such that if $x\in I$ and
$y<x$, then $y\in I$. Dually, a {\it filter} is a
subset $F$ of $P$ such that if $x\in I$ and $y>x$, then $y\in I$. The {\it principal order ideal} $\down x$
of an element $x\in P$ is the set of all $y\in P$ such that $y\leq x$; similarly, the {\it order ideal generated
by $X = \{x_1,x_2,\ldots,x_k\}\subseteq P$} is the set $\down X$
of all $y\in P$ such that $y\leq x_i$ for some $1\leq i\leq k$.
(One defines the dual notion of a {\it principal filter} $\up x$ and {\it filter $\up X$
generated by $X$} in the obvious way.)

Whenever $x\leq y$, we can define the {\it interval} $[x,y]$ to be the subposet $\{z\in P: x\leq z\leq y\}$.
If the interval $[x,y]$ is the two-element set $\{x,y\}$, then we say that $y$ {\it covers} $x$.
Given two elements $x,y\in P$, we define an {\it upper bound} to be an element $z\in P$ such that
$z\geq x$ and $z\geq y$. Similarly, we can define a {\it lower bound} of two elements, and by induction,
we obtain a notion of upper and lower bounds of any finite subset of $P$. If an upper bound $z$ of $x$ and $y$
is such that for any upper bound $w$ of $x$ and $y$, we have $z\leq w$, then $z$ is said to be the {\it least
upper bound} or {\it join}
of $x$ and $y$ and we may denote $z=x\vee y$. Dually, we define the {\it greatest lower bound} or {\it meet}
$x\wedge y$ of two elements $x$ and $y$. In either case, the existence of such an element
implies its uniqueness.

\bye

### 2021-06-11 misc/proofs/coprime-probability.tex
Spacing and punctuation fix
Lattice of arithmetic progressions

--- 2019-07-25 misc/proofs/coprime-probability.tex
@@
 If two integers $u,v$ have no prime factors in common (i.e.\ $\gcd(u,v) = 1$), we say they are {\it relatively prime} and write $u\perp v$.
-For any positive integer $n$, let $q_n$ denote the number of ordered pairs of integers $(u,v)$ such that $1\leq u,v\leq n$and $u\perp v$.
+For any positive integer $n$, let $q_n$ denote the number of ordered pairs of integers $(u,v)$ such that $1\leq u,v\leq n$ and $u\perp v$.
 Then the probability that any pair of integers $u,v$ lying in the range $[1\ ..\ n]$ are relatively prime is $q_n/n^2$.
@@
 We will want to put asymptotic bounds on these rearranged sums. First, note that for all positive real $x$,
-$$x^2 - \lfloor x\rfloor^2 = O(x).\oldno 2$$
-For $x^2 - \lfloor x\rfloor ^2$ can be rewritten $(x + \lfloor x\rfloor)(x-\lfloor x\rfloor)$, which is less than $2x$.
+$$x^2 - \lfloor x\rfloor^2 = O(x),\oldno 2$$
+for $x^2 - \lfloor x\rfloor ^2$ can be rewritten $(x + \lfloor x\rfloor)(x-\lfloor x\rfloor)$, which is less than $2x+1$.
 Next, we use standard formulas to analyse the asymptotic behaviour of the series $\sum_{k>n}(n/k)^2$:
@@
 Since the magnitude of $\mu(d)$ is 1 (whenever $d$ is square-free) and the sign of $\mu(d)$ depends on the number of primes that make up $d$, this can be re-expressed as the alternating sum of binomial coefficients
-$$\sum_{d\backslash n}\mu(d) = {k \choose 0} - {k\choose 1} + {k\choose 2} - \cdots + (-1)^k{k\choose k}.$$
-which equals 0. (We will not prove this claim, but one may be convinced of this fact by trying it out on a couple of rows of Pascal's Triangle.)\slug
+$$\sum_{d\backslash n}\mu(d) = {k \choose 0} - {k\choose 1} + {k\choose 2} - \cdots + (-1)^k{k\choose k} = (1-1)^k = 0.\noskipslug$$
 
@@
 \proclaim Lemma C.
-$$\biggl(\sum_{k\geq 1} {\mu(k) \over k^2}\biggr) \biggl( \sum_{m \geq 1} {1\over m^2}\biggr) = 1$$.
+$$\biggl(\sum_{k\geq 1} {\mu(k) \over k^2}\biggr) \biggl( \sum_{m \geq 1} {1\over m^2}\biggr) = 1.$$
 

### 2021-06-29 about/vita.tex
Lattice of arithmetic progressions
Indepno

@@
 \subitem (with Anna M. Brandenberger and Luc Devroye)
-Root estimation in Galton--Watson trees. {\sl arXiv preprint 2007.05681}.
+Root estimation in Galton--Watson trees. {\sl arXiv preprint 2007.05681}, 20~pp.
 Submitted July 2020.
 \subitem (with Rosie Y. Zhao)
-Arithmetic subsequences in a random ordering of an additive set. {\sl arXiv preprint 2012.12339}.
+Arithmetic subsequences in a random ordering of an additive set. {\sl arXiv preprint 2012.12339}, 11~pp.
 Submitted December 2020.
 \subitem (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
-Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl arXiv preprint 2105.12046}.
+Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl arXiv preprint 2105.12046}, 11~pp.
 Submitted May 2021.
-
-\goodbreak
+\subitem (with Jad Hamdan) The lattice of arithmetic progressions. {\sl arXiv preprint 2106.05949}, 8~pp.
+Submitted June 2021.
+\subitem (with Luc Devroye and Rosie Y. Zhao) The independence number of a Bienaym\'e--Galton--Watson tree
+and related parameters. {\sl arXiv preprint 2106.14389}, 20~pp.
+Submitted June 2021.
 
@@
 trees, two on generating functions and elementary analytic combinatorics, and two on graph regularity.)
-[S1, S3]
+[S1, S3, S5]
 

### 2021-08-20 misc/literate/lp-balls.tex
Small changes here and there

\input cwebmac
\def\cwebtitle#1{\gdef\title{\expandafter\uppercase\expandafter{#1}}}
\def\namedatethis{
  \def\startsection{
    \leftline{\sc Written by Marcel K. Goh. Last updated \today\ at \hours}%
\bigskip
    \let\startsection=\stsec\stsec
  }
}
\cwebtitle{lp-balls}

\input fontmac
\input epsf

\def\bb{{\bf b}}
\def\setZ{{\bf Z}}
\def\RR{{\bf R}}
\def\norm#1{\vert\!\vert#1\vert\!\vert}

\namedatethis

\N{1}{1}Introduction. This literate program contains various functions for
experimenting with points
in the unit ball of $L^p$-space over $\RR^d$. We will maintain a set of points
in this space, and provide
functionality for randomly generating new ones, as well as printing the current
state of the point set.

\fi

\M{2}This is the main outline of the program. We have a couple of global
variables storing $p\in (0,\infty]$,
the dimension $d$, the set \PB{\\{points}} of points we are working with (and
the maximum size \PB{\\{max\_points}}
of this array). The three parameters will be supplied by the user via
command-line arguments. One specifies
the case $p=\infty$ by supplying a negative number for $p$.
We represent vectors in $\RR^d$ as \CEE/ \PB{\&{double}} arrays of length
$d+1$. While perhaps a bit wasteful,
this allows us to index from $1$ to $d$ rather than from $0$ to $d-1$.

\Y\B\8\#\&{include} \.{<float.h>}\6
\8\#\&{include} \.{<math.h>}\6
\8\#\&{include} \.{<stdio.h>}\6
\8\#\&{include} \.{<stdlib.h>}\6
\8\#\&{include} \.{<time.h>}\7
\&{double} \|p;\6
\&{double} \\{p\_inv};\6
\&{int} \|d;\6
\&{int} \\{max\_points};\6
\&{int} \\{num\_points};\6
\&{double} ${}{*}{*}\\{points}{}$;\7
\X12:Print points to PostScript file\X;\6
\X3:Point set functions\X;\6
\X4:Random variate generation\X;\6
\X15:Committee-generating algorithm\X;\6
\X19:Inner product computation\X;\6
\X20:Draw loci when $p=\infty$\X;\7
\&{int} \\{main}(\&{int} \\{argc}${},\39{}$\&{char} ${}{*}\\{argv}[\,]){}$\1\1%
\2\2\6
${}\{{}$\1\6
\X24:Gather input and initialise global variables\X;\6
\X5:Seed the pseudorandom number generator\X;\6
\&{if} ${}(\|d\E\T{2}\W\|p<\T{0}){}$\1\5
${}\\{random\_locus}(\T{5},\39\.{"locus.ps"},\39\T{150},\39\T{30});{}$\2\6
\4${}\}{}$\2\par
\fi

\N{1}{3}The point set.
This section contains functions on points in $L^p$-space.
First, we provide facilities to add new points, delete all points, and to list
the current point set on
the console.
We use the current size of the point set, stored in the variable \PB{\\{num%
\_points}} to help index into
the \PB{\\{points}} array. Clearing the point set is done by setting \PB{\\{num%
\_points}} to $0$, so everything
in the array at and after the index \PB{\\{num\_points}} possibly contains
garbage values.

\Y\B\4\X3:Point set functions\X${}\E{}$\6
\&{void} \\{add\_point}(\&{double} ${}{*}\\{point}){}$\1\1\2\2\6
${}\{{}$\1\6
\&{if} ${}(\\{num\_points}\G\\{max\_points}){}$\5
${}\{{}$\1\6
\\{printf}(\.{"I\ failed\ to\ add\ poi}\)\.{nt\ because\ array\ is\ }\)\.{full.%
\\n"});\6
\&{return};\6
\4${}\}{}$\2\6
\&{for} (\&{int} \|i${}\K\T{1};{}$ ${}\|i\Z\|d;{}$ ${}\PP\|i){}$\1\5
${}\\{points}[\\{num\_points}][\|i]\K\\{point}[\|i];{}$\2\6
${}\PP\\{num\_points};{}$\6
\4${}\}{}$\2\7
\&{void} \\{clear\_points}(\,)\1\1\2\2\6
${}\{{}$\1\6
${}\\{num\_points}\K\T{0};{}$\6
\4${}\}{}$\2\7
\&{void} \\{list\_points}(\,)\1\1\2\2\6
${}\{{}$\1\6
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{num\_points};{}$ ${}\PP\|i){}$\5
${}\{{}$\1\6
\\{printf}(\.{"("});\6
\&{for} (\&{int} \|j${}\K\T{1};{}$ ${}\|j\Z\|d;{}$ ${}\PP\|j){}$\5
${}\{{}$\1\6
${}\\{printf}(\.{"\%.2f"},\39\\{points}[\|i][\|j]);{}$\6
\&{if} ${}(\|j\I\|d){}$\1\5
\\{printf}(\.{",\ "});\2\6
\4${}\}{}$\2\6
\\{printf}(\.{")\\n"});\6
\4${}\}{}$\2\6
\4${}\}{}$\2\par
\U2.\fi

\N{1}{4}Random variate generation.
The main goal is to generate a point uniformly at random from the interior of
the unit ball in $L^p$-space
over $\RR^d$.
To do this, we
will require an exponential random variable with mean $1$.

\Y\B\4\X4:Random variate generation\X${}\E{}$\6
\X6:Uniform and exponential variates\X;\6
\X11:Normal random variate\X;\6
\X7:Gamma random variate\X;\6
\X8:Uniform random point in unit ball\X;\par
\U2.\fi

\M{5}First, in the \PB{\\{main}} function, we initialise
a pseudorandom number generator with the current time.

\Y\B\4\X5:Seed the pseudorandom number generator\X${}\E{}$\6
\&{time\_t} \|t;\7
\\{srand}((\&{unsigned}) \\{time}${}({\AND}\|t));{}$\6
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\T{10};{}$ ${}\PP\|i){}$\5
${}\{{}$\1\6
\\{rand}(\,);\6
\4${}\}{}$\2\par
\U2.\fi

\M{6}To generate our exponential random variable, we use von Neumann's
algorithm,
as described by L.~Devroye on p.~126 of {\sl Non-Uniform Random Variate
Generation} (New York: Springer, 1986).

\Y\B\4\X6:Uniform and exponential variates\X${}\E{}$\6
\&{double} \\{uniform01}(\,)\1\1\2\2\6
${}\{{}$\1\6
\&{return} ((\&{double}) \\{rand}(\,))${}/\.{RAND\_MAX};{}$\6
\4${}\}{}$\2\7
\&{double} \\{exponential1}(\,)\1\1\2\2\6
${}\{{}$\1\6
\&{int} \|Z${}\K\T{0};{}$\6
\&{double} \|Y;\6
\&{int} \|k;\7
\&{do}\5
${}\{{}$\1\6
${}\|Y\K\\{uniform01}(\,);{}$\6
${}\|k\K\T{1};{}$\7
\&{double} \|W${}\K\|Y;{}$\6
\&{int} \\{stop}${}\K\T{0};{}$\7
\&{do}\5
${}\{{}$\1\6
\&{double} \|U${}\K\\{uniform01}(\,);{}$\7
\&{if} ${}(\|U>\|W){}$\5
${}\{{}$\1\6
${}\\{stop}\K\T{1};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\|W\K\|U;{}$\6
${}\PP\|k;{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\5
\&{while} ${}(\R\\{stop});{}$\6
${}\PP\|Z;{}$\6
\4${}\}{}$\2\5
\&{while} ${}(\|k\MOD\T{2}\E\T{0});{}$\6
\&{return} ${}((\&{double})(\|Z-\T{1}))+\|Y;{}$\6
\4${}\}{}$\2\par
\U4.\fi

\M{7}We will also make use of the gamma distribution, which, for a parameter
$a>0$, has density
$$f(x) = {1\over \Gamma(a)}x^{a-1}e^{-x}.$$
We use an algorithm of R.~C.~H.~Cheng [{\sl Applied Statistics} {\bf 26}
(1977), 71--75], described on
p.~413 of {\sl Non-Uniform Random Variate Generation}, which works for $a\ge
1$. (Note that there is
a mistake in the printed algorithm, but the corrigenda on Devroye's website
contain the required amendments.)

\Y\B\4\X7:Gamma random variate\X${}\E{}$\6
\&{double} \\{gamma\_dist}(\&{double} \|a)\1\1\2\2\6
${}\{{}$\1\6
\&{double} \|b${}\K\|a-\\{log}(\T{4});{}$\6
\&{double} \\{lambda}${}\K\\{sqrt}(\T{2}*\|a-\T{1});{}$\6
\&{double} \|c${}\K\|a+\\{lambda};{}$\6
\&{int} \\{accept}${}\K\T{0};{}$\6
\&{double} \|U${},{}$ \|V${},{}$ \|Y${},{}$ \|X${},{}$ \|Z${},{}$ \|R;\6
\&{double} \|S${}\K\T{4.5}*\|Z-(\T{1}+\\{log}(\T{4.5}));{}$\7
\&{do}\5
${}\{{}$\1\6
${}\|U\K\\{uniform01}(\,);{}$\6
${}\|V\K\\{uniform01}(\,);{}$\6
${}\|Y\K(\T{1.0}/\\{lambda})*\\{log}(\|V/(\T{1}-\|V));{}$\6
${}\|X\K\|a*\\{exp}(\|Y);{}$\6
${}\|Z\K\|U*\|V*\|V;{}$\6
${}\|R\K\|b+\|c*\|Y-\|X;{}$\6
${}\\{accept}\K(\|R\G\|S);{}$\6
\&{if} ${}(\R\\{accept}){}$\1\5
${}\\{accept}\K(\|R\G\\{log}(\|Z));{}$\2\6
\4${}\}{}$\2\5
\&{while} ${}(\R\\{accept});{}$\6
\&{return} \|X;\6
\4${}\}{}$\2\par
\U4.\fi

\M{8}To get a uniform random point from the unit ball in $L^p(\RR^d)$, we
sample $d$ independent
random variables, call them $X_1,\ldots, X_d$ from the density
$$f(x) = {1\over 2\Gamma(1+1/p)} e^{-\vert x\vert ^p}.$$
This is called an exponential power distribution, and Devroye notes in {\sl
Non-Uniform Random Variate Generation}
that if $X = VY^{1/p}$ with $V$ uniformly distributed on $[-1,1]$ and $Y$ is
gamma-distributed with parameter
$1+1/p$, then $X$ has the density $f(x)$.
We also sample an exponential random variable $Y$ with mean $1$, and then
output the random vector
$${(X_1, \ldots, X_d) \over (Y+ \sum_{i=1}^d \vert X_i\vert^p)^{1/p}}.$$
This method and extensions thereof are described by F.~Barthe, O.~Guedon,
S.~Mendelson, and A.~Naor
[{\sl Annals of Probability} {\bf 33} (2005), 480--513].
The function we write returns a point by modifying an
array that is passed as an argument.

\Y\B\4\X8:Uniform random point in unit ball\X${}\E{}$\6
\&{void} \\{random\_point}(\&{double} ${}{*}\\{point}){}$\1\1\2\2\6
${}\{{}$\1\6
\&{double} ${}{*}\|X\K\\{malloc}((\|d+\T{1})*\&{sizeof}(\&{double}));{}$\7
\&{for} (\&{int} \|i${}\K\T{1};{}$ ${}\|i\Z\|d;{}$ ${}\PP\|i){}$\1\5
${}\|X[\|i]\K\T{0.0};{}$\2\6
\&{if} ${}(\|p<\T{0}){}$\5
${}\{{}$\1\6
\X9:The case $p=\infty$\X;\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\X10:The case $p\ne\infty$\X;\6
\4${}\}{}$\2\6
\\{free}(\|X);\6
\4${}\}{}$\2\par
\U4.\fi

\M{9}When $p=\infty$, we simply return a point uniformly from the box
$[-1,1]^d$.

\Y\B\4\X9:The case $p=\infty$\X${}\E{}$\6
\&{for} (\&{int} \|i${}\K\T{1};{}$ ${}\|i\Z\|d;{}$ ${}\PP\|i){}$\1\5
${}\\{point}[\|i]\K\T{2}*\\{uniform01}(\,)-\T{1}{}$;\2\par
\U8.\fi

\M{10}In all the other cases, we sample using the method described above.

\Y\B\4\X10:The case $p\ne\infty$\X${}\E{}$\6
\&{for} (\&{int} \|i${}\K\T{1};{}$ ${}\|i\Z\|d;{}$ ${}\PP\|i){}$\5
${}\{{}$\1\6
\&{double} \|V${}\K\T{2}*\\{uniform01}(\,)-\T{1};{}$\7
${}\|X[\|i]\K\|V*\\{pow}(\\{gamma\_dist}(\T{1}+\\{p\_inv}),\39\\{p\_inv});{}$\6
\4${}\}{}$\2\7
\&{double} \|Y${}\K\\{exponential1}(\,);{}$\6
\&{double} \\{pow\_sum}${}\K\|Y;{}$\7
\&{for} (\&{int} \|i${}\K\T{1};{}$ ${}\|i\Z\|d;{}$ ${}\PP\|i){}$\1\5
${}\\{pow\_sum}\MRL{+{\K}}\\{pow}(\\{fabs}(\|X[\|i]),\39\|p);{}$\2\7
\&{double} \\{scale\_factor}${}\K\T{1.0}/(\\{pow}(\\{pow\_sum},\39\\{p%
\_inv}));{}$\7
\&{for} (\&{int} \|i${}\K\T{1};{}$ ${}\|i\Z\|d;{}$ ${}\PP\|i){}$\1\5
${}\\{point}[\|i]\K\\{scale\_factor}*\|X[\|i]{}$;\2\par
\U8.\fi

\M{11}I originally needed a normal random deviate to handle the case $p=2$, so
I added this function,
but it is no longer necessary since we now have a general function. I left it
here just for kicks.
We generate the normal using the Box-Muller transform,
due to G.~E.~P.~Box and M.~E.~Muller [{\sl Annals of Mathematical Statistics} {%
\bf 29} (1958), 610--611].
If $U_1$ and $U_2$ are two independent random variables uniformly distributed
in $[0,1]$, then
$$V_1 = \sqrt{-2\ln U_1}\cos(2\pi U_2)\qquad\hbox{and}\qquad V_2 = \sqrt{-2\ln
U_1}\sin(2\pi U_2)$$
are independent normal random variables with mean $0$ and variance $1$. To make
the return type
of our function simpler, we
simply output one of these values (so we will end up calling this function
twice as often as is actually
necessary).

\Y\B\4\X11:Normal random variate\X${}\E{}$\6
\&{double} \\{normal01}(\,)\1\1\2\2\6
${}\{{}$\1\6
\&{double} \.{U1}${}\K\\{uniform01}(\,);{}$\6
\&{double} \.{U2}${}\K\\{uniform01}(\,);{}$\6
\&{double} \\{scale}${}\K\\{sqrt}({-}\T{2.0}*\\{log}(\.{U1}));{}$\7
\&{return} \\{scale}${}*\\{cos}(\T{2}*\.{M\_PI}*\.{U2});{}$\6
\4${}\}{}$\2\par
\U4.\fi

\N{1}{12}File output.
When $d=2$, it is easy to plot our set of points graphically. Our program does
this by generating a PostScript
file.

\Y\B\4\X12:Print points to PostScript file\X${}\E{}$\6
\&{void} \\{plot\_single\_point}(\&{FILE} ${}{*}\\{file},\39{}$\&{double} %
\\{red}${},\39{}$\&{double} \\{green}${},\39{}$\&{double} \\{blue}${},\39{}$%
\&{double} \|x${},\39{}$\&{double} \|y)\1\1\2\2\6
${}\{{}$\1\6
${}\\{fprintf}(\\{file},\39\.{"\%f\ \%f\ \%f\ setrgbcolo}\)\.{r\ \%f\ \%f\ dot%
\\n"},\39\\{red},\39\\{green},\39\\{blue},\39\|x,\39\|y);{}$\6
\4${}\}{}$\2\7
\&{void} \\{to\_postscript}(\&{const} \&{char} ${}{*}\\{filename},\39{}$\&{int}
\\{radius})\1\1\2\2\6
${}\{{}$\1\6
\&{if} ${}(\|d\I\T{2}){}$\5
${}\{{}$\1\6
\\{printf}(\.{"I\ cannot\ output\ Pos}\)\.{tScript\ unless\ d\ equ}\)\.{als\ 2!%
\\n"});\6
\&{return};\6
\4${}\}{}$\2\7
\&{FILE} ${}{*}\\{file}\K\\{fopen}(\\{filename},\39\.{"w"});{}$\7
\X13:Preamble\X;\6
\X14:Plot the points array\X;\6
${}\\{fprintf}(\\{file},\39\.{"showpage\\n"});{}$\6
\\{fclose}(\\{file});\6
\4${}\}{}$\2\par
\U2.\fi

\M{13}We first add a bare-bones preamble to the file to draw the axes and
declare various PostScript functions.
The axes are drawn with center at $(\PB{\\{radius}}, \PB{\\{radius}})$

\Y\B\4\X13:Preamble\X${}\E{}$\6
\&{char} ${}{*}\\{preamble}\K\.{"\%!PS\\n/dot\ \{\ 1.5\ 0\ }\)\.{360\ arc\
closepath\ fi}\)\.{ll\ \}\ def\\n/square\ \{\ }\)\.{/r\ exch\ def\ /y\ exch\ }%
\)\.{def\ /x\ exch\ def\\nnew}\)\.{path\ x\ r\ sub\ y\ r\ sub}\)\.{\ moveto\ 0\
r\ 2\ mul\ rl}\)\.{ineto\\nr\ 2\ mul\ 0\ rli}\)\.{neto\ 0\ r\ 2\ mul\ neg\ r}\)%
\.{lineto\\nr\ 2\ mul\ neg\ }\)\.{0\ rlineto\ closepath\ }\)\.{fill\}\ def%
\\n0.5\ setli}\)\.{newidth\\n"};{}$\7
${}\\{fprintf}(\\{file},\39\.{"\%s"},\39\\{preamble});{}$\6
${}\\{fprintf}(\\{file},\39\.{"\%d\ \%d\ translate\\n"},\39\\{radius},\39%
\\{radius});{}$\6
${}\\{fprintf}(\\{file},\39\.{"newpath\ 0\ \%d\ moveto}\)\.{\ 0\ \%d\ lineto\
"},\39{-}\\{radius},\39\\{radius});{}$\6
${}\\{fprintf}(\\{file},\39\.{"\%d\ 0\ moveto\ \%d\ 0\ li}\)\.{neto\ stroke%
\\n"},\39{-}\\{radius},\39\\{radius}){}$;\par
\Us12\ET23.\fi

\M{14}Next, we plot each of the elements of the \PB{\\{points}} array, with a
colour gradient to indicate the relative
orderings of points. The first point in the array is drawn in \PB{\\{colour1}},
the final point is drawn in
\PB{\\{colour2}}, and points in between have their colours interpolated
accordingly.

\Y\B\4\X14:Plot the points array\X${}\E{}$\6
\&{double} \\{colour1}[\,]${}\K\{\T{1.0},\39\T{0.0},\39\T{0.0}\}{}$;\C{ red }\6
\&{double} \\{colour2}[\,]${}\K\{\T{0.0},\39\T{0.0},\39\T{1.0}\}{}$;\C{ blue }\7
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{num\_points};{}$ ${}\PP\|i){}$\5
${}\{{}$\1\6
\&{double} \|t${}\K{}$((\&{double}) \|i)${}/((\&{double})(\\{num\_points}-%
\T{1}));{}$\6
\&{double} \|r${}\K(\T{1.0}-\|t)*\\{colour1}[\T{0}]+\|t*\\{colour2}[\T{0}];{}$\6
\&{double} \|g${}\K(\T{1.0}-\|t)*\\{colour1}[\T{1}]+\|t*\\{colour2}[\T{1}];{}$\6
\&{double} \|b${}\K(\T{1.0}-\|t)*\\{colour1}[\T{2}]+\|t*\\{colour2}[\T{2}];{}$\7
${}\\{plot\_single\_point}(\\{file},\39\|r,\39\|g,\39\|b,\39\\{points}[\|i][%
\T{1}]*\\{radius},\39\\{points}[\|i][\T{2}]*\\{radius});{}$\6
\4${}\}{}$\2\par
\U12.\fi

\N{1}{15}Growing a committee by consensus voting.
Consider the point set as representing a group of people (each person is a
vector of real numbers
each describing a different trait). The committee grows in discrete time steps.
At each step, two
candidates are presented to the committee, and each committee member votes for
the candidate closest
to itself in $L^p$-distance. A candidate can only win by {\it consensus}; that
is, if any two committee
members vote for different candidates, then the election is inconclusive and no
new member is added.

\Y\B\4\X15:Committee-generating algorithm\X${}\E{}$\6
\X16:Distance computation\X;\6
\X17:Consensus algorithm\X;\par
\U2.\fi

\M{16}First we need to compute distances between two points $x = (x_1,\ldots,
x_d)$
and $y = (y_1,\ldots, y_d)$ in $L^p(\RR^d)$. When $p\neq\infty$, this is given
by the formula
$$\norm{x-y}_p = \Bigl(\sum_{i=1}^d \vert x_i-y_i\vert ^p\Bigr)^{1/p},$$
and when $p=\infty$, we simply take the maximum of the coordinate-wise
distances:
$$\norm{x-y}_\infty = \max_{1\leq i\leq d} \vert x_i - y_i\vert$$

\Y\B\4\X16:Distance computation\X${}\E{}$\6
\&{double} \\{p\_dist}(\&{double} ${}{*}\|x,\39{}$\&{double} ${}{*}\|y){}$\1\1%
\2\2\6
${}\{{}$\1\6
\&{if} ${}(\|p<\T{0}){}$\5
${}\{{}$\1\6
\&{double} \\{max}${}\K\.{DBL\_MIN};{}$\7
\&{for} (\&{int} \|i${}\K\T{1};{}$ ${}\|i\Z\|d;{}$ ${}\PP\|i){}$\5
${}\{{}$\1\6
\&{double} \\{diff}${}\K\\{fabs}(\|x[\|i]-\|y[\|i]);{}$\7
\&{if} ${}(\\{diff}>\\{max}){}$\1\5
${}\\{max}\K\\{diff};{}$\2\6
\4${}\}{}$\2\6
\&{return} \\{max};\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
\&{double} \\{sum}${}\K\T{0};{}$\7
\&{for} (\&{int} \|i${}\K\T{1};{}$ ${}\|i\Z\|d;{}$ ${}\PP\|i){}$\1\5
${}\\{sum}\MRL{+{\K}}\\{pow}(\\{fabs}(\|x[\|i]-\|y[\|i]),\39\|p);{}$\2\6
\&{return} \\{pow}${}(\\{sum},\39\T{1.0}/\|p);{}$\6
\4${}\}{}$\2\6
\4${}\}{}$\2\par
\U15.\fi

\M{17}Now we present the algorithm for growing the committee, which takes a
parameter indicating how many
rounds of voting should be undertaken. We start by clearing the point set and
initialising the committee with a point chosen uniformly at random from the
unit ball (this counts
as the first round of voting). In each round,
we compute distances from each point to the two candidates and keep track of
whether each candidate has
votes. If at any point, both candidates have votes, we can continue to the next
election, since that round
is inconclusive. The return value is the size of the committee after all rounds
have elapsed.

\Y\B\4\X17:Consensus algorithm\X${}\E{}$\6
\&{int} \\{consensus}(\&{int} \\{max\_t}${},\39\&{double}({*}\\{dist}){}$(%
\&{double} ${}{*},\39{}$\&{double} ${}{*})){}$\1\1\2\2\6
${}\{{}$\1\6
\\{clear\_points}(\,);\7
\&{double} ${}{*}\\{cand1}\K\\{malloc}((\|d+\T{1})*\&{sizeof}(\&{double}));{}$\6
\&{double} ${}{*}\\{cand2}\K\\{malloc}((\|d+\T{1})*\&{sizeof}(\&{double}));{}$\7
\\{random\_point}(\\{cand1});\6
\\{add\_point}(\\{cand1});\6
\&{for} (\&{int} \|t${}\K\T{0};{}$ ${}\|t<\\{max\_t}-\T{1};{}$ ${}\PP\|t){}$\5
${}\{{}$\1\6
\\{random\_point}(\\{cand1});\6
\\{random\_point}(\\{cand2});\7
\&{int} \\{voted1}${}\K\T{0};{}$\6
\&{int} \\{voted2}${}\K\T{0};{}$\7
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{num\_points};{}$ ${}\PP\|i){}$\5
${}\{{}$\1\6
\&{double} \\{dist1}${}\K\\{dist}(\\{points}[\|i],\39\\{cand1});{}$\6
\&{double} \\{dist2}${}\K\\{dist}(\\{points}[\|i],\39\\{cand2});{}$\7
\&{if} ${}(\\{dist1}>\\{dist2}){}$\5
${}\{{}$\1\6
${}\\{voted1}\K\T{1};{}$\6
\4${}\}{}$\2\6
\&{else}\5
${}\{{}$\1\6
${}\\{voted2}\K\T{1};{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{voted1}\W\\{voted2}){}$\1\5
\&{break};\2\6
\4${}\}{}$\2\6
\&{if} ${}(\\{voted1}\W\R\\{voted2}){}$\1\5
\\{add\_point}(\\{cand1});\2\6
\&{if} ${}(\\{voted2}\W\R\\{voted1}){}$\1\5
\\{add\_point}(\\{cand2});\2\6
\4${}\}{}$\2\6
\\{free}(\\{cand1});\6
\\{free}(\\{cand2});\6
\&{return} \\{num\_points};\6
\4${}\}{}$\2\par
\U15.\fi

\N{1}{18}Illustrations and examples.
Here are some examples of the output of the committee-generating algorithm. In
each case, $1\,000,000$
rounds of voting were conducted. In each figure, redder points were added
earlier and bluer points were
added later.

For $p=\infty$, a committee of 20 members was formed:
$$\epsfbox{infinity.ps}$$
For $p=1$, a committee of $16$ members was formed:
$$\epsfbox{one.ps}$$
And for $p=2$, a committee of $33$ members was formed:
$$\epsfbox{two.ps}$$
These committee sizes seemed representative of typical behaviour, in that
running the program multiple times
with the same parameters did not produce wildly different values in any of the
cases.

\fi

\N{1}{19}Orthogonal committees.
We now restrict ourselves to the case $p = 2$, when $L^p(\RR^d)$ is an
inner-product space. We can perform
the same experiment, but instead of measuring distance with the $L^p$-norm, we
can measure distance
between two vectors as the absolute value of their inner product.

\Y\B\4\X19:Inner product computation\X${}\E{}$\6
\&{double} \\{inner\_product}(\&{double} ${}{*}\|x,\39{}$\&{double} ${}{*}%
\|y){}$\1\1\2\2\6
${}\{{}$\1\6
\&{if} ${}(\|p\I\T{2}){}$\5
${}\{{}$\1\6
\\{printf}(\.{"I\ can\ only\ take\ inn}\)\.{er\ products\ when\ p\ =}\)\.{\ 2!%
\\n"});\6
\&{return} \T{0.0};\6
\4${}\}{}$\2\7
\&{double} \\{sum}${}\K\T{0.0};{}$\7
\&{for} (\&{int} \|i${}\K\T{1};{}$ ${}\|i\Z\|d;{}$ ${}\PP\|i){}$\1\5
${}\\{sum}\MRL{+{\K}}\|x[\|i]*\|y[\|i];{}$\2\6
\&{return} \\{sum};\6
\4${}\}{}$\2\7
\&{double} \\{abs\_inner\_product}(\&{double} ${}{*}\|x,\39{}$\&{double} ${}{*}%
\|y){}$\1\1\2\2\6
${}\{{}$\1\6
\&{return} \\{fabs}${}(\\{inner\_product}(\|x,\39\|y));{}$\6
\4${}\}{}$\2\7
\&{double} \\{one\_minus\_inner\_product}(\&{double} ${}{*}\|x,\39{}$\&{double}
${}{*}\|y){}$\1\1\2\2\6
${}\{{}$\1\6
\&{return} \T{1.0}${}-\\{inner\_product}(\|x,\39\|y);{}$\6
\4${}\}{}$\2\par
\U2.\fi

\N{1}{20}Locus of points further to committee than candidate.
In this section, we consider the case when $p=\infty$ and $d=2$. Given a
committee $G$ of points and a candidate
$c$, we will draw the locus of all points $c'$ such that the committee $G$ will
reach a consensus and elect
$c$ over $c'$.

\Y\B\4\X20:Draw loci when $p=\infty$\X${}\E{}$\6
\X21:Draw squares in PostScript\X;\6
\X22:Locus computation\X;\6
\X23:Generate a drawing with random points\X;\par
\U2.\fi

\M{21}First, we supply a facility to draw $L^\infty$ balls, which look like
squares, in PostScript.

\Y\B\4\X21:Draw squares in PostScript\X${}\E{}$\6
\&{void} \\{draw\_square}(\&{FILE} ${}{*}\\{file},\39{}$\&{double} \\{red}${},%
\39{}$\&{double} \\{green}${},\39{}$\&{double} \\{blue}${},\39{}$\&{double} %
\|x${},\39{}$\&{double} \|y${},\39{}$\&{double} \|r)\1\1\2\2\6
${}\{{}$\1\6
${}\\{fprintf}(\\{file},\39\.{"\%f\ \%f\ \%f\ setrgbcolo}\)\.{r\ \%f\ \%f\ \%f\
square\\n"},\39\\{red},\39\\{green},\39\\{blue},\39\|x,\39\|y,\39\|r);{}$\6
\4${}\}{}$\2\par
\U20.\fi

\M{22}Next, given a set of points $G = \{g_1, \ldots, g_n\}$
representing a committee as well as a candidate point $c$, the locus we seek is
$$[0,1]^2\setminus \bigcup_{i=1}^n B_\infty\bigl(g_i, d_\infty(g_i, c)\bigr),$$
where $d_\infty(x,y) = \norm{x-y}_\infty$ and $B_\infty(x,r)$ is the set of all
points $y$ with $d_\infty(x,y) <r$.
The way we draw this is to first fill in $[0,1]^2$ in red, then superimpose
various balls in blue. The remaining
red area is the locus of all points that is further from each committee member
than the candidate is.

\Y\B\4\X22:Locus computation\X${}\E{}$\6
\&{void} \\{draw\_locus}(\&{FILE} ${}{*}\\{file},\39{}$\&{int} \\{committee%
\_size}${},\39{}$\&{double} ${}{*}{*}\\{committee},\39{}$\&{double} ${}{*}%
\\{candidate},\39{}$\&{int} \\{radius})\1\1\2\2\6
${}\{{}$\1\6
${}\\{draw\_square}(\\{file},\39\T{1},\39\T{0},\39\T{0},\39\T{0},\39\T{0},\39%
\\{radius}){}$;\C{ background drawn in red }\6
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{committee\_size};{}$ ${}\PP\|i){}$%
\5
${}\{{}$\C{ complement of locus drawn in blue }\1\6
${}\\{draw\_square}(\\{file},\39\T{0.2},\39\T{0},\39\T{1},\39\\{committee}[%
\|i][\T{1}]*\\{radius},\39\\{committee}[\|i][\T{2}]*\\{radius},\39\\{p\_dist}(%
\\{committee}[\|i],\39\\{candidate})*\\{radius});{}$\6
\4${}\}{}$\2\6
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{committee\_size};{}$ ${}\PP\|i){}$%
\5
${}\{{}$\C{ draw committee in black }\1\6
${}\\{plot\_single\_point}(\\{file},\39\T{0},\39\T{0},\39\T{0},\39%
\\{committee}[\|i][\T{1}]*\\{radius},\39\\{committee}[\|i][\T{2}]*%
\\{radius});{}$\6
\4${}\}{}$\2\6
${}\\{plot\_single\_point}(\\{file},\39\T{0},\39\T{1},\39\T{0},\39%
\\{candidate}[\T{1}]*\\{radius},\39\\{candidate}[\T{2}]*\\{radius}){}$;\C{
candidate in green }\6
${}\\{draw\_square}(\\{file},\39\T{1},\39\T{1},\39\T{1},\39\T{2}*\\{radius},\39%
\T{0},\39\\{radius}){}$;\C{ clean up with white squares }\6
${}\\{draw\_square}(\\{file},\39\T{1},\39\T{1},\39\T{1},\39\T{0},\39\T{2}*%
\\{radius},\39\\{radius});{}$\6
${}\\{draw\_square}(\\{file},\39\T{1},\39\T{1},\39\T{1},\39\T{2}*\\{radius},\39%
\T{2}*\\{radius},\39\\{radius});{}$\6
\4${}\}{}$\2\par
\U20.\fi

\M{23}Lastly, we sample random points and generate the corresponding drawing.

\Y\B\4\X23:Generate a drawing with random points\X${}\E{}$\6
\&{void} \\{random\_locus}(\&{int} \\{committee\_size}${},\39{}$\&{const} %
\&{char} ${}{*}\\{filename},\39{}$\&{int} \\{radius}${},\39{}$\&{int} \\{num%
\_pages})\1\1\2\2\6
${}\{{}$\1\6
\&{double} ${}{*}\\{candidate}\K\\{malloc}((\|d+\T{1})*\&{sizeof}(%
\&{double}));{}$\6
\&{double} ${}{*}{*}\\{committee}\K\\{malloc}(\\{committee\_size}*{}$%
\&{sizeof}(\&{double} ${}{*}));{}$\7
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{committee\_size};{}$ ${}\PP\|i){}$%
\1\5
${}\\{committee}[\|i]\K\\{malloc}((\|d+\T{1})*\&{sizeof}(\&{double}));{}$\2\6
\&{if} ${}(\|d\I\T{2}){}$\5
${}\{{}$\1\6
\\{printf}(\.{"I\ cannot\ output\ Pos}\)\.{tScript\ unless\ d\ equ}\)\.{als\ 2!%
\\n"});\6
\&{return};\6
\4${}\}{}$\2\7
\&{FILE} ${}{*}\\{file}\K\\{fopen}(\\{filename},\39\.{"w"});{}$\7
\X13:Preamble\X;\6
\&{for} (\&{int} \\{page}${}\K\T{0};{}$ ${}\\{page}<\\{num\_pages};{}$ ${}\PP%
\\{page}){}$\5
${}\{{}$\1\6
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{committee\_size};{}$ ${}\PP\|i){}$%
\1\5
\\{random\_point}(\\{committee}[\|i]);\2\6
\\{random\_point}(\\{candidate});\6
\&{if} ${}(\\{page}\I\T{0}){}$\1\5
${}\\{fprintf}(\\{file},\39\.{"\%d\ \%d\ translate\\n"},\39\\{radius},\39%
\\{radius});{}$\2\6
${}\\{draw\_locus}(\\{file},\39\\{committee\_size},\39\\{committee},\39%
\\{candidate},\39\\{radius});{}$\6
${}\\{fprintf}(\\{file},\39\.{"showpage\\n"});{}$\6
\4${}\}{}$\2\6
\\{fclose}(\\{file});\6
\\{free}(\\{candidate});\6
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{committee\_size};{}$ ${}\PP\|i){}$%
\1\5
\\{free}(\\{committee}[\|i]);\2\6
\4${}\}{}$\2\par
\U20.\fi

\N{1}{24}Handling user input.
These components of the \PB{\\{main}} function deal with command-line input.
The program takes up to three command-line
arguments in the following order: \PB{\|p}, \PB{\|d}, and \PB{\\{max\_points}}.
If some or all of
these arguments are missing, we default to \PB{$\|p\K\T{2}$}, \PB{$\|d\K%
\T{2}$}, and \PB{$\\{max\_points}\K\T{300}$}.

\Y\B\4\X24:Gather input and initialise global variables\X${}\E{}$\6
$\|p\K\T{2.0};{}$\6
${}\|d\K\T{2};{}$\6
${}\\{max\_points}\K\T{300};{}$\6
${}\\{num\_points}\K\T{0};{}$\6
\&{if} ${}(\\{argc}\G\T{2}){}$\5
${}\{{}$\1\6
${}\|p\K\\{atof}(\\{argv}[\T{1}]);{}$\6
${}\\{p\_inv}\K\T{1.0}/\|p;{}$\6
\4${}\}{}$\2\6
\&{if} ${}(\\{argc}\G\T{3}){}$\5
${}\{{}$\1\6
${}\|d\K\\{atoi}(\\{argv}[\T{2}]);{}$\6
\&{if} ${}(\|d<\T{0}){}$\5
${}\{{}$\1\6
\\{printf}(\.{"I\ expect\ d\ to\ be\ >=}\)\.{\ 0.\\n"});\6
\&{return} \T{1};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
\&{if} ${}(\\{argc}\G\T{4}){}$\5
${}\{{}$\1\6
${}\\{max\_points}\K\\{atoi}(\\{argv}[\T{3}]);{}$\6
\&{if} ${}(\\{max\_points}<\T{10}){}$\5
${}\{{}$\1\6
\\{printf}(\.{"I\ expect\ max\_points}\)\.{\ to\ be\ >=\ 0.\\n"});\6
\&{return} \T{1};\6
\4${}\}{}$\2\6
\4${}\}{}$\2\6
${}\\{points}\K\\{malloc}(\\{max\_points}*{}$\&{sizeof}(\&{double} ${}{*}));{}$%
\6
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{max\_points};{}$ ${}\PP\|i){}$\1\5
${}\\{points}[\|i]\K\\{malloc}((\|d+\T{1})*\&{sizeof}(\&{double}));{}$\2\6
${}\\{printf}(\.{"Program\ started\ wit}\)\.{h\ p\ =\ \%f,\ d\ =\ \%d,\ an}\)%
\.{d\ max\_points\ =\ \%d\\n"},\39\|p,\39\|d,\39\\{max\_points}){}$;\par
\U2.\fi

\N{1}{25}Miscellaneous tests.
Here are collected some snippets that were used during testing phases.

\Y\B\4\X25:Sample mean test\X${}\E{}$\6
\&{double} \\{sum}${}\K\T{0.0};{}$\6
\&{int} \\{num\_samples}${}\K\T{0};{}$\7
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{num\_samples};{}$ ${}\PP\|i){}$\5
${}\{{}$\1\6
\&{double} \\{sample}${}\K\\{exponential1}(\,);{}$\7
${}\\{sum}\MRL{+{\K}}\\{sample};{}$\6
${}\\{printf}(\.{"\%f\\n"},\39\\{sample});{}$\6
\4${}\}{}$\2\6
${}\\{printf}(\.{"Sample\ mean:\ \%f\\n"},\39\\{sum}/\\{num\_samples}){}$;\par
\fi

\M{26}\B\X26:Generate random point cloud\X${}\E{}$\6
\&{double} ${}{*}\\{point}\K\\{malloc}((\|d+\T{1})*\&{sizeof}(\&{double}));{}$\7
\&{for} (\&{int} \|i${}\K\T{0};{}$ ${}\|i<\\{max\_points};{}$ ${}\PP\|i){}$\5
${}\{{}$\1\6
\\{random\_point}(\\{point});\6
\\{add\_point}(\\{point});\6
\4${}\}{}$\2\6
\\{free}(\\{point});\par
\fi

\M{27}\B\X27:Test $L^p$\X${}\E{}$\6
\&{int} \\{rounds}${}\K\\{max\_points}*\\{max\_points};{}$\7
${}\\{printf}(\.{"L\^p-distance\ voting}\)\.{\ for\ \%d\ rounds\ produ}\)\.{ced%
\ a\ committee\ with}\)\.{\ \%d\ members.\\n"},\39\\{rounds},\39\\{consensus}(%
\\{rounds},\39\\{p\_dist})){}$;\par
\fi

\M{28}\B\X28:Test orthogonal\X${}\E{}$\6
\&{int} \\{rounds}${}\K\\{max\_points};{}$\7
${}\\{printf}(\.{"Orthogonal\ voting\ f}\)\.{or\ \%d\ rounds\ produce}\)\.{d\ a%
\ committee\ with\ \%}\)\.{d\ members.\\n"},\39\\{rounds},\39\\{consensus}(%
\\{rounds},\39\\{abs\_inner\_product})){}$;\par
\fi

\M{29}\B\X29:Test close inner product\X${}\E{}$\6
\&{int} \\{rounds}${}\K\\{max\_points};{}$\7
${}\\{printf}(\.{"Close\ inner\ product}\)\.{\ voting\ for\ \%d\ round}\)\.{s\
produced\ a\ committ}\)\.{ee\ with\ \%d\ members.\\}\)\.{n"},\39\\{rounds},\39%
\\{consensus}(\\{rounds},\39\\{one\_minus\_inner\_product})){}$;\par
\fi

\N{1}{30}Index.
\fi

\inx
\fin
\con

### 2021-08-20 misc/literate/lp-balls.c

@@ -1,0 +1,437 @@
+/*2:*/
+#line 31 "./lp-balls.w"
+
+#include <float.h> 
+#include <math.h> 
+#include <stdio.h> 
+#include <stdlib.h> 
+#include <time.h> 
+
+
+
+double p;
+double p_inv;
+int d;
+int max_points;
+int num_points;
+double**points;
+
+/*12:*/
+#line 236 "./lp-balls.w"
+
+void plot_single_point(FILE*file,double red,double green,double blue,double x,double y){
+fprintf(file,"%f %f %f setrgbcolor %f %f dot\n",red,green,blue,x,y);
+}
+void to_postscript(const char*filename,int radius){
+if(d!=2){
+printf("I cannot output PostScript unless d equals 2!\n");
+return;
+}
+FILE*file= fopen(filename,"w");
+/*13:*/
+#line 255 "./lp-balls.w"
+
+char*preamble= "%!PS\n\
+/dot { 1.5 0 360 arc closepath fill } def\n\
+/square { /r exch def /y exch def /x exch def\n\
+newpath x r sub y r sub moveto 0 r 2 mul rlineto\n\
+r 2 mul 0 rlineto 0 r 2 mul neg rlineto\n\
+r 2 mul neg 0 rlineto closepath fill} def\n\
+0.5 setlinewidth\n";
+fprintf(file,"%s",preamble);
+fprintf(file,"%d %d translate\n",radius,radius);
+fprintf(file,"newpath 0 %d moveto 0 %d lineto ",-radius,radius);
+fprintf(file,"%d 0 moveto %d 0 lineto stroke\n",-radius,radius);
+
+/*:13*/
+#line 246 "./lp-balls.w"
+;
+/*14:*/
+#line 272 "./lp-balls.w"
+
+double colour1[]= {1.0,0.0,0.0};
+double colour2[]= {0.0,0.0,1.0};
+for(int i= 0;i<num_points;++i){
+double t= ((double)i)/((double)(num_points-1));
+double r= (1.0-t)*colour1[0]+t*colour2[0];
+double g= (1.0-t)*colour1[1]+t*colour2[1];
+double b= (1.0-t)*colour1[2]+t*colour2[2];
+plot_single_point(file,r,g,b,points[i][1]*radius,points[i][2]*radius);
+}
+
+/*:14*/
+#line 247 "./lp-balls.w"
+;
[…]

### 2021-09-15 about/vita.tex
Small changes here and there
Updated research page
Added various things

@@
 
+\newcount\papercount
+\papercount=1
+\def\papitem{\pubitem P\the\papercount. \global\advance\papercount by 1}
 \newcount\submittedcount
@@
 
+\sectheader{Papers}
+
+\pubbegin P
+\papitem (with Rosie Y. Zhao)
+Arithmetic subsequences in a random ordering of an additive set. {\sl arXiv preprint 2012.12339}, 11~pp.
+To appear in {\sl Integers: Electronic Journal of Combinatorial Number Theory}, 2021.
+
 \sectheader{Submitted Papers}
 
-\pubbegin S
-\subitem (with Anna M. Brandenberger and Luc Devroye)
+\begingroup\parindent=10pt
+\thing (with Anna M. Brandenberger and Luc Devroye)
 Root estimation in Galton--Watson trees. {\sl arXiv preprint 2007.05681}, 20~pp.
-Submitted July 2020.
-\subitem (with Rosie Y. Zhao)
-Arithmetic subsequences in a random ordering of an additive set. {\sl arXiv preprint 2012.12339}, 11~pp.
-Submitted December 2020.
-\subitem (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
+\thing (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
 Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl arXiv preprint 2105.12046}, 11~pp.
-Submitted May 2021.
-\subitem (with Jad Hamdan) The lattice of arithmetic progressions. {\sl arXiv preprint 2106.05949}, 8~pp.
-Submitted June 2021.
-\subitem (with Luc Devroye and Rosie Y. Zhao) The independence number of a Bienaym\'e--Galton--Watson tree
+\thing (with Jad Hamdan and Jonah Saks)
+The lattice of arithmetic progressions. {\sl arXiv preprint 2106.05949}, 8~pp.
+\thing (with Luc Devroye and Rosie Y. Zhao)
+The independence number of a Bienaym\'e--Galton--Watson tree
 and related parameters. {\sl arXiv preprint 2106.14389}, 20~pp.
-Submitted June 2021.
+\endgroup
 
@@
 trees, two on generating functions and elementary analytic combinatorics, and two on graph regularity.)
-[S1, S3, S5]
 
@@
 Work involved writing of code in OCaml and PostScript that generates illustrations of certain quotient
-spaces, in various models of hyperbolic geometry. [R3]
+spaces, in various models of hyperbolic geometry.
 \medbreak
@@
 of the interactive proof assistant Harpoon and wrote OCaml code as part of
-ongoing work on the functional programming language Beluga. [R2]
+ongoing work on the functional programming language Beluga.
 \medbreak
@@
 OPythn includes support for lists, dictionaries, named and anonymous functions, objects, and classes.
-It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing. [R1]
+It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing.
 
 \sectheader{Work and Volunteer Experience}
+
+\leftright{\bf Teaching assistant}{\bf September 2021 -- present}
+\leftright{\sl School of Computer Science, McGill University}{\sl Montr\'eal, QC, Canada}
+\smallskip
+Teaching assistant for COMP 690, a graduate-level course on the probabilistic analysis of algorithms.
+Hold office hours twice a week and responsible for grading of assignments.
+\medbreak
 
@@
 \parindent=10pt
-\thing Scored 10/120 on the William Lowell Putnam Mathematical Competition in 2019
-(the median score that year was 2/120).
-\smallskip
 \thing Recipient of 0x\$3.40 in Knuth reward cheques.
 \smallskip
-\thing Contributed sequences A335562, A338550, A338993, A339941, A339942, and A341822
+\thing Contributed sequences A335562, A338550, A338993, A339941, A339942, A341822, and A347580
 to the On-line Encyclopedia of Integer Sequences.

### 2021-10-18 about/vita.tex
New post and slight revamp
Typos
Added paper
RSA

@@
 
+\sectheader{Graduate Coursework}
+
+Higher algebra (categories, groups, modules), functional analysis, quadratic and modular forms, combinatorics.
+
 \sectheader{Research Interests}
@@
 \papitem (with Rosie Y. Zhao)
-Arithmetic subsequences in a random ordering of an additive set. {\sl arXiv preprint 2012.12339}, 11~pp.
-To appear in {\sl Integers: Electronic Journal of Combinatorial Number Theory}, 2021.
+Arithmetic subsequences in a random ordering of an additive set.
+{\sl Integers: Electronic Journal of Combinatorial Number Theory} {\bf 21} (2021), \#A89, 19~pp.
+[arXiv:2012.12339]
+\papitem (with Anna M. Brandenberger and Luc Devroye)
+Root estimation in Galton--Watson trees. To appear in {\sl Random Structures and Algorithms}, 20~pp.
+[arXiv 2007.05681]
 
@@
 \begingroup\parindent=10pt
-\thing (with Anna M. Brandenberger and Luc Devroye)
-Root estimation in Galton--Watson trees. {\sl arXiv preprint 2007.05681}, 20~pp.
+\smallskip
 \thing (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
 Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl arXiv preprint 2105.12046}, 11~pp.
+\smallskip
 \thing (with Jad Hamdan and Jonah Saks)
-The lattice of arithmetic progressions. {\sl arXiv preprint 2106.05949}, 8~pp.
+The lattice of arithmetic progressions. {\sl arXiv preprint 2106.05949}, 15~pp.
+\smallskip
 \thing (with Luc Devroye and Rosie Y. Zhao)
@@
 
+\goodbreak
 \sectheader{Reports}
@@
 university events such as frosh, sports games, and formals. Attend team training sessions twice a month to keep
-first-aid skills up-to-date. Most recent first responder certification: September 2021.
+first-aid skills up-to-date. Most recent first responder certification: September 2020.
 \medbreak

### 2021-11-15 about/vita.tex
Reverse chronological
Small edits to orthogonal groups

@@
 \papercount=1
-\def\papitem{\pubitem P\the\papercount. \global\advance\papercount by 1}
+\def\papitem{\pubitem P\the\papercount. \global\advance\papercount by -1}
 \newcount\submittedcount
 \submittedcount=1
-\def\subitem{\pubitem S\the\submittedcount. \global\advance\submittedcount by 1}
+\def\subitem{\pubitem S\the\submittedcount. \global\advance\submittedcount by -1}
 \newcount\reportcount
 \reportcount=1
-\def\repitem{\pubitem R\the\reportcount. \global\advance\reportcount by 1}
+\def\repitem{\pubitem R\the\reportcount. \global\advance\reportcount by -1}
 
@@
 \ifx\Umathchar\undefined\else% Chinese characters do not appear when using pdfTeX
-\font\chinese="STKaiti" at 12pt%
+\font\chinese="AR PL UKai HK" at 12pt%
 \font\twelverm=cmr12%
@@
 
-\sectheader{Graduate Coursework}
+\sectheader{Graduate-level Coursework}
 
-Higher algebra (categories, groups, modules), functional analysis, quadratic and modular forms, combinatorics.
+Higher algebra (categories, groups, modules), functional analysis, quadratic and modular forms, combinatorics,
+topology (including some algebraic topology), probabilistic analysis of algorithms.
 
@@
 
+\papercount=2
 \pubbegin P
+\papitem (with Anna M. Brandenberger and Luc Devroye)
+Root estimation in Galton--Watson trees. To appear in {\sl Random Structures and Algorithms}, 24~pp.
+[arXiv:2007.05681]
 \papitem (with Rosie Y. Zhao)
@@
 [arXiv:2012.12339]
-\papitem (with Anna M. Brandenberger and Luc Devroye)
-Root estimation in Galton--Watson trees. To appear in {\sl Random Structures and Algorithms}, 20~pp.
-[arXiv 2007.05681]
 
@@
 \begingroup\parindent=10pt
-\smallskip
-\thing (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
-Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl arXiv preprint 2105.12046}, 11~pp.
+\thing (with Luc Devroye and Rosie Y. Zhao)
+The independence number of a Bienaym\'e--Galton--Watson tree
+and related parameters. {\sl arXiv preprint 2106.14389}, 20~pp.
 \smallskip
@@
 \smallskip
-\thing (with Luc Devroye and Rosie Y. Zhao)
-The independence number of a Bienaym\'e--Galton--Watson tree
-and related parameters. {\sl arXiv preprint 2106.14389}, 20~pp.
+\thing (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
+Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl arXiv preprint 2105.12046}, 11~pp.
 \endgroup
@@
 
+\reportcount=4
 \pubbegin R
+\repitem Finding regularity in Tlingit verb prefixes. Semester project report, McGill University
+(Montr\'eal, Qu\'ebec, April 2021), 7 pp.
+\repitem Grid-building algorithms on manifolds. Summer research report, McGill University (Montr\'eal, Qu\'ebec,
+August 2020), 10 pp.
+\repitem Typechecking proof scripts: making interactive proof assistants robust.
+Honours project report, McGill University (Montr\'eal, Qu\'ebec, December 2019), 10 pp.
 \repitem The OPythn programming language. Software project report, Charles University (Prague,
 Czech Republic, June 2019), 10 pp.
-\repitem Typechecking proof scripts: making interactive proof assistants robust.
-Honours project report, McGill University (Montr\'eal, Qu\'ebec, December 2019), 10 pp.
-\repitem Grid-building algorithms on manifolds. Summer research report, McGill University (Montr\'eal, Qu\'ebec,
-August 2020), 10 pp.
-\repitem Finding regularity in Tlingit verb prefixes. Semester project report, McGill University
-(Montr\'eal, Qu\'ebec, April 2021), 7 pp.
 
@@
 Ongoing research on branching processes, headed by Luc Devroye. Studied estimation problems
-on Galton-Watson trees. Attended informal seminars over the year, discussing
+on Galton-Watson trees. Attended informal seminars, discussing
 various topics related to branching
-processes. (Gave a presentation at five of these: one on root estimation in Galton-Watson
-trees, two on generating functions and elementary analytic combinatorics, and two on graph regularity.)
+processes and other topics in probability and combinatorics.
+(Gave a presentation at six of these: one on root estimation in Galton-Watson
+trees, two on generating functions and elementary analytic combinatorics, two on graph regularity, and
+one concerning two of Erd\H os's proofs on prime numbers.)
 

### 2021-12-17 misc/expo/exotic.tex
Added more notes and video

\input fontmac
\input mathmac

\input epsf
\input color
\input soul.sty

\def\norm#1{|\!|#1|\!|}
\def\threenorm#1{|\!|\!|#1|\!|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\B{{\cal B}}
\def\supp{\op{supp}}

\widemargins
\bookheader{TSIRELSON'S SPACE}{MARCEL K. GOH}

\maketitle{Tsirelson's space\qquad and other exotic constructions}{}{Marcel K. Goh}{16 December 2021}

\floattext4.5 \ninebf Abstract.
\ninepoint
This set of expository notes was
written as a final report for the class MATH 567 Introduction to Functional Analysis,
taught by Prof.~Gantumur Tsogtgerel at McGill University in the Fall 2021 semester.
It builds up the necessary theory to describe Tsirelson's example of an infinite-dimensional
Banach space with no subspace isomorphic to $c_0$ or $l_p$ for $1\le p<\infty$, and
modifications thereof.

\advsect Preliminary notions

In this section we state definitions and well-known results that we shall require later on.
Proofs are not given for many of the facts given here; they can be found in introductory textbooks on
linear analysis (see, e.g.,~\ref{bollobas},~\ref{lindenstrauss},~\ref{schechter}).
The reader comfortable with the terminology of Banach space theory
may wish to skip this section, returning only upon finding an unfamiliar concept or definition.

Let $K$ denote either the real or complex field and let $V$ be a vector space over $K$. A
{\it norm} is a function $\norm{\cdot}:V\to\RR$ satisfying
\medskip
\item{i)} $\norm v \ge 0$ for all $v\in V$ and $\norm v = 0$ if and only if $v = 0$;
\smallskip
\item{ii)} $\norm{\alpha \cdot v} = |\alpha|\cdot\norm v$ for all $\alpha\in K$ and $v\in V$; and
\smallskip
\item{iii)} $\norm{v+w}\le \norm v + \norm w$ for all $v,w\in V$.
\medskip
A vector space equipped with such a function is called a {\it normed vector space}, and one can
define a metric $d$ on the space by setting $d(x,y) = \norm{x-y}$. Thus one has a notion of
Cauchy sequences in these spaces, and if a normed vector space $X$ is complete
(every Cauchy sequence in $X$ has a limit in $X$), we say that $X$ is a {\it Banach space}.
Any two norms on a finite-dimensional normed vector space are equivalent, in that they induce the
same topology. This fact can be used to show that any finite-dimensional normed vector space over
$\RR$ or $\CC$ is a Banach space, since these fields are complete.

The field $\RR$ or $\CC$ under the norm given by the absolute value $|\cdot|$ is of course the
simplest Banach space. Other basic examples include the following.
\medskip
\item{i)} Consider an $n$-dimensional vector space over $K$. Writing an element $x$ of this space
as $x = (x_1, \ldots, x_n)$, this vector space can be endowed with the {\it $p$-norm}
$$\norm{x}_p = \Bigl(\sum_{i=1}^n |x_i|^p\Bigr)^{1/p}.$$
for $1\le p<\infty$. This defines a Banach space which we shall denote by $l_p^n$. As $p\to\infty$,
this approaches the {\it maximum norm}
$$\norm{x}_\infty = \max_{1\le i\le n} |x_i|,$$
and we denote this Banach space by $l_\infty^n$.
\smallskip
\item{ii)} For $1\le p< \infty$, the $p$-norm for infinite sequences $x = (x_n)_{n=1}^\infty$ is given by
$$\norm{x}_{l_p} = \Bigl(\sum_{i=1}^\infty |x_i|^p \Bigr)^{1/p}.$$
As $p\to\infty$, this approaches the {\it supremum norm}
$$\norm{x}_{l_\infty} = \sup_{i\ge 1} |x_i|.$$
For $1\le p\le \infty$, the set $l_p$ of all infinite sequences $x$ with $\norm{x}{l_p}$ finite is a Banach space;
when $p=\infty$, these are the bounded sequences.
\smallskip
\item{iii)} Let $c_0$ be the normed vector subspace of $l_\infty$ consisting of sequences that tend to zero.
It can be shown that this subspace is closed and thus itself a Banach space.
\medskip
The latter two examples are infinite-dimensional, meaning that in these cases one can exhibit an
infinite sequence of vectors that are all linearly independent.

\medskip\boldlabel Hamel bases.
Assuming the axiom of choice, one can
find a basis for any vector space. This is a set of elements $\{e_i\}_{i\in I}$ such that any
$x\in X$ can be expressed uniquely as $\sum_{i\in F} a_ie_i$ for a finite set $F\subseteq I$
and scalars $a_i$, both depending on $x$.
When dealing with infinite-dimensional spaces, such a basis
is called a {\it Hamel basis} and are not awfully useful. To see why, we first need to establish (a variant of)
the Baire category theorem, as well as simple lemma.

\parenproclaim Theorem B (Baire category theorem). Let $X$ be a metric space that equals the union of
countably many closed sets. If $X$ is complete, then at least one of the closed sets has nonempty interior.\slug

\proclaim Lemma \advthm. Let $X$ be a normed vector space and let $Y$ be a subspace of $X$. If $Y$ has nonempty
interior, then $Y=X$.

\proof Suppose that for some $r>0$ and $y\in Y$, the ball $B_r(y) = \{x\in X : \norm{x-y} < r\}$ is contained
in $Y$. Now let $x\in X$ and note that the vector
$$z = y + {r\over 2}\cdot{x\over\norm x}$$
is at distance $r/2$ from $y$, meaning that $z\in B_r(y)\subseteq Y$. Since $Y$ is a subspace, we find that
$$x = {2\norm x\over r}(z-y)$$
is in $Y$, completing the proof.\slug

We now show that Hamel bases of Banach spaces cannot be countably infinite.

\proclaim Lemma \advthm. Let $X$ be an infinite-dimensional normed vector space. If $X$ is also complete
(and thus a Banach space), then any Hamel basis of $X$ is uncountable.

\proof Suppose that $\{e_1\}_{i=1}^\infty$ is a countable Hamel basis for $X$. For integers $n\ge 1$,
let $X_n$ denote the linear span of $\{e_1,\ldots,e_n\}$. Each normed vector space $X_n$ is finite-dimensional
and thus complete, which implies that each $X_n$ is closed in $X$. We now show that every $X_n$ has empty
interior, which completes the proof.\slug

Since many infinite-dimensional
Banach spaces have the cardinality of the continuum to begin with,
the above lemma tells us that Hamel bases can essentially be as complicated as the underlying space.
(In fact, it has been shown that
the cardinality of the Hamel basis of a Banach space always equals the cardinality of the space
itself~\ref{halbeisen2000}.)

\medskip\boldlabel Completions.
Consider now the infinite-dimensional vector space $c_{00}$ of sequences that are eventually zero. For all
integers $i\ge 1$, let $e_i$ denote the sequence that is $1$ at index $i$ and $0$ elsewhere. The set
$\{e_i\}_{i=1}^\infty$ is countable and clearly
a Hamel basis for $c_{00}$, so we conclude that $c_{00}$, under any norm, cannot be a Banach space. However,
for every normed vector space $V$, we can always find a Banach space $X$ such that $V$ is dense in $X$.
We do this by letting $X$ be the completion of $V$ as a metric space and then defining scalar multiplication
and addition as follows. If $x,y\in X$ are such that $x_n\to y$ and $y_n\to y$, then $\lambda x+\mu y$
shall simply be defined as $\lim_{n\to \infty} (\lambda x_n + \mu y_n)$. We leave it as an easy exercise
for the reader to check
that this makes $X$ a normed vector space under the norm $\norm x = \lim_{n\to\infty}\norm{x_n}$, and that
$X$ is complete with respect to this norm.

\medskip\boldlabel Dual spaces and the weak topology.
Let $X$ be a normed vector space.
A linear operator from $X$ to its base field is called a {\it linear functional},
and is continuous if and only if it is bounded.
The set of bounded linear functionals on a normed vector space $X$ is a Banach space under the operator norm
$$\norm{T} = \sup_{x\in X} {\norm {Tx}\over \norm{x}}.$$
This space is called the {\it dual space
of $X$} and is denoted $X^*$. If $(x_n)$ is a sequence of vectors in $X$ such that for some $x\in X$,
$x^*(x_n)\to x^*(x)$ for all $x^*\in X^*$, then we say that $(x_n)$ {\it converges weakly to $x$}.
If $(x_n)$ converges weakly to $0$, then we call $(x_n)$ {\it weakly null}. The topology that
this definition of convergence
induces on $X$ is called the {\it weak topology}; it is the coarsest topology such that every element
of $X^*$ is still a continuous function.
% When distinguishing the topology given by the norm from the
% weak one, we shall call the norm topology the {\it strong topology}.
% Since $x_n\to x$ implies that $x_n$ converges weakly
% to $x$, the weak topology is, in general, strictly coarser than the strong topology. However, there are
% many ``weak-to-strong'' principles in the theory of Banach spaces, the most celebrated of which is the following.
% 
% \parenproclaim Theorem U (Uniform boundedness principle). Let $X$ be a Banach space, $Y$ a normed vector space,
% and ${\cal T}$ a collection of linear operators from $X$ to $Y$. If
% $$\sup_{T\in{\cal T}} \bignorm{T(x)}_Y < \infty,$$
% then
% $$\sup_{T\in {\cal T}} \sup_{x\in X} {\bignorm{T(x)}_Y\over \norm{x}_X} <\infty.\noskipslug$$
% 
% The uniform boundedness principle can be used to show that any set of vectors
% that is bounded in the weak topology is also bounded in the strong topology.
% This fact will come in handy in a proof later on.

\advsect Schauder bases

We saw in the previous section that Hamel bases are not very useful for performing analysis on infinite-dimensional
normed vector spaces.
A better notion in the infinite-dimensional setting is that of a {\it Schauder basis}.
This is a countable sequence of vectors $(e_i)_{i=1}^\infty$
such that every $x\in X$ has a unique representation
$$x = \sum_{i=1}^\infty a_ie_i$$
for some sequence $(a_i)_{i=1}^\infty$ of scalars, where convergence of the infinite sum is defined in
terms of the metric induced by the norm. From here on out, we shall sometimes write ``basis'' to mean ``Schauder
basis'', for brevity's sake.
A basis $(e_i)$ for which there exists $C$ such that
$1/C \le \norm{e_i} \le C$ for all $i$ is called {\it seminormalised} and a seminormalised basis
whose corresponding constant $C$ equals $1$ is said to be {\it normalised}.
A sequence of vectors $(x_i)$
that is a basis for the closure of its linear span is called a {\it basic sequence}.

\medskip\boldlabel Equivalent bases.
A basis $(x_i)$ for a Banach space $X$ and a basis $(y_i)$ for a Banach space $Y$ are
said to be {\it equivalent} if the map $T$ sending $x_i$ to $y_i$ extends to a linear homeomorphism
between $X$ and $Y$. Of course, this relation is reflexive, symmetric, and transitive.
One can also show this to be the same as saying that there are constants $0<m\le M$ such that
for any sequence $(a_i)$ of scalars,
$$m\cdot\Norm{\sum_{i=1}^\infty a_ix_i}\le\Norm{\sum_{i=1}^\infty a_iy_i}\le
 M\cdot\Norm{\sum_{i=1}^\infty a_ix_i}.$$
To every Banach space $X$ with a basis $(e_i)_{i=1}^\infty$,
we can associate a sequence $(P_n)_{n=1}^\infty$ of bounded linear operators given by
$$P_n\Bigl(\sum_{i=1}^\infty a_ie_i\Bigr) = \sum_{i=1}^n a_ie_i.$$
The quantity $\sup_{n\ge 1} \norm{P_n}$ is finite and called the {\it basis constant} of $(e_i)$.
A basis $(e_i)$ with basis constant equal to $1$ is said to be {\it monotone}. Every Schauder basis
is monotone with respect to the norm $\threenorm{x} = \sup_n \norm{P_n x}$, and this norm happens to be
equivalent to the original one. Hence, given a basis, we can always pass to a norm under which the basis
is monotone.

\medskip\boldlabel Block basic sequences.
Let $(e_i)$ be a basic sequence in a Banach space. A sequence of vectors $(x_j)$, each of the form
$$x_j = \sum_{i=n_j+1}^{n_{j+1}} a_ie_i$$
where the $a_i$ are scalars and $n_1<n_2<\cdots$ is an increasing sequence of positive integers, is
called a {\it block basic sequence} or {\it block basis} of $(e_i)$.
Defining the {\it support} $\supp x$ of a vector
$ x = \sum_{i=1}^\infty a_i e_i$ with respect to a basis $(e_i)$
to be the set of indices $i$ for which $a_i$ is nonzero, a block basis is a sequence of vectors $x_j$ for
which $\supp x_j$ is finite for all $j$ and $\max \supp x_j < \min \supp x_{j+1}$ for all $j\ge 1$.

\medskip\boldlabel Unconditional bases.
Note that the definition of a Schauder basis may rely very much on the ordering
imposed on its elements. There are some bases for which this is not the case. The following conditions are
equivalent for a basis $(e_i)$ of a Banach space $X$.
\medskip
\item{i)} For every permutation $\pi$ of the positive integers, $\bigl(e_{\pi(i)}\bigr)_{i=1}^\infty$ is
a basis of $X$.
\smallskip
\item{ii)} If a sum of the form $\sum_{i=1}^\infty a_i e_i$ converges, then any reordering of its terms
produces a sum which converges to the same value.
\smallskip
\item{iii)} There exists a constant $C$ such that for all pairs of scalar sequences $(a_i)$ and $(b_i)$
with $|a_i|\le |b_i|$ for all $i$,
$$\Norm{\sum_{i=1}^\infty a_ie_i} \le C\Norm{\sum_{i=1}^\infty b_ie_i}.$$
\medskip
The proof of this equivalence can be found in Section~1.c of~\ref{lindenstrauss}. A basis satisfying any
(and thus all) of these properties is called an {\it unconditional basis}, and a basis is said to be
{\it $C$-unconditional} if it satisfies (iii) for some specific constant $C$.

\medskip\boldlabel Symmetric and subsymmetric bases. In the sequence spaces $c_0$ or $l_p$ for $1\le p<\infty$,
the canonical basis consists of vectors that have a $1$ at some index and zeroes everywhere else. Since
each coordinate is weighted equally in $l_\infty$ or $l_p$ norm, this basis is equivalent to any of its
permutations. Such a basic sequence in a Banach space is said to be {\it symmetric}. Note that
this is a bit stronger than the condition we saw above for
a basic sequence $(e_i)$ to be unconditional. In that case, we only needed $\bigl(e_{\pi(i)}\bigr)$ to be a
basic sequence for all permutations $\pi$.

An unconditional basis $(e_i)$ of a Banach space $X$ is called {\it subsymmetric} if for all $i_1<i_2<\cdots$,
the subsequence $\bigl(e_{i_k}\bigr)$ is equivalent to $(e_i)$. Note that this time, the requirement that
the basis be unconditional is not redundant. It can be shown, however, that every symmetric basis
is subsymmetric (see Prop.~3.a.3 of~\ref{lindenstrauss}); we shall use this fact in the next section.

Another fact that we'll need later regards a certain dichotomy in the family of subsymmetric basic sequences.

\newcount\dichotomy
\dichotomy=\thmcount
\proclaim Lemma \advthm. Let $(x_i)$ be a bounded subsymmetric basic sequence. Then $(x_i)$ is either equivalent
to the standard unit vector basis for $l_1$ or else $(x_i)$ is weakly null.

\proof If $(x_i)$ is weakly null, then we are done. If not, then we know that there exists some
$f\in X^*$ and $\delta>0$ such that $f(x_i) \ge \delta$ for all $i$. (If not, then $(x_i)$ would have a subsequence
that is weakly null, and thus by subsymmetry the sequence itself is weakly null.)
By scaling $f$ we can assume that it has
operator norm equal to $1$. For any $n$, and scalars $(b_1,\ldots,b_n)$ we have
$$\Norm{\sum_{i=1}^n |b_i| x_i} =\norm f\cdot \Norm{\sum_{i=1}^n |b_i| x_i} \ge f\Bigl(\sum_{i=1}^n |b_i|x_i\Bigr)
= \sum_{i=1}^n |b_i|f(x_i)\ge \delta \sum_{i=1}^n |b_i|.$$
Now let $C$ be the constant given by the unconditionality of $(x_i)$ and setting $a_i = |b_i|$, we have
$$\Norm{\sum_{i=1}^n |b_i|x_i} \le C \cdot\Norm{\sum_{i=1}^n b_ix_i}.$$
Putting everything together establishes
$$\Norm{\sum_{i=1}^n b_ix_i} \ge {1\over C} \Norm{\sum_{i=1}^n |b_i|x_i}\ge
{\delta\over C} \sum_{i=1}^n |b_i|$$
for all $n$, and taking $n$ to infinity gives a corresponding lower bound on all infinite sequences $(b_i)$.

The upper bound is a simple consequence of the triangle inequality. Let $M$ be the constant bounding $(x_i)$
from above. Then since $\norm{x_i}\le M$ for all $i$, for an arbitrary infinite sequence $(b_i)$ of scalars we have
$$\Norm{\sum_{i=1}^\infty b_ix_i}\le \sum_{i=1}^\infty |b_i|\cdot\norm{x_i}\le M \sum_{i=1}^\infty |b_i|,$$
which combines with the lower bound to give
$${\delta\over C} \sum_{i=1}^\infty |b_i| \le \Norm{\sum_{i=1}^\infty b_ix_i} \le M \sum_{i=1}^\infty |b_i|.$$
Hence $(x_i)$ is equivalent to the unit vector basis of $l_1$.
\slug

\advsect Tsirelson's space and its dual

For many decades, the infinite-dimensional
Banach spaces known to functional analysts all contained a subspace linearly homeomorphic
to $c_0$ or $l_p$ for $1\le p<\infty$. The first infinite-dimensional space without such a subspace was
constructed by B.~S.~Tsirelson in 1974~\ref{tsirelson},
and that same year T.~Figiel and W.~B.~Johnson gave a more explicit
characterisation of its dual~\ref{figieljohnson}.
In time, the dual of the original space has come to be known as $T$, with
Tsirelson's original space denoted $T^*$. Properties of Tsirelson's space and variations thereof are collected
in a book by P.~G.~Casazza and T.~J.~Shura~\ref{casazzashura}. We shall describe the space $T$ of
Figiel and Johnson in this section, following Casazza and Shura's exposition but supplying some of the proofs
that are omitted there.

Let $E$ and $E'$ be subsets of the positive integers. We write $E<F$ to mean that the maximum element of $E$
is less than the minimum element of $F$ and likewise we write $E\le F$ to mean the same thing, with
``less than'' replaced by ``at most''. A sequence of finitely many sets $E_1,E_2,\ldots,E_k$ of positive integers
is said to be {\it admissible} if
$$ \{k\} \le E_1 < E_2 < \cdots < E_k.$$
For $x\in c_{00}$ and a subset $E$ of positive integers, we write $Ex$ to denote the sequence that equals
$x_i$ at index $i$ for all $i\in E$ and $0$ elsewhere. For a fixed $x\in c_{00}$ there are only
finitely many $E$ such that $Ex$ is nonzero, since $x$ is finitely supported. Fixing a norm $\norm\cdot$
on $c_{00}$ now,
we say that
$$\sum_{i=1}^k \norm{E_i x}$$
is an {\it admissible sum for $x$} if $E_1,\ldots,E_k$ is admissible.

The following proposition establishes a sequence of norms on $c_{00}$ in terms of admissible sequences
of sets.

\newcount\inductive
\inductive=\thmcount
\proclaim Proposition \advthm.
For $x\in c_{00}$, we let $\norm{x}_0 = \norm{x}_{l_\infty}$
and for $n\ge 1$, we inductively set
$$\norm{x}_n = \max \biggl\{\norm{x}_{n-1}, {1\over 2}\max \sum_{i=1}^k \norm{E_i x}_{n-1}\biggr\},$$
where the inner maximum is taken over the finitely many admissible sequences $E_1, E_2, \ldots, E_k$ for which
the admissible sum for $x$ is nonzero. Then the following statements hold.
\medskip
\item{i)} For all $n\ge 0$, $\norm{\cdot}_n$ is a norm.
\smallskip
\item{ii)} For all $n\ge 0$ and $x\in c_{00}$, $\norm{x}_n\le \norm{x}_{l_1}$.
\smallskip
\item{iii)} The limit $\lim_{n\to \infty} \norm{x}_n$ exists and defines a norm on $c_{00}$.
\medskip

\proof The proof of (i) is inductive; the base case $n=0$ corresponds to $\norm\cdot _{l_\infty}$, which
we already know to be a norm. Now assume the statement proven for $n-1$. The only property that isn't quite
immediate is the triangle inequality. Note that for $x,y\in c_{00}$ nonzero,
if $E(x+y)$ is nonzero and $Ex=0$, then $E(x+y) = Ey$. So
$$\max^{(x+y)} \sum_{i=1}^k \norm{E_i (x+y)}_{n-1} \le
   \max^{(x)}\sum_{i=1}^k \norm{E_ix}_{n-1}+ \max^{(y)} \sum_{i=1}^k \norm{E_iy}_{n-1},$$
where $\max^{(x)}$ ranges over all admissible sequences whose corresponding admissible sum for $x$ is nonzero.
Then we have
$$\eqalign{
\norm{x+y}_n &= \max\biggl\{\norm{x+y}_{n-1}, {1\over 2}\max^{(x+y)}\sum_{i=1}^k \norm{E_i (x+y)}_{n-1}\biggr\}\cr
&\le \max\biggl\{\norm{x}_{n-1} +\norm{y}_{n-1},
{1\over 2}\max^{(x+y)} \sum_{i=1}^k \bigl(\norm{E_i x}_{n-1} + \norm{E_i y}_{n-1}\bigr)\biggr\} \cr
&\le \max\biggl\{\norm{x}_{n-1}, {1\over 2}\max^{(x)}\sum_{i=1}^k \norm{E_i x}_{n-1}\biggr\} \cr
&\qquad\qquad+\max\biggl\{\norm{y}_{n-1}, {1\over 2}\max^{(y)}\sum_{i=1}^k \norm{E_i y}_{n-1}\biggr\}\cr
&= \norm{x}_n + \norm{y}_n,\cr
}$$
by linearity of the maximum function.

The claim (ii) is rather obvious, and again the proof is by induction, with the base case $\norm{x}_{l_\infty}
\le \norm{x}_{l_1}$. Now let $n\ge 1$ and suppose the statement proven for $n-1$. If $\norm{x}_n = \norm{x}_{n-1}$
we are done, and if not, then
$$\norm{x}_n = {1\over 2}\max^{(x+y)}\sum_{i=1}^k \norm{E_i (x+y)}_{n-1}
\le {1\over 2}\max^{(x+y)}\sum_{i=1}^k \norm{E_i (x+y)}_{l_1}$$
and since any admissible sequence consists of disjoint sets, each coordinate $x_i$ in $x$ contributes $|x_i|$
to the sum on the right-hand side at most once. Hence in this case $\norm{x}_n \le \norm{x}_{l_1}$ as well.

Since the norms $\norm{x}_n$ are nondecreasing in $n$ for any fixed $x$ and bounded above by $\norm{x}_{l_1}$,
this value must converge to a limit as $n\to\infty$. The fact that $\norm{x} = \lim_{n\to\infty} \norm{x}_n$
is a norm is easy to see (the triangle inequality follows from linearity of the limit). We have thus proven
(iii).\slug

The above proposition asserts that $c_{00}$ endowed with the norm $\norm\cdot$ is a normed vector space. We
then define Tsirelson's space $T$ to be the completion of $c_{00}$ with respect to this norm. For each index
$i$, let $t_i$ be the vector whose $i$th coordinate is $1$ and whose other coordinates are all $0$. The
sequence $(t_i)$ forms a normalised $1$-unconditional Schauder basis for $T$. This can be proven by inductively
verifying that it holds for the completion of $c_{00}$ with respect to $\norm\cdot _n$ for all $n$.

We also have the identity
$$\norm{x} = \max\biggl\{ \norm{x}_{l_\infty}, {1\over 2}\sup \sum_{i=1}^k \norm{E_i x}\biggr\},$$
in which the supremum is taken over all $k$ and all possible admissible sequences $E_1,\ldots,E_k$.
This could actually have been taken as the {\it definition} of the Tsirelson norm (it is the
unique norm with this property), which
might appear to be nonsensical, because the norm appears on both sides of the definition.
However, it is actually perfectly reasonable, since the $E_i x$ on the right hand side have
$|\supp E_i x| < |\supp x|$ and we can assume by induction that the norm is already defined for these vectors.
The base cases are the vectors that have only one nonzero coordinate, for which we must take the $l_\infty$-norm.
In any case, let us prove that this definition is equivalent to the norm that we constructed above.

\newcount\recursive
\recursive=\thmcount
\proclaim Proposition \advthm. For any $x\in T$, we have
$$\norm{x} = \max\biggl\{ \norm{x}_{l_\infty}, {1\over 2}\sup \sum_{i=1}^k \norm{E_i x}\biggr\},$$
where the supremum is taken over all $k$ and all admissible sequences $E_1,\ldots,E_k$.

\proof Since $c_{00}$ is dense in $T$, it suffices to prove the proposition for $x\in c_{00}$. If
$\norm x = \norm{x}_{l_\infty}$, then for every $n\in \NN$ and admissible sequence $E_1,\ldots,E_k$, we
have
$$ {1\over 2}\sum_{i=1}^k \norm{E_i x}_n \le \norm{x}_n.$$
Then by linearity of the supremum, we have
$${1\over 2}\sum_{i=1}^k \norm{E_i x} \le {1\over 2}\sum_{i=1}^k \sup_{n\ge 0} \norm{E_i x}_n
= \sup_{n\ge 0}{1\over 2}\sum_{i=1}^k \norm{E_i x}_n \le \sup_{n\ge 0} \norm{x}_n = \norm{x},$$
which shows that the identity holds in this case. If $\norm x\ne \norm_{l_\infty}$, then for all $\eps >0$
there is $n\ge 1$ such that $\norm{x}< \norm{x}_n + \eps$ and $\norm{x}_n > \norm{x}_{n-1}$. This means
there is some $k'$ and $E_1,\ldots,E_{k'}$ such that
$$ \norm {x}_n = {1\over 2} \sum_{i=1}^{k'} \norm{E_i x}_{n-1}.$$
It follows that
$$\eqalign{
\norm x - \eps &< \norm{x}_n = {1\over 2} \sum_{i=1}^{k'} \norm{E_i x}_{n-1}
\le {1\over 2} \sum_{i=1}^{k'} \norm{E_i x}
\le \sup{1\over 2} \sum_{i=1}^{k} \norm{E_i x}.
}$$
For an upper bound on the supremum, we pick $k''$ and an admissible sequence $E_1,\ldots, E_{k''}$ such that
$$\sup{1\over 2} \sum_{i=1}^{k} \norm{E_i x} < {1\over 2} \sum_{i=1}^{k''} \norm{E_i x} + \eps.$$
We then pick $n'$ such that
$$\sup{n\ge 1}{1\over 2} \sum_{i=1}^{k''} \norm{E_i x}_n <
{1\over 2} \sum_{i=1}^{k''} \norm{E_i x}_{n'} + \eps.$$
This yields the upper bound
$$\eqalign{
{1\over 2} \sum_{i=1}^{k} \norm{E_i x} &< {1\over 2} \sum_{i=1}^{k''} \norm{E_i x} + \eps \cr
&= {1\over 2} \sum_{i=1}^{k''} \sup_{n\ge 1} \norm{E_i x}_n + \eps \cr
&= \sup_{n\ge 1} {1\over 2} \sum_{i=1}^{k''} \norm{E_i x}_n + \eps \cr
&< {1\over 2} \sum_{i=1}^{k''} \norm{E_i x}_{n'} + 2\eps \cr
&\le \norm{x}_{n'+1} + 2\eps\cr
&\le \norm{x} + 2\eps.\cr
}$$
We have shown that
$$\norm{x}-\eps < \sup{1\over 2} \sum_{i=1}^{k} \norm{E_i x} < \norm{x} + 2\eps,$$
which, since $\eps$ was arbitrary, gives us equality.\slug

This recursive identity is the only definition one really uses when performing any serious computations
with Tsirelson's norm. In particular, it asserts that we have,
for any $x\in T$ and admissible sequence $E_1,\ldots,E_k$,
$$\sum_{i=1}^k \norm{E_i x} \le 2 \norm x.$$
As a demonstration of the utility of the identity, consider the proof of the
following proposition, which gives us an idea of
how Tsirelson's norm behaves on subsets of a seminormalised block basis.

\newcount\tsirelsonblock
\tsirelsonblock=\thmcount
\proclaim Proposition \advthm. Let $(t_i)$ denote the usual unit vector basis of Tsirelson's space
and let $k$ be a positive integer and let $M\ge 1$. For any $k$ blocks $(y_j)_{j=1}^k$,
where each $y_j$ satisfies $1/M \le \norm{y_j}\le M$ and is of the form
$$y_j = \sum_{i=n_j+1}^{n_{j+1}} a_i t_i$$
for some scalars $a_i$ and $k-1\le n_1<n_2 <\cdots < n_{k+1}$, we have
$${1\over 2M} \sum_{j=1}^k |b_j| \le \Norm{\sum_{j=1}^k b_jy_j} \le M\sum_{j=1}^k |b_j|$$
for all $k$-tuples of scalars $(b_1,\ldots, b_k)$.

\proof By the triangle inequality and the fact that $\norm{y_j} \le M$ for all $j$, we have
$$\Norm{\sum_{j=1}^k b_jy_j} \le \sum_{j=1}^k \norm{b_jy_j} \le  M\sum_{j=1}^k |b_j|.$$
For the lower bound, we let $x = \sum_{j=1}^k b_jy_j$ and let
$$E_j = \{n_j+1, n_j+2, \ldots, n_{j+1}-1\}$$
for $1\le j\le k$. Then
$$\norm x \ge {1\over 2}\sum_{j=1}^k \norm{E_j x} = {1\over 2}\sum_{j=1}^k \norm{b_j y_j}
  \ge {1\over 2M}\sum_{j=1}^k |b_j|.\noskipslug$$

We are almost ready to prove that $T$ does not contain a copy of $l_1$. We just need one more lemma,
which is well-known and due to R.~C.~James~\ref{james1964}.

\parenproclaim Lemma J (James, {\rm 1964}). If a normed linear space contains a subspace isomorphic
to $l_1$, then for any $\delta>0$, there is a normalised block basic sequence $(u_j)_{j=0}^\infty$ such that
$$(1-\delta)\sum_{j=0} |a_j| \le \Norm{\sum_{j=0}^\infty a_ju_j} \le \sum_{j=0}^\infty |a_j|$$
for every sequence $(a_j)_{j=0}^\infty$ of scalars such that at least one $a_j$ is nonzero.

\proof The statement holds trivially when $\delta\ge 1$, so let $0<\delta<1$.
The subspace $B$ isomorphic to $l_1$ contains a sequence $(x_i)_{i=0}^\infty$ for which there
exist constants $m$ and $M$ such that
$$ m\sum_{i=0}^\infty |a_i| \le \Norm{\sum_{i=0}^\infty a_i x_i} \le M\sum_{i=0}^\infty |a_i|$$
for all sequences $(a_i)_{i=0}^\infty$ of scalars. For $n\ge 1$, let
$$ K_n = \inf \biggl\{ \Norm{ \sum_{i=n}^\infty a_ix_i } : \sum_{i=0}^\infty |a_i| = 1\biggr\} $$
and set $K$ to $\lim_{n\to \infty} K_n$. We of course have $m\le K\le M$. Let $\phi >1$ and $0<\theta<1$
be such that $1-\delta \le \theta/\phi$. Choose $n_0<n_1<n_2$ such that $K_{n_0+1} \ge \theta K$ and for which
there is a block
basic sequence $(y_j)_{j=0}^\infty$ with each $y_j$ of the form
$$ y_j = \sum_{j=n_j+1}^{n_{j+1}} a_i x_i $$
for some scalars $a_i$ (depending on $j$) and with $\norm{y_j} \le \phi K$ for all $j$.
Let $(a_j)_{j=0}^\infty$ be an arbitrary sequence of scalars.
The normalised
block basis consisting of $u_j = y_j/\norm{y_j}$ satisfies
$$\Norm{\sum_{j=0}^\infty a_ju_j} \le \sum_{j=0}^\infty |a_j|$$
by the triangle inequality. By the construction of the block basis,
each $y_j$ has $\supp y_j > \{n_0\}$, so by the definition of $K_{n_0+1}$,
$$\Norm{\sum_{j=0}^\infty a_jy_j} = \Norm{\sum_{j=n_0+1}^\infty a_jy_j} \ge K_{n_0+1} \sum_{j=0}^\infty |a_j|
\ge \theta K \sum_{j=0}^\infty |a_j|.$$
Then since $\norm{y_j} \le \phi K$ for all $K$ and $\theta/\phi \ge 1-\delta$, we have
$$\Norm{\sum_{j=0}^\infty a_j u_j} = \normm{\sum_{j=0}^\infty a_j {y_j\over \norm{y_j}}}
\ge {1\over \phi K} \Norm{\sum_{j=0}^\infty a_j y_j} \ge (1-\delta)\sum_{j=0}^\infty |a_j|,$$
which proves the lower bound.\slug

Note that in James' original statement of the lemma, the $u_j$ are just vectors contained in the closed unit
ball. However, the construction in his proof actually produces a block basic sequence, and we
normalised it in our own version of the proof above.
We are now set to show that $l_1$ does not embed into $T$.

\newcount\nolone
\nolone=\thmcount
\proclaim Theorem \advthm. Tsirelson's space $T$ does not contain a subspace isomorphic to $l_1$.

\proof Suppose, towards a contradiction, that $T$ did contain a subspace isomorphic to $l_1$. Then, applying
Lemma~J with $\delta = 1/9$, there exists a normalised block basic sequence $(y_j)_{j=0}^\infty$
such that for all sequences $(a_j)_{j=0}^\infty$ of scalars,
$${8\over 9} \sum_{j=0}^\infty |a_j| \le \Norm{\sum_{j=0}^\infty a_jy_j} \le\sum_{j=0}^\infty |a_j|.$$
In particular, if we let $r\ge 1$ be an integer and let $(a_j)_{j=0}^\infty$ be the sequence with
$$a_j = \cases{ 1,& if $j=0$;\cr 1/r,& if $1\le j\le r$;\cr 0,& if $j>r$,}$$
then we have $\sum_{j=0}^\infty |a_j| = 2$. This yields
$$\Norm{ y_0 + {1\over r}\sum_{j=1}^r y_j} = \Norm{\sum_{j=0}^\infty a_jy_j} \ge
{8\over 9}\sum_{j=0}^\infty |a_j| = {16\over 9}$$
for all integers $r\ge 1$.

Now let $\{k\}\le E_1 < E_2 < \cdots < E_k$ be an arbitrary admissible sequence of sets, and let
$r_0 =\max\supp y_0$. If $k > r_0$, then
$$\sum_{i=1}^k \normm{ E_i \Bigl( y_0 + {1\over r}\sum_{j=1}^r y_j\Bigr)}
=\sum_{i=1}^k \normm{ E_i \Bigl({1\over r}\sum_{j=1}^r y_j\Bigr)} \le 2\cdot\Norm{{1\over r}\sum_{j=1}^r y_j}
\le 2.$$
If $k\le r_0$, then we set
$$S = \bigl\{1\le j\le r : \norm{E_i y_j}\ne 0\ \hbox{for at least two values of}\ i\bigr\}$$
and
$$T = \bigl\{1\le j\le r: \norm{E_i y_j}\ne 0\ \hbox{for at most one value of}\ i\bigr\}.$$
Note that if $j\in S$, then $y_j$ straddles the border between $E_i$ and $E_{i+1}$ for some $i$, and
no other $y_{j'}$ can do so, since the supports of the $y_j$ are disjoint. So the cardinality of $S$ is
at most $k-1$. It is also clear that $S$ and $T$ are disjoint and $|S|+|T| = r$, so
$$2|S| + |T| \le 2(k-1) + r- k+1 = r+k-1.$$
Since the $y_j$ are unit vectors, we have
$$\eqalign{
\sum_{i=1}^k \normm{E_i \Bigl({1\over r}\sum_{j=1}^\infty y_i\Bigr)} &=
\sum_{j\in S} \sum_{i=1}^k \norm{E_iy_j} + \sum_{j\in T} \sum_{i=1}^k \norm{E_iy_j}\cr
&\le  \sum_{j\in S} 2\cdot \norm{y_j} + \sum_{j\in T} \norm{y_j},\cr
&\le r+k-1\cr
}$$
by rearranging sums.
Dividing by $r$ and adding the unit vector $y_0$ into the mix, we have
$$\eqalign{
\sum_{i=1}^k \normm{ E_i \Bigl( y_0 + {1\over r} \sum_{j=1}^r y_j\Bigr)}
  &\le\sum_{i=1}^k \norm{E_i y_0} + {1\over r}\sum_{i=1}^k\normm{E_i\Bigl({1\over r}\sum_{j=1}^\infty y_i\Bigr)}\cr
&\le 2 + {r+k-1\over r} \cr
&\le 3 + {n_0\over r}.\cr
}$$
Selecting some $r\ge 2n_0$, the right-hand side is at most $7/2$, meaning that
$$\Norm{y_0 + {1\over r} \sum_{j=1}^r y_j} \le 7/4,$$
and we have $7/4 \ge 16/9$, the contradiction we sought.\slug

In the proof that $c_0$ and the other $l_p$ spaces do not embed into $T$ either, we shall make use of
the following theorem, a version of what is called the Bessaga--Pe\l czy\'nski Selection Principle
(see~\ref{bessagapelczynski}, Theorem 3).

\parenproclaim Theorem S (Bessaga--Pe\l czy\'nski, {\rm 1958}).
Let $X$ be a Banach space with basis $(e_i)$ and let $(e_i^*)$ be the coefficient
functionals given by
$$ e_j^*\Big(\sum_{i=1}^\infty a_i e_i\Bigr) = a_j.$$
If $(y_i)_{i=1}^\infty$ is a sequence of vectors in $X$ such that
\medskip
\item{i)} $\lim_{i\to\infty} \norm{y_i} > 0$; and
\smallskip
\item{ii)} $\lim_{i\to\infty} e_j^*(y_i) = 0$ for all indices $j\ge 1$,
\medskip\noindent
then there exists a subsequence of $(y_i)$ which is equivalent to a block basis with respect to the original
basis $(e_i)$.\slug

To prove that $c_0$ and $l_p$ do not embed into $T$, we shall actually prove the following stronger result.

\proclaim Theorem \advthm. Tsirelson's space $T$ does not contain a seminormalised subsymmetric basic
sequence.

\proof Towards a contradiction, suppose that $(y_i)$ is a seminormalised subsymmetric basic sequence in $T$.
Since the sequence is seminormalised, it is bounded from above, so by Lemma~{\the\dichotomy}, either $(y_i)$
is equivalent to the canonical basis for $l_1$ or $(y_i)$ is weakly null. We already proved Theorem~{\the\nolone}
above, which deems the first scenario impossible, so $(y_i)$ must be weakly null and thus in particular
satisfies condition (ii) of Theorem S. In addition, since $(y_i)$ is seminormalised, it is uniformly bounded
away from zero, so we are in the position to conclude from Theorem S that there is a subsequence of $(y_i)$
which is equivalent to a block basic sequence $(x_i)$ against the unit vector basis $(t_i)$ of $T$.
We shall now pass to subsequences both in $(x_i)$
and in $(y_i)$. First, pass to a subsequence and reindex $(x_i)$ to ensure that $\supp x_i > \{i\}$ for all $i$.
Now we pass to a subsequence of $(y_i)$ that is equivalent to this new $x_i$ and reindex; this is equivalent
to the original sequence by subsymmetry.

Since $y_i$ is seminormalised and there is a linear homeomorphism mapping each $y_i\mapsto x_i$,
$(x_i)$ is seminormalised as well; that is, there is
$M\ge 1$ such that $1/M \le \norm{x_i}\le M$.
Now since $\supp x_i > \{i\}$ for all $i$,
we can apply Proposition~{\the\tsirelsonblock} with the constant $M$ on any $n$ blocks
$x_{i_1}, x_{i_2}, \ldots, x_{i_n}$ of $(x_i)$
to find that for all $n$-tuples $(b_1,\ldots, b_n)$ of scalars,
$${1\over 2M}\sum_{k=1}^n |b_k| \le \Norm{\sum_{k=1}^n b_kx_{i_k}} \le M\sum_{k=1}^n |b_k|.$$
We thus find that for any $n$, $(x_i)$ (and hence $(y_i)$)
contains a subsequence of length $n$ which is equivalent to the standard basis of $l_1^n$. Complete
each of these finite sequences to infinite subsequences. Recall that $(y_i)$ is equivalent to all of them,
so for all $n$, the first $n$ elements of $(y_i)$ are equivalent to the standard basis of $l_1^n$. Thus
$(y_i)$ is equivalent to the standard basis of $l_1$ and we know this cannot happen.\slug

\proclaim Corollary \advthm. Tsirelson's space $T$ does not contain $c_0$ or $l_p$ for $1<p<\infty$.

\proof The unit vector bases of these spaces are normalised (hence seminormalised) subsymmetric
basic sequences.\slug

\advsect Banach spaces inspired by Tsirelson's space

Tsirelson's example marked a bit of a turning point in the theory of Banach spaces. It is important not
only because it settled a long-standing open problem, but because the construction that underlied it turned
out to be modifiable in various ways to tackle other problems. In this section we shall give a brief survey
of some ``Tsirelson-inspired'' Banach spaces. The aim of this section is breadth rather than rigour,
so we shall not give proofs as we did in the previous section for Tsirelson's space. Instead, we are
more interested in the constructions themselves, how they build upon each other, and the properties
they satisfy.

\medskip\boldlabel Schlumprecht's arbitrarily distortable space. Let $X$ be an infinite-dimen\-sional Banach
space under the norm $\norm\cdot$
For $\lambda > 1$, we say that $X$ is {\it $\lambda$-distortable} if there exists an equivalent norm
$\threenorm\cdot$ on $X$ such that for each infinite-dimensional subspace $Y$ of $X$, we have
$$\sup\biggl\{{\threenorm{y_1}\over\threenorm{y_2}}:y_1,y_2\in Y,\,\norm{y_1} = \norm{y_2} = 1\biggr\}\ge\lambda.$$
The norm $\threenorm\cdot$ is called a {\it $\lambda$-distortion} if this is true.
The space $X$ is called {\it distortable} if it is $\lambda$-distortable for some $\lambda>1$, and
{\it arbitrarily distortable} if it is distortable for every $\lambda>1$. In 1991, T.~Schlumprecht
proved the existence of an arbitrarily distortable Banach space~\ref{schlumprecht1991}.

Schlumprecht's construction relies on a function $f : [1,\infty)\to [1,\infty)$ satisfying
\medskip
\item{i)} $f(1) = 1$ and $f(x)<x$ for all $x>1$;
\smallskip
\item{ii)} $f$ is strictly increasing to infinity;
\smallskip
\item{iii)} $lim_{x\to\infty} f(x)/x^q = 0$ for all $q>0$;
\smallskip
\item{iv)} the function $g(x) = x/f(x)$ is concave for $x\ge 1$; and
\smallskip
\item{v)} $f(x)f(y) \ge f(xy)$ for all $x,y\ge 1$.
\medskip
Of course, we need such a function to actually exist, but it is easily seen that $f(x) = \log_2(x+1)$ does.
We then define a sequence $\threenorm{\cdot}_n$ of norms on $c_{00}$ by setting
$\threenorm{x}_0 = \norm{x}_{l_\infty}$
and for $n\ge 1$ we put
$$\threenorm{x}_{n} = \max_{k\ge 2} \max {1\over f(k)} \sum_{i=1}^k \threenorm{E_i x}_{n-1},$$
where the inner maximum is over all subsets $E_1<E_2<\cdots <E_k$ of positive integers (unlike in
Tsirelson's example, we do not need ${k}\le E_1$ this time) for which $E_i x$ is nonzero for some $1\le i\le k$.
Schlumprecht's space $X$ is then defined to be the completion of $c_{00}$ with respect to the limit
$\norm\cdot$ of these norms.

Let $(e_i)$ denote the sequence of vectors in $c_{00}$, where $e_i$ has a $1$ at the $i$th coordinate and
zeroes elsewhere. It is an unconditional basis for $X$ just as it was for $T$. But whereas $T$ contained no
seminormalised subsymmetric basic sequences, $(e_i)$ is actually a normalised subsymmetric basis of $X$.
Just as in Tsirelson's example, one can show that the norm on $X$ satisfies the
recursive identity
$$\norm x = \max\Bigl\{\norm{x}_{l_\infty}, \sup_{k\ge 2} \sup {1\over f(k)} \sum_{i=1}^k \norm{E_i x}\Bigr\},$$
where the inner supremum runs over all subsets $E_1 < E_2<\cdots <E_k$ of positive integers.
Note that $X$ also does not contain $c_0$ or $l_p$ for $1\le p<\infty$,
and the fact that $X$ is arbitrarily
distortable is established by the following theorem.

\proclaim Theorem D. Let $X$ denote Schlumprecht's space and $\norm\cdot$ its norm. For any positive integer
$k$ and $x\in X$, let
$$\norm{x}_k = \sup {1\over f(k)} \sum_{i=1}^k \norm{E_i x}.$$
For every $k$, every $\eps>0$, and every infinite-dimensional subspace $Y$ of $X$,
there are unit vectors $y_1, y_2\in Y$ such that
$$\norm{y_1}_k \ge 1-\eps\qquad\hbox{\rm and}\qquad \norm{y_2}_k \le {1+\eps\over f(k)}.$$
In particular, this means that $\norm\cdot _k$ is an $f(k)$-distortion on $X$.\slug

\medbreak
\boldlabel The unconditional basic sequence problem.
In Section~2 we defined what it means for a basis or a basic sequence to be unconditional.
Both Tsirelson's space and Schlumprecht's space have unconditional bases, and so do all classical Banach spaces.
It was unknown for many years whether every infinite-dimensional Banach space contains an unconditional basic
sequence. Soon after Schlumprecht presented his example in the summer of 1991,
W.~T.~Gowers and B.~Maurey independently constructed
Banach spaces without an unconditional basic sequences. They subsequently discovered that both examples
were fundamentally the same, and ended up publishing jointly~\ref{gowersmaurey1993}.

Recall that we say that $X$ is the {\it direct sum} of $Y$ and $Z$ and write $X = Y\oplus Z$ if
the map $Y\times Z\to X$ sending $(y,z)\mapsto y+z$ is a linear homeomorphism. An infinite-dimensional
Banach space $X$ is said to be {\it decomposable} if $X$ can be written as $X = Y\oplus Z$ where
both $Y$ and $Z$ are also infinite-diimensional. An infinite-dimensional
Banach space $X$ without any decomposable subspaces is called
{\it hereditarily indecomposable}. If a space $X$ has an unconditional basic sequence $(x_i)_{i=1}^\infty$, then
letting $Y$ be the closed linear span of $(x_{2i-1})_{i=1}^\infty$ and $Z$ be the closed linear span
of $(x_{2i})_{i=1}^\infty$, we see that $X$ is not hereditarily indecomposable. So to produce an
infinite-dimensional space without an unconditional basic sequence, it suffices to produce a hereditarily
indecomposable one, and that is what Gowers and Maurey did.

Before describing their construction, we need to establish a fair bit of terminology and notation.
Let $J$ be
a set of positive integers such that if $m,n\in J$ with $m<n$, then $\log\log\log n \ge 4m^2$. We also
assume that the smallest element of $J$ is greater than $256$. We write
$J$ in increasing order as $\{j_1, j_2,\ldots\}$, let $K$ be the subset of elements with odd index, and let $L$
be the subset of elements with even index. We recall here that a sequence $y=(y_i)\in l_1$
defines a linear functional $f_y$ on $c_{00}$ with $\norm{f_y} = \norm{y}_{l_1}$ given by
$$f_y\bigl((x_i)\bigr) = \sum_{i=1}^k x_i y_i.$$
Any bounded linear functional $f$ on $c_{00}$ can be associated to an element of $l_1$ by evaluating $f$
at each of the standard basis elements of $c_{00}$, so the topological dual of $c_{00}$ is $l_1$.

We say that a sequence $x_1,x_2,\ldots,x_n$ of scalar sequences is {\it successive} and write
$x_1<x_2<\ldots<x_n$ if $\supp x_i < \supp x_{i+1}$ for all $1\le i<n$.
Let $f = \log_2(1+x)$, as in Schlumprecht's example. Let $Q\subseteq$ denote
the set of finitely-supported sequences whose coordinates are all rational numbers in the range $[-1,1]$.
Let $\sigma$ be an injection from the set of finite successive sequences of elements of $Q$ to $L$ such
that for all such sequences $x_1, \ldots, x_s$ is such a sequence, $S = \sigma(x_1,\ldots,x_s)$, and
$x = x_1 + \cdots + x_s$, then
$$\left|\supp z\right| \le {\sqrt{f\bigl(S^{1/40}\bigr)}\over 20}.$$
This injection is used to define functionals on any normed space with $c_{00}$ as the underlying set.
Let $X=(c_{00}, \norm\cdot)$ be such a normed space. For $m\in \NN$, we let $A^*_m(X)$ be the set of functionals
of the form
$${1\over f(m)} \sum_{i=1}^m x_i^*,$$
where the $x_i^*$ are functionals with $\norm{x_i^*}\le 1$ for all $i$ and $x_1^* < \cdots < x_m^*$
(as sequences). Note that members in $A^*_m$ have operator norm at most $1$ for any $m$. For $k\in \NN$,
let $\Gamma_k^X$ be the set of sequences $(g_1,\ldots, g_k)$ such that $g_i\in Q$ for all $i$,
$g_1\in A^*_{j_{2k}}(X)$, and $g_{i+1}\in A^*_{\sigma(g_1,\ldots,g_i}$ for all $1\le i<k$. We call the
elements of $\Gamma_k^X$ {\it special sequences}. Now let $B_k^*(X)$ be the set of functionals of the form
$$ {1\over \sqrt{f(k)}} \sum_{j=1}^k g_j $$
such that $(g_1, \ldots, g_k)$ is a special sequence. An element of $B_k^*(X)$ is called a {\it special
functional}.

The space without an unconditional basic sequence that Gowers and Maurey exhibit is also constructed via
a sequence of norms, as Tsirelson's and Schlumprecht's spaces were. Let $\norm{\cdot}_0 = \norm{\cdot}_{l_\infty}$
and
set $X_0 = (c_{00}, \norm\cdot_0)$. For $n\ge 1$, we let $X_n = (c_{00}, \norm\cdot _n)$,
where $\norm{x}_n$ be the maximum of
$$\sup_{m\in \NN} \sup {1\over f(m)} \sum_{i=1}^m \norm{E_i x}_{n-1},$$
where the second supremum is over all sequences of subsets $E_1 < \cdots < E_m$, and
$$\sup_{k\in K} \sup_{g\in B^*_k(X_{N-1})} \sup_{E\subseteq \NN} \bigl|g(Ex)\bigr|.$$
This is a nondecreasing sequence of norms that is bounded above by the $l_1$-norm, so it has a limit, which
we call $\norm \cdot$. The completion $X$ of $(c_{00}, \norm\cdot)$ is the space we're after, though the proof
that $X$ is hereditarily indecomposable would run us aground.

An operator $T:X\to Y$ is {\it strictly singular} if there is some $c>0$ such that $\norm{Tx} \ge c\norm x$ for
all $x\in X$. Gowers and Maurey show that every operator from
$X$ to $X$ can be written as a scalar multiple of the identity, plus a strictly singular operator. This
implies that there are, in some sense, very few operators on $X$. A subsequent paper of Gowers and Maurey,
published in 1997,
describes a more general method of producing
Banach spaces whose spaces of operators are small~\ref{gowersmaurey1997}. Recall that an operator
is said to be {\it compact} if it sends bounded sets in the domain to sets with compact closure in the image.
Every compact operator is strictly singular.
In their original
1993 paper, the authors ask if there exists a Banach space $X$ such that every operator from $X$ to $X$
is a scalar multiple of the identity, plus a compact operator. This question was answered positively
S.~A.~Argyros and R.~G.~Haydon in 2009; the paper was published in 2011~\ref{argyroshaydon}.

\medskip\boldlabel The Banach hyperplane problem. In a book published in 1932,
S.~Banach~\ref{banach} asked whether an infinite-dimensional Banach space $X$ over the real
numbers is always isomorphic to $X\oplus \RR$. This amounts to asking whether every infinite-dimensional
Banach space $X$ is such that
that every subspace of codimension $1$ is isomorphic to $X$ itself. Since a subspace of codimension $1$ is
sometimes called a {\it hyperplane}, this question came to be known as the hyperplane problem.
It turns out that the answer to the question is no, and there is an infinite-dimensional
Banach space that is not isomorphic
to any of its hyperplanes.

This problem was solved by W.~T.~Gowers soon after he solved the unconditional basic sequence problem.
His construction, which appeared in print in 1994, is extremely similar~\ref{gowers1994}.
We shall thus retain the definitions from the previous
subsection and immediately define this space. As the reader probably suspects at this point, the construction
is inductive.
Set $X_0 = (c_{00}, \norm\cdot_0)$. For $n\ge 1$, we let $X_n = (c_{00}, \norm\cdot _n)$,
where $\norm{x}_n$ be the maximum of
$$\sup_{m\in \NN} \sup {1\over f(m)} \sum_{i=1}^m \norm{E_i x}_{n-1},$$
where the inner supremum is over all sequences of subsets $E_1 < \cdots < E_m$, and
$$\sup_{k\in K} \sup_{g\in B^*_k(X_{N-1})} \bigl|g(x)\bigr|.$$
The notation we use here is not exactly the same as in the original paper, but using the same letters
as before makes it clear that the only change from the space without an unconditional basic sequence
is the replacement of $\sup_{E\subseteq \NN} \bigl|g(Ex)\bigr|$ has changed to the simpler $\bigl|g(x)\bigr|$.
The space $X$ is the completion of $c_{00}$ with respect to the limit of the norms $\norm{x}_n$ above.
Perhaps surprisingly, this new space $X$ has an unconditional basis. It is noted in Gowers' paper
that the Gowers--Maurey
space of the previous section is also a counterexample to the hyperplane problem, though is is
more difficult to prove.
The proof that $X$ is not isomorphic to any of its hyperplanes is done by showing that $X$
satisfies the hypothesis of the following lemma, which is attributed to P.~G.~Casazza.

\proclaim Lemma C. If $X$ is a Banach space in which no two equivalent sequences $(y_i)$ and $(z_i)$ satisfy
$y_i < z_i < y_{i+1}$ for all $i$, then $X$ is not isomorphic to any proper subspace.\slug

\section References

\bye

### 2021-12-17 misc/study-notes/math596notes.tex
Added more notes and video

\input fontmac
\input mathmac

\def\AA{{\bf A}}
\def\FF{{\bf F}}
\def\HH{{\bf H}}
\def\PP{{\bf P}}
\def\H{{\cal H}}
\def\M{\op{\rm M}}
\def\calM{\mathop{\cal M}\nolimits}
\def\calS{\mathop{\cal S}\nolimits}
\def\T{{\rm T}}
\def\O{\op{\rm O}}
\def\SO{\op{\rm SO}}
\def\SL{\op{\rm SL}}
\def\GL{\op{\rm GL}}
\def\ord{\op{\rm ord}}
\def\disc{\op{\rm disc}}
\def\im{\op{\rm im}}
\def\Aut{\op{\rm Aut}}
\def\Stab{\op{\rm Stab}}
\def\Hom{\op{\rm Hom}}
\def\ortho{\mathbin{\widehat\oplus}}
\def\bar{\overline}
\def\pmod#1{\;({\rm mod}\;#1)}
\outer\def\nineproclaim #1. #2\par{\medbreak
  {\noindent\ninepoint{\bf#1.\enspace}{\sl#2\par}}%
  \ifdim\lastskip<\medskipamount \removelastskip\penalty55\fi}
\long\def\nineproof#1\slug{{\noindent\ninepoint\proof#1%
\quad\hbox{\kern1.5pt\vrule width2.5pt height6pt depth0.5pt\kern1.5pt}\medskip}}
\long\def\ninesolution#1\slug{{\noindent\ninepoint\solution#1%
\quad\hbox{\kern1.5pt\vrule width2.5pt height6pt depth0.5pt\kern1.5pt}\medskip}}
\def\smallmatrix#1#2#3#4{
    \big({#1\atop #3}{#2\atop #4}\big)
}
\def\longto{\longrightarrow}
\def\hat{\widehat}

\widemargins
\bookheader{MARCEL K. GOH}{MATH {\eightrm 596} NOTES}

% hacky divides
\def\divides{\mathrel\backslash}
\catcode`@=11
\def\notdivides{\mathrel{\mathpalette\c@ncel\divides}}
\catcode`@=12 % at signs are no longer letters

\font\elevenbf=cmbx12 at 11 pt

\maketitle{\elevenbf MATH 596: Quadratic and modular forms}{notes
by}{Marcel K. Goh}{10 December 2021}

\floattext5 \ninebf Disclaimer. \ninepoint
These notes were taken for the class MATH 596, given at McGill University by
Prof.~Henri Darmon at McGill University during the Fall 2021 semester. Over the course of the term,
students were asked to present solutions to exercises. I have indicated when this occurred by
attaching students' names to their respective solutions. However, in some cases, the solution I recorded
here is not word-for-word the one presented, as I sometimes found a modification that I understood better.
An exercise solution that is unattributed does
not necessarily indicate that it is
completely my work, since I spent a lot of time discussing the material with my classmates.
But any error that appears in the notes
or exercise solutions, whether typographical or mathematical, are due to me and me alone.

\bigskip

\noindent
Let $V$ be a module over a commutative ring $R$. A function $Q:V\to R$ is a {\it quadratic form} if
it satisfies
\medskip
\item{i)} $Q(ax) = a^2Q(x)$ for all $a\in R$ and $x\in V$; and
\smallskip
\item{ii)} the function $(x,y)\mapsto Q(x+y)-Q(x)-Q(y)$ is a bilinear form.
\medskip
We call $(V,Q)$ a {\it quadratic module}; if $R$ is a field, $V$ is a vector space and we call
$(V,Q)$ a {\it quadratic space} instead. When
$R$ is a field of characteristic not equal to $2$, we can let
$$x\cdot y = {1\over 2}\bigl(Q(x+y)-Q(x)-Q(y)\bigr).$$
This defines a symmetric bilinear form on $V$, and we have $Q(x) = x\cdot x$ and there is a one-to-one
correspondence between symmetric bilinear forms and quadratic forms (which is not true if the characteristic
of the field is equal to $2$).

Pick a basis $(e_i)_{i=1}^n$ of $V$. The matrix $A = (a_{ij})$ where $a_{ij} = e_i\cdot e_j$ is a symmetric
matrix, and for $x = \sum_i^n x_i e_i \in V$,
$$Q(x) = \sum_{i=1}^n \sum_{j=1}^n a_{ij} x_i x_j.$$
If we switch to a new basis with the invertible change-of-basis matrix $B$, then the matrix of $Q$
with respect to this new basis is the matrix $B A B^\T$, which has determinant $\det(A)\det(B)^2$.
We see then that for a quadratic form $Q$, the determinant of the matrix $A$ corresponding to $Q$ in
any basis is unique up to multiplication by an element of $(k^\times)^2$;
we call this the {\it discriminant} of $Q$ and denote it by $\disc Q$.

Two elements $x$ and $y$ are {\it orthogonal} if $x\cdot y = 0$. For a subset $W\subseteq V$, we define
a vector subspace
$$W^\perp = \{ x\in V : x\cdot y = 0\ \hbox{for all}\ y\in W\}.$$
Two vector subspaces $W_1$ and $W_2$ of $V$ are said to be {\it orthogonal} if $W_1\subseteq {W_2}^\perp$
(this is a symmetric relation).
We call $V^\perp$ the {\it radical} of $V$ and say that $V$ is {\it nondegenerate} if $V^\perp = \{0\}$.
The codimension of $V^\perp$ is called the {\it rank} of $V$.
If $V$ is the direct sum of vector subspaces $W_1, \ldots, W_n$ and the $W_i$ are pairwise orthogonal,
then we say that $V$ is the {\it orthogonal direct sum} of the $W_i$, and write
$$V = W_1\ortho \cdots \ortho W_n.$$

An element $x\in V$ is said to be {\it isotropic} if $x\cdot x = 0$, and a subspace is {\it isotropic}
if all of its elements are.
If no nonzero element in a subspace $W$ is isotropic, the $W$ is said to be
{\it anisotropic}. The linear span of two basis vectors $e$ and $f$
with $e\cdot e = f\cdot f = 0$ and $e\cdot f = 1$ is called a {\it hyperbolic plane} and is often denoted
$H$.

\nineproclaim
Exercise 1. Let $V$ be a nondegenerate quadratic space. Show that any two maximal isotropic
spaces of $V$ have the same dimension, $t$, called the {\it Witt index} of $V$. Show that $V$ is isomorphic
to an orthogonal direct sum of $t$ hyperbolic spaces and an anisotropic space $W$ of dimension $n-2t$
(where $n = \dim V$).

\nineproof (Hazem Hassan and Arihant Jain.)
We start with the claim that any two maximal isotropic spaces have the same dimension. First observe
that if $U$ is a maximal isotropic subspace, then any $u\in U^\perp$ with $u\cdot u =0$ must also be in $U$,
otherwise we could extend $U$ and contradict maximality. So let $U_1$ and $U_2$ be maximal isotropic
subspaces. If $U_1 = U_2$ we are done; otherwise,
consider $U_1\times U_2\to k$ that maps $(u_1, u_2)\mapsto u_1\cdot u_2$. If $u_1$
is such that $u_1\cdot u_2 = 0$ for all $u_2\in U_2$, then by the observation above, $u_1\in U_2$. So the
left kernel of this map is $U_1\cap U_2$ and a similar argument shows that this is also the right kernel.
This means the map
$${U_1\over U_1\cap U_2} \times  {U_2\over U_1\cap U_2} \to k$$
is a perfect pairing; that is, $U_1/(U_1\cap U_2)\to \Hom_k\bigl(U_2/(U_1\cap U_2), k\bigr)$ is an isomorphism
and vice versa. So $\dim U_1 = \dim U_2$.

Now we show that we can write $V = W \ortho H_1 \ortho \cdots \ortho H_t$ where $W$ is anisotropic, $t$
is the dimension of every maximal isotropic subspace of $V$, and the $H_i$
are hyperbolic spaces. Let $H_i = ke_i\oplus kf_i$ where $e_i\cdot e_i = 0 = f_i\cdot f_i$ and $e_i\cdot f_i = 1$.
Let $U = ke_1\oplus \cdots \oplus ke_t$. Then we see that $U$ is isotropic, since $U^\perp = U\ortho W$
and for $v = u+w$ in this space, $v\cdot v = (u+w)\cdot (u+w) = w\cdot w$ is only zero if $w=0$. Hence $U$
is maximal and isotropic, proving that every maximal isotropic subspace has dimension $t$.\slug

\boldlabel Quadratic spaces over $\RR$. Note that $\RR^\times/(\RR^\times)^2 = \{\RR_{\ge 0}, \RR_{\le 0}\}$.
If $V$ is a nondegenerate quadratic space over $\RR$, then it has
an orthogonal basis
$$e_1,\ldots, e_r, e_{r+1}, \ldots, e_{r+s}$$
with $e_j\cdot e_j = 1$ for $1\le j\le r$
and $e_j\cdot e_j = -1$ for $r+1\le j\le r+s$. The pair $(r,s)$ is called the {\it signature} of the quadratic
space. The Witt index of $V$ is $t= \min\{r,s\}$, and if $V = W\ortho H^t$, where $H^t$ is the orthogonal
direct sum of $h$ copies of a hyperbolic plane, then $\dim W = |r-s|$. When $r>s$,
$W$ is positive definite and when $r<s$, $W$ is negative definite.

The orthogonal group $\O(V)$ of a quadratic space $V$ over $R$ is the space
$$\O(V) = \{g \in \Aut(V) : gv\cdot gw = v\cdot w\ \hbox{for all}\ v,w\in V\}.$$
Letting $n=\dim V$ and $(r,s)$ be the signature of the space, this can be viewed as the set of $n\times n$
matrices
$$\bigg\{A\in \M_n(\RR) : A^\T A = AA^\T = \pmatrix{I_r&0\cr 0&-I_s}\bigg\},$$
where the $I_r$ and $I_s$ that appear in the block matrix are the $r\times r$ and $s\times s$ identity
matrices, respectively. The condition implies that $\det(A)^2 = 1$, meaning that $\det A = \pm 1$ for all
$A\in \O(V)$. The subgroup of $A\in \O(V)$ with $\det(A) = 1$ is called $\SO(V)$. It was proven in class
by a standard inductive argument that $\O(V)$ is a real manifold (and thus a {\it Lie group})
of dimension $n(n-1)/2$, and so is the subgroup $\SO(V)$.

% \proclaim Proposition D. For a quadratic space $V$ over $\RR$ with $\dim V = n$,
% the orthogonal group $\O(V)$ is a real manifold of dimension $n(n-1)/2$.
% 
% \proof By induction on $n$. When $n=1$, $\O(V) = \{\pm 1\}$, which has dimension $0$. Now suppose that
% $V$ has dimension $n>1$. Let $X = \{v\in V : v\cdot v = 1\}$ and note that $\O(V)$ acts transitively on $X$.
% Since $X$ is the set of all $(x_1, \ldots, x_n)\in \RR^n$ with $Q(x_1, \ldots, x_n) = 1$, we see that
% $\dim X = n-1$. Fix $v\in X$ and note that $=\Stab_{{\rm O}(V)}(v) = \O(V')$ where $V' = (\RR v)^\perp$ is
% a quadratic subspace of dimension $n-1$. By the
% orbit-stabiliser theorem, $X$ can be identified with the quotient space $\O(V)/\O(V')$. Using the
% inductive hypothesis, we have
% $$\dim \O(V) = \dim\O(V) + \dim X = {(n-1)(n-2)\over 2} + n-1 = {n(n-1)\over 2}.\noskipslug$$

\medskip\boldlabel Hamilton quaternions.
The {\it Hamilton quaternions} are members of the set $\HH = \RR + \RR i + \RR j+ \RR k$ (since the working
field here is $\RR$, we will temporarily allow ourselves the use of the letter $k$ for one of the generators
of the space) where we have the
relation $i^2 = j^2 = k^2 = ijk = -1$. If $a = x + yi + zj + wk$ then $\bar a = x - yi-zj-wk$ and
the {\it norm} of $a$ is $N(a) = a\bar a = x^2 + y^2 + z^2 + w^2$. We see from this that every nonzero element $a$
has a multiplicative inverse, namely, $a^{-1} = \bar a/N(a)$ with $N(a^{-1}) = N(a)^{-1}$.
Thus $\HH^\times = \HH\setminus\{0\}$. The {\it trace} of $a$ is $(a+\bar a)/2$.

\medskip\boldlabel Infinitesimals and the tangent space of the identity. Here we give a somewhat informal
description of the tangent space of the identity element in a topological group. Let $M$ be a topological
group with identity element $1$ that is also a subgroup of the multiplicative group of an algebra $A$.
The tangent space $T_1M$ at the point $1$ is the space of all $a\in A$ such that $1+\eps a$ is still in $M$,
and here we encapsulate the fact that $\eps$ should be very small by letting $\eps$ be a nonzero formal
parameter whose square is zero. The tangent space is closed under addition, since if $a$ and $b$ are elements
of $T_1M$, then
$$(1+\eps a)(1+\eps b) = 1+\eps a + \eps b + \eps^2 ab = 1 + \eps(a+b)$$
and we see that $a+b$ is also in the space. It is also clear that $T_1 M$ is closed under multiplication
by scalars in the field, so that $T_pM$ is a real vector space. The tangent space of a point $(p,q)\in M\times N$
is the product of spaces $T_pM\times T_q M$.

Let $M$ and $N$ be two such groups with identities $1_M$ and $1_N$ respectively and let
$\phi : M\to N$ be a group homomorphism. We define the {\it differential} $d\phi$ to be the map
from $T_{1_M} M$ to $T_{1_N} N$ such that the diagram
$$\matrix{
T_{1_M} M & \buildrel d\phi\over\longrightarrow & T_{1_N} N \cr
\quad\Big\downarrow\pi_N & & \quad\Big\downarrow\pi_N \cr
M & \buildrel \phi\over\longrightarrow & N \cr
}$$
commutes, where $\pi_M(a) = 1_M +\epsilon a$ and $\pi_N(b) = 1_N + \epsilon b$. The commutativity of the
above diagram tells us that $d\phi(a)$ is given by the formula $1_N + \epsilon d\phi(a) = \phi(1_M+\epsilon a)$
for $a\in T_{1_M} M$.

Tangent spaces are relevant because the inverse function theorem tells us that if $\phi:\RR^n\to \RR^n$ is
a differentiable map and its derivative (Jacobian) at a point $x$ is invertible, then there exists an open
neighbourhood $U$ of $x$ that is homeomorphic to its image $\phi(U)$. If $\phi$ is also a group homomorphism,
then we have the following lemma.

\proclaim Lemma T. Let $\phi:G\to H$ be a homomorphism of topological groups and suppose that there is an
open neighbourhood $S\subseteq \phi(G)$ that contains the identity of $H$. If $G$ is connected, then
$\phi(G)$ is the connected component containing the identity in $H$.

\proof Since the function $h\mapsto h^{-1}$ is a continuous involution, $S^{-1}$ is also open and so is
$S\cap S^{-1}$, which still contains the identity. Thus we may assume without loss of generality that $S$
is closed under inverses. Let $\langle S\rangle$ be the smallest subgroup of $\phi(G)$ with $S\subseteq
\langle S\rangle$. It is easy to see that $\langle S\rangle = \bigcup_{h\in \langle S\rangle} hS$. For all
$h\in \phi(G)$, left multiplication by $h$ is a homeomorphism from $\phi(G)$ to itself, to $\langle S\rangle$
is open.

It remains to show that $\langle S\rangle$ is closed, so let $h\in \langle S\rangle^c$ be given. If
$hs \in \langle S\rangle$ for some $s\in S$, then $h = hss^{-1}\in \langle S\rangle$. So $hS$ is an open
neighbourhood of $h$ contained in $\langle S\rangle^c$, proving that $\langle S\rangle$ is indeed closed.
Since $\langle S\rangle$ is nonempty, open, and closed in the connected group $\phi(G)$, $\langle S\rangle
=\phi(G)$ and $\phi(G)$ is the connected component of the identity.\slug

Finally, we will take the following lemma on faith.

\proclaim Lemma C. If $G$ is a Lie group with a connected compact Lie subgroup $H$ such that $G/H$ is also
connected, then $G$ is connected.\slug

\nineproclaim
Exercise 2. Describe $\O(V)$ and $\SO(V)$ when $V$ is a nondegenerate
quadratic space of dimension $4$ over $\RR$. How many connected components does the full orthogonal group have?

\ninesolution (Marcel Goh and Jad Hamdan.)
We will write $\SO(r,s)$ to denote $\SO(V)$ when $V$ has signature $(r,s)$, and write $\SO(n)$ for $\SO(n,0)$.
The question has two largely unrelated parts. First we describe the components of the identity
in the three separate cases. First we deal with the case $(r,s) = (4,0)$. Then
$$Q(x,y,z,w) = x^2 + y^2 + z^2 + w^2$$
and we can identify $(V,Q)$ with $(\HH, n)$. Note that the group $\HH^\times\times\HH^\times$ acts on $V$ by
setting $(g,h)*v = gvh^{-1}$. The norm of $(g,h)*v$ is $N(g) N(h^{-1}) N(v)$, so for an element $(g,h)$
to preserve the norm, it is necessary and sufficient that $N(g) = N(h)$. We can also assume that $g$ and $h$
have norm $1$, since if $\lambda = N(g) = N(h)$, then $g=\lambda g'$ and $h=\lambda h'$ for some unit quaternions
$g'$ and $h'$ and
$$(g,h)*v = (\lambda g', \lambda h')*v = \lambda g'v\lambda^{-1} h'^{-1} = g'vh'^{-1} = (g',h')*v.$$
In particular, if $g$ and $h$ are both real, then $(g,h)$ sends any $v\in V$ to itself. Thus, letting
$\HH_1$ denote the set of quaternions with norm $1$, we have the exact sequence
$$ 1\longrightarrow \bigl\{(-1,-1),(1,1)\bigr\}\longrightarrow \HH_1\times \HH_1
\buildrel\phi\over\longrightarrow \O(V).$$
Note that the last map $\phi$ in the sequence is not surjective. In fact, the image of $\phi$ is contained
in $\SO(4)$, since if we represent the action of $(g,h) = (a_1 + a_2i + a_3j + a_4k, b_1+b_2i+b_3j+b_4k)$
on a quaterion $v$ as the action of a matrix on a vector in $\RR^4$, then the matrix of the transformation
is $AB$, where
$$ A = \pmatrix {
a_1 & -a_2 & -a_3 & -a_4 \cr
a_2 & a_1 & -a_4 & a_3 \cr
a_3 & a_4 & a_1 & -a_2 \cr
a_4 & -a_3 & a_2 & a_1 \cr
}\qquad\hbox{and}\qquad
B = \pmatrix {
b_1 & -b_2 & -b_3 & -b_4 \cr
b_2 &  b_1 &  b_4 & -b_3 \cr
b_3 & -b_4 &  b_1 &  b_2 \cr
b_4 &  b_3 & -b_2 &  b_1 \cr
}.$$
We have $AA^\T = I$ and $BB^\T = I$ so $(AB)(AB)^\T = I$, as prescribed for a member of $\O(V)$. But we
also see that $\det(A) = N(g) = 1 = N(H) = \det(B)$, so that $\det(AB) = 1$ as well.

To show that $\SO(4)\subseteq \im(\phi)$, we need to find the tangent space of the identity in $\HH_1$.
if $1+\eps(x+yi+zj+wk)$ has norm $1$, then we must have
$$(1+\eps x)^2 + \eps^2 y^2 + \eps^2 z^2 + \eps^2 w^2 = (1+\eps x)^2 = 1,$$
so the trace $x$ of the quaternion must be zero. Thus $T_1\HH$ is $3$-dimensional and $T_1(\HH_1\times \HH_1)$
is $6$-dimensional. On the other hand, the tangent space of the identity matrix $I$ in $\SO(4)$
is the set of all matrices with $A^\T = -A$, since the condition that $(I+\eps A)(I+\eps A)^\T$
implies that $I+\eps(A+A^\T) = I$. This is a $6$-dimensional space as well, since the diagonal of $A$ must
have all entries zero, and the rest of the matrix is determined by the choice of the six remaining entries
in the upper triangle. Thus to show that $d\phi$ is invertible, it suffices to show it is injective.

Now for $(a,b)\in T_1(\HH_1\times \HH_1)$, we note that $\phi(1+\eps a, 1+\eps b)$ is a map on $\HH$
that sends $v$ to $(1+\eps a)v(1+\eps b)^{-1}$. Since $b$ has zero trace, $(1+\eps b)^{-1} = 1-\eps b$
and
$$\phi(1+\eps a, 1+\eps b)(v) = (1+\eps a)v(1+\eps b)^{-1} = (v+\eps av)(1-\eps b) = v + \eps(av-vb).$$
Then since $\eps d\phi(a,b)$ equals $\phi(1+\eps a, 1+\eps b)$ minus the identity endomorphism, we have
$d\phi(a,b)(v) = av-vb$. If $(a,b)\in \ker(d\phi)$, then $av-vb=0$ for all $v\in \HH$, and taking $v=1$ in
particular, we have $a=b$. Thus $av=va$ for all $v\in \HH$ and we see that $a\in \RR$. But we assumed that
$a$ has trace zero, so $(a,b) = (0,0)$. We have shown that $d\phi$ is bijective, so by Lemma~T, the
image of $\phi$ is the connected component of the identity in $\O(4)$.

In the case $(r,s) = (3,1)$, we have
$Q(x,y,z,w) = x^2 + y^2 + z^2 - w^2$ but we can perform a change of basis with $u = z+w$ and $v=z-w$
to get $Q(x,y,u,v) = x^2 + y^2 + uv$. Since $x^2 + y^2 = (x+iy)(x-iy)$, we can identify $(V,Q)$ with the set
of all matrices
$$\biggl\{ \pmatrix{x+iy & u \cr v &  iy-x} : x,y,u,v\in \RR\biggr\},$$
with the negative determinant as the norm. Letting $M^* = \det(M) M^{-1}$, we find that $V$ is precisely
the set of $M\in M_2({\bf C})$ with $M^* = -\overline M$. We have $(AB)^* = B^*A^*$ for
$A,B\in M_2({\bf C})$,
and this operation is also linear. The group $\SL_2(\CC)$ acts on $V$ by $g* M = gM\overline g^{-1}$.
Indeed,
$$(g* M)^* = (gM\overline g^{-1})^* = \overline g M^* g^* = - \overline g\overline M g^* = -
\overline{gM\overline g^*} -\overline{g* M},$$
so $g* M$ is in $V$. Note that the center of $\SL_2(\CC)$ is $\{\pm I\}$, and we have the
exact sequence
$$ 1\longrightarrow \{\pm I\} \longrightarrow \SL_2(\CC) \buildrel\phi\over\longrightarrow {\rm O}(V).$$
Once again, the image of $\phi$ is connected.
Now we find the tangent space of $\SL_2(\CC)$ at the identity. An element
$$A = \pmatrix{ a & b \cr c & d }\in \M_2(\CC)$$
in this space satisfies
$I + \epsilon A\in \SL_2(\CC)$, so $\det (I+\epsilon A) = 1+\epsilon a + \epsilon d = 1$.
This implies that $d=-a$. In other words, the trace of $A$ is zero, and in particular we see that $A^* = -A$.
Now we examine what $I+\epsilon A$ does to a matrix $M\in V$. We note first that
$$(I+\epsilon \overline A)^{-1} =
\pmatrix{ 1 + \epsilon \overline a & \epsilon \overline b \cr
\epsilon\overline c & 1-\epsilon \overline a }^{-1}
=\pmatrix{ 1 - \epsilon \overline a & -\epsilon \overline b \cr
-\epsilon\overline c & 1+\epsilon \overline a } = I+\epsilon \overline{A^*}$$
So
$$(I+\epsilon A)M\overline{(I+\epsilon A)}^{-1} = (M+\epsilon AM)(I+\epsilon \overline{A^*})
= M + \epsilon AM + \epsilon M \overline{A^*} = M+ \epsilon(AM-M\overline A),$$
telling us that $d\phi(A)$ takes matrices $M$ to $AM-M\overline A$. We now investigate what it means for
$A$ to be in the kernel of $d\phi$. If $MA - M\overline A$ for all matrices $M$, then taking
$M = I$, we see that $A = \overline A$, meaning that $A$ has all real entries and $AM = MA$ for all
$M$. This implies that $A$ is a scalar multiple of the identity and since it has zero trace, $A$ must be $0$.
We have found that $\ker(d\phi) = 0$, so the connected component of
the identity is isomorphic to $\SL_2(\CC) / \{\pm I\}$.

The third case $(r,s) = (2,2)$ feels a bit like a combination of the two cases above.
We have $Q(x,y,z,w) = x^2 + y^2 - z^2 - w^2$ and with the substitutions
$x = x+z, y=x-z, z = w+y, w= w-y$ (the variables on the left-hand side are not the same as the ones on the
right-hand side), we have
$$Q(x,y,z,w) = xy-zw,$$
so we can identify $(V,Q)$ with $(M_2({\bf R}), \det)$. The group ${\rm GL}_2({\bf R})\times{\rm GL}_2({\bf R})$
defines an action on $M_2({\bf R})$ given by $(g,h)* M = gMh^{-1}$. For $(g,h)$ to preserve the determinant
we must have $\det g = \det h$. We can also require that $g$ and $h$ have determinant $1$, because for any
$\lambda\in {\bf R}$ we have $\det(\lambda g) = \det(\lambda h)$ and
$$ \lambda g M(\lambda h)^{-1} = \lambda^2 g M \lambda^{-2} h^{-1} = gMh^{-1}.$$
Thus we have the exact sequence
$$1 \to \{(I,I), (-I, -I)\} \to {\rm SL}_2({\bf R})\times{\rm SL}_2({\bf R}) \buildrel\phi\over\to {\rm O}(V).$$
The computation we performed above for the tangent space of $\SL_2(\CC)$ works when the entries
are real as well, so we find that the tangent space of ${\rm SL}_2({\bf R})\times{\rm SL}_2({\bf R})$
is the set of $(A,B)$ such that ${\rm tr}\,A = {\rm tr}\,B = 0$. In particular, since the trace of $B$
is zero, we can write
$$ B = \pmatrix{a & b \cr c & -a} $$
and the fact that $\det(I + \epsilon B) = 1$ implies that
$$(I + \epsilon B)^{-1} = \pmatrix{ 1+a & b \cr c & 1-a }^{-1} = \pmatrix{1-a & -b \cr -c & 1+a}
= I - \epsilon B.$$
So
$$\phi(I+\epsilon A, I+\epsilon B)(M) = (I+\epsilon A)M(I+\epsilon B)^{-1}
= (M + \epsilon AM)(I-\epsilon B) = M+\epsilon (AM-MB),$$
and $d\phi(A,B)(M) = AM-MB$.
We argue as before to find that for $(A,B)\in \ker(d\phi)$,
$A=B$ and $A$ is a scalar of the identity with trace zero and thus $(A,B) = (0,0)$.
So $d\phi$ is injective and the image of $\phi$ is isomorphic to
$${\rm SL}_2({\bf R})\times{\rm SL}_2({\bf R})/\{(I,I), (-I,-I)\}.$$

On to the second part of the question. The claim is that $\O(4)$ has two connected components and
that $\O(3,1)$ and $\O(2,2)$ both have four. Since $\SO(V)$ is a subgroup
of index $2$ in the group $\O(V)$, it suffices to that $\SO(V)$ is connected in the definite case and that
it has two connected components in the other two cases. We do this by induction, building up from
smaller-dimensional instances.

Let $V$ be a $4$-dimensional quadratic space with signature $(r,s)$. Note that $\SO(r,s)$ acts
transitively on the set $X = \{x\in V : x\cdot x = 1\}$, which is the orbit of the point $e_1 = (1,0,0,0)$.
A matrix in $\Stab(e_1)$ has $e_1$ as its first row and column, so it must have the form
$$\pmatrix{1&0\cr 0&M}$$
for some $M\in \SO(r-1,s)$. So by the orbit-stabiliser theorem, we have a diffeomorphism
$\SO(r,s)/\SO(r-1,s)\cong X$. We will proceed by induction.

The base cases are $\SO(1)$ and $\SO(1,1)$; the former is $\{1\}$, which clearly has one connected component.
On the other hand, $\SO(1,1)$ consists of orthogonal $2\times 2$ matrices $M=\smallmatrix abcd$
with $\det(M)=1$ and
$$M^{-1} = \pmatrix{1&0\cr 0&-1} M^{\T}\pmatrix{1&0\cr 0&-1}.$$
This condition implies that $a = d$ and $b=c$, so we can write
$$\SO(1,1) = \biggl\{\pmatrix{a&b\cr b&a}\in \SL_2(\RR) : a^2-b^2 =1; a,b\in \RR\biggr\}$$
establishing a bijection from $\SO(1,1)$ to the algebraic set $x^2-y^2=1$, a hyperbola with two branches
on either side of the $y$-axis. This is in fact a homeomorphism, showing that $\SO(1,1)$ has two connected
components.

In the definite case $(r,s)=(4,0)$ the induction is straightforward. The set $X$ of elements of norm $1$
in a quadratic space of signature $(n,0)$ is the unit sphere $S^{n-1}$, which is connected. Having shown
that $\SO(1)$ is connected and that the quotient $\SO(n)/\SO(n-1) \cong S^{n-1}$ is connected for all $n>1$,
we apply Lemma~C inductively to conclude that $\SO(n)$ is connected for all $n$.

For the indefinite cases, the induction is a bit more involved. The set of elements of unit norm in
a quadratic space of signature $(1,2)$, $(1,3)$, or $(2,2)$ is the set of tuples $(x,y,z)$ or $(x,y,z,t)$
satisfying
$$x^2-y^2-z^2 = 1,\quad x^2-y^2-z^2-t^2=1,\quad \hbox{or} \quad x^2+y^2-z^2-t^2 = 1$$
respectively, where are all two-sheeted hyperboloids each with two connected components. In each case,
let $X^+$ be the sheet containing $(0,\ldots,1)$. If we let $\SO^+(r,s)$ be the set of matrices of
$\SO(r,s)$ that preserve $X^+$, one can show that $\SO^+(r,s)$ is a subgroup of $\SO(r,s)$ with
index $2$. Using an orbit-stabiliser argument analogous to the one above, we find that
$\SO^+(r,s)/\SO^+(r-1,s)\cong X^+$.

The set of unit elements in a space of signature $(1,1)$ is a hyperbola. We have also shown that $\SO(1,1)$
is homeomorphic to a hyperbola by explicit computation. Using this identification, we find that the set
$\SO^+(1,1)$ is homeomorphic to a branch of $\SO(1,1)$ and is therefore connected. By repeated application
of Lemma~C and the fact that $\SO(r,s)\cong \SO(s,r)$ for all integers $r,s$, we conclude that
$\SO^+(1,3)$ and $\SO^+(2,2)$ are connected. So $\O(1,3)$ and $\O(2,2)$ both have four connected components,
and we already showed that $\O(4,0)$ has two, finishing the exercise.\slug

\boldlabel The Hilbert symbol. For this discussion, let $k$ denote either $\RR$ or $\QQ_p$. For
$a,b\in k^\times$, we define the {\it Hilbert symbol} $(a,b)$ by setting
$$(a,b) = \cases{1,& if $ax^2 + by^2 = z^2$ has a nonzero solution $(x,y,z)$ in $k^3$;\cr -1& otherwise.}$$
Note that $(a,b) = (a, c^2b)$ for any element $c\in k^\times$, since the square can be absorbed into the
variable. For $a,b\in k^\times$, further properties of the Hilbert symbol include
\medskip
\item{i)} $(a,b) = (b,a)$ and $(a, b^2) = 1$;
\smallskip
\item{ii)} $(a,-a) = 1$ and $(a,1-a) = 1$;
\smallskip
\item{iii)} if $(a,b) = 1$ then $(ac,b) = (c,b)$ for all $c\in k^\times$; and
\smallskip
\item{iv)} $(a,b) = (a,-ab) = \bigl(a, (1-a)b\bigr)$.
\medskip
Furthermore, it can be shown that the Hilbert symbol is bilinear; that is, $(ac,b) = (a,b)(c,b)$ for all
$a,b,c\in k^\times$.

\medskip
\boldlabel Quadratic forms over $\QQ_p$. Let $(V,Q)$ be a quadratic space of rank $n$ over $\QQ_p$ and pick
an orthogonal basis $(e_1, \ldots, e_n)$. Letting $a_i = e_i\cdot e_i$, we have $\disc(Q) = a_1\cdots a_n$.
The {\it Hasse-Witt invariant} of $V$ is the product
$$\eps(V) = \prod_{i<j} (a_i, a_j);$$
we saw in class that this does not depend on the choice of orthogonal basis. We also saw that two quadratic
forms over $\QQ_p$ are equivalent (this means their respective matrices $A$ and $A'$ are related by
$A' = BAB^\T$ for some invertible matrix $B$)
if and only if they have the same rank, discriminant, and Hasse-Witt invariant.

\proclaim Theorem Z. Let $(V,Q)$ be a quadratic space of rank $n$ over $\QQ_p$. Writing
$d=\disc(Q)\in \QQ_p/(\QQ_p^\times)^2$
and letting $\eps = \eps(V)$ be the Hasse-Witt invariant of the space, there is a
nonzero vector $x\in V$ with $Q(x) = x\cdot x = 0$ if and only if
\medskip
\item{i)} $n=2$ and $d=-1$;
\smallskip
\item{ii)} $n=3$ and $(-1,-d)=\eps$;
\smallskip
\item{iii)} $n=4$ and either $d\ne 1$ or else $d=1$ and $\eps = (-1,-1)$; or
\smallskip
\item{iv)} $n\ge 5$.\slug
\medskip

\nineproclaim
Exercise 3. Show that the Hilbert symbol $(a,b)$ for $a,b\in \QQ_p^\times$ is equal to $-1$
if and only if the idoneous central simple algebra over $\QQ_p$ defined by
$$B=\QQ_p + \QQ_pi + \QQ_pj + \QQ_pk,$$
where $i^2 = a$, $j^2 = b$, and $ij = -ji = k$, is a division algebra, and that it is isomorphic
to the matrix algebra $\M_2(\QQ_p)$ if $(a,b) = 1$.

\nineproof (Sun Kai Leung and Paul-Antoine Seitz.)
Call this central simple algebra $A$ and let $q = x+yi+zj+wk \in A$. Defining $\bar q = x-yi-zj-wk$,
a routine computation gives $q\bar q = x^2 - ay^2 - bz^2 +ab w^2\in \QQ_p$. This is a quadratic form
in the variables $x$, $y$, $z$, and $w$ of discriminant $(ab)^2 \sim 1$. The Hasse-Witt invariant of
the associated quadratic space is
$$\eqalign{
(1,-a)(1,-b)(1,ab)(-a,-b)(-a,ab)(-b,ab) &= (1,ab)(1,ab)(-a,-ab^2)(-b,ab) \cr
&= (-a,-a)(-b,ab) \cr
&= (-1,-a)(a,-a)(-b,b)(-b,a) \cr
&= (-1,-1)(-1,a)(-b,a) \cr
&= (-1,-1)(a,b).\cr
}$$
If $(a,b) = -1$, by part (iii) of Theorem Z, we have $q\bar q \ne 0$ for all nonzero $q\in A$, so every
$q\ne 0$ has an inverse $q^{-1} = \bar q/(q\bar q)$. If $(a,b) = 1$, then there is some
$(x,y,z,w)\ne (0,0,0,0)$
such that $x^2 - ay^2 - bz^2 +ab w^2 = 0$. Letting $q = x+yi+zj+wk$, we have $q\bar \ne 0$ and
$q\bar q = 0$, meaning that $q$ is a zero divisor and thus $A$ is not a division algebra. This settles
the first part of the question.

In the case that $(a,b)=1$, there is a nonzero vector $(x,y,z)\in \QQ_p$ such that $z^2 = ax^2 + by^2$.
We must have $z\ne 0$ and at most one of $x$ and $y$ is zero. Without loss of generality, suppose that
$x\ne 0$. Defining
$$A = {1\over x}\pmatrix{z & -by \cr y & -z}\qquad\hbox{and}\qquad B = {1\over x}\pmatrix{0 & b \cr 1 & 0},$$
we have
$$A^2 = {1\over x^2}\pmatrix{z^2 - by^2 & -zby + zby \cr zy-zy & -by^2+z^2} = \pmatrix{a & 0 \cr 0 & a} = aI$$
and $B^2 = bI$, while
$$AB = {1\over x}\pmatrix{z&-by\cr y&z}\pmatrix{0&b\cr 1&0}
={1\over x}\pmatrix{-by&bz\cr -z&by}
= - {1\over x}\pmatrix{0&b\cr 1&0}\pmatrix{z&-by\cr y&z} = -BA,$$
so that the map sending $i\mapsto A$, $j\mapsto B$, is an isomorphism of $\QQ_p$-algebras between
$A$ and this matrix algebra. The proof concludes by noting that the span of $\{I, A, xB, xAB\}$ is all
of $\M_2(\QQ_p)$, since the matrix
$$\pmatrix{1&0&0&1\cr 0&b&1&0\cr z&-by&y&-z\cr -by&bz&-z&by}$$
has determinant $4b(by^2-z^2) = -4ax^2b\ne 0$. In particular, the set $\{I,A,B,AB\}$ spans all of $\M_2(\QQ_p)$
as well.\slug

% \boldlabel Hensel's lemma. We define the {\it $p$-adic valuation} $v_p : \QQ_p\to \ZZ \cup \{\infty\}$ by
% $$v_p(x) = \cases{ n,& if $x = p^n u$, where $u\in (\ZZ/p\ZZ)^\times$;\cr \infty,& if $x=0$.}$$
% The subring $\ZZ_p\subseteq \QQ_p$ is exactly the set of $x\in Q_p$ with $v_p(x) \in \NN\cup \{0,\infty\}$.
% The {\it $p$-adic absolute value} is given by $|x|_p = p^{-v_p(x)}$ and we define a metric space on $\QQ_p$
% by setting the distance function to $d_p(x,y) = |x-y|_p$. Under this metric, $\ZZ_p$ is complete,
% i.e., every Cauchy sequence has a limit in the space.
% 
% Given a polynomial $f(x_1, \ldots, x_m)$ with coefficients in $\ZZ_p$ and a solution to
% $f(x_1, \ldots, x_m) \equiv 0\pmod{p^n}$, we would like to be able to lift this solution to a solution
% with coefficients in $\ZZ_p$. We have the following lemma, which deals with polynomials in a single variable.
% 
% \proclaim Lemma G. Let $f$ be a polynomial in one variable with coefficients in $\ZZ_p$ and let $f'$
% be its derivative. Let $x\in \ZZ_p$ with $f(x)\equiv 0\pmod{p^n}$ and $v_p\bigl(f'(x)\bigr)=k$ for integers
% $n$ and $k$ satisfying $0\le 2k<n$. Then there is $y\in \ZZ_p$ such that
% \medskip
% \item{i)} $f(y)\equiv 0\pmod{p^{n+1}}$;
% \smallskip
% \item{ii)} $v_p\bigl(f'(y)\bigr) = k$; and
% \smallskip
% \item{iii)} $y\equiv x\pmod{p^{n-k}}$.
% \medskip
% 
% \proof We shall set $y=x+p^{n-k}z$ for some $z\in \ZZ_p$ to be chosen later. This certainly satisfies condition
% (iii). By Taylor's formula, we have
% $$f(y) = f(x) = p^{n-k}z f'(x) + p^{2n-2k}z^2 h(y)$$
% for some $h(y)\in \ZZ_p$. By hypothesis, $f(x) = p^nb$ for some $b\in \ZZ_p$ and $f'(x) = p^k c$ for some
% invertible element $c\in \ZZ_p$. So we can rewrite
% $$f(y) = p^n(b+zc) + p^{2n-2k}z^2 h(y),$$
% and by choosing $z$ such that $b+zc\equiv 0\pmod p$, we have condition (i), since $2n-2k>n$.
% Lastly, we apply Taylor's formula to $f'$ to get $f'(y)\equiv p^kc\pmod{p^{n-k}}$ and because $n-k>k$, we
% have condition (ii).\slug

\boldlabel Lemma Henselianum. {\it Ordo $p$-adicus} $v_p : \QQ_p\to \ZZ \cup \{\infty\}$ est functio
$$v_p(x) = \cases{ n,& si $x = p^n u$, ubi $u\in (\ZZ/p\ZZ)^\times$;\cr \infty,& si $x=0$.}$$
Subanulus $\ZZ_p\subseteq \QQ_p$ est copia elementorum $x\in \QQ_p$ cum $v_p(x) \in \NN\cup \{0,\infty\}$.
{\it Norma $p$-adica} est $|x|_p = p^{-v_p(x)}$ et definimus spatiam metricum super $\QQ_p$
cum functio distantiae $d_p(x,y) = |x-y|_p$. Sub illam functionem $\ZZ_p$ completum est;
i.e., omnis sequentia Cauchiana limitem in spatio habet.

Polynomium $f(x_1, \ldots, x_m)$ cum coefficientibus in $\ZZ_p$ et solutionem aequatione
$f(x_1, \ldots, x_m) \equiv 0\pmod{p^n}$ dantur, et volumus levare hanc solutionem ad solutionem
cum coefficientibus in $\ZZ_p$. Lemma sequens de polynomiis unae incognitae est.

\proclaim Lemma G. Sit $f$ polynomium unae incognitae cum coefficientibus in $\ZZ_p$ et sit $f'$
derivativum eius. Sit $x\in \ZZ_p$ cum $f(x)\equiv 0\pmod{p^n}$ et $v_p\bigl(f'(x)\bigr)=k$, ubi
$n$ et $k$ numeri integri sunt, et $0\le 2k<n$. Tum existit $y\in \ZZ_p$ satisfaciens
\medskip
\item{i)} $f(y)\equiv 0\pmod{p^{n+1}}$;
\smallskip
\item{ii)} $v_p\bigl(f'(y)\bigr) = k$; et
\smallskip
\item{iii)} $y\equiv x\pmod{p^{n-k}}$.
\medskip

\noindent{\it Demonstratio.}\enspace
Ponimus $y=x+p^{n-k}z$, ubi $z\in \ZZ_p$ eligetur postmodum. Hoc elementum $y$ postulationem
(iii) satisfacit. Ex formula Tayloriana sequitur
$$f(y) = f(x) = p^{n-k}z f'(x) + p^{2n-2k}z^2 h(y),$$
ubi $h(y)$ elementum anuli $\ZZ_p$ est. Ex hypothesi $f(x) = p^nb$ cum $b\in \ZZ_p$ et $f'(x) = p^k c$ ubi
$c\in \ZZ_p$ est elementum vertibile. Tum
$$f(y) = p^n(b+zc) + p^{2n-2k}z^2 h(y)$$
scribere possumus,
et eligentes $z$ cum $b+zc\equiv 0\pmod p$, habemus postulationem (i), quia $2n-2k>n$.
Ex formula Tayloriana cum $f'$ habemus $f'(y)\equiv p^kc\pmod{p^{n-k}}$ et cum $n-k>k$,
habemus propositionem (ii).\slug

Applicatio repetita huius lemmatis et completudo spatii lemma Henselianum producunt.

\proclaim Theorema H. Sit $f$ polynomium $m$ incognitarum cum coefficientibus in $\ZZ_p$. Sit
$x = (x_1, \ldots, x_m)\in (\ZZ_p)^m$ ut
$f(x)\equiv 0\pmod{p^n}$ et $v_p\bigl(f_j(x)\bigr) = k$,
ubi $1\le j\le m$, $k$ et $n$ sunt numeri integri satisfacientes
$0\le 2k< n$, et $f_j$ derivativum partiale
functionis $f$ est in incognita $j^{\hbox{\eightsl a}}$.
Tum existit $y\in (\ZZ_p)^m$ cum $f(y) = 0$ et satisfaciens $y\equiv x$ secundum modulum $p^{n-k}$.

\noindent{\it Demonstratio.}\enspace
Primo assumimus $m=1$. Ex Lemmate~G cum $x^{(0)} = x$ invenimus $x^{(1)}\in \ZZ_p$ ut
$x^{(0)}$ et $x^{(1)}$ congrui sunt secundum modulum $p^{n-k}$ et
satisfaciens $f\bigl(x^{(1)}\bigr)\equiv 0\pmod{p^{n+1}}$ et $v_p\bigl(f'(x^{(1)})\bigr)=k$.
Tum applicamus Lemma~G cum $x^{(1)}$. Sic inductiva mente construimus sequentiam $x^{(0)},x^{(1)},\ldots$
ubi indici generali $q$
$$x^{(q+1)}\equiv x^{(q)}\pmod{p^{n+q-k}}\qquad\hbox{et}\qquad f\bigl(x^{(q)}\bigr)\equiv 0\pmod{p^{n+q}}$$
habemus.
Haec sequentia Cauchiana est, quia omni integri $q,r>0$ habemus
$d_p\bigl(x^{(q+r)} - x^{(q)}\bigr) \le p^{-n-q+k}$.
Tum sequentia verge ad limitem $y\in \ZZ_p$ satisfaciens $f(y) = 0$ et $y\equiv x\pmod{p^{n-k}}$.

Conditio indicis $j$ casum generale ad casum $m=1$ reducit.
Consideramus
polynomium $\tilde f$ unae incognitae cum formula
$$\tilde f(x) = f(x_1, \ldots, x_{j-1}, x, x_{j+1}, \ldots, x_m).$$
Deinde casum $m=1$
propositionis applicare possumus cum $\tilde f$;
ex illo invenimus $y_j\in \ZZ_p$ satisfaciens $y_j\equiv x_j\pmod{p^{n-k}}$ et
$\tilde f(y_j) = 0$. Ponimus $y_i = x_i$ omni indici $i\ne j$ et elementum $y = (y_i)$ est
solutio desiderata.\slug

\nineproclaim
Exercise 4. Let $p$ be an odd prime.
Show that the quadratic form $ax^2 + by^2 + cz^2$ with coefficients in $\QQ_p$,
in which $a$, $b$, and $c$ belong to $\ZZ_p^\times$, has a nontrivial zero, i.e., the associated quadratic
space over $\QQ_p$ is not anisotropic.

\nineproof (Davide Accadia and Niccol\`o Bosio.)
First, we show that the equation $ax^2 + by^2 + cz^2 = 0$ has a nonzero solution in $\ZZ/p\ZZ$.
We pick $z\in \ZZ/p\ZZ$ so that $c' = -cz^2 \ne 0$, reducing our problem to finding a solution $(x,y)$
to the equation $ax^2 = c'-by^2$. Since $a$, $b$, and $c$ are all nonzero in $\ZZ/p\ZZ$, there
are $(p-1)/2 + 1$ elements in $\ZZ/p\ZZ$ of the form
$ax^2$ and $(p-1)/2+1$ elements of the form $c'-by^2$, so these two sets must intersect in at least one
element, giving us a nontrivial solution $(x,y,z)$ to the equation.

It now remains to apply Theorem H to this solution $(x,y,z)$ with $m=3$, $n=1$, and $k=0$. (The gradient
vector of $f$ is $(2x,2y,2z)$, one of whose components must have $p$-adic valuation equal to $0$
since $(x,y,z)\ne(0,0,0)$ and $p\ne 2$.)\slug

\boldlabel Integral lattices. We now turn to quadratic forms defined over the integers $\ZZ$. A
{\it unimodular lattice}
is a free abelian group of rank $n$ with a symmetric bilinear form $x\cdot y$ such that
\medskip
\item{i)} The homomorphism $L\to \Hom(L,\ZZ)$ that sends $x\mapsto (y\mapsto x\cdot y)$ is an isomorphism.
\smallskip
\item{ii)} If $(e_i)$ is a basis of $L$ over $\ZZ$, then the determinant of the matrix $(e_i\cdot e_j)$
is $\pm 1$.
\medskip
The set $\Hom(L,\ZZ)$, also denoted $L^\vee$,
is called the {\it dual} of $L$, and the condition above explains why
unimodular lattices are sometimes called self-dual. An element $x\in L$ can be identified with an element
of $L^\vee$ if $x\cdot y$ is integer for all $y\in L$, and this is true for all $x$ in a unimodular lattice.
For any ring $S$ admitting a homomorphism $\ZZ\to S$, we obtain an $S$-module $L\otimes S$ by extending
the scalars from $\ZZ$ to $S$. Two lattices $L_1$ and $L_2$ are said to be {\it locally isomorphic}
if $L_1\otimes \ZZ_p\cong L_2\otimes \ZZ_p$ for all primes $p$ and $L_1\otimes \ZZ_p\cong L_2\otimes \ZZ_p$.
The set of lattices that are locally isomorphic to a given lattice $L$ is called the {\it genus} of $L$.
We saw in class that two lattices in the same genus must have the same discriminant.
Since $V = L\otimes \RR$ is a quadratic space over $\RR$, it has a well
defined signature $(r,s)$. As in the real case, we say that $L$ is {\it positive definite} if $s=0$,
{\it negative definite} if $r=0$, and {\it indefinite} otherwise. Given a quadratic module $V$, we will sometimes
denote the corresponding quadratic space over a different ring $R$ by $V_R$.

We say that a lattice $L$ is {\it even} or {\it of type {\rm II}} if the quadratic form associated to $L$
takes only even values, and we say that $L$ is {\it odd} or {\it of type {\rm I}} otherwise. We saw in
class that there is an element $u\in L$, unique modulo reduction modulo $2$, such that
$u\cdot x = x\cdot x \pmod 2$ for all $x\in L$. The image of $u\cdot u$ in $\ZZ/8\ZZ$ is an invariant of
$L$, denoted $\sigma(L)$. If $L$ is even, then $\sigma(L) = 0$.

Let $\langle 1\rangle$ denote the $1$-dimensional quadratic space with $Q(x)=x^2$ and let $\langle-1\rangle$
denote the quadratic space with $Q(x) = -x^2$ (their bilinear forms are $x\cdot y = xy$ and $x\cdot y = -xy$
respectively).
Note that unless $r$ and $s$ are both zero, the direct sum
$\langle 1\rangle^r \ortho \langle -1\rangle^s$ is an odd lattice. In class we saw the following structure
theorem for indefinite unimodular lattices.

\proclaim Theorem S. Let $L$ be a unimodular indefinite lattice of signature $(r,s)$.
If $L$ is odd, then $L\cong \langle 1\rangle^r \ortho \langle -1\rangle^s$. If $L$ is even, then
$r-s \equiv 0\pmod 8$ and there is only one lattice in this case as well, up to isomorphism.\slug

In the solution to the next exercise, we will also use the following lemma.

\proclaim Lemma P. Let $L$ and $L'$ be $\ZZ_p$-lattices of discriminant $d$ with pairing matrices $A$
and $A'$ respectively. Let $\lambda = 1$ if $p$ is odd and $3$ if $p=2$.
If there exists $T\in \M_n(\ZZ_p)$ such that $T^\T AT = A'\pmod {p^\lambda}$, then there is $X$ in $\M_n(\ZZ_p)$
such that $X^\T A X = A'$.\slug

In most of the course, we have assumed that $p\ne 2$. But for the next exercise, the definition of a
genus of a lattice requires that we consider $\ZZ_2$-lattices. We will thus require the following lemma,
whose proof can be found in Serre (1973).

\proclaim Lemma E. An element $x\in \QQ_2^\times$ is a square if and only if $x$ can be written as $2^n u$ where
$n$ is even and $u\equiv 1\pmod 8$.\slug

\nineproclaim
Exercise 5. Show that all even unimodular lattices of a given signature $(r,s)$ are in the same
genus. Show that all odd unimodular lattices are in the same genus. Give an example of two quadratic forms
of the same discriminant that lie in different genera.

\nineproof (Mart\'\i\ Roset.)
Theorem S allows us to consider only the definite case, and without loss of generality, we can further assume
that both lattices are positive definite.
Suppose that $L_1$ and $L_2$ are in the same genus. Then $L_1\otimes \ZZ_2 \cong L_2\otimes \ZZ_2$
and we find that $L_1\otimes \FF_2\cong L_2\otimes \FF_2$. In these lattices over $\FF_2$, either
all vectors have zero length, in which case $L_1$ and $L_2$ were both even, or some vector has nonzero length,
in which case both $L_1$ and $L_2$ were odd.

Now for the other direction of the proof, suppose that $L_1$ and $L_2$ are unimodular positive
definite integral lattices. Since $L_i \ortho \langle -1\rangle$ is odd, unimodular, and indefinite,
it is isomorphic to $L'=\langle 1\rangle^n \ortho \langle -1\rangle$, which has $\disc(L')=-1$ and
$\sigma(L') \equiv n-1\pmod 8$. The discriminant is multiplicative and the $\sigma$-invariant additive
under orthogonal sum, so $\disc(L_i) = 1$ and $\sigma(L_i) \equiv n\pmod 8$.
Furthermore, $L_1\otimes \RR \cong \langle 1\rangle ^n \cong L_2\otimes \RR$, so it remains to show
that $L_1\otimes \ZZ_p \cong L_2\otimes \ZZ_p$ for all primes $p$.

When $p$ is odd, we see that $L_1\otimes \FF_p$ and $L_2\otimes \FF_p$ have the same rank and discriminant,
so by Lemma~P we find that in fact $L_1\otimes \ZZ_p \cong L_2\otimes \ZZ_p$. In the case $p=2$, we introduce
some new notation. Let $\langle d\rangle$ denote the quadratic form of rank $1$ given by $Q(x) = dx^2$. We
will also abuse notation and denote a quadratic form by its pairing matrix.
We then use the fact that any unimodular $\ZZ_2$-lattice is the orthogonal sum of copies of
$$\langle 1\rangle,\quad\langle 3\rangle,\quad\langle 5\rangle,\quad\langle 7\rangle,\quad\pmatrix{0&1\cr 1&0},
\quad\hbox{and}\quad
\pmatrix{2&1\cr 1&2}.$$
We have the relations
\medskip
\item{i)} $\langle 1\rangle^2 \cong \langle 5\rangle^2$ and $\langle 3\rangle^2\cong \langle 7\rangle^2$;
\smallskip
\item{ii)} $\langle 3\rangle \ortho \langle 5\rangle \ortho \langle 7\rangle \cong
\langle 1\rangle \ortho \langle 3\rangle^2$;
\smallskip
\item{iii)} $\langle 1\rangle^4 \cong \langle 7\rangle^4$;
\smallskip
\item{iv)} $\langle d\rangle \ortho A \cong \langle d\rangle \ortho \langle 1\rangle
\ortho \langle -1\rangle$ for all
$d \in \{1,3,5,7\}$ and $A\in \bigl\{ \smallmatrix 0110,\smallmatrix 2112\bigr\}$; and
\smallskip
\item{v)} $\smallmatrix 0110^2 \cong \smallmatrix 2112^2$.
\medskip
Now consider $L_i\otimes \ZZ_2$. Since $L_i$ is even, it is an orthogonal sum of copies of
$\smallmatrix 0110$ and $\smallmatrix 2112$, since any copies of $\langle d\rangle$ would cause the lattice
to be odd. The discriminant of $L_i\otimes \ZZ_2$ is $1$ and the discriminants of $\smallmatrix 0110$
and $\smallmatrix 2112$ are $-1$ and $3$ respectively, so we find that
$$L_i\otimes \ZZ_2 \cong \pmatrix{0&1\cr 1&0}^{r_1} \ortho \pmatrix{2&1\cr 1&2}^{r_2}$$
with both $r_1$ and $r_2$ even, so by property (v) above we have
$L_i\otimes \ZZ_2 \cong \smallmatrix 0110 ^{r_1+r_2}$.
If the $L_i$ are odd, we have
$$L_i\otimes \ZZ_2 \cong \langle 1\rangle^{r_1} \ortho\langle 3\rangle^{r_3}
\ortho\langle 5\rangle^{r_5} \ortho\langle 7\rangle^{r_7}.$$
Since the discriminant is $1$, either $r_3$, $r_5$, and $r_7$ are all even or they are all odd.
If they are all odd, we the relation (ii) to make them all even (increasing $r_1$ in the process).
Then we can use the relations in (i) to see that
$$L_i\otimes \ZZ_2\cong \langle 1\rangle ^{r_1} \ortho \langle 7\rangle ^{r_7},$$
for some $r_1$ and $r_7$ possibly different from above. Since $L_i$ is positive definite of rank $r_1+r_7$,
$\sigma(L_i)$ must be congruent to $r_1+r_7$ modulo $8$, but on the other hand, from the right-hand
side we must have $r_1+r_7\equiv r_1+7r_7\pmod 8$. This means that $r_7$ is either $0$ or $4$ modulo $8$,
and we can use relation (iii) to find that $L_1\otimes \ZZ_2 \cong \langle 1\rangle^n\cong L_2\otimes \ZZ_2$.

To get lattices with the same discriminant lying in different genera, take an even lattice of discriminant
$1$, say the space $E_8$ we constructed in class, and an odd lattice of discriminant $1$ of the same rank (we
can take $\langle 1\rangle ^8$ as the example corresponding to $E_8$).
\slug

\boldlabel Modular forms and theta series. Let $\H$ denote the complex upper half-plane; that is, the set
of $z\in \CC$ with $\Im z > 0$.
Given an lattice $L$ of rank $n$, we define the {\it theta series}
of $L$ to be the sum
$$\theta_L(z) = \sum_{v\in L} e^{\pi i (v\cdot v) z},$$
where $z\in \H$.
It is easy to see that $\theta_L(z+2) = \theta_L(z)$, and when $L$ is unimodular, one can also show that
$$\theta_L\Bigl({-1\over z}\Bigr) = \Bigl({z\over i}\Bigr)^{n/2} \theta_L(z).$$
If $L$ is even, then $v\cdot v$ is always even, so we actually have $\theta_L(z+1) = \theta_L(z)$, and
we saw in class that the dimension $n$ of an even unimodular lattice is always a multiple of $8$, so
the factor of $1/\sqrt i$ disappears and we have $\theta_L(-1/z) = z^{n/2} \theta_L(z)$. The significance
of this becomes apparent when we note that the group $\SL_2(\ZZ)$ of integral matrices with determinant $1$
acts by M\"obius transformations on the upper half-plane; for an element $g = \smallmatrix abcd\in \SL_2(\ZZ)$
and $z\in \H$, we have
$$g*z = {az+b\over cz+d}.$$
Since the matrices
$$\pmatrix{1&1\cr 0&1}\qquad\hbox{and}\qquad\pmatrix{0&-1\cr 1&0}$$
generate $\SL_2(\ZZ)$, the formulas above show that the theta series of an even unimodular lattice
is invariant under the action of $\SL_2(\ZZ)$, up to a factor of $z^{n/2}$.

A {\it modular form of weight $k$} on a subgroup $\Gamma\subseteq \SL_2(\ZZ)$ (which may be
$\SL_2(\ZZ)$ itself)
is a holomorphic function $f:\H\to \CC$ satisfying
$$f\Bigl({az+b\over cz+d}\Bigr) = (cz+d)^k f(z)$$
and with an expansion
$$f(z) = \sum_{n=0}^\infty a_n e^{2\pi i nz},$$
for some $a_n\in \CC$. The space of all such functions is denoted $\calM_k(\Gamma)$.

\nineproclaim Exercise 6. This exercise deals with odd unimodular lattices, which were largely left out of
our discussion in class. In the following, $q = e^{2\pi i z}$ and $y = \Im z$.
\medskip
\item{a)} Let $L$ be an odd unimodular lattice of rank $2k$. Show that the theta series $\theta_L(z)$
is invariant under the group $\Gamma(2)$ consisting of matrices in $\SL_2(\ZZ)$ that are congruent to the
identity modulo $2$.
\smallskip
\item{b)} To avoid having theta series with fractional powers of $q$, it is useful to redefine $\theta_L(q)$
by the rules
$$\theta_L(z) = \sum_{v\in L} e^{2\pi i(v\cdot v)z}\qquad\hbox{and}\qquad
\theta_L(q) = \sum_{v\in L} q^{v\cdot v} = \sum_{n=0}^\infty r_L(n)q^n,$$
where $r_L(n)$ denotes the number of vectors $v\in L$ with $v\cdot v = n$. Show that $\theta_L(q)$ is a
modular form of weight $k$ on the subgroup $\Gamma_0(4)$ consisting of matrices in $\SL_2(\ZZ)$ that are
upper triangular modulo $4$.
\smallskip
\item{c)} Although the Eisenstein series $E_2$ defined by
$$E_2(q) = 1-24\sum_{n=1}^\infty \sigma_1(n)q^n,$$
where $\sigma_1(n) = \sum_{d\divides n} d$, fails to be invariant under the action of $\SL_2(\ZZ)$, the
modification
$$E_2^*(z) = E_2(z) - {3\over \pi y}$$
is a modular form of weight $2$, though no longer holomorphic. Use this fact to show that the series
$$E_2^{(2)} = E_2(z) - 2E_2(2z) = E_2(q) - 2E_2(q^2)$$
and
$$E_2^{(4)} = E_2(z) - 4E_2(2z) = E_2(q) - 2E_2(q^4)$$
are (holomorphic) modular forms of weight $2$ on $\Gamma_0(4)$.
\smallskip
\item{d)} Show that any modular form of weight $k$ on $\Gamma_0(4)$ has exactly $k/2$ zeroes on any fundamental
region. Use this to conclude that $\calM_2\bigl(\Gamma_0(4)\bigr)$ is $2$-dimensional, and thus spanned by
the two Eisenstein series $E_2^{(2)}$ and $E_2^{(4)}$.
\smallskip
\item{e)} Use the result of (d) to calculate the number of vectors of odd length in any odd unimodular
quaternary quadratic form.
\smallskip
\item{f)} Write the theta series attached to the standard quaternary lattice $\ZZ^4$ with the standard dot
product, and the theta series attached to the lattice
$$D_4 = \Big\{ (a,b,c,d) \in \ZZ^4 \cup  \Bigl(\ZZ+{1\over 2}\Bigr)^4
: a+b+c+d\in 2\ZZ\Big\}.$$
Deduce a closed form expression for the number of vectors of a given
length in each of these two lattices.
\medskip

\nineproof For part (a), we first show that $\Gamma(2)$ is generated by the three matrices
$$\pmatrix{1&2\cr 0&1},\quad\pmatrix{1&0\cr 2&1},\quad\hbox{and}\quad\pmatrix{-1&0\cr 0&-1}$$
To see this, note that for a general matrix $\smallmatrix abcd\in \Gamma(2)$,
$$\pmatrix{a&b\cr c&d}\pmatrix{1&-2\cr 0&1} = \pmatrix{a & b-2a\cr c&d-2c}$$
and
$$\pmatrix{a&b\cr c&d}\pmatrix{1&0\cr -2&1} = \pmatrix{a-2b & b\cr c-2d&d}.$$
Suppose that $b\ne 0$.
Since $|a|$ is odd and $|b|$ is even, they are not equal. if $|a|$ is larger, we use the division algorithm
to find $q,r$ such that $|a| = |2b|q+r$, where $|r| < |b|$, and then apply the second transformation $q$
times to strictly reduce the absolute value of the top-left matrix entry. If $|b|$ is larger, we reduce
the absolute value of the top-right entry in a similar fashion. We can keep doing this until $b=0$, in which
case the matrix must be some integer power of $\smallmatrix 1021$, after possibly multiplying by
$\smallmatrix {-1}00{-1}$.

Thus it suffices to prove invariance under the three generators of $\Gamma(2)$. The negative identity matrix
acts as the identity M\"obius transformation, and we already saw that theta series are invariant under
the transformation $z\mapsto z+2$. The last generator is $\smallmatrix 1021$, which also poses no problem,
since
$$\theta_L\Bigl({z\over 2z+1}\Bigr) = \Bigl({i\over z}\Bigr)^k \theta_L\Bigl({-2z-1\over z}\Bigr)
= \Bigl({i\over z}\Bigr)^k \theta_L\Bigl({-1\over z}\Bigr) =
\Bigl({i\over z}\Bigr)^k\Bigl({z\over i}\Bigr)^k\theta_L(z) = \theta_L(z).$$

We start part (b) by claiming that $\Gamma(2)$ and $\Gamma_0(4)$ are conjugate in $\SL_2(\QQ)$,
by the element $\bigl( {2\atop 0}{0\atop 1}\bigr)$. Indeed,
for any $\gamma = \smallmatrix abcd \in \SL_2(\ZZ)$
$$\pmatrix{ 1/2 & 0\cr 0&1} \pmatrix {a&b\cr c&d}\pmatrix{2&0\cr 0&1} = 
\pmatrix{ a/2 & b/2\cr 0&1}\pmatrix{2&0\cr 0&1} = \pmatrix{a&b/2\cr 2c&d},$$
and if $\gamma\in \Gamma(2)$ to begin with, then $2c$ is a multiple of $4$ so the result is in
$\Gamma_0(4)$. On the other hand,
$$\pmatrix{ 2 & 0\cr 0&1} \pmatrix {a&b\cr c&d}\pmatrix{1/2&0\cr 0&1} = 
\pmatrix{ 2 & b\cr 0&1}\pmatrix{1/2&0\cr 0&1} = \pmatrix{a&b\cr c/2&d}.$$
In this case, if $\gamma\in \Gamma_0(4)$, then $bc$ is even so $a$ and $d$ must be odd for
$ad-bc$ to equal $1$. Since $c$ was a multiple of $4$, we have $c/2$ even and of course,
so is $2b$, so the result is in $\Gamma(2)$.
This gives a set of generators for $\Gamma_0(4)$; since
$$\pmatrix{ 1/2 & 0\cr 0&1} \pmatrix {1&2\cr 0&1}\pmatrix{2&0\cr 0&1} = \pmatrix{ 1&1\cr 0&1}$$
and
$$\pmatrix{ 1/2 & 0\cr 0&1} \pmatrix {1&0\cr 2&1}\pmatrix{2&0\cr 0&1} = \pmatrix{ 1&0\cr 4&1},$$
the matrices
$$\pmatrix{1&1\cr 0&1},\quad\pmatrix{1&0\cr 4&1},\quad\hbox{and}\quad\pmatrix{-1&0\cr 0&-1}$$
comprise a set of generators for $\Gamma_0(4)$.

We shall call the modified theta series in part (b) $\theta'_L(z)$, to know when we have an extra factor
of two and when we do not. Invariance under $z\mapsto z+1$ is easy, since
$$\theta'_L(z+1) = \theta_L(2z+2) = \theta_L(2z) = \theta'_L(z).$$
For the other nontrivial transformation, we first note that
$$\theta'_L\Bigl({-1\over 4z}\Bigr) = \theta_L\Bigl({-1\over 2z}\Bigr) =
\Bigl({2z\over i}\Bigr)^k\theta_L(2z) = \Bigl({2z\over i}\Bigr)^k\theta'_L(z).$$
We now have all we need to show that $\theta'_L(z)$ is a modular form of weight $k$ on $\Gamma_0(4)$, because
for the M\"obius transformation $z\mapsto z/(4z+1)$ we have
$$\eqalign{
\theta'_L\Bigl({z\over 4z+1}\Bigr) &= \theta'_L\biggl({-1\over 4\bigl(-1/(4z) - 1\bigr)}\biggr) \cr
&= \biggl(2i \Bigl({1\over 4z}+1\Bigr)\biggr)^k \theta'_L\Bigl({-1\over 4z} -1\Bigr) \cr
&= \biggl(2i \Bigl({1\over 4z}+1\Bigr)\biggr)^k \theta'_L\Bigl({-1\over 4z}\Bigr) \cr
&= \biggl(2i \Bigl({2z\over i}\Bigr)\Bigl({1\over 4z}+1\Bigr)\biggr)^k \theta'_L(z)\cr
&= (4z+1)^k \theta'_L(z).\cr
}$$

For part (c), we begin by noting that since
$$E_2(z) - N E_2(Nz) = E_2(z) - {3\over \pi y} - N E_2(Nz) - N{3\over N\pi y} = E_2^*(z) - NE_2^*(Nz),$$
for any $N\ge 2$, in particular both $E_2^{(2)}$ and $E_2^{(4)}$ are holomorphic and
we are done if we can show that $E_2^*(2z)$ and $E_2^*(4z)$ are
modular forms of weight $2$ on $\Gamma_0(4)$. In fact, for all $N\ge 2$
we shall show that if $g(z)$ is a modular form of weight $2$ on $\SL_2(\ZZ)$, then $f(z) = g(Nz)$
is a modular form of weight $k$ on $\Gamma_0(N)$. (This would settle the question since $\Gamma_0(4)\subseteq
\Gamma_0(2)$ and any modular form on $\Gamma_0(2)$ is automatically a modular form on $\Gamma_0(4)$.)
Well, letting $\smallmatrix abcd\in \Gamma_0(N)$, we have
$$f\Bigl({az+b\over cz+d}\Bigr) = g\Bigl( N{az+b\over cz+d}\Bigr)
= g\Bigl({Naz+Nb\over cz+d}\Bigr) = g\Big( {a(Nz) + Nb\over (c/N)(Nz) + d} \Bigr).$$
But since $c$ is a multiple of $N$, the matrix $\smallmatrix a{Nb}{c/N}d$ has integral entries and has
determinant $ad - (Nb)(c/N) = ad-bc= 1$. So $g$ is weight-$2$ invariant under its action, and
$$f\Bigl({az+b\over cz+d}\Bigr) = g\Big({a(Nz) + Nb\over (c/N)(Nz) + d} \Bigr)
= \bigl((c/N)(Nz)+d\bigr)^2 g(Nz) = (cz+d)^2f(z).$$

To begin part (d), we first show that $\Gamma_0(4)$ is a subgroup of
index $6$. We already found that $\Gamma(2)$ and $\Gamma_0(4)$ are conjugate subgroups of $\SL_2(\QQ)$, which
means they have the same index in $\SL_2(\ZZ)$. Furthermore, we a homomorphism from $\SL_2(\ZZ)$ to
$\SL_2(\ZZ/2\ZZ)$ given by reduction of entries modulo $2$, giving the short exact sequence
$$1\longto\Gamma(2)\longto\SL_2 (\ZZ)\longto \SL_2(\ZZ/2\ZZ)\longto 1.$$
An element $\smallmatrix abcd$ of $\SL_2(\ZZ/2\ZZ)$ must have $ad-bc = 1$ in $\ZZ/2\ZZ$,
so either $ad =1$ and $bc=0$ or vice versa. In each case there are three ways to make a product equal to $0$,
so the cardinality of $\SL_2(\ZZ/2\ZZ)$ is $6$.
We saw in class that any fundamental domain of the action of $\SL_2(\ZZ)$ on the upper half-plane has
$k/12$ zeroes of a modular form (counting fractional zeroes on the boundary). So let $F$ be a modular form
of weight $12$ on $\SL_2(\ZZ)$ with its single zero at, say, $2i$ (chosen because it is right in the middle
of the usual fundamental domain for $\SL_2(\ZZ)\backslash \H$). Since $\Gamma_0(4)$ is an index-$6$ subgroup
of $\SL_2(\ZZ)$, we know that $F$ has $6$ zeroes on $\Gamma_0(4)\backslash \H$ and $F^k$ has $6k$ zeroes
on the same region. Now let $g\in \calM_2\bigl(
\Gamma_0(4)\bigr)$ be given, with $m$ zeroes on $\Gamma_0(4)\backslash \H$. We want to show that $m=k/2$.
Well, whatever $m$ is, we know that $g^{12}$ has $12m$ zeroes. Now we consider
the function $g^{12}/F^k$, which is meromorphic on the compact Riemann surface $\Gamma_0(4)\backslash \H$
and thus has as many zeroes as poles. The number of zeroes it has is $12m$ and the number of poles it
has is $6k$. So $m= k/2$, which is what we wanted to show.

In particular, any modular form of weight $2$ on $\Gamma_0(4)$ has exactly one zero.
Evaluating $E_2(q)$ at $q=0$ (equivalent to evaluating $E_2(z)$ at $z=i\infty$) gives a value of $1$,
so $E_2^{(2)}(0) = -1$ and $E_2^{(4)}(0) = -3$. From here we see that $F(z) = 3E_2^{(2)}(z) - E_2^{(4)}(z)$ has
a zero at $i\infty$, meaning that it has no zeroes on the upper half-plane $\H$. So, given
any modular form $f$ on $\Gamma_0(4)$ of weight $2$, we can subtract a multiple of $E_2^{(2)}$ to get
a form that has a zero at $i\infty$, and then divide by $F$ to get a modular form of weight $0$, which must
be a constant. Symbolically, this amounts to showing that any $f$ satisfies
$$ {f-\lambda E_2^{(2)}\over F} = \mu$$
for some $\lambda,\mu\in \CC$; that is, $f = \lambda E_2^{(2)} + \mu F$.
Since $E_2^{(2)}$ and $E_2^{(4)}$ are linearly independent,
we have shown that $\calM_2\bigl(\Gamma_0(4)\bigr)$ is $2$-dimensional, bringing us to the end of part (d).

For part (e), we have the general formula
$$\sigma_1(n/N) - N\sigma_1(n/N) = \sum_{d\divides n} d - \sum_{d\divides (n/N)} Nd = \sum_{d\divides n}d -
\sum_{N\divides d\divides n}d = \sum_{N\notdivides d\divides n}d.$$
We saw that for
any lattice $L$ of rank $2k$, $\theta_L'(q)$ is a modular form of weight $k$ on $\Gamma_0(4)$, so
by the result of the previous section, we can write $\theta_L' = \lambda E_2^{(2)} + \mu E_2^{(4)}$
and by equating coefficients, we find that the number of vectors of length $n\ge 1$ in $L$ is
$$r_L(n) = \lambda \sum_{2\notdivides d\divides n} d + \mu\sum_{4\notdivides d\divides n} d.$$
We can determine $\lambda$ and $\mu$ from $r_L(0)$ and $r_L(1)$.
Since the constant terms of $E_2^{(2)}$ and $E_2^{(4)}$ are $-1$ and $-3$ respectively, we have
$r_L(0) = -\lambda - 3\mu$. So we have the simultaneous equations
$$ \pmatrix{r_L(0)\cr r_L(1)} = \pmatrix{ -1&-3\cr -24 &-24} \pmatrix{\lambda\cr \mu},$$
which we can solve to get
$$ \lambda = {r_L(0)\over 2} - {r_L(1)\over 16} \qquad\hbox{and}\qquad
\mu = {r_L(1)\over 48}-{r_L(0)\over 2}.$$

Finally, we have come to part (f).
We start by noting that both of these lattices have exactly one vector of length $0$. There are $8$ ways
to make a vector of length $1$ in $\ZZ^4$, since there are four slots to put either a $1$ or $-1$.
There are also $8$ ways to make a vector of length $1$ in $D_4$, since the $16$ vectors
of the form $(\pm 1/2, \pm 1/2, \pm 1/2, \pm 1/2)$ have norm $1$, but only half of them have an even number
of $-1/2$s, which is necesssary for the sum of the coordinates to be even. Invoking part (e) now, we
find that $\lambda = 0$ and $\mu = -1/3$. Hence both lattices have theta series $-E_2^{(4)}/3$.
\slug

\nineproclaim Exercise 7. Let $G$ be a group acting transitvely on a set $X$. Let $x_0\in X$ and let $S$
be a subset of $G$ such that for all $x\in X$, there exists $g\in S$ with $gx_0 = x$. Show
that $G$ is generated by $S$ together with the stabiliser of $x_0$.

\nineproof Let $g\in G$. There exists $x\in X$ (namely, $g^{-1}x$) such that $gx = x_0$.
Now by the property of
$S$, there is $s\in S$ with $sx_0 = x$. So we see that $gs x_0 = x_0$, meaning that $gs\in \Stab(x_0)$.
But this means that $g = gs \cdot s^{-1}$, and since $G$ is closed under inverses,
$G = G^{-1} = \bigl(\Stab(x_0) S^{-1}\bigr)^{-1} = S\Stab(x_0)$.
\slug

\boldlabel The projective line. For the next exercise, we define the {\it projective line} $\PP^1(k)$
over a field
$k$ to be the set $k^2\setminus\bigl\{(0,0)\bigr\}$
quotiented by the equivalence relation $\sim$ that deems
$(x,y)\sim (\lambda x,\lambda y)$ for any nonzero scalar $\lambda\in k$. We denote the equivalence class
of $(x,y)$ by $[x:y]$. If $y\ne 0$, we see that $[x:y] = [z:1]$ for some $z\in k$. The element $[1:0]$
is the only element that cannot be written in this way; it is called {\it the point at infinity}.

\nineproclaim Exercise 8. Let $k$ be a field. Show that $\SL_2(k)$ is generated by matrices of the form
$$\pmatrix{ 1&t\cr 0&1}\qquad\hbox{and}\qquad \pmatrix{a&0\cr 0&a^{-1}}$$
for $t\in K$ and $a\in K^\times$, along with $w = \smallmatrix 01{-1}0$. More precisely, show that
$\SL_2(k) = B \sqcup BwB$, where $B$ is the subgroup of upper triangular matrices. This is known as
the {\it Bruhat decomposition} of $\SL_2(k)$.

\nineproof The group $\SL_2(k)$ acts on the set $\PP^1(k)$ by matrix multiplication;
that is,
$$\pmatrix{a&b\cr c&d} [x:y] = [ax+by : cz+dy].$$
To see that this action is transitive, note that $\GL_2(k)$ is transitive on the set of nonzero vectors in
$k^2$, and scaling the matrix to insist that it has determinant $1$, we do not change its output in
the projective line. Note that
$$Bw[1:0] = \biggl\{ \pmatrix{a&b\cr 0&d}\pmatrix{0&1\cr -1&0} : a,b,d\in k\biggr\} =
\bigl\{ [-b: -d] : b,d\in k\bigr\}.$$
Since $ad=1$, $d$ cannot be zero, and we can divide out by $-d$ to find that this is the set
$\bigl\{[z:1] : z\in k\bigr\}$.
In particular, the identity matrix $I$ is not a member of $Bw$, but
for any $p\in \PP^1(k)$, there exists $\gamma\in S =  Bw \sqcup \{I\}$ such that
$\gamma [1:0] = p$. Note also that any upper triangular matrix fixes the point $[1:0]$, and if any
$\gamma\in \SL_2(\ZZ)$ fixes $[1:0]$, then its bottom-left entry must be zero, so $\Stab\bigl([1:0]\bigr) = B$.
Applying the result of Exercise 7 with $x_0 = [1:0]$ gives us
$$ \SL_2(k) = S\cdot \Stab\bigl([1:0]\bigr) = \bigl(Bw\sqcup \{I\}\bigr) B = BwB\sqcup B,$$
which is what we wanted.
\slug

\nineproclaim Exercise 9. Show that every element of $\SL_2(\RR)$ can be written uniquely in the form
$$\pmatrix{1&x\cr 0&1}\pmatrix{y^{1/2}&0\cr 0&y^{-1/2}}\pmatrix{\cos\theta&\sin\theta\cr -\sin\theta&
\cos\theta}$$
for $x\in \RR$, $y\in \RR^{>0}$, and $\theta\in [0,2\pi)$. This is known as the {\it Iwasawa
decomposition} of $\SL_2(\RR)$.

\nineproof
The set $\SL_2(\RR)$ acts on the upper half-plane $\H$ by M\"obius transformations.
We let
$$ S = \biggl\{ \pmatrix{1 & x \cr 0&1}\pmatrix{y^{1/2}&0\cr 0&y^{-1/2}} : x\in \RR,\,y>0 \biggr\},$$
and note that
$$ \pmatrix{y^{1/2} & xy^{1/2} \cr 0 & y^{-1/2}} * i = {y^{1/2} i + xy^{1/2}\over y^{-1/2}} = x+iy,$$
so since $x\in \RR$ and $y>0$ we see that for any $z\in \H$, there is some element of $S$ such that $S*i = z$.
Now the stabiliser of $i$ is the set of all $\smallmatrix abcd\in SL_2(\RR)$ with
$$ {ai+b\over ci+d} = i.$$
This equation implies that $a=d$ and $b=-c$, which, along with the condition $ad-bc=1$, gives
$a^2+b^2 = 1$ and we see that
the set of all such matrices can be parametrised by letting $a = \cos\theta$ and
$b=\sin\theta$. We are now in the fortunate position to apply Exercise 7 once again.

We have shown that for any $\gamma = \smallmatrix abcd\in \SL_2(\ZZ)$, there are $x$, $y$, and $\theta$
such that
$$\pmatrix{1&x\cr 0&1}\pmatrix{y^{1/2}&0\cr 0&y^{-1/2}}\pmatrix{\cos\theta&\sin\theta\cr -\sin\theta&\cos\theta}
\pmatrix{a&b\cr c&d} = \pmatrix{1&0\cr 0&1}.$$
(For convenience, we construct $\gamma^{-1}$ instead of $\gamma$.) Given only
the triple $(x,y,\theta)$, we can work backwards to find the matrix $\smallmatrix abcd$ above.
Since the bottom-left entry on the right-hand side is zero,
we have $c\cos\theta = a\sin\theta$, whence $c = a\tan\theta$. In the next multiplication, we find that
$ya^2+yc^2 = 1$, so
$$a = {1\over \sqrt{ y(1+\tan\theta)}}\qquad\hbox{and}\qquad c = {\tan\theta\over \sqrt{ y(1+\tan\theta)}}.$$
Lastly, from the equations
$ab+cd = -x(a^2+c^2)$ and $ad-bc = 1$,
we find that
$$b={c-ax(a^2+c^2)\over a^2+c}\qquad\hbox{and}\qquad d = {1+bc\over a},$$
finishing the proof of uniqueness.
\slug

\boldlabel The Fourier transform. Let $f$ be a Schwartz function on a finite-dimensional vector space $V$
over $\RR$. (We will not concern ourselves with exactly what a Schwartz function is; one should think of it as
having nice decay properties at infinity.) The {\it Fourier transform} $\hat f$ of $f$ is the integral
$$ \hat f(y) = \int_V f(x) e^{-2\pi i (x\cdot y)}\d x.$$
The Fourier transform over other quadratic spaces is given similarly, though in these settings we have
to change what we mean by ``nice'' function. On $\QQ_p$, we can take compactly supported functions.
Of course, when we integrate over $\QQ_p$, we'll need to know what measure we are integrating against.
Luckily, there is a translation-invariant, countably additive measure $\mu$
on $\QQ_p$ called the {\it Haar measure},
which we shall use without worrying about the details of its construction. Now that we have a measure
(Lebesgue or Haar) on $V$, we can define the {\it covolume} of $L$ as the measure of a fundamental
region of $V/L$.

\medskip\boldlabel Additive characters. An {\it additive character} is a homomorphism from an abel\-ian
group $Z$ to the unit circle; that is, for $x,y\in Z$ we have $\chi(x+y) = \chi(x)\chi(y)$.
A character is said to be {\it trivial} if it assigns the value $1$ to
every member of the group. The following lemma is a simple consequence of definitions, but is extremely
useful.

\proclaim Lemma Z. Let $\chi$ be a nontrivial character on an additive
group $Z$. Then
$$\sum_{x\in G} \chi(x) = 0.$$

\proof Since $\chi$ is nontrivial, there must be some $x_0\in Z$
with $\chi(x)\ne 1$. Then writing
$$\sum_{x\in G} \chi(x) = \sum_{x\in G} \chi(x_0+x) = \chi(x_0) \sum_{x\in G} \chi(x),$$
we see that this sum must be zero.\slug

Before proceeding, we stop to prove a miscellaneous lemma about containment of
$\ZZ_p$-lattices. It will be useful in a couple of the exercises below.

\proclaim Lemma B. Let $L$ and $L'$ be $\ZZ_p$-lattices
of the same rank. There exists a positive integer $m$ such that $p^m L\subseteq L'$.

\proof Being free abelian groups, $L$ and $L'$ both have $\ZZ_p$-bases; call the matrices with these
bases as columns $B$ and $B'$ respectively.
These bases are also $\QQ_p$-bases for the associated quadratic space $V_{\QQ_p}$. The matrix
$X$ such that $B = XB'$ has finitely many entries, all of which are $p$-adic
rationals. Thus there is an $m$ for which $p^m X$ has entries in $\ZZ_p$. Now given a vector
$v = p^m w\in p^m L$, we can express it as a linear combination of vectors in $B'$ by multiplying it by
$X$ on the left. Since $p^mX$ and $w$ both have entries consisting entirely of $p$-adic integers,
$Xv = X p^m w = (p^m X) w$ also consists entirely of $p$-adic integers. Hence we can express $v$ as a
$\ZZ_p$-linear combination of vectors in $L'$.\slug

\nineproclaim Exercise 10.
Show that the characteristic function on $\ZZ_p$ is equal to its Fourier transform. More
generally, let $V$ be a quadratic space over $\QQ_p$ and let $L$ be a $\ZZ_p$-sublattice of $V$. Show that
the Fourier transform of $L$ is $\mu(L)$ times the characteristic function of the $\ZZ_p$-dual lattice;
that is,
$$\hat{\one_L} = \mu(L) \one_{L^\vee}.$$

\nineproof
Expanding the definition of the Fourier transform gives
$$\hat{\one_L} (x) = \int_V \one_L(y) e^{-2\pi i(x\cdot y)} \d y = \int_L e^{-2\pi i(x\cdot y)}\d y.$$
If $x$ is an element of the dual, then $x\cdot y$ is an integer for all $y\in L$, so the integrand
is $1$ and the integral equals $\mu(L)$. If not, then
$\chi(y) = e^{-2\pi i (x\cdot y)}$ is a nontrivial character. The kernel of $\chi$ is a sublattice
of $L$, so by Lemma B, there is $n$ such that
$\ker\chi \supseteq p^nL$, so by translation-invariance of the measure $\mu$,
$$\eqalign{
\int_L e^{-2\pi i(x\cdot y)}\d y
& = \sum_{v\in L/p^nL} \int_{v+p^nL} \chi(y) \d y \cr
& = \sum_{v\in L/p^nL} \chi(v)\int_{p^nL} \chi(y) \d y \cr
&= \sum_{v\in L/p^nL} \chi(v) \int_{p^nL} 1\d y \cr
&= \mu(p^nL)\sum_{v\in L/p^n L}\chi(v).\cr
}$$
But this is zero because the sum of a nontrivial character over a group is zero, by Lemma Z.
\slug

\boldlabel Ad\`eles. We let
$$\hat\ZZ = \prod_p \ZZ_p,$$
\vskip-5pt\noindent
where the product is taken over all primes $p$.
We then define $\AA_\ZZ = \RR\times \hat\ZZ$. The {\it ring of ad\`eles} $\AA_\QQ$ is the tensor product
$\AA_\QQ = \QQ\otimes_\ZZ \AA_\ZZ$. An element in this ring is an infinite tuple consisting
of a real number and one $p$-adic
rational for each $p$, all but finitely many of which are $p$-adic integers.

\nineproclaim Exercise 11. Let $L$ be a lattice of covolume $1$ in a quadratic space $V$. The Poisson
summation formula on $V_\RR$ asserts that
$$\sum_{v\in L} \phi(v) = \sum_{v\in L^\vee} \hat\phi(v),$$
for all Schwartz functions $\phi$ on $V_\RR$, where $L^\vee$ is the dual lattice of $V$. The ad\`elic
Poisson summation formula asserts that
$$\sum_{v\in V} \phi(v) = \sum_{v\in V} \hat\phi(v),$$
for all ad\`elic Schwartz functions $\phi$ on $V_{\AA_\QQ}$. Show that the ad\`elic Poisson
summation formula implies its more familiar analogue on $V_\RR$.

\nineproof
Given a Schwartz function $\phi$ on $\RR$, we take our Schwartz function
on the ad\`ele ring to be $f = \phi \otimes \one_{L\otimes \hat\ZZ}$. We have
$$\sum_{v\in L} \phi(v) = \sum_{v\in V} \one_{L\otimes \hat\ZZ}(v) \phi(v) = \sum_{v\in V} f(v).$$
By the ad\`elic Poisson summation formula, the right-hand side becomes
$$\sum_{v\in V} f(v) = \sum_{v\in V} \hat f(v),$$
but by the previous exercise, the Fourier transforms at the $p$-adic components are characteristic functions
of dual lattices. So
$$\sum_{v\in V} \hat f(v) = \sum_{v\in V} \one_{L^\vee\otimes \hat\ZZ} \hat\phi(v) = \sum_{v\in L^\vee}
\hat\phi(v),$$
which is what we were aiming to show.
\slug

\nineproclaim Exercise 12. Let $G = \GL_n(\QQ_p)$ and let $X = \GL_n(\QQ_p)/\GL_n(\ZZ_p)$. Show that
the action of $G$ on $X$ satisfies the finiteness assumption that was made in class when we discussed
Hecke operators, i.e., that the stabiliser of any $x\in X$ acts on $X$ with finite orbits.

\nineproof We want to show that for all $x\in X$, the set
$\Stab_G(x) y$ is finite. Note that $\Stab_G(e) = \GL_n(\ZZ_p)$, and for general $x\in X$, we have
$$\Stab_G(x) = x \GL_n(\ZZ) x^{-1}.$$
So for $y\in X$, $\Stab(x) y = x\GL_n(\ZZ_p) x^{-1} y$ and we are done if we can show that
$\GL_n(\ZZ_p) z$ is finite for any $z\in X$.
By applying Lemma~B with $L = z\ZZ_p^n$ and $L' = \ZZ_p^n$, we obtain $m$ such that
$zp^m \ZZ_p^n \subseteq \ZZ_p^n$. Then we apply Lemma~B again with $L = \ZZ_p^n$ and $L' = zp^m \ZZ_p^n$
to get $k\ge m$ such that $p^{k}\ZZ_p^n\subseteq p^m z\ZZ_p^n$.

Now letting $L = p^m \ZZ_p^n$, we can associate to any $x\in X$ the sublattice $xL$
of $\QQ_p^n$. If $x$ and $x'$ are different,
then these sublattices are different, so our problem reduces to showing that the set of lattices of the
form
$\gamma zL$, where $\gamma\in \Stab_G(x)$, is finite. Since $\gamma$ comes from a conjugate of $\GL_n(\ZZ_p)$,
it fixes $\ZZ_p^n$, so have the chain of inclusions
$$p^{k}\ZZ_p^n \subseteq \gamma (zL) \subseteq \ZZ_p^n,$$
showing that
any such lattice of the prescribed form
is a subgroup of $\ZZ_p^n$ containing $p^{k}\ZZ_p^n$. By the correspondence theorem,
these subgroups are in bijection with elements of
$$ \ZZ_p^n/\bigl(p^{k}\ZZ_p^n\bigr) \cong \bigl(\ZZ/p^{k}\ZZ\bigr)^n,$$
which is finite.
\slug

\nineproclaim Exercise 13. Let $L$ be a unimodular $\ZZ_p$-lattice in a quadratic space $V$ over $\QQ_p$,
and let $G$ be the orthogonal group over $\ZZ_p$ attached to $L$. Show that $G(\QQ_p)$ acts
transitively on the set of pairs $(L_1,L_2)$ of unimodular lattices satisfying $L_1/(L_1\cap L_2)\cong
\ZZ/p\ZZ$.

\nineproof (Reginald Lybbert.)
We write $L\sim_p L'$ if $L/(L\cap L') \cong \ZZ/p\ZZ$ and we take it as a fact (it was shown
in class) that this is a symmetric relation. We shall also assume that $p\ne 2$, for the sake of everyone's
sanity.
Since $G(\QQ_p)$ acts transitively on lattices, we can assume that the first lattice in each tuple is
the same.
It is then enough to show that for any $(L,L_1)$ and $(L,L_2)$, we can find a map stabilising $L$ that sends
$L_1$ to $L_2$. (We will allow ourselves the use of $L$ for a general unimodular
lattice, not necessarily the one in the definition of $G$.) Note first that $pL_1\subseteq L$, since it
is contained in the kernel of $L_1\to \ZZ/p\ZZ$, which is $L_1\cap L$. The kernel of $L\to L/pL$ is of
course $pL$, so the kernel of $\phi: pL_1\to L/pL$ is $pL1\cap pL$. So we have
$$\phi(pL_1)\cong {pL_1\over pL_1\cap pL} \cong {L_1\over L_1\cap L} \cong {\ZZ\over p\ZZ}.$$
This is some line in $L/pL$, and it is isotropic because the length of any element in $L_1$ is integer.
Multiplying by $p$, the length of the element in $L/pL$ is a multiple of $p$ and thus $0$. In fact, there
is a correspondence between $p$-neighbours of $L$ and isotropic lines in $L/pL$. Given an isotropic line
in $L/pL$, we see that it's preimage must lie in $L\cap L'$ for some lattices $L$ and $L'$. But we know
what $L$ is, so we can deduce the lattice $L'$ for which the isotropic line is $\phi(pL')$. This shows
the relation is one-to-one.

To show that the relation is surjective, we take an isotropic line in $L/pL$, spanned by an element $\bar v\in L$;
by Hensel's lemma (since $p\ne 2$, the gradient vector of the quadratic form is nonzero),
we can lift this to a vector $v\in L$ with $v\cdot v = 0$. Consider
$$L_v = \ZZ_p\cdot {1\over p}v + \{w\in L : w\cdot v \equiv 0\pmod p\}.$$
The claim is that $L_v$ is a $p$-neighbour of $L$. To show that it has rank $n$, note that $L_v\subseteq (1/p)L$
and $pL\subseteq L_v$, giving us $\QQ_p^n \subseteq L_v\otimes \QQ_p\subseteq \QQ_p^n$. Next, consider
the map $L_v\to \ZZ/p\ZZ$ sending $(a/p)v + w$ to $a\bmod p$. It is certainly injective and its kernel is
$L\cap L_v$. It remains to show that $L_v$ is unimodular, which takes a bit of work. Since $v$ is an isotropic
vector in a unimodular lattice $L$, we can find a vector $u$ with $u\cdot v = 0$
such that $\QQ_p u \ortho \QQ_p v$ is a hyperbolic plane. Let $\langle u,v\rangle$ denote the span of
$u$ and $v$ in the lattice $L$, and let $\langle u,v\rangle^\perp_L$ denote its complement with respect to $L$.
We have
$$L = \langle u,v\rangle \ortho \langle u,v\rangle^\perp_L,$$
where since both $L$ and $\langle u,v\rangle$ are unimodular, we conclude that $\langle u,v\rangle^\perp_L$ is.
Note that
$$L_v = \Bigl\langle pu,{v\over p}\Bigr\rangle \ortho \Bigl\langle pu,{v\over p}\Bigr\rangle^\perp_{L_v}.$$
Since $(v/p)\cdot(v/p) = 0 = (p u)\cdot (pu)$ and $(v/p)\cdot (pu) = 1$, the first summand is unimodular.
It thus remains to prove that the second summand is. We shall in fact show that
$\langle u,v\rangle^\perp_L = \langle v/p, pu\rangle^\perp_{L_v}$. Take $w\in L$ such that $w\cdot v=w\cdot u=0$.
Then $w\in L_v$ and $w\cdot (v/p) = w\cdot pu = 0$. On the other hand, if $w\in L_v$ with
$w\cdot (v/p) = w\cdot pu = 0$, then writing $w = (a/p)v + x$ for some $x\in L$, we can write
$x = \lambda v + \mu u + y$ where $y\in \langle u,v\rangle^\perp_L$, by the decomposition of $L$ we found earlier.
We know that $w\cdot v = p(w\cdot v/p) = 0$ and $w\cdot u = (1/p)w\cdot(pu) = 0$, so $w$ is actually equal to $y$
above, and thus is in $L$.

We have now shown that we can write
$$L = \langle u_1,v_1\rangle \ortho \langle u_1,v_1\rangle_L^\perp =
\langle u_2,v_2\rangle \ortho       \langle u_2,v_2\rangle_L^\perp$$
where
$$L_1 = \Bigl\langle pu_1,{v_1\over p}\Bigr\rangle \ortho
\Bigl\langle pu_1,{v_1\over p}\Bigr\rangle^\perp_{L_v}$$
and
$$L_2 = \Bigl\langle pu_2,{v_2\over p}\Bigr\rangle \ortho
\Bigl\langle pu_2,{v_2\over p}\Bigr\rangle^\perp_{L_v}.$$
The claim is that the map $g\in G(\QQ_p)$ sending
$u_1\mapsto u_2$ and $v_1\mapsto v_2$ fixes $L$ and sends $L_1$ to $L_2$.
It is clear that $\langle u_1,v_1\rangle \cong \langle u_2, v_2\rangle = g\langle u_1,v_1\rangle$.
Then by Witt's cancellation theorem
over $\ZZ_p$, we have $\langle u_1,v_1\rangle_L^\perp \cong \langle u_2,v_2\rangle_L^\perp =
g\langle u_1,v_1\rangle_L^\perp$ as well, which
completes the proof that $gL_1 = L_2$. (We proved Witt's cancellation theorem in class for fields but not lattices
in general. However, it holds for lattices over $\ZZ_p$ where $p$ is odd,
which can be seen by reducing the lattice modulo $p$
and using the cancellation theorem for $\FF_p$; this was proved by B.~W.~Jones in 1942.)
\slug

\boldlabel Symplectic space and the Heisenberg group.
A {\it symplectic space} is a vector space $V$ over a field $k$ endowed with a bilinear form
$\langle\cdot,\cdot\rangle : V\times V\to k$, which is alternating in that $\langle v,w\rangle = 
-\langle w,v\rangle$ for all $v,w\in V$ and which is nondegenerate, i.e., $\langle u,v\rangle = 0$
for all $v\in V$ if and only if $u=0$. A key example is taking $W = k^2$, with
$$\bigl\langle (a_1,b_1),(a_2,b_2)\bigl\rangle = a_1b_2-a_2b_1.$$

The {\it Heisenberg group} of a symplectic space $W$ is the set $k\times W$ endowed with
the group law
$$(t_1, w_1)(t_2, w_2) = \bigl(t_1+t_2 +\langle w_1, w_2\rangle, w_1+w_2\bigr).$$
Of course, in the case that $W$ is $k^2$, this boils down to the group law
$$(t_1, v_1, w_1)(t_2, v_2, w_2) = (t_1+t_2 + v_1w_2 - v_2w_1, v_1+v_2, w_1+w_2)$$
on triples in $k^3$.

\nineproclaim Exercise 14. Write down the character table of the Heisenberg group $H(W)$ where $W$
is the two-dimensional symplectic space over the field with $p$ elements.

\ninesolution
There are $p$ elements in the centre of $H(W)$, namely the elements $(t,0,0)$ for $t\in k$, so
there are $p$ conjugacy classes $\bigl\{ (t,0,0)\bigr\}$ of one element each. The other
conjugacy classes are of the form
$$\bigl\{  (t,v_1, v_2) : t\in k\bigr\}$$
for $(v_1, v_2) \ne (0,0)$; there are $p^2-1$ of these classes, and each of them contains $p$ elements,
so we have accounted for all $p + p(p^2-1) = p^2$ elements of $H(W)$. So there are $p^2+p-1$ characters as well.
We shall show that the character table is the following:
$$
\vbox{\offinterlineskip\eightpoint
    \halign{ 
        \hfil $#$\hfil\quad&\hfil $#$\hfil\quad&\quad $#$ \hfil\quad
                 &\quad\hfil $#$ \hfil\quad&\quad\hfil $#$\hfil\cr
        \hbox{Quantity} & \hbox{Dimension} & \hbox{Indexed by}
              & \bigl\{(t,0,0)\bigr\} : t\in k & \bigl\{ (*,v,w)\bigr\}:(v,w)\ne (0,0) \cr
        \noalign{\medskip}
        \noalign{\hrule}
        \noalign{\medskip}
        p^2 & 1 & (m,n)\in k^2 & 1 & {\zeta_p}^{mv + nw} \cr
        \noalign{\medskip}
        p-1 & p & n\in k\setminus \{0\} & p{\zeta_p}^{nt} & 0 \cr
    }
}$$

Let $\zeta_p$ be a primitive $p$th root of unity. There are $p^2$ representations of degree $1$, indexed
by $(m,n)\in k^2$, each mapping $(t,v,w) \mapsto {\zeta_p}^{mv+nw}$. For the other $p-1$ representations,
note that there are $p-1$ nontrivial characters $\psi_n : k\to \CC$, given by $\psi_n(t) = {\zeta_p}^{nt}$,
where $n\in k\setminus \{0\}$.
Each of these gives an action of $H(W)$ on $\calS(V)$, since for $f\in \calS(V)$, we can let
$$\bigl((t,0,0)* f\bigr)(x) = \psi_n(t) f(x)\qquad\hbox{and}\qquad
\bigl((0,v,w)* f\bigr)(x) = \psi_n(-2v\cdot x) f(x+w),$$
It remains to find the trace of these representations on the conjugacy classes. To do so, we take
as a basis for $\calS(W)$ the $p$ delta functions
$$\delta_y(x) = \cases{1,& if $x=y$;\cr 0,& otherwise}$$
for $y\in V$. In the case where an
element of the form $(t,0,0)$ acts on the space, each $\delta_y$ is taken to $\psi_n(t) \delta_y$, so
the matrix has $\psi_n(t)$ down the main diagonal and the trace is $p\psi_n(t) = p{\zeta_p}^{nt}$.

For the conjugacy class $\bigl\{(*,v,w)\bigr\}$,
we shall show that the trace is zero.
These elements of $H(W)$ send
$\delta_y$ to the function $x\mapsto \psi_n(-2v\cdot x) \delta_y(w+x) = \psi_n(-2v\cdot x)\delta_{y-w}$.
This means that each row of this transformations's matrix has exactly one nonzero entry, but it cannot be on
the main diagonal, having been shifted cyclically by $w$ places. Thus the trace is zero. If $w=0$, then all
of the nonzero entries are still on the main diagonal, but then the trace becomes
$$\sum_{x\in V} \psi_n(-2v\cdot x) = \sum_{x\in V} \psi_n(x) = 0,$$
by Lemma~Z and the nondegeneracy of the dot product on $\FF_p$.\slug

\section References

\bye

### 2021-12-17 misc/expo/exotic.tex
Typos

@@
 
-\maketitle{Tsirelson's space\qquad and other exotic constructions}{}{Marcel K. Goh}{16 December 2021}
+\maketitle{Tsirelson's space and other exotic constructions}{}{Marcel K. Goh}{16 December 2021}
 
@@
 $$\norm{x}_{l_\infty} = \sup_{i\ge 1} |x_i|.$$
-For $1\le p\le \infty$, the set $l_p$ of all infinite sequences $x$ with $\norm{x}{l_p}$ finite is a Banach space;
+For $1\le p\le \infty$, the set $l_p$ of all infinite sequences $x$ with $\norm{x}_{l_p}$ finite is a Banach space;
 when $p=\infty$, these are the bounded sequences.
@@
 let $X_n$ denote the linear span of $\{e_1,\ldots,e_n\}$. Each normed vector space $X_n$ is finite-dimensional
-and thus complete, which implies that each $X_n$ is closed in $X$. We now show that every $X_n$ has empty
-interior, which completes the proof.\slug
+and thus complete, which implies that each $X_n$ is closed in $X$. But each of these subspaces is proper,
+so by the lemma above, so all of these subspaces have empty interior, which contradicts the Baire category
+theorem.\slug
 

### 2022-02-14 misc/solutions/prob-method-sols.tex
Added prob method solutions

\input fontmac
\input mathmac

\maketitle{The Probabilistic Method}{exercise solutions by}{Marcel K. Goh}{14 February 2022}

\bigskip

\section 1. The Basic Method

\proclaim Exercise 1.1. Prove that if there is a real $p$, with $0\le p\le 1$ such that
$${n\choose k}p^{k\choose 2} + {n\choose t}(1-p)^{t\choose 2} < 1,$$
then the Ramsey number $r(k,t)$ satisfies $r(k,t)>n$. Using this, show that
$$r(4,t) \ge \Omega\bigl(t^{3/2}/ (\ln t)^{3/2}\bigr).$$

\proof We follow the proof of Proposition~1.1.1 in the book. We consider a random graph on
$n$ vertices, where each edge is present with probability $p$. Let $K$ be the event that
there is a clique of size $k$ in the graph, and let $I$ be the event that there is an
independent set of size $t$ in the graph. By the union bound,
$$\pr\{ K \cup I\} \le \pr\{K\} + \pr\{I\} \le \sum_{|S| = k} p^{k\choose 2} + \sum_{|S|=t} (1-p)^{t\choose 2}
= {n\choose k}p^{k\choose 2} + {n\choose t}(1-p)^{t\choose 2} < 1.$$
This means that $\pr\{ \neg K \cap \neg I\} > 0$ and since the sample space is finite,
there exists a graph on $n$ vertices with
no clique of size $k$ and no independent set of size $t$ and therefore $r(k,t) > n$.

Next we show that $r(4,t) > (et/\ln t)^{3/2}$ for large enough $t$. Note that
$${n\choose 4} p^{k\choose 2} + {n\choose t}(1-p)^{t\choose 2}\le
n^4p^6 + {e^tn^t\over t^t}(1-p)^{t^2/4},$$
by the inequality
$$ {n^k\over k^k} \le {n\choose k} \le {e^k n^k\over k^k}.$$
Setting $n = et^{3/2} / (\ln t)^{3/2}$, we have
$$\eqalign{
n^4p^6 + {e^tn^t\over t^t}(1-p)^{t^2/4} &=
\biggl({tp\over e\ln t}\biggr)^6 + {e^t t^{3t/2} \over t^t e^{3t/2} (\ln t)^{3t/2}} (1-p)^{t^2/4} \cr
&=\biggl({tp\over e\ln t}\biggr)^6 + {t^{t/2} \over e^{t/2} (\ln t)^{3t/2}} (1-p)^{t^2/4} \cr
&\le \biggl({tp\over e\ln t}\biggr)^6 + \biggl({t(1-p)^{t/2} \over e(\ln t)^3}\biggr)^{t/2} \cr
&\le \biggl({tp\over e\ln t}\biggr)^6 + \biggl({t \over e^{pt/2+1}(\ln t)^3}\biggr)^{t/2},\cr
}$$
where in the last line we used the inequality $1-p \le e^{-p}$.
Choosing $p = 2\ln t/t$, we simply need $t$ large enough such that
$$ \biggl({t\over e^{\ln t+1} (\ln t)^3}\biggr)^{t/2} = 
\biggl({1\over e(\ln t)^3}\biggr)^{t/2} < 1- {2\over e^6},$$
which can be done since the left-hand side goes to $0$.\slug

\bye

### 2022-02-14 misc/solutions/prob-method-sols.tex
Finished 1.2

@@
 
-Next we show that $r(4,t) > (et/\ln t)^{3/2}$ for large enough $t$. Note that
+Next we show that $r(4,t) > \bigl(t/(e\ln t)\bigr)^{3/2}$ for large enough $t$. Note that
 $${n\choose 4} p^{k\choose 2} + {n\choose t}(1-p)^{t\choose 2}\le
 n^4p^6 + {e^tn^t\over t^t}(1-p)^{t^2/4},$$
-by the inequality
+by the inequalities
 $$ {n^k\over k^k} \le {n\choose k} \le {e^k n^k\over k^k}.$$
-Setting $n = et^{3/2} / (\ln t)^{3/2}$, we have
+Setting $n = t^{3/2} / (e\ln t)^{3/2}$, we have
 $$\eqalign{
@@
 $$ \biggl({t\over e^{\ln t+1} (\ln t)^3}\biggr)^{t/2} = 
-\biggl({1\over e(\ln t)^3}\biggr)^{t/2} < 1- {2\over e^6},$$
+\biggl({1\over e(\ln t)^3}\biggr)^{t/2} < 1- \biggl({2\over e}\biggr)^6,$$
 which can be done since the left-hand side goes to $0$.\slug
 
+\proclaim Exercise 1.2. Suppose $n\ge 4$ and let $H$ be an $n$-uniform hypergraph with at most
+$4^{n-1}/3^n$ edges. Prove that there is a colouring of the vertices of $H$ by $4$ colours so that in
+every edge all $4$ colours are represented.
+
+\proof Let each vertex of $H$ be independently given one of the four colours uniformly at random. (If
+$H$ is infinite, it does not matter what colour we give to vertices that do not appear in any edge,
+so it suffices to consider $H$ finite, which makes the sample space finite.)
+Given some edge $e$ of $H$ with $n$ vertices, there are $4^n$ total ways that $e$ may be coloured,
+and for each of the four colours, $3^n$ total ways that $e$ may be coloured using only the other three colours.
+Let $K(e)$ denote the event that $e$ does not contain all four colours. By the inclusion-exclusion principle,
+$$\pr\bigl\{K(e)\bigr\} = 4\cdot 3^n - 6\cdot 2^n + 4$$
+Since $6\cdot 2^n \ge 96 > 4$, the probability that a given edge {\it does not} contain all four colours
+is (much) less than $3^n/4^{n-1}$. By the union bound,
+$$\pr\Bigl\{ \bigcup_{e\in E(H)} K(e)\Bigr\} \le \sum_{e\in E(H)} \pr\bigl\{K(e)\bigr\}
+< {4^{n-1}\over 3^n} \cdot {3^n\over 4^{n-1}} = 1.$$
+Since the sample space is finite this implies that there is some colouring of the vertices of $H$
+in which every edge has all four colours.\slug
+
 \bye

### 2022-02-14 misc/solutions/prob-method-sols.tex
Exercise 4.1

@@
 \input mathmac
+
+\def\given{\,|\,}
 
@@
 
+\section 4. The Second Moment
+
+\proclaim Exercise 4.1. Let $X$ be a random variable taking integral nonnegative values, let $\ex\{X^2\}$
+denote the expectation of its square, and let $\var\{X\}$ denote its variance. Prove that
+$$\pr\{ X = 0\} \le {\var\{X\} \over \ex\{X^2\}}.$$
+
+\proof Since $X$ is integer and nonnegative, we have $\pr\{X = 0\} = 1-\pr\{X\ge 1\}$ and since
+$\var\{X\} = \ex\{X^2\} - \ex\{X\}^2$, to get our result it suffices to show that
+$$\pr\{X\ge 1\} \ge {\ex\{X\}^2 \over \ex\{X^2\}}.$$
+We start by noting that
+$$\ex\{X\} = \sum_{k=0}^\infty k \pr\{X=k\}
+= \pr\{X\ge 1\}\sum_{k=1}^\infty {k\pr\{X=k\}\over \pr\{X\ge 1\}}
+= \pr\{X\ge 1\}\ex\{ X \given X\ge 1\}.$$
+Since the function $x\mapsto x^2$ is convex, we have, by Jensen's inequality,
+$$\ex\{X\}^2 = \pr\{X\ge 1\}^2 \ex\{X\given X\ge 1\}^2 \le \pr\{X\ge 1\}^2 \ex\{X^2 \given X\ge 1\}
+= \pr\{X\ge 1\} \ex\{X^2\}.$$
+Dividing both sides by $\ex\{X^2\}$ gives us what we want.\slug
+
 \bye

### 2022-02-15 misc/solutions/prob-method-sols.tex
Tried 1.3 and did 1.4

@@
 \def\given{\,|\,}
+\def\biggiven{\,\big|\,}
 
-\maketitle{The Probabilistic Method}{exercise solutions by}{Marcel K. Goh}{14 February 2022}
+\maketitle{The Probabilistic Method}{exercise solutions by}{Marcel K. Goh}{15 February 2022}
+
+\floattext5 \ninebf Note. \ninepoint
+A star indicates a starred problem in the text; these are supposed to be harder.
+I have decided to put my work here even when I haven't solved the problem yet, so that when I come back
+later I can see what I've tried. I won't put the black square
+\hbox{\kern1.5pt\vrule width2.5pt height6pt depth1.5pt\kern1.5pt}
+until I think the proof
+is actually done. Also I wrote $\log$ everywhere to mean $\log_e$
+even though the text uses $\ln$.
 
@@
 then the Ramsey number $r(k,t)$ satisfies $r(k,t)>n$. Using this, show that
-$$r(4,t) \ge \Omega\bigl(t^{3/2}/ (\ln t)^{3/2}\bigr).$$
+$$r(4,t) \ge \Omega\bigl(t^{3/2}/ (\log t)^{3/2}\bigr).$$
 
@@
 
-Next we show that $r(4,t) > \bigl(t/(e\ln t)\bigr)^{3/2}$ for large enough $t$. Note that
+Next we show that $r(4,t) > \bigl(t/(e\log t)\bigr)^{3/2}$ for large enough $t$. Note that
 $${n\choose 4} p^{k\choose 2} + {n\choose t}(1-p)^{t\choose 2}\le
@@
 $$ {n^k\over k^k} \le {n\choose k} \le {e^k n^k\over k^k}.$$
-Setting $n = t^{3/2} / (e\ln t)^{3/2}$, we have
+Setting $n = t^{3/2} / (e\log t)^{3/2}$, we have
 $$\eqalign{
 n^4p^6 + {e^tn^t\over t^t}(1-p)^{t^2/4} &=
-\biggl({tp\over e\ln t}\biggr)^6 + {e^t t^{3t/2} \over t^t e^{3t/2} (\ln t)^{3t/2}} (1-p)^{t^2/4} \cr
-&=\biggl({tp\over e\ln t}\biggr)^6 + {t^{t/2} \over e^{t/2} (\ln t)^{3t/2}} (1-p)^{t^2/4} \cr
-&\le \biggl({tp\over e\ln t}\biggr)^6 + \biggl({t(1-p)^{t/2} \over e(\ln t)^3}\biggr)^{t/2} \cr
-&\le \biggl({tp\over e\ln t}\biggr)^6 + \biggl({t \over e^{pt/2+1}(\ln t)^3}\biggr)^{t/2},\cr
+\biggl({tp\over e\log t}\biggr)^6 + {e^t t^{3t/2} \over t^t e^{3t/2} (\log t)^{3t/2}} (1-p)^{t^2/4} \cr
+&=\biggl({tp\over e\log t}\biggr)^6 + {t^{t/2} \over e^{t/2} (\log t)^{3t/2}} (1-p)^{t^2/4} \cr
+&\le \biggl({tp\over e\log t}\biggr)^6 + \biggl({t(1-p)^{t/2} \over e(\log t)^3}\biggr)^{t/2} \cr
+&\le \biggl({tp\over e\log t}\biggr)^6 + \biggl({t \over e^{pt/2+1}(\log t)^3}\biggr)^{t/2},\cr
 }$$
 where in the last line we used the inequality $1-p \le e^{-p}$.
-Choosing $p = 2\ln t/t$, we simply need $t$ large enough such that
-$$ \biggl({t\over e^{\ln t+1} (\ln t)^3}\biggr)^{t/2} = 
-\biggl({1\over e(\ln t)^3}\biggr)^{t/2} < 1- \biggl({2\over e}\biggr)^6,$$
+Choosing $p = 2\log t/t$, we simply need $t$ large enough such that
+$$ \biggl({t\over e^{\log t+1} (\log t)^3}\biggr)^{t/2} = 
+\biggl({1\over e(\log t)^3}\biggr)^{t/2} < 1- \biggl({2\over e}\biggr)^6,$$
 which can be done since the left-hand side goes to $0$.\slug
@@
 
+\proclaim \llap{*}Exercise 1.3. Prove that for two independent, identically distributed real random
+variables $X$ and $Y$,
+$$\pr\bigl\{|X-Y|\le 2\bigr\} \le 3\pr\bigl\{|X-Y|\le 1\bigr\}.$$
+
+\proof [Not done. Need to use independence somehow. I think by symmetry
+it is enough to show that $\pr\{1 < X-Y\le 2\}
+\le \pr\bigl\{|X-Y|\le 1\bigr\}$ or $\pr\bigl\{X-Y > 1 \biggiven |X-Y|\le 2\bigr\}\le 1/3$.
+In the second expression,
+$X$ and $Y$ are no longer independent.]
+
+\proclaim \llap{*}Exercise 1.4. Let $G=(V,E)$ be a graph with $n$ vertices and minimum degree $\delta>10$.
+Prove that there is a partition of $V$ into two disjoint subsets $A$ and $B$ so that
+$|A|\le O\bigl((n\log n)/\delta\bigr)$, and each vertex of $B$ has at least one neighbour in $A$ and at least
+one neighbour in $B$.
+
+\proof We follow the construction of the dominating set from the proof of Theorem~1.2.2, since the
+required $A$ here is a dominating set with some extra conditions. Let $p$ be chosen later and let $X$
+be a random set of vertices obtained by independently selecting each $v\in V$ with probability $p$.
+Then as in the proof from the textbook, we let $Y$ be the set of all $v\in V\setminus X$ that have no
+neighbours in $X$. At this point $X\cup Y$ is a dominating set, but we are not yet done constructing
+$A$, since there may still be elements in $V\setminus (X\cup Y)$ all of whose neighbours belong to $X\cup Y$.
+Let $Z\subseteq V\setminus (X\cup Y)$ denote all of these elements. For any given $v\in V$, we
+have $\pr\{v\in X\} = p$ and
+$$\pr\{v\in Y\} = (1-p)^{\deg(v) + 1} \le (1-p)^{\delta+1} \le e^{-p(\delta+1)},$$
+since for $v\in Y$ we need $v$
+itself as well as all $\deg(v)$ of its neighbours to not be in $X$. Lastly,
+$$\pr\{v\in Z\} = (1-p)\prod_{w\in N(v)} \bigl(p+(1-p)^{\deg(w)+1}\bigr)
+\le (1-p)\bigl(p+(1-p)^{\delta+1}\bigr)^\delta \le (1-p)(p+e^{-p(\delta+1)})^\delta.$$
+Now we compute
+$$\eqalign{
+\ex\bigl\{|A|\bigr\} &= \sum_{v\in V} v\in \{X\cup Y\cup Z\} \cr
+&\le n\bigl( p + e^{-p(\delta+1)} + (1-p)(p+e^{-p(\delta+1)})^\delta \bigr).\cr
+}$$
+Since $p+e^{-p(\delta+1)} < 1$, we can remove the $\delta$ from its exponent for the looser but
+simpler bound
+$$\eqalign{
+\ex\bigl\{|A|\bigr\} &\le np + ne^{-p(\delta+1)} + np - np^2 + ne^{-p(\delta+1)} - npe^{-p(\delta+1)}\cr
+&=(2-p)(np+ne^{-p(\delta+1)}).\cr
+}$$
+Setting $p = \log(\delta+1)/(\delta+1)$ just as in the dominating set proof, we have
+$$e^{-p(\delta+1)} = {1\over \delta+1}$$
+and
+$$\ex\bigl\{|A|\bigr\} \le \biggl(2-{\log(\delta+1)\over \delta+1}\biggr)\biggl( n{\log(\delta+1)\over \delta+1}
++ {n\over \delta+1}\biggr) = 2n{\log(\delta+1)\over\delta+1} + o\biggl({n\log\delta\over\delta}\biggr),$$
+which has the required asymptotics. It remains to choose an $A$ with $|A|$ at least this average.\slug
+
 \section 4. The Second Moment

### 2022-02-16 misc/solutions/prob-method-sols.tex
Did 4.4

@@
 
+\proclaim Exercise 4.4. Let $X$ be a random variable with expectation $\ex\{X\} =0$ and variance
+$\var\{X\} = \sigma^2$. Prove that for all $\lambda > 0$,
+$$\pr\{X\ge \lambda\} \le {\sigma^2\over \sigma^2 + \lambda^2}.$$
+
+\proof Let $\lambda > 0$. Note that if $X\ge \lambda$, then $X$ is also positive so in particular,
+for any $a>0$, $X+a$ and $\lambda + a$ are both positive. This implies that $(X+a)^2\ge (\lambda+a)^2$.
+Note that since $\var\{X\} = \ex\{X^2\} - \ex\{X\}^2 = \sigma^2$ and $\ex\{X\}=0$, we have
+$\ex\{X^2\} = \sigma^2$, which in turn implies that
+$$\ex\bigl\{(X+a)^2\bigr\} = \ex\{X^2\} + 2a\ex\{X\} + \ex\{a^2\} = \sigma^2+a^2.$$
+Now we take the infimum over all $a>0$ and apply Markov's inequality to get
+$$\pr\{X\ge\lambda\} \le \inf_{a > 0} \pr\bigl\{(X+a)^2 \ge (\lambda+a)^2\bigr\} \le
+\inf_{a>0} {\sigma^2 + a^2\over (\lambda + a)^2}.$$
+We now optimise over $a>0$. We have
+$${d\over da} {\sigma^2 + a^2\over (\lambda + a)^2}
+  = {(\lambda+a)^2 2a - (\sigma^2+a^2)2(\lambda+a)\over (\lambda + a)^4},$$
+which is zero when $a = \sigma^2/\lambda > 0$. Plugging in this value of $a$ gives
+$$\pr\{X\ge \lambda\} \le {\sigma^2 + \sigma^4/\lambda^2\over (\lambda + \sigma^2/\lambda)^2}
+= {\sigma^2(\lambda + \sigma^2/\lambda) \over \lambda(\lambda + \sigma^2/\lambda)^2}
+= {\sigma^2 \over \sigma^2+\lambda^2},$$
+exactly what we need.\slug
+
 \bye

### 2022-02-16 misc/solutions/prob-method-sols.tex
Two small problems

@@
 
+\proclaim Exercise 1.7. Let $F$ be a finite collection of binary strings of finite lengths and assume no
+member of $F$ is a prefix of another one. Let $N_i$ denote the number of strings of length $i$ in $F$.
+Prove that
+$$\sum_{i\ge 1} {N_i\over 2_i} \le 1.$$
+
+\proof This inequality looks an awful bit like the LYM inequality, so our proof is informed by the proof
+of that statement. Let $m$ be the maximum length of any string in $F$ (such a maximum exists because $F$
+is finite). Let $b_1, b_2, \ldots, b_m$ be a sequence of independent random variables with
+$\pr\{b_i = 0\} = \pr\{b_i = 1\} = 1/2$ for all $1\le i\le m$. Consider the set of strings
+$$C = \{b_1, b_1b_2, \ldots, b_1b_2\cdots b_m\}.$$
+This is a random set, so we may study the random quantity $|C\cap F|$. Note that $C$ contains
+exactly one bitstring of each length, and every string of length $i$ has an equal chance of being in $C$,
+namely $2^{-i}$. So
+$$\ex\bigl\{|C\cap F|\bigr\} = \sum_{s\in F} \pr\{s\in C\} = \sum_{s\in F} 2^{-|s|}
+= \sum_{i\ge 1} {N_i\over 2^i},$$
+where $|s|$ is the length of the string $s$. In the last equality we simply grouped strings of the same
+length together. On the other hand, $|C\cap F|$ cannot possibly be greater than $1$, since for any two
+members of $C$, one must be a prefix of the other. So $\ex\bigl\{|C\cap F|\bigr\} \le 1$ as well,
+and this observation completes the proof.\slug
+
+\section 2. Linearity of Expectation
+
+\proclaim Exercise 2.1. Suppose $n\ge 2$ and let $H=(V,E)$ be an $n$-uniform hypergraph with $|E| = 4^{n-1}$
+edges. Show that there is a colouring of $V$ by $4$ colours so that no edge is monochromatic.
+
+\proof Let $\phi$ be a random function from $V$ to $[4]$, so that all $4^n$ possible colourings are equally
+likely. Let $X$ be the number of monochromatic edges in $H$ under this colouring; that is,
+$$X = \sum_{e \in E} \indic{e\ \hbox{is monochromatic}}.$$
+The probability that a given edge $e = \{v_1,\ldots,v_n\}$ is monochromatic is $1/4^{n-1}$, since
+$\phi(v_1)$ can be anything, but then $\phi(v_2)$ through to $\phi(v_n)$ each have a $1/4$ chance
+of matching $\phi(v_1)$. So
+$$\ex\{X\} = \sum_{e\in E} \pr\{e\ \hbox{is monochromatic}\} = {4^{n-1}\over 4^{n-1}}
+= 1.$$
+We can produce a colouring with $4^{n-1}$ monochromatic edges by setting $\phi(v) = 1$ for all $v\in V$,
+so $\pr\{X > 1 \} > 0$. Thus for $\ex\{X\}$ to equal $1$, we must also have $\pr\{X < 1\} > 0$, and since
+$X$ is nonnegative, this means that $\pr\{X = 0\} > 0$ and there exists some colouring with no monochromatic
+edges.\slug
+
 \section 4. The Second Moment

### 2022-02-16 misc/solutions/prob-method-sols.tex
Exercise 3.1

@@
 
-\proclaim Exercise 1.7. Let $F$ be a finite collection of binary strings of finite lengths and assume no
+\proclaim Exercise 1.8. Let $F$ be a finite collection of binary strings of finite lengths and assume no
 member of $F$ is a prefix of another one. Let $N_i$ denote the number of strings of length $i$ in $F$.
@@
 edges.\slug
+
+\goodbreak
+\section 3. Alterations
+
+\proclaim Exercise 3.1. As shown in Section~3.1, the Ramsey number $R(k,k)$ satisfies
+$$R(k,k) > n-{n\choose k}2^{1-{k\choose 2}}$$
+for every integer $n$. Conclude that
+$$R(k,k) \ge \bigl(1-o(1)\bigr) {k\over e} 2^{k/2}.$$
+
+\proof We set $n= (k-1) 2^{k/2} /e$ in the above, to get the bound
+$$R(k,k) > {k\over e} 2^{k/2} - {1\over e} 2^{k/2} - {(k-1) 2^{k/2} /e \choose k} 2^{1-{k\choose 2}}.$$
+Now we apply the inequality ${n\choose k} \le e^kn^k/k^k$, which yields
+$$\eqalign{
+R(k,k) &> {k\over e} 2^{k/2} - {1\over e} 2^{k/2} - {e^k(k-1)^k 2^{k^2/2} /e^k \over k^k} 2^{1-k^2/2+k/2}\cr
+&= {k\over e} 2^{k/2} - {1\over e} 2^{k/2} - \biggl(1-{1\over k}\biggr)^k 2^{1+k/2}. \cr
+}$$
+Since $(1-1/k)^k\to 1/e$ as $k\to\infty$,
+$$R(k,k) > {k\over e}2^{k/2} \biggl( 1 - {1\over k} - {2e(1-1/k)^k \over k}\biggr)
+= {k\over e}2^{k/2} \bigl(1-o(1)\bigr).\noskipslug$$
 

### 2022-02-18 misc/solutions/prob-method-sols.tex
2.7

@@
 \def\biggiven{\,\big|\,}
+\def\F{{\cal F}}
 
-\maketitle{The Probabilistic Method}{exercise solutions by}{Marcel K. Goh}{15 February 2022}
+\maketitle{The Probabilistic Method}{exercise solutions by}{Marcel K. Goh}{17 February 2022}
 
@@
 
+\proclaim Exercise 2.7. Let $\F$ be a family of subsets of $[n] = \{1,2,\ldots,n\}$ and suppose there are
+no $A,B\in \F$ satisfying $A\subseteq B$. Let $\sigma\in S_n$ be a random permutation of the elements of
+$[n]$ and consider the random variable
+$$ X = \bigl| \bigl\{ i : \{\sigma(1),\sigma(2),\ldots,\sigma(i)\}\in \F\bigr\}\bigr|.$$
+By considering the expectation of $X$, prove that $|\F|\le {n\choose \lfloor n/2\rfloor}$.
+
+\proof This is the proof that I said Exercise~1.8 reminded me of. Let $S_i = \{\sigma(1),\ldots,\sigma(i)\}$
+and let
+$$ C = \{\emptyset, S_1, S_2,\ldots, S_n\}.$$
+If there are $i<j$ such that $S_i$ and $S_j$ are both in $\F$, then
+$\F$ violates the condition that there are no
+$A,B\in \F$ with $A\subseteq B$. So $X\le 1$.
+On the other hand, $C$ contains
+exactly one set of size $i$ for each $0\le i\le n$ and the probability that $S_i$ is a given
+element of $X^{(i)}$ is the same. So letting $N_i$ be the number of sets of size $i$ in $\F$, we have
+$$1\ge \ex\{X\} = \sum_{A\in \F} {n\choose |A|}^{-1} = \sum_{i=0}^n N_i {n\choose i}^{-1}\ge
+{|\F|\over {n\choose \lfloor n/2\rfloor}},$$
+where the last inequality holds because ${n\choose k}$ is maximised when $k=\lfloor n/2\rfloor$
+and because the sum of the $N_i$ is $|\F|$.\slug
+
 \goodbreak

### 2022-02-19 misc/solutions/prob-method-sols.tex
1.5

@@
 
+\proclaim \llap{*}Exercise 1.5. Let $G=(V,E)$ be a graph on $n\ge 10$ vertices and suppose that if we add
+to $G$ any edge not in $G$, the number of copies of a complete graph on $10$ vertices in it increases.
+Show that the number of edges of $G$ is at least $8n-36$.
+
+\proof The figure $8n-36$ in the conclusion obscures the fact that we should use Theorem 1.3.3. In
+fact,
+$$8n-36 = {n^2-n-n^2+17n-72\over 2} = {n(n-1) - (n-8)(n-9)\over 2} = {n\choose 2}-{n-8\choose 2}.$$
+Let $F = V^{(2)}\setminus E = \{A_1, A_2,\ldots, A_h\}$ be the set of non-edges in $G$. Since the addition
+of any non-edge causes the number of $K_{10}$ subgraphs to increase, we know that for each $A_i$ there
+exists a set $X_i$ of size $8$ and disjoint from $A_i$ such that the subgraph induced on
+the vertices $A_i\cup X_i$ would be a $K_{10}$ if $A_i$ happened to be in $E$ (which it isn't).
+Set $B_i = V\setminus (A_i\cup S_i)$. Clearly, $A_i\cap B_i = \emptyset$ for all $i$. Also,
+since $A_i\cup S_i$ is complete except for one missing edge $A_i$, for any $j\ne i$, we have
+$A_j\not\subseteq A_i\cup S_i$. This means that $A_j\cap B_i\neq\emptyset$ for all $j\ne i$. Hence
+$\bigl\{(A_i,B_i)\bigr\}_{i=1}^h$ is a $(2,n-10)$-system as defined in the chapter, and by
+Theorem~1.3.3, $|F| = h\le {2+n-10\choose 2}$. We conclude that $|E|\ge {n\choose 2}-{n-8\choose 2}$.\slug
+
 \proclaim Exercise 1.8. Let $F$ be a finite collection of binary strings of finite lengths and assume no

### 2022-02-19 misc/solutions/prob-method-sols.tex
3.3

@@
 
+\proclaim Exercise 3.3. Prove that every $3$-uniform hypergraph with $n$ vertices and $m\ge n/3$ edges
+contains an independent set (i.e., a set of vertices containing no edges) of size at least
+$${2n^{3/2}\over 3\sqrt 3\sqrt m}.$$
+
+\proof We mirror the proof of Theorem~3.2.1. Let $(X\cap \F)$ be a $3$-uniform hypergraph with
+$|X| = n$ and $|\F| = m \ge n/3$. Let $S\subseteq X$ be a random subset obtained by independently
+putting each $x\in X$ in $S$ with probability $p$, to be chosen later. Of course, $\ex\bigl\{|X|\bigr\} = np$.
+What is the expected number of edges contained in $S$? Well, each edge $\{x,y,z\}\in \F$ has probability
+$p^3$ of being contained in $S$, so we expect $mp^3$ edges in the subgraph induced by $S$. If
+$Y$ is the set of these edges, we now want to choose $p$ to
+minimise the quantity $\ex\bigl\{|X|-|Y|\bigr\} = np-mp^3$. The derivative
+is $n-3mp^2$ so we ought to set $p=\sqrt{n/3m}$, obtaining
+$$\ex\bigl\{ |X|-|Y|\bigr\} = {n^{3/2}-m\sqrt n\over \sqrt{3m}}.$$
+Since $m\ge n/3$, we have
+$$\ex\bigl\{ |X|-|Y|\bigr\} \ge {n^{3/2}\over \sqrt{3m}}-{n^{3/2}\over 3\sqrt{3m}}={2n^{3/2}\over 3\sqrt{3m}}.$$
+This means there exists some set $S\subseteq X$ with the number of vertices minus the number of edges
+at least $2n^{3/2}/\bigl(3\sqrt{3m}\bigr)$. Removing one vertex from each edge, we arrive at an independent
+set of vertices $S'$ with cardinality at least this quantity.\slug
+
 \section 4. The Second Moment

### 2022-02-24 about/vita.tex
Added manderz
New semester
Lots of new links

@@
 \def\sectheader#1{{\bigskip\elevensc #1}\smallskip\hrule\medskip}
-\def\leftright#1#2{{#1}\hfill{#2}\par}
+\def\leftright#1#2{\hbox to\hsize{{#1}\hfill{#2}}\par}
+\def\starleftright#1#2{\hbox to\hsize{\llap{*}{#1}\hfil{#2}}\par}
 \def\CEE{{\mc C}}
@@
 \smallskip
-Supervisor: Luc Devroye.
+Supervisor: Luc Devroye. CGPA: 4.00/4. Thesis topic: ``Structural properties of Galton--Watson trees.''
 \medbreak
@@
 Analysis of random discrete structures, analysis and design of algorithms, enumerative combinatorics,
-additive combinatorics, algebra.
+additive combinatorics.
 
@@
 
-\papercount=2
+\papercount=3
 \pubbegin P
+\papitem (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
+Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. To appear in {\sl Discrete Mathematics
+and Theoretical Computer Science}, 11~pp. [arXiv:2105.12046]
 \papitem (with Anna M. Brandenberger and Luc Devroye)
@@
 The lattice of arithmetic progressions. {\sl arXiv preprint 2106.05949}, 15~pp.
-\smallskip
-\thing (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
-Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl arXiv preprint 2105.12046}, 11~pp.
 \endgroup
@@
 
-\sectheader{Work and Volunteer Experience}
+\sectheader{Work and Volunteer Experience {\rm($*$ indicates a paid position)}}
 
-\leftright{\bf Teaching assistant}{\bf September 2021 -- present}
+\leftright{\bf Mentor}{\bf January 2021 -- present}
+\leftright{\bf Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
+\smallskip
+Meet with two undergraduate students as part of the Directed Reading Program
+on a weekly basis to give them an introduction to research-level
+mathematics in a casual setting. Focused on topics in extremal combinatorics related to the increasing
+triples problem.
+\medbreak
+
+\starleftright{\bf Teaching assistant}{\bf September -- December 2021}
 \leftright{\sl School of Computer Science, McGill University}{\sl Montr\'eal, QC, Canada}
@@
 
-\leftright{\bf First responder}{\bf October 2017 -- present}
+\leftright{\bf First responder}{\bf October 2017 -- September 2021}
 \leftright{\sl McGill Student Emergency Response Team}{\sl Montr\'eal, QC, Canada}
@@
 
-\leftright{\bf Grader}{\bf September 2019 -- April 2021}
+\starleftright{\bf Grader}{\bf September 2019 -- April 2021}
 \leftright{\sl Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
@@
 
-\leftright{\bf Helpdesk tutor}{\bf September 2018 -- April 2021}
+\starleftright{\bf Helpdesk tutor}{\bf September 2018 -- April 2021}
 \leftright{\sl Computer Science Undergraduate Society Helpdesk}{\sl Montr\'eal, QC, Canada}
@@
 
-\leftright{\bf Painter}{\bf May -- August 2018}
+\starleftright{\bf Painter}{\bf May -- August 2018}
 \leftright{\sl Bakir Contracting Corp.}{\sl Edmonton, AB, Canada}
@@
 
-\leftright{\bf Infantryman}{\bf August 2015 -- August 2017}
+\starleftright{\bf Infantryman}{\bf August 2015 -- August 2017}
 \leftright{\sl Singapore Armed Forces}{\sl Singapore}
@@
 
-\leftright{\bf Tutor}{\bf September 2014 -- June 2015}
+\starleftright{\bf Tutor}{\bf September 2014 -- June 2015}
 \leftright{\sl \'Ecole Secondaire Beaumont Composite High School}{\sl Beaumont, AB, Canada}
@@
 
-\leftright{\bf Summer camp counsellor}{\bf June -- August 2014}
+\starleftright{\bf Summer camp counsellor}{\bf June -- August 2014}
 \leftright{\sl YoWoChAs Outdoor Education Centre}{\sl Fallis, AB, Canada}

### 2022-03-09 misc/solutions/prob-method-sols.tex
Added two new questions that I did on break

@@
 \def\biggiven{\,\big|\,}
+\def\A{{\cal A}}
+\def\B{{\cal B}}
+\def\C{{\cal C}}
 \def\F{{\cal F}}
+\def\G{{\cal G}}
 
-\maketitle{The Probabilistic Method}{exercise solutions by}{Marcel K. Goh}{17 February 2022}
+\maketitle{The Probabilistic Method}{exercise solutions by}{Marcel K. Goh}{9 March 2022}
 
@@
 
+\section 6. Correlation Inequalities
+
+\proclaim Exercise 6.1. Let $G$ be a graph and let $P$ denote the probability that a random subgraph
+of $G$ obtained by picking each edge of $G$ with probability $1/2$ independently is connected (and spanning).
+Let $Q$ denote the probability that in a random two-colouring of $G$, where each edge is chosen,
+randomly and independently, to be either red or blue, the red graph and the blue graph are both connected
+(and spanning). Is $Q\le P^2$?
+
+\solution The answer is yes. Let $\A$ be the family of subsets $A\subseteq E$ such that $G' = (V,A)$
+is connected. Let $\C$ be the family of subsets $C\subseteq E$ such that $G'' = (V, E\setminus C)$ is
+connected. Adding any edge to $A\in \A$ keeps the set in $\A$, and removing any edge from $C\in \C$ keeps
+the set in $\C$, so $\A$ is monotone increasing and $\C$ is monotone decreasing. Note that if $S$ is a
+random subset of edges obtained by taking each edge independently with probability $1/2$, then
+$\pr\{S = A\} = \pr\{S = E\setminus A\}$ for any fixed $A\subseteq E$, so we find that
+$ P = \pr\{\A\} = \pr\{\C\}$. On the other hand, $Q$ is the probability that both $S$ and $E\setminus S$
+induce connected subgraphs; i.e., $Q = \pr\{S\in \A\cap \C\}$. Applying Proposition~6.3.1 now gives
+$$Q = \pr\{ \A\cap \C\} \le \pr\{\A\} \pr\{\C\} = P^2,$$
+as desired.\slug
+
+\proclaim Exercise 6.2. A family of subsets $\G$ is called {\it intersecting} if $G_1\cap G_2\ne \emptyset$
+for all $G_1,G_2\in \G$. Let $\F_1, \F_2, \ldots, \F_k$ be $k$ intersecting families of subsets of $X = [n]$.
+Prove that
+$$\biggl|\bigcup_{i=1}^k \F_i\biggr| \le 2^n - 2^{n-k}.$$
+
+\proof We do this by induction on $k$. If $k=1$, then the upper bound of $2^{n-1}$ on the size of an
+intersecting family $\F$ is clear from the fact that $\F$ may contain at most one element of each pair
+$(A,X\setminus A)$, and there are $2^{n-1}$ such pairs.
+For the induction step, we split
+$$\bigcup_{i=1}^k \F_i = \bigcup_{i=1}^{k-1} \F_i \cup \F_k;$$
+call the first term on the right-hand side $\A$ and let $\C = 2^X\setminus \F_k$, so that
+$$\bigcup_{i=1}^{k-1} \F_i \cup \F_k = \A \cup \F_k = (\A \cap \C) \cup \F_k.$$
+Since we are trying to form an upper bound on this quantity,
+we might as well assume that $\A$ is maximal; i.e., adding any set to $\F_i$ for $1\le i< k$ causes
+$\F_i$ to no longer be intersecting. The claim is that $\A$ is monotone increasing. To see this, suppose
+that $A\supseteq B$ and $B\in \A$. This means that $B\in \F_i$ for some $1\le i<k$, and so for any $C\in \F_i$,
+$A\cap C \supseteq B\cap C\ne \emptyset$, meaning that $A\in \F_i\subseteq \A$ as well, otherwise we could
+add $A$ to $\A$ and contradict maximality. We have actually shown
+that the union of any $m$ maximal intersecting families is monotone increasing, so in particular $\F_k$ is
+as well. But this means that $\C = 2^X\setminus \F_k$ is monotone decreasing, since if $C\notin \F_k$
+and $B\subseteq C$ with $B\in \F_k$, then $\F_k$ would not be monotone increasing. We are now in a position
+to apply the cardinality version of Proposition~6.3.1. It gives
+$$\biggl|\bigcup_{i=1}^k \F_i\biggr| = |\A\cap \C| + |\F_k|
+\le {|\A|\cdot|\C|\over 2^n} + |\F_k|
+\le \biggl(1-{1\over 2^{k-1}}\biggr) |\C| + |\F_k|,$$
+where in the last inequality we used the induction hypothesis $|\A| \le 2^n - 2^{n-k+1}$. Since
+$|\C| + |\F_k| = 2^n$ and the coefficient of the latter is larger than the coefficient of the former
+in the above, the entire expression is largest when $|\F_k| = |\C| = 2^{n-1}$. In this case, we have
+$$\biggl|\bigcup_{i=1}^k \F_i\biggr| \le \biggl(1-{1\over 2^{k-1}}\biggr) 2^{n-1} + 2^{n-1} = 2^n-2^{n-k},$$
+completing the induction and the proof.\slug
+
 \bye

### 2022-03-21 misc/expo/fourier.tex
Started Fourier

\input fontmac
\input mathmac

\def\FF{{\bf F}}
\def\TT{{\bf T}}
\def\bar{\overline}
\def\hat{\widehat}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}

\widemargins
\bookheader{FOURIER ANALYSIS IN ADDITIVE COMBINATORICS}{MARCEL K. GOH}

\maketitle{Fourier analysis in additive combinatorics}{notes by}{Marcel K. Goh}{}

\floattext4.5 \ninebf Note.
\ninepoint
I am compiling this set of expository notes primarily to solidify these bread-and-butter
concepts in
my own mind. Sources that are far more comprehensive exist on the web, so I'm not sure
how useful this will be to anyone else. Currently, the bulk of these notes come from
a series of YouTube lectures of Timothy Gowers (given in early 2022), though I intend
in future to add more relevant topics as I learn about them.

\bigskip

\advsect Preliminary notions

Let $Z$ be a finite abelian group. A {\it character} on $Z$ is a homomorphism from
$Z$ to the multiplicative group $\C\setminus\{0\}$. If $\chi$ is such a homomorphism, then
$\bigl|\chi(x)\bigr| = 1$ for every $x\in Z$, so we can actually regard $\chi$ as being
a function from $Z$ to the unit circle $\TT = \{z\in \C : |z| = 1\}$. The pointwise product
of two characters gives another character, and the function $\chi_0$ that sends every $x\in Z$
to $1$ has the property that if $\chi$ is any character, then $\chi\chi_0 = \chi = \chi_0\chi$.
Lastly, we note that multiplication is commutative and
for any character $\chi$, the product of $\chi$ with $\bar\chi$
gives $\chi_0$, so the set of characters on $Z$ is an abelian group. This is called the
{\it dual group} or sometimes {\it Pontryagin dual} of $Z$ and is denoted $\hat Z$.

We define an inner product on the space $\CC^Z$ of functions from $Z$ to $\CC$ by setting
$$\langle f,g\rangle = \ex_x f(x) \bar{g(x)},$$
where $\ex_{x\in Z} F(x) = |Z|^{-1} \sum_{x\in Z} F(x)$.
We can also make $\hat Z$ into an inner product space by letting
$$\langle \hat f, \hat g\rangle = \sum_{\chi\in \hat Z} \hat f(\chi)\bar{\hat g(\chi)};$$
note that this time we do not normalise by dividing by $|Z|$.
Two functions $f$ and $g$ in an inner product space are said to be {\it orthogonal} if $\langle f,g\rangle = 0$.
The first lemma we'll prove concerns certain orthogonality relations.

\parenproclaim Lemma {\advthm} (Orthogonality relations). Let $Z$ be a finite abelian group.
\medskip
\item{a)} If $\chi_1$ and $\chi_2$ are characters on $Z$, then
$$\langle \chi_1,\chi_2 \rangle = \ex_{x\in Z} \chi_1(x)\bar{\chi_2(x)}
  = \cases{ 1, & if $\chi_1 = \chi_2$;\cr 0, & if $\chi_1 \ne\chi_2$.}.$$
\smallskip
\item{b)} For $x,y\in Z$, we have
$$\sum_{\chi\in \hat Z} \chi(x)\bar{\chi(y)}
  = \cases{ |\hat Z|, & if $x=y$;\cr 0, & if $x\ne y$,}$$
\medskip

\proof Let $\chi_0$ denote the trivial character. We have
$$\ex_{x\in Z} \chi_0(x) = \ex_{x\in Z} 1 = 1.$$
On the other hand, let $\chi$ be a nontrivial character and let $u\in Z$ be such that $\chi(u)\ne 1$. Then from
$$\ex_{x\in Z} \chi(x) = \ex_{x\in Z} \chi(u+x) = \ex_{x\in Z}\chi(u)\chi(x) = \chi(u)\ex_{x\in Z}\chi(x),$$
it follows that $\ex_{x\in Z}\chi(x) = 0$. Now consider $\chi_1\bar{\chi_2}$. If $\chi_1 = \chi_2$, then
this is the trivial character and from our first observation, $\langle \chi_1,\chi_2\rangle = 0$. Otherwise,
we are in the second case and the inner product is zero. This proves part (a).

Part (b) is proven similarly. Note that $\chi(x)\bar{\chi(y)} = \chi(x-y)$. If $x=y$, then we have
$$\sum_{\chi\in\hat Z} \chi(x-y) = \sum_{\chi\in\hat Z} \chi(0) = |\hat Z|.$$
If $x\ne y$, then there is some $\psi\in \hat Z$ such that $\psi(x-y)\ne 1$, and then writing
$$\sum_{\chi\in \hat Z} \chi(x-y) = \sum_{\chi\in\hat Z} \psi(x-y)\chi(x-y)
= \psi(x-y)\sum_{\chi\in\hat Z} \chi(x-y),$$
we see that $\sum_{\chi\in\hat Z} \chi(x-y)$ must be zero.\slug

Since the space of functions from $Z\to \CC$ has dimension $|Z|$, there are at most $|Z|$ characters.
Every finite abelian group can be written as a direct product of cyclic groups, and we shall use this fact
to show that there are exactly $|Z|$ characters; i.e., $|\hat Z| = |Z|$. For brevity, let $\ZZ_n$ stand for $\ZZ/n\ZZ$ for all $n\in\NN$,
and for $\alpha\in \RR$, let $e(\alpha) = e^{2\pi i\alpha}$.

\proclaim Lemma \advthm.
The set of characters on a finite abelian group $Z$ spans the space of functions from $Z$ to $\CC$.

\proof Write $Z \cong \ZZ_{n_1} \times \ZZ_{n_2}\times \cdots\times \ZZ_{n_r}$. For every
$u = (u_1,\ldots, u_r)\in Z$, the function $\chi_u : Z\to \CC$ that takes
$$(x_1,\ldots, x_r) \mapsto \prod_{i=1}^r e\biggl( {u_ix_i\over n_i}\biggr)$$
is a character (this is easy to show, since $e(x_i + x_i') = e(x_i)e(x_i')$). If $u\ne u'$,
then $\chi_u\ne\chi_{u'}$ and $\langle \chi_u,\chi_{u'}\rangle$ by the previous lemma. Hence
the set $\{\chi_u : u\in Z\}$ comprises $|Z|$ linearly independent elements in a space of dimension $|Z|$,
which must be spanning.\slug

Note that the map $u\mapsto \chi_u$ gives an isomorphism from $Z$ to $\hat Z$.
The two main examples we shall consider are when $Z = \ZZ_n$ and when $Z = \FF_p^n$. In the first
case, $\chi_u$ takes $x\mapsto e(ux/N)$ and in the second case $\chi_u$ takes $x\mapsto e(u\cdot x/p)$.
(Recall that if $u = (u_1,\ldots, u_n)$ and $x = (x_1,\ldots, x_n)$ then the dot product $u\cdot x$
is the sum $\sum_{i=1}^n u_ix_i$.)

For a finite set $X$, we shall denote by $L_p(X)$ the normed vector space of all functions $f:X\to \CC$
under the norm
$$\norm{f}_p = \bigl( \ex_{x\in X} \bigl| f(x)\bigr|^p \bigr)^{1/p}.$$
The notation $l_p(X)$ denotes the same set, but with the norm
$$\norm{f}_p = \Bigl( \sum_{x\in X} \bigl| f(x)\bigr|^p \Bigr)^{1/p}$$
instead. Note that $\norm{f}_2^2 = \langle f,f\rangle$ and $\norm{f}_\infty = \max_{x\in X} \bigl|f(x)\bigr|$.

\medskip\boldlabel The Fourier transform.
If $f:Z\to \CC$, the {\it Fourier transform} of $f$ is the function $\hat f : \hat Z\to\CC$ given by
$$\hat f(\chi) = \langle f,\chi\rangle = \ex_{x\in Z} f(x) \bar{\chi(x)}.$$
We shall now state and prove an identity which is extremely useful in Fourier analysis.

\parenproclaim Lemma {\advthm} (Parseval's identity). Let $Z$ be a finite abelian group
and let $f$ and $g$ be functions from $Z$ to $\CC$.
If $\hat f$ and $\hat g$ are the Fourier transforms of $f$ and $g$ respectively, then
$\langle f,g\rangle = \langle \hat f, \hat g\rangle$.

\proof First, we expand all the definitions and rearrange sums to obtain
$$\eqalign{
\langle \hat f, \hat g\rangle &= \sum_{\chi\in \hat Z} \hat f(\chi)\bar{\hat g(\chi)} \cr
&=  \sum_{\chi\in \hat Z} \ex_{x\in Z} f(x)\bar{\chi(x)} \ex_{y\in Z} g(y)\bar{\chi(y)} \cr
&=   \ex_{x\in Z} f(x) \ex_{y\in Z} g(y) \sum_{\chi\in \hat Z}\bar{\chi(x)}\bar{\chi(y)}.\cr
}$$
By the second orthogonality relation and the fact that $|\hat Z| = |Z|$, the inner sum equals $|Z|$ when
$y = x$ and $0$ otherwise. Thus for any given $x$, we have
$$\ex_{y\in Z} g(y) \sum_{\chi\in \hat Z}\chi(x)\bar{\chi(y)} = g(x),$$
which is exactly what we need for the whole thing to equal $\langle f,g\rangle$.\slug

We also have the following inversion formula that recovers the original function $f$ from $\hat f$.

\parenproclaim Lemma {\advthm} (Fourier inversion formula). Let $Z$ be a finite abelian group and
let $f:Z\to \CC$. We have
$$f(x) = \sum_{\chi\in \hat Z} \hat f(\chi)\chi(x).$$

\proof We expand
$$\sum_{\chi\in \hat Z} \hat f(\chi)\chi(x) = \sum_{\chi\in \hat Z} \ex_{y\in Z} f(y) \bar{\chi(y)} \chi(x)
= \ex_{y\in Z} f(y) \sum_{\chi\in \hat Z} \chi(x)\bar{\chi(y)},$$
and we noted in the proof of Parseval's theorem that the right-hand side is exactly $f(x)$.\slug

The inversion formula tells us that two functions from $Z\to \CC$ are equal if and only if their Fourier
transforms are equal. For the next lemma, which concerns dilates of a function $f: Z\to \CC$, we will
establish some notation. If $x\in Z$, we define $nx$ recursively for all integers $n\ge 0$ by $0x = 0$
and $nx = x + (n-1)x$. We then set $nz = - (-n)x$ for all negative integers. Since the group operation
on characters is multiplication, we will not write $n\chi$ but instead $\chi^n$ for $\chi$ multiplied
with itself $n$ times.

\parenproclaim Lemma {\advthm} (Dilation rule). Let $Z$ be a finite abelian group and let $a\in \ZZ$
be an integer that is coprime to $|Z|$. Denote the multiplicative inverse of $a$ modulo $|Z|$ by $a^{-1}$.
Letting $f_a : Z\to \CC$ be the function given by $f_a(x) = f(a^{-1} x)$, we have
$\hat{f_a}(\chi) = \hat f(\chi^a)$.

\proof We have
$$\hat{f_a}(\chi) = \ex_{x\in Z} f_a(x) \bar{\chi(x)} = \ex_{x\in Z} f(a^{-1} x)\bar{\chi(x)}.$$
Since $x\mapsto ax$ is a bijection from $Z$ to itself, we can replace $x$ formally by $ax$ in the above
to get
$$\hat{f_a}(\chi) = \ex_{x\in Z} f(x)\bar{\chi(ax)} = \ex_{x\in Z} f(x) \bar{\chi(ax)}^a
= \hat f(\chi^a).\noskipslug$$

\medskip\boldlabel Convolutions.
For functions $f$ and $g$ from a finite abelian group $Z$ to $\CC$, the {\it convolution} $f*g$ is
defined by
$$ (f*g)(x) = \ex_{y+z = x} f(y)g(z).$$
If instead $\hat f$ and $\hat g$ are functions from $\hat Z$ to $\CC$, then
$$ (\hat f * \hat g)(\chi) = \sum_{\chi_1\chi_2 = \chi} \hat f(\chi_1)\hat g(\chi_2).$$

\parenproclaim Lemma {\advthm} (Convolution law). Let $Z$ is a finite abelian group and $f,g : \Z\to \CC$.
For all $\chi\in \hat Z$, we have $\hat{f*g} (\chi) = \hat f(\chi)\hat g(\chi)$.

\proof We start by expanding
$$\hat{f*g}(\chi) = \ex_{x\in Z} (f*g)(x) \bar{\chi(x)} = \ex_{x\in Z} \ex_{y+z=x} f(y)g(z)\bar{\chi(y)\chi(z)}.$$
But note that $x$ does not appear in the summand anymore, meaning that we can simply rewrite this as
an expectation
over {\it all} $y$ and $z$ (their sum will equal $x$ for some $x\in Z$). Thus we can conclude that
$$\hat{f*g}(\chi) = \ex_{y\in Z}\ex_{z\in Z} f(y)g(z)\bar{\chi(y)\chi(z)} = \hat f(\chi)\hat g(\chi),$$
which is what we wanted to show.\slug

If $A$ is a subset of a finite abelian group $Z$, we associate to $A$ the {\it characteristic function}
$\one_A : Z\to \CC$ given by
$$\one_A(x) = \cases{ 1, & if $x\in A$;\cr 0, & otherwise.}$$
When it will not cause confusion, we will abuse notation and write $A(x)$ instead of $\one_A(x)$.
With the definitions and results stated above, we are now in a position to translate properties of $A$ to
Fourier-analytic statements about the function $\one_A$. First off, note that
$$\bignorm{\hat{\one_A}}_2^2 = \norm{\one_A}_2^2 = \ex_{x\in Z} \one_A(x)^2 = \ex_{x\in Z} \one_A(x)
= {|A|\over |Z|}.$$
If $|A|/|Z| = \delta$, then we will say that $A$ has {\it density} $\delta$. Another way of expressing $\delta$
in terms of $\one_A$ is
$$\hat{\one_A}(\chi_0) = \ex_{x\in Z} \one_A(x)\chi_0(x) = \ex_{x\in Z} \one_A(x) = \delta,$$
where $\chi_0$ is the trivial character.

The fact that $\langle \one_A, \one_A\rangle = \delta$ means that if we sample $x$ and $y$ from $Z$ and
condition that $x=y$, the probability that $x$ (or $y$) is in $A$ is $\delta$. The reason for stating this
obvious fact in such a stilted way is that it generalises by way of the convolution operation we defined
earlier; that is, what if we are interested in the probability that a $4$-tuple $(x,y,z,w)$ satisfying $x+y=z+w$
is a member of $A^4$? Well, the convolution law gives
$$\eqalign{
\ex_{x+y=z+w} A(x)A(y)A(z)A(w) &= \ex_{u\in Z} \ex_{x+y=u}\ex_{z+w=u} A(x)A(y)A(z)A(w) \cr
&= \ex_{u\in Z} (\one_A*\one_A)(u)^2 \cr
&= \norm{\one_A*\one_A}_2^2 \cr
&= \bignorm{\hat{\one_A*\one_A}}_2^2 \cr
&= \bignorm{\bigl(\hat{\one_A}\bigr)^2}_2^2 \cr
&= \sum_{\chi\in \hat Z} \bigl| \hat{\one_A}(\chi)\bigr|^4 \cr
&= \norm{\one_A}_4^4, \cr
}$$
and more generally, the probability that a tuple $(a_1,\ldots,a_k, b_1,\ldots,b_k)\in Z^{2k}$
with $a_1+\cdots+a_k=b_1+\cdots+b_k$ is a member of $A^{2k}$ is $\norm{\one_A}_{2k}^{2k}$. In any case,
the number of $(x,y,z,w)\in A^4$ with $x+y = z+w$ is so important in additive combinatorics that is has a name.
It is called the {\it additive energy} of $A$ and is denoted $E(A)$; what we showed above is that
$E(A) = |Z|^3 \norm{\one_A}_4^4$.

\section References

\bye

### 2022-03-22 misc/expo/fourier.tex
Couple of mistakes

@@
 is a character (this is easy to show, since $e(x_i + x_i') = e(x_i)e(x_i')$). If $u\ne u'$,
-then $\chi_u\ne\chi_{u'}$ and $\langle \chi_u,\chi_{u'}\rangle$ by the previous lemma. Hence
+then $\chi_u\ne\chi_{u'}$ and $\langle \chi_u,\chi_{u'}\rangle = 0$ by the previous lemma. Hence
 the set $\{\chi_u : u\in Z\}$ comprises $|Z|$ linearly independent elements in a space of dimension $|Z|$,
@@
 Fourier-analytic statements about the function $\one_A$. First off, note that
-$$\bignorm{\hat{\one_A}}_2^2 = \norm{\one_A}_2^2 = \ex_{x\in Z} \one_A(x)^2 = \ex_{x\in Z} \one_A(x)
+$$\bignorm{\hat{\one_A}}_2^2 = \norm{\one_A}_2^2 = \ex_{x\in Z} A(x)^2 = \ex_{x\in Z} A(x)
 = {|A|\over |Z|}.$$
@@
 in terms of $\one_A$ is
-$$\hat{\one_A}(\chi_0) = \ex_{x\in Z} \one_A(x)\chi_0(x) = \ex_{x\in Z} \one_A(x) = \delta,$$
+$$\hat{\one_A}(\chi_0) = \ex_{x\in Z} A(x)\chi_0(x) = \ex_{x\in Z} A(x) = \delta,$$
 where $\chi_0$ is the trivial character.
 
-The fact that $\langle \one_A, \one_A\rangle = \delta$ means that if we sample $x$ and $y$ from $Z$ and
+The fact that $\norm{\one_A}_2^2 = \delta$ means that if we sample $x$ and $y$ from $Z$ and
 condition that $x=y$, the probability that $x$ (or $y$) is in $A$ is $\delta$. The reason for stating this
-obvious fact in such a stilted way is that it generalises by way of the convolution operation we defined
+obvious fact in such a stilted fashion is that it generalises by way of the convolution operation we defined
 earlier; that is, what if we are interested in the probability that a $4$-tuple $(x,y,z,w)$ satisfying $x+y=z+w$
@@
 the number of $(x,y,z,w)\in A^4$ with $x+y = z+w$ is so important in additive combinatorics that is has a name.
-It is called the {\it additive energy} of $A$ and is denoted $E(A)$; what we showed above is that
+It is called the {\it additive energy} of $A$ and is denoted $E(A)$; we just showed above is that
 $E(A) = |Z|^3 \norm{\one_A}_4^4$.

### 2022-03-26 misc/expo/additive.tex
It is 2022 omg

\input fontmac
\input mathmac

\def\FF{{\bf F}}
\def\TT{{\bf T}}
\def\bar{\overline}
\def\hat{\widehat}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}

\widemargins
\bookheader{NOTES ON ADDITIVE COMBINATORICS}{MARCEL K. GOH}

\maketitle{Notes on additive combinatorics}{by}{Marcel K. Goh}{23 March 2022}

\floattext4.5 \ninebf Note.
\ninepoint
I am compiling this set of expository notes primarily to solidify these fundamental results of
additive combinatorics in
my own mind. Sources that are far more comprehensive exist on the web, so I'm not sure
how useful this will be to anyone else. Currently, the bulk of these notes come from
a series of YouTube lectures of Timothy Gowers (given in early 2022), though I intend
in future to add more relevant topics as I learn about them.

\bigskip

\advsect Pl\"unnecke's theorem

Let $A$ and $B$ be finite subsets of an abelian group. We define the {\it sumset}
$A+B = \{a+b : a\in A,\, b\in B\}$ as well as the {\it difference set}
$A-B = \{a-b : a\in A,\, b\in B\}$. For a nonnegative integer $r$, the {\it $r$-fold iterated sumset}
$rA$ is defined recursively by $0A = \{0\}$ and $rA = A+(r-1)A$. The goal of this first section
is to show that if the cardinality of
a sumset $|A+B|$ is small compared to the size of the set $A$, then the difference
$rB-sB$ of two iterated sumsets must also be small relative to the size of $A$,
in some sense. This result is called Pl\"unnecke's
theorem as it follows from a statement given by H.~Pl\"unnecke in~1969, though this particular
formulation of it was stated and proved by I.~Z.~Ruzsa in 1989.

\edef\thmplunnecke{\the\thmcount}
\parenproclaim Theorem {\advthm} (Pl\"unnecke, {\rm 1969;} Ruzsa, {\rm 1989}). Let $A_0$ and
$B$ be finite subsets of an abelian group and suppose that $|A_0+B|\le K_0|A_0|$ for some constant $K_0$.
Then there exist $A\subseteq A_0$ and $K\le K_0$ such that
for any nonnegative integers $r$ and $s$, not both zero, we have $|rB-sB| \le K^{r+s}|A|$. In particular,
this means that $|rB-sB| \le {K_0}^{r+s}|A_0|$.

Ruzsa used Menger's theorem from graph theory to prove this theorem, but we will use a shorter
argument from a 2012 paper by G.~Petridis. It hinges on the following lemma. In this proof (and the
rest of these notes), we will write $A+x$ instead of $A+\{x\}$ for convenience.

\edef\lempetridis{\the\thmcount}
\parenproclaim Lemma {\advthm} (Petridis, {\rm 2012}). Let $A$ and $B$ be subsets of an abelian
group and let $K$ be such that $|A_0+B| = K_0|A_0|$. Then there exist $A\subseteq A_0$ and
$K\le K_0$ such that $|A+B+C| = K|A+C|$ for every finite subset $C$ of the group.

\proof We pick $A\subseteq A_0$ nonempty such that the ratio $|A+B|/|A|$, which we shall
set as $K$, is minimised. So $|A+B| = K|A|$ and $|A'+B|\ge K|A'|$ for all $A'\subseteq A_0$.
Now we shall show by induction on $C$ that $|A+B+C|\le K|A+C|$ for all $C$. If $C$ is empty we
are done, of course. Now assume the result holds for $C$ and let $x\notin C$ be arbitrary. We have
$$\bigl| A+(C\cup \{x\}) \bigr| = |A+C| + |A+x| - |A'+x|$$
where $A' = \{a\in A : a+x \in A+C\}$. But adding $x$ to everything in a set does not change its
cardinality, so we have
$$|A+C| + |A+x|-|A'+x| = |A+C| + |A| + |A'|.$$
Also, any element of $A'+B+x$ is an element of $A+B+C$, since $a'+x\in A+C$ for all $a'\in A'$.
Applying the induction hypothesis now gives
$$\eqalign{
\bigl| A + B + (C\cup \{x\}) \bigr| &\le |A+B+C| + |A+B+x| - |A'+B+x| \cr
&\le K|A+B| + K|A| = K|A'| \cr
&= K\bigl| A+(C\cup \{x\})\bigr|,
}$$
which completes the proof.\slug

This lemma has an immediate corollary that already has some resemblance to the version of Pl\"unnecke's
theorem that we eventually want to prove.

\proclaim Corollary {\advthm}. If $|A_0 + B|\le K_0|A_0|$, then there is $A\subseteq A_0$ nonempty
and $K\le K_0$ such that $|A+rB| \le K^r |A|$ for all nonnegative integers $r$.

\proof The previous lemma gives $A$ and $K$ such that $|A+B+C|\le K|A+C|$ for all $C$. We now induce
on $r$. The case $r=0$ is obvious. Now for $r>0$ applying the lemma again with $C = (r-1)B$ and
then invoking the induction hypothesis gives
$$|A+rB| = |A+B+(r-1)B| \le K|A+(r-1)B| \le K^r|A|.\noskipslug$$

To get from this corollary to the full theorem, we will use a certain inequality of Rusza (1996), which
is often called the Ruzsa triangle inequality due to its similarity to the triangle inequality in
metric spaces.

\parenproclaim Lemma {\advthm} (Ruzsa triangle inequality). Let $U$, $V$, and $W$ be finite
subsets of an abelian group. Then
$$|U|\cdot |V-W| \le |U-V|\cdot |U-W|.$$

\proof For each $x\in V-W$, pick $v(x)\in V$ and $w(x)\in W$ such that $v(x)-w(x) = x$ (there may
be multiple choices for each of $v(x)$ and $w(x)$, but we arbitrarily pick one).
Define $\phi : U\times (V-W) \to (U-V)\times (U-W)$ by $\phi(u,x) = \bigl(u-v(x), u-w(x)\bigr)$.
If $\bigl(u_1 - v(x_1), u_1 - w(x_2)\bigr) = \bigl(u_2 - v(x_2), u_2 - w(x_2)\bigr)$, then
subtracting the second coordinate of each pair from the first, we find that
$$x_1 = v(x_1) - w(x_1) = v(x_2) - w(x_2) = x_2,$$
and this also implies that $u_1 = u_2$. So $\phi$ gives an injection from $U\times (V-1)$
to $(U-V)\times (U-W)$.\slug

If we define $d(U,V) = |U-V| / \bigl(|U|^{1/2}|V|^{1/2}\bigr)$, then this lemma states
that $d(U,W) \le d(V,U)d(U,W)$. Taking logarithms gives a genuine additive triangle inequality,
but the space of finite subsets of an abelian group does not form a metric space, since
$|U-U|/|U|$ does not equal $1$ in general. In any case, we are now able to prove Ruzsa's
version of Pl\"unnecke's inequality.

\medskip
\noindent{\it Proof of Theorem~{\thmplunnecke}.}\enspace Let $A$ and $K$ be given by
Lemma~{\lempetridis}. By the corollary to that lemma, we have $|A+rB| \le K^r|A|$ and
$|A+sB| \le K^r|A|$. Cardinalities are not changed by taking additive inverses of
everything, so $|{-A}-rB| \le K^r|A|$ and $|{-A}-sB|\le K^r|A|$. Now an application of Ruzsa's triangle
inequality with $U = -A$, $V = rB$ and $W = sB$ yields
$$|A|\cdot|rB-sB| = |{-A}|\cdot|rB-sB| \le |{-A}-rB|\cdot|{-A}-sB| \le K^{r+s}|A|^2,$$
and dividing both sides by $|A|$ proves the theorem.\slug

\advsect Khovanskii's theorem

The goal of this section is to prove the following wonderful theorem of A.~Khovanskii.

\edef\thmkhovanskii{\the\thmcount}
\parenproclaim Theorem {\advthm} (Khovanskii, {\rm 1992}). Let $A$ be a finite
subset of an abelian group and let $f_A(n) = |nA|$ for each $n\in \NN$. Then there exists
$n_A$ and a polynomial $p_A$ such that $f_A(n) = p_A(n)$ for all $n\ge n_A$.

Consider the example when $A = \{0,1,b-1,b\} \subseteq \ZZ$. Then letting $[a,b] = \{x\in \ZZ : a\le x\le b\}$,
we have
$$nA = [0, n] \cup [b-1,\ldots, b+n-1] \cup [2b-2, 2b+n-2] \cup\cdots\cup [n(b-1),nb].$$
This has size $(n+1)^2$ as long as $n< b-1$, but when $n\ge b-1$, $nA$ is simply the set
of integers in the range $[0,bn]$, which has size $bn+1$.

We shall present a combinatorial proof of Khovanskii's theorem,
given by M.~B.~Nathanson and I.~Z.~Ruzsa in 2002.
Let $A = \{a_1,\ldots, a_r\}$ and
for $x = (x_1,\ldots, x_r)\in {\NN_0}^r$, write
$$\sigma(x) = \sum_{i=1}^r x_i\qquad\hbox{and}\qquad
\alpha(x) = \sum_{i=1}^r a_ix_i.$$
Then
$$nA = \{ \alpha(x) : x\in {\NN_0}^r,\, \sigma(x) = n\}.$$
Let $\prec$ be the lexicographic order on ${\NN_0}^r$; that is, $(x_1,\ldots,x_r)\prec (y_1,\ldots,y_r)$
if there is some $i\in [r]$ such that $x_i<y_i$ and $x_j = y_j$ for all $1\le j<i$.
Say that $x$ is {\it useless} if there is $x'\prec x$ such that $\sigma(x') = \sigma(x)$ and
$\alpha(x') = \alpha(x)$; otherwise, say that $x$ is {\it useful}. It is easy to see that
$|nA|$ is exactly the number of useful $x$ with $\sigma(x) = n$. Now let $\le$ denote
the usual partial order on ${\NN_0}^r$, where $(x_1,\ldots,x_r)\le (y_1,\ldots,y_r)$
if $x_i\le y_i$ for all $1\le i\le r$.

\proclaim Lemma {\advthm}. If $x\le y$ and $x$ is useless, then $y$ is useless.

\proof Since $x$ is useless, we can find $x'\prec x$ with $\sigma(x') = \sigma(x)$ and $\alpha(x') = \alpha(x)$.
Let $y' = y+x'-x$. Since $x\le y$, $y-x$ is in ${\NN_0}^r$ and therefore so is $y'$. Then since
$x_i' < x_i$ for the first $i$ at which the two tuples differ, the first coordinate at which $y'$ and
$y$ differ is also $i$ and we have $y_i'<y_i$. Hence $y'\prec y$. Lastly,
$$\sigma(y') = \sigma(y+x'-x) = \sigma(y) + \sigma(x') - \sigma(x) = \sigma(y)$$
and the same is true with $\sigma$ replaced by $\alpha$, so $y$ is useless.\slug

For each useless $x$ we can find a useless $x'$, minimal in the $\le$ order, such that $x'\le x$.
This lemma gives a converse; it says that if $x'$ is a minimal useless element and $x'\le x$, then
$x$ is also useless.

\edef\lemminimalfinite{\the\thmcount}
\proclaim Lemma {\advthm}. For any nonempty $U\subseteq {\NN_0}^r$, the set of minimal elements of
$U$ (with respect to the $\le$ order) is finite.

\proof We perform induction on $r$. If $r=1$, then it is clear that $U$ has exactly one minimal element.
Now let $r>1$ and let $u$ be a minimal element of $u$. Then for every $v\in U$ with $u\not\le v$,
there exists some $i$ and $t< u_i$ such that $v_i = t$. Let
$$V(i,t) = \{v\in U : v_i = t\}$$
for all $1\le i\le r$ and $0\le t<u_i$. Each set $V(i,t)$ is order-isomorphic to a subset of ${\NN_0}^{r-1}$,
so by induction it has finitely many minimal elements. The number of $V(i,t)$ is $r\prod_{i=1}^r u_i$,
and every minimal element of $u$ is either $u$ or a minimal element of some $V(i,t)$, so there are
finitely many minimal elements in $U$.\slug

We will need just one more lemma before we are able to prove Khovanskii's theorem. It shows that
for any fixed $r$-tuple $x'$, the number of $x$ with $x'\le x$ and $\sigma(x) = n$
is given by a polynomial in the variable $n$, with coefficients depending on $x'$ and $r$.

\proclaim Lemma {\advthm}. Let $x'\in {\NN_0}^r$ and for $n\ge \sigma(x')$, let
$$C(x', n) = \{n\in {\NN_0}^r : \sigma(x) = n,\,x'\le x\}.$$
Then
$$\bigl| C(x',n)\bigr| = {n-\sigma(x') + r-1 \choose r-1}.$$

\proof Given $x\in C(x',n)$, map it to $x-x'$. This is a bijection from $C(x',n)$ to $C(0, n-\sigma(x')$.
Now let $m = n-\sigma(x')$. The map $(x_1,\ldots,x_r) \mapsto (x_1,+1,\ldots, x_r+1)$ is a bijection
from $C(0,m)$ to the set of $y\in \NN^r$ such that $y_1+\cdots+y_r = m+r$. Now the map
$$(y_1,\ldots,y_r) \mapsto \{y_1, y_1+y_2, \ldots, y_1+\cdots+y_r\}$$
is a bijection from the previous set to the set of subsets of $[m+r]$ that contain $m+r$. (It is
inverted by the map $\{a_1, \ldots, a_r\} \mapsto (a_1, a_2-a_1, \ldots, a_r-a_{r-1})$.) But this final
set has size ${m+r-1\choose r-1}$, so we are done.\slug

Without further ado, let us now prove Khovanskii's theorem.

\medskip\noindent {\it Proof of Theorem~{\thmkhovanskii}.}\enspace Let $U$ be the set of all
useless $r$-tuples. By Lemma~{\lemminimalfinite}, the set $X\subseteq U$ of all minimal useless
$r$-tuples is finite. Recall that $|nA|$ is the number of useful $x$ such that $\sigma(x) = n$,
so using the notation of the previous lemma, we have
$$|nA| = \bigl| C(0,n)\bigr| - \Bigl| \bigcup_{x'\in X} C(x',n)\Bigr|
=\bigl| C(0,n)\bigr| - \sum_{Y\subseteq X} (-1)^{|Y|} \Bigl| \bigcap_{x'\in Y} C(x',n)\Bigr| $$
But for any $Y\subseteq X$, $\bigcap_{x'\in Y} C(x',n) = C(x_Y, n)$, where $x_Y$ has for its
$i$th coordinate the integer $\max_{y\in Y} y_i$. Hence every term in the inclusion-exclusion
formula depends polynomially on $n$ once
$$n\ge \sum_{1\le i\le r} \max_{x\in X} x_i.\noskipslug$$

\advsect The Balog--Szemer\'edi--Gowers theorem

Let $A$ be a finite nonempty subset of an abelian group. An {\it additive quadruple} in
$A$ is a quadruple $(x,y,z,w)\in A^4$ such that $x+y = z+w$. The {\it additive energy}
$E(A)$ of $A$ is the number of additive quadruples in $A$. If $(x,y,z,w)$ is an additive
quadruple, then $w = x+y-z$, so $E(A)\le |A|^3$. The following lemma roughly states that if
a set has a small sumset, then it has large additive energy.

\proclaim Lemma {\advthm}. Let $A$ be a finite subset of an abelian group and let $C$ be such
that $|A+A|\le C|A|$. Then $E(A)\ge |A|^3/C$.

\proof Define $f : A+A\to \NN$ by letting $f(d)$ equal the number of pairs $(a,b)\in A^2$ such
that $a+b =d$. Then
$$\sum_{d\in A+A} f(d) = |A|^2\qquad\hbox{and}\qquad \sum_{d\in A+A} f(d)^2 = E(A).$$
By the Cauchy--Schwarz inequality,
$$|A|^2 = \sum_{d\ge 1} 1\cdot f(d)
\le \Bigl(\sum_{d\ge 1} 1\Bigr)^{1/2} \Bigl( \sum_{d\ge 1} f(d)^2\Bigr)^{1/2}
\le C^{1/2} |A|^{1/2} E(A)^{1/2}.$$
From here, squaring the whole inequality and then dividing through by $C|A|$ gives the statement
we want to prove.\slug

The converse of this lemma does not hold. For example, we could take the disjoint
union of an arithmetic progression with a geometric progression of the same length.
This set has a large additive energy,
because of the arithmetic progression, but it also has a large sumset, because of the geometric progression.
However, we can pass to a large subset that has a small sumset (namely, the arithmetic progression), and this
is essentially the statement of the Balog--Szemer\'edi--Gowers theorem. We shall state it later on in the
section, after we have given the graph-theoretic definitions and lemmas needed in its proof.

Let $G$ be a bipartite graph with finite vertex sets $X$ and $Y$. For a vertex $v\in X\cup Y$,
let $d(v) = \deg(v)$ denote the {\it degree} of $x$, that is, the number of vertices adjacent to $v$.
For $x_1,x_2\in X$, let $d(x_1,x_2)$ denote the number of paths of length $2$ from $x_1$ to $x_2$. This
is called the {\it codegree} of $x_1$ and $x_2$. The {\it density} of $G$ is the number of edges of $G$
divided by the maximum number of vertices it could have, namely $|X|\cdot|Y|$.

\proclaim Lemma {\advthm}. Let $G$ be a bipartite graph with bipartition $X\cup Y$ and let the density
of $G$ be $\delta$. Then for every $c>0$ there is a subset $X'\subseteq X$ with $|X'|\ge \delta|X|/\sqrt 2$
such that the number of pairs $(x_1,x_2)\in (X')^2$ with $d(x_1,x_2) < c|Y|$ is at most $2c|X|^2/\delta^2$.

\proof Let $y\in Y$ be chosen uniformly at random. Letting $X' = N(y)$, we have
$$\ex\bigl\{ |X'|\bigr\} = \ex_{y\in Y} \bigl\{ d(y)\bigr\} = {1\over |Y|} \sum_{y\in Y} d(y)
= {|E(G)|\over |Y|} = \delta|X|.$$
By convexity of the map $x\mapsto x^2$, we then find that $\ex\bigl\{ |X'|^2\bigr\} \ge \delta^2|X|^2$.
Let
$$B = \bigl\{ (x_1,x_2)\in X^2 : d(x_1,x_2) < c|Y|\bigr\}.$$
The probability that any $(x_1, x_2)\in B$ is also in $(X')^2$ is the same as the probability that
$y$ is connected to both $x_1$ and $x_2$; this is less than $c|Y|$ by the construction of $B$.
So by linearity of expectation, $\ex\bigl\{ |B\cap (X')^2| \bigr\} < c|X|^2$ and in particular, we
have
$$\biggl\{ |X'|^2 - {\delta^2\over 2c} |B\cap(X')^2|\biggr\}
 >\biggl(\delta^2 - {\delta^2\over 2c}\cdot c\biggr)|X|^2 = {\delta^2\over 2}|X|^2.$$
So there is $X'$ such that $|X'| - \bigl(\delta^2/(2c)\bigr) |B\cap (X')^2| \ge \delta^2|X|^2/2$.
This set $X'$ has cardinality at least $\delta |X|/\sqrt 2$ and $|B\cap (X')^2| \le 2c|X'|^2/\delta^2$.\slug

This lemma says that we can find a fairly large subset of $X$ such that almost every pair in this
subset is joined by at least $c|Y|$ paths of length $2$. We would really like to remove the ``almost''
in this statement, but to do so, we will have to pass to paths of length $3$.
First, we need a little lemma about sums of real numbers between $0$ and $1$.

\proclaim Lemma {\advthm}. Let $a_1,\ldots, a_n\in [0,1]$. If $\sum_{i=1}^n a_i \ge an$, then
the number of $i$ such that $a_i\ge b$ is at least
$$ \biggl( {a-b\over 1-b} \biggr) n.$$

\proof Let $I$ be the set of $i$ such that $a_i\ge b$. Then we may write
$$ an \le \sum_{i=1}^n a_i = \sum_{i\in I} a_i + \sum_{i\notin I} a_i
\le |I| + (n-|I|)b.$$
It remains to isolate for $|I|$.\slug

For example, if $b = a/2$, then the number of $i$ with $a_i\ge b$ must be at least $an/2$, and more
generally, if $a= 1-\theta$ and $b=1-\eta$, then the number of such $i$ is at least $(1-\theta/\eta)n$.
If $\eta = 2\theta$, then $a_i\ge 1-2\theta$ at least half the time.
Next we prove the lemma alluded to earlier, concerning paths of length $3$.

\proclaim Lemma {\advthm}. Let $G$ be a bipartite graph with vertex sets $X$ and $Y$ and density $\delta$.
Then there are subsets $X'\subseteq X$ and $Y'\subseteq Y$ such that
$$|X'| \ge {\delta^2\over 8\sqrt 2} |X|\qquad\hbox{and}\qquad
|Y'|\ge {\delta\over 4}|Y|$$
and every $x\in X'$ and $y\in Y'$ are joined by at least $(\delta^6/2^{13}) |X|\cdot |Y|$ paths
of length $3$ in $G$.

\proof By letting $n = |X|$ and setting the $a_i$ equal to the degrees of the vertices in $X$, we have
$a = \delta|Y|$
and we can choose $b=a/2$ in the previous lemma
to deduce that $X$ has a subset $X_1$ such that every vertex in $X_1$ has degree at least
$\delta|Y|/2$ and $|X_1|\ge \delta|X|/2$. Note that the density of the bipartite subgraph with
vertex set $X_1\cup Y$ is $\ge \delta/2$,
so we may invoke the lemma about paths of length $2$ with $\delta$ set to $\delta/2$. It tells us that
for any $c>0$, $X_1$ has a subset $X_2$ of size $\delta|X_1|/2^{3/2}$ such that for all but at most
$8c|X_2|^2/\delta^2$ pairs $(x_1,x_2)\in {X_2}^2$, we have $d(x_1,x_2) \ge c|Y|$.

Now let $\Gamma$ be a graph with vertex set $X_2$ and an edge between $x_1$ and $x_2$ if and only if
$\delta(x_1,x_2)\ge c|Y|$. Note that by this definition, there may be loops in $\Gamma$. In any case,
the average degree in $\Gamma$ is at least $(1-8c/\delta^2)|X_2|$. Invoking the previous lemma again,
at least half the vertices in $X_2$ have degree at least $(1-16c/\delta^2)|X_2|$. Let $X'$ be the set of
all such vertices. Then
$$|X'| \ge {|X_2|\over 2} \ge {\delta|X_1|\over 4\sqrt 2} \ge {\delta^2 |X|\over 8\sqrt 2}.$$
Every vertex in $X_2$ has degree at least $\delta |Y|/2$, so the bipartite subgraph on $X_2\cup Y$
has density at least $\delta/2$ and the average degree of $y\in Y$ is at least $\delta|X_2|/2$. The
previous lemma comes in handy once again, telling us that at least $\delta|Y|/4$ vertices in $Y$ has
degree at least $\delta|X_2|/4$. Let $Y'$ be the set of these vertices.

For any $x\in X'$ and $y\in Y'$, we know that $x$ has at least $(1-16c/\delta^2)|X_2|$ neighbours in $\Gamma$
and $y$ has at least $\delta|X_2|/4$ neighbours in $X_2$. There must be at least
$(\delta/4-16c/\delta^2)|X_2|$ neighbours in commmon. Letting $c=\delta^3/128$, the number of neighbours
in common becomes $\delta|X_2|/8$, and the number of paths of length $3$ from $x$ to $y$ is at least
$$ {\delta |X_2|\over 8} |X_2| \cdot c|Y| = {\delta\over 8}\cdot {\delta^3\over 128}
\cdot {\delta^2\over 4\sqrt 2} |X|\cdot |Y| \ge {\delta^6\over 2^{13}}|X|\cdot |Y|,$$
as foretold.\slug

Let $A$ be a finite subset of an abelian group. We say that an element $d\in A-A$ is {\it $\theta$-popular}
if $\bigl|\bigl\{ (a,b)\in A^2 : b-a=d\bigr\}\bigr|\ge \theta |A|$.

\proclaim Lemma {\advthm}. Let $A$ be a finite subset of an abelian group and let $E(A)$ denote
the additive energy of $A$. If $E(A)\ge c|A|^3$, then there are at least $c|A|/2$ elements of $A-A$
that are $(c/2)$-popular.

\proof Define $f : A-A\to \NN$ by $f(d) = \bigl|\bigl\{ (a,b)\in A^2 : b-a=d\bigr\}\bigr|$.
Note that
$$\sum_{d\in A-A} f(d) = |A|^2\qquad\hbox{and}\qquad \sum_{d\in A-A} f(d)^2 = E(A).$$
Let $P$ be the set of $(c/2)$-popular differences and let $U = (A-A)\setminus P$. Then
$$\eqalign{
c|A|^3 &\le E(A)\cr
&= \sum_{d\in A-A} f(d)^2 \cr
&= \sum_{d\in P} f(d)^2 + \sum_{d\in U} f(d)^2 \cr
&\le |P|\cdot |A|^2 + {c|A|\over 2} \sum_{d\in U} f(d) \cr
&= |P|\cdot |A|^2 + {c|A|^3\over 2}, \cr
}$$
which implies that $|A|\ge c|A|/2$.\slug

We are, at long last, ready to state and prove the Balog--Szemer\'edi--Gowers theorem. It is
named for A.~Balog and E.~Szemer\'edi, who first proved it in 1994, as well as for W.~T.~Gowers,
who gave an alternative proof in 1998 that yielded power-type bounds, a vast improvement to the tower-type
bounds in Balog and Szemer\'edi's proof.

\parenproclaim Theorem {\advthm} (Balog--Szemer\'edi, {\rm 1994;} Gowers, {\rm 1998}).
Let $A$ be a finite subset of an abelian group with $E(A)\ge c|A|^3$. Then $A$ has a subset
$A'$ of size at least $c'|A|$ such that $|A'-A'|\le C|A'|$, where $c'$ and $C$ have a power-type
dependence on $c$. In particular, we may take
$$ c' = {c^4\over 2^8}\qquad\hbox{and}\qquad C = {2^{68}\over c^{36}}.$$

\proof Define a bipartite graph $G$ with both vertex sets being copies of $A$ (though
we'll call one of these copies $B$ to avoid any confusion). Join $a\in A$ to
$b\in B$ if and only if $b-a$ is $(c/2)$-popular. Each $(c/2)$-popular difference leads to at least
$c|A|/2$ edges, so by the lemma, there are at least $c^2|A|^2/4$ edges. Let $\delta = c^2/4$.
By the lemma concerning paths of length $3$, there exist $A'\subseteq A$ and $B'\subseteq B$
with
$$ |A'| \ge {\delta^2\over 8\sqrt 2} |A|\qquad\hbox{and}\qquad
|B'|\ge {\delta \over 4} |A|$$
such that every $a\in A$ and $b\in B$ are joined by at least $\delta^6|A|^2/2^{13}$ paths of length $3$
in $G$.

Given a path $(a,u,v,b)$ of length $3$, we have $b-a = u-a - (u-v) + b-v$, where $u-a$, $u-v$, and $b-v$
are all $(c/2)$-popular. Hence the number of ways of writing $b-a$ as $r_1-s_1 - (r_2-s_2) + r_3 - s_3$
is at least
$${\delta^6\over 2^{13}} |A|^2 \biggl( {c\over 2} |A|\biggr)^3 = {c^{15}\over 2^{28}} |A|^5.$$
But the number of possible $(r_1,s_1,r_2,s_2,r_3,s_3)$ is $|A|^6$. So
$$ |B'-A'|\cdot {c^{15}\over 2^{28}} |A|^5 \le |A|^6 $$
and $|B'-A'|\le 2^{28}|A|\over c^{15}$. Now the Ruzsa triangle inequality with $U = B'$ and
$V = W = A'$ tells us that $|B'|\cdot |A'-A'| \le |B'-A'|^2$, so
$$|A'-A'|\le {|B'-A'|^2\over |B'|} \le {2^{56}|A|^2\over c^{30}}\cdot {2^4\over c^2|A|}={2^{60}\over c^{32}}|A|.$$
But recall that
$$|A'| \ge {\delta^2\over 8\sqrt 2} |A| \ge {c^4\over 2^8}|A|,$$
so $|A'-A'| \le 2^{68}|A'|/c^{36}$.\slug

\advsect Basics of discrete Fourier analysis

Let $Z$ be a finite abelian group. A {\it character} on $Z$ is a homomorphism from
$Z$ to the multiplicative group $\C\setminus\{0\}$. If $\chi$ is such a homomorphism, then
$\bigl|\chi(x)\bigr| = 1$ for every $x\in Z$, so we can actually regard $\chi$ as being
a function from $Z$ to the unit circle $\TT = \{z\in \C : |z| = 1\}$. The pointwise product
of two characters gives another character, and the function $\chi_0$ that sends every $x\in Z$
to $1$ has the property that if $\chi$ is any character, then $\chi\chi_0 = \chi = \chi_0\chi$.
Lastly, we note that multiplication is commutative and
for any character $\chi$, the product of $\chi$ with $\bar\chi$
gives $\chi_0$, so the set of characters on $Z$ is an abelian group. This is called the
{\it dual group} or sometimes {\it Pontryagin dual} of $Z$ and is denoted $\hat Z$.

We define an inner product on the space $\CC^Z$ of functions from $Z$ to $\CC$ by setting
$$\langle f,g\rangle = \ex_x f(x) \bar{g(x)},$$
where $\ex_{x\in Z} F(x) = |Z|^{-1} \sum_{x\in Z} F(x)$.
We can also make $\hat Z$ into an inner product space by letting
$$\langle \hat f, \hat g\rangle = \sum_{\chi\in \hat Z} \hat f(\chi)\bar{\hat g(\chi)};$$
note that this time we do not normalise by dividing by $|Z|$.
Two functions $f$ and $g$ in an inner product space are said to be {\it orthogonal} if $\langle f,g\rangle = 0$.
The first lemma we'll prove concerns certain orthogonality relations.

\parenproclaim Lemma {\advthm} (Orthogonality relations). Let $Z$ be a finite abelian group.
\medskip
\item{a)} If $\chi_1$ and $\chi_2$ are characters on $Z$, then
$$\langle \chi_1,\chi_2 \rangle = \ex_{x\in Z} \chi_1(x)\bar{\chi_2(x)}
  = \cases{ 1, & if $\chi_1 = \chi_2$;\cr 0, & if $\chi_1 \ne\chi_2$.}.$$
\smallskip
\item{b)} For $x,y\in Z$, we have
$$\sum_{\chi\in \hat Z} \chi(x)\bar{\chi(y)}
  = \cases{ |\hat Z|, & if $x=y$;\cr 0, & if $x\ne y$,}$$
\medskip

\proof Let $\chi_0$ denote the trivial character. We have
$$\ex_{x\in Z} \chi_0(x) = \ex_{x\in Z} 1 = 1.$$
On the other hand, let $\chi$ be a nontrivial character and let $u\in Z$ be such that $\chi(u)\ne 1$. Then from
$$\ex_{x\in Z} \chi(x) = \ex_{x\in Z} \chi(u+x) = \ex_{x\in Z}\chi(u)\chi(x) = \chi(u)\ex_{x\in Z}\chi(x),$$
it follows that $\ex_{x\in Z}\chi(x) = 0$. Now consider $\chi_1\bar{\chi_2}$. If $\chi_1 = \chi_2$, then
this is the trivial character and from our first observation, $\langle \chi_1,\chi_2\rangle = 0$. Otherwise,
we are in the second case and the inner product is zero. This proves part (a).

Part (b) is proven similarly. Note that $\chi(x)\bar{\chi(y)} = \chi(x-y)$. If $x=y$, then we have
$$\sum_{\chi\in\hat Z} \chi(x-y) = \sum_{\chi\in\hat Z} \chi(0) = |\hat Z|.$$
If $x\ne y$, then there is some $\psi\in \hat Z$ such that $\psi(x-y)\ne 1$, and then writing
$$\sum_{\chi\in \hat Z} \chi(x-y) = \sum_{\chi\in\hat Z} \psi(x-y)\chi(x-y)
= \psi(x-y)\sum_{\chi\in\hat Z} \chi(x-y),$$
we see that $\sum_{\chi\in\hat Z} \chi(x-y)$ must be zero.\slug

Since the space of functions from $Z\to \CC$ has dimension $|Z|$, there are at most $|Z|$ characters.
Every finite abelian group can be written as a direct product of cyclic groups, and we shall use this fact
to show that there are exactly $|Z|$ characters; i.e., $|\hat Z| = |Z|$. For brevity, let $\ZZ_n$ stand for $\ZZ/n\ZZ$ for all $n\in\NN$,
and for $\alpha\in \RR$, let $e(\alpha) = e^{2\pi i\alpha}$.

\proclaim Lemma \advthm.
The set of characters on a finite abelian group $Z$ spans the space of functions from $Z$ to $\CC$.

\proof Write $Z \cong \ZZ_{n_1} \times \ZZ_{n_2}\times \cdots\times \ZZ_{n_r}$. For every
$u = (u_1,\ldots, u_r)\in Z$, the function $\chi_u : Z\to \CC$ that takes
$$(x_1,\ldots, x_r) \mapsto \prod_{i=1}^r e\biggl( {u_ix_i\over n_i}\biggr)$$
is a character (this is easy to show, since $e(x_i + x_i') = e(x_i)e(x_i')$). If $u\ne u'$,
then $\chi_u\ne\chi_{u'}$ and $\langle \chi_u,\chi_{u'}\rangle = 0$ by the previous lemma. Hence
the set $\{\chi_u : u\in Z\}$ comprises $|Z|$ linearly independent elements in a space of dimension $|Z|$,
which must be spanning.\slug

Note that the map $u\mapsto \chi_u$ gives an isomorphism from $Z$ to $\hat Z$.
The two main examples we shall consider are when $Z = \ZZ_n$ and when $Z = \FF_p^n$. In the first
case, $\chi_u$ takes $x\mapsto e(ux/N)$ and in the second case $\chi_u$ takes $x\mapsto e(u\cdot x/p)$.
(Recall that if $u = (u_1,\ldots, u_n)$ and $x = (x_1,\ldots, x_n)$ then the dot product $u\cdot x$
is the sum $\sum_{i=1}^n u_ix_i$.)

For a finite set $X$, we shall denote by $L_p(X)$ the normed vector space of all functions $f:X\to \CC$
under the norm
$$\norm{f}_p = \bigl( \ex_{x\in X} \bigl| f(x)\bigr|^p \bigr)^{1/p}.$$
The notation $l_p(X)$ denotes the same set, but with the norm
$$\norm{f}_p = \Bigl( \sum_{x\in X} \bigl| f(x)\bigr|^p \Bigr)^{1/p}$$
instead. Note that $\norm{f}_2^2 = \langle f,f\rangle$ and $\norm{f}_\infty = \max_{x\in X} \bigl|f(x)\bigr|$.

\medskip\boldlabel The Fourier transform.
If $f:Z\to \CC$, the {\it Fourier transform} of $f$ is the function $\hat f : \hat Z\to\CC$ given by
$$\hat f(\chi) = \langle f,\chi\rangle = \ex_{x\in Z} f(x) \bar{\chi(x)}.$$
We shall now state and prove an identity which is extremely useful in Fourier analysis.

\parenproclaim Lemma {\advthm} (Parseval's identity). Let $Z$ be a finite abelian group
and let $f$ and $g$ be functions from $Z$ to $\CC$.
If $\hat f$ and $\hat g$ are the Fourier transforms of $f$ and $g$ respectively, then
$\langle f,g\rangle = \langle \hat f, \hat g\rangle$.

\proof First, we expand all the definitions and rearrange sums to obtain
$$\eqalign{
\langle \hat f, \hat g\rangle &= \sum_{\chi\in \hat Z} \hat f(\chi)\bar{\hat g(\chi)} \cr
&=  \sum_{\chi\in \hat Z} \ex_{x\in Z} f(x)\bar{\chi(x)} \ex_{y\in Z} g(y)\bar{\chi(y)} \cr
&=   \ex_{x\in Z} f(x) \ex_{y\in Z} g(y) \sum_{\chi\in \hat Z}\bar{\chi(x)}\bar{\chi(y)}.\cr
}$$
By the second orthogonality relation and the fact that $|\hat Z| = |Z|$, the inner sum equals $|Z|$ when
$y = x$ and $0$ otherwise. Thus for any given $x$, we have
$$\ex_{y\in Z} g(y) \sum_{\chi\in \hat Z}\chi(x)\bar{\chi(y)} = g(x),$$
which is exactly what we need for the whole thing to equal $\langle f,g\rangle$.\slug

We also have the following inversion formula that recovers the original function $f$ from $\hat f$.

\parenproclaim Lemma {\advthm} (Fourier inversion formula). Let $Z$ be a finite abelian group and
let $f:Z\to \CC$. We have
$$f(x) = \sum_{\chi\in \hat Z} \hat f(\chi)\chi(x).$$

\proof We expand
$$\sum_{\chi\in \hat Z} \hat f(\chi)\chi(x) = \sum_{\chi\in \hat Z} \ex_{y\in Z} f(y) \bar{\chi(y)} \chi(x)
= \ex_{y\in Z} f(y) \sum_{\chi\in \hat Z} \chi(x)\bar{\chi(y)},$$
and we noted in the proof of Parseval's theorem that the right-hand side is exactly $f(x)$.\slug

The inversion formula tells us that two functions from $Z\to \CC$ are equal if and only if their Fourier
transforms are equal. For the next lemma, which concerns dilates of a function $f: Z\to \CC$, we will
establish some notation. If $x\in Z$, we define $nx$ recursively for all integers $n\ge 0$ by $0x = 0$
and $nx = x + (n-1)x$. We then set $nz = - (-n)x$ for all negative integers. Since the group operation
on characters is multiplication, we will not write $n\chi$ but instead $\chi^n$ for $\chi$ multiplied
with itself $n$ times.

\parenproclaim Lemma {\advthm} (Dilation rule). Let $Z$ be a finite abelian group and let $a\in \ZZ$
be an integer that is coprime to $|Z|$. Denote the multiplicative inverse of $a$ modulo $|Z|$ by $a^{-1}$.
Letting $f_a : Z\to \CC$ be the function given by $f_a(x) = f(a^{-1} x)$, we have
$\hat{f_a}(\chi) = \hat f(\chi^a)$.

\proof We have
$$\hat{f_a}(\chi) = \ex_{x\in Z} f_a(x) \bar{\chi(x)} = \ex_{x\in Z} f(a^{-1} x)\bar{\chi(x)}.$$
Since $x\mapsto ax$ is a bijection from $Z$ to itself, we can replace $x$ formally by $ax$ in the above
to get
$$\hat{f_a}(\chi) = \ex_{x\in Z} f(x)\bar{\chi(ax)} = \ex_{x\in Z} f(x) \bar{\chi(ax)}^a
= \hat f(\chi^a).\noskipslug$$

\medskip\boldlabel Convolutions.
For functions $f$ and $g$ from a finite abelian group $Z$ to $\CC$, the {\it convolution} $f*g$ is
defined by
$$ (f*g)(x) = \ex_{y+z = x} f(y)g(z).$$
If instead $\hat f$ and $\hat g$ are functions from $\hat Z$ to $\CC$, then
$$ (\hat f * \hat g)(\chi) = \sum_{\chi_1\chi_2 = \chi} \hat f(\chi_1)\hat g(\chi_2).$$

\parenproclaim Lemma {\advthm} (Convolution law). Let $Z$ is a finite abelian group and $f,g : \Z\to \CC$.
For all $\chi\in \hat Z$, we have $\hat{f*g} (\chi) = \hat f(\chi)\hat g(\chi)$.

\proof We start by expanding
$$\hat{f*g}(\chi) = \ex_{x\in Z} (f*g)(x) \bar{\chi(x)} = \ex_{x\in Z} \ex_{y+z=x} f(y)g(z)\bar{\chi(y)\chi(z)}.$$
But note that $x$ does not appear in the summand anymore, meaning that we can simply rewrite this as
an expectation
over {\it all} $y$ and $z$ (their sum will equal $x$ for some $x\in Z$). Thus we can conclude that
$$\hat{f*g}(\chi) = \ex_{y\in Z}\ex_{z\in Z} f(y)g(z)\bar{\chi(y)\chi(z)} = \hat f(\chi)\hat g(\chi),$$
which is what we wanted to show.\slug

If $A$ is a subset of a finite abelian group $Z$, we associate to $A$ the {\it characteristic function}
$\one_A : Z\to \CC$ given by
$$\one_A(x) = \cases{ 1, & if $x\in A$;\cr 0, & otherwise.}$$
When it will not cause confusion, we will abuse notation and write $A(x)$ instead of $\one_A(x)$.
With the definitions and results stated above, we are now in a position to translate properties of $A$ to
Fourier-analytic statements about the function $\one_A$. First off, note that
$$\bignorm{\hat{\one_A}}_2^2 = \norm{\one_A}_2^2 = \ex_{x\in Z} A(x)^2 = \ex_{x\in Z} A(x)
= {|A|\over |Z|}.$$
If $|A|/|Z| = \delta$, then we will say that $A$ has {\it density} $\delta$. Another way of expressing $\delta$
in terms of $\one_A$ is
$$\hat{\one_A}(\chi_0) = \ex_{x\in Z} A(x)\chi_0(x) = \ex_{x\in Z} A(x) = \delta,$$
where $\chi_0$ is the trivial character.

The fact that $\norm{\one_A}_2^2 = \delta$ means that if we sample $x$ and $y$ from $Z$ and
condition that $x=y$, the probability that $x$ (or $y$) is in $A$ is $\delta$. The reason for stating this
obvious fact in such a stilted fashion is that it generalises by way of the convolution operation we defined
earlier; that is, what if we are interested in the probability that a $4$-tuple $(x,y,z,w)$ satisfying $x+y=z+w$
is a member of $A^4$? Well, the convolution law gives
$$\eqalign{
\ex_{x+y=z+w} A(x)A(y)A(z)A(w) &= \ex_{u\in Z} \ex_{x+y=u}\ex_{z+w=u} A(x)A(y)A(z)A(w) \cr
&= \ex_{u\in Z} (\one_A*\one_A)(u)^2 \cr
&= \norm{\one_A*\one_A}_2^2 \cr
&= \bignorm{\hat{\one_A*\one_A}}_2^2 \cr
&= \bignorm{\bigl(\hat{\one_A}\bigr)^2}_2^2 \cr
&= \sum_{\chi\in \hat Z} \bigl| \hat{\one_A}(\chi)\bigr|^4 \cr
&= \norm{\one_A}_4^4, \cr
}$$
and more generally, the probability that a tuple $(a_1,\ldots,a_k, b_1,\ldots,b_k)\in Z^{2k}$
with $a_1+\cdots+a_k=b_1+\cdots+b_k$ is a member of $A^{2k}$ is $\norm{\one_A}_{2k}^{2k}$. Recall that
the number of $(x,y,z,w)\in A^4$ with $x+y = z+w$
is the additive energy $E(A)$ of $A$; we just showed above that
$E(A) = |Z|^3 \norm{\one_A}_4^4$.

\section References

\bye

### 2022-03-27 misc/expo/additive.tex
Photos

@@
 
+\advsect Freiman homomorphisms and Ruzsa embedding lemmas
+
+We have been talking a lot about finite subsets of abelian groups, so it's about time we gave
+a special name for these sets. An {\it additive set} is a pair $(A,Z)$ where $Z$ is an abelian
+group and $A$ is a finite subset of $Z$. In almost all cases we shall suppress mention of $Z$
+and speak simply of an additive set $A$.
+
+Let $A$ and $B$ be additive sets (not necessarily in the same group). A function $\phi:A\to B$
+is a {\it Freiman homomorphism of order $k$} if for any $2k$ elements $a_1,\ldots,a_k,b_1,\ldots,b_k\in A$
+satisfying
+$a_1+\cdots +a_k = b_1+\cdots+b_k$, we also have $\phi(a_1)+\cdots+\phi(a_k) =\phi(b_1)+\cdots+\phi(b_k)$.
+The existence of such a $\phi$ tells us that the additive relationships among
+the elements of $A$ are approximately the same as the additive relationships in $B$. Note that a Freiman
+homomorphism of order $k$ is also a Freiman homomorphism of order $k'$ for all $k'>k$.
+If we write simply ``Freiman
+homomorphism'', then we mean a Freiman homomorphism of order $2$, and we will sometimes write ``$k$-homomorphism''
+instead of ``Freiman homomorphism of order $k$''.
+
+A $k$-homomorphism $\phi:A\to B$ induces a map from $kA$ to $kB$. Indeed, if we define $\psi: kA\to kB$
+by
+$$\psi(a_1+\cdots+a_k) = \phi(a_1) + \cdots + \phi(a_k),$$
+then the definition of $k$-homomorphism ensures that $\phi$ is well-defined. More generally, if $\phi$
+is a $k(r+s)$-homomorphism, then defining $\psi : rA-sA \to rB-sB$ by
+$$\phi(a_1+\cdots+a_r-a_1'-\cdots-a_s') = \phi(a_1)+\cdots+\phi(a_r) -\phi(a_1')-\cdots-\phi(a_s'),$$
+it is easy to see that $\psi$ is a $k$-homomorphism. It is also clear that if $\phi$ is the
+restriction of a group homomorphism to $A$, then $\phi$ is a $k$-homomorphism for every $k$, and
+if $\phi:A\to B$ and $\psi:B\to C$ are $k$-homomorphisms, then $\psi\circ \phi : A\to C$ is a $k$-homomorphism
+as well.
+
+If $\phi : A\to B$ is a $k$-homorphism with an inverse that is also a $k$-homorphism, then we say that
+$\phi$ is a {\it Freiman isomorphism of order $k$} or, more briefly, a {\it $k$-isomorphism}. These
+maps preserve a lot of the properties that are of interest in additive combinatorics. Suppose that $A$
+is $k$-isomorphic to $B$. Then
+\medskip
+\item{i)} $|kA| = |kB|$;
+\smallskip
+\item{ii)} $|rA-sA| = |rB-sB|$ for all $r$ and $s$ with $r+s\le k$;
+\smallskip
+\item{iii)} $E(A) = E(B)$ if $k=2$; and
+\smallskip
+\item{iv)} if $k=2$ and $A$ contains an arithmetic progression of length $r$, then so does $B$.
+\medskip
+Of these statements, the only slightly nontrivial one is (iv). It follows from the fact that an
+arithmetic progression is a sequence $(a_1,\ldots,a_r)$ with $a_i+a_i = a_{i-1}+a_{i+1}$ for all $1<i<r$.
+Taking $\phi$ of both sides gives
+$$\phi(a_i) + \phi(a_i) = \phi(a_i+a_i) = \phi(a_{i-1}+a_{i+1}) = \phi(a_{i-1}) + \phi(a_{i+1}).$$
+Let us now state and prove two important ``embedding lemmas''. These lemmas tell us that if
+$kA-kA$ is not too large, then the additive set $A$ is $k$-isomorphic to a subset of a group that
+is simpler, in some sense.
+
+\parenproclaim Lemma {\advthm} (Ruzsa embedding lemma in $\FF_p^N$). Let $k\ge 1$ and $C>0$.
+Let $A\subseteq \FF_p^N$ be such that $|kA-kA|\le C|A|$. Then $A$ is isomorphic to a subset of
+$\FF_p^n$, where $p^{n-1} < C|A|$.
+
+\proof Let $X$ be a subspace of $\FF_p^N$ chosen uniformly at random from all subspaces of
+codimension $n$, where $n$ will be specified later. Let $q : \FF_p^N \to \FF_p^N/X$ be the
+quotient map taking $v$ to $v+X$. Since $q$ is a group homomorphism, its restriction to $A$
+is a Freiman homomorphism
+of order $k$. The only way that the restriction of $q$ to $A$ can fail to be a {\it $k$-isomorphism}
+is if there exist $a_1,\ldots,a_k,b_1,\ldots,b_k\in A$ such that $a_1+\cdots+a_k \ne b_1+\cdots+b_k$
+but $q(a_1)+ \cdots + q(a_k) = q(b_1)+\cdots+q(b_k)$. This is equivalent to
+$a_1+\cdots+a_k - b_1-\cdots-b_k$ being in $X$. For any nonzero $y\in \FF_p^N$, we have
+$$\pr\{y\in X\} = {p^{n-n}-1\over p^N-1} < {1\over p^n},$$
+and since $|kA-kA|\le C|A|$,
+$$\pr\biggl\{\bigcup_{y\in (kA-kA)\setminus\{0\}} \{y\in X\}\biggr\} \le \sum_{y\in (kA-kA)\setminus 0}
+\pr\{y\in X\} < {C|A|\over p^n},$$
+the probability that $q$ is a $k$-isomorphism is nonzero so long as $p^n\ge C|A|$. Taking the minimal such $n$,
+we have $p^{n-1} < C|A|$ and
+can conclude there is a subspace $X$ of dimension $n$ that contains a set $k$-isomorphic to $A$.\slug
+
 \advsect Basics of discrete Fourier analysis
@@
 gives $\chi_0$, so the set of characters on $Z$ is an abelian group. This is called the
-{\it dual group} or sometimes {\it Pontryagin dual} of $Z$ and is denoted $\hat Z$.
+{\it dual group} (or sometimes the {\it Pontryagin dual}) of $Z$ and is denoted $\hat Z$.
 
@@
 \proof Write $Z \cong \ZZ_{n_1} \times \ZZ_{n_2}\times \cdots\times \ZZ_{n_r}$. For every
-$u = (u_1,\ldots, u_r)\in Z$, the function $\chi_u : Z\to \CC$ that takes
+$u = (u_1,\ldots, u_r)\in Z$, the function $\chi_u : Z\to \CC$ that maps
 $$(x_1,\ldots, x_r) \mapsto \prod_{i=1}^r e\biggl( {u_ix_i\over n_i}\biggr)$$
@@
 The two main examples we shall consider are when $Z = \ZZ_n$ and when $Z = \FF_p^n$. In the first
-case, $\chi_u$ takes $x\mapsto e(ux/N)$ and in the second case $\chi_u$ takes $x\mapsto e(u\cdot x/p)$.
+case, $\chi_u$ maps $x\mapsto e(ux/N)$ and in the second case $\chi_u$ maps $x\mapsto e(u\cdot x/p)$.
 (Recall that if $u = (u_1,\ldots, u_n)$ and $x = (x_1,\ldots, x_n)$ then the dot product $u\cdot x$

### 2022-03-30 about/vita.tex
It is 2022 omg
New paper

@@
 \smallskip
-Supervisor: Luc Devroye. CGPA: 4.00/4. Thesis topic: ``Structural properties of Galton--Watson trees.''
+Supervisor: Luc Devroye. CGPA: 4.00/4.
+Thesis: ``Structural properties of conditional Galton--Watson trees.''
 \medbreak
@@
 
+\begingroup\frenchspacing
 \papercount=3
@@
 \papitem (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
-Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. To appear in {\sl Discrete Mathematics
-and Theoretical Computer Science}, 11~pp. [arXiv:2105.12046]
+Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl Discrete Mathematics
+and Theoretical Computer Science} {\bf 24},1 (2022), \#7, 16~pp. [arXiv:2105.12046]
 \papitem (with Anna M. Brandenberger and Luc Devroye)
@@
 \endgroup
+\endgroup%end the frenchspacing
 

### 2022-03-30 misc/expo/additive.tex
New paper

@@
 $a_1+\cdots+a_k - b_1-\cdots-b_k$ being in $X$. For any nonzero $y\in \FF_p^N$, we have
-$$\pr\{y\in X\} = {p^{n-n}-1\over p^N-1} < {1\over p^n},$$
+$$\pr\{y\in X\} = {p^{N-n}-1\over p^N-1} < {1\over p^n},$$
 and since $|kA-kA|\le C|A|$,
@@
 can conclude there is a subspace $X$ of dimension $n$ that contains a set $k$-isomorphic to $A$.\slug
+
+Next, we will prove an embedding lemma for subsets of $\ZZ$. In this setting the ``simpler'' group
+that we pass to is a cyclic group, but first we need a lemma that essentially says that passing
+to a cyclic group preserves the additive structure.
+
+\proclaim Lemma {\advthm}. Let $k$ and $n$ be positive integers and let $A\subseteq \ZZ$ be a set of
+diameter less than $n/k$; i.e., $\max A - \min A < n/k$. Then the map $\phi : A\to \ZZ_n$ that
+sends $a\mapsto a\bmod n$ is a Freiman isomorphism of order $k$.
+
+\proof The function $\phi$ is the restriction of a group homomorphism to $A$, so it is a Freiman
+homomorphism of every order. Now suppose that $\min A = r$ and $\max A = s$, so that $s-r < n/k$.
+Then if $a_1,\ldots,a_k,b_1,\ldots,b_k$ are such that $\phi(a_1)+\cdots+\phi(a_k) = \phi(b_1)+\cdots+\phi(b_k)$,
+then $a_1+\cdots+a_k -b_1-\cdots-b_k$ is at most $k(s-r) < n$ but also a multiple of $n$.
+Hence this difference is zero and $a_1 + \cdots + a_k = b_1+\cdots+b_k$.\slug
+
+As promised, here is the embedding lemma for subsets $A\subseteq \ZZ$. This time, we cannot simply
+say anything about the whole set $A$, but we can about a large subset of it.
+
+\parenproclaim Lemma {\advthm} (Ruzsa embedding lemma in $\ZZ$). Let $k\in \NN$ and $C\in \RR$.
+If $A\subseteq \ZZ$ with $|kA-kA|\le C|A|$, then there exists $A'\subseteq A$ with $|A'|\ge |A|/k$
+such that $A'$ is $k$-isomorphic to a subset of $\ZZ/N\ZZ$, where $N$ is a prime between
+$2C|A|$ and $4C|A|$.\slug
+
+\proof It is easily checked that translation is a Freiman isomorphism of every order, so we may assume
+that $A\subseteq \NN_0$. Let $p$ be a prime bigger than $k\max A$ and consider the
+sequence of functions
+$$\ZZ \buildrel{q_1}\over\longrightarrow \ZZ_p \buildrel{\mu_r}\over\longrightarrow \ZZ_p
+\buildrel\iota\over\longrightarrow \ZZ \buildrel{q_2}\over\longrightarrow \ZZ_N,$$
+where $q_1$ is reduction modulo $p$, $\mu_r$ is multiplication by a random nonzero element $r\in \ZZ_p$,
+$\iota$ takes an element of $\ZZ_p$ to its representative in $[0,p)$, and $q_2$ is reduction mod $N$.
+The only map here that is not a group homomorphism is $\iota$, so to show that the composition
+$\phi = q_2\circ \iota\circ \mu_r\circ q_1$ restricted to $A$ is a $k$-homomorphism, it suffices to worry
+about $\iota$. Split the interval $\{0,1,\ldots,p-1\}$ into $k$ subintervals $J_1,\ldots,J_k$,
+each of diameter less than $p/k$ and let $I_j = \iota^{-1}(J_j)$ for each $1\le j\le k$. By the previous
+lemma, the map $\iota^{-1}$ restricted to any $J_j$ is a $k$-isomorphism, so $\iota$ restricted
+to $I_j$ is a $k$-isomorphism as well. Now let
+$$A_{j,r} = \bigl\{a\in A : \mu_r\bigl(q_1(a)\bigr) \in I_j\bigr\}$$
+and note that $\mu_r\bigl(q_1(a)\bigr) = ra \bmod p$. Thus the composition $\phi$ is a $k$-homomorph\-ism
+when restricted to any $A_{j,r}$.
+
+We shall now show that with positive probability, the restriction of $\phi$ to every $A_{j,r}$ is
+a $k$-isomorphism. For fixed $r$, the restriction to $A_{j,r}$ is not a $k$-isomorphism
+if there exist $j$ and $a_1,\ldots,a_k,b_1,\ldots,b_k\in A_{j,r}$ such that
+$a_1+\cdots+a_k \ne b_1+\cdots+b_k$ but $\phi(a_1)+\cdots+\phi(a_k) = \phi(b_1)+\cdots+\phi(b_k)$.
 

### 2022-04-13 about/vita_professional.tex
Shovelling trouble and new title for indepno

--- 2022-03-30 about/vita.tex
@@
 
-\leftright{Citizen of Canada and Singapore}{3660 Rue Hutchison, Apt.\ 4}
+\leftright{+1\thinspace\up(825\up)\thinspace 440-0681}{3660 Rue Hutchison, Apt.\ 4}
 \leftright{Email: marcel\_goh@yahoo.ca}{Montr\'eal, QC H2X 2H3, Canada}
-\leftright{Webpage: marcelgoh.ca}{+1\thinspace\up(825\up)\thinspace 440-0681}
+\leftright{Webpage: marcelgoh.ca}{}
 
@@
 \begingroup\frenchspacing
-\papercount=3
+\papercount=4
 \pubbegin P
+\papitem (with Luc Devroye and Rosie Y. Zhao)
+On the peel number and the leaf-height of a Galton--Watson tree.
+To appear in {\sl Combinatorics, Probability and Computing}, 20~pp. [arXiv:2106.14389]
 \papitem (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
@@
 \begingroup\parindent=10pt
-\thing (with Luc Devroye and Rosie Y. Zhao)
-The independence number of a Bienaym\'e--Galton--Watson tree
-and related parameters. {\sl arXiv preprint 2106.14389}, 20~pp.
-\smallskip
 \thing (with Jad Hamdan and Jonah Saks)
@@
 
+\sectheader{Selected Software Projects}
+
+\leftright{\bf Tlingit verb parsing}{\bf February -- April 2021}
+\smallskip
+A rule-based program that finds regularity in Tlingit verbs, written in Python.
+
+\noindent[{\tt https://github.com/marcelgoh/tlingit-verb-parsing}]
+\medbreak
+
+\leftright{\bf Sorting algorithms on manifolds}{\bf May -- August 2020}
+\smallskip
+Research project under the supervision of Michael Lipnowski.
+Studied algorithms that investigate the topology of group actions on locally symmetric spaces.
+Work involved writing of code in OCaml and PostScript that generates illustrations of certain quotient
+spaces, in various models of hyperbolic geometry.
+
+\noindent[{\tt https://github.com/marcelgoh/manifold-sorting}]
+\medbreak
+
+\leftright{\bf Bytecode compiler and interpreter}{\bf February -- June 2019}
+\smallskip
+Under the supervision of Adam Dingle, wrote a bytecode compiler and virtual machine for a subset of Python.
+OPythn includes support for lists, dictionaries, named and anonymous functions, objects, and classes.
+It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing.
+
+\noindent[{\tt https://github.com/marcelgoh/opythn}]
+\medbreak
+
+\leftright{\bf Alonzo $\lambda$-calculus interpreter}{\bf May -- June 2019}
+\smallskip
+Wrote an interpreter for the untyped $\lambda$-calculus that performs $\alpha$-conversion
+and $\beta$-reduction. Includes an interactive environment that allows users to bind names to
+terms. Implemented in Haskell.
+
+\noindent[{\tt https://github.com/marcelgoh/alonzo}]
+\medbreak
+
+\leftright{\bf Escape-time fractals}{\bf May 2019}
+\smallskip
+Made an interactive graphical program to generate and explore fractals. Implemented using the SDL2
+graphics library and written in \CEE. (HackPrague 2019)
+
+\noindent[{\tt https://github.com/mattonicorp/mattoni}]
+\medbreak
+
+\leftright{\bf Nim}{\bf January 2019}
+\smallskip
+Implemented the game of Nim graphically using Processing. Worked on game logic and created the soundtrack
+using an $8$-bit sequencer. (ConUHacks IV)
+
+\noindent[{\tt https://github.com/conudihedral4/nim}]
+\medbreak
+
+\leftright{\bf CourseTalk}{\bf November 2018}
+\smallskip
+Created a web chat app using Node.js and React.js. (HackPrinceton Fall 2018)
+
+\noindent[{\tt https://github.com/marcelgoh/hackprinceton-2018}]
+\medbreak
+
 \sectheader{Research Experience}
@@
 one concerning two of Erd\H os's proofs on prime numbers.)
-
-\medbreak
-\leftright{\bf Sorting algorithms on manifolds}{\bf May -- August 2020}
-\leftright{\sl Summer research project}{\sl McGill University}
-\smallskip
-Research project under the supervision of Michael Lipnowski.
-Studied algorithms that investigate the topology of group actions on locally symmetric spaces.
-Work involved writing of code in OCaml and PostScript that generates illustrations of certain quotient
-spaces, in various models of hyperbolic geometry.
 \medbreak
@@
 \medbreak
-
-\leftright{\bf Bytecode compiler and interpreter}{\bf February -- June 2019}
-\leftright{\sl Individual software project}{\sl Charles University}
-\smallskip
-Semester-long individual software project under the supervision of Adam Dingle.
-Created OPythn, a bytecode compiler and virtual machine for a subset of Python.
-OPythn includes support for lists, dictionaries, named and anonymous functions, objects, and classes.
-It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing.
 

### 2022-04-13 misc/expo/tensorrank.tex
Shovelling trouble and new title for indepno

--- 2022-03-26 misc/expo/additive.tex
@@
 \widemargins
-\bookheader{NOTES ON ADDITIVE COMBINATORICS}{MARCEL K. GOH}
+\bookheader{THE ANALYTIC RANK OF A TENSOR}{MARCEL K. GOH}
 
-\maketitle{Notes on additive combinatorics}{by}{Marcel K. Goh}{23 March 2022}
-
-\floattext4.5 \ninebf Note.
-\ninepoint
-I am compiling this set of expository notes primarily to solidify these fundamental results of
-additive combinatorics in
-my own mind. Sources that are far more comprehensive exist on the web, so I'm not sure
-how useful this will be to anyone else. Currently, the bulk of these notes come from
-a series of YouTube lectures of Timothy Gowers (given in early 2022), though I intend
-in future to add more relevant topics as I learn about them.
-
-\bigskip
-
-\advsect Pl\"unnecke's theorem
-
-Let $A$ and $B$ be finite subsets of an abelian group. We define the {\it sumset}
-$A+B = \{a+b : a\in A,\, b\in B\}$ as well as the {\it difference set}
-$A-B = \{a-b : a\in A,\, b\in B\}$. For a nonnegative integer $r$, the {\it $r$-fold iterated sumset}
-$rA$ is defined recursively by $0A = \{0\}$ and $rA = A+(r-1)A$. The goal of this first section
-is to show that if the cardinality of
-a sumset $|A+B|$ is small compared to the size of the set $A$, then the difference
-$rB-sB$ of two iterated sumsets must also be small relative to the size of $A$,
-in some sense. This result is called Pl\"unnecke's
-theorem as it follows from a statement given by H.~Pl\"unnecke in~1969, though this particular
-formulation of it was stated and proved by I.~Z.~Ruzsa in 1989.
-
-\edef\thmplunnecke{\the\thmcount}
-\parenproclaim Theorem {\advthm} (Pl\"unnecke, {\rm 1969;} Ruzsa, {\rm 1989}). Let $A_0$ and
-$B$ be finite subsets of an abelian group and suppose that $|A_0+B|\le K_0|A_0|$ for some constant $K_0$.
-Then there exist $A\subseteq A_0$ and $K\le K_0$ such that
-for any nonnegative integers $r$ and $s$, not both zero, we have $|rB-sB| \le K^{r+s}|A|$. In particular,
-this means that $|rB-sB| \le {K_0}^{r+s}|A_0|$.
-
-Ruzsa used Menger's theorem from graph theory to prove this theorem, but we will use a shorter
-argument from a 2012 paper by G.~Petridis. It hinges on the following lemma. In this proof (and the
-rest of these notes), we will write $A+x$ instead of $A+\{x\}$ for convenience.
-
-\edef\lempetridis{\the\thmcount}
-\parenproclaim Lemma {\advthm} (Petridis, {\rm 2012}). Let $A$ and $B$ be subsets of an abelian
-group and let $K$ be such that $|A_0+B| = K_0|A_0|$. Then there exist $A\subseteq A_0$ and
-$K\le K_0$ such that $|A+B+C| = K|A+C|$ for every finite subset $C$ of the group.
-
-\proof We pick $A\subseteq A_0$ nonempty such that the ratio $|A+B|/|A|$, which we shall
-set as $K$, is minimised. So $|A+B| = K|A|$ and $|A'+B|\ge K|A'|$ for all $A'\subseteq A_0$.
-Now we shall show by induction on $C$ that $|A+B+C|\le K|A+C|$ for all $C$. If $C$ is empty we
-are done, of course. Now assume the result holds for $C$ and let $x\notin C$ be arbitrary. We have
-$$\bigl| A+(C\cup \{x\}) \bigr| = |A+C| + |A+x| - |A'+x|$$
-where $A' = \{a\in A : a+x \in A+C\}$. But adding $x$ to everything in a set does not change its
-cardinality, so we have
-$$|A+C| + |A+x|-|A'+x| = |A+C| + |A| + |A'|.$$
-Also, any element of $A'+B+x$ is an element of $A+B+C$, since $a'+x\in A+C$ for all $a'\in A'$.
-Applying the induction hypothesis now gives
-$$\eqalign{
-\bigl| A + B + (C\cup \{x\}) \bigr| &\le |A+B+C| + |A+B+x| - |A'+B+x| \cr
-&\le K|A+B| + K|A| = K|A'| \cr
-&= K\bigl| A+(C\cup \{x\})\bigr|,
-}$$
-which completes the proof.\slug
-
-This lemma has an immediate corollary that already has some resemblance to the version of Pl\"unnecke's
-theorem that we eventually want to prove.
-
-\proclaim Corollary {\advthm}. If $|A_0 + B|\le K_0|A_0|$, then there is $A\subseteq A_0$ nonempty
-and $K\le K_0$ such that $|A+rB| \le K^r |A|$ for all nonnegative integers $r$.
-
-\proof The previous lemma gives $A$ and $K$ such that $|A+B+C|\le K|A+C|$ for all $C$. We now induce
-on $r$. The case $r=0$ is obvious. Now for $r>0$ applying the lemma again with $C = (r-1)B$ and
-then invoking the induction hypothesis gives
-$$|A+rB| = |A+B+(r-1)B| \le K|A+(r-1)B| \le K^r|A|.\noskipslug$$
-
-To get from this corollary to the full theorem, we will use a certain inequality of Rusza (1996), which
-is often called the Ruzsa triangle inequality due to its similarity to the triangle inequality in
-metric spaces.
-
-\parenproclaim Lemma {\advthm} (Ruzsa triangle inequality). Let $U$, $V$, and $W$ be finite
-subsets of an abelian group. Then
-$$|U|\cdot |V-W| \le |U-V|\cdot |U-W|.$$
-
-\proof For each $x\in V-W$, pick $v(x)\in V$ and $w(x)\in W$ such that $v(x)-w(x) = x$ (there may
-be multiple choices for each of $v(x)$ and $w(x)$, but we arbitrarily pick one).
-Define $\phi : U\times (V-W) \to (U-V)\times (U-W)$ by $\phi(u,x) = \bigl(u-v(x), u-w(x)\bigr)$.
-If $\bigl(u_1 - v(x_1), u_1 - w(x_2)\bigr) = \bigl(u_2 - v(x_2), u_2 - w(x_2)\bigr)$, then
-subtracting the second coordinate of each pair from the first, we find that
-$$x_1 = v(x_1) - w(x_1) = v(x_2) - w(x_2) = x_2,$$
-and this also implies that $u_1 = u_2$. So $\phi$ gives an injection from $U\times (V-1)$
-to $(U-V)\times (U-W)$.\slug
-
-If we define $d(U,V) = |U-V| / \bigl(|U|^{1/2}|V|^{1/2}\bigr)$, then this lemma states
-that $d(U,W) \le d(V,U)d(U,W)$. Taking logarithms gives a genuine additive triangle inequality,
-but the space of finite subsets of an abelian group does not form a metric space, since
-$|U-U|/|U|$ does not equal $1$ in general. In any case, we are now able to prove Ruzsa's
-version of Pl\"unnecke's inequality.
-
-\medskip
-\noindent{\it Proof of Theorem~{\thmplunnecke}.}\enspace Let $A$ and $K$ be given by
-Lemma~{\lempetridis}. By the corollary to that lemma, we have $|A+rB| \le K^r|A|$ and
-$|A+sB| \le K^r|A|$. Cardinalities are not changed by taking additive inverses of
-everything, so $|{-A}-rB| \le K^r|A|$ and $|{-A}-sB|\le K^r|A|$. Now an application of Ruzsa's triangle
-inequality with $U = -A$, $V = rB$ and $W = sB$ yields
-$$|A|\cdot|rB-sB| = |{-A}|\cdot|rB-sB| \le |{-A}-rB|\cdot|{-A}-sB| \le K^{r+s}|A|^2,$$
-and dividing both sides by $|A|$ proves the theorem.\slug
-
-\advsect Khovanskii's theorem
-
-The goal of this section is to prove the following wonderful theorem of A.~Khovanskii.
-
-\edef\thmkhovanskii{\the\thmcount}
-\parenproclaim Theorem {\advthm} (Khovanskii, {\rm 1992}). Let $A$ be a finite
-subset of an abelian group and let $f_A(n) = |nA|$ for each $n\in \NN$. Then there exists
-$n_A$ and a polynomial $p_A$ such that $f_A(n) = p_A(n)$ for all $n\ge n_A$.
-
-Consider the example when $A = \{0,1,b-1,b\} \subseteq \ZZ$. Then letting $[a,b] = \{x\in \ZZ : a\le x\le b\}$,
-we have
-$$nA = [0, n] \cup [b-1,\ldots, b+n-1] \cup [2b-2, 2b+n-2] \cup\cdots\cup [n(b-1),nb].$$
-This has size $(n+1)^2$ as long as $n< b-1$, but when $n\ge b-1$, $nA$ is simply the set
-of integers in the range $[0,bn]$, which has size $bn+1$.
-
-We shall present a combinatorial proof of Khovanskii's theorem,
-given by M.~B.~Nathanson and I.~Z.~Ruzsa in 2002.
-Let $A = \{a_1,\ldots, a_r\}$ and
-for $x = (x_1,\ldots, x_r)\in {\NN_0}^r$, write
-$$\sigma(x) = \sum_{i=1}^r x_i\qquad\hbox{and}\qquad
-\alpha(x) = \sum_{i=1}^r a_ix_i.$$
-Then
-$$nA = \{ \alpha(x) : x\in {\NN_0}^r,\, \sigma(x) = n\}.$$
-Let $\prec$ be the lexicographic order on ${\NN_0}^r$; that is, $(x_1,\ldots,x_r)\prec (y_1,\ldots,y_r)$
-if there is some $i\in [r]$ such that $x_i<y_i$ and $x_j = y_j$ for all $1\le j<i$.
-Say that $x$ is {\it useless} if there is $x'\prec x$ such that $\sigma(x') = \sigma(x)$ and
-$\alpha(x') = \alpha(x)$; otherwise, say that $x$ is {\it useful}. It is easy to see that
-$|nA|$ is exactly the number of useful $x$ with $\sigma(x) = n$. Now let $\le$ denote
-the usual partial order on ${\NN_0}^r$, where $(x_1,\ldots,x_r)\le (y_1,\ldots,y_r)$
-if $x_i\le y_i$ for all $1\le i\le r$.
-
-\proclaim Lemma {\advthm}. If $x\le y$ and $x$ is useless, then $y$ is useless.
-
-\proof Since $x$ is useless, we can find $x'\prec x$ with $\sigma(x') = \sigma(x)$ and $\alpha(x') = \alpha(x)$.
-Let $y' = y+x'-x$. Since $x\le y$, $y-x$ is in ${\NN_0}^r$ and therefore so is $y'$. Then since
-$x_i' < x_i$ for the first $i$ at which the two tuples differ, the first coordinate at which $y'$ and
-$y$ differ is also $i$ and we have $y_i'<y_i$. Hence $y'\prec y$. Lastly,
-$$\sigma(y') = \sigma(y+x'-x) = \sigma(y) + \sigma(x') - \sigma(x) = \sigma(y)$$
-and the same is true with $\sigma$ replaced by $\alpha$, so $y$ is useless.\slug
-
-For each useless $x$ we can find a useless $x'$, minimal in the $\le$ order, such that $x'\le x$.
-This lemma gives a converse; it says that if $x'$ is a minimal useless element and $x'\le x$, then
-$x$ is also useless.
-
-\edef\lemminimalfinite{\the\thmcount}
-\proclaim Lemma {\advthm}. For any nonempty $U\subseteq {\NN_0}^r$, the set of minimal elements of
-$U$ (with respect to the $\le$ order) is finite.
-
-\proof We perform induction on $r$. If $r=1$, then it is clear that $U$ has exactly one minimal element.
-Now let $r>1$ and let $u$ be a minimal element of $u$. Then for every $v\in U$ with $u\not\le v$,
-there exists some $i$ and $t< u_i$ such that $v_i = t$. Let
-$$V(i,t) = \{v\in U : v_i = t\}$$
-for all $1\le i\le r$ and $0\le t<u_i$. Each set $V(i,t)$ is order-isomorphic to a subset of ${\NN_0}^{r-1}$,
-so by induction it has finitely many minimal elements. The number of $V(i,t)$ is $r\prod_{i=1}^r u_i$,
-and every minimal element of $u$ is either $u$ or a minimal element of some $V(i,t)$, so there are
-finitely many minimal elements in $U$.\slug
-
-We will need just one more lemma before we are able to prove Khovanskii's theorem. It shows that
-for any fixed $r$-tuple $x'$, the number of $x$ with $x'\le x$ and $\sigma(x) = n$
-is given by a polynomial in the variable $n$, with coefficients depending on $x'$ and $r$.
-
-\proclaim Lemma {\advthm}. Let $x'\in {\NN_0}^r$ and for $n\ge \sigma(x')$, let
-$$C(x', n) = \{n\in {\NN_0}^r : \sigma(x) = n,\,x'\le x\}.$$
-Then
-$$\bigl| C(x',n)\bigr| = {n-\sigma(x') + r-1 \choose r-1}.$$
-
-\proof Given $x\in C(x',n)$, map it to $x-x'$. This is a bijection from $C(x',n)$ to $C(0, n-\sigma(x')$.
-Now let $m = n-\sigma(x')$. The map $(x_1,\ldots,x_r) \mapsto (x_1,+1,\ldots, x_r+1)$ is a bijection
-from $C(0,m)$ to the set of $y\in \NN^r$ such that $y_1+\cdots+y_r = m+r$. Now the map
-$$(y_1,\ldots,y_r) \mapsto \{y_1, y_1+y_2, \ldots, y_1+\cdots+y_r\}$$
-is a bijection from the previous set to the set of subsets of $[m+r]$ that contain $m+r$. (It is
-inverted by the map $\{a_1, \ldots, a_r\} \mapsto (a_1, a_2-a_1, \ldots, a_r-a_{r-1})$.) But this final
-set has size ${m+r-1\choose r-1}$, so we are done.\slug
-
-Without further ado, let us now prove Khovanskii's theorem.
-
-\medskip\noindent {\it Proof of Theorem~{\thmkhovanskii}.}\enspace Let $U$ be the set of all
-useless $r$-tuples. By Lemma~{\lemminimalfinite}, the set $X\subseteq U$ of all minimal useless
-$r$-tuples is finite. Recall that $|nA|$ is the number of useful $x$ such that $\sigma(x) = n$,
-so using the notation of the previous lemma, we have
-$$|nA| = \bigl| C(0,n)\bigr| - \Bigl| \bigcup_{x'\in X} C(x',n)\Bigr|
-=\bigl| C(0,n)\bigr| - \sum_{Y\subseteq X} (-1)^{|Y|} \Bigl| \bigcap_{x'\in Y} C(x',n)\Bigr| $$
-But for any $Y\subseteq X$, $\bigcap_{x'\in Y} C(x',n) = C(x_Y, n)$, where $x_Y$ has for its
-$i$th coordinate the integer $\max_{y\in Y} y_i$. Hence every term in the inclusion-exclusion
-formula depends polynomially on $n$ once
-$$n\ge \sum_{1\le i\le r} \max_{x\in X} x_i.\noskipslug$$
-
-\advsect The Balog--Szemer\'edi--Gowers theorem
-
-Let $A$ be a finite nonempty subset of an abelian group. An {\it additive quadruple} in
-$A$ is a quadruple $(x,y,z,w)\in A^4$ such that $x+y = z+w$. The {\it additive energy}
-$E(A)$ of $A$ is the number of additive quadruples in $A$. If $(x,y,z,w)$ is an additive
-quadruple, then $w = x+y-z$, so $E(A)\le |A|^3$. The following lemma roughly states that if
-a set has a small sumset, then it has large additive energy.
-
-\proclaim Lemma {\advthm}. Let $A$ be a finite subset of an abelian group and let $C$ be such
-that $|A+A|\le C|A|$. Then $E(A)\ge |A|^3/C$.
-
-\proof Define $f : A+A\to \NN$ by letting $f(d)$ equal the number of pairs $(a,b)\in A^2$ such
-that $a+b =d$. Then
-$$\sum_{d\in A+A} f(d) = |A|^2\qquad\hbox{and}\qquad \sum_{d\in A+A} f(d)^2 = E(A).$$
-By the Cauchy--Schwarz inequality,
-$$|A|^2 = \sum_{d\ge 1} 1\cdot f(d)
-\le \Bigl(\sum_{d\ge 1} 1\Bigr)^{1/2} \Bigl( \sum_{d\ge 1} f(d)^2\Bigr)^{1/2}
-\le C^{1/2} |A|^{1/2} E(A)^{1/2}.$$
-From here, squaring the whole inequality and then dividing through by $C|A|$ gives the statement
-we want to prove.\slug
-
-The converse of this lemma does not hold. For example, we could take the disjoint
-union of an arithmetic progression with a geometric progression of the same length.
-This set has a large additive energy,
-because of the arithmetic progression, but it also has a large sumset, because of the geometric progression.
-However, we can pass to a large subset that has a small sumset (namely, the arithmetic progression), and this
-is essentially the statement of the Balog--Szemer\'edi--Gowers theorem. We shall state it later on in the
-section, after we have given the graph-theoretic definitions and lemmas needed in its proof.
-
-Let $G$ be a bipartite graph with finite vertex sets $X$ and $Y$. For a vertex $v\in X\cup Y$,
-let $d(v) = \deg(v)$ denote the {\it degree} of $x$, that is, the number of vertices adjacent to $v$.
-For $x_1,x_2\in X$, let $d(x_1,x_2)$ denote the number of paths of length $2$ from $x_1$ to $x_2$. This
-is called the {\it codegree} of $x_1$ and $x_2$. The {\it density} of $G$ is the number of edges of $G$
-divided by the maximum number of vertices it could have, namely $|X|\cdot|Y|$.
-
-\proclaim Lemma {\advthm}. Let $G$ be a bipartite graph with bipartition $X\cup Y$ and let the density
-of $G$ be $\delta$. Then for every $c>0$ there is a subset $X'\subseteq X$ with $|X'|\ge \delta|X|/\sqrt 2$
-such that the number of pairs $(x_1,x_2)\in (X')^2$ with $d(x_1,x_2) < c|Y|$ is at most $2c|X|^2/\delta^2$.
-
-\proof Let $y\in Y$ be chosen uniformly at random. Letting $X' = N(y)$, we have
-$$\ex\bigl\{ |X'|\bigr\} = \ex_{y\in Y} \bigl\{ d(y)\bigr\} = {1\over |Y|} \sum_{y\in Y} d(y)
-= {|E(G)|\over |Y|} = \delta|X|.$$
-By convexity of the map $x\mapsto x^2$, we then find that $\ex\bigl\{ |X'|^2\bigr\} \ge \delta^2|X|^2$.
-Let
-$$B = \bigl\{ (x_1,x_2)\in X^2 : d(x_1,x_2) < c|Y|\bigr\}.$$
-The probability that any $(x_1, x_2)\in B$ is also in $(X')^2$ is the same as the probability that
-$y$ is connected to both $x_1$ and $x_2$; this is less than $c|Y|$ by the construction of $B$.
-So by linearity of expectation, $\ex\bigl\{ |B\cap (X')^2| \bigr\} < c|X|^2$ and in particular, we
-have
-$$\biggl\{ |X'|^2 - {\delta^2\over 2c} |B\cap(X')^2|\biggr\}
- >\biggl(\delta^2 - {\delta^2\over 2c}\cdot c\biggr)|X|^2 = {\delta^2\over 2}|X|^2.$$
-So there is $X'$ such that $|X'| - \bigl(\delta^2/(2c)\bigr) |B\cap (X')^2| \ge \delta^2|X|^2/2$.
-This set $X'$ has cardinality at least $\delta |X|/\sqrt 2$ and $|B\cap (X')^2| \le 2c|X'|^2/\delta^2$.\slug
-
-This lemma says that we can find a fairly large subset of $X$ such that almost every pair in this
-subset is joined by at least $c|Y|$ paths of length $2$. We would really like to remove the ``almost''
-in this statement, but to do so, we will have to pass to paths of length $3$.
-First, we need a little lemma about sums of real numbers between $0$ and $1$.
-
-\proclaim Lemma {\advthm}. Let $a_1,\ldots, a_n\in [0,1]$. If $\sum_{i=1}^n a_i \ge an$, then
-the number of $i$ such that $a_i\ge b$ is at least
-$$ \biggl( {a-b\over 1-b} \biggr) n.$$
-
-\proof Let $I$ be the set of $i$ such that $a_i\ge b$. Then we may write
-$$ an \le \sum_{i=1}^n a_i = \sum_{i\in I} a_i + \sum_{i\notin I} a_i
-\le |I| + (n-|I|)b.$$
-It remains to isolate for $|I|$.\slug
-
-For example, if $b = a/2$, then the number of $i$ with $a_i\ge b$ must be at least $an/2$, and more
-generally, if $a= 1-\theta$ and $b=1-\eta$, then the number of such $i$ is at least $(1-\theta/\eta)n$.
-If $\eta = 2\theta$, then $a_i\ge 1-2\theta$ at least half the time.
-Next we prove the lemma alluded to earlier, concerning paths of length $3$.
-
-\proclaim Lemma {\advthm}. Let $G$ be a bipartite graph with vertex sets $X$ and $Y$ and density $\delta$.
-Then there are subsets $X'\subseteq X$ and $Y'\subseteq Y$ such that
-$$|X'| \ge {\delta^2\over 8\sqrt 2} |X|\qquad\hbox{and}\qquad
-|Y'|\ge {\delta\over 4}|Y|$$
-and every $x\in X'$ and $y\in Y'$ are joined by at least $(\delta^6/2^{13}) |X|\cdot |Y|$ paths
-of length $3$ in $G$.
-
-\proof By letting $n = |X|$ and setting the $a_i$ equal to the degrees of the vertices in $X$, we have
-$a = \delta|Y|$
-and we can choose $b=a/2$ in the previous lemma
-to deduce that $X$ has a subset $X_1$ such that every vertex in $X_1$ has degree at least
-$\delta|Y|/2$ and $|X_1|\ge \delta|X|/2$. Note that the density of the bipartite subgraph with
-vertex set $X_1\cup Y$ is $\ge \delta/2$,
-so we may invoke the lemma about paths of length $2$ with $\delta$ set to $\delta/2$. It tells us that
-for any $c>0$, $X_1$ has a subset $X_2$ of size $\delta|X_1|/2^{3/2}$ such that for all but at most
-$8c|X_2|^2/\delta^2$ pairs $(x_1,x_2)\in {X_2}^2$, we have $d(x_1,x_2) \ge c|Y|$.
-
-Now let $\Gamma$ be a graph with vertex set $X_2$ and an edge between $x_1$ and $x_2$ if and only if
-$\delta(x_1,x_2)\ge c|Y|$. Note that by this definition, there may be loops in $\Gamma$. In any case,
-the average degree in $\Gamma$ is at least $(1-8c/\delta^2)|X_2|$. Invoking the previous lemma again,
-at least half the vertices in $X_2$ have degree at least $(1-16c/\delta^2)|X_2|$. Let $X'$ be the set of
-all such vertices. Then
-$$|X'| \ge {|X_2|\over 2} \ge {\delta|X_1|\over 4\sqrt 2} \ge {\delta^2 |X|\over 8\sqrt 2}.$$
-Every vertex in $X_2$ has degree at least $\delta |Y|/2$, so the bipartite subgraph on $X_2\cup Y$
-has density at least $\delta/2$ and the average degree of $y\in Y$ is at least $\delta|X_2|/2$. The
-previous lemma comes in handy once again, telling us that at least $\delta|Y|/4$ vertices in $Y$ has
-degree at least $\delta|X_2|/4$. Let $Y'$ be the set of these vertices.
-
-For any $x\in X'$ and $y\in Y'$, we know that $x$ has at least $(1-16c/\delta^2)|X_2|$ neighbours in $\Gamma$
-and $y$ has at least $\delta|X_2|/4$ neighbours in $X_2$. There must be at least
-$(\delta/4-16c/\delta^2)|X_2|$ neighbours in commmon. Letting $c=\delta^3/128$, the number of neighbours
-in common becomes $\delta|X_2|/8$, and the number of paths of length $3$ from $x$ to $y$ is at least
-$$ {\delta |X_2|\over 8} |X_2| \cdot c|Y| = {\delta\over 8}\cdot {\delta^3\over 128}
-\cdot {\delta^2\over 4\sqrt 2} |X|\cdot |Y| \ge {\delta^6\over 2^{13}}|X|\cdot |Y|,$$
-as foretold.\slug
-
-Let $A$ be a finite subset of an abelian group. We say that an element $d\in A-A$ is {\it $\theta$-popular}
-if $\bigl|\bigl\{ (a,b)\in A^2 : b-a=d\bigr\}\bigr|\ge \theta |A|$.
-
-\proclaim Lemma {\advthm}. Let $A$ be a finite subset of an abelian group and let $E(A)$ denote
-the additive energy of $A$. If $E(A)\ge c|A|^3$, then there are at least $c|A|/2$ elements of $A-A$
-that are $(c/2)$-popular.
-
-\proof Define $f : A-A\to \NN$ by $f(d) = \bigl|\bigl\{ (a,b)\in A^2 : b-a=d\bigr\}\bigr|$.
-Note that
-$$\sum_{d\in A-A} f(d) = |A|^2\qquad\hbox{and}\qquad \sum_{d\in A-A} f(d)^2 = E(A).$$
-Let $P$ be the set of $(c/2)$-popular differences and let $U = (A-A)\setminus P$. Then
-$$\eqalign{
-c|A|^3 &\le E(A)\cr
-&= \sum_{d\in A-A} f(d)^2 \cr
-&= \sum_{d\in P} f(d)^2 + \sum_{d\in U} f(d)^2 \cr
-&\le |P|\cdot |A|^2 + {c|A|\over 2} \sum_{d\in U} f(d) \cr
-&= |P|\cdot |A|^2 + {c|A|^3\over 2}, \cr
-}$$
-which implies that $|A|\ge c|A|/2$.\slug
-
-We are, at long last, ready to state and prove the Balog--Szemer\'edi--Gowers theorem. It is
-named for A.~Balog and E.~Szemer\'edi, who first proved it in 1994, as well as for W.~T.~Gowers,
-who gave an alternative proof in 1998 that yielded power-type bounds, a vast improvement to the tower-type
-bounds in Balog and Szemer\'edi's proof.
-
-\parenproclaim Theorem {\advthm} (Balog--Szemer\'edi, {\rm 1994;} Gowers, {\rm 1998}).
-Let $A$ be a finite subset of an abelian group with $E(A)\ge c|A|^3$. Then $A$ has a subset
-$A'$ of size at least $c'|A|$ such that $|A'-A'|\le C|A'|$, where $c'$ and $C$ have a power-type
-dependence on $c$. In particular, we may take
-$$ c' = {c^4\over 2^8}\qquad\hbox{and}\qquad C = {2^{68}\over c^{36}}.$$
-
-\proof Define a bipartite graph $G$ with both vertex sets being copies of $A$ (though
-we'll call one of these copies $B$ to avoid any confusion). Join $a\in A$ to
-$b\in B$ if and only if $b-a$ is $(c/2)$-popular. Each $(c/2)$-popular difference leads to at least
-$c|A|/2$ edges, so by the lemma, there are at least $c^2|A|^2/4$ edges. Let $\delta = c^2/4$.
-By the lemma concerning paths of length $3$, there exist $A'\subseteq A$ and $B'\subseteq B$
-with
-$$ |A'| \ge {\delta^2\over 8\sqrt 2} |A|\qquad\hbox{and}\qquad
-|B'|\ge {\delta \over 4} |A|$$
-such that every $a\in A$ and $b\in B$ are joined by at least $\delta^6|A|^2/2^{13}$ paths of length $3$
-in $G$.
-
-Given a path $(a,u,v,b)$ of length $3$, we have $b-a = u-a - (u-v) + b-v$, where $u-a$, $u-v$, and $b-v$
-are all $(c/2)$-popular. Hence the number of ways of writing $b-a$ as $r_1-s_1 - (r_2-s_2) + r_3 - s_3$
-is at least
-$${\delta^6\over 2^{13}} |A|^2 \biggl( {c\over 2} |A|\biggr)^3 = {c^{15}\over 2^{28}} |A|^5.$$
-But the number of possible $(r_1,s_1,r_2,s_2,r_3,s_3)$ is $|A|^6$. So
-$$ |B'-A'|\cdot {c^{15}\over 2^{28}} |A|^5 \le |A|^6 $$
-and $|B'-A'|\le 2^{28}|A|\over c^{15}$. Now the Ruzsa triangle inequality with $U = B'$ and
-$V = W = A'$ tells us that $|B'|\cdot |A'-A'| \le |B'-A'|^2$, so
-$$|A'-A'|\le {|B'-A'|^2\over |B'|} \le {2^{56}|A|^2\over c^{30}}\cdot {2^4\over c^2|A|}={2^{60}\over c^{32}}|A|.$$
-But recall that
-$$|A'| \ge {\delta^2\over 8\sqrt 2} |A| \ge {c^4\over 2^8}|A|,$$
-so $|A'-A'| \le 2^{68}|A'|/c^{36}$.\slug
-
-\advsect Basics of discrete Fourier analysis
-
-Let $Z$ be a finite abelian group. A {\it character} on $Z$ is a homomorphism from
-$Z$ to the multiplicative group $\C\setminus\{0\}$. If $\chi$ is such a homomorphism, then
-$\bigl|\chi(x)\bigr| = 1$ for every $x\in Z$, so we can actually regard $\chi$ as being
-a function from $Z$ to the unit circle $\TT = \{z\in \C : |z| = 1\}$. The pointwise product
-of two characters gives another character, and the function $\chi_0$ that sends every $x\in Z$
-to $1$ has the property that if $\chi$ is any character, then $\chi\chi_0 = \chi = \chi_0\chi$.
-Lastly, we note that multiplication is commutative and
-for any character $\chi$, the product of $\chi$ with $\bar\chi$
-gives $\chi_0$, so the set of characters on $Z$ is an abelian group. This is called the
-{\it dual group} or sometimes {\it Pontryagin dual} of $Z$ and is denoted $\hat Z$.
-
-We define an inner product on the space $\CC^Z$ of functions from $Z$ to $\CC$ by setting
-$$\langle f,g\rangle = \ex_x f(x) \bar{g(x)},$$
-where $\ex_{x\in Z} F(x) = |Z|^{-1} \sum_{x\in Z} F(x)$.
-We can also make $\hat Z$ into an inner product space by letting
-$$\langle \hat f, \hat g\rangle = \sum_{\chi\in \hat Z} \hat f(\chi)\bar{\hat g(\chi)};$$
-note that this time we do not normalise by dividing by $|Z|$.
-Two functions $f$ and $g$ in an inner product space are said to be {\it orthogonal} if $\langle f,g\rangle = 0$.
-The first lemma we'll prove concerns certain orthogonality relations.
-
-\parenproclaim Lemma {\advthm} (Orthogonality relations). Let $Z$ be a finite abelian group.
-\medskip
-\item{a)} If $\chi_1$ and $\chi_2$ are characters on $Z$, then
-$$\langle \chi_1,\chi_2 \rangle = \ex_{x\in Z} \chi_1(x)\bar{\chi_2(x)}
-  = \cases{ 1, & if $\chi_1 = \chi_2$;\cr 0, & if $\chi_1 \ne\chi_2$.}.$$
-\smallskip
-\item{b)} For $x,y\in Z$, we have
-$$\sum_{\chi\in \hat Z} \chi(x)\bar{\chi(y)}
-  = \cases{ |\hat Z|, & if $x=y$;\cr 0, & if $x\ne y$,}$$
-\medskip
-
-\proof Let $\chi_0$ denote the trivial character. We have
-$$\ex_{x\in Z} \chi_0(x) = \ex_{x\in Z} 1 = 1.$$
-On the other hand, let $\chi$ be a nontrivial character and let $u\in Z$ be such that $\chi(u)\ne 1$. Then from
-$$\ex_{x\in Z} \chi(x) = \ex_{x\in Z} \chi(u+x) = \ex_{x\in Z}\chi(u)\chi(x) = \chi(u)\ex_{x\in Z}\chi(x),$$
-it follows that $\ex_{x\in Z}\chi(x) = 0$. Now consider $\chi_1\bar{\chi_2}$. If $\chi_1 = \chi_2$, then
-this is the trivial character and from our first observation, $\langle \chi_1,\chi_2\rangle = 0$. Otherwise,
-we are in the second case and the inner product is zero. This proves part (a).
-
-Part (b) is proven similarly. Note that $\chi(x)\bar{\chi(y)} = \chi(x-y)$. If $x=y$, then we have
-$$\sum_{\chi\in\hat Z} \chi(x-y) = \sum_{\chi\in\hat Z} \chi(0) = |\hat Z|.$$
-If $x\ne y$, then there is some $\psi\in \hat Z$ such that $\psi(x-y)\ne 1$, and then writing
-$$\sum_{\chi\in \hat Z} \chi(x-y) = \sum_{\chi\in\hat Z} \psi(x-y)\chi(x-y)
-= \psi(x-y)\sum_{\chi\in\hat Z} \chi(x-y),$$
-we see that $\sum_{\chi\in\hat Z} \chi(x-y)$ must be zero.\slug
-
-Since the space of functions from $Z\to \CC$ has dimension $|Z|$, there are at most $|Z|$ characters.
-Every finite abelian group can be written as a direct product of cyclic groups, and we shall use this fact
-to show that there are exactly $|Z|$ characters; i.e., $|\hat Z| = |Z|$. For brevity, let $\ZZ_n$ stand for $\ZZ/n\ZZ$ for all $n\in\NN$,
-and for $\alpha\in \RR$, let $e(\alpha) = e^{2\pi i\alpha}$.
-
-\proclaim Lemma \advthm.
-The set of characters on a finite abelian group $Z$ spans the space of functions from $Z$ to $\CC$.
-
-\proof Write $Z \cong \ZZ_{n_1} \times \ZZ_{n_2}\times \cdots\times \ZZ_{n_r}$. For every
-$u = (u_1,\ldots, u_r)\in Z$, the function $\chi_u : Z\to \CC$ that takes
-$$(x_1,\ldots, x_r) \mapsto \prod_{i=1}^r e\biggl( {u_ix_i\over n_i}\biggr)$$
-is a character (this is easy to show, since $e(x_i + x_i') = e(x_i)e(x_i')$). If $u\ne u'$,
-then $\chi_u\ne\chi_{u'}$ and $\langle \chi_u,\chi_{u'}\rangle = 0$ by the previous lemma. Hence
-the set $\{\chi_u : u\in Z\}$ comprises $|Z|$ linearly independent elements in a space of dimension $|Z|$,
-which must be spanning.\slug
-
-Note that the map $u\mapsto \chi_u$ gives an isomorphism from $Z$ to $\hat Z$.
-The two main examples we shall consider are when $Z = \ZZ_n$ and when $Z = \FF_p^n$. In the first
-case, $\chi_u$ takes $x\mapsto e(ux/N)$ and in the second case $\chi_u$ takes $x\mapsto e(u\cdot x/p)$.
-(Recall that if $u = (u_1,\ldots, u_n)$ and $x = (x_1,\ldots, x_n)$ then the dot product $u\cdot x$
-is the sum $\sum_{i=1}^n u_ix_i$.)
-
-For a finite set $X$, we shall denote by $L_p(X)$ the normed vector space of all functions $f:X\to \CC$
-under the norm
-$$\norm{f}_p = \bigl( \ex_{x\in X} \bigl| f(x)\bigr|^p \bigr)^{1/p}.$$
-The notation $l_p(X)$ denotes the same set, but with the norm
-$$\norm{f}_p = \Bigl( \sum_{x\in X} \bigl| f(x)\bigr|^p \Bigr)^{1/p}$$
-instead. Note that $\norm{f}_2^2 = \langle f,f\rangle$ and $\norm{f}_\infty = \max_{x\in X} \bigl|f(x)\bigr|$.
-
-\medskip\boldlabel The Fourier transform.
-If $f:Z\to \CC$, the {\it Fourier transform} of $f$ is the function $\hat f : \hat Z\to\CC$ given by
-$$\hat f(\chi) = \langle f,\chi\rangle = \ex_{x\in Z} f(x) \bar{\chi(x)}.$$
-We shall now state and prove an identity which is extremely useful in Fourier analysis.
-
-\parenproclaim Lemma {\advthm} (Parseval's identity). Let $Z$ be a finite abelian group
-and let $f$ and $g$ be functions from $Z$ to $\CC$.
-If $\hat f$ and $\hat g$ are the Fourier transforms of $f$ and $g$ respectively, then
-$\langle f,g\rangle = \langle \hat f, \hat g\rangle$.
-
-\proof First, we expand all the definitions and rearrange sums to obtain
-$$\eqalign{
-\langle \hat f, \hat g\rangle &= \sum_{\chi\in \hat Z} \hat f(\chi)\bar{\hat g(\chi)} \cr
-&=  \sum_{\chi\in \hat Z} \ex_{x\in Z} f(x)\bar{\chi(x)} \ex_{y\in Z} g(y)\bar{\chi(y)} \cr
-&=   \ex_{x\in Z} f(x) \ex_{y\in Z} g(y) \sum_{\chi\in \hat Z}\bar{\chi(x)}\bar{\chi(y)}.\cr
-}$$
-By the second orthogonality relation and the fact that $|\hat Z| = |Z|$, the inner sum equals $|Z|$ when
-$y = x$ and $0$ otherwise. Thus for any given $x$, we have
-$$\ex_{y\in Z} g(y) \sum_{\chi\in \hat Z}\chi(x)\bar{\chi(y)} = g(x),$$
-which is exactly what we need for the whole thing to equal $\langle f,g\rangle$.\slug
-
-We also have the following inversion formula that recovers the original function $f$ from $\hat f$.
-
-\parenproclaim Lemma {\advthm} (Fourier inversion formula). Let $Z$ be a finite abelian group and
-let $f:Z\to \CC$. We have
-$$f(x) = \sum_{\chi\in \hat Z} \hat f(\chi)\chi(x).$$
-
-\proof We expand
-$$\sum_{\chi\in \hat Z} \hat f(\chi)\chi(x) = \sum_{\chi\in \hat Z} \ex_{y\in Z} f(y) \bar{\chi(y)} \chi(x)
-= \ex_{y\in Z} f(y) \sum_{\chi\in \hat Z} \chi(x)\bar{\chi(y)},$$
-and we noted in the proof of Parseval's theorem that the right-hand side is exactly $f(x)$.\slug
-
-The inversion formula tells us that two functions from $Z\to \CC$ are equal if and only if their Fourier
-transforms are equal. For the next lemma, which concerns dilates of a function $f: Z\to \CC$, we will
-establish some notation. If $x\in Z$, we define $nx$ recursively for all integers $n\ge 0$ by $0x = 0$
-and $nx = x + (n-1)x$. We then set $nz = - (-n)x$ for all negative integers. Since the group operation
-on characters is multiplication, we will not write $n\chi$ but instead $\chi^n$ for $\chi$ multiplied
-with itself $n$ times.
-
-\parenproclaim Lemma {\advthm} (Dilation rule). Let $Z$ be a finite abelian group and let $a\in \ZZ$
-be an integer that is coprime to $|Z|$. Denote the multiplicative inverse of $a$ modulo $|Z|$ by $a^{-1}$.
-Letting $f_a : Z\to \CC$ be the function given by $f_a(x) = f(a^{-1} x)$, we have
-$\hat{f_a}(\chi) = \hat f(\chi^a)$.
-
-\proof We have
-$$\hat{f_a}(\chi) = \ex_{x\in Z} f_a(x) \bar{\chi(x)} = \ex_{x\in Z} f(a^{-1} x)\bar{\chi(x)}.$$
-Since $x\mapsto ax$ is a bijection from $Z$ to itself, we can replace $x$ formally by $ax$ in the above
-to get
-$$\hat{f_a}(\chi) = \ex_{x\in Z} f(x)\bar{\chi(ax)} = \ex_{x\in Z} f(x) \bar{\chi(ax)}^a
-= \hat f(\chi^a).\noskipslug$$
-
-\medskip\boldlabel Convolutions.
-For functions $f$ and $g$ from a finite abelian group $Z$ to $\CC$, the {\it convolution} $f*g$ is
-defined by
-$$ (f*g)(x) = \ex_{y+z = x} f(y)g(z).$$
-If instead $\hat f$ and $\hat g$ are functions from $\hat Z$ to $\CC$, then
-$$ (\hat f * \hat g)(\chi) = \sum_{\chi_1\chi_2 = \chi} \hat f(\chi_1)\hat g(\chi_2).$$
-
-\parenproclaim Lemma {\advthm} (Convolution law). Let $Z$ is a finite abelian group and $f,g : \Z\to \CC$.
-For all $\chi\in \hat Z$, we have $\hat{f*g} (\chi) = \hat f(\chi)\hat g(\chi)$.
-
-\proof We start by expanding
-$$\hat{f*g}(\chi) = \ex_{x\in Z} (f*g)(x) \bar{\chi(x)} = \ex_{x\in Z} \ex_{y+z=x} f(y)g(z)\bar{\chi(y)\chi(z)}.$$
-But note that $x$ does not appear in the summand anymore, meaning that we can simply rewrite this as
-an expectation
-over {\it all} $y$ and $z$ (their sum will equal $x$ for some $x\in Z$). Thus we can conclude that
-$$\hat{f*g}(\chi) = \ex_{y\in Z}\ex_{z\in Z} f(y)g(z)\bar{\chi(y)\chi(z)} = \hat f(\chi)\hat g(\chi),$$
-which is what we wanted to show.\slug
-
-If $A$ is a subset of a finite abelian group $Z$, we associate to $A$ the {\it characteristic function}
-$\one_A : Z\to \CC$ given by
-$$\one_A(x) = \cases{ 1, & if $x\in A$;\cr 0, & otherwise.}$$
-When it will not cause confusion, we will abuse notation and write $A(x)$ instead of $\one_A(x)$.
-With the definitions and results stated above, we are now in a position to translate properties of $A$ to
-Fourier-analytic statements about the function $\one_A$. First off, note that
-$$\bignorm{\hat{\one_A}}_2^2 = \norm{\one_A}_2^2 = \ex_{x\in Z} A(x)^2 = \ex_{x\in Z} A(x)
-= {|A|\over |Z|}.$$
-If $|A|/|Z| = \delta$, then we will say that $A$ has {\it density} $\delta$. Another way of expressing $\delta$
-in terms of $\one_A$ is
-$$\hat{\one_A}(\chi_0) = \ex_{x\in Z} A(x)\chi_0(x) = \ex_{x\in Z} A(x) = \delta,$$
-where $\chi_0$ is the trivial character.
-
-The fact that $\norm{\one_A}_2^2 = \delta$ means that if we sample $x$ and $y$ from $Z$ and
-condition that $x=y$, the probability that $x$ (or $y$) is in $A$ is $\delta$. The reason for stating this
-obvious fact in such a stilted fashion is that it generalises by way of the convolution operation we defined
-earlier; that is, what if we are interested in the probability that a $4$-tuple $(x,y,z,w)$ satisfying $x+y=z+w$
-is a member of $A^4$? Well, the convolution law gives
-$$\eqalign{
-\ex_{x+y=z+w} A(x)A(y)A(z)A(w) &= \ex_{u\in Z} \ex_{x+y=u}\ex_{z+w=u} A(x)A(y)A(z)A(w) \cr
-&= \ex_{u\in Z} (\one_A*\one_A)(u)^2 \cr
-&= \norm{\one_A*\one_A}_2^2 \cr
-&= \bignorm{\hat{\one_A*\one_A}}_2^2 \cr
-&= \bignorm{\bigl(\hat{\one_A}\bigr)^2}_2^2 \cr
-&= \sum_{\chi\in \hat Z} \bigl| \hat{\one_A}(\chi)\bigr|^4 \cr
-&= \norm{\one_A}_4^4, \cr
-}$$
-and more generally, the probability that a tuple $(a_1,\ldots,a_k, b_1,\ldots,b_k)\in Z^{2k}$
-with $a_1+\cdots+a_k=b_1+\cdots+b_k$ is a member of $A^{2k}$ is $\norm{\one_A}_{2k}^{2k}$. Recall that
-the number of $(x,y,z,w)\in A^4$ with $x+y = z+w$
-is the additive energy $E(A)$ of $A$; we just showed above that
-$E(A) = |Z|^3 \norm{\one_A}_4^4$.
-
-\section References
+\maketitle{The analytic rank of a tensor}{by}{Marcel K. Goh}{10 April 2022}
 

### 2022-04-21 misc/expo/tensorrank.tex
Writing stuff

@@
 \def\FF{{\bf F}}
-\def\TT{{\bf T}}
+\def\T{{\rm T}}
+\def\bias{\op{\rm bias}}
+\def\prk{\op{\rm prk}}
+\def\ark{\op{\rm ark}}
 \def\bar{\overline}
@@
 \def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
+\def\x{{\bf x}}
+\def\y{{\bf y}}
 
@@
 
-\maketitle{The analytic rank of a tensor}{by}{Marcel K. Goh}{10 April 2022}
+\maketitle{The analytic rank of a tensor}{by}{Marcel K. Goh}{13 April 2022}
+
+\floattext4.5 \ninebf Note. \ninepoint These notes are more or less a retelling of
+some results in a 2019 paper entitled ``The analytic rank of tensors and its applications'' by
+S.~Lovett. I go more deeply into definitions (for my own sake) so for other students this may be easier to
+follow than the original paper.
+
+\bigskip
+
+\advsect Introduction
+
+\noindent
+There are various compatible definitions of the {\it rank} of a matrix. The one that
+extends most easily to the context of tensors, which we define later, is the following.
+An $m\times n$ matrix $A$ is said to be {\it rank one} if there exist vectors $u\in \FF^m$
+and $v\in \FF^n$ such that $A = uv^\T$. The {\it rank} of a general matrix $A$ is the minimum
+number $k$ such that we can write $A = A_1 + \cdots + A_k$, where $A_i$ is a rank one matrix
+for all $1\le i\le k$.
+
+In a first course on linear algebra, one usually learns that the rank of a matrix is the
+dimension of its column (or row) space. To see that the above definition of rank is equivalent,
+let $A$ be a rank-$k$ matrix and let $B = \{b_1, \ldots, b_k\}$
+be a basis of its column space. Since every column of $A$
+can be written as a linear combination of vectors in $B$, so there is a $k\times n$ matrix $C$
+such that $A = BC$. Now letting $c_1, \ldots, c_k$ be the rows of $C$, we have
+$$A = b_1{c_1}^\T + \cdots + b_k{c_k}^\T,$$
+so the rank of $A$ is at most $k$. On the other hand, if
+$$A = u_1{v_1}^\T + \cdots + u_{k'}{v_{k'}}^\T,$$
+for some $k'< k$, then for all $x \in \FF^n$, we have
+$$Ax = u_1{v_1}^\T x + \cdots + u_{k'}{v_{k'}}^\T x.$$
+Since ${v_i}^\T x$ is a scalar for all $1\le i\le k'$, we conclude that $u_1, \ldots, u_{k'}$
+span the image of $A$ and the column space of $A$ is at most $k'<k$.
+
+Any $m\times n$ matrix $A$ gives a bilinear map from $\FF^m \times \FF^n$ to $\FF$ by taking
+$(x,y)\mapsto xAy$. We extend this to more than two vector spaces by defining an {\it order-$d$ tensor}
+to be a multilinear map $T : V_1\times\cdots\times V_d\to \FF$, where $V_i$ is a vector space over $\FF$
+for all $1\le i\le d$. From here on out,
+we restrict ourselves to the case where each $V_i$ has the same dimension $n$ and can thus be identified
+with $\FF^n$. Then there exist $n^d$ scalars $\{T_{j_1,\ldots,j_d}\}_{j_1,\ldots,j_d\in [n]}$
+such that for all $x_1,\ldots,x_d\in \FF^n$,
+$$ T(x_1, \ldots, x_d) = \sum_{j_1,\ldots,j_d\in [n]} T_{j_1,\ldots,j_d} x_{1,j_1}\cdots x_{d,j_d},$$
+where $x_{i, k}$ denotes the $k$th component of the vector $x_i$. There is thus a one-to-one correspondence
+between order-$d$ tensors and $d$-dimensional arrays of scalars (in our setting each dimension has size
+$n$).
+If $T$ is an order-$d$ tensor and $T'$ is an order-$d'$ tensor, then we can form a tensor of order $d+d'$
+from the $(d+d')$-dimensional array of scalars
+$$\{T_{i_1,\ldots,i_d}T'_{j_1,\ldots,j_{d'}}\}_{i_1,\ldots,i_{d},j_1,\ldots,j_{d'}\in [n]}.$$
+This tensor is denoted $T\otimes T'$ and is called the {\it tensor product} of $T$ and $T'$.
+
+Now we say that an order-$d$ tensor $T$ is {\it partition rank one} if there exists $A\subseteq [d]$
+with $0 < |A| < d$, as well as an order-$|A|$ tensor $T_1$ and an order-$(d-|A|)$ tensor $T_2$ such
+that $T$ can be written as
+$$ T(x_1, \ldots, x_d) = T_1(x_i : i\in A)T_2(x_i : i\notin A).$$
+The {\it partition rank} $\prk(T)$ of a
+general tensor $T$ is the minimum $k$ such that $T$ can be written as a sum
+of $k$ partition rank one tensors. Note that in the case $d=2$ this reduces to the ordinary matrix rank.
+
+\medskip\boldlabel The cap set problem. The partition rank was introduced to study the cap-set problem,
+and here we shall sketch how it applies. A {\it cap set} is a subset $A\subseteq \FF_3^n$ such that for
+every triple $(x,y,z)\in A$ of pairwise distinct elements, $x+y+z\ne 0$. In 1984, T.~C.~Brown
+and J.~P.~Buhler showed that, loosely speaking, cap sets have zero density.
+
+\parenproclaim Theorem C (Brown--Buhler{\rm, 1986}). For every $\delta>0$
+there exists $n$ such that every subset $A\subseteq \FF_3^n$ with $|A|\ge \delta 3^n$ contains three pairwise
+distinct elements $x$, $y$, and $z$ with $x+y+z=0$.
+
+A later paper by R.~Meshulam gave better quantitative
+bounds on $n$ with respect to $\delta$; namely, it was shown that we need only take $n > 2/\delta$.
+This means that if $A$ is a cap set in $\FF_3^n$, then $|A| \le 2\cdot 3^n/n$. However, it was long
+suspected that this bound could be improved to $|A| \le O(c^n)$ for some $c<3$. This was finally proved
+in 2017 by J.~S.~Ellenberg and D.~Gijswijt, and T.~Tao showed in a blog post (dated 18 May 2016)
+that thir proof can be restated in terms of the partition rank of a function in $3$ variables. This
+can actually be modified so the function is a $3$-tensor, but to just get the general idea, let
+us extend our definition of partition rank to general functions of three variables temporarily.
+
+Given $A\subseteq \FF_3^n$, let $T : V^3 \to \FF_3$, where $V = {\FF_3}^{\FF_3^n}$, be given by
+$$f(e_a,e_b,e_c = \cases{1, & if $a+b+c=0$;\cr 0, & otherwise,}$$
+for basis vectors $e_v$ and extended to all other vectors by linearity. (The function $e_v$ has
+$e_v(v) = 1$ and $e_v(x)=0$ for all $x\ne v$.)
+Now for a tensor $T : (\FF^X)^d \to \FF$, we say that a set $A\subseteq X$ is an {\it independent
+set in $T$} if for all $i_1,\ldots,i_d\in A$, the condition that the coefficient $T_{i_1,\ldots,i_d}$
+be nonzero is equivalent to $i_1=\cdots=i_d$.
+We then give an upper bound on the size of a cap set by proving that
+\medskip
+\item{i)} if $A$ contains no nontrivial solutions to $x+y+z=0$, then $A$ is an independent set in $T$;
+\smallskip
+\item{ii)} if $A$ is an independent set in $T$ then $\prk(T)\ge |A|$; and
+\smallskip
+\item{iii)} the partition rank of $T$ is low.
+\medskip
+In these notes, we aim to show that this general strategy may be performed with the partition rank
+replaced by something called the analytic rank.
+
+\medskip\boldlabel The analytic rank.
+In a 2011 paper, W.~T.~Gowers and J.~Wolf introduced another definition of rank that is Fourier-analytic
+in nature. Now we require the field $\FF$ to be finite, and let $\chi:\FF\to\CC$ be any nontrivial
+additive character. Recall that for such a character, $\ex_{a\in \FF} \chi(a) = 0$.
+The {\it bias} of a tensor $T : V^d\to \FF$ is the average
+$$\bias(T) = \ex_{x\in V^d} \chi\bigl(T(x)\bigr).$$
+Note that if $T$ is a linear form (i.e., an order-$1$ tensor) that is not identically zero,
+then $\bias(T) = 0$, since we can bring the sum inside all three functions and the sum over all elements
+a vector space over a finite field is zero. If $T$ is identically zero then $\bias(T) = 1$.
+Now to see that the bias of a tensor is always in $(0,1]$, note that if we fix any $(x_2, \ldots, x_d)\in V^{d-1}$,
+then $T(x_1,x_2,\ldots,x_d)$ becomes a linear form (order-$1$ tensor) in $x_1$ and
+$$\eqalign{
+\bias(T)
+&= \ex_{x_2,\ldots,x_d\in V} \ex_{x_1\in V} \chi\bigl(T(x_1,\ldots,x_d)\bigr)\cr
+&=\pr_{x_2,\ldots,x_d\in V} \bigl\{ T(x_1,\ldots,x_d) \equiv 0\bigr\},\cr
+}$$
+from our earlier observation about order-$1$ tensors.
+
+The {\it analytic rank} is defined to be the quantity
+$$\ark(T) = -\log_{|\FF|} \bias(T);$$
+since $\bias(T) \in (0,1]$ we have $\ark(T)\ge 0$. In the case of order-$2$ tensors, the analytic rank
+is once again equivalent to ordinary matrix rank. To see this, suppose that $T : (\FF^n)^2\to \FF$
+is defined as $T(x,y) = \sum_{i=1}^r x_iy_i$. Then $\bias(T)$ is the probability that, fixing
+$y$, the linear form $T(x,y)$ is identically zero. This is equivalent to every coordinate of $y$ being
+zero, which happens with probability $1/|\FF|^r$, and hence we see that $\ark(T) = r$.
+
+\advsect Subadditivity of analytic rank
+
+The goal of this section is to prove that if $T$ and $S$ are tensors, then $\ark(T+S)\le \ark(T)+\ark(S)$.
+Our first small lemma is the following.
+
+\proclaim Lemma \advthm. Let $W_0, W_1,\ldots,W_n : \FF^m\to \FF$ be functions. Let
+functions $A,B : \FF^n\times \FF^m\to \FF$ be given by
+$$A(x,y) = \sum_{i=1}^n x_i W_i(y)\qquad\hbox{and}\qquad B(x,y) = A(x,y) + W_0(y).$$
+Then
+$$\bigl| \bias(B)\bigr| \le \bias(A).$$
+
+\proof We expand
+$$\bias(B) = \ex_{x\in \FF^n, y\in \FF_m} \chi\bigl(B(x,y)\bigr)
+= \ex_{y\in \FF^m} \one_{W_1(y) = \cdots = W_n(y) = 0} \cdot \chi\bigl(W_0(y)\bigr)$$
+and by the triangle inequality,
+$$\bigl|\bias(B)\bigr| \le \ex_y \one_{W_1(y) = \cdots = W_n(y) = 0} = \bias(A).\noskipslug$$
+
+This lemma is used to prove a bound on the bias of a certain sum of tensors.
+First, we introduce some notation. With some $d$ fixed, we let $\x = (x_1,\ldots,x_d)$ and
+similarly for $\y = (y_1,\ldots,y_d)$. Then for
+$I\subseteq [d]$, define $I^c = [d]\setminus I$ and let $\x_I = (x_i : i\in I)$.
+
+\proclaim Lemma \advthm. Let $d\ge 1$ and for each $I\subseteq [d]$, let $R_I : V^I\to \FF$
+be an order-$|I|$ tensor. Consider the function
+$$R(\x) = \sum_{I\subseteq [d]} R_I(x_I).$$
+Then
+$$\bigl|\bias(R)\bigr| \le \bias(R_{[d]}).$$
+
+\proof Fix some $i\in [d]$ and write $R(\x)$ as
+$$R(\x) = \sum_{I \ni i} R_I(x_I) + \sum_{I\not\ni i} R_I(x_I).$$
+Setting $x = x_i$ and $y = \x_{[d]\setminus\{i\}}$, the first sum has the form
+$$\sum_{i=1}^n x_iW_i(y)$$ and the second sum does not depend on $x$ at all, so we can set it
+to be $W_0(y)$. The previous lemma now tells us that
+$$\bigl|\bias(R)\bigr| \le \bias\Bigl(\sum_{I\ni i} R_I(\x_I)\Bigr).$$
+Now iterate this with $i=d$ all the way down to $i=1$ (replacing $d$ by $d-1$ each time)
+to get the statement of the lemma.\slug
+
+\proclaim Theorem \advthm. Let $T,S: V^d\to \FF$ be order-$d$ tensors. Then
+$$\ark(T+S) \le \ark(T) + \ark(S).$$
+
+\section References
 

### 2022-04-23 misc/expo/tensorrank.tex
Added Matej

@@
 \def\y{{\bf y}}
+\def\b{{\bf b}}
 
@@
 
-\maketitle{The analytic rank of a tensor}{by}{Marcel K. Goh}{13 April 2022}
+\maketitle{The analytic rank of a tensor}{by}{Marcel K. Goh}{22 April 2022}
 
@@
 some results in a 2019 paper entitled ``The analytic rank of tensors and its applications'' by
-S.~Lovett. I go more deeply into definitions (for my own sake) so for other students this may be easier to
+S.~Lovett. I rearranged results in an order that makes more sense to me.
+I also go more deeply into definitions (for my own sake) and skip fewer steps,
+so for other students this may be easier to
 follow than the original paper.
@@
 
+\edef\normbound{\the\thmcount}
 \proclaim Lemma \advthm. Let $d\ge 1$ and for each $I\subseteq [d]$, let $R_I : V^I\to \FF$
@@
 
+Before proving subadditivity, we introduce another bit of notation. For $I\subseteq [d]$,
+let
+$$T_I(\x,\y) = T(\x_I, y_{I^c}) = T(z_1, \ldots, z_d),$$
+where $z_i = x_i$ if $i\in I$ and $z_i = y_i$ if $i\notin I$. After expanding out the definition
+of multilinearity, we see that $T(\x+\y)$ decomposes as
+$$ T(\x+\y) = \sum_{I\subseteq[d]} T_I(\x,\y).$$
+
+\edef\subadditivity{\the\thmcount}
 \proclaim Theorem \advthm. Let $T,S: V^d\to \FF$ be order-$d$ tensors. Then
 $$\ark(T+S) \le \ark(T) + \ark(S).$$
+
+\proof It is enough to show that
+$$\bias(T+S) \ge \bias(T)\bias(S).$$
+We express
+$$\eqalign{
+\bias(T)\bias(S) &= \Bigl(\ex_{\x\in V^d} \chi\bigl(T(\x)\bigr)\Bigr)
+                    \Bigl(\ex_{\y\in V^d} \chi\bigl(T(\y)\bigr)\Bigr)\cr
+&= \ex_{\x\in X} \ex_{\y\in Y} \chi\bigl( T(\x) + S(\y)\bigr) \cr
+&= \ex_{\x\in X} \ex_{\y\in Y} \chi\bigl( T(\x) + S(\x+\y)\bigr) \cr
+&= \bias\bigl( T(\x) + S(\x+\y)\bigr), \cr
+}$$
+and by our earlier decomposition the tensor product of a sum, this gives us
+$$\bias(T)\bias(S)
+= \bias \Bigl( T(\x) + \sum_{I\subseteq [d]} S_I(\x,\y) \Bigr)
+= \bias \Bigl( (T+S)(\x) + \sum_{I\ne [d]} S_I(\x,\y) \Bigr), $$
+Setting $\y = \b\in V^d$ to maximise the norm of this right-hand side, we have
+$$\bias(T)\bias(S) \le \Bigl| \bias\Bigl( (T+S)(\x) + \sum_{I\ne [d]} S_I(\x, \b)\Bigr)\Bigr|.$$
+Here $S_I(\x,\b)$ is an order-$|I|$ tensor in the variables $\x_I$. Thus we may apply
+the previous lemma with
+$$R_{[d]}(\x) = (T+S)(\x)\qquad\hbox{and}\qquad R_I(\x_I) = S_I(\x,\b)\ \hbox{for}\ I\ne [d]$$
+to get $\bias(T)\bias(S) \le \bias(T+S)$.\slug
+
+As an application of this theorem, we show that common roots of tensors on a common input are
+positively correlated.
+
+\proclaim Corollary \advthm. Let $T_1,\ldots,T_m, S_1,\ldots,S_n : V^d\to \FF$ be order-$d$ tensors.
+Then
+$$\eqalign{
+\pr_{x\in V^d}&\bigl\{ T_1(x) = \cdots = T_m(x) = S_1(x) = \cdots = S_n(x) = 0\bigr\}\ge \cr
+&\pr_{x\in V^d}\bigl\{ T_1(x) = \cdots = T_m(x) = 0\}\cdot
+\pr_{x\in V^d}\bigl\{ S_1(x) = \cdots = S_n(x) = 0\}
+}$$
+
+\proof Define order-$(d+1)$ tensors $T$ and $S$ by setting
+$$T(x_0,x_1,\ldots,x_d) = \sum_{i=1}^m x_{0,i} T_i(x_1,\ldots, x_d)$$
+and
+$$S(x_0,x_1,\ldots,x_d) = \sum_{i=m+1}^{m+n} x_{0,i} S_{i-m}(x_1,\ldots,x_d).$$
+The left-hand side of the desired inequality is $\bias(T+S)$ and the right-hand side is
+$\bias(T)\bias(S)$.\slug
+
+\advsect Analytic rank and partition rank
+
+In this section, we shall discuss properties of the analytic rank that allow it to be used as
+the partition rank was used above in the cap-set problem. First, we show that the analytic rank
+is bounded above by the partition rank.
+
+\proclaim Theorem \advthm. Let $T:V^d\to\FF$ be an order-$d$ tensor. Then $\ark(T)\le \prk(T)$.
+
+\proof By Theorem~\subadditivity, we only need to consider the case where $T$ has partition rank one.
+So we assume that $T:V^d\to \FF$ factors as
+$$T(\x) = T_1(\x_A)T_2(\x_B),$$
+where $A\cup B = [d]$ is a nontrivial partition of $[d]$. We shall show that $\ark(T)\le 1 = \prk(T)$ by
+showing that $\bias(T) \ge 1/|\FF|$. For $a,b\in \FF$ define the function
+$$ F_{a,b}(\x) = \bigl(T_1(\x_A) + a\bigr) \bigl( T_2(\x_B) + b\bigr).$$
+Letting $R_{[d]} = T$, $R_{\emptyset} = ab$, $R_A = bT_1(\x_A)$, $R_B = aT_2(\x_B)$, and $R_I$ be the
+zero tensor (of the corresponding order)
+for all other $I\subseteq [d]$, we see that $R = F_{a,b}$ and
+by Lemma~\normbound, $\bigl|\bias(F_{a,b})\bigr| \le \bias(T)$.
+On the other hand, we have
+$$\eqalign{
+\ex_{a,b\in \FF} \bias(F_{a,b}) &= \ex_{a,b\in \FF} \ex_{\x\in V^d}
+   \chi\bigl( (T_1(\x_A) + a)(T_2(\x_B) + b) \bigr) \cr
+&= \ex_{a,b\in F} \chi(ab) \cr
+&= \pr_{b\in \FF} \{b=0\} \cr
+&= {1\over |\FF|}. \cr
+}$$
+So we may find $a,b$ such that $\bias(T) \ge \bigl| \bias(F_{a,b})\bigr| \ge 1/|\FF|$, which
+is what we wanted.\slug
+
+Since we are often interested in finding some upper bound on the partition rank in applications,
+this theorem suggests that it should actually be easier to upper bound the analytic
+rank, though {\it ad hoc} methods to do so have not currently been developed.
+
+In applications, it is also important that the partition rank of a tensor $T$
+is bounded below by the size of an
+independent set in $T$. Towards showing a version of this for the analytic rank, we first show
+that the analytic rank does not increase when the tensor is restricted to a subspace.
+
+\proclaim Lemma \advthm. Let $T : V^d \to \FF$ be an order-$d$ tensor and let $U\subseteq V$ be a
+subspace. If $T|_U : U^d\to \FF$ is the restriction of $T$ to $U$, then $\ark(T|_U) \le \ark(T)$.
+
+\proof Let $W\subseteq V$ be a subspace so that $V = U\oplus W$. Then any $v\in V$ has a unique
+expression as $v = u+w$ for $u\in U$ and $w\in W$, so
+$$\bias(T) = \ex_{u_1,\ldots,u_d\in U} \ex_{w_1,\ldots,w_d\in W}\chi\bigl(T(u_1+w_1,\ldots,u_d+w_d)\bigr).$$
+Fix an arbitrary choice of $w_1,\ldots,w_d\in W$. Then
+$$T(u_1+w_1,\ldots, u_d+w_d) = \sum_{I\subseteq [d]} T_I(u_I, w_{I^c}),$$
+where $T_I(u_I, w_{I^c}) = T(z_1,\ldots,z_d)$ where $z_i = u_i$ if $i\in I$ and $z_i = w_i$ if $i\notin I$.
+By Lemma~\normbound, we have
+$$\eqalign{
+\bigl| \ex_{u_1,\ldots,u_d\in U} \chi\bigl( T(u_1+w_1,\ldots,u_d+w_d) \bigr) \bigr|
+&\le \ex_{u_1,\ldots,u_d\in U} \chi\bigl( T(u_1,\ldots,u_d)\bigr) \cr
+&= \bias(T|_U) \cr
+}$$
+Since this is for all $w_1,\ldots, w_d$, averaging the left-hand side over all choices and
+applying the triangle inequality gives us what we want.\slug
+
+We are now able to show that for any tensor $T$ and independent set $A$ in $T$, we have
+$\ark(T) \ge c|A|$ for $c$ depending on $d$ and $|\FF|$.
+
+\proclaim Theorem \advthm. Let $T : (\FF^n)^d\to \FF$ be an order-$d$ tensor. Assume that
+$A\subseteq [n]$ is an independent set in $T$. Then $\ark(T)\ge c|A|$ where
+$$c = -\log_{|\FF|} \biggl( 1-\biggl( 1-{1\over y}\biggr)^{d-1}\biggr),$$
+so that $c\ge 2^{-d}$ and $c\ge 1-\log(d-1)/\log |\FF|$.
+
+\proof Let $S : (F^A)^d\to \FF$ be the restriction of $T$ to $\FF^A$.
+By the previous lemma, $\ark(S)$ is a lower bound on $\ark(T)$, so we shall prove the theorem
+by expressing $\ark(S)$ as a function of $|A|$. We have
+$$\bias(A) = \ex_{x_1,\ldots,x_d\in \FF^A} \chi\Bigl( \sum_{i\in A} c_i \prod_{j\in [d]} x_{i,j}\Bigr),$$
+where, since $A$ is an independent set, $c_i\ne 0$ for all $i\in A$. But we also know that
+$$\bias(S) =
+\pr_{x_2,\ldots,x_d\in \FF^A} \Bigl\{ \bigcap_{i\in A} \prod_{2\le j\le d} \{x_{i,j} = 0\}\Bigr\},$$
+and this is easily seen to be
+$$\bias(S) = \biggl( 1-\biggl( 1-{1\over |\FF|}\biggr)^{d-1} \biggr)^{|A|}.$$
+Taking the logarithm of both sides, we see that $\ark(S) = c|A|$, where $c$ is as in the theorem
+statement.
+
+By convexity,
+$$c \ge -\log_2 (1-2^{-(d-1)}) \ge 2^{-(d-1)}.$$
+Also, as long as $|\FF|\ge d$, we have
+$$c \ge -\log_{|\FF|}\bigl((d-1)/|\FF|\bigr) = 1-\log_{|\FF|} (d-1).\noskipslug$$
 

### 2022-04-26 misc/expo/tensorrank.tex
Tensorrank typos

@@
 and here we shall sketch how it applies. A {\it cap set} is a subset $A\subseteq \FF_3^n$ such that for
-every triple $(x,y,z)\in A$ of pairwise distinct elements, $x+y+z\ne 0$. In 1984, T.~C.~Brown
-and J.~P.~Buhler showed that, loosely speaking, cap sets have zero density.
+every triple $(x,y,z)\in A$ of pairwise distinct elements, $x+y+z\ne 0$. It was shown by T.~C.~Brown
+and J.~P.~Buhler that, loosely speaking, cap sets have zero density.
 
@@
 Given $A\subseteq \FF_3^n$, let $T : V^3 \to \FF_3$, where $V = {\FF_3}^{\FF_3^n}$, be given by
-$$f(e_a,e_b,e_c = \cases{1, & if $a+b+c=0$;\cr 0, & otherwise,}$$
+$$T(e_a,e_b,e_c) = \cases{1, & if $a+b+c=0$;\cr 0, & otherwise,}$$
 for basis vectors $e_v$ and extended to all other vectors by linearity. (The function $e_v$ has
@@
 \proof We expand
-$$\bias(B) = \ex_{x\in \FF^n, y\in \FF_m} \chi\bigl(B(x,y)\bigr)
-= \ex_{y\in \FF^m} \one_{W_1(y) = \cdots = W_n(y) = 0} \cdot \chi\bigl(W_0(y)\bigr)$$
+$$\bias(B) = \ex_{x\in \FF^n, y\in \FF^m} \chi\bigl(B(x,y)\bigr)
+= \ex_{y\in \FF^m} \indic{W_1(y) = \cdots = W_n(y) = 0} \cdot \chi\bigl(W_0(y)\bigr)$$
 and by the triangle inequality,
-$$\bigl|\bias(B)\bigr| \le \ex_y \one_{W_1(y) = \cdots = W_n(y) = 0} = \bias(A).\noskipslug$$
+$$\bigl|\bias(B)\bigr| \le \ex_y \indic{W_1(y) = \cdots = W_n(y) = 0} = \bias(A).\noskipslug$$
 
@@
 let
-$$T_I(\x,\y) = T(\x_I, y_{I^c}) = T(z_1, \ldots, z_d),$$
+$$T_I(\x,\y) = T(\x_I, \y_{I^c}) = T(z_1, \ldots, z_d),$$
 where $z_i = x_i$ if $i\in I$ and $z_i = y_i$ if $i\notin I$. After expanding out the definition
@@
                     \Bigl(\ex_{\y\in V^d} \chi\bigl(T(\y)\bigr)\Bigr)\cr
-&= \ex_{\x\in X} \ex_{\y\in Y} \chi\bigl( T(\x) + S(\y)\bigr) \cr
-&= \ex_{\x\in X} \ex_{\y\in Y} \chi\bigl( T(\x) + S(\x+\y)\bigr) \cr
+&= \ex_{\x\in V^d} \ex_{\y\in V^d} \chi\bigl( T(\x) + S(\y)\bigr) \cr
+&= \ex_{\x\in V^d} \ex_{\y\in V^d} \chi\bigl( T(\x) + S(\x+\y)\bigr) \cr
 &= \bias\bigl( T(\x) + S(\x+\y)\bigr), \cr
@@
 $A\subseteq [n]$ is an independent set in $T$. Then $\ark(T)\ge c|A|$ where
-$$c = -\log_{|\FF|} \biggl( 1-\biggl( 1-{1\over y}\biggr)^{d-1}\biggr),$$
+$$c = -\log_{|\FF|} \biggl( 1-\biggl( 1-{1\over |\FF|}\biggr)^{d-1}\biggr),$$
 so that $c\ge 2^{-d}$ and $c\ge 1-\log(d-1)/\log |\FF|$.
@@
 by expressing $\ark(S)$ as a function of $|A|$. We have
-$$\bias(A) = \ex_{x_1,\ldots,x_d\in \FF^A} \chi\Bigl( \sum_{i\in A} c_i \prod_{j\in [d]} x_{i,j}\Bigr),$$
+$$\bias(S) = \ex_{x_1,\ldots,x_d\in \FF^A} \chi\Bigl( \sum_{i\in A} c_i \prod_{j\in [d]} x_{i,j}\Bigr),$$
 where, since $A$ is an independent set, $c_i\ne 0$ for all $i\in A$. But we also know that
 $$\bias(S) =
-\pr_{x_2,\ldots,x_d\in \FF^A} \Bigl\{ \bigcap_{i\in A} \prod_{2\le j\le d} \{x_{i,j} = 0\}\Bigr\},$$
+\pr_{x_2,\ldots,x_d\in \FF^A} \Bigl\{ \bigcap_{i\in A} \{x_{i,2}\cdots x_{i,d} = 0\}\Bigr\},$$
 and this is easily seen to be

### 2022-07-01 about/vita.tex
New paper

@@
 
-\leftright{Citizen of Canada and Singapore}{3660 Rue Hutchison, Apt.\ 4}
-\leftright{Email: marcel\_goh@yahoo.ca}{Montr\'eal, QC H2X 2H3, Canada}
+\leftright{Citizen of Canada and Singapore}{2-9602 74th Avenue NW}
+\leftright{Email: marcel\_goh@yahoo.ca}{Edmonton, AB T6E 1E6}
 \leftright{Webpage: marcelgoh.ca}{+1\thinspace\up(825\up)\thinspace 440-0681}
@@
 \begingroup\parindent=10pt
-\thing (with Luc Devroye and Rosie Y. Zhao)
-The independence number of a Bienaym\'e--Galton--Watson tree
-and related parameters. {\sl arXiv preprint 2106.14389}, 20~pp.
+\thing (with Jonah Saks)
+Alternating-sum statistics for certain sets of integers. {\sl arXiv:2206.12535} (2022), 12~pp.
 \smallskip
 \thing (with Jad Hamdan and Jonah Saks)
-The lattice of arithmetic progressions. {\sl arXiv preprint 2106.05949}, 15~pp.
+The lattice of arithmetic progressions. {\sl arXiv preprint 2106.05949} (2021), 15~pp.
 \endgroup

### 2022-07-01 about/vita_professional.tex
New paper

@@
 
-\leftright{+1\thinspace\up(825\up)\thinspace 440-0681}{3660 Rue Hutchison, Apt.\ 4}
-\leftright{Email: marcel\_goh@yahoo.ca}{Montr\'eal, QC H2X 2H3, Canada}
-\leftright{Webpage: marcelgoh.ca}{}
+\leftright{+1\thinspace\up(825\up)\thinspace 440-0681}{2-9602 74 Avenue NW, Apt.\ 4}
+\leftright{Email: marcel\_goh@yahoo.ca}{Edmonton, AB T6E 1E6}
+\leftright{Webpage: marcelgoh.ca}{Canada}
 
@@
 
-\sectheader{Awards and Funding}
+\sectheader{Skills}
 
-\leftright{\bf NSERC Canada Graduate Scholarship -- Master's \rm (\$17,500)}{\bf May 2021 -- April 2022}
+{\bf Programming Languages}\par
+\CEE, OCaml, Python, Java, PostScript, Haskell, Scheme Lisp,  Standard ML, {\tt CWEB}, {\tt MIXAL}, MIPS Assembly.
+\medbreak
+
+{\bf Technologies}\par
+\UNIX, Vim, \TeX, Git.
+\medbreak
+
+{\bf Languages}\par
+Fluent: English, French. Proficient: Mandarin, Italian.
+\medbreak
+
+\sectheader{Selected Software Projects}
+
+\leftright{\bf Tlingit verb parsing}{}
 \smallskip
-\leftright{\bf NSERC Undergraduate Student Research Award \rm (\$7,125)}{\bf May -- August 2020}
+A rule-based program that finds regularity in Tlingit verbs, written in Python.
+
+\noindent[{\tt https://github.com/marcelgoh/tlingit-verb-parsing}]
+\medbreak
+
+\leftright{\bf Sorting algorithms on manifolds}{}
 \smallskip
-\leftright{\bf Governor General's Academic Medal -- Bronze}{\bf June 2015}
+Research project under the supervision of Michael Lipnowski.
+Studied algorithms that investigate the topology of group actions on locally symmetric spaces.
+Work involved writing of code in OCaml and PostScript that generates illustrations of certain quotient
+spaces, in various models of hyperbolic geometry.
 
-\sectheader{Graduate-level Coursework}
+\noindent[{\tt https://github.com/marcelgoh/manifold-sorting}]
+\medbreak
 
-Higher algebra (categories, groups, modules), functional analysis, quadratic and modular forms, combinatorics,
-topology (including some algebraic topology), probabilistic analysis of algorithms.
+\leftright{\bf Bytecode compiler and interpreter}{}
+\smallskip
+Under the supervision of Adam Dingle, wrote a bytecode compiler and virtual machine for a subset of Python.
+OPythn includes support for lists, dictionaries, named and anonymous functions, objects, and classes.
+It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing.
 
-\sectheader{Research Interests}
+\noindent[{\tt https://github.com/marcelgoh/opythn}]
+\medbreak
 
-Analysis of random discrete structures, analysis and design of algorithms, enumerative combinatorics,
-additive combinatorics.
+\leftright{\bf Alonzo $\lambda$-calculus interpreter}{}
+\smallskip
+Wrote an interpreter for the untyped $\lambda$-calculus that performs $\alpha$-conversion
+and $\beta$-reduction. Includes an interactive environment that allows users to bind names to
+terms. Implemented in Haskell.
+
+\noindent[{\tt https://github.com/marcelgoh/alonzo}]
+\medbreak
+
+\leftright{\bf Escape-time fractals}{}
+\smallskip
+Made an interactive graphical program to generate and explore fractals. Implemented using the SDL2
+graphics library and written in \CEE. (HackPrague 2019)
+
+\noindent[{\tt https://github.com/mattonicorp/mattoni}]
+\medbreak
+
+\leftright{\bf Nim}{}
+\smallskip
+Implemented the game of Nim graphically using Processing. Worked on game logic and created the soundtrack
+using an $8$-bit sequencer. (ConUHacks IV)
+
+\noindent[{\tt https://github.com/conudihedral4/nim}]
+\medbreak
+
+\leftright{\bf CourseTalk}{}
+\smallskip
+Created a web chat app using Node.js and React.js. (HackPrinceton Fall 2018)
+
+\noindent[{\tt https://github.com/marcelgoh/hackprinceton-2018}]
+\medbreak
 
@@
 \begingroup\parindent=10pt
+\thing (with Jonah Saks)
+Alternating-sum statistics for certain sets of integers. {\sl arXiv:2206.12535} (2022), 12~pp.
+\smallskip
 \thing (with Jad Hamdan and Jonah Saks)
-The lattice of arithmetic progressions. {\sl arXiv preprint 2106.05949}, 15~pp.
+The lattice of arithmetic progressions. {\sl arXiv:2106.05949} (2021), 15~pp.
 \endgroup
@@
 
-\sectheader{Selected Software Projects}
+\sectheader{Awards and Funding}
 
-\leftright{\bf Tlingit verb parsing}{\bf February -- April 2021}
+\leftright{\bf NSERC Canada Graduate Scholarship -- Master's \rm (\$17,500)}{\bf May 2021 -- April 2022}
 \smallskip
-A rule-based program that finds regularity in Tlingit verbs, written in Python.
+\leftright{\bf NSERC Undergraduate Student Research Award \rm (\$7,125)}{\bf May -- August 2020}
+\smallskip
+\leftright{\bf Governor General's Academic Medal -- Bronze}{\bf June 2015}
 
-\noindent[{\tt https://github.com/marcelgoh/tlingit-verb-parsing}]
-\medbreak
+\sectheader{Graduate-level Coursework}
 
-\leftright{\bf Sorting algorithms on manifolds}{\bf May -- August 2020}
-\smallskip
-Research project under the supervision of Michael Lipnowski.
-Studied algorithms that investigate the topology of group actions on locally symmetric spaces.
-Work involved writing of code in OCaml and PostScript that generates illustrations of certain quotient
-spaces, in various models of hyperbolic geometry.
+Higher algebra (categories, groups, modules), functional analysis, quadratic and modular forms, combinatorics,
+topology (including some algebraic topology), probabilistic analysis of algorithms.
 
-\noindent[{\tt https://github.com/marcelgoh/manifold-sorting}]
-\medbreak
+\goodbreak
+\sectheader{Research Interests}
 
-\leftright{\bf Bytecode compiler and interpreter}{\bf February -- June 2019}
-\smallskip
-Under the supervision of Adam Dingle, wrote a bytecode compiler and virtual machine for a subset of Python.
-OPythn includes support for lists, dictionaries, named and anonymous functions, objects, and classes.
-It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing.
-
-\noindent[{\tt https://github.com/marcelgoh/opythn}]
-\medbreak
-
-\leftright{\bf Alonzo $\lambda$-calculus interpreter}{\bf May -- June 2019}
-\smallskip
-Wrote an interpreter for the untyped $\lambda$-calculus that performs $\alpha$-conversion
-and $\beta$-reduction. Includes an interactive environment that allows users to bind names to
-terms. Implemented in Haskell.
-
-\noindent[{\tt https://github.com/marcelgoh/alonzo}]
-\medbreak
-
-\leftright{\bf Escape-time fractals}{\bf May 2019}
-\smallskip
-Made an interactive graphical program to generate and explore fractals. Implemented using the SDL2
-graphics library and written in \CEE. (HackPrague 2019)
-
-\noindent[{\tt https://github.com/mattonicorp/mattoni}]
-\medbreak
-
-\leftright{\bf Nim}{\bf January 2019}
-\smallskip
-Implemented the game of Nim graphically using Processing. Worked on game logic and created the soundtrack
-using an $8$-bit sequencer. (ConUHacks IV)
-
-\noindent[{\tt https://github.com/conudihedral4/nim}]
-\medbreak
-
-\leftright{\bf CourseTalk}{\bf November 2018}
-\smallskip
-Created a web chat app using Node.js and React.js. (HackPrinceton Fall 2018)
-
-\noindent[{\tt https://github.com/marcelgoh/hackprinceton-2018}]
-\medbreak
+Analysis of random discrete structures, analysis and design of algorithms, enumerative combinatorics,
+additive combinatorics.
 
@@
 
-\sectheader{Skills}
-
-{\bf Programming Languages}\par
-\CEE, OCaml, Python, Java, PostScript, Haskell, Scheme Lisp,  Standard ML, {\tt CWEB}, {\tt MIXAL}, MIPS Assembly.
-\medbreak
-
-{\bf Technologies}\par
-\UNIX, Vim, \TeX, Git.
-\medbreak
-
-{\bf Languages}\par
-Fluent: English, French. Proficient: Mandarin, Italian.
-\medbreak
-
 \sectheader{Other}

### 2022-07-12 about/vita.tex
Added thesis

@@
 
-\reportcount=4
+\reportcount=5
 \pubbegin R
+\repitem Structural properties of conditional Galton--Watson trees. M.Sc.~thesis, McGill University
+(Montr\'eal, Qu\'ebec, August 2022), $\hbox{\rm vi} + 75$~pp.
 \repitem Finding regularity in Tlingit verb prefixes. Semester project report, McGill University
@@
 
-\leftright{\bf Mentor}{\bf January 2021 -- present}
+\starleftright{\bf Visiting teaching assistant}{\bf July 2022}
+\leftright{\bf Bocconi University}{\sl Como, CO, Italy}
+\smallskip
+Teaching assistant for a two-week summer school on random structures and combinatorial statistics,
+which was
+organised by Bocconi University in collaboration with Oxford University
+and Imperial College London. Led problem sessions in the afternoons, with
+around forty graduate-level students were in attendance.
+\medbreak
+
+\leftright{\bf Mentor}{\bf January -- April 2022}
 \leftright{\bf Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
@@
 mathematics in a casual setting. Focused on topics in extremal combinatorics related to the increasing
-triples problem.
+triples problem, as well as topics related to Khovanskii's theorem.
 \medbreak

### 2022-09-07 about/vita_small.tex
No longer a student

--- 2022-07-12 about/vita.tex
@@
 
-\sectheader{Personal and Contact Information}
+\sectheader{Contact Information}
 
-\leftright{Citizen of Canada and Singapore}{2-9602 74th Avenue NW}
+\leftright{Mobile phone: +1\thinspace\up(825\up)\thinspace 440-0681}{2-9602 74 Avenue NW}
 \leftright{Email: marcel\_goh@yahoo.ca}{Edmonton, AB T6E 1E6}
-\leftright{Webpage: marcelgoh.ca}{+1\thinspace\up(825\up)\thinspace 440-0681}
+
+\sectheader{Languages}
+
+Fluent: English, French. Proficient: Mandarin, Italian.
 
@@
 \smallskip
-Supervisor: Luc Devroye. CGPA: 4.00/4.
+Supervisor: Luc Devroye. CGPA: 4.0/4.
 Thesis: ``Structural properties of conditional Galton--Watson trees.''
@@
 
-\sectheader{Awards and Funding}
-
-\leftright{\bf NSERC Canada Graduate Scholarship -- Master's \rm (\$17,500)}{\bf May 2021 -- April 2022}
-\smallskip
-\leftright{\bf NSERC Undergraduate Student Research Award \rm (\$7,125)}{\bf May -- August 2020}
-\smallskip
-\leftright{\bf Governor General's Academic Medal -- Bronze}{\bf June 2015}
-
-\sectheader{Graduate-level Coursework}
-
-Higher algebra (categories, groups, modules), functional analysis, quadratic and modular forms, combinatorics,
-topology (including some algebraic topology), probabilistic analysis of algorithms.
-
-\sectheader{Research Interests}
-
-Analysis of random discrete structures, analysis and design of algorithms, enumerative combinatorics,
-additive combinatorics.
-
-\sectheader{Papers}
-
-\begingroup\frenchspacing
-\papercount=4
-\pubbegin P
-\papitem (with Luc Devroye and Rosie Y. Zhao)
-On the peel number and the leaf-height of a Galton--Watson tree.
-To appear in {\sl Combinatorics, Probability and Computing}, 20~pp. [arXiv:2106.14389]
-\papitem (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
-Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl Discrete Mathematics
-and Theoretical Computer Science} {\bf 24},1 (2022), \#7, 16~pp. [arXiv:2105.12046]
-\papitem (with Anna M. Brandenberger and Luc Devroye)
-Root estimation in Galton--Watson trees. To appear in {\sl Random Structures and Algorithms}, 24~pp.
-[arXiv:2007.05681]
-\papitem (with Rosie Y. Zhao)
-Arithmetic subsequences in a random ordering of an additive set.
-{\sl Integers: Electronic Journal of Combinatorial Number Theory} {\bf 21} (2021), \#A89, 19~pp.
-[arXiv:2012.12339]
-
-\sectheader{Submitted Papers}
-
-\begingroup\parindent=10pt
-\thing (with Jonah Saks)
-Alternating-sum statistics for certain sets of integers. {\sl arXiv:2206.12535} (2022), 12~pp.
-\smallskip
-\thing (with Jad Hamdan and Jonah Saks)
-The lattice of arithmetic progressions. {\sl arXiv preprint 2106.05949} (2021), 15~pp.
-\endgroup
-\endgroup%end the frenchspacing
-
-\goodbreak
-\sectheader{Reports}
-
-\reportcount=5
-\pubbegin R
-\repitem Structural properties of conditional Galton--Watson trees. M.Sc.~thesis, McGill University
-(Montr\'eal, Qu\'ebec, August 2022), $\hbox{\rm vi} + 75$~pp.
-\repitem Finding regularity in Tlingit verb prefixes. Semester project report, McGill University
-(Montr\'eal, Qu\'ebec, April 2021), 7 pp.
-\repitem Grid-building algorithms on manifolds. Summer research report, McGill University (Montr\'eal, Qu\'ebec,
-August 2020), 10 pp.
-\repitem Typechecking proof scripts: making interactive proof assistants robust.
-Honours project report, McGill University (Montr\'eal, Qu\'ebec, December 2019), 10 pp.
-\repitem The OPythn programming language. Software project report, Charles University (Prague,
-Czech Republic, June 2019), 10 pp.
-
-\sectheader{Research Experience}
-
-\leftright{\bf Probabilistic analysis of branching processes}{\bf May 2020 -- present}
-\leftright{\sl Research group}{\sl McGill University}
-\smallskip
-Ongoing research on branching processes, headed by Luc Devroye. Studied estimation problems
-on Galton-Watson trees. Attended informal seminars, discussing
-various topics related to branching
-processes and other topics in probability and combinatorics.
-(Gave a presentation at six of these: one on root estimation in Galton-Watson
-trees, two on generating functions and elementary analytic combinatorics, two on graph regularity, and
-one concerning two of Erd\H os's proofs on prime numbers.)
-
-\medbreak
-\leftright{\bf Sorting algorithms on manifolds}{\bf May -- August 2020}
-\leftright{\sl Summer research project}{\sl McGill University}
-\smallskip
-Research project under the supervision of Michael Lipnowski.
-Studied algorithms that investigate the topology of group actions on locally symmetric spaces.
-Work involved writing of code in OCaml and PostScript that generates illustrations of certain quotient
-spaces, in various models of hyperbolic geometry.
-\medbreak
-
-\leftright{\bf Interactive proofs}{\bf September -- December 2019}
-\leftright{\sl Honours research project}{\sl McGill University}
-\smallskip
-Semester-long research project in the Computation and Logic Group, supervised
-by Brigitte Pientka.
-Proved a theorem in constructive logic concerning the formal verification
-of the interactive proof assistant Harpoon and wrote OCaml code as part of
-ongoing work on the functional programming language Beluga.
-\medbreak
-
-\leftright{\bf Bytecode compiler and interpreter}{\bf February -- June 2019}
-\leftright{\sl Individual software project}{\sl Charles University}
-\smallskip
-Semester-long individual software project under the supervision of Adam Dingle.
-Created OPythn, a bytecode compiler and virtual machine for a subset of Python.
-OPythn includes support for lists, dictionaries, named and anonymous functions, objects, and classes.
-It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing.
-
 \sectheader{Work and Volunteer Experience {\rm($*$ indicates a paid position)}}
@@
 organised by Bocconi University in collaboration with Oxford University
-and Imperial College London. Led problem sessions in the afternoons, with
+and Imperial College London. Led problem sessions in the afternoons;
 around forty graduate-level students were in attendance.
@@
 \leftright{\bf Mentor}{\bf January -- April 2022}
-\leftright{\bf Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
+\leftright{\sl Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
 \smallskip
-Meet with two undergraduate students as part of the Directed Reading Program
+Met with two undergraduate students as part of the Directed Reading Program
 on a weekly basis to give them an introduction to research-level
@@
 Teaching assistant for COMP 690, a graduate-level course on the probabilistic analysis of algorithms.
-Hold office hours twice a week and responsible for grading of assignments.
+Held office hours twice a week and responsible for grading of assignments.
 \medbreak
@@
 
-\sectheader{Skills}
+\sectheader{Awards and Funding}
 
-{\bf Programming Languages}\par
-\CEE, OCaml, Python, Java, PostScript, Haskell, Scheme Lisp,  Standard ML, {\tt CWEB}, {\tt MIXAL}, MIPS Assembly.
-\medbreak
+\leftright{\bf NSERC Canada Graduate Scholarship -- Master's \rm (\$17,500)}{\bf May 2021 -- April 2022}
+\smallskip
+\leftright{\bf NSERC Undergraduate Student Research Award \rm (\$7,125)}{\bf May -- August 2020}
+\smallskip
+\leftright{\bf Governor General's Academic Medal -- Bronze}{\bf June 2015}
 
-{\bf Technologies}\par
-\UNIX, Vim, \TeX, Git.
-\medbreak
+\sectheader{Papers}
 
-{\bf Languages}\par
-Fluent: English, French. Proficient: Mandarin, Italian.
-\medbreak
+\begingroup\frenchspacing
+\papercount=5
+\pubbegin P
+\papitem (with Jad Hamdan and Jonah Saks)
+The lattice of arithmetic progressions. To appear in {\sl Australasian Journal of Combinatorics}, 15~pp.
+[arXiv:2106.05949]
+\papitem (with Luc Devroye and Rosie Y. Zhao)
+On the peel number and the leaf-height of a Galton--Watson tree.
+To appear in {\sl Combinatorics, Probability and Computing}, 20~pp. [arXiv:2106.14389]
+\papitem (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
+Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl Discrete Mathematics
+and Theoretical Computer Science} {\bf 24},1 (2022), \#7, 16~pp. [arXiv:2105.12046]
+\papitem (with Anna M. Brandenberger and Luc Devroye)
+Root estimation in Galton--Watson trees. {\sl Random Structures and Algorithms} {\bf 61},3 (2022), 520--542.
+[arXiv:2007.05681]
+\papitem (with Rosie Y. Zhao)
+Arithmetic subsequences in a random ordering of an additive set.
+{\sl Integers: Electronic Journal of Combinatorial Number Theory} {\bf 21} (2021), \#A89, 19~pp.
+[arXiv:2012.12339]
 
-\sectheader{Other}
+\sectheader{Submitted Papers}
 
-\parindent=10pt
-\thing Recipient of 0x\$3.40 in Knuth reward cheques.
-\smallskip
-\thing Contributed sequences A335562, A338550, A338993, A339941, A339942, A341822, and A347580
-to the On-line Encyclopedia of Integer Sequences.
-\medbreak
-
-\filbreak
+\begingroup\parindent=10pt
+\thing (with Jonah Saks)
+Alternating-sum statistics for certain sets of integers. {\sl arXiv:2206.12535} (2022), 12~pp.
+\endgroup
+\endgroup%end the frenchspacing
 

### 2022-10-06 about/vita_academic.tex
Added Australasia and cows

--- 2022-07-12 about/vita.tex
@@
 
-\sectheader{Personal and Contact Information}
-
-\leftright{Citizen of Canada and Singapore}{2-9602 74th Avenue NW}
-\leftright{Email: marcel\_goh@yahoo.ca}{Edmonton, AB T6E 1E6}
-\leftright{Webpage: marcelgoh.ca}{+1\thinspace\up(825\up)\thinspace 440-0681}
-
 \sectheader{Education}
@@
 Supervisor: Luc Devroye. CGPA: 4.00/4.
-Thesis: ``Structural properties of conditional Galton--Watson trees.''
+Thesis: Structural properties of conditional Galton--Watson trees.
 \medbreak
@@
 \begingroup\frenchspacing
-\papercount=4
+\papercount=5
 \pubbegin P
+\papitem (with Jad Hamdan and Jonah Saks)
+The lattice of arithmetic progressions.
+{\sl Australasian Journal of Combinatorics}\/ {\bf 84},3 (2022), 357--374.
+[arXiv:2106.05949]
 \papitem (with Luc Devroye and Rosie Y. Zhao)
@@
 Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl Discrete Mathematics
-and Theoretical Computer Science} {\bf 24},1 (2022), \#7, 16~pp. [arXiv:2105.12046]
+and Theoretical Computer Science}\/ {\bf 24},1 (2022), \#7, 16~pp. [arXiv:2105.12046]
 \papitem (with Anna M. Brandenberger and Luc Devroye)
-Root estimation in Galton--Watson trees. To appear in {\sl Random Structures and Algorithms}, 24~pp.
+Root estimation in Galton--Watson trees. {\sl Random Structures and Algorithms}\/ {\bf 61},3 (2022), 520--542.
 [arXiv:2007.05681]
@@
 Arithmetic subsequences in a random ordering of an additive set.
-{\sl Integers: Electronic Journal of Combinatorial Number Theory} {\bf 21} (2021), \#A89, 19~pp.
+{\sl Integers: Electronic Journal of Combinatorial Number Theory}\/ {\bf 21} (2021), \#A89, 19~pp.
 [arXiv:2012.12339]
@@
 Alternating-sum statistics for certain sets of integers. {\sl arXiv:2206.12535} (2022), 12~pp.
-\smallskip
-\thing (with Jad Hamdan and Jonah Saks)
-The lattice of arithmetic progressions. {\sl arXiv preprint 2106.05949} (2021), 15~pp.
 \endgroup

### 2022-10-06 misc/expo/meditations.tex
Added Australasia and cows

\input fontmac

\def\quo#1.{\par\textindent{\it #1.}}

\widemargins
\bookheader{MEDITATIONS OF A MATHEMATICS STUDENT}{MARCEL K. GOH}

\maketitle{Meditations of a mathematics graduate student}{}{Marcel K. Goh}{\sl after Marcus Aurelius}

\vskip24pt

\centerline{\bf I}
\medskip

\quo 1. From my grandfather I learned the multiples of five and how they may be used to tell the time.
\quo 2. From my mother, how my fingers can be used to recall the multiples of nine.
\quo 3. From Luc, how to do research: how to formulate problems and seek out their solutions.
From him too I learned that, weather permitting, mathematics should be done and taught outdoors; and that
incorporating exercise and fresh air into one's day often sharpens one's productivity.
\quo 4. With my friends Rosie, Anna, Jad, and Jonah (among others), that the joy of
discovery is something easily shared. I also learned through their example
the value of hard work, and how diligence
and devotion to one's studies can be translated into erudition and academic success.
\quo 5. From the books and writings of Knuth, not to chase scholarly trends, and how indeed the popularity
of a given subfield is often in direct correlation with the shallowness of the research therein. Knuth also
impressed on me the importance of striking a balance between theory and practice in all my academic endeavours.

\bigskip\vskip\parskip

\centerline{\bf II}
\medskip

\quo 1. Recall how early in your university days you were discouraged by the ease with which
your peers seemed to grasp course concepts, and later discovered this impression to be largely baseless, that
their skills were not drastically better than your own. Take care, even in times of greatest insecurity,
not to overrepresent your
knowledge when speaking to a acquaintance,
for such a fa\c{c}ade will only impede your own learning, and by boasting you are liable to discourage others.

\bigskip
\rightline{\sl Last updated 12.IX.2022, Edmonton, Alberta.}

\bye

### 2022-10-20 misc/expo/uncertainty.tex
Writing section

\input fontmac
\input mathmac

\def\FF{{\bf F}}
\def\TT{{\bf T}}
\def\bar{\overline}
\def\hat{\widehat}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\supp{\op{{\rm supp}}}
\def\smallsupp{\op{{\sevenrm supp}}}
\def\divides{\backslash}

\widemargins
\bookheader{THE DISCRETE FOURIER UNCERTAINTY PRINCIPLE}{MARCEL K. GOH}

\maketitle{The discrete Fourier uncertainty principle}{by}{Marcel K. Goh}{6 October 2022}

\bigskip

\advsect Introduction

Let $Z$ be a finite abelian group. A {\it character} on $Z$ is a homomorphism from $Z$ to the multiplicative
group $\CC\setminus \{0\}$. It is easily seen that $|\chi(x)|$ must equal $1$ for all $x\in Z$.
The set of characters forms a group, which we shall call $\hat Z$.
Now if $Z = \ZZ_{n_1}\times \ZZ_{n_2} \times\cdots\times \ZZ_{n_r}$, then for every $u = (u_1,\ldots,u_r)\in Z$
the function $\chi_u : Z\to \CC$ given by
$$\chi_u(x_1,\ldots,x_r) = \prod_{i=1}^r \exp\biggl( {2\pi i u_i x_i\over n_i} \biggr)$$
is a character, and in fact the map $u\mapsto \chi_u$ gives an isomorphism of groups from $Z$ to $\hat Z$.

The space of functions from $Z$ to $\CC$ can be made into an inner product space by setting
$$\langle f,g\rangle = \ex_{x\in Z} f(x)\bar{g(x)},$$
where $\ex_{x\in Z} F(x) = |Z|^{-1} \sum_{x\in Z} F(x)$, and likewise we define an inner product on
the space of functions from $\hat Z$ to $\CC$ by putting
$$\langle \hat f, \hat g\rangle = \sum_{\chi \in \hat Z} \hat f(\chi) \bar{\hat g(\chi)}.$$
For $f:Z\to \CC$, the {\it Fourier transform} of $f$ is the function $\hat f : \hat Z\to \CC$ given by
$$ \hat f(\chi) = \langle f,\chi\rangle
= \ex_{x\in Z} f(x) \bar{\chi(x)}.$$
Of course, we can associate to any $\alpha\in Z$ the character $\chi_\alpha\in \hat Z$, so we may write
$\hat f(\alpha)$ to mean $\hat f(\chi_\alpha)$, and this is called the {\it Fourier coefficient
of $f$ at $\alpha$}.

We have the following important formulas, whose proofs
can be found in any book on Fourier analysis.

\parenproclaim Theorem P (Parseval--Plancherel identity). Let $Z$ be a finite abelian group and
let $f,g : Z\to \CC$. If $\hat f$ and $\hat g$ are the Fourier transforms of $f$ and $g$ respectively,
then $\langle f,g \rangle = \langle \hat f, \hat g\rangle$.\slug

\parenproclaim Theorem I (Fourier inversion formula). Let $Z$ be a finite abelian group and let $f:Z\to \CC$.
Then
$$f(x) = \sum_{\chi\in \hat Z} \hat f(\chi) \chi(x).\noskipslug$$

Recall also the {\it Cauchy--Schwarz inequality}, which wears many disguises but in our context says that
$$ \Bigl( \sum_{x\in Z} |f(x)|\cdot |g(x)| \Bigr)^2
\le \Bigl(\sum_{x\in Z} |f(x)|^2\Bigr)\Bigl( \sum_{x\in Z} |g(x)|^2\Bigr)$$
for all $f,g : Z\to \CC$.

\advsect The uncertainty principle

The {\it support} of a function $f:Z\to \CC$ is the set $\{x\in Z : f(x)\ne 0\}$. We will write
$\norm{f}_0$ for the size $|\supp(f)|$ of the support, and it is also convenient to write
$\norm{f}_\infty$ for the quantity $\max_{x\in Z} |f(x)|$. (These are defined analogously for
functions on $\hat Z$.) The uncertainty principle states that
the support of $f:Z\to \CC$ and the support of its Fourier transform $\hat f:\hat Z\to \CC$ cannot both
be small. We will make this fact quantitative very soon. First off, let us prove a lemma.

\proclaim Lemma \advthm. Let $f$ be a function from an abelian group $Z$ to $\CC$ and let $\hat f$
be its Fourier transform. Then
$$\norm{\hat f}_\infty \le \ex_{x\in Z} |f(x)|.$$

\proof Let $\chi\in \hat Z$ be given. We have, by the definition of Fourier transform
and the triangle inequality,
$$|\hat f(\chi)| = \Bigl| \ex_{x\in Z} f(x) \bar{\chi(x)} \Bigr|
\le \ex_{x\in Z} \bigl| f(x) \bar{\chi(x)} \bigr|,$$
but since $|\chi(x)| = 1$ for all $x$, this is exactly the right-hand side of the lemma statement
and we are done since $\chi$ was arbitrary.\slug

We now state and prove the Fourier uncertainty principle.

\parenproclaim Theorem~{\advthm} (Fourier uncertainty principle).
Let $Z$ be a finite abelian group and $\hat Z$ be its dual. If $f : Z\to \CC$ is not identically zero
and $\hat f: \hat Z\to \CC$ is its Fourier transform, then
$$\norm{f}_0 \cdot \norm{\hat f}_0 \ge |Z|.$$

\proof By the previous lemma and the definition of the support,
$$\norm{\hat f}_\infty \le \ex_{x\in Z} |f(x)| = {1\over |Z|} \sum_{x\in Z} |f(x)|
= {1\over |Z|} \sum_{x\in \smallsupp(f)} |f(x)|.$$
We then use the Cauchy--Schwarz inequality to obtain
$$\sum_{x\in \smallsupp(f)} |f(x)| \le \sqrt{\sum_{x\in \smallsupp(f)} |f(x)|}\sqrt{\sum_{x\in \smallsupp(f)} 1^2}
= \sqrt{\norm{f}_0 \sum_{x\in Z} |f(x)|^2},$$
and so far we have shown that
$$\norm{\hat f}_\infty \le {1\over |Z|} \sqrt{\norm{f}_0 \sum_{x\in Z} |f(x)|^2}.$$
But by the Parseval--Plancherel identity, we have
$$\sum_{x\in Z} |f(x)|^2 = |Z| \sum_{\chi\in \hat Z} |\hat f(\chi)|^2
\le |Z|\cdot \norm{\hat f}_0 \cdot \norm{\hat f}_\infty^2, $$
and plugging this in above, we have
$$\norm{\hat f}_\infty \le \norm{\hat f}_\infty \sqrt{ \norm{f}_0\cdot \norm{\hat f}_0 \over |Z|}.$$
Since $f$ is not the zero function, we can divide both sides by $\norm{\hat f}_\infty$, square the
inequality, then rearrange to get the theorem statement.\slug

It can be shown that we have equality above if and only if $f$ is (some multiple of) the characteristic
function of a coset of a subgroup of $Z$.

So far so good, but for $Z = \ZZ_p$ a much stronger uncertainty principle holds, and the rest of these notes
will be dedicated to establishing the algebraic machinery needed to prove it.

\advsect Cyclotomic polynomials

Let $n$ be a positve integer. An {\it $n$th root of unity} is any complex number $\omega$ such
that $\omega^n = 1$. Note that if $d$ divides $n$, then any $\omega$ with $\omega^d = 1$
also satisfies $\omega^n = 1$, so in some sense this number should be associated to $d$ and not
$n$. An $n$th root
of unity is called {\it primitive} if it is not an $m$th root of unity for any $1\le m<n$. (Thus
any $n$th root of unity is a primitive $d$th root of unity for exactly one $d$ dividing $n$.) The {\it $n$th
cyclotomic polynomial}, which we shall denote by $\Phi_n$, is given by
$$\Phi_n(z) = \prod_{\omega} (z-\omega),$$
where in the product, $\omega$ runs over the primitive $n$th roots of unity.
As some small examples, we have $\Phi_1(z) = z-1$, $\Phi_2(z) = z+1$, $\Phi_3(z) = z^2 + z + 1$,
and $\Phi_4(z) = z^2 + 1$. Observe that so far, all the coefficients have been polynomial, a fact which
is not obvious from the definition but can be shown by induction (and indeed we shall).

In the proof of the next lemma we will also require the {\it von Mangoldt function} $\Lambda(n)$,
which is defined on positive integers by the rule
$$\Lambda(n) = \cases{ \log p, & if $n=p^k$ for some prime $p$ and some integer $k\ge 1$;\cr 0, & otherwise.}$$
By the fundamental theorem of arithmetic, any integer $n$ can be factored into
$n = {p_1}^{e_1} {p_2}^{e_2} \cdots {p_s}^{e_s}$, and taking logarithms of both sides we see that
$$ \log n = \sum_{i=1}^s e_i \log p_i = \sum_{d\divides n} \Lambda(d).$$

\proclaim Lemma \advthm. Let $n\ge 1$. The $n$th cyclotomic polynomial $\Phi_n$ is monic
with integer coefficients and
we have
$$\Phi_n(1) = \cases{ 0, & if $n=1$;\cr p,
& if $n=p^k$ for some integer $k\ge 1$;\cr 1, & otherwise.}$$

\proof Let $\Omega_n$ be the set of all $n$th roots of unity, primitive or not. Then the polynomial
$z^n-1$ factors as
$$z^n-1 = \prod_{\omega\in \Omega_n} (z-\omega).$$
Now since every $n$th root of unity is a primitive
$d$th root of unity for exactly one $d$ dividing $n$, we can group
roots together and write
$$z^n - 1 = \prod_{d\divides n,} \Phi_d(z).$$

Let us prove the formula for $\Phi_n(1)$ first. Of course, $\Phi_1(1) = 1-1 = 0$. Then for $n>1$,
$${z^n-1\over \Phi_1(z)} = \lim_{z\to 1} {z^n-1 \over z-1} = \lim_{z\to 1} {nz^{n-1} \over 1} = n,$$
giving us the formula
$$ n = \prod_{d\divides n,\,d>1} \Phi_d(1) .$$
Taking logarithms of both sides, we have
$$\log n = \sum_{d\divides n,\,d>1} \log \Phi_d(1),$$
and by the formula above for the von Mangoldt function $\Lambda$, as well as the fact that $\Lambda(1) = 0$,
we have
$$\sum_{d\divides n,\,d>1} \Lambda(d) = \sum_{d\divides n,\,d>1} \log \Phi_d(1).$$
The claim is that these two sums are actually equal term-by-term. When $n$ is prime, the statement above
already shows that $\log \Phi_p(1) = \Lambda(p) = \log p$,
and supposing the claim proven for all $m<n$, we cancel all smaller terms in the formula
to conclude that $\Lambda(n) = \log\Phi_n(1)$, which is what we needed to show.

Now we prove that $\Phi_n$ has integer coefficients. Again, the proof starts with the decomposition
of $z^n-1$ into linear factors, which this time we write as
$$z^n - 1 = \Phi_n(z) \prod_{d\divides n,\, d<n} \Phi_d(z).$$
With the base case $\Phi_1(z) = z-1$, strong induction would prove the claim if we can show that in a factorisation
$$z^n-1 = (a_0 + a_1z + \cdots + a_rz^r) (b_0 + b_1z + \cdots + b_sz^s),$$
the hypotheses $b_s = 1$ and $b_j$ being integer for all $1\le j< s$ implies that the coefficients $a_i$
are all integer for $1\le i\le r$ and that this polynomial is monic as well. The fact that $a_r = 1$ is
obvious. Then since $b_0$ is an integer and $a_0b_0 = -1$, both $a_0$ and $b_0$
must be $\pm 1$. Now assume that for some $t\ge 0$, $a_i$ is integral for all $1\le i\le t$, and consider
the coefficient of $z^{t+1}$ of the left-hand side. Call this coefficient $c_{t+1}$ and note that it is
an integer (in fact, it is either $0$ or $1$, but that is unimportant). We expand
$$c_{t+1} = a_{t+1} b_{0} + a_t b_1 + \cdots + a_0 b_{t+1},$$
and rearrange to obtain
$$a_{t+1} = {c_{t+1} - a_t b_1 - a_{t-1} b_2 - \cdots - a_0 b_{t+1}\over b_0},$$
from which we conclude by induction on $t$ that
$$a_{t+1} = \pm (c_{t+1} - a_t b_1 - a_{t-1} b_2 - \cdots - a_0 b_{t+1})$$
is an integer. This also completes the induction on $n$, so we have shown that $\Phi_n$ is a monic polynomial
with integer coefficients for all $n$.

\advsect Irreducibility of cyclotomic polynomials

A polynomial $p(z)$ with integer coefficients is said to be {\it irreducible over $\ZZ$} if it cannot be expressed
as a product of two nonconstant polynomials in $\ZZ[z]$. This section will be devoted to proving that
the cyclotomic polynomials $\Phi_n$ are irreducible over $\ZZ$.

\proclaim Theorem \advthm. The $n$th cyclotomic polynomial is irreducible over $\ZZ$.

\proof Suppose, towards a contradiction, that $\Phi_n = fg$ for nonconstant $f$ and $g$ in $\ZZ[z]$. Then
we can partition the primitive $n$ roots of unity into two disjoint nonempty classes $A$ and $B$ such that
$$f(z) = \prod_{\omega\in A} (z-\omega)\qquad\hbox{and}\qquad g(z) = \prod_{\omega\in B} (z-\omega).$$
Since any two primitive roots are powers of one another, there exists $\omega\in A$ and an integer $m>1$ such
that $\omega^m \in B$. Factor $m$ into primes $m = p_1p_1\cdots p_k$. Let $\omega_0 = \omega$ and
for $1\le i\le k$ let $\omega_i = \omega^{p_1p_2\cdots p_i}$. Let $j$ be the
smallest integer such that $\omega_j\in B$. (Since $\omega_0\in A$ and $\omega_k = \omega^m\in B$, such
a $j$ must exist.) Now formally replacing $\omega$ by $\omega^{p_1\cdots p_{j-1}}$ and setting $p=p_j$,
we have found some $\omega\in A$ and some prime $p$ such that $\omega^p\in B$.

This means that $\omega$ is a root of both $f(z)$ and $g(z^p)$. Let $h(z)$ be the greatest common divisor
of $f(z)$ and $g(z^p)$. By the Euclidean algorithm there exist polynomials $r(z)$ and $s(z)$
such that
$$ h(z) = f(z)r(z) + g(z^p)s(z),$$
showing that $h(z)$ has $\omega$ as a root and, in particular, is not constant.
Now we work modulo $p$.
By Fermat's little theorem, $a^p = a$ for all $a\in \FF_p$, so we have
$h(z^p) = h(z) = h(z)^p$ and $z^{np}-1 = z^n-1 = (z^n-1)^p$.
Now since $\Phi_n(z^p) = f(z^p)g(z^p) = f(z)^pg(z^p)$, we find that in $\FF_p$,
the polynomial $h(z)^{p+1}$ divides $\Phi_n(z^p)$, and because $\Phi_n(z^p)$ divides $z^{np}-1 = (z^n -1)^p$,
we see that $h(z)^{p+1}$ divides $(z^n-1)^p$ as well. This means that $h(z)^2$ divides $z^n-1$.
Putting $p(z) = z^n-1$, this means that there is some polynomial $q$ such that $p = h^2q$. Then we
find that $nz^{n-1} = p' = 2h h'q + h^2 q'$ is divisible by $h$, and thus $z^n-1$ and $nz^{n-1}$
have a nonconstant common factor.

On the other hand, letting $n^{-1}$ be the multiplicative inverse of $n$ in $\FF_p$, we can
run the Euclidean algorithm on $z^n-1$ and $nz^{n-1}$:
$$\eqalign{
z^n - 1 &= (n^{-1} z) (nz^{n-1}) + (-1) \cr
nz^{n-1} &= (-1)(-nz^{n-1}) + 0,\cr
}$$
discovering that the greatest common divisor of these two polynomials is $1$.
This contradiction shows that $\Phi_n(z)$ is irreducible over $\ZZ$.\slug

\section References

\bye

### 2022-11-07 misc/expo/additive.tex
Notes changed

@@
 \langle \hat f, \hat g\rangle &= \sum_{\chi\in \hat Z} \hat f(\chi)\bar{\hat g(\chi)} \cr
-&=  \sum_{\chi\in \hat Z} \ex_{x\in Z} f(x)\bar{\chi(x)} \ex_{y\in Z} g(y)\bar{\chi(y)} \cr
-&=   \ex_{x\in Z} f(x) \ex_{y\in Z} g(y) \sum_{\chi\in \hat Z}\bar{\chi(x)}\bar{\chi(y)}.\cr
+&=  \sum_{\chi\in \hat Z} \ex_{x\in Z} f(x)\bar{\chi(x)} \bar{\ex_{y\in Z} g(y)\bar{\chi(y)}} \cr
+&=   \ex_{x\in Z} f(x) \bar{\ex_{y\in Z} \bar{g(y)} \sum_{\chi\in \hat Z}\chi(x)\bar{\chi(y)}}.\cr
 }$$
@@
 $y = x$ and $0$ otherwise. Thus for any given $x$, we have
-$$\ex_{y\in Z} g(y) \sum_{\chi\in \hat Z}\chi(x)\bar{\chi(y)} = g(x),$$
+$$\ex_{y\in Z} \bar{g(y)} \sum_{\chi\in \hat Z}\chi(x)\bar{\chi(y)} = \bar{g(x)},$$
 which is exactly what we need for the whole thing to equal $\langle f,g\rangle$.\slug

### 2022-11-07 misc/expo/uncertainty.tex
Notes changed

@@
 \def\supp{\op{{\rm supp}}}
+\def\sgn{\op{{\rm sgn}}}
 \def\smallsupp{\op{{\sevenrm supp}}}
 \def\divides{\backslash}
+\def\del{\partial}
 
@@
 group $\CC\setminus \{0\}$. It is easily seen that $|\chi(x)|$ must equal $1$ for all $x\in Z$.
-The set of characters forms a group, which we shall call $\hat Z$.
-Now if $Z = \ZZ_{n_1}\times \ZZ_{n_2} \times\cdots\times \ZZ_{n_r}$, then for every $u = (u_1,\ldots,u_r)\in Z$
+The set of characters forms a group, which we shall denote by $\hat Z$. This is the Pontryagin dual of $Z$.
+Letting $\ZZ_n$ be the $n$-element cyclic group,
+if $Z = \ZZ_{n_1}\times \ZZ_{n_2} \times\cdots\times \ZZ_{n_r}$, then for every $u = (u_1,\ldots,u_r)\in Z$
 the function $\chi_u : Z\to \CC$ given by
@@
 of $f$ at $\alpha$}.
+
+It is not difficult to prove that any two distinct characters are orthogonal in the space
+of functions from $Z$ to $\CC$. Furthermore, for any $x\in Z$ we can define a function $F_x : \hat Z\to \CC$
+by $F_x(\chi) = \chi(x)$, and one can similarly show that if $x\ne y$, then $\langle F_x, F_y\rangle = 0$.
+So since both of the vector spaces $\CC^Z$ and $\CC^{\hat Z}$ have dimension $n$, we have found orthogonal
+bases for these spaces, namely $\{\chi_u : u\in Z\}$ and $\{F_x : x\in Z\}$ respectively.
 
@@
 As some small examples, we have $\Phi_1(z) = z-1$, $\Phi_2(z) = z+1$, $\Phi_3(z) = z^2 + z + 1$,
-and $\Phi_4(z) = z^2 + 1$. Observe that so far, all the coefficients have been polynomial, a fact which
+and $\Phi_4(z) = z^2 + 1$. Observe that so far, all the coefficients have been integers, a fact which
 is not obvious from the definition but can be shown by induction (and indeed we shall).
@@
 is an integer. This also completes the induction on $n$, so we have shown that $\Phi_n$ is a monic polynomial
-with integer coefficients for all $n$.
+with integer coefficients for all $n$.\slug
 
@@
 
-A polynomial $p(z)$ with integer coefficients is said to be {\it irreducible over $\ZZ$} if it cannot be expressed
+A polynomial $f(z)$ with integer coefficients is said to be {\it irreducible over $\ZZ$} if it cannot be expressed
 as a product of two nonconstant polynomials in $\ZZ[z]$. This section will be devoted to proving that
-the cyclotomic polynomials $\Phi_n$ are irreducible over $\ZZ$.
+the cyclotomic polynomials $\Phi_n$ are irreducible over $\ZZ$. First we need a lemma in the ring
+of formal polynomials $\FF_p[z]$.
+
+\proclaim Lemma \advthm. Let $f(z)$ be a polynomial with coefficients in $\FF_p$. Then $f(z^p) = f(z)^p$
+in $\FF_p[z]$.
+
+\proof Let $f(z) = a_0 + a_1z + \cdots + a_m z^m$. We have
+$$ f(z)^p = (a_0 + a_1z + \cdots + a_mz^m)^p = \sum_{k_1+\cdots+k_m = p} {p \choose k_1,\ldots, k_m}
+\prod_{j=1}^m (a_jz^j) ^{k_j}$$
+by the multinomial theorem. But unless some $k_i = p$ and all the others are $0$,
+there is a $p$ in the numerator of
+the multinomial coefficient that does not appear in the denominator. Hence in $\FF_p$ we have
+$${p\choose k_1,\ldots,k_m} = \cases{1, & if $k_i = p$ for some $i$;\cr 0, & otherwise.}$$
+Applying Fermat's little theorem, which states that $a^p = a$ in $\FF_p$, we have
+$$f(z)^p = \sum_{i=1}^m (a_iz^i)^p = \sum_{i=1}^m a_i (z^i)^p = f(z^p),$$
+which is what we wanted to show.\slug
 
@@
 smallest integer such that $\omega_j\in B$. (Since $\omega_0\in A$ and $\omega_k = \omega^m\in B$, such
-a $j$ must exist.) Now formally replacing $\omega$ by $\omega^{p_1\cdots p_{j-1}}$ and setting $p=p_j$,
-we have found some $\omega\in A$ and some prime $p$ such that $\omega^p\in B$.
+a $j$ must exist.) Now letting $\omega' = \omega^{p_1\cdots p_{j-1}}$ and setting $p=p_j$,
+we have found some $\omega'\in A$ and some prime $p$ such that $\omega^p\in B$.
 
-This means that $\omega$ is a root of both $f(z)$ and $g(z^p)$. Let $h(z)$ be the greatest common divisor
+This means that $\omega'$ is a root of both $f(z)$ and $g(z^p)$. Let $h(z)$ be the greatest common divisor
 of $f(z)$ and $g(z^p)$. By the Euclidean algorithm there exist polynomials $r(z)$ and $s(z)$
@@
 $$ h(z) = f(z)r(z) + g(z^p)s(z),$$
-showing that $h(z)$ has $\omega$ as a root and, in particular, is not constant.
-Now we work modulo $p$.
-By Fermat's little theorem, $a^p = a$ for all $a\in \FF_p$, so we have
-$h(z^p) = h(z) = h(z)^p$ and $z^{np}-1 = z^n-1 = (z^n-1)^p$.
-Now since $\Phi_n(z^p) = f(z^p)g(z^p) = f(z)^pg(z^p)$, we find that in $\FF_p$,
+showing that $h(z)$ has $\omega'$ as a root and, in particular, is not constant.
+Now we consider everything as polynomials in $\FF_p[z]$, which is a unique factorisation domain.
+Applying the previous lemma twice, we have
+$h(z^p) = h(z)^p$ and $z^{np}-1 = (z^n-1)^p$ in this ring.
+Now since $\Phi_n(z^p) = f(z^p)g(z^p) = f(z)^pg(z^p)$, we find that in $\FF_p[z]$,
 the polynomial $h(z)^{p+1}$ divides $\Phi_n(z^p)$, and because $\Phi_n(z^p)$ divides $z^{np}-1 = (z^n -1)^p$,
@@
 
+\advsect Vandermonde determinants
+
+In our journey towards proving a stronger uncertainty principle over $\FF_p$, we will require special
+polynomials called
+{\it Vandermonde determinants}. These are indexed by $n\ge 1$ and defined by
+$$\Delta_n(z_1,\ldots,z_n) = \prod_{i=1}^n \prod_{j=i+1}^n (z_j-z_i).$$
+The next lemma justifies the name ``determinant''.
+
+\proclaim Lemma \advthm. Let $z_1,\ldots,z_n$ be indeterminates.
+Letting
+$$V = \pmatrix{
+1 & z_1 & {z_1}^2 & \cdots & {z_1}^{n-1} \cr
+1 & z_2 & {z_2}^2 & \cdots & {z_2}^{n-1} \cr
+\vdots & \vdots & \vdots & \ddots & \vdots \cr
+1 & z_n & {z_n}^2 & \cdots & {z_n}^{n-1} \cr
+},$$
+we have $\Delta_n(z_1,\ldots,z_n) = \det V$.
+
+\proof By the Leibniz formula for determinants, we have
+$$\det V = \sum_{\pi\in {\frak S}_n} \sgn(\pi) \prod_{i=1}^n {z_i}^{\pi(i) - 1},$$
+where ${\frak S}_n$ is the symmetric group of all permutations on $n$ letters. (The factor $\sgn(\pi)$ is
+$1$ if the permutation $\pi$ factors as a product of an even number of transpositions,
+and $-1$ if it factors as an odd number of transpositions.) Since the $i$th column of $V$
+contains only monomials of degree $i-1$, every term of $\det V$ is a monomial in which the sum
+of the degrees over all coefficients is $0 + 1 + \cdots n-1 = n(n-1)/2$.
+
+In the formula
+$$\Delta_n(z_1,\ldots,z_n) = \prod_{i=1}^n \prod_{j=i+1}^n (z_j-z_i),$$
+note that since the product runs over ${n\choose 2} = n(n-1)/2$ linear factors, every term in this sum
+is a monomial of total degree $n(n-1)/2$ as well. Because $\det V$ is equal to zero if any two
+of the $z_i$ are equal, $\det V$ is divisible by
+the linear polynomial $z_j-z_i$ for all $i < j$. Repeating this for all such $i$ and $j$, we
+conclude that $\Delta_n(z_1,\ldots,z_n)$ divides $\det V$; that is,
+$\det V = \Delta_n f$ for some polynomial $f$. But since both of them consist purely of
+monomials of total degree $n$, $f$ must be a constant polynomial. To find out what this
+constant factor is, note that the term corresponding to the identity permutation in the Leibniz formula
+is $z_2{z_3}^2{z_4}^3\cdots {z_n}^{n-1}$, and expanding
+$$\Delta_n(z_1,\ldots,z_n) = (z_2-z_1)(z_3-z_2)(z_3-z_1)(z_4-z_3)\cdots(z_n-z_{1}),$$
+a moment's scrutiny reveals that this term is also $z_2{z_3}^2{z_4}^3\cdots {z_n}^{n-1}$, and hence
+$f$ must equal $1$, which is what we needed.\slug
+
+\newcount\vandermonde
+\vandermonde=\thmcount
+\proclaim Lemma \advthm. Let $n_1,\ldots, n_k$ be positive integers and let $P\in \ZZ[z_1,\ldots,z_k]$
+be the polynomial given by
+$$ P(z_1,\ldots,z_n) = \sum_{\pi\in {\frak S}_k} \sgn(\pi) \prod_{i=1}^k {z_i}^{n_{\pi(i)}}.$$
+Then we may factor $P = \Delta_k Q$, where $Q\in \ZZ[z_1,\ldots,z_k]$ is such that
+$$Q(1,1,\ldots,1) = \Delta_k(n_1, \ldots, n_k) / \Delta_k(1,\ldots,k).$$
+
+\proof By the Leibniz formula, $P(z_1,\ldots,z_n)$ is the determinant of
+the $k\times k$ matrix whose entry in the $i$th row and $j$th column is $z_j^{n_i}$. As in the previous
+proof, $P$ is divisible by $z_j-z_i$ for all $i<j$ and dividing out these linear factors, we obtain
+a polynomial $Q$ such that $P = \Delta_k Q$.
+
+It remains to compute $Q(1,1,\ldots,1)$. To do so, we make use of the normalised differentiation operators
+$D_i = z_i(\del /\del z_i)$.
+It is easy to see that these operators obey the product rule $D_i(fg) = fD_ig+ D_if g$. Since
+$$D_i({z_1}^{n_1} \cdots {z_k}^{n_k}) = n_i ({z_1}^{n_1} \cdots {z_k}^{n_k}),$$
+this monomial is an eigenfunction of $D_i$ with eigenvalue $n_i$. Now consider the polynomial
+$$(D_2 {D_3}^2 {D_4}^3 \cdots {D_k}^{k-1}) P = 
+(D_2 {D_3}^2 {D_4}^3 \cdots {D_k}^{k-1}) \Bigl(Q \cdot \prod_{i<j} (z_j-z_i)\Bigr),$$
+evaluated at $1$. There are $k(k-1)/2$ differentiation operators to be applied and the same
+number of linear factors on the right-hand side. Repeatedly applying the product rule, we obtain
+$2^{k(k-1)/2}$ terms, but
+the only terms that survive when evaluated at $(1,\ldots,1)$ are the ones in which each
+linear factor $z_j-z_i$ is acted upon by either $D_j$ or $D_i$, yielding $z_j$ or $-z_i$ respectively.
+
+Note that there are $k-1$ instances of the operator $D_k$ to be applied, and there are only
+$k-1$ factors with the variable $z_k$ appearing, namely the
+factors of the form $z_k-z_i$ for some $i<k$. So all those operators must hit those factors (yielding
+$z_k$), and there are $(k-1)!$ ways for this to happen. With those out of the way, there are now
+$k-2$ instances of $D_{k-1}$ to be applied, and the only undifferentiated factors with the variable $z_{k-1}$
+appearing are the factors of the form $z_{k-1}-z_i$, of which there are $k-2$. So there are $(k-2)!$
+ways for this to happen. Continuing in this manner, we see that
+$$(D_2 {D_3}^2 {D_4}^3 \cdots {D_k}^{k-1} P) (1,\ldots,1) = 0!1!2!\cdots(k-1)! Q(1,\ldots,1).$$
+But note that
+$$\Delta_k(1,\ldots,k) = \prod_{i=1}^k \prod_{j=i+1}^k (j-i) = \prod_{i=1}^k (i-1)!
+= 0!1!2!\cdots(k-1)!,$$
+so
+$$(D_2 {D_3}^2 {D_4}^3 \cdots {D_k}^{k-1} P) (1,\ldots,1) = \Delta_k(1,\ldots,k) Q(1,\ldots,1).$$
+
+But from the definition of $P$ and the observation that $z_i^{n_{\pi(i)}}$ is an eigenfunction
+of the operator $D_i$ with eigenvalue $n_{\pi(i)}$, we directly compute
+$$(D_2 {D_3}^2 {D_4}^3 \cdots {D_k}^{k-1} P) (z_1,\ldots,z_k) =
+\sum_{\pi\in {\frak S}_n} \sgn(\pi) \prod_{i=1}^k {n_{\pi(i)}}^{i-1} z_i^{n_{\pi(i)}}.$$
+Evaluating at $z_1=\cdots=z_k = 1$ and noting that the sum over all $\pi$ in ${\frak S}_n$ also
+runs over all $\pi^{-1}$ in ${\frak S}_n$ ($\pi$ and $\pi^{-1}$ have the same sign), we see that
+$$(D_2 {D_3}^2 {D_4}^3 \cdots {D_k}^{k-1} P) (1,\ldots,1) = \Delta_k(n_1,\ldots,n_k).$$
+Combining this with our earlier computation gives the conclusion
+$$Q(1,\ldots,1) = \Delta_k(n_1,\ldots,n_k) / \Delta_k(1,\ldots,k) ,$$
+which is what we wanted to prove.\slug
+
+\advsect Chebotar\"ev's lemma
+
+Armed with the irreducibility of cyclotomic polynomials and the computation from the last section,
+we can now prove a useful lemma concerning matrices of $q$th roots of unity, where $q$ is a prime
+power. First we prove
+a criterion for the nonvanishing of polynomials on $q$th roots of unity.
+
+\proclaim Lemma \advthm. Let $p$ be a prime and $q$ an integer power of $p$. Let $Q\in \ZZ[z_1,\ldots,z_k]$
+be such that $Q(1,\ldots,1)$ is not divisible by $p$. Then for any $k$-tuple $(\omega_1,\ldots,\omega_k)$
+of $q$th roots of unity, $Q(\omega_1,\ldots,\omega_k)\ne 0$.
+
+\proof We proceed by contraposition. Suppose that $\omega_1,\ldots,\omega_k$ exist such
+that $Q(\omega_1,\ldots,\omega_k) = 0$. Letting $\omega$ be a primitive root of unity, there
+are integers $n_1,\ldots,n_k$ such that for all $i$, $\omega_i = \omega^{n_i}$. Let $R\in \ZZ[z]$
+be given by $R(z) = Q(z^{n_1},\ldots,z^{n_k})$. Then $R(\omega) = 0$. Thus $R(z)$ has a root in common
+with the cyclotomic polynomial $\Phi_q(z)$. But we showed earlier that
+this polynomial is irreducible, implying
+that $\Phi_q(z)$ divides $R(z)$. So $Q(1,\ldots,1) = R(1)$ is divisible by $\Phi_q(1) = p$.\slug
+
+We are now able to prove the following useful result, which is named after N.~Chebotar\"ev.
+
+\parenproclaim Lemma {\advthm} (Chebotar\"ev, {\rm 1926}). Let $q$ be a prime power, let $1\le k<p$,
+and let $\omega_1,\ldots,\omega_k$ be distinct $q$th roots of unity. Let $n_1,\ldots,n_k$
+be integers that are all distinct modulo $p$. Then the $k\times k$ matrix whose entry in the $i$th
+row and $j$th column is ${\omega_i}^{n_j}$ has nonzero determinant.
+
+\proof Let $P(z_1,\ldots,z_k)$ be the determinant of the matrix $({z_i}^{n_j})_{1\le i,j\le k}$. This
+is the polynomial from Lemma~{\the\vandermonde}, and that lemma says that we can factor
+$P = \Delta_k Q$, where $Q$ is a polynomial with integer coefficients such that
+$$Q(1,\ldots,1) = \Delta_k(n_1,\ldots,n_k)/\Delta_k(1,\ldots,k).$$
+We want to show that $P(\omega_1,\ldots,\omega_k)$ is not zero. Since
+$\omega_i$ are all distinct, $\Delta_k(\omega_1,\ldots,\omega_k)$ is a product of ${k\choose 2}$
+nonzero elements, and in particular is nonzero. So we need only show that $Q(\omega_1,\ldots,\omega_k)\ne 0$
+and by the previous lemma, it suffices to show that $Q(1,\ldots,1)$ is not divisible by $p$.
+But the numerator in the formula for $Q(1,\ldots,1)$, namely $\Delta_k(n_1,\ldots,n_k)$ is a product
+of differences $n_j-n_i$ for all $1\le i<j\le k$, and these differences were all assumed to be nonzero
+modulo $p$. Thus their product is nonzero modulo $p$; in other words, their product is not divisible by $p$.
+This completes the proof.\slug
+
+\advsect Tao's improved uncertainty principle
+
+Chebotar\"ev's lemma is all we need to prove Tao's improved Fourier uncertainty principle for
+functions $f:\ZZ_p\to \CC$. First we state a corollary of that lemma.
+
+\proclaim Corollary \advthm. Let $p$ be a prime and let $A$ and $B$ be subsets of $\ZZ_p$
+with $|A| = |B|$. The linear transformation $\CC^A\to \CC^B$ define by $T f = \hat f|_B$ (that is,
+we restrict the Fourier transform of $f$ to $B$) is invertible. (We write, for instance, $\CC^A$
+to denote functions from $A$ to $\CC$, or in other words, functions $f:\ZZ_p\to \CC$ such that
+$\supp(f)\subseteq A$.)
+
+\proof Write $Z = \ZZ_p$.
+Recall that the sets $\{\chi_u : u\in Z\}$ and $\{F_x : x\in Z\}$, as defined in the introduction, are
+orthogonal bases for $\CC^Z$ and $\CC^{\hat Z}$ respectively. In the first basis, by the Fourier
+inversion formula, the function $f$ is represented by the vector
+$\bigl(\hat f(0), \ldots, \hat f(p-1))\bigr)$, and in the second basis, by the definition of
+Fourier transform, the function $\hat f$ is represented by the vector $p^{-1}\bigl(f(0),\ldots, f(p-1)\bigr)$.
+If we set $\omega = e^{2\pi i/p}$, then the Fourier inversion formula can be expressed as
+$$\eqalign{
+\pmatrix{ f(0) \cr f(1)\cr \vdots \cr f(p-1) }
+&= \pmatrix{ \chi_0(0) & \chi_1(0) & \cdots & \chi_{p-1}(0) \cr
+\chi_0(1) & \chi_1(1) & \cdots & \chi_{p-1}(1) \cr
+\vdots & \vdots & \ddots & \vdots \cr
+\chi_0(p-1) & \chi_1(p-1) & \cdots & \chi_{p-1}(p-1) \cr}
+\pmatrix{ \hat f(0) \cr \hat f(1)\cr \vdots \cr \hat f(p-1) }\cr
+&=
+\pmatrix{(\omega^0)^0 & \omega^0 & \cdots & ({\omega^0})^{p-1} \cr
+(\omega^1)^0 & \omega^1 & \cdots & ({\omega^1})^{p-1} \cr
+\vdots & \vdots & \ddots & \vdots \cr
+(\omega^{p-1})^0 & \omega^{p-1} & \cdots & ({\omega^{p-1}})^{p-1} \cr}
+\pmatrix{ \hat f(0) \cr \hat f(1)\cr \vdots \cr \hat f(p-1) }.
+}$$
+Hence, letting $k= |A| = |B|$, the matrix of $T$ is some $k\times k$ minor of the matrix above.
+But now letting the $z_i$ be the $i$th powers of $\omega$ for $i\in A$ and letting $n_j = j$
+for all $j\in B$, we see that this matrix satisfies the hypotheses of Chebotar\"ev's lemma, and must
+therefore be invertible.\slug
+
+\parenproclaim Theorem {\advthm} (Tao, {\rm 2005}). Let $p$ be a prime number. If $f:\ZZ_p\to \CC$
+is a nonzero function, then
+$$\norm{f}_0 + \norm{\hat f}_0 \ge p+1.$$
+Conversely, if $A$ and $B$ are two nonempty subsets of $\ZZ_p$ such that $|A|+|B|\ge p+1$,
+then there exists a function $f$ such that $\supp(f) = A$ and $\supp(\hat f) = B$.
+
+\proof Suppose, for a contradiction, that $f$ is such that $\norm{f}_0 + \norm{\hat f}_0 \le p$. Letting
+$A = \supp(f)$, we can then find a set $B$ with $|B| = |A|$ that is disjoint from $\supp({\hat f})$.
+Now the Fourier transform of $f$ restricted to $B$ must be zero. But applying the corollary wit
+with $A$ and $B$, we see that $T$ should have nonzero determinant, which gives a contradiction since
+we have just found that $Tf = 0$ for some $f\ne 0$.
+
+Now we prove the converse. First let us handle the case where $|A|+|B| = p+1$. In this situation we
+may choose $A'$ with $|A'| = |A|$ such that $A'\cup B = \ZZ_p$ and
+$|A'\cap B| = 1$; say $A'\cap B = \{x\}$. Now, apply
+the corollary with $A$ and $A'$ to find that $T : \CC^A \to \CC^{A'}$ is invertible. Letting $g\in \CC^{A'}$
+be a function with $g(x)\ne 0$ and $g(y) = 0$ for all $y\in A'\setminus\{x\}$, we can find $f\in \CC^A$
+such that $Tf = g$, that is, the restriction of $\hat f$ to $A'$ equals $g$. Now, since
+$\supp({\hat f}) \subseteq {A'}^c \cup \{x\} = B$, we have $\norm{\hat f}_0 \le |B|$ and then since
+$\supp(f)\subseteq A$, we have $\norm{f}_0 \le |A|$. But in order not to contradict
+what we proved in the previous paragraph, we must have $\supp(f) = A$ and $\supp({\hat f}) = B$.
+
+Now assume that $|A|+|B|>p+1$. Consider the set
+$$S = \{(A', B') : A'\subseteq A, B'\subseteq B, |A|+|B| = p+1\}.$$
+This set is finite, so let us index its elements $(A_1,B_1),\ldots, (A_s, B_s)$. From
+the previous paragraph, there exist functions $f_1,\ldots,f_s$ such that for all $1\le i\le s$,
+$\supp(f_i) = A_i$ and $\supp(\hat{f_i}) = B_i$. Now let
+$$f = \lambda_1 f_1 + \cdots \lambda_s f_s$$
+for some scalars $\lambda_i\in \CC$ to be chosen later. It is clear that $\supp(f) \subseteq A$
+and, since the Fourier transform is linear, we also have
+$\supp(\hat f) \subseteq B$. This is true regardless of our choices for the $\lambda_i$. Now we must prove
+that we can pick the $\lambda_i$ so that $A\subseteq \supp(f)$ and $B\subseteq \supp(\hat f)$. For
+$x\in A$, let
+$$ V_x = \Bigl\{ (\lambda_1,\ldots,\lambda_s) \in \CC^s : \sum_{i=1}^s \lambda_i f_i(x) = 0 \Bigr\}.$$
+This is a subspace of codimension $1$ in $\CC^s$, since we have $s$ degrees of freedom and one nontrivial
+linear constraint. Similarly, for all $x\in B$, let
+$$ W_x = \Bigl\{ (\lambda_1,\ldots,\lambda_s) \in \CC^s : \sum_{i=1}^s \lambda_i \hat{f_i}(x) = 0 \Bigr\}.$$
+Now
+$$\bigcup_{x\in A} V_x \cup \bigcup_{x\in B} W_x$$
+is a finite union of subspaces of codimension $1$. Thus its complement is nonempty and we can choose
+$\lambda_1,\ldots,\lambda_s$ such that the resulting $f$ has $f(x) \ne 0$ for all $x\in A$
+and ${\hat f}(x)\ne 0$ for all $x\in B$. This completes the proof.\slug
+
 \section References

### 2022-11-11 misc/expo/uncertainty.tex
End of uncertainty

@@
 $A = \supp(f)$, we can then find a set $B$ with $|B| = |A|$ that is disjoint from $\supp({\hat f})$.
-Now the Fourier transform of $f$ restricted to $B$ must be zero. But applying the corollary wit
+Now the Fourier transform of $f$ restricted to $B$ must be zero. But applying the corollary
 with $A$ and $B$, we see that $T$ should have nonzero determinant, which gives a contradiction since
@@
 
+\advsect The Cauchy--Davenport theorem
+
+We now use Tao's uncertainty principle for cyclic groups of prime order to prove the Cauchy--Davenport theorem.
+First, we need to define the {\it convolution} of two functions $f,g:Z \to \CC$. This is denoted $f * g$
+and given by
+$$ (f*g)(x) = \ex_{y+z = x} f(y)g(z).$$
+The following easy theorem describes the nice behaviour of convolutions under Fourier transform.
+
+\parenproclaim Theorem C (Convolution law). Let $Z$ be a finite abelian group and $f,g: Z\to \CC$.
+For all $\chi\in \hat Z$, we have $\hat{f * g} (\chi) = \hat f(\chi)\hat g(\chi)$.
+
+\proof We expand
+$$\hat{f*g}(\chi) = \ex_{x\in Z} (f*g)(x)\bar{\chi(x)} = \ex_{x\in Z} \ex_{y+z=x} f(y)g(z)\bar{\chi(y)\chi(z)}.$$
+But note that $x$ does not appear anywhere in the summand, so we can rewrite this as an expectation over
+{\it all} $y$ and $z$ (their sum must equal $x$ for some $x\in Z$). So
+$$\hat{f*g}(\chi) = \ex_{y\in Z}\ex_{z\in Z} f(y)\bar{\chi(y)} g(z)\bar{\chi(z)}
+=\hat f(\chi)\hat g(\chi),$$
+which is what we wanted.\slug
+
+Now let $Z$ be a finite abelian group and let $A$ and $B$ be subsets of $Z$. We define the {\it sumset}
+$A+B$ to be the set of all sums $a+b$ where $a\in A$ and $b\in B$. This is related to the
+convolution operation above, since if $\supp (f) = A$
+and $\supp (g) = B$, then $\supp(f*g)\subseteq A+B$. Also notice that the convolution law gives
+$\supp\bigl(\hat{f*g}\bigr) = \supp(\hat f)\cap \supp(\hat g)$.
+
+\parenproclaim Theorem {\advthm} (Cauchy--Davenport theorem). Let $A$ and $B$ be nonempty subsets of
+$\ZZ_p$. Then we have
+$$|A+B| \ge \min\bigl( |A|+|B| - 1, p\bigr).$$
+
+\proof Since $A$ is nonempty, we can build a nonempty set $X\subseteq \ZZ_p$ with $p+1-|A|$ elements, so that
+$\ZZ_p \setminus X$ has $|A|-1$ elements. Then since $B$ is nonempty, if $|X|+1 > |B|$, then
+we can pick $Y'\subseteq X$ with $|X|+1-|B|$ elements, and let $Y = Y' \cup (\ZZ_p \setminus X)$, so that
+$$|Y| = p+1 - |A| +1 -|B| + |A|-1 = p+1-|B|.$$
+If instead $|X| + 1 \le |B|$, then we let $Y'$ be any singleton subset of $X$ and
+since $p-|X| \ge p+1-|B|$ we can find a subset $Y''$ of
+$\ZZ_p\setminus X$ of size $p-|B|$. Letting $Y = Y' \cup Y''$, we also have $|Y| = p+1-|B|$ in this case,
+and in both cases we have
+$$|X\cap Y| = \max\bigl(|X|+1-|B|, 1\bigr) = \max\bigl( |X|+|Y|-p, 1\bigr).$$
+
+Now since $|A|+|X| = |B|+|Y| = p+1$, by Tao's uncertainty principle we can find $f : \ZZ_p\to \CC$ with
+$\supp(f) = A$ and $\supp(\hat f) = X$, as well as $g:\ZZ_p\to \CC$ with
+$\supp(g) = B$ and $\supp(\hat g) = Y$. As we saw before, the convolution $f*g$ has support contained in $A+B$ and
+the support of its Fourier transform equals $X\cap Y$. By the other direction of Tao's uncertainty principle,
+we must have
+$$|A+B|+|X\cap Y| \ge |\supp(f*g)| + \bigl|\supp\bigl( \hat{f*g}\bigr)\bigr| \ge p+1.$$
+But since $|X|+|Y|-p = p+2-|A|-|B|$,
+$$|A+B| \ge p+1 - \max\bigl( |X|+|Y|-1, 1\bigr) = \min(|A|+|B|-1, p),$$
+exactly what we wanted to show.\slug
+
 \section References

### 2022-11-26 about/vita.tex
Mountain Pass

@@
 \def\repitem{\pubitem R\the\reportcount. \global\advance\reportcount by -1}
+\newcount\fictioncount
+\fictioncount=1
+\def\ficitem{\pubitem F\the\fictioncount. \global\advance\fictioncount by -1}
 
@@
 
+\sectheader{Published Short Fiction}
+
+\begingroup\frenchspacing
+\fictioncount=1
+\pubbegin F
+\ficitem ``Mountain Pass,'' {\sl Prairie Journal of Canadian Literature} {\bf 79} (2023), 48--55.
+\endgroup%end the frenchspacing
+
 \sectheader{Skills}

### 2023-01-13 about/vita.tex
Update vita

@@
 
+\starleftright{\bf After school care supervisor}{\bf January 2023 -- present}
+\leftright{\bf Suzuki Charter School}{\sl Edmonton, AB, Canada}
+\smallskip
+Supervise and facilitate activities (e.g., arts and crafts, dodgeball, drawing/colouring) for children
+in grades 1 through 6.
+\medbreak
+
+\starleftright{\bf Software developer}{\bf January 2023 -- present}
+\leftright{\bf Levven}{\sl Edmonton, AB, Canada}
+\smallskip
+Develop firmware for the manufacture and function of in-home IoT gateways, smart switches,
+and related electronic products. Code primarily in \CEE\ and Python.
+\medbreak
+
 \starleftright{\bf Visiting teaching assistant}{\bf July 2022}
@@
 Led a six-person team consisting of a medic, signaller, sensor, and two-machine gunners within a rifle platoon.
+Most recent reserve training: December 2022.
 \medbreak

### 2023-03-06 about/vita.tex
Small updates
New books and updated resume

@@
 On the peel number and the leaf-height of a Galton--Watson tree.
-To appear in {\sl Combinatorics, Probability and Computing}, 20~pp. [arXiv:2106.14389]
+{\sl Combinatorics, Probability and Computing}\/ {\bf 32},1 (2023), 68--90.
+[arXiv:2106.14389]
 \papitem (with Anna M. Brandenberger, Luc Devroye, and Rosie Y. Zhao)
 Leaf multiplicity in a Bienaym\'e--Galton--Watson tree. {\sl Discrete Mathematics
-and Theoretical Computer Science}\/ {\bf 24},1 (2022), \#7, 16~pp. [arXiv:2105.12046]
+and Theoretical Computer Science}\/ {\bf 24},1 (2022), \#7, 16~pp.
+[arXiv:2105.12046]
 \papitem (with Anna M. Brandenberger and Luc Devroye)
@@
 \begingroup\parindent=10pt
+\thing An uncertainty principle for M\"obius inversion on posets. {\sl arXiv:2302.02466} (2023), 7~pp.
+\smallskip
 \thing (with Jonah Saks)
-Alternating-sum statistics for certain sets of integers. {\sl arXiv:2206.12535} (2022), 12~pp.
+On the homology of several number-theoretic set families. {\sl arXiv:2206.12535} (2022), 16~pp.
 \endgroup
@@
 Develop firmware for the manufacture and function of in-home IoT gateways, smart switches,
-and related electronic products. Code primarily in \CEE\ and Python.
+and related electronic products. Code primarily in \CEE\ and Python. Contributed to the open-source
+Mongoose OS codebase.
 \medbreak

### 2023-03-25 misc/expo/meditations.tex
Meditations

@@
 
-\def\quo#1.{\par\textindent{\it #1.}}
+\newcount\quocount
+
+\def\quo{\par\textindent{\it \the\quocount\global\advance\quocount by 1.}}
 
@@
 
-\quo 1. From my grandfather I learned the multiples of five and how they may be used to tell the time.
-\quo 2. From my mother, how my fingers can be used to recall the multiples of nine.
-\quo 3. From Luc, how to do research: how to formulate problems and seek out their solutions.
+\quocount=1
+\quo From my grandfather I learned the multiples of five and how they may be used to tell the time.
+\quo From my mother, how my fingers can be used to recall the multiples of nine.
+\quo From my uncle, the long division algorithm for integers. When I was older my uncle also introduced me
+to basic concepts of philosophy, and he was an early inspiration for me to view
+scholarly endeavour for its own sake as not only acceptable, but also noble.
+\quo From Luc, how to do research: how to formulate problems and seek out their solutions.
 From him too I learned that, weather permitting, mathematics should be done and taught outdoors; and that
 incorporating exercise and fresh air into one's day often sharpens one's productivity.
-\quo 4. With my friends Rosie, Anna, Jad, and Jonah (among others), that the joy of
+\quo With my friends Rosie, Anna, Jad, and Jonah (among others), that the joy of
 discovery is something easily shared. I also learned through their example
@@
 and devotion to one's studies can be translated into erudition and academic success.
-\quo 5. From the books and writings of Knuth, not to chase scholarly trends, and how indeed the popularity
+\quo From the books and writings of Knuth, not to chase scholarly trends, and how indeed the popularity
 of a given subfield is often in direct correlation with the shallowness of the research therein. Knuth also
@@
 
-\quo 1. Recall how early in your university days you were discouraged by the ease with which
+\quocount=1
+\quo Recall how early in your university days you were discouraged by the ease with which
 your peers seemed to grasp course concepts, and later discovered this impression to be largely baseless, that
@@
 for such a fa\c{c}ade will only impede your own learning, and by boasting you are liable to discourage others.
+\quo Avoid saying that you will do something, be it of personal or professional nature, to a friend or
+family member or a professor or a colleague, etc., and then not following through on it.
+It is always better to do or accomplish the thing and then announce it after the fact. But undoubtedly the best
+practice is to make promises and then fulfill them, to declare clear intentions and act upon them.
 
 \bigskip
-\rightline{\sl Last updated 12.IX.2022, Edmonton, Alberta.}
+\rightline{\sl Last updated 22.III.2023, Edmonton, Alberta.}
 

### 2023-05-02 about/vita.tex
The Vigil

@@
 
-\leftright{\bf Probabilistic analysis of branching processes}{\bf May 2020 -- present}
+\leftright{\bf Probabilistic analysis of branching processes}{\bf May 2020 -- June 2022}
 \leftright{\sl Research group}{\sl McGill University}
 \smallskip
-Ongoing research on branching processes, headed by Luc Devroye. Studied estimation problems
+Research on branching processes, headed by Luc Devroye. Studied estimation problems
 on Galton-Watson trees. Attended informal seminars, discussing
@@
 \smallskip
-Supervise and facilitate activities (e.g., arts and crafts, dodgeball, drawing/colouring) for children
+Supervise and facilitate activities (e.g., arts and crafts, soccer, board games, drawing/colouring) for children
 in grades 1 through 6.
@@
 \begingroup\frenchspacing
-\fictioncount=1
+\fictioncount=2
 \pubbegin F
+\ficitem ``The Vigil,'' {\sl Ricepaper Magazine} (2023), online.
 \ficitem ``Mountain Pass,'' {\sl Prairie Journal of Canadian Literature} {\bf 79} (2023), 48--55.
@@
 \smallskip
-\thing Contributed sequences A335562, A338550, A338993, A339941, A339942, A341822, and A347580
+\thing Contributed sequences A335562, A338550, A338993, A339941, A339942, A341822, A347580,
+A355145--A355147, and A360285
 to the On-line Encyclopedia of Integer Sequences.

### 2023-05-05 misc/expo/additive.tex
Bogolyubov

@@
 \def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
+\def\mone{-\!\one}  % aesthetic quibble
 
@@
 
-\maketitle{Notes on additive combinatorics}{by}{Marcel K. Goh}{23 March 2022}
+\maketitle{Notes on additive combinatorics}{by}{Marcel K. Goh}{5 May 2023}
 
@@
 
+\advsect Bohr sets and Bogolyubov's lemma
+
+Let $G$ be a finite abelian group, let $\chi_1, \ldots, \chi_k$ be characters on $G$, and let
+$\delta>0$. The {\it Bohr set} $B(\chi_1,\ldots,\chi_k;\delta)$ is the set
+$$\bigl\{ x\in G : \chi_i(x)\in e\bigl([-\delta,\delta]\bigr)\ \hbox{for all}\ 1\le i\le k\bigr\}.$$
+(As before, we write $e(\alpha)$ for $e^{2\pi i\alpha}$.) We say that such a Bohr set has {\it dimension} $k$
+and {\it width} $\delta$.
+
+If $G= \FF_p^n$, we can write $\chi_i = \chi_{u_i}$ where $\chi_{u_i}(x) = e(u_i\cdot x/p$. If $\delta<1/p$,
+then $\chi_{u_i}(x)\in e\bigl([-\delta,\delta]\bigr)$ if and only if
+$(u_i\cdot x) / p \bmod 1\in [-\delta,\delta]$. But $u_i\cdot x$ is an integer, so this is true if and only
+if $u_i\cdot x = 0$, so consequently
+$$B(\chi_1,\ldots,\chi_k;\delta) = \{ x\in \FF_p^n : u_i\cdot x = 0\ \hbox{for all}\ 1\le i\le k\},$$
+which is a subspace cut out by $k$ linear conditions, i.e., of codimension at most $k$.
+
+\parenproclaim Lemma {\advthm} (Bogolyubov). Let $G$ be a finite abelian group. Let $A\subseteq G$ be a
+subset of density $\alpha$. Then $2A-2A$ contains a Bohr set of dimension at most $1/\alpha^2$ and width
+$1/4$.
+
+\proof For $x\in G$, let
+$$\eqalign{
+f(x) &= \bigl( \one_A * \one_A * (\mone_A) * (\mone_A)\bigr)(x) \cr
+&= \ex_{p+q-r-s=x} \one_A(p)\one_A(q)\one_A(r)\one_A(s).
+}$$
+Then $x\in 2A-2A$ if and only if $f(x)\ne 0$. We want to find a Bohr set on which $f$ does not vanish.
+
+Note first that for any character $\chi$,
+$$\hat{\mone_A}(\chi) = \ex_x \one_A(-x)\bar{\chi(x)} = \ex_x \one_A(x)\bar{\chi(-x)} = \ex_x\one_A(x)\chi(x)
+=\bar{\hat{\one_A}(\chi)}.$$
+(The last identity relies on the fact that $\one_A(x)$ is real.) This implies that
+$\hat{\one_A}(\hat{\mone_A}) = |\one_A|^2$. Now we apply the Fourier inversion
+formula and the convolution law to expand
+$$\eqalign{
+f(x) &= \sum_\chi \hat{\one_A*\one_A*(\mone_A)*(\mone_A)}(\chi)(\chi)(x) \cr
+&= \sum_\chi \hat{\one_A}(\chi)^2 \hat{\mone_A}(\chi)^2 \chi(x) \cr
+&= \sum_\chi \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x). \cr
+}$$
+Let
+$$K = \bigl\{ \chi\in \hat G : \bigl| \hat{\one_A}(\chi) \bigr| \ge \alpha^{3/2}\bigr\}$$
+and let $B$ be the Bohr set $B(K;1/4)$. Then the expression above for $f(x)$ can be split into
+$$f(x) =
+\bigl| \hat{\one_A}(\chi_0) \bigr|^4
++ \sum_{\chi\in K} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x)
++ \sum_{\chi\notin K\cup \{\chi_0\}} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x).$$
+We shall assume that $x\in B$ and deal with each of the three terms separately.
+
+The first term is easy; since $A$ has density $\alpha$, $\hat{\one_A}(\chi_0) = \alpha$, and thus the first
+time is $\alpha^4$. For the second term, note that if $\chi\in K$ and $x\in B$, then
+$\chi(x)\in e\bigl([-1/4, 1/4]\bigr)$. This means that $\chi(x)$ has argument between $-\pi/2$ and $\pi/2$,
+and {\it a fortiori} has nonnegative real part. Now, if $\chi\notin K$ and $\chi\ne \chi_0$,
+then $\bigl|\hat{\one_A}\bigr| < \alpha^{3/2}$, so we take the absolute value of the third term
+and apply the triangle inequality and Parseval's identity to obtain
+$$\Bigl|\sum_{\chi\notin K\cup \{\chi_0\}} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x)\Bigr|
+\le \sum_{\chi\notin K\cup\{\chi_0\}} \bigl|\hat{\one_A}(\chi)\bigr|^4
+< \sum_{\chi\in \hat G} \alpha^3 \bigl| \hat{\one_A}(\chi)\bigr|^2
+= \alpha^3\langle \hat{\one_A},\hat{\one_A}\rangle.$$
+But by Parseval's identity, the right-hand side equals $\alpha^3 \langle \one_A,\one_A\rangle = \alpha^4$,
+so we find that the real part of the third term is more than $-\alpha^4$. This means that $\Re f(x) > 0$,
+so $f(x)\ne 0$ for all $x\in B$; that is, $B\subseteq 2A-2A$.
+
+It remains to give an upper bound on the size of $K$. We have
+$$\alpha = \sum_{\chi\in \hat G} \bigl| \one_A(\chi)\bigr|^2 \ge
+\sum_{\chi\in K} \bigl|\hat{\one_A(\chi)}\bigr|^2 \ge |K|\alpha^3,$$
+meaning that $|K|\le 1/\alpha^2$.\slug
+
+When $A$ is a subset of density $\alpha$ in $G = \FF_p^n$, Bogolyubov's lemma tells us that $2A-2A$ contains
+a subspace of codimension at most $1/\alpha^2$.
+
 \section References

### 2023-05-13 misc/expo/additive.tex
Roth's thm lemma 1

@@
 
-If $\phi : A\to B$ is a $k$-homorphism with an inverse that is also a $k$-homorphism, then we say that
+If $\phi : A\to B$ is a $k$-homomorphism with an inverse that is also a $k$-homomorphism, then we say that
 $\phi$ is a {\it Freiman isomorphism of order $k$} or, more briefly, a {\it $k$-isomorphism}. These
@@
 
-Let $G$ be a finite abelian group, let $\chi_1, \ldots, \chi_k$ be characters on $G$, and let
+Let $Z$ be a finite abelian group, let $\chi_1, \ldots, \chi_k$ be characters on $Z$, and let
 $\delta>0$. The {\it Bohr set} $B(\chi_1,\ldots,\chi_k;\delta)$ is the set
-$$\bigl\{ x\in G : \chi_i(x)\in e\bigl([-\delta,\delta]\bigr)\ \hbox{for all}\ 1\le i\le k\bigr\}.$$
+$$\bigl\{ x\in Z : \chi_i(x)\in e\bigl([-\delta,\delta]\bigr)\ \hbox{for all}\ 1\le i\le k\bigr\}.$$
 (As before, we write $e(\alpha)$ for $e^{2\pi i\alpha}$.) We say that such a Bohr set has {\it dimension} $k$
@@
 
-If $G= \FF_p^n$, we can write $\chi_i = \chi_{u_i}$ where $\chi_{u_i}(x) = e(u_i\cdot x/p$. If $\delta<1/p$,
+If $Z= \FF_p^n$, we can write $\chi_i = \chi_{u_i}$ where $\chi_{u_i}(x) = e(u_i\cdot x/p)$. If $\delta<1/p$,
 then $\chi_{u_i}(x)\in e\bigl([-\delta,\delta]\bigr)$ if and only if
@@
 
-\parenproclaim Lemma {\advthm} (Bogolyubov). Let $G$ be a finite abelian group. Let $A\subseteq G$ be a
+\parenproclaim Lemma {\advthm} (Bogolyubov). Let $Z$ be a finite abelian group. Let $A\subseteq Z$ be a
 subset of density $\alpha$. Then $2A-2A$ contains a Bohr set of dimension at most $1/\alpha^2$ and width
@@
 
-\proof For $x\in G$, let
+\proof For $x\in Z$, let
 $$\eqalign{
@@
 Let
-$$K = \bigl\{ \chi\in \hat G : \bigl| \hat{\one_A}(\chi) \bigr| \ge \alpha^{3/2}\bigr\}$$
+$$K = \bigl\{ \chi\in \hat Z : \bigl| \hat{\one_A}(\chi) \bigr| \ge \alpha^{3/2}\bigr\}$$
 and let $B$ be the Bohr set $B(K;1/4)$. Then the expression above for $f(x)$ can be split into
@@
 \le \sum_{\chi\notin K\cup\{\chi_0\}} \bigl|\hat{\one_A}(\chi)\bigr|^4
-< \sum_{\chi\in \hat G} \alpha^3 \bigl| \hat{\one_A}(\chi)\bigr|^2
+< \sum_{\chi\in \hat Z} \alpha^3 \bigl| \hat{\one_A}(\chi)\bigr|^2
 = \alpha^3\langle \hat{\one_A},\hat{\one_A}\rangle.$$
@@
 It remains to give an upper bound on the size of $K$. We have
-$$\alpha = \sum_{\chi\in \hat G} \bigl| \one_A(\chi)\bigr|^2 \ge
+$$\alpha = \sum_{\chi\in \hat Z} \bigl| \one_A(\chi)\bigr|^2 \ge
 \sum_{\chi\in K} \bigl|\hat{\one_A(\chi)}\bigr|^2 \ge |K|\alpha^3,$$
@@
 
-When $A$ is a subset of density $\alpha$ in $G = \FF_p^n$, Bogolyubov's lemma tells us that $2A-2A$ contains
+When $A$ is a subset of density $\alpha$ in $Z = \FF_p^n$, Bogolyubov's lemma tells us that $2A-2A$ contains
 a subspace of codimension at most $1/\alpha^2$.
+
+\advsect Freiman's theorem in vector spaces over fields of prime order
+
+Let $Z$ be an additive group and let $A$ be a finite subset of $Z$. It is possible that $|A+A| = |A|$,
+but this happens if and only if $A$ is a coset of a subgroup of $Z$ (which means that $Z$ must be finite).
+If $A$ is a finite subset of integers, then without loss of generality we can shift $A$ so that $\min A = 0$.
+if $A$ only contains $0$ then $|A+A|$ is $1 = 2|A|-1$, and if $|A| > 1$, then let $m = \max A$ and
+note that $A+A$ contains $A$ and $m+A$, and $A\cap (m+A) = \{m\}$, so in this case we also have $|A+A|\ge 2|A|-1$.
+
+It is not hard to see that if $A$ is an arithmetic progression
+in the integers, then $A+A$ is exactly $2|A|-1$,
+but for instance $A = \{1,\ldots,n\}\subseteq \ZZ$ is not contained in any nontrivial subgroup of $\ZZ$ and
+it also has $|A+A| = 2|A|-1$.
+So next we might amend our conjecture to say that if $|A+A|$ is small, then $A$ is a large subset of an arithmetic
+progression. But this is also false. Consider the set
+$$\{0,1,2,\quad 10,11,12,\quad 20,21,22,\quad 30,31,32\}.$$
+In some sense this is a ``progression of progressions,'' or a $2$-dimensional arithmetic progression. Such
+a set is the projection of some rectangle in $\ZZ^2$, and from this observation it is not hard to see
+that $|A+A|\le 4|A|$. If we increase $d$, we can still say that $|A+A|\le 2^d|A|$.
+
+Once we add these $d$-dimensional arithmetic progressions into consideration, it turns out that we can have
+an inverse statement. Concretely, if $|A+A|\le C|A|$, then there exists a $d$-dimensional arithmetic
+progression $P$ such that $|P|\le K|A|$ and $A\subseteq P$, where $d$ and $K$ depend only on $C$ (otherwise
+we could simply embed $A$ into $[\min A, \max A]$).
+
+This statement is called Freiman's theorem, as it was first proved by Freiman over the integers.
+In these notes we will prove an analogue of it for subsets of $\FF_p^N$, in which ``$d$-dimensional arithmetic
+progression'' is replaced by ``subspace''. First we need a lemma.
+
+\parenproclaim Lemma {\advthm} (Ruzsa covering lemma). Let $Z$ be an abelian group and let $A$ and
+$B$ be finite subsets of $Z$. Then there is a set $K\subseteq A$ of size at most $|A+B|/|B|$ such that
+$A\subseteq K+B-B$.
+
+\proof Let $K = \{a_1,\ldots,a_k\}$ be a maximal subset of $A$ with the property that the sets $a_i+B$
+are disjoint. For all $a\in A$ there is $i$ such that $(a+B) \cap (a_i+B) \ne \emptyset$, otherwise we
+could add $a$ to $K$ and contradict maximality. Equivalently, we can find $b$ and $b'$ such that
+$a+b = a_i + b'$ so $a\in a_i + B - B$. Hence $A\subseteq K+B-B$. Lastly, $|K|\cdot |B| = |K+B| \le |A+B|$.\slug
+
+Next, we prove two simple lemmas about Freiman isomorphisms, the second concerning
+how they behave on subspaces of $\FF_p^n$.
+
+\edef\leminducedfreiman{\the\thmcount}
+\proclaim Lemma {\advthm}. A Freiman isomorphism of order $k(r+s)$ from $A$ to $B$ induces a Freiman
+isomorphism of order $k$ from $rA-sA$ to $rB-sB$.
+
+\proof This proof is conceptually very easy but notationally a total nightmare.
+Let $\phi$ be a $k(r+s)$ isomorphism from $A$ to $B$. Then
+$$x_1+\cdots x_{k(r+s)} = y_1+\cdots+y_{k(r+s)}$$
+if and only if
+$$\phi(x_1)+\cdots \phi(x_{k(r+s)}) = \phi(y_1)+\cdots+\phi(y_{k(r+s)})$$
+Now let $g_1,\ldots,g_k, h_1,\ldots,h_k$ be $2k$ elements in $rA-sA$. Then the expression
+$$g_1+\cdots g_k = h_1+\cdots+h_k$$
+has a sum of $kr$ elements of $A$ and $ks$ elements of $-A$ on each side. We can subtract
+off the elements of $-A$ on each side to get $k(r+s)$ elements of $A$ on either side, and the equality
+holds if and only if it holds with $\phi$ applied to every term. So letting
+$$g_i = g'_{i,1}+\cdots+g'_{i,r}-g''_{i,1}-\cdots-g''_{i,s}$$
+and
+$$h_i = h'_{i,1}+\cdots+h'_{i,r}-h''_{i,1}-\cdots-h''_{i,s}$$
+for all $i$, we find that $g_1+\cdots g_k = h_1+\cdots+h_k$ holds if and only if
+$$\eqalign{
+\phi(g'_{1,1})+\cdots+&\phi(g'_{k,r}) + \phi(h''_{1,1})+\cdots+\phi(h''_{k,s}) \cr
+&= \phi(h'_{1,1})+\cdots+\phi(h'_{k,r}) + \phi(g''_{1,1})+\cdots+\phi(g''_{k,s}).
+}$$
+Subtracting all the double-prime terms from both sides yields
+$$\eqalign{
+\phi(g'_{1,1})+\cdots+&\phi(g'_{k,r}) - \phi(g''_{1,1})-\cdots-\phi(g''_{k,s}) \cr
+&= \phi(h'_{1,1})+\cdots+\phi(h'_{k,r})  - \phi(h''_{1,1})-\cdots-\phi(h''_{k,s}),
+}$$
+and letting
+$$\psi(g_i) = \phi(g'_{i,1}) + \cdots + \phi(g'_{i,r}) - \phi(g''_{i,1}) -\cdots-\phi(g''_{i,s})$$
+and
+$$\psi(h_i) = \phi(h'_{i,1}) + \cdots + \phi(h'_{i,r}) - \phi(h''_{i,1}) -\cdots-\phi(h''_{i,s})$$
+for all $i$, we see that $g_1+\cdots g_k = h_1+\cdots+h_k$ holds if and only if
+$$\psi(g_1)+\cdots \psi(g_k) = \psi(h_1)+\cdots+\psi(h_k).$$
+It is invertible, since we can just apply $\phi^{-1}$ termwise to an element of $rB-sB$, so
+this is the $k$-isomorphism we need between $rA-sA$ and $rB-sB$.\slug
+
+\edef\lemaffinefreiman{\the\thmcount}
+\proclaim Lemma {\advthm}. Let $X$ be a linear subspace of $\FF_p^n$ and let $\phi$ be a $2$-isomorphism
+from $X$ to a set $A\subseteq \FF_p^n$. Then $A$ is an affine subspace of $\FF_p^n$ of dimension $\dim X$.
+
+\proof We shall show that $Y = A - \phi(0)$ is a linear subspace of $\FF_p^n$. Clearly it contains $0$,
+since $A$ contains $\phi(0)$. Now let $y_1,y_2\in Y$ so that $y_1 + \phi(0)$ and $y_2+\phi(0)$ are both in $A$.
+Let $x_1 = \phi^{-1}\bigl(y_1+\phi(0)\bigr)$, $x_2 = \phi^{-1}\bigl(y_1+\phi(0)\bigr)$, and let $x$ denote
+their sum in $X$. Then since $x_1 + x_2 = x+0$, we have
+$$y_1 + \phi(0) + y_2 + \phi(0) = \phi(x) + \phi(0).$$
+Subtracting $\phi(0)$ from both sides, we find that $y_1 + y_2 + \phi(0) = \phi(x)$, meaning
+that $y_1 + y_2 \in A - \phi(0) = Y$. This proves that $Y$ is closed under finite sums, and since scalar
+multiplication over a finite field can always be re\"expressed as a finite sum, we find that
+$A$ is an affine subspace of $\FF_p^n$ in bijection with $X$, that is, of dimension $\dim X$.\slug
+
+We are now able to prove Freiman's theorem for $\FF_p^n$, whose proof amounts to little more than
+a concatenation of several previous results.
+
+\parenproclaim Theorem {\advthm} (Freiman's theorem in $\FF_p^n$). Let $A\subseteq \FF_p^n$ and suppose
+that $|A+A|\le C|A|$. Then there is a subspace $X$ of $\FF_p^n$ such that $A\subseteq X$ and $|X|\le C'|A|$,
+where $C'$ depends only on $C$ and $p$.
+
+\proof By Pl\"unnecke's theorem, $|8A-8A| \le C^{16} |A|$, hence by Ruzsa's embedding lemma, $A$
+is $8$-isomorphic to a subset $A_1\subseteq \FF_p^m$ where $p^m \le pC^{16}|A|$. By Bogolyubov's lemma
+with $\alpha = p^{-1}C^{-16}$, $2A_1 - 2A_1$ contains a Bohr set $B(K;1/4)$ of dimension at most
+$p^2C^{32}$. But letting $\delta < 1/(2p) \le 1/4$, and $X = B(K;\delta)$, we have $X\subseteq B(K;1/4)$
+and $X$ is a linear subspace of $\FF_p^m$ of codimension at most $p^2C^{32}$.
+
+Now since $A$ is $8$-isomorphic to $A_1$, we can apply Lemma~{\leminducedfreiman} to deduce that
+$2A-2A$ is $2$-isomorphic to $2A_1 - 2A_1$. It follows that
+$2A-2A$ contains a set $Y$ that is $2$-isomorphic to the linear subspace $X$. By Lemma~{\lemaffinefreiman},
+this set $Y$ is an affine subspace, of dimension at least $m-p^2C^{32}$, and thus of cardinality
+at least $p^{m-p^2C^{32}} \ge p^{p^2C^{32}}|A|$ (here we used the fact that $A_1\subseteq \FF_p^m$ and
+$A_1$ is Freiman isomorphic to $A$).
+
+Ruzsa's covering lemma tells us that there exists $L$ of size at most $|A+Y|/|Y|$ such that $A\subseteq L+Y-Y$.
+But since $Y$ is an affine subspace, $Y-Y$ is a linear subspace; setting $V = Y-Y$, we have
+$$|V| = |Y| \le |2A-2A| \le C^4|A|,$$
+which follows from another application of Pl\"unnecke's theorem.
+Then by possibly adding $|L|$ basis vectors, $L+V$ is contained in a linear subspace of size at most $p^{|L|} |V|
+\le p^{|L|} C^4|A|$.
+On the other hand, $L$ has size at most
+$${|A+2A-2A|\over |V|} \le {C^5|A| \over p^{-p^2C^{32}} |A|} = C^5p^{p^2C^{32}},$$
+where yet again we have used Pl\"unnecke's theorem.
+Hence we conclude that $A$ is contained in a linear subspace of size at most
+$$ p^{C^5 p^{p^2C^{32}}} C^4|A|.\noskipslug$$
+
+\advsect Roth's theorem
+
+In the previous section we showed that any set $A$ with $A+A$ small
+must be a dense subset of a structured set. We didn't prove this over $\ZZ$, but we saw briefly that
+the notion of ``structured set'' in that case was a $d$-dimensional arithmetic progression. Now we switch
+gears and try to find arithmetic progressions in subsets of $\ZZ$. The first major theorem in this vein was
+proved by B.~L.~van der Waerden in 1927.
+
+\parenproclaim Theorem W (van der Waerden, {\rm 1927}). For all $k$ and $r$, there is $n$ such
+that if we partition $[n]$ into disjoint sets $C_1\cup \cdots \cup C_r$ then some $C_i$ contains an
+arithmetic progression of length $k$.\slug
+
+A stronger statement is given by the celebrated 1975 theorem of E.~Szemer\'edi.
+
+\parenproclaim Theorem S (Szemer\'edi, {\rm 1975}). For every $\delta>0$ and $k\in \NN$ there
+exists $n$ such that if $A\subseteq [n]$ and $|A|\ge \delta n$, then $A$ contains an arithmetic progression
+of length $k$.\slug
+
+By setting $\delta = 1/r$ and invoking the pigeonhole principle, we see that
+Szemer\'edi's theorem implies van der Waerden's theorem. We will not prove either of these statements in
+these notes. The first has a relatively elementary proof, but not one that is very additive-combinatorial
+in nature, while the second is far too complex to prove in full generality here. We will, however,
+prove Szemer\'edi's theorem in the case $k=3$. This special case is called Roth's theorem as it was
+proved by K.~Roth in 1953.
+Before we get there we first establish three lemmas. In their proofs we will write, e.g., $A$ instead
+of $\one_A$, and identify
+$r$ with its corresponding element $\chi_r\in \hat{\ZZ_N}$, writing $\hat f(r)$ instead of $\hat f(\chi_r)$.
+
+\proclaim Lemma {\advthm}. Let $N$ be odd and let $A$, $B$, and $C$ be subsets of $\ZZ_N$ with densities
+$\alpha$, $\beta$, and $\gamma$ respectively. Suppose that $N > 2/(\alpha\beta\gamma)$. Then either
+there exists $x$ and $d\ne 0$ such that $(x,x+d,x+2d)\in A\times B\times C$, or there exists $r\ne 0$
+such that $|\hat A(r)| \ge \alpha \beta^{1/2} \gamma^{1/2} / 2$.
+
+\proof Consider the quantity
+$$\eqalign{
+\ex_{x,d} A(x)B(x+d)C(x+2d) &= \ex_{x+z=2y} A(x)C(z)B(y) \cr
+&= \ex_u \bigl( \ex_{x+z=u} A(x)C(z)\bigr) B(u/2).
+}$$
+Using the notation $f_a(x) = f(x/a)$ for a dilation, we see that this expectation is actually the
+inner product $\langle A *C, B_2\rangle$, which by Parseval's identity is equal to
+$\langle \hat{A*C}, \hat{B_2}\rangle$.
+Recall that the characters of $\ZZ_N$ are given by $\chi_a(x) = e(ax/N)$. Applying the dilation rule gives
+$$ \hat{B_2}(r) = \hat{B_2}(\chi_r) = \hat B({\chi_r}^2) = \hat B(\chi_{2r})=\hat B(2r)=(\hat B)_{2^{-1}}(r)$$
+where $2^{-1}$ denotes the inverse of $2$ modulo $N$.
+This, combined with the convolution law, gives
+$\langle \hat{A*C}, \hat{B_2}\rangle = \bigl\langle \hat A\hat C, (\hat B)_{2^{-1}}\bigr\rangle$.
+But since $B$ is real-valued,
+$$\hat B(-r) = \ex_x B(x)\bar{\chi_{-r}(x)} = \ex_x B(x)\chi_r(x) = \bar{\ex_x B(x)\bar{\chi_r(x)}}
+= \bar{\hat B(r)},$$
+so putting everything together yields
+$$\eqalign{
+\ex_{x,d} A(x)B(x+d)C(x+2d) &= \bigl\langle \hat A\hat C, (\hat B)_{2^{-1}}\bigr\rangle \cr
+&= \sum_r \hat A(r)\hat C(r) \bar{\hat B(2r)} \cr
+&= \sum_r \hat A(r)\hat C(r) \hat B(-2r) \cr
+&= \alpha\beta\gamma + \sum_{r\ne 0} \hat A(r) \hat C(r) \hat B(-2r). \cr
+}$$
+Note that by the triangle and Cauchy--Schwarz inequalities,
+$$\eqalign{
+\Bigl|\sum_{r\ne 0} \hat A(r) \hat C(r) \hat B(-2r)\Bigr|
+&\le \max_{r\ne 0} \bigl|\hat A(r)\bigr| \sum_{r\in \ZZ_N} \bigl|\hat B(-2r)\bigr|\cdot\bigl|\hat C(r)\bigr| \cr
+&\le \max_{r\ne 0} \bigl|\hat A(r)\bigr|
+\Bigl(\sum_{r\in \ZZ_N} \bigl|\hat B(-2r)\bigr|^2\Bigr)^{1/2}
+\Bigl(\sum_{r\in \ZZ_N} \bigl|\hat C(r)\bigr|^2\Bigr)^{1/2} \cr
+&= \bignorm{\hat B}_2 \bignorm{\hat C}_2 \max_{r\ne 0} \bigl| \hat A(r)\bigr| \cr
+&= \beta^{1/2} \gamma^{1/2} \max_{r\ne 0} \bigl| \hat A(r)\bigr|.\cr
+}$$
+(We used the fact that $N$ is odd in the third line above.) So
+$$
+\ex_{x,d} A(x)B(x+d)C(x+2d) \ge \alpha\beta\gamma
+-\beta^{1/2} \gamma^{1/2} \max_{r\ne 0} \bigl| \hat A(r)\bigr|,$$
+meaning that if $\max_{r\ge 0} \bigl| \hat A(r)\bigr| < \alpha \beta^{1/2}\gamma^{1/2} / 2$, then
+$\ex_{x,d} A(x)B(x+d)C(x+2d) > \alpha\beta\gamma/2$. The further assumption that $\alpha\beta\gamma/2 > 1/N$
+allows us to conclude that there is some $x$ for which $\ex_{d} A(x)B(x+d)C(x+2d) > 1$, and thus some
+$d\ge 0$ such that $(x,x+d,x+2d)\in A\times B\times C$.\slug
 

### 2023-05-14 misc/expo/additive.tex
Second lemma of Roth and new book

@@
 
+In the next lemma, we define the {\it diameter} of $X\subseteq \CC$ to be the quantity
+$\sup_{z_1,z_2\in X} |z_1-z_2|$.
+
+\proclaim Lemma {\advthm}. For every $\theta\in \RR$, $\eps>0$, and $n\in \NN$ sufficiently large, the
+set $[n]$ can be partitioned
+into disjoint arithmetic progressions $P_1, \ldots, P_m$, each of size at least $\sqrt{\eps n/(32 \pi)}$,
+such that the diameter of $e(\theta P_i)\le \eps$ for all $1\le i\le m$.
+
+\proof Let $t$ be a positive integer to be chosen later. Dividing the unit circle into $t$ segments of
+length $2\pi/t$, by the pigeonhole principle there must be some segment that contains more than one of
+the $t+1$ points $e(0), e(\theta), \ldots, e(t\theta)$. Thus there are $u$ and $v$ with $0\le u<v\le r$ such that
+$\bigl|e(u\theta) - e(v\theta)\bigr|\le 2\pi/t$. Let $d = v-u$. By multiplying both $e(u\theta)$ and
+$e(v\theta)$ by $e(-u\theta)$, we find that $\bigl|1-e(d\theta)\bigr| \le 2\pi/t$ as well. Then by telescoping and
+applying the triangle inequality, we may expand
+$$\bigl|1-e(sd\theta)\bigl| \le \sum_{k=1}^s \bigl|e\bigl((k-1)d\theta-e(kd\theta)\bigr|.$$
+But by applying rotations as above, each of the terms in the sum can is equal to $\bigl|1-e(d\theta)\bigr|$,
+implying that $\bigl|1-e(sd\theta)\bigl| \le 2\pi s/t$.
+
+The reason this is important is that
+whenever $P$ is any arithmetic progression with common difference $d$, any $x,y\in P$ will
+have $\bigl|e(x\theta) - e(y\theta)\bigr| = \bigl|1-e(sd\theta)\bigr|$ for some $s$ that is at most
+the length of $P$. Hence if $P$ is an arithmetic progression with common difference $d$ and length
+at most $\eps t/(2\pi)$, then the diameter of $e(\theta P)$ will be at most $\eps$.
+
+Take $[n]$ and partition it into residue classes modulo $d$. Within each residue class, we would like
+to further group the elements into ``intervals'' of size roughly $\eps t/(2\pi)$. First we need to make
+sure that the residue classes are large enough. If a residue class has size at least $\eps t/(2\pi)$,
+then it can be subdivided into intervals of size between $\eps t/(4\pi)$ and $\eps t/(2\pi)$.
+To ensure this, we note that the residue classes have size $\lfloor n/d\rfloor \ge \lfloor n/t\rfloor$,
+so we shall set $t \le \sqrt{2\pi n / \eps}$. With this choice we find that each $P_i$ has size
+$$\biggl\lfloor {\eps t\over 4\pi}\biggr\rfloor \le  \Biggl\lfloor {\eps\over 4\pi} \sqrt{2\pi n\over \eps}
+\Biggr\rfloor = \Biggl\lfloor \sqrt{\eps n\over 8\pi} \Biggr\rfloor \le \sqrt{\eps n\over 32\pi}.$$
+
 \section References

### 2023-05-15 misc/expo/additive.tex
Finished Roth

@@
 
+\edef\lemABC{\the\thmcount}
 \proclaim Lemma {\advthm}. Let $N$ be odd and let $A$, $B$, and $C$ be subsets of $\ZZ_N$ with densities
@@
 
-\proclaim Lemma {\advthm}. For every $\theta\in \RR$, $\eps>0$, and $n\in \NN$ sufficiently large, the
+\proclaim Lemma {\advthm}. For every $\theta\in \RR$, $n\in \NN$, and $\eps > 0$, the
 set $[n]$ can be partitioned
@@
 $$\bigl|1-e(sd\theta)\bigl| \le \sum_{k=1}^s \bigl|e\bigl((k-1)d\theta-e(kd\theta)\bigr|.$$
-But by applying rotations as above, each of the terms in the sum can is equal to $\bigl|1-e(d\theta)\bigr|$,
+But by applying rotations as above, each term in the sum is equal to $\bigl|1-e(d\theta)\bigr|$,
 implying that $\bigl|1-e(sd\theta)\bigl| \le 2\pi s/t$.
@@
 $$\biggl\lfloor {\eps t\over 4\pi}\biggr\rfloor \le  \Biggl\lfloor {\eps\over 4\pi} \sqrt{2\pi n\over \eps}
-\Biggr\rfloor = \Biggl\lfloor \sqrt{\eps n\over 8\pi} \Biggr\rfloor \le \sqrt{\eps n\over 32\pi}.$$
+\Biggr\rfloor = \Biggl\lfloor \sqrt{\eps n\over 8\pi} \Biggr\rfloor \le \sqrt{\eps n\over 32\pi}.\noskipslug$$
+
+We use this lemma immediately to prove another one, which is key for the density-increment argument
+of Roth's proof.
+
+\edef\lemdensityincrement{\the\thmcount}
+\proclaim Lemma {\advthm}. Let $A\subseteq \ZZ_N$ be a set of density $\delta$, let $\eta>0$ and suppose
+there exists $r$ such that $\bigl| \hat A(r)\bigr| \ge \eta$. Then there is a progression $P$ of length
+at least $\sqrt{\eta N/(64\pi)}$ such that $|A\cap P| \ge (\delta + \eta/4)|P|$.
+
+\proof Define $f(x) = A(x)-\delta$. Then $\sum_{x\in \ZZ_n} f(x) = 0$ and
+$$ \hat f(\chi) = \ex_x (A(x) - \delta)\bar{\chi(x)} = \ex_x A(x)\bar{\chi(x)} - \delta \ex_x \bar{\chi(x)} $$
+If $\chi = \chi_0$ then this whole expression equals $0$; otherwise the second expectation disappears and
+$\hat f(\chi)$ just equals $\hat A(\chi)$. Let $r$ be such that $\bigl| \hat f(r)\bigr| = \bigl| \hat A(r)\bigr|
+\ge \eta$. Then
+$$ \eta N \le N\bigl| \hat f(r)\bigr| = N \Bigl| \sum_{x\in \ZZ_N} f(x) e(-rx/N)\Bigr|.$$
+By the previous lemma with $\theta = -r/N$ and $\eps = \eta/2$,
+we can partition $[N]$ into progressions $P_1,\ldots,P_m$, each of size at least $\sqrt{\eta n / (64\pi)}$
+such that every set $e(-rP_i/N)$ has diameter at most $\eta/2$. Thus by the triangle inequality,
+$$\eta N \le \sum_{i=1}^m \Bigl| \sum_{x\in P_i} f(x) e(-rx/N) \Bigr|.$$
+For each $i$, pick some element $x_i\in P_i$. Then
+$$\eqalign{
+\eta N &\le \sum_{i=1}^m \Bigl| \sum_{x\in P_i} f(x) e(-rx_i/N) \Bigr|
+  +\sum_{i=1}^m \Bigl| \sum_{x\in P_i} f(x) \bigl(e(-rx/N) - e(-rx_i/N) \bigr)\Bigr| \cr
+&\le \sum_{i=1}^m \bigl|e(-rx_i/N)\bigr|\cdot\Bigl| \sum_{x\in P_i} f(x) \Bigr| + {\eta N\over 2},
+}$$
+whence
+$$ \sum_{i=1}^m \Bigl| \sum_{x\in P_i} f(x) \Bigr| \ge {\eta N\over 2} = {\eta\over 2} \sum_{i=1}^m |P_i|.$$
+But $\sum_{i=1}^m \sum_{x\in P_i} f(x) = 0$, so we may add these terms into the sum to obtain
+$$ \sum_{i=1}^m \biggl(\sum_{x\in P_i} f(x) + \Bigl| \sum_{x\in P_i} f(x) \Bigr|\biggr)
+\sum_{i=1}^m {\eta\over 2}|P_i|,$$
+and deduce that there must be some $i$ such that
+$$ \sum_{x\in P_i} f(x) + \Bigl| \sum_{x\in P_i} f(x) \Bigr| \ge {\eta\over 2} |P_i|.$$
+For this $i$ we have
+$$|A\cap P_i| - \delta|P_i| =
+\sum_{x\in P_i} A(x) - \delta|P_i| = \sum_{x\in P_i} f(x) \ge {\eta\over 4}|P_i|,$$
+which implies that $|A\cap P_i| \ge (\delta + \eta/4)|P_i|$.\slug
+
+We are now equipped to prove Roth's theorem.
+
+\parenproclaim Theorem {\advthm} (Roth, {\rm 1953}). For every $\delta>0$ there exists
+$N\le \exp\bigl(\exp(300/\delta)\bigr)$ such that
+every $A\subseteq [N]$ with $|A|\ge \delta N$ contains an arithmetic progression of length $3$.
+
+\proof Let $N$ be an integer to be specified later. Let $A_0$ be a subset of $N$ of density
+$\delta = \delta$If $N$ is even, we can express $N = L_0+M_0$
+where both $L_0$ and $M_0$ are odd integers at least $N/4$. Either
+$$\bigl|A_0 \cap [L_0]\bigr| \ge \delta |L_0|\ \hbox{or}
+\ \bigl|A_0\cap [L_0+1, N]\bigr| \ge \delta M_0$$
+In the first case let $N_0 = L_0$ and set $A = A_0\cap [L_0]$, and in the second case set $N_0=M_0$
+and set $A = \bigl(A_0\cap [L_0+1,N]\bigr) - \{L_0\}$.
+
+Now we reach another crossroads. If $\bigl| A\cap (N_0/3, 2N_0/3]\bigr| < \delta N_0/5$, then one of
+$\bigl| A\cap [1,N_0/3]\bigr|$ or $\bigl| A\cap (2N_0/3,N_0]\bigr|$ must be greater than $2\delta N_0 / 5$.
+Either way, we get a progression $P$ of length roughly $N_0/3$ such that $|A\cap P| \ge \delta_1|P|$,
+where $\delta_1$ is roughly $6\delta/5$. Thus we may iterate the argument with $A = A\cap P$, $N$
+roughly equal to $N_0/3$, and $\delta = \delta_1$. Here we can be very sloppy with bounds and say that
+the new choice of $N$ is at least $N/24$, and the new choice of $\delta$ is at least $11\delta/10$.
+
+It is the case $\bigl| A\cap (N_0/3, 2N_0/3]\bigr| \ge \delta N_0/5$ in which we must use the lemmas
+above. Let $B = C = A\cap [N_0/3, 2N_0/3)$. This choice for $B$ and $C$ ensures that if $d\ne 0$ and
+$(x,x+d,x+2d)\in A\times B\times C$ where $A$, $B$, and $C$ are regarded as subsets of $\ZZ_{N_0}$, then
+$x,x+d,x+2d$ is still an arithmetic progression in $\ZZ$. This is clear because if
+$y,z\in [N_0/3,2N_0/3)\subseteq \ZZ$ and $d$ is the common difference $y-z$, then $|d| < N_0/3$, which
+ensures that $x\in [N_0]$.
+
+So if $A\times B\times C$ contains a triple $(x,x+d,x+2d)$ with $d\ne 0$, then we have an arithmetic
+progression of length $3$ in $[N_0]$ (and thus in $[N]$), so we are done. If not,
+let $\alpha$, $\beta$, and $\gamma$ be the densities of $A$, $B$, and $C$ respectively.
+$\alpha \ge \delta$ and $\beta,\gamma\ge \delta/5$. By Lemma~{\lemABC}, so long as
+$N_0 > 50/\delta^3 \ge 2/(\alpha\beta\gamma)$, there is some $r\ne 0$ such that
+$\hat A(r) \ge \alpha\sqrt{\beta\gamma} \ge \delta^2/10$. By Lemma~{\lemdensityincrement} with
+$\eta = \delta^2/10$, there is a progression $P$ in $A$ of size at least $\delta \sqrt{N_0/(640\pi)}$
+such that $|A\cap P| \ge (\delta + \delta^2/40)|P|$.
+
+By scaling and translating, we obtain a subset of $\bigl[|P|\bigr]$ of size at least $(\delta+\delta^2/40)|P|$,
+so we can repeat the argument with $A$ set to $A = A\cap P$ and $N$ set to $|P|$. It is clear that
+we cannot iterate forever, since $\delta \le 1$, so we will eventually find an arithmetic progression
+of length $3$. It remains to calculate how many iterations it takes for this to happen, which in turn tells
+us an upper bound for $N$.
+
+Well, as long as $\delta \le 4/21$, we have $\delta + \delta^2/40 \le 11\delta/10$, so the
+second case has the smaller density increment. In this case, note that if we iterate $40/\delta$ times,
+then the density at least doubles. Then to double it again we need to iterate $40/(2\delta) = 20/\delta$
+times, and so on. By repeating this argument, we only need to iterate at most
+$$ {40\over\delta} \biggl(1 + {1\over 2} + {1\over 4} + \cdots \biggr) = {80\over \delta}$$
+times and we will eventually have to find an arithmetic progression of length $3$, otherwise $\delta$ would
+become greater than $1$. We can even be extra pedantic here and say that this only gets us to $\delta >4/21$,
+and then $20$ more iterations is sufficient to get to $\delta > 1$. (This is because $\log_{11/10}(21/4) < 20$).
+
+In the argument we required $N_0 > 50/\delta^3$, which means that $N$ must always be greater than
+$200/\delta^3$. To further simplify our calculations, we note that
+$$\delta \sqrt{N_0\over 640\pi } \ge {\delta\sqrt N\over 100} \ge N^{1/3}$$
+as long as $N\ge (100/\delta)^6$. Requiring this condition ensures that in any iteration, the next
+value of $N$ is at least $N^{1/3}$. So to eventually find a $3$-term arithmetic
+progression, we need to start with an $N$ such that $N^{3^{-80/\delta+20}} \ge (100/\delta)^6$.
+Taking the logarithm of both sides, we get
+$$ 3^{-80/\delta+20} \log N \ge 6\log(100/\delta),$$
+and then taking another logarithm yields
+$$ \log\log N \ge \biggl({80\over\delta} + 20\biggr)\log 3 +\log 6 + \log\log(100/\delta),$$
+We'll be lazy here and set $\log\log N\ge 300/\delta$ as
+a sufficient condition, which is the bound claimed in the theorem statement.\slug
 

### 2023-06-15 about/vita.tex
senang diri

@@
 
-\starleftright{\bf After school care supervisor}{\bf January 2023 -- present}
+\starleftright{\bf After school care supervisor}{\bf January -- June 2023}
 \leftright{\bf Suzuki Charter School}{\sl Edmonton, AB, Canada}
 \smallskip
-Supervise and facilitate activities (e.g., arts and crafts, soccer, board games, drawing/colouring) for children
-in grades 1 through 6.
+Supervised and facilitated activities (e.g., arts and crafts, soccer, board games, drawing/colouring, movie
+days) for children in grades 1 through 6.
 \medbreak
 
-\starleftright{\bf Software developer}{\bf January 2023 -- present}
+\starleftright{\bf Software developer}{\bf January -- May 2023}
 \leftright{\bf Levven}{\sl Edmonton, AB, Canada}
 \smallskip
-Develop firmware for the manufacture and function of in-home IoT gateways, smart switches,
+Developed firmware for the manufacture and function of in-home IoT gateways, smart switches,
 and related electronic products. Code primarily in \CEE\ and Python. Contributed to the open-source
@@
 \smallskip
-Meet with two undergraduate students as part of the Directed Reading Program
+Met with two undergraduate students as part of the Directed Reading Program
 on a weekly basis to give them an introduction to research-level
@@
 On call on a weekly basis to provide emergency medical care at campus residences overnight as well as at
-university events such as frosh, sports games, and formals. Attend team training sessions twice a month to keep
+university events such as frosh, sports games, and formals. Attended team training sessions twice a month to keep
 first-aid skills up-to-date. Most recent first responder certification: September 2020.
@@
 \smallskip
-Hold twice-weekly office hours to tutor students in a variety of undergraduate courses.
-Topics covered include elementary data structures and algorithms, command-line scripting,
+Held twice-weekly office hours to tutor students in a variety of undergraduate courses.
+Topics covered included elementary data structures and algorithms, command-line scripting,
 and functional programming. Recipient of the Tomlinson Engagement Award for Mentoring.
@@
 \begingroup\frenchspacing
-\fictioncount=2
+\fictioncount=3
 \pubbegin F
+\ficitem ``Senang Diri,'' {\sl Existere} {\bf 42},2 (2023), 17--25.
 \ficitem ``The Vigil,'' {\sl Ricepaper Magazine} (2023), online.

### 2023-08-15 misc/expo/kelleymeka.tex
Working on Kelley--Meka

\input fontmac
\input mathmac

\def\eps{\epsilon}
\def\FF{{\bf F}}
\def\bar#1{\overline{#1}}
\def\hat#1{\widehat{#1}}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}

\widemargins
\bookheader{KELLEY AND MEKA'S PROOF OF ROTH'S THEOREM}{MARCEL K. GOH}

\maketitle{Kelley and Meka's proof of Roth's theorem}{by}{Marcel K. Goh}{5 August 2023}

\bigskip

\advsect Introduction

Let $A$ be a subset of $\ZZ$. We want to know how dense $A$ can be before it must contain a subset
$\{x,y,z\}\subseteq A$ with $x+z = 2y$, that is, an arithmetic progression of length $3$.

\medskip
\boldlabel Basic definitions and elementary lemmas.
We will use $G$ primarily to refer to a finite abelian group, on which we have the normalised counting measure.
For functions $f,g:G\to \CC$ we have the inner product
$$\langle f,g\rangle = \ex_{x\in G} f(x)\bar{g(x)}$$
and the $L_p$ norm
$$ \norm{f}_p = \Bigl( \ex_{x\in G} \bigl| f(x) \bigr|^p \Bigr)^{1/p}.$$
In $L_p$ spaces we have the useful H\"older inequality
$$ \bigl| \langle f,g\rangle \bigr| \le \norm{f}_p  \cdot \norm{g}_{1-p},$$
for $p,q\in [1,\infty]$ with $1/p + 1/q = 1$.
Assuming now that $f$ and $g$ are $\RR$-valued, we also have the convolution
$$ (f*g)(x) = \ex_{y\in G} f(y) g(x-y)$$
and the difference convolution
$$ (f\circ g)(x) = \ex_{y\in G} f(y) g(x+y)$$
that are related by the following adjoint property.

\parenproclaim Lemma~{\advthm} (Adjoint property).
Let $G$ be a finite abelian group and let $f,g,h:G\to \RR$. Then
$$\langle f,g*h\rangle = \langle h\circ f, g\rangle.$$

\proof First expand
$$\eqalign{
\langle f, g*h \rangle &= \ex_{x\in G} f(x) (g*h)(x) \cr
&= \ex_{x\in G} f(x) \ex_{y\in G} g(y) h(x-y) \cr
&= \ex_{y\in G} g(y) \ex_{x\in G} f(x) h(x-y). \cr
}$$
Then substituting $z = x-y$ so that $x = z+y$ yields
$$\eqalign{
\langle f, g*h\rangle &= \ex_{y\in G} g(y) \ex_{z\in G} f(z+y) h(z) \cr
&= \ex_{z\in G} (h\circ f)(z) g(z) \cr
&= \langle h\circ f, g\rangle .\noskipslug\cr
}$$

For a group $G$ the dual group $\hat G$ is the set of all homomorphisms from $G\to \CC^\times$.
The Fourier transform of $f:G\to \RR$ is the function $\hat f:\hat G\to \CC$ given by
$$ \hat f(\chi) = \ex_{x\in G} f(x)\chi(-x).$$
The following lemma describes how the convolution and difference convolution behave under the Fourier
transform.

\parenproclaim Lemma~{\advthm} (Convolution laws).
Let $G$ be a finite abelian group and let $f,g:G\to \RR$. Then the following identities hold:
\medskip
\item{i)} $\hat{f*g} = \hat f\cdot \hat g$
\smallskip
\item{ii)} $\hat{f\circ g} = \bar{\hat f} \cdot \hat g$
\medskip\noindent
In particular, $\hat {f\circ f} = |\hat f|^2$.

\proof Expand
$$\hat{f*g}(\chi) = \ex_{x\in G} (f*g)(\chi)\chi(-x)$$
and multiply the right-hand side by $1 = \chi(-y)\chi(y)$ to get
$$\hat{f*g}(\chi) = \ex_{x\in G} \ex_{y\in G} f(y)g(x-y) \chi(-y) \chi(y-x).$$
Then we may interchange the order of summation and substitute $z = x-y$ to arrive at
$$\hat{f*g}(\chi) = \ex_{y\in G} \ex_{z\in G} f(y)g(z) \chi(-y) \chi(-z) = \hat f(\chi)\hat g(\chi),$$
which proves (i). For part (ii), we expand and multiply by the same $1$ to get
$$\widehat{f\circ g}(\chi) = \ex_{x\in G} (f\circ g)(\chi)\chi(-x)
= \ex_{x\in G} \ex_{y\in G} f(y)g(x+y) \chi(y) \chi(-x-y).$$
We again interchange the order of summation; this time substituting $z = x+y$ gives us
$$\eqalign{
\hat{f\circ g}(\chi) &= \ex_{y\in G} \ex_{z\in G} f(y)g(z) \chi(y)\chi(-z) \cr
&= \bar{\ex_{y\in G} f(y)\chi(-y) } \ex_{z\in G} g(z)\chi(-z) \cr
&=  \bar{\hat f(\chi)}\hat g(\chi),\cr
}$$
which is what we wanted.\slug

For sets $A$ and $X$, let $\mu_X(A) = |A\cap X|/|X|$ denote the
relative density of $A$ in $X$, and if $X$ is understood to be a subset of a larger set $G$, then we
use $\mu_X$ also to denote the normalised indicator function given by
$$\mu_X(x) = \cases{ 1/\mu_G(X), & if $x\in X$;\cr 0, & otherwise.}.$$
The scaling is done so that $\norm{\mu_X}_1 = 1$ for any $X\subseteq G$, as can easily be checked.

\advsect H\"older lifting and unbalancing

Here we state and prove two lemmas that are general enough to apply in both the integer and finite-field
settings.

\parenproclaim Lemma~{\advthm} (H\"older lifting). Let $\eps\ge 0$ and let $A$ and $C$ be subsets of
a finite abelian group $G$, where $C$ has relative density $\gamma$. Then at least one of
the following two statements holds.
\medskip
\item{i)} $\bigl|\langle \mu_A *\mu_A, \mu_C\rangle\bigr| \le \eps$
\smallskip
\item{ii)} $\norm{\mu_A\circ \mu_A - 1}_p \ge \eps/2$ for some $p = O\bigl(\log(1/\gamma)\bigr)$.

\proof Bilinearity of the inner product gives
$$\langle \mu_A *\mu_A -1, \mu_C \rangle = \langle \mu_A * \mu_A ,\mu_C\rangle + \langle -1,\mu_C\rangle
= \langle\mu_A*\mu_A,\mu_C\rangle -1,$$
so if the first statement does not hold, then for $q = 1/(1-1/p)$, we have, by H\"older's inequality,
$$\eqalign{
\eps &< \bigl|\langle \mu_A *\mu_A, \mu_C\rangle\bigr|
\le \norm{\mu_A*\mu_A-1}_p \Bigl( \ex_{x\in G} \bigl| \mu_C(x)\bigr|^q\Bigr)^{1/q} \cr
&\le \norm{\mu_A*\mu_A-1}_p \gamma^{1/q-1} \le \norm{\mu_A*\mu_A-1}_p \gamma^{-1/p}.\cr
}$$
Letting $p$ be an even integer greater than $\log_2(1/\gamma)$, we have $\log\gamma \ge p\log (1/2)$,
whence $\gamma^{1/p} \ge 1/2$, which gives the inequality
$$ \norm{\mu_A*\mu_A - 1}_p \ge {\epsilon\over 2}.$$
Lastly, observe that since $p$ is even,

\advsect The finite-field case

First, we give bounds for the size of a $3$-AP-free subset of $\FF_q^n$. As is common with problems of this
sort, the finite-field case is a simpler prototype of the integer one.

\section References

\bye

### 2023-08-28 about/vita.tex
Density increment and misc things

@@
 
-\leftright{Citizen of Canada and Singapore}{2-9602 74th Avenue NW}
-\leftright{Email: marcel\_goh@yahoo.ca}{Edmonton, AB T6E 1E6}
+\leftright{Citizen of Canada and Singapore}{5872 Esplanade Avenue}
+\leftright{Email: marcel\_goh@yahoo.ca}{Montreal, QC H2T 3A3}
 \leftright{Webpage: marcelgoh.ca}{+1\thinspace\up(825\up)\thinspace 440-0681}
@@
 
+\leftright{\bf Ph.D.\ Mathematics}{\bf September 2023 -- present}
+\leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
+\smallskip
+Supervisors: Andrew Granville and Hamed Hatami.
+\medbreak
 \leftright{\bf M.Sc.\ Mathematics}{\bf May 2021 -- present}
@@
 
+\leftright{\bf NSERC Canada Graduate Scholarship -- Doctoral
+\rm (\$105,000)}{\bf September 2023 -- August 2026}
+\smallskip
+\leftright{\bf FRQNT Bourse de doctorat en recherche
+\rm (\$100,000)}{\bf [Declined]}
+\smallskip
 \leftright{\bf NSERC Canada Graduate Scholarship -- Master's \rm (\$17,500)}{\bf May 2021 -- April 2022}
@@
 
-\sectheader{Graduate-level Coursework}
-
-Higher algebra (categories, groups, modules), functional analysis, quadratic and modular forms, combinatorics,
-topology (including some algebraic topology), probabilistic analysis of algorithms.
-
 \sectheader{Research Interests}
 
-Analysis of random discrete structures, analysis and design of algorithms, enumerative combinatorics,
-additive combinatorics.
+Additive combinatorics, analytic number theory, random trees, analysis of algorithms.
 
@@
 
+\starleftright{\bf Teaching assistant}{\bf September 2021 -- present}
+\leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
+\smallskip
+Teaching assistant in the following courses. Duties include giving tutorials, holding office hours, and
+grading assignments and exams.
+\begingroup\parindent=10pt
+\smallskip
+\thing Fall 2023: MATH 240 Discrete Structures
+\smallskip
+\thing Fall 2021: COMP 690 Probabilistic Analysis of Algorithms
+\endgroup
+\medbreak
+
 \starleftright{\bf After school care supervisor}{\bf January -- June 2023}
@@
 triples problem, as well as topics related to Khovanskii's theorem.
-\medbreak
-
-\starleftright{\bf Teaching assistant}{\bf September -- December 2021}
-\leftright{\sl School of Computer Science, McGill University}{\sl Montr\'eal, QC, Canada}
-\smallskip
-Teaching assistant for COMP 690, a graduate-level course on the probabilistic analysis of algorithms.
-Held office hours twice a week and responsible for grading of assignments.
 \medbreak

### 2023-08-28 misc/expo/kelleymeka.tex
Density increment and misc things

@@
 
-\maketitle{Kelley and Meka's proof of Roth's theorem}{by}{Marcel K. Goh}{5 August 2023}
+\maketitle{Kelley and Meka's proof of Roth's theorem}{by}{Marcel K. Goh}{23 August 2023}
 
@@
 
-\advsect Introduction
+\advsect Definitions and elementary facts
 
-Let $A$ be a subset of $\ZZ$. We want to know how dense $A$ can be before it must contain a subset
-$\{x,y,z\}\subseteq A$ with $x+z = 2y$, that is, an arithmetic progression of length $3$.
-
-\medskip
-\boldlabel Basic definitions and elementary lemmas.
-We will use $G$ primarily to refer to a finite abelian group, on which we have the normalised counting measure.
+We will use $G$ primarily to refer to a finite abelian group.
 For functions $f,g:G\to \CC$ we have the inner product
@@
 In $L_p$ spaces we have the useful H\"older inequality
-$$ \bigl| \langle f,g\rangle \bigr| \le \norm{f}_p  \cdot \norm{g}_{1-p},$$
-for $p,q\in [1,\infty]$ with $1/p + 1/q = 1$.
+$$ \bigl| \langle f,g\rangle \bigr| \le \norm{f}_p  \cdot \norm{g}_q,$$
+for $p,q\in [1,\infty)$ with $1/p + 1/q = 1$.
 Assuming now that $f$ and $g$ are $\RR$-valued, we also have the convolution
@@
 and the difference convolution
-$$ (f\circ g)(x) = \ex_{y\in G} f(y) g(x+y)$$
+$$ (f\circ g)(x) = \ex_{y\in G} f(y) g(x+y).$$
+It is easy to check that for all $x\in G$, $(f*g)(x) = (g*f)(x)$, but with the difference convolution
+we have $(f\circ g)(x) = (g\circ f)(-x)$.
 that are related by the following adjoint property.
 
-\parenproclaim Lemma~{\advthm} (Adjoint property).
+\parenproclaim Proposition~{\advthm} (Adjoint property).
 Let $G$ be a finite abelian group and let $f,g,h:G\to \RR$. Then
@@
 $$ \hat f(\chi) = \ex_{x\in G} f(x)\chi(-x).$$
-The following lemma describes how the convolution and difference convolution behave under the Fourier
+The following proposition describes how the convolution and difference convolution behave under the Fourier
 transform.
 
-\parenproclaim Lemma~{\advthm} (Convolution laws).
+\edef\convolutions{\the\thmcount}
+\parenproclaim Proposition~{\advthm} (Convolution laws).
 Let $G$ be a finite abelian group and let $f,g:G\to \RR$. Then the following identities hold:
@@
 
+When we convolve two functions $\hat f$ and $\hat g$ on the dual group, we
+take a sum instead of an expectation:
+$$(\hat f\circ \hat g)(\chi) = \sum_{\psi\in G} \hat f(\psi)\hat g(\chi\psi^{-1}).$$
+The same goes in the definition of the inner product $\langle \hat f, \hat g\rangle$.
+
+Let $f^{*k}$ denote the $k$-fold
+convolution of a function $f$.
+The next proposition interprets $k$-norms in terms of $k$-fold convolutions of the Fourier transform.
+
+\edef\knorms{\the\thmcount}
+\proclaim Proposition \advthm. Let $G$ be a finite abelian group, let $k\ge 1$ be an integer, and let
+$\chi_0$ denote the identity element of the dual group $\hat G$ of $G$. We have the identity
+$$\ex_{x\in G} f(x)^k = {\hat f} ^{*p} (\chi_0).$$
+
+\proof
+Expand by the Fourier inversion formula to get
+$$\eqalign{
+\ex_{x\in G} f(x)^k &= \ex_{x\in G} \Bigl(\sum_{\chi\in \hat G} \hat f(\chi) \chi(x)\Bigr)^k \cr
+&= \ex_{x\in G} \sum_{\chi_1 \in \hat G} \cdots \sum_{\chi_k \in \hat G} \hat f(\chi_1)\cdots \hat f(\chi_k)
+\chi_1(x)\cdots\chi_k(x) \cr
+&= \sum_{\chi_1 \in \hat G} \cdots \sum_{\chi_k \in \hat G} \hat f(\chi_1)\cdots \hat f(\chi_k)
+\ex_{x\in G} \chi_1\cdots\chi_k(x). \cr
+}$$
+By orthogonality of characters, the inner expectation is zero when $\chi_1\cdots\chi_k \ne \chi_0$, so we have
+$$\ex_{x\in G} f(x)^k = \sum_{\chi_1\cdots\chi_k = \chi_0} \hat f(\chi_1)\cdots \hat f(\chi_k) =
+{\hat f}^{*p}(\chi_0).\noskipslug$$
+
 For sets $A$ and $X$, let $\mu_X(A) = |A\cap X|/|X|$ denote the
@@
 The scaling is done so that $\norm{\mu_X}_1 = 1$ for any $X\subseteq G$, as can easily be checked.
+We denote the ordinary indicator function by $\one_X = \mu_G(X)\mu_X$, and sometimes
+write $\one_x$ for the indicator function $\one_{\{x\}}$ of a singleton set. Lastly, we also sometimes use the
+same symbol to denote the indicator function of a statement; i.e., $\indic{P}$ is $1$ if the statement $P$
+is true and $0$ if it is false.
+
+It is easy to check that if $\mu$ has $\norm{\mu}_1 = 1$, then so does $\mu*\mu$ and $\mu\circ\mu$.
+We shall say that $\mu:G\to \RR_{\ge 0}$ is a {\it probability measure} on $G$ if $\norm{\mu}_1 = 1$. The
+following proposition concerns such measures.
+
+\edef\normonefourier{\the\thmcount}
+\proclaim Proposition \advthm. Let $G$ be a finite abelian group.
+If $\mu:G\to \RR_{\ge 0}$ is a probability measure, then
+$$ \hat{\mu-1} = \hat \mu (1-\one_{\chi_0}).$$
+
+\proof We expand
+$$\eqalign{
+\hat{\mu-1}(\chi_0) &=  \ex_{x\in G} \bigl(\mu(x)-1\bigr) \chi(-x) \cr
+&= \ex_{x\in G} \mu(x)\chi(-x) - \ex_{x\in G} \chi(-x).\cr
+&= \hat\mu(\chi) - \ex_{x\in G} \chi(-x).\cr
+}$$
+If $\chi\ne\chi_0$, then the expectation vanishes, and
+if $\chi = \chi_0$, then the expection clearly equals $1$, and $\mu(\chi_0) = \norm{\mu}_1 = 1$, so the whole
+expression is zero.\slug
+
+If $\mu:G\to \RR_{\ge 0}$ is a probability measure and $f,g:G\to \CC$ we write
+$$\langle f,g\rangle_\mu = \ex_{x\in G} \mu(x)f(x)\bar{g(x)}$$
+for the inner product relative to $\mu$, and for $1\le p<\infty$ we write
+$$\norm{f}_{p(\mu)} = \Bigl( \ex_{x\in G} \mu(x) \bigl| f(x)\bigr|^p \Bigr)^{1/p}$$
+for the $L_p$ norm relative to $\mu$. The following basic proposition establishes the monotonicity of
+$L_p$ norms with respect to $p$.
+
+\edef\monotonenorm{\the\thmcount}
+\parenproclaim Proposition~{\advthm} (Monotonicity of $L_p$ norms). Let $G$ be a finite abelian group.
+Let $\mu:G\to \RR_{\ge 0}$ be a probability measure and let $f:G\to \CC$.
+For $1\le p<q<\infty$, we have
+$$\norm{f}_{p(\mu)} \le \norm{f}_{q(\mu)}.$$
+
+\proof Let $r = q/p > 1$ and let $s = r/(r-1)$ so that $1/r+1/s = 1$.
+We have
+$$\ex_{x\in G} \mu(x) \bigl| f(x)\bigr|^p = \ex_{x\in G} \mu(x) \bigl| f(x)\bigr|^{q/r} \cdot 1$$
+Now by H\"older's inequality, we have
+$$\eqalign{
+\ex_{x\in G} \mu(x) \bigl| f(x)\bigr|^p &\le
+\Bigl(\ex_{x\in G} \mu(x) \bigl| f(x)\bigr|^q \Bigr)^{1/r} \Bigl( \ex_{x\in G} 1^s\Bigr)^{1/s} \cr
+&= \Bigl(\ex_{x\in G}\mu(x)\bigl| f(x)\bigr|^q \Bigr)^{p/q}. \cr
+}$$
+Taking $p$th roots of both sides now produces the inequality we wanted.\slug
+
+For convenience, when $\mu:G\to \RR_{\ge 0}$ is a probability measure and $X\subseteq G$, we write
+$$\mu(X) = \norm{\one_X}_{1(\mu)} = \ex_{x\in G} \mu(x) \one_X(x),$$
+and refer to this quantity as the {\it density of $X$ relative to $\mu$}.
 
@@
 
-Here we state and prove two lemmas that are general enough to apply in both the integer and finite-field
-settings.
+With preliminaries out of the way, we begin the proof of Kelley and Meka~\ref{kelleymeka2023}, as described
+and reworked by Bloom and Sisask~\ref{bloomsisask2023}.
+In this section we perform the first two steps of the proof,
+which are identical in both the integer and finite-field settings.
 
+\edef\holderlifting{\the\thmcount}
 \parenproclaim Lemma~{\advthm} (H\"older lifting). Let $\eps\ge 0$ and let $A$ and $C$ be subsets of
@@
 \medskip
-\item{i)} $\bigl|\langle \mu_A *\mu_A, \mu_C\rangle\bigr| \le \eps$
+\item{i)} $\bigl|\langle \mu_A *\mu_A, \mu_C\rangle - 1\bigr| \le \eps$
 \smallskip
@@
 
-\proof Bilinearity of the inner product gives
+\proof Linearity of the inner product in the first argument gives
 $$\langle \mu_A *\mu_A -1, \mu_C \rangle = \langle \mu_A * \mu_A ,\mu_C\rangle + \langle -1,\mu_C\rangle
@@
 $$\eqalign{
-\eps &< \bigl|\langle \mu_A *\mu_A, \mu_C\rangle\bigr|
+\eps &< \bigl|\langle \mu_A *\mu_A-1, \mu_C\rangle\bigr|
 \le \norm{\mu_A*\mu_A-1}_p \Bigl( \ex_{x\in G} \bigl| \mu_C(x)\bigr|^q\Bigr)^{1/q} \cr
@@
 Letting $p$ be an even integer greater than $\log_2(1/\gamma)$, we have $\log\gamma \ge p\log (1/2)$,
-whence $\gamma^{1/p} \ge 1/2$, which gives the inequality
+whence $\gamma^{1/p} \ge 1/2$. This gives the inequality
 $$ \norm{\mu_A*\mu_A - 1}_p \ge {\epsilon\over 2}.$$
-Lastly, observe that since $p$ is even,
+Since $p$ is even,
+$$ \norm{\mu_A*\mu_A-1}_p^p = \ex_{x\in G} \bigl|(\mu_A*\mu_A-1)(x)\bigr|^p
+= \ex_{x\in G} (\mu_A*\mu_A-1)(x)^p,$$
+and we can apply Proposition~{\knorms} to get
+$$\norm{g}_p^p = \hat g^{*p} (\chi_0),$$
+where we have put $g = \mu_A*\mu_A-1$.
+It was noted earlier that $\mu_A*\mu_A$ has $1$-norm equal to $1$, so we can apply
+Propositions~{\normonefourier} and~{\convolutions} in that order to get
+$$\norm{\mu_A*\mu_A-1}_p^p
+= \bigl(\hat{\mu_A*\mu_A}(1-\one_{\chi_0})\bigr)^{*p} (\chi_0)
+= \bigl(\hat{\mu_A}^2(1-\one_{\chi_0})\bigr)^{*p} (\chi_0).$$
+Repeating this whole process with $\mu_A\circ\mu_A$ in place of $\mu_A*\mu_A$ produces the very similar identity
+$$\norm{\mu_A\circ\mu_A-1}_p^p = \bigl(|\hat{\mu_A}|^2(1-\one_{\chi_0})\bigr)^{*p} (\chi_0),$$
+from which we conclude that
+$$\norm{\mu_A\circ\mu_A-1}_p^p \ge \norm{\mu_A*\mu_A-1}_p^p \ge {\epsilon\over 2}.\noskipslug$$
+
+This lemma tells us that if $\langle \mu_A*\mu_A,\mu_C\rangle \ge 1/2$, then $\norm{\mu\circ \mu_A-1}_p \ge 1/4$
+for some $p = O\bigl(\log(1/\gamma)\bigr)$. This information can then be fed to the following general lemma.
+
+\edef\unbalancing{\the\thmcount}
+\parenproclaim Lemma~{\advthm} (Unbalancing of spectrally nonnegative functions). Let $\eps\in (0,1)$
+and let $\nu : G\to \RR_{\ge 0}$ be a probability measure with $\hat \nu\ge 0$. If $f : G\to \RR$ has
+$\hat f\ge 0$ and $\norm{f}_{p(\nu)} \ge \eps$ for some $p\ge 1$, then
+$$ \norm{f+1}_{p'(\nu)} \ge 1 + {\eps\over 2}$$
+for some $p' = O\bigl(\eps^{-1} \log(\eps^{-1}) p\bigr)$.
+
+\proof Proposition~{\monotonenorm} tells us that $\norm{f}_{p(\mu)}$ is monotonically increasing in $p$, so
+without loss of generality we can pick $p$ odd and at least $5$. As usual, we denote the identity in $\hat G$
+by $\chi_0$. Using the Fourier inversion formula
+and orthogonality of characters as we did in the proof of Proposition~{\knorms}, we observe that
+$$\eqalign{
+\norm{f}_{p(\nu)}^p &= \ex_{x\in G}
+\Bigl(\sum_{\chi\in \hat G} \hat\nu(\chi)\chi(x)\Bigr)\Bigl(\sum_{\chi\in \hat G} \hat f(\chi)\chi(x)\Bigr)^k \cr
+&= \sum_{\chi_1\in \hat G} \cdots \sum_{\chi_{k+1}\in \hat G}
+\hat f(\chi_1)\cdots\hat f(\chi_k) \hat \nu(\chi_{k+1} )\ex_{x\in G} \chi_1(x)\cdots\chi_{k+1}(x) \cr
+&= \sum_{\chi_1\cdots\chi_{k+1} =\chi_0} \hat f(\chi_1) \cdots \hat f(\chi_k) \hat\nu(\chi_{k+1} )\cr
+&= \hat\nu * {\hat f}^{*p}(\chi_0). \cr
+}$$
+Let $P = \{x\in G : f(x)\ge 0\}$ and let $g(x) = \max\{f(x), 0\}$. It is easy to see that
+$2g(x) = f(x) + \bigl|f(x)\bigr|$, so
+$$\eqalign{
+2\langle \one_P, f^p\rangle_\nu &= 2 \ex_{x\in G} \nu(x)\one_P(x)f(x)^p \cr
+&= 2\ex_{x\in G} \nu(x) g(x) f(x)^{p-1}  \cr
+&= \langle 2g,f^{p-1}\rangle_\nu \cr
+&= \ex_{x\in G} \nu(x) f(x)^p + \bigl\langle |f|, f^{p-1}\bigr\rangle_\nu \cr
+&= \hat\nu * {\hat f}^{*p}(\chi_0)+ \bigl\langle |f|, |f|^{p-1}\bigr\rangle_\nu, \cr
+}$$
+where in the last line we used the fact that $f$ is real-valued as well as evenness of $p-1$.
+Since the Fourier transforms of $f$ and $\nu$ are both nonnegative, the first term is nonnegative, so
+$$\langle \one_P, f^p\rangle_\nu \ge {\bigl\langle |f|, |f|^{p-1}\bigr\rangle_\nu \over 2}
+= {\norm{f}_{p(\nu)}^p\over 2} \ge {\eps^p\over 2} .$$
+Now let $T = \{x\in P : f(x) \ge 3\eps / 4\}$. Then
+$$\eqalign{
+\langle \one_{P\setminus T}, f^p \rangle &= \langle \one_P - \one_T, f^p \rangle \cr
+&= \langle \one_P , f^p \rangle - \langle \one_T, f^p \rangle \cr
+&\ge {\eps\over 2} - \ex_{x\in G} \one_T(x)f(x)^p \cr
+&\ge {\eps\over 2} - \ex_{x\in G} 3\eps/4 \cr
+&= {\eps\over 4},\cr
+}$$
+so $\langle \one_T, f^p \rangle_\nu \ge \eps^p/4$ and we have
+$$\eqalign{
+{\eps\over 4} &\le \langle \one_T,f^p\rangle_\nu \cr
+&= \ex_{x\in G} \bigl( \nu(x)^{1/2}\one_T(x)\bigr) \bigl(\nu(x)^{1/2} f(x)^p\bigr) \cr
+&= \Bigl( \ex_{x\in G} \nu(x)\one_T(x)^2\Bigr)^{1/2} \Bigl(\ex_{x\in G}\nu(x) f(x)^{2p}\Bigr)^{1/2} \cr
+&= \Bigl( \ex_{x\in G} \nu(x)\one_T(x)\Bigr)^{1/2} \Bigl(\ex_{x\in G} \nu(x) |f(x)|^{2p}\Bigr)^{p/(2p)} \cr
+&= \nu(T)^{1/2} \norm{f}_{2p(\nu)}^p\cr
+}$$
+by the Cauchy--Schwarz inequality.
+
+Now if $\norm{f}_{2p(\nu)}^p > 2$, then we could take $p' = 2p$, so assume that this norm is at most $2$.
+By the triangle inequality, we have
+$$\norm{f}_{2p(\nu)} \le \norm{-1}_{2p(\nu)} + \norm{f+1}_{2p(\nu)} \le 3,$$
+hence
+$$ \nu(T)^{1/2} 3^p \le {\eps^p\over 4}.$$
+Since we assumed that $p\ge 5$, and since $4 < 1024/243$, we have $4^{1/p} < 4/3$ and thus
+$$ \nu(T) \ge {\eps^{2p} \over 16 \cdot 3^{2p}} = \biggl( {\eps \over 4^{1/p} \cdot 3} \biggr)^{2p}
+> \biggl({\eps\over 4}\biggr)^{2p}.$$
+This allows us to bound
+$$\eqalign{
+\norm{f+1}_{p'(\nu)}
+&= \Bigl( \ex_{x\in G} \nu(x) \bigl| f(x)+1\bigr|^{p'}\Bigr)^{1/p'} \cr
+&\ge \Bigl( \ex_{x\in G} \nu(x) \one_T(x) \bigl| f(x)+1\bigr|^{p'}\Bigr)^{1/p'} \cr
+&\ge \bigl( \nu(T) (1+3\eps/4)^{p'} \bigr)^{1/p'} \cr
+&> \biggl( {\eps \over 4} \biggr)^{2p/p'} \biggl(1+{3\over 4}\eps\biggr). \cr
+}$$
+Now if $p' \ge (8p/\eps)\log(4/\eps) = O\bigl(\eps^{-1} \log(\eps^{-1}) p\bigr)$, then
+$\eps/4 \ge (2p/p')\log(4/\eps)$ and thus
+$$ -{2p\over p'} \log\biggl({4\over \eps}\biggr) \ge -{\eps\over 4}.$$
+Taking $e$ to the power of both sides gives us
+$$\biggl({4\over \eps}\biggr) ^{-2p/p'} \ge e^{-\eps/4} \ge 1-{\eps\over 4},$$
+and plugging this in above gives the bound
+$$\norm{f+1}_{p'(\nu)} > \biggl(1-{\eps\over 4}\biggr)\biggl(1+{3\over 4}\eps\biggr)
+= 1+{\eps \over 2} -{3\eps^2\over 16} \ge 1+{\eps\over 2},$$
+which is what we needed.\slug
+
+\advsect Dependent random choice
+
+The next lemma uses a dependent random choice argument to pass the information from the previous
+step down to high density subsets, which allows us to iterate the argument.
+
+\parenproclaim Lemma~{\advthm} (Dependent random choice). Let $G$ be a finite abelian group and let $A$
+be a subset of $G$ with density $\alpha$. Let $B_1,B_2\subseteq G$ and $\mu = \mu_{B_1}\circ \mu_{B_2}$.
+For any function $f:G\to \RR_{\ge 0}$ there exist sets $A_1\subseteq B_1$ and $A_2\subseteq B_2$ with
+densities satisfying
+$$\min\bigl\{\mu_{B_1}(A_1), \mu_{B_2}(A_2)\bigr\}
+\ge {1\over 4} \alpha^{2p} \norm{\mu_A\circ\mu_A}_{p(\mu)}^{2p}k.$$
+and such that
+$$ \langle \mu_{A_1}\circ \mu_{A_2}, f\rangle \le 2 {\bigl\langle (\mu_A\circ\mu_A)^p, f\bigr\rangle_\mu
+\over \norm{\mu_A\circ\mu_A}_{p(\mu)}^p}.$$
+
+\proof For $s = (s_1,\ldots,s_p)\in G^p$ let
+$A_1(s) = B_1\cap (A+s_1) \cap \cdots \cap (A+s_p)$, and define $A_2(s)$ analogously.
+First we expand
+$$\eqalign{
+\bigl \langle (\mu_A\circ \mu_A)^p, f\bigr\rangle_{\mu}
+&= \ex_{x\in G} \mu(x) (\mu_A\circ \mu_A)(x)^p f(x) \cr
+&=  \ex_{x\in G} \ex_{y\in G} \mu_{B_1}(y) \mu_{B_2}(x+y)  (\mu_A\circ \mu_A)(x)^p f(x)\cr
+&= {1\over |B_1|\cdot |B_2|} \sum_{x\in G} \sum_{y\in G}
+\one_{B_1}(y) \one_{B_2}(x+y) (\mu_A\circ\mu_A)(x)^p f(x).\cr
+}$$
+Renaming $b_1 = y$ and performing the change of variable $b_2 = x+b_1 = x+y$, we have
+$$\eqalign{
+\bigl \langle (\mu_A\circ \mu_A)^p, f\bigr\rangle_{\mu}
+&= {1\over |B_1||B_2|} \displaystyle\sum_{\scriptstyle b_1\in G\atop \scriptstyle b_2\in G}
+\one_{B_1}(b_1) \one_{B_2}(b_2) (\mu_A\circ\mu_A)(b_2-b_1)^p f(b_2-b_1)\cr
+&= \ex_{b_1\in B_1, b_2\in B_2} (\mu_A\circ\mu_A)(b_2-b_1)^p f(b_2-b_1)\cr
+&= \ex_{b_1\in B_1, b_2\in B_2} \Bigl( \alpha^{-2}\ex_{y\in G} \one_A(y) \one_A(b_2-b_1+y)\Bigr)^p f(b_2-b_1).\cr
+}$$
+Now since $y\in A$ if and only if $b_1\in A+b_1-y$ and $b_2-b_1+y \in A$ if and only if $b_2 \in A+b_1-y$, so
+writing $t = b_1-y$ and changing variables, we have
+$$\eqalign{
+\bigl \langle (\mu_A\circ \mu_A)^p, f\bigr\rangle_{\mu}
+&= \ex_{b_1\in B_1, b_2\in B_2}
+ \Bigl( \alpha^{-2}\ex_{t\in G} \one_{A+t}(b_1) \one_{A+t}(b_2)\Bigr)^p f(b_2-b_1)\cr
+&= \alpha^{-2p}\ex_{b_1\in B_1, b_2\in B_2} \ex_{s\in G^p} \one_{A_1(s)}(b_1) \one_{A_2(s)}(b_2) f(b_2-b_1).\cr
+}$$
+Putting $y=b_2-b_1$ so that $b_2 = y+b_1$, we have
+$$\eqalign{
+\bigl \langle (\mu_A\circ \mu_A)^p, f\bigr\rangle_{\mu}
+&= \alpha^{-2p} \ex_{s\in G^p} \ex_{b_1\in B_1} {|G|\over|B_2|} \ex_{y\in G}
+ \one_{A_1(s)}(b_1) \one_{A_2(s)}(y+b_1) f(y)\cr
+&= {|G|\over \alpha^{2p}|B_2|} \ex_{s\in G^p} \ex_{b_1\in B_1}
+\bigl\langle \one_{A_1(s)}\circ \one_{A_2(s)}, f\bigr\rangle\cr
+&= {|G|^2\over \alpha^{2p}|B_1|\cdot|B_2|} \ex_{s\in G^p} 
+\bigl\langle \one_{A_1(s)}\circ \one_{A_2(s)}, f\bigr\rangle.\cr
+}$$
+Thus we let $\beta_i = |B_i|/|G|$ and $\alpha_i(s) = {|A_i(s)|\over|G|}$
+for $i\in\{1,2\}$, and then apply the above in the case where $f$ is the constant function $1$ to obtain
+$$\eqalign{
+\bignorm{(\mu_A\circ \mu_A)^p, f}_{p(\mu)}^p
+&= {|G|^2\over \alpha^{2p}|B_1|\cdot|B_2|} \ex_{s\in G^p} 
+\ex_{x\in G} \ex_{y\in G} \one_{A_1(s)}(y) \one_{A_2(s)}(x+y)\cr
+&= {1\over \alpha^{2p}\beta_1\beta_2} \ex_{s\in G^p} 
+\ex_{y\in G} \one_{A_1(s)}(y) \ex_{x\in G}\one_{A_2(s)}(x+y)\cr
+&= {1\over \alpha^{2p}\beta_1\beta_2} \ex_{s\in G^p} \alpha_1(s)\alpha_2(s)\cr
+}$$
+The constants out front do not depend on $f$, so we see that
+$$ {\bigl\langle (\mu_A\circ\mu_A)^p, f\bigr\rangle_{\mu}\over \norm{\mu_A\circ\mu_A}_{p(\mu)}^p}
+= {\ex_{s\in G^p} \bigl\langle \one_{A_1(s)}\circ \one_{A_2(s)}, f\bigr\rangle
+\over \ex_{s\in G^p} \alpha_1(s)\alpha_2(s)};$$
+call this quotient $\eta$ for brevity.
+Now we consider the quantity
+$$\ex_{s\in G^p} \ex_{x\in G} \one_{A_1(s)}(x).$$
+For a given $s = (s_1,\ldots,s_p)\in G^p$ and $x\in G$, the corresponding $\one_{A_1(s)}(x)$ term is $1$ if and
+only if $x\in B_1$ and $x-s_i\in A$ for all $1\le i\le p$. Hence we have
+$$\ex_{s\in G^p} \ex_{x\in G} \one_{A_1(s)}(x) = {|B_1|\cdot |A|^p\over |G|^{p+1}} = \alpha^p\beta_1.$$
+The analogous identity holds for $A_2(s)$. So, letting
+$$M = {1\over 2} \alpha^p (\beta_1\beta_2)^{1/2} \norm{\mu_A\circ\mu_A}^p_{p(\mu)},$$
+we have
+$$\eqalign{
+\ex_{s\in G^p} \indic{\alpha_1(s)\alpha_2<M^2} \alpha_1(s)\alpha_2(s)
+ &< \ex_{s\in G^p} M \sqrt{\alpha_1(s)\alpha_2(s)} \cr
+ &\le \Bigl( \ex_{s\in G^p} M \alpha_1(s) \Bigr)^{1/2} \Bigl( \ex_{s\in G^p} M \alpha_2(s) \Bigr)^{1/2} \cr
+ &= M \Bigl( \ex_{s\in G^p} \ex_{x\in G} \one_{A_1(s)}(x) \Bigr)^{1/2} \cr
+ &\hskip70pt\Bigl( \ex_{s\in G^p} \ex_{x\in G} \one_{A_2(s)}(x) \Bigr)^{1/2} \cr
+ &= M\alpha^p \sqrt{\beta_1\beta_2} \cr
+ &= {1\over 2} \alpha^{2p}\beta_1\beta_2 \norm{\mu_A\circ\mu_A}^p_{p(\mu)} \cr
+&= {1\over 2} \ex_{s\in G^p} \alpha_1(s)\alpha_2(s) \cr
+}$$
+and consequently
+$$\ex_{s\in G^p} \indic{\alpha_1(s)\alpha_2\ge M^2} \alpha_1(s)\alpha_2(s)
+> {1\over 2} \ex_{s\in G^p} \alpha_1(s)\alpha_2(s).$$
+So we have
+$$\eqalign{
+\ex_{s\in G^p} \bigl\langle \one_{A_1(s)}\circ\one_{A_2(s)}, f\bigr\rangle
+&= \eta \ex_{s\in G^p} \alpha_1(s)\alpha_2(s) \cr
+  &< 2\eta \ex_{s\in G^p} \alpha_1(s)\alpha_2(s)\indic{\alpha_1(s)\alpha_2(s)\ge M^2}, \cr
+}$$
+and thus there must be some $s$ such that
+$$\bigl\langle \one_{A_1(s)}\circ\one_{A_2(s)}, f\bigr\rangle
+< 2\eta \alpha_1(s)\alpha_2(s)\indic{\alpha_1(s)\alpha_2(s)\ge M^2}.$$
+Since $f(x)\ge 0$ for all $x$, the left-hand side is nonnegative, meaning that the right-hand side cannot be $0$.
+Thus such an $s$ must satisfy $\alpha_1(s)\alpha_2(s)\ge M^2$. Letting $A_1 = A_1(s)$ and
+$A_2 = A_2(s)$ for this particular $s$, we have
+$${|A_1|\cdot|A_2|\over |G|^2} \ge
+{1\over 4} \alpha^{2p} {|B_1|\cdot|B_2|\over |G|^2}\norm{\mu_A\circ\mu_A}_{p(\mu)}^{2p},$$
+whence
+$$\mu_{B_1}(A_1)\mu_{B_2}(A_2) \ge {1\over 4} \alpha^{2p} \norm{\mu_A\circ\mu_A}_{p(\mu)}^{2p},$$
+so neither $\mu_{B_1}(A_1)$ nor $\mu_{B_2}(A_2)$ can be less than the right-hand side.
+
+On the other hand, letting $\alpha_1 = \alpha_1(s)$ and $\alpha_2 = \alpha_2(s)$, we also have
+$$\eqalign{
+\langle\mu_{A_1}\circ\mu_{A_2}, f\rangle &= \ex_{x\in G} \ex_{y\in G} \mu_{A_1}(y)\mu_{A_2}(x+y) f(x) \cr
+&= {\alpha_1}^{-1}{\alpha_2}^{-1} \ex_{x\in G} \ex_{y\in G} \one_{A_1}(y)\one_{A_2}(x+y) f(x) \cr
+&= {\alpha_1}^{-1}{\alpha_2}^{-1} \langle \one_{A_1}\circ\one_{A_2},f\rangle  \cr
+&< 2\eta \cr
+&= 2 {\bigl\langle (\mu_A\circ\mu_A)^p, f\bigr\rangle_\mu \over \norm{\mu_A\circ\mu_A}_{p(\mu)}^p}, \cr
+}$$
+which proves the lemma.\slug
+
+This lemma is slightly more general than we shall require. The version that suffices for all our applications
+is the following.
+
+\edef\dependentrandom{\the\thmcount}
+\proclaim Lemma~\advthm. Let $G$ be a finite abelian group, let $p\ge 1$ be an integer, and let
+$\eps,\delta>0$. Let $B_1$ and $B_2$ be subsets of $G$ and let $\mu = \mu_{B_1}\circ \mu_{B_2}$.
+If $A\subseteq G$ has density $\alpha$ and
+$$S = \bigl\{x\in G : (\mu_A\circ\mu_A)(x) > (1-\eps)\norm{\mu_A\circ\mu_A}_{p(\mu)}\bigr\},$$
+then there exist $A_1\subseteq B_1$ and $A_2\subseteq B_2$ with densities satisfying
+$$\min\bigl\{ \mu_{B_1}(A_1), \mu_{B_2}(A_2)\bigr\}
+= \Omega\bigl( (\alpha \norm{\mu_A\circ\mu_A}_{p(\mu)})^{2p+O(\eps^{-1}\log(\delta^{-1}))}\bigr).$$
+such that
+$$\langle \mu_{A_1}\circ\mu_{A_2}, \one_S\rangle \ge 1-\delta.$$
+
+\proof Let $p'$ be the smallest even integer at least $p+\eps^{-1}\log(\delta^{-1})$.
+By the previous lemma applied to the set
+$\one_{G\setminus S}$, there exist sets $A_1\subseteq B_1$ and $A_2\subseteq B_2$ with densities
+satisfying
+$$\eqalign{
+\min\bigl\{\mu_{B_1}(A_1), \mu_{B_2}(A_2)\bigr\} &\ge
+{1\over 4} \alpha^{2p'} \norm{\mu_A\circ\mu_A}_{p'(\mu)}^{2p'}  \cr
+&\ge {1\over 4}\bigl(\alpha\norm{\mu_A\circ\mu_A}_{p(\mu)}\bigr)^{2p+2\eps^{-1}\log(\delta^{-1})+O(1)} \cr
+&= \Omega\bigl( (\alpha \norm{\mu_A\circ\mu_A}_{p(\mu)})^{2p+O(\eps^{-1}\log(\delta^{-1}))}\bigr)\cr
+}$$
+such that
+$$\langle \mu_{A_1}\circ\mu_{A_2}, \one_{G\setminus S}\rangle
+\le {\bigl\langle (\mu_A\circ\mu_A)^{p'}, \one_{G\setminus S}\bigr\rangle_{\mu} \over
+\norm{\mu_A\circ\mu_A}^{p'}_{p'(\mu)}}.$$
+Our construction of $S$ ensures that
+$${\bigl\langle (\mu_A\circ\mu_A)^{p'}, \one_{G\setminus S}\bigr\rangle_{\mu} \over
+\norm{\mu_A\circ\mu_A}^{p'}_{p'(\mu)}} \le (1-\eps)^p,$$
+and since $p'\ge \eps^{-1}\log(\delta^{-1})$, we have
+$$(1-\eps)^p\le e^{-\eps p} \le \delta.$$
+Putting everything together, we have
+$$\langle \mu_{A_1}\circ\mu_{A_2}, \one_{G\setminus S}\rangle \le \delta,$$
+so that
+$$\langle \mu_{A_1}\circ\mu_{A_2}, \one_S\rangle
+= 1- \langle \mu_{A_1}\circ\mu_{A_2}, \one_{G\setminus S}\rangle \ge 1-\delta,$$
+which completes the proof.\slug
 
@@
 
-First, we give bounds for the size of a $3$-AP-free subset of $\FF_q^n$. As is common with problems of this
-sort, the finite-field case is a simpler prototype of the integer one.
+We now use the methods of Kelley and Meka to give upper bounds on the size of a subset of $\FF_q^n$
+without any three-term arithmetic progressions.
+First, we restate the dependent random choice lemma in the special case that applies to this finite field
+context.
+
+\edef\deprandcorollary{\the\thmcount}
+\proclaim Corollary~{\advthm}. Let $p\ge 1$ be an integer and $\eps\in (0,1/2]$. If $A\subseteq G$ is such that
+$\norm{\mu_A\circ\mu_A}_p \ge 1+\eps$ and
+$S = \bigl\{x\in G: (\mu_A\circ\mu_A)(x) > 1+\eps/2\bigr\}$, then there are
+subsets $A_1$ and $A_2$ of $G$, each of density $\Omega\bigl( \alpha^{2p+O(\eps^{-1}\log(\eps^{-1}))}\bigr)$,
+such that
+$$\langle \mu_{A_1}\circ\mu_{A_2}, \one_S\rangle \ge 1-\eps/8.$$
+
+\proof Let $S' = \bigl\{x\in G : (\mu_A\circ\mu_A)(x) > (1+\eps)\norm{\mu_A\circ\mu_A}_p\bigr\}$, and apply
+Lemma~{\dependentrandom} with the same $p$ and $\eps$, but $\delta$ set to $\eps/8$, $S$ set to $S'$,
+and $B_1 = B_2 = G$ so that $\mu = \mu_{B_1} = \mu_{B_2}$ is the uniform measure on $G$. Hence the
+sets $A_1$ and $A_2$ given by the lemma will each have density
+$$\Omega\bigl((\alpha(1+\eps))^{2p+O(\eps^{-1}\log(8/\eps))}\bigr)
+=\Omega\bigl(\alpha^{2p+O(\eps^{-1}\log(\eps^{-1}))}\bigr)$$
+in $G$, and since
+$$(1-\eps)\norm{\mu_A\circ\mu_A}_p \ge (1-\eps)(1+\eps) = 1-\eps^2 \ge 1-\eps/2,$$
+we have $S'\subseteq S$ and thus
+$$\langle \mu_{A_1}\circ\mu_{A_2}, \one_S\rangle \ge
+\langle \mu_{A_1}\circ\mu_{A_2}, \one_{S'}\rangle \ge 1-\eps/8.\noskipslug$$
+
+There is a theorem that we need which, for now, we shall just state without proof. (This is Theorem 3.2
+of~\ref{schoensisask2016}.)
+[TODO: Add whole section proving this.]
+
+\edef\almostperiodicity{\the\thmcount}
+\parenproclaim Theorem~{\advthm} (Schoen--Sisask, {\rm 2016}).
+Let $\eps\in (0,1)$, let $S\subseteq \FF_q^n$, and
+let $A_1,A_2\subseteq \FF_q^n$ be subsets
+of relative density at least $\alpha$. There is a subspace $V$ of codimension
+$O\bigl(\eps^{-2}\log(\eps^{-1}\alpha^{-1})^2\log(\alpha^{-1})^2\bigr)$
+such that
+$$\bigl| \langle \mu_V*\mu_{A_1}*\mu_{A_2}, \one_S\rangle -
+ \langle \mu_{A_1}*\mu_{A_2}, \one_S\rangle\bigr| \le \eps.$$
+
+The following lemma encapsulates the density increment argument that underlies the proof of Roth's theorem.
+
+\parenproclaim Lemma~{\advthm} (Density increment). Let $\eps\in (0,1)$ and let $A$ and $C$
+be subsets of $G =\FF_q^n$ with relative densities $\alpha$ and $\gamma$, respectively. Then either
+\medskip
+\item{i)} $\bigl|\langle \mu_A*\mu_A,\mu_C\rangle-1\bigr| \le \eps$; or
+\smallskip
+\item{ii)} there is a subspace $V$ of codimension
+$$O\Bigl(\eps^{-2}\bigl(\log(1/\gamma)+\eps^{-1}\log(\eps^{-1})\bigr)^4\log(1/\alpha)^4\Bigr)$$
+such that $\max_{x\in G}\;(\one_A*\mu_V)(x) \ge \bigl(1+\eps/32\bigr)\alpha$.
+\medskip
+
+\proof Suppose that (i) fails. Then by Lemma~{\holderlifting} there is some $p = O\bigl(\log(1/\gamma)\bigr)$
+such that $\norm{\mu_A\circ\mu_A - 1}_p\ge \eps/ 2$. By the second convolution law (part (ii) of
+Proposition~{\convolutions}), $\mu_A\circ\mu_A$ is a nonnegative function, and note also that if
+$\nu$ is the uniform measure on $\FF_q^n$, then for any $\chi:\hat{\FF_q^n}\to \CC$,
+$$\hat\nu(\chi) = \ex_{x\in G} \nu(x)\chi(-x) = \cases{q^{-n}, & if $\chi$ is the trivial character;
+\cr 0, & otherwise.}$$
+This implies in particular that $\hat\nu \ge 0$,
+so by Lemma~{\unbalancing} applied with $f=\mu_A\circ\mu_A$ and the uniform measure for $\nu$, we find
+that $\norm{\mu_A\circ\mu_A}_{p'}\ge 1+\eps/4$ for some
+$p' = O\bigl(2\log(2) p\bigr)= O\bigl(\log(1/\gamma)\bigr)$.
+Let $C(\eps) = \eps^{-1}\log(\eps^{-1})$. By Corollary~{\deprandcorollary},
+there are sets $A_1,A_2\subseteq G$, each of density
+$\Omega(\alpha^{2p'+O(C(\eps))})$, such that
+$$\langle \mu_{A_1}\circ\mu_{A_2}, \one_S\rangle \ge 1 - \eps/32,$$
+where $S = \bigl\{ x \in G:(\mu_A\circ\mu_A)(x)\ge 1+\eps/8\bigr\}$. Feeding $-A_1$, $A_2$, and $S$
+into Theorem~{\almostperiodicity} with $\eps/32$ gives us a subspace $V$ of codimension
+$$\eqalign{
+O\Bigl(\eps^{-2}&\log\bigl(\eps^{-1}\alpha^{-2p'-O(C(\eps))}\bigr)^2
+\log\bigl(\alpha^{-2p'-O(C(\eps))}\bigr)^2\Bigr) \cr
+&= O\Bigl(\eps^{-2} (2p'+C(\eps))^2\log(1/\alpha)^2
+\bigl((2p'+C(\eps))\log(1/\alpha)+\log(\eps^{-1})\bigr)^2
+\Bigr)\cr
+&=O\Bigl(\eps^{-2}\bigl(\log(1/\gamma)+C(\eps)\bigr)^4\log(1/\alpha)^4\Bigr) \cr
+}$$
+such that
+$$\bigl| \langle \mu_V*\mu_{-A_1}*\mu_{A_2}, \one_S\rangle -
+ \langle \mu_{-A_1}*\mu_{A_2}, \one_S\rangle\bigr| \le \eps/32.$$
+It is easily checked that $\mu_{-A_1}*\mu_{A_2} = \mu_{A_1}\circ\mu_{A_2}$, so we find that
+$$\bigl\langle \mu_V*(\mu_{A_1}\circ\mu_{A_2}), \one_S\bigr\rangle \ge 1-\eps/16.$$
+Now we observe that
+$$\bignorm{(\mu_{A_1}\circ\mu_{A_2})\circ\mu_{A_2}}_1
+= \ex_{z\in G}\mu_{A_1}(z)\ex_{y\in G} \mu_{A_2}(y+z) \ex_{x\in G} \mu_A(x+y) = 1,$$
+so that,
+$$\eqalign{
+\max_{x\in G}\;(\mu_V*\one_A)(x) &= \alpha\max_{x\in G}\;(\mu_V*\mu_A)(x) \cr
+&= \alpha \bignorm{(\mu_{A_1}\circ\mu_{A_2})\circ\mu_A}_1 \max_{x\in G}\;(\mu_V*\mu_A)(x)\cr
+&\ge \alpha\bigl\langle \mu_V*\mu_A, (\mu_{A_1}\circ\mu_{A_2})\circ\mu_A\bigr\rangle\cr
+&= \alpha\bigl\langle \mu_V*\mu_A * (\mu_{A_1}\circ\mu_{A_2}), \mu_A\bigr\rangle\cr
+&= \alpha\bigl\langle \mu_V* (\mu_{A_1}\circ\mu_{A_2}), \mu_A\circ\mu_A\bigr\rangle,\cr
+}$$
+where in the last two lines we have employed the adjoint property
+$\langle f,g*h \rangle = \langle h\circ f,g\rangle$, as well as the commutative properties
+$f*g = g*f$ and $\langle f,g\rangle = \langle g,f\rangle$,
+all of which hold for functions $f,g,h : G\to \RR$. But by the construction of $S$, we have
+$\one_S(x)(\mu_A\circ\mu_A)(x) \ge (1+\eps/8)\one_S$, so
+$$\eqalign{
+\bigl\langle \mu_V* (\mu_{A_1}\circ\mu_{A_2}), \mu_A\circ\mu_A\bigr\rangle
+&\ge \bigl\langle \mu_V* (\mu_{A_1}\circ\mu_{A_2}), \one_S(\mu_A\circ\mu_A)\bigr\rangle \cr
+&\ge (1+\eps/8)\bigl\langle \mu_V* (\mu_{A_1}\circ\mu_{A_2}), \one_S\bigr\rangle \cr
+&\ge (1+\eps/8)(1-\eps/16) \cr
+&\ge 1+\eps/32, \cr
+}$$
+hence we conclude that
+$$\max_{x\in G}\;(\one_A*\mu_V)(x) \ge (1+\eps/32)\alpha\noskipslug$$
 

### 2023-08-30 misc/expo/kelleymeka.tex
More kelley meka

@@
 we have $(f\circ g)(x) = (g\circ f)(-x)$.
-that are related by the following adjoint property.
+We also have the following adjoint property.
 
@@
 
-For a group $G$ the dual group $\hat G$ is the set of all homomorphisms from $G\to \CC^\times$.
+For a group $G$ the dual group $\hat G$ is the set of all homomorphisms from $G$ to $\CC^\times$.
 The Fourier transform of $f:G\to \RR$ is the function $\hat f:\hat G\to \CC$ given by
@@
 
-\advsect H\"older lifting and unbalancing
+\advsect H\"older lifting and unbalancing for finite groups
 
@@
 and reworked by Bloom and Sisask~\ref{bloomsisask2023}.
-In this section we perform the first two steps of the proof,
-which are identical in both the integer and finite-field settings.
+In this section we perform the first two steps of the proof, in the general setting of finite groups.
 
@@
 of~\ref{schoensisask2016}.)
-[TODO: Add whole section proving this.]
 
@@
 
+\edef\densityincrement{\the\thmcount}
 \parenproclaim Lemma~{\advthm} (Density increment). Let $\eps\in (0,1)$ and let $A$ and $C$
@@
 
+\parenproclaim Theorem~{\advthm} (Finite field). Let $q$ be a power of an odd prime and let $A$ be
+a subset of $G = \FF_q^n$ of cardinality $\alpha q^n$. The number of (possibly trivial)
+three-term arithmetic progressions contained in $A$ is at least
+$${\alpha^3\over 2}q^{2n-O(\log(1/\alpha)^9)}.$$
+Hence if $A\subseteq \FF_q^n$ contains no nontrivial three-term arithmetic progressions, then
+$\alpha \le e^{-\Omega(n^{1/9})} = O\bigl(q^{-\Omega(n^{1/9})}\bigr)$.
+
+\proof Let $C = 2\cdot A = \{2a : a\in A\}$, so that $\gamma = |C|/q^n = \alpha$. By Lemma~{\densityincrement}
+applied to $A$ and $C$ with parameter $\eps = 1/2$, we find that either
+$\langle \mu_A*\mu_A,\mu_C\rangle \ge 1/2$ or there is a subspace $V$ of codimension
+$O\bigl(\log(1/alpha)^8\bigr)$ such that $\max_{x\in G}\;(\one_A*\mu_V)(x)\ge (1+\eps/64)\alpha$.
+
+In the second case, there exists some $x\in G$ such that
+$$\ex_{y\in G} \one_A(y)\mu_V(x-y) \ge (1+\eps/64)\alpha.$$
+But $x-y\in V$ if and only if $y-x\in V$ if and only if $y\in V+x$, so we find that
+$$|A\cap V+x| \ge (1+\eps/64)\alpha|V|.$$
+Now $A$ has exactly the same number of three-term arithmetic progressions as $A-x$, so we can invoke
+Lemma~{\densityincrement} again with $V$ in place of $G$ and $A-x$ in place of $A$, but note that $\alpha$
+has been replaced by $(1+\eps/64)\alpha$, so this iteration can only happen
+$\log_{1+\eps/64}(1/\alpha) = O\bigl(\log(1/\alpha)\bigr)$ times
+before the second case of the lemma becomes impossible, since $\alpha\le 1$. Hence we deduce that
+there is some subspace $V$ of codimension $O\bigl(\log(1/\alpha)^9\bigr)$ and some translate $A+x'$ of $A$
+such that
+$$\ex_{x\in V} \ex_{y\in V} \mu_{A'}(y)\mu_{A'}(x-y)\mu_{2\cdot A'}(x) \ge {1\over 2},$$
+where $A' = (A+x)\cap V$ and
+the relative densities are taken with respect to the subspace $V$. Expanding further, this implies that
+$${|V|^3\over |A'|^3|V|^2} \sum_{x\in V}\sum_{y\in V}
+\one_{A'}(y) \one_{A'}(x-y)\one_{2\cdot A'}(x) \ge {1\over 2};$$
+that is,
+$$\bigl| \{ (x,y)\in (2\cdot A) \times A : x-y\in A'\}\bigr| \ge {|A'|^3|V|^5\over 2|V|^3}
+\ge {\alpha^3\over 2}q^{2n-O(\log(1/\alpha)^9)}.$$
+Renaming variables, this counts the number of pairs $(x,z) \in A'\times A'$ such that $x+z = 2y$ for some
+$y\in A'$. Since this equation implies that $z-y = y-x$, the above expression
+counts the number of three-term arithmetic
+progressions in $A'$, including the $|A'|$ trivial instances of $x=y=z$. This proves the first part of the
+theorem.
+
+For the last part of the theorem statement,
+suppose that $A$ does not contain any nontrivial three-term arithmetic progressions.
+Then
+$${\alpha^3\over 2}q^{2n-O(\log(1/\alpha)^9)} \le \bigl| \{ (x,y)\in (2\cdot A) \times A : x-y\in A'\}\bigr|
+\le |A'| \le \alpha q^n,$$
+whence
+$$q^n \le {2q^{O(\log(1/\alpha)^9)}\over \alpha^2},$$
+and taking $q$th logs of both sides yields
+$$n \le \log_q\biggl({2q^{O(\log(1/\alpha)^9)}\over \alpha^2}\biggr) = O\bigl(\log(1/\alpha)^9\bigr).$$
+Letting $C$ be the constant implicit in the $O$-notation, we invert this to obtain the bound
+$$\alpha\le e^{-n^{1/9}/C^{1/9}} = e^{-\Omega(n^{1/9})},$$
+which is what we wanted.\slug
+
 \section References

### 2023-09-21 misc/expo/kelleymeka.tex
Update bio

@@
 which proves (i). For part (ii), we expand and multiply by the same $1$ to get
-$$\widehat{f\circ g}(\chi) = \ex_{x\in G} (f\circ g)(\chi)\chi(-x)
+$$\widehat{f\circ g}(\chi) = \ex_{x\in G} (f\circ g)(x)\chi(-x)
 = \ex_{x\in G} \ex_{y\in G} f(y)g(x+y) \chi(y) \chi(-x-y).$$
@@
 $\chi_0$ denote the identity element of the dual group $\hat G$ of $G$. We have the identity
-$$\ex_{x\in G} f(x)^k = {\hat f} ^{*p} (\chi_0).$$
+$$\ex_{x\in G} f(x)^k = {\hat f} ^{*k} (\chi_0).$$
 
@@
 $$\eqalign{
-\hat{\mu-1}(\chi_0) &=  \ex_{x\in G} \bigl(\mu(x)-1\bigr) \chi(-x) \cr
+\hat{\mu-1}(\chi) &=  \ex_{x\in G} \bigl(\mu(x)-1\bigr) \chi(-x) \cr
 &= \ex_{x\in G} \mu(x)\chi(-x) - \ex_{x\in G} \chi(-x).\cr
@@
 \parenproclaim Lemma~{\advthm} (Unbalancing of spectrally nonnegative functions). Let $\eps\in (0,1)$
-and let $\nu : G\to \RR_{\ge 0}$ be a probability measure with $\hat \nu\ge 0$. If $f : G\to \RR$ has
+and let $\nu : G\to \RR_{\ge 0}$ have $\norm{\mu}_1 = 1$ and $\hat \nu\ge 0$. If $f : G\to \RR$ has
 $\hat f\ge 0$ and $\norm{f}_{p(\nu)} \ge \eps$ for some $p\ge 1$, then
@@
 \norm{f}_{p(\nu)}^p &= \ex_{x\in G}
-\Bigl(\sum_{\chi\in \hat G} \hat\nu(\chi)\chi(x)\Bigr)\Bigl(\sum_{\chi\in \hat G} \hat f(\chi)\chi(x)\Bigr)^k \cr
-&= \sum_{\chi_1\in \hat G} \cdots \sum_{\chi_{k+1}\in \hat G}
-\hat f(\chi_1)\cdots\hat f(\chi_k) \hat \nu(\chi_{k+1} )\ex_{x\in G} \chi_1(x)\cdots\chi_{k+1}(x) \cr
-&= \sum_{\chi_1\cdots\chi_{k+1} =\chi_0} \hat f(\chi_1) \cdots \hat f(\chi_k) \hat\nu(\chi_{k+1} )\cr
+\Bigl(\sum_{\chi\in \hat G} \hat\nu(\chi)\chi(x)\Bigr)\Bigl(\sum_{\chi\in \hat G} \hat f(\chi)\chi(x)\Bigr)^p \cr
+&= \sum_{\chi_1\in \hat G} \cdots \sum_{\chi_{p+1}\in \hat G}
+\hat f(\chi_1)\cdots\hat f(\chi_p) \hat \nu(\chi_{p+1} )\ex_{x\in G} \chi_1(x)\cdots\chi_{p+1}(x) \cr
+&= \sum_{\chi_1\cdots\chi_{p+1} =\chi_0} \hat f(\chi_1) \cdots \hat f(\chi_p) \hat\nu(\chi_{p+1} )\cr
 &= \hat\nu * {\hat f}^{*p}(\chi_0). \cr
@@
 $$\eqalign{
-\langle \one_{P\setminus T}, f^p \rangle &= \langle \one_P - \one_T, f^p \rangle \cr
-&= \langle \one_P , f^p \rangle - \langle \one_T, f^p \rangle \cr
-&\ge {\eps\over 2} - \ex_{x\in G} \one_T(x)f(x)^p \cr
-&\ge {\eps\over 2} - \ex_{x\in G} 3\eps/4 \cr
-&= {\eps\over 4},\cr
+\langle \one_T, f^p \rangle_\nu &= \langle \one_P - \one_{P\setminus T}, f^p \rangle_\nu \cr
+&\ge \langle\one_P, f^p\rangle_\nu - \langle \one_{P\setminus T}, f^p\rangle_\nu \cr
+&\ge {\eps^p\over 2}- \ex_{x\in G} \one_{P\setminus T} f(x)^p\nu(x)\cr
+&> {\eps^p\over 2}- \ex_{x\in G} (3\eps/4)^p \nu(x)\cr
+&\ge {\eps^p\over 4}, \cr
 }$$
-so $\langle \one_T, f^p \rangle_\nu \ge \eps^p/4$ and we have
+where in the last line we used the fact that $(3/4)^p \le (3/4)^5 < 243/1024 < 4$.
+From this we deduce
 $$\eqalign{
-{\eps\over 4} &\le \langle \one_T,f^p\rangle_\nu \cr
+{\eps^p\over 4} &\le \langle \one_T,f^p\rangle_\nu \cr
 &= \ex_{x\in G} \bigl( \nu(x)^{1/2}\one_T(x)\bigr) \bigl(\nu(x)^{1/2} f(x)^p\bigr) \cr
-&= \Bigl( \ex_{x\in G} \nu(x)\one_T(x)^2\Bigr)^{1/2} \Bigl(\ex_{x\in G}\nu(x) f(x)^{2p}\Bigr)^{1/2} \cr
+&\le \Bigl( \ex_{x\in G} \nu(x)\one_T(x)^2\Bigr)^{1/2} \Bigl(\ex_{x\in G}\nu(x) f(x)^{2p}\Bigr)^{1/2} \cr
 &= \Bigl( \ex_{x\in G} \nu(x)\one_T(x)\Bigr)^{1/2} \Bigl(\ex_{x\in G} \nu(x) |f(x)|^{2p}\Bigr)^{p/(2p)} \cr
@@
 
-Now if $\norm{f}_{2p(\nu)}^p > 2$, then we could take $p' = 2p$, so assume that this norm is at most $2$.
+Now if $\norm{f+1}_{2p(\nu)} > 2$, then we could take $p' = 2p$, so assume that this norm is at most $2$.
 By the triangle inequality, we have
@@
 hence
-$$ \nu(T)^{1/2} 3^p \le {\eps^p\over 4}.$$
-Since we assumed that $p\ge 5$, and since $4 < 1024/243$, we have $4^{1/p} < 4/3$ and thus
+$$ \nu(T)^{1/2} 3^p \ge {\eps^p\over 4}.$$
+Once again using the fact that $4 < 1024/243$, we have $4^{1/p} < 4/3$ and thus
 $$ \nu(T) \ge {\eps^{2p} \over 16 \cdot 3^{2p}} = \biggl( {\eps \over 4^{1/p} \cdot 3} \biggr)^{2p}
@@
 that $\norm{\mu_A\circ\mu_A}_{p'}\ge 1+\eps/4$ for some
-$p' = O\bigl(2\log(2) p\bigr)= O\bigl(\log(1/\gamma)\bigr)$.
+$p' = O\bigl((2/\eps)\log(2/\eps) p\bigr)= O\bigl(\eps^{-1}\log(\eps^{-1})\log(1/\gamma)\bigr)$.
 Let $C(\eps) = \eps^{-1}\log(\eps^{-1})$. By Corollary~{\deprandcorollary},
@@
 Hence if $A\subseteq \FF_q^n$ contains no nontrivial three-term arithmetic progressions, then
-$\alpha \le e^{-\Omega(n^{1/9})} = O\bigl(q^{-\Omega(n^{1/9})}\bigr)$.
+$\alpha \le q^{-\Omega(n^{1/9})}$.
 
@@
 $\langle \mu_A*\mu_A,\mu_C\rangle \ge 1/2$ or there is a subspace $V$ of codimension
-$O\bigl(\log(1/alpha)^8\bigr)$ such that $\max_{x\in G}\;(\one_A*\mu_V)(x)\ge (1+\eps/64)\alpha$.
+$O\bigl(\log(1/\alpha)^8\bigr)$ such that $\max_{x\in G}\;(\one_A*\mu_V)(x)\ge (1+\eps/64)\alpha$.
 
@@
 and taking $q$th logs of both sides yields
-$$n \le \log_q\biggl({2q^{O(\log(1/\alpha)^9)}\over \alpha^2}\biggr) = O\bigl(\log(1/\alpha)^9\bigr).$$
-Letting $C$ be the constant implicit in the $O$-notation, we invert this to obtain the bound
-$$\alpha\le e^{-n^{1/9}/C^{1/9}} = e^{-\Omega(n^{1/9})},$$
+$$n \le \log_q\biggl({2q^{O(\log(1/\alpha)^9)}\over \alpha^2}\biggr) = O\bigl(\log(1/\alpha)^9\bigr)
+= O\bigl(\log_q(1/\alpha)^9\bigr).$$
+Letting $C$ be the constant implicit in the last big-$O$ bound, we invert this to obtain
+$$\alpha\le q^{-n^{1/9}/C^{1/9}} = q^{-\Omega(n^{1/9})},$$
 which is what we wanted.\slug

### 2023-09-28 misc/expo/kelleymeka.tex
Fish

@@
 \def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
+\def\Bohr{\op{\rm Bohr}}
 
@@
 \maketitle{Kelley and Meka's proof of Roth's theorem}{by}{Marcel K. Goh}{23 August 2023}
+
+\floattext4.5 \ninebf Note.
+\ninepoint This exposition of Kelley and Meka's proof closely follows that of Bloom and Sisask,
+but with many extra details supplied. I hope that it will be useful to other students at the Ph.D.-student
+level.
 
@@
 $$\min\bigl\{\mu_{B_1}(A_1), \mu_{B_2}(A_2)\bigr\}
-\ge {1\over 4} \alpha^{2p} \norm{\mu_A\circ\mu_A}_{p(\mu)}^{2p}k.$$
+\ge {1\over 4} \alpha^{2p} \norm{\mu_A\circ\mu_A}_{p(\mu)}^{2p}.$$
 and such that
@@
 $${\bigl\langle (\mu_A\circ\mu_A)^{p'}, \one_{G\setminus S}\bigr\rangle_{\mu} \over
-\norm{\mu_A\circ\mu_A}^{p'}_{p'(\mu)}} \le (1-\eps)^p,$$
+\norm{\mu_A\circ\mu_A}^{p'}_{p'(\mu)}} \le (1-\eps)^{p'},$$
 and since $p'\ge \eps^{-1}\log(\delta^{-1})$, we have
-$$(1-\eps)^p\le e^{-\eps p} \le \delta.$$
+$$(1-\eps)^{p'}\le e^{-\eps p'} \le \delta.$$
 Putting everything together, we have
@@
 
+\advsect Bohr sets
+
+To transfer the ideas of the finite-field proof over to the integer case, we will need the machinery of Bohr
+sets. These are, in some sense, the analogue in general abelian groups to subspaces in $\FF_q^n$.
+
+Let $G$ be a finite abelian group, $\Gamma$ be a nonempty subset of $\hat G$, and let $\nu:\Gamma\to [0,2]$.
+The {\it Bohr set} $B = \Bohr_\nu(\Gamma)$ corresponding to this data is the set
+$$\Bohr_\nu(\Gamma) = \bigl\{x\in G: |1-\gamma(x)| \le \nu(\gamma)\ \hbox{for all}\ \gamma\in \Gamma\bigr\}.$$
+The set $\Gamma$ is called the {\it frequency set} of $B$, and $\nu$ is its {\it width function}.
+We shall also say that $B$ has rank $d$ if $|\Gamma| = d$.
+Bohr sets contain $0$, since $\gamma(0) = 1$ for all characters $\gamma$, and $x\in B$ if and only if $-x\in B$,
+since $|1-\gamma(x)| = |1-\gamma(-x)|$.
+Note that the set $\Bohr_\nu(\Gamma)$ does not uniquely determine the pair $(\Gamma,\nu)$.
+When we write $B'\subseteq B$ for Bohr sets $B'$ and $B$, we mean an inclusion of sets, and do not intend
+to say anything about the corresponding frequency sets or width functions.
+
+Let $B = \Bohr_\nu(\Gamma)$ be a Bohr set and let $\rho>0$. Let $\nu_\rho:\Gamma\to[0,2]$ be the width function
+given by $\nu_\rho(\gamma) = \max\bigl\{\rho\cdot\nu(\gamma),2\bigr\}$. We define the {\it dilate} of
+$B$ by $\rho$
+to be the Bohr set $\Bohr_{\rho\nu}(\Gamma)$. If $\rho\le 1$, $\rho \nu(\gamma)\le \nu$, meaning that
+the condition $|1-\gamma(x)|\le \rho\nu(\gamma)$ is now stronger. so $B_\rho\subseteq B$, and similarly
+$B\subseteq B_\rho$ if $\rho\ge 1$.
+
+We now introduce a definition that characterises when a Bohr set is approximately closed under
+addition. Let $B$ be a Bohr set of rank $d$. We say that $B$ is {\it regular} if for all $\kappa$
+with $|\kappa|\le 1/(100d)$,
+$$\bigl(1-100d|\kappa|\bigr)|B| \le |B_{1+\kappa}| \le \bigl(1+100d|\kappa|\bigr)|B|.$$
+
+The following proposition will be used frequently in the following sections.
+
+\proclaim Proposition \advthm. Let $G$ be a finite abelian group and let $k$ be an integer with
+$\gcd\bigl(k,|G|\bigr) = 1$. If $B = \Bohr_\nu(\Gamma)$ is a regular Bohr set of rank $d$, then
+$k\cdot B$ is also a regular Bohr set of rank $d$.
+
+\proof We will construct the frequency set and width function of $k\cdot B$.
+Note first that if we have $kx = kx'$, then $k(x-x') = 0$. But since $\gcd\bigl(k,|G|\bigr) = 1$,
+$k(x-x') = 0$ implies that $x-x' = 0$, so $x=x'$ and thus the function $x\mapsto kx$ permutes
+$G$. We will write its inverse as $x\mapsto x/k$. For all $\gamma\in\Gamma$,
+let $\gamma_k$ be given by $\gamma_k(x) = \gamma(kx)$, and let $\gamma_{k^{-1}}(x) = \gamma(x/k)$.
+We let
+$$\Gamma' = \{\gamma_{k^{-1}} : \gamma\in \Gamma\}$$
+and let $\nu':\Gamma'\to [0,2]$ be given by $\nu'(\gamma') = \gamma'_k$.
+Then we have a bijection $\gamma \mapsto \gamma'$ such that $\nu(\gamma) = \nu'(\gamma')$,
+so we see that
+$$\eqalign{
+k\cdot B &= \bigl\{ kx\in G : \gamma(x) \le \nu(\gamma)\ \hbox{for all}\ \gamma\in \Gamma\bigr\} \cr
+&= \bigl\{ x\in G : \gamma_{k^{-1}}(x) \le \nu(\gamma)\ \hbox{for all}\ \gamma\in \Gamma\bigr\} \cr
+&= \bigl\{ x\in G : \gamma'(x) \le \nu'(\gamma')\ \hbox{for all}\ \gamma'\in \Gamma'\bigr\} \cr
+&= \Bohr_{\nu'}(\Gamma'), \cr
+}$$
+so $k\cdot B$ is a Bohr set of rank $d$ as well.
+
+To check regularity, it suffices to show that $\bigl| (k\cdot B)_{1+\kappa}\bigr| = |B_{1+\kappa}|$.
+We actually have the stronger fact that
+$$\eqalign{
+(k\cdot B)_{1+\kappa}&=\bigl\{x\in G:\gamma'(x)\le(1+\kappa)\nu'(\gamma')\hbox{ for all }\gamma'\in \Gamma'\bigr\}\cr
+&=\bigl\{x\in G:\gamma_{k^{-1}}(x)\le(1+\kappa)\nu(\gamma)\hbox{ for all }\gamma\in \Gamma\bigr\}\cr
+&=\bigl\{kx\in G:\gamma(x)\le(1+\kappa)\nu(\gamma)\hbox{ for all }\gamma\in \Gamma\bigr\}\cr
+&= k\cdot B_{1+\kappa} .\noskipslug
+}$$
+
 \section References

### 2023-11-01 about/vita.tex
Updates

@@
 
-Additive combinatorics, analytic number theory, random trees, analysis of algorithms.
+Additive combinatorics, enumerative combinatorics, analysis of random discrete structures.
 
@@
 \begingroup\frenchspacing
-\papercount=5
+\papercount=6
 \pubbegin P
+\papitem (with Jonah Saks)
+On the homology of several number-theoretic set families. To appear in
+{\sl Enumerative Combinatorics and Applications} (2024), 17~pp.
+[arXiv:2206.12535]
 \papitem (with Jad Hamdan and Jonah Saks)
@@
 
-\sectheader{Submitted Papers}
-
-\begingroup\parindent=10pt
-\thing An uncertainty principle for M\"obius inversion on posets. {\sl arXiv:2302.02466} (2023), 7~pp.
-\smallskip
-\thing (with Jonah Saks)
-On the homology of several number-theoretic set families. {\sl arXiv:2206.12535} (2022), 16~pp.
-\endgroup
-\endgroup%end the frenchspacing
-
 \goodbreak
@@
 
-\sectheader{Research Experience}
+\goodbreak
 
-\leftright{\bf Probabilistic analysis of branching processes}{\bf May 2020 -- June 2022}
-\leftright{\sl Research group}{\sl McGill University}
-\smallskip
-Research on branching processes, headed by Luc Devroye. Studied estimation problems
-on Galton-Watson trees. Attended informal seminars, discussing
-various topics related to branching
-processes and other topics in probability and combinatorics.
-(Gave a presentation at six of these: one on root estimation in Galton-Watson
-trees, two on generating functions and elementary analytic combinatorics, two on graph regularity, and
-one concerning two of Erd\H os's proofs on prime numbers.)
+\sectheader{Published Short Fiction}
 
-\medbreak
-\leftright{\bf Sorting algorithms on manifolds}{\bf May -- August 2020}
-\leftright{\sl Summer research project}{\sl McGill University}
-\smallskip
-Research project under the supervision of Michael Lipnowski.
-Studied algorithms that investigate the topology of group actions on locally symmetric spaces.
-Work involved writing of code in OCaml and PostScript that generates illustrations of certain quotient
-spaces, in various models of hyperbolic geometry.
-\medbreak
-
-\leftright{\bf Interactive proofs}{\bf September -- December 2019}
-\leftright{\sl Honours research project}{\sl McGill University}
-\smallskip
-Semester-long research project in the Computation and Logic Group, supervised
-by Brigitte Pientka.
-Proved a theorem in constructive logic concerning the formal verification
-of the interactive proof assistant Harpoon and wrote OCaml code as part of
-ongoing work on the functional programming language Beluga.
-\medbreak
-
-\leftright{\bf Bytecode compiler and interpreter}{\bf February -- June 2019}
-\leftright{\sl Individual software project}{\sl Charles University}
-\smallskip
-Semester-long individual software project under the supervision of Adam Dingle.
-Created OPythn, a bytecode compiler and virtual machine for a subset of Python.
-OPythn includes support for lists, dictionaries, named and anonymous functions, objects, and classes.
-It is implemented in OCaml using Ocamllex for lexing and Menhir for parsing.
+\fictioncount=3
+\pubbegin F
+\ficitem ``Senang Diri,'' {\sl Existere} {\bf 42},2 (2023), 17--25.
+\ficitem ``The Vigil,'' {\sl Ricepaper Magazine} (2023), online.
+\ficitem ``Mountain Pass,'' {\sl Prairie Journal of Canadian Literature} {\bf 79} (2023), 48--55.
+\endgroup%end the frenchspacing
 
@@
 \starleftright{\bf After school care supervisor}{\bf January -- June 2023}
-\leftright{\bf Suzuki Charter School}{\sl Edmonton, AB, Canada}
+\leftright{\sl Suzuki Charter School}{\sl Edmonton, AB, Canada}
 \smallskip
@@
 \starleftright{\bf Software developer}{\bf January -- May 2023}
-\leftright{\bf Levven}{\sl Edmonton, AB, Canada}
+\leftright{\sl Levven}{\sl Edmonton, AB, Canada}
 \smallskip
@@
 \starleftright{\bf Visiting teaching assistant}{\bf July 2022}
-\leftright{\bf Bocconi University}{\sl Como, CO, Italy}
+\leftright{\sl Bocconi University}{\sl Como, CO, Italy}
 \smallskip
@@
 
-\leftright{\bf Mentor}{\bf January -- April 2022}
-\leftright{\bf Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
+\leftright{\bf DRP Mentor}{\bf January -- April 2022}
+\leftright{\sl Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
 \smallskip
@@
 
-\sectheader{Published Short Fiction}
-
-\begingroup\frenchspacing
-\fictioncount=3
-\pubbegin F
-\ficitem ``Senang Diri,'' {\sl Existere} {\bf 42},2 (2023), 17--25.
-\ficitem ``The Vigil,'' {\sl Ricepaper Magazine} (2023), online.
-\ficitem ``Mountain Pass,'' {\sl Prairie Journal of Canadian Literature} {\bf 79} (2023), 48--55.
-\endgroup%end the frenchspacing
-
 \sectheader{Skills}
@@
 {\bf Programming Languages}\par
-\CEE, OCaml, Python, Java, PostScript, Haskell, Scheme Lisp,  Standard ML, {\tt CWEB}, {\tt MIXAL}, MIPS Assembly.
+\CEE, OCaml, Python, Java, PostScript, Haskell, Scheme Lisp,  Standard ML, {\tt CWEB}, {\tt MIXAL},
+{\mc MIPS} Assembly.
 \medbreak

### 2023-11-01 misc/expo/kelleymeka.tex
Updates

@@
 
-\maketitle{Kelley and Meka's proof of Roth's theorem}{by}{Marcel K. Goh}{23 August 2023}
+\maketitle{Kelley and Meka's proof of Roth's theorem}{by}{Marcel K. Goh}{12 October 2023}
 
@@
 \parenproclaim Lemma~{\advthm} (H\"older lifting). Let $\eps\ge 0$ and let $A$ and $C$ be subsets of
-a finite abelian group $G$, where $C$ has relative density $\gamma$. Then at least one of
+a finite abelian group $G$, where $C$ has relative density at least $\gamma$. Then at least one of
 the following two statements holds.
@@
 
-This lemma tells us that if $\langle \mu_A*\mu_A,\mu_C\rangle \ge 1/2$, then $\norm{\mu\circ \mu_A-1}_p \ge 1/4$
+This lemma tells us that if $\langle \mu_A*\mu_A,\mu_C\rangle \ge 1/2$, then
+$\norm{\mu_A\circ \mu_A-1}_p \ge 1/4$
 for some $p = O\bigl(\log(1/\gamma)\bigr)$. This information can then be fed to the following general lemma.
@@
 $$\eqalign{
-\bignorm{(\mu_A\circ \mu_A)^p, f}_{p(\mu)}^p
+\bignorm{\mu_A\circ \mu_A}_{p(\mu)}^p
 &= {|G|^2\over \alpha^{2p}|B_1|\cdot|B_2|} \ex_{s\in G^p} 
@@
 $$\eqalign{
-\ex_{s\in G^p} \indic{\alpha_1(s)\alpha_2<M^2} \alpha_1(s)\alpha_2(s)
+\ex_{s\in G^p} \indic{\alpha_1(s)\alpha_2(s)<M^2} \alpha_1(s)\alpha_2(s)
  &< \ex_{s\in G^p} M \sqrt{\alpha_1(s)\alpha_2(s)} \cr
@@
 \proof Let $p'$ be the smallest even integer at least $p+\eps^{-1}\log(\delta^{-1})$.
-By the previous lemma applied to the set
+By the previous lemma applied to the function
 $\one_{G\setminus S}$, there exist sets $A_1\subseteq B_1$ and $A_2\subseteq B_2$ with densities

### 2023-12-17 misc/expo/entropy.tex
Ahoy

\input fontmac
\input mathmac

\def\eps{\epsilon}
\def\FF{{\bf F}}
\def\bar#1{\overline{#1}}
\def\hat#1{\widehat{#1}}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\Bohr{\op{\rm Bohr}}
\def\Eta{\op{\bf H}}
\def\II{\op{\bf I}}
\def\given{\mathbin{|}}

\widemargins
\bookheader{ENTROPY AND THE FREIMAN--RUZSA THEOREM}{MARCEL K. GOH}

\maketitle{Entropy and the polynomial Freiman--Ruzsa theorem}{by}{Marcel K. Goh}{23 November 2023}

\floattext4.5 \ninebf Note.
\ninepoint This is a gentle introduction to the notion of entropy as it is used
in combinatorics, with a view towards understanding the new proof of the polynomial Freiman--Ruzsa
conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. The preliminary portion of
these notes are largely transcribed from lectures given by W.~T.~Gowers.

\bigskip

\advsect The Khintchine--Shannon axioms

Let $X$ be a discrete random variable. Its entropy $\Eta\{X\}$ is a real number
(or $\infty$) that measures the ``information content'' of $X$. For example, if $X$ is
a constant random variable, then $\Eta\{X\}$ should be zero (we do not gain any information
from knowing the value of $X$), and if $X$ is
uniformly distributed on $\{0,1\}^n$, then $\Eta\{X\}$ should be proportional to $n$, since $X$ is
determined by
$n$ bits of information. It satisfies the following axioms, which are
sometimes called the Khinchine--Shannon axioms.
\medskip
\item{a)} ({\it Invariance.}) If $X$ takes values in $A$, $Y$ takes values in $B$,
$\phi:A\to B$ is a bijection, and $\pr\{Y = \phi(a)\} = \pr\{X = a\}$ for all $a\in A$,
then $\Eta\{X\} = \Eta\{Y\}$.
\smallskip
\item{b)} ({\it Extensibility.}) If $X$ takes values in $A$ and $Y$ takes values in $B$ for
a set $B$ such that $A\subseteq B$, and furthermore $\pr\{Y=a\} = \pr\{X=a\}$ for all $a\in A$,
then $\Eta\{X\} = \Eta\{Y\}$.
\smallskip
\item{c)} ({\it Continuity.}) The quantity $\Eta\{X\}$ depends continuously on the probabilities
$\pr\{X=a\}$.
\smallskip
\item{d)} ({\it Maximisation.}) Over all possible random variables $X$ taking values in a finite
set $A$, the quantity $\Eta\{X\}$ is maximised for the uniform distribution.
\smallskip
\item{e)} ({\it Additivity.})
For $X$ taking values in $A$ and $Y$ taking values in $B$, we have the formula
$$ \Eta\{X,Y\} = \Eta\{X\} + \Eta\{ Y \given X\},$$
where $\Eta\{X,Y\} = \Eta\bigl\{ (X,Y)\bigr\}$ and
$$ \Eta\{Y\given X\} = \sum_{x\in A} \pr\{X=x\} \Eta\{ Y\given X = x\}.$$
\medskip

For now, we shall take it on faith that there really exists a function on random variables satisfying
these axioms. Later on, when we prove this, we will find that actually, the axioms only define entropy
up to a multiplicative constant, so we shall add the following axiom.

\medskip
\item{f)} ({\it Normalisation.}) If $X$ is uniformly distributed on $\{0,1\}$, then
$\Eta\{X\} = \log_2(e)$.
\medskip

It is actually more common (for obvious reasons, especially in computer science)
for one to set the entropy of a uniform random variable on $\{0,1\}$
to $1$, but we shall
follow the convention of Gowers et al., since the eventual goal of these notes is to understand their proof
of the polynomial Freiman--Ruzsa conjecture.

Immediate from the definition of the conditional entropy $\Eta\{Y\given X\}$ given
in the additivity axiom is that if
$X$ and $Y$ are independent, then $\Eta\{Y\given X\} = \Eta\{Y\}$. But since the uniform distribution
on $\{0,1\}^n$ is the joint distribution $(X_1,\ldots,X_n)$ where the $X_i$ are independent and
uniformly distributed on $\{0,1\}$, we can additionally use the normalisation axiom to conclude that
$$\Eta\{X\} = \Eta\{X_1,\ldots,X_n\} = \sum_{i=1}^n \Eta\{X_i\} = \log_2(e)\cdot n,$$
by induction. Induction also yields the chain rule
$$\Eta\{X_1,\ldots,X_n\} = \Eta\{X_1\} + \Eta\{X_2\given X_1\} + \cdots+\Eta\{X_n \given X_1,\ldots,X_{n-1}\}.$$

The first proposition establishes the intuitive fact
that the entropy of a uniform random variable supported on a set $A$ is at most
the entropy of a uniform random variable supported on a superset $B$ of $A$.

\proclaim Proposition \advthm. Let $A\subseteq B$ with $B$ finite,
let $X$ be uniformly distributed on $A$, and let $Y$ be uniformly distributed on $B$.
Then $\Eta\{X\} \le \Eta\{Y\}$, with equality if and only if $A = B$.

\proof

The next proposition is that $\Eta\{X\}$ is always nonnegative. The following
proof is attributed to S.~Prendiville.

\proclaim Proposition \advthm. For any random variable $X$ taking values in a finite set $A$,
we have $\Eta\{X\} \ge 0$.

\proof First we consider the case that there exists $n\in \NN$ such that for all $x\in A$, there is some
$c_x$ such that $\pr\{X = x\} = c_x/n$. Let $Y$ be uniform on $[n]$ and let $E_x$ be a partition
of $[n]$ such that $|E_x| = c_x$ for all $x\in X$. Then let $Z$ be given by $Z = x$ if $Y\in E_x$.
This random variable $Z$ takes values in $A$ and has the exact same distribution as $X$, so
by the invariance axiom, $\Eta\{X\} = \Eta\{Z\}$.
This means that $Z = f(Y)$ for some function $f$, meaning that there is a bijection between
values $y$ taken by $Y$ and values $\bigl(y, f(y)\bigr)$ taken by $(Y,Z)$. So we may apply the invariance
axiom again to conclude that $\Eta\{Y,Z\} = \Eta\{Y\}$.

Now by the additivity axiom, $\Eta\{Y,Z\} = \Eta\{Y\given Z\} + \Eta\{Z\}$, so we
have $\Eta\{Y\} = \Eta\{Y\given Z\} + \Eta\{Z\}$.

If $X$ is a random variable such that $X = f(X)$ for some random variable $Y$, then we say
that $X$ is {\it determined by} $Y$ or $Y$ {\it determines} $X$. The following proposition formalises
the idea that we get more information from $Y$ than we get from $X$.

\proclaim Proposition \advthm. Let $X$ and $Y$ be random variables such that $Y$ determines $X$. Then
$$\Eta\{ X\} \le \Eta\{Y\}.$$

\proof The

For random variables $X$ and $Y$, the {\it mutual information} $\II\{X : Y\}$ is defined
by the equivalent formulas
$$\eqalign{
\II\{ X : Y\} &= \Eta\{X\} + \Eta\{Y\} - \Eta\{X,Y\} \cr
&= \Eta\{X\} - \Eta\{X\given Y\} \cr
&= \Eta\{Y\} - \Eta\{Y\given X\}. \cr
}$$
It measures, roughly speaking, how much information one can get from one variable
by looking at the other one. From the formula it is clear that $\II\{X : Y\}  = 0$
if and only if $X$ and $Y$ are independent, and we also have the
inequality $\Eta\{X \given Y\} \le \Eta\{X\}$. Given a triple $(X,Y,Z)$ of random variables,
we can apply this with $(X\given Z = z)$ to obtain the inequality
$$ \Eta\{ X\given Y,Z\} \le \Eta\{X\given Z\},$$
which is called {\it submodularity}. An equivalent statement is
$$\Eta\{X,Y,Z\} + \Eta\{Z\} \le \Eta\{X,Z\} + \Eta\{Y,Z\}.$$
If $Z$ takes values in a set $C$, the {\it conditional mutual information} is defined by
$$\eqalign{
\II\{ X : Y \given Z\} &= \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
&= \Eta\{ X\given Z\} + \Eta\{Y\given Z\} - \Eta\{ X, Y \given Z \} \cr
&= \Eta\{ X, Z\} -2\Eta\{ Z\} + \Eta\{Y,Z\} - \Eta\{ X, Y ,Z \} + \Eta\{Z\}\cr
&= \Eta\{ X, Z\} + \Eta\{Y,Z\} - \Eta\{ X, Y ,Z \} - \Eta\{Z\},\cr
}$$
so we see that submodularity is equivalent to the statement $\II\{X : Y\given Z\} \ge 0$.
Analogously to the unconditional case, we have equality if and only if $X$ and $Y$ are
independent when conditioned on $Z$.

\medskip\boldlabel The formula for entropy.
Now we finally give the formula for the entropy of a random variable. For a discrete random
variable $X$, we let $p_X : S\to [0,1]$ denote the probability mass function of $X$, given by
$p_X(x) = \pr\{X = x\}$ for all $x\in S$. Unless subscripted, $\log$ always denotes the natural logarithm.

\proclaim Proposition \advthm. Let $X$ be a random variable taking values in a finite set $S$.
Then $\Eta\{X\}$ satisfies axioms (a) through (e) as well as the additional normalisation axiom (f)
if and only if
$$\Eta\{X\} = \ex\bigl\{ 1/\log p_X(X) \bigr\} = \sum_{x\in S} p_X(x) \log {1\over p_X(s)},$$
where we adopt the convention that $0\log(1/0) = 0$.

\proof [TODO: Prove this.]\slug

Equipped now with a formula for the entropy $\Eta\{X\}$ of a random variable $X$, we can prove further useful
facts. We saw in the previous section that the entropy of a uniform random variable whose range is $\{0,1\}^n$
is $\log_2(e)\cdot n = \log(2^n)$, and we know, by the maximisation axiom, that this is the largest
the entropy of any $X$ taking values in $\{0,1\}^n$ can be. The following proposition generalises
this notion.

\proclaim Proposition \advthm. For any random variable $X$ taking values in a finite set $S$,
we have
$$\Eta\{X\} \le \log |S|.$$

\proof By Jensen's inequality, we have
$$\eqalign{
\Eta\{X\} &= \ex\bigl\{ \log(1/p_X(X)\bigr\} \cr
&\le \log \ex\bigl\{ 1/p_X(X)\bigr\} \cr
&= \log \sum_{x\in S} \pr\{X = x\} {1\over \pr\{X = x\}} \cr
&= \log |S|.\noskipslug \cr
}$$

\section References

\bye

### 2023-12-26 misc/expo/additive.tex
Furstenberg--Sarkozy lemma

@@
 
+\advsect The Furstenberg--S\'ark\"ozy theorem
+
+In this section, we'll prove a bound on the size of {\it square-difference-free sets}. These are sets
+$A\subseteq \NN$ such that $A-A$ does not contain any of the squares $\{1,4,9,16,\ldots\}$.
+The proof, which is due to B.~Green, T.~Tao, and T.~Ziegler,
+makes use of a density increment argument similar to the one used in the proof of
+Roth's theorem, but does not use any Fourier analysis. We begin with the following lemma, which allows
+us to increment the density of a set containing a square in its difference set.
+
+\proclaim Lemma~\advthm. There exists a real constant $c>0$ such that the following holds.
+Suppose that there is $N\in \NN$ and $A\subseteq [N]$ of density $\delta = |A|/N$ such that
+$A-A$ does not contain any positive square numbers. Then there exists an arithmetic progression
+$P$ of length $\lfloor N^{1/3}\rfloor$ such that $|A\cap P| \ge (\delta + c\delta^3)|P|$.
+
+Suppose that for
+some $\delta_0>0$ such that for arbitrarily large $N\in \NN$, we can find $A\subseteq [N]$
+of size $|A|\ge \delta_0 N$ such that $A-A$ does not contain a square. Then for
+arbitrarily large $N'\in \NN$, there exists a subset
+$A'\subseteq [N']$ such that $|A'|\ge (\delta_0 + c{\delta_0}^3)N'$ and $A'-A'$ does not contain a square.
+
+\proof Suppose $\delta_0>0$ is such that for arbitrarily large $N$ we can find $A\subseteq [N]$
+of density $\delta = |A|/N \ge \delta_0$ such that $A-A$ does not contain a square. Then for all
+$n\in [N]$, we have
+$$ \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]} \one_A(n) \one_A\bigl( n+(r+h)^2\bigr) = 0.$$
+Now for $n\in [N]$, note that $n+(r+h)^2$ is in $[N]$ for all
+$n\le N-(N^{1/3}-N^{1/100})^2 = N-O(N^{2/3})$. This means that
+$$ \ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]}
+\one_{[N]}\bigl( n+(r+h)^2\bigr) = 1 - O(N^{-1/3}).$$
+By further requiring that $n\in A$ inside the expectation, we have
+$$ \ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]}
+\one_A(n)\one_{[N]}\bigl( n+(r+h)^2\bigr) = \delta - O(N^{-1/3}),$$
+since the proportion of $n\in [N]$ that are also in $A$ is $\delta$.
+Lastly, by a simple change of variables we have
+$$\eqalign{
+\ex_{n\in [N]} \ex_{r\in [N^{1/3}]} &\ex_{h\in [N^{1/100}]} \one_A\bigl(n+(r+h)^2\bigr) \cr
+&=\ex_{m\in A} \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]} \one_{[N]}\bigl(m-(r+h)^2\bigr) \cr
+&= \delta - O(N^{-1/3}). \cr
+}$$
+Set $f = \one_A - \delta\one_{[N]}$. From the above computations we deduce that
+$$\eqalign{
+\ex_{n\in [N]} &\ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]}
+f(n) f\bigl( n+(r+h)^2\bigr) \cr
+&= 
+\ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]}\cr
+&\qquad\qquad\Bigl(\one_A(n)\one_A\bigl(n+(r+h)^2\bigr)
+-\delta\one_A(n)\one_{[N]}\bigl(n+(r+h)^2\bigr) \cr
+&\qquad\qquad\qquad\qquad\qquad- \delta\one_A\bigl(n+(r+h)^2\bigr)
++ \delta^2\one\bigl(n+(r+h)^2\bigr)\Bigr) \cr
+&= 0 - \delta^2 - \delta^2 +\delta^2 + O(N^{-1/3}) \cr
+&= -\delta^2 + O(N^{-1/3}) .\cr
+}$$
+The triangle inequality then gives
+$$ \ex_{n\in [N]} \bigl| f(n)\bigr| \ex_{r\in [N^{1/3}]} \Bigl| \ex_{h\in [N^{1/100}]} f\bigl( n+(r+h)^2\bigr)
+\Bigr| \ge \delta^2-O(N^{-1/3}) \gg \delta^2,$$
+where here and in the rest of the proof, the Vinogradov notation hides an absolute constant that works
+for all $N\in \NN$. We also have
+$$\eqalign{
+\ex_{n\in [N]} \bigl| f(n)\bigr| ^2
+&= \ex_{n\in [N]} \Bigl(\one_A(n) - 2\delta \one_A(n)\one_{[N]}(n) + \delta^2 \one_{[N]}(n)\Bigr)\cr
+&= \delta-\delta^2, \cr
+}$$
+so by the Cauchy--Schwarz inequality,
+$$\eqalign{
+\delta^4-O(N^{1/3}) &\le
+\biggl(\ex_{n\in [N]} \bigl| f(n)\bigr| \ex_{r\in [N^{1/3}]} \Bigl| \ex_{h\in [N^{1/100}]} f\bigl( n+(r+h)^2\bigr)
+\Bigr|\biggr)^2 \cr
+&\le \ex_{n\in [N]} \bigl| f(n) \bigr|^2 
+\ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \Bigl| \ex_{h\in N^{1/100}} f\bigl(n+(r+h)^2\bigr)\Bigr|^2\cr
+&= (\delta-\delta^2)
+\ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \Bigl| \ex_{h\in N^{1/100}} f\bigl(n+(r+h)^2\bigr)\Bigr|^2,\cr
+}$$
+and therefore
+$$
+\ex_{n\in [N]} \ex_{r\in [N^{1/3}]} \Bigl| \ex_{h\in N^{1/100}} f\bigl(n+(r+h)^2\bigr)\Bigr|^2
+\gg {\delta^4\over \delta-\delta^2} \ge \delta^3.$$
+We can rewrite this as
+$$
+\ex_{h,h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]} 
+f\bigl(n+(r+h)^2\bigr) f\bigl(n+(r+h')^2\bigr) \gg \delta^3.
+$$
+Replacing $n$ by $n-(r+h)^2$ in the innermost expectation introduces an error term of at most $O(N^{-1/3})$
+and deleting the diagonal case $h=h'$ introduces an error term of at most $O(N^{-1/100})$,
+so we have
+$$\eqalign{
+\ex_{h,h'\in N^{1/100}}&\ex_{n\in [N]}\ex_{r\in [N^{1/3}]} 
+f\bigl(n+(r+h)^2\bigr) f\bigl(n+(r+h')^2\bigr) \cr
+&\le
+\ex_{h,h'\in N^{1/100}}\Bigl|\ex_{n\in [N]}\ex_{r\in [N^{1/3}]} 
+f\bigl(n+(r+h)^2\bigr) f\bigl(n+(r+h')^2\bigr) \Bigr|\cr
+&\le
+\ex_{h,h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}  \cr
+&\qquad\qquad f(n) f\bigl(n-(r+h)^2+(r+h')^2\bigr) + O(N^{-1/3})\cr
+&= \ex_{h,h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}  \cr
+&\qquad\qquad f(n) f\bigl(n- (h'-h)(2r+h'-h)\bigr) + O(N^{-1/3})\cr
+&= \ex_{h,h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}  \cr
+&\qquad\qquad f(n) f\bigl(n- (h'-h)(2r+h'-h)\bigr) + O(N^{-1/3})\cr
+&= \ex_{h\ne h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]} \cr
+&\qquad\qquad f(n) f\bigl(n- (h'-h)(2r+h'-h)\bigr) + O(N^{-1/100}),\cr
+}$$
+whence
+$$\ex_{h\ne h'\in N^{1/100}}\Bigl|\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}
+f(n) f\bigl(n- (h'-h)(2r+h'-h)\bigr) \Bigr| \gg \delta^3 $$
+by the triangle inequality. From this we see that
+there are distinct $h$ and $h'$ in $[N^{1/100}]$ with
+$$\Bigl|\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}
+f(n) f\bigl(n- (h'-h)(2r+h'-h)\bigr)\Bigr| \gg \delta^3$$
+Here another application of the triangle inequality and the fact that $\bigl|f(n)\bigr| \le 1$ together
+give
+$$\ex_{n\in [N]} \Bigl|\ex_{r\in [N^{1/3}]}
+f\bigl(n- (h'-h)(2r+h'-h)\bigr)\Bigr|\gg \delta^3.$$
+Now we will perform another shift on $n$, this time by $(h'-h)(h'+h) = O(N^{1/50})$. In the expectation
+over $n$, this produces an error term of at most $O(N^{-49/50})$, which we can hide behind the Vinogradov notation
+as well. Letting $d = 2(h'-h)$, we have
+$$\eqalign{
+\ex_{n\in [N]} \Bigl| &\ex_{r\in [N^{1/3}]} f(n+dr)\Bigr| \cr
+&= \ex_{n\in [N]} \Bigl| \ex_{r\in [N^{1/3}]} f\bigl(n+2(h'-h)r\bigr)\Bigr| \cr
+&\ge \ex_{n\in [N]} \Bigl| \ex_{r\in [N^{1/3}]} f\bigl(n+(h'-h)(2r+h+h')\bigr)\Bigr| - O(N^{-49/50}) \cr
+&\gg \delta^3 \cr
+}$$
+But recall that we have
+$$\sum_{n\in [N]} f(n) = 0,$$
+so since $d \le 2N^{1/100}$ and $r\le N^{1/3}$, we have
+$$\sum_{n\in [N]} f(n+dr) = O(N^{1/3+ 1/100})$$
+for all $r\in [N^{1/3}]$, and hence
+$$\ex_{n\in [N]} \ex_{f\in [N^{1/3}]} f(n+dr) = O(N^{-2/3+ 1/100}).$$
+Then since $\bigl(x+|x|\bigr)/2 = \max(x,0)$ for all $x\in \RR$, we have
+$$ \eqalign{
+\ex_{n\in [N]} \max\Bigl( \ex_{r\in [N^{1/3}]} &f(n+dr), 0\Bigr) \cr
+&\gg {\delta^3 + O(N^{-2/3+1/100})\over 2} \cr
+&\gg \delta^3.\cr
+}$$
+By artificially zeroing out all terms corresponding to the largest $2N^{1/3+1/100}$ elements of the interval $[N]$,
+we introduce an error
+term of $N^{-2/3+1/100}$, which disappears behind the asymptotic notation.
+Then by the pigeonhole principle we may find $n\in [N-2N^{1/3+1/100}]$ such that
+$$\ex_{r\in [N^{1/3}]} f(n+dr) \gg \delta^3.$$
+Let $P\subseteq [N]$ be the arithmetic progression $\bigl\{n + dr : r\in [N^{1/3}]\bigr\}$ and let
+$c$ be the absolute constant implied by the Vinogradov notation. We have
+$$\ex_{x\in P} f(x) \ge c\delta ^3,$$
+and can unravel the definition of $f$ to obtain
+$${|A\cap P|\over |P|} \ge \delta + c\delta^3,$$
+which is what we wanted.\slug
+
 \section References

### 2023-12-26 misc/expo/entropy.tex
Furstenberg--Sarkozy lemma

@@
 
-\maketitle{Entropy and the polynomial Freiman--Ruzsa theorem}{by}{Marcel K. Goh}{23 November 2023}
+\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{23 November 2023}
 
@@
 \ninepoint This is a gentle introduction to the notion of entropy as it is used
-in combinatorics, with a view towards understanding the new proof of the polynomial Freiman--Ruzsa
+in additive combinatorics, with a view towards understanding the new proof of the polynomial Freiman--Ruzsa
 conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. The preliminary portion of
@@
 
-Immediate from the definition of the conditional entropy $\Eta\{Y\given X\}$ given
-in the additivity axiom is that if
-$X$ and $Y$ are independent, then $\Eta\{Y\given X\} = \Eta\{Y\}$. But since the uniform distribution
-on $\{0,1\}^n$ is the joint distribution $(X_1,\ldots,X_n)$ where the $X_i$ are independent and
-uniformly distributed on $\{0,1\}$, we can additionally use the normalisation axiom to conclude that
-$$\Eta\{X\} = \Eta\{X_1,\ldots,X_n\} = \sum_{i=1}^n \Eta\{X_i\} = \log_2(e)\cdot n,$$
-by induction. Induction also yields the chain rule
+Notationally, we would expect that $\Eta\{Y\given X\} = \Eta\{Y\}$ if $X$ and $Y$ are independent.
+This is the first proposition we will carefully prove using the axioms above.
+
+\proclaim Let $X$ and $Y$ be independent random variables. Then
+$\Eta\{Y\given X\} = \Eta\{Y\}$ and consequently
+$\Eta\{X,Y\} = \Eta\{X\} + \Eta\{Y\}$.
+
+\proof Suppose $X$ takes values in a finite set $A$. Then for all $x\in A$, the distribution of
+$Y$ and $Y$ given that $X = x$ is the same, so
+$$\Eta\{Y \given X\} = \sum_{x\in A} \pr\{X = x\} \Eta\{Y\given X = x\} = \sum_{x\in A} \pr\{X = x\} \Eta\{Y\}
+=\Eta\{Y\}.$$
+The second version of the statement is a consequence of the additivity axiom.\slug
+
+We will sometimes use the notation $X^n$ to denote the vector $(X_1,\ldots,X_n)$ where the $X_i$
+are independent copies of the random variable $X$.
+
+\proclaim Corollary \advthm. For a discrete random variable $X$, we have $\Eta\{X^n\} = n\Eta\{X\}$.
+
+\proof Perform induction on $n$ and use the previous proposition.\slug
+
+In the case where $X$ is the uniform distribution on $\{0,1\}$, and $X_1,\ldots,X_n$ are independent
+and distributed as $X$,
+we can additionally use the normalisation axiom to conclude that
+$$\Eta\{X^n\} = \Eta\{X_1,\ldots,X_n\} = \sum_{i=1}^n \Eta\{X_i\} = \log_2(e)\cdot n,$$
+by induction. For general random variables $X_1,\ldots,X_n$, induction also yields the {\it chain rule}
 $$\Eta\{X_1,\ldots,X_n\} = \Eta\{X_1\} + \Eta\{X_2\given X_1\} + \cdots+\Eta\{X_n \given X_1,\ldots,X_{n-1}\}.$$
 
-The first proposition establishes the intuitive fact
+Next we show that a constant random variable has zero entropy. This reflects the idea that we get
+no information from sampling a random variable if it always takes the same value no matter what.
+
+\proclaim Proposition \advthm. Let $X$ be a discrete random variable.
+Then $\Eta\{X\} = 0$ if and only if it takes exactly one value.
+
+\proof First suppose that $X$ takes only one value.
+Let $a$ be the value of $X$ such that $\pr\{X=a\} = 1$.
+Then $(X,X)$ equals $(a,a)$ with probability $1$ as well,
+so $\Eta\{X\} = \Eta\{X,X\}$ by the invariance axiom. But it can easily be checked that
+$X$ and $(X,X)$ are independent (we have, for example, $\pr\bigl\{X = a, (X,X)= (a,a)\bigr\} =
+\pr\{X = a\}\pr\bigl\{(X,X) = (a,a)\bigr\}$), so $\Eta\{X,X\} = 2\Eta\{X\}$. Thus we conclude
+that $\Eta\{X\} = 0$.
+
+Now suppose that $X$ takes more than one value; let $A$ be the set of $a$ such that
+$\pr\{X = a\} > 0$.
+\slug
+
+Now we establish the intuitive fact
 that the entropy of a uniform random variable supported on a set $A$ is at most
@@
 
-\proof
+\proof By the extensibility axiom, $\Eta\{X\}$ is not affected if we regard $X$ as a function
+taking values in $B$. Then by the maximisation axiom, $\eta\{X\}\le \eta\{Y\}$, since $Y$ is
+uniform on $B$. If $A = B$, then it is clear that $\Eta\{X\} = \Eta\{Y\}$, since $X$ and $Y$
+are the same random variable.
+Now say $|A| = m$ and $|B| = n$ with $m<n$. If $m=1$, then by the previous proposition
+we have $\Eta\{X\} = 0$, and $Y$ takes at least two values,
 

### 2023-12-27 misc/expo/additive.tex
Lisa

@@
 
-\proclaim Lemma~\advthm. There exists a real constant $c>0$ such that the following holds.
+\proclaim Lemma~\advthm.
 Suppose that there is $N\in \NN$ and $A\subseteq [N]$ of density $\delta = |A|/N$ such that
 $A-A$ does not contain any positive square numbers. Then there exists an arithmetic progression
-$P$ of length $\lfloor N^{1/3}\rfloor$ such that $|A\cap P| \ge (\delta + c\delta^3)|P|$.
-
-Suppose that for
-some $\delta_0>0$ such that for arbitrarily large $N\in \NN$, we can find $A\subseteq [N]$
-of size $|A|\ge \delta_0 N$ such that $A-A$ does not contain a square. Then for
-arbitrarily large $N'\in \NN$, there exists a subset
-$A'\subseteq [N']$ such that $|A'|\ge (\delta_0 + c{\delta_0}^3)N'$ and $A'-A'$ does not contain a square.
-
-\proof Suppose $\delta_0>0$ is such that for arbitrarily large $N$ we can find $A\subseteq [N]$
-of density $\delta = |A|/N \ge \delta_0$ such that $A-A$ does not contain a square. Then for all
-$n\in [N]$, we have
+$P$ of length $\lfloor N^{1/3}\rfloor$ and step size at most $2N^{1/100}$
+such that $|A\cap P| \ge (\delta + c\delta^3)|P|$, where $c$ does not depend on $\delta$ or $N$.
+
+\proof By hypothesis, $n\in [N]$, we have
 $$ \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]} \one_A(n) \one_A\bigl( n+(r+h)^2\bigr) = 0.$$
@@
 &\qquad\qquad f(n) f\bigl(n-(r+h)^2+(r+h')^2\bigr) + O(N^{-1/3})\cr
-&= \ex_{h,h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}  \cr
-&\qquad\qquad f(n) f\bigl(n- (h'-h)(2r+h'-h)\bigr) + O(N^{-1/3})\cr
 &= \ex_{h,h'\in N^{1/100}}\ex_{n\in [N]}\ex_{r\in [N^{1/3}]}  \cr
@@
 for all $r\in [N^{1/3}]$, and hence
-$$\ex_{n\in [N]} \ex_{f\in [N^{1/3}]} f(n+dr) = O(N^{-2/3+ 1/100}).$$
+$$\ex_{n\in [N]} \ex_{r\in [N^{1/3}]} f(n+dr) = O(N^{-2/3+ 1/100}).$$
 Then since $\bigl(x+|x|\bigr)/2 = \max(x,0)$ for all $x\in \RR$, we have
@@
 
+Recall that at the analogous point in the proof of Roth's theorem, we were able to shift and
+rescale our progression to become an interval, and then iterate the argument until the density
+increased above $1$. Unfortunately,
+now we can't simply scale the progression $P$ to transform it into a {\it bona fide} interval, since then
+square differences between elements in $A$ may no longer be square and vice versa. For
+this reason, we will need another lemma that is suited for iteration.
+
+\proclaim Lemma~\advthm.
+Suppose that there is $N\in \NN$ and $A\subseteq [N]$ of density $\delta = |A|/N$ such that
+$A-A$ does not contain any positive square numbers. Then there is a set $A'\subseteq [N^{1/4}]$
+with $A' \ge (\delta + c'\delta^3) N^{1/4}$, where $c$ is an absolute constant, such that $A'-A'$
+also does not contain any squares.
+
+\proof By the previous lemma, there is a progression $P\subseteq [N]$ of length $\lfloor N^{1/3}$
+and step size $1\le d \le 2N^{1/100}$ such that $|A\cap P|/|P| \ge \delta + c\delta^3$ for an absolute
+constant $c$.
+
 \section References

### 2023-12-31 misc/expo/additive.tex
2024 tomorrow

@@
 
-\advsect Basics of discrete Fourier analysis
+\advsect Rudiments of discrete Fourier analysis
 
@@
 \parenproclaim Theorem {\advthm} (Roth, {\rm 1953}). For every $\delta>0$ there exists
-$N\le \exp\bigl(\exp(300/\delta)\bigr)$ such that
+$N\le \bigl\lceil\exp\bigl(\exp(300/\delta)\bigr)\bigr\rceil$ such that
 every $A\subseteq [N]$ with $|A|\ge \delta N$ contains an arithmetic progression of length $3$.
@@
 
+We can flip the roles of $\delta$ and $N$ to arrive at the following statement,
+equivalent to what we just proved: {\sl Any subset of $[N]$
+of size at least $300/\log\log N$ contains a nontrivial arithmetic progression of length $3$.}
+
 \advsect The Furstenberg--S\'ark\"ozy theorem
@@
 $A-A$ does not contain any positive square numbers. Then there exists an arithmetic progression
-$P$ of length $\lfloor N^{1/3}\rfloor$ and step size at most $2N^{1/100}$
+$P$ of length $\lfloor N^{1/3}\rfloor$
 such that $|A\cap P| \ge (\delta + c\delta^3)|P|$, where $c$ does not depend on $\delta$ or $N$.
 
-\proof By hypothesis, $n\in [N]$, we have
+The exponents $1/3$ and $1/100$ in this statement are not really important, and the proof
+may be modified to replace these with other small constants if desired.
+
+\proof By hypothesis, for any $n\in [N]$ we have
 $$ \ex_{r\in [N^{1/3}]} \ex_{h\in [N^{1/100}]} \one_A(n) \one_A\bigl( n+(r+h)^2\bigr) = 0.$$
@@
 
-Recall that at the analogous point in the proof of Roth's theorem, we were able to shift and
+Recall that at this analogous point in the proof of Roth's theorem, we were able to shift and
 rescale our progression to become an interval, and then iterate the argument until the density
@@
 square differences between elements in $A$ may no longer be square and vice versa. For
-this reason, we will need another lemma that is suited for iteration.
-
+this reason, we will need another lemma.
+
+\edef\lemfurstenbergsarkozy{\the\thmcount}
 \proclaim Lemma~\advthm.
-Suppose that there is $N\in \NN$ and $A\subseteq [N]$ of density $\delta = |A|/N$ such that
-$A-A$ does not contain any positive square numbers. Then there is a set $A'\subseteq [N^{1/4}]$
-with $A' \ge (\delta + c'\delta^3) N^{1/4}$, where $c$ is an absolute constant, such that $A'-A'$
-also does not contain any squares.
-
-\proof By the previous lemma, there is a progression $P\subseteq [N]$ of length $\lfloor N^{1/3}$
-and step size $1\le d \le 2N^{1/100}$ such that $|A\cap P|/|P| \ge \delta + c\delta^3$ for an absolute
-constant $c$.
+Suppose that there is $N\in \NN$ and a square-difference-free
+subset $A\subseteq [N]$ of density $\delta = |A|/N$. Then there is a square-difference-free
+subset $A'\subseteq [N^{1/4}]$
+of size $|A'| \ge (\delta + c'\delta^3) N^{1/4}$, where $c$ is an absolute constant.
+
+\proof By the previous lemma, there is a progression $P\subseteq [N]$ of length $\lfloor N^{1/3}\rfloor$
+such that $|A\cap P|/|P| \ge \delta + c\delta^3$ for an absolute
+constant $c$. We can partition this progression into subprogressions of length $\lfloor N^{1/4}\rfloor$
+and step size $d^2$, with a little bit left over at the end that is of size $O(N^{1/4})$. The density
+of $A$ on the union of the progressions is at least $\delta + c\delta^3 - O(N^{-1/12})$,
+which is at least $\delta + c'\delta^3$ for some new absolute constant $c'$. By the pigeonhole principle,
+$A$ has density at least $\delta + c'\delta^3$ on one of the subprogressions, call it
+$$P' = \bigl\{ n' + d^2r' : r'\in [N^{1/4}]\bigr\}.$$
+Now consider the set $A' = \bigl\{ r'\in [N^{1/4}] : n' + d^2 r'\in A\bigr\}$. We claim that
+$A'$ does not contain two elements that differ by a square. For if $A'$ did contain an integer $m$ and $m+s^2$
+for some $s\ne 0$, then we would have $n' + d^2m$ and $m'+d^2(m+s^2)$ in
+$A$, but these two elements differ by $(ds)^2$, and we know that $A$ does not contain two elements
+that differ by a square. This completes the proof.\slug
+
+We are now ready to state and prove a version of the Furstenberg--S\'ark\"ozy theorem.
+
+\parenproclaim Theorem~{\advthm} (Furstenberg, {\rm 1977;} S\'ark\"ozy, {\rm 1978}).
+There is an absolute constant $C$ such that the following holds.
+For all $\delta > 0$ there exists $N \le \bigl\lceil \exp\bigl(\exp(C/\delta^2)\bigr)\bigr\rceil$ such that
+any subset $A\subseteq [N]$
+with $|A|\ge \delta N$ contains two elements whose difference is a positive square integer.
+
+\proof If the theorem does not hold, then there is a square-difference-free subset $A\subseteq [N]$ of density
+at least $\delta$. Then we can repeatedly apply Lemma~{\lemfurstenbergsarkozy}, at each iteration replacing
+$\delta$ by $\delta + c'\delta^3$, where $c'$ is an absolute constant, and replacing $N$ by
+$\lfloor N^{1/4}\rfloor$.
+After $1/(c\delta^2)$ iterations of the increment $\delta\mapsto \delta + c\delta^3$,
+the density at least doubles. Then to double it again we need to iterate $1/\bigl((c(2\delta)^2\bigr)
+= 1/(4c\delta^2)$ times, and so on. In general, after iterating at most
+$$ {1\over c\delta^2} \biggl( 1 + {1\over 4} + {1\over 16} + \cdots\biggr)
+= {4\over 3c\delta^2}$$
+times, the density of the set $A$ will exceed $1$, a contradiction.
+
+To avoid any messiness arising from $N$ eventually equalling $1$, we will insist that in the final
+round of the iteration,
+$N\ge 100$. This means that in the very first step we must have $N^{4^{-4/(3c\delta^2)}} \ge 100$.
+Taking the logarithm of both sides, we need
+$$ 4^{-4/(3c\delta^2)} \log N \ge \log 100,$$
+and then taking the logarithm again makes this equivalent to
+$$\log\log N \ge {4\over 3c'\delta^2} \log 4 + \log\log 100.$$
+There is an absolute constant $C$ such that the condition $\log\log N \ge C/\delta^2$ implies the one above,
+so we see that a sufficient condition on $N$ for this whole argument to work is
+$$ N \ge \exp\bigl(\exp(C/\delta^2)\bigr).\noskipslug$$
+
+What we just proved can equivalently be stated as follows: {\sl There exists a constant $C'$
+such that for all $N$, any subset of $[N]$ with density at least $C'/\sqrt{\log\log N}$ contains
+two elements that differ by a nonzero square.}
 

### 2024-01-26 misc/expo/additive.tex
Hazem created

@@
 \parenproclaim Lemma {\advthm} (Petridis, {\rm 2012}). Let $A$ and $B$ be subsets of an abelian
-group and let $K$ be such that $|A_0+B| = K_0|A_0|$. Then there exist $A\subseteq A_0$ and
-$K\le K_0$ such that $|A+B+C| = K|A+C|$ for every finite subset $C$ of the group.
+group and let $K_0$ be such that $|A_0+B| = K_0|A_0|$. Then there exist $A\subseteq A_0$ and
+$K\le K_0$ such that $|A+B+C| \le K|A+C|$ for every finite subset $C$ of the group.
 

### 2024-01-26 misc/expo/entropy.tex
Hazem created

@@
 \def\II{\op{\bf I}}
+\def\dd{\op{\bf d}}
 \def\given{\mathbin{|}}
@@
 \widemargins
-\bookheader{ENTROPY AND THE FREIMAN--RUZSA THEOREM}{MARCEL K. GOH}
+\bookheader{ENTROPY AND ADDITIVE COMBINATORICS}{MARCEL K. GOH}
 
-\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{23 November 2023}
+\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{16 January 2024}
 
@@
 
-\proclaim Let $X$ and $Y$ be independent random variables. Then
+\proclaim Proposition \advthm. Let $X$ and $Y$ be independent random variables. Then
 $\Eta\{Y\given X\} = \Eta\{Y\}$ and consequently
@@
 =\Eta\{Y\}.$$
-The second version of the statement is a consequence of the additivity axiom.\slug
+The second version of the statement follows from the additivity axiom.\slug
 
@@
 
+\edef\corxn{\the\thmcount}
 \proclaim Corollary \advthm. For a discrete random variable $X$, we have $\Eta\{X^n\} = n\Eta\{X\}$.
@@
 $$\Eta\{X^n\} = \Eta\{X_1,\ldots,X_n\} = \sum_{i=1}^n \Eta\{X_i\} = \log_2(e)\cdot n,$$
-by induction. For general random variables $X_1,\ldots,X_n$, induction also yields the {\it chain rule}
-$$\Eta\{X_1,\ldots,X_n\} = \Eta\{X_1\} + \Eta\{X_2\given X_1\} + \cdots+\Eta\{X_n \given X_1,\ldots,X_{n-1}\}.$$
+by induction.
 
-Next we show that a constant random variable has zero entropy. This reflects the idea that we get
-no information from sampling a random variable if it always takes the same value no matter what.
+A similar induction on general random variables $X_1,\ldots,X_n$ gives the following statement,
+often known as the {\it chain rule}.
+
+\parenproclaim Proposition~{\advthm} (Chain rule).
+Let $X_1,\ldots,X_n$ be random variables. Then
+$$\Eta\{X_1,\ldots,X_n\} = \Eta\{X_1\} + \Eta\{X_2\given X_1\}
++ \cdots+\Eta\{X_n \given X_1,\ldots,X_{n-1}\}.\noskipslug$$
+
+Next we establish the intuitive fact
+that the entropy of a uniform random variable supported on a set $A$ is at most
+the entropy of a uniform random variable supported on a superset $B$ of $A$.
+
+\edef\propunifineq{\the\thmcount}
+\proclaim Proposition \advthm. Let $A\subseteq B$ with $B$ finite,
+let $X$ be uniformly distributed on $A$, and let $Y$ be uniformly distributed on $B$.
+Then $\Eta\{X\} \le \Eta\{Y\}$, with equality if and only if $A = B$.
+
+\proof By the extensibility axiom, $\Eta\{X\}$ is not affected if we regard $X$ as a function
+taking values in $B$. Then by the maximisation axiom, $\Eta\{X\}\le \Eta\{Y\}$, since $Y$ is
+uniform on $B$. If $A = B$, then it is clear that $\Eta\{X\} = \Eta\{Y\}$, since $X$ and $Y$
+are the same random variable.
+On the other hand, say $|A| = m$ and $|B| = n$ with $m<n$. If $m=1$, then by the previous proposition
+we have $\Eta\{X\} = 0$, and by normalisation and invariance, $\Eta\{Y\} = \log_2(e)$.\slug
+
+If $Y$ is a random variable such that $Y = f(X)$ for some random variable $Y$ and some function $f$,
+then we say that $Y$ is {\it determined by} $X$ or $X$ {\it determines} $Y$.
+We want to show that $\Eta\{Y\}\le \Eta\{X\}$, which reflects the idea that we get more information
+from $X$ than from $Y$. This, rather annoyingly, seems to require two steps.
+
+\proclaim Lemma \advthm. If $Y = f(X)$ then $\Eta\{X\} = \Eta\{Y\} + \Eta\{X\given Y\}$.
+
+\proof There is a a bijection between values $x$ taken
+by $Y$ and values $\bigl(x,f(x)\bigr)$ taken by $(X,Y)$, so we have
+$$\Eta\{X\} = \Eta\{X,Y\} = \Eta\{Y\} + \Eta\{X\given Y\}$$
+by additivity.\slug
+
+We are now done if we can show that entropy is nonnegative.
+This can also be derived straight from the axioms. The following proof is due to S.~Eberhard.
+
+\proclaim Proposition \advthm. Let $X$ be a discrete random variable taking values in a finite
+set $A$. Then $\Eta\{X\} \ge 0$.
+
+\proof First we will work in the case where there exists $n$ such that $\pr\{X = a\}$ is a multiple of
+$1/n$ for all $a\in A$. Let $Y$ be uniformly distributed on $[n]$ and let $\{E_a\}_{a\in A}$
+be a partition of $[n]$ such that $|E_a| = n\pr\{X= a\}$ for all $a\in A$, and let $Z = a$
+if $Y\in E_a$. This definition makes $Z$ and $X$ identically distributed, so $\Eta\{Z\} = \Eta\{X\}$
+by the invariance axiom, and it suffices to prove $\Eta\{Z\} \ge 0$.
+
+Since $Z$ is determined by $Y$, we have $\Eta\{Y\} = \Eta\{Z\} + \Eta\{Y\given Z\}$ by the previous lemma.
+Furthermore, for every $a\in A$, the conditional entropy $\Eta\{Y \given Z = a\}$ is uniformly
+distributed on a set of size at most $n$, whereas is uniformly distributed on $n$, so by
+Proposition~{\propunifineq}, we have $\pr\{Y \given Z = a\} \le \Eta\{Y\}$. This implies
+that $\Eta\{Y\} \ge \Eta\{Y\given Z\}$, which in turn gives us $\Eta\{Z\}\ge 0$.\slug
+
+\proclaim Corollary \advthm. If $Y = f(X)$ then $\Eta\{X\} \ge \Eta\{Y\}$.\slug
+
+Now we show that a random variable has zero entropy if and only if it is constant.
+This reflects the idea that the variables from which we get no information are those which take
+the same value no matter what.
 
@@
 so $\Eta\{X\} = \Eta\{X,X\}$ by the invariance axiom. But it can easily be checked that
-$X$ and $(X,X)$ are independent (we have, for example, $\pr\bigl\{X = a, (X,X)= (a,a)\bigr\} =
-\pr\{X = a\}\pr\bigl\{(X,X) = (a,a)\bigr\}$), so $\Eta\{X,X\} = 2\Eta\{X\}$. Thus we conclude
+$X$ and $(X,X)$ are independent (we have
+$$\pr\bigl\{X = a, (X,X)= (a,a)\bigr\} = \pr\{X = a\}\pr\bigl\{(X,X) = (a,a)\bigr\}$$
+for instance), so $\Eta\{X,X\} = 2\Eta\{X\}$. Thus we conclude
 that $\Eta\{X\} = 0$.
@@
 Now suppose that $X$ takes more than one value; let $A$ be the set of $a$ such that
-$\pr\{X = a\} > 0$.
-\slug
+$\pr\{X = a\} > 0$ and let $\alpha = \max_{a\in A} \pr\{X = a\}$. For all $n$ let $X^n$ denote
+the tuple of $n$ independent copies of $X$; the maximum probability of any particular value (in $A^n$)
+that $X^n$ takes is $\alpha^n$. But $\alpha < 1$ since $X$ takes more than one value, so for any
+$\eps > 0$ we can find $n$ such that $\alpha^n < \eps$. This means that we can partition $A^n$ into
+two disjoint sets $E$ and $F$ such that $\pr\{X^n \in E\}$ and $\pr\{X^n \in F\}$ are both
+in the range $[1/2-\eps, 1/2+\eps]$.
 
-Now we establish the intuitive fact
-that the entropy of a uniform random variable supported on a set $A$ is at most
-the entropy of a uniform random variable supported on a superset $B$ of $A$.
+Let $Y$ be the random variable taking the value $0$ if $X^n\in E$ and $1$ if $X^n\in F$. Then
+by Corollary~{\corxn}, $\Eta\{X^n\} = n\Eta\{X\}$, and since $X^n$ determines $Y$,
+$$\Eta\{X^n\} = \Eta\{Y\} + \Eta\{X^n\given Y\} \ge \Eta\{Y\}.$$
+But $\Eta\{Y\} > 0$ for $\eps$ small enough, the normalisation and continuity axioms. So
+$\Eta\{X\} \ge \Eta\{Y\}/n > 0$ as well.\slug
 
-\proclaim Proposition \advthm. Let $A\subseteq B$ with $B$ finite,
-let $X$ be uniformly distributed on $A$, and let $Y$ be uniformly distributed on $B$.
-Then $\Eta\{X\} \le \Eta\{Y\}$, with equality if and only if $A = B$.
-
-\proof By the extensibility axiom, $\Eta\{X\}$ is not affected if we regard $X$ as a function
-taking values in $B$. Then by the maximisation axiom, $\eta\{X\}\le \eta\{Y\}$, since $Y$ is
-uniform on $B$. If $A = B$, then it is clear that $\Eta\{X\} = \Eta\{Y\}$, since $X$ and $Y$
-are the same random variable.
-Now say $|A| = m$ and $|B| = n$ with $m<n$. If $m=1$, then by the previous proposition
-we have $\Eta\{X\} = 0$, and $Y$ takes at least two values,
-
-The next proposition is that $\Eta\{X\}$ is always nonnegative. The following
-proof is attributed to S.~Prendiville.
-
-\proclaim Proposition \advthm. For any random variable $X$ taking values in a finite set $A$,
-we have $\Eta\{X\} \ge 0$.
-
-\proof First we consider the case that there exists $n\in \NN$ such that for all $x\in A$, there is some
-$c_x$ such that $\pr\{X = x\} = c_x/n$. Let $Y$ be uniform on $[n]$ and let $E_x$ be a partition
-of $[n]$ such that $|E_x| = c_x$ for all $x\in X$. Then let $Z$ be given by $Z = x$ if $Y\in E_x$.
-This random variable $Z$ takes values in $A$ and has the exact same distribution as $X$, so
-by the invariance axiom, $\Eta\{X\} = \Eta\{Z\}$.
-This means that $Z = f(Y)$ for some function $f$, meaning that there is a bijection between
-values $y$ taken by $Y$ and values $\bigl(y, f(y)\bigr)$ taken by $(Y,Z)$. So we may apply the invariance
-axiom again to conclude that $\Eta\{Y,Z\} = \Eta\{Y\}$.
-
-Now by the additivity axiom, $\Eta\{Y,Z\} = \Eta\{Y\given Z\} + \Eta\{Z\}$, so we
-have $\Eta\{Y\} = \Eta\{Y\given Z\} + \Eta\{Z\}$.
-
-If $X$ is a random variable such that $X = f(X)$ for some random variable $Y$, then we say
-that $X$ is {\it determined by} $Y$ or $Y$ {\it determines} $X$. The following proposition formalises
-the idea that we get more information from $Y$ than we get from $X$.
-
-\proclaim Proposition \advthm. Let $X$ and $Y$ be random variables such that $Y$ determines $X$. Then
-$$\Eta\{ X\} \le \Eta\{Y\}.$$
-
-\proof The
-
+\medskip\boldlabel Mutual information.
 For random variables $X$ and $Y$, the {\it mutual information} $\II\{X : Y\}$ is defined
@@
 if and only if $X$ and $Y$ are independent, and we also have the
-inequality $\Eta\{X \given Y\} \le \Eta\{X\}$. Given a triple $(X,Y,Z)$ of random variables,
+inequality $\Eta\{X \given Y\} \le \Eta\{X\}$; that is, {\sl conditioning cannot increase the
+entropy of a random variable}. Given a triple $(X,Y,Z)$ of random variables,
 we can apply this with $(X\given Z = z)$ to obtain the inequality
@@
 }$$
-so we see that submodularity is equivalent to the statement $\II\{X : Y\given Z\} \ge 0$.
+so we see that submodularity of entropy is equivalent to the statement that $\II\{X : Y\given Z\} \ge 0$.
 Analogously to the unconditional case, we have equality if and only if $X$ and $Y$ are
@@
 
-\medskip\boldlabel The formula for entropy.
-Now we finally give the formula for the entropy of a random variable. For a discrete random
-variable $X$, we let $p_X : S\to [0,1]$ denote the probability mass function of $X$, given by
-$p_X(x) = \pr\{X = x\}$ for all $x\in S$. Unless subscripted, $\log$ always denotes the natural logarithm.
+\advsect Group-valued random variables
 
-\proclaim Proposition \advthm. Let $X$ be a random variable taking values in a finite set $S$.
-Then $\Eta\{X\}$ satisfies axioms (a) through (e) as well as the additional normalisation axiom (f)
-if and only if
-$$\Eta\{X\} = \ex\bigl\{ 1/\log p_X(X) \bigr\} = \sum_{x\in S} p_X(x) \log {1\over p_X(s)},$$
-where we adopt the convention that $0\log(1/0) = 0$.
+Now we will examine the case where the random variables in question take values in an abelian group $G$,
+meaning we can take sums $X+Y$ and differences $X-Y$ of them. Note that if we condition on $Y$,
+then the values taken $X+Y$ are in bijection with values taken by $X$. This leads to the following
+proposition.
 
-\proof [TODO: Prove this.]\slug
+\edef\maxsumsetbound{\the\thmcount}
+\proclaim Proposition \advthm. Let $X$ and $Y$ be random variables each taking finitely
+many values in an abelian group $G$.
+We have
+$$\max\bigl( \Eta\{X\}, \Eta\{Y\} \bigr) - \II\{X : Y\} \le \Eta\{ X \pm Y\}.$$
+Furthermore, for any random variable $Z$, we have the conditional version
+$$\max\bigl( \Eta\{X\given Z\}, \Eta\{Y\given Z\} \bigr) - \II\{X : Y\given Z\} \le \Eta\{ X \pm Y\given Z\}$$
+of the same statement.
 
-Equipped now with a formula for the entropy $\Eta\{X\}$ of a random variable $X$, we can prove further useful
-facts. We saw in the previous section that the entropy of a uniform random variable whose range is $\{0,1\}^n$
-is $\log_2(e)\cdot n = \log(2^n)$, and we know, by the maximisation axiom, that this is the largest
-the entropy of any $X$ taking values in $\{0,1\}^n$ can be. The following proposition generalises
-this notion.
+\proof Since conditioning does not increase entropy, we have
+$$\Eta\{ X\pm Y\} \ge \Eta\{ X\pm Y \given Y\},$$
+and since the probabilities $\pr\{ X + Y = z \given Y = y\} = \pr\{ X = z-y \given Y = y\}$ for all $z\in G$,
+by invariance we have
+$$\Eta\{ X\pm Y\} \ge \Eta\{X\given Y\} = \Eta\{X\} - \II\{X : Y\}.$$
+Repeating the same argument but exchanging the roles of $X$ and $Y$, we get
+$$\Eta\{ X\pm Y\} \ge \Eta\{Y\given X\} = \Eta\{Y\} - \II\{X : Y\},$$
+so
+$$\Eta\{ X\pm Y\} \ge \max\bigl(\Eta\{X\}, \Eta\{Y\}\bigr) - \II\{X : Y\}.$$
+Now let $Z$ be any random variable with finite support.
+$$\eqalign{
+\Eta\{ X\pm Y\given Z\} &= \sum_{z\in G} \pr\{Z = z\} \Eta\{ X \pm Y \given Z = z\} \cr
+&\ge \Bigl(\max\bigl(\Eta\{X\given Z\}, \Eta\{Y\given Z\}\bigr) - \II\{X : Y\given Z\} \Bigr)
+\sum_{z\in G} \pr\{Z = z\} \cr
+&= \max\bigl(\Eta\{X\given Z\}, \Eta\{Y\given Z\}\bigr) - \II\{X : Y\given Z\} ,\cr
+}$$
+which completes the proof.\slug
 
-\proclaim Proposition \advthm. For any random variable $X$ taking values in a finite set $S$,
-we have
-$$\Eta\{X\} \le \log |S|.$$
+\proclaim Corollary \advthm. If $X$ and $Y$ are independent, then
+$$\max\bigl( \Eta\{X\}, \Eta\{Y\} \bigr) \le \Eta\{ X \pm Y\}.$$
 
-\proof By Jensen's inequality, we have
+\proof The mutual information $\II\{X : Y\}$ is zero whenever $X$ and $Y$ are independent.\slug
+
+\medskip\boldlabel Entropic Ruzsa distance.
+In additive combinatorics, whenever we have two finite subsets $A$ and $B$
+of the same abelian group, we can compute the Ruzsa distance
+$$d(A,B) = \log {|A-B|\over \sqrt{|A|\cdot|B|}}$$
+between them. (This satisfies all the axioms of a metric except the one requiring $d(A,A) = 0$ for all sets
+$A$.)
+
+The entropic analogue of the Ruzsa distance is defined as follows.
+For finitely supported random variables $X$ and $Y$ taking values in the same abelian group,
+we let $X'$ and $Y'$ be independent copies of $X$
+and $Y$, respectively, and define the {\it entropic Ruzsa distance} by
+$$\dd\{X,Y\} = \Eta\{X'-Y'\} - {\Eta\{X'\} \over 2} - {\Eta\{Y'\}\over 2}.$$
+This definition only depends on the individual distributions of $X$ and $Y$ and does not require them to
+have the same sample space. Once again, we don't necessarily have $\dd\{X,X\} = 0$, but we do have
+the triangle inequality, which shall now prove.
+
+\proclaim Proposition \advthm. Let $X$, $Y$, and $Z$ be random variables with finite support in the same
+abelian group. Then
+$$\dd\{X,Z\} \le \dd\{X,Y\} + \dd\{Y,Z\},$$
+which is equivalent to
+$$\Eta\{X'-Z'\} \le \Eta\{X'-Y'\} + \Eta\{Y'-Z'\} - \Eta\{Y'\}$$
+for $X'$, $Y'$, and $Z'$ independent and distributed as $X$, $Y$, and $Z$, respectively.
+
+\proof That the two statements are equivalent is easily obtained by expanding the definition of
+entropic Ruzsa distance and cancelling some terms. So without loss of generality, we may assume that
+that $X$, $Y$, and $Z$ are independent and just prove the second statement.
+
+By submodularity, we have $\II\bigl\{ (X-Y : Z) \given X-Z\bigr\} \ge 0$, so
+\edef\eqfirstsubmodularity{\the\eqcount}
 $$\eqalign{
-\Eta\{X\} &= \ex\bigl\{ \log(1/p_X(X)\bigr\} \cr
-&\le \log \ex\bigl\{ 1/p_X(X)\bigr\} \cr
-&= \log \sum_{x\in S} \pr\{X = x\} {1\over \pr\{X = x\}} \cr
-&= \log |S|.\noskipslug \cr
+0 &\le \II\bigl\{ (X-Y : Z) \given X-Z\bigr\} \cr
+&\le \Eta\{X-Y \given X-Z\} + \Eta\{ Z\given X-Z\} - \Eta\{ X-Y, Z \given X-Z\} \cr
+&\le \Eta\{X-Y, X-Z\} + \Eta\{ Z, X-Z\} - \Eta\{ X-Y, Z, X-Z\} - \Eta\{X-Z\}.\cr
+}\adveq$$
+Now, since the values $(x-y,x-z)$ taken by $(X-Y,X-Z)$ are in bijection with values
+$(x-z,y-z)$ taken by $(X-Z, Y-Z)$ via the map $(v,w) \mapsto (w, w-v)$,
+by the invariance axiom we have
+$$\Eta\{X-Y, X-Z\} = \Eta\{X-Z, Y-Z\},$$
+and
+$$\Eta\{X-Y, X-Z\} \le \Eta\{X-Y\} + \Eta\{Y-Z\}$$
+follows by submodularity. Similar invocations of the invariance axiom give
+$$\Eta\{Z, X-Z\} = \Eta\{X,Z\}$$
+and
+$$\Eta\{X-Y, Z, X-Z\} = \Eta\{X,Y,Z\} = \Eta\{X,Z\} + \Eta\{Y\},$$
+where in the latter statement the second equality follows from the fact that $(X,Y)$ and $Z$ are
+independent.
+Substituting these three inequalities into~\refeq{\eqfirstsubmodularity}, we have
+$$0\le \Eta\{X-Y\} + \Eta\{Y-Z\} +\Eta\{X,Z\} -\Eta\{X,Z\} + \Eta\{Y\} - \Eta\{X-Z\},$$
+whence
+$$ \Eta\{X-Z\} \le \Eta\{X-Y\} + \Eta\{Y-Z\} + \Eta\{Y\},$$
+which completes the proof.\slug
+
+We also define a conditional version of the entropic Ruzsa distance. If $X$ and $Y$ are $G$-valued
+random variables with finite support and $Z$ and $W$ are any random variables with finite supports
+$A$ and $B$ respectively, then we define
+$$\dd\{ X\given Z ; Y\given W\} = \sum_{z\in A} \sum_{w\in B} \pr\{Z = z\} \pr\{W=w\}
+\dd\bigl\{ (X\given Z = z) ; (Y\given W=w)\bigr\}.$$
+If $(X',Z')$ and $(Y',W')$ are independent copies of $(X,Z)$ and $(Y,W)$ respectively, then this
+distance is also given by the formula
+$$\dd\{ X\given Z ; Y\given W\} = \Eta\{X'-Y'\given Z',W'\} - {\Eta\{X'\given Z'\}\over 2}
+- {\Eta\{Y'\given W'\}\over 2}.$$
+
+\advsect The Pl\"unnecke--Ruzsa inequality
+
+In additive combinatorics, one of the most useful sumset inequalities is the following.
+
+\parenproclaim Theorem A (Pl\"unnecke--Ruzsa inequality). Let $A$ and $B$ be finite subsets of an
+abelian group and suppose that $|A+B|\le K|A|$ for some constant $K$. Then for any integers
+$r,s\ge 0$, not both zero, we have $|rB-sB| \le K^{r+s}|A|$.\slug
+
+In this section we will develop an entropic analogue of this statement, in which sets are replaced
+by random variables of finite support and cardinality is replaced with the exponential of
+entropy. First, a technical lemma.
+
+\proclaim Lemma \advthm. Let $X$, $Y$, and $Z$ be independent random variables taking values in a common
+abelian group. Then
+$$\Eta\{X+Y+Z\} - \Eta\{X+Y\} \le \Eta\{Y+Z\} -\Eta\{Y\}.$$
+
+\proof By submodularity, the quantity $\II\{ X : Z\given X+Y+Z\}$ is nonnegative, so we have
+$$\eqalign{
+0&\le \II\{ X : Z\given X+Y+Z\} \cr
+&= \Eta\{X, X+Y+Z\} + \Eta\{Z, X+Y+Z\}\cr
+&\qquad\qquad\qquad\qquad- \Eta\{X,Z,X+Y+Z\} - \Eta\{X+Y+Z\}.\cr
 }$$
+Since $X$, $Y$, and $Z$ are independent, we have
+$$\Eta\{X,X+Y+Z\} = \Eta\{X,Y+Z\} = \Eta\{X\} + \Eta\{Y+Z\},$$
+where in the first equality we use invariance. By similar reasoning we have
+$$\Eta\{Z, X+Y+Z\} = \Eta\{Z\} + \Eta\{X+Y\}$$
+and
+$$\Eta\{X,Z,X+Y+Z\} = \Eta\{X\} + \Eta\{Y\} + \Eta\{Z\}.$$
+Plugging these three identities into the inequality above yields
+$$\eqalign{
+0&\le \Eta\{X\} + \Eta\{Y+Z\} + \Eta\{Z\} + \Eta\{X+Y\} \cr
+&\qquad\qquad -\Eta\{X\} - \Eta\{Y\} - \Eta\{Z\} - \Eta\{X+Y+Z\} \cr
+&= \Eta\{Y+Z\} + \Eta\{X+Y\} - \Eta\{Z\} - \Eta\{X+Y+Z\},\cr
+}$$
+whence the claim follows upon rearranging.\slug
+
+From here we are not far from proving the entropic Pl\"unnecke--Ruzsa inequality, a result of T.~Tao.
+
+\proclaim Theorem \advthm. Let $X,Y_1,\ldots,Y_m$ be independent random variables of finite entropy
+taking values in an abelian group $G$, such that
+$$\Eta\{X+Y_i\} \le \Eta\{X\} + \log K_i$$
+for all $1\le i\le m$ and some scalars $K_1,\ldots,K_m\ge 1$. Then
+$$\Eta\{X+Y_1+\cdots+Y_m\} \le \Eta\{X\} + \log(K_1\cdots K_m).$$
+
+\proof We prove the claim by induction on $m$. If $m=1$, then we are done by hypothesis. Now
+suppose that $\Eta\{X+Y_1+\cdots+Y_{m-1}\} \le \Eta\{X\} + \log(K_1\cdots K_{m-1})$. Then by
+the previous lemma, the induction hypothesis, and the hypothesis on $\Eta\{X+Y_m\}$, we have
+$$\eqalign{
+\Eta\{Y_1 + \cdots + Y_{m-1} + X + Y_m\}
+&\le \Eta\{Y_1 + \cdots + Y_{m-1} + X\} \cr
+&\qquad\qquad\qquad\qquad+ \Eta\{X+Y_m\} - \Eta\{X\}\cr
+&\le \Eta\{X\} + \log(K_1\cdots K_{m-1}) + \log K_m\cr
+&\le \Eta\{X\} + \log(K_1\cdots K_m),\cr
+}$$
+which is what we sought to prove.\slug
+
+We can make this look bit more like the version of the Pl\"unnecke--Ruzsa inequality above
+by using the triangle inequality.
+
+\parenproclaim Corollary {\advthm} (Entropic Plunnecke--Ruzsa inequality). Let $X$ and $Y$ be random variables
+with $\Eta\{X+Y\} \le \Eta\{X\} + \log K$. Then for any $r,s\ge 0$ not both zero, we have
+$$\Eta\{ Y_1 + \cdots + Y_r - Z_1 - \cdots - Z_s \} \le \Eta\{X\} + (r+s)\log K,$$
+where $Y_1,\ldots,Y_r,Z_1,\ldots,Z_s$ are independent copies of $Y$.
+
+\proof By the entropic Ruzsa triangle inequality, we have
+$$\eqalign{
+&\Eta\{ Y_1 + \cdots + Y_r - Z_1 - \cdots - Z_s \} \le \cr
+&\qquad\qquad\Eta\{Y_1 + \cdots + Y_r + X\} + \Eta\{-X - Z_1 - \cdots - Z_s\} - \Eta\{-X\}. \cr
+}$$
+The values of $-X$ are in bijection with values of $X$, and the values of $-X-Z_1-\cdots-Z_s$
+are in bijection with the values of $X+Z_1+\cdots+Z_s$ (with the same probabilities in both cases),
+so by the invariance axiom, we have
+$$\eqalign{
+&\Eta\{ Y_1 + \cdots + Y_r - Z_1 - \cdots - Z_s \} \le \cr
+&\qquad\qquad\Eta\{X+Y_1 + \cdots + Y_r\} + \Eta\{X+ Z_1 + \cdots + Z_s\} - \Eta\{X\},\cr
+}$$
+and we can apply the the previous theorem twice to get
+$$\eqalign{
+\Eta\{ Y_1 + \cdots + Y_r - Z_1 - \cdots - Z_s \}  &\le \Eta\{X\} + \log (K^r) + \log (K^s),\cr
+&= \Eta\{X\} + (r+s)\log K.\noskipslug\cr
+}$$
+
+% \medskip\boldlabel The formula for entropy.
+% Now we finally give the formula for the entropy of a random variable. For a discrete random
+% variable $X$, we let $p_X : S\to [0,1]$ denote the probability mass function of $X$, given by
+% $p_X(x) = \pr\{X = x\}$ for all $x\in S$. Unless subscripted, $\log$ always denotes the natural logarithm.
+% 
+% \proclaim Proposition \advthm. Let $X$ be a random variable taking values in a finite set $S$.
+% Then $\Eta\{X\}$ satisfies axioms (a) through (e) as well as the additional normalisation axiom (f)
+% if and only if
+% $$\Eta\{X\} = \ex\bigl\{ 1/\log p_X(X) \bigr\} = \sum_{x\in S} p_X(x) \log {1\over p_X(s)},$$
+% where we adopt the convention that $0\log(1/0) = 0$.
+% 
+% \proof [TODO: Prove this.]\slug
+% 
+% Equipped now with a formula for the entropy $\Eta\{X\}$ of a random variable $X$, we can prove further useful
+% facts. We saw in the previous section that the entropy of a uniform random variable whose range is $\{0,1\}^n$
+% is $\log_2(e)\cdot n = \log(2^n)$, and we know, by the maximisation axiom, that this is the largest
+% the entropy of any $X$ taking values in $\{0,1\}^n$ can be. The following proposition generalises
+% this notion.
+% 
+% \proclaim Proposition \advthm. For any random variable $X$ taking values in a finite set $S$,
+% we have
+% $$\Eta\{X\} \le \log |S|.$$
+% 
+% \proof By Jensen's inequality, we have
+% $$\eqalign{
+% \Eta\{X\} &= \ex\bigl\{ \log(1/p_X(X)\bigr\} \cr
+% &\le \log \ex\bigl\{ 1/p_X(X)\bigr\} \cr
+% &= \log \sum_{x\in S} \pr\{X = x\} {1\over \pr\{X = x\}} \cr
+% &= \log |S|.\noskipslug \cr
+% }$$
 

### 2024-02-16 misc/expo/entropy.tex
Progress in entropy

@@
 \def\Eta{\op{\bf H}}
+\def\Etaseven{\op{\sevenbf H}}
 \def\II{\op{\bf I}}
 \def\dd{\op{\bf d}}
+\def\ee{\op{\bf e}}
 \def\given{\mathbin{|}}
+\def\argmax{\limitop{arg$\;$max}}
+\def\advthm{\the\sectcount.\the\thmcount\global\advance \thmcount by 1}
+\def\advsect{\global\advance\sectcount by 1\section\the\sectcount\global\thmcount=1. }
+\sectcount=0
 
@@
 
-\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{16 January 2024}
+\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{15 February 2024}
 
-\floattext4.5 \ninebf Note.
-\ninepoint This is a gentle introduction to the notion of entropy as it is used
-in additive combinatorics, with a view towards understanding the new proof of the polynomial Freiman--Ruzsa
-conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. The preliminary portion of
-these notes are largely transcribed from lectures given by W.~T.~Gowers.
+\floattext4.5 \ninebf Abstract.
+\ninepoint These expository notes give a gentle introduction to the notion of entropy as it is used
+in additive combinatorics, with a view towards understanding the proof of the polynomial Freiman--Ruzsa
+conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. Much of the first section has been
+transcribed from lectures given by W.~T.~Gowers.
 
@@
 
-For now, we shall take it on faith that there really exists a function on random variables satisfying
-these axioms. Later on, when we prove this, we will find that actually, the axioms only define entropy
+We shall take it on faith that there really exists a function on random variables satisfying
+these axioms.
+In fact, the axioms only define entropy
 up to a multiplicative constant, so we shall add the following axiom.
+(It is very possible you have met this function somewhere on your travels, but you will not find it anywhere in
+these notes.)
 
@@
 \item{f)} ({\it Normalisation.}) If $X$ is uniformly distributed on $\{0,1\}$, then
-$\Eta\{X\} = \log_2(e)$.
+$\Eta\{X\} = 1$.
 \medskip
-
-It is actually more common (for obvious reasons, especially in computer science)
-for one to set the entropy of a uniform random variable on $\{0,1\}$
-to $1$, but we shall
-follow the convention of Gowers et al., since the eventual goal of these notes is to understand their proof
-of the polynomial Freiman--Ruzsa conjecture.
 
@@
 We will sometimes use the notation $X^n$ to denote the vector $(X_1,\ldots,X_n)$ where the $X_i$
-are independent copies of the random variable $X$.
+are independent copies of the random variable $X$. We have the following three corollaries, which
+are each proved by induction. The second also requires the normalisation axiom,
+and the third is often known as the {\it chain rule}.
 
-\edef\corxn{\the\thmcount}
-\proclaim Corollary \advthm. For a discrete random variable $X$, we have $\Eta\{X^n\} = n\Eta\{X\}$.
+\edef\corxn{\the\sectcount.\the\thmcount}
+\proclaim Corollary \advthm. We have $\Eta\{X^n\} = n\Eta\{X\}$.\slug
 
-\proof Perform induction on $n$ and use the previous proposition.\slug
+\edef\cortwon{\the\sectcount.\the\thmcount}
+\proclaim Corollary \advthm. If $X$ is uniformly distributed on a set of size $2^n$, then
+$$\Eta\{X\} = n.\noskipslug$$
 
-In the case where $X$ is the uniform distribution on $\{0,1\}$, and $X_1,\ldots,X_n$ are independent
-and distributed as $X$,
-we can additionally use the normalisation axiom to conclude that
-$$\Eta\{X^n\} = \Eta\{X_1,\ldots,X_n\} = \sum_{i=1}^n \Eta\{X_i\} = \log_2(e)\cdot n,$$
-by induction.
-
-A similar induction on general random variables $X_1,\ldots,X_n$ gives the following statement,
-often known as the {\it chain rule}.
-
-\parenproclaim Proposition~{\advthm} (Chain rule).
+\parenproclaim Corollary~{\advthm} (Chain rule).
 Let $X_1,\ldots,X_n$ be random variables. Then
@@
 
-\edef\propunifineq{\the\thmcount}
+\edef\propunifineq{\the\sectcount.\the\thmcount}
 \proclaim Proposition \advthm. Let $A\subseteq B$ with $B$ finite,
@@
 taking values in $B$. Then by the maximisation axiom, $\Eta\{X\}\le \Eta\{Y\}$, since $Y$ is
-uniform on $B$. If $A = B$, then it is clear that $\Eta\{X\} = \Eta\{Y\}$, since $X$ and $Y$
+uniform on $B$.
+
+If $A = B$, then it is clear that $\Eta\{X\} = \Eta\{Y\}$, since $X$ and $Y$
 are the same random variable.
+
 On the other hand, say $|A| = m$ and $|B| = n$ with $m<n$. If $m=1$, then by the previous proposition
-we have $\Eta\{X\} = 0$, and by normalisation and invariance, $\Eta\{Y\} = \log_2(e)$.\slug
+we have $\Eta\{X\} = 0$, and by normalisation and invariance, $\Eta\{Y\} = 1$.
+When $m\ge 2$, pick $k$ such that $m^k \le n^{k-1}$, so that $|A^k| \le |B^{k-1}|$.
+Then by Corollary~{\corxn} and the inequality we showed in the first paragraph of this proof, we have
+$$n\Eta\{X\} = \Eta\{X^n\} \le \Eta\{X^{n-1}\} = (n-1)\Eta\{Y\},$$
+whence $\Eta\{X\} < \Eta\{Y\}$.\slug
+
+The link between entropy and additive combinatorics is rather a fine one. It is based on the following
+observation. (In this and the rest of the notes, $\lg$ denotes the base-$2$ logarithm. This base can
+be changed by modifying the normalisation axiom.)
+
+\edef\proplogeq{\the\sectcount.\the\thmcount}
+\proclaim Proposition \advthm. Let $X$ be a uniform random variable on a finite set $A$.
+Then
+$$\Eta\{X\} = \lg |A|.$$
+
+\proof For any positive integer $n$ we can let
+$X^n$ denote a tuple of independent copies of $X$; Corollary~{\corxn} tells us $\Eta\{X^n\} = n\Eta\{X\}$.
+Let $m$ be such that $2^m \le |A|^n \le 2^{m+1}$ so that
+$${m\over n} \le \lg |A| \le {(m+1)\over n}.$$
+Let $Y$ be uniform on a set of size $2^m$,
+and let $Z$ be uniform on a set of size $2^{m+1}$. Then by Corollary~{\cortwon} we have
+$\Eta\{Y\} = m $ and $\Eta\{Z\} = (m+1)$. Then by Proposition~{\propunifineq}
+we have
+$${m\over n} \le \Eta\{X\} \le {(m+1)\over n}.$$
+In other words, $\Eta\{X\}$ satisfies the same bounds as $\lg |A|$. Taking $n$ large, we can make
+these bounds arbitrarily tight, proving the claim.\slug
+
+The maximisation axiom gives the following corollary.
+
+\edef\corlogineq{\the\sectcount.\the\thmcount}
+\proclaim Corollary \advthm. Let $X$ be a random variable supported on a finite set $A$.
+Then
+$$\Eta\{X\} \le \lg |A|.\noskipslug$$
+
+Hence the entropy $\Eta\{X\}$ is at most the exponential of the size of its support. As we will see,
+simply replacing (logarithms of) cardinalities with entropies, we at useful ``entropic analogues''
+of combinatorial statements. But first, we need more lemmas.
 
@@
 We want to show that $\Eta\{Y\}\le \Eta\{X\}$, which reflects the idea that we get more information
-from $X$ than from $Y$. This, rather annoyingly, seems to require two steps.
+from $X$ than from $Y$. This, rather annoyingly, seems to require a couple of steps.
 
@@
 \proof There is a a bijection between values $x$ taken
-by $Y$ and values $\bigl(x,f(x)\bigr)$ taken by $(X,Y)$, so we have
+by $X$ and values $\bigl(x,f(x)\bigr)$ taken by $(X,Y)$, so we have
 $$\Eta\{X\} = \Eta\{X,Y\} = \Eta\{Y\} + \Eta\{X\given Y\}$$
-by additivity.\slug
+by invariance and additivity.\slug
 
 We are now done if we can show that entropy is nonnegative.
-This can also be derived straight from the axioms. The following proof is due to S.~Eberhard.
+This is a corollary of the following lemma. The following proof is a modification of
+one due to S.~Eberhard.
 
-\proclaim Proposition \advthm. Let $X$ be a discrete random variable taking values in a finite
-set $A$. Then $\Eta\{X\} \ge 0$.
+\edef\lemmaxprob{\the\sectcount.\the\thmcount}
+\proclaim Proposition \advthm. Let $X$ be a discrete random variable supported on a set $A$ and let
+$$a^* = \argmax_{a\in A} \pr\{X = a\}.$$
+Then
+$$\pr\{X = a^*\} \ge 2^{-\Etaseven\{X\}}.$$
 
@@
 
+For every $a\in A$, the conditional entropy $\Eta\{Y \given Z = a\}$ is uniformly
+distributed on a set of size $|E_a|$. From our choice of $a^*$
+we have $|E_{a^*}| \ge |E_a|$ for all $a\in A$. Hence by Proposition~{\proplogeq}, we have
+$$\Eta\{Y\given Z\} = \sum_{a\in A} \pr\{X = a\} \Eta\{Y\given X = a\}
+\sum_{a\in A} \pr\{X = a\} \log |E_a| \le \log |E_{a^*}|.$$
 Since $Z$ is determined by $Y$, we have $\Eta\{Y\} = \Eta\{Z\} + \Eta\{Y\given Z\}$ by the previous lemma.
-Furthermore, for every $a\in A$, the conditional entropy $\Eta\{Y \given Z = a\}$ is uniformly
-distributed on a set of size at most $n$, whereas is uniformly distributed on $n$, so by
-Proposition~{\propunifineq}, we have $\pr\{Y \given Z = a\} \le \Eta\{Y\}$. This implies
-that $\Eta\{Y\} \ge \Eta\{Y\given Z\}$, which in turn gives us $\Eta\{Z\}\ge 0$.\slug
+So by another invocation of Proposition~{\proplogeq}, we have
+$$\eqalign{
+\Eta\{Z\} &= \Eta\{Y\} - \Eta\{Y\given Z\} \cr
+&\ge \log n - \log |E_{a^*}| \cr
+&\ge \log \biggl({n\over |E_{a^*}|}\biggr) \cr
+&= \log\biggl({1\over \pr\{X = a^*\}}\biggr), \cr
+}$$
+and hence $2^{-\Etaseven\{X\}} = 2^{-\Etaseven\{Z\}} \le \pr\{X = a^*\}$.
 
+The general case follows from the continuity axiom.\slug
+
+This proof came dangerously close to deriving the formula for entropy, but we will not need any such formula,
+so we will refrain from mentioning it.
+From the fact that $\pr\{X = a^*\} \le 1$ we can immediately conclude that entropy is nonnegative.
+
+\edef\corentropynonneg{\the\sectcount.\the\thmcount}
+\proclaim Corollary \advthm. Let $X$ be a discrete random variable taking values in a finite
+set $A$. Then $\Eta\{X\} \ge 0$.\slug
+
+This observation completes the proof that a random variable has a smaller entropy than one by which
+it is determined.
+
+\edef\cordetermines{\the\sectcount.\the\thmcount}
 \proclaim Corollary \advthm. If $Y = f(X)$ then $\Eta\{X\} \ge \Eta\{Y\}$.\slug
 
-Now we show that a random variable has zero entropy if and only if it is constant.
+Next we show that a random variable has zero entropy if and only if it is constant.
 This reflects the idea that the variables from which we get no information are those which take
@@
 by looking at the other one. From the formula it is clear that $\II\{X : Y\}  = 0$
-if and only if $X$ and $Y$ are independent, and we also have the
-inequality $\Eta\{X \given Y\} \le \Eta\{X\}$; that is, {\sl conditioning cannot increase the
-entropy of a random variable}. Given a triple $(X,Y,Z)$ of random variables,
-we can apply this with $(X\given Z = z)$ to obtain the inequality
-$$ \Eta\{ X\given Y,Z\} \le \Eta\{X\given Z\},$$
-which is called {\it submodularity}. An equivalent statement is
-$$\Eta\{X,Y,Z\} + \Eta\{Z\} \le \Eta\{X,Z\} + \Eta\{Y,Z\}.$$
-If $Z$ takes values in a set $C$, the {\it conditional mutual information} is defined by
+if $X$ and $Y$ are independent. In general, we still have the inequality $\II\{X : Y\} \ge 0$,
+which a corollary of the following lemma, which expresses the intuitive fact that conditioning
+cannot increase the entropy of a random variable.
+The proof, which is due to C.~West, uses an argument similar to the one
+we used to prove Lemma~{\lemmaxprob}.
+
+\proclaim Proposition {\advthm}. Let $X$ and $Y$ be discrete random variables. Then
+$$\Eta\{X\given Y\} \le \Eta\{X\}.$$
+
+\proof Let $A$ be the support of $X$ and $B$ be the support of $Y$.
+First we consider the case that $X$ is uniform on $A$ (so $A$ is finite). Then by the definition
+of conditional entropy,
+$$\Eta\{X\given Y\} = \sum_{b\in B} \pr\{Y = b\} \Eta\{X \given Y= b\}.$$
+But for each $b$, the random variable $(X\given Y = b)$ takes values in $A$, so its entropy is
+bounded above by $\Eta\{X\}$ by the maximisation axiom. Hence $\Eta\{X\given Y\} \le \Eta\{X\}$.
+
+Next, suppose that $A$ and $B$ are both finite and suppose further that $\pr\{Y = b\}$
+is rational for all $b$. Then there is an integer $n$ and integers $\{m_b\}_{b\in B}$
+such that $\pr\{Y= b\} = m_b/n$ for all $b\in B$. Now partition $[n]$ into sets $\{E_b\}_{b\in B}$,
+where $|E_b| = m_b$ for all $b\in B$. We define a random variable $Z$ by sampling uniformly at
+random from $E_b$ if $Y = b$, and doing so independently of $(X\given Y=b)$. The result is
+a random variable $Z$ that is uniform on $[n]$ and which is independent of $X\given Y$.
+Furthermore, since $Z$ determines $Y$, we have $\Eta\{Z\} = \Eta\{Y,Z\}$ by the invariance
+axiom. Hence
+$$\eqalign{
+\Eta\{X\given Y\} &= \Eta\{X\given Y, Z\} \cr
+&= \Eta\{X,Y,Z\} - \Eta\{Y,Z\} \cr
+&= \Eta\{X,Z\} - \Eta\{Z\} \cr
+&= \Eta\{X\given Z\} \cr
+&\le \Eta\{X\},\cr
+}$$
+where the inequality on the last line follows from the previous paragraph.
+
+The general case follows from the continuity axiom and the fact that any discrete random
+variable can be approximated by discrete random variables with finite support and rational
+probabilities.\slug
+
+By the additivity axiom, the previous proposition is equivalent to
+$$\Eta\{X,Y\} \le \Eta\{X\} + \Eta\{Y\}.$$
+and we shall use this to prove the following submodularity inequality.
+
+\parenproclaim Proposition {\advthm} (Submodularity).
+Suppose $X$, $Y$, $Z$, and $W$ are random variables such that $(Z,W)$ determines
+$X$, $Z$ determines $Y$, and $W$ also determines $Y$. Then
+$$\Eta\{X\} + \Eta\{Y\} \le \Eta\{Z\} + \Eta\{W\}.$$
+
+\proof The hypotheses give the three inequalities
+$$\Eta\{X\} \le \Eta\{Z,W\}, \quad \Eta\{Y\}\le \Eta\{Z\},\quad\hbox{and}\quad \Eta\{Y\}\le \Eta\{W\}.$$
+From this we see that
+$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z,W\} + \Eta\{Z\} + \Eta\{Y\}.$$
+But then since conditioning does not increase entropy, we have
+$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z\} + 2\Eta\{Y\},$$
+whence dividing both sides by $2$ completes the proof.\slug
+
+The submodularity inequality is often stated in terms of a triple of random variables in terms of
+the {\it conditional mutual information}, which is defined by
+$$\II\{ X : Y \given Z\} = \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\}.$$
+
+\proclaim Proposition \advthm. Let $X$, $Y$, and $Z$ be discrete random variables. Then
+$$\Eta\{X,Y,Z\} + \Eta\{Z\} \le \Eta\{X,Z\} + \Eta\{Y,Z\},$$
+which is equivalent to
+$$\II\{X : Y\given Z\} \ge 0.$$
+
+\proof It is clear from the additivity axiom that both sides in the first
+inequality are equal to $\Eta\{X\} + \Eta\{Y\} + 2\Eta\{Z\}$
+if and only if $X$ and $Y$ are independent conditional on $Z$.
+
+To prove the first inequality, note that $(X,Y,Z)$ is jointly determined by $(X,Z)$ and $(Y,Z)$,
+and $Z$ is determined by both $(X,Z)$ and $(Y,Z)$ separately, then apply the previous proposition.
+Now by the definition of conditional mutual information,
 $$\eqalign{
@@
 }$$
-so we see that submodularity of entropy is equivalent to the statement that $\II\{X : Y\given Z\} \ge 0$.
-Analogously to the unconditional case, we have equality if and only if $X$ and $Y$ are
-independent when conditioned on $Z$.
+and this proves that the first statement is equivalent to the second.\slug
 
@@
 
-\edef\maxsumsetbound{\the\thmcount}
+\edef\maxsumsetbound{\the\sectcount.\the\thmcount}
 \proclaim Proposition \advthm. Let $X$ and $Y$ be random variables each taking finitely
@@
 of the same abelian group, we can compute the Ruzsa distance
-$$d(A,B) = \log {|A-B|\over \sqrt{|A|\cdot|B|}}$$
+$$d(A,B) = \lg {|A-B|\over \sqrt{|A|\cdot|B|}}$$
 between them. (This satisfies all the axioms of a metric except the one requiring $d(A,A) = 0$ for all sets
@@
 
+\medskip\boldlabel The sum-difference inequality.
+The Ruzsa triangle inequality bounds the size of a difference set by passing through a different subset. There
+is another inequality that relates the size of a sumset with the size of a difference set.
+If $A$ and $B$ are nonempty finite subsets of an abelian group $G$, then
+$$|A+B| \le {|A-B|^3\over |A||B|}.$$
+If we replace cardinalities by exponentials of entropies, then we obtain the statement of the following
+proposition.
+
+\proclaim Proposition \advthm.
+Let $X$ and $Y$ be independent random variables taking values in the same abelian group. Then
+$$\Eta\{X+Y\} \le 3\Eta\{X-Y\} - \Eta\{X\} - \Eta\{Y\}.$$
+
+Before we proceed to the proof, we establish the following definition, which will be needed later in these
+notes as well. Let $X$ and $Y$ be random variables (not necessarily independent). We say that
+$X_1$ and $Y_1$ are {\it conditionally independent trials of $X$ and $Y$ relative to $Z$} if for all
+$z$ in the range
+of $Z$, the random variables distributed as $(X_1\given Z=z)$ and $(Y_1\given Z=z)$ are independent,
+$(X_1\given Z=z)$ has the same distribution as $(X\given Z=z)$,
+and similarly for $Y_1$ and $Y$.
+In particular, if $X=Y$ and $X_1$ and $X_2$ are conditionally independent trials of $X$ relative
+to $Z$, we have
+$$\Eta\{X_1,X_2\given Z\} = \Eta\{X_1\given Z\} + \Eta\{X_2\given Z\} = 2\Eta\{X\given Z\},$$
+by additivity and independence. From this we obtain
+\edef\eqcondindep{\the\eqcount}
+$$\Eta\{X_1,X_2,Z\} = 2\Eta\{X\given Z\} + \Eta\{Z\} = 2\Eta\{X,Z\} - \Eta\{Z\}.\adveq$$
+It is also important to observe that $(X_1,Z)$ and $(X_2,Z)$ both have the same distributions
+as $(X,Z)$.
+
+\proof Let $(X_1,Y_1)$ and $(X_2,Y_2)$ be conditionally independent trials of $(X,Y)$ relative
+to $X-Y$. Since $(X,Y)$ determines $X-Y$, we have $X_1-Y_1 = =X-Y= X_2-Y_2$. Let $(X_3,Y_3)$ be another
+trial of $(X,Y)$ independent of $(X_1, X_2, Y_1, Y_2)$. Then
+$$X_3 + Y_3 = X_3 + Y_3 + X_1 - Y_1 - X_2 + Y_2 = (X_3-Y_2) - (X_1-Y_3) + X_2 + Y_1,$$
+so $(X_3-Y_2, X_1-Y_3, X_2, Y_1)$ and $(X_3, Y_3)$ each determine $X_3+ Y_3$. On the other hand,
+$(X_3-Y_2, X_1-Y_3, X_2, Y_1)$ and $(X_3,Y_3)$ together determine the sextuple $(X_1,X_2,X_3,Y_1,Y_2,Y_3)$,
+so by the submodularity inequality, we have
+\edef\eqsumdiff{\the\eqcount}
+$$\eqalign{
+\Eta\{X_1,X_2,X_3,Y_1,Y_2,&Y_3\} + \Eta\{X_3+Y_3\} \cr
+&\le \Eta\{X_3-Y_2, X_1-Y_3, X_2, Y_1\} + \Eta\{X_3, Y_3\}.\cr
+}\adveq$$
+We have
+$$\Eta\{X_3,Y_3\} = \Eta\{X,Y\} = \Eta\{X\} + \Eta\{Y\}$$
+by independence of
+$X$ and $Y$, and
+since $(X_3,Y_3)$ and $(X_1,X_2,Y_1,Y_2)$ are independent, we have
+$$\eqalign{
+\Eta\{X_1,X_2,X_3,Y_1,Y_2,Y_3\} &= \Eta\{X_1,X_2,Y_1,Y_2\} + \Eta\{X_3,Y_3\} \cr
+&= \Eta\{X_1, Y_1,X_2,Y_2,X-Y\} + \Eta\{X\} + \Eta\{Y\} \cr
+&= 2\Eta\{X,Y,X-Y\} - \Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\} \cr
+&= 2\Eta\{X,Y\} - \Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\}\cr
+&= 3\Eta\{X\} + 3\Eta\{Y\} - \Eta\{X-Y\},\cr
+}$$
+where in the third line we applied~\refeq{\eqcondindep}. On the other hand,
+$$\Eta\{X_3 + Y_3\} = \Eta\{X+Y\}$$
+and
+$$\eqalign{
+\Eta\{X_3-Y_2, &X_1-Y_3, X_2, Y_1\} \cr
+&\le \Eta\{X_3-Y_2\} + \Eta\{X_1-Y_3\} + \Eta\{X_2\} + \Eta\{Y_1\} \cr
+&= 2\Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\}.
+}$$
+Substituting everything into~\refeq{\eqsumdiff} yields
+$$\eqalign{
+3\Eta\{X\} + 3\Eta\{Y\} - &\Eta\{X-Y\} + \Eta\{X+Y\} \cr
+&\le 2\Eta\{X-Y\} + 2\Eta\{X\} + 2\Eta\{Y\},
+}$$
+and the desired inequality follows upon rearrangement of terms.\slug
+
+The entropic sum-difference inequality can also be stated in terms of entropic Ruzsa distances;
+independence is not necessary here because independent trials are baked into the definition
+of entropic Ruzsa distance.
+
+\proclaim Corollary \advthm. Let $X$ and $Y$ be discrete random variables taking values in the same
+abelian group. Then
+$$\dd\{X, -Y\} \le 3\dd\{X,Y\}.\noskipslug$$
+
 \advsect The Pl\"unnecke--Ruzsa inequality
@@
 
-\parenproclaim Theorem A (Pl\"unnecke--Ruzsa inequality). Let $A$ and $B$ be finite subsets of an
+\parenproclaim Theorem {\advthm} (Pl\"unnecke--Ruzsa inequality). Let $A$ and $B$ be finite subsets of an
 abelian group and suppose that $|A+B|\le K|A|$ for some constant $K$. Then for any integers
@@
 
-% \medskip\boldlabel The formula for entropy.
-% Now we finally give the formula for the entropy of a random variable. For a discrete random
-% variable $X$, we let $p_X : S\to [0,1]$ denote the probability mass function of $X$, given by
-% $p_X(x) = \pr\{X = x\}$ for all $x\in S$. Unless subscripted, $\log$ always denotes the natural logarithm.
-% 
-% \proclaim Proposition \advthm. Let $X$ be a random variable taking values in a finite set $S$.
-% Then $\Eta\{X\}$ satisfies axioms (a) through (e) as well as the additional normalisation axiom (f)
-% if and only if
-% $$\Eta\{X\} = \ex\bigl\{ 1/\log p_X(X) \bigr\} = \sum_{x\in S} p_X(x) \log {1\over p_X(s)},$$
-% where we adopt the convention that $0\log(1/0) = 0$.
-% 
-% \proof [TODO: Prove this.]\slug
-% 
-% Equipped now with a formula for the entropy $\Eta\{X\}$ of a random variable $X$, we can prove further useful
-% facts. We saw in the previous section that the entropy of a uniform random variable whose range is $\{0,1\}^n$
-% is $\log_2(e)\cdot n = \log(2^n)$, and we know, by the maximisation axiom, that this is the largest
-% the entropy of any $X$ taking values in $\{0,1\}^n$ can be. The following proposition generalises
-% this notion.
-% 
-% \proclaim Proposition \advthm. For any random variable $X$ taking values in a finite set $S$,
-% we have
-% $$\Eta\{X\} \le \log |S|.$$
-% 
-% \proof By Jensen's inequality, we have
-% $$\eqalign{
-% \Eta\{X\} &= \ex\bigl\{ \log(1/p_X(X)\bigr\} \cr
-% &\le \log \ex\bigl\{ 1/p_X(X)\bigr\} \cr
-% &= \log \sum_{x\in S} \pr\{X = x\} {1\over \pr\{X = x\}} \cr
-% &= \log |S|.\noskipslug \cr
-% }$$
+\advsect The Balog--Szemer\'edi--Gowers theorem
+
+If $A$ and $B$ are subsets of the same abelian group such that $|A+B|$ is much smaller than $|A|\cdot|B|$,
+then it stands to reason that there must be a lot of redundancy in $A+B$; that is, many elements of $A+B$ can
+be expressed as $a+b$ in lots of different ways. To capture this notion, we can define the
+{\it additive energy} between two sets $A$ and $B$ to be
+$$E(A,B) = \bigl| \bigl\{ (a,a',b,b') \in A\times A\times B\times B : a + b = a'+b'\bigr\}\bigr\}.$$
+
+We now define an entropic version of $E(A,B)$. Let $X$ and $Y$ be discrete random variables taking values
+in the same
+abelian group. Let $(X_1, Y_1)$ and $(X_2, Y_2)$ be conditionally independent trials of $(X,Y)$ relative to
+$X+Y$. These  $X_1 + Y_1 = X_2 + Y_2 = A+B$.
+The {\it entropic additive energy} between $X$ and $Y$ is
+$$ \ee\{X,Y\} = \Eta\{X_1, Y_1, X_2, Y_2\}.$$
+This definition makes clear the analogy between this value and the additive energy of sets, but
+by conditional independence and the fact that $(X_1, Y_1, X_2, Y_2)$ determines $X_1+Y_1 = X+Y$,
+we can rewrite
+$$\eqalign{
+\ee\{X,Y\} &= \Eta\{X_1, Y_1, X_2, Y_2, X+Y\} \cr
+&= 2\Eta\{X,Y,X+Y\} - \Eta\{X+Y\} \cr
+&= 2\Eta\{X,Y\} - \Eta\{X+Y\},\cr
+}$$
+where in the second equality we applied~\refeq{\eqcondindep}. This formula is something we will use
+often, as it makes no direct mention of the variables $(X_1, Y_1)$ and $(X_2,Y_2)$.
+
+Quantifying the idea that small sumset must imply large additive energy, we have the following
+proposition.
+
+\edef\propinversebalog{\the\sectcount.\the\thmcount}
+\proclaim Proposition \advthm.
+Let $A$ and $B$ be finite subsets of an abelian group.
+If $|A+B|\le K|A|^{1/2} |B|^{1/2}$ for some constant $K$, then we have
+$$E(A,B) \ge {1\over K} |A|^{3/2} |B|^{3/2}.\noskipslug$$
+
+Somewhat surprisingly, if we convert
+these statements into their entropic analogues in the na\"ive way, as we've been doing,
+the implication goes the other way!
+However, we have a weak equivalence (with worse constants in one direction)
+under the further assumption that the random variables in
+question are not too dependent.
+
+\edef\propnaive{\the\sectcount.\the\thmcount}
+\proclaim Proposition \advthm. Let $X$ and $Y$ be discrete random variables taking values in the
+same abelian group. If
+\global\edef\eqenergybound{\the\eqcount}
+$$\ee\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2}\Eta\{Y\} - \log K,\adveq$$
+for some constant $K$, then
+\global\edef\eqsumbound{\the\eqcount}
+$$\Eta\{ X+Y\} \le {1\over 2} \Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K.\adveq$$
+If one adds the further assumption that
+$\Eta\{X,Y\} \ge \Eta\{X\} + \Eta\{Y\} - C$,
+then~{\rm \refeq{\eqsumbound}} implies~{\rm \refeq{\eqenergybound}} with a worse constant, namely,
+we may only conclude
+\global\edef\eqworsebound{\the\eqcount}
+$$\ee\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2}\Eta\{Y\} - \log K - 2C.\adveq$$
+In particular, if $X$ and $Y$ are independent, then we can recover~{\rm \refeq{\eqenergybound}}
+from~{\rm \refeq{\eqsumbound}}.
+
+\proof Assuming the lower bound on the additive energy, we have
+$$2\Eta\{X,Y\} - \Eta\{X+Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} - \log K,$$
+so
+$$\eqalign{
+\Eta\{X+Y\} &\le 2\Eta\{X,Y\} - {3\over 2} \Eta\{X\} - {3\over 2} \Eta\{Y\} + \log K \cr
+&\le 2\Eta\{X\} + 2\Eta\{Y\} - {3\over 2} \Eta\{X\} - {3\over 2} \Eta\{Y\} + \log K \cr
+&= {1\over 2} \Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K .\cr
+}$$
+On the other hand, assuming this upper bound on $\Eta\{X+Y\}$, we have
+$$\eqalign{
+\ee\{X,Y\} &= 2\Eta\{X,Y\} - \Eta\{X+Y\}  \cr
+&\ge 2\Eta\{X,Y\} - {1\over 2} \Eta\{X\} - {1\over 2} \Eta\{Y\} - \log K ,\cr
+}$$
+and if $2\Eta\{X,Y\} \ge 2 \Eta\{X\} + 2\Eta\{Y\} - 2C$, then~\refeq{\eqworsebound} follows
+directly.\slug
+
+The fact that the implication goes the ``wrong'' way may seem somewhat baffling at first. We will now get
+to the bottom of this.
+If $A$ and $B$ are subsets of a finite abelian group, we can let $X$ and $Y$ be the uniform
+distributions on $A$ and $B$, respectively. We have been operating under the belief that $\Eta\{X+Y\}$
+should correspond (up to taking powers or logarithms) to the size of $A+B$. But this is not true, since
+$X$ and $Y$ may be given a joint distribution that is not uniform on $A\times B$, even if
+its marginals are uniform on $A$ and $B$.
+
+For example, let $A$ and $B$ be subsets of $G$ and consider any regular bipartite graph $H$ on the
+vertex set $A\cup B$. Let $(X,Y)$ be defined by sampling an edge from $H$ uniformly at random, letting
+$X$ be its endpoint in $A$ and $Y$ be its endpoint in $B$. Since the graph is regular, $X$ is uniform
+on $A$ and $Y$ is uniform on $B$, but $X+Y$ can only take values $a+b$ where $(a,b)$ is an edge of $H$.
+In other words, $X+Y$ samples from the {\it partial sumset}
+$$ A +_H B = \bigl\{ a + b : (a,b)\in E(H)\bigr\},$$
+where elements that are represented more times as the sum of edge endpoints are given a greater weight.
+The way to properly recover the ordinary sumset $A+B$ is to let $H$ be all of $A\times B$, in which case
+$X$ and $Y$ are independent. The extra assumption we added in Proposition~{\propnaive} is analogous to
+stipulating that $|H| \ge K |A|\cdot |B|$, so that the resulting $X$ and $Y$ are ``nearly'' independent.
+
+Simply put, in the entropic setting the bound~\refeq{\eqenergybound} is
+stronger than~\refeq{\eqsumbound} because the entropy $\Eta\{X,Y\}$ of the joint distribution appears in
+the formula for $\ee\{X,Y\}$, whereas~\refeq{\eqsumbound} says nothing whatsoever about this joint distribution.
+
+The converse to Proposition~{\propinversebalog} does not hold in general; that is, large additive energy
+does not necessarily imply a small sumset. However, there does exist a partial converse,
+which says that if sets $A$ and $B$ have a large additive energy, then there are dense subsets $A'\subseteq A$
+and $B'\subseteq B$ such that the sumset $|A+B|$ is small. This is the celebrated
+Balog--Szemer\'edi--Gowers theorem.
+
+\parenproclaim Theorem {\advthm} (Balog--Szemer\'edi--Gowers theorem). Let $A$ be a finite
+subset of an abelian group with $E(A,B) \ge c |A|^{3/2} |B|^{3/2}$. Then there are subsets
+$A'\subseteq A$ and $B'\subseteq B$ with $|A'| \ge c' |A|$ and $|B'|\ge c''|B|$ such that
+$$|A'+B'| \le C |A|^{1/2}|B|^{1/2},$$
+where $c'$, $c''$, and $C$ depend only on $c$.
+
+In the entropy setting, the operation on random variables that corresponds to taking subsets is
+to conditioning. (As a sanity check, recall that conditioning never increases entropy, just
+as taking subsets never increases cardinality.) The Balog--Szemer\'edi--Gowers theorem gives
+us subsets between which we can take a {\it bona fide} sumset, so its entropic analogue
+should return conditionings $X'$ and $Y'$ of $X$ and $Y$ relative to some random variable $Z$,
+such that
+\medskip
+\item{i)} $X'$ and $Y'$ are conditionally independent relative to $Z$;
+\smallskip
+\item{ii)} the entropies $\Eta\{X'\given Z\}$ and
+$\Eta\{Y'\given Z\}$ are not too small compared to their unconditioned analogues; and
+\smallskip
+\item{iii)} $\Eta\{X'+Y'\given Z\}$ is small.
+\medskip
+In fact, the conditioning we shall perform is exactly the one used to define additive energy.
+
+First, we need a lemma, which we state separately since it will also be used later in the proof of
+the polynomial Freiman--Ruzsa theorem.
+
+\edef\lematwo{\the\sectcount.\the\thmcount}
+\proclaim Lemma \advthm. Let $X$ and $Y$ be discrete random variables taking values in the same abelian
+group. Let $(X_1, Y_1)$ and $(X_2,Y_2)$ be conditionally
+independent trials of $(X,Y)$ relative to $X+Y$. Then we have
+$$\max\bigl(\Eta\{X_1- X_2\}, \Eta\{X_1-Y_2\}\bigr) \le \Eta\{X+Y\} - \II\{X:Y\}.$$
+The right-hand side of this expression can also be written $2\Eta\{X\} + 2\Eta\{Y\} - \ee\{X,Y\}$.
+
+\proof First we perform the proof for $X_1 - Y_2$.
+Submodularity gives us
+$$\Eta\{X_1, Y_1, X_1-Y_2\} + \Eta\{X_1-Y_2\} \le
+\Eta\{ X_1, X_1-Y_2\} + \Eta\{Y_1, X_1-Y_2\}.$$
+Since $X_1 + Y_1 = X+Y = X_2 + Y_2$, given $(X_1, Y_1, X_1-Y_2)$ we can recover
+the values of $X_2$ and $Y_2$. So $(X_1, Y_1, X_1-Y_2)$ and $(X_1,Y_1,X_2,Y_2)$ determine each
+other and hence
+$$\Eta\{X_1, Y_1, X_1-Y_2\} = \Eta\{X_1, Y_1, X_2, Y_2\} = 2\Eta\{X,Y\} - \Eta\{X+Y\}.$$
+On the other side of the inequality, we have
+$$\Eta\{ X_1, X_1-Y_2\} = \Eta\{X_1, Y_2\} \le \Eta\{X\} + \Eta\{Y\},$$
+and similarly
+$$\Eta\{ Y_1, X_1-Y_2\} = \Eta\{Y_1, X_2-Y_1\} = \Eta\{X_2, Y_1\} \le \Eta\{X\} + \Eta\{Y\}.$$
+Therefore,
+$$\eqalign{
+\Eta\{X_1-Y_2\} &\le \Eta\{X+Y\} - 2\Eta\{X,Y\} + 2\Eta\{X\} + 2\Eta\{Y\} \cr
+&= \Eta\{X+Y\} - \II\{X : Y\}.\cr
+}$$
+The same holds with the roles of $X_2$ and $Y_2$ exchanged.\slug
+
+\parenproclaim Theorem {\advthm} (Entropic Balog--Szemer\'edi--Gowers theorem).
+Let $X$ and $Y$ be discrete random variables taking values in the same abelian group, and
+suppose that
+$$\ee\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} + \log K$$
+for some constant $K$.
+Then letting $(X_1, Y_1)$ and $(X_2,Y_2)$ be conditionally
+independent trials of $(X,Y)$ relative to $X+Y$, we have
+$$ \Eta\{ X_1 \given X+Y\} \ge \Eta\{X\} - 2\log K$$
+and
+$$ \Eta\{ Y_2 \given X+Y\} \ge \Eta\{Y\} - 2\log K.$$
+Furthermore, the variables $X_1$ and $Y_2$ are conditionally independent relative to $X+Y$, and
+we have
+$$\Eta\{ X_1 + Y_2 \given X+Y\} \le {1\over 2}\Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K.$$
+
+\proof Using the coupling $X+Y = X_1+Y_1 = X_2 + Y_2$, we have
+$$\eqalign{
+\Eta\{X_1\given X + Y\} &= \Eta\{X_1, X_1+Y_1\} - \Eta\{X+Y\} \cr
+&= \Eta\{X,Y\} - \Eta\{X+Y\} \cr
+&= \ee\{X,Y\} - \Eta\{X,Y\} \cr
+&\ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} - \log K - \Eta\{X,Y\} \cr
+&\ge {1\over 2} \Eta\{X\} + {1\over 2}\Eta\{Y\} -\log K.\cr
+}$$
+where in the fourth line we used the hypothesis on $\ee\{X,Y\}$, and in the last line we observed
+that $\Eta\{X\} + \Eta\{Y\} - \Eta\{X,Y\} \ge 0$. The bound
+$$\Eta\{Y_2, X+Y\} \ge {1\over 2} \Eta\{X\} + {1\over 2}\Eta\{Y\} -\log K$$
+is shown in the exact same way; only the first step differs.
+
+Now, taking the sum of both these bounds, we arrive at
+$$\Eta\{X_1\given X+Y\} + \Eta\{Y_2 \given X+Y\} \ge \Eta\{X\} + \Eta\{Y\} - 2\log K.$$
+From this one deduces
+$$\eqalign{
+\Eta\{X_1\given X+Y\} &\ge \Eta\{X\} + \Eta\{Y_2\} - \Eta\{Y_2\given X+Y\}  - 2\log K \cr
+&\ge \Eta\{X\} - 2\log K.\cr
+}$$
+The corresponding lower bound on $\Eta\{Y_2\given X+Y\}$ is proved similarly.
+
+It remains to prove the upper bound on $\Eta\{X_1 + Y_2 \given X+Y\}$.
+Note that $(X_1, Y_2, X+Y)$ and $(X_1 - X_2, X+Y)$ jointly determine $(X_1, X_2, X+Y)$.
+Then given $X_1 - X_2$ and $X+Y$ we can calculate
+$$X_1 + Y_2 = X_1 - X_2 + X_2 + Y_2 = (X_1 - X_2) + (X+Y),$$
+so $(X_1, Y_2, X+Y)$ and $(X_1 - X_2, X+Y)$
+each separately determine $(X_1 + Y_2, X+Y)$. Hence the submodularity inequality yields
+$$ \Eta\{X_1, X_2, X+Y\} + \Eta\{X_1+Y_2, X+Y\} \le \Eta\{X_1, Y_2, X+Y\} + \Eta\{X_1- X_2, X+Y\}.$$
+From $(X_1, X_2, X+Y)$
+we can calculate $Y_1 = X+Y-X_1$ and $Y_2 = X+Y-X_2$, so this triple and the triple
+$(X_1, X_2, Y_1, Y_2)$ determine each other. So the first term above is simply the additive energy
+between $X$ and $Y$; that is
+$$ \Eta\{X_1, X_2, X+Y\} = \Eta\{X_1, X_2, Y_1, Y_2\} = 2\Eta\{X,Y\} - \Eta\{X+Y\}.$$
+Now since $X_1$ and $Y_2$ are conditionally independent relative to $X+Y$, we have
+$$\eqalign{
+\Eta\{X_1, Y_2, X+Y\} &= \Eta\{X_1, X+Y\} + \Eta\{Y_2, X+Y\} - \Eta\{X+Y\} \cr
+&= \Eta\{X, X+Y\} + \Eta\{Y, X+Y\} - \Eta\{X+Y\} \cr
+&= 2\Eta\{X,Y\} - \Eta\{X+Y\} \cr
+}$$
+For the last term above we split
+$$\Eta\{X_1 - X_2, X+Y\} = \Eta\{X_1 - X_2 \given X+Y\} - \Eta\{X+Y\}.$$
+Putting everything together, we obtain
+$$\eqalign{
+2\Eta\{X,Y\} - \Eta\{X+&Y\} + \Eta\{X_1 + Y_2, X+Y\} \cr
+&\le 2\Eta\{X,Y\} + \Eta\{X_1-X_2\given X+Y\} - 2\Eta\{X+Y\},\cr
+}$$
+so that
+$$\Eta\{ X_1 + Y_2\given X+Y\} \le \Eta\{X_1 - X_2\given X+Y\} - 2\Eta\{X+Y\}\le \Eta\{X_1 - X_2\}.$$
+The previous lemma then gives
+$$\Eta\{ X_1 + Y_2\given X+Y\} \le 2\Eta\{X\} + 2\Eta\{Y\} - \ee\{X,Y\},$$
+and from our lower bound on $\ee\{X,Y\}$, we conclude that
+$$\Eta\{ X_1 + Y_2\given X+Y\} \le {1\over 2} \Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K,$$
+which is what we wanted to show.\slug
+
+\advsect The Freiman--Ruzsa theorem
+
+Both Pl\"unnecke's theorem and the Balog--Szemer\'edi--Gowers theorem have to do with the ratio $|A+A|/|A|$
+of a finite set $A$. Let us now give this ratio a name. It is called the {\it doubling constant} of $A$.
+Pl\"unnecke's theorem says that if the doubling constant is at most $K$,
+then the ratio $|rA-sA|/|A|$ is at most $K^{r+s}$. The Balog--Szemer\'edi--Gowers theorem, on the other hand,
+says that sets $A$ with large additive energy contain large subsets $A$ and $A''$ such that
+$|A' + A''|/|A|$ is bounded from above by some constant.
+
+It is natural to ask whether there is some way to characterise the structure of sets $A$ with small doubling
+constant (in some asymptotic sense). One reason why $A$ might have $|A+A|\le K|A|$ is if $A$ is contained
+in some subgroup $H$, since subgroups are closed under addition. If $H$ is not much larger than $A$,
+then this would explain why the doubling constant of $A$ is small. It turns out that the converse of
+this statement holds; that is, if $A$ has small doubling constant, then it has high density as a subset
+of a subgroup. This is called the Freiman--Ruzsa theorem.
+
+\parenproclaim Theorem {\advthm} (Freiman--Ruzsa theorem). Let $A\subseteq G = (\ZZ/r\ZZ)^n$ and suppose
+that $|A+A|\le K|A|$. Then there is a subgroup $H$ of $G$ such that $A\subseteq H$ and $|H|\le K'|A|$,
+where $K'$ depends only on $K$ and $r$.
+
+But subspaces may not be a very efficient way of covering sets with small doubling. One can easily
+cook up examples where there is enough linear independence in $A$ so that the smallest subspace
+containing $A$ has size exponential in $|A|$.
+
+Returning to first
+principles, observe that if the doubling
+constant of $A$ is $1$, then the structure of $A$ is completely determined, as shown by the following proposition.
+
+\proclaim Proposition \advthm. Let $A$ be a finite subset of an abelian group $G$. If $|A+A| = |A|$,
+then $A$ is a coset of a subgroup $H$.
+
+\proof First we assume that $A$ contains $0$. Then
+$$A = A + \{0\} \subseteq A+A,$$
+but since $|A+A| = |A|$ this implies that $A+A = A$. Now let $H = \{h\in G : A + h = A$. The above observation
+shows that $A\subseteq H$. But if $h\in H$, then $A = A+h$ contains the element $0+h = h$, so in
+fact $H=A$.
+
+If $x,y\in H$, then $A+x = A$ and $A+y=A$, so adding $y$ to both sides of the second identity we obtain
+$A-y = A$, and adding $x$ to both sides of this, we get
+$$A + x-y = A+x = A,$$
+so $x-y\in H$.
+This along with the fact that $0\in H$ shows that $H$ is a subgroup of $G$.
+
+Thus in the case where $A$ contains $0$, we see that $A$ is actually a subgroup of $G$, and
+in the general case we may translate $A$ without changing $|A+A|$, so $A$ is the translate of a subgroup;
+that is, $A$ is a coset of $H$.\slug
+
+So in the case that $A$ is a union of not too many cosets, we also expect $A+A$ to be quite small, since
+each individual coset does not grow under addition (the only possible growth comes from additions between cosets).
+
+It is believed that the converse of this statement holds and yields a more efficient version
+of the Freiman--Ruzsa theorem; that is,
+if $A$ has doubling constant bounded above by $K$, then $A$ is contained in a union of cosets,
+where the number of cosets needed can be taken to be no more than polynomial in the doubling constant.
+This conjecture is due to K.~Marton.
+
+\proclaim Conjecture \advthm. Let $A\subseteq \bigl(\ZZ/r\ZZ\bigr)^n$ have $|A+A|\le K|A|$
+for some constant $K$. Then there is a subgroup $H$ with $|H|\le |A|$ such that $A$
+is contained in a union of $K^C$ cosets of $H$, where $C$ is a constant that can depend
+on $r$ but not on $n$ or $K$.
+
+In 2023, the first special case of this conjecture was proved by W.~T.~Gowers, B.~Green, F.~Manners, and
+T.~Tao.
+
+\edef\thmpfr{\the\sectcount.\the\thmcount}
+\parenproclaim Theorem {\advthm} (Gowers--Green--Manners--Tao, {\rm 2023}).
+There is a constant $C$ such that the following holds.
+Let $A\subseteq \FF_2^n$ have $|A+A|\le K|A|$
+for some constant $K$. Then there is a subgroup $H$ with $|H|\le |A|$ such that $A$
+is contained in a union of $2K^C$ cosets of $H$.
+
+The proof, which is almost entirely information-theoretic in nature, will be our main concern for the
+remainder of these notes.
 

### 2024-02-22 misc/expo/entropy.tex
PFR progress

@@
 \def\ee{\op{\bf e}}
+\def\P{{\cal P}}
 \def\given{\mathbin{|}}
-\def\argmax{\limitop{arg$\;$max}}
+\def\argmax{\limitop{\rm arg$\,$max}}
+\def\dTV{d_{\rm TV}}
 \def\advthm{\the\sectcount.\the\thmcount\global\advance \thmcount by 1}
@@
 
-\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{15 February 2024}
+\maketitle{Entropy and additive combinatorics}{}{Marcel K. Goh}{15 February 2024}
 
@@
 
+The continuity axiom presupposes a topology on the set of all distributions on the target
+space of $X$. This topology will only be important here and there, so we will not give a complete
+description of it now, but simply introduce the relevant properties as needed later on.
+
 We shall take it on faith that there really exists a function on random variables satisfying
@@
 up to a multiplicative constant, so we shall add the following axiom.
-(It is very possible you have met this function somewhere on your travels, but you will not find it anywhere in
+(It is very possible you have met the formula for entropy
+somewhere on your travels, but you will not find it anywhere in
 these notes.)
@@
 
-\edef\lemmaxprob{\the\sectcount.\the\thmcount}
+\edef\propmaxprob{\the\sectcount.\the\thmcount}
 \proclaim Proposition \advthm. Let $X$ be a discrete random variable supported on a set $A$ and let
@@
 The proof, which is due to C.~West, uses an argument similar to the one
-we used to prove Lemma~{\lemmaxprob}.
+we used to prove Proposition~{\propmaxprob}.
 
@@
 The general case follows from the continuity axiom and the fact that any discrete random
-variable can be approximated by discrete random variables with finite support and rational
-probabilities.\slug
+variable, regarded as a vector in $l_1(\RR)$,
+by vectors with finite support, and these in turn can be approximated by vectors of finite support
+and rational coordinates.\slug
 
@@
 
+\edef\cormaxsumineq{\the\sectcount.\the\thmcount}
 \proclaim Corollary \advthm. If $X$ and $Y$ are independent, then
@@
 This definition only depends on the individual distributions of $X$ and $Y$ and does not require them to
-have the same sample space. Once again, we don't necessarily have $\dd\{X,X\} = 0$, but we do have
+have the same sample space. When $X$ and $Y$ are independent, the Ruzsa distance enjoys some
+nice properties.
+
+\edef\propabsdist{\the\sectcount.\the\thmcount}
+\proclaim Proposition \advthm. Let $X$ and $Y$ be independent taking values in the same group. Then
+$$\bigl| \Eta\{X\} - \Eta\{Y\} \bigr| \le 2\dd\{X,Y\}$$
+and
+$$\max\bigl( \Eta\{X-Y\} - \Eta\{X\}, \Eta\{X-Y\} - \Eta\{Y\} \bigr) \le 2\dd\{X,Y\}$$
+
+\proof By independence we have
+$$2\dd\{X,Y\} = 2\Eta\{X-Y\} - \Eta\{X\} - \Eta\{Y\}\over 2,$$
+Corollary~{\cormaxsumineq} tells us that $2\Eta\{X-Y\} \ge 2\Eta\{X\}$, so this is at least $\Eta\{X\}-\Eta\{Y\}$.
+The same corollary also says that $2\Eta\{X-Y\} \ge 2\Eta\{Y\}$, which allows us to add the absolute value bars
+to the inequality.
+
+The second inequality is proved using the same corollary, but only applying it to one of the $\Eta\{X-Y\}$
+terms.\slug
+
+Similarly to the Ruzsa distance on sets, we don't necessarily have $\dd\{X,X\} = 0$, but we do have
 the triangle inequality, which shall now prove.
@@
 {\it additive energy} between two sets $A$ and $B$ to be
-$$E(A,B) = \bigl| \bigl\{ (a,a',b,b') \in A\times A\times B\times B : a + b = a'+b'\bigr\}\bigr\}.$$
+$$E(A,B) = \bigl| \bigl\{ (a,a',b,b') \in A\times A\times B\times B : a + b = a'+b'\bigr\}\bigr|.$$
 
@@
 
+\edef\propdoublingone{\the\sectcount.\the\thmcount}
 \proclaim Proposition \advthm. Let $A$ be a finite subset of an abelian group $G$. If $|A+A| = |A|$,
@@
 
-If $x,y\in H$, then $A+x = A$ and $A+y=A$, so adding $y$ to both sides of the second identity we obtain
+If $x,y\in H$, then $A+x = A$ and $A+y=A$, so subtracting $y$ from both sides of the second identity we obtain
 $A-y = A$, and adding $x$ to both sides of this, we get
@@
 
-The proof, which is almost entirely information-theoretic in nature, will be our main concern for the
-remainder of these notes.
+The proof of this theorem will occupy the
+remainder of these notes. As a first step, we shall show that
+Theorem~{\thmpfr} follows from an entirely information-theoretic statement.
+In this and the rest of the notes, for $A$ finite
+we use the notation $U_A$ to denote the random variable that is uniform on $A$.
+
+\edef\thmsubgroup{\the\sectcount.\the\thmcount}
+\proclaim Theorem \advthm. There is a constant $C'$ such that the following holds.
+Let $G = \FF_2^n$ and let $X_1^*$ and $X_2^*$ be random variables taking values in
+$G$. There is some subgroup $H\subseteq G$ such that
+$$\dd\{ X_1^*, U_H\} + \dd\{ X_2^*, U_H\} \le C'\dd\{X_1^*, X_2^*\}.$$
+
+In the proof that this theorem implies the previous one, we shall need the Ruzsa covering lemma,
+whose proof is short enough that we include it for completeness.
+
+\edef\lemruzsacovering{\the\sectcount.\the\thmcount}
+\parenproclaim Lemma {\advthm} (Ruzsa covering lemma). Let $G$ be an abelian group and let $A$ and
+$B$ be finite subsets of $Z$. Then there is a set $L\subseteq A$ of size at most $|A+B|/|B|$ such that
+$A\subseteq L+B-B$.
+
+\proof Let $L = \{a_1,\ldots,a_l\}$ be a maximal subset of $A$ with the property that the sets $a_i+B$
+are disjoint. For all $a\in A$ there is $i$ such that $(a+B) \cap (a_i+B) \ne \emptyset$, otherwise we
+could add $a$ to $L$ and contradict maximality. Equivalently, we can find $b$ and $b'$ such that
+$a+b = a_i + b'$ so $a\in a_i + B - B$. Hence $A\subseteq L+B-B$. Lastly, $|L|\cdot |B| = |K+B| \le |A+B|$.\slug
+
+We now show that if Theorem~{\thmsubgroup} holds for some constant $C'$, then
+Theorem~{\thmpfr} holds with $C=C'+1$.
+
+\medskip\noindent{\it Proof of Theorem~{\thmpfr} assuming Theorem~{\thmsubgroup}.}\enspace
+Let $A\subseteq \FF_2$ and $K\ge 1$
+such that $|A+A|\le K|A|$. Let $U_A$ be the uniform distribution on $A$, so that $\Eta\{U_A\} = \lg|A|$.
+Letting $U_A'$ and $U_A''$ be two independent copies of $U_A$, the sum $U_A' - U_A''$ belongs to $|A+A|$,
+so $\Eta\{U_A' + U_A''\} \le \lg|A+A|$, and we have
+$$\dd\{U_A, U_A\} = \Eta\{U_A' + U_A''\} - \Eta\{U_A\} \le \lg|A+A|-\lg|A| \le \lg K,$$
+since $U_A' + U_A'' = U_A' - U_A''$ in $\FF_2^n$. By Theorem~{\thmsubgroup}, we can find
+$H\subseteq G$ such that
+$$2\dd\{ U_A, U_H\} \le C'\lg K,$$
+where $U_H$ is taken to be uniform on $H$ independently of $U_A$. Independence and the definition
+of Ruzsa distance let us conclude
+$$\Eta\{U_A-U_H\} \le {\lg|A| + \lg|H| + C'\lg K\over 2},$$
+from which applying Proposition~{\propmaxprob} allows us to procure some $x_0 \in \FF_2^n$ such that
+$$\pr\{ U_A - U_H = x_0\} \ge {1\over |A|^{1/2} |H|^{1/2} K^{C'/2}}.$$
+Since $U_A$ and $U_H$ are uniformly and independently random on $A$ and $H$ respectively, we rewrite this as
+$${\bigl| A\cap (H+x_0)\bigr| \over |A|\cdot|H|}  \ge {1\over |A|^{1/2} |H|^{1/2} K^{C'/2}},$$
+whence
+$$\bigl| A\cap (H+x_0)\bigr|  \ge {|A|^{1/2} |H|^{1/2} \over K^{C'/2}}.$$
+By the Ruzsa covering lemma applied to the sets $A$ and $A\cap (H+x_0)$, there is a set $L\subseteq A$ of size
+$$\eqalign{
+|L| &\le {\bigl| A + (A\cap (H+x_0))\bigr|\over \bigl| A\cap (H+x_0)\bigr|} \cr
+&\le {|A + A|\over \bigl| A\cap (H+x_0)\bigr|} \cr
+&\le {K|A|\cdot K^{C'/2} \over |A|^{1/2} |H|^{1/2} } \cr
+&= {K^{C'/2+1}|A|^{1/2} \over |H|^{1/2} } \cr
+}$$
+such that
+$$A \subseteq L+\bigl(A\cap (H+x_0)\bigr) - \bigl(A\cap (H+x_0)\bigr) \subseteq L+H.$$
+Proposition~{\propabsdist} tells us that
+$$\bigl| \lg|H| - \lg|A|\bigr| = \bigl| \Eta\{U_H\} - \Eta\{H_A\}\bigr| \le 2\dd\{ U_A, U_H\} \le C'\lg K,$$
+so
+$$\max\biggl( {|H|\over |A|}, {|A|\over |H|}\biggr) \le K^{C'}$$
+and we have
+$$|L| \le K^{C'+1}.$$
+If $|H|\le |A|$ then we are done (and the factor of $2$ is unnecessary in the theorem statement),
+since $A$ is a subset of $|L|$ cosets of $H$.
+
+If not, then we can pick a
+subgroup $H'$ of $H$ with $|A|/2 \le |H'|\le |A|$ by iteratively taking hyperplanes until the condition is met.
+The group $H$ may be covered by $|H|/|H'| \le 2|H|/|A|$ translates of $H'$, so $A$ can be covered by
+$${2|H|\over |A|}|L| \le {2|H|\over |A|}\cdot {K^{C'/2+1}|A|^{1/2} \over |H|^{1/2} } 
+= 2 {C'/2+1} {|H|^{1/2} \over |A|^{1/2}} \le 2K{C'+1}$$
+translates of $H'$.\slug
+
+\advsect A compactness argument
+
+We have reduced the proof of the Freiman--Ruzsa theorem to proving the inform\-ation-theoretic
+Theorem~{\thmsubgroup}. In this section, we shall reduce the proof to a different statement via a compactness
+argument. First we recall some basic topological properties of probability measures.
+
+Fix a finite set $G$ of size $n$. Ordering the elements of $G$ from $1$ to $n$, one can associate
+to each probability distribution $\mu$ on $G$ the vector
+$$\bigl( \mu(\{1\}), \mu(\{2\}), \ldots, \mu(\{n\})\bigr)$$
+in $\RR^n$. (Starting now we will write $\mu_i$ instead of $\mu(\{i\})$.)
+Let $\P(G)$ denote the set of all probability distributions on $G$.
+This is a metric space under the {\it total variation
+distance} $\dTV$, where if $\mu = (\mu_1,\ldots,\mu_n)$
+and $\nu = (\nu_1,\ldots,\nu_n)$ are two probability distributions on $G$, then
+$$\dTV(\mu,\nu) = \sum_{i=1}^n |\mu_i - \nu_i|.$$
+Regarding the space as a subset of $\RR^n$, this norm is equivalent to the $L_1$ norm; hence
+the total variation metric on $\P(G)$ is equivalent to the Euclidean metric on $\RR^n$.
+Since probabilities are in $[0,1]$, the space $\P(G)$ is a subset of the closed set $[0,1]^n$,
+and since $\P(G)$ is the inverse image of the set $\{1\}$ under the continuous function
+$$(x_1,\ldots,x_n) \mapsto x_1 + \cdots + x_n,$$
+we conclude that $\P(G)$ is a compact topological space. In passing, we note here that this is
+the topology under which the continuity axiom asserts that $\Eta\{X\}$ is continuous.
+
+Next we show that the Ruzsa distance is also continuous with respect to this topology.
+
+\edef\propruzsacontinuous{\the\sectcount.\the\thmcount}
+\proclaim Proposition \advthm. Let $G$ be a finite abelian group.
+The Ruzsa distance $\dd$, regarded as a functional
+from $\P(G)\times \P(G)\to \RR$, is continuous.
+
+Let $X$ and $Y$ be random variables; without loss of generality, we may assume they are independent,
+so that
+$$\dd\{X,Y\} = \Eta\{X-Y\} - {\Eta\{X\} \over 2} - {\Eta\{Y\}\over 2}.$$
+The entropy functional is continuous by axiom, so from here it suffices to show that
+the function $f : \P(G)\times \P(G) \to \P(G)$ mapping $(X,Y)\mapsto X-Y$ is continuous.
+
+Since all norms on finite-dimensional spaces are equivalent, we may work with
+$\P(G)\times \P(G)$ and $\P(G)$ as normed spaces using the $L_1$ norm on $\RR^{2n}$ and $\RR^n$
+respectively.
+Let $\mu_g = \pr\{X = g\}$, $\nu_g = \pr\{ Y = g\}$, and $\xi_g = \pr\{X-Y = g\}$ for all $g\in G$.
+Given another pair $(X', Y')$ of random variables, define $\mu_g'$, $\nu_g'$, and $\xi_g'$ accordingly.
+Now let $\eps > 0$ and suppose that the vectors $(\mu, \nu)$ and $(\mu', \nu')$
+have $\bignorm{(\mu,\nu), (\mu',\nu')}_1 < \delta$ for some $\delta>0$ to be specified later; writing
+this out in full, we have
+$$\sum_{g\in G} \big(|\mu_g - \mu_g'| + |\nu_g - \nu_g|\bigr) < \delta.$$
+Observe that
+$$\xi_g = \pr\{X-Y= g\} = \sum_{h\in G} \pr\{X = g+h\} \pr\{Y = h\} = \sum_{h\in G} \mu_{g+h}\nu_h$$
+and similarly for $\xi_g'$. From this we can expand and bound
+$$\eqalign{
+\norm{\xi, \xi'}_1 &= \sum_{g\in G} |\xi_g - \xi_g'| \cr
+&= \sum_{g\in G} \Bigl| \sum_{h\in G} \mu_{g+h}\nu_h - \sum_{h\in G} \mu_{g+h}\nu_h\Bigr| \cr
+&\le \sum_{g\in G} \sum_{h\in G} \bigl( |\mu_{g+h}\nu_h - \mu_{g+h}'\nu_h| +
+|\mu_{g+h}'\nu_h - \mu_{g+h}'\nu_h'|\bigl) \cr
+&= \sum_{h\in G} \nu_h \sum_{g\in G} |\mu_{g+h} - \mu_{g+h}'|
++ \sum_{g\in G} \mu_g \sum_{h\in G} |\nu_h - \nu_h'| \cr
+&= \sum_{g\in G} |\mu_{g} - \mu_{g}'| + \sum_{h\in G} |\nu_h - \nu_h'| \cr
+&< \delta,
+}$$
+so in fact we can set $\delta = \eps$.\slug
+
+So much for topology. Returning to entropies, we now prove an entropic version of
+Proposition~{\propdoublingone}.
+
+\edef\propentropydoublingone{\the\sectcount.\the\thmcount}
+\proclaim Proposition \advthm. Let $X$ be a random variable taking values in an abelian group $G$.
+Then $\dd\{X, -X\} = 0$ if and only if $X$ is uniformly distributed on a coset of a finite subgroup of $G$.
+
+\proof If $X$ is the uniform distribution on a coset $a+H$ of a finite subgroup $H$, then letting $X'$ be an
+independent copy of $X$, the random variable $X+X'$ is uniform on the coset $a+a+H$, which has the
+same size as $|a+H|$, so $\Eta\{X+X'\} = \Eta\{X\}$, meaning that
+$$\dd\{X, -X\} = \Eta\{X+X'\} - {\Eta\{X\}\over 2} - {\Eta\{X'\} \over 2} = 0.$$
+On the other hand, suppose that $\Eta\{X+X'\} = \Eta\{X\}$, where $X'$ is an independent copy of $X$.
+Then
+$$\eqalign{
+\Eta\{X+X'\} &= 2\Eta\{X\} - \Eta\{X\} \cr
+&= \Eta\{X, X'\} - \Eta\{X'\} \cr
+&= \Eta\{X+X', X'\} - \Eta\{X'\} \cr
+&= \Eta\{X+X'\given X'\},\cr
+}$$
+so $X+X'$ is independent of $X'$.
+
+Let $A$ denote the support of $X$. We have shown that the distribution
+$(X + X' \given X' = x)$ is the same
+regardless of our choice of $x\in A$. Then for all $x,y\in A$, the distribution of $X+x$ is the
+same as that of $X+y$, so the distribution of
+of $X+x-y$ is the same as that of $X$. This implies that $A$ is finite,
+$X$ is the uniform distribution on $A$,
+and adding an element $h\in A-A$ to the set $A$ simply permutes the set.
+
+As in the proof
+of Proposition~{\propdoublingone}, let $H = \{h\in G : A+h = A\}$. We proved back then that $H$ is a subgroup
+of $G$, and the conclusion of the previous paragraph
+asserts that $A-A \subseteq H$. But letting $h\in H$, the fact that
+$h+A = A$ implies that $h\in A-A$. We conclude that
+$A-A$ is a finite subgroup of $G$, whence $A$ is a coset of a finite subgroup of $G$.\slug
+
+From this, we are able to prove the special case of Theorem~{\thmsubgroup} in which the Ruzsa distance between
+the two given random variables is zero.
+
+\edef\lemzerosubgroup{\the\sectcount.\the\thmcount}
+\proclaim Lemma \advthm. Let $X_1$ and $X_2$ be random variables taking values in $\FF_2^n$ with
+$\dd\{X_1,X_2\} = 0$. Then there exists a subgroup $H$ of $\FF_2^n$ such that
+$$\dd\{X_1, U_H\} = \dd\{X_2,U_H\} = 0.$$
+
+\proof The triangle inequality gives $\dd\{X_1,X_1\} = 0$, and since $-x = x$ in $\FF_2^n$,
+we also have $\dd\{X_1, - X_1\} = 0$. By the previous proposition, there is a subgroup $H$ such that
+$X_1$ has the same distribution as $H$. So $\dd\{X_1, U_H\} = 0$ and by the triangle inequality once again,
+$\dd\{X_2, U_H\} = 0$ as well.\slug
+
+Recall that in the previous section, we reduced the proof of the polynomial Freiman--Ruzsa theorem
+to a statement involving some random variables $X_1^*$ and $X_2^*$. For the remainder of these notes,
+we shall consider these variables to be fixed, and we also fix $\eta = 1/9$ for short.
+Defining the functional
+$$\tau(X_1, X_2) = \dd\{X_1, X_2\} - \eta\dd\{X_1^*, X_1\} - \eta\dd\{X_2^*, X_2\},$$
+we can then reduce the proof of Theorem~{\thmsubgroup} (and consequently the entire proof) to showing the
+following proposition.
+
+\edef\propmaintau{\the\sectcount.\the\thmcount}
+\proclaim Proposition \advthm. Let $X_1$ and $X_2$ be two $\FF_2^n$-valued random variables with
+$\dd\{ X_1, X_2\} > 0$. Then there are $\FF_2^n$-valued random variables $X_1'$ and $X_2'$ such that
+$$\tau\{X_1', X_2'\} < \tau\{X_1, X_2\}.$$
+
+If we can prove this proposition, then we have Theorem~{\thmsubgroup} with a certain constant $C'$ which
+we shall make explicit during the proof.
+
+\medskip\noindent{\it Proof of Theorem~{\thmsubgroup} assuming Proposition~{\propmaintau}}.\enspace
+We proved earlier that $\P(\FF_2^n)$ is a compact topological space, thus so its Cartesian product with
+itself. By Proposition~{\propruzsacontinuous}, the Ruzsa distance functional is continuous on this space,
+which implies that $\tau$ is as well. Hence $\tau$ attains its infimum on the space $\P(\FF_2^n)^2$;
+let $X_1$ and $X_2$ be two distributions such that $\tau\{X_1, X_2\}$ is minimal. By the contrapositive
+of Proposition~{\propmaintau}, we must have $\dd\{X_1, X_2\} = 0$, so by Lemma~{\lemzerosubgroup},
+there exists a subgroup $H$ of $\FF_2^n$ such that
+$$\dd\{X_1, U_H\} = \dd\{X_2, U_H\} = 0.$$
+This means that
+$$\dd\{X_1^*, U_H\} \le \dd\{X_1^*, X_1\} + \dd\{ X_1, U_H\} = \dd\{X_1^*, X_1\}$$
+and
+$$\dd\{X_1^*, X_1\} \le \dd\{X_1^*, U_H\} + \dd\{U_H, X_1\} = \dd\{X_1^*, U_H\},$$
+so $\dd\{X_1^*, U_H\} = \dd\{X_1^*, X_1\}$. The same holds for $X_2^*$ and $X_2$.
+Then since $\dd\{X_1, X_2\} = 0$ and by minimality of $\tau\{X_1, X_2\}$, we obtain
+$$\eqalign{
+\eta\dd\{X_1^*, U_H\} + \eta\dd\{X_2^*, U_H\} &= \eta\dd\{X_1, U_H\} + \eta\dd\{X_2, U_H\} \cr
+&= \tau\{X_1, X_2\} \cr
+&\le \tau\{X_1^*, X_2^*\} \cr
+&= \dd\{X_2^*, X_1^*\} + \eta\dd\{X_1^*, X_2^*\} + \eta\dd\{X_2^*, X_1^*\} \cr
+&= (1+2\eta) \dd\{X_1^*, X_2^*\}. \cr
+}$$
+From our choice of $\eta = 1/9$ we can multiply both sides by $9$ to get
+$$\dd\{X_1^*, U_H\} + \dd\{X_2^*, U_H\} \le 11\dd\{X_1^*, X_2^*\},$$
+and furthermore, since
+$$\bigl| \dd\{X_1, U_H\} - \dd\{X_2, U_H\} \bigr| \le \dd\{X_1, X_2\}$$
+by two invocations of the triangle inequality, neither $\dd\{X_1^*, U_H\}$ nor
+$\dd\{X_2^*, U_H\}$ can be greater than $6\dd\{X_1^*, X_2^*\}$.\slug
+
+For convenience, we restate what we have just proved with the constants plugged in.
+
+\proclaim Theorem~{\thmsubgroup}$\mathbold'$.
+Let $X_1^*$ and $X_2^*$ be random variables taking values in
+$\FF_2^n$. There is some subgroup $H\subseteq \FF_2^n$ such that
+$$\dd\{ X_1^*, U_H\} + \dd\{ X_2^*, U_H\} \le 11\dd\{X_1^*, X_2^*\},$$
+and furthermore,
+$$\max\bigl( \dd\{X_1^*, U_H\}, \dd\{X_2^*, U_H\}\bigr) \le 6\dd\{X_1^*, X_2^*\}.\noskipslug$$
 

### 2024-03-06 misc/expo/entropy.tex
Done first estimate in PFR

@@
 
-\maketitle{Entropy and additive combinatorics}{}{Marcel K. Goh}{15 February 2024}
+\maketitle{Entropy and additive combinatorics}{}{Marcel K. Goh}{\sl Department of Mathematics and Statistics,
+McGill University}
 
@@
 \ninepoint These expository notes give a gentle introduction to the notion of entropy as it is used
-in additive combinatorics, with a view towards understanding the proof of the polynomial Freiman--Ruzsa
-conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. Much of the first section has been
-transcribed from lectures given by W.~T.~Gowers.
+in additive combinatorics, moving at a leisurely pace through the entropic analogues of Pl\"unnecke's
+theorem and the Balog--Szemer\'edi--Gowers theorem before tackling the recent
+proof of the polynomial Freiman--Ruzsa
+conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. Effort has been put into making this
+document as self-contained as possible, and extra proof details have been supplied in the hope
+that these notes may be accessible to the average graduate student or enterprising undergraduate.
 
@@
 determined by
-$n$ bits of information. It satisfies the following axioms, which are
+$n$ bits of information.
+It satisfies the following axioms, which are
 sometimes called the Khinchine--Shannon axioms.
@@
 these axioms.
-In fact, the axioms only define entropy
-up to a multiplicative constant, so we shall add the following axiom.
 (It is very possible you have met the formula for entropy
@@
 these notes.)
+In fact, the axioms only define entropy
+up to a multiplicative constant, so we shall add the following axiom.
 
@@
 Notationally, we would expect that $\Eta\{Y\given X\} = \Eta\{Y\}$ if $X$ and $Y$ are independent.
-This is the first proposition we will carefully prove using the axioms above.
+This is the first proposition we will carefully prove, using only the axioms. The axiomatic description above,
+as well as many of the proofs in
+this section, are transcribed (with some adaptations here and there) from lectures given by W.~T.~Gowers.
 
@@
 
-Next we establish the intuitive fact
+Next we establish the intuitive statement
 that the entropy of a uniform random variable supported on a set $A$ is at most
@@
 
-The link between entropy and additive combinatorics is rather a fine one. It is based on the following
+The thread connecting entropy and additive combinatorics is rather a precarious one.
+As noted by I.~Ruzsa~\ref{ruzsa2008}, there are scenarios in which combinatorial inequalities and
+their corresponding entropy ones
+are equivalent, scenarios in which they both hold but their equivalence cannot be established,
+and further scenarios where an inequality holds in one world but cannot (or has not) been proven in the other.
+
+The ``dictionary'' that allows us to translate
+statements between cardinality inequalities and entropy inequalities is based, in part, on the following
 observation. (In this and the rest of the notes, $\lg$ denotes the base-$2$ logarithm. This base can
@@
 the {\it conditional mutual information}, which is defined by
-$$\II\{ X : Y \given Z\} = \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\}.$$
+$$\eqalign{
+\II\{ X : Y \given Z\} &= \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
+&= \Eta\{X\given Z\} - \Eta\{X\given Y, Z\} \cr
+&= \Eta\{Y\given Z\} - \Eta\{Y\given X, Z\}.\cr
+}$$
 
@@
 $$\Eta\{ X\pm Y\} \ge \Eta\{X\given Y\} = \Eta\{X\} - \II\{X : Y\}.$$
-Repeating the same argument but exchanging the roles of $X$ and $Y$, we get
+Repeating the same argument but exchanging the r\^oles of $X$ and $Y$, we get
 $$\Eta\{ X\pm Y\} \ge \Eta\{Y\given X\} = \Eta\{Y\} - \II\{X : Y\},$$
@@
 between them. (This satisfies all the axioms of a metric except the one requiring $d(A,A) = 0$ for all sets
-$A$.)
+$A$.) This ``distance'' is called the {\it Ruzsa distance} as it was first defined by I.~Ruzsa~\ref{ruzsa1996}.
 
@@
 $$\dd\{X,Y\} = \Eta\{X'-Y'\} - {\Eta\{X'\} \over 2} - {\Eta\{Y'\}\over 2}.$$
-This definition only depends on the individual distributions of $X$ and $Y$ and does not require them to
-have the same sample space. When $X$ and $Y$ are independent, the Ruzsa distance enjoys some
+This definition, first established by T.~Tao~\ref{tao2010},
+only depends on the individual distributions of $X$ and $Y$ and does not require them to
+have the same sample space. When $X$ and $Y$ are independent, the entropic Ruzsa distance enjoys some
 nice properties.
@@
 \proof By independence we have
-$$2\dd\{X,Y\} = 2\Eta\{X-Y\} - \Eta\{X\} - \Eta\{Y\}\over 2,$$
+$$2\dd\{X,Y\} = 2\Eta\{X-Y\} - \Eta\{X\} - \Eta\{Y\}.$$
 Corollary~{\cormaxsumineq} tells us that $2\Eta\{X-Y\} \ge 2\Eta\{X\}$, so this is at least $\Eta\{X\}-\Eta\{Y\}$.
@@
 
+The following inequality relates the conditional and unconditional versions of Ruzsa distance.
+
+\edef\lemfiveone{\the\sectcount.\the\thmcount}
+\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma 5.1). Let $(X,Z)$, and $(Y,W)$ be random
+variables, with $X$ and $Y$ taking values in the same abelian group. Then
+$$\dd\{ X\given Z ; Y\given W\} \le \dd\{X,Y\} + {1\over 2} \II\{X:Z\} + {1\over 2}\II\{Y:W\}.$$
+
+\proof Letting $(X',Z')$ and $(Y',W')$ be independent copies of the given random variables, we
+expand
+$$\eqalign{
+\dd\{ X\given Z ; Y\given W\}
+&= \Eta\{X'-Y'\given Z',W'\} - {\Eta\{X'\given Z'\}\over 2} - {\Eta\{Y'\given W'\}\over 2} \cr
+&\le \Eta\{X'-Y'\} - {\Eta\{X',Z'\}-\Eta\{Z'\}\over 2} \cr
+&\hskip130pt- {\Eta\{Y', W'\}-\Eta\{W'\}\over 2} \cr
+&= \dd\{X-Y\} + {\Eta\{X'\} + \Eta\{Z'\} -\Eta\{X',Z'\}\over 2} \cr
+&\hskip90pt+ {\Eta\{Y'\} + \Eta\{W'\} - \Eta\{Y', W'\}\over 2} \cr
+&= \dd\{X-Y\} + {1\over 2}\II\{X:Z\} + {1\over 2}\II\{Y:W\},\cr
+}$$
+by independence and the fact that conditioning does not increase entropy.\slug
+
 \medskip\boldlabel The sum-difference inequality.
@@
 \proof Let $(X_1,Y_1)$ and $(X_2,Y_2)$ be conditionally independent trials of $(X,Y)$ relative
-to $X-Y$. Since $(X,Y)$ determines $X-Y$, we have $X_1-Y_1 = =X-Y= X_2-Y_2$. Let $(X_3,Y_3)$ be another
+to $X-Y$. Since $(X,Y)$ determines $X-Y$, we have $X_1-Y_1 =X-Y= X_2-Y_2$. Let $(X_3,Y_3)$ be another
 trial of $(X,Y)$ independent of $(X_1, X_2, Y_1, Y_2)$. Then
@@
 
-\proclaim Lemma \advthm. Let $X$, $Y$, and $Z$ be independent random variables taking values in a common
+\edef\lemaone{\the\sectcount.\the\thmcount}
+\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma A.1).
+Let $X$, $Y$, and $Z$ be independent random variables taking values in a common
 abelian group. Then
@@
 }$$
+
+We conclude this section by recording two consequences of Lemma~{\lemaone} and will
+be of use in the proof of the polynomial Freiman--Ruzsa theorem.
+
+\edef\lemfivetwo{\the\sectcount.\the\thmcount}
+\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma 5.2).
+Let $X$, $Y$, and $Z$ be random variables taking values in an abelian group,
+with $Y$ and $Z$ independent. Then
+$$\eqalign{
+\dd\{X, Y-Z\} - \dd\{X, Y\} &\le {\Eta\{Y-Z\} - \Eta\{Y\}\over 2} \cr
+&= {1\over 2} \dd\{Y, Z\} + {1\over 4}\Eta\{Z\} - {1\over 4}\Eta\{Y\}\cr
+}$$
+and
+$$\eqalign{
+\dd\{X; Y\given Y-Z\} - \dd\{X, Y\} &\le {\Eta\{Y-Z\} - \Eta\{Z\}\over 2} \cr
+&= {1\over 2} \dd\{Y, Z\} + {1\over 4}\Eta\{Z\} - {1\over 4}\Eta\{Z\}.\cr
+}$$
+
+\proof For the first inequality, let $X'$ be a copy of $X$ that is independent of $(Y,Z)$, so that
+$$\eqalign{
+\dd\{X, Y-Z\} - \dd\{X, Y\}
+&= \Eta\{X'-Y+Z\} - {1\over 2}\Eta\{X'\} - {1\over 2} \Eta\{Y-Z\} \cr
+&\hskip60pt-\Eta\{X'-Y\} + {1\over 2}\Eta\{X'\} + {1\over 2}\Eta\{Y\}. \cr
+&= \Eta\{X'-Y+Z\} - {1\over 2} \Eta\{Y-Z\} \cr
+&\hskip110pt-\Eta\{X'-Y\} + {1\over 2}\Eta\{Y\}.\cr
+}$$
+Lemma~\ref{\lemaone} with $X'$ in place of $X$ and $-Y$ in place of $Y$ yields
+$$\Eta\{X'-Y+Z\} - \Eta\{X'-Y\} \le \Eta\{Y-Z\} -\Eta\{Y\},$$
+so
+$$\dd\{X, Y-Z\} - \dd\{X, Y\} \le {1\over 2} \Eta\{Y-Z\} - {1\over 2}\Eta\{Y\},$$
+as desired. The alternate version of the statement follows directly from the definition of Ruzsa distance,
+since $Y$ and $Z$ are independent.
+
+For the other inequality, we apply Lemma~\ref{\lemfiveone} with $W = Y-Z$ (and the variable $Z$ in that lemma
+set to anything that is independent of $X$) to obtain
+$$\eqalign{
+\dd\{X; Y\given Y-Z\} - \dd\{X,Y\} &\le {1\over 2} \II\{Y:Y-Z\} \cr
+&= {1\over 2}\bigl(\Eta\{Y\} + \Eta\{Y-Z\} - \Eta\{Y, Y-Z\}\bigr) \cr
+&= {1\over 2}\bigl(\Eta\{Y\} + \Eta\{Y-Z\} - \Eta\{Y, Z\}\bigr) \cr
+&= {1\over 2}\bigl(\Eta\{Y-Z\} - \Eta\{Z\}\bigr),\cr
+}$$
+where in the last line we used independence of $Y$ and $Z$. Independence also gives the alternative version
+of the right-hand-side expression.\slug
+
+Changing variables in the first inequality and then adding the second inequality gives us the following.
+
+\edef\lemsevenone{\the\sectcount.\the\thmcount}
+\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma 7.1).
+Let $X$, $Y$, $Z$, and $W$ be random variables taking values in the same abelian group, with $Y$, $Z$,
+and $W$ independent. Then
+$$\eqalign{
+\dd\{&X; Y-Z\given Y-Z-W\} - \dd\{X,Y\} \cr
+&\le {1\over 2} \bigl( \Eta\{Y-Z-W\} + \Eta\{Y-Z\} - \Eta\{Y\} - \Eta\{W\}\bigr) .\cr
+}$$
+
+\proof The second inequality of Lemma~\ref{\lemfivetwo} (setting $Y \gets Y-Z$ and $Z\gets W$) yields
+$$\dd\{X; Y-Z\given Y-Z-W\} - \dd\{X, Y-Z\} \le {\Eta\{Y-Z-W\} - \Eta\{W\}\over 2},$$
+and the first inequality (without any change of variables) is
+$$\dd\{X, Y-Z\} - \dd\{X, Y\} \le {\Eta\{Y-Z\} - \Eta\{Y\}\over 2}.$$
+Adding these two inequalites proves the lemma.\slug
 
@@
 
-The fact that the implication goes the ``wrong'' way may seem somewhat baffling at first. We will now get
+The fact that the implication goes the ``wrong'' way may seem somewhat baffling. Let us get
 to the bottom of this.
@@
 In the entropy setting, the operation on random variables that corresponds to taking subsets is
-to conditioning. (As a sanity check, recall that conditioning never increases entropy, just
+conditioning. (As a sanity check, recall that conditioning never increases entropy, just
 as taking subsets never increases cardinality.) The Balog--Szemer\'edi--Gowers theorem gives
@@
 \edef\lematwo{\the\sectcount.\the\thmcount}
-\proclaim Lemma \advthm. Let $X$ and $Y$ be discrete random variables taking values in the same abelian
+\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma A.2).
+Let $X$ and $Y$ be discrete random variables taking values in the same abelian
 group. Let $(X_1, Y_1)$ and $(X_2,Y_2)$ be conditionally
@@
 }$$
-The same holds with the roles of $X_2$ and $Y_2$ exchanged.\slug
+The same holds with $Y_2$ replaced by $X_2$.\slug
 
@@
 cook up examples where there is enough linear independence in $A$ so that the smallest subspace
-containing $A$ has size exponential in $|A|$.
-
-Returning to first
-principles, observe that if the doubling
+containing $A$ has size exponential in $|A|$, but $A$ still has small doubling.
+
+On the other hand, in the extreme case that the doubling
 constant of $A$ is $1$, then the structure of $A$ is completely determined, as shown by the following proposition.
@@
 
+\edef\sectcompactness{\the\sectcount}
 \advsect A compactness argument
@@
 \medskip\noindent{\it Proof of Theorem~{\thmsubgroup} assuming Proposition~{\propmaintau}}.\enspace
-We proved earlier that $\P(\FF_2^n)$ is a compact topological space, thus so its Cartesian product with
+We proved earlier that $\P(\FF_2^n)$ is a compact topological space, thus so is its Cartesian product with
 itself. By Proposition~{\propruzsacontinuous}, the Ruzsa distance functional is continuous on this space,
@@
 
-For convenience, we restate what we have just proved with the constants plugged in.
+For posterity, let us restate what we have just proved with the constants plugged in.
 
@@
 
+\advsect Entropy distance under homomorphisms
+
+Our goal has been reduced to showing that if $X_1$ and $X_2$ have $\dd\{X_1, X_2\} > 0$, then
+there exist $X_1'$ and $X_2'$ with $\tau\{X_1', X_2'\} < \tau\{X_1, X_2\}$. Remember that
+the fixed variables $X_1^*$ and $X_2^*$ still play a r\^ole in this statement, since
+they feature in the definition of the functional $\tau$.
+
+We have elected to first prove all the technical lemmas
+then stitch them into a complete proof.
+We employ this ``bottom-up'' approach not because
+it is necessarily more natural, but because the original proof is written
+in a rather ``top-down'' style, and the reader may find it enlightening to have both expositions at his
+or her disposal.
+
+Having said this, the remainder of this section will be devoted to the study of how entropy
+behaves under homomorphism. This appeared as Proposition~1.4
+in~\ref{revisited2023}, and reproved in~\ref{pfr2023} with the
+explicit error term shown below.
+
+\edef\propfourone{\the\sectcount.\the\thmcount}
+\parenproclaim Proposition {\advthm} ({\rm\ref{pfr2023},} Proposition 4.1).
+Let $\pi : G\to G'$ be a homomorphism of abelian groups and let
+$Z_1$ and $Z_2$ be $G$-valued random variables. Then
+$$\dd\{Z_1, Z_2\} \ge \dd\bigl\{ \pi(Z_1), \pi(Z_2)\bigr\}
++ \dd\bigl\{ Z_1\given \pi(Z_1); Z_2\given \pi(Z_2)\bigr\}.$$
+If $Z_1$ and $Z_2$ are assumed to be independent, then the two sides differ by
+$$\II\bigl\{ Z_1 - Z_2 : \bigl( \pi(Z_1), \pi(Z_2)\bigr) \mathbin{\big|} \pi(Z_1-Z_2)\bigr\}.$$
+
+\proof Let $Z_1'$ and $Z_2'$ be independent copies of $Z_1$ and $Z_2$. Then
+$$\eqalign{
+\dd\bigl\{ Z_1\given &\pi(Z_1); Z_2\given \pi(Z_2)\bigr\} \cr
+&= \Eta\bigl\{Z_1'-Z_2'\given \pi(Z_1'), \pi(Z_2') \bigr\}
+ - {1\over 2}\Eta\bigl\{ Z_1' \given \pi(Z_1')\bigr\} - {1\over 2}\Eta\bigl\{ Z_2' \given \pi(Z_2')\bigr\} \cr
+&\le \Eta\bigl\{Z_1'-Z_2'\given \pi(Z_2') \bigr\}
+ - {1\over 2}\Eta\bigl\{ Z_1' \given \pi(Z_1')\bigr\} - {1\over 2}\Eta\bigl\{ Z_2' \given \pi(Z_2')\bigr\} \cr
+&= \Eta\bigl\{Z_1'-Z_2'\given \pi(Z_1'- Z_2') \bigr\}
+ - {1\over 2}\Eta\bigl\{ Z_1' \given \pi(Z_1')\bigr\} - {1\over 2}\Eta\bigl\{ Z_2' \given \pi(Z_2')\bigr\} \cr
+&= \Eta\{Z_1'-Z_2'\} - \Eta\bigl\{ \pi(Z_1'-Z_2')\bigr\} \cr
+&\qquad\qquad
++{1\over 2}\Eta\{Z_1'\} - {1\over 2}\Eta\bigl\{ \pi(Z_1')\bigr\} 
++{1\over 2}\Eta\{Z_2'\} - {1\over 2}\Eta\bigl\{ \pi(Z_2')\bigr\} \cr
+&= \dd\{Z_1, Z_2\} - \dd\bigl\{ \pi(Z_1),\pi(Z_2)\bigr\},\cr
+}$$
+where the inequality follows from submodularity, and in the second-last equality we used (three times)
+the identity
+$$\Eta\bigl\{X\given \pi(X)\bigr\} = \Eta\{X\} - \Eta\bigl\{\pi(X)\bigr\},$$
+which holds for any $G$-valued random variable $X$, since $X$ determines $\pi(X)$.
+
+If $Z_1$ and $Z_2$ are independent, then the difference between the two sides is
+$$\eqalign{
+\Eta\bigl\{&Z_1-Z_2\given \pi(Z_1-Z_2) \bigr\}-\Eta\bigl\{Z_1-Z_2\given \pi(Z_1), \pi(Z_2) \bigr\} \cr
+&= \Eta\bigl\{Z_1-Z_2\given \pi(Z_1-Z_2) \bigr\}-\Eta\bigl\{Z_1-Z_2\given \pi(Z_1), \pi(Z_2), \pi(Z_1-Z_2)
+\bigr\}.\cr
+}$$
+But one of the definitions of conditional mutual information is
+$$\II\{X:Y\given Z\} = \Eta\{X\given Z\} - \Eta\{X\given Y, Z\},$$
+so letting $X = Z_1-Z_2$, $Y=\bigl(\pi(Z_1), \pi(Z_2)\bigr)$, and $Z=\pi(Z_1-Z_2)$, we
+see that the two sides of the inequality differ by exactly the conditional mutual information term
+claimed above.\slug
+
+We will use this proposition in the following special case.
+
+\edef\corfourtwo{\the\sectcount.\the\thmcount}
+\parenproclaim Corollary {\advthm} ({\rm\ref{pfr2023},} Corollary 4.2).
+Let $G$ be an abelian group and let $Y_1$, $Y_2$, $Y_3$, and $Y_4$ be independent
+$G$-valued random variables. Then
+$$\eqalign{
+\dd\{Y_1, Y_2\} + \dd\{Y_3,Y_4\} = \dd\{Y_1 - Y_3;&\ Y_2 - Y_4\}
++ \dd\{ Y_1\given Y_1 - Y_3;\ Y_2\given Y_2-Y_4\}\cr
+&+ \II\{Y_1 - Y_2 : Y_2-Y_4 \given Y_1 - Y_2 - Y_3 + Y_4\}. \cr
+}$$
+
+\proof We apply this with the subtraction homomorphism $\pi : G\times G\to G$ given by $\pi(x,y) = x-y$.
+Then the previous proposition applied to $Z_1 = (Y_1, Y_3)$ and $Z_2 = (Y_2, Y_4)$ yields the equality
+$$\eqalign{
+\dd\bigl\{&(Y_1, Y_3);\ (Y_2, Y_4)\bigr\} \cr
+&= \dd\bigl\{ Y_1-Y_3, Y_2-Y_4\bigr\}
++ \dd\bigl\{ (Y_1, Y_3)\given Y_1-Y_3;\ (Y_2,Y_4)\given Y_2-Y_4\bigr\}\cr
+&\hskip40pt+\II\bigl\{ (Y_1-Y_2, Y_3-Y_4) : (Y_1-Y_3, Y_2-Y_4)\mathbin{\big|} Y_1 - Y_2 - Y_3 - Y_4\bigr\} \cr
+&= \dd\bigl\{ Y_1-Y_3, Y_2-Y_4\bigr\} + \dd\bigl\{ Y_1\given Y_1-Y_3;\ Y_2\given Y_2-Y_4\bigr\}.\cr
+&\hskip40pt+\II\bigl\{ Y_1-Y_2 : Y_2-Y_4 \mathbin{\big|} Y_1-Y_2-Y_3+Y_4\bigr\} \cr
+}$$
+where in the last line we used the fact that $(Y_1-Y_2, Y_1-Y_2-Y_3+Y_4)$ determines $Y_3-Y_4$ as well
+as the fact that $(Y_2-Y_4, Y_1-Y_2-Y_3+Y_4)$ determines $Y_1-Y_3$.
+But by independence,
+$$\eqalign{
+\dd\bigl\{(Y_1, Y_3);\ (Y_2, Y_4)\bigr\}
+&= \Eta\{ Y_1-Y_2, Y_3-Y_4\} - {\Eta\{Y_1, Y_3\}\over 2} - {\Eta\{Y_2, Y_4\}\over 2} \cr
+&= \dd\{Y_1, Y_2\} + \dd\{Y_3, Y_4\},\cr
+}$$
+so the corollary is proved.\slug
+
+\advsect Sums and fibres
+
+In Section~{\sectcompactness}, we showed that to prove the polynomial Freiman--Ruzsa theorem,
+it suffices to produce, for every pair $(X_1, X_2)$ of random variables taking values in $G = \FF_2^n$
+with $\dd\{X_1, X_2\}>0$,
+a pair $(X_1',X_2')$ with $\tau\{X_1', X_2'\} < \tau\{X_1, X_2\}$. (This was Proposition~{\propmaintau}.)
+Recall that the definition of $\tau$ involves the ``global'' variables $X_1^*$ and $X_2^*$, as well as the global
+constant $\eta = 1/9$.
+
+To prove this statement, it is somewhat more convenient to work with its contrapositive. That is,
+we shall assume that $(X_1, X_2)$ has $\dd\{X_1, X_2\} = k$ and that
+this pair minimises $\tau$; that is, $\tau\{X_1', X_2'\} \ge \tau\{X_1,X_2\}$
+for all $(X_1',X_2')$.
+Without loss of generality we may further assume that $X_1$ and $X_2$ are independent.
+Our aim is to show that $k=0$.
+The first lemma we prove amounts to little more than expansion of definitions, but is worth writing
+explicitly nonetheless.
+
+\edef\lemionehelper{\the\sectcount.\the\thmcount}
+\proclaim Lemma \advthm. Suppose $(X_1, X_2)\in \P(G)^2$
+is a minimiser of $\tau$ with $\dd\{X_1, X_2\} = k$.
+Then for all $(X_1',X_2')\in \P(G)^2$,
+$$\dd\{X_1', X_2'\} \ge k-\eta\bigl(\dd\{X_1^*, X_1'\} - \dd\{X_1^*, X_1\}
++ \dd\{X_2^*, X_2'\} - \dd\{X_2^*, X_2\}\bigr),$$
+and for all $(X_1',X_2')\in \P(G)^2$ and any discrete random variables $Y_1$ and $Y_2$, we have
+$$\eqalign{
+\dd\{&X_1'\given Y_1; X_2'\given Y_2\} \cr
+&\ge k-\eta\bigl(\dd\{X_1^*; X_1'\given Y_1\} - \dd\{X_1^*, X_1\}
++ \dd\{X_2^*; X_2'\given Y_2\} - \dd\{X_2^*, X_2\} \bigr). \cr
+}$$
+
+\proof The first inequality follows directly from $\tau\{X_1',X_2'\} \ge \tau\{X_1, X_2\}$ and
+the definition of $\tau$. For the second inequality, suppose that $Y_1$ has support $A_1$ and
+$Y_2$ has support $A_2$ and sum the inequalities
+$\tau\{X_1\given Y_1 = y_1;\ X_2\given Y_2 = y_2\}\ge \tau\{X_1,X_2\}$,
+weighted by $\pr\{Y_1 = y_1, Y_2 = y_2\}$ for all $(y_1, y_2)\in A_1\times A_2$.\slug
+
+The task now is to investigate specific
+choices of $(X_1',X_2')$ and see what information we can glean about a minimising pair $(X_1, X_2)$.
+The next lemma proves our first inequality in this direction.
+
+\edef\lemionebound{\the\sectcount.\the\thmcount}
+\proclaim Lemma \advthm. Let $G = \FF_2^n$
+and suppose $(X_1, X_2)\in \P(G)^2$ is a minimiser of $\tau$ with $\dd\{X_1, X_2\} = k$.
+Then letting
+$$I_1 = \II\{ X_1 + X_2 : \bar X_1 + X_2 \given X_1 + X_2 + \bar X_1 + \bar X_2\},$$
+where
+$X_1$ and $\bar X_1$ be copies of $X_1$ and $X_2$ and $\bar X_2$ be copies of $X_2$ such that $X_1$,
+$\bar X_1$, $X_2$, and $\bar X_2$ are all independent,
+we have
+$$I_1 \le 2\eta k.$$
+Furthermore, we have
+$$\Eta\{X_1 + X_2 + \bar X_1 + \bar X_2\} \le {1\over 2} \Eta\{X_1\} + {1\over 2} \Eta\{X_2\}+(2+\eta)k - I_1.$$
+
+\proof Since
+$$ k = \dd\{X_1, X_2\} = \dd\{X_1, \bar X_2\} = \dd\{X_2, \bar X_1\},$$
+applying Lemma~{\lemfivetwo} four times gives us the inequalities
+$$\eqalign{
+\dd\{X_1^*; X_1 + \bar X_2\}-\dd\{X_1^*, X_1\}&\le{1\over 2}k + {1\over 4}\Eta\{X_2\}-{1\over 4}\Eta\{X_1\},\cr
+\dd\{X_2^*; X_2 + \bar X_1\}-\dd\{X_2^*, X_2\}&\le{1\over 2}k + {1\over 4}\Eta\{X_1\}-{1\over 4}\Eta\{X_2\},\cr
+\dd\{X_1^*; X_1\given X_1-\bar X_2\}-\dd\{X_1^*,X_1\}
+  &\le{1\over 2}k + {1\over 4}\Eta\{X_1\} - {1\over 4}\Eta\{X_2\},\cr
+}$$
+and
+$$\dd\{X_2^*; X_2\given X_2-\bar X_1\}-\dd\{X_2^*,X_2\}
+  \le{1\over 2}k + {1\over 4}\Eta\{X_2\} - {1\over 4}\Eta\{X_1\}.$$
+By summing these four inequalites, we obtain
+\edef\eqfoursum{\the\eqcount}
+$$\eqalign{
+2k &\ge \dd\{X_1^*; X_1 + \bar X_2\} + \dd\{X_2^*; X_2 + \bar X_1\} \cr
+&\qquad\qquad+\dd\{X_1^*; X_1\given X_1 + \bar X_2\} + \dd\{X_2^*; X_2\given X_2 + \bar X_1\} \cr
+&\qquad\qquad-2\dd\{X_1^*, X_1\} - 2\dd\{X_2^*, X_2\} .\cr
+}\adveq$$
+On the other hand,
+Corollary~{\corfourtwo} with $(Y_1, Y_2, Y_3, Y_4)$ set to $(X_1, X_2, \bar X_2, \bar X_1)$
+gives us
+$$\eqalign{
+\dd\{X_1, X_2\} + \dd\{\bar X_2,\bar X_1\} = \dd\{&X_1 - \bar X_2;\ X_2 - \bar X_1\} \cr
+&+ \dd\{ X_1\given X_1 - \bar X_2;\ X_2\given X_2-\bar X_1\}\cr
+&+ \II\{X_1 - X_2 : X_2-\bar X_1 \given X_1 - X_2 - \bar X_2 + \bar X_1\}. \cr
+}$$
+which can be rewritten
+\edef\eqioneidentity{\the\eqcount}
+$$\eqalign{
+2k &=
+\dd\{X_1 + \bar X_2;\ X_2 + \bar X_1\} +\dd\{ X_1\given X_1 + \bar X_2;\ X_2\given X_2+\bar X_1\}\cr
+&\qquad\qquad+ \II\{X_1 + X_2 : X_2+\bar X_1 \given X_1 + X_2 + \bar X_1 + \bar X_2\}\cr
+&=  \dd\{X_1 + \bar X_2;\ X_2 + \bar X_1\} + \dd\{ X_1\given X_1 + \bar X_2;\ X_2\given X_2+\bar X_1\}+I_1\cr
+}\adveq$$
+since $\dd\{X_1,X_2\} = k$ and we are working in $G = \FF_2^n$.
+By Lemma~{\lemionehelper}, we have
+$$\eqalign{
+\dd\{X_1 + \bar X_2; X_2 + \bar X_1\} &\ge k-\eta\bigl(\dd\{X_1^*, X_1 + \bar X_2\}-\dd\{X_1^*, X_1\}\cr
+&\qquad + \dd\{X_2^*, X_2 + \bar X_1\}- \dd\{X_2^*, X_2\}\bigr) \cr
+}\adveq$$
+and
+\edef\eqioneconditional{\the\eqcount}
+$$\eqalign{
+\dd\{X_1\given X_1 + \bar X_2;X_2\given X_2 + \bar X_1\}
+&\ge k-\eta\bigl(\dd\{X_1^*; X_1\given  X_1 + \bar X_2\} - \dd\{X_1^*, X_1\}\cr
+&\qquad+ \dd\{X_2^*, X_2\given X_2 + \bar X_1\}- \dd\{X_2^*, X_2\} \bigr).\cr
+}\adveq$$
+Substituting these two inequalities into~\refeq{\eqioneidentity} yields
+$$\eqalign{
+2k &\ge I_1 + 2k - \eta\bigl( \dd\{X_1^*, X_1 + \bar X_2\} + \dd\{X_1^*, X_1\given X_1+\bar X_2\} \cr
+&\qquad\qquad\qquad\qquad+ \dd\{X_2^*, X_2 + \bar X_1\} + \dd\{X_2^*, X_2\given X_2+\bar X_2\},\cr
+&\qquad\qquad\qquad\qquad -2\dd\{X_1^*, X_1\} - 2\{X_2^*, X_2\} \bigl)\cr
+}$$
+whence
+$$\eqalign{
+I_1&\le \eta\bigl( \dd\{X_1^*, X_1 + \bar X_2\} + \dd\{X_1^*, X_1\given X_1+\bar X_2\} \cr
+&\qquad\qquad\qquad\qquad+ \dd\{X_2^*, X_2 + \bar X_1\} + \dd\{X_2^*, X_2\given X_2+\bar X_2\} ,\cr
+&\qquad\qquad\qquad\qquad -2\dd\{X_1^*, X_1\} - 2\{X_2^*, X_2\} \bigl)\cr
+}$$
+and we can substitute the inequality~\refeq{\eqfoursum} to get $I_1\le 2\eta k$.
+
+To prove the other claim, we substitute~\refeq{\eqioneconditional} into~\refeq{\eqioneidentity}
+to obtain
+$$\eqalign{
+k&\ge I_1 + \dd\{X_1 + \bar X_2 ; X_2 + \bar X_1\}
+  -\eta\bigl(\dd\{X_1^*; X_1\given  X_1 + \bar X_2\} - \dd\{X_1^*, X_1\}\cr
+&\hskip150pt+ \dd\{X_2^*, X_2\given X_2 + \bar X_1\}- \dd\{X_2^*, X_2\} \bigr) \cr
+&\ge I_1 + \dd\{X_1 + \bar X_2; X_2 + \bar X_1\} -\eta k, \cr
+}$$
+where in the second line we have used two of the four inequalities begotten by Lemma~{\lemfivetwo}.
+So $(1+\eta)k - I_1 \ge \dd\{X_1 + \bar X_2; X_2 + \bar X_1\}$, which we expand to
+$$\eqalign{
+(1-\eta)k-I_1 &\ge \Eta\{X_1 + X_2 + \bar X_1 + \bar X_2\} - {1\over 2}\Eta\{X_1 + \bar X_2\} - 
+{1\over 2} \Eta\{X_2+\bar X_1\} \cr
+&=  \Eta\{X_1 + X_2 + \bar X_1 + \bar X_2\}-\dd\{X_1, X_2\} - {1\over 2} \Eta\{X_1\}-{1\over 2}\Eta\{X_2\}\cr
+}$$
+by independence and the fact that subtraction is addition in $G$. Since
+$\dd\{X_1, X_2\}=k$, rearranging completes
+the proof.\slug
+
+We can interpret this lemma as saying that if setting $X_1' = X_1 + \bar X_2$ and $X_2' = X_2 + \bar X_1$
+gives the inequality $\tau\{X_1',X_2'\} \ge \tau\{X_1,X_2\}$, and furthermore if the same holds when
+setting $(X_1',X_2')$ to any pair
+$$\bigl(X_1 \given X_1+\bar X_2 = v_1, X_2\given X_2+\bar X_1 = v_2\bigr)$$
+of ``fibres'', where $(v_1, v_2)\in G^2$,
+then we have a bound on a certain mutual information quantity $I_1$.
+
+Now we perform a similar analysis, where instead of taking sums $X_1+\bar X_2$ and
+$X_2+\bar X_1$ across variables, now we sum copies of the {\it same} variables.
+
 \section References

### 2024-03-18 misc/expo/entropy.tex
Ahoy

@@
 if $Y\in E_a$. This definition makes $Z$ and $X$ identically distributed, so $\Eta\{Z\} = \Eta\{X\}$
-by the invariance axiom, and it suffices to prove $\Eta\{Z\} \ge 0$.
+by the invariance axiom.
 
@@
 $$\Eta\{Y\given Z\} = \sum_{a\in A} \pr\{X = a\} \Eta\{Y\given X = a\}
-\sum_{a\in A} \pr\{X = a\} \log |E_a| \le \log |E_{a^*}|.$$
+= \sum_{a\in A} \pr\{X = a\} \log |E_a| \le \log |E_{a^*}|.$$
 Since $Z$ is determined by $Y$, we have $\Eta\{Y\} = \Eta\{Z\} + \Eta\{Y\given Z\}$ by the previous lemma.
@@
 $$\eqalign{
-\II\{ X : Y \given Z\} &= \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
+\II\{ X : Y \given Z\} &= \sum_{z\in C} \pr\{Z = z\} \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
 &= \Eta\{X\given Z\} - \Eta\{X\given Y, Z\} \cr
@@
 $$\II\{X : Y\given Z\} \ge 0.$$
+Equality holds if and only if $X$ and $Y$ are independent conditional on $Z$.
 
@@
 $$\eqalign{
-\II\{ X : Y \given Z\} &= \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
+\II\{ X : Y \given Z\} &= \sum_{z\in C} \pr\{Z=z\} \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
 &= \Eta\{ X\given Z\} + \Eta\{Y\given Z\} - \Eta\{ X, Y \given Z \} \cr
@@
 meaning we can take sums $X+Y$ and differences $X-Y$ of them. Note that if we condition on $Y$,
-then the values taken $X+Y$ are in bijection with values taken by $X$. This leads to the following
+then the values taken by $X+Y$ are in bijection with values taken by $X$. This leads to the following
 proposition.
@@
 by the invariance axiom we have
-$$\Eta\{X-Y, X-Z\} = \Eta\{X-Z, Y-Z\},$$
-and
-$$\Eta\{X-Y, X-Z\} \le \Eta\{X-Y\} + \Eta\{Y-Z\}$$
-follows by submodularity. Similar invocations of the invariance axiom give
+$$\Eta\{X-Y, X-Z\} = \Eta\{X-Z, Y-Z\} \le \Eta\{X-Z\} + \Eta\{Y-Z\}.$$
+Similar invocations of the invariance axiom give
 $$\Eta\{Z, X-Z\} = \Eta\{X,Z\}$$
@@
 
-\proof Since
+\proof
+Since
 $$ k = \dd\{X_1, X_2\} = \dd\{X_1, \bar X_2\} = \dd\{X_2, \bar X_1\},$$
@@
 By summing these four inequalites, we obtain
-\edef\eqfoursum{\the\eqcount}
+\edef\eqionefoursum{\the\eqcount}
 $$\eqalign{
@@
 }$$
-and we can substitute the inequality~\refeq{\eqfoursum} to get $I_1\le 2\eta k$.
+and we can substitute the inequality~\refeq{\eqionefoursum} to get $I_1\le 2\eta k$.
 
@@
   -\eta\bigl(\dd\{X_1^*; X_1\given  X_1 + \bar X_2\} - \dd\{X_1^*, X_1\}\cr
-&\hskip150pt+ \dd\{X_2^*, X_2\given X_2 + \bar X_1\}- \dd\{X_2^*, X_2\} \bigr) \cr
+&\hskip150pt+ \dd\{X_2^*; X_2\given X_2 + \bar X_1\}- \dd\{X_2^*, X_2\} \bigr) \cr
 &\ge I_1 + \dd\{X_1 + \bar X_2; X_2 + \bar X_1\} -\eta k, \cr
@@
 }$$
-by independence and the fact that subtraction is addition in $G$. Since
-$\dd\{X_1, X_2\}=k$, rearranging completes
+by independence and the fact that subtraction is addition in $G$. Rearranging terms completes
 the proof.\slug
@@
 
+\edef\lemitwobound{\the\sectcount.\the\thmcount}
+\proclaim Lemma \advthm. Let $G = \FF_2^n$
+and suppose $(X_1, X_2)\in \P(G)^2$ is a minimiser of $\tau$ with $\dd\{X_1, X_2\} = k$.
+Then letting
+$$I_2 = \II\{ X_1 + X_2 : X_1 + \bar X_1 \given X_1 + X_2 + \bar X_1 + \bar X_2\},$$
+where
+$X_1$ and $\bar X_1$ be copies of $X_1$ and $X_2$ and $\bar X_2$ be copies of $X_2$ such that $X_1$,
+$\bar X_1$, $X_2$, and $\bar X_2$ are all independent,
+we have
+$$I_2 \le 2\eta k + {2\eta(2\eta k- I_1)\over 1-\eta}.$$
+
+\proof
+Applying Lemma~{\lemfivetwo} four times gives us the inequalities
+\edef\eqitwofourineq{\the\eqcount}
+$$\eqalign{
+\dd\{X_1^*; X_1 + \bar X_1\}-\dd\{X_1^*, X_1\}&\le{1\over 2}\dd\{X_1, X_1\},\cr
+\dd\{X_2^*; X_2 + \bar X_2\}-\dd\{X_2^*, X_2\}&\le{1\over 2}\dd\{X_2, X_2\},\cr
+\dd\{X_1^*; X_1\given X_1-\bar X_1\}-\dd\{X_1^*,X_1\} &\le {1\over 2}\dd\{X_1, X_1\},\cr
+}\adveq$$
+and
+$$\dd\{X_2^*; X_2\given X_2-\bar X_2\}-\dd\{X_2^*,X_2\} \le{1\over 2}\dd\{X_2,X_2\}.\adveq$$
+These inequalities and Lemma~{\lemionehelper} together give
+\edef\eqitwounconditional{\the\eqcount}
+$$\eqalign{
+\dd\{X_1 + \bar X_1; X_2 + \bar X_2\} &\ge k-\eta\bigl(\dd\{X_1^*, X_1 + \bar X_1\}-\dd\{X_1^*, X_1\}\cr
+&\qquad + \dd\{X_2^*, X_2 + \bar X_2\}- \dd\{X_2^*, X_2\}\bigr) \cr
+&\ge k-{\eta\over 2}\dd\{X_1, X_1\} -{\eta\over 2} \dd\{X_2, X_2\} \cr
+}\adveq$$
+and
+\edef\eqitwoconditional{\the\eqcount}
+$$\eqalign{
+\dd\{X_1\given X_1 + \bar X_1;X_2\given X_2 + \bar X_2\}
+&\ge k-\eta\bigl(\dd\{X_1^*; X_1\given  X_1 + \bar X_1\} - \dd\{X_1^*, X_1\}\cr
+&\qquad+ \dd\{X_2^*, X_2\given X_2 + \bar X_2\}- \dd\{X_2^*, X_2\} \bigr)\cr
+&\ge k-{\eta\over 2}\dd\{X_1, X_1\} -{\eta\over 2} \dd\{X_2, X_2\}. \cr
+}\adveq$$
+Corollary~{\corfourtwo}, with $(Y_1, Y_2, Y_3, Y_4)$ set to $(X_2, X_1, \bar X_2, \bar X_1)$ this time,
+yields
+\edef\eqitwoidentity{\the\eqcount}
+$$\eqalign{
+2k &=
+\dd\{X_2 + \bar X_2;\ X_1 + \bar X_1\} +\dd\{ X_2\given X_2 + \bar X_2;\ X_1\given X_1+\bar X_1\}\cr
+&\qquad\qquad+ \II\{X_1 + X_2 : X_1+\bar X_1 \given X_1 + X_2 + \bar X_1 + \bar X_2\}\cr
+&=  \dd\{X_1 + \bar X_1;\ X_2 + \bar X_2\} + \dd\{ X_1\given X_1 + \bar X_1;\ X_2\given X_2+\bar X_2\}+I_2,\cr
+}\adveq$$
+into which we substitute~\refeq{\eqitwounconditional} and~\refeq{\eqitwoconditional} to obtain
+\edef\eqitwoalmostdone{\the\eqcount}
+$$ I_2 \le \eta\bigl( \dd\{X_1, X_1\} + \dd\{X_2,X_2\}\bigr).$$
+It remains to bound $\dd\{X_1, X_1\} + \dd\{X_2,X_2\}$.
+
+Since $\Eta\{X_1 + \bar X_1\}= \Eta\{X_1\} + \dd\{X_1, X_1\}$ and similarly for $X_2$, we may expand
+$$\eqalign{
+\dd\{X_1 + \bar X_1; X_2 + \bar X_2\} &= \Eta\{X_1 + \bar X_1 + X_2 + \bar X_2\} \cr
+&\qquad\qquad-{1\over 2} \Eta\{X_1+\bar X_1\} - {1\over 2} \Eta\{X_2+\bar X_2\} \cr
+&= \Eta\{X_1 + \bar X_1 + X_2 + \bar X_2\} - {1\over 2}\Eta\{X_1\}
+-{1\over 2} \Eta\{X_2\} \cr
+&\qquad\qquad-{1\over 2} \dd\{X_1,X_1\} - {1\over 2} \Eta\{X_2,X_2\}.\cr
+}$$
+But recall that the second conclusion of Lemma~{\lemionebound} was
+$$\Eta\{X_1 + X_2 + \bar X_1 + \bar X_2\} \le {1\over 2} \Eta\{X_1\} + {1\over 2} \Eta\{X_2\}+(2+\eta)k - I_1,$$
+meaning that
+$$\dd\{X_1+\bar X_1; X_2+\bar X_2\}\le (2+\eta)k - {1\over 2}\dd\{X_1, X_1\} -{1\over 2}\dd\{X_2,X_2\}-I_1.$$
+Chaining this with~\refeq{\eqitwounconditional} yields
+$$ k-{\eta\over 2}\dd\{X_1,X_1\} - {\eta\over 2}\dd\{X_2,X_2\} \le
+(2+\eta)k-{1\over 2}\dd\{X_1, X_1\} -{1\over 2}\dd\{X_2,X_2\}-I_1,$$
+which simplifies to
+$$\dd\{X_1,X_1\} + \dd\{X_2,X_2\} \le {2k +2\eta k - 2I_1\over 1-\eta} = 2k + {2(2\eta k -I_1)\over 1-\eta} $$
+and hence
+$$I_2 \le 2\eta k + {2\eta(2\eta k -I_1)\over 1-\eta}.\noskipslug$$
+
 \section References

### 2024-03-29 misc/expo/entropy.tex
Done entropy notes

@@
 }$$
-Lemma~\ref{\lemaone} with $X'$ in place of $X$ and $-Y$ in place of $Y$ yields
+Lemma~{\lemaone} with $X'$ in place of $X$ and $-Y$ in place of $Y$ yields
 $$\Eta\{X'-Y+Z\} - \Eta\{X'-Y\} \le \Eta\{Y-Z\} -\Eta\{Y\},$$
@@
 
-For the other inequality, we apply Lemma~\ref{\lemfiveone} with $W = Y-Z$ (and the variable $Z$ in that lemma
+For the other inequality, we apply Lemma~{\lemfiveone} with $W = Y-Z$ (and the variable $Z$ in that lemma
 set to anything that is independent of $X$) to obtain
@@
 
-\proof The second inequality of Lemma~\ref{\lemfivetwo} (setting $Y \gets Y-Z$ and $Z\gets W$) yields
+\proof The second inequality of Lemma~{\lemfivetwo} (setting $Y \gets Y-Z$ and $Z\gets W$) yields
 $$\dd\{X; Y-Z\given Y-Z-W\} - \dd\{X, Y-Z\} \le {\Eta\{Y-Z-W\} - \Eta\{W\}\over 2},$$
@@
 $$\dd\{X, Y-Z\} - \dd\{X, Y\} \le {\Eta\{Y-Z\} - \Eta\{Y\}\over 2}.$$
-Adding these two inequalites proves the lemma.\slug
+Adding these two inequalities proves the lemma.\slug
 
@@
 independent trials of $(X,Y)$ relative to $X+Y$. Then we have
-$$\max\bigl(\Eta\{X_1- X_2\}, \Eta\{X_1-Y_2\}\bigr) \le \Eta\{X+Y\} - \II\{X:Y\}.$$
+$$\max\bigl(\Eta\{X_1- X_2\}, \Eta\{X_1-Y_2\}\bigr) \le \Eta\{X+Y\} + 2\II\{X:Y\}.$$
 The right-hand side of this expression can also be written $2\Eta\{X\} + 2\Eta\{Y\} - \ee\{X,Y\}$.
@@
 
+Let us also extract the specific consequence of Lemma~{\lematwo} that we need for the polynomial
+Freiman--Ruzsa theorem.
+
+\edef\lematwopfr{\the\sectcount.\the\thmcount}
+\proclaim Lemma \advthm. Let $X$ and $Y$ be $G$-valued random variables, let $Z = X+Y$, and let $C$
+denote the support of $Z$. Then
+$$\eqalign{
+\sum_{z\in C} \pr\{Z=z\} \dd\bigl\{ &(X\given Z=z); (Y\given Z=z)\bigr\} \cr
+&\le 2\II\{X:Y\}+2\Eta\{Z\}-\Eta\{X,Y\}. \cr
+}$$
+
+\proof Let $(X_1, Y_1)$ and $(X_2, Y_2)$ be as in the proof of Lemma~{\lematwo}. Note that
+$X_1$ and $Y_2$ are conditionally independent copies of $X$ and $Y$ relative to $Z$, so it suffices
+to show that
+\edef\eqxoneytwoa{\the\eqcount}
+$$\eqalign{
+\Eta\{X_1 - Y_2\given Z\} - {1\over 2} &\Eta\{X_1\given Z\} - {1\over 2} \Eta\{Y_2\given Z\} \cr
+&\le 2\II\{X:Y\}+2\Eta\{Z\}-\Eta\{X,Y\}. \cr
+}\adveq$$
+Lemma~{\lematwo} tells us that $\Eta\{X_1 - Y_2\} \le \Eta\{Z\} + 2\II\{X:Y\}$, so that
+\edef\eqxoneytwob{\the\eqcount}
+$$\Eta\{X_1 - Y_2\given Z\} \le \Eta\{Z\} + 2\II\{X:Y\},\adveq$$
+since conditioning does not increase entropy. Then
+we expand
+\edef\eqxoneytwoc{\the\eqcount}
+$$\Eta\{X_1\given Z\} = \Eta\{X_1, X_1 + Y_1\} - \Eta\{Z\} = \Eta\{X,Y\} - \Eta\{Z\},\adveq$$
+and $\Eta\{Y_2\given Z\}$ is equal to this quantity as well. Subtracting~\refeq{\eqxoneytwoc}
+from~\refeq{\eqxoneytwob} gives us~\refeq{\eqxoneytwoa}, which completes the proof.\slug
+
 \advsect The Freiman--Ruzsa theorem
@@
 
+The reason we need the factor of two is that if $A$ is almost all of a subgroup, then $K$ is very close
+to $1$ and the largest
+subgroup of $\FF_2^n$ has cardinality just slightly above half that of $A$. That creates the possibility
+that $K^C$ is much less than $2$, making it impossible to satisfy the statement of the conjecture.
+Adding the factor of two solves this issue because if $A\subseteq H_0$ with $|A| \ge (1-\eps)|H_0|$, then
+we may take a subgroup $H$ of $H_0$ of index $2$ and cover $A$ by two cosets of $H$.
+
 The proof of this theorem will occupy the
@@
 \parenproclaim Lemma {\advthm} (Ruzsa covering lemma). Let $G$ be an abelian group and let $A$ and
-$B$ be finite subsets of $Z$. Then there is a set $L\subseteq A$ of size at most $|A+B|/|B|$ such that
+$B$ be finite subsets of $G$. Then there is a set $L\subseteq A$ of size at most $|A+B|/|B|$ such that
 $A\subseteq L+B-B$.
@@
 \medskip\noindent{\it Proof of Theorem~{\thmpfr} assuming Theorem~{\thmsubgroup}.}\enspace
-Let $A\subseteq \FF_2$ and $K\ge 1$
+Let $A\subseteq \FF_2^n$ and $K\ge 1$
 such that $|A+A|\le K|A|$. Let $U_A$ be the uniform distribution on $A$, so that $\Eta\{U_A\} = \lg|A|$.
@@
 $${2|H|\over |A|}|L| \le {2|H|\over |A|}\cdot {K^{C'/2+1}|A|^{1/2} \over |H|^{1/2} } 
-= 2 {C'/2+1} {|H|^{1/2} \over |A|^{1/2}} \le 2K{C'+1}$$
+= 2K^{C'/2+1} {|H|^{1/2} \over |A|^{1/2}} \le 2K^{C'+1}$$
 translates of $H'$.\slug
@@
 $$\dd\{X, -X\} = \Eta\{X+X'\} - {\Eta\{X\}\over 2} - {\Eta\{X'\} \over 2} = 0.$$
-On the other hand, suppose that $\Eta\{X+X'\} = \Eta\{X\}$, where $X'$ is an independent copy of $X$.
+On the other hand, suppose that $\Eta\{X+X'\} = 2\Eta\{X\}$, where $X'$ is an independent copy of $X$.
 Then
@@
 \proof The triangle inequality gives $\dd\{X_1,X_1\} = 0$, and since $-x = x$ in $\FF_2^n$,
-we also have $\dd\{X_1, - X_1\} = 0$. By the previous proposition, there is a subgroup $H$ such that
-$X_1$ has the same distribution as $H$. So $\dd\{X_1, U_H\} = 0$ and by the triangle inequality once again,
+we also have $\dd\{X_1, - X_1\} = 0$. By the previous proposition, there is a coset $S$ of a
+subgroup $H$ such that
+$X_1$ has the same distribution as $U_S$. So
+$$\dd\{X_1, U_H\} = \Eta\{U_S - U_H\} - {1\over 2}\Eta\{U_S\} - {1\over 2}\Eta\{U_H\} = 0,$$
+since $U_S-U_H$ is uniform on $S$ and $|H| = |S|$. Then by the triangle inequality once again,
 $\dd\{X_2, U_H\} = 0$ as well.\slug
@@
 If we can prove this proposition, then we have Theorem~{\thmsubgroup} with a certain constant $C'$ which
-we shall make explicit during the proof.
+shall be made explicit during the proof.
 
@@
 and furthermore, since
-$$\bigl| \dd\{X_1, U_H\} - \dd\{X_2, U_H\} \bigr| \le \dd\{X_1, X_2\}$$
+$$\bigl| \dd\{X_1^*, U_H\} - \dd\{X_2^*, U_H\} \bigr| \le \dd\{X_1^*, X_2^*\}$$
 by two invocations of the triangle inequality, neither $\dd\{X_1^*, U_H\}$ nor
@@
 2k &\ge \dd\{X_1^*; X_1 + \bar X_2\} + \dd\{X_2^*; X_2 + \bar X_1\} \cr
-&\qquad\qquad+\dd\{X_1^*; X_1\given X_1 + \bar X_2\} + \dd\{X_2^*; X_2\given X_2 + \bar X_1\} \cr
+&\qquad\qquad+\dd\{X_1^*; X_1\given X_1 - \bar X_2\} + \dd\{X_2^*; X_2\given X_2 - \bar X_1\} \cr
 &\qquad\qquad-2\dd\{X_1^*, X_1\} - 2\dd\{X_2^*, X_2\} .\cr
@@
 &\qquad\qquad\qquad\qquad+ \dd\{X_2^*, X_2 + \bar X_1\} + \dd\{X_2^*, X_2\given X_2+\bar X_2\},\cr
-&\qquad\qquad\qquad\qquad -2\dd\{X_1^*, X_1\} - 2\{X_2^*, X_2\} \bigl)\cr
+&\qquad\qquad\qquad\qquad -2\dd\{X_1^*, X_1\} - 2\dd\{X_2^*, X_2\} \bigl)\cr
 }$$
@@
 $$\eqalign{
-(1-\eta)k-I_1 &\ge \Eta\{X_1 + X_2 + \bar X_1 + \bar X_2\} - {1\over 2}\Eta\{X_1 + \bar X_2\} - 
+(1+\eta)k-I_1 &\ge \Eta\{X_1 + X_2 + \bar X_1 + \bar X_2\} - {1\over 2}\Eta\{X_1 + \bar X_2\} - 
 {1\over 2} \Eta\{X_2+\bar X_1\} \cr
@@
 
+\advsect Endgame
+
+We now approach the phase of the proof which the authors of~\ref{pfr2023} theatrically call the `endgame.'
+There is just one more lemma we need before we can give a full proof of Proposition~{\propmaintau}.
+
+\edef\lemseventwo{\the\sectcount.\the\thmcount}
+\proclaim Lemma~\advthm. Let $X_1$ and $Y_1$ be any $\FF_2^n$-valued random variables,
+and let $T_1$, $T_2$, and $T_3$ be $\FF_2^n$-valued random variables
+such that $T_1 + T_2 + T_3 = 0$ holds identically. Putting
+$$\delta = \II\{T_1 : T_2\} + \II\{T_1:T_3\} + \II\{T_2 : T_3\}$$
+and letting $\psi$ be the functional given by
+$$\psi\{T_1', T_2'\}
+= \dd\{T_1', T_2'\} + \eta\bigl( \dd\{X_1^*,T_1'\}-\dd\{X_1^*,X_1\}+\dd\{X_2^*,T_2'\}-\dd\{X_2^*, X_2\}\bigr),$$
+there exist random variables $T_1'$ and $T_2'$ such that
+$$\psi\{T_1', T_2'\} \le \biggl(1+{\eta\over 3}\biggr)\delta
++ {\eta\over 3}\sum_{i=1}^2\sum_{j=1}^3 \bigl(\dd\{X_i^*, T_j\}-\dd\{X_i^*, X_i\}\bigr)$$
+
+\proof From the fact that any two of the $T_j$ determine the full triple $(T_1, T_2, T_3)$,
+we have
+$$\Eta\{T_1, T_2\} = \Eta\{T_1, T_3\} = \Eta\{T_2,T_3\} = \Eta\{T_1, T_2, T_3\}.$$
+Then since $T_1 + T_2 = T_3$ we may apply Lemma~{\lematwopfr} with
+$(X,Y,Z) = (T_1, T_2, T_3)$ to bound
+$$\eqalign{
+\sum_{t_3\in \FF_2^n} \pr\{T_3 &= t_3\} \dd\bigl\{ (T_1\given T_3 = t_3);
+(T_2\given T_3 = t_3)\bigr\} \cr
+&\le 2\II\{T_1:T_2\} + 2\Eta\{T_3\} - \Eta\{T_1,T_2\} \cr
+&= 2\Eta\{T_1\} + 2\Eta\{T_2\} + 2\Eta\{T_3\} - 3\Eta\{T_1,T_2\} \cr
+&= \II\{T_1,T_2\} + \II\{T_1, T_3\} + \II\{T_2,T_3\} \cr
+&= \delta.
+}$$
+Letting $Z$ be a random variable independent of both $X_1^*$ and $X_2^*$, we apply Lemma~{\lemfiveone}
+with $(X,Z,Y,W) = (X_1^*, Z, T_1, T_3)$ to obtain
+$$\eqalign{
+\sum_{t_3\in\FF_2^n}\pr\{T_3 = t_3\}\bigl(\dd\bigl\{X_1^*;(T_1\given T_3&=t_3)\bigr\}-\dd\{X_1^*, X_1\}\bigr) \cr
+&= \dd\{ X_1^*, T_1\given T_3\} - \dd\{X_1^*, X_1\} \cr
+&\le \dd\{X_1^*, T_1\} + {1\over 2}\II\{T_1:T_3\} - \dd\{X_1^*, X_1\}, \cr
+}$$
+and applying the same lemma with $(X,Z,Y,W) = (X_2^*, Z, T_2, T_3)$ yields
+$$\eqalign{
+\sum_{t_3\in\FF_2^n}\pr\{T_3 = t_3\}\bigl(\dd\bigl\{X_2^*;(T_2\given T_3&=t_3)\bigr\}-\dd\{X_2^*, X_2\}\bigr) \cr
+&\le \dd\{X_2^*, T_2\} + {1\over 2}\II\{T_2:T_3\} - \dd\{X_2^*, X_2\}. \cr
+}$$
+Putting the three observations together, we have
+$$\eqalign{
+&\sum_{t_3\in\FF_2^n}\pr\{T_3 = t_3\}\psi\bigl\{(T_1\given T_3=t_3);(T_2\given T_3=t_3)\bigr\} \cr
+&\hskip10pt = \sum_{t_3\in\FF_2^n}\pr\{T_3 = t_3\}\Bigl(\dd\bigl\{(T_1\given T_3=t_3);(T_2\given T_3=t_3)\bigr\} \cr
+&\hskip103pt + \eta\dd\bigl\{X_1^*;(T_1\given T_3=t_3)\bigr\}+ \eta\dd\bigl\{X_2^*;(T_2\given T_3=t_3)\bigr\} \cr
+&\hskip192pt - \eta\dd\{X_1^*, X_1\} - \eta\dd\{X_2^*, X_2\} \Bigr) \cr
+&\hskip10pt \le \delta + \eta\bigl( \dd\{X_1^*, T_1\} - \dd\{X_1^*, X_1\}+\dd\{X_2^*, T_2\}-\dd\{X_2^*, X_2\}\cr
+&\hskip180pt 
++ {1\over 2} \II\{T_1 : T_3\} + {1\over 2}\II\{T_2:T_3\} \bigr), \cr
+}$$
+Choosing some $t_3$ in the support of $T_3$ such that the value of $\phi$ inside
+the sum is minimal and then setting $T_{1,3}' = (T_1\given T_3 = t_3)$ and
+$T_{2,3}' = (T_2\given T_3 = t_3)$, we have
+$$\eqalign{
+\psi\bigl\{T_{1,3}', T_{2,3}'\bigr\} \le \delta + \eta\bigl( \dd\{X_1^*, T_1\} - \dd\{X_1^*, X_1\}
+&+ \dd\{X_2^*, T_2\} - \dd\{X_2^*, X_2\} \cr
+& + {1\over 2} \II\{T_1 : T_3\} + {1\over 2}\II\{T_2:T_3\} \bigr)\cr
+}$$
+We can now repeat this for all six permutations $(\alpha, \beta, \gamma)$ of $(1,2,3)$ and average the
+corresponding bounds for $(T_\alpha, T_\beta, T_\gamma)$ to obtain
+$$\eqalign{
+{1\over 6} \sum_{(\alpha,\beta,\gamma)\in {\frak S}_3}
+\psi\bigl\{T_{\alpha,\gamma}', T_{\beta,\gamma}'\bigr\}
+\le \delta &- \eta\dd\{X_1^*, X_1\} - \eta\dd\{X_2^*, X_2\} \cr
+&+ {\eta\over 6}\sum_{(\alpha,\beta,\gamma)\in {\frak S}_3}
+\bigl( \dd\{X_1^*, T_\alpha\} + \dd\{X_2^*, T_\beta\} \cr
+&\hskip68pt+ {1\over 2} \II\{T_\alpha : T_\gamma\} + {1\over 2}\II\{T_\beta:T_\gamma\} \bigr).\cr
+}$$
+Since each of $\{1,2,3\}$ appears twice as $\alpha$ and twice as $\beta$, we have
+$$\sum_{(\alpha,\beta,\gamma)\in {\frak S}_3} \bigl(\dd\{X_1^*, T_\alpha\} + \dd\{X_2^*, T_\beta\}\bigr)
+= 2\sum_{i=1}^2\sum_{j=1}^3 \dd\{X_i^*, T_j\}$$
+and
+$$\sum_{(\alpha,\beta,\gamma)\in {\frak S}_3}
+\biggl({1\over 2} \II\{T_\alpha : T_\gamma\} + {1\over 2}\II\{T_\beta:T_\gamma\}\biggr) = 2\delta.$$
+Consequently, the average of
+$\psi\bigl\{T_{\alpha,\gamma}', T_{\beta,\gamma}'\bigr\}$
+over all permutations $(\alpha,\beta,\gamma)$ is bounded above by
+$$\biggl(1+{\eta\over 3}\biggr)\delta
+- \eta\dd\{X_1^*, X_1\} -\eta\dd\{X_2^*, X_2\} + {\eta\over 3}\sum_{i=1}^2\sum_{j=1}^3 \dd\{X_i^*, T_j\},$$
+so the result follows by letting $(T_1', T_2')$ equal the pair $(T_{\alpha,\gamma}', T_{\beta,\gamma}')$
+for the choice of $(\alpha,\beta,\gamma)$ that minimises
+$\psi\bigl\{T_{\alpha,\gamma}', T_{\beta,\gamma}'\bigr\}$.\slug
+
+We are now able to prove Proposition~{\propmaintau}, which we shall restate in the contrapositive.
+Just as in previous sections, the functional $\tau$ depends on the fixed random variables $X_1^*$ and $X_2^*$
+as well as the choice of constant $\eta = 1/9$.
+
+\proclaim Proposition~{\propmaintau}$\mathbold'$. Let $X_1$ and $X_2$ be $\FF_2^n$-valued random variables
+with the property that $\tau\{X_1', X_2'\} \ge \tau\{X_1, X_2\}$ for all random variables
+$X_1'$ and $X_2'$ on $\FF_2^n$. Then $\dd\{X_1, X_2\} = 0$.
+
+\proof Let $k=\dd\{X_1, X_2\}$. Let $X_1$ and $\bar X_1$ be copies of $X_1$ and $X_2$ and $\bar X_2$ copies
+of $X_2$ such that $X_1$, $\bar X_1$, $X_2$, and $\bar X_2$ are all independent. Let $I_1$ and $I_2$
+be as in the previous section, and let
+$$I_3 = \II\{\bar X_1 + X_2 : X_1 + \bar X_1\given X_1 + X_2 + \bar X_1 + \bar X_2\},$$
+so that $I_3 = I_2$ (which follows from interchanging the r\^oles of $X_1$ and $\bar X_1$).
+For brevity of notation, let
+$$U = X_1+X_2,\qquad V=\bar X_1 + X_2,\qquad W = X_1 + \bar X_1,$$
+and
+$$S = X_1 + X_2 + \bar X_1 + \bar X_2.$$
+Then
+$$I_1 = \II\{ U : V\given S\},\qquad I_2 = \II\{W:U\given S\},\qquad\hbox{and}\qquad I_3 = \II\{V:W\given S\}.$$
+Lemma~{\lemionebound} gave us
+$$I_3 = I_2 \le 2\eta k + {2\eta(2\eta k-I_1)\over 1-\eta},$$
+so setting
+$$\bar\delta = \II\{U:V\given S\} + \II\{W:U\given S\} + \II\{V:W\given S\},$$
+we have
+\edef\eqbardeltabound{\the\eqcount}
+$$\eqalign{
+\bar \delta &\le I_1+4\eta k+{4\eta(2\eta k - I_1)\over 1-\eta} \cr
+&= {I_1 - \eta I_n + 4\eta k - 4\eta^2 k + 6\eta^2 k - 4\eta I_1\over 1-\eta} \cr
+&= 6\eta k + {(1-5\eta)I_1 + 4\eta k + 4\eta^2 k - 6\eta k + 6\eta^2 k \over 1-\eta} \cr
+&= 6\eta k + {(1-5\eta)I_1 - 2\eta k + 10\eta^2 k \over 1-\eta} \cr
+&= 6\eta k - {1-5\eta \over 1-\eta}(2\eta k - I_1) \cr
+}\adveq$$
+
+Now we peform invocations of Lemma~{\lemsevenone} to obtain bounds on various Ruzsa distances. Setting
+$(X,Y,Z,W) = (X_1^*, X_1, X_2, \bar X_1+\bar X_2)$, gives
+$$\eqalign{
+\dd\{X_1^*; U\given S\} - \dd\{X_1^*, X_1\}
+&\le {1\over 2}\bigl( \Eta\{S\} + \Eta\{U\} - \Eta\{X_1\} - \Eta\{\bar X_1+\bar X_2\}\bigr)\cr
+&= {1\over 2} \bigl( \Eta\{S\} - \Eta\{X_1\}\bigr),
+}$$
+where in the second line we used the fact that
+$$\Eta\{U\} = \Eta\{X_1 + X_2\} = \Eta\{\bar X_1 + \bar X_2\}.$$
+Similarly, we have
+$$\eqalign{
+\dd\{X_2^*; U\given S\} - \dd\{X_2^*, X_2\} &= {1\over 2} \bigl( \Eta\{S\} - \Eta\{X_2\}\bigr), \cr
+\dd\{X_1^*; V\given S\} - \dd\{X_1^*, X_1\} &= {1\over 2} \bigl( \Eta\{S\} - \Eta\{X_1\}\bigr), \cr
+\dd\{X_2^*; V\given S\} - \dd\{X_2^*, X_2\} &= {1\over 2} \bigl( \Eta\{S\} - \Eta\{X_2\}\bigr), \cr
+}$$
+and
+$$\dd\{X_1^*, W\given S\} - \dd\{X_1^*, X_1\}
+= {1\over 2} \bigl( \Eta\{S\} + \Eta\{W\} - \Eta\{X_1\} - \Eta\{W'\}\bigr),$$
+where $W' = X_2 + \bar X_2$. To address the asymmetry in the last bound, we note that for any fixed value
+$s$ taken by $S$, we have $W' = W + s$, so $\dd\{X_2^*; W\given S\} = \dd\{X_2^*; W'\given S\}$. Now
+applying Lemma~{\lemsevenone} to $W'$ yields
+$$\dd\{X_1^*; W\given S\} - \dd\{X_1^*, X_1\}
+= {1\over 2} \bigl( \Eta\{S\} + \Eta\{W\} - \Eta\{X_1\} - \Eta\{W'\}\bigr).$$
+The sum of these six inequalities is
+\edef\eqsixsum{\the\eqcount}
+$$\sum_{i=1}^2 \sum_{A\in \{U,V,W\}} \bigl( \dd\{X_i^*; A\given S\} - \dd\{X_i^*, X_i\}\bigr)
+\le 3\Eta\{S\} - {3\over 2} \Eta\{X_1\} - {3\over 2} \Eta\{X_2\}.\adveq$$
+The second part of Lemma~{\lemionebound} states that
+$$\Eta\{S\} \le {1\over 2} \Eta\{X_1\} + {1\over 2} \Eta\{X_2\} + (2+\eta)k - I_1,$$
+which can be plugged into~\refeq{\eqsixsum} to give
+\edef\eqdistancebounds{\the\eqcount}
+$$\eqalign{
+\sum_{i=1}^2 \sum_{A\in \{U,V,W\}} \bigl( \dd\{X_i^*; A\given S\} - \dd\{X_i^*, X_i\}\bigr)
+&\le 3(2+\eta)k - 3I_1 \cr
+&\le 6k -3\eta k + 6\eta k - 3I_1 \cr
+&\le (6-3\eta)k + 3(2\eta k - I_1).\cr
+}\adveq$$
+
+Let $\psi$ be defined as in Lemma~{\lemseventwo}. By Lemma~{\lemionehelper}, $k\le \psi\{T_1', T_2'\}$
+for all random variables $T_1'$ and $T_2'$, so
+for any random variables $(T_1, T_2, T_3)$, the pair $(T_1',T_2')$ furnished by Lemma~{\lemseventwo}
+satisfies
+$$k\le \psi\{T_1', T_2'\} \le \biggl(1+{\eta\over 3}\biggr)\delta
++ {\eta\over 3}\sum_{i=1}^2\sum_{j=1}^3 \bigl(\dd\{X_i^*, T_j\}-\dd\{X_i^*, X_i\}\bigr).$$
+In particular, since $(U+V+W\given S=s)$ is identically zero for all possible values $s$ of $S$,
+we can set
+$$T_1 = (U\given S=s),\qquad T_2 = (V\given S=s), \qquad\hbox{and}\qquad T_3 = (W\given S=s)$$
+and then average the above inequality over all $s\in \FF_2^n$, weighted by $\pr\{S=s\}$ to obtain
+$$k\le \biggl(1+{\eta\over 3}\biggr)\bar\delta
++ {\eta\over 3}\sum_{i=1}^2\sum_{A\subseteq\{U,V,W\}} \bigl(\dd\{X_i^*, A\given S\}-\dd\{X_i^*, X_i\}\bigr).$$
+We now substitute the upper bounds~\refeq{\eqbardeltabound} and~\refeq{\eqdistancebounds} to get
+$$\eqalign{
+k &\le \biggl(1+{\eta\over 3}\biggr)\biggl(6\eta k - {1-5\eta \over 1-\eta}(2\eta k - I_1)\biggr)
++ {\eta\over 3}\biggl( (6-3\eta)k + 3(2\eta k - I_1)\biggr) \cr
+&= (8\eta - \eta^2)k + \biggl(\eta-\Bigl(1+{\eta\over 3}\Bigr){1-5\eta\over 1-\eta} \biggr)
+(2\eta k - I_1).\cr
+}$$
+With the choice $\eta = 1/9$,
+$$\eta-\Bigl(1+{\eta\over 3}\Bigr){1-5\eta\over 1-\eta} = -{11\over 27} \le 0$$
+and $2\eta k - I_1\ge 0$ by Lemma~{\lemionebound}, hence one concludes that
+$$ k \le (8\eta + \eta^2)k = {73\over 81}k,$$
+which is nonsense unless $k=0$.\slug
+
 \section References

### 2024-04-30 about/vita.tex
eca
Ahoy

@@
 \smallskip
-Supervisors: Andrew Granville and Hamed Hatami.
+Supervisor: Hamed Hatami.
 \medbreak
-\leftright{\bf M.Sc.\ Mathematics}{\bf May 2021 -- present}
+\leftright{\bf M.Sc.\ Mathematics}{\bf May 2021 -- August 2022}
 \leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
@@
 \papitem (with Jonah Saks)
-On the homology of several number-theoretic set families. To appear in
-{\sl Enumerative Combinatorics and Applications} (2024), 17~pp.
+On the homology of several number-theoretic set families.
+{\sl Enumerative Combinatorics and Applications}\/ {4},2 (2024), \#S2R12, 11~pp.
 [arXiv:2206.12535]
@@
 \goodbreak
-\sectheader{Reports}
+\sectheader{Reports and Theses}
 
@@
 
-\sectheader{Published Short Fiction}
+\sectheader{Published Fiction}
 
@@
 
+\leftright{\bf Co-Editor}{\bf December 2023 -- present}
+\leftright{\sl Ahoy Literary Magazine}{\sl Montr\'eal, QC, Canada}
+\smallskip
+Founding co-editor of a semiannual print magazine that publishes poetry, prose, and art.
+In charge of English short fiction submissions and the overall typesetting. Share duties with the other
+co-editors related to marketing, emails, and sales.
+\medbreak
+
 \starleftright{\bf Teaching assistant}{\bf September 2021 -- present}
@@
 \smallskip
-\thing Fall 2023: MATH 240 Discrete Structures
+\thing Winter 2024, Fall 2023: MATH 240 Discrete Structures
 \smallskip
@@
 \endgroup
+\medbreak
+
+\leftright{\bf DRP Mentor}{\bf January 2022 -- April 2024}
+\leftright{\sl Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
+\smallskip
+Met with two undergraduate students as part of the Directed Reading Program
+on a weekly basis to give them an introduction to research-level
+mathematics in a casual setting. Focused on topics in extremal combinatorics related to the increasing
+triples problem, as well as topics related to Khovanskii's theorem.
 \medbreak
@@
 around forty graduate-level students were in attendance.
-\medbreak
-
-\leftright{\bf DRP Mentor}{\bf January -- April 2022}
-\leftright{\sl Department of Mathematics and Statistics, McGill University}{\sl Montr\'eal, QC, Canada}
-\smallskip
-Met with two undergraduate students as part of the Directed Reading Program
-on a weekly basis to give them an introduction to research-level
-mathematics in a casual setting. Focused on topics in extremal combinatorics related to the increasing
-triples problem, as well as topics related to Khovanskii's theorem.
 \medbreak

### 2024-04-30 misc/expo/entropy.tex
Ahoy

@@
 \dd\{X; Y\given Y-Z\} - \dd\{X, Y\} &\le {\Eta\{Y-Z\} - \Eta\{Z\}\over 2} \cr
-&= {1\over 2} \dd\{Y, Z\} + {1\over 4}\Eta\{Z\} - {1\over 4}\Eta\{Z\}.\cr
+&= {1\over 2} \dd\{Y, Z\} + {1\over 4}\Eta\{Y\} - {1\over 4}\Eta\{Z\}.\cr
 }$$
@@
 &= 2\Eta\{T_1\} + 2\Eta\{T_2\} + 2\Eta\{T_3\} - 3\Eta\{T_1,T_2\} \cr
-&= \II\{T_1,T_2\} + \II\{T_1, T_3\} + \II\{T_2,T_3\} \cr
+&= \II\{T_1:T_2\} + \II\{T_1:T_3\} + \II\{T_2:T_3\} \cr
 &= \delta.
@@
 }$$
-Choosing some $t_3$ in the support of $T_3$ such that the value of $\phi$ inside
+Choosing some $t_3$ in the support of $T_3$ such that the value of $\psi$ inside
 the sum is minimal and then setting $T_{1,3}' = (T_1\given T_3 = t_3)$ and
@@
 \bar \delta &\le I_1+4\eta k+{4\eta(2\eta k - I_1)\over 1-\eta} \cr
-&= {I_1 - \eta I_n + 4\eta k - 4\eta^2 k + 6\eta^2 k - 4\eta I_1\over 1-\eta} \cr
+&= {I_1 - \eta I_n + 4\eta k - 4\eta^2 k + 8\eta^2 k - 4\eta I_1\over 1-\eta} \cr
 &= 6\eta k + {(1-5\eta)I_1 + 4\eta k + 4\eta^2 k - 6\eta k + 6\eta^2 k \over 1-\eta} \cr
@@
 
-Now we peform invocations of Lemma~{\lemsevenone} to obtain bounds on various Ruzsa distances. Setting
+Now we perform invocations of Lemma~{\lemsevenone} to obtain bounds on various Ruzsa distances. Setting
 $(X,Y,Z,W) = (X_1^*, X_1, X_2, \bar X_1+\bar X_2)$, gives
@@
 $$\eqalign{
-\dd\{X_2^*; U\given S\} - \dd\{X_2^*, X_2\} &= {1\over 2} \bigl( \Eta\{S\} - \Eta\{X_2\}\bigr), \cr
-\dd\{X_1^*; V\given S\} - \dd\{X_1^*, X_1\} &= {1\over 2} \bigl( \Eta\{S\} - \Eta\{X_1\}\bigr), \cr
-\dd\{X_2^*; V\given S\} - \dd\{X_2^*, X_2\} &= {1\over 2} \bigl( \Eta\{S\} - \Eta\{X_2\}\bigr), \cr
+\dd\{X_2^*; U\given S\} - \dd\{X_2^*, X_2\} &\le {1\over 2} \bigl( \Eta\{S\} - \Eta\{X_2\}\bigr), \cr
+\dd\{X_1^*; V\given S\} - \dd\{X_1^*, X_1\} &\le {1\over 2} \bigl( \Eta\{S\} - \Eta\{X_1\}\bigr), \cr
+\dd\{X_2^*; V\given S\} - \dd\{X_2^*, X_2\} &\le {1\over 2} \bigl( \Eta\{S\} - \Eta\{X_2\}\bigr), \cr
 }$$
@@
 $$\dd\{X_1^*, W\given S\} - \dd\{X_1^*, X_1\}
-= {1\over 2} \bigl( \Eta\{S\} + \Eta\{W\} - \Eta\{X_1\} - \Eta\{W'\}\bigr),$$
+\le {1\over 2} \bigl( \Eta\{S\} + \Eta\{W\} - \Eta\{X_1\} - \Eta\{W'\}\bigr),$$
 where $W' = X_2 + \bar X_2$. To address the asymmetry in the last bound, we note that for any fixed value
@@
 applying Lemma~{\lemsevenone} to $W'$ yields
-$$\dd\{X_1^*; W\given S\} - \dd\{X_1^*, X_1\}
-= {1\over 2} \bigl( \Eta\{S\} + \Eta\{W\} - \Eta\{X_1\} - \Eta\{W'\}\bigr).$$
+$$\dd\{X_2^*; W\given S\} - \dd\{X_2^*, X_2\}
+\le {1\over 2} \bigl( \Eta\{S\} + \Eta\{W\} - \Eta\{X_2\} - \Eta\{W'\}\bigr).$$
 The sum of these six inequalities is
@@
 &\le 6k -3\eta k + 6\eta k - 3I_1 \cr
-&\le (6-3\eta)k + 3(2\eta k - I_1).\cr
+&= (6-3\eta)k + 3(2\eta k - I_1).\cr
 }\adveq$$
@@
 $$k\le \biggl(1+{\eta\over 3}\biggr)\bar\delta
-+ {\eta\over 3}\sum_{i=1}^2\sum_{A\subseteq\{U,V,W\}} \bigl(\dd\{X_i^*, A\given S\}-\dd\{X_i^*, X_i\}\bigr).$$
++ {\eta\over 3}\sum_{i=1}^2\sum_{A\in\{U,V,W\}} \bigl(\dd\{X_i^*, A\given S\}-\dd\{X_i^*, X_i\}\bigr).$$
 We now substitute the upper bounds~\refeq{\eqbardeltabound} and~\refeq{\eqdistancebounds} to get

### 2024-06-27 misc/expo/kelleymeka.tex
Added many photos

@@
 It is easy to check that if $\mu$ has $\norm{\mu}_1 = 1$, then so does $\mu*\mu$ and $\mu\circ\mu$.
-We shall say that $\mu:G\to \RR_{\ge 0}$ is a {\it probability measure} on $G$ if $\norm{\mu}_1 = 1$. The
-following proposition concerns such measures.
+The following proposition concerns functions with this $1$-norm.
 
@@
 \proclaim Proposition \advthm. Let $G$ be a finite abelian group.
-If $\mu:G\to \RR_{\ge 0}$ is a probability measure, then
+If $\mu:G\to \RR_{\ge 0}$ has $\norm{\mu}_1 = 1$, then
 $$ \hat{\mu-1} = \hat \mu (1-\one_{\chi_0}).$$
@@
 If $\chi\ne\chi_0$, then the expectation vanishes, and
-if $\chi = \chi_0$, then the expection clearly equals $1$, and $\mu(\chi_0) = \norm{\mu}_1 = 1$, so the whole
+if $\chi = \chi_0$, then the expectation
+clearly equals $1$, and $\mu(\chi_0) = \norm{\mu}_1 = 1$, so the whole
 expression is zero.\slug
 
-If $\mu:G\to \RR_{\ge 0}$ is a probability measure and $f,g:G\to \CC$ we write
+For $\mu:G\to \RR_{\ge 0}$ with $\norm{\mu}_1 = 1$ and arbitrary $f,g:G\to \CC$ we write
 $$\langle f,g\rangle_\mu = \ex_{x\in G} \mu(x)f(x)\bar{g(x)}$$
@@
 \parenproclaim Proposition~{\advthm} (Monotonicity of $L_p$ norms). Let $G$ be a finite abelian group.
-Let $\mu:G\to \RR_{\ge 0}$ be a probability measure and let $f:G\to \CC$.
+Let $\mu:G\to \RR_{\ge 0}$ have $\norm{\mu}_1 = 1$ and let $f:G\to \CC$.
 For $1\le p<q<\infty$, we have
@@
 
-For convenience, when $\mu:G\to \RR_{\ge 0}$ is a probability measure and $X\subseteq G$, we write
+For convenience, when $\mu:G\to \RR_{\ge 0}$ has $\norm{\mu}_1 = 1$ and $X\subseteq G$, we write
 $$\mu(X) = \norm{\one_X}_{1(\mu)} = \ex_{x\in G} \mu(x) \one_X(x),$$
@@
 Lemma~{\dependentrandom} with the same $p$ and $\eps$, but $\delta$ set to $\eps/8$, $S$ set to $S'$,
-and $B_1 = B_2 = G$ so that $\mu = \mu_{B_1} = \mu_{B_2}$ is the uniform measure on $G$. Hence the
+and $B_1 = B_2 = G$ so that $\mu = \mu_{B_1} = \mu_{B_2}$ is the constant function $1$ on $G$. Hence the
 sets $A_1$ and $A_2$ given by the lemma will each have density
@@
 Proposition~{\convolutions}), $\mu_A\circ\mu_A$ is a nonnegative function, and note also that if
-$\nu$ is the uniform measure on $\FF_q^n$, then for any $\chi:\hat{\FF_q^n}\to \CC$,
-$$\hat\nu(\chi) = \ex_{x\in G} \nu(x)\chi(-x) = \cases{q^{-n}, & if $\chi$ is the trivial character;
+$\nu$ is the constant function $1$ on $\FF_q^n$, then for any $\chi:\hat{\FF_q^n}\to \CC$,
+$$\hat\nu(\chi) = \ex_{x\in G} \nu(x)\chi(-x) = \cases{1, & if $\chi$ is the trivial character;
 \cr 0, & otherwise.}$$
 This implies in particular that $\hat\nu \ge 0$,
-so by Lemma~{\unbalancing} applied with $f=\mu_A\circ\mu_A$ and the uniform measure for $\nu$, we find
+so by Lemma~{\unbalancing} applied with $f=\mu_A\circ\mu_A$ and the constant function $1$ for $\nu$, we find
 that $\norm{\mu_A\circ\mu_A}_{p'}\ge 1+\eps/4$ for some

### 2024-07-25 teaching/math240f24/notes.tex
two lectures done

\input fontmac
\input mathmac
\widemargins
\bookheader{\tenpoint\sc marcel goh}{\tenpoint\sc math {\oldstyle 240} fall {\oldstyle 2024}}

\def\datestamp#1#2 {\llap{{\oldstyle #1}{\rm.}{\sc #2}\hskip12pt}}
\def\bar{\overline}
\def\symdiff{\mathrel{\triangle}}

\maketitle{MATH 240 Fall 2024}{notes by}{Marcel Goh}{}

\floattext 4.5 \ninebf A note on these notes.
\ninepoint After each class, this document will be updated with the new material that was
just covered. The datestamps in the left margin indicate when the notes from each day start.
Subsections labelled with a $*$ are optional. This document is based on notes by
Jeremy Macdonald, but any errors are likely my own. Please email me if you find any.

\bigskip

\advsect Set theory

\datestamp{1}{}
A {\it set} is a collection of distinct objects, called its {\it elements} or its {\it members}. If
$x$ is a member of set $A$, then we write $x\in A$, and if $x$ is not an element of $A$,
then we write $x\notin A$.
Sets can be written by listing out its elements. For example,
$$\{1,4,7,10,\sqrt{782}\}$$
and
$$\bigl\{\{1,2\}, \pi, \{4\}\bigr\}$$
are both sets (the second example shows that sets can themselves contain other sets). The order of the elements
is not important, and duplicate elements are ignored (so $\{1,2\} = \{2,1\} = \{1,1,2\}$).
The notation using $\{$ and $\}$
is useful for defining small, concrete examples, but expressing large sets can become
very cumbersome. The first way one can describe larger sets is to use the $\ldots$ symbol and the power of
suggestion. For instance, anyone faced with the notation
$$A = \{1,3,5,7,9,\ldots\}$$
can quickly guess that this set is supposed to contain all the positive odd integers. We can also use $\ldots$
to define finite sets. Most Canadians will be able to tell you that the set
$$\{\hbox{\rm Alberta}, \hbox{\rm British Columbia}, \ldots, \hbox{\rm Yukon}\}$$
of provinces and territories contains $13$ elements.
But this notation inherently produces some ambiguity. For example, since the sequence of positive palindromic
binary numbers starts $1,3,5,7,9,15,17,21,27,\ldots$,
we are left with some doubt as to whether the set $A$ above should be the set of odd numbers or the set of
palindromic binary numbers.

But there is another, less ambiguous way to define large sets. It is called set-builder notation
and it refers to any construction of the form
$$\bigl\{x \in U : P(x)\bigr\},$$
where $x$ is a variable, $U$ is a set,
and $P$ is a statement about $x$. The resulting set contains {\sl all $x$ such that $P(x)$
holds}. For example, letting $\NN$ denote the set $\{0,1,2,3,\ldots\}$ of counting numbers (more on this later),
to define the set of all odd numbers, we can write
$$A = \{ x\in \NN : \hbox{there exists}\ k\in \NN\ \hbox{such that}\ x = 2k + 1\}.$$
Note that the statement $P(x)$ must contain $x$, but it may also contain other previously defined symbols, as well
as new symbols defined within the statement (such as $k$ in the example above).

\medskip\boldlabel Special sets of numbers.
There are certain infinite sets of numbers that are used so often as to be given special bold notation.
Back in elementary school, you learned to use the counting numbers $0, 1, 2, 3, \ldots$. We already saw this
set in the previous paragraph; in the business,
this set is known as the {\it natural numbers}, because if you go on a nature hike you
can use them to count the number of bluebells, donkeys, etc.~that you see.
(Many mathematicians use the symbol $\NN$ to denote this set without zero. When in doubt,
clarify with the person you're talking to; in this class, $0\in \NN$.)

Sometime towards the start of junior high you were introduced to the concept of natural numbers. The set
$$\ZZ=\{ \ldots, -3, -2, -1, 0, 1, 2, 3, \ldots\}$$
is called the set of {\it integers} or {\it whole numbers}.
(The word {\it integer} just means ``whole'' in Latin; cf.~French {\it entier}. We use the letter $\ZZ$
because of the German word {\it Zahl} meaning ``number.'')

Even before you learned about negative numbers, you probably learned about fractions. They can be defined
in set-builder notation as a collection of ratios of integers, where the denominator is not zero:
$$\QQ = \{ p/q : p, q\in \ZZ, q\ne 0\}.$$
(The nitpicky reader will notice that $p/q$ is not stipulated to be a member of any set here. This is because
a rigorous definition of $\QQ$ involves quotienting out by an equivalence relation, which we don't know how to
do yet.)
This is the set of {\it rational numbers}.
Remember that sets only contain distinct elements, so $2/3$, $4/6$, $6/9$, etc.~are all considered the
same rational number.

Lastly, we have the set $\RR$ of real numbers. Constructing this set using only notions from set theory
and logic is quite the byzantine task and well outside the scope of this course, but you can think of
$\RR$ as the set of decimal numbers with a finite number of digits to the left of the decimal point, and
a possibly infinite number of digits to the right of the decimal point.

A set of numbers near and dear to many mathematicians' hearts is the set of {\it prime} numbers $P$,
defined by
$$P = \bigl\{ p\in \NN : p\ge 2,\ \hbox{and if $p = ab$ then $\{a,b\} = \{1,p\}$}\bigr\}.$$
(You may have met a different definition of prime numbers in the past. Pause a moment and convince
yourself that the statement above defines the set of prime numbers as you know it.)

\medskip\boldlabel Set inclusion.
When we use set builder notation $B = \bigl\{x\in A : P(x)\bigr\}$, every element of $B$ is necessarily
a member of the set $A$ as well, since $B$ is defined to be the set of all $x$ in $A$ satisfying
$P(x)$. This is one way in which we can obtain
a {\it subset} of another set. More generally, we write $B\subseteq A$ if every element of $B$ is also
an element of $A$, and $B\supseteq A$ if every element of $A$ is an element of $B$. Sometimes, if $B\supseteq A$,
we say that $B$ {\it contains} $A$, or $B$ {\it includes} $A$. For example, we have the chain
$$\NN\subseteq \ZZ \subseteq \QQ\subseteq \RR$$
for the special sets of numbers defined earlier.

Symbols like $\subseteq$, $\supseteq$, and $=$ (that are
used to produce statements) can be
negated with a slash; for example, if there is some element of $B$ that is not an element of $A$,
then we write $B\not\subseteq A$.

The concept of set inclusion is important, because
the most common way to prove that two sets $A$ and $B$ are equal is to show that $A$ is a subset of $B$,
then show that $B$ is a subset of $A$. We illustrate this with the following example.

\proclaim Proposition \advthm. The sets
$$A = \{ x\in \ZZ : \hbox{there exists}\ k\in \ZZ\ \hbox{such that}\ x = 2k + 1\}$$
and
$$A = \{ x\in \ZZ : \hbox{there exists}\ l\in \ZZ\ \hbox{such that}\ x = 2l + 5\}$$
are equal. (Both are different ways of expressing the set of all odd integers.)

\proof Let $x\in A$. Then there exists $k\in\ZZ$ such that $x = 2k+1$. Letting $l = k-2$, we find that
$l$ is an integer (since $k$ was). Furthermore,
$$x = 2k+1 = 2k - 4 + 5 = 2(k-2)+5 = 2l+5.$$
We have found $l$ such that $x = 2l+5$, so $x\in B$. This shows that $A\subseteq B$.

On the other hand, let $x\in B$, so that there exists $l\in \ZZ$ such that $x = 2l+5$. Now we let $k = l+2$;
$k\in \ZZ$ since $l\in \ZZ$. We have
$$x = 2l+5 = 2l + 4 + 1 = 2(l+2) + 1 = 2k+1.$$
This shows that $x\in A$, and we have proved that $B\subseteq A$. This combined with the previous paragraph
shows that $A = B$.\slug

The above result is not so important, but pay attention to the structure of this proof. It is called a ``proof
by double inclusion,'' since we have shown that $A$ includes $B$ and $B$ includes $A$.

\medskip\boldlabel Set operations.
Now we describe a number of operations that may be performed on sets to produce other sets. They can all be
built up from the following two operations.
\medskip
\item{$\bullet$} The {\it union} $A\cup B$ of two sets $A$ and $B$ is the set of all elements that are either in
$A$ or in $B$ (or both).
\smallskip
\item{$\bullet$} The {\it intersection} $A\cap B$ of $A$ and $B$ is the set of all elements that are in both $A$
and $B$.
\medskip
As an example, if $A = \{1,2,4\}$ and $B = \{1,3,5\}$, then $A\cup B = \{1,2,3,4,5\}$ and $A\cap B = \{1\}$.

To avoid logical difficulties, we always assume that the sets we're working with are a subset of some
larger ambient set $U$, often called the {\it universe}.
Once we know what $U$ is, we may define the {\it complement}
of a set $A$ to be the set $\bar A$ of all the elements in $U$ except those that are in $A$.
So if $U = \{1,2,3,4,5\}$ in the example above, then $\bar A = \{3,5\}$ and $\bar B = \{2,4\}$.
What about $\bar{A\cup B}$? Well since $A\cup B$ is all of $U$, its complement must be empty, and we can denote
it $\{\}$. This is one valid notation for the {\it empty set}. The other is $\emptyset$.

Now is a good time to define the cardinality $|A|$ of a set $A$.
This is the number of elements in it, so $|A| = |B| = 3$
in our example, and $|A\cup B| = 5$, etc. We have $|\emptyset| = 0$, and it is possible for the cardinality of
a set to be infinity; for example, $|\NN| = \infty$. We also have $|\RR| = \infty$, but this infinity
is, in some sense, larger than $|\NN|$. (More on that later.)

Next, we define the {\it difference} $B\setminus A$ of two sets. This is the set of all elements
in $B$ that are {\it not} in $A$. So, using the complement notation we just learned about,
we can express $B\setminus A = B \cap \bar A$. It is not necessary that $A$ be a subset of $B$. In
the small example above, we have $A\setminus B = \{2,4\}$ and $B\setminus A = \{3,5\}$.

Lastly, we define the {\it symmetric difference} $A \symdiff B$ of two sets $A$ and $B$ to be the set of all
elements that are either in $A$ or in $B$ {\it but not both}. Invoking the above example one last time, we
have $A\symdiff B = \{2,3,4,5\}$. To practise using all the different operations we just learned,
convince yourself that the following are all valid ways to express the symmetric difference:
$$A\symdiff B = (A\cup B)\setminus (A\cap B) = (A\setminus B) \cup (B\setminus A) =
\bar{\bar{A\cup B} \cup (A\cap B)}$$
\datestamp{2}{} More set identities abound. We state the following proposition without proof; you should try going
though this list and convincing yourself that each identity holds, for all sets $A$, $B$, and $C$.
(This is a great way of practising proofs by double inclusion.)

\proclaim Proposition 2. Let $A$, $B$, and $C$ be subsets of a universe $U$. Then
\medskip
\item{i)} $A\cap U = A$ and $A\cup \emptyset = A$;
\smallskip
\item{ii)} $A\cup U = U$ and $A\cap \emptyset = \emptyset$;
\smallskip
\item{iii)} $A\cup A = A$ and $A\cap A = A$;
\smallskip
\item{iv)} $\bar{\bar A} = A$;
\smallskip
\item{v)} $A\cup B = B\cup A$ and $A\cap B = B\cap A$;
\smallskip
\item{vi)} $A\cup (B\cup C) = (A\cup B) \cup C$ and $A\cap (B\cap C) = (A\cap B)\cap C)$;
\smallskip
\item{vii)} $A\cup (B\cap C) = (A\cup B) \cap (A\cup C)$ and $A\cap (B\cup C) = (A\cap B)\cup (A\cap C)$;
\smallskip
\item{viii)} $A\cup (A\cap B) = A$ and $A\cap (A\cup B) = A$; and
\smallskip
\item{ix)} $A\cup \bar A = U$ and $A\cap \bar A = \emptyset$.\slug
\medskip

Some of these laws have names: (v) is called the commutative law, (vi) is called the associative
law, (vii) is the distributive law, (viii) is called the absorption law, and (ix) is the complement law.

\medskip\boldlabel \llap{*}Analogy with addition and multiplication.
Some of these laws bear some resemblance to laws about numbers that you already know. As an exercise,
replace $\cup$ with $+$ (addition), $\cap$ with $\cdot$ (multiplication), $\bar A$ with $-A$ (negation),
$U$ with $1$, and $\emptyset$ with $0$ in all the formulas
above, and now assume that $A$, $B$, and $C$ are arbitrary real numbers. Which identities still hold in
the number setting, and which ones don't? As a more advanced exercise, try replacing $\cup$
with $\symdiff$ in the identities above (some statements will have to be tweaked a bit so that they're
actually true, some won't). Now do the same
replacement as before, except replace $\symdiff$ with $+$. You will find that many more identities
carry over.

\medskip\boldlabel De Morgan's laws. There are two important laws relating complements with union and intersection.
We shall state them as a proposition, this time giving a proof (of one of them).

\parenproclaim Proposition {\advthm} (De Morgan's laws). Let $A$ and $B$ be sets. Then
\medskip
\item{i)} $\bar{A\cup B} = \bar A \cap \bar B$; and
\smallskip
\item{ii)} $\bar{A\cap B} = \bar A \cup \bar B$.
\medskip

\proof
Let $x\in \bar{A\cup B}$. This means that $x$ does not belong to the union of $A$ and $B$,
$x$ cannot be in $A$, nor can it be in $B$. Since $x\notin A$, $x\in \bar A$, and since $x\notin B$,
$x\in \bar B$. Therefore, $x\in \bar A\cap \bar B$. This shows that $\bar{A\cup B}\subseteq \bar A\cap \bar B$.

Now assume that $x\in \bar A\cap \bar B$. So $x\in \bar A$ and $x\in \bar B$, meaning that $x\notin A$
and $x\notin B$. Since $x$ is in neither $A$ nor $B$, it is also not a member of the union $A\cup B$.
We conclude that $x\in \bar{A\cup B}$. We have shown that $\bar A\cap \bar B \subseteq \bar{A\cup B}$,
which fact, combined with the previous paragraph, completes the proof of (i).

The proof of (ii) is similar and left to the reader as an exercise.\slug

Armed with all of these laws, we are able to perform lots of mechanical set manipulations to simplify
expressions. For example, consider the expression
$$\bigl( (A\setminus B)\cup A\bigr) \cap \bar{A\cap \bar B}.$$
Since $A\setminus B = A\cap \bar B$ and invoking the second De Morgan law on the right of the intersection
yields
$$\bigl( (A\cap \bar B)\cup A\bigr) \cap (\bar A \cup B).$$
Now, we can use absorption on the left-hand side to obtain
$$A \cap (\bar A \cup B),$$
and then distributing gives us
$$(A\cap \bar A) \cup (A\cap B) = \emptyset \cup (A\cap B) = A\cap B.$$
We thus see that the nasty expression $\bigl( (A\setminus B)\cup A\bigr) \cap \bar{A\cap \bar B}$ is
simply another way of writing $A\cap B$.

\medskip\boldlabel The Cartesian product and power set. From the real line $\RR$, we can geometrically construct
the Cartesian plane $\RR^2$ by lining up parallel copies of $\RR$, one for each element of the original line
and all parallel to the original line.
Notationally, $\RR^2$ is the set of all ordered pairs $(a,b)$, where $a,b\in \RR$. Generalising this,
for any sets $A$ and $B$ we can define the {\it Cartesian product} $A\times B$
to be the set
$$A \times B = \bigl\{ (a,b) : a\in A, b\in B\bigr\}.$$
We sometimes write $A^2$ for $A\times A$, and more generally $A^n$ for the $n$-fold Cartesian product
of $A$ with itself. (This explains the notation $\RR^2$ for the Cartesian plane, and $\RR^n$ for the
$n$-dimensional vector space over $\RR$.)
Note that $A\times B$ is not equal to $B\times A$ in general.

\proclaim Proposition \advthm. If $A$ and $B$ are finite sets, then $|A\times B| = |A|\cdot |B|$.

\proof The set $A\times B$ consists of all ordered pairs $(a,b)$ where $a\in A$ and $b\in B$. There are
$|A|$ choices for $a$, and for $a$, there are $|B|$ ways to pair it with a $b$ from $B$. So there
are $|A|\cdot |B|$ pairs in total.\slug

Now we define the {\it power set}. For a set $A$, this is the set of all subsets of $A$, and is
commonly denoted by ${\cal P}(A)$ or $2^A$. (We will use the latter notation in these notes.)
In set-builder notation, we have
$$2^A = \{X\subseteq U : X\subseteq A\}.$$
As an example, if $A = \{1,2,3\}$, then
$$2^A = \bigl\{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\bigr\}.$$
Note that even though, say, $1\in A$, we do not have $1\in 2^A$. We do, however, have $\{1\}\in 2^A$.

Another example is $2^\ZZ$, the set of all subsets of integers. If $P$ is the set of primes, then
$P\in 2^\ZZ$. Also in $2^\ZZ$ is the set $S = \{n^2 : n\in \ZZ\}$ of square numbers.

There is a way of encoding subsets with strings of $0$s and $1$s. Suppose we have a set
$$\{-3, 1, 7, 19, 23\}.$$
Fixing this order of the elements, for any arbitrary subset of this set, we can associate to it a binary string.
Consider the subset $\{1, 19, 23\}$. This corresponds to the binary string $01011$: since the first element,
$3$, is not in the subset, we write a $0$. Then $1$ is in the subset, so we write a $1$, and so on.
This is a reversible process. Given a binary string of length $5$, say, $00101$, we can reconstruct the
subset that corresponds to it. The first two $0$s mean that $-3$ and $1$ do not belong to the set,
but $7$ does, $19$ doesn't, and $23$ does. So the subset is $\{7, 23\}$. In this way we see that there
is a one-to-one correspondence between the elements of $2^A$ and binary strings of length $|A|$.
We'll use this fact in the proof of the following proposition.

\proclaim Proposition \advthm. If $A$ is finite, then
$|2^A| = 2^{|A|}$.

\proof We just saw that there is a one-to-one correspondence between elements of $2^A$ and binary strings
of length $|A|$. So it suffices to count the number of binary strings of length $|A|$. Well, each digit
can be either $0$ or $1$, and there are $|A|$ digits, so the number of strings is
$$\underbrace{2\cdot 2\cdot \cdots \cdot 2}_{|A|\ \hbox{\sevenrm times}} = 2^{|A|}.\noskipslug$$

\medskip\boldlabel Counterexamples in proofs.
We finish this subsection with a little example problem. {\sl Is it true that $2^A \cup 2^B = 2^{A\cup B}$
for all sets $A$ and $B$?}

Let's start by trying to prove the statement is true. As usual, we will attempt a double-inclusion proof.
Let $2^A\cup 2^B$. This means $X$ is either a subset of $A$ or it is a subset of $B$. Either way,
$X$ is a subset of $A\cup B$, so $X\in 2^{A\cup B}$. So far so good; we have proved that
$2^A\cup 2^B \subseteq 2^{A\cup B}$.

Now we try the other direction. Let $X\in 2^{A\cup B}$. So $X$ is a subset of $A\cup B$. From here
we want to say that $X$ must be a subset of $A$ or it must be a subset of $B$, but is that necessarily true?
It is possible that $X$ is contained slightly in $A$ and slightly in $B$. So we have failed
to prove that $2^{A\cup B} \subseteq 2^A \cup 2^B$ in general. But just because we have failed to prove
that something is true doesn't mean we have proved it is false!

To actually prove that $2^{A\cup B} \subseteq 2^A \cup 2^B$ doesn't hold in general, we need to find
a {\it counterexample}. That is, we need to construct sets $A$ and $B$ such that the statement is false.
In this case, we can let $A = \{1,2\}$, $B = \{3,4\}$, so that $A\cup B = \{1,2,3,4\}$.
Then the set $\{1,3\}$ is a subset of $A\cup B$ but is not a subset of $A$ and it is not a subset of $B$.
In other words, $\{1,3\}\in 2^{A\cup B}$ but $\{1,3\} \notin 2^A\cup 2^B$, proving that
$2^{A\cup B} \not\subseteq 2^A \cup 2^B$.

\medskip\boldlabel\datestamp{3}{} Russell's paradox. Earlier, we said that the sets we are working with
need to be a subset of a universe $U$, which has already been proven to be a set. We gave lots of
ways to make sets out of new sets, such as the union and intersection operations, etc. Starting with
the empty set $\emptyset$ is a set, it is possible to define the set of natural numbers as follows.
We can define $0 = \emptyset$, $1 = \{0\}$, and $2 = \{0,1\}$, and so on.
Now we take the set of all of these, and call this $\NN$.
(We can also define addition and multiplication on these set-theoretic ``numbers''
so that they behave like addition and multiplication do on $\NN$.) From here we can do more funky stuff
to define $\ZZ$, $\QQ$, and $\RR$, and prove that these are all sets (you can see this in a higher-level
course on set theory, if you're interested). So there isn't much of a problem
with all the sets we have played with so far; they are all subsets of things that are already known to be sets.

Ungodly things can happen if we don't stick by these rules. An example, due to Bertrand Russell, is the ``set''
$$R = \{\hbox{sets}\ X : X \notin X \}.$$
In plain English, $R$ is defined to be the set of all sets that contain themselves.
This is not a subset of any known thing, so by our criterion above we would not consider it a set.
But supposing it is, let us ask ourselves the following question. Does $R$ contain itself?
If it does not, then $R\notin R$, so $R$ would be a set that satisfied the condition of $R$, so $R\in R$.
But on the other hand, if $R\in R$, then $R$ violates the condition defining $R$, so $R\notin R$.
Round and round we go in a circle of contradiction.

Such is the price of meddling with ``sets'' that aren't subsets of known sets.

\advsect Propositional logic

A {\it proposition} is a statement that is true or false. For example ``$8$ is even'' is a statement
we know to be true, and ``$8$ is prime'' is a statement we know to be false. The statement
``$n$ is prime'' is not a proposition because its truth or falsity depends on what $n$ is. (But if
$n$ is previously fixed to stand for some value, then this would be a proposition.) The statement
``$2^{2^240} - 1$ is prime'' is a proposition, because it is either true or false (even though you or I
might not know which one it is).

A {\it propositional variable} or a {\it boolean variable} is a variable which can take either the
value $0$ or $1$, where $0$ means ``false'' and $1$ means ``true.'' Usually we use letters $p$, $q$,
and $r$ to denote propositional variables.

\bye

### 2024-08-27 teaching/math240f24/notes.tex
Notes draft

@@
 \input mathmac
+
+\input epsf
 \widemargins
@@
 \def\symdiff{\mathrel{\triangle}}
+\def\To{\Rightarrow}
+\def\Gets{\Leftarrow}
 
@@
 just covered. The datestamps in the left margin indicate when the notes from each day start.
-Subsections labelled with a $*$ are optional. This document is based on notes by
+Subsections labelled with a $*$ are optional. This document is heavily based on notes by
 Jeremy Macdonald, but any errors are likely my own. Please email me if you find any.
@@
 
-Next, we define the {\it difference} $B\setminus A$ of two sets. This is the set of all elements
+Next, we define the {\it difference} $B\setminus A$ (sometimes $B-A$) of two sets.
+This is the set of all elements
 in $B$ that are {\it not} in $A$. So, using the complement notation we just learned about,
@@
 Lastly, we define the {\it symmetric difference} $A \symdiff B$ of two sets $A$ and $B$ to be the set of all
-elements that are either in $A$ or in $B$ {\it but not both}. Invoking the above example one last time, we
-have $A\symdiff B = \{2,3,4,5\}$. To practise using all the different operations we just learned,
-convince yourself that the following are all valid ways to express the symmetric difference:
+elements that are either in $A$ or in $B$ {\it but not both}.
+Invoking the above example one last time, we
+have $A\symdiff B = \{2,3,4,5\}$.
+We can express it as using unions, intersections,
+and complements as
+\edef\eqsymdiffdef{\the\eqcount}
+$$A \symdiff B = (A\cup B) \cap \bar{A \cap B}.\adveq$$
+To practise using all the different operations we just learned,
+convince yourself that the following are three more valid ways to express the symmetric difference:
 $$A\symdiff B = (A\cup B)\setminus (A\cap B) = (A\setminus B) \cup (B\setminus A) =
@@
 
-\proclaim Proposition 2. Let $A$, $B$, and $C$ be subsets of a universe $U$. Then
+\edef\propsetlaws{\the\thmcount}
+\proclaim Proposition \advthm. Let $A$, $B$, and $C$ be subsets of a universe $U$. Then
 \medskip
@@
 
-Some of these laws have names: (v) is called the commutative law, (vi) is called the associative
-law, (vii) is the distributive law, (viii) is called the absorption law, and (ix) is the complement law.
+These laws have names, some of which we will use more often than others:
+(i) is called the identity law, (ii) the domination law, (iii) the idempotent law,
+(iv) the law of double negation, (v) the commutative law, (vi) the associative law,
+(vii) the distributive law, (viii) the absorption law, and (ix) the complement law.
 
@@
 we know to be true, and ``$8$ is prime'' is a statement we know to be false. The statement
-``$n$ is prime'' is not a proposition because its truth or falsity depends on what $n$ is. (But if
-$n$ is previously fixed to stand for some value, then this would be a proposition.) The statement
-``$2^{2^240} - 1$ is prime'' is a proposition, because it is either true or false (even though you or I
+``$n$ is prime'' is not a proposition because its truth or falsity depends on what $n$ is.
+The statement
+``$2^{2^{240}} - 1$ is prime'' is a proposition, because it is either true or false (even though you or I
 might not know which one it is).
@@
 value $0$ or $1$, where $0$ means ``false'' and $1$ means ``true.'' Usually we use letters $p$, $q$,
-and $r$ to denote propositional variables.
+and $r$ to denote propositional variables. The simplest logical operator is negation, defined
+by the table
+$$
+\vbox{
+\tabskip=1em plus.2em minus .2em
+\halign{
+\hfil$#$\hfil & \hfil$#$\hfil \cr
+p & \neg p \cr
+\noalign{\medskip}
+\noalign{\hrule}
+\noalign{\medskip}
+0 & 1 \cr
+1 & 0 \cr
+}}.
+$$
+This is also called the {\mc NOT} operator, since if $p$ is true, then $\neg p$
+is false, and vice versa.
+Next is {\it conjunction}, which has the table
+$$
+\vbox{
+\tabskip=1em plus.2em minus .2em
+\halign{
+\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
+p & q & p\wedge q \cr
+\noalign{\medskip}
+\noalign{\hrule}
+\noalign{\medskip}
+0 & 0 & 0 \cr
+0 & 1 & 0 \cr
+1 & 0 & 0 \cr
+1 & 1 & 1 \cr
+}}.
+$$
+This is also called the {\mc AND} operator, because $p\wedge q$ is true if and only if $p$ and
+$q$ are both true. The {\mc OR} operator, also called {\it disjunction}, has the table
+$$
+\vbox{
+\tabskip=1em plus.2em minus .2em
+\halign{
+\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
+p & q & p\vee q \cr
+\noalign{\medskip}
+\noalign{\hrule}
+\noalign{\medskip}
+0 & 0 & 0 \cr
+0 & 1 & 1 \cr
+1 & 0 & 1 \cr
+1 & 1 & 1 \cr
+}}.
+$$
+We see that $p\vee q$ is true if $p$ or $q$ is true (or both). The symbol $\vee$ is meant to recall
+the Latin word {\it vel}, meaning ``or.'' (One of the most important early treatises on mathematical
+logic and set theory was {\sl Arithmetices principia, nova methodo exposita}, published in Latin in 1889
+by Giuseppe Peano. It established the now-standard axiomatisation of the natural numbers.)
+
+On the other hand, in English, we often use the word ``or''
+to mean an {\it exclusive} or; that is, either $p$ and $q$ are true but not both. In mathematics, on the
+other hand, ``or'' is usually {\it inclusive}, so both $p$ and $q$ are allowed to hold at the same time.
+It is possible to express the exclusive disjunction (often called {\mc XOR})
+by a table, however. We will use the symbol $\oplus$ for this operator, and its table looks like this:
+$$
+\vbox{
+\tabskip=1em plus.2em minus .2em
+\halign{
+\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
+p & q & p\oplus q \cr
+\noalign{\medskip}
+\noalign{\hrule}
+\noalign{\medskip}
+0 & 0 & 0 \cr
+0 & 1 & 1 \cr
+1 & 0 & 1 \cr
+1 & 1 & 0 \cr
+}}
+$$
+So $p\oplus q$ is true if $p$ is true or $q$ is true, but not both.
+
+A {\it formula} is an expression containing propositional variables, $0$, $1$, logical operators, and
+parentheses. The formula must syntactically make sense; for instance, $0 (\vee \wedge q \neg $ is
+not a formula. Just as in ordinary mathematical notation,
+parentheses are used to clarify which operators should be evaluated first. We will assume that negation
+applies first, but an expression such as $p\vee q \wedge r$ is ambiguous. (In many programming languages,
+conjunction has higher priority than disjunction, but in this class, just add parentheses to clarify.)
+
+Above we have illustrated the basic logical operators
+by writing out their {\it truth tables}. These are tables that give the value of
+a formula for all possible values of its variables. We can write truth tables for more complex formulas as
+well:
+$$
+\vbox{
+\tabskip=1em plus.2em minus .2em
+\halign{
+\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
+p & q & p\wedge q & \neg(p \wedge q) & \neg(p\wedge q) \oplus q \cr
+\noalign{\medskip}
+\noalign{\hrule}
+\noalign{\medskip}
+0 & 0 & 0 & 1 & 1 \cr
+0 & 1 & 0 & 1 & 0 \cr
+1 & 0 & 0 & 1 & 1 \cr
+1 & 1 & 1 & 0 & 1 \cr
+}}
+$$
+Strictly speaking, the third and fourth columns are not necessary, but these intermediary columns help us
+verify the accuracy of the following ones.
+
+Two formulas $f_1$ and $f_2$ are said to be {\it logically equivalent} if they have the same truth table;
+that is, they produce the same output if given the same input. In this case we write $f_1 \equiv f_2$.
+For example, let $f_1 = \neg(p\wedge q) \oplus q$, the formula whose truth table is illustrated above.
+Now let $f_2 = p \vee \neg q$. Its truth table is
+$$
+\vbox{
+\tabskip=1em plus.2em minus .2em
+\halign{
+\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
+p & q & \neg q & p\vee \neg q \cr
+\noalign{\medskip}
+\noalign{\hrule}
+\noalign{\medskip}
+0 & 0 & 1 & 1 \cr
+0 & 1 & 0 & 0 \cr
+1 & 0 & 1 & 1 \cr
+1 & 1 & 0 & 1 \cr
+}},
+$$
+so we conclude that $f_1 \equiv f_2$. As a larger example, suppose we want to find all
+values of $p$, $q$, and $r$ such that
+$$ f = (p\vee q)\wedge (\neg q \vee \neg r)$$
+evaluates to $1$. The truth table
+$$
+\vbox{
+\tabskip=1em plus.2em minus .2em
+\halign{
+\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil &
+\hfil$#$\hfil & \hfil$#$\hfil  \cr
+p & q & r & p\vee q & \neg q \vee \neg r & f \cr
+\noalign{\medskip}
+\noalign{\hrule}
+\noalign{\medskip}
+0 & 0 & 0 & 0 & 1 & 0 \cr
+0 & 0 & 1 & 0 & 1 & 0 \cr
+0 & 1 & 0 & 1 & 1 & 1 \cr
+0 & 1 & 1 & 1 & 0 & 0 \cr
+1 & 0 & 0 & 1 & 1 & 1 \cr
+1 & 0 & 1 & 1 & 1 & 1 \cr
+1 & 1 & 0 & 1 & 1 & 1 \cr
+1 & 1 & 1 & 1 & 0 & 0 \cr
+}}
+$$
+shows that $f$ evaluates to $1$ precisely when
+$$(p,q,r) \in \bigl\{(0,1,0), (1,0,0), (1,0,1), (1,1,0)\bigr\}.$$
+To write out the truth table for a formula with $n$ variables, we need $2^n$ rows, so this
+method is unsuitable for formulas with more than three or four variables.
+
+\medskip\boldlabel Simplifying logical formulas. Just as we have rules for simplifying set expressions,
+there are ways to turn complicated logical formulas into simpler ones that are logically equivalent.
+What might surprise you is that the rules turn out to be exactly the same! To see the basis for this
+correspondence, consider the definition of a union of sets $A$ and $B$. In set-builder notation,
+this is
+$$A \cup B = \{ x\in U : x\in A\ \hbox{or}\ x\in B\}.$$
+The ``or'' in the definition suggests that $\cup$ is intimately related to the $\vee$ operation in
+propositional logic. Repeating this process, we have the ``dictionary''
+$$
+\vbox{
+\tabskip=3em plus.005em minus .005em
+\halign{
+\hfil{\ninepoint #}\hfil & \hfil{\ninepoint #}\hfil \cr
+Set theory & Propositional logic\cr
+\noalign{\medskip}
+\noalign{\hrule}
+\noalign{\medskip}
+sets $A$, $B$ & variables $p$, $q$ \cr
+unions $A\cup B$ & disjunctions $p\vee q$ \cr
+intersections $A\cap B$ & conjunctions $p\wedge q$ \cr
+complements $\bar A$ & negations $\neg p$ \cr
+symmetric differences $A\symdiff B$ & exclusive disjunctions $p\oplus q$ \cr
+the empty set $\emptyset$ & $0$ \cr
+the universe $U$ & $1$ \cr
+}}
+$$
+Exploiting this connection, we have the following analogue of Proposition~{\propsetlaws}.
+
+\proclaim Proposition 2. Let $p$, $q$, and $r$ be propositional variables. Then
+\medskip
+\item{i)} $p\wedge 1 \equiv p$ and $p\vee 0 \equiv p$;
+\smallskip
+\item{ii)} $p \vee 1 \equiv 1$ and $p \wedge 0 \equiv 0$;
+\smallskip
+\item{iii)} $p \vee p \equiv p$ and $p\wedge p \equiv p$;
+\smallskip
+\item{iv)} $\neg\neg p \equiv p$;
+\smallskip
+\item{v)} $p \vee q \equiv q \vee p$ and $p\wedge q \equiv q\wedge p $;
+\smallskip
+\item{vi)} $p\vee (q\vee r) \equiv (p\vee q) \vee r$ and $p\wedge (q\wedge r) \equiv (p\wedge q)\wedge r)$;
+\smallskip
+\item{vii)} $p\vee (q\vee r) \equiv (p\vee q) \wedge (p\vee r)$ and
+$p\wedge (q\vee r) \equiv (p\wedge q)\vee (p\wedge r)$;
+\smallskip
+\item{viii)} $p\vee (p\wedge q) \equiv p$ and $p\wedge (p\vee q) \equiv p$; and
+\smallskip
+\item{ix)} $p\cup \bar p \equiv 1$ and $p\wedge \bar p \equiv 0$.\slug
+\medskip
+
+Since they are essentially the same as their set equivalents, the names of these laws are the same as in the
+realm of sets.
+We also have the propositional equivalent of De Morgan's law, which states that
+\medskip\begingroup\sl
+\item{x)} $\neg(p\vee q) \equiv \neg p \wedge \neg q$ and $\neg(p\wedge q) \equiv \neg p \vee \neg q$.\endgroup%sl
+\medskip
+
+Using these rules, we can now show that $\neg(p\wedge q) \oplus q \equiv p\vee\neg q$, which we
+already saw earlier from their truth tables. First we observe that
+\edef\eqxordef{\the\eqcount}
+$$p\oplus q \equiv (p\vee q)\wedge \neg(p\wedge q),\adveq$$
+which is analogous to the identity~\refeq{\eqsymdiffdef} for symmetric differences. So
+$$\eqalign{
+\neg(p\wedge q)\oplus q&\equiv\bigl(\neg(p\wedge q)\vee q\bigr)\wedge\neg\bigl(\neg(p\wedge q)\wedge q\bigr)\cr
+&\equiv (\neg p \vee\neg q \vee q) \wedge \bigl((p\wedge q) \vee \neg q\bigr) \cr
+&\equiv (\neg p \vee 1) \wedge \bigl((p\vee \neg q) \wedge (q\vee \neg q)\bigr) \cr
+&\equiv 1 \wedge \bigl((p\vee \neg q) \wedge 1\bigr) \cr
+&\equiv p\vee \neg q, \cr
+}$$
+where in the first line we use~\refeq{\eqxordef}, in the second line we use De Morgan's law twice, in
+the third line we use the complement and distributive laws, the fourth line we use the domination and
+complement laws, and in the last line we use the identity law (twice). (It is not super important
+to remember the names of all these laws, as long as you remember their statements, but you may actually find
+it is easier to remember the names along with the statements, just as it might be easier to remember faces of
+people you've met if you also know their names.)
+
+\medskip\boldlabel\datestamp{4}{} Conditional and biconditional. We now examine the {\it conditional}
+logical relation {\mc IF} $p$ {\mc THEN} $q$. It is denoted by $p\To q$ and its truth table is given by
+$$
+\vbox{
+\tabskip=1em plus.2em minus .2em
+\halign{
+\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
+p & q & p\To q \cr
+\noalign{\medskip}
+\noalign{\hrule}
+\noalign{\medskip}
+0 & 0 & 1 \cr
+0 & 1 & 1 \cr
+1 & 0 & 0 \cr
+1 & 1 & 1 \cr
+}}.
+$$
+Within the conditional statement, $p$ is called the {\it antecedent}; this is the assumption. The
+statement that is asserted, conditional that the antecendent holds, is called the {\it consequent} $q$.
+You can quickly check that the relation $p\To q$ is equivalent to $\neg p \vee q$. This fact is useful
+when performing mechanical simplifications.
+In English, the statement ``if $p$ then $q$'' asserts a causal relation between $p$ and $q$.
+Take a moment and reconcile this idea with the truth table above.
+It is normal to get a little tripped up if it's your first time seeing this table.
+In the first row, $q$ doesn't even happen, so it might feel weird to say that $p$
+has ``caused'' $q$ to happen in this case, and
+in the second row, $p$ is not true
+and $q$ is true, so it might seem odd that we have set $p\To q$ to true, because there seems to be no relation
+between $p$ and $q$.
+But it should make sense if you think
+of $p$ as a precondition to a promise $q$, and then considering whether the promise is broken. As an example,
+suppose your friend says: ``If it snows tomorrow I'll work in Trottier with you.'' If it doesn't snow
+and she doesn't pull up, she hasn't technically broken her promise. If it doesn't
+snow and she shows up, then she still hasn't broken her promise. The only way she can break her promise is if
+it snows and she doesn't come to Trottier; this situation corresponds to the only $0$-row in the truth table.
+
+The conditional is a very important logical operator to understand, because most theorem statements assume
+some hypothesis and claim some conclusion. You will be asked to prove statements of this form, so it is
+important to understand the logical nature of the statements to begin with.
+
+The last relation is called the {\it biconditional}, and it asserts that variables $p$ and $q$ are logically
+equivalent. That is, $p$ happens if and only if $q$ happens. It's truth table
+$$
+\vbox{
+\tabskip=1em plus.2em minus .2em
+\halign{
+\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
+p & q & p\Leftrightarrow q \cr
+\noalign{\medskip}
+\noalign{\hrule}
+\noalign{\medskip}
+0 & 0 & 1 \cr
+0 & 1 & 0 \cr
+1 & 0 & 0 \cr
+1 & 1 & 1 \cr
+}}
+$$
+is pretty self-explanatory; in it, $p\Leftrightarrow q$ is true whenever $p$ and $q$ have the same truth value.
+The name ``biconditional'' is suggested by (part of) the following proposition.
+
+\proclaim Proposition {\advthm}. Let $p$ and $q$ be propositional variables. Then
+$$\eqalign{
+p \Leftrightarrow q &\equiv (p\To q) \wedge (q\To p) \cr
+&\equiv \bigl( (\neg p)\vee q\bigr) \wedge \bigl( (\neg q)\vee p\bigr) \cr
+&\equiv (p\wedge q) \vee \bigl( (\neg p) \wedge (\neg q)\bigr).\cr
+}$$
+
+\proof We leave the first equivalence to the reader (writing out the truth table is one way of proving it).
+The second equivalence follows from our earlier observation that $p\To q \equiv \neg p \vee q$, and
+the third equivalence follows from the distributive, complement, and identity laws.\slug
+
+A formula $f$ is called a
+\medskip
+\item{i)} {\it tautology} if $f\equiv 1$, i.e., $f$ always evaluates to true;
+\smallskip
+\item{ii)} {\it contradiction} if $f\equiv 0$, i.e., $f$ always evaluates to false;
+\smallskip
+\item{iii)} {\it contingency} if $f$ can evaluate to both $1$ and $0$, depending on the
+values of its variables;
+\smallskip
+\item{iv)} {\it satisfiable} if $f$ evaluates to $1$ for at least one input; and
+\smallskip
+\item{v)} {\it falsifiable} if $f$ evaluates to $0$ for at least one input.
+\medskip
+An example of a tautology is $p\vee\neg p$ and an example of a contradiction is $p\wedge \neg p$.
+This follows from the complement laws. To say that something is satisfiable is precisely to
+say that it is not a contradiction, and to say that something is falsifiable is equivalent to saying
+that it is not a tautology. Contingencies are those formulas that are both satisfiable
+and falsifiable (in other words, formulas that are neither tautologies nor contradictions).
+
+Suppose we are asked which of the above definitions the formula
+$$f \equiv \bigl( p\wedge (p\To q)\bigr) \To q$$
+satisfies. This formula only has two variables, so it is easy enough to use a truth table
+for this purpose, but we will take the opportunity to practise simplifying the expression
+symbolically. First of all, let's change all conditionals of the form $r \To s$,
+to disjunctions of the form $\neg r \vee s$. This gives us
+$$f\equiv \neg\bigl( p\wedge (\neg p \vee q)\bigr) \vee q.$$
+Now we use the distributive law to distribute the conjunction over the innermost disjunction,
+obtaining
+$$f\equiv \neg\bigl((p\wedge \neg p) \vee (p\wedge q)\bigr) \vee q;$$
+by the complement and identity laws in that order, this simplifies to
+$$f\equiv \neg(p\wedge q) \vee q.$$
+Now De Morgan's law and associativity give
+$$f\equiv (\neg p \vee \neg q) \vee q \equiv \neg p \vee (\neg q \vee q),$$
+and thus
+$$f\equiv \neg p \vee 1 \equiv 1,$$
+by the complement and domination laws in that order. We conclude that $f$ is a tautology,
+which also means that it is satisfiable.
+
+The fact that $\bigl( p\wedge (p\To q)\bigr) \To q$ is a tautology symbolically justifies
+the argument that of $p$ is true and $p\To q$ is true, then we should be able to conclude $q$.
+This form of argument is called {\it modus ponens}, and it dates back to ancient times. You probably
+use {\it modus ponens} all the time in everyday life without knowing it, and we will
+certainly use it in this class a lot.
+
+\medskip\boldlabel Encoding problems in propositional logic.
+Many algorithmic and logical problems can be encoded in propositional logic (and then later
+solved by a computer program). For example, suppose we want to play $4\times 4$ Sudoku. In this
+game, we have a $4\times 4$ grid and we want to fill it with the numbers $1$ through $4$ such that
+\medskip
+\item{i)} every row contains $1$ through $4$;
+\smallskip
+\item{ii)} every column contains $1$ through $4$; and
+\smallskip
+\item{iii)} the four subsquares each contain $1$ through $4$.
+\medskip
+In a given instance of the game, some cells are already filled in. The puzzle is: {\sl Is there a
+solution and if so, what is it?}
+\midinsert
+$$\epsfbox{sudoku.ps}$$
+\vskip5pt
+\caption{An example $4\times 4$ Sudoku game.}
+\endinsert
+\goodbreak
+To represent a Sudoku game in propositional logic, we can define boolean variables
+$p_{i,j,k}$, where $i$, $j$, and $k$ range over $\{1,2,3,4\}$. (So there are $4^3$ variables
+in total.) We shall set
+$$ p_{i,j,k} = \cases{1, & if the number $k$ is in row $i$ and column $j$;\cr 0, & otherwise.}$$
+We'll  number the rows increasing from the top and the columns increasing
+from left to right.
+For example, in Fig.~1 there is a $2$ in row $1$ and column $3$, so $p_{1,3,2} = 1$.
+
+Now we set to work encoding the conditions of a Sudoku grid in propositional logic:
+\medskip
+\item{i)} To stipulate that every row contain $1$ through $4$, we first define auxiliary variables
+$$r_{i,k} = p_{i,1,k} \vee p_{i,2,k} \vee p_{i,3,k} \vee p_{i,3,k},$$
+for $i,k\in \{1,2,3,4\}$.
+With these helper variables, we now see that
+$$ r_{1,1} \wedge r_{1,2} \wedge r_{1,3} \wedge r_{1,4} $$
+encodes the requirement that row $1$ contains one of each number.
+We do the same for rows $2$, $3$, and $4$ as well, and then combine with {\mc AND}.
+\smallskip
+\item{ii)} We do a similar thing as in part (i) for each of the four columns.
+\smallskip
+\item{iii)} Ditto for subsquares.
+\smallskip
+\item{iv)} We need to set the initial values of the grid. For the grid in Fig.~1, we have
+the formula
+$$p_{1,3,2} \wedge p_{3,1,1} \wedge p_{3,2,3}.$$
+\smallskip
+\item{v)} Lastly, we need to make sure that there is not more than one number per cell. To do
+this, for each cell $(i,j)\in \{1,2,3,4\}^2$ we write
+$$ p_{i,j,1} \To (\neg p_{i,j,2} \wedge \neg p_{i,j,3} \wedge \neg p_{i,j,4}),$$
+and so on (four conditionals in total). Of course we'll need to {\mc AND} all these together.
+\medskip
+Now we combine the formulas from each of these five steps into one long formula $f$ such that
+$f$ is satisfiable if and only if the grid has a solution, and the values for $p_{i,j,k}$ give
+a solution.
+
+Defining all of these variables was a rather arduous and cumbersome process, and not entirely worth
+it for a $4\times 4$ game of Sudoku (which can just be solved by eyeballing the grid).
+But one could imagine writing a general computer program to encode larger and larger grids.
+In fact, there are lots of problems that can be {\it reduced} to the problem of
+determining if a boolean formula is satisfiable. This means that if we have a program
+capable of taking a formula $f$ as input and spitting out whether or not it is satisfiable (in a
+reasonable amount of time), then there are lots of real-world problems that this program
+could be applied to.
+
+This problem is called the {\it boolean satisfiability problem}, often abbreviated {\mc SAT}. One
+way of solving it for any given $f$ is to just compute its truth table. We already know the
+downside of this approach: if $f$ has $n$ variables, then its truth table will have $2^n$ rows.
+Given a few minutes, you are certainly capable of writing down a formula $f$ that has $300$
+variables, call them $p_1, \ldots, p_{300}$. The truth table of $f$ will have $2^{300}$ rows,
+which is more than the number of atoms in the observable universe. You can learn a lot more
+about {\mc SAT} in a higher-level class on computational complexity (e.g., {\mc COMP 360/362}).
+
+\advsect Predicate logic
+
+\datestamp{5}{}
+Propositional logic allows us to work with simple declarations, but this isn't powerful enough
+to express some deeper mathematical concepts. For this purpose, we now introduce the notion of
+a {\it predicate}. This is a statement involving some number of variables, each of which may take
+values coming from a universe $U$, such that the statement evaluates to either true or false {\sl once
+all variables are assigned values}. The statement $P(n)$ given by ``$n$ is prime''
+is an example of this, where $n$
+can take any value in the universe $\ZZ$. An example with two variables is the
+predicate $L(x,y)$ defined by ``$x$ is less than $y$.''
+(A more commonly-used notation for this predicate is ``$x<y$.'')
+
+Predicates contain variables, but at the moment we don't have any way of introducing new variables into
+a statement. This is done using two different {\it quantifiers}.
+The first is the
+{\it universal quantifier}, denoted $\forall$ and meaning ``for all.'' The statement $\forall n:P(n)$
+is true if and only if $P(n)$ is true for every possible value that $n$ can take. The second quantifier
+is the {\it existential quantifier}, written ``$\exists$'' and with the meaning ``there exists.''
+The statement $\exists n:P(n)$ is true if and only if there is (at least) one value that $n$ can take
+such that $P(n)$ is true. The colon doesn't really have any mathematical meaning in these formulas;
+they just visually set the quantifiers apart from the predicates that follow.
+
+For instance, taking $P(n)$ to be the statement ``$n$ is prime,'' where the universe $U$ is $\NN$,
+the statement $\exists n P(n)$ is true and the statement $\forall n:P(n)$ is false.
+What about the statement $\forall x\,\exists y:y< x$? Well, if the universe $U$ is taken to be $\NN$,
+then the statement is false, because setting $x$ equal to $0$, there is no element $y$ of $\NN$ such that
+$y<0$. But if $U = \ZZ$, then the statement is true, since for every integer $x$, we can put $y = x-1$,
+so that $y<x$.
+
+Now we practise translating converting mathematical statements written in English into formulas
+in predicate logic. Suppose we want to write: ``Every integer is even or odd.'' The universe here is
+is the set $\ZZ$ of integers. The word ``every'' has the same meaning as ``for all,'' so right
+off the bat, we can reexpress the statement as: ``For all $n\in \ZZ$, $n$ is even or $n$ is odd.''
+In symbols, this is
+$$ \forall n\,(n\ \hbox{even} \vee n\ \hbox{odd}).$$
+Lastly, we need to figure out how to express the property of being even or being odd.
+An integer $n$ is even if an only if it is a multiple of two; that is, if there is some integer $k$
+such that $2k = n$. Likewise, an integer is odd if and only if it is one more than a multiple of two.
+The corresponding formula is $\exists k : 2k+1 = n$.
+So our statement can be expressed
+$$\forall n\, \bigl((\exists k : n = 2k)\vee (\exists k : n = 2k+1)\bigr).$$
+The variable $k$ appears twice in this formula, but its first instance is independent of its second instance,
+because of the parentheses. (Readers who write computer programs will be familiar with the concept of
+a variable ``going out of scope.'') So there is nothing wrong with this formula, but to be extra
+clear that the first $k$ is different from the second $k$, why don't we replace it with a different letter?
+Thus we arrive at
+$$\forall n\, \bigl((\exists k : n = 2k)\vee (\exists l : n = 2l+1)\bigr),$$
+a formula in predicate logic that means ``every integer is even or odd.''
+
+\medskip\boldlabel Restrictions using quantifiers. It is not true that every real number has a multiplicative
+inverse, since one cannot divide by zero. However, the statement ``every nonzero real number has a multiplicative
+inverse'' is true. How should we write this as a formula in predicate logic?
+One way is to use the conditional: over the universe $U = \RR$, we could write
+$$\forall x : (x\ne 0 \To \exists y : xy = 1).$$
+Another is to use subscripts: in the same universe, we write
+$$\forall x_{x\ne 0}\, \exists y : xy = 1.$$
+Using subscripts is slightly informal, since we didn't formally define above what a subscript is supposed
+to mean, but it is something that you might encounter. The last way is to simply restrict the universe itself:
+in the universe $U = \RR\setminus\{0\}$, the statement
+$$\forall x\, \exists y : xy = 1$$
+is true.
+
+\medskip\boldlabel Multiple quantifiers. Withing a formula, quantifiers cannot be interchanged
+willy-nilly. The order of $\forall$ and $\exists$ matters! They are read from left to right.
+Consider the following examples, over the universe $U = \RR$.
+The statement
+$$ \forall x\, \exists y : x+y = 0$$
+is true, since for each given $x$ we can take $y$ to be $-x$. On the other hand,
+$$ \exists y\, \forall x : x+y = 0$$
+is false, since it would mean that there is some integer that adds up to zero with {\it any} integer.
+Of course, sometimes switching the order of quantifiers, doesn't change the truth value of a statement.
+Both
+$$\exists y\, \forall x: xy = 0$$
+and
+$$\forall x\, \exists y: xy = 0$$
+are true, since in the first case, we can take $y = 0$, and in the second case, we can set $y$ to $0$ no
+matter what $x$ is given.
+
+So we know that the order of $\forall$ and $\exists$ matters in general, but repeated instances
+of the {\it same} quantifier {\it can} be interchanged. For instance,
+$$\forall x\, \forall y : x^2 + y^4 \ge 0$$
+is the same as
+$$\forall y\, \forall x : x^2 + y^4 \ge 0,$$
+and we can even write $\forall x,y : x^2 + y^4 \ge 0$, to introduce both variables simultaneously.
+
+\medskip\boldlabel Negating quantifiers. The universal quantifier is sort of like a big chain of conjunctions
+that goes over the whole of the universe. For example, in the universe $\NN$, the
+statement $\forall n : P(n)$ is equivalent to
+$$P(1) \wedge P(2) \wedge P(3) \wedge \cdots,$$
+if this were a valid propositional formula (it isn't because we don't allow propositional formulas to be
+infinite). Likewise, the existential quantifier $\exists n : P(n)$ is equivalent to
+$$P(1)\vee P(2) \vee P(3) \vee\cdots.$$
+We know, by De Morgan's laws, that negating a big series of conjunctions requires us to
+flip all the {\mc AND}s to {\mc OR}s. So, once again abusing notation somewhat, we expect
+$$\neg\bigl(P(1) \wedge P(2) \wedge P(3) \wedge \cdots\bigr)
+\equiv \neg P(1) \vee \neg P(2)\vee \neg P(3)\vee\cdots.$$
+Thus we conclude that
+$$\neg\bigl(\forall n : P(n)\bigr) \equiv \exists n : \neg P(n).$$
+You can play the same game with the other De Morgan's law to show that
+$$\neg \bigl(\exists n : P(n)\bigr) \equiv \forall n : \neg P(n).$$
+Going back to our example of $P(n)$ denoting ``$n$ is prime,'' the statement $\neg \bigl( \forall n : P(n)\bigr)$
+is true, since not all integers $n$ are prime, and we have just shown that this is equivalent to the
+statement $\exists n : \neg P(n)$; that is, there exists $n$ such that $n$ is not prime.
+
+We end this section with a longer example. Let's express the statement, ``There is a nonzero
+real number such that every real number is not its inverse or is negative.'' In the universe $\RR$,
+the formula corresponding to this statement is
+$$ \exists x : ( x\ne 0 \wedge (\forall y : xy\ne 1 \vee y< 0)\bigr) .$$
+(Work it out yourself!) Is this statement true or false? It turns out that it is true. You might be able
+to stare at the formula long enough to convince yourself of this fact, but another way to see
+that it's true is to note that its negation is false. Let's do this now (it's a good excuse to
+practise negating a formula). We have
+$$\eqalign{
+\neg\Bigl(\exists x :( x\ne 0 \wedge (\forall y : x&y\ne 1 \vee y< 0)\bigr)\Bigr) \cr
+&\equiv \forall x : \neg ( x\ne 0 \wedge (\forall y : xy\ne 1 \vee y< 0)\bigr) \cr
+&\equiv \forall x : \bigr(x= 0 \vee \neg(\forall y : xy\ne 1 \vee y< 0)\bigr) \cr
+&\equiv \forall x : \bigr(x= 0 \vee \exists y : \neg (xy\ne 1 \vee y< 0)\bigr) \cr
+&\equiv \forall x : \bigr(x= 0 \vee \exists y : (xy = 1 \wedge y\ge 0)\bigr).\cr
+}$$
+This negated statement is false, since if $x = -2$, then $x=0$ doesn't hold, so the left-hand
+side of the {\mc OR} isn't true, and there is no $y$ such that $-2 y = 1$ and $y\ge 0$ are both
+true, since the only $y$ satisfying $-2y = 1$ is $-1/2$.
+
+Negating a formula in predicate logic is entirely mechanical.
+The $\neg$ symbol moves from left to right like a bulldozer that flips quantifiers and negates
+predicates it finds along the way, until eventually its job is done and it disappears.
+More broadly, we write mathematical statements in formal logic to make things more precise and mechanical.
+This can be useful to humans, since English is often ambiguous whereas the notation we just established
+is not. It can be useful to machines as well,
+since, as we just saw, manipulating a formula is something that can very easily be automated.
+
+\advsect Proofs
+
+\datestamp{6}{}
+We're getting to the fun part of the course now.
+Further back in these notes, we already proved a few statements about sets. This was just a taste of
+what's to come, as the main focus of this course is to teach you how to prove mathematical statements.
+These will all be statements that can ultimately be stated in predicate logic, so using the tools
+from the previous section, you can boil them down to their logical skeleton. In this section, we will
+formally define what it means to prove a statement in predicate logic.
+
+To prove a statement, we always process the quantifiers from left to right. Whenever we encounter something
+of the form $\exists x: P(x)$, we are allowed to choose the value for $x$ (from the given universe), and we
+just need to show that $P(x)$ holds for that value of $x$. Here's an example.
+
+\proclaim Proposition \advthm. There exists an integer $m>0$ and an integer $n<0$ such that $m^2 + n^2 = 25$.
+
+\noindent We've written the statement in words, and we will continue to do so for all the propositions
+in these notes because we are humans and not cyborgs,
+but notice that the underlying predicate logical formula here is
+$$\exists m\,\exists n : m>0 \wedge n< 0 \wedge m^2 + n^2 = 25,$$
+with $U = \ZZ$. For the remainder of this section, we'll continue to write out the formulaic equivalents
+of propositions, to practise converting between the two worlds.
+
+\proof After a moment's contemplation, we notice that $9 + 16 = 25$. So we need a positive integer that
+squares to to $9$ and a negative number that squares to $16$. Hence we may pick $m = 3$ and $n=-4$,
+and $m^2 + n^2 = 25$.\slug
+
+On the other hand, to prove a statement of the form $\forall x : P(x)$, we are not allowed to choose the
+value of $x$. Instead, we imagine it is given to us, and we still have to prove $P(x)$, no matter what
+the $x$ might be. Here's what we mean.
+
+\proclaim Proposition \advthm. For all $x\in \QQ$, there exists $y\in \ZZ$ such that
+$xy \in \ZZ$.
+
+\noindent The formula this time is
+$$\forall x\, \exists y : y\in \ZZ \wedge xy\in \ZZ,$$
+over $U = \QQ$.
+
+\proof Let $x \in \QQ$. Then $x$ can be expressed as the ratio of two integers; write $x = m/n$ with
+$m,n\in \ZZ$. Then, setting $y=n\in \ZZ$, we have $xy = (m/n)\cdot n = m\in \ZZ$.\slug
+
+Notice how we introduced the variable $x$ in the above proof.
+Because we're trying to prove a ``for all'' statement, we use
+the word ``let,'' to indicate that $x$ is given to us by some higher power. In fact, we should sort
+of view this higher power as possibly being malicious. A very useful way to think about writing proof is
+to imagine a game in which you are trying to prove a statement in predicate logic, and a supernatural adversary
+is attempting to thwart you. Let's say this predicate has four variables, so the statement is
+$$\exists x \, \forall y \, \forall z\, \exists w : P(x,y,z,w).$$
+The variables in the statement are introduced from left to right. Each time you see a $\exists$
+symbol, it's your turn. In the example above, you get to set the variable $x$ to any element of the
+given universe (keeping in mind that your eventual goal is to prove $P(x,y,z,w)$). Each time there
+is a $\forall$ symbol, it's the adversary's turn, and you should be prepared for whatever he throws at you.
+So in our example, the adversary may set $y$ and $z$ to anything in the universe, and he knows you picked
+$x$. Lastly, you get to pick $w$. If $P(x,y,z,w)$ is true, you win, and if not, the adversary wins.
+Writing a proof is equivalent to describing a winning strategy for the player against the adversary.
+
+Many mathematical statements introduce some hypotheses, then assert some conclusion. Thus they are some
+kind of statement of the form $p\To q$. To prove this kind of statement,
+we assume that $p$ holds, then prove that $q$ is true. This is because the only way a statement for $p\To q$
+to be false is if $p$ is true and $q$ is false, so we're showing this cannot happen. Take a look at
+the following example.
+
+\edef\propoddimpliessquareodd{\the\thmcount}
+\proclaim Proposition \advthm. If $n$ is an odd integer, then $n^2$ is also odd.
+
+\noindent
+The underlying formula is $\forall n\bigl( (\exists k : n= 2k+1) \To (\exists l : n^2 = 2l+1)\bigr)$,
+in the universe $U=\ZZ$.
+
+\proof Let $n$ be an arbitrary odd integer, so that there exists $k$ such that $n = 2k+1$. Then
+$$n^2 = (2k+1)^2 = 4k^2 + 4k + 1 = 2(2k^2 + 2k)+1,$$
+so setting $l = 2k^2 + 2k$, which is an integer, we have $n^2 = 2l+1$. Hence $n^2$ is odd.\slug
+
+To disprove a statement, we simply prove that its negation is true. For example, suppose we want
+to disprove the statement $\exists x\, \forall y : x+y = 0$ (this was a statement we encountered
+in the previous section). Its negation is
+$$\forall x\, \exists y : x+y \ne 0,$$
+and to prove it we let $x$ be given, set $y = -x + 1$, and observe that $x+y = x + (-x+1) = 1$, which is
+not equal to $0$.
+
+Now we define two words that sound similar but are very different, logically speaking.
+The {\it converse}
+of a conditional statement $p\To q$ is the statement $q\To p$. The
+statement $p\To q$ and its converse both hold if and only if
+the biconditional statement $p\Leftrightarrow q$ holds.
+On the other hand, the {\it contrapositive} to $p\To q$ is the statement
+$\neg q \To \neg p$. What's the deal with this silly-looking conditional? Well, it turns out to actually
+be equivalent to $p\To q$. Check it out:
+$$\eqalign{
+p\To q &\equiv \neg p \vee q \cr
+&\equiv q \vee \neg p \cr
+&\equiv \neg (\neg q) \vee \neg p \cr
+&\equiv \neg q \To \neg p \cr
+}$$
+So to prove a statement of the form $p\To q$, it is sufficient to prove $\neg q\To \neg p$. Sometimes,
+this can make the proof a lot easier. Here's an example. (It is the converse of
+Proposition~{\propoddimpliessquareodd}.)
+
+\edef\propsquareoddimpliesodd{\the\thmcount}
+\proclaim Proposition \advthm. If $n$ is an integer such that $n^2$ is odd, then $n$ is odd.
+
+\noindent{\it Proof attempt.}\enspace Assume $n^2$ is odd, so that $n^2 = 2k+1$ for some integer $k$.
+Then $n = \sqrt{2k+1}$.
+\medskip
+
+Where do we go from here? It's not even clear that $\sqrt{2k+1}$ is an integer, let alone an odd one.
+The contrapositive comes to the rescue.
+
+\medskip\noindent{\it Proof of Proposition~{\propsquareoddimpliesodd}.}\enspace
+We proceed by contraposition.
+Suppose that $n$ is not odd; that is, $n$ is even. So there
+exists an integer $k$ such that $n=2k$. Then $n^2 = (2k)^2 = 4k^2$. Setting $l = 2k^2$, we have
+$n^2 = 2l$, so $n^2$ is even. Hence $n^2$ is not odd.\slug
+
+\medskip\boldlabel Proofs by contradiction. We now discuss a powerful method of proof, which dates
+back to at least the
+ancient Greeks and referred to for much of Western history by its Latin name,
+{\it reductio ad absurdum}. It proceeds to prove
+a statement $p$ by assuming its negation $\neg p$
+holds, then deriving a contradiction, i.e., a statement whose truth value is $0$.
+By doing this, one will have proved that $\neg p\To 0$, which in turn shows that $p$ is true, since
+$$\neg p \To 0 \equiv \neg(\neg p) \vee 0 \equiv p\vee 0 \vee p.$$
+Here is an example of a proof by contradiction.
+
+\proclaim Proposition \advthm. There is no least positive rational number.
+
+\noindent In $U = \QQ$, one way to formulate this is
+$$\neg \bigl(\exists x: x > 0 \wedge (\forall y : y > 0 \To x \le y)\bigr).$$
+
+\proof Suppose, towards a contradiction, that there exists some rational $x>0$ such that for all $y\in \QQ$
+with $y > 0$, $x\le y$. Then, we can apply this property with $y = x/2$ to see that $x \le x/2$.
+Since $x$ is positive, we can divide this inequality by $x$ on both sides to obtain $1\le 1/2$. But this
+is absurd, since $1/2 < 1$. This contradiction completes the proof.\slug
+
+Here is another classical example of a proof by contradiction. It was known to the Pythagoreans.
+
+\edef\propsqrttwo{\the\thmcount}
+\proclaim Proposition \advthm. The number $\sqrt 2$ is irrational.
+
+\proof Suppose, towards a contradiction, that $\sqrt 2$ is rational. Then there are integers $p$ and $q$
+such that $\sqrt 2 = p/q$, and furthermore, we can assume that $p$ and $q$ do not have any common factors,
+since if they did, we could divide them out and the ratio would remain the same.
+Squaring both sides of the equation, we have $2 = p^2/q^2$, or in other words, $2q^2 = p^2$. This implies
+that $p^2$ is even, so by (the contrapositive of) Proposition~{\propsquareoddimpliesodd}, $p$ is even as well.
+That means we can write $p=2r$ for some integer $r$, and substitute this new information
+into the above equation to obtain $2q^2 = (2r)^2 = 4r^2$. Dividing out by $2$ yields $q^2 = 4r^2$,
+so $q^2$ is even, and so is $q$.
+
+We have deduced that $p$ and $q$ are both even. On the other hand, we assumed they had no common factors.
+This is a contradiction.\slug
+
+Earlier we described the composition of a proof as playing a game with a supernatural adversary.
+To prove something by contradiction, then is is akin to starting the game
+by letting adversary believe he has already won,
+and then working from that assumption to derive an impossibility. Quite the devious strategem.
+
+\bigskip
+\begingroup\obeylines\eightssi
+\hfill {\eightss Reductio ad absurdum}, which Euclid loved so much,
+\hfill is one of a mathematicians finest weapons.
+\hfill It is a far finer gambit than any chess gambit:
+\hfill a chess player may offer the sacrifice of a pawn
+\hfill or even a piece,
+\hfill but a mathematician offers the {\eightss game}.
+\eightss
+\smallskip
+\hfill --- G.~H.~HARDY, {\eightssi A Mathematician's Apology} (1940)
+\endgroup%\obeylines
+\bigskip\goodbreak
+
+\medskip\boldlabel Case analysis.
+For all $n\ge 2$ and all propositions $p_1, \ldots, p_n, q$, we have the equivalence
+$$(p_1\vee p_2 \vee \cdots \vee p_n) \To q \equiv (p_1\To q) \wedge (p_2\To q) \wedge \cdots \wedge
+(p_n\To q).$$
+(As an exercise, prove this when $n=2$. Later on, when we learn about mathematical induction,
+you'll be able to prove it in general.)
+
+Why is this useful? Well, if we have two or more propositions $p_1, p_2, \ldots, p_n$
+such that at least one of them must hold, that is,
+$$p_1\vee p_2 \vee \cdots \vee p_n \equiv 1,$$
+then if we are able to show that
+$$p_1 \To q,\quad p_2\To q,\quad \ldots,\quad\hbox{and}\quad p_n\To q$$
+all hold, then we will have shown
+$$1\To q \equiv \neg 1 \vee q \equiv 0 \vee q \equiv q.$$
+Let's see an example of this in action.
+
+\proclaim Proposition \advthm. For all integers $n\ge 0$, $1 + (-1)^n (2n-1)$ is a multiple
+of $4$.
+
+\proof Let an integer $n\ge 0$ be given. We know that $n$ is even or $n$ is odd.
+
+If $n$ is even, then $n = 2k$ for some integer $k$, and
+$$1 + (-1)^{2k} \bigl( 2(2k)-1\bigr) = 1 + 1^k (4k-1) = 4k,$$
+which is a multiple of $4$.
+
+If $n$ is odd, then $n=2k+1$ for some integer $k$, and
+$$1 + (-1)^{2k+1}\bigl(2(2k+1) - 1\bigr) = 1- (4k+2-1) = 1-(4k-1) = -4k,$$
+which is also a multiple of $4$.\slug
+
+Sometimes, the cases in which to split the proof are not so obvious.
+
+\proclaim Proposition \advthm. There exist irrational numbers $a$ and $b$ such that $a^b$ is rational.
+
+\proof Consider the number $(\sqrt 2)^{\sqrt 2}$. It is either rational or irrational.
+
+If it is rational, then we can set $a = b = \sqrt 2$, which we already know to be irrational, by
+Proposition~{\propsqrttwo}.
+
+If it is irrational, then we can set $a =(\sqrt 2)^{\sqrt 2}$ and $b=\sqrt 2$. We compute
+$$a^b = \bigl((\sqrt 2)^{\sqrt 2}\bigr)^{\sqrt 2} = (\sqrt 2)^{\sqrt 2 \cdot \sqrt 2}
+= (\sqrt 2)^2 = 2,$$
+which is rational.\slug
+
+Note that in this proof, we did not need to know whether $(\sqrt 2)^{\sqrt 2}$ is rational or not;
+we just know that it must either be rational or irrational.
+(In fact, it is known that $(\sqrt 2)^{\sqrt 2}$ is irrational, but the usual proof of this
+requires some notions from complex analysis as well as Galois theory, subjects that are
+typically not encountered until at least the second or third year of a mathematics degree.)
+
+\medskip\boldlabel\datestamp{7}{} Mathematical induction. Sometimes we want to prove an infinite
+number of statements, indexed by the natural numbers. If the complexity of the statements
+sort of ``grows'' in $n$ (a vague notion to be made precise soon), the following theorem holds.
+
+\parenproclaim Theorem {\advthm} (Principle of mathematical induction). Let $P(n)$ be a family of predicates
+indexed by $n\in \NN$. Let $m\in \NN$. If
+\medskip
+\item{i)} $P(m)$ holds; and
+\smallskip
+\item{ii)} for all $n\ge m$,
+$$\bigl( P(m)\wedge P(m+1)\wedge \cdots \wedge P(n) \bigr) \To P(n+1)$$
+\medskip
+then $P(n)$ holds for all $n\ge m$.
+
+In the following proof, we use the fact that every nonempty subset of the natural numbers has
+a least element, the proof of which belongs to a higher-level set theory course.
+
+\proof Assume that (i) and (ii) both hold. Our goal is to prove that $P(n)$ is true for all
+$n\ge m$, so we shall suppose, towards a contradiction, that there is some $n\ge m$ such that
+$\neg P(n)$ holds. (As an exercise, convince yourself, via symbolic manipulations, that
+$$\neg \forall n : n\ge m \To P(n) \equiv \exists n : n\ge m \wedge \neg P(n).)$$
+In other words, the set of $n\ge m$ such that $\neg P(n)$ holds is nonempty, so it has a least element,
+call it $k$. So $k\ge m$, and we cannot have $k=m$ due to (i), so $k>m$. Furthermore, by the minimality
+of $k$, the statements $P(m)$, $P(m+1)$, all the way up to $P(k-1)$ are true. By (ii) though,
+this implies that $P(k)$ holds: a contradiction.\slug
+
+When proving something by induction, we need to prove two things: one of the form (i) and another
+of the form (ii). The former is called the {\it base case} and the latter the {\it induction step}
+or {\it inductive step}. The assumption to the left of the $\To$ symbol in (ii)
+is called the {\it induction hypothesis} or {\it inductive hypothesis}.
+Technically, the theorem above is called the principle of {\it strong} induction;
+the principle of {\it weak} induction has
+\medskip
+\item{ii)$'$} for all $n\ge m$, $P(n)\To P(n+1)$,
+\medskip
+instead of the stronger (ii). It turns out that both types of induction are actually equivalent,
+so we'll use both interchangeably. More often than not, the hypothesis (ii)$'$ is perfectly sufficient,
+and in these cases we'll simply use weak induction so as not to clutter our proofs with lots of unused
+hypotheses.
+
+Here is an example. Suppose we want to find (and prove) a formula for the sum of the first $n$ odd numbers.
+The first thing to do is to try some small cases. When $n = 1$, the sum is just $1$. When $n=2$,
+the sum is $1+3 = 4$, when $n=3$, we have $1+3+5 = 9$, and for $n=4$, we compute $1+3+5+7 = 16$.
+So the pattern goes $1,4,9,16,\ldots$, which leads us to conjecture that the sum of the first $n$
+odd numbers might equal $n^2$. In fact, this is true, and we shall prove it by induction.
+
+\proclaim Proposition \advthm. For all integers $n\ge 1$,
+$$\sum_{i=1}^n (2i-1) = n^2.$$
+
+\proof By induction on $n$. First we prove the base case, $n=1$. We have
+$$\sum_{i=1}^1 (2i-1) = 2-1 = 1 = 1^2.$$
+Now for the inductive step. Let $n\ge 1$ and assume that
+$$ \sum_{i=1}^n (2i-1) = n^2.$$
+Then
+$$\eqalign{
+\sum_{i=1}^{n+1} (2i-1) &= \sum_{i=1}^n (2i-1) + \bigl(2(n+1) - 1\bigr) \cr
+&= n^2 + 2n+2 - 1 \cr
+&= n^2 + 2n + 1\cr
+&= (n+1)^2,\cr
+}$$
+where it is in the second line that we used the inductive hypothesis.\slug
+
+The first thing to notice about doing a proof by induction is that the proof method itself doesn't
+tell you what it is you should prove. You have to guess at the correct statement first. Also,
+proofs by induction are often ``unenlightening,'' in that they often don't reveal the fundamental reasons
+why something might be true. (The previous proposition can be illustrated by a rather simple
+picture, which is not a proof, but is somewhat more elucidating that the induction proof.)
+
+A longer example now. Suppose we have a pile of $n$ stones, with $n\ge 1$. We have a job, which
+can be described by a pseudo-algorithm.
+
+\algbegin Algorithm S (Divide stones). The input to
+this algorithm is an integer $n\ge 1$, representing a number of stones.
+We have a list {\tt PILES} of integers representing a collection
+of piles of stones, as well as an integer variable {\tt BALANCE}.
+Initialise ${\tt PILES} \Gets 1$ (one pile with $n$ stones), and set ${\tt BALANCE}\gets 0$.
+This algorithm splits the stones into $n$ piles of $1$ stone each, accumulating profits
+into {\tt BALANCE} along the way.
+\algstep S1. [Done?] If every element of {\tt PILES} is $1$, terminate the algorithm and output
+{\tt BALANCE}.
+\algstep S2. [Choose a pile.] Select some element $m>2$
+from {\tt PILES} and remove it from the list. (The variable $m$ is the number of stones in this pile.)
+\algstep S3. [Split.] Let $k$ and $l$ be two numbers with $k+l = m$. We append ${\tt PILES}\Gets k$ and
+${\tt PILES} \Gets l$, and increment ${\tt BALANCE} \gets {\tt BALANCE} + k\cdot l$. (We have split
+pile $n$ into two piles of size $k$ and $l$, and the payout for doing so is $k\cdot l$ dollars.)
+\algstep S4. [Loop.] Go to step S1.\slug
+
+This is not, strictly speaking, an algorithm, since we didn't specify how the algorithm
+should choose the integers $k$ and $l$ that add up
+to $m$ in step S3. However, running through a few instances with, say, $n=6$, on paper, using whatever
+choices of split you like in every iteration of step S3, you'll find that the algorithm always terminates
+with ${\tt BALANCE} = 15$. Trying a few different starting values of $n$ might lead you to conjecture
+the following proposition, which we will prove by (strong) induction.
+
+\proclaim Proposition \advthm. For a given input $n$, Algorithm S always outputs ${\tt BALANCE} = n(n-1)/2$,
+regardless of the choice of split at any given iteration of step S3.
+
+\proof By induction on $n$. For the base case $n=1$, note that we immediately output ${\tt BALANCE} = 0$
+in the very first step of the algorithm, and $0 = 1(1-1)/2$.
+
+Now for the inductive step, let $n\ge 1$ and suppose that for $1\le k \le n$, the payout for running
+algorithm S on input $l$ is $l(l-1)/2$. Suppose we have a pile of $n+1$ stones. We shall divide
+it into piles of size $k$ and size $n+1-k$. The total payout will be the pay for this division,
+namely $k(n+1-k)$, plus the pay for further subdividing the two piles. Thus by the induction
+hypothesis applied twice, the total payout will be
+$$\eqalign{
+k(n+1-k) + &{k(k-1)\over 2} + {(n+1-k)(n-k)\over 2} \cr
+&= {2nk + 2k - 2k^2 + k^2 - k + n^2 + n -2nk - k + k^2\over 2} \cr
+&= {n^2 + n\over 2} \cr
+&= {(n+1)\bigl((n+1)-1\bigl)\over 2},\cr
+}$$
+which is the expected formula for $n+1$.\slug
+
+\advsect Functions
+
+A {\it function} $f$ from a set $A$ to a set $B$ is a subset $f\subseteq A\times B$ such that
+for every $a\in A$, there is exactly one $b\in B$ such that $(a,b)\in f$. If $(a,b)\in f$,
+we write $f(a) = b$. The set $A$ is called the {\it domain} and the set $B$ is the {\it codomain}.
+The notation $f:A\to B$ is a way of concisely writing ``$f$ is a function with domain $A$ and
+codomain $B$.''
+
+Here are some examples and non-examples.
+\medskip
+\item{i)} The function $f:\RR\to\RR$ given by $f(x) = x^2$ is
+$$f = \bigl\{ (a,b)\in \RR^2 : b = a^2 \bigl\},$$
+when written in set-builder notation.
+\smallskip
+\item{ii)} On the other hand, the set
+$$g = \bigl\{ (a,b)\in \RR^2 : a = b^2 \bigl\}$$
+is not a function, since $(1,1)$ and $(1,-1)$ are both in $g$. Furthermore,
+there is no element of $g$ with $-1$ as its first coordinate.
+\smallskip
+\item{iii)} If $X = \{1,4,5\}$ and $Y = \{0,1,2,3,4,5\}$, then
+$$\bigl\{ (1,0), (2, 4), (5,5)\bigr\}$$
+is a function.
+\smallskip
+\item{iv)} The set
+$$h = \bigl\{ (x,y)\in \RR^2 : y = 1/x \bigl\}$$
+is not a function, since there is no $y$ such that $(0,y)\in h$. However
+if we amend the domain and consider
+$$h = \bigl\{ (x,y)\in \bigl(\RR\setminus \{0\}\bigr)\times \RR : y = 1/x \bigl\},$$
+then in fact, $h:\RR\setminus \{0\} \to \RR$ is a function.
+\medskip
+\datestamp{8}{}
+The {\it range} or {\it image} of a function $f:A\to B$ is
+$$f(A) = \bigl\{b\in B : \hbox{there exists}\ a\in A\ \hbox{such that}\ b = f(a)\bigr\}.$$
+These are all the values $f$ actually outputs. For instance, if we let $f : \ZZ\to \NN$
+be given by $f(n) = n^2$, then
+$$\eqalign{
+f(\ZZ) &= \{\ldots, f(-2), f(-1), f(0) , f(1), f(2),\ldots\} \cr
+&= \{\ldots, 9, 4, 1, 0 1, 4, 9,\ldots\} \cr
+&= \{0, 1, 4, 9, 16, 25, \ldots\}. \cr
+}$$
+A function $f:A\to B$ is called {\it surjective} or {\it onto} if $f(A) = B$, that is,
+for every $b\in B$ there exists some $a\in A$ such that $f(a) = b$. An example of a
+surjective function is $f:\QQ\to \QQ$ sending $x\mapsto x/2$. This is because for any $q\in \QQ$,
+we can set $r = 2q$, and
+$$f(r) = {r\over 2} = {2q\over 2} = q.$$
+A function $f:A\to B$ is called {\it injective} or {\it one-to-one} if for all $a_1,a_2\in A$
+with $a_1 \ne a_2$, we also have $f(a_1)\ne f(a_2)$. Equivalently, $f$ is injective if
+$f(a_1) = f(a_2)$ implies that $a_1 = a_2$ for all $a_1, a_2 \in A$. For instance, the function
+$f:\ZZ\to \NN$ that sends $n\mapsto n^2$ is not injective since $f(-1) = f(1)$, but $-1 \ne 1$.
+On the other hand, if we modify the domain, considering $f:\NN\to \NN$ sending $n\mapsto n^2$,
+then now $f$ is injective, since if $f(m) = f(n)$, then $m^2 = n^2$, and there is only one positive
+integer that squares to any given integer, so $m=n$.
+
+Hence we see that any function can be transformed into an injective one, in principle, by shrinking its
+domain (though this new function might no longer have the properties you liked in the original one),
+and any function $f:A\to B$ can be made surjective by changing its codomain to its range, i.e.,
+letting $B = f(A)$.
+
+\medskip\boldlabel The pigeonhole principle. We now take a brief pause to introduce one of
+the most fundamental laws in discrete mathematics. It is called the {\it pigeonhole principle},
+because one formulation of it is, {\sl If $n$ pigeons nest in $n-1$ holes, then there is at least
+one hole that contains more than one pigeon.}
+First we'll prove a more general statement.
+
+\proclaim Theorem \advthm. Let $a_1, a_2, \ldots, a_m$ be a finite sequence (repeats allowed) of
+real numbers. Let
+$$a = {1\over m}\sum_{i=1}^m a_i$$
+be the average value of the sequence and let $m$ be the maximum value the sequence attains. Then $m \ge a$.
+
+\proof We have
+$$a = {1\over m}\sum_{i=1}^m a_i \le {1\over m} \sum_{i=1}^m m = {m^2\over m} = m.\noskipslug$$
+
+This statement can be summed up in one sentence: {\sl The maximum is at least the average.}
+
+A function $f$ is called {\it bijective} (or a {\it one-to-one correspondence})
+if it is injective and surjective. Bijections are important because of the following proposition.
+
+\proclaim Proposition \advthm. Let $A$ and $B$ be finite sets. Then
+\medskip
+\item{i)} there exists a bijection $f:A\to B$ if and only if $|A| = |B|$; and
+\smallskip
+\item{ii)} if $|A| = |B|$ and $f:A\to B$ then $f$ is injective if and only if $f$ is surjective.
+\medskip
+
+\proof Suppose $|A| = |B| = n$. Then choose an ordering $a_1, \ldots, a_n$ of $A$ and an ordering
+$b_1, \ldots, b_n$ of $B$. Let $f(a_i) = b_i$ for all $1\le i\le n$. By construction, this is a bijection.
+On the other hand, suppose $|A|\ne |B|$. If $A<B$, then $f$ cannot be surjective, since the image of
+$f$ has size at most $|A| < |B|$ (at least one element of $B$ must be missed). If $A>B$, then since
+$f(A)$ must
 

### 2024-08-29 teaching/math240f24/draft.tex
Teaching

--- 2024-08-27 teaching/math240f24/notes.tex
@@
 Sets can be written by listing out its elements. For example,
-$$\{1,4,7,10,\sqrt{782}\}$$
+$$\bigl\{1,4,7,10,\sqrt{782}\bigr\}$$
 and
@@
 
+\edef\propcardpowerset{\the\thmcount}
 \proclaim Proposition \advthm. If $A$ is finite, then
@@
 \medskip
-\datestamp{8}{}
+
+\medskip\boldlabel\datestamp{8}{} Injective and surjective functions.
 The {\it range} or {\it image} of a function $f:A\to B$ is
@@
 \medskip\boldlabel The pigeonhole principle. We now take a brief pause to introduce one of
-the most fundamental laws in discrete mathematics. It is called the {\it pigeonhole principle},
-because one formulation of it is, {\sl If $n$ pigeons nest in $n-1$ holes, then there is at least
-one hole that contains more than one pigeon.}
-First we'll prove a more general statement.
+the most fundamental laws in discrete mathematics, called the pigeonhole principle.
+We begin with the following intuitive theorem.
 
@@
 
-This statement can be summed up in one sentence: {\sl The maximum is at least the average.}
-
-A function $f$ is called {\it bijective} (or a {\it one-to-one correspondence})
+This theorem can be summed up in one sentence: {\sl The maximum is at least the average.}
+Don't underestimate this theorem even though its proof was a one-liner! It is often used to prove
+highly nontrivial results.
+(As an exercise, prove the similar statement: {\sl The minimum is at most the average.})
+From here we are now equipped to prove (a general version of) the pigeonhole principle.
+
+\edef\thmgeneralpigeonhole{\the\thmcount}
+\proclaim Theorem \advthm. Let $A$ and $B$ be finite sets with $|A| = m$ and $|B|=n$.
+For every function $f:A\to B$, then there is some $b\in B$ such that there are at least
+$\lceil m/n\rceil$ elements $a\in A$ with $f(a) = b$.
+
+\proof For $1\le i \le n$, let $r_i$ be the number of $a\in A$
+such that $f(a) = b_i$. This is a sequence that adds up to $m$, since every $a$ in $A$ maps to
+exactly one element of $B$. So the sequence has average $m/n$, and by the previous theorem,
+there must be some $j$ such that $r_j \ge m/n$. But the $r_i$ are all actually integers (since
+cardinalities of finite sets are integers), meaning that $r_j \ge \lceil m/n \rceil$. Letting
+$b = b_j$ completes the proof.\slug
+
+The reason this is called the pigeonhole principle is because of the following special case.
+
+\parenproclaim Corollary {\advthm} (Pigeonhole principle). Let $n\ge 2$. If $n$ pigeons nest in $n-1$ holes,
+there is at least one hole that contains at least two pigeons.
+
+\proof Let $A$ be the set of pigeons and $B$ the set of pigeonholes. Let $f$ be any function sending
+the set of pigeons to the set of holes. By the previous theorem, there is some hole with at least
+$\lceil n/(n-1)\rceil = 2$ pigeons in it. (This is because $1 < n/(n-1) < 2$ for all integers $n\ge 2$.)\slug
+
+\medskip\boldlabel Bijections.
+A function $f$ is called {\it bijective} (or a {\it bijection}, or a {\it one-to-one correspondence})
 if it is injective and surjective. Bijections are important because of the following proposition.
@@
 \proof Suppose $|A| = |B| = n$. Then choose an ordering $a_1, \ldots, a_n$ of $A$ and an ordering
-$b_1, \ldots, b_n$ of $B$. Let $f(a_i) = b_i$ for all $1\le i\le n$. By construction, this is a bijection.
-On the other hand, suppose $|A|\ne |B|$. If $A<B$, then $f$ cannot be surjective, since the image of
-$f$ has size at most $|A| < |B|$ (at least one element of $B$ must be missed). If $A>B$, then since
-$f(A)$ must
+$b_1, \ldots, b_n$ of $B$. Let $f(a_i) = b_i$ for all $1\le i\le n$. By construction, this is a bijection,
+proving one direction of (i).
+
+On the other hand, suppose $|A|\ne |B|$ (so we prove this direction by contraposition).
+If $A<B$, then $f$ cannot be surjective, since the image of
+$f$ has size at most $|A| < |B|$ (at least one element of $B$ must be missed). If $A>B$, then
+$|A|/|B| > 1$, so by Theorem~{\thmgeneralpigeonhole}, there is some element $b\in B$
+such that the number of $a\in A$ mapping to $b$ is at least $\lceil |A|/|B| \rceil \ge 2$.
+This means that $f$ is not injective. We have proved part (i).
+
+To prove part (ii), let $|A| = |B|$ and let $f:A\to B$. First we assume that $f$ is injective.
+We enumerate $A = \{a_1, a_2, \ldots, a_n\}$. Then
+$$f(A) = \bigl\{ f(a_1), f(a_2), \ldots, f(a_n)\bigr\} \subseteq B.$$
+All of the $f(a_i)$ are distinct, since if $f(a_i) = f(a_j)$, then $a_i = a_j$. So
+$\bigl| f(A)\bigr| = |A| = n$, and $f(A)$ is a size $n$ subset of $B$, which also has size $n$.
+Hence $f(A) = B$; that is, $f$ is surjective.
+
+Lastly, suppose $f$ is not injective (again we are using contraposition). So there
+are $a_i$ and $a_j$ such that $a_i\ne a_j$ but $f(a_i) = f(a_j)$. So
+$$\bigl|f(A)\bigr| = \bigl|\bigl\{ f(a_1), f(a_2), \ldots, f(a_n)\bigr\}\bigr| < n = |B|,$$
+so $f(A) \ne B$ and $f$ is not surjective.\slug
+
+Item (i) of the previous proposition should be entirely intuitive, especially if we use the alternative
+nomenclature ``one-to-one correspondence'' instead of ``bijection.'' (In fact, we already implicitly
+used (i) in these notes, in the proof of Proposition~{\propcardpowerset}.) Item (ii) is perhaps not
+as immediate, but should become clear if you work it out with a picture.
+
+\medskip\boldlabel Bijections.
+A function $f:A\to B$ is called {\it invertible} if there exists $g:B\to A$ such that
+\medskip
+\item{i)} for all $b\in B$, $f\bigl(g(b)\bigr) = b$; and
+\smallskip
+\item{ii)} for all $a\in A$, $g\bigl(f(a)\bigr) = a$.
+\medskip
+If $g$ exists, it can be shown that $g$ must be unique, so we write $g = f^{-1}$ and speak of
+{\it the} inverse of $f$.
+
+\proclaim Proposition \advthm. Let $f:A\to B$ be a function. Then $f$ is invertible
+if and only if $f$ is bijective.
+
+\proof First we assume that $f$ is invertible. So there exists an inverse $g$ of $f$. For each $b\in B$,
+setting $a = g(b)$ we have
+$$f(a) = f\bigl(g(b)\bigr) = b.$$
+This proves that $f$ is surjective. To show that $f$ is injective, suppose that $f(a_1) = f(a_2)$.
+By applying $g$ on both sides, we have $g\bigl(f(a_1)\bigr) = g\bigl(f(a_2)\bigr)$, whence
+$a_1 = a_2$, by definition of $g$.
+
+Now assume that $f$ is bijective. We need to define $g:B\to A$. Well, gven any $b\in B$, there is
+some $a$ such that $f(a) = b$, from surjectivity of $f$, and this $a$ is unique, since $f$ is
+injective. So set $g(b) = a$ (and repeat this process for every $b\in B$). We have
+$f\bigl(g(b)\bigr) = f(a) = b$, and for every $a\in A$, by definition of $g$ the
+element $g\bigl(f(a)\bigr)$ is the unique element in $A$ that gets brought to $f(a)$ by $f$,
+has to be $a$ itself.\slug
+
+Sometimes to prove that two sets have the same cardinality, it is easier to prove that there
+exists a bijection (as we already saw in the example of Proposition~{\propcardpowerset}),
+and sometimes to prove that a function is a bijection, it is easier to show that it has an inverse,
+rather than messing around with the definitions of injective and surjective.
+Here's an example.
+
+\proclaim Proposition. Let $X$ be a finite nonempty set. Let $E$ be the set of all subsets of $X$ with
+even cardinality, and let $F$ be the set of all subsets of $X$ with odd cardinality.
+Then $|E| = |F|$.
+
+\proof We shall construct a function $f:E\to F$. Fix one specific $x\in X$;
+we can do this because $X\ne \emptyset$. Now, for all $A\in E$, let
+$$f(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
+Note that since $|A|$ is even for all $A\in E$, the cardinality of $f(A)$ is odd (in the first
+case it is $|A|-1$ and in the second case it is $|A|+1$. This shows that $f$ is indeed a function
+with codomain $F$. Now to prove $|E| = |F|$ we will show that $f$ is bijective, which we
+do by showing that $f$ is an inverse (as an exercise, you might try to prove bijectivity from
+the definitions of injective and surjective).
+
+We define $g : F\to E$ by
+$$g(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
+As before, $\bigl|g(A)\bigr|$ is even, since $A$ is assumed to be a member of $F$ now. Now
+for any $A\in E$,
+$$\eqalign{
+g\bigl(f(A)\bigr) &= \cases{g\bigl(A\setminus\{x\}\bigr), & if $x\in A$;\cr
+g\bigl(A\cup \{x\}\bigr), & if $x\notin A$\cr} \cr
+&= \cases{\bigl(A\setminus\{x\}\bigr)\cup \{x\}, & if $x\in A$;\cr
+\bigl(A\cup \{x\}\bigr)\setminus \{x\} , & if $x\notin A$\cr} \cr
+&= \cases{A, & if $x\in A$;\cr
+A, & if $x\notin A$\cr} \cr
+&= A.\cr
+}$$
+The proof that $f\bigl(g(A)\bigr) = A$ is similar. Thus $g$ is the inverse of $f$.\slug
+
+\advsect Cardinality
+
+\datestamp{9}{}
+Earlier, we defined the cardinality of a set to be the number of elements it contains. What, then,
+is the cardinality of $\NN$? How about $\RR$? You might say $\infty$, but this is not a number
+(at least, it's not an element of $\NN$, the way all cardinalities of finite sets are).
+So perhaps we should amend our question to the following: {\sl When do infinite sets have the same size?}
+Our experience with functions leads us to the answer: {\sl When there exists a bijection between
+them}. We shall say that $A$ and $B$ are {\it equipotent} (or {\it equinumerous}, or {\it have
+the same cardinality}) if there exists a bijection between $A$ and $B$. In this case
+we write $|A| = |B|$.
+
+As an example, the sets $\NN$ and $\NN\setminus\{0\}$ are equipotent, since $f$
+given by $n\mapsto n+1$ is a bijection $\NN\to\NN\setminus \{0\}$. (Check that
+$f^{-1}(m) = m-1$ is its inverse.)
+
+It is even possible to remove an infinite number of elements from $\NN$ and end up with
+something still equipotent with $\NN$. To see this, let $E$ be the set of nonnegative even integers,
+and consider the function $f :\NN\to E$ sending $n\mapsto 2n$. This is injective because if
+$2m = 2n$, then dividing out by $2$ on both sides yields $m=n$. It is surjective because if
+$n\in E$, then $n=2k$ for some $k\in\NN$, by definition, and $f(k) = 2k = n$.
 

### 2024-08-29 teaching/math240f24/draft.tex
Lecture 1

@@
 
-\datestamp{1}{}
+\datestamp{29}{viii}
 A {\it set} is a collection of distinct objects, called its {\it elements} or its {\it members}. If
@@
 and
-$$A = \{ x\in \ZZ : \hbox{there exists}\ l\in \ZZ\ \hbox{such that}\ x = 2l + 5\}$$
+$$B = \{ x\in \ZZ : \hbox{there exists}\ l\in \ZZ\ \hbox{such that}\ x = 2l + 5\}$$
 are equal. (Both are different ways of expressing the set of all odd integers.)
@@
 (vii) the distributive law, (viii) the absorption law, and (ix) the complement law.
+\bye
 
@@
 
-Such is the price of meddling with ``sets'' that aren't subsets of known sets.
+Such is the price of meddling with ``sets'' that aren't subsets of known sets. This also shows that there
+is no such thing as ``the set of all sets.''
 

### 2024-09-04 teaching/math240f24/draft.tex
Fixed formatting
Lecture 3 september

@@
 \bar{\bar{A\cup B} \cup (A\cap B)}$$
-\datestamp{2}{} More set identities abound. We state the following proposition without proof; you should try going
+More set identities abound. We state the following proposition without proof; you should try going
 though this list and convincing yourself that each identity holds, for all sets $A$, $B$, and $C$.
@@
 (vii) the distributive law, (viii) the absorption law, and (ix) the complement law.
-\bye
 
@@
 
-\medskip\boldlabel De Morgan's laws. There are two important laws relating complements with union and intersection.
+\medskip\boldlabel\datestamp{03}{ix} De Morgan's laws.
+There are two important laws relating complements with union and intersection.
 We shall state them as a proposition, this time giving a proof (of one of them).
@@
 $2^{A\cup B} \not\subseteq 2^A \cup 2^B$.
+\bye
 
@@
 A {\it function} $f$ from a set $A$ to a set $B$ is a subset $f\subseteq A\times B$ such that
-for every $a\in A$, there is exactly one $b\in B$ such that $(a,b)\in f$. If $(a,b)\in f$,
+for every $a\in A$, there is exactly one $b\in B$ such that $(a,b)\in f$. (If there is no $b\in B$,
+or more than one, we say that $f$ is not well-defined.
+If $(a,b)\in f$,
 we write $f(a) = b$. The set $A$ is called the {\it domain} and the set $B$ is the {\it codomain}.
@@
 
+So we can find subset of $\NN$ equipotent with it. It turns out we can also find supersets of $\NN$ with
+the same property.
+
+\proclaim Theorem \advthm. We have $|\NN| = |\ZZ|$.
+
+\proof We define $f : \NN\to \ZZ$ by
+$$f(n) = \cases{{n\over 2}, & if $n$ is even;\cr {-{n+1\over 2}}, & if $n$ is odd.}$$
+We shall show that $f$ is a bijection.
+
+To show that $f$ is injective, there are four cases. Note first that if $n$ is even, then $f(n)\ge 0$,
+and if $n$ is odd, then $f(n) < 0$. So if $f(m) = f(n)$ for some $m,n\in \NN$, then $f(m)$ and $f(n)$ must
+either both be negative, or both be nonnegative. Either way, $m$ and $n$
+are either both even or they are both odd. If $m$ and $n$ are both even, then from $f(m) = f(n)$ we derive
+$${m\over 2} = {n\over 2},$$
+whence multiplying by $2$ on both sides we see that $m=n$. If $m$ and $n$ are both odd, then
+$$-{m+1\over 2} = -{n+1\over 2},$$
+so, multiplying by $-2$ and subtracting $1$ from both sides we have $m=n$ in this case as well.
+
+Now we show that $f$ is surjective. Let $k\in \ZZ$. If $k\ge 0$, then consider $n=2k$. We have
+$$f(n) = f(2k) = {2k\over 2} = k.$$
+If $k<0$, then consider $n=-2k-1$. (Check that this is an element of $\NN$.) Then
+$$f(n) = f(-2k-1) = -{-2k-1+1\over 2} = -{-2k\over 2} = k.$$
+This shows that $f$ is surjective, so $f$ is in fact bijective and we conclude that $\NN=\ZZ$.\slug
+
+We say that a set $A$ is {\it countably infinite} if there exists a bijection $f:\NN\to A$, that is,
+if $|\NN| = |A|$.
+A set is said to be {\it countable} if it is either finite or countably infinite.
+Otherwise it is called {\it uncountable}.
+The previous theorem shows that $\ZZ$ is countably infinite.
+If a set $A$ is countable, then there exists an enumeration
+$$A = \{a_0, a_1, a_2, \ldots\}.$$
+This is useful in the proof of the following theorem.
+
+\proclaim Theorem \advthm. If $A$ and $B$ are countably infinite sets,
+then $A\times B$ is also countably infinite.
+
+\noindent{\it Proof sketch.}\enspace
+Enumerate $A = \{a_1, a_2, \ldots\}$ and $B = \{b_1, b_2, \ldots\}$. Now given
+$(a_i, b_j)\in A\times B$ for some $i,j\in \NN$, we can let
+$$f(a_i, b_j) = {(i+j+1)(i+j+2)\over 2} - i - 1.$$
+Work out the values of $f(a_i, b_j)$ for small choices of $i$ and $j$, and draw a picture to convince yourself that
+$f : A\times B \to\NN$ is a bijection. (Rigorously proving this fact
+is quite tedious, so we omit the details.) This means that $f^{-1}$
+is a bijection from $\NN$ to $A\times B$, so $A\times B$ is countably infinite.\slug
+
+\proclaim Corollary \advthm. We have $|\ZZ\times \ZZ| = |\NN|$.
+
+\proof
+We proved earlier that $\ZZ$ is countably infinite, so we may apply the previous theorem with $A = B = \ZZ$.\slug
+
+One can also show that $|\QQ| = |\NN|$. We will not include the proof, but the idea is that there
+is a surjective map from $\ZZ\times\bigl( \ZZ \setminus \{0\}\bigr)\to \QQ$ given by $(m,n)\mapsto m/n$.
+On the other hand, there are infinite sets that are not countably infinite. An {\it infinite binary
+string} is a function $f:\NN\to \{0,1\}$. (By listing out
+$$ f(0), f(1), f(2), \ldots$$
+you get an infinite string of $0$s and $1$.)
+
+\proclaim The set $A$ of all infinite binary strings is uncountable.
+
+\proof Certainly $A$ is not finite, so we need to show that $|\NN|\ne A$.
+
+Let $f : \NN\to A$. We shall show that $f$ is not surjective. For all $m,n\in \NN$, let $a_{m,n}$
+be the $n$th bit of $f(m)$. Consider the infinite binary string
+$$s = (1-a_{0,0},1-a_{1,1}, 1-a_{2,2}, 1-a_{3,3}, \ldots).$$
+This string $s$ cannot equal $f(m)$ for any $m\in \NN$, since given an arbitrary $m$,
+the $m$th bit of $f(m)$ is $a_{m,m}$, whereas the $m$th bit of $s$ is $1-a{m,m}$. Hence there is some
+$s\in A$ such that $f(m)\ne s$ for all $m\in \NN$. So $f$ is not surjective.
+
+Since there can be no surjection $f:\NN\to A$, {\it a fortiori} we cannot have a bijection $\NN\to A$.
+We conclude that $|A| \ne |\NN|$.\slug
+
+The set of all infinite binary strings is in bijection with the set of all real numbers in the interval
+$[0,1)$. If we're being extra pedantic, we need to forbid an infinite trailing string of all $1$s,
+since, e.g., $0.0111\ldots = 0.1000\ldots$ after carrying the $1$s (akin to the fact
+that, e.g., $0.0999\ldots = 0.1$). After dealing with this detail, we will have shown that
+$\bigl| [0,1]\bigr| \ne |\NN|$.
+
 \bye

### 2024-09-07 teaching/math240f24/draft.tex
Lecture 3

@@
 $2^{A\cup B} \not\subseteq 2^A \cup 2^B$.
-\bye
-
-\medskip\boldlabel\datestamp{3}{} Russell's paradox. Earlier, we said that the sets we are working with
+
+\medskip\boldlabel\datestamp{05}{ix} Russell's paradox. Earlier, we said that the sets we are working with
 need to be a subset of a universe $U$, which has already been proven to be a set. We gave lots of
 ways to make sets out of new sets, such as the union and intersection operations, etc. Starting with
-the empty set $\emptyset$ is a set, it is possible to define the set of natural numbers as follows.
+the assumption that the empty set $\emptyset$ is a set,
+it is possible to define the set of natural numbers as follows.
 We can define $0 = \emptyset$, $1 = \{0\}$, and $2 = \{0,1\}$, and so on.
@@
 \smallskip
-\item{ix)} $p\cup \bar p \equiv 1$ and $p\wedge \bar p \equiv 0$.\slug
+\item{ix)} $p\vee \bar p \equiv 1$ and $p\wedge \bar p \equiv 0$.\slug
 \medskip
@@
 
-\medskip\boldlabel\datestamp{4}{} Conditional and biconditional. We now examine the {\it conditional}
+\medskip\boldlabel Conditional and biconditional. We now examine the {\it conditional}
 logical relation {\mc IF} $p$ {\mc THEN} $q$. It is denoted by $p\To q$ and its truth table is given by
@@
 is pretty self-explanatory; in it, $p\Leftrightarrow q$ is true whenever $p$ and $q$ have the same truth value.
+\bye
 The name ``biconditional'' is suggested by (part of) the following proposition.

### 2024-09-10 teaching/math240f24/draft.tex
Lecture 4

@@
 \smallskip
-\item{vii)} $p\vee (q\vee r) \equiv (p\vee q) \wedge (p\vee r)$ and
+\item{vii)} $p\vee (q\wedge r) \equiv (p\vee q) \wedge (p\vee r)$ and
 $p\wedge (q\vee r) \equiv (p\wedge q)\vee (p\wedge r)$;
@@
 
-\medskip\boldlabel Conditional and biconditional. We now examine the {\it conditional}
+\medskip\boldlabel\datestamp{10}{ix} Conditional and biconditional. We now examine the {\it conditional}
 logical relation {\mc IF} $p$ {\mc THEN} $q$. It is denoted by $p\To q$ and its truth table is given by
@@
 is pretty self-explanatory; in it, $p\Leftrightarrow q$ is true whenever $p$ and $q$ have the same truth value.
-\bye
 The name ``biconditional'' is suggested by (part of) the following proposition.
@@
 about {\mc SAT} in a higher-level class on computational complexity (e.g., {\mc COMP 360/362}).
+\bye
 
@@
 
-\proclaim Proposition. Let $X$ be a finite nonempty set. Let $E$ be the set of all subsets of $X$ with
+\proclaim Proposition \advthm. Let $X$ be a finite nonempty set. Let $E$ be the set of all subsets of $X$ with
 even cardinality, and let $F$ be the set of all subsets of $X$ with odd cardinality.
@@
 the same property.
-
 \proclaim Theorem \advthm. We have $|\NN| = |\ZZ|$.
@@
 $$f(n) = f(-2k-1) = -{-2k-1+1\over 2} = -{-2k\over 2} = k.$$
-This shows that $f$ is surjective, so $f$ is in fact bijective and we conclude that $\NN=\ZZ$.\slug
+This shows that $f$ is surjective, so $f$ is in fact bijective and we conclude that $|\NN|=|\ZZ|$.\slug
 
@@
 The previous theorem shows that $\ZZ$ is countably infinite.
-If a set $A$ is countable, then there exists an enumeration
+
+Sometimes it is difficult to come up with a bijection directly. Instead, we would like to find
+an injection from $A$ to $B$ (which, in some sense, shows that $|A|\le |B|$), and then an injection
+from $B$ to $A$. This is made possible by the following useful theorem, named for Ernst Schr\"oder
+and Felix Bernstein, who independently proved it in 1898. The proof is a bit difficult, so it's
+technically outside the scope of the course. For fun, you might try to do it as an exercise.
+
+\parenproclaim Theorem {\advthm}. (Schr\"oder--Bernstein theorem). If there exists an injective
+function $f:A\to B$ and another injective function $g:B\to A$, then there is a bijection
+$h:A\to B$.
+
+\proof We present the proof as a (difficult) exercise. Here is the roadmap.
+Call $b\in B$ {\it unattached} if there is no $a\in A$ such that $f(a) = b$.
+Let $h:B\to B$ be given by $h(b)  = f\bigl(g(b)\bigr)$. Given $b,b'\in B$, we shall say that
+$b$ is a {\it peer} of $b'$ if either $b = b'$ or there exists some $n\in \NN$ such
+that
+$$ b = \underbrace{h(h(\cdots(h}_{n\ \hbox{\sevenrm times.}}(b'))\cdots))$$
+Say that $b\in B$ is a {\it PAE} if it is the peer of an unattached element. (So unattached
+elements are automatically PAEs, by setting $n=0$.)
+\medskip
+\item{a)} Show that if $a\in A$ is such that
+$f(a)$ is a PAE, then
+there is a unique element $b^*\in B$ such that $g(b^*) = a$, and that
+this element is a PAE.
+\medskip
+\noindent By part (a), if $f(a)$ is a PAE, it makes sense to speak of $g^{-1}(a)$. It is
+the element $b^*\in B$ such that $g(b^*) = a$. From here we define
+$$ r(a) = \cases{ g^{-1}(a), & if $f(a)$ is a PAE;\cr
+f(a), & otherwise.}$$
+\medskip
+\item{b)} Show that if $b\in B$ is a PAE
+then so is $f\bigl(g(b)\bigr)$.
+\item{c)} Show that $r$ is surjective. [{\it Hint}: Do a proof by cases. Every $b\in B$ is either
+a PAE or it is not a PAE.]
+\smallskip
+\item{d)} We already proved in part (a) that if $f(a)$ is a PAE, then so is $r(a)$. Prove
+the converse of this statement.
+\smallskip
+\item{e)} Show that $r$ is injective, and therefore bijective. [{\it Hint}:
+Assume $r(a_1) = r(a_2)$, and do a proof by cases again. Part (d) will be useful here.]\slug
+\medskip
+
+Once again, this proof is outside the scope of the course. Don't lose sleep over it if you can't do it.
+Feel free to email me with questions if you get stuck.
+
+Using the Schr\"oder--Bernstein theorem, we now show that the Cartesian product of two countable sets is also
+countable.
+In the proof, we shall employ the fact that if a set $A$ is countable, then there exists an enumeration
 $$A = \{a_0, a_1, a_2, \ldots\}.$$
-This is useful in the proof of the following theorem.
-
+(If $f$ is the bijection given by the definition, we can set $a_0 = f(0)$, $a_1 = f(1)$, and so on.)
+
+In the following proof, we'll also use the Fundamental Theorem of Arithmetic, which we state now, and prove
+later in the course (in the number theory section). It says that
+any integer can be factored as a product of primes, a theorem you should have learned in grade school.
+
+\parenproclaim Theorem {\advthm} (Fundamental Theorem of Arithmetic). Every positive
+integer $n\ge 2$ can be factored into a product
+$${p_1}^{v_1} {p_2}^{v_2}\cdots {p_n}^{v_n},$$
+where $n\ge 0$ is an integer and, for $1\le i\le n$, $p_i$ is a prime and $v_i$ is a positive integer.
+This factorisation is unique up to the order of the primes.\slug
+
+\edef\thmcartesiancountable{\the\thmcount}
 \proclaim Theorem \advthm. If $A$ and $B$ are countably infinite sets,
@@
 
-\noindent{\it Proof sketch.}\enspace
-Enumerate $A = \{a_1, a_2, \ldots\}$ and $B = \{b_1, b_2, \ldots\}$. Now given
+\proof
+Enumerate $A = \{a_0, a_2, \ldots\}$ and $B = \{b_0, b_2, \ldots\}$. Now given
 $(a_i, b_j)\in A\times B$ for some $i,j\in \NN$, we can let
-$$f(a_i, b_j) = {(i+j+1)(i+j+2)\over 2} - i - 1.$$
-Work out the values of $f(a_i, b_j)$ for small choices of $i$ and $j$, and draw a picture to convince yourself that
-$f : A\times B \to\NN$ is a bijection. (Rigorously proving this fact
-is quite tedious, so we omit the details.) This means that $f^{-1}$
-is a bijection from $\NN$ to $A\times B$, so $A\times B$ is countably infinite.\slug
-
+$$f(a_i, b_j) = 2^i3^j.$$
+We show that this defines an injective function. Suppose that $f(a_i, b_j) = f(a_{i'}, f(b_{j'}).$
+Then $2^i3^j = 2^{i'} 3^{j'}$, and by the uniqueness of prime factorisations in the Fundamental
+Theorem of Arithmetic, we see that $i = i'$ and $j=j'$.
+
+Now we produce an injection $g : \NN\to A\times B$ by simply setting
+$$g(n) = (a_n, b_0).$$
+It is clear this is injective, since if $g(m) = g(n)$, then $(a_m, b_0) = (a_n, b_0)$, and
+that implies that $m=n$.\slug
+
+\edef\corzedtwocountable{\the\thmcount}
 \proclaim Corollary \advthm. We have $|\ZZ\times \ZZ| = |\NN|$.
@@
 
-One can also show that $|\QQ| = |\NN|$. We will not include the proof, but the idea is that there
-is a surjective map from $\ZZ\times\bigl( \ZZ \setminus \{0\}\bigr)\to \QQ$ given by $(m,n)\mapsto m/n$.
-On the other hand, there are infinite sets that are not countably infinite. An {\it infinite binary
-string} is a function $f:\NN\to \{0,1\}$. (By listing out
-$$ f(0), f(1), f(2), \ldots$$
-you get an infinite string of $0$s and $1$.)
-
-\proclaim The set $A$ of all infinite binary strings is uncountable.
+This corollary concerning $\ZZ\times \ZZ$ allows us to prove that $\QQ$ is countable as well.
+
+\proclaim Theorem. The set $\QQ$ of rational numbers is countable.
+
+\proof We define $f:\QQ\to \ZZ\times \ZZ$ as follows. For an element $q\in \QQ$, we let $q = a/b$ the fraction $q$
+written in lowest terms (where $a\in \ZZ$ and $b\in \NN\setminus\{0\}$). Then we set $f(q) = (a,b)$.
+This is an injective function because if $f(q) = f(q')$, then writing $q = a/b$ and $q' = a'/b'$ in lowest
+terms, we have $(a,b) = (a',b')$, so $q = a/b = a'/b' = q'$.
+
+To define an injection $g: \ZZ\times \ZZ\to \QQ$, we recycle the injection we had from the proof
+of Theorem~{\thmcartesiancountable}. We already know that $\ZZ$ is countable, so fix an enumeration
+$\ZZ = \{a_0, a_1, a_2, \ldots\}$. Then let $g(a_i, a_j) = 2^i 3^j$. The range of $g$ is a subset of $\NN$,
+so {\it a fortiori} it is a subset of $\QQ$, and we already showed before that it is an injection.
+
+We have shown that $|\QQ| = |\ZZ\times \ZZ|$, which in turn shows that $|\QQ| = |\NN|$, after
+applying Corollary~{\corzedtwocountable}.\slug
+
+So far, we have just given lots of examples of countably infinite sets. Finally, we give an example
+of a set that is not countably infinite.
+
+\proclaim Theorem \advthm. The set $A$ of all infinite binary strings is uncountable.
 
@@
 since, e.g., $0.0111\ldots = 0.1000\ldots$ after carrying the $1$s (akin to the fact
-that, e.g., $0.0999\ldots = 0.1$). After dealing with this detail, we will have shown that
+that, e.g., $0.0999\ldots = 0.1$). After dealing with this detail, one will have shown that
 $\bigl| [0,1]\bigr| \ne |\NN|$.
 
+\advsect Relations
+
+\datestamp{10}{} A {\it relation} on a set $X$ is a subset $R\subseteq X\times X$. If $(a,b)\in R$
+we write $aRb$ and say ``$a$ is related to $b$.'' Here are some examples.
+\medskip
+\item{i)} The set $< = \bigl\{(a,b)\in \RR\times\RR : a<b\bigr\}$ is a relation. For instance, we can
+write $(2,\pi)\in <$, or $2<\pi$ (which is the more common notation for this relation).
+
 \bye

### 2024-09-13 teaching/math240f24/draft.tex
BCS 2025

@@
 about {\mc SAT} in a higher-level class on computational complexity (e.g., {\mc COMP 360/362}).
-\bye
 
@@
 
-\datestamp{5}{}
+\datestamp{12}{ix}
 Propositional logic allows us to work with simple declarations, but this isn't powerful enough
@@
 
-\datestamp{6}{}
 We're getting to the fun part of the course now.
@@
 Writing a proof is equivalent to describing a winning strategy for the player against the adversary.
+\bye
 

### 2024-09-17 teaching/math240f24/draft.tex
Lecture proof techniques

@@
 the word ``let,'' to indicate that $x$ is given to us by some higher power. In fact, we should sort
-of view this higher power as possibly being malicious. A very useful way to think about writing proof is
+of view this higher power as possibly being malicious. A very useful way to think about writing a proof is
 to imagine a game in which you are trying to prove a statement in predicate logic, and a supernatural adversary
@@
 Writing a proof is equivalent to describing a winning strategy for the player against the adversary.
-\bye
-
+
+\medskip\noindent\boldlabel\datestamp{17}{ix} Proving conditional statements.
 Many mathematical statements introduce some hypotheses, then assert some conclusion. Thus they are some
@@
 which is also a multiple of $4$.\slug
+\bye
 

### 2024-09-19 teaching/math240f24/draft.tex
Sophia Howard created

@@
 By doing this, one will have proved that $\neg p\To 0$, which in turn shows that $p$ is true, since
-$$\neg p \To 0 \equiv \neg(\neg p) \vee 0 \equiv p\vee 0 \vee p.$$
+$$\neg p \To 0 \equiv \neg(\neg p) \vee 0 \equiv p\vee 0 \equiv p.$$
 Here is an example of a proof by contradiction.
@@
 If $n$ is odd, then $n=2k+1$ for some integer $k$, and
-$$1 + (-1)^{2k+1}\bigl(2(2k+1) - 1\bigr) = 1- (4k+2-1) = 1-(4k-1) = -4k,$$
+$$1 + (-1)^{2k+1}\bigl(2(2k+1) - 1\bigr) = 1- (4k+2-1) = 1-(4k+1) = -4k,$$
 which is also a multiple of $4$.\slug
-\bye
-
-Sometimes, the cases in which to split the proof are not so obvious.
+
+\noindent
+\datestamp{19}{ix} \hskip\parindent Sometimes, the cases into which one might split the proof are not so obvious.
 
@@
 
-\medskip\boldlabel\datestamp{7}{} Mathematical induction. Sometimes we want to prove an infinite
+\medskip\boldlabel Mathematical induction. Sometimes we want to prove an infinite
 number of statements, indexed by the natural numbers. If the complexity of the statements
@@
 \smallskip
-\item{iii)} If $X = \{1,4,5\}$ and $Y = \{0,1,2,3,4,5\}$, then
+\item{iii)} If $X = \{1,2,5\}$ and $Y = \{0,1,2,3,4,5\}$, then
 $$\bigl\{ (1,0), (2, 4), (5,5)\bigr\}$$
-is a function.
+is a function from $X$ to $Y$.
 \smallskip
@@
 \medskip
+\bye
 

### 2024-09-25 teaching/math240f24/draft.tex
transfer to dorothea2

@@
 \def\Gets{\Leftarrow}
+\font\titlefont=cmbx12 at 14 pt
+\def\cont#1#2{#1\hskip-2pt\leaders\hbox to 8pt{\hss.\hss}\hfil\hbox to 1em{\hss#2}}
+\def\divides{\mid}
+
+\vglue24pt
 
 \maketitle{MATH 240 Fall 2024}{notes by}{Marcel Goh}{}
+
+\bigskip
 
@@
 
+\vskip60pt
+
+\vskip \medskipamount \hbox to\hsize {\relax \unhcopy \strutbox \cont
+{{\tenssbx I. Foundations \kern 3pt}}{2}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
+%\smallskip
+%\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
+%\smallskip
+%\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
+
+\vfill\eject
+\begingroup\headline{\hfil}\footline{\hfil}
+\vglue160pt
+\centerline{\titlefont I. FOUNDATIONS}
+\vskip220pt
 \bigskip
+\begingroup\obeylines\eightssi
+\hfill Die Mathematik ist in ihrer Entwickelung v\"ollig frei
+\hfill und nur an die selbstredende R\"ucksicht gebunden,
+\hfill dass ihre Begriffe sowohl in sich widerspruchslos sind,
+\hfill als auch in festen durch Definitionen geordeneten Beziehungen
+\hfill zu den vorher gebildeten,
+\hfill bereits vorhandenen und bew\"ahrten Begriffen stehen.
+\eightss
+\smallskip
+\hfill --- GEORG CANTOR, {\eightssi Grundlagen einer allgemeinen Mannigfaltigkeits­lehre} (1883)
+\endgroup%\obeylines
+\bigskip\goodbreak
+\vfill\eject
 
@@
 yourself that the statement above defines the set of prime numbers as you know it.)
+\endgroup%\headline{\hfil}\footline{\hfil}
 
@@
 of $p$ as a precondition to a promise $q$, and then considering whether the promise is broken. As an example,
-suppose your friend says: ``If it snows tomorrow I'll work in Trottier with you.'' If it doesn't snow
+suppose your friend says, If it snows tomorrow I'll work in Trottier with you. If it doesn't snow
 and she doesn't pull up, she hasn't technically broken her promise. If it doesn't
@@
 Now we practise translating converting mathematical statements written in English into formulas
-in predicate logic. Suppose we want to write: ``Every integer is even or odd.'' The universe here is
+in predicate logic. Suppose we want to write, Every integer is even or odd. The universe here is
 is the set $\ZZ$ of integers. The word ``every'' has the same meaning as ``for all,'' so right
-off the bat, we can reexpress the statement as: ``For all $n\in \ZZ$, $n$ is even or $n$ is odd.''
+off the bat, we can reexpress the statement as, For all $n\in \ZZ$, $n$ is even or $n$ is odd.
 In symbols, this is
@@
 $$\forall n\, \bigl((\exists k : n = 2k)\vee (\exists l : n = 2l+1)\bigr),$$
-a formula in predicate logic that means ``every integer is even or odd.''
+a formula in predicate logic that means, Every integer is even or odd.
 
@@
 
-In the following proof, we use the fact that every nonempty subset of the natural numbers has
-a least element, the proof of which belongs to a higher-level set theory course.
+In the following proof, we use the {\it well-ordering principle}, which is the
+fact that every nonempty subset of the natural numbers has
+a least element. You can see the proof of the well-ordering principle in a higher-level set theory course.
 
@@
 \medskip
-\bye
 
@@
 with codomain $F$. Now to prove $|E| = |F|$ we will show that $f$ is bijective, which we
-do by showing that $f$ is an inverse (as an exercise, you might try to prove bijectivity from
+do by showing that $f$ is an inverse (as an exercise, you might instead try to prove bijectivity from
 the definitions of injective and surjective).
@@
 
-\proof We present the proof as a (difficult) exercise. Here is the roadmap.
+\noindent\llap{*}{\it Proof.}\enspace We present the proof as a (difficult) exercise. Here is the roadmap.
 Call $b\in B$ {\it unattached} if there is no $a\in A$ such that $f(a) = b$.
@@
 Once again, this proof is outside the scope of the course. Don't lose sleep over it if you can't do it.
-Feel free to email me with questions if you get stuck.
+Feel free to ask questions at office hours if you get stuck.
 
@@
 \proof
-Enumerate $A = \{a_0, a_2, \ldots\}$ and $B = \{b_0, b_2, \ldots\}$. Now given
+Enumerate $A = \{a_0, a_1, \ldots\}$ and $B = \{b_0, b_1, \ldots\}$. Now given
 $(a_i, b_j)\in A\times B$ for some $i,j\in \NN$, we can let
@@
 
-\proclaim Theorem. The set $\QQ$ of rational numbers is countable.
+\proclaim Theorem \advthm. The set $\QQ$ of rational numbers is countable.
 
@@
 \medskip
-\item{i)} The set $< = \bigl\{(a,b)\in \RR\times\RR : a<b\bigr\}$ is a relation. For instance, we can
-write $(2,\pi)\in <$, or $2<\pi$ (which is the more common notation for this relation).
+\item{i)} The set $L = \bigl\{(a,b)\in \RR\times\RR : a<b\bigr\}$ is a relation. For instance, we can
+write $(2,\pi)\in L$, or $2L\pi$, or $2<\pi$ (which is the more common notation for this relation).
+\smallskip
+\item{ii)} The set $E = \bigl\{ (a,b) \in \RR\times \RR : a = b\bigr\}$ is a relation (it is the ``equals''
+relation).
+\smallskip
+\item{iii)} On $\ZZ$, the set
+$$R = \bigl\{(-1, 4), (8, -3), (0, 0), (0,1)\bigr\}$$
+is a relation.
+\smallskip
+\item{iv)} For any set $A$, a function $f:A\to A$ is by definition a subset of $A\times A$, and hence
+is an example of a relation.
+\smallskip
+\item{v)} Let $H$ be the set of all humans, define $M\subseteq H\times H$ by setting
+$(h_1, h_2)$ if and only if $h_1$ is married to $h_2$. For instance,
+$$(\hbox{Michelle Obama}, \hbox{Barack Obama})\in M.$$
+On the other hand, it is unfortunately the case that
+$$(\hbox{Marcel Goh}, \hbox{Taylor Swift})\notin M.$$
+\medskip
+
+\medskip\boldlabel Properties of relations. Let $R\subseteq A\times A$ be a relation. $R$ is called
+\medskip
+\item{i)} {\it reflexive} if for all $a\in A$, $aRa$;
+\smallskip
+\item{ii)} {\it symmetric} if for all $a,b\in A$ with $aRb$, we also have $bRa$; and
+\smallskip
+\item{iii)} {\it transitive} if for all $a,b,c\in A$ with $aRb$ and $bRc$, we also have $aRc$.
+\medskip
+Going back to the examples we had above, the relation $L$ is transitive but neither reflexive nor symmetric,
+the relation $E$ satisfies all three properties, the relation $R$ satisfies none of them, and the relation
+$M$ is symmetric but neither reflexive nor transitive. (Convince yourself of all of these facts.)
+
+As a more involved example, let $X$ be any set with $|X|\ge 2$. On $2^X$, define a relation $R$ by
+$$(A,B)\in R \Leftrightarrow A\cap B\ne \emptyset$$
+for all $A,B\subseteq X$. Which of the three properties above does $R$ satisfy?
+
+Is it reflexive? Well, is it true that for all $A\subseteq X$, $A\cap A \ne \emptyset$? The answer
+is no, since we have $\emptyset \cap \emptyset = \emptyset$, so we have $(\emptyset, \emptyset)\notin R$.
+
+Is $R$ symmetric? Well, if $A\cap B \ne \emptyset$, then since $\cap$ is commutative, we have
+$B\cap A \ne \emptyset$ as well, so $(A,B)\in R$ implies that $(B,A)\in R$. In other words, yes, $R$ is symmetric.
+
+Is $R$ transitive? If $A\cap B\ne \emptyset$ and $B\cap C\ne \emptyset$, does that necessarily mean
+that $A\cap C\ne\emptyset$? The answer is no. Here's the proof. Since $|X|\ge 2$ we can find $x,y\in X$
+with $x\ne y$. Let $A = \{x\}$, $B = \{x,y\}$, and $C = \{y\}$. Then $A\cap B = \{x\}\ne \emptyset$,
+$B\cap C = \{y\}\ne \emptyset$, but alas $A\cap C = \emptyset$.
+
+\medskip\boldlabel Equivalence relations and classes. If $R$ is reflexive, symmetric, and transitive, then we say
+that $R$ is an {\it equivalence relation}. Here are two examples (as an exercise, prove that both of
+these are equivalence relations).
+\medskip
+\item{i)} Let $F$ be the set of all formulas in propositional logic with the variables $p$, $q$, and $r$,
+and operations $\neg$, $\wedge$, and $\vee$. The relation $\equiv$ defined by $f_1 \equiv f_2$
+if and only if $f_1$ and $f_2$ have the same truth table is an equivalence relation.
+\smallskip
+\item{ii)} On the set $\RR^2$, define the relation $R \subseteq \RR^2\times \RR^2$ by letting
+$(x,y) R (w,z)$ if and only if $\sqrt{x^2 + y^2} = \sqrt{w^2 + z^2}$. This is also an equivalence relation
+\medskip
+If $R$ is an equivalence relation, often we shall write $x\sim y$ to mean $xRy$. Sometimes we might
+even just say that $\sim$ {\it is} the equivalence relation.
+
+Let $R\subseteq A\times A$ be an equivalence relation. Define the {\it equivalence class} of $a\in A$
+to be the set
+$$[a] = \{b\in A : a\sim b\}.$$
+This is the set of all $b\in A$ that are related to $a$.
+
+Take for instance the example $F$ above of all formulas under the relation $\equiv$. The equivalence class
+of the formula $p\vee\neg p$ is the set of all formulas whose truth table contains only $1$s, that is,
+the set of all tautologies.
+And for the example $R$ above, the
+equivalence class of the point $(1,3)$ is the set
+$$\bigl[ (1,3)\bigr] = \bigl\{ (x,y)\in \RR^2 : \sqrt\{10\} = \sqrt{x^2 + y^2}\bigr\};$$
+that is, the circle of radius $\sqrt{10}$ centred about the origin.
+
+We have the following proposition concerning equivalence relations.
+
+\proclaim Proposition \advthm. Let $R$ be an equivalence relation on $A$. Then
+\medskip
+\item{i)} for all $x\in A$, $x\in [x]$;
+\smallskip
+\item{ii)} for all $x,y\in A$, $x\sim y$ if and only if $[x] = [y]$; and
+\smallskip
+\item{iii)} for all $x,y\in A$, $x\not\sim y$ if and only if $[x] \cap [y] =  \emptyset$.
+\medskip
+
+\proof Let $x\in A$. Since $R$ is reflexive, $x\sim x$, so we have $x\in [x]$. This proves (i).
+
+For part (ii), first we prove the ``only if'' direction. Let $x,y\in A$ be such that $x\sim y$. To prove that
+$[x]\subseteq [y]$, we let $z\in [x]$; so $x\sim z$. By symmetry, we have $z\sim x$, and this
+combined with $x\sim y$ allow us to deduce that $z\sim y$, by transitivity. Then by symmetry again,
+we have $y\sim z$, so $z\in [y]$. {\it Mutatis mutandis}, i.e., by swapping the roles of $x$ and $y$,
+we also have $[y]\subseteq [x]$. Hence $[x] = [y]$.
+
+Now for the ``if'' direction of (ii). Let $x,y\in A$ be such that
+$[x] = [y]$. By (i), we have $x\in [x]$, but
+then this means that $x\in [y]$. By definition of $[y]$, this means that $y\sim x$, so $x\sim y$
+by symmetry.
+
+On to part (iii). We prove both implications by contraposition (that is, we negate both sides of the
+statement). Let $x,y\in A$ be such that $[x]\cap [y] \ne \emptyset$. This means there is some $z\in A$ such
+that $z\in [x]$ and $z\in [y]$; so $x\sim z$ and $y\sim z$. By symmetry, $z\sim y$,
+so by transitivity, we have $x\sim y$.
+
+On the other hand, suppose that $x,y\in A$ satisfy $x\sim y$.
+By (i) again, we have $[x] = [y]$, so $[x] \cap [y] = [x] \ne \emptyset$,
+where we know that $[x]\ne \emptyset$ because it contains at least $x$.\slug
+
+\noindent
+\datestamp{11}{} \hskip\parindent Let $A$ be a set. A {\it partition} of $A$ is a set $P$ of subsets
+of $A$ (i.e., $P\subseteq 2^A$), such that
+\medskip
+\item{i)} for all $x\in A$ there exists $S\in P$ such that $x\in S$;
+\smallskip
+\item{ii)} for all $S_1, S_2\in P$ with $S_1\ne S_2$, the intersection $S_1\cap S_2$ is empty; and
+\smallskip
+\item{iii)} $\emptyset\notin P$.
+\medskip
+Here are some examples.
+\medskip
+\item{i)} Let $E$ be the set of all even integers and $F$ the set of all odd integers. Then $\{E,F\}$
+is a partition of $\ZZ$.
+\smallskip
+\item{ii)} The set $\bigl\{ (-\infty,0), \{0\}, (0,\infty)\bigr\}$ is a partition of $\RR$.
+\medskip
+If $\sim$ is an equivalence relation on a set $A$, then we can define the {\it quotient}
+of $A$ by $\sim$ as the set of all equivalence classes of $A$ under $\sim$. We denote this set by
+$$A/\sim = \bigl\{ [x] : x\in A\bigr\}.$$
+We use the previous proposition to prove that equivalence relations give rise to partitions.
+
+\proclaim Proposition \advthm. Let $A$ be a set and $\sim$ an equivalence relation on $A$.
+Then $A/\sim$ is a partition of $A$.
+
+\proof By part (i) the previous proposition, every $x\in A$ belongs to the equivalence class $[x]$.
+Then, by part (ii) of the previous proposition, we know that for any equivalence classes
+$[x]$ and $[y]$ such that $[x]\ne [y]$, we must have $x\not\sim y$, and by part (iii) of the previous
+proposition, we deduce that $[x]\cap [y]=\emptyset$. This shows that $A/\sim$ satisfies the second
+part of the definition of partition. Lastly, we note that $\emptyset\notin A/\sim$, since
+every element of $A/\sim$ is equal to $[x]$ for some $x\in A$, and must thus contain at least
+the element $x$.\slug
+
+Let us now revisit the examples of equivalence relations from last class, and see what partitions
+they give rise to. In the example $F$ of propositional formulas with variables $p$, $q$, and $r$,
+under the equivalence relation $\equiv$, the set of equivalence classes is the set of all possible
+truth tables on three variables. Each such truth table has $8$ rows, so there are $2^8$ equivalence
+classes. In other words, $|F/\equiv| = 2^8$.
+
+How about the relation $R$ defined on $\RR^2$ where
+$$(x,y)\sim (w,z) \qquad\hbox{if and only if}\qquad \sqrt{x^2+y^2} = \sqrt{w^2 + z^2}?$$
+Well, each $\bigl[(x,y)\bigr]$ is the circle of radius $\sqrt{x^2 + y^2}$ centred
+around the origin, so $\RR^2/\sim$ is the set of all circles in the plane centred at $(0,0)$.
+
+\vfill\eject
+\begingroup\headline{\hfil}\footline{\hfil}
+\vglue160pt
+\centerline{\titlefont II. NUMBER THEORY}
+\vskip220pt
+\bigskip
+\begingroup\obeylines\eightssi
+\hfill Die Mathematik ist die K\"onigin der Wissenschaften
+\hfill und die Zahlentheorie ist die K\"onigin der Mathematik.
+\eightss
+\smallskip
+\hfill --- {\eightssi Attributed to} C. F. GAUSS {\eightssi (1777--1855)}
+\endgroup%\obeylines
+\bigskip\goodbreak
+\vfill\eject
+
+\advsect Division
+
+Let $a,b\in \ZZ$. We say that $a$ {\it divides} $b$ (or $b$ is a {\it multiple} of $a$, or $b$ is {\it divisible}
+by $a$, or $a$ is a {\it factor} of $b$) if there exists $n\in \ZZ$ such that $b=na$. In this
+case we write $a\divides b$. For example, $2\divides 10$, since $10 = 5\cdot 2$, but
+$3$ does not divide $10$, since there does not exist $n\in \ZZ$ such that $10 = 3n$.
+This defines a relation on $\ZZ$.
+
+Note that for all $n\in \ZZ$, $n\divides 0$, since $0 = 0\cdot n$. We also have
+$1\divides n$ for all $n\in \ZZ$< since $n = n\cdot 1$. It is true that $0\divides 0$, since we
+have, say, $0 = 1\cdot 0$, but for all nonzero $n\in \NN$, $0$ does not divide $n$, since all multiples
+of $0$ equal $0$ (and thus cannot equal $n$). Further properties of the ``divides'' relation
+are given by the next proposition.
+
+\proclaim Proposition \advthm. For all $a,b,c,d\in \ZZ$,
+\medskip
+\item{i)} if $a\divides b$, then $a\divides bc$;
+\smallskip
+\item{ii)} if $a\divides b$ and $a\divides c$, then $a\divides (b+c)$;
+\smallskip
+\item{iii)} if $a\divides b$ and $b\divides c$, then $a\divides c$;
+\smallskip
+\item{iv)} if $a\divides b$ and $b\ne 0$, then $|a|\le |b|$; and
+\smallskip
+\item{v)} if $a\divides b$ and $b\divides a$, then $|a| = |b|$.
+\medskip
+
+\proof We leave parts (i) and (ii) as exercises to the reader.
+
+For part (iii), if $a\divides b$ and $b\divides c$,
+then there are integers $k$ and $l$ such that $b = ka$ and $c = lb$. Then $c = kla$, so $a\divides c$
+(since $kl$ is also an integer).
+
+For part (iv), suppose that $a\divides b$ and $b\ne 0$. Then $b = ka$ for some $k\in \ZZ$,
+and $k\ne 0$ since $b\ne 0$. This means that $|b| = |k|\cdot |a|$, but $|k|\ge 1$,
+so $|b|\ge |a|$.
+
+For part (v), assume that $a\divides b$ and $b\divides a$. If $a\ne 0$ and $b\ne 0$, then we may apply
+part (iv) twice to get $|a|\le |b|$ and $|b|\le |a|$, which together imply $|a| = |b|$. If $b=0$,
+then $0\divides a$ so $a = 0$ as well, and $|a| = |b|$. Likewise, if $a = 0$, then $0\divides b$,
+so $b=0$ and in this case as well, $|a| = |b|$.\slug
+
+In grade school, you learned how to divide an integer by another one, obtaining a quotient and a remainder.
+We state this as a theorem. Its proof (which we shall consider outside the scope of the course,
+but which we include as optional reading for those interested)
+relies on the well-ordering principle (which we've used
+already in these notes)
+and the {\it Archimedean property} of $\RR$, which states that for every $x\in \RR$ there
+exists $n\in \NN$ such that $n > x$. (One can learn the proof of the Archimedean property from an introductory
+course in analysis, e.g., {\mc MATH} 242/254.)
+
+\parenproclaim Theorem {\advthm} (Division algorithm). Let $a,b\in \ZZ$ with $b> 0$. Then there
+exist unique integers $q$ and $r$ such that $a = bq+r$ and $0\le r\le |b|$.
+
+\noindent\llap{*}{\it Proof.}\enspace First we show that such integers $q$ and $r$
+exist. If $b\divides a$ then $a = kb$ for some $k\in \ZZ$,
+and we can set $q = k$ and $r = 0$.
+
+If $a$ is not divisible by $b$, then consider the numbers
+$$ \ldots, a - 3b, a-2b, a-b, a, a+b, a+2b, a+3b, \ldots.$$
+Let $S$ be the set of these integers that are positive. Symbolically, we have
+$$ S = \{a-kb : k\in \ZZ\ \hbox{and}\ a-kb\ge 0\}.$$
+By the Archimedean property, there is some $n\in \NN$ such that $n> -a$, which
+implies that $nb \ge n > -a$ (here we use the fact that if $b>0$ and $b$ is an integer, then
+$b\ge 1$). From this we derive $a > -nb$, and hence $a+nb = a - (-n)b > 0$. This shows
+that $S$ is nonempty.
+
+Since $S$ is a nonempty subset of $\NN$, by the well-ordering principle it has a least
+element, call it $r$. By definition of $S$, there must be some integer $q$ such that
+$r = a-qb$, so $a = bq + r$. We now claim that $0<r<b$.
+
+We know that $r> 0$, since all elements of $S$ are positive by definition. Suppose,
+for a contradiction, that $r\ge b$. Then $a - bq = r \ge b$, and so
+$$0 \le r-b = a - qb - b = a-(q+1)b.$$
+Since $q+1$ is an integer, by definition of $S$, either $r-b$ is an element of $S$,
+or $r-b = 0$. Since $r$ was defined to be the minimal element of $S$, it cannot be the
+case that $r-b$ is in $S$. So $r-b = 0$. But this means that $0 = a-(q+1)b$; that is,
+$a = (q+1)b$, contradicting our assumption that $b$ does not divide $a$. The contradiction
+allows us to conclude that $0<r<b$ (in the case that $b$ does not divide $a$). In general,
+we have shown that $0\le r<b$.
+
+\endgroup%\headline{\hfil}\footline{\hfil}
+Lastly, we need to prove that $q$ and $r$ are uniquely determined by the integers $a$ and
+$b$. Suppose that
+$$a = bq_1 + r_1\qquad\hbox{and}\qquad 0\le r_1 < b$$
+and
+$$a = bq_2 + r_2\qquad\hbox{and}\qquad 0\le r_2 < b,$$
+for some integers $q_1$, $q_2$, $r_1$, and $r_2$. We shall show that $q_1 = q_2$
+and $r_1 = r_2$. Well, suppose that $r_1\ne r_2$, for a contradiction. Without loss
+of generality we can assume that $r_1 < r_2$. Then subtracting the two equations, we obtain
+$$ 0 = a-a = (bq_1 + r_1) - (bq_2 + r_2) = b(q_1-q_2) + (r_1 - r_2).$$
+This means that
+$$r_2 - r_1 = b(q_1 - q_2),$$
+so we find that $b\divides (r_2 - r_1)$. By part (iv) of the previous proposition,
+we obtain $|b| \le |r_2-r_1|$, and we can simply write $b \le r_2 -r_1$, since both of
+these quantities are positive. But this is a contradiction, since
+$$ 0\le r_1 < r_2 < b,$$
+yields $r_2 - r_1 < b$. The contradiction shows that that $r_2 = r_1$. Substituting this
+into the relation $r_2 - r_1 = b(q_1 - q_2)$, we get $0 = b(q_1 - q_2)$ and conclude that
+$q_1 - q_2 = 0$, since $b>0$.\slug
+
+Let $a$ and $b$ be integers, not both zero. Their {\it greatest common divisor}, $\gcd(a,b)$
+is defined to be the greatest positive integer $d$ such that $d\divides a$ and $d\divides b$.
+Note that $\gcd(0,0)$ is not defined, since all positive integers $d$ satisfy $d\divides 0$.
+On the other hand $\gcd(x,0)$ is simply $|x|$, and $\gcd(x,1) = 1$. Lastly, we don't
+need to worry about negative signs when computing greatest common divisors; i.e.,
+$\gcd(\pm x, \pm y) = \gcd\bigl(|x|, |y|\bigr)$.
+
+Now we ask ourselves, How do we compute greatest common divisors in general? The answer
+lies in one of the oldest algorithms known to humankind. It appears in Euclid's {\sl Elements},
+written around 300~{\sc b.c.}
+
+\algbegin Algorithm E (Euclid's algorithm). Given two nonnegative integers $a$ and $b$, not both zero,
+this algorithm outputs $\gcd(a,b)$.
+\algstep E1. If $b=0$, then output $a$ and terminate the algorithm.
+\algstep E2. Since $b\ne 0$, by the division algorithm we my write $a = qb+r$, where $0\le r < b$.
+Set $a\gets b$, $b\gets r$, and return to step E1.\slug
+
+The algorithm will eventually terminate, since the stopping criterion is that $b$ be equal to $0$,
+and in step E2 we replace $b$ with a number that is strictly closer to $0$. But will it terminate
+with the correct answer? Well, we know step E1 is correct,
+because of our earlier observation that $\gcd(a,0) = a$ (whenever $a$
+is positive). On the other hand, it is not at all evident that step E2 will eventually output
+$\gcd(a,b)$, since we actually overwrite the values of $a$ and $b$ in the step! The following lemma
+clarifies the situation.
+
+\proclaim Lemma \advthm. Let $a,b,q,r\in \ZZ$ be integers such that $a = qb+r$. Then
+$\gcd(a,b) = \gcd(b,r)$.
+
+\proof We shall show that for any $d\in \ZZ$,
+$$ d\divides a\ \hbox{and}\ d\divides b\qquad\hbox{if and only if}\qquad
+d\divides b\ \hbox{and}\ d\divides r.$$
+For the forward implication, suppose that $a = kd$ and $b=ld$ for some $k,l\in \ZZ$. Substituting
+this into the identity $a=qb+r$ yields $kd = ldq+r$, whence $r = c(k-lq)$, so we conclude that $c\divides r$.
+(This is because $k-lq\in \ZZ$.)
+
+For the reverse implication, suppose that $b = ld$ and $r = md$ for some $k,l\in \ZZ$. Substituting
+this into $a=qb+r$, we have $a = ldq + md$, so $a = d(lq+m)$, which means that $c\divides a$.
+
+We have proved that $a$ and $b$ have the same common factors as $b$ and $r$, so they must have the
+same greatest common divisor.\slug
+
+Now that we are secure in the fact that Euclid's algorithm will indeed terminate with the correct output,
+let us now see it in action. Suppose we want to find $\gcd(30,112)$. We perform successive divisions
+replacing the pair $(a,b)$ with a new pair $(b,r)$ each time:
+$$\eqalign{
+30 &= 0\cdot 112 + 30 \cr
+112 &= 3\cdot 30 + 22 \cr
+30 &= 1\cdot 22 + 8 \cr
+22 &= 2\cdot 8 + 6 \cr
+8 &= 1\cdot 6 + 2 \cr
+6 &= 3\cdot 2 + 0\cr
+}$$
+We stop once the remainder $r$ equals $0$, and the answer is $\gcd(b,r) = \gcd(b,0) = b$. (So in the
+example above, the final answer is $2$.)
 

### 2024-09-25 teaching/math240f24/draft.tex
lecture sept 24

@@
 
-\medskip\boldlabel\datestamp{8}{} Injective and surjective functions.
+\medskip\boldlabel\datestamp{24}{ix} Injective and surjective functions.
 The {\it range} or {\it image} of a function $f:A\to B$ is
@@
 
-\proclaim Theorem \advthm. Let $a_1, a_2, \ldots, a_m$ be a finite sequence (repeats allowed) of
+\proclaim Theorem \advthm. Let $a_1, a_2, \ldots, a_n$ be a finite sequence (repeats allowed) of
 real numbers. Let
-$$a = {1\over m}\sum_{i=1}^m a_i$$
+$$a = {1\over n}\sum_{i=1}^n a_i$$
 be the average value of the sequence and let $m$ be the maximum value the sequence attains. Then $m \ge a$.
@@
 \proof We have
-$$a = {1\over m}\sum_{i=1}^m a_i \le {1\over m} \sum_{i=1}^m m = {m^2\over m} = m.\noskipslug$$
+$$a = {1\over n}\sum_{i=1}^n a_i \le {1\over n} \sum_{i=1}^n m = {mn\over n} = m.\noskipslug$$
 
@@
 
-\proof For $1\le i \le n$, let $r_i$ be the number of $a\in A$
+\proof Enumerate $B = \{b_1, b_2, \ldots, b_n\}$. For $1\le i \le n$, let $r_i$ be the number of $a\in A$
 such that $f(a) = b_i$. This is a sequence that adds up to $m$, since every $a$ in $A$ maps to
@@
 The proof that $f\bigl(g(A)\bigr) = A$ is similar. Thus $g$ is the inverse of $f$.\slug
+\bye
 

### 2024-09-27 teaching/math240f24/draft.tex
sep 26

@@
 \hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
-%\smallskip
-%\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
 %\smallskip
@@
 The proof that $f\bigl(g(A)\bigr) = A$ is similar. Thus $g$ is the inverse of $f$.\slug
-\bye
 
@@
 
-\datestamp{9}{}
+\datestamp{26}{ix}
 Earlier, we defined the cardinality of a set to be the number of elements it contains. What, then,
@@
 
-So we can find subset of $\NN$ equipotent with it. It turns out we can also find supersets of $\NN$ with
+So we can find subsets of $\NN$ equipotent with it. It turns out we can also find supersets of $\NN$ with
 the same property.
@@
 
-To show that $f$ is injective, there are four cases. Note first that if $n$ is even, then $f(n)\ge 0$,
+Note first that if $n$ is even, then $f(n)\ge 0$,
 and if $n$ is odd, then $f(n) < 0$. So if $f(m) = f(n)$ for some $m,n\in \NN$, then $f(m)$ and $f(n)$ must
@@
 $$f(a_i, b_j) = 2^i3^j.$$
-We show that this defines an injective function. Suppose that $f(a_i, b_j) = f(a_{i'}, f(b_{j'}).$
+We show that this defines an injective function. Suppose that $f(a_i, b_j) = f(a_{i'}, b_{j'}).$
 Then $2^i3^j = 2^{i'} 3^{j'}$, and by the uniqueness of prime factorisations in the Fundamental
@@
 This string $s$ cannot equal $f(m)$ for any $m\in \NN$, since given an arbitrary $m$,
-the $m$th bit of $f(m)$ is $a_{m,m}$, whereas the $m$th bit of $s$ is $1-a{m,m}$. Hence there is some
+the $m$th bit of $f(m)$ is $a_{m,m}$, whereas the $m$th bit of $s$ is $1-a_{m,m}$. Hence there is some
 $s\in A$ such that $f(m)\ne s$ for all $m\in \NN$. So $f$ is not surjective.
@@
 We conclude that $|A| \ne |\NN|$.\slug
+
+The proof above was published by Georg Cantor in 1891, and hence is known as {\it Cantor's diagonal argument}.
+The technique has since been used to prove many other things.
 
@@
 $\bigl| [0,1]\bigr| \ne |\NN|$.
+\bye
 

### 2024-10-01 teaching/math240f24/draft.tex
lecture october 1

@@
 \hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
-%\smallskip
-%\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
 
@@
 $\bigl| [0,1]\bigr| \ne |\NN|$.
-\bye
 
@@
 
-\datestamp{10}{} A {\it relation} on a set $X$ is a subset $R\subseteq X\times X$. If $(a,b)\in R$
+\datestamp{01}{x} A {\it relation} on a set $X$ is a subset $R\subseteq X\times X$. If $(a,b)\in R$
 we write $aRb$ and say ``$a$ is related to $b$.'' Here are some examples.
@@
 equivalence class of the point $(1,3)$ is the set
-$$\bigl[ (1,3)\bigr] = \bigl\{ (x,y)\in \RR^2 : \sqrt\{10\} = \sqrt{x^2 + y^2}\bigr\};$$
+$$\bigl[ (1,3)\bigr] = \bigl\{ (x,y)\in \RR^2 : \sqrt{10}= \sqrt{x^2 + y^2}\bigr\};$$
 that is, the circle of radius $\sqrt{10}$ centred about the origin.
@@
 
-\proof Let $x\in A$. Since $R$ is reflexive, $x\sim x$, so we have $x\in [x]$. This proves (i).
+\proof Let $x\in A$. Since $R$ is reflexive, $x\sim x$, so we have $x\in [x]$. This proves part (i).
 
@@
 On the other hand, suppose that $x,y\in A$ satisfy $x\sim y$.
-By (i) again, we have $[x] = [y]$, so $[x] \cap [y] = [x] \ne \emptyset$,
-where we know that $[x]\ne \emptyset$ because it contains at least $x$.\slug
-
-\noindent
-\datestamp{11}{} \hskip\parindent Let $A$ be a set. A {\it partition} of $A$ is a set $P$ of subsets
+By (ii), we have $[x] = [y]$, so $[x] \cap [y] = [x] \ne \emptyset$,
+where we know that $[x]\ne \emptyset$ because it contains at least $x$ (again, using (i)).\slug
+
+Let $A$ be a set. A {\it partition} of $A$ is a set $P$ of subsets
 of $A$ (i.e., $P\subseteq 2^A$), such that
@@
 \medskip
+\bye
 If $\sim$ is an equivalence relation on a set $A$, then we can define the {\it quotient}
 of $A$ by $\sim$ as the set of all equivalence classes of $A$ under $\sim$. We denote this set by
-$$A/\sim = \bigl\{ [x] : x\in A\bigr\}.$$
+$$A/{\sim} = \bigl\{ [x] : x\in A\bigr\}.$$
 We use the previous proposition to prove that equivalence relations give rise to partitions.
@@
 \proclaim Proposition \advthm. Let $A$ be a set and $\sim$ an equivalence relation on $A$.
-Then $A/\sim$ is a partition of $A$.
+Then $A/{\sim}$ is a partition of $A$.
 

### 2024-10-03 teaching/math240f24/draft.tex
Lecture 3 october

@@
 
-\vskip \medskipamount \hbox to\hsize {\relax \unhcopy \strutbox \cont
+\vskip \medskipamount \hbox to\hsize
+{\relax \unhcopy \strutbox \cont
 {{\tenssbx I. Foundations \kern 3pt}}{2}}
@@
 \hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
+\medskip
+\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
+{{\tenssbx II. Number theory \kern 3pt}}{41}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
 
@@
 
-\edef\propsqrttwo{\the\thmcount}
-\proclaim Proposition \advthm. The number $\sqrt 2$ is irrational.
+\edef\thmsqrttwo{\the\thmcount}
+\proclaim Theorem \advthm. The number $\sqrt 2$ is irrational.
 
@@
 If it is rational, then we can set $a = b = \sqrt 2$, which we already know to be irrational, by
-Proposition~{\propsqrttwo}.
+Theorem~{\thmsqrttwo}.
 
@@
 
+\edef\thmfta{\the\thmcount}
 \parenproclaim Theorem {\advthm} (Fundamental Theorem of Arithmetic). Every positive
 integer $n\ge 2$ can be factored into a product
-$${p_1}^{v_1} {p_2}^{v_2}\cdots {p_n}^{v_n},$$
-where $n\ge 0$ is an integer and, for $1\le i\le n$, $p_i$ is a prime and $v_i$ is a positive integer.
+$${p_1}^{v_1} {p_2}^{v_2}\cdots {p_m}^{v_m},$$
+where $m\ge 0$ is an integer, $p_1,p_2,\ldots,p_m$ are distinct primes, and
+$v_1,v_2,\ldots,v_m$ are positive integers.
 This factorisation is unique up to the order of the primes.\slug
@@
 \medskip
-\bye
+
+\noindent\datestamp{03}{x} \hskip\parindent
 If $\sim$ is an equivalence relation on a set $A$, then we can define the {\it quotient}
@@
 $$A/{\sim} = \bigl\{ [x] : x\in A\bigr\}.$$
-We use the previous proposition to prove that equivalence relations give rise to partitions.
+We use the previous proposition to prove that quotients of sets by equivalence relations
+are partitions.
 
@@
 
+\medskip\boldlabel Euclid's algorithm.
 Now we ask ourselves, How do we compute greatest common divisors in general? The answer
@@
 replacing the pair $(a,b)$ with a new pair $(b,r)$ each time:
+\edef\eqeuclid{\the\eqcount}
 $$\eqalign{
@@
 6 &= 3\cdot 2 + 0\cr
-}$$
+}\adveq$$
 We stop once the remainder $r$ equals $0$, and the answer is $\gcd(b,r) = \gcd(b,0) = b$. (So in the
@@
 
+\medskip\boldlabel B\'ezout's identity.
+The following theorem allows us to express the greatest common divisor as a linear combination
+of the two integers in question.
+
+\edef\thmbezout{\the\thmcount}
+\parenproclaim Theorem {\advthm} (B\'ezout's identity). Let $a$ and $b$ be nonzero integers
+with greatest common divisor $\gcd(a,b)$. Then there exist integers $s$ and $t$ such that
+$$\gcd(a,b) = sa+tb.$$
+Moreover, $\gcd(a,b)$ is the least positive integer that can be expressed as an
+integer linear combination of $a$ and $b$.
+
 \bye
+
+\proof Let
+$$S = \{ s'a+t'b : s',t'\in \ZZ,\ ax+by > 0\}.$$
+This set is nonempty, since if $a$ is negative then $(-1)a+0b\in S$ and if $a$ is positive,
+then $1a+0b\in S$. Since $S$ is a nonempty set of positive integers, it has
+a least element, by the well-ordering principle. Call this integer $d = sa+tb$ (for some
+specific choices of $s,t\in \ZZ$);
+the claim is that
+$d = \gcd(a,b)$.
+
+By the division algorithm, we may write
+$$a = dq+r$$
+where $q$ and $r$ are integers with $0\le r<d$. But we can write
+$$\eqalign{
+r &= a-qd \cr
+&= a - q(sa+tb) \cr
+&= (1-qs)a + (qt)b,\cr
+}$$ so $r\in S\cup \{0\}$. But $r < d$, so if $r\in S$, then $d$ would not be the smallest
+element of $S$. So we must have $r = 0$. This implies that $d$ is a divisor of $a$.
+Repeating this argument with $b$ instead of $a$, we find that $d$ divides $b$ as well.
+
+We have shown that $d$ is a common divisor of $a$ and $b$. It remains to show that it is the greatest
+one. That is, we must show that if $c\divides a$ and $c\divides b$, then $c\le d$. Well, if
+$a = kc$ and $b=lc$, then the identity
+$$d = sa+tb = skc + tlc = (sk+tl)c,$$
+shows that $d$ is a multiple of $c$ as well. Since $d>0$, this means that $c\le d$.\slug
+
+This theorem is named for \'Etienne B\'ezout, who proved an analogous result (with polynomials
+instead of integers) in 1779, but the result above for integers has been known since at least
+the 1600s.
+
+To actually find the integers $s$ and $t$ such that $\gcd(a,b) = sa + tb$, we
+first perform the Euclidean algorithm, keeping track of all our intermediate steps. Then we combine
+all the information from each step to work out what $s$ and $t$ are.
+For example, in the earlier example showing that $2 = \gcd(112,30)$, we start with
+$$ 2 = 8- 1\cdot 6,$$
+which is (a rearrangement of) the fifth line of~\refeq{\eqeuclid}.
+Then the fourth line of~\refeq{\eqeuclid} says that $6=22-2\cdot 8$, so
+$$ 2 = 8 - 1\cdot (22-2\cdot 8) = 8 - 22 + 2\cdot 8 = 3\cdot 8 - 22.$$
+The third line of~\refeq{\eqeuclid} tells us that $8 = 30-1\cdot 22$, which gives
+$$ 2 = 3\cdot (30-1\cdot 22) - 22 = 3\cdot 30 - 3\cdot 22 - 22 = 3\cdot 30 - 4\cdot 22.$$
+We now have the number $30$ appearing in the expression, we just need to get rid of the $22$
+and replace it with $112$. To do this we use the second line of~\refeq{\eqeuclid}, which
+says that $22 = 112-3\cdot 30$. We end up with
+$$ 2= 3\cdot 30 - 4(112-3\cdot 30) = 3\cdot 30 - 4\cdot 112 + 12\cdot 30 = (-4)\cdot 112 + 15\cdot 30.$$
+So in the case that $a = 112$ and $b=30$, we have $d = 2$, $s = -4$, and $t = 15$.
+
+We now summarise this section on B\'ezout's identity with a little scenario.
+Imagine a robot that lives on the line of integers. It starts at the point $0$
+and can hop in steps of $a$ or $b$ (in either direction). Theorem~{\thmbezout} tells us
+that the point $d = \gcd(a,b)$ is reachable by the robot. The next proposition characterises
+the set of {\it all} points that the robot can get to.
+
+\proclaim Proposition \advthm. Let $a$ and $b$ be nonzero integers. The set
+$$X = \{ s'a + t'b : s',t'\in \ZZ\}$$
+is exactly the set of multiples of $d = \gcd(a,b)$.
+
+\proof By B\'ezout's identity, there exist integers $s$ and $t$ such that $d = sa+tb$.
+First let $n\in \ZZ$ be a multiple of $d$. Then there is $k\in \ZZ$ such that $n=kd$,
+and we have
+$$ n = kd = d(sa+tb) = (ds)a + (dt)b,$$
+which means that $n\in X$ (since $ds$ and $dt$ are both integers).
+
+Conversely, suppose that $n\in X$, so $n = s'a + t'b$ for some $s',t'\in \ZZ$. Then since
+$d$ divides $a$ and $d$ divides $b$, we can write $a = ld$ and $b=md$ for some integers $l,m\in \ZZ$.
+So we have
+$$ n= s'a + t'b = s'ld + t'md = (s'l+t'm)d,$$
+which shows that $d\divides n$, since $s'l+t'm$ is an integer.\slug
+
+As a corollary, if $\gcd(a,b) = 1$, then it is possible for the robot to reach every integer!
+The situation in which $\gcd(a,b) = 1$ is very special, so much so that we have a name for it.
+We say that integers $a$ and $b$ are {\it relatively prime} or {\it coprime} if $\gcd(a,b) = 1$.
+By B\'ezout's identity, $a$ and $b$ are relatively prime if and only if
+there are integers $s$ and $t$ such that $1 = sa + tb$. This gives a very quick and easy way to
+check if certain numbers are relatively prime. For example, we have the following proposition.
+
+\proclaim Proposition \advthm. For all integers $n>1$, $n$ and $n+1$ are relatively prime.
+
+\proof We have $1= 1(n+1) + (-1)n$.\slug
+
+\advsect Primes
+
+An integer $p$ is {\it prime} if $p\ge 2$ and for all $d\in \NN$ with $d\divides p$,
+we either have $d=1$ or $d = p$. An integer $n$ is {\it composite}
+if $n\ge 2$ and $n$ is not prime. (By negating the definition of prime, we
+see that $n\ge 2$ is composite if and only if there exist $a,b\in \{2,\ldots,n-1\}$ such that
+$n = ab$.) Note that the integers $0$ and $1$ are neither prime nor composite.
+
+The following theorem gives another characterisation of prime numbers.
+
+\proclaim Theorem \advthm. An integer $p$ with $p\ge 2$ is prime if and only if
+for all $a,b\in \NN$, $p\divides ab$ implies that $p\divides a$ or $p\divides b$.
+
+\proof First we prove the forward implication.
+Suppose that $p$ is prime and let $a,b\in \NN$ be such that $p\divides ab$. So
+there exists an integer $k$ such that $ab = kp$.
+Consider $\gcd(a,p)$. Since the
+only divisors if $p$ are $1$ and $p$, this must be $1$ or $p$. If it is $p$,
+then $p\divides a$ and we are done. So we restrict our attention to the case that
+$\gcd(a,p) =1$, and our goal is to prove $p\divides b$. By B\'ezout's identity,
+there are integers $s$ and $t$ such that $1= sa+tp$, so multiplying both sides by $b$,
+we have
+$$ b =bsa+btp = skp + btp = (sk+bt)p,$$
+so $p\divides b$.
+
+We prove the reverse implication by contraposition.
+Now suppose that $p$ is composite. So there exist integers $2\le a,b\le p-1$
+such that $p=ab$. We want to show that $p$ does not divide $a$ and $p$ does not divide $b$.
+We shall do this by showing that $u = a/p$ and $v = b/p$ are both not integers. Well
+since $p = ab = upb$, by dividing through by $p$ we arrive at
+$1 = ub$, and since $b\ge 2$, $u = 1/b$ is between $0$ and $1$. This shows that $p$ does
+not divide $a$. On the other hand, since $p = ab = apv$, we have $1 = av$,
+and since $a\ge 2$, this means that $v = a/1$ is between $0$ and $1$. Hence
+$p$ does not divide $b$.\slug
+
+To illustrate that $p$ really does have to be prime for this theorem to hold, consider
+$p=6$, $a = 2$ and $b=15$. We have $6 \divides 30 = 2\cdot 15$, but $6$ divides neither $2$ nor $15$.
+By induction, the theorem extends to arbitrary finite products.
+
+\proclaim Corollary \advthm. Let $p$ be prime and $n$ be a positive integer. If $a_1,a_2,\ldots,a_n$
+are integers such that $p \divides a_1a_2\cdots a_n$, then $p\divides a_i$ for some
+$1\le i\le n$.
+
+\proof By induction on $n$. When $n=1$, there is nothing to prove,
+for if $p\divides a_1$, then $p\divides a_i$ for $i = 1$.
+
+Now suppose the statement holds for $n$. Assume that $p\divides a_1a_2\ldots a_n a_{n+1}$.
+By setting $a = a_1a_2\cdots a_n$ and $b=a_{n+1}$, we have $p\divides ab$, so by the previous theorem,
+either $p\divides a$ or $p\divides b$. If $p\divides b$, then $p\divides a_i$ for $i=n+1$,
+and if $p\divides a$, then $p\divides a_1\cdots a_n$, so by the induction hypothesis,
+$p\divides a_i$ for some $1\le i\le n$.\slug
+
+Back in Section~6, we used the prime factorisation of integers in a proof, but didn't prove
+that statement itself. It's finally time to do so. Recall that we statement we used
+is that any integer $n$ can be factored into a product
+$$n = {p_1}^{v_1} {p_2}^{v_2} \cdots {p_k}^{v_k}$$
+where $p_1 < p_2<\cdots<p_k$ are primes and $v_1,v_2,\ldots,v_n$ are positive integers. By renumbering
+the primes, allowing them to possibly equal one another, we have the following equivalent statement.
+
+\parenproclaim Theorem {\thmfta}$'$ (Fundamental Theorem of Arithmetic, again). Every integer
+$n\ge 2$ can be expressed as a product
+$$n = p_1p_2\cdots p_k$$
+where $p_1\le p_2\le \cdots \le p_k$ are prime numbers. Furthermore, this factorisation
+is unique.
+
+\noindent\datestamp{13}{} {\it Proof.}\enspace The proof that such a decomposition of $n$ exists is
+by (strong) induction. The base case is $n=2$. This is already a prime factorisation, since $2$ is prime.
+
+For the inductive step, let $n\ge 2$ and assume such a decomposition exists
+for all $2\le i\le n$. Now consider $n+1$. If $n+1$ is prime, then by setting $p_1 = n+1$,
+we have a prime factorisation of $n+1$ without getting out of bed. If $n+1$ is not prime,
+then $n+1 = ab$ for some integers $2\le a,b\le n$. By the induction hypothesis, $a$ and $b$
+can be factored into primes; that is, $a = p_1,\ldots,p_l$
+and $b = q_1,\ldots,q_m$ for some primes $p_1,\ldots,p_l,q_1,\ldots,q_m$. So
+$$n+1 = ab = p_1\cdots p_l q_1\cdots q_m$$
+is a factorisation of $n+1$ into primes.
+It remains to arrange the $p$s and $q$s in nondecreasing order.
+
+Now we prove that the prime factorisation of an integer is unique. In other words, $n\ge 2$
+decomposes into
+$$n = p_1p_2\cdots p_s$$
+and
+$$n = q_1q_2\cdots q_t$$
+for primes $p_1,\ldots,p_s,q_1,\ldots,q_t$,
+then $s=t$ and $p_i = q_i$ for all $1\le i\le s$. We also prove this statement by induction,
+but this time it is on the integer $s$. If $s=1$, then
+$$p_1 = n = q_1q_2\cdots q_t.$$
+Note that $t$ has to equal $1$ here, since otherwise $q_1$ and $q_2$ would both
+divide $p$ and satisfy $2\le q_1,q_2\le p_1$, contradicting the fact that $p_1$ is prime.
+So $p_1 = q_1$ and we are done.
+
+Next, suppose that the uniqueness
+statement holds for $s$ (i.e., for any integer that decomposes
+into a product of $s$ primes, the decomposition is unique) and we want to show that it holds
+for $s+1$. We assume that
+$$n = p_1p_2\cdots p_{s+1}$$
+and
+$$n = q_1q_2\cdots q_t,$$
+where $p_1\le p_2\le \cdots \le p_{s+1}$ and $q_1\le q_2\le\cdots\le q_t$. We
+have $p_{s+1}\divides n = q_1q_2\cdots q_t$, so by the preceding corollary, there
+exists $1\le i\le t$ such that $p_{s+1}\divides q_i$, and since $q_i$ is prime, this means
+that $p_{s+1} = q_i\le q_t$. Similarly, $q_t$ divides $n = q_1\ldots p_{s+1}$, so it
+divides $p_j$ for some $1\le j\le s+1$. This means that $q_t = p_j \le p_{s+1}$. Hence
+$q_t = p_{s+1}$. But
+$$p_1p_2 \cdots p_s p_{s+1} = q_1q_2 \cdots q_t,$$
+so we may divide out by $p_{s+1} = q_t$ on both sides to get
+$${n\over p_{s+1}} = p_1p_2\cdots p_s = q_1 q_2 \cdots q_{t-1}.$$
+By the induction hypothesis, the decomposition of $n/p_{s+1}$ is unique, so $t-1 = s$
+and $p_i = q_i$ for all $1\le i\le t-1$. This proves that $s+1 = t$ and
+$p_i = q_i$ for all $1\le i\le s+1$.\slug
+
+We can use the Fundamental Theorem of Arithmetic to prove the following generalisation
+of Theorem~{\thmsqrttwo}.
+
+\proclaim Theorem \advthm. Let $k$ and $n$ be positive integers. Then either
+$\root k\of n$ is an integer or it is irrational.
+
+\proof We prove this by contraposition, showing that if $\root k\of n$ is rational, then it
+must be an integer. Suppose that $\root k\of n$ is rational; so there exist integers
+$a$ and $b$ with $b\ne 0$ such that $\root k\of n = a/b$. Without loss of generality,
+we may choose $a$ and $b$ such that $\gcd(a,b) = 1$.
+Writing
+$$a = {p_1}^{v_1} {p_2}^{v_2}\cdots {p_r}^{v_r}$$
+and
+$$b = {q_1}^{w_1} {q_2}^{w_2}\cdots {q_s}^{w_s},$$
+for primes $p_1<p_2<\cdots<p_r$, primes $q_1<q_2<\cdots<q_s$, and positive integers
+$v_1,v_2,\ldots,v_r,w_1,w_2,\ldots,w_s$, all the primes $p_i$ and $q_j$ must be different,
+since if there was a prime in common, call it $p = p_i = q_j$, we would have $p\divides \gcd(a,b)$,
+contradicting the fact that $\gcd(a,b) = 1$.
+
+Taking the identity $\root k\of n = a/b$ to the power of $k$ yields
+$$n = {a^k\over b^k} = {{p_1}^{kv_1} {p_2}^{kv_2}\cdots {p_r}^{kv_r}
+\over {q_1}^{k w_1} {q_2}^{k w_2}\cdots {q_s}^{k w_s}}.$$
+Since there do not exist $1\le i\le r$ and $1\le j\le s$ such that $p_i=q_j$, this
+fraction is in reduced form. But since $n$ is an integer, that means the denominator equals $1$.
+In other words, $n= a^k$, so $\root k\of n = a$ is an integer.\slug
+
+Let's see how to use this theorem. Suppose we want to know if the number $\root 6\of 18$ is
+irrational or not. Well, since $1^6 = 1$ and $2^6 = 64$, we have $1^6 < 18 < 2^6$,
+meaning that $1<\root 6\of 18 < 2$. (This is because the function $f:\RR\to\RR$ defined
+by $f(x) = x^6$ is increasing on the interval $[1,2]$.) Hence $\root 6\of 18$ is not
+an integer, and by the theorem it must be irrational.
+
+\bye

### 2024-10-08 teaching/math240f24/draft.tex
Bezout and FTA

@@
 \def\divides{\mid}
+\def\percent{\mathrel{\%}}
 
@@
 \hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
 
@@
 \medskip\boldlabel\datestamp{05}{ix} Russell's paradox. Earlier, we said that the sets we are working with
-need to be a subset of a universe $U$, which has already been proven to be a set. We gave lots of
+need to be a subset of a universe $U$, which has already been proved to be a set. We gave lots of
 ways to make sets out of new sets, such as the union and intersection operations, etc. Starting with
@@
 $[x]$ and $[y]$ such that $[x]\ne [y]$, we must have $x\not\sim y$, and by part (iii) of the previous
-proposition, we deduce that $[x]\cap [y]=\emptyset$. This shows that $A/\sim$ satisfies the second
-part of the definition of partition. Lastly, we note that $\emptyset\notin A/\sim$, since
-every element of $A/\sim$ is equal to $[x]$ for some $x\in A$, and must thus contain at least
+proposition, we deduce that $[x]\cap [y]=\emptyset$. This shows that $A/{\sim}$ satisfies the second
+part of the definition of partition. Lastly, we note that $\emptyset\notin A/{\sim}$, since
+every element of $A/{\sim}$ is equal to $[x]$ for some $x\in A$, and must thus contain at least
 the element $x$.\slug
@@
 truth tables on three variables. Each such truth table has $8$ rows, so there are $2^8$ equivalence
-classes. In other words, $|F/\equiv| = 2^8$.
+classes. In other words, $|F/{\equiv}| = 2^8$.
 
@@
 Well, each $\bigl[(x,y)\bigr]$ is the circle of radius $\sqrt{x^2 + y^2}$ centred
-around the origin, so $\RR^2/\sim$ is the set of all circles in the plane centred at $(0,0)$.
+around the origin, so $\RR^2/{\sim}$ is the set of all circles in the plane centred at $(0,0)$.
 
@@
 
-\bye
-
-\proof Let
+\noindent\datestamp{08}{x} {\it Proof.}\enspace
+Let
 $$S = \{ s'a+t'b : s',t'\in \ZZ,\ ax+by > 0\}.$$
@@
 We now summarise this section on B\'ezout's identity with a little scenario.
-Imagine a robot that lives on the line of integers. It starts at the point $0$
+Imagine a frog that lives on a doubly-infinite line of lilypads,
+indexed by the integers. It starts at the point $0$
 and can hop in steps of $a$ or $b$ (in either direction). Theorem~{\thmbezout} tells us
-that the point $d = \gcd(a,b)$ is reachable by the robot. The next proposition characterises
-the set of {\it all} points that the robot can get to.
+that the lilypad $d = \gcd(a,b)$ is reachable by the frog. The next proposition characterises
+the set of {\it all} lilypads that the frog can get to.
 
@@
 
+\edef\thmpdivides{\the\thmcount}
 \proclaim Theorem \advthm. An integer $p$ with $p\ge 2$ is prime if and only if
@@
 
-\noindent\datestamp{13}{} {\it Proof.}\enspace The proof that such a decomposition of $n$ exists is
+\proof The proof that such a decomposition of $n$ exists is
 by (strong) induction. The base case is $n=2$. This is already a prime factorisation, since $2$ is prime.
@@
 $p_i = q_i$ for all $1\le i\le s+1$.\slug
+\bye
 
@@
 
+Next, we show that there are infinitely many primes, a theorem first proved by Euclid.
+
+\proclaim Theorem \advthm. There are infinitely many prime numbers.
+
+\proof Suppose, towards a contradiction, that there are finitely many prime numbers, call
+them $p_1, p_2, \ldots, p_m$. Let $n = p_1p_2\cdots p_m$, and consider the integer $n+1$. Either
+it is prime or it is not. If $n+1$ is prime, then we already have a contradiction, since $n+1 > p_i$
+for all $1\le i\le m$, and we assumed that the $p_1,\ldots,p_m$ were all the primes. If $n+1$
+is not prime, then it is divisible by some prime in our list, call it $p_i$. Hence we can write
+$n = kp_i$ for some integer $k$, and we have
+$$ k p_i = n+1 = p_1 p_2\cdots p_m + 1.$$
+Rearranging this a bit, we have
+$$ 1 = kp_i - p_1 p_2\cdots p_m = p_i(k - p_1 p_2 \cdots p_{i-1} p_{i+1} \cdots p_m);$$
+in other words, $p_i$ divides $1$. This is a contradiction as $1$ is not divisible by any integer
+greater than $1$.\slug
+
+Though the set $P$ of prime numbers is infinite, it does sort of get ``sparser'' as one heads off
+towards infinity. This is quantified by the following theorem, proved independently
+in~1896 by Jacques Hadamard and Charles Jean de la Vall\'ee Poussin.
+
+\parenproclaim Theorem {\advthm} (Prime number theorem). For $x\in \RR$, let
+$$\pi(x) = \bigl| \{ p \le x : p\ \hbox{prime}\} \bigr|.$$
+Then $pi(x) \sim x/\ln x$, in the sense that
+$$\lim_{x\to\infty} {\pi(x)\over x/\ln x} = 1.$$
+
+The proof is long and arduous, requiring a lot of background in complex analysis.
+It is often taught in a first-year graduate class in analytic number theory.
+Here is an easy corollary of the prime number theorem.
+
+\proclaim Corollary \advthm. Let $n$ be a positive integer and let $m$ be chosen
+uniformly at random from the set $\{1,2,\ldots, n\}$. Then
+$$(\ln n)\pr\{m\ \hbox{prime}\}\to 1$$
+as $n\to\infty$. In other words, the probability that $m$ is prime is asymptotically
+$1/\ln n$.
+
+\proof Since $m$ is chosen uniformly at random from $\{1,\ldots,n\}$,
+the probability $\pr\{ m\ \hbox{prime}\}$ equals $\pi(n)/ n$. So
+$$\lim_{n\to\infty} (\ln n) \pr\{m\ \hbox{prime}\} = \lim_{n\to\infty} {\pi(n) \ln n\over n} 
+= 1,$$
+by the prime number theorem.\slug
+
+As an example, if we choose a $30$-digit number at random (so $n=10^{30} - 1$), the
+probability that this number is prime is roughly $1/\ln(10^{30}) = 1/(30\ln 10) = 0.0145$
+or $1.45\%$.
+
+\advsect Modular Arithmetic
+
+Fix $n\ge 1$, and let $a,b\in\ZZ$. We say {$a$ is congruent to $b$ modulo $n$}
+if $n\divides a-b$, i.e., if $a-b = kn$ for some $k\in \ZZ$. Write $a \equiv b\pmod n$
+or $a \equiv_n b$.
+
+Take for example $n=12$. We have, e.g., $4\equiv 16\pmod 14$, since $4-16 = -12 = (-1)12$,
+but, e.g., $7\not\equiv 17$, since $7-17 = -10$, and $12$ does not divide $-10$. This
+situation should be a familiar one, since we are used to working with numbers modulo $12$ when
+telling the time.
+
+For any fixed $n$, the set of all $(a,b)\in \ZZ^2$ with $a\equiv_n b$ is a relation on $\ZZ$.
+In fact, we have the following proposition.
+
+\proclaim Proposition \advthm.
+For all fixed $n$, the relation $\equiv_n$ is an equivalence relation on the set $\ZZ$.
+
+\proof We must show that $\equiv_n$ is reflexive, symmetric, and transitive.
+
+Let $a\in \ZZ$. We have $a-a = 0 = 0\cdot n$, so $a\equiv a\pmod n$, proving reflexivity.
+
+Let $a,b\in \ZZ$ and suppose that $a\equiv_n b$, so there exists $k$ such that $a-b=kn$.
+Then $b-a = (-k)n$, so $b\equiv_n a$. This proves symmetry.
+
+Lastly, let $a,b,c\in \ZZ$ be such that $a\equiv_n b$ and $b\equiv_n c$. So there
+exist integers $k,l\in \ZZ$ such that $a-b = kn$ and $b-c = ln$. Then
+$$a - c = (a-b) + (b-c) = kn +ln = (k+l)n,$$
+which shows that $a\equiv_n c$.\slug
+
+Since $\equiv_n$ is an equivalence relation, it partitions $\ZZ$ into equivalence classes.
+We shall denote by $[a]_n$ the equivalence class of $a$ modulo $n$. This is the set
+$$[a]_n = \{b\in \ZZ : a-b = kn\ \hbox{for some}\ k\in \ZZ\}
+= \{a + ln : l\in \ZZ\}.$$
+For instance, when $n = 3$, the set
+$$[0]_3 = \{\ldots, -6, -3, 0, 3, 6, \ldots\}$$
+is just the set of all multiples of $3$, and we also have
+$$[1]_3 = \{\ldots, -5, -2, 1, 4, 7,\ldots\}$$
+and
+$$[2]_3 = \{\ldots, -4, -1, 2, 5, 8,\ldots\}.$$
+These are all of the equivalence classes, since, for instance,
+$$[4]_3 = \{\ldots, -2, 1, 4, 7, 10,\ldots\} = [1]_3.$$
+
+Lastly, we shall touch upon the {\it modulo} operator, which is a feature of many programming
+languages. Let $a\in \ZZ$ and $b\ge 1$. Let $a \percent b = r$ where $q$ and $r$ are the integers
+given by the division algorithm. (That is, $a = qb+r$ where $0\le r\le b$.)
+
+\proclaim Proposition \advthm. Fix an integer $n\ge 2$.
+Let $a,b\in \ZZ$. Then $a\equiv b\pmod n$ if and only
+if
+$$a\percent n = b\percent n.$$
+
+\noindent\datestamp{14}{} {\it Proof.}\enspace
+Suppose that $a\equiv_n b$, so that there exists $k\in \NN$ such that $a-b = kn$.
+Then, let $a = nq_1 + r_1$ and $b = nq_2 + r_2$ from the division algorithm.
+(So $r_1 = a\percent n$ and $r_2 = b\percent n$, and we have $0\le r_1,r_2 < n$.)
+Write
+$$kn = a - b = (q_1-q_2) n + (r_1 - r_2),$$
+which we can rearrange to
+$$r_1 - r_2 = kn - (q_1 - q_2)n = (k-q_1 + q_2)n.$$
+This implies that $r_1 - r_2$ divides $n$, but
+since $r_1, r_2\in [0,n)$, the quantity $r_1 - r_2$ is in the range $[-n-1, n)$. Hence
+the only way it can divide $n$ is for $r_1 - r_2 = 0$. We conclude that $a\percent n = b\percent n$.
+
+For the other direction, we once again
+let $a = nq_1 + r_1$ and $b = nq_2 + r_2$ from the division algorithm.
+Now the assumption is that $r_1 = a\percent n = b\percent n = r_2$. So
+$$a - b = a - b = (q_1-q_2) n + (r_1 - r_2) = (q_1-q_2)n.$$
+This shows that $a\equiv b\pmod n$.\slug
+
+This proposition is useful in practice. For example, suppose we want to know whether
+$22\equiv 8\pmod 3$. We calculate $22 = 7\cdot 3 + 1$, so $22\percent 3 = 1$; meanwhile
+$8 = 2\cdot 3 + 2$, so $8\percent 3 = 2$. By the proposition, this means that
+$22\not\equiv 8\pmod 3$.
+
+The proposition also implies that for all $a\in \ZZ$ and $n\ge 2$ one has $[a]_n = [a\percent n]$.
+From the division algorithm, we know that $a\percent n$ is an element in the range $[0,n)$;
+it is equal to the integer $r$ in that range such that we may write $a = qn+r$ for some
+integer $q$. We may choose to denote the whole equivalence class by this element $r$. The set
+$$\ZZ/n\ZZ = \ZZ/{\equiv_n} = \bigl\{ [0], [1], \ldots, [n-1] \bigr\}$$
+is called the {\it ring of integers modulo $n$} or the {\it cyclic group on $n$ elements}.
+
+\medskip\boldlabel Computations modulo {\mathbold n}.
+We can do addition and multiplication modulo $n$ by first doing addition and multiplication
+and then taking the remainder with respect to $n$. For example, when $n=4$ we have the
+addition table
+$$
+\vbox{\ninepoint\tabskip7pt\offinterlineskip
+\halign{
+\hfil $#$ \hfil & \hfil\vrule# \hfil \enskip &
+\hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil \cr
++ && 0 & 1 & 2 & 3 \cr
+& height7pt &&&& \cr
+\noalign{\hrule}
+& height7pt &&&& \cr
+0 && 0 & 1 & 2 & 3 \cr
+& height7pt &&&& \cr
+1 && 1 & 2 & 3 & 0 \cr
+& height7pt &&&& \cr
+2 && 2 & 3 & 0 & 1 \cr
+& height7pt &&&& \cr
+3 && 3 & 0 & 1 & 2 \cr
+}}
+$$
+and the multiplication table
+$$
+\vbox{\ninepoint\tabskip7pt\offinterlineskip
+\halign{
+\hfil $#$ \hfil & \hfil\vrule# \hfil \enskip &
+\hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil \cr
+\cdot && 0 & 1 & 2 & 3 \cr
+& height7pt &&&& \cr
+\noalign{\hrule}
+& height7pt &&&& \cr
+0 && 0 & 0 & 0 & 0 \cr
+& height7pt &&&& \cr
+1 && 0 & 1 & 2 & 3 \cr
+& height7pt &&&& \cr
+2 && 0 & 2 & 0 & 2 \cr
+& height7pt &&&& \cr
+3 && 0 & 3 & 2 & 1 \cr
+}}.
+$$
+For $x,y\in \ZZ$, if $xy = 0$, then either $x=0$ or $y=0$.
+On the other hand, from the assumption $xy \equiv 0\pmod n$ we cannot conclude
+in general that $x\equiv 0\pmod n$ or $y\equiv 0\pmod n$. As an example, when $n=4$
+and $x=y=2$, neither $x$ nor $y$ is congruent to $0$ modulo $4$, but their product is $0$ modulo $4$.
+An element $a\in \ZZ$ with $a\not\equiv 0\pmod n$ is said to be a {\it zero divisor}
+if there exists $b\in \ZZ$ with $b\not\equiv 0\pmod n$ such that $ab \equiv 0\pmod n$.
+
+Remember that the numbers in the tables above are elements of $\ZZ/n\ZZ$ (in the case $n=4$),
+hence they are not really integers so much as equivalence classes of integers, which are sets!
+So it is not immediate that addition and multiplication are well-defined modulo some fixed integer $n$.
+For instance, to say that $[3]_4 \cdot [1]_4 = [3]_4$, it is not enough to just observe that
+$3\cdot 1 = 3$. We need to show that for any integers $a\in [3]_4$ and $b\in [1]_4$, we have
+$a\cdot b\in [3]_4$. This is the essence of the following proposition.
+
+\proclaim Proposition \advthm. Let $a\equiv c\pmod n$ and $b\equiv d\pmod n$. Then
+\medskip
+\item{i)} $a+b\equiv c+d \pmod n$;
+\smallskip
+\item{ii)} $ab\equiv cd\pmod n$; and
+\smallskip
+\item{iii)} $a^m \equiv c^m \pmod n$ for all $m\in \NN$.
+\medskip
+
+\proof By assumption, there are integers $k$ and $l$ such that $a-c = kn$ and $b-d = ln$.
+
+To prove (i), we compute
+$$(a+b)-(c+d) = a-c + b-d = kn+ln = (k+l)n,$$
+so $a+b\equiv c+d\pmod n$.
+
+For part (ii), we compute
+$$\eqalign{
+ab-cd &= ab-cb+cb - cd \cr
+&= b(a-c) + c(b-d) \cr
+&= bkn + cln \cr
+&= (bk + cl) n ,\cr
+}$$
+so $ab\equiv cd \pmod n$.
+
+We prove part (iii) by induction. The base case is $m=0$; we have $a^0 = 1 = c^0$,
+so $a^0\equiv c^0\pmod n$. Now let $m\ge 0$ and assume that $a^m\equiv c^m\pmod n$.
+Combining this with the hypothesis $a\equiv c\pmod n$, we use part (ii) to
+conclude that $a\cdot a^m \equiv c\cdot c^m\pmod n$, which simplifies
+to $a^{m+1}\equiv c^{m+1}\pmod n$.\slug
+
+We can use this to perform the modulo operation on large numbers without necessarily
+knowing the large numbers themselves. For example, let $m=21^{10} - 3\cdot 13 + 14$
+and suppose we want to know $m\percent 6$.
+
+First we observe that $14\equiv 2\pmod 6$
+and $13\equiv 1\pmod 6$. So
+$$ m = 21^{10} - 3\cdot 13 + 14 \equiv_6 21^{10} - 3\cdot 1 + 2 \equiv_6 21^{10} - 1.$$
+Now we deal with the exponent by first taking the base of the exponent modulo $6$.
+We have $21\equiv 3 \pmod 6$; hence
+$$m \equiv_6 3^{10} -1.$$
+To reduce the exponent,
+we factorise $10=2\cdot 5$, yielding
+$$m \equiv_6 (3^2)^5 - 1.$$
+But again we can reduce the base here, since $9\equiv_6 3$, which gives us
+$$m \equiv_6 3^5-1.$$
+Now since the exponent $5$ is prime, we aren't able to reduce it by factorising. However,
+we do have $5 = 2+2+1$, and, using once again the fact that $3^2 = 9\equiv_6 3$, we obtain
+$$m \equiv_6 3^2\cdot 3^2 \cdot 3 - 1 \equiv_6 3\cdot 3\cdot 3 - 1 \equiv_6 3\cdot 3 - 1
+\equiv_6 3- 1 \equiv_6 2.$$
+
+\medskip\boldlabel Inverses modulo {\mathbold n}. Considering $2$ as a rational
+number, we say that the inverse of $2$ is $1/2$, since $2\cdot (1/2) = 1$.
+We write $2^{-1} = 1/2$. We shall use the same terminology in the ring of integers modulo $n$.
+An element $a\in \ZZ$ is said to be {\it invertible modulo $n$} if there exists $b\in \ZZ$
+such that $ab \equiv 1\pmod n$. In this case we shall say that $b$ is an {\it inverse} of $a$.
+In fact, inverses are unique (in $\ZZ/n\ZZ$).
+
+\proclaim Proposition \advthm. Let $a,n\in \ZZ$ with $n\ge 2$. Then if $ab \equiv 1\pmod n$
+and $ac\equiv 1\pmod n$, then $b\equiv c\pmod n$.
+
+\proof We have
+$$b \equiv b\cdot 1 \equiv b(ac) \equiv (ab)c \equiv 1\cdot c \equiv c\pmod n.\noskipslug$$
+
+Hence we may speak of {\it the} inverse of an element $a$, which we shall denote by $a^{-1}$.
+We have $a^{-1}a \equiv 1 \equiv a a^{-1}\pmod n$.
+
+For instance, when $n=5$, the fact that
+$$2\cdot 3 \equiv_5 6\equiv_5 1$$
+tells us that $2$ (and $3$, for that matter) are invertible modulo $5$.
+When $n=6$, is $2$ still invertible? Here $n$ is small enough that we can just try all the cases.
+We have $2\cdot 0 = 0$, $2\cdot 1 = 2$, $2\cdot 2 = 4$, $2\cdot 3 = 6\equiv_6 = 0$
+$2\cdot 4 = 8 \equiv_6 2$, and $2\cdot 5 = 10\equiv_6 4$. Since none of these were equivalent
+to $1$ modulo $6$, we see that $2$ is not invertible modulo $6$.
+
+It was quite a tedious process to exhaustively show that $2$ is not invertible modulo $6$. The following
+theorem gives a much easier-to-use criterion for invertibility.
+
+\proclaim Theorem \advthm. Let $a,n\in \ZZ$ with $n\ge 2$. Then
+\medskip
+\item{i)} $a$ is invertible modulo $n$ if and only if $\gcd(a,n) = 1$; and
+\smallskip
+\item{ii)} if $a$ is invertible, then there is a unique integer $b\in [0,n-1]$ such that
+$ab\equiv 1\pmod n$. Namely, if
+$$1 = sa + tn$$,
+then we can set $b = s\percent n$.
+\medskip
+
+\proof The proof of (i) is a fairly simple corollary of B\'ezout's theorem.
+f $a$ is invertible, then $ab\equiv 1\pmod n$, so there exists $k\in \ZZ$ such that
+$1-ab=kn$. Rearranging this, we have $1 = ab+kn$, and by B\'ezout's theorem (specifically,
+the final statement of Theorem~{\thmbezout}, this implies that $\gcd(a,n) = 1$.
+On the other hand, if $\gcd(a,n) = 1$, then there exists $k\in \ZZ$ such that
+$1 = ab + kn$. This means that $1-ab = kn$, so $ab\equiv 1\pmod n$.
+
+For part (ii), suppose that $s,t\in \ZZ$ are such that $1 = sa+tn$. Then
+taking this equation modulo $n$, we have
+$$ 1 = sa+tn \equiv_n sa + t\cdot 0 \equiv_n sa.$$
+Letting $b = s\percent n$, we have $1 = ba$, since $s$ and $b$ belong to the same
+equivalence class modulo $n$, and $b$ is an integer in $[0,n-1]$ by the division algorithm.
+This proves existence of such a $b$. To show uniqueness, suppose that $b$ and $c$ are both
+integers in $[0,n-1]$ with $ab \equiv 1\pmod n$ and $ac\equiv 1\pmod n$. Then
+$ab\equiv ac\pmod n$, and multiplying on both sides by $b$, we have $bab\equiv bac\pmod n$.
+But $ba\equiv 1\pmod n$, so cancelling this out in the equation, we have $b\equiv c\pmod n$.
+In other words, $n\divides b-c$. But since $b,c\in [0,n-1]$, we have $b-c\in [-n+1, n-1]$,
+and the only divisor of $n$ in this range is $0$. Hence $b-c = 0$ and we conclude that $b=c$.\slug
+
+We can use this theorem to systematically find inverses of integers modulo other integers. For instance,
+suppose we want to find the inverse of $17$ modulo $20$. First we find $\gcd(20,17)$ by Euclid's
+algorithm:
+$$\eqalign{
+20 &= 1\cdot 17 + 3 \cr
+17 &= 5\cdot 3 + 2 \cr
+3 &= 1\cdot 2 + 1 \cr
+1 &= 1\cdot 1 + 0 \cr
+}$$
+The fact that $\gcd(20,17) = 1$ tells us that $17$ indeed has an inverse. Now we go backwards to
+express $1$ as an integer linear combination of $20$ and $17$:
+$$\eqalign{
+1 &= 3-1\cdot 2 \cr
+&= 3-(17-5\cdot 3) \cr
+&= -17 + 6\cdot 3 \cr
+&= -17 + 6(20-17) \cr
+&= 6\cdot 20 - 7\cdot 17 \cr
+}$$
+So the inverse of $17$ modulo $20$ is $-7 \equiv_{20} 13$. (We can verify that
+$$17\cdot 13\equiv (-3)(-7)\equiv 21 \equiv 1\pmod {20}.$$
+As a trick for computations, use the range $-n/2$ to $n/2$ instead of $0$ to $n-1$ to keep the
+numbers a bit smaller. It might help you avoid errors.)
+
+Let's find the full list of invertible integers modulo $20$, as well as their inverses. First,
+we list the integers in the range $[0,19]$ that are relatively prime with $20$:
+$$1,3,7,9,11,13,17,19$$
+If some number is in this list, then its inverse must also be in this list (since inverses to
+a given element must themselves be invertible).
+The elements $1$ and $19$ are their own inverses, since $1\cdot 1 = 1$ and
+$19\cdot 19 \equiv_{20} (-1)(-1) = 1$.
+We already saw above that $13$ and $17$ are inverses to each other. That leaves $3$, $7$, $9$, and $11$.
+First we see that $3\cdot 7 = 21 \equiv_{20} 1$, so $3$ and $7$ are inverses to each other.
+Then we note that $9\cdot 9 = 81 \equiv_{20} 1$, so $9$ is its own inverse, leaving us to conclude
+that $11$ must also be its own inverse as well---either by the fact that we've ruled
+out all other possibilities, or by the computation
+$$11\cdot 11 \equiv_{20} (-9)(-9)  = 81 \equiv_{20} 1.$$
+
+\noindent\datestamp{14}{} \hskip\parindent
+We now investigate what happens in the particular case where we the modulus $n$ is prime.
+In this situation we have the following proposition.
+
+\edef\propinvertiblemodp{\the\thmcount}
+\proclaim Proposition \advthm. Let $p$ be prime. Then
+\medskip
+\item{i)} every $x\in \ZZ$ with $x\not\equiv 0\pmod p$ is invertible modulo $p$; and
+\smallskip
+\item{ii)} for all $a,b\in \ZZ$ with $ab\equiv 0\pmod p$ one has
+$a\equiv 0\pmod p$ or $b\equiv 0\pmod p$.
+\medskip
+
+\proof Since $p$ only has factors $1$ and $p$, $\gcd(x,p)$ must either be $1$ or $p$.
+But the condition $x\not\equiv 0\pmod p$ implies that $p$ does not divide $x$.
+Hence $\gcd(x,p) = 1$, proving part (i).
+
+For part (ii), assume that $ab\equiv 0\pmod p$, so $p\divides ab$. But $p$ is prime, so
+by Theorem~{\thmpdivides}, either $p\divides a$ or $p\divides b$. In the first
+case we have $a\equiv 0\pmod p$, and in the second case, $b\equiv 0\pmod p$.\slug
+
+\medskip\boldlabel Solving congruences modulo {\mathbold n}.
+Suppose we want to solve for all integers $x$ satisfying $x^2 \equiv x\pmod n$, first for $n=6$,
+then for $n=7$.
+
+For the case where $n=6$, it suffices to consider $x\in \{0,1,\ldots, n-1\}$,
+since if $x^2\equiv x\pmod n$ and $x\equiv a\pmod n$, then $a^2\equiv a\pmod n$.
+So we simply try all $x\in \{0,1,\ldots, 5\}$. We have $0^2 = 0$, $1^2 = 1$,
+$2^2 = 4 \not\equiv_6 2$, $3^2 = 9\equiv_6 3$, $4^2 = 16\equiv_6 4$, and
+$5^2 \equiv_2 (-1)^2 = 1\not\equiv_6 5$. We conclude that $0$, $1$, $3$, and $4$ are solutions
+to this congruence modulo $6$, and more broadly, any integer $y$ of the form
+$$y = x+6k$$
+where $k\in \ZZ$ and $x\in \{0,1,3,4\}$, is a solution $x^2\equiv x\pmod 6$.
+
+Now we tackle the case where $n=7$. In fact, the solution is no different
+if $n$ were any other prime.
+First we subtract $x$ from both sides, obtaining the congruence $x^2 - x\equiv 0\pmod 7$,
+then we factorise the left side to get $x(x-1)\equiv 0\pmod 7$. Now by the previous proposition,
+since $7$ is prime if $x(x-1)\equiv 0\pmod 7$, we must have either $x\equiv 0\pmod 7$ or
+$x-1\equiv 0\pmod 7$, so the only solutions are $x\equiv_7 0$ and $x\equiv_7 1$ (and anything
+in their equivalence classes).
+
+Replacing $7$ with any other prime $p$ in the above paragraph yields a proof of the following
+proposition.
+
+\proclaim Proposition \advthm. Let $p$ be a prime. Then $a^2\equiv a\pmod p$ if and only if
+$a$ is either congruent to $0$ or $1$ modulo $p$.\slug
+
+Here is a similar proposition.
+
+\edef\propselfinversemodp{\the\thmcount}
+\proclaim Proposition \advthm. Let $p$ be a prime and let $a\not\equiv 0\pmod p$ (so
+that $a$ is invertible. Then $a \equiv a^{-1}\pmod p$ if and only
+if $a$ is either congruent to $1$ or $-1$ modulo $p$.
+
+\proof If $a = 1$ or $a=-1$, then $a^2 = 1$, so $a^2\equiv 1$, and multiplying by $a^{-1}$ on both sides,
+we have $a\equiv a^{-1}\pmod p$.
+
+Now suppose that $a\equiv a^{-1}\pmod p$. Multiplying by $a$ on both sides, we get $a^2\equiv 1\pmod p$;
+then, subtracting $1$ from both sides and factoring the resulting polynomial, we get
+$$(a+1)(a-1)\equiv 0\pmod p.$$
+This implies that either $a+1\equiv 0\pmod p$ or $a-1\equiv 0\pmod p$. In the first case, $a \equiv_p -1$,
+and in the second case, $a\equiv_p 1$.\slug
+
+We finish off this section with an important theorem, first stated by Pierre de Fermat in~1640.
+He did not supply a proof; the first published proof of this theorem was given by
+Leonhard Euler in~1736.
+
+\edef\thmflt{\the\thmcount}
+\parenproclaim Theorem {\advthm} (Fermat's little theorem). Let $a$ and $p$ be integers with $p$ prime.
+If $a\not\equiv 0\pmod p$, then $a^{p-1}\equiv 1\pmod p$.
+
+Before proving this theorem, we first state and prove a lemma. Recall that we define the {\it factorial}
+of $n\in \NN$ to be the product $n! = 1\cdot 2\cdots (n-1) n$.
+
+\proclaim Lemma \advthm. For all prime numbers $p$, the integer $(p-1)!$ is congruent to $-1$ modulo $p$.
+
+\proof If $p=2$, we have $(p-1)! = 1\equiv -1\pmod 2$, and if $p=3$, then $(p-1)! = 2 \equiv -1 \pmod 3$.
+
+For the rest of the proof, assume that $p\ge 5$.
+By Proposition~{\propselfinversemodp}, each element in the set $S = \{2,3,\ldots,p-3,p-2\}$
+is not its own inverse modulo $p$, so for each element $s\in S$, there is some other element $s'\in S$ with
+$s'\ne s$ such that $ss'\equiv 1\pmod p$. This means that
+$$ 2\cdot 3 \cdots (p-3)(p-2) \equiv 1\pmod p.$$
+From this, we see that
+$$(p-1)! = 1\cdot 2\ldots (p-2)(p-1) = \bigl(2\cdot 3 \cdots (p-3)(p-2)\bigr)(p-1)
+\equiv_p (p-1)\equiv_p -1.\noskipslug$$
+
+We are now able to prove Fermat's little theorem.
+
+\medskip\noindent{\it Proof of Theorem~{\thmflt}}.\enspace
+Since $a\not\equiv 0\pmod p$, it is invertible modulo $p$ by Proposition~{\propinvertiblemodp}.
+Denote its inverse by $a^{-1}$.
+Let $G = (\ZZ/p\ZZ)\setminus\{0\}$ for short. Define a function $f:G\to G$ by letting $f(x) = a\cdot x$,
+where we consider the result modulo $p$ (and hence the result is an element of $\ZZ/p\ZZ$).
+This function
+is well defined: since $a$ and $x$ are both not zero modulo $p$, their product will be an element of $G$.
+If $f(x_1) = f(x_2)$, then
+$$a\cdot x_1 \equiv a\cdot x_2\pmod p,$$
+so multiplying
+both sides by $a^{-1}$, we have $x_1 \equiv x_2\pmod p$.
+This proves that $f$ is injective. Now let $y\in G$. We want
+to find $x$ with $f(x) \equiv y\pmod p$.
+To do so, simply set $x \equiv a^{-1}y\pmod p$. Then
+$$f(x) = f(a^{-1}y) = a(a^{-1}y) \equiv y \pmod p.$$
+This proves that $f$ is surjective, and hence bijective.
+
+The bijection $f$ shows that the set
+$$\bigl\{a, 2a, \ldots,(p-2)a, (p-1)a\bigr\},$$
+taken as a subset of $\ZZ/p\ZZ$ (i.e., we take each element modulo $p$, in the range $[0,\ldots p-1]$),
+is exactly the same as the set
+$$\{1, 2, \ldots, p-2, p-1\},$$
+just that the order of elements might be permuted. Hence we have
+$$a(2a)\cdots\bigl((p-2)a\bigr)\bigl((p-1)a\bigr) \equiv (p-1)!\pmod p.$$
+Combining all the $a$ factors on the left-hand side, we get
+$$a^{p-1} (p-1)!\equiv (p-1)!\pmod p.$$
+But by the previous lemma, $-1$ is the inverse of $(p-1)!$ modulo $p$,
+so multiplying both sides of this congruence by $-1$, we get
+$$a^{p-1}\equiv 1\pmod p,$$
+which is what we wanted to show.\slug
+
+\bigskip
+\begingroup\obeylines\eightssi
+\hfill Tout nombre premier mesure infalliblement
+\hfill une des puissances - 1 de quelque progression que ce soit,
+\hfill \& l'exposant de ladite puissance est so\^us-multiple du nombre premier donn\'e - 1.
+\hfill Et apr\'es qu'on a trouv\'e la premiere puissance qui satisfait \`a la question,
+\hfill toutes celles dont les exposans sont multiples de l'exposant de la premiere
+\hfill satisfont de m\'eme \`a la question.
+\eightss
+\smallskip
+\hfill --- PIERRE DE FERMAT, {\eightssi in a letter to Bernard Fr\'enicle de Bessy} (1640)
+\endgroup%\obeylines
+\bigskip\goodbreak
+
 \bye

### 2024-10-11 teaching/math240f24/draft.tex
lecture oct 10

@@
 \hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}
 
@@
 $p_i = q_i$ for all $1\le i\le s+1$.\slug
-\bye
-
+
+\noindent\datestamp{10}{x} \hskip\parindent
 We can use the Fundamental Theorem of Arithmetic to prove the following generalisation
@@
 $$\pi(x) = \bigl| \{ p \le x : p\ \hbox{prime}\} \bigr|.$$
-Then $pi(x) \sim x/\ln x$, in the sense that
+Then $\pi(x) \sim x/\ln x$, in the sense that
 $$\lim_{x\to\infty} {\pi(x)\over x/\ln x} = 1.$$
@@
 
-\advsect Modular Arithmetic
+\advsect Modular arithmetic
 
@@
 
-\noindent\datestamp{14}{} {\it Proof.}\enspace
+\proof
 Suppose that $a\equiv_n b$, so that there exists $k\in \NN$ such that $a-b = kn$.
@@
 Now the assumption is that $r_1 = a\percent n = b\percent n = r_2$. So
-$$a - b = a - b = (q_1-q_2) n + (r_1 - r_2) = (q_1-q_2)n.$$
+$$a - b = (q_1-q_2) n + (r_1 - r_2) = (q_1-q_2)n.$$
 This shows that $a\equiv b\pmod n$.\slug
-
+\bye
+
+\noindent\datestamp{14}{} \hskip\parindent
 This proposition is useful in practice. For example, suppose we want to know whether
@@
 taking this equation modulo $n$, we have
-$$ 1 = sa+tn \equiv_n sa + t\cdot 0 \equiv_n sa.$$
+$$ 1 = sa+tn \equiv_n sa + 0t \equiv_n sa.$$
 Letting $b = s\percent n$, we have $1 = ba$, since $s$ and $b$ belong to the same
@@
 
-\noindent\datestamp{14}{} \hskip\parindent
+\noindent\datestamp{15}{} \hskip\parindent
 We now investigate what happens in the particular case where we the modulus $n$ is prime.
@@
 
+\edef\lemhalfwilson{\the\thmcount}
 \proclaim Lemma \advthm. For all prime numbers $p$, the integer $(p-1)!$ is congruent to $-1$ modulo $p$.
@@
 
+As a matter of interest, the intermediary lemma we proved is one direction of Wilson's theorem, proved
+by John Wilson in 1770.
+
+\parenproclaim Theorem {\advthm} (Wilson's theorem).
+For all integers $n\ge 2$, the congruence
+$$(n-1)!\equiv -1\pmod n$$
+holds if and only if $n$ is prime.
+
+\proof We already proved the ``if'' direction earlier, as Lemma~{\lemhalfwilson}.
+We leave the ``only if'' direction of the proof as an exercise for the reader.\slug
+
 \bigskip
@@
 
-\bye
+\advsect Applications of number theory
+
+In this section we present a potpourri of interesting ways number theory is applied to make your life better.
+
+\medskip\boldlabel {\ninepoint\bf ISBN} book identifiers. Every published book has an
+{\mc ISBN} code that serves as its
+unique identifier. This code is $10$ digits long for books published before 2007, but three new digits
+have been added for books published after 2007. In this section we'll deal with the simpler case of $10$ digits.
+
+What we want to do is to bake some redundancy into the codes, so that if a single digit is typed wrong,
+a computer will be able to tell the user that the code is invalid. This allows the user to correct their
+mistake. The way we do this, in a ten-digit code $d_{10}d_9d_8\cdots d_2d_1$,
+is to have only the first nine digits encode the book. The last digit is a {\it check digit},
+chosen so that
+\edef\eqvalidisbn{\the\eqcount}
+$$\sum_{i=1}^{10} i d_i \equiv 0 \pmod{11}.\adveq$$
+In other words,
+$$d_1\equiv -\sum_{i=2}^{10} id_i \pmod{11}.$$
+If $d_1$ needs to equal $10$, we use the symbol `X'.
+
+How does this solve our problem? Well, if a hapless person takes a valid {\mc ISBN} code
+and mangles it by either getting one digit wrong, or by swapping two adjacent digits,
+the result is an invalid {\mc ISBN} code, and the computer will be able to flag it.
+We formalise this statement as the following theorem.
+
+\proclaim Theorem \advthm. Let $d_{10} d_9 d_8\cdots d_2 d_1$ be a valid {\mc ISBN}
+code; that is, a code that satisfies~\refeq{\eqvalidisbn}. Then any code obtained by either
+\medskip
+\item{i)} changing exactly one digit $d_i$, for some $1\le i\le 10$, or
+\smallskip
+\item{ii)} swapping {\it distinct} adjacent digits $d_i$ and $d_{i-1}$, for some $2\le i\le 10$,
+\medskip
+is not a valid {\mc ISBN} codes (it does not satisfy~\refeq{\eqvalidisbn}).
+
+\proof First we prove (i). Suppose that $c_{10} c_9 c_8\cdots c_2c_1$ is a code that is
+equal to $d_{10} d_9 d_8\cdots d_2 d_1$ except at one place $1\le j\le 10$. So $c_j\ne d_j$,
+but for all $1\le i\le 10$ with $i\ne j$, we have $c_i = d_i$. Then
+$$\eqalign{
+\sum_{i=1}^{10} i c_i &\equiv_11 \sum_{i=1}^{10} i c_i - 0 \cr
+&\equiv_11 \sum_{i=1}^{10} i c_i - \sum_{i=1}^{10} id_i \cr
+&= \sum_{i=1}^{10} i(c_i - d_i) \cr
+&= j(c_j-d_j). \cr
+}$$
+But $j\in \{1,\ldots, 10\}$ and $(c_j-d_j) \in [-10, 10]$. So neither $j$ nor $c_j-d_j$
+are congruent to $0$ modulo $11$. Since $11$ is prime, their product cannot be congruent to $0$
+modulo $11$. We conclude that $\sum_{i=1}^{10} i c_i\not\equiv 0\pmod 11$; i.e., this is
+not a valid {\mc ISBN} code.
+
+To prove part (ii), suppose that $c_{10} c_9 c_8\cdots c_2c_1$ is a code that is
+equal to $d_{10} d_9 d_8\cdots d_2 d_1$, except that for some $2\le j\le 10$, we
+have $c_j = d_{j-1}$ and $c_{j-1}=d_j$ (the adjacent digits are swapped). Furthermore,
+assume that $d_j\ne d_{j-1}$, since swapping adjacent digits that are the same doesn't
+do anything to the code. Then
+$$\eqalign{
+\sum_{i=1}^{10} i c_i &\equiv_11 \sum_{i=1}^{10} i c_i - 0 \cr
+&\equiv_11 \sum_{i=1}^{10} i c_i - \sum_{i=1}^{10} id_i \cr
+&= \sum_{i=1}^{10} i(c_i - d_i) \cr
+&= (j-1)c_{j-1} + jc_j - (j-1)d_{j-1} - jd_j \cr
+&= (j-1)d_j + jd_{j-1} - (j-1)d_{j-1} - jd_j \cr
+&= d_{j-1} - d_j.\cr
+}$$
+But since $d_{j-1}$ and $d_j$ are both in the range $\{0,\ldots, 10\}$, their difference
+is in the range $[-10,10]$, and we assumed that $d_{j-1}\ne d_j$, so this difference is not zero.
+We conclude that $d_{j-1}-d_j\not\equiv 0\pmod 11$, so the code is not valid.\slug
+
+Something similar is done with credit card numbers, though we don't take digits modulo $11$
+(or else we'd have to use the digit 'X', which is cumbersome).
+Instead, a slightly different method called Luhn's algorithm is used. The calculation
+used to find the check digit is a bit different (let's not get into it),
+but the result is that we can still detect
+single-digit errors, and we can detect all swapping errors except for the transposition
+$09\leftrightarrow 90$.
+
+\medskip\boldlabel Divisibility tests.
+In grade school you might have learned the following divisiblity rule:
+{\sl To tell if a number is divisible by $3$, you sum up all of its digits. If this is a
+multiple of $3$, then the original number was a multiple of $3$, and if not, then the original
+number wasn't.} The same holds with $3$ replaced by $9$.
+
+Using the number theory we've learned, we are now able to prove this fact.
+
+\proclaim Proposition \advthm. Suppose that $n\in \NN$ is written in base-$10$ digits as
+$$n = d_k d_{k-1} d_{k-2}\cdots d_2 d_1 d_0,$$
+for some integer $k\ge 0$ and $d_i\in \{0,\ldots,9\}$ for $0\le i\le k$.
+Then $n$ is divisible by $3$ if and only if
+$$\sum_{i=0}^k d_i$$
+is also divisible by $3$.
+
+\proof Since $10\equiv 1\pmod 3$, any power of $10$ is also congruent to $1$ modulo $3$. So we have
+$$n = \sum_{i=0}^k d_i 10^i \equiv_3 \sum_{i=0}^k d_i,$$
+and we see that $n$ is divisible by $3$ if and only if the sum on the right-hand side is.\slug
+
+This works with $3$ replaced by $9$ since we also have $10\equiv 1\pmod 9$.

### 2024-10-16 teaching/math240f24/draft.tex
Constance Rooke

@@
 
-Let's see how to use this theorem. Suppose we want to know if the number $\root 6\of 18$ is
+Let's see how to use this theorem. Suppose we want to know if the number $\root 6\of{18}$ is
 irrational or not. Well, since $1^6 = 1$ and $2^6 = 64$, we have $1^6 < 18 < 2^6$,
-meaning that $1<\root 6\of 18 < 2$. (This is because the function $f:\RR\to\RR$ defined
-by $f(x) = x^6$ is increasing on the interval $[1,2]$.) Hence $\root 6\of 18$ is not
+meaning that $1<\root 6\of{18} < 2$. (This is because the function $f:\RR\to\RR$ defined
+by $f(x) = x^6$ is increasing on the interval $[1,2]$.) Hence $\root 6\of{18}$ is not
 an integer, and by the theorem it must be irrational.

### 2024-10-17 about/vita.tex
two lectures done
BCS 2025
fixed dyslexia

@@
   \hangindent=\itemindent\hangafter=1\ignorespaces}
+\def\starpubitem#1.{\ifitempar\smallskip\noindent\else\itempartrue
+  \hskip-\parindent\fi
+  \llap{*}\hbox to\itemindent{\bf\hfil #1.\kern.25em}%
+  \hangindent=\itemindent\hangafter=1\ignorespaces}
 
@@
 \def\ficitem{\pubitem F\the\fictioncount. \global\advance\fictioncount by -1}
+\def\starficitem{\starpubitem F\the\fictioncount. \global\advance\fictioncount by -1}
 
@@
 \leftright{\bf NSERC Canada Graduate Scholarship -- Doctoral
-\rm (\$105,000)}{\bf September 2023 -- August 2026}
+\rm (\$115,000)}{\bf September 2023 -- August 2026}
 \smallskip
@@
 
-\sectheader{Published Fiction}
+\sectheader{Published Creative Writing {\rm($*$ indicates nonfiction)}}
 
-\fictioncount=3
+\fictioncount=4
 \pubbegin F
+\starficitem ``Lanterns,'' Winner of the 2024 Constance Rooke Creative Nonfiction Prize. To appear in
+{\sl The Malahat Review}
 \ficitem ``Senang Diri,'' {\sl Existere} {\bf 42},2 (2023), 17--25.
-\ficitem ``The Vigil,'' {\sl Ricepaper Magazine} (2023), online.
+\ficitem ``The Vigil,'' {\sl Ricepaper Magazine} (2023), online. Reprinted in
+{\sl Best Canadian Stories 2025} (Windsor, Ont.: Biblioasis, 2024), 244--253.
 \ficitem ``Mountain Pass,'' {\sl Prairie Journal of Canadian Literature} {\bf 79} (2023), 48--55.
@@
 \sectheader{Work and Volunteer Experience {\rm($*$ indicates a paid position)}}
+
+\starleftright{\bf Course lecturer}{\bf August 2024 -- present}
+\leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
+\smallskip
+Instructor for a section of MATH 240 Discrete Structures. Duties include preparing and delivering lectures,
+designing assignment and exam questions, and holding office hours.
+\medbreak
 
@@
 
-\starleftright{\bf Teaching assistant}{\bf September 2021 -- present}
+\starleftright{\bf Teaching assistant}{\bf September 2021 -- April 2024}
 \leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}

### 2024-10-23 teaching/math240f24/draft.tex
administrative details
lecture 22 oct

@@
 This shows that $a\equiv b\pmod n$.\slug
-\bye
-
-\noindent\datestamp{14}{} \hskip\parindent
+
+\noindent\datestamp{22}{x} \hskip\parindent
 This proposition is useful in practice. For example, suppose we want to know whether
@@
 
+\edef\propinversesunique{\the\thmcount}
 \proclaim Proposition \advthm. Let $a,n\in \ZZ$ with $n\ge 2$. Then if $ab \equiv 1\pmod n$
@@
 
+\edef\thminvertiblemodn{\the\thmcount}
 \proclaim Theorem \advthm. Let $a,n\in \ZZ$ with $n\ge 2$. Then
@@
 $ab\equiv 1\pmod n$. Namely, if
-$$1 = sa + tn$$,
+$$1 = sa + tn,$$
 then we can set $b = s\percent n$.
@@
 equivalence class modulo $n$, and $b$ is an integer in $[0,n-1]$ by the division algorithm.
-This proves existence of such a $b$. To show uniqueness, suppose that $b$ and $c$ are both
-integers in $[0,n-1]$ with $ab \equiv 1\pmod n$ and $ac\equiv 1\pmod n$. Then
-$ab\equiv ac\pmod n$, and multiplying on both sides by $b$, we have $bab\equiv bac\pmod n$.
-But $ba\equiv 1\pmod n$, so cancelling this out in the equation, we have $b\equiv c\pmod n$.
-In other words, $n\divides b-c$. But since $b,c\in [0,n-1]$, we have $b-c\in [-n+1, n-1]$,
+This proves existence of such a $b$. To show uniqueness, we suppose that $b$ and $c$ are both
+elements of $[0,n-1]$ with $ab\equiv 1 \pmod n$ and $ac\equiv 1\pmod n$. By
+Proposition~{\propinversesunique}, we have $b\equiv c\pmod n$;
+in other words, $n\divides b-c$. But since $b,c\in [0,n-1]$, we have $b-c\in [-n+1, n-1]$,
 and the only divisor of $n$ in this range is $0$. Hence $b-c = 0$ and we conclude that $b=c$.\slug
@@
 
-\noindent\datestamp{15}{} \hskip\parindent
 We now investigate what happens in the particular case where we the modulus $n$ is prime.
@@
 case we have $a\equiv 0\pmod p$, and in the second case, $b\equiv 0\pmod p$.\slug
+
+\bye
 
@@
 This works with $3$ replaced by $9$ since we also have $10\equiv 1\pmod 9$.
+
+\medskip\boldlabel Computing large powers modulo {\mathbold p}.
+The next application is not a precise mathemathical statement; rather, it's an observation
+that Fermat's little theorem can allow us to compute what large powers are modulo a prime
+without having to compute the number itself. We'll illustrate this by two examples.
+
+First we compute $25^{134} \pmod{11}$. We begin by noting that
+$$25\equiv 3\pmod{11},$$
+so what we really want is $3^{134}\pmod{11}$. By Fermat's little theorem, $3^{10}\equiv 1\pmod{11}$,
+so we may cast out multiples of $10$ from $134$ to arrive at the easy computation
+$$3^{134} \equiv 3^{10\cdot 13+4} \equiv (3^{10})^{13}\cdot 3^4 \equiv 3^2 \cdot 3^2
+\equiv 9\cdot 9\equiv (-2)(-2) \equiv 4\pmod{11}.$$
+
+Now let's try $25{134} \pmod{14}$. Since $14$ is not prime, we can no longer use Fermat's little theorem.
+The trick we shall use is to square $25$ repeatedly. First note that $25\equiv -3 \pmod{14}$,
+so $25^2 \equiv 9\pmod{14}$. Then, squaring again, we get
+$$25^4 \equiv 9^2 \equiv (-5)^2 \equiv 25\equiv -3\pmod{14}.$$
+This means that repeated squaring of $25$ cycles between $-3$ and $9$ modulo $14$, so we immediately
+have
+$$25^8\equiv 9,\qquad 25^{16} \equiv -3,\qquad 25^{32}\equiv 9,\qquad 25^{64} \equiv -3,$$
+and
+$$25^{128}\equiv 9$$
+modulo $14$. Now writing the exponent $134$ as a sum of powers of two allows us to conclude that
+$$ 25^{134} = 25^{128+4+2} = 25^{128}\cdot 25^4 \cdot 25^2 \equiv 9\cdot (-3)\cdot 9 \equiv (-3)(-3)
+\equiv 9\pmod{14}.$$
+
+\medskip\boldlabel Fermat's primality test.
+Fermat's little theorem says that if $n$ is prime, then for all integers $a$ with $a\not\equiv 0\pmod n$
+one has $a^{n-1}\equiv 1\pmod n$. The contrapositive of this statement is that if there
+is some integer $a$ with $a\not\equiv 0\pmod n$ and $a^{n-1}\not\equiv 1\pmod n$, then $n$ is not prime.
+This leads us to the following algorithm that can (usually) tell us when a number is not prime.
+
+\algbegin Algorithm F (Fermat's primality test). Given an integer $n\ge 4$,
+this algorithm will attempt to declare that
+$n$ is not prime. If the algorithm fails to do so, then the test is inconclusive ($n$ may either
+be prime or not).
+\algstep F1. [Initialise.] Set $a \gets 2$. (We can skip the step $a=1$ since $a^{n-1}$ will
+always equal $1$ in this case.)
+\algstep F2. [Compute power.] Set $b\gets a^{n-1} \% n$. (So $0\le b< n$.)
+\algstep F3. [Not prime?] If $b \neq 1$, output ``Not prime,'' and terminate the algorithm.
+\algstep F4. [Loop?] Set $a\gets a+1$. If $a=n$, output ``Inconclusive,'' and terminate the algorithm.
+Otherwise, go to step F2.\slug
+
+Let's see a couple of examples. Suppose we want to find out if $9$ is prime or not (we both know it isn't
+but play along for a second). In the very first step loop, when $a=2$, we have
+$$2^{9-1} = 2^8 = (2^4)^2 = 16^2 \equiv (-2)^2 = 4\pmod 9,$$
+so since $4 \not\equiv 1\pmod 9$, we conclude right away that $9$ is not prime.
+
+A harder one. Is $341$ prime? We start with $a=2$. Noting that $3\cdot 341 = 1023$, we have
+$$2^{340} = (2^{10})^{34} = 1024^{34} \equiv 1^{34}\equiv 1\pmod{341},$$
+so the test is inconclusive after the first loop. Fine. If we try $a=3$ next, we will
+find that $3^{340}\equiv 56\not\equiv 1\pmod{341}$, so we conclude that $341$ is not prime.
+(Those wishing to practise computing large powers modulo $n$ might like to verify the fact
+that $3^{340}\equiv 56\pmod {341}$. Use the squaring trick.)
+
+If $n$ causes Algorithm~F to output ``Inconclusive,'' it is very likely that $n$ is prime. However,
+there are rare cases of composite numbers $n$ that cause Algorithm~F to output ``Inconclusive.''
+These are called {\it Carmichael numbers}. The smallest one is $561$.
+
+\medskip\boldlabel Vernam's one-time pad.
+Alice and Bob want to exchange secret messages over a public channel without anyone
+being able to decipher them. A message can be encoded as a binary string $M$.
+Here is one way that Alice and Bob can securely send encrypted messages to each other.
+It is called Vernam's one-time pad, named for G.~Vernam, an {\mc AT\&T} engineer who patented
+the method in 1917.
+
+First, they need to meet in person and generate a long sequence of random bits. (How can
+one generate random numbers? This is an interesting quasi-philosophical question.) Each of
+them keeps a copy of this sequence, call it $S$.
+
+Now suppose Alice and Bob separate and Alice wants to send the message $M = 0110$ to Bob. Of course,
+she shouldn't send this string directly, as it could be intercepted by foes. She needs to use
+the sequence $S$. Suppose
+the first four bits of $S$ are $1101$. To encrypt her message, Alice adds each bit of $M$
+to the corresponding bit of $S$, modulo $2$. We will denote this by the $\oplus$ operation.
+The encrypted message is $M'=0110\oplus 1101 = 1011$. She sends it to Bob.
+
+Bob receives the string $M' = 1011$. To decode it and recover the message $M$, Bob simply has to
+perform the $\oplus$ operation using the (same) first four bits of $S$ that Alice used
+to encode the message in the first place. Indeed, we see that $1101\oplus 1101 = 0110 = M$.
+The reason that this works is that for any binary string $S$, we have $S\oplus S = 0$,
+so performing the $\oplus$ operation twice always returns the original string.
+
+If they want to send more messages, they simply discard the bits they used before and continue
+using the next few bits of $S$.
+
+The advantage to this approach is that an adversarial third-party, reading $M$', gains
+no information about the original string $M$, provided they do not know the sequence $S$.
+In other words, it is entirely secure.
+The disadvantages of Vernam's one-time pad are that Alice and Bob must meet in person beforehand,
+they need to find a way to generate random bits securely, and they need to periodically
+repeat this process to replenish the bits once they run out.
+
+\medskip\boldlabel The RSA cryptosystem.
+Now we move on to a more sophisticated method of encrypting and decrypting messages.
+It is called the {\it {\mc RSA} cryptosystem}, named for R.~Rivest, A.~Shamir, and L.~Adleman,
+who described the algorithm in 1977.
+
+Here's the scenario. Alice wants people to be able to send her messages such that
+only she can decode them. To do so, she needs to create a {\it public key}, which allows
+people to encrypt their messages before sending them to her, and then a {\it private key}
+that will allow her to decrypt messages. (Only she knows the private key.)
+
+To set things up, Alice needs two large primes, call them $p$ and $q$. (Something like $200$
+bits is sufficient.) Then she computes $n=pq$. Now, she picks some integer $k$ with $\gcd(k,n) = 1$.
+By Theorem~{\thminvertiblemodn}, there is some integer $s$ such that $ks=1$. Alice publishes
+the integer $n$ and the public key $k$. The private key is $s$; this she keeps to herself.
+
+Now Bob wants to send a secret message $M$ to Alice. Considering $M$
+as an integer (in binary), a requirement for the algorithm to work is $M<n$. Bob computes
+$\bar M = M^k\% n$ and sends $\bar M$ to $A$. (We have $0\le \bar M < n$ by the division
+algorithm.)
+
+Alice receives $\bar M$. She computes ${\bar M}^s$. It turns out that this is $M$.
+To prove this, we need a lemma.
+
+\proclaim Lemma \advthm. Let $a$ and $b$ be integers.
+\medskip
+\item{i)} For all positive integers $m$ and $n$,
+if $a\equiv b\pmod{mn}$, then $a\equiv b\pmod m$ and $a\equiv b\pmod n$.
+\medskip
+In the case that $m$ and $n$ are different primes, the converse holds; that is, the following
+holds.
+\medskip
+\item{ii)} For all primes $p$ and $q$ with $p\ne q$, if $a\equiv b\pmod p$
+and $a\equiv b\pmod q$, then $a\equiv b\pmod{pq}$.
+\medskip
+
+\proof Assume that $a\equiv b\pmod{mn}$, so there exists $k\in \ZZ$ such that
+$a-b = kmn$. This means that $a-b\equiv 0\pmod m$ and $a-b\equiv 0\pmod n$,
+so $a\equiv b\pmod m$ and $a\equiv b\pmod n$.
+
+To prove (ii), suppose that there exist $k,l\in \ZZ$ such that $a-b = kp$ and $a-b = lq$.
+We have $kp=lq$, so $p\divides lq$. Because $p$ is prime, $p$ must divide either $l$ or $q$.
+But $q$ is prime as well, so the only divisors of $q$ are $1$ and $q$. Hence $p\divides l$,
+and we may write $l = rp$ for some $r\in \ZZ$. So we have
+$$ a-b = lq = rpq,$$
+implying that $a-b\equiv 0\pmod{pq}$; that is, $a\equiv b\pmod{pq}$.\slug
+
+Now we shall prove that the encrypted message can be decrypted into the original message.
+
+\parenproclaim Theorem {\advthm} ({\mc RSA} encryption). Let $p$ and $q$ be distinct primes and
+let $n=pq$.
+
+\bye

### 2024-10-24 teaching/math240f24/draft.tex
notes oct 24

@@
 
-\bye
-
-\medskip\boldlabel Solving congruences modulo {\mathbold n}.
+\medskip\boldlabel\datestamp{24}{x} Solving congruences modulo {\mathbold n}.
 Suppose we want to solve for all integers $x$ satisfying $x^2 \equiv x\pmod n$, first for $n=6$,
@@
 
-Now we tackle the case where $n=7$. In fact, the solution is no different
-if $n$ were any other prime.
+Now we tackle the case where $n=7$. In fact, the solution would be no different
+should $n$ be any other prime.
 First we subtract $x$ from both sides, obtaining the congruence $x^2 - x\equiv 0\pmod 7$,
@@
 $$\eqalign{
-\sum_{i=1}^{10} i c_i &\equiv_11 \sum_{i=1}^{10} i c_i - 0 \cr
-&\equiv_11 \sum_{i=1}^{10} i c_i - \sum_{i=1}^{10} id_i \cr
+\sum_{i=1}^{10} i c_i &\equiv_{11} \sum_{i=1}^{10} i c_i - 0 \cr
+&\equiv_{11} \sum_{i=1}^{10} i c_i - \sum_{i=1}^{10} id_i \cr
 &= \sum_{i=1}^{10} i(c_i - d_i) \cr
@@
 is in the range $[-10,10]$, and we assumed that $d_{j-1}\ne d_j$, so this difference is not zero.
-We conclude that $d_{j-1}-d_j\not\equiv 0\pmod 11$, so the code is not valid.\slug
+We conclude that $d_{j-1}-d_j\not\equiv 0\pmod{11}$, so the code is not valid.\slug
 
@@
 
-\medskip\boldlabel Computing large powers modulo {\mathbold p}.
+\medskip\boldlabel Computing large powers modulo {\mathbold n}.
 The next application is not a precise mathemathical statement; rather, it's an observation
@@
 always equal $1$ in this case.)
-\algstep F2. [Compute power.] Set $b\gets a^{n-1} \% n$. (So $0\le b< n$.)
+\algstep F2. [Compute power.] Set $b\gets a^{n-1} \percent n$. (So $0\le b< n$.)
 \algstep F3. [Not prime?] If $b \neq 1$, output ``Not prime,'' and terminate the algorithm.
@@
 Otherwise, go to step F2.\slug
+
+\bye
 
@@
 To set things up, Alice needs two large primes, call them $p$ and $q$. (Something like $200$
-bits is sufficient.) Then she computes $n=pq$. Now, she picks some integer $k$ with $\gcd(k,n) = 1$.
-By Theorem~{\thminvertiblemodn}, there is some integer $s$ such that $ks=1$. Alice publishes
+bits is sufficient.) Then she computes $n=pq$. Now, she picks some integer $k$ with
+$$\gcd\bigl(k,(p-1)(q-1)\bigr) = 1.$$
+By Theorem~{\thminvertiblemodn}, there is some integer $s$ such that $ks\equiv 1\pmod{(p-1)(q-1)}$.
+Alice publishes
 the integer $n$ and the public key $k$. The private key is $s$; this she keeps to herself.
@@
 as an integer (in binary), a requirement for the algorithm to work is $M<n$. Bob computes
-$\bar M = M^k\% n$ and sends $\bar M$ to $A$. (We have $0\le \bar M < n$ by the division
+$\bar M = M^k\percent n$ and sends $\bar M$ to $A$. (We have $0\le \bar M < n$ by the division
 algorithm.)
@@
 
-Now we shall prove that the encrypted message can be decrypted into the original message.
+Now we shall prove that the encrypted message $\bar M = M^k$
+can be decrypted into the original message.
 
 \parenproclaim Theorem {\advthm} ({\mc RSA} encryption). Let $p$ and $q$ be distinct primes and
-let $n=pq$.
+let $n=pq$. Let $k$ and $s$ be such that $ks\equiv 1\pmod n$. Then for all integers $0\le M < n$,
+$(M^k)^s \equiv M\pmod n$.
+
+\proof Without loss of generality, we can take $k$ and $s$ both positive. From
+$ks\equiv 1\pmod{(p-1)(q-1)}$,
+we know there is an integer $l$ such that $ks-1 =l (p-1)(q-1)$. Since $k,s>0$, this identity
+means that $l>0$ as well. Now, since $n=pq$, we have
+$$(M^k)^s \equiv M^{ks} \equiv M^{1+l(p-1)(q-1)} \equiv M(M^{p-1})^{l(q-1)} \pmod{pq}.$$
+By part (i) of the previous lemma, this means we have the same congruence modulo $p$; i.e.,
+$$(M^k)^s \equiv M(M^{p-1})^{l(q-1)} \pmod p,$$
+and by Fermat's little theorem $M^{p-1}\equiv 1$ for all $M\not\equiv 0\pmod p$, so this simplifies to
+$$\eqalign{
+(M^k)^s &\equiv_p \cases{M, & if $M\not\equiv 0\pmod p$; \cr 0, & otherwise} \cr
+&\equiv_p M.\cr
+}$$
+{\it Mutatis mutandis} (by the roles of $p$ and $q$ in the above), we have
+$(M^k)^s \equiv_q M$ as well, so part (ii) of the previous lemma can now be applied to give
+$$(M^k)^s \equiv M \pmod{pq}.\noskipslug$$
+
+Here is an example of the {\mc RSA} algorithm in action. Alice sets $p=3$ and $q=11$,
+so that $n=33$. This means that $(p-1)(q-1) = 2\cdot 10 = 20$. Using $k=7$ (which we can
+do since $\gcd(7,20) = 1$, we have $s=3$, since $3\cdot 7 = 21\equiv 1\pmod{20}$.
+Alice publishes the integer $n=33$ as well as the public key $k=7$. She keeps the number $s=3$ private.
+
+The possible messages $M$ one can send to Alice are $0,1,\ldots, 32$, since we need $0\le M<n$.
+Suppose Bob wants to send the message $00010$ to Alice; that is, we have $M=2$. Bob computes
+$$M^k = 2^7 = 2^5\cdot 2^2 = 32\cdot 4 \equiv (-1)\cdot 4 \equiv 29\pmod{33},$$
+and sends Alice $\bar M = 29$.
+
+Now Alice computes
+$$\bar M^s = 29^3 \equiv (-4)^3 = -64 \equiv -64 + 66 = 2\pmod{33}$$
+and recovers the original message $M=2$.
+
+Now imagine that there is a third person, Charlie, who wants to thwart Alice and Bob.
+The integer $n$ and the public key $n$ are known to all,
+including Charlie. If he wants to decrypt Bob's message, all Charlie needs to know is the
+the private key $s$. In this case, since $n$ is very small, the Adversary can simply
+factor $33 = 3\cdot 11$ to find out that $(p-1)(q-1) = 20$, and then use Euclid's Algorithm
+to find the inverse of $7$ modulo $20$, and he will be able to decrypt all messages between
+Alice and Bob.
+
+In real life, {\mc RSA} is implemented using very large primes $p$ and $q$.
+This makes it difficult, but not impossible, to break the encryption. To put things
+in perspective, it took a team of researchers $900$ core-years of computing time to
+factorise the $240$-digit ($795$-bit) modulus $n$ that is used
+ in {\mc RSA}-240 encryption. These researchers estimated
+that increasing the number of bits to $1024$ would make their factorisation program
+take $500$ times as long to crack the encryption.
+
+However, this does not mean there doesn't exist a fast algorithm for factorising
+integers. We just haven't found one yet (or if people have, they are keeping the algorithm
+to themselves). Most computer scientists believe that this is impossible though. It
+is conjectured
+that integer factorisation is a difficult problem---that is, there is no algorithm to
+factorise a $b$-bit integer in time polynomial in $b$. Until this conjecture is proved,
+all of the data we rely on encryption to protect could theoretically be in jeopardy.
+
+\vfill\eject
+\begingroup\headline{\hfil}\footline{\hfil}
+\vglue160pt
+\centerline{\titlefont III. GRAPH THEORY}
+\vskip220pt
+\bigskip
+\begingroup\obeylines\eightssi
+\hfill Quamobrem, cum nuper problematis cuiusdam mentio esset facta,
+\hfill quod quidem ad geometriam pertinere videbatur,
+\hfill at ita erat comparatum, ut neque determinationem quantitatum requieret,
+\hfill neque solutionem calculi quantitatum ope admitteret,
+\hfill id ad geometriam situs referre haud dubitavi:
+\hfill praesertim quod in eius solutione solus situs in considerationem veniat,
+\hfill calculus vero nullius prorsus sit usus.
+\eightss
+\smallskip
+\hfill --- LEONHARD EULER {\eightssi ``Solutio problematis ad geometriam situs pertinentis''} (1741)
+\endgroup%\obeylines
+\bigskip\goodbreak
+\vfill\eject
+
+\advsect Definitions and basic notions
+
+A {\it graph} is a pair $G = (V,E)$ where $V$ is a nonempty set and
+$$E \subseteq \bigl\{ \{u,v\} : u,v\in V\bigr\}.$$
+The elements of $V$ are called {\it vertices} and the elements of $E$ are
+called edges.
+If $e = \{u,v\}\in E$, we say that $u$ and $v$ are {\it adjacent}, and
+$e$ is {\it incident on} $u$ and $v$. For brevity, we often write $uv$ instead
+of $\{u,v\}$.
+
+For instance, we might have $V = \{1,2,3,4\}$ and
+$$E = \{12, 13, 14, 24\}.$$
+We can draw the graph $G = (V,E)$ as follows:
+\midinsert
+$$\epsfbox{firstexample.ps}$$
+\endinsert
+\goodbreak
+More examples of graphs are pictured in the figure below.
+
+\midinsert
+$$\epsfbox{smallgraphs.ps}$$
+\vskip5pt
+\caption{Some small graphs.}
+\endinsert
+\goodbreak
+
+We can define graphs in a more complicated manner. Let $G = (V,E)$
+be the graph with $V = \ZZ/5\ZZ$ and $ab\in E$ if and only if
+$$a + b \equiv 1\pmod 5\qquad\hbox{or}\qquad a+b \equiv 3\pmod 5.$$
+Here is a drawing of $G$:
+\midinsert
+$$\epsfbox{modfiveexample.ps}$$
+\endinsert
+\goodbreak
 

### 2024-10-29 teaching/math240f24/draft.tex
lecture 29 oct

@@
 \def\percent{\mathrel{\%}}
-
-\vglue24pt
+\def\norm#1{|\!|#1|\!|}
+\def\dist{\op{\rm dist}}
 
@@
 \hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 11.\kern .5em Applications of
+number theory}{60}}
+\medskip
+\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
+{{\tenssbx III. Graph theory \kern 3pt}}{68}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 12.\kern .5em Definitions and
+basic notions}{69}}
 
@@
 
+\edef\propfrogs{\the\thmcount}
 \proclaim Proposition \advthm. Let $a$ and $b$ be nonzero integers. The set
@@
 
-\bye
-
-Let's see a couple of examples. Suppose we want to find out if $9$ is prime or not (we both know it isn't
+\noindent\datestamp{29}{x} \hskip\parindent
+Let's see a couple of examples.
+Suppose we want to find out if $9$ is prime or not (we both know it isn't
 but play along for a second). In the very first step loop, when $a=2$, we have
@@
 if $a\equiv b\pmod{mn}$, then $a\equiv b\pmod m$ and $a\equiv b\pmod n$.
-\medskip
+\medskip\noindent
 In the case that $m$ and $n$ are different primes, the converse holds; that is, the following
@@
 A {\it graph} is a pair $G = (V,E)$ where $V$ is a nonempty set and
-$$E \subseteq \bigl\{ \{u,v\} : u,v\in V\bigr\}.$$
+$$E \subseteq \bigl\{ \{u,v\} : u,v\in V, u\ne v\bigr\}.$$
+The graph $G$ is said to be {\it finite} if and only if $V$ is finite.
 The elements of $V$ are called {\it vertices} and the elements of $E$ are
@@
 $e$ is {\it incident on} $u$ and $v$. For brevity, we often write $uv$ instead
-of $\{u,v\}$.
-
-For instance, we might have $V = \{1,2,3,4\}$ and
+of $\{u,v\}$. Note that an edge is defined to be a $2$-element set, not an ordered
+pair, so $uv = vu$.
+
+Let's start with a small example. Let $V = \{1,2,3,4\}$ and
 $$E = \{12, 13, 14, 24\}.$$
@@
 \goodbreak
-More examples of graphs are pictured in the figure below.
-
-\midinsert
-$$\epsfbox{smallgraphs.ps}$$
-\vskip5pt
-\caption{Some small graphs.}
-\endinsert
-\goodbreak
-
-We can define graphs in a more complicated manner. Let $G = (V,E)$
-be the graph with $V = \ZZ/5\ZZ$ and $ab\in E$ if and only if
+%More examples of graphs are pictured in the figure below.
+%\midinsert
+%$$\epsfbox{smallgraphs.ps}$$
+%\vskip5pt
+%\caption{Some small graphs.}
+%\endinsert
+%\goodbreak
+
+Instead of listing out all the vertices,
+we can define graphs by setting conditions for when a pair of vertices should
+be connected by an edge.
+For instance, let $G = (V,E)$
+be the graph with $V = \ZZ/5\ZZ$ and, for $a,b\in V$ with $a\ne b$,
+we put $ab\in E$ if and only if
 $$a + b \equiv 1\pmod 5\qquad\hbox{or}\qquad a+b \equiv 3\pmod 5.$$
@@
 \goodbreak
-
+Turns out this graph was simpler than it might have seemed at first glance.
+Here is an alternate drawing of $G$:
+\midinsert
+$$\epsfbox{modfiveexamplesimple.ps}$$
+\endinsert
+\goodbreak
+Note that when we refer to $G$, we are talking
+about a set of vertices and edges, not any particular drawing!
+There are infinitely many valid drawings of any graph.
+
+Suppose $G = (V,E)$ has a finite number of vertices. Then we can put the elements
+of $V$ in bijection with $\{1,\ldots, n\}$. Without loss of generality, we can simply assume
+that $\{1,\ldots,n\}$ {\it is} the vertex set of $G$.
+The {\it adjacency matrix} of $G$ is then defined to be the symmetric $n\times n$ matrix
+obtained by setting
+$$A_{ij} = \cases{ 1, & if $ij\in E$,\cr 0, & otherwise.}$$
+\endgroup%\headline{\hfil}\footline{\hfil}
+For instance, the adjacency matrix of the graph on $\ZZ/5\ZZ$ defined earlier is
+$$ A = 
+\pmatrix{
+0 & 1 & 0 & 0 & 1 \cr
+1 & 0 & 0 & 1 & 0 \cr
+0 & 0 & 0 & 0 & 1 \cr
+0 & 1 & 0 & 0 & 0 \cr
+1 & 0 & 1 & 0 & 0 \cr
+}.$$
+
+We now define certain important families of graphs. For any integer $n\ge 1$, we define
+the {\it complete graph} $K_n$ to be the graph on $n$ vertices with every possible edge present.
+Small examples are illustrated in Fig.~2.
+\midinsert
+$$\epsfbox{kn.ps}$$
+\vskip5pt
+\caption{Complete graphs $K_n$ for small $n$.}
+\endinsert
+\goodbreak
 \bye
+For an integer $n\ge 1$,
+we define the {\it Hamming cube} or {\it hypercube} to be the graph $Q_n$
+whose vertex set is $\{0,1\}^n$, the set of all binary strings of length $n$,
+and whose edge set is the exactly the set of pairs of strings that differ by exactly
+one bit. Small examples are illustrated in Fig.~3. (As an exercise, try to figure out
+which strings correspond to which vertices!)
+\midinsert
+$$\epsfbox{qn.ps}$$
+\vskip5pt
+\caption{Hypercubes $Q_n$ for small $n$.}
+\endinsert
+\goodbreak
+
+\medskip\boldlabel Degrees and {\mathbold k}-regularity.
+The {\it neighbours} of a vertex $v$ are all $u\in V$ such that $uv\in E$.
+The {\it degree} of a vertex $v$, denoted by $\deg(v)$, is the number of neighbours
+that $v$ has. A graph is said to be {\it $k$-regular} for some $k\in \NN$ if every $v\in V$ has
+degree $k$. For example, let $C_n$ denote the graph on $V = \ZZ/n\ZZ$ where
+we connect $a\in V$ and $b\in V$ if and only if $a-b\equiv 1\pmod n$
+or $a-b\equiv -1\pmod n$. Every vertex $a$ has exactly two neighbours, namely,
+the congruence classes of $a-1$ and $a+1$ modulo $n$. So $C_n$ is $2$-regular for all $n$.
+In fact, complete graphs and hypercubes are also regular, for in $K_n$, each vertex
+is connected to each of the $n-1$ other vertices, and in $Q_n$, each string is
+connected to the $n$ other strings obtained by flipping each of its bits.
+
+The following theorem relates vertex degrees to the number of edges.
+
+\edef\thmdegformula{\the\thmcount}
+\proclaim Theorem \advthm. Let $G = (V,E)$ be a finite graph. Then
+$$\sum_{v\in V} \deg(v) = 2|E|.$$
+
+\proof Consider each edge $e = uv \in E$. It contributes $+2$ to the right-hand side.
+On the left-hand side, it contributes $+1$ to the degree $\deg(u)$ of $u$,
+and $+1$ to the degree $\deg(v)$ of $v$.\slug
+
+The proof of this theorem is an excellent example of a proof by double counting, where
+we have proved an identity by showing that both sides are different ways
+of counting the same thing. This theorem has the following corollary.
+
+\parenproclaim Corollary {\advthm} (Handshaking lemma).
+In every finite simple graph, the number of vertices having odd degree is even.
+
+\proof We partition $V = V_o \cup V_e$, where $V_o$ comprises all vertices of
+odd degree and $V_e$ comprises all vertices of even degree. By the previous theorem,
+we have
+$$2|E| = \sum_{v\in V} \deg (v) = \sum_{v\in V_o} \deg(v) + \sum_{v\in V_e} \deg(v).$$
+Taking this whole identity modulo $2$, we have
+$$0 \equiv \sum_{v\in V_o} 1 + \sum_{v\in V_e} 0\pmod 2.$$
+Hence
+$$|V_o| = \sum_{v\in V_o} 1 \equiv 0\pmod 2,$$
+which is what we wanted.\slug
+
+This proposition is often called the handshaking lemma (sometimes this is also used
+to refer to the preceding theorem),
+since it implies that at any gathering, the number of people who shake hands with
+an odd number of people is even. From Theorem~{\thmdegformula} we also able to derive
+a corollary that counts the number of edges in $k$-regular graphs.
+
+\proclaim Corollary \advthm. Let $G = (V,E)$ be $k$-regular. Then
+$$|E| = {k|V|\over 2}.$$
+
+\proof By Theorem~{\thmdegformula}, we have
+$$2|E| = \sum_{v\in V}\deg (v),$$
+but $\deg(v) = k$ for all $v\in V$ by $k$-regularity. Hence
+$$2|E| = \sum_{v\in V} k = k|V|,$$
+and the result follows upon rearranging.\slug
+
+\medskip\boldlabel\datestamp{18}{} Walks, paths, and cycles.
+A {\it walk} in $G = (V,E)$ is a sequence $\sigma$ of vertices
+$$\sigma = (v_0, v_1, \ldots, v_n)$$
+such that $v_i v_{i+1}\in E$ for all $0\le i \le n-1$. The {\it endpoints}
+of $\sigma$ are $v_0$ and $v_n$, and the {\it length} of $\sigma$, denoted $|\sigma|$,
+is the number of edges traversed, namely, $n$.
+The walk is said to be {\it closed} if $v_0 = v_n$ and {\it open} otherwise.
+A walk is called a {\it path} if no vertex is repeated.
+
+\proclaim Theorem \advthm. Let $G = (V,E)$ be a graph. If $u$ and $v$ are vertices
+such that there exists a walk from $u$ to $v$, then there exists a path from $u$ to $v$.
+
+\proof We perform a minimality argument.
+Let $\sigma = (v_0, v_1,\ldots, v_n)$ be a walk from $u$ to $v$ (so
+$u = v_0$ and $v = v_n$) of shortest length. We claim that $\sigma$ is in fact a path.
+Indeed, suppose for a contradiction that it is not a path; then there is some repeated vertex,
+so there exist $i,j\in\{0,1,\ldots,n\}$ such that $i<j$ and $v_i = v_j$. Hence
+there is no need to visit any of the vertices between $v_i$ and $v_j$ in $\sigma$,
+since $v_i = v_j$ is connected to $v_{j+1}$. Concretely, consider the walk
+$$\sigma' = (v_0, v_1,\ldots, v_i, v_{j+1}, \ldots, v_n).$$
+Note that $|\sigma'| = |\sigma| - (j-i)$, and $j-i > 0$, so $\sigma'$ is a shorter walk
+from $u$ to $v$. But this contradicts our choice of $\sigma$ as a walk from
+$u$ to $v$ of shortest length. We conclude that $\sigma$ is a path.\slug
+
+A {\it cycle} is a walk $\sigma = (v_0,v_1,\ldots,v_n)$ of length at least $3$
+with $v_0 = v_n$ and no vertex repeated except $v_0 = v_n$.
+
+\edef\propoddwalkoddcycle{\the\thmcount}
+\proclaim Proposition \advthm. Let $G = (V,E)$. If $G$ contains a closed walk
+of odd length, then it contains a cycle of odd length.
+
+\proof The idea is similar to the proof of the previous theorem. We perform a
+minimality argument. Let
+$\sigma = (v_0, v_1,\ldots,v_n)$ be an odd-length closed walk in $G$, and choose
+this walk to have minimal odd length (i.e, any shorter walk has even length). We shall
+prove that $\sigma$ is a cycle.
+
+For a contradiction, suppose that $\sigma$ is not a cycle, so that
+there exist indices $i$ and $j$ with $i<j$ such that $v_i = v_j$.
+Consider the two closed walks
+$$\sigma_1 = (v_0, v_1, \ldots, v_i, v_{j+1}, \ldots, v_n)$$
+and
+$$\sigma_2 = (v_i, v_{i+1}, \ldots, v_j).$$
+Both are shorter than $\sigma$, so by the minimality of $\sigma$,
+$|\sigma_1|$ and $|\sigma_2|$ are even. But this implies that
+$|\sigma| = |\sigma_1|+|\sigma_2|$ is even. This contradicts the hypothesis that $|\sigma|$
+is odd.
+
+We conclude that $\sigma$ is a cycle.\slug
+
+\medskip\boldlabel Connectedness.
+We say a graph $G = (V,E)$ is {\it connected} if for all $u,v\in V$ there exists
+a walk (and hence a path) from $u$ to $v$. All of the graphs we have
+drawn so far have been connected. Here is a graph that is disconnected:
+\midinsert
+$$\epsfbox{disconnected.ps}$$
+\endinsert
+\noindent This picture is meant to depict one graph on $8$ vertices, not three
+different graphs.
+
+The small examples of $K_n$ and $Q_n$ in Figs.~2 and~3 were all connected,
+so it is natural to assume that $K_n$ and $Q_n$ are connected for all $n$.
+This is indeed the case.
+
+\proclaim Proposition \advthm. For all $n\ge 1$, the graphs $K_n$ and $Q_n$
+are connected.
+
+\proof Any vertices $u$ and $v$ in the vertex set of $K_n$ belong to the
+edge $uv$, so we have the path $(u, v)$ of length $1$ between $u$ and $v$.
+This shows that $K_n$ is connected.
+
+Now let $u$ and $v$ be any two vertices in $K_n$. Suppose there are $m$
+bits that differ between $u$ and $v$. Then we can flip them one by one to
+change $u$ into $v$. This represents a path of length $m$ from $u$ to $v$, since there
+is an edge in $Q_n$ between any two strings in $Q_n$ that differ at exactly one
+bit.\slug
+
+Let $G = (V,E)$ and
+let $u,v\in V$. We say that {$v$ is reachable from $u$} if there exists a walk
+from $u$ to $v$. If we define a relation $\sim$ on $V$ by setting
+$u\sim v$ if and only if $v$ is reachable from $u$, then $\sim$ is an equivalence
+relation (prove this as an exercise!) and the equivalence
+classes are called the {\it connected components} of $G$.
+
+Here is an example. Let $G = (\ZZ, E)$, where for $i,j\in \ZZ$ with $i<j$, we have
+$ij\in E$ if and only if $j-i \in \{9,15\}$. That is, every edge either
+hops forwards or backwards in the number line in increments of $9$ or $15$.
+Starting at $n\in \ZZ$, we can reach any vertex $m$ that is of the form
+$$ m = n + 15s + 9t$$
+for some $s,t\in \ZZ$. By Proposition~{\propfrogs} (this was the proposition
+about frogs and lilypads, if that jogs your memory), the integers
+representable
+as $15s+9t$ for some $s,t\in \ZZ$ are exactly the multiples of
+$\gcd(9,15) = 3$. So in fact, from $n$ one can reach any integer $m$
+of the form $n+3l$ for some $l\in \ZZ$; that is, one can reach any integer
+$m$ with $m\equiv n\pmod 3$. So the three connected components in this
+graph are $[0]_3$, $[1]_3$, and $[2]_3$, the equivalence classes of integers
+modulo $3$.
+
+\advsect Triangles and bipartite graphs
+
+A {\it subgraph} of $G = (V,E)$ is a graph $G' = (V',E')$ such that
+$V'\subseteq V$ and $E'\subseteq E$ where for all $e = uv\in E'$, we have
+$u,v\in V'$. For example, the {\it triangle graph} $C_3$ is a subgraph of $K_4$ (draw it!).
+In fact, any graph on $n$ vertices or less is a subgraph of $K_n$, since
+$K_n$ contains all possible edges among $n$ vertices.
+
+We now consider an extremal question concerning
+subgraphs. An extremal question is one of the form, {\sl What is the
+extremal (maximum or minimum) number of objects we can have, subject to some restriction?}
+The extremal question we ask is, {\sl What is the maximum number of edges that a graph
+on $n$ vertices can have, assuming that it does not contain a triangle as a subgraph?}
+To answer this question, we need the following inequality, which is one of the
+most important inequalities in all of mathematics. It is named for A.~Cauchy,
+who in 1821 proved the version we shall record below,
+and H.~Schwarz, who proved a version for integrals in 1888.
+
+\parenproclaim Theorem {\advthm} (Cauchy--Schwarz inequality). For all
+$u_1,\ldots,u_n, v_1,\ldots,v_n\in \RR$, we have
+$$\biggl(\sum_{i=1}^n u_i v_i\biggr)^2
+\le \biggl( \sum_{i=1}^n {u_i}^2\biggr) \biggl( \sum_{i=1}^n {v_i}^2\biggr).$$
+
+In the proof below, you will need to recall from your high school days that
+for a quadratic polynomial $ax^2 + bx + c$, the {\it discriminant} is
+defined to be the quantity $b^2 - 4ac$. (It appears under the square-root sign
+in the quadratic formula.) If it is negative, the polynomial has no real roots,
+if it is positive, the polynomial has two real roots, and if it is zero, the polynomial
+has one real root.
+
+\proof If $u_i = 0$ for all $1\le i\le n$, then both sides are $0$ and the inequality
+is trivially true. So suppose that at least one of the $u_i$ is nonzero. Then
+$$p(x) = (u_1 x + n_1)^2 + (u_2 x + v_2)^2 + \cdots + (u_n x+v_n)^2$$
+is a quadratic polynomial in the variable $x$. Rewriting $p(x)$ in summation notation,
+we have
+$$\eqalign{
+p(x) &= \sum_{i=1}^n (u_ix + v_i)^2\cr
+&= \sum_{i=1}^n \bigl( u_i^2 x^2 + 2u_i v_i + v_i^2\bigr)\cr
+ &= \biggl(\sum_{i=1}^n u_i^2 \biggr) x^2 +
+2\biggl(\sum_{i=1}^n u_i v_i\biggr) x + \biggl( \sum_{i=1}^n {v_i}^2\biggr).\cr
+}$$
+Since it is defined as a sum of squares, $p(x)\ge 0$ for all $x\in \RR$, so it has
+at most one real root.
+This means its discriminant is nonpositive; in other words,
+$$4\biggl(\sum_{i=1}^n u_i v_i\biggr)^2
+- 4\biggl( \sum_{i=1}^n {u_i}^2\biggr) \biggl( \sum_{i=1}^n {v_i}^2\biggr) \le 0.$$
+Dividing through by $4$ and then rearranging gives exactly the inequality we sought.\slug
+
+The Cauchy--Schwarz inequality can also be proved using some results you might
+know from linear algebra.
+Defining vectors $u,v\in \RR^n$ by
+$u = (u_1,\ldots,u_n)$ and $v = (v_1,\ldots,v_n)$, the left-hand side of the
+Cauchy--Schwarz inequality is
+$(u\cdot v)^2$ and the right-hand side is $\norm{u}^2\norm{v}^2$. But,
+letting $\theta$ is the angle between $u$ and $v$,
+then $u\cdot v = \cos\theta \norm{u}\norm{v}$, and applying the fact that
+$|\cos\theta|\le 1$, we easily deduce the desired inequality.
+
+We are now able to derive the maximum number of edges in a triangle-free graph on
+$n$ vertices. The following theorem was proved by W.~Mantel in 1907.
+
+\parenproclaim Theorem {\advthm}. (Mantel's theorem). Let $G = (V,E)$ be a graph
+not containing a triangle as a subgraph. Then
+$$|E| \le \biggl\lfloor {|V|^2\over 4}\biggr\rfloor.$$
+
+\proof Consider the sum
+$$\sum_{uv\in E} \bigl(\deg(u)+\deg(v)\bigr).$$
+The term $\deg(u)$ appears in the sum exactly once for every edge incident on $u$;
+that is, it appears $\deg(u)$ times. This is true for all $u\in V$, so
+we conclude that
+$$\sum_{uv\in E} \bigl(\deg(u)+\deg(v)\bigr) = \sum_{u\in V} \deg(u)^2.$$
+On the other hand, since $G$ contains no triangle, for every pair of vertices $u$
+and $v$, the set of neighbours of $u$ is disjoint from the set of neighbours of $v$.
+So $\deg(u)+\deg(v)\le |V|$, and we have
+$$\sum_{u\in V} \deg(u)^2 = \sum_{uv\in E} \bigl(\deg(u)+\deg(v)\bigr)\le |V|\cdot |E|.$$
+By Theorem~{\thmdegformula} and the Cauchy--Schwarz inequality in that order,
+we have
+$$\eqalign{
+\bigl(2|E|\bigr)^2 &= \Bigl(\sum_{u\in V} \deg(u)\Bigr)^2 \cr
+&= \Bigl(\sum_{u\in V} \deg(u)\cdot 1\Bigr)^2 \cr
+&\le \Bigl(\sum_{u\in V} \deg(u)^2\Bigr) \Bigl(\sum_{u\in V} 1^2\Bigr) \cr
+&= |V| \Bigl(\sum_{u\in V} \deg(u)^2\Bigr).\cr
+}$$
+Hence
+$$4|E|^2 \le |V| \Bigl(\sum_{u\in V} \deg(u)^2\Bigr) \le |V|^2\cdot |E|.$$
+Rearranging this gives us
+$$|E|\le {|V|^2\over 4},$$
+and we can take the floor function on the right-hand side, since $|E|$ must be an
+integer.\slug
+
+So if $G = (V,E)$
+has $|E|\ge \lfloor |V|^2/4\rfloor$, there must be a triangle subgraph in $G$.
+Now let $|V|$ be even, so that $|V|^2/4$ is an integer. Does there exist a graph
+$G$ with exactly $|V|^2/4$ edges such that $G$ does not contain a triangle
+as a subgraph? This brings us to the definition of a bipartite graph.
+
+\medskip\boldlabel\datestamp{19}{} Bipartite graphs.
+A graph $G = (V,E)$ is called {\it bipartite} if there is a partition
+$V = A\cup B$ of the vertex set (so $A\cap B = \emptyset$) called
+the {\it bipartition} such that each edge has one endpoint in $A$ and one endpoint
+in $B$. For example, hypercubes are bipartite.
+
+\proclaim Proposition \advthm. For all $n\ge 1$, the hypercube $Q_n$ is bipartite.
+
+\proof Let $Q_n = (V,E)$. Every element $s\in V$ corresponds to a string
+$S = (s_1, \ldots, s_n)$ of bits, where $s_i\in \{0,1\}$ for all $1\le i\le n$.
+Define
+$$V_0 = \bigl\{ s\in V : s_1 + \cdots + s_n \equiv 0 \pmod 2\bigr\}$$
+and
+$$V_1 = \bigl\{ s\in V : s_1 + \cdots + s_n \equiv 1 \pmod 2\bigr\}.$$
+It is clear that $V_0 \cup V_1 = V$ and $V_0 \cap V_1 = \emptyset$, so
+this is a bipartition of the vertex set, and for every $e = s_1s_2\in E$,
+the strings $s_1$ and $s_2$ differ in exactly one bit, so if $s_1\in V_0$,
+then $s_2\in V_1$, and vice versa.\slug
+
+The {\it complete bipartite graph} $K_{m,n}$ is a graph with $V = V_m \cup V_n$, where
+$|V_m|=m$ and $|V_n| = n$, and $E$ is the set $\{uv : u\in V_m, v\in V_n\}$ of
+all possible edges between $V_m$ and $V_n$. Fig.~4 illustrates two small examples
+of complete bipartite graphs.
+
+\midinsert
+$$\epsfbox{kmn.ps}$$
+\vskip5pt
+\caption{The complete bipartite graphs $K_{2,3}$ and $K_{4,3}$.}
+\endinsert
+\goodbreak
+When $n$ is even, the graph $K_{n/2,n/2}$ has $n$ vertices and exactly $n^2/4$ edges.
+When $n$ is odd, the graph $K_{(n+1)/2, (n-1)/2}$ has $n$ vertices and
+$$ {n+1\over 2} \cdot {n-1\over 2} = {n^2-1\over 4} = {n^2\over 4}-{1\over 4}
+= \biggl\lfloor {n^2\over 4} \biggr\rfloor$$
+edges. (The last equality here follows from the fact that any odd $n$ is congruent to either $1$ or
+$3$ modulo $4$, which means its square is congruent to $1$ modulo $4$.)
+
+The calculation above shows that $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$
+always has $n$ vertices and $\lfloor n^2/4\rfloor$ edges. It is the largest possible number
+of edges such that Mantel's theorem does not apply; that is, from Mantel's theorem alone,
+we do not know whether
+there must be a triangle in $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$, but adding
+a single edge to $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$, Mantel's theorem tells
+us there must be a triangle in the resulting graph.
+
+We will actually be able to prove that $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$ does not
+contain any triangles. This means that the lower bound of $\lfloor n^2/4\rfloor$ in Mantel's theorem
+is the best possible one.
+
+First we need a lemma.
+
+\proclaim Lemma \advthm. A graph $G=(V,E)$ is bipartite if and only if all of its connected components
+are bipartite.
+
+\proof Let $G_1 = (V_1, E_1), G_2 = (V_2,E_2),\ldots, G_n = (V_k, E_k)$ be the
+connected components of $G$.
+
+If each connected component is bipartite, say,
+$V_i = A_i\cup B_i$ is a bipartition of $V_i$ for all $1\le i\le n$, then
+$A = \bigcup_{i=1}^n A_i$ and $B = \bigcup_{i=1}^n B_i$ is a bipartition of $V$,
+and every $e\in E$ has one endpoint in $A$ and one endpoint in $B$.
+
+Now suppose that there is some component $G_i$ that is not bipartite.
+Now let $V = A\cup B$
+be any partition of $V$ into disjoint nonempty sets. Then $A_i = A\cap V_i$
+and $B_i = B\cap V_i$ form a partition of $V_i$
+into two nonempty sets. But $G_i$ is not bipartite, so there is some $e\in E$
+that either has both endpoints in $A_i$ or both endpoints in $B_i$. This means
+that $G$ is not bipartite either.\slug
+
+As a matter of fact, we shall prove something much stronger than just the fact that
+$K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$ does not contain any triangles.
+The following theorem characterises the family of bipartite graphs as {\it exactly} those graphs not
+containing odd-length cycles.
+Its proof shall require the following metric imposed on a graph $G$. For all $u,v\in V$
+define the {\it distance} $\dist(u,v)$ between $u$ and $v$ to be the length of the shortest path
+from $u$ to $v$. If no such path between $u$ and $v$ exists, we set $\dist(u,v) = \infty$.
+
+\proclaim Theorem \advthm. A graph is bipartite if and only if it does not
+contain any cycles of odd length.
+
+\proof First, assume that $G=(V,E)$ is bipartite; let $V = A\cup B$ be the bipartition
+of the vertex set. Let $\sigma$ be a cycle in $G$. Each edge changes side (either
+$A$ to $B$ or $B$ to $A$), so in order for the starting and ending vertices of this
+cycle to be the same, $|\sigma|$ must be even.
+
+Now assume that $G = (V,E)$ has no odd-length cycles. To show that $G$ is bipartite,
+it suffices to show that each of its connected components is bipartite (by the
+previous lemma), so without
+loss of generality we may assume that $G$ is connected.
+This means that $\dist(u,v)< 0$ for all $u,v\in V$.
+
+Select one vertex $h\in V$ and set
+$$ V_0 = \{v\in V : \dist(h,v)\equiv 0\pmod 2\}$$
+and
+$$ V_1 = \{v\in V : \dist(h,v)\equiv 1\pmod 2\}.$$
+Let $e = uv\in E$.
+Consider a closed walk $\sigma$ that follows a shortest path from $u$ to $h$, then a shortest
+path from $h$ to $v$, and finally the edge $e$ going from $v$ to $u$. We have
+$$|\sigma| = \dist(u,h) + \dist(h,v) + 1.$$
+But since $G$ contains no odd-length cycles, by (the contrapositive of)
+Proposition~{\propoddwalkoddcycle}, any closed walk in $G$ must have even length.
+This means that $\dist(u,h)$ and $\dist(h,v)$ are not the same modulo $2$; that is,
+either $u\in V_0$ and $v\in V_1$ or vice versa.\slug
+
+\advsect Trees
+
+A graph $G$ is called a {\it forest} if it has no cycles. A connected forest is called
+a {\it tree}. A vertex of degree $1$ in a forest is called a {\it leaf}. Fig.~5 contains
+an example of a tree.
+
+\midinsert
+$$\epsfbox{spidery.ps}$$
+\vskip5pt
+\caption{A tree with $25$ leaves (count them!).}
+\endinsert
+\goodbreak
+Our first proposition
+concerning trees is that every tree containing at least one edge has at least two leaves.
+
+\edef\proptwoleaves{\the\thmcount}
+\proclaim Proposition \advthm. Let $F$ be a forest with at least one edge (and hence at least
+two vertices). There are at least two leaves in $F$.
+
+\proof Let $\sigma = (v_0,v_1,\ldots,v_n)$ be a path of maximum length in $F$.
+Since there is at least one edge in $F$, we have $|\sigma|\ge 1$. The claim is that
+$\deg(v_0) = \deg(v_n) = 1$. Indeed, for a contradiction, assume that one of these
+two endpoints has degree at least two. Without loss of generality, suppose it is
+$v_0$ (we can reverse the path if needed to make this the case). So there exists
+$u\in V$ such that $uv_0$ is an edge. If $u = v_i$ for some $1\le i\le n$, then
+there is a cycle $(v_i, v_0, v_1,\ldots, v_i)$, contradicting the fact that $F$ is a forest.
+So for all $1\le i\le n$, $u\ne v_i$. This means that
+$(u,v_0, \ldots, v_n)$ is a strictly longer path than $\sigma$, contradicting the maximality
+of $\sigma$.\slug
+
+The following gives a characterisation of trees.
+
+\proclaim Proposition \advthm. A graph $G = (V,E)$ is a tree if and only if
+for all $u,v\in V$, there exists a unique path from $u$ to $v$ in $G$.
+
+\proof First we assume that $G$ is a tree. Let $u,v\in V$. Since $G$ is a tree,
+it is connected, so there must exist some path $\sigma$ from $u$ to $v$ in $G$.
+Now we prove that this path is unique. For a contradiction, let $\sigma'$
+be another, different, path from $u$ to $v$ in $G$. Both $\sigma$ and $\sigma'$
+start at $u$, but eventually they are different. Let $v_1$ be the last vertex they have
+in common (this vertex could be $u$).
+Eventually, $\sigma$ and $\sigma'$ must meet again, since they both
+end at $v$. Let $v_2$ be the first time they meet again, after $v_1$ (we could have $v_2 = v$). Now
+let $\sigma$ be the path obtained by following $\sigma$ from $v_1$ to $v_2$, then
+following $\sigma'$ backwards to $v_1$. This is a cycle, by choice of $v_1$ and $v_2$.
+But this gives a contradiction, since there are no cycles in $G$.
+
+Now suppose that for all $u,v\in V$, there exists a unique path from $u$ to $v$ in $G$.
+The fact that there exists a path at all between every two vertices immediately implies
+that $G$ is connected. It remains to prove that there is no cycle in $G$. But this
+is clear, since if $\sigma$ is a cycle in $G$, then picking any two distinct vertices in $\sigma$,
+we have two different paths between those two vertices (one obtained by following $\sigma$
+clockwise, and one obtained by following it anticlockwise).\slug
+
+Another characterisation of trees is that they are exactly the connected graphs with $n-1$
+nodes. To prove this, we first need a lemma.
+
+\parenproclaim Lemma {\advthm} (Detour lemma). Let $G$ be a connected graph and let
+$\sigma$ be a cycle in $G$. If $G'$ is the graph obtained by deleting one edge
+of $\sigma$, then $G'$ is also connected.
+
+\proof Let $e = v_1v_2$ be the deleted edge. Let $u,v$ be any two vertices in $G$.
+Since $G$ is connected, there exists a walk $\tau$ from $u$ to $v$ in $G$.
+We want to show that there is a path from $u$ to $v$ in $G'$. If $\tau$ does not use
+the edge $e$, then it is also a valid path in $G'$ and we are done. If it does use $e$,
+then to get from $v_1$ to $v_2$,
+we detour around $\sigma$ in the other direction, then continue from $v_2$ to $v$, obtaining
+a walk in $G'$ from $u$ to $w$.\slug
+
+\proclaim\datestamp{20}{} Theorem \advthm. A graph $G = (V,E)$ is a tree if and only if
+$|E| = |V|-1$ and $G$ is connected.
+
+\proof We prove the forward implication by induction on $n=|V|$, the number of vertices.
+Concretely, the statement we shall prove is, {\sl For all $n\ge 1$, for all $G=(V,E)$ with
+$|V| = n$, if $G$ is a tree then $|E| = |V|-1$ and $G$ is connected.}
+
+The base case is $n=1$. Then the only possible graph $G$ is simply $\bullet$, a single vertex.
+This is a tree since it is connected and contains no cycles. We have $|E| = 0 = 1-1 = |V|-1$.
+
+For the inductive step, let $n\ge 1$ and assume the statement holds for $n$. Let $G = (V,E)$
+be a graph with $|V| = n+1$, and assume that $G$ is a tree. Since $n\ge 1$,
+$|V| \ge 2$, so there are at least two vertices, and connectedness of $G$ means that there
+is at least one edge in $G$. Hence $G$ is a forest with at least one edge, and by
+Proposition~{\proptwoleaves}, there are at least two leaves in $G$. Pick one of these leaves
+and call it $u$. Form a new graph $G' = (V', E')$ by removing $u$ and the single edge
+incident on it; we have
+
+\bye

### 2024-11-03 teaching/math240f24/draft.tex
hallowe'en lecture

@@
 \goodbreak
-\bye
-For an integer $n\ge 1$,
+\noindent\datestamp{31}{x} \hskip\parindent For an integer $n\ge 1$,
 we define the {\it Hamming cube} or {\it hypercube} to be the graph $Q_n$
@@
 
-\medskip\boldlabel\datestamp{18}{} Walks, paths, and cycles.
+\medskip\boldlabel Walks, paths, and cycles.
 A {\it walk} in $G = (V,E)$ is a sequence $\sigma$ of vertices
@@
 We conclude that $\sigma$ is a cycle.\slug
+\bye
 
@@
 
-Another characterisation of trees is that they are exactly the connected graphs with $n-1$
-nodes. To prove this, we first need a lemma.
+Another characterisation of trees is that they are exactly the connected graphs on $n$ vertices
+with $n-1$ edges. To prove this, we first need a lemma.
 

### 2024-11-06 teaching/math240f24/draft.tex
lecture notes 5 nov

@@
 We conclude that $\sigma$ is a cycle.\slug
-\bye
-
-\medskip\boldlabel Connectedness.
+
+\medskip\boldlabel\datestamp{05}{xi} Connectedness.
 We say a graph $G = (V,E)$ is {\it connected} if for all $u,v\in V$ there exists
@@
 
-\medskip\boldlabel\datestamp{19}{} Bipartite graphs.
+\medskip\boldlabel Bipartite graphs.
 A graph $G = (V,E)$ is called {\it bipartite} if there is a partition
@@
 always has $n$ vertices and $\lfloor n^2/4\rfloor$ edges. It is the largest possible number
-of edges such that Mantel's theorem does not apply; that is, from Mantel's theorem alone,
+of edges such that Mantel's theorem does not apply. In other words, from Mantel's theorem alone,
 we do not know whether
 there must be a triangle in $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$, but adding
-a single edge to $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$, Mantel's theorem tells
-us there must be a triangle in the resulting graph.
-
-We will actually be able to prove that $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$ does not
+a single edge to $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$ results in a graph that must
+have a triangle by Mantel's theorem.
+\bye
+
+Soon we will actually be able to prove that $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$ does not
 contain any triangles. This means that the lower bound of $\lfloor n^2/4\rfloor$ in Mantel's theorem

### 2024-11-08 teaching/math240f24/draft.tex
lecture 8 nov

@@
 basic notions}{69}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 13.\kern .5em Triangles and bipartite
+graphs}{74}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 14.\kern .5em Trees}{78}}
 
@@
 have a triangle by Mantel's theorem.
-\bye
-
-Soon we will actually be able to prove that $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$ does not
+
+We will actually be able to prove that $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$ does not
 contain any triangles. This means that the lower bound of $\lfloor n^2/4\rfloor$ in Mantel's theorem
@@
 
-\proclaim Lemma \advthm. A graph $G=(V,E)$ is bipartite if and only if all of its connected components
+\proclaim\datestamp{08}{xi} Lemma \advthm.
+A graph $G=(V,E)$ is bipartite if and only if all of its connected components
 are bipartite.
@@
 
+\bye
+
 Another characterisation of trees is that they are exactly the connected graphs on $n$ vertices

### 2024-11-15 teaching/math240f24/draft.tex
lectures this week

@@
 
-\bigskip
-
 \floattext 4.5 \ninebf A note on these notes.
@@
 
-\vskip60pt
+\vskip48pt
 
@@
 \hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 14.\kern .5em Trees}{78}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 15.\kern .5em Eulerian
+trails and circuits}{81}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 16.\kern .5em Planar graphs}{83}}
 
@@
 
-\proclaim Proposition \advthm. Let $X$ be a finite nonempty set. Let $E$ be the set of all subsets of $X$ with
-even cardinality, and let $F$ be the set of all subsets of $X$ with odd cardinality.
-Then $|E| = |F|$.
-
-\proof We shall construct a function $f:E\to F$. Fix one specific $x\in X$;
+\proclaim Proposition \advthm. Let $X$ be a finite nonempty set.
+Let $E$ be the set of all subsets of $X$ with
+even cardinality, and let $D$ be the set of all subsets of $X$ with odd cardinality.
+Then $|E| = |D|$.
+
+\proof We shall construct a function $f:E\to D$. Fix one specific $x\in X$;
 we can do this because $X\ne \emptyset$. Now, for all $A\in E$, let
@@
 case it is $|A|-1$ and in the second case it is $|A|+1$. This shows that $f$ is indeed a function
-with codomain $F$. Now to prove $|E| = |F|$ we will show that $f$ is bijective, which we
+with codomain $D$. Now to prove $|E| = |D|$ we will show that $f$ is bijective, which we
 do by showing that $f$ is an inverse (as an exercise, you might instead try to prove bijectivity from
@@
 
-We define $g : F\to E$ by
+We define $g : D\to E$ by
 $$g(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
-As before, $\bigl|g(A)\bigr|$ is even, since $A$ is assumed to be a member of $F$ now. Now
+As before, $\bigl|g(A)\bigr|$ is even, since $A$ is assumed to be a member of $D$ now. Now
 for any $A\in E$,
@@
 Let $S$ be the set of these integers that are positive. Symbolically, we have
-$$ S = \{a-kb : k\in \ZZ\ \hbox{and}\ a-kb\ge 0\}.$$
+$$ S = \{a-kb : k\in \ZZ\ \hbox{and}\ a-kb\ge 0\}\phantom{D}.$$
 By the Archimedean property, there is some $n\in \NN$ such that $n> -a$, which
@@
 
-\proclaim\datestamp{08}{xi} Lemma \advthm.
+\proclaim\datestamp{07}{xi} Lemma \advthm.
 A graph $G=(V,E)$ is bipartite if and only if all of its connected components
@@
 
-\bye
-
 Another characterisation of trees is that they are exactly the connected graphs on $n$ vertices
@@
 
-\parenproclaim Lemma {\advthm} (Detour lemma). Let $G$ be a connected graph and let
+\parenproclaim\datestamp{12}{xi}
+Lemma {\advthm} (Detour lemma). Let $G$ be a connected graph and let
 $\sigma$ be a cycle in $G$. If $G'$ is the graph obtained by deleting one edge
@@
 
-\proclaim\datestamp{20}{} Theorem \advthm. A graph $G = (V,E)$ is a tree if and only if
+\proclaim Theorem \advthm. A graph $G = (V,E)$ is a tree if and only if
 $|E| = |V|-1$ and $G$ is connected.
@@
 $|V| = n$, if $G$ is a tree then $|E| = |V|-1$ and $G$ is connected.}
-
 The base case is $n=1$. Then the only possible graph $G$ is simply $\bullet$, a single vertex.
 This is a tree since it is connected and contains no cycles. We have $|E| = 0 = 1-1 = |V|-1$.
-
 For the inductive step, let $n\ge 1$ and assume the statement holds for $n$. Let $G = (V,E)$
@@
 and call it $u$. Form a new graph $G' = (V', E')$ by removing $u$ and the single edge
-incident on it; we have
-
+incident on it; we have $|V'| = n$ and $|E'| = |E|-1$. The graph $G'$ is a tree, because
+for each $v,w\in V'$, there is a unique path from $v$ to $w$ in $G$, and this path
+does not pass through $u$ (since $u$ is a leaf). Hence
+$$ |E| = |E'| + 1  = |V'| - 1 + 1 = |V|-1,$$
+where in the second equality we have used the induction hypothesis.
+
+Now assume that $|E| = |V|-1$ and $G$ is connected. We want to show that $G$ has no cycles.
+For a contradiction, suppose it does; call it $\sigma$. Remove an edge from $\sigma$. By
+the detour lemma, this resulting graph is still connected. If this graph still has a cycle,
+remove an edge from it again, and repeat this process until no cycles remain. This graph
+$G'$ still has $V$ as its vertex set, but it has a new edge set $E'$ with $|E'|< |E|$.
+But $G'$ is now a tree, since it contains no cycles but is still connected, so by
+the previous paragraph, we have $|E'| = |V|-1 = |E|$. The contradiction completes the proof.\slug
+
+We are now able to answer the following extremal question: {\sl How many edges can a graph
+on $n$ vertices have if it does not contain a cycle (i.e., if it is a forest?}
+
+\proclaim Corollary \advthm. Let $F = (V,E)$ be a forest. Then $|E| \le |V|-1$.
+
+\proof Let $k\ge 1$ denote the number of connected components in $F$, and number the
+connected components $C_1, C_2,\ldots,C_k$.
+Join $C_1$ to $C_2$ by an edge, $C_2$ to $C_3$ by an
+edge, and so on. This creates a connected graph $G = (V, E')$ with $|E'| = |E| + k-1$, and $G$ must
+be a tree, since adding these edges does not introduce a cycle. By the previous
+theorem, $|E'| = |V|-1$, so $|E| \le |V|-1$.\slug
+
+\advsect Eulerian trails and circuits
+
+Recall that a walk that does not repeat any vertex is called a path. A similar definition
+is that of a {\it trail}: this is a walk that does not repeat any {\it edge}.
+A walk is called an {\it Eulerian trail} if it uses every edge of $G$ exactly once, and an
+{\it Eulerian circuit} if it is an Eulerian trail and it is closed (its endpoints are
+the same vertex).
+
+The term ``Eulerian'' refers to L.~Euler, who in 1735 answered the following question:
+{\sl Is it possible to cross every bridge in K\"onigsberg (a city in Prussia),
+without having to repeat any bridges?} The situation is illustrated in Fig.~6.
+\midinsert
+$$\epsfxsize=220pt\epsfbox{konigsberg.ps}$$
+\vskip5pt
+\caption{The seven bridges of K\"onigsberg in 1735.}
+\endinsert
+\goodbreak
+To represent this map as a graph $G = (V,E)$, we will have to allow multiple edges between vertices.
+There are a couple ways to formalise this set-theoretically, but we will dispense with the details
+here and simply think of $E$ as a special type of set
+that allows for copies of the same edge to be included multiple times. The graph $G$ representing
+K\"onigsberg is depicted in Fig.~7.
+\midinsert
+$$\epsfbox{konigsberggraph.ps}$$
+\vskip5pt
+\caption{The graph representing the seven bridges of K\"onigsberg.}
+\endinsert
+\goodbreak
+What Euler actually proves in his paper (which did not appear in print until 1741) is that
+it is impossible to cross every bridge in K\"onigsberg exactly once; i.e., $G$ does not
+contain an Eulerian trail.
+First we characterise the graphs that contain Eulerian circuits, as we'll need this to later
+characterise graphs with Eulerian trails.
+
+\proclaim Theorem \advthm. Let $G = (V,E)$ be a connected graph with a finite number of vertices
+and edges, where multiple edges between the same two vertices are allowed. Then
+$G$ has an Eulerian circuit if and only if all vertices have even degree.
+
+\proof Suppose $G$ has an Eulerian circuit $\sigma = (v_0, v_1, \ldots, v_n)$ where
+$v_0 = v_n$. For each $v_i\ne v_0$, each visit uses two edges, one edge to enter $v_i$ and
+one to exit it, so $\deg(v_i)$ is even for all $0\le i < n$. For $v_0 = v_n$, there is one
+edge that leaves it at the beginning, two edges for each visit during the circuit, and then
+one returning to it at the end, so $\deg(v_0)$ is even as well.
+
+We prove the reverse implication by induction on $m = |E|$. That is, we prove
+that for all integers $m\ge 0$, if $G = (V,E)$ with $|E| = m$ and all vertices in $G$ have
+even degree, then $G$ has an Eulerian circuit.
+
+For the base case $m=0$, the stipulation
+that $G$ be connected forces
+$G = \bullet$, and this graph has an Eulerian circuit (the trivial walk).
+
+Now let $m\ge 0$ and suppose the statement is true for all integers less than or equal to $m$.
+Let $G = (V,E)$ with $|E| = m+1$. Since there is at least one edge in $G$ now, $G$ cannot be
+a tree. This is because every tree with at least one edge contains at least two leaves,
+and leaves have (odd) degree $1$. Let $\sigma$ be a cycle in $G$ and form $G' = (V,E')$
+be removing every edge in $\sigma$. This graph may no longer be connected, but since we have removed
+a cycle, every vertex in $G'$ either has the same degree as in $G$ (if the vertex
+was not in $\sigma$), or degree two less than in $G$ (if the vertex was in $\sigma$). So all
+vertices in $G'$ have even degree. Let $k$ be the number of connected components in
+$G'$; number them $H_1,\ldots,H_k$. Each has at most $m$ edges, is connected, and all vertices
+have even degree, so each $H_i$ contains an Eulerian circuit, by induction. We build an Eulerian
+circuit in $G$ as follows.
+\medskip
+\item{i)} Start at any vertex of $\sigma$.
+\smallskip
+\item{ii)} Follow $\sigma$.
+\smallskip
+\item{iii)} Each time we reach a vertex of a connected component $H_i$ of $G'$ for the first
+time, follow its Eulerian circuit.
+\smallskip
+\item{iv)} Once we return to the beginning of this circuit, continue along $\sigma$.
+\smallskip
+\item{v)} Repeat this process until we return to the vertex of $\sigma$ that we started at.
+\medskip
+This gives an Eulerian circuit of $G$.\slug
+
+Now we characterise Eulerian trails.
+
+\proclaim Theorem \advthm. Let $G = (V,E)$ be a graph (with multiple edges allowed). Then $G$ contains
+an Eulerian trail that isn't an Eulerian circuit
+if and only if exactly two vertices of $G$ have odd degree.
+
+\proof Suppose that $G$ contains an Eulerian trail $\sigma = (v_0,v_1,\ldots,v_n)$
+in which $v_0\ne v_n$. By a similar reasoning to the first paragraph of the previous proof,
+every vertex in the walk has even degree, except both $v_0$ and $v_n$ must have odd degree,
+since starting at $v_0$ we must leave for the first time via one edge, and at the end we enter $v_n$
+one last time without exiting.
+
+Now assume that $G$ contains exactly two odd-degree vertices. By the previous theorem, there does
+not exist an Eulerian circuit in $G$. Let $u$ and $v$ be the two odd-degree vertices.
+Add a new edge $e = uv$, even if there was already an edge between $u$ and $v$ before. We get $G'$ in
+which every vertex has even degree. So $G'$ must have an Eulerian circuit, call it $\sigma$.
+It contains the edge $e$, but we can remove $e$ to get an Eulerian trail in the original graph $G$.\slug
+
+Since the graph representing the K\"onigsberg bridge problem has four vertices, all of odd degree,
+the two theorems we just proved tell us that it contains neither an Eulerian trail nor an
+Eulerian circuit. Hence it is impossible to cross every bridge in K\"onigsberg exactly once.
+
+\advsect Planar graphs
+
+\datestamp{14}{xi} A graph $G$ is called {\it planar} if we can draw $G$ in the plane wihout edges
+crossing. If we can, such a drawing is called a {\it planar embedding} of $G$.
+The graph $K_4$ is planar; its representation in Fig.~2 does not make this evident, but
+a planar embedding is given in Fig.~8. Likewise, the hypercube $Q_3$ was drawn with edge crossings
+in Fig.~3, but it is also planar.
+\midinsert
+$$\epsfbox{planarembeddings.ps}$$
+\vskip5pt
+\caption{Planar embeddings of $K_4$ and $Q_3$, with numbered faces}
+\endinsert
+\goodbreak
+Planarity is a property that passes down to subgraphs, as the following proposition shows.
+
+\edef\propsubgraphplanar{\the\thmcount}
+\proclaim Proposition \advthm. If $G$ is planar then any subgraph $H$ of $G$ is also planar.
+
+\proof If $G$ is drawn in the plane without edges crossing, and we erase some edges and vertices
+to create a drawing of $H$, we cannot introduce any edge crossings along the way.\slug
+
+A planar embedding of a graph $G$ determines regions, or {\it faces} bounded by the edges,
+including one region ``outside'' the graph.
+We define the {\it Euler characteristic}
+of a planar embedding of $G = (V,E)$ to be the quantity
+$$\chi = |V| - |E| + f,$$
+where $f$ is the number of faces in the planar embedding.
+For the examples in Fig.~8, we have
+$$\chi = 4 - 6 + 4 = 2$$
+in the case of $K_4$, and
+$$\chi = 8 - 12 + 6 = 2$$
+in the case of $Q_3$. This might seem like a funny coincidence. The following theorem shows that
+it is not.
+
+\edef\thmeulersformula{\the\thmcount}
+\parenproclaim Theorem {\advthm} (Euler's formula). Let $G = (V,E)$ be a connected planar graph,
+let $f$ be the number of faces in a planar embedding of $G$. Then the Euler
+characteristic $\chi$ of $G$ equals $2$.
+
+The proof of this formula requires a theorem whose statement seems so ridiculously intuitive
+that it seems silly that it even needs to be stated as a theorem. Despite its innocuous
+appearance, is surprisingly difficult to prove, so much so that its original 1887 proof by
+C.~Jordan was thought by many mathematicians to be so lacking in detail as to be wrong, and
+the first widely accepted proof did not appear until 1905 (due to O.~Veblen). More recently
+some mathematicians have analysed Jordan's original proof and found that it was more or less
+correct to begin with, just not presented in a satisfactory way. In any case, here is the theorem.
+
+\parenproclaim Theorem {\advthm} (Jordan curve theorem). Every closed curve in the plane
+$\RR^2$ that does not intersect itself divides the plane into two distinct regions.\slug
+
+The two regions that the closed curve creates are simply the inside of the curve and the outside
+of the curve. Using this theorem, we can now prove Euler's formula.
+
+\medskip\noindent{\it Proof of Theorem~\thmeulersformula}.\enspace
+We proceed by induction on $m = |E|$. In the case that $m=0$, $G$ can only be $\bullet$,
+in which case we have $\chi = 1 - 0 + 1 = 2$.
+
+Now let $m\ge 0$ and assume that the theorem holds for all natural numbers at most $m$.
+Let $G = (V,E)$ be a connected planar graph with $|E| = m+1$. Draw $G$ in the plane and let
+$f$ be the number of faces in the embedding. There are two cases, according to whether
+$G$ is a tree or not.
+If $G$ is a tree, then $f=1$, since $G$ has no cycles. But we also have $|E| = |V|-1$, so
+$$\chi = |V|-|E|+f = |V|-\bigl(|V|-1\bigr) + 1 = 2.$$
+If $G$ is not a tree, then $G$ has a cycle, call it $\sigma$. Form $G'$ by removing
+exactly one edge $e$ of $\sigma$; so
+$$ |E'| = |E| - 1 = m+1-1 = m.$$
+By the detour lemma, $G'$ is still connected, and it is planar, since removing an edge doesn't
+create any crossings. By the inductive hypothesis, the Euler characteristic $\chi'$ of $G'$
+satisfies
+$$\chi' = |V| - |E'| + f' = 2,$$
+where $f'$ is the number of faces in the drawing of $G$. Returning to the drawing
+of $G$, the cycle $\sigma$ is a closed curve without intersections in the plane, so
+it divides the plane into two distinct regions (each of which may have multiple faces within them).
+In particular, the removed edge $e$ forms a border between two distinct faces, and in $G'$,
+these two faces combine into one face. Hence $f = f'+1$. Putting everything together,
+we calculate that the Euler characteristic $\chi$ of $G$ satisfies
+$$\eqalign{
+\chi &= |V| - |E| + f \cr
+&= |V| - \bigl(|E'| + 1\bigr) + (f' + 1) \cr
+&= |V| - |E'| + f' \cr
+&= 2.\noskipslug\cr
+}$$
+
+Euler's formula tells us that the number of faces in any drawing of $G$ depends entirely on
+the number of vertices and edges that $G$ has. We can use this information to give conditions
+for a graph to be non-planar.
+
+\edef\thmnonplanarconditions{\the\thmcount}
+\proclaim Theorem \advthm. Let $G = (V,E)$ be a connected planar graph with $|V|\ge 5$. Then
+$$|E| \le 3|V| - 6.$$
+Under the further assumption that $G$ contains no triangles, we have the better bound
+$$|E| \le 2|V| - 4.$$
+
+\proof The proof is by double counting. Let $R$ be the set of all regions into which $G$
+divides the plane, so that $|R| = f$, the number of faces. Consider the set
+$$S = \bigl\{ (e, r) \subseteq E\times R : \hbox{the edge}\ e\ \hbox{touches the region}\ r\bigr\}.$$
+We will count $|S|$ in two ways. First off, each edge $e\in E$ touches at most two regions,
+so $|S|\le 2|E|$. On the other hand, every region is bounded by a cycle, and cycles
+have at least $3$ edges, so each region $r\in R$ touches at least $3$ edges. In other words,
+$|S|\ge 3f$. Chaining these two inequalities, we have
+$$3f \le |S| \le 2|E|.$$
+But by Euler's formula, we have $|V| - |E| + f = 2$, so $f = 2-|V|+|E|$, and substituting this above,
+we get
+$$3\bigl( 2-|V|+|E|\bigr) \le 2|E|.$$
+Distributing and rearranging terms yields the desired inequality $|E| \le 3|V|-6$.
+
+If we have the further assumption that there are no triangles in $G$, then every region must
+touch at least $4$ edges, allowing us to conclude the stronger bound $4f \le 2|E|$. Then we proceed
+as above to get
+$$4\bigl( 2-|V|+|E|\bigr) \le 2|E|,$$
+which can be manipulated to give $|E|\le 2|V|-4$.\slug
+
+The condition that $|V|\ge 5$ is a complete non-issue, since every graph with fewer than $5$
+vertices is a subgraph of $K_4$, and we can apply Proposition~{\propsubgraphplanar},
+since we already know that $K_4$ is planar. As a corollary of the (contrapositive of the)
+previous theorem, we can give two important examples of nonplanar graphs.
+
+\proclaim Corollary \advthm.  The complete graph $K_5$ and the complete bipartite graph $K_{3,3}$
+are nonplanar.
+
+\proof If $(V,E) = K_5$ we have $|V| = 5$ so $3|V| - 6 = 9$,
+but $|E| = 5\cdot 4 / 2 = 10$. So by the contrapositive of the previous theorem $K_5$ cannot
+be planar.
+
+The graph $K_{3,3}$ does not contain any triangles, so if $K_{3,3}$ is planar, then the
+stronger bound $|V| \le 2|V|-4$ must hold. But in $(V,E) = K_{3,3}$, we have $|V| = 6$,
+so $2|V|-4 = 8$, and we have $|E| = 9$. So $K_{3,3}$ is also nonplanar.\slug
+
+\midinsert
+$$\epsfbox{k5k33.ps}$$
+\vskip5pt
+\caption{The nonplanar graphs $K_5$ and $K_{3,3}$.}
+\endinsert
+\goodbreak
+
+This corollary, combined with the contrapositive of Proposition~{\propsubgraphplanar},
+shows that $K_n$ is nonplanar for all $n\ge 5$, and $K_{m,n}$ is non-planar whenever
+$m$ and $n$ are both at least $3$. More generally, any graph that contains $K_5$ or $K_{3,3}$
+as a subgraph is nonplanar.
+
+However, subgraphs are not exactly the right notion to be considering
+when talking about planarity. To see why, consider the graph in Fig.~10.
+\midinsert
+$$\epsfbox{k5dot.ps}$$
+\vskip5pt
+\caption{The graph $K_5$, with an extra vertex subdividing an edge.}
+\endinsert
+\goodbreak
+It is nonplanar, since if it had a planar embedding, then by contracting one of the
+edges incident on the extra vertex and bending other edges accordingly,
+we would obtain a planar embedding of $K_5$.
+On the other hand, $K_5$ is not a subgraph of this graph.
+We thus introduce the following notion.
+We say that a graph $H$ is a {\it graph minor} of a graph $G$ if $H$ can be obtained
+from $G$ by repeatedly either
+\medskip
+\item{i)} deleting an edge;
+\smallskip
+\item{ii)} deleting a vertex; or
+\smallskip
+\item{iii)} contracting an edge $uv$ by removing it and merging $u$ and $v$ into a single vertex
+(and also combining any resulting multiple edges into a single edge).
+\medskip
+Any number of the first two operations simply creates a subgraph. It is the contraction operation
+that produces interesting examples of graph minors. It is easy to see pictorially that
+contracting edges preserves planarity (we will not prove this rigorously), so {\sl any graph
+minor of a planar graph is planar}. We can use this to show, for instance, that the Petersen graph,
+depicted in Fig.~11, is nonplanar, since by contracting each of the five edges connecting
+the inner star pentagram to the outer pentagon, we obtain $K_5$. (Note that
+Theorem~{\thmnonplanarconditions} does not directly apply to the Petersen graph.)
+\midinsert
+$$\epsfbox{petersen.ps}$$
+\vskip5pt
+\caption{The Petersen graph.}
+\endinsert
+\goodbreak
 \bye
+
+\noindent\datestamp{22}{} \hskip\parindent
+Not only are all graphs that have $K_5$ or $K_{3,3}$ as a minor nonplanar, it turns out
+that these are the only nonplanar graphs!
+
+\parenproclaim Theorem {\advthm} (Wagner's theorem). A graph $G$ is nonplanar if and only
+if either $K_5$ or $K_{3,3}$ is a graph minor of $G$.\slug
+
+This theorem was proven in 1937 by K.~Wagner, but it is closely related to a theorem that was
+proved earlier (in 1930) by K.~Kuratowski.
+It deals with the notion of a {\it subdivision} of a graph $G$:
+this is any graph obtained by repeatedly placing vertices on edges of $G$, subdividing an edge
+into two edges each time. (Fig.~10 was an example of a subdivision of $K_5$.) We now state Kuratowski's
+theorem.
+
+\parenproclaim Theorem {\advthm} (Kuratowski's theorem). A graph $G$ is nonplanar if and only
+if it contains as a subgraph a subdivision of either $K_5$ or $K_{3,3}$.\slug
+
+It is easy to
+see that if $H$ is a subdivision of $G$, then $G$ is a minor of $H$, since we can reobtain
+$G$ by contracting all the edges created by the subdivision operations. So the two theorems
+above are certainly very closely related, though it is not immediately obvious if one should imply
+the other. The truth is that they are equivalent, because it can be shown that any graph with
+either $K_5$ or $K_{3,3}$ as a minor also has a subgraph that is a subdivision of one of them.
+
+\vfill\eject
+\begingroup\headline{\hfil}\footline{\hfil}
+\vglue160pt
+\centerline{\titlefont IV. COMBINATORICS}
+\vskip220pt
+\bigskip
+\begingroup\obeylines\eightssi
+\hfill Quare merito suo utilissima censenda est Ars, {\eightss Combinatoria} dicta,
+\hfill quae huic mentis nostrae defectui medetur,
+\hfill docetque sic enumerare modos omnes possibiles,
+\hfill secundum quos res plures permisceri, transponi vel conjungi invicem possunt,
+\hfill ut certi, simus, nos nullum eorum praetermisisse,
+\hfill qui instituto nostro conducere valent.
+\eightss
+\smallskip
+\hfill --- JAKOB BERNOULLI, {\eightssi Ars Conjectandi} (1713)
+\endgroup%\obeylines
+\bigskip\goodbreak
+\vfill\eject
+
+\advsect Basic counting techniques
+
+Combinatorics is a branch of mathematics that, broadly speaking, deals with counting (or estimating)
+the cardinalities of finite sets. Let us now recollect the counting techniques that we know.
+If two sets $A$ and $B$ are disjoint, i.e., $A\cap B = \emptyset$,
+then the number of elements in the union $A\cup B$ is simply $|A| + |B|$.
+The cardinality of the power set $A\times B$ is $|A|\cdot |B|$, since for each of the $|A|$ elements
+$a\in A$, there are $|B|$ choices for $b\in B$ to pair with it.
+Lastly, if $U$ is a universe and $A\subseteq U$, then $\bar A$ is the set of all elements
+of the universe that don't belong to $A$, so $|\bar A| = |U|-|A|$.
+
+\medskip\boldlabel Counting strings.
+Back in the proof of Proposition~{\propcardpowerset}, we counted the number of binary strings of
+length $n$ to be $2^n$. More generally, if there are $s$ different symbols and $t$ different
+slots in which to put them, then the number of strings is $s^t$. This can be phrased
+in terms of the number of functions from a set $T$ of size $t$ to a set $S$ of size $s$.
+Indeed, people often use the notation $S^T$ to denote the set of all functions from $T$ to $S$,
+and the notation should remind you that $|S^T| = |S|^{|T|}$.
+
+For example, using the symbol set $S = \{a,b,c,d\}$, the number of strings of length $5$
+is $|S|^5 = 4^5$. What about the number of strings of length {\it at most $5$}, including
+one empty string of length $0$? This is the number of strings of length $0$, plus the number
+of strings of length $1$, and so on up to and including length $5$; that is, this equals
+$$4^0 + 4^1 + 4^2 + 4^3 + 4^4 + 4^5.$$
+
+\bye

### 2024-11-15 teaching/math240f24/draft.tex
Added link

@@
 \goodbreak
-\bye
 
@@
 
+To count strings of length $5$ and containing at least one $a$, we let the universe
+be all strings of length $5$ and let $A$ be the set of all strings without
+any $a$, i.e., using only the letters $b$, $c$, and $d$. Then the number of strings
+containing at least one $a$ is
+$$|\bar A| = |U| - |A| = 4^5 - 3^5.$$
+
+Next we count the number of strings of length $5$, but we require that adjacent symbols
+be different. This one is best calculated by counting the number of choices we have at
+each slot. There are $4$ choices for the first symbol, but after
+that we only have choices for each letter, since they cannot match the previous one.
+Hence the count is $4\cdot 3^4$.
+
+Lastly, to count the number of strings that must have at least one of the patterns
+$aa$, $bb$, $cc$, or $dd$, we simply note that this is the complement of what
+we counted in the previous paragraph, so the answer is $4^5 - 4\cdot 3^4$.
+
+\medskip\boldlabel The principle of inclusion and exclusion. As our next toy problem,
+suppose we want to count the number of strings of length $6$, using the symbols
+$a$, $b$, $c$, and $d$, that start with $a$ or end with $aa$ (or both). The number
+of strings that start with $a$ is $4^5$, since we have no choice for the first symbol,
+but $4$ choices for each of the remaining five symbols; likewise, the number
+of strings that end with $aa$ is $4^4$. Adding these two up, we get $4^5+4^4$.
+There's a problem though. We have accidentally double-counted the strings that
+both start with $a$ {\it and} end with $aa$. How many of these are there? Well, the first, fifth, and
+sixth slots all have to be $a$, but there are four choices for each of the remaining
+three slots, for a total of $4^3$. Subtracting this from our subtotal yields
+the correct answer: $4^5+4^4-4^3$.
+\endgroup%\headline{\hfil}\footline{\hfil}
+
+This technique works with any two sets $A$ and $B$,
+i.e., $|A\cup B| = |A| + |B| - |A\cap B|$. For three sets $A$, $B$, and $C$,
+we have
+$$|A\cup B\cup C| = |A| + |B| + |C| - |A\cap B| - |A\cap C| - |B\cap C| + |A\cap B\cap C|$$
+(draw a Venn diagram and convince yourself that this is correct).
+For a general number of sets, we have the following theorem.
+
+\proclaim Theorem {\advthm} (Principle of inclusion and exclusion).
+
 \bye

### 2024-11-20 teaching/math240f24/draft.tex
Lecture notes 19 nov

@@
 
-\vskip48pt
+\vskip36pt
 
@@
 \hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 16.\kern .5em Planar graphs}{83}}
+\medskip
+\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
+{{\tenssbx IV. Combinatorics \kern 3pt}}{88}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 17.\kern .5em Basic counting techniques}{89}}
 
@@
 between $p$ and $q$.
+
 But it should make sense if you think
-of $p$ as a precondition to a promise $q$, and then considering whether the promise is broken. As an example,
+of $p$ as a precondition to a promise $q$, and then consider
+whether the promise is broken. As an example,
 suppose your friend says, If it snows tomorrow I'll work in Trottier with you. If it doesn't snow
@@
 
+\edef\propoddevensubsets{\the\thmcount}
 \proclaim Proposition \advthm. Let $X$ be a finite nonempty set.
@@
 
+\edef\thmpnt{\the\thmcount}
 \parenproclaim Theorem {\advthm} (Prime number theorem). For $x\in \RR$, let
@@
 
-\noindent\datestamp{22}{} \hskip\parindent
+\noindent\datestamp{19}{xi} \hskip\parindent
 Not only are all graphs that have $K_5$ or $K_{3,3}$ as a minor nonplanar, it turns out
@@
 
-\advsect Basic counting techniques
+\advsect Counting techniques
 
@@
 Indeed, people often use the notation $S^T$ to denote the set of all functions from $T$ to $S$,
-and the notation should remind you that $|S^T| = |S|^{|T|}$.
+and the notation should remind you that $|S^T| = |S|^{|T|}$. Identifying the number
+$2$ with the set $\{0,1\}$, the power set $2^X$ is in bijection with the set $\{0,1\}^X$, so
+this notation for the set of all functions is consistent with our earlier notation for
+power set.
 
@@
 the correct answer: $4^5+4^4-4^3$.
-\endgroup%\headline{\hfil}\footline{\hfil}
 
@@
 For a general number of sets, we have the following theorem.
-
-\proclaim Theorem {\advthm} (Principle of inclusion and exclusion).
-
+\endgroup%\headline{\hfil}\footline{\hfil}
+
+\edef\thminclusionexclusion{\the\thmcount}
+\parenproclaim Theorem {\advthm} (Principle of inclusion and exclusion). Let $n\ge 2$
+be an integer and let $A_1,\ldots, A_n$ be finite sets. Then
+$$\biggl| \bigcup_{i=1}^n A_i\biggr|
+=\sum_{k=1}^n (-1)^{k+1}\biggl(\sum_{1\le i_1<\cdots<i_k\le n}|A_{i_1}\cap\cdots\cap A_{i_k}|\biggr).$$
+
+This sum of sums might look intimidating, but breaking down what it says bit by bit,
+we see that it is simply a sum over all possible nonempty intersections of the $A_i$, with a sign
+according to how many of the $A_i$ we are intersecting. We will prove this theorem soon,
+after we state and prove the binomial theorem. We first define, for nonnegative integers
+$n$ and $k$, the {\it binomial coefficient} ${n\choose k}$ to be the number of ways of choosing
+$k$ distinct elements out of a pool of $n$ distinct elements. (The quantity ${n\choose k}$
+is read ``$n$ choose $k$.'') We have the formula
+$${n\choose k} = {n!\over k!(n-k)!} = {n(n-1)\cdots (n-k+1)\over k(k-1)\cdots 1}.$$
+This is because we can pick $k$ elements out of an $n$-element set by permuting all the $n$ elements
+(there are $n!$ ways to do this), and then simply looking at the first $k$ elements. But the order
+of the $k$ elements we choose doesn't matter, so we arrive at the same elements no matter which
+of the $k!$ possible orders they might appear in. Same goes for the $(n-k)!$ orders the elements
+that we {\it didn't} pick can appear in.
+
+The name ``binomial coefficient'' is justified by the following theorem, first stated and
+proved by Al-Karaji in around the tenth century.
+
+\parenproclaim Theorem {\advthm} (Binomial theorem). For all $x,y\in \RR$ and positive integers $n$,
+$$(x+y)^n = \sum_{k=0}^n {n\choose k} x^k y^{n-k}.$$
+
+\proof Expanding the left-hand side
+$$(x+y)^n = (x+y)(x+y)\cdots (x+y),$$
+we see that there will be $2^n$ different terms after repeatedly using the distributive law.
+Each term will be a product of $x$ to some power and $y$ to some power, where the powers add up to $n$.
+In other words, each term looks like $x^ky^{n-k}$ for some $0\le k\le n$. The number of times
+that the term $x^k y^{n-k}$ appears in the expansion
+is the number of ways to choose $k$ items out of a pool of $n$, namely,
+${n\choose k}$. This gives exactly the right-hand side.\slug
+
+We are now able to prove the principle of inclusion and exclusion.
+
+\medskip\noindent{\it Proof of Theorem~\thminclusionexclusion.}\enspace
+Let $n\ge 2$ and let $A_1,\ldots,A_n$ be finite sets.
+Recall that the identity we want to prove is
+$$\biggl| \bigcup_{i=1}^n A_i\biggr|
+=\sum_{k=1}^n (-1)^{k+1}\biggl(\sum_{1\le i_1<\cdots<i_k\le n}|A_{i_1}\cap\cdots\cap A_{i_k}|\biggr).$$
+Let $x$ be an element that belongs to the union $\bigcup_{i=1}^n A_i$. Clearly, this contributes
+$1$ to the left-hand side, so we must show that it contributes exactly $1$ to the right-hand side
+as well. We know that $x$ is a member of at least one of the $A_i$;
+let $s\ge 1$ be the number of sets $A_i$ that contain $x$.
+
+Each time $x$ is a member of
+some $A_i$, it contributes $+1$ to the right-hand side. This happens ${s\choose 1}$ times.
+Each time it is a member of $A_i$ and
+$A_j$ for $i<j$, $x$ contributes $-1$ to the-right hand side; there are ${s\choose 2}$ of these terms.
+Generally, if $x$ is a member
+of $A_1\cap A_2\cap\cdots\cap A_r$, it contributes $+1$ if $r$ is odd and $-1$ if $r$ is even,
+and this term is repeated ${s\choose r}$ times. Hence $x$ contributes
+$${s\choose 1} - {s\choose 2} + {s\choose 3} - \cdots + (-1)^{s+1} {s\choose s}$$
+to the right-hand side.
+
+But, by the binomial theorem,
+$$0 = (1-1)^s = {s\choose 0} - {s\choose 1} + {s\choose 2} -\cdots + (-1)^s{s\choose s}.$$
+We know that ${s\choose 0} = s!/(0! s!) = 1$, so by rearranging we find that
+$${s\choose 1} - {s\choose 2} + {s\choose 3} - \cdots + (-1)^{s+1} {s\choose s} = {s\choose 0} = 1,$$
+which means that the contribution of $x$ to the right-hand side is $1$.
+We are done since $x$ is arbitrary.\slug
+
+As an application of this principle, let $X = \{1,2,\ldots, 100\}$ and suppose we want
+to count the number of $n\in X$
+with $\gcd(n,30) = 1$. To accomplish this, it turns out to be easier to count the number
+of $n\in X$ that have $\gcd(n,30)\ge 2$ (and then we must subtract this number from $|X| = 100$).
+For any positive integer $r$, let
+$$A_r = \{n\in X : r\divides n\}.$$
+Since $30 = 2\cdot 3\cdot 5$, an integer $n\in X$ has $\gcd(n,30)\ge 2$ if and only if
+$n\in A_2$, $n\in A_3$, or $n\in A_5$.
+By the principle of inclusion and exclusion,
+$$|A_2\cup A_3\cup A_5| = |A_2| + |A_3| + |A_5|
+- |A_2\cap A_3| - |A_2\cap A_5| - |A_3\cap A_5| + |A_2\cap A_3\cap A_5|.$$
+The number of even integers in $X$ is $100/2 = 50$; similarly, the number
+of integers divisible by $3$ in $X$ is $\lfloor 100/3\rfloor = 33$. In general,
+$|A_r| = \lfloor 100/r\rfloor$. Furthermore, since $\gcd(2,3) = 1$,
+the intersection $A_2\cap A_3$ is simply $A_6$, and by analogous reasoning, we see that
+$$\eqalign{
+|A_2\cup A&_3\cup A_5| \cr
+&= |A_2| + |A_3| + |A_5| - |A_6| - |A_{10}| - |A_{15}| + |A_{30}| \cr
+&= \biggl\lfloor {100\over 2}\biggr\rfloor
++\biggl\lfloor {100\over 3}\biggr\rfloor
++\biggl\lfloor {100\over 5}\biggr\rfloor
+-\biggl\lfloor {100\over 6}\biggr\rfloor
+-\biggl\lfloor {100\over 10}\biggr\rfloor
+-\biggl\lfloor {100\over 15}\biggr\rfloor
++\biggl\lfloor {100\over 30}\biggr\rfloor\cr
+&= 50 + 33 + 20 - 16 - 10 - 6 + 3 \cr
+&= 74.
+}$$
+We conclude that the number of $n\in X$ such that $\gcd(n,30) = 1$ is
+$100-74 = 26$.
+
+\advsect Permutations and combinations
+
+Suppose we have a pool of $100$ hockey players and we want to dress $20$
+of them for a game, giving them the numbers $1$ through $20$. How many ways are there to do this?
+Well, there are $100$ choices for the player with jersey number $1$, then after
+picking that player, there are $99$ choices for the player with jersey number $2$, and so on,
+so the total number is
+$$100\cdot 99\cdot 98 \cdots 82\cdot 81 = {100!\over 80!}.$$
 \bye
+
+\noindent\datestamp{23}{} \hskip\parindent
+In general, we define a {\it $k$-permutation} of a set $X$ to be an ordered subset
+of $X$ (i.e., the order of the selection matters) of size $k$. If $|X| = n$,
+the number of $k$-permutations of $X$ is denoted $P(n,k)$ and equals
+$$ P(n,k) = n\cdot (n-1)\cdots (n-k+1) = {n!\over (n-k)!}.$$
+This number is also sometimes called the {\it falling factorial} and denoted $n^{\underline k}$.
+It is the same as the number of functions from $\{1,2\ldots, k\}$ to $X$ that are injective.
+(Pause a moment to think about it if you need to.) Recall that if $|S| = s$ and $|T| = t$,
+the total cardinality of $S^T$ is $s^t$. Well, the number of
+{\it injective} functions from $T$ to $S$ is $P(s,t) = s^{\underline{t}}$.
+
+An $n$-permutation of a set of size $n$ is also just called a {\it permutation} of
+$n$. There are
+$$P(n,n) = {n!\over (n-n)!} = n!$$
+of these. (Note that $0! = 1$, by definition.)
+
+To illustrate various counting techniques associated with permutations, consider the following
+situation. We have ten books: three comic books, three books of poetry, and four novels.
+How many ways are there to order all of them on a shelf? The answer is the number of permutations
+on $10$ objects, namely, $10! = 3628800$. How about just putting five of them on the shelf?
+Then the answer
+is $P(10,5) = 10!/5! = 30240$. Lastly, what if we want to put all of them on the shelf,
+but group all the comic books together, all the poetry books together, and all the novels together.
+There are $3!$ ways to order the comic books amongst themselves, $3!$ for the poetry books
+as well, and then $4!$ for the novels. Then we must arrange the ordered bundles of books
+on the shelf; there are three bundles so there are $3!$ ways to do this. Thus the total
+number of ways is $3!3!4!3! = 5184$.
+
+Returning to the hockey-team example, suppose we still want to pick a team of $20$ out of
+$100$ possible players, but now we don't number the players; that is, we only care about
+the sets of the players themselves, not the ordering amongst them. The number $P(100,20)$
+is $20!$ times too big, since there are $20!$ different orderings of the team we choose,
+and each of them is only counted once in our new scenario. So the answer is
+$${P(100,20)\over 20!} = {100!\over 20!(100-20)!} = {100\choose 20}.$$
+We already had an intimate encounter with
+the quantity ${n\choose k}$ in the previous section. Defining a
+{\it $k$-combination} of a set $X$ to be the number of ways of choosing $k$ elements of of
+$X$, the number of $k$-combinations of a set of size $n$ is ${n\choose k}$.
+
+\medskip\boldlabel Counting grid paths.
+An application of $k$-combinations is counting the number of paths in a grid.
+If you have ever had the pleasure or misfortune of visiting
+Edmonton, Alberta, you will know that most of the city's roads
+are organised in a numbered grid pattern. Avenues run east to west and streets run north to south.
+Suppose we have just enjoyed a cortado in the trendy caf\'e
+at the corner of 82nd Avenue and 104th Street,
+and now we want to get to the sandwich shop at the corner of 85th Avenue and 109th Street.
+To do so we will have to walk three blocks north and five blocks west. How many ways are there
+to do this?
+\midinsert
+$$\epsfbox{edmonton.ps}$$
+\vskip5pt
+\caption{One possible path from the caf\'e to the sandwich shop.}
+\endinsert
+\goodbreak
+Well, the journey is eight blocks in length, so each possible path can be expressed as
+a string of eight symbols, call them {\mc N} for north and {\mc W} for west. We also know there
+must be exactly three {\mc N}s and five {\mc W}s; for example, one possible path, pictured in
+Fig.~12, is {\mc NWWWNWWN}. There are ${8\choose 3}$ different ways to choose where the {\mc N}s
+go in the string, after which the positions of the {\mc W}s is completely determined. Alternatively,
+we can pick where the {\mc W}s go, in ${8\choose 5}$ ways. Thus the number of possible paths
+from the caf\'e to the sandwich shop is
+$${8\choose 3} = {8!\over 3! 5!} = {8\choose 5} = 56.$$
+Here we see an important principle in action. {\sl The number of $k$-combinations of a set
+of size $n$ is the same as the number of $(n-k)$-combinations of the same set.}
+This is because choosing $k$ items is
+the same as choosing which $n-k$ items {\it not} to take.
+In general, the number of ways of walking $m$ blocks in one direction and $n$ blocks in a perpendicular
+direction is
+$${m+n\choose m} = {m+n\choose n}.$$
+
+\medskip\boldlabel Stars and bars.
+Suppose there is a shop selling marbles that come in six colours. How many ways are there to select
+a bag of four marbles? Imagine that the marbles are laid out in six different bins all in a row,
+one for each colour. We stand at the first bin and may take some number of them. Mark down a
+star for each one that we take. Then we step forward to the next bin; record this action by
+drawing a bar. If we take some marbles here, mark down a star for each marble taken. Then we step forward
+again, so mark down a bar. At the end, we must have taken exactly four marbles and switched bins
+five times, so there must be exactly four stars and five bars, e.g., the sequence
+$$\mid\quad\mid\star\mid\star\mid\quad\mid\star\star$$
+corresponds to taking one marble of the third colour, one of the fourth, and two of the sixth.
+There are nine slots, four of which have to be stars, so
+the number of ways to pick four marbles out of six bins is ${9\choose 4} = 126$. In general
+the number of ways to pick $m$ marbles out of $b$ bins is ${m+b-1\choose m}$. This method of
+counting is called the {\it stars-and-bars} technique.
+
+\medskip\boldlabel A night at the casino.
+Suppose we are playing a high-stakes card game at the local gambling
+house and it is somehow pertinent for us to calculate how many five-card hands there
+are with at least two hearts. A na\"\i ve attempt would run as follows. We pick two cards
+to be hearts, in ${13\choose 2}$ ways, then there are $50$ cards left to choose from,
+and we can pick any three of those, in ${50\choose 3}$ ways. This seems logical, except
+that we have counted some hands twice! For instance, suppose that in the first step we pick
+our two hearts to be $\tt2\heartsuit$ and $\tt6\heartsuit$, then for our remaining three
+cards we pick $\tt8\heartsuit$, $\tt3\diamondsuit$, and $\tt Q\spadesuit$. We can get the exact
+same hand by first picking our two hearts to be $\tt6\heartsuit$ and $\tt8\heartsuit$,
+and then choosing $\tt2\heartsuit$, $\tt3\diamondsuit$,
+and $\tt Q\spadesuit$ for the remaining three cards.
+
+One correct way of getting the correct count is to count the number of hands with
+exactly two hearts, exactly three hearts, exactly four hearts, and exactly five hearts. Since
+these possibilities all define disjoint sets, we can simply add them up at the end.
+The number of ways of drawing exactly two hearts is
+$${13\choose 2}{52-13\choose 3} = {13\choose 2}{39\choose 3} = 712842,$$
+since there are $39$ non-heart cards and we must select three of them.
+Similarly, the number of ways of drawing exactly three hearts is
+$${13\choose 3}{52-13\choose 2} = {13\choose 3}{39\choose 2} = 211926;$$
+we also calculate
+$${13\choose 4}{39\choose 1} = 27885\qquad\hbox{and}\qquad
+{13\choose 5}{30\choose 0} = 1287.$$
+Hence the number of five-card hands with at least two hearts is
+$$712842 + 211926 + 27885 + 1287 = 953940.$$
+This number has little practical value unless we know the number of five-card hands
+in total; this is ${52\choose 5} = 2598960$. We can then calculate the probability
+of a five-card hand containing at least two hearts to be $953940/2598960 \approx 36.7\%$.
+
+Deriving the total number of five-card hands suggests a different way to get
+the same count.
+We could take this total and subtract the number
+of hands with no hearts, as well as the number of hands with exactly one heart. This gives
+$${52\choose 5} - {13\choose 0}{39\choose 5} - {13\choose 1}{39\choose 4}
+= 2598960 - 575757 - 1069263 = 953940,$$
+the same figure we obtained in the previous paragraph.
+
+\medskip\boldlabel\datestamp{24}{} Identities involving binomial coefficients.
+We already observed earlier that the number of ways of picking $k$ objects out of a
+set of size $n$ is the same as the number of picking $n-k$ objects out of the same set.
+This can be seen algebraically by expanding
+$${n\choose k} = {n!\over k!(n-k)!} = {n!\over (n-k)!\bigl(n-(n-k)\bigr)!} = {n\choose n-k}.$$
+Here is another identity involving binomial coefficients.
+
+\proclaim Proposition \advthm. Let $n\ge 0$ be an integer. Then
+$$\sum_{k=0}^n {n\choose k}^2 = {2n\choose n}.$$
+
+\proof Let $X$ be set of size $2n$. The right-hand side counts the number of subsets of
+$X$ of size $n$. We count this in a different way. Let $A$ and $B$ be such that $|A| =|B| = n$,
+and $A\cup B = X$ (so we must have $A\cap B = \emptyset$). Then choosing a subset
+of $X$ is the same as choosing $k$ elements of $A$, where $0\le k\le n$,
+and then choosing the remaining $n-k$ elements from $B$. In other words,
+$${2n\choose n} = \sum_{k=0}^n {n\choose k}{n\choose n-k}
+= {n\choose k}^2.\noskipslug$$
+
+More identities can be obtained by laying out binomial coefficients in the triangle
+$$\matrix{
+&&&{0\choose 0}&&&\cr
+&&{1\choose 0}&&{1\choose 1}&&\cr
+&{2\choose 0} && {2\choose 1} && {2\choose 2}&\cr
+{3\choose 0} && {3\choose 1} && {3\choose 2} && {3\choose 3} \cr
+\vdots && \vdots && \vdots && \vdots && \cr
+}$$
+Adding a few more rows and writing out the actual values of the binomial coefficients
+yields Fig.~13. This triangle is named after B.~Pascal, who wrote a treatise regarding some of
+its properties that was published posthumously in 1665. (The triangle is known by different
+names in some different countries, as it was studied by Al-Karaji in the tenth or
+eleventh century, by O.~Khayy\'am about a hundred years later, by Yang Hui in
+the thirteenth century, and by Tartaglia in 1556, among many others.)
+\midinsert
+$$\matrix{
+&&&&&&&& 1 &&\cr
+&&&&&&& 1 && 1 &&\cr
+&&&&&& 1 && 2 && 1 &\cr
+&&&&& 1 && 3 && 3  && 1 \cr
+&&&& 1 && 4 && 6 && 4 && 1 \cr
+&&& 1 && 5 && 10 && 10 && 5 && 1 \cr
+&& 1 && 6 && 15 && 20 && 15 && 6 && 1 \cr
+& 1 && 7 && 21 && 35 && 35 && 21 && 7 && 1 \cr
+1 && 8 && 28 && 56 && 70 && 56 && 28 && 8 && 1 \cr
+\vdots &&\vdots && \vdots && \vdots && \vdots && \vdots && \vdots && \vdots &&\vdots \cr
+}$$
+\vskip5pt
+\caption{The first nine rows of Pascal's triangle}
+\endinsert
+\goodbreak
+This triangle can easily be drawn by hand by writing $1$s along its boundary, and then filling it
+it using the observation that every number is the sum of the two directly above it in the previous
+row. Of course, this rule has to be proved, so we shall do so right now.
+
+\parenproclaim Theorem {\advthm} (Pascal's identity). Let $n\ge 0$ and $k\ge 1$ be integers.
+Then
+$${n\choose k} = {n-1\choose k} + {n-1\choose k-1}.$$
+
+\proof Set $X = \{1,2,\ldots,n\}$. We count the number of sets $Y\subseteq X$ there are
+of size $k$.
+Either $1\notin Y$ or $1\in Y$. If $1\notin Y$, then there are ${n-1\choose k}$ possibilities
+for what $Y$ can be. If $1\in Y$, then there are still $k-1$ other elements of $Y$, and $n-1$
+elements to choose from, so there are ${n-1\choose k-1}$ possiblities in this case. Hence
+the right-hand side counts the number of subsets of $X$ with cardinality $k$. But this is
+exactly what the left-hand side counts as well.\slug
+
+Another thing we notice about Pascal's triangle is that the first row sums to $1$, the second
+row sums to $2$, the third to $4$, the fourth to $8$, and so on. We might therefore
+conjecture that the $n$th row of Pascal's triangle should always sum up to $2^n$. Indeed, we
+have the following easy corollary of the binomial theorem.
+
+\proclaim Proposition \advthm. For all $n\in \NN$,
+$$\sum_{k=0}^n {n\choose k} = 2^n.$$
+
+\proof By the binomial theorem with $x=y=1$, we have
+$$2^n = (1+1)^n = \sum_{k=0}^n {n\choose k} 1^k 1^{n-k} = \sum_{k=0}^n {n\choose k}.$$
+The proposition can also be proven by noting that the left-hand side counts the number
+of subsets of $\{1,2,\ldots,n\}$ of size $0$, plus the number of subsets of size $1$,
+and so on up to all subsets of size $n$. But adding these all up simply gives the total number
+of subsets of $\{1,2,\ldots, n\}$, which we know to be $2^n$.\slug
+
+The binomial theorem with a different choice of $x$ and $y$ allows us to reprove a statement
+we showed a long time ago, that for any finite nonempty set $X$, the number of subsets of $X$ with
+odd cardinality is the same as the number of subsets of $X$ with even cardinality.
+
+\medskip\noindent{\it Another proof of Proposition~{\propoddevensubsets}}.\enspace
+We re-establish the notation of the proposition statement. Let $X$ be a finite nonempty set,
+let $E$ be the set of subsets of $X$ with even cardinality, and let $O$ be the set of
+subsets with odd cardinality. By the binomial theorem with $x=-1$ and $y = 1$, we have
+$$0 = \bigl(-1+1\bigr)^n = \sum_{k=0}^n {n\choose k} (-1)^k 1^{n-k}
+= {n\choose 0} - {n\choose 1} + {n\choose 2} -\cdots + (-1)^n{n\choose n}.$$
+Hence
+$${n\choose 0} + {n\choose 2} +\cdots = {n\choose 1} + {n\choose 3} + \cdots,$$
+but the left-hand side of this is just $|E|$, and the right-hand side is $|O|$.\slug
+
+\medskip\boldlabel The freshman's dream. Back when you were in high school, you might
+have made the following computational error:
+$$(x+y)^2 = x^2 + y^2$$
+It turns out you were not wrong, provided $x$ and $y$ were integers
+and you were working modulo $2$. In general, one has the following consequence of the binomial theorem.
+
+\parenproclaim Theorem {\advthm} (Freshman's dream). Let $p$ be a prime number
+and let $x,y\in \ZZ$. Then
+$$(x+y)^p \equiv x^p + y^p \pmod p.$$
+
+\proof By the binomial theorem,
+$$(x+y)^p = \sum_{k=0}^p {p\choose k} x^k y^{p-k}
+= x^p + y^p + \sum_{k=1}^{p-1} {p\choose k}x^k y^{p-k}.$$
+Taking this equation modulo $p$,
+we are done if we can show that for all $1\le k\le p-1$, ${p\choose k} \equiv 0\pmod p$.
+
+Let $1\le k\le p-1$ and expand
+$${p\choose k} = {p!\over k!(p-k)!}$$
+Multiplying both sides by $k!(p-k)!$, we get
+$$k!(p-k)!{p\choose k} = p! = p(p-1)!.$$
+This is an integer and the
+right-hand side is a multiple of $p$, so the left-hand side is as well. Then since $p$ divides
+the product $k!(p-k)!{p\choose k}$, either $p$ divides $k!(p-k)!$ or it divides $p\choose k$.
+But
+$$k!(p-k)! = k(k-1)\cdots 2\cdot 1\cdot (p-k)(p-k-1)\cdots 2\cdot 1,$$
+and all of the factors on the right-hand side are positive integers less than $p$. So $p$ cannot divide
+any of them, and hence $p$ does not divide $k!(p-k)!$. We conclude that $p\divides {p\choose k}$.\slug
+
+\medskip\boldlabel A day spent selling milk.
+The binomial theorem can be used to find the value of a specific coefficient in a polynomial,
+without necessarily expanding the whole thing out. To illustrate this, consider the following
+scenario. Imagine that a farmer runs a roadside
+milk stand. Each time a customer visits they either
+\medskip
+\item{i)} pay her $\$6$ and buy one bottle of white, chocolate, or strawberry milk; or
+\smallskip
+\item{ii)} sell her back a glass bottle for $\$1$.
+\medskip
+So after one customer visits, there are three ways in which the farmer could have
+made $\$6$, and one way in which she could have made $-\$1$. Let us arrange these
+possibilities as a function of $x$; we write $3x^6 + x^{-1}$.
+
+What benefit is there
+of doing this? Well, suppose two customers come by. There are three ways for the
+first customer to give her $\$6$, and three ways for the second customer to give
+her $\$6$, so the number of ways for her to make $\$12$ after two customers visit is
+$9$ (where the order of the flavours she sells matters---selling chocolate then strawberry
+is different from selling strawberry then chocolate). It is also possible for her to
+net $\$5$ from two customers arriving. She could buy a bottle from the first customer
+and then sells the second a bottle of milk, or vice versa. There are three ways for
+each possibility, so six ways in total that she can make $\$5$. Lastly, there is only
+one way for her to have made $-\$2$. This is if she buys back a bottle from both of them.
+Thus the function of $x$ that describes all the fiscal possibilities of the farmer is
+$$9x^{12} + 6x^{5} + x^{-2} = (3x^6 + x^{-1})^2.$$
+By a similar reasoning, the number of ways for the farmer to make $x$ dollars after
+$n$ customers come by is $(3x^6 + x^{-1})^n$.
+When $n$ gets large we don't necessarily want to expand out this entire function
+(it has $2^n$ terms), but the binomial theorem allows this representation to be useful
+nonetheless.
+
+Suppose the farmer expects $15$ customers to come by tomorrow.
+Let's calculate the number of ways that the farmer can make $\$48$ from this.
+By the binomial theorem we have
+$$\eqalign{
+(3x^6 + x^{-1})^{15} &= \sum_{k=0}^{15} {15\choose k} (3x^6)^k (x^{-1})^{15-k} \cr
+&= \sum_{k=0}^{15} {15\choose k} 3^k x^{6k-15+k} \cr
+&= \sum_{k=0}^{15} {15\choose k} 3^k x^{7k-15}, \cr
+}$$
+and the number of ways to make $\$48$ from $15$ customers
+is the coefficient of $x^{48}$ on the right-hand side. So we want to know
+if $7k-15$ ever equals $48$ for any integers $0\le k\le 15$;
+setting $k=9$ does the trick. Hence on the right-hand side, when $k=9$,
+we have the term ${15\choose 9} 3^9 x^{48}$, meaning that the number of ways of making $\$48$ after
+$15$ customers come by is ${15\choose 9} 3^9 = 98513415$.
+
+A sum of powers of $x$ whose coefficients encode some combinatorial meaning (the number of ways
+to do something) is called a {\it generating function}. Using generating functions
+is one of the most beautiful and powerful techniques in combinatorics; the
+topic is usually covered in a second course on discrete mathematics (e.g., {\mc MATH}~340).
+
+\bigskip
+{\obeylines\eightssi
+\hfill A generating function is a clothesline
+\hfill on which we hang up a sequence for display.
+\eightss
+\smallskip
+\hfill --- HERBERT S. WILF, {\eightssi generatingfunctionology} (1989)}
+\bigskip\goodbreak
+
+\advsect Applications of the pigeonhole principle
+
+Back in Section~5, we proved the pigeonhole principle: {\sl If $n$ pigeons nest in $n-1$ holes,
+then at least one hole contains more than one pigeon.} We now present some examples illustrating
+its use.
+
+\medskip\boldlabel Sheep.
+Here is the first example. Suppose five sheep are placed in a square enclosure with a side
+length of $20$ metres. One can prove that any point in time, there are always two sheep within
+a distance of $\sqrt{200}\approx 14.14$ metres from one another.
+\midinsert
+$$\epsfbox{sheep.ps}$$
+\vskip5pt
+\caption{Five sheep in an enclosure.}
+\endinsert
+\goodbreak
+To use the pigeonhole principle, we need to decide what will be our ``pigeons'' and what will
+be our ``holes''. In this case, we can divide the enclosure into four square quadrants,
+each of side length $10$. We take our ``pigeons'' to be the sheep and our ``holes''
+to be the quadrants. Since there are five sheep and four quadrants, there is at least one quadrant
+that contains more than one sheep. But the quadrants are squares with side lengths of $10$ metres,
+so the farthest any two sheep can stand apart from each other while being in the same quadrant
+is the length of the diagonal. This is $\sqrt{10^2 + 10^2} = \sqrt{200}$ by the Pythagorean
+theorem.
+
+\medskip\boldlabel\datestamp{25} Subset sums.
+Let $A\subseteq \{1,2,\ldots, 100\}$ with $|A|=10$. Prove
+that there has to be subsets $X\subseteq A$ and $Y\subseteq A$ with $X\ne Y$ such that
+$$\sum_{x\in X} x = \sum_{y\in Y} y.$$
+We define $f:2^A\to \NN$ by letting
+$$f(X) = \sum_{x\in X} x.$$
+(So $f(\emptyset) = 0$.) What is the maximum possible sum? Well, this happens when
+$A = \{100,99, 98,\ldots, 91\}$ and $X = A$, in which case
+$$f(X) = \sum_{x\in X} x \le 100 + 99 + 98 + \cdots + 91 \le 1000.$$
+The minimum possible sum is $0$, when $X = \emptyset$,
+Hence let $2^A$ be the set of pigeons and let $\{0,1,2,\ldots,1000\}$ be the set of holes
+There are $2^{|A|} = 2^{10} = 1024$ pigeons and $1001$ holes, so by the pigeonhole principle
+there must be two pigeons that nest in the same hole; in other words, $f$ is not injective.
+Hence there are $X,Y\in 2^A$ with $X\ne Y$ such that $f(X) = f(Y)$.
+
+\medskip\boldlabel Theoretical limits to data compression. For $n\in \NN$, let $B_n$ be the set of all
+binary strings of length $n$. Let $B$ be the set of all binary strings of any finite length,
+so that
+$$B = B_0 \cup B_1 \cup B_2 \cup \cdots =  \bigcup_{n=0}^\infty B_n.$$
+Consider a string $x\in B$ as a file (containing some data encoded by the $0$s and $1$s).
+Let $L(x)$ be the length of $x$, i.e., the number
+of bits needed to encode $x$. Consider an algorithm $\alpha$ that takes a file as input and
+outputs a file. We can view $\alpha$ as a function $\alpha : B\to B$. Now imagine that the goal
+of $\alpha$ is to {\it compress} the file $x$; that is, we would ideally like to feed
+a file $x$ to $\alpha$ and get a file $y$ that is shorter in length.
+A compression algorithm $\alpha$ is {\it lossless} that for any $y$ in the range of $\alpha$,
+we can recover $x$ such that $\alpha(x) = y$. In other words, $\alpha$ is {\it lossless}
+if and only if it is injective when considered as a function $\alpha : B\to B$.
+Examples of widely-used file formats that feature lossless compression algorithms are
+{\mc ZIP}, {\mc PNG}, and {\mc GIF}.
+
+What we will prove is that any compression algorithm that decreases the file size for some inputs
+must increase the file size for other inputs. (For example, there exist some files such that
+compressing them into a {\mc ZIP} file causes them to get larger!)
+This is formalised in the following theorem.
+
+\proclaim Theorem \advthm. Let $\alpha : B\to B$ be a lossless compression algorithm such
+that for at least one $x\in B$, we have $L\bigl(\alpha(x)\bigr) < L(x)$.
+Then there exists $y\in B$ such that $L\bigl(\alpha(y)\bigr) > L(y)$.
+
+\proof For a contradiction, assume that there exists $\alpha: B\to B$ that is injective
+such that for at least some $x\in B$, $L\bigl(\alpha(x)\bigr) < L(x)$, and for all
+$y\in B$, $L\bigl(\alpha(x)\bigr) \le L(x)$.
+
+Pick $x\in B$ to be of minimal length such that
+$L\bigl(\alpha(x)\bigr) < L(x)$. Set $n = L(x)$ and $m = L\bigl(\alpha(x)\bigr)$, so that
+$n>m$. By minimality of $x$, if $y\in B$ with $L(y)<n$, then $L\bigl(\alpha(y)\bigr) \ge L(y)$,
+which, combined with our earlier assumption, means that
+$L\bigl(\alpha(y)\bigr) = L(y)$ for all $y\in B$ with $L(y)<n$.
+
+Define a new algorithm that has a restricted domain and codomain. We let
+$\alpha' : \bigl(B_m \cup \{x\}\bigr) \to B_m$ be given by $\alpha'(x) = \alpha$.
+The codomain is correct because $\alpha(y) \in B_m$ for all $y\in B_m$ and $\alpha(x)\in B_m$
+as well, by choice of $x$. Let $B_m\cup\{x\}$ be the set of pigeons and let $B_m$
+be the set of pigeonholes. Since there are $|B_m|+1$ pigeons and $|B_m|$ holes, there must
+be some $y,y'\in B_m\cup\{x\}$ with $y\ne y$ such that $\alpha'(y) = \alpha'(y')$. But this
+means that $\alpha(y) = \alpha(y')$ as well. Hence $\alpha$ is not injective: a contradiction.\slug
+
+\advsect Recurrences
+
+Let $(a_n) = a_0, a_1, a_2, \ldots$ be a sequence of real numbers (in all
+the examples we consider, they will be integers). A {\it recurrence relation}
+for the sequence is a formula for $a_n$ in terms of some of the $a_m$ with $m<n$.
+For instance, the {\it Fibonacci sequence} is the sequence $F_n$ defined by $F_0 = 0$,
+$F_1 = 1$, and $F_n = F_{n-1} + F_{n-2}$ for $n\ge 1$. The first few terms of $F_n$ are
+$$ 0 , 1, 1, 2, 3, 5, 8, 13, 21, \ldots.$$
+The sequence is named for the Italian mathematician Fibonacci, who in his 1202 book {\sl Liber Abaci}
+uses these numbers to describe the growth of a population of rabbits under certain idealised
+conditions. The sequence was known to Indian linguists many centuries earlier, as it pertains
+to the number different ways to arrange long and short Sanskrit syllables into lines of poetry.
+
+\medskip\boldlabel Finding recurrence relations.
+First we show an example of how to find a recurrence relation for a given sequence.
+For $n\ge 1$, let $C_n$ be the number of binary strings of length $n$ not containing $00$.
+We have $C_1 = 2$ ($0$ and $1$ are both fine) and $C_2 = 3$ ($01$, $10$, and $11$ are all fine,
+but $00$ is not).
+Now let $n\ge 2$ and consider how to derive a count for $C_n$, assuming
+we know the smaller values of $n$. Consider a string of length $n$ with no $00$.
+Either the string ends in $1$ or it ends in $10$, since if it ends with a $0$ the second-last
+bit must be $1$. In the first case, there are $n-1$ bits preceding the $1$, and the only requirement
+is that this string of $n-1$ bits not contain $00$. So there are $C_{n-1}$ possibilities in the
+first case. In the second case, there are $n-2$ bits preceding the $10$, and the requirement
+again is that the string $n-2$ bits not contain $00$. Hence in the second case there are
+$C_{n-2}$ possibilities. Adding up the two possibilities, we have
+$$C_n = C_{n-1} + C_{n-2}.$$
+This is the exact same recurrence relation as for the Fibonacci numbers! The only difference
+is that the answer is shifted (e.g., $C_1 = 3 = F_3$). So in general, $C_n = F_{n+2}$
+for all integers $n\ge 1$.
+
+Why don't we do another one. Let $A_n$ be the number of strings of length $n$ using the symbols
+$0$, $1$, $2$, $3$, such that there are an even number of $1$s. We have $A_1 = 3$, and
+we can calculate $A_2 = 10$, since there are $16$ possible strings, and the forbidden
+ones are $12$, $13$, $14$, $21$, $31$, and $41$. Now let $n\ge 2$ and consider a
+string of length $n$ with an even number of $1$s. Either the last number is $1$ or it is not.
+In the first case, there are $n-1$ bits preceding the $1$, and we need them to have an {\it odd}
+number of bits. The number of ways of doing this is $4^{n-1} - A_{n-1}$ (subtract the
+ones with an even number of $1$s from the total). In the second case, the last digit
+is $0$, $2$, or $3$, and in each case we just need the preceding $n-1$ bits to have an even
+number of $1$s, in $A_n$ ways. Hence there are $3A_{n-1}$ possibilities for this case.
+Adding everything up, we have
+$$A_n = 3A_{n-1} + 4^{n-1} - A_{n-1} = 2A_{n-1} + 4^{n-1}.$$
+
+\medskip\boldlabel Solving recurrences. Suppose we have a recurrence relation for a sequence $a_n$.
+We would like to ``solve'' this recurrence by deriving a formula for $a_n$ that isn't recursive,
+that is, one that does not contain any instances of $a_m$ for $m<n$ on the right-hand side.
+
+For example, in the very simple case where the recurrence has only one term, we have the following
+solution.
+
+\edef\prophomoone{\the\thmcount}
+\proclaim Proposition \advthm. Suppose $(a_n)$ is a sequence with $a_0 = c$ for some $c\in \RR$
+and, for all $n\ge 0$, $a_n = b\cdot a_{n-1}$ for some $b\in \RR$.
+Then for all $n\ge 0$, we have
+$$ a_n = c b^n.$$
+
+\proof By induction on $n$. For $n=0$, we have
+$$ a_0 = c = cb^0.$$
+(We define $0^0 = 1$). Now suppose the formula proven for $n$ and observe that
+$$ a_{n+1} = b\cdot a_n = b\cdot cb^n = cb^{n+1}.\noskipslug$$
+
+That was too easy. In general, we will consider recurrences of the form
+$$ a_n = f_1(n) a_{n-1} + f_2(n)a_{n-2} + \cdots + f_k(n) a_{n-k} + g(n)$$
+where $k\ge 1$ is an integer and $f_1, f_2, \ldots, f_k, g : \NN \to \RR$ are functions.
+These recurrences are said to be {\it linear of degree $k$}. If $g(n) = 0$,
+the recurrence is called {\it homogeneous}; otherwise, it is called {\it non-homogeneous}.
+
+For a non-homogeneous recurrence, the same recurrence without the $g(n)$ term is called
+the {\it associated homogeneous recurrence}.
+If, for all $1\le i\le k$, the function $f_i(n)$ is a constant function, then the recurrence
+is said to have {\it constant coefficients}.
+
+A sequence $(p_n)$ that satisfies the recurrence is called a {\it particular solution}.
+A {\it general solution} is a formula describing all possible solutions using some parameters.
+If we specify values for the first few terms $a_0, a_1, a_2,\ldots$, these are called
+{\it initial conditions} for the recurrence.
+
+\medskip\boldlabel\datestamp{26}{} Non-homogeneous recurrences of degree 1.
+So far, the only recurrences we know how to solve are
+linear homogeneous recurrences of degree $1$.
+As a step up in difficulty, let us now consider non-homogeneous recurrences of degree $1$.
+
+\edef\thmnonhomoone{\the\thmcount}
+\proclaim Theorem \advthm. Consider the non-homogeneous recurrence
+$$a_n = f(n)a_{n-1} + g(n),$$
+where $f,g:\NN\to \RR$. If $(p_n)$ is any particular solution to the recurrence
+and $(h_n)$ is a general solution to the associated homogenous recurrence, i.e.,
+$$ h_n = f(n) h_{n-1}$$
+for all $n\ge 1$, then the general solution for the recurrence is
+$$a_n = h_n + p_n.$$
+
+\proof Let $(b_n)$ be any solution to the recurrence and
+let $(p_n)$ be a particular solution to the recurrence, so that for $n\ge 1$ we have
+$$b_n = f(n)b_{n-1}+g(n)\qquad\hbox{and}\qquad p_n = f(n) p_{n-1} + g(n).$$
+Let $(c_n)$ be the sequence given by $c_n = b_n - p_n$. Note that for all $n\ge 1$,
+$$\eqalign{
+f(n) c_{n-1} &= f(n)(b_{n-1} - p_{n-1}) \cr
+&= f(n)b_{n-1} - f(n)p_{n-1} \cr
+&= b_n - g(n) - \bigl( p_n - g(n)\bigr) \cr
+&= b_n - p_n \cr
+&= c_n. \cr
+}$$
+So $(c_n)$ is a solution to the associated homogeneous recurrence. In other words, it is
+obtained from $(h_n)$ by specifying values for its parameters. Hence, letting
+$$a_n = h_n + p_n,$$
+the solution $b_n = c_n + p_n$ is derived by specifying specific values for parameters
+in the general solution $(a_n)$.\slug
+
+Here, then, is the general method for solving a non-homogeneous recurrence of degree $1$:
+\medskip
+\item{i)} Solve the associated homogeneous recurrence, using Proposition~{\prophomoone},
+to get a general solution $(h_n)$ with parameters.
+\smallskip
+\item{ii)} Find a solution $p_n$ to the recurrence. To do so, we often find it useful
+to guess that $p_n$ should have a similar form to $g(n)$. We'll see an example of this soon.
+\smallskip
+\item{iii)} Theorem~{\thmnonhomoone} tells us that the general solution to our recurrence is
+$$a_n = h_n + p_n.$$
+\smallskip
+\item{iv)} There will be unspecified parameters in the general solution, so if possible, we solve
+for these parameters by calculating some initial conditions.
+
+To illustrate this method, let us revisit the example from earlier about the number $A_n$ of
+strings of $0$, $1$, $2$, and $3$ that have an even number of $1$s. We derived
+the recurrence
+$$A_n = 2A_{n-1} + 4^{n-1}.$$
+Now let's solve it using Proposition~{\prophomoone} and Theorem~{\thmnonhomoone}.
+First, we find the general solution to the associated homogeneous
+recurrence $A_n = 2A_{n-1}$. By Proposition~{\prophomoone}, this is
+$$h_n = \alpha 2^n,$$
+where $\alpha\in \RR$ is a parameter that we can fill in by looking at initial conditions,
+but we'll skip that for now.
+Now we want a particular solution $p_n$. Since $g(n) = 4^{n-1}$, we guess that
+$p_n = \beta 4^n$ for some $\beta\in \RR$. We want $p_n = 2p_{n-1} + 4^{n-1}$, so
+$$ \beta 4^n = 2\beta 4^{n-1} + 4^{n-1}.$$
+Dividing through by $4^{n-1}$ yields
+$$4\beta = 2\beta + 1,$$
+so we can take $\beta = 1/2$, and $p_n = (1/2)4^n$ is a particular solution.
+By Theorem~{\thmnonhomoone}, the general solution to our original recurrence is
+$$\alpha \cdot 2^n  + {1\over 2} 4^n.$$
+Now, earlier we derived the initial condition $A_1 = 3$, so
+$$3 = A_1 = \alpha\cdot 2^1 + {1\over 2} 4^1.$$
+Solving this, we obtain $\alpha = {1\over 2}$. So for $n\ge 1$, the final formula for $A_n$ is
+$$A_n = {1\over 2} 2^n + {1\over 2} 4^n.$$
+We verify that
+$$A_2 = {1\over 2} 2^2 + {1\over 2} 4^2 = 2+8 = 10,$$
+which matches the count we performed earlier.
+
+In this example, we guessed that the particular solution
+$p_n$ should take the form $\beta 4^n$ for some $\beta$, since $g(n) = 4^{n-1}$. This sort of
+guess works for certain other classes of $g(n)$ as well. The general principle is outlined in Table~1.
+
+\topinsert
+$$\vcenter{\vbox{
+\eightpoint
+\tabskip=2em plus2em minus.5em
+\halign{
+\hfil#\hfil & \hfil#\hfil \cr
+\noalign{\hrule}
+\noalign{\medskip}
+If $g(n)$ is \dots & \dots then take $p_n$ to be \dots \cr
+\noalign{\medskip}
+\noalign{\hrule}
+\noalign{\medskip}
+a constant $q$, & a constant $r$. \cr
+\noalign{\medskip}
+a linear function $q_0 + q_1n$, & a linear function $r_0 + r_1 n$. \cr
+\noalign{\medskip}
+an exponential function $qt^n$, & an exponential function with the same base $rt^n$.\cr
+\noalign{\medskip}
+\noalign{\hrule}
+}
+}}$$
+\vskip5pt
+{\centerline{\ninepoint{\bf Table~1.\enspace}Particular solutions $p_n$ for certain classes of $g(n)$.}}
+\bigskip
+\endinsert\goodbreak
+
+Theorem~{\thmnonhomoone} is a special case of a more general theorem, whose proof is beyond the
+scope of this course.
+
+\edef\thmnonhomo{\the\thmcount}
+\proclaim Theorem \advthm. Consider the recurrence
+$$ a_n = f_1(n) a_{n-1} + f_2(n)a_{n-2} + \cdots + f_k(n) a_{n-k} + g(n)$$
+for some $k\ge 1$ and $f_1, f_2,\ldots, f_k,g : \NN\to \RR$. If $(p_n)$ is any particular solution
+and $(h_n)$ is a solution to the associated homogeneous recurrence, i.e.,
+$$h_n = f_1(n) a_{n-1} + f_2(n)a_{n-2} + \cdots + f_k(n) a_{n-k}$$
+for all $n\ge 1$, then the general solution for $(a_n)$ is given by $a_n = h_n + p_n$.\slug
+
+\medskip\boldlabel Recurrences with constant coefficients.
+We now turn to the case where the $f_i$ are all constant functions. This means
+there are constants $c_1, c_2, \ldots, c_k$ such that $f_1(n) = c_1$ for all $n$,
+$f_2(n) = c_2$ for all $n$, and so on.
+In light of Theorem~{\thmnonhomo}, we can
+restrict our attention to homogeneous recurrences, since we can solve non-homogeneous
+recurrences by solving their associated homogeneous recurrences and then applying the theorem.
+Homogeneous recurrences with constant coefficients look like
+$$a_n = c_1 a_{n-1} + c_2 a_{n-2} + \cdots + c_k a_{n-k},$$
+for some degree $k\ge 1$. We define the {\it characteristic equation} of such a recurrence to be the
+equation
+$$x^k = c_1 x^{k-1} + c_2 x^{k-2} + \cdots + c_k,$$
+and the {\it characteristic polynomial} is the polynomial obtained by subtracting
+the right-hand side from the left-hand side:
+$$x^k - c_1 x^{k-1} + c_2 x^{k-2} + \cdots + c_k$$
+It turns out that this
+polynomial entirely determines the general solution to the recurrence. We shall state
+and prove this theorem in the special case that $k=2$ and the two roots of the characteristic
+polynomial are distinct.
+
+\edef\thmdegtwodistinct{\the\thmcount}
+\proclaim Theorem \advthm. Consider the recurrence given by
+$$a_n = c_1 a_{n-1} + c_2 a_{n-2}.$$
+If the characteristic polynomial of this recurrence has two distinct roots $r_1$ and $r_2$, then
+the general solution of the recurrence is
+$$a_n = \alpha_1 {r_1}^n + \alpha_2 {r_2}^n.$$
+Supposing we know the initial conditions $a_0$ and $a_1$, we can solve for
+$\alpha_1$ and $\alpha_2$ via the identities
+$$\alpha_1 = {a_0 r_2 - a_1\over r_2 - r_1}
+\qquad\hbox{and}\qquad
+\alpha_2 = {a_1 - a_0 r_1\over r_2-r_1}.$$
+
+\proof The characteristic polynomial of the recurrence is
+$$ x^2 - c_1 x + c_2.$$
+For any root $r$ of this polynomial, we have
+$$ r^2 = c_1 r + c_2,$$
+so for any $n\ge 2$ we can multiply both sides by $r^{n-2}$
+to get
+$$ r^n = c_1 r^{n-1} + c_2 r^{n-2}.$$
+Hence the sequence $(r^n)$ is a solution to the recurrence. Conversely,
+if $r$ is such that $(r^n)$ is a solution to the recurrence, then
+$r^n = c_1 r^{n-1} + c_2 r^{n-2}$ for all $n\ge 2$, so we can divide by $r^{n-2}$
+to get $r^2 = c_1 r + c_2$; that is, $r$ is a root of the characteristic polynomial.
+
+What we have found so far is that $a_n = {r_1}^n$ and $a_n = {r_2}^n$ are both solutions to the
+recurrence, and that they are the only two solutions that take the form $r^n$ for some $r$.
+Now we must show that for any scalars $\alpha_1$ and $\alpha_2$,
+$$f_n = \alpha_1 {r_1}^n + \alpha_2 {r_2}^n$$
+is also a solution to the recurrence. Well,
+$$\eqalign{
+c_1 f_{n-1} + c_2 f_{n-2} &=
+c_1\bigl(\alpha_1 {r_1}^{n-1} + \alpha_2 {r_2}^{n-1}\bigr)
++ c_2\bigl(\alpha_1 {r_1}^{n-2} + \alpha_2 {r_2}^{n-2}\bigr) \cr
+&= c_1\alpha_1 {r_1}^{n-1} + c_1 \alpha_2 {r_2}^{n-1} + c_2\alpha_1 {r_1}^{n-2}
++ c_2\alpha_2 {r_2}^{n-2} \cr
+&= \alpha_1 \bigl( c_1 {r_1}^{n-1} + c_2 {r_1}^{n-2} \bigr)
++ \alpha_2 \bigl( c_1 {r_2}^{n-1} + c_2 {r_2}^{n-2} \bigr) \cr
+&= \alpha_1 {r_1}^n + \alpha_2 {r_2}^n \cr
+&= f_n,
+}$$
+So the claimed expression is in fact the general solution to the recurrence.
+
+Lastly,
+we must show that we can solve for $\alpha_1$ and $\alpha_2$ once we know
+the initial conditions $a_0$ and $a_1$. We have
+$$ a_0 = \alpha_1 {r_1}^0 + \alpha_2 {r_2}^0 = \alpha_1 + \alpha_2$$
+and
+$$ a_1 = \alpha_1 r_1 + \alpha_2 r_2.$$
+We can arrange these equations into the matrix equation
+$$\pmatrix{ 1 & 1 \cr r_1 & r_2 } \pmatrix{ \alpha_1 \cr \alpha_2 } = \pmatrix{ a_0 \cr a_1 }.$$
+This equation is solvable if and only if the determinant $r_2 - r_1$ of the matrix
+$\bigl( {1\atop r_1} {1\atop r_2} \bigr)$
+is nonzero. But we assumed $r_1\ne r_2$, so this is the case. Using the formula
+for the inverse of a $2\times 2$ matrix, we find that the inverse of
+$\bigl( {1\atop r_1} {1\atop r_2} \bigr)$ is
+$${1\over r_1-r_2} \pmatrix{ r_2 & -1 \cr -r_1 & 1 },$$
+whence
+$$\pmatrix{\alpha_1\cr \alpha_2} = {1\over r_2-r_1} \pmatrix{ r_2 & -1 \cr -r_1 & 1 }
+\pmatrix{a_0 \cr a_1}.$$
+We conclude that
+$$\alpha_1 = {a_0 r_2 - a_1\over r_2 - r_1}
+\qquad\hbox{and}\qquad
+\alpha_2 = {a_1 - a_0 r_1\over r_2-r_1},$$
+which is what we wanted to show.
+\slug
+
+Let us use this theorem to derive a non-recursive formula for the Fibonacci numbers.
+Recall that they satisfy the recurrence $F_n = F_{n-1} + F_{n-2}$, so the characteristic
+polynomial is
+$$x^2 - x - 1.$$
+By the quadratic formula, the two roots of this polynomial are
+$$r_1 = {1 - \sqrt 5\over 2}\qquad\hbox{and}\qquad r_2 = {1+\sqrt 5\over 2}.$$
+The roots are distinct; in fact, we have
+$$r_2 - r_1 = {1+\sqrt 5 - (1-\sqrt 5)\over 2} = {2\sqrt 5\over 2} = \sqrt 5.$$
+In any case, Theorem~{\thmdegtwodistinct} applies, and we have the general
+solution
+$$F_n = \alpha_1 {r_1}^n + \alpha_2 {r_2}^n.$$
+We have $F_0 = 0$ and $F_1 = 1$, so using the formulas given by the theorem,
+we calculate
+$$\alpha_1 = {0\cdot r_2 - 1\over \sqrt 5} = {-1\over \sqrt 5}$$
+and
+$$\alpha_2 = {1 - 0\cdot r_2\over \sqrt 5} = {1\over \sqrt 5}.$$
+So
+$$F_n = {-1\over \sqrt 5}\biggl({1 - \sqrt 5\over 2}\biggr)^n
++ {1\over \sqrt 5}\biggl({1+\sqrt 5\over 2}\biggr)^n.$$
+From this formula, it is not even clear that $F_n \in \NN$ for all $n$, though it must be
+since the first two terms are integer and each term is the sum of the two previous entries.
+
+The quantity $r_1 = (1+\sqrt 5)/2\approx 1.618$ is called the {\it golden ratio}, and is
+often denoted $\varphi$.
+We have $r_2 = (1-\sqrt 5)/2 \approx -0.618$, so $(-1/\sqrt 5){r_2}^n$ is very small in absolute value
+for large $n$.
+This means that
+$$ F_n \sim {1\over \sqrt 5} \biggl( {1+\sqrt 5\over 2}\biggr)^n \approx {1.61^n\over \sqrt 5}.$$
+In other words,
+$$\lim_{n\to\infty} {-1\over \sqrt 5}\biggl({1 - \sqrt 5\over 2}\biggr)^n = 0,$$
+so
+$$\lim_{n\to \infty} {F_{n+1}\over F_n} = {1+\sqrt 5\over 2}.$$
+Incidentally, it is not just the first few decimal places in the approximations of $r_1$ and $r_2$
+that match. As an exercise, prove that $r_2 = 1/r_1 = 1-r_1$.
+
+The case where the characteristic polynomial has a repeated root is not covered by
+Theorem~{\thmdegtwodistinct}. Instead, we have the following theorem, which we will
+not have time to prove in this course (though the proof is by no means difficult).
+
+\proclaim Theorem \advthm. Consider the recurrence given by
+$$a_n = c_1 a_{n-1} + c_2 a_{n-2}.$$
+If the characteristic polynomial of this recurrence has a repeated root $r$,
+then the general solution of the recurrence is
+$$a_n = \alpha_1 r^n + \alpha_2 nr^n.\noskipslug$$
+
+Theorem~{\thmdegtwodistinct} can also be generalised to higher degrees $k$. Its proof is
+morally the same as the one we gave for the case $k=2$.
+
+\proclaim Theorem \advthm. Let $k\ge 2$ be an integer and consider the recurrence given by
+$$a_n = c_1 a_{n-1} + c_2 a_{n-2} + \cdots + c_k a_{n-k}.$$
+If the characteristic polynomial of this recurrence has $k$ roots
+$r_1,r_2,\ldots, r_k$ that are pairwise distinct, then
+the general solution of the recurrence is
+$$a_n = \alpha_1 {r_1}^n + \alpha_2 {r_2}^n + \cdots + \alpha_k {r_k}^n.\noskipslug$$
+
+\bye

### 2024-11-22 teaching/math240f24/draft.tex
lecture 21 nov

@@
 \hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 17.\kern .5em Basic counting techniques}{89}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 18.\kern .5em Permutations and
+combinations}{92}}
 
@@
 that either has both endpoints in $A_i$ or both endpoints in $B_i$. This means
-that $G$ is not bipartite either.\slug
+that $A\cup B$ is not a valid bipartition, and since $A$ and $B$ were arbitrary,
+we conclude that
+$G$ is not bipartite either.\slug
 
@@
 $$100\cdot 99\cdot 98 \cdots 82\cdot 81 = {100!\over 80!}.$$
-\bye
-
-\noindent\datestamp{23}{} \hskip\parindent
+
+\noindent\datestamp{21}{xi} \hskip\parindent
 In general, we define a {\it $k$-permutation} of a set $X$ to be an ordered subset
@@
 
-\medskip\boldlabel\datestamp{24}{} Identities involving binomial coefficients.
+\medskip\boldlabel Identities involving binomial coefficients.
 We already observed earlier that the number of ways of picking $k$ objects out of a
@@
 
-Another thing we notice about Pascal's triangle is that the first row sums to $1$, the second
+The only number that appears infinitely many times is $1$, since any $n\in \ZZ$
+can only appear in the first $n+1$ rows of the triangle (if $m>n$, then for all
+$0\le k\le m$, ${m\choose k}$ is either $1$ or at least $m$). A fun conjecture to think
+about is Singmaster's conjecture, put forth by D.~Singmaster in 1971: {\sl There
+exists some number $C$ such that every integer $n\ge 2$ appears in Pascal's triangle
+at most $C$ times.} Whether this statement is true is still an open problem.
+The integer (besides $1$) known to appear most often in Pascal's triangle is
+$3003$. It appears eight times, and no other integer is known to appear eight times.
+
+But let us not be too diverted by lofty conjectures. Returning to things we can prove,
+another thing we notice about Pascal's triangle is that the first row sums to $1$, the second
 row sums to $2$, the third to $4$, the fourth to $8$, and so on. We might therefore
@@
 any of them, and hence $p$ does not divide $k!(p-k)!$. We conclude that $p\divides {p\choose k}$.\slug
-
-\medskip\boldlabel A day spent selling milk.
+\bye
+
+\medskip\boldlabel\datestamp{24}{} A day spent selling milk.
 The binomial theorem can be used to find the value of a specific coefficient in a polynomial,

### 2024-11-27 misc/expo/additiveout.tex
lecture nov 26

--- 2024-01-26 misc/expo/additive.tex
@@
 
-\bye
+\parskip=0pt
+\hyphenpenalty=-1000 \pretolerance=-1 \tolerance=1000
+\doublehyphendemerits=-100000 \finalhyphendemerits=-100000
+\frenchspacing
+\def\bref#1{[#1]}
+\def\beginref{\noindent}
+\def\endref{\medskip}
+\vskip\parskip
+
+\beginref
+Antal Balog
+and Endre Szemer\'edi,
+``A statistical theorem of set addition,''
+{\sl Combinatorica}\/
+{\bf 14}
+(1994),
+263--268.
+\endref
+\beginref
+William Timothy Gowers,
+``A new proof of Szemer\'edi's theorem for arithmetic progressions of length four,''
+{\sl Geometric and Functional Analysis}\/
+{\bf 8}
+(1998),
+529--551.
+\endref
+\beginref
+Askold Khovanskii,
+``Newton polyhedron, Hilbert polynomial, and sums of finite sets,''
+{\sl Functional Analysis and its Applications}\/
+{\bf 26}
+(1992),
+276--281.
+\endref
+\beginref
+Melvyn Bernard Nathanson
+and Imre Zolt\'an Ruzsa,
+``Polynomial growth of sumsets in abelian semigroups,''
+{\sl Journal de Th\'eorie des Nombres de Bordeaux}\/
+{\bf 14}
+(2002),
+553--560.
+\endref
+\beginref
+Giorgis Petridis,
+``New proofs of Pl\"unnecke-type estimates for product sets in groups.,''
+{\sl Combinatorica}\/
+{\bf 32}
+(2012),
+721--733.
+\endref
+\beginref
+Helmut Pl\"unnecke,
+{\sl Eigenschaften und Absch\"atzungen von Wirkungsfunktionen}
+(Bonn:
+Berichte der Gesellschaft f\"ur Mathematik und Datenverarbeitung,
+1969).
+\endref
+\beginref
+Klaus Friedrich Roth,
+``On certain sets of integers,''
+{\sl Journal of the London Mathematical Society}\/
+{\bf 28}
+(1953),
+104--109.
+\endref
+\beginref
+Imre Zolt\'an Ruzsa,
+``An application of graph theory to additive number theory,''
+{\sl Scientia, Series A}\/
+{\bf 3}
+(1989),
+97--109.
+\endref
+\beginref
+Imre Zolt\'an Ruzsa, ``Sums of finite sets,'' {\sl Number Theory: New York Seminar 1991--1995} (1996).
+\endref
+\beginref
+Endre Szemer\'edi,
+``On sets of integers containing no $k$ elements in arithmetic progression,''
+{\sl Acta Arithmetica}\/
+{\bf 27}
+(1975),
+199--245.
+\endref
+\beginref
+Terence Tao
+and Van Ha Vu,
+{\sl Additive Combinatorics}
+(Cambridge:
+Cambridge University Press,
+2006).
+\endref
+\beginref
+Bartel Leendert van~der~Waerden,
+``Beweis einer Baudetschen Vermutung,''
+{\sl Nieuw Archief voor Wiskunde}\/
+{\bf 15}
+(1927),
+212--216.
+\endref
+\beginref
+\goodbreak\bye

### 2024-11-27 teaching/math240f24/draft.tex
lecture nov 26

@@
 combinations}{92}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 19.\kern .5em Applications of the
+pigeonhole principle}{100}}
+\smallskip
+\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 20.\kern .5em Recurrences}{102}}
 
@@
 any of them, and hence $p$ does not divide $k!(p-k)!$. We conclude that $p\divides {p\choose k}$.\slug
-\bye
-
-\medskip\boldlabel\datestamp{24}{} A day spent selling milk.
+
+\medskip\boldlabel\datestamp{26}{xi} A day spent selling milk.
 The binomial theorem can be used to find the value of a specific coefficient in a polynomial,
@@
 
-\medskip\boldlabel\datestamp{25} Subset sums.
+\medskip\boldlabel Subset sums.
 Let $A\subseteq \{1,2,\ldots, 100\}$ with $|A|=10$. Prove
@@
 conditions. The sequence was known to Indian linguists many centuries earlier, as it pertains
-to the number different ways to arrange long and short Sanskrit syllables into lines of poetry.
+to the number of different ways to arrange long and short Sanskrit syllables into lines of poetry.
+\bye
 

### 2024-11-29 teaching/math240f24/draft.tex
lecture 28 nov

@@
 \font\titlefont=cmbx12 at 14 pt
-\def\cont#1#2{#1\hskip-2pt\leaders\hbox to 8pt{\hss.\hss}\hfil\hbox to 1em{\hss#2}}
+\def\cont#1#2{#1\hskip-2pt\leaders\hbox to 8pt{\hss.\hss}\hfil\hbox to 2em{\hss#2}}
 \def\divides{\mid}
@@
 For instance, the {\it Fibonacci sequence} is the sequence $F_n$ defined by $F_0 = 0$,
-$F_1 = 1$, and $F_n = F_{n-1} + F_{n-2}$ for $n\ge 1$. The first few terms of $F_n$ are
+$F_1 = 1$, and $F_n = F_{n-1} + F_{n-2}$ for $n\ge 2$. The first few terms of $F_n$ are
 $$ 0 , 1, 1, 2, 3, 5, 8, 13, 21, \ldots.$$
 The sequence is named for the Italian mathematician Fibonacci, who in his 1202 book {\sl Liber Abaci}
-uses these numbers to describe the growth of a population of rabbits under certain idealised
+used these numbers to describe the growth of a population of rabbits under certain idealised
 conditions. The sequence was known to Indian linguists many centuries earlier, as it pertains
 to the number of different ways to arrange long and short Sanskrit syllables into lines of poetry.
-\bye
-
-\medskip\boldlabel Finding recurrence relations.
+
+\medskip\boldlabel\datestamp{28}{xi} Finding recurrence relations.
 First we show an example of how to find a recurrence relation for a given sequence.
@@
 
-\medskip\boldlabel\datestamp{26}{} Non-homogeneous recurrences of degree 1.
+\medskip\boldlabel Non-homogeneous recurrences of degree 1.
 So far, the only recurrences we know how to solve are
@@
 
-\medskip\boldlabel Recurrences with constant coefficients.
+\bye
+
+\medskip\boldlabel\datestamp{26}{} Recurrences with constant coefficients.
 We now turn to the case where the $f_i$ are all constant functions. This means
@@
 
+%\midinsert
+%$$\epsfbox{bijection.ps}$$
+%\vskip5pt
+%\caption{A planted plane tree with root label $5$ and its corresponding walk in $T_4$.}
+%\endinsert
+%\goodbreak
+
 \bye

### 2024-12-04 teaching/math240f24/draft.tex
Last lecture

@@
 
-\bye
-
-\medskip\boldlabel\datestamp{26}{} Recurrences with constant coefficients.
+\medskip\boldlabel\datestamp{03}{xii} Recurrences with constant coefficients.
 We now turn to the case where the $f_i$ are all constant functions. This means
@@
 $$a_n = \alpha_1 {r_1}^n + \alpha_2 {r_2}^n.$$
-Supposing we know the initial conditions $a_0$ and $a_1$, we can solve for
-$\alpha_1$ and $\alpha_2$ via the identities
+Supposing we know the initial conditions $a_0$ and $a_1$, we have the identities
 $$\alpha_1 = {a_0 r_2 - a_1\over r_2 - r_1}
@@
 
-The case where the characteristic polynomial has a repeated root is not covered by
-Theorem~{\thmdegtwodistinct}. Instead, we have the following theorem, which we will
-not have time to prove in this course (though the proof is by no means difficult).
-
-\proclaim Theorem \advthm. Consider the recurrence given by
-$$a_n = c_1 a_{n-1} + c_2 a_{n-2}.$$
-If the characteristic polynomial of this recurrence has a repeated root $r$,
-then the general solution of the recurrence is
-$$a_n = \alpha_1 r^n + \alpha_2 nr^n.\noskipslug$$
-
-Theorem~{\thmdegtwodistinct} can also be generalised to higher degrees $k$. Its proof is
+Theorem~{\thmdegtwodistinct} can be generalised to higher degrees $k$. Its proof is
 morally the same as the one we gave for the case $k=2$.
@@
 
+Returning to the setting $k=2$,
+the case where the characteristic polynomial has a repeated root is not covered by
+Theorem~{\thmdegtwodistinct}. Instead, we have the following theorem.
+
+\edef\thmdegtworepeated{\the\thmcount}
+\proclaim Theorem \advthm. Consider the homogeneous recurrence
+$$a_n = c_1 a_{n-1} + c_2 a_{n-2}.$$
+If the characteristic polynomial of this recurrence has a repeated root $r$,
+then the general solution of the recurrence is
+$$a_n = \alpha_1 r^n + \alpha_2 nr^n.$$
+Supposing we know the initial conditions $a_0$ and $a_1$, we have the identities
+$$ \alpha_1 = a_0\qquad\hbox{and}\qquad \alpha_2 = {a_1 - a_0 r\over r}.$$
+
+\proof By the same logic as in the proof of Theorem~{\thmdegtwodistinct}, we know that
+$r^n$ is a solution to the recurrence. We show that $b_n = nr^n$ is also a solution.
+Since $r$ is a repeated root to the characteristic polynomial $x^2 - c_1x - c_2$, we have
+$$x^2 - c_1 x - c_2 = (x-r)^2 = x^2 - 2rx + r^2,$$
+so by comparing coefficients we conclude that $c_1 = 2r$ and $c_2 = -r^2$.
+Now by expanding
+$$\eqalign{
+c_1 b_{n-1} + c_2 b_{n-2} &= c_2(n-1)r^{n-1} + c_2 (n-2) r^{n-2} \cr
+&= r^{n-2} \bigl( c_1 (n-1) r + c_2(n-2)\bigr) \cr
+&= r^{n-2} (c_1 n r - c_1 r + c_2 n - 2c_2 ) \cr
+&= r^{n-2} (2r^2 n - 2r^2 -r^2 n + 2r^2 ) \cr
+&= nr^n \cr
+&= b_n \cr
+}$$
+we see that $b_n = nr^n$ satisfies the recurrence as well. By similar reasoning
+as in the proof of Theorem~{\thmdegtwodistinct}, any linear combination of the two solutions
+$(r^n)$ and $(nr^n)$ is also a solution.
+
+Now we prove the further claim. Suppose we know $a_0$ and $a_1$. Then
+$$a_0 = \alpha_1 r^0 + \alpha_2 0r^0 = \alpha_1,$$
+and
+$$ a_1 = \alpha_1 r^1 + \alpha_2 r^1 = a_0 r + \alpha_2 r,$$
+and hence
+$$ \alpha_1 = a_0\qquad\hbox{and}\qquad \alpha_2 = {a_1 - a_0 r\over r}.\noskipslug$$
+
+Let's illustrate this theorem with a final example. We solve the recurrence
+$$a_n = 4 a_{n-1} - 4 a_{n-2} + n,$$
+with the initial conditions $a_0 = 1$ and $a_1 = 3$.
+First we solve the homogeneous recurrence
+$$ h_n = 4h_{n-1} + 4 h_{n-2}.$$
+The characteristic polynomial is $x^2 - 4x + 4 = (x-2)^2$, so by the above theorem,
+$$h_n = \alpha_1 2^n + \alpha_2 n 2^n.$$
+Now we need a particular solution $p_n$. We have $g(n) = n$, which is a linear function
+in $n$, so we guess that $p_n = cn+d$ for some real numbers $c$ and $d$. Then
+from the identity $p_n = 4p_{n-1} - 4p_{n-2} + n$, we have
+$$\eqalign{
+cn+d &=  4\bigl( c(n-1) + d\bigr) - 4 \bigl( c(n-2)+d\bigr) + n \cr
+0 &= -cn - d + 4cn - 4c + 4d - 4cn + 8c - 4d + n \cr
+0 &= (1-c) n + (4c-d).
+}$$
+So $1-c = 0$ and $4c-d = 0$, whence $c = 1$ and $d=4$. So $p_n = n+4$.
+This gives us the general solution
+$$ a_n = \alpha_1 2^n + \alpha_2 n 2^n + n + 4.$$
+We cannot use Theorem~{\thmdegtworepeated} to get $\alpha_1$ and $\alpha_2$, since
+that theorem concerned homogeneous recurrences. Instead we directly compute
+$$ 1 = a_0 = \alpha_1 2^0 + \beta\cdot 0 + 0 + 4 = \alpha_1 + 4,$$
+and derive $\alpha_1 = -3$, then find that
+$$3 = a_1 = \alpha_1 2 + \alpha_1 \cdot 2 + 1 + 4
+= -6 + 2\alpha_2 + 4.$$
+Hence $4=2\alpha_2$ and $\alpha_2 = 2$, which gives us a final solution of
+$$a_n = 3\cdot 2^n + 2n\cdot 2^n + n + 4$$
+to our recurrence.
+
 %\midinsert

### 2024-12-07 misc/expo/additive.tex
second convolution law

@@
 
-\parenproclaim Lemma {\advthm} (Convolution law). Let $Z$ is a finite abelian group and $f,g : \Z\to \CC$.
-For all $\chi\in \hat Z$, we have $\hat{f*g} (\chi) = \hat f(\chi)\hat g(\chi)$.
+\parenproclaim Lemma {\advthm} (Convolution laws). Let $Z$ be a finite abelian group
+and $f,g : Z \to \CC$.
+For all $\chi\in \hat Z$, we have $\hat{f*g} (\chi) = \hat f(\chi)\hat g(\chi)$,
+and $\hat{fg}(\chi) = (\hat f * \hat g)(\chi)$.
 
@@
 $$\hat{f*g}(\chi) = \ex_{y\in Z}\ex_{z\in Z} f(y)g(z)\bar{\chi(y)\chi(z)} = \hat f(\chi)\hat g(\chi),$$
+which is what we wanted to show.
+
+To prove the other convolution law, we instead expand
+$$\eqalign{
+(\hat f * \hat g)(\chi) &= \sum_{\psi \in \hat G} \hat f(\psi) g(\chi\psi^{-1})  \cr
+&= \sum_{\psi\in \hat G} \ex_{x\in G}f(x)\bar{\psi(x)}\ex_{y\in G} g(x+y) \bar{\chi(x+y)\psi^{-1}(x+y)} \cr
+&= \ex_{x\in G} f(x) \ex_{y\in G} g(x+y) \bar{\chi(x+y)} \sum_{\psi\in \hat G} \bar{\psi^{-1}(y)}.\cr
+}$$
+The innermost sum equals $|\hat Z|$ when $y=0$ and is $0$ otherwise. Hence
+$$\eqalign{
+(\hat f * \hat g)(\chi) &= \ex_{x\in G} f(x) \ex_{y\in G} g(x+y) \bar{\chi(x+y)} \one_{[y=0]} |\hat Z| \cr
+&= \ex_{x\in G} f(x) g(x) \bar{\chi(x)} \cr
+&= \hat{fg}(\chi), \cr
+}$$
 which is what we wanted to show.\slug

### 2025-02-12 about/vita.tex
Updates for 2025

@@
 On the homology of several number-theoretic set families.
-{\sl Enumerative Combinatorics and Applications}\/ {4},2 (2024), \#S2R12, 11~pp.
+{\sl Enumerative Combinatorics and Applications}\/ {\bf 4},2 (2024), \#S2R12, 11~pp.
 [arXiv:2206.12535]
@@
 \pubbegin F
-\starficitem ``Lanterns,'' Winner of the 2024 Constance Rooke Creative Nonfiction Prize. To appear in
-{\sl The Malahat Review}
+\starficitem ``Lanterns,'' {\sl The Malahat Review} {\bf 229} (2025), 7--13.
+Winner of the 2024 Constance Rooke Creative Nonfiction Prize.
 \ficitem ``Senang Diri,'' {\sl Existere} {\bf 42},2 (2023), 17--25.
@@
 \smallskip
-Met with two undergraduate students as part of the Directed Reading Program
+Met with undergraduate students as part of the Directed Reading Program
 on a weekly basis to give them an introduction to research-level
-mathematics in a casual setting. Focused on topics in extremal combinatorics related to the increasing
+mathematics in a casual setting.
+\begingroup\parindent=10pt
+\smallskip
+\thing Winter 2024: Topics in additive combinatorics, with a particular focus on Roth's theorem.
+\smallskip
+\thing Winter 2022: Topics in extremal combinatorics related to the increasing
 triples problem, as well as topics related to Khovanskii's theorem.
+\endgroup
 \medbreak
@@
 Developed firmware for the manufacture and function of in-home IoT gateways, smart switches,
-and related electronic products. Code primarily in \CEE\ and Python. Contributed to the open-source
+and related electronic products. Worked primarily in \CEE\ and Python. Contributed to the open-source
 Mongoose OS codebase.
@@
 and Imperial College London. Led problem sessions in the afternoons, with
-around forty graduate-level students were in attendance.
+around forty graduate-level students in attendance.
 \medbreak
@@
 \smallskip
-Grading of assignments in the following courses:
+Graded assignments in the following courses:
 \begingroup\parindent=10pt
@@
 \smallskip
-Oversaw academic affairs within SUMS council and acted as liaison between the undergraduate community and
+Oversaw academic affairs within {\mc SUMS} council and acted as liaison between the undergraduate community and
 mathematics faculty. Duties included representing the student body at department meetings,

### 2025-02-12 misc/expo/entropy.tex
Updates for 2025

@@
 \def\eps{\epsilon}
-\def\FF{{\bf F}}
+\def\FF{\hbox{\bf F}}
 \def\bar#1{\overline{#1}}
@@
 \def\dd{\op{\bf d}}
-\def\ee{\op{\bf e}}
+\def\AA{\op{\bf A}}
 \def\P{{\cal P}}
@@
 We are now done if we can show that entropy is nonnegative.
-This is a corollary of the following lemma. The following proof is a modification of
+This is a corollary of the following lemma, whose proof is a modification of
 one due to S.~Eberhard.
@@
 From this we see that
-$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z,W\} + \Eta\{Z\} + \Eta\{Y\}.$$
+$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z,W\} + \Eta\{Z\} + \Eta\{W\}.$$
 But then since conditioning does not increase entropy, we have
-$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z\} + 2\Eta\{Y\},$$
+$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z\} + 2\Eta\{W\},$$
 whence dividing both sides by $2$ completes the proof.\slug
@@
 The {\it entropic additive energy} between $X$ and $Y$ is
-$$ \ee\{X,Y\} = \Eta\{X_1, Y_1, X_2, Y_2\}.$$
+$$ \AA\{X,Y\} = \Eta\{X_1, Y_1, X_2, Y_2\}.$$
 This definition makes clear the analogy between this value and the additive energy of sets, but
@@
 $$\eqalign{
-\ee\{X,Y\} &= \Eta\{X_1, Y_1, X_2, Y_2, X+Y\} \cr
+\AA\{X,Y\} &= \Eta\{X_1, Y_1, X_2, Y_2, X+Y\} \cr
 &= 2\Eta\{X,Y,X+Y\} - \Eta\{X+Y\} \cr
@@
 \global\edef\eqenergybound{\the\eqcount}
-$$\ee\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2}\Eta\{Y\} - \log K,\adveq$$
+$$\AA\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2}\Eta\{Y\} - \log K,\adveq$$
 for some constant $K$, then
@@
 \global\edef\eqworsebound{\the\eqcount}
-$$\ee\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2}\Eta\{Y\} - \log K - 2C.\adveq$$
+$$\AA\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2}\Eta\{Y\} - \log K - 2C.\adveq$$
 In particular, if $X$ and $Y$ are independent, then we can recover~{\rm \refeq{\eqenergybound}}
@@
 $$\eqalign{
-\ee\{X,Y\} &= 2\Eta\{X,Y\} - \Eta\{X+Y\}  \cr
+\AA\{X,Y\} &= 2\Eta\{X,Y\} - \Eta\{X+Y\}  \cr
 &\ge 2\Eta\{X,Y\} - {1\over 2} \Eta\{X\} - {1\over 2} \Eta\{Y\} - \log K ,\cr
@@
 stronger than~\refeq{\eqsumbound} because the entropy $\Eta\{X,Y\}$ of the joint distribution appears in
-the formula for $\ee\{X,Y\}$, whereas~\refeq{\eqsumbound} says nothing whatsoever about this joint distribution.
+the formula for $\AA\{X,Y\}$, whereas~\refeq{\eqsumbound} says nothing whatsoever about this joint distribution.
 
@@
 $$\max\bigl(\Eta\{X_1- X_2\}, \Eta\{X_1-Y_2\}\bigr) \le \Eta\{X+Y\} + 2\II\{X:Y\}.$$
-The right-hand side of this expression can also be written $2\Eta\{X\} + 2\Eta\{Y\} - \ee\{X,Y\}$.
+The right-hand side of this expression can also be written $2\Eta\{X\} + 2\Eta\{Y\} - \AA\{X,Y\}$.
 
@@
 suppose that
-$$\ee\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} + \log K$$
+$$\AA\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} + \log K$$
 for some constant $K$.
@@
 &= \Eta\{X,Y\} - \Eta\{X+Y\} \cr
-&= \ee\{X,Y\} - \Eta\{X,Y\} \cr
+&= \AA\{X,Y\} - \Eta\{X,Y\} \cr
 &\ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} - \log K - \Eta\{X,Y\} \cr
@@
 }$$
-where in the fourth line we used the hypothesis on $\ee\{X,Y\}$, and in the last line we observed
+where in the fourth line we used the hypothesis on $\AA\{X,Y\}$, and in the last line we observed
 that $\Eta\{X\} + \Eta\{Y\} - \Eta\{X,Y\} \ge 0$. The bound
@@
 The previous lemma then gives
-$$\Eta\{ X_1 + Y_2\given X+Y\} \le 2\Eta\{X\} + 2\Eta\{Y\} - \ee\{X,Y\},$$
-and from our lower bound on $\ee\{X,Y\}$, we conclude that
+$$\Eta\{ X_1 + Y_2\given X+Y\} \le 2\Eta\{X\} + 2\Eta\{Y\} - \AA\{X,Y\},$$
+and from our lower bound on $\AA\{X,Y\}$, we conclude that
 $$\Eta\{ X_1 + Y_2\given X+Y\} \le {1\over 2} \Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K,$$
@@
 
+\advsect Sum-product phenomena
+
+We prove the following lemma.
+
+\proclaim Lemma \advthm.
+Let $X$ and $Y$ be random variables taking values in $\FF_p^*$ for some prime $p$.
+There exists $u\in \FF_p^*$ such that
+$$\Eta\{X,uY\} \ge {\Eta\{X,Y\}\over 2^{\Etaseven\{X,Y\}}} - 1,$$
+
+\proof Let $U$ be a random variable uniform on $\FF_p^*$ and independent of both $X$ and $Y$.
+Independence gives us the identity $\Eta\{X+UY\} = \Eta\{U\} = \log(p-1)$.
+We let $(X_1, Y_1)$, $(X_2,Y_2)$, and $U$ be conditionally independent trials
+of $(X,Y)$, $(X,Y)$, and $U$ respectively, relative to $X+UY$.
+Put
+$$H = \Eta\{X_1, Y_1, X_2, Y_2, U\}.$$
+Let $W = 1$ if $X_1 = X_2$ and $W=0$ otherwise. Then since $X_1 + UY_1 = X+UY$, we have
+$$\eqalign{
+H &= \Eta\{X_1, Y_1, X_2, Y_2, W, X+UY\} \cr
+&= \Eta\{U\} + \Eta\{W\} + \Eta\{X_1, Y_1, X_2, Y_2 \given W, X+UY\}.\cr
+}$$
+If $X_1 = X_2$, then $Y_1 = Y_2$, so by conditional independence,
+$$\eqalign{
+H &= \Eta\{U\} + \Eta\{W\} +  \pr\{X_1 = X_2\given X+UY\} \Eta\{X_1, Y_1 \given X+UY\} \cr
+&\qquad+ \pr\{X_1 \ne X_2 \given X+UY\} \Eta\{X_1,Y_1, X_2,Y_2\given X+UY\}.\cr
+&= \Eta\{U\} + \Eta\{W\} + \pr\{X_1 = X_2\}\bigl( \Eta\{X,Y\given X+UY\}\bigr) \cr
+&\qquad
++\pr\{X_1 \ne X_2\given X+UY\}\bigl( 2\Eta\{X,Y\given X+UY\} \bigr) \cr
+&= \Eta\{U\} + \Eta\{W\} + \Eta\{X,Y\given X+UY\}\cr
+&\qquad+\pr\{X_1 \ne X_2\given X+UY\}\Eta\{X,Y \given X+UY\} \cr
+}$$
+But
+$$\Eta\{X,Y\given X+UY\} = \Eta\{X,Y, X+UY\} - \Eta\{X+UY\} = \Eta\{X,Y\}$$
+by independence of $(X,Y)$ and $U$, so
+$$H =\bigl(1+\pr\{X_1\ne X_2\given X+UY\}\}\bigr) \Eta\{X,Y\} + \Eta\{U\} + \Eta\{W\}.$$
+
+On the other hand, by invertibility of each $u\in \FF_p^*$, we have
+$$\eqalign{
+H - \Eta\{U\} &= \Eta\{X_1, Y_1, X_2, Y_2\given U\} \cr
+&= {1\over p-1} \sum_{u\in \FF_p^*} \Eta\{X_1, Y_1, X_2, Y_2\given U=u\} \cr
+&= {1\over p-1} \sum_{u\in \FF_p^*} \Eta\{X_1, uY_1, X_2, uY_2\} \cr
+&= {1\over p-1} \sum_{u\in \FF_p^*} \AA\{X, uY\}.\cr
+}$$
+Consequently there exists $u\in \FF_p^*$ such that
+$$\eqalign{
+\AA\{X,uY\} &\le H-\Eta\{U\} \cr
+&= \bigl(1+\pr\{X_1\ne X_2\given X+UY\}\bigr) \Eta\{X,Y\} + \Eta\{W\} \cr
+&\le \bigl(1+\pr\{X_1\ne X_2\given X+UY\}\bigr) \Eta\{X,Y\} + 1, \cr
+}$$
+and for this choice of $u$ we have
+$$\Eta\{X+uY\} \ge \pr\{X_1 = X_2\given X+UY\} \Eta\{X,Y\} - 1.$$
+But
+$$\pr\{X_1 = X_2\given X+UY\} = \pr\bigl\{ (X_1,Y_1) = (X_2, Y_2) \given X+UY\bigr\},$$
+and by [TODO: add lemma], we have
+$$\eqalign{
+\pr\{ (X_1,Y_1) = (X_2, Y_2) \given X+UY\} &\ge 2^{-\Etaseven\{X, Y \given X+UY\}} \cr
+&= 2^{-\Etaseven\{X, Y\}},\cr
+}$$
+which immediately yields the desired inequality.\slug
+
 \section References

### 2025-03-14 misc/expo/entropy.tex
Ben created

@@
 
-\proof The hypotheses give the three inequalities
-$$\Eta\{X\} \le \Eta\{Z,W\}, \quad \Eta\{Y\}\le \Eta\{Z\},\quad\hbox{and}\quad \Eta\{Y\}\le \Eta\{W\}.$$
-From this we see that
-$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z,W\} + \Eta\{Z\} + \Eta\{W\}.$$
-But then since conditioning does not increase entropy, we have
-$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z\} + 2\Eta\{W\},$$
-whence dividing both sides by $2$ completes the proof.\slug
-
-The submodularity inequality is often stated in terms of a triple of random variables in terms of
-the {\it conditional mutual information}, which is defined by
-$$\eqalign{
-\II\{ X : Y \given Z\} &= \sum_{z\in C} \pr\{Z = z\} \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
-&= \Eta\{X\given Z\} - \Eta\{X\given Y, Z\} \cr
-&= \Eta\{Y\given Z\} - \Eta\{Y\given X, Z\}.\cr
-}$$
+\proof Since $(Z,W)$ determines $X$, for each possible value $y$ that $Y$ takes,
+$(Z,W\given Y= y)$ determines $(X\given Y=y)$. Summing over the support of $Y$ and
+applying Corollary~{\cordetermines} gives us
+$$ \Eta\{X\given Y\} \le \Eta\{Z,W \given Y\} \le \Eta\{Z\given Y\} + \Eta\{W\given Y\}.$$
+Expanding the conditional entropies and rearranging yields
+$$\Eta\{X,Y\} + \Eta\{Y\} \le \Eta\{Z,Y\} + \Eta\{W,Y\}.$$
+But $Z$ and $W$ each determine $Y$, so $\Eta\{Z,Y\} = \Eta\{Z\}$ and
+$\Eta\{W,Y\} = \Eta\{W\}$. Hence
+$$\Eta\{X\} + \Eta\{Y\} \le \Eta\{X,Y\} + \Eta\{Y\} \le \Eta\{Z\} + \Eta\{W\},$$
+as desired.\slug
 
@@
 
-By submodularity, we have $\II\bigl\{ (X-Y : Z) \given X-Z\bigr\} \ge 0$, so
-\edef\eqfirstsubmodularity{\the\eqcount}
-$$\eqalign{
-0 &\le \II\bigl\{ (X-Y : Z) \given X-Z\bigr\} \cr
-&\le \Eta\{X-Y \given X-Z\} + \Eta\{ Z\given X-Z\} - \Eta\{ X-Y, Z \given X-Z\} \cr
-&\le \Eta\{X-Y, X-Z\} + \Eta\{ Z, X-Z\} - \Eta\{ X-Y, Z, X-Z\} - \Eta\{X-Z\}.\cr
-}\adveq$$
-Now, since the values $(x-y,x-z)$ taken by $(X-Y,X-Z)$ are in bijection with values
-$(x-z,y-z)$ taken by $(X-Z, Y-Z)$ via the map $(v,w) \mapsto (w, w-v)$,
-by the invariance axiom we have
-$$\Eta\{X-Y, X-Z\} = \Eta\{X-Z, Y-Z\} \le \Eta\{X-Z\} + \Eta\{Y-Z\}.$$
-Similar invocations of the invariance axiom give
-$$\Eta\{Z, X-Z\} = \Eta\{X,Z\}$$
-and
-$$\Eta\{X-Y, Z, X-Z\} = \Eta\{X,Y,Z\} = \Eta\{X,Z\} + \Eta\{Y\},$$
-where in the latter statement the second equality follows from the fact that $(X,Y)$ and $Z$ are
-independent.
-Substituting these three inequalities into~\refeq{\eqfirstsubmodularity}, we have
-$$0\le \Eta\{X-Y\} + \Eta\{Y-Z\} +\Eta\{X,Z\} -\Eta\{X,Z\} + \Eta\{Y\} - \Eta\{X-Z\},$$
-whence
-$$ \Eta\{X-Z\} \le \Eta\{X-Y\} + \Eta\{Y-Z\} + \Eta\{Y\},$$
-which completes the proof.\slug
+By submodularity, we have
+$$\Eta\{X,Y,Z\} + \Eta\{X-Z\} \le \Eta\{X-Y, Y-Z\} + H\{Z, X-Z\}.$$
+But $\Eta\{Z, X-Z\} = \Eta\{X,Z\}$, and
+$\Eta\{X,Y,Z\} = \Eta\{X,Z\} + \Eta\{Y\}$ by independence, so
+$$ \Eta\{X-Z\} + \Eta\{Y\} \le \Eta\{X-Y\} + \Eta\{Y-Z\}$$
+and the result follows.\slug
 
@@
 
-We prove the following lemma.
-
-\proclaim Lemma \advthm.
-Let $X$ and $Y$ be random variables taking values in $\FF_p^*$ for some prime $p$.
-There exists $u\in \FF_p^*$ such that
-$$\Eta\{X,uY\} \ge {\Eta\{X,Y\}\over 2^{\Etaseven\{X,Y\}}} - 1,$$
-
-\proof Let $U$ be a random variable uniform on $\FF_p^*$ and independent of both $X$ and $Y$.
-Independence gives us the identity $\Eta\{X+UY\} = \Eta\{U\} = \log(p-1)$.
-We let $(X_1, Y_1)$, $(X_2,Y_2)$, and $U$ be conditionally independent trials
-of $(X,Y)$, $(X,Y)$, and $U$ respectively, relative to $X+UY$.
-Put
-$$H = \Eta\{X_1, Y_1, X_2, Y_2, U\}.$$
-Let $W = 1$ if $X_1 = X_2$ and $W=0$ otherwise. Then since $X_1 + UY_1 = X+UY$, we have
-$$\eqalign{
-H &= \Eta\{X_1, Y_1, X_2, Y_2, W, X+UY\} \cr
-&= \Eta\{U\} + \Eta\{W\} + \Eta\{X_1, Y_1, X_2, Y_2 \given W, X+UY\}.\cr
-}$$
-If $X_1 = X_2$, then $Y_1 = Y_2$, so by conditional independence,
-$$\eqalign{
-H &= \Eta\{U\} + \Eta\{W\} +  \pr\{X_1 = X_2\given X+UY\} \Eta\{X_1, Y_1 \given X+UY\} \cr
-&\qquad+ \pr\{X_1 \ne X_2 \given X+UY\} \Eta\{X_1,Y_1, X_2,Y_2\given X+UY\}.\cr
-&= \Eta\{U\} + \Eta\{W\} + \pr\{X_1 = X_2\}\bigl( \Eta\{X,Y\given X+UY\}\bigr) \cr
-&\qquad
-+\pr\{X_1 \ne X_2\given X+UY\}\bigl( 2\Eta\{X,Y\given X+UY\} \bigr) \cr
-&= \Eta\{U\} + \Eta\{W\} + \Eta\{X,Y\given X+UY\}\cr
-&\qquad+\pr\{X_1 \ne X_2\given X+UY\}\Eta\{X,Y \given X+UY\} \cr
-}$$
-But
-$$\Eta\{X,Y\given X+UY\} = \Eta\{X,Y, X+UY\} - \Eta\{X+UY\} = \Eta\{X,Y\}$$
-by independence of $(X,Y)$ and $U$, so
-$$H =\bigl(1+\pr\{X_1\ne X_2\given X+UY\}\}\bigr) \Eta\{X,Y\} + \Eta\{U\} + \Eta\{W\}.$$
-
-On the other hand, by invertibility of each $u\in \FF_p^*$, we have
-$$\eqalign{
-H - \Eta\{U\} &= \Eta\{X_1, Y_1, X_2, Y_2\given U\} \cr
-&= {1\over p-1} \sum_{u\in \FF_p^*} \Eta\{X_1, Y_1, X_2, Y_2\given U=u\} \cr
-&= {1\over p-1} \sum_{u\in \FF_p^*} \Eta\{X_1, uY_1, X_2, uY_2\} \cr
-&= {1\over p-1} \sum_{u\in \FF_p^*} \AA\{X, uY\}.\cr
-}$$
-Consequently there exists $u\in \FF_p^*$ such that
-$$\eqalign{
-\AA\{X,uY\} &\le H-\Eta\{U\} \cr
-&= \bigl(1+\pr\{X_1\ne X_2\given X+UY\}\bigr) \Eta\{X,Y\} + \Eta\{W\} \cr
-&\le \bigl(1+\pr\{X_1\ne X_2\given X+UY\}\bigr) \Eta\{X,Y\} + 1, \cr
-}$$
-and for this choice of $u$ we have
-$$\Eta\{X+uY\} \ge \pr\{X_1 = X_2\given X+UY\} \Eta\{X,Y\} - 1.$$
-But
-$$\pr\{X_1 = X_2\given X+UY\} = \pr\bigl\{ (X_1,Y_1) = (X_2, Y_2) \given X+UY\bigr\},$$
-and by [TODO: add lemma], we have
-$$\eqalign{
-\pr\{ (X_1,Y_1) = (X_2, Y_2) \given X+UY\} &\ge 2^{-\Etaseven\{X, Y \given X+UY\}} \cr
-&= 2^{-\Etaseven\{X, Y\}},\cr
-}$$
-which immediately yields the desired inequality.\slug
+In this section, we prove an entropy variant of the Katz--Tao lemma~\ref{katztao}, which states
+that if $R$ is a commutative ring, $R^*$ the set of elements in $R$ that are not zero divisors,
+and $A\subseteq R^*$ has $|A+A|\le K|A|$ and $|A\cdot A|\le K|A|$, then there is a large subset
+$A'$ of $A$ such that $|A'\cdot A' - A'\cdot A'| \le CK^C|A'|$, where $C$ is an absolute constant.
+
+Crucial to the argumentation in this section is
+the cancellative property of elements that are not zero divisors (if $xy = xz$ then $x(y-z) = 0$,
+so for all $x\in R$ that is not a zero divisor, the function $y\mapsto xy$ is injective).
+We require a special case of a rather technical lemma due to A.~M\'ath\'e and W.~L.~O'Regan.
+We furnish a slightly simpler proof.
+
+\proclaim Lemma \advthm. Let $X$, $Y$, $Z$, and $W$ be random variables taking values in $R^*$ for some
+commutative ring $R$. Then
+$$\eqalign{
+\Eta\{XY + Z&W\} + \Eta\{X\} + \Eta\{Y\} + 2\Eta\{Z\} + 2\Eta\{W\} \cr
+&\le \Eta\{X+Y\} + \Eta\{Z-W\} + \Eta\{XZ\} + 2\Eta\{YZ\} + 2\Eta\{ZW\}.\cr
+}$$
+
+\proof Let $S = \bigl( XY, ZW, (X+Z)Y, Z(Y-W)\bigr)$ for short. The tuples $(S,Y)$ and
+$(X,Y,Z,W)$ determine each other, so $\Eta\{S,Y\} = \Eta\{X,Y,Z,W\}$. By submodularity,
+$$\Eta\{S,Y\} + \Eta\{XY\} \le \Eta\{S\} + \Eta\{X,Y\},$$
+so
+\edef\eqfirstsub{\the\eqcount}
+$$\eqalign{
+\Eta\{S\} &\ge \Eta\{X,Y,Z,W\} + \Eta\{XY\} - \Eta\{X,Y\} \cr
+&= \Eta\{Z,W\} + \Eta\{XY\}, \cr
+}\adveq$$
+by independence.
+Another application of submodularity yields
+$$\Eta\{XY+ZW\} + \Eta\{S\} \le \Eta\{XY, ZW\} + \Eta\bigl\{ (X+Z)Y, Z(Y-W)\bigr\},$$
+since $(X+Z)Y + Z(Y-W) = XY+ZW$. Substituting in~\refeq{\eqfirstsub}, we get
+\edef\eqsecondsub{\the\eqcount}
+$$\Eta\{XY+ZW\} + \Eta\{Z,W\} \le \Eta\{ZW\} + \Eta\bigl\{ (X+Z)Y\bigr\} + \Eta\bigl\{Z(Y-W)\bigr\},\adveq$$
+where we have cancelled a $\Eta\{XY\}$ term from both sides.
+Lastly, note that
+$$\eqalign{
+\Eta\bigl\{ (X+Z)Y\bigr\} &\le \Eta\{XZ, YZ\} + \Eta\{X+Z,Y\} - \Eta\{X,Y,Z\} \cr
+&\le \Eta\{XZ\} + \Eta\{YZ\} + \Eta\{X+Z\} - \Eta\{X\} - \Eta\{Z\} \cr
+}$$
+by submodularity, and likewise
+$$\eqalign{
+\Eta\bigl\{ Z(Y-W)\bigr\} &\le \Eta\{YZ, ZW\} + \Eta\{Z,Y-W\} - \Eta\{Y,Z,W\} \cr
+&\le \Eta\{YZ\} + \Eta\{ZW\} + \Eta\{Y-W\} - \Eta\{Y\} - \Eta\{W\}.\cr
+}$$
+Substituting both of these inequalities into~\refeq{\eqsecondsub}, gives exactly
+the desired inequality.\slug
+
+\parenproclaim Lemma {\advthm} (Entropic Katz--Tao lemma).
+Let $X$ be a random variable taking values in $F^*$ for some field $F$, and let
+$X_1$, $X_2$, $X_3$, and $X_4$ be independent copies of $X$. If
+$$\max\bigl(\Eta\{X_1+X_2\}, \Eta\{X_1X_2\}\bigr) \le \Eta\{X\} + \log K .$$
+for some constant $K$, then
+$$\Eta\{X_1 X_2 + X_3X_4\} \le \Eta\{X\} + 8\log K.\adveq$$
+
+\proof The previous lemma with $X\gets X_1$, $Y\gets X_2$, $Z\gets X_3$ and $W\gets X_4$
+gives
+$$\Eta\{X_1X_2+X_3X_4\} \le \Eta\{X_1+X_2\} + \Eta\{X_1-X_2\} + 5\Eta\{X_1X_2\} - 6\Eta\{X\},$$
+after applying the fact that the $X_i$ are identically distributed.
+By the Ruzsa triangle inequality, we have
+$$\Eta\{X_1 - X_2\} \le \Eta\{X_1 + X_3\} + \Eta\{X_2 + X_3\} - \Eta\{X_3\},$$
+so
+$$\Eta\{X_1X_2+X_3X_4\} \le 3\Eta\{X_1+X_2\} + 5\Eta\{X_1X_2\} - 7\Eta\{X\}.$$
+Finally, we use the hypothesis
+$\max\bigl(\Eta\{X_1+X_2\}, \Eta\{X_1X_2\}\bigr) \le \Eta\{X\} + \log K$ to arrive
+at the required bound.\slug
 

### 2025-04-04 about/vita.tex
uncertainty acceptance

@@
 \begingroup\frenchspacing
-\papercount=6
+\papercount=7
 \pubbegin P
+\papitem An uncertainty principle for M\"obius inversion on posets.
+To appear in {\sl Contributions to Discrete Mathematics}.
+[arXiv:2302.02466]
 \papitem (with Jonah Saks)
@@
 {\sl Best Canadian Stories 2025} (Windsor, Ont.: Biblioasis, 2024), 244--253.
+Reprinted in {\sl Infusion} (Toronto, Ont.: Dark Helix Press, 2025), 37--47.
 \ficitem ``Mountain Pass,'' {\sl Prairie Journal of Canadian Literature} {\bf 79} (2023), 48--55.
@@
 \sectheader{Work and Volunteer Experience {\rm($*$ indicates a paid position)}}
-
-\starleftright{\bf Course lecturer}{\bf August 2024 -- present}
-\leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
-\smallskip
-Instructor for a section of MATH 240 Discrete Structures. Duties include preparing and delivering lectures,
-designing assignment and exam questions, and holding office hours.
-\medbreak
 
@@
 co-editors related to marketing, emails, and sales.
+\medbreak
+
+\starleftright{\bf Course lecturer}{\bf August -- December 2024}
+\leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
+\smallskip
+Instructor for a section of MATH 240 Discrete Structures. Duties included preparing and delivering lectures,
+designing assignment and exam questions, and holding office hours.
 \medbreak

### 2025-04-04 misc/expo/entropy.tex
uncertainty acceptance

@@
 
-On the other hand, say $|A| = m$ and $|B| = n$ with $m<n$. If $m=1$, then by the previous proposition
-we have $\Eta\{X\} = 0$, and by normalisation and invariance, $\Eta\{Y\} = 1$.
-When $m\ge 2$, pick $k$ such that $m^k \le n^{k-1}$, so that $|A^k| \le |B^{k-1}|$.
-Then by Corollary~{\corxn} and the inequality we showed in the first paragraph of this proof, we have
-$$n\Eta\{X\} = \Eta\{X^n\} \le \Eta\{X^{n-1}\} = (n-1)\Eta\{Y\},$$
+On the other hand, say $|A| = m$ and $|B| = n$ with $m<n$. We want to conclude the
+strict inequality $\Eta\{X\} < \Eta\{Y\}$.
+Pick $k$ such that $m^k \le n^{k-1}$, so that $A^k$ can be embedded (as a set) into $B^{k-1}$.
+By invariance we can view $X^k$ as a random variable that is uniform on a subset of $B^{k-1}$.
+so by the inequality we showed in the first paragraph of this proof, as well as Corollary~{\corxn},
+we have
+$$n\Eta\{X\} = \Eta\{X^n\} \le \Eta\{Y^{n-1}\} = (n-1)\Eta\{Y\},$$
 whence $\Eta\{X\} < \Eta\{Y\}$.\slug
+
+The argument in the last paragraph of the previous proof is an instance of the ``tensor power trick,''
+in which one amplifies a known inequality by applying it to suitably high powers of the objects
+involved.
 
@@
 But for each $b$, the random variable $(X\given Y = b)$ takes values in $A$, so its entropy is
-bounded above by $\Eta\{X\}$ by the maximisation axiom. Hence $\Eta\{X\given Y\} \le \Eta\{X\}$.
+bounded above by $\Eta\{X\}$ by the maximisation axiom. Hence $\Eta\{X\given Y\} \le \Eta\{X\}$,
+which is equivalent to $\Eta\{Y\given X\} \le \Eta\{Y\}$.
 

### 2025-06-19 misc/expo/additive.tex
Ben created
update

@@
 \edef\lempetridis{\the\thmcount}
-\parenproclaim Lemma {\advthm} (Petridis, {\rm 2012}). Let $A$ and $B$ be subsets of an abelian
+\parenproclaim Lemma {\advthm} (Petridis, {\rm 2012}). Let $A_0$ and $B$ be subsets of an abelian
 group and let $K_0$ be such that $|A_0+B| = K_0|A_0|$. Then there exist $A\subseteq A_0$ and
@@
 \bigl| A + B + (C\cup \{x\}) \bigr| &\le |A+B+C| + |A+B+x| - |A'+B+x| \cr
-&\le K|A+B| + K|A| = K|A'| \cr
+&\le K|A+B| + K|A| - K|A'| \cr
 &= K\bigl| A+(C\cup \{x\})\bigr|,
@@
 
+\edef\corpetridis{\the\thmcount}
 \proclaim Corollary {\advthm}. If $|A_0 + B|\le K_0|A_0|$, then there is $A\subseteq A_0$ nonempty
@@
 $$x_1 = v(x_1) - w(x_1) = v(x_2) - w(x_2) = x_2,$$
-and this also implies that $u_1 = u_2$. So $\phi$ gives an injection from $U\times (V-1)$
+and this also implies that $u_1 = u_2$. So $\phi$ gives an injection from $U\times (V-W)$
 to $(U-V)\times (U-W)$.\slug
 
-If we define $d(U,V) = |U-V| / \bigl(|U|^{1/2}|V|^{1/2}\bigr)$, then this lemma states
+If we define $d(U,V) = |U-V| / \bigl(|U|^{1/2}|V|^{1/2}\bigr)$, then this lemma states:q
 that $d(U,W) \le d(V,U)d(U,W)$. Taking logarithms gives a genuine additive triangle inequality,
@@
 and dividing both sides by $|A|$ proves the theorem.\slug
+
+\medskip\boldlabel Sumsets with distinct summands.
+The Pl\"unnecke--Ruzsa inequality can be generalised from iterated sumsets of one fixed set $B$
+to apply to sumsets $B_1 + \ldots + B_m$ of different sets. We start with a lemma.
+
+\proclaim Lemma \advthm. Let $G$ be an abelian group and let $A, B_1,\ldots, B_m$
+be subsets of $G$ such that for all $1\le i\le m$, $|A+B_i|\le K_i|A|$ for some integer constants
+$K_1,\ldots, K_m$. Then there is a subset $A'\subseteq A$ such that
+$$|A' + B_1 + \cdots + B_m| \le m^m K_1\cdots K_m |A'|.$$
+
+\proof
+Let $e_1,\ldots,e_m$ denote
+the standard basis of $\ZZ^m$, set $N_i = K_1\cdots K_m / K_i$ for all $1\le i \le m$,
+and let $B_i' = B_i \times \{e_i, 2e_i, \ldots, N_ie_i\} \subseteq G\times \ZZ^m$ for all
+$1\le i\le m$. By abuse of notation, consider $A = A\times \{0\}$ to be a subset of $G\times \ZZ^m$
+as well, and let $B = \bigcup_{i=1}^m B_i'$. Then
+$$|A+B| \le \sum_{i=1}^m |A+B_i'| = \sum_{i=1}^m N_i|A+B_i| \le |A|\sum_{i=1}^m N_i K_i.$$
+By Corollary~{\corpetridis} there exists $A'\subseteq A$ such that
+$$ |A'+ m B| \le \biggl( \sum_{i=1}^m N_i K_i \biggr)^m |A'|,$$
+so we have
+$$\eqalign{
+(K_1\cdots K_m)^{m-1} |A' + B_1 + \cdots + B_m| &= N_1\cdots N_k |A' + B_1 + \cdots + B_m| \cr
+&= |A'+B_1' + \cdots + B_m'| \cr
+&\le |A' + mB| \cr
+&\le \biggl( \sum_{i=1}^m N_i K_i\biggr)^m |A'| \cr
+&\le (m K_1 \cdots K_m )^m |A'|. \cr
+}$$
+Cancelling out a factor of $(K_1\cdots K_m)^{m-1}$ from both sides proves the result.\slug
+
+If we just want to bound the size of $|B_1+\cdots + B_m|$, we can remove the factor
+of $m^m$ on the right-hand side by employing a variant of the ``tensor power trick.''
+
+\proclaim Theorem \advthm. Let $A, B_1, \ldots, B_m$ be subsets of an abelian group $G$ such that
+$|A+B_i| \le K_i|A|$ for some integer constants $K_1,\ldots,K_m$. Then
+$$|B_1 + \cdots + B_m| \le K_1 \cdots K_m |A|.$$
+
+\proof Let $d$ be a positive integer. We have
+$$|A^d + B_i^d| = |A+B_i|^d \le \bigl(K_i |A|\bigr)^d = K_i^d |A^d|$$
+for all $1\le i\le m$, so by the previous lemma, there is a subset $A'$ of $A^d$ such that
+$$|A' + B_1^d + \cdots + B_m^d| \le m^m (K_1\cdots K_m)^d |A'|
+\le m^m (K_1\cdots K_m)^d |A|^d.$$
+But this implies that
+$$|B_1+\cdots + B_m|^d \le |A' + B_1^d + \cdots + B_m^d| \le m^m (K_1\cdots K_m)^d |A|^d.$$
+Taking $d$th roots yields
+$$|B_1 + \cdots +B_m| \le m^{m/d} K_1 \cdots K_m |A|,$$
+and this holds for all $d$, so by taking $d$ to infinity we can make $m^{m/d}$ as close to $1$ as
+we like.\slug
 

### 2025-06-19 misc/expo/entropy.tex
update

@@
 $$\Eta\{X^n\} = \Eta\{Y\} + \Eta\{X^n\given Y\} \ge \Eta\{Y\}.$$
-But $\Eta\{Y\} > 0$ for $\eps$ small enough, the normalisation and continuity axioms. So
-$\Eta\{X\} \ge \Eta\{Y\}/n > 0$ as well.\slug
+But $\Eta\{Y\}$ can be taken as close to $1$ as we like by choosing $\eps$ small enough,
+by the normalisation and continuity axioms. So
+$\Eta\{X\} \ge \Eta\{Y\}/n > 0$.\slug
 

### 2025-07-16 about/vita.tex
new accepted paper
updated cv

@@
 \smallskip
+\leftright{\bf Constance Rooke CNF Prize
+\rm (\$1,250)}{\bf October 2024}
+\smallskip
 \leftright{\bf FRQNT Bourse de doctorat en recherche
@@
 \begingroup\frenchspacing
-\papercount=7
+\papercount=8
 \pubbegin P
+\papitem (with Tsun-Ming Cheung and Luc Devroye)
+A note on plane trees with decreasing labels.
+To appear in {\sl Online Journal of Analytic Combinatorics}.
+[arXiv:2502.14596]
 \papitem An uncertainty principle for M\"obius inversion on posets.
@@
 
+\sectheader{Submitted}
+
+\begingroup\frenchspacing\parindent=10pt
+\thing (with Hamed Hatami)
+Block structure in boolean matrices of small factorization norm.
+{\sl arXiv:2507.00872} (2025), 14~pp.
+\smallskip
+\thing (with Hamed Hatami)
+Block complexity and idempotent Schur multipliers.
+{\sl arXiv:2506.21752} (2025), 17~pp.
+\smallskip
+\thing
+On an entropic analogue of additive energy.
+{\sl arXiv:2406.18798} (2025), 19~pp.
+\endgroup
+
 \goodbreak
@@
 \starficitem ``Lanterns,'' {\sl The Malahat Review} {\bf 229} (2025), 7--13.
-Winner of the 2024 Constance Rooke Creative Nonfiction Prize.
+Winner of the 2024 Constance Rooke Creative Nonfiction Prize. Honourable Mention in the Personal Journalism
+category of Canada's 2025 National Magazine Awards.
 \ficitem ``Senang Diri,'' {\sl Existere} {\bf 42},2 (2023), 17--25.

### 2025-07-25 misc/expo/idempotent.tex
More typos in blocky schur post

\input fontmac
\input mathmac

\def\eps{\epsilon}
\def\FF{\hbox{\bf F}}
\def\bar#1{\overline{#1}}
\def\hat#1{\widehat{#1}}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\argmax{\limitop{\rm arg$\,$max}}
\def\dTV{d_{\rm TV}}

\def\advthm{\the\sectcount.\the\thmcount\global\advance \thmcount by 1}
\def\advsect{\global\advance\sectcount by 1\section\the\sectcount\global\thmcount=1. }
\sectcount=0

\widemargins
\bookheader{IDEMPOTENT THEOREMS AND MATRICES}{MARCEL K. GOH}

\maketitle{Idempotent theorems and matrices}{}{Marcel K. Goh}{\sl Department of Mathematics and Statistics,
McGill University}

\floattext4.5 \ninebf Abstract.
\ninepoint These expository notes move at a leisurely pace though background material needed to
appreciate the Cohen idempotent theorem and its quantitative version, due to Green and Sanders.
We introduce Schur multipliers and investigate a conjectured analogue in this setting.

\bigskip

\advsect The structure of locally compact abelian groups

{\sc A topological abelian group} is a Hausdorff space that is also an abelian group, such that
the map from $G\times G$ to $G$ sending $(x,y) \mapsto x-y$ is continuous (where the topology
on $G\times G$ is the product topology).
If, in addition, every point in $G$ is contained in some compact neighbourhood of $G$, then
$G$ is said to be a {\it locally compact abelian} group.

The following proposition shows that by passing to closures, we can always assume subgroups to be closed.

\proclaim Proposition \advthm. Let $G$ be a locally compact abelian group and $H$ a subgroup of $G$.
Then the closure $\bar H$ of $H$ is also a subgroup of $G$.

\proof Let $x,y\in \bar H$ and suppose, for a contradiction, that $xy\notin \bar H$. :

A topological space is called {\it discrete} if for every point $p$ in it, the singleton set $\{p\}$
is open.

Let $G$ be a locally compact abelian group.
The set of all measures $\mu$ on $G$ forms a topological semigroup under the convolution operation
$$\mu * \nu = \int_G

\section References

\bye

### 2025-09-13 about/vita.tex
Started 242 tutorials

@@
 
-\leftright{\bf Co-Editor}{\bf December 2023 -- present}
+\leftright{\bf Editor-in-Chief}{\bf December 2023 -- present}
 \leftright{\sl Ahoy Literary Magazine}{\sl Montr\'eal, QC, Canada}
 \smallskip
-Founding co-editor of a semiannual print magazine that publishes poetry, prose, and art.
+Founding member and one of two editors-in-chief of
+a semiannual print magazine that publishes poetry, prose, and art.
 In charge of English short fiction submissions and the overall typesetting. Share duties with the other
-co-editors related to marketing, emails, and sales.
+editors related to emails, sales, and magazine launches.
+\medbreak
+
+\starleftright{\bf Teaching assistant}{\bf September 2021 -- present}
+\leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
+\smallskip
+Teaching assistant in the following courses. Duties include giving tutorials, holding office hours, and
+grading assignments and exams.
+\begingroup\parindent=10pt
+\smallskip
+\thing Fall 2025: MATH 242 Analysis 1
+\smallskip
+\thing Winter 2024, Fall 2023: MATH 240 Discrete Structures
+\smallskip
+\thing Fall 2021: COMP 690 Probabilistic Analysis of Algorithms
+\endgroup
 \medbreak
@@
 designing assignment and exam questions, and holding office hours.
-\medbreak
-
-\starleftright{\bf Teaching assistant}{\bf September 2021 -- April 2024}
-\leftright{\sl McGill University}{\sl Montr\'eal, QC, Canada}
-\smallskip
-Teaching assistant in the following courses. Duties include giving tutorials, holding office hours, and
-grading assignments and exams.
-\begingroup\parindent=10pt
-\smallskip
-\thing Winter 2024, Fall 2023: MATH 240 Discrete Structures
-\smallskip
-\thing Fall 2021: COMP 690 Probabilistic Analysis of Algorithms
-\endgroup
 \medbreak
@@
 \smallskip
-Led children aged 4--15 through various activities (e.g. archery, canoeing, zipline) at a sleepaway camp.
+Led children aged 4--15 through various activities (e.g., archery, canoeing, zipline) at a sleepaway camp.
 \medbreak

### 2025-09-13 misc/expo/idempotent.tex
Started 242 tutorials

@@
 \def\dTV{d_{\rm TV}}
+\def\d{\;d}
 
@@
 
+\advsect Topological preliminaries
+
+A {\it topology} on a set $S$ is a collection of subsets of $S$ that contains the empty set and
+is closed under the formation of unions and finite intersections. If $S$ has a topology defined on it,
+then $S$ is said to be a {\it topological space}, and the members of its topology are called
+{\it open sets}. The complements of open sets are called {\it closed sets}.
+
+\edef\propopencriterion{\the\thmcount}
+\proclaim Proposition \advthm. A set $A$ is open if and only if for every $x\in A$, there is an open
+set $U_x$ with $x\in U_x$ and $U_x\subseteq A$.
+
+\proof If $A$ is open, then for any $x\in A$ we can simply take $U = A$. Conversely, if for every
+point $x\in A$ there is some open $U_x\subseteq A$ with $x\in U_x$, then $A = \bigcup_{x\in A} U_x$,
+so $A$ is open, being a union of open sets.\slug
+
+The largest
+open set contained in a set $A$ is the {\it interior} of $A$, and the smallest closed set containing
+$A$ is its {\it closure}, denoted $\bar A$. If $A\bar A$ is the whole set $S$, then we say $A$ is
+{\it dense} in $S$, and if some countable set is dense in $S$, then $S$ is a {\it separable} space.
+
+A set $A$ is said to be a {\it neighbourhood} of a point $p$ if $p$ belongs to the interior of $A$.
+The space $S$ is called {\it Hausdorff} if for all points $x,y\in S$ with $x\ne y$, there is some
+open neighbourhood $U$ of $X$ and open neighbourhood $V$ of $y$ such that $U\cap V=\emptyset$.
+A set $A$ is {\it compact} if every open cover of $A$ has a finite subcover. If every point of $S$
+has some compact neighbourhood, then $S$ is called {\it locally compact}.
+
+If the singleton set $\{p\}$ is open, then $p$ is called an {\it isolated point} of $S$, and if
+$p$ is isolated for all points $p\in S$, then $S$ is a {\it discrete space}.
+
+\medskip\boldlabel Limit points of nets.
+A relation $\preceq$ on a set $I$ is a {\it preorder} if it is reflexive and transitive.
+A set $I$ is said to be {\it directed} if there exists a preorder $\preceq$ on it such that
+any two elements have an upper bound; that is,
+for all $i,j\in I$, there exists $k\in I$ such that $i\preceq k$ and $j\preceq k$.
+
+Let $I$ be a directed set and $S$ any set. A function $f:I\to S$ is called
+a {\it net}. We usually index the elements of the net by elements of $I$; writing $x_i = f(i)$ for all
+$i\in I$, the net $f$ can be expressed as $(x_i)_{i\in I}$. When $I = \NN$ under the usual ordering
+$\le$, we recover the usual definition of a {\it sequence} of points in $S$.
+We say that $(x_i)_{i\in I}$ is {\it eventually} in $A$ if there exists some $j\in I$ such that for all
+$i \succeq j$, $x_i\in A$. A point $x\in S$ is called a {\it limit point} of $(x_i)_{i\in I}$ if
+for every open neighbourhood $U$ of $x$, the net $(x_i)_{i\in I}$ is eventually in $U$. In this
+case we write $x_i\to x$.
+
+\proclaim Proposition \advthm. If $S$ is a Hausdorff space, then every net $(x_i)_{i\in I}$ valued
+in $S$ has at most one limit point.
+
+\proof Let $S$ be a topological space and suppose there is a net $(x_i)_{i\in I}$ with two distinct
+limit points $x$ and $y$. Let $U$ be an open neighbourhood of $x$ and $V$ an open neighbourhood of $Y$.
+By definition of limit point, there is some $i\in I$ such that $x_k \in U$ for all $k\succeq i$,
+and similarly there is some $j\in I$ such that $x_i\in V$ for all $k\succeq j$. But since $I$
+is a directed set, there must exist some $k\in I$ with $k\succeq i$ and $k\succeq j$, and we find
+that $x_k\in U\cap V$. Since $U$ and $V$ were arbitrary, we conclude that $S$ is not Hausdorff.\slug
+
+In light of the proposition, so long as the ambient space is Hausdorff it makes sense to write
+$\lim_{i\in I} x_i = x$ whenever the limit point of $(x_i)_{i\in I}$ is $x$. We may also choose
+to write $x_i \to x$.
+
+We now introduce a common choice of directed set that features in some of the proofs below.
+
+\edef\propnbhdnet{\the\thmcount}
+\proclaim Proposition \advthm.
+Let $x$ be a point in a topological space $S$ and let $N(x)$ be the set of all open neighbourhoods
+of $x$. Define an order on $N(x)$ by declaring $U\preceq V$ if and only if $U\supset V$. Then the order
+$\preceq$ is directed, and $x$ is a limit point of any
+net $(x_U)_{U\in N(x)}$ satisfying $x_U\in U$ for all $U\in N(x)$.
+
+\proof The set $N(x)$ is clearly directed, since for every $U,V\in N(x)$,
+$U \supset U\cap V$ and $V\supset U\cap V$. Now let $(x_U)_{U\in N(x)}$ be any net with
+$x_U\in U$ for all $U\in N(x)$. Pick any open set $V$ containing $x$. This set $V$ is a member of
+$N(x)$, and for any $U\in N(x)$ with $U\succeq V$, we must have $x_U\in U \subseteq V$.\slug
+
+One can characterise closed sets in terms of limit points of nets.
+
+\proclaim Proposition \advthm. A set $A$ is closed if and only if whenever a
+net with $(x_i)_{i\in I}$ satisfying $x_i\in A$ for all $i\in I$ possesses a limit point $x$,
+we must have $x\in A$.
+
+\proof Suppose that $A$ is closed and let $(x_i)_{i\in I}$ be a net with limit point $x$
+and $x_i\in A$ for all $i\in I$. For a contradiction, suppose that $x\notin A$. Then $x \in A^c$,
+which is open. By the definition of limit point, there exists $j$ such that $x_i\in A^c$ for all
+$j\succeq i$. This contradicts the assumption that $x$ is a limit point of $(x_i)_{i\in I}$.
+
+Now we assume the second condition and prove that $A^c$ is open.
+Let $N(x)$ be the set of all open neighbourhoods of $x$.
+Towards a contradiction, assume
+that there is some $x\in A^c$ such that for all $U\in N(x)$, we have $U\cap A \ne \emptyset$.
+So for all $U\in N(x)$, we may pick $x_U\in U\cap A$, and the net $(x_U)_{U\in N(x)}$
+has $x$ as its limit point, by Proposition~{\propnbhdnet}. But $x_U\in A$ for all $U\in N(x)$,
+so by our assumption, we must have $x\in A$. This contradiction shows that
+any $x\in A^c$ must have some open neighbourhood $U$ that is contained in $A^c$. Hence by
+Proposition~{\propopencriterion}, $A$ is closed.\slug
+
 \advsect The structure of locally compact abelian groups
 
-{\sc A topological abelian group} is a Hausdorff space that is also an abelian group, such that
-the map from $G\times G$ to $G$ sending $(x,y) \mapsto x-y$ is continuous (where the topology
-on $G\times G$ is the product topology).
-If, in addition, every point in $G$ is contained in some compact neighbourhood of $G$, then
+A topological abelian group is an abelian group $G$ endowed with the topology of a Hausdorff
+space, such that
+the map from $G\times G$ to $G$ sending $(x,y) \mapsto x-y$ is continuous. (The topology
+on $G\times G$ is taken to be the product topology).
+If, in addition, $G$ is also locally compact, then
 $G$ is said to be a {\it locally compact abelian} group.
@@
 
-\proof Let $x,y\in \bar H$ and suppose, for a contradiction, that $xy\notin \bar H$. :
-
-A topological space is called {\it discrete} if for every point $p$ in it, the singleton set $\{p\}$
-is open.
+\proof Let $x,y\in \bar H$ and let $Z$ be any neighbourhood of $x-y$. Since multiplication
+is continuous (and by the definition of the product topology), there is an open neighbourhood
+$X$ of $x$ and $Y$ of $y$ such that $X-Y\subseteq Z$. By the definition of closure,
+there are points $a\in X\cap H$ and $b\in Y\cap H$. The point $a-b$ is in $H$ because $H$ is a subgroup,
+and thus $a-b\in Z\cap H$.\slug
 
@@
 The set of all measures $\mu$ on $G$ forms a topological semigroup under the convolution operation
-$$\mu * \nu = \int_G
+$$\mu * \nu(A)  = \int_G \chi_A(x+y) \d \mu(x) \d\nu(x),$$
+where $\chi_A$ denotes the characteristic function of $A$.
 

### 2025-09-13 teaching/math242f25/tut1.tex
Started 242 tutorials

\input fontmac
\input mathmac

\def\To{\Rightarrow}

\maketitle{Math 242 Tutorial 1}{prepared by}{Marcel Goh}{11 September 2025}

\bigskip

\proclaim Problem 1. Let $p$ and $q$ be statements. Show that $p\To q$ is equivalent to $\neg q \To \neg p$.

\proof We have
$$\eqalign{
p\To q &\equiv \neg p \vee q \cr
&\equiv q \vee \neg p \cr
&\equiv \neg (\neg q) \vee \neg p \cr
&\equiv \neg q \To \neg p \noskipslug\cr
}$$

\proclaim Problem 2. Let $n$ be a positive integer. Show that if $n$ is not a perfect square, then $\sqrt n$
is irrational.

\proof By contrapositive. Suppose $\sqrt n$ is rational. Then there exist $a,b\in \ZZ$ (with $b\ne 0$)
such that $\sqrt n = a/b$. Without loss of generality, we can assume that $\gcd(a,b) = 1$. We see that
$n = a^2/b^2$, but since $n$ is an integer, and $a^2$ and $b^2$ have no common factors, we must have
$b^2 = 1$. This forces $n=a^2$, which shows that $n$ is a perfect square.\slug

\proclaim Problem 3. Prove the identity
$$\sum_{k=1}^n k^3 = \biggl({n(n+1)\over 2}\biggr)^2$$
for all integers $n\ge 1$.

\proof By induction on $n$. For the base case $n=1$, we have
$$\sum_{k=1}^1 k^3 = 1^3 = 1^2 = \biggl( {1\cdot 2\over 2}\biggr)^2 = 
\biggl( {1\cdot (1+1)\over 2}\biggr)^2.$$
Now suppose the statement holds for some $n$. For $n+1$ we have
$$\eqalign{
\sum_{k=1}^{n+1} k^3 &= (n+1)^3 + \sum_{k=1}^n k^3 \cr
&= (n+1)^3 + \biggl({n(n+1)\over 2}\biggr)^2 \cr
&= {4(n+1)(n+1)^2 + n^2 (n+1)^2\over 4} \cr
&= {(n^2 + 4n + 4)^2(n+1)^2 \over 4} \cr
&= \biggl({(n+1)\bigl((n+1)+1\bigr)\over 2}\biggr)^2,\cr
}$$
where in the second line, we used the induction hypothesis.\slug

% \proclaim Problem 4. Prove the Cauchy--Schwarz inequality, the statement that for all
% $u_1,\ldots,u_n, v_1,\ldots,v_n\in \RR$,
% $$\biggl(\sum_{i=1}^n u_i v_i\biggr)^2
% \le \biggl( \sum_{i=1}^n {u_i}^2\biggr) \biggl( \sum_{i=1}^n {v_i}^2\biggr).$$
% [{\it Hint:} Recall from your high school days that
% for a quadratic polynomial $ax^2 + bx + c$, the {\it discriminant} is
% defined to be the quantity $b^2 - 4ac$. (It appears under the square-root sign
% in the quadratic formula.) If it is negative, the polynomial has no real roots,
% if it is positive, the polynomial has two real roots, and if it is zero, the polynomial
% has one real root.]
% 
% \proof If $u_i = 0$ for all $1\le i\le n$, then both sides are $0$ and the inequality
% is trivially true. So suppose that at least one of the $u_i$ is nonzero. Then
% $$p(x) = (u_1 x + n_1)^2 + (u_2 x + v_2)^2 + \cdots + (u_n x+v_n)^2$$
% is a quadratic polynomial in the variable $x$. Rewriting $p(x)$ in summation notation,
% we have
% $$\eqalign{
% p(x) &= \sum_{i=1}^n (u_ix + v_i)^2\cr
% &= \sum_{i=1}^n \bigl( u_i^2 x^2 + 2u_i v_i + v_i^2\bigr)\cr
%  &= \biggl(\sum_{i=1}^n u_i^2 \biggr) x^2 +
% 2\biggl(\sum_{i=1}^n u_i v_i\biggr) x + \biggl( \sum_{i=1}^n {v_i}^2\biggr).\cr
% }$$
% Since it is defined as a sum of squares, $p(x)\ge 0$ for all $x\in \RR$, so it has
% at most one real root.
% This means its discriminant is nonpositive; in other words,
% $$4\biggl(\sum_{i=1}^n u_i v_i\biggr)^2
% - 4\biggl( \sum_{i=1}^n {u_i}^2\biggr) \biggl( \sum_{i=1}^n {v_i}^2\biggr) \le 0.$$
% Dividing through by $4$ and then rearranging gives exactly the inequality we sought.\slug

\proclaim Problem 4. Suppose that a local McDonalds branch sells Chicken McNuggets in meals of $4$
or $9$. Prove that for all $n\ge 36$, it is possible to buy exactly $n$ Chicken McNuggets.

\proof Formulated precisely in mathematical language, what we are trying to show is that for every
integer $n\ge 36$, there exist nonnegative integers $a$ and $b$ such that $n=9a + 4b$.
We perform a proof by strong induction. Note first that
$$ 36 = 9\cdot 4 + 4\cdot 0, \qquad 37 = 9\cdot 1 + 4\cdot 7, \qquad
38 = 9\cdot 2 + 4\cdot 5,$$
and
$$39 = 9\cdot 3 + 4\cdot 3.$$
Now let $n\ge 40$ and assume that the statement holds for all integers in the range $[36, n]$.
In particular, it holds for $n-4$. So there exist nonnegative integers $a$ and $b$ such that
$n-4 = 9a + 4b$, and we see that $n = 9a + 4(b+1)$.\slug

\proclaim Problem 5. Show that there exist irrational numbers $a$ and $b$ such that $a^b$ is rational.

\proof Consider the number $(\sqrt 2)^{\sqrt 2}$. It is either rational or irrational.

If it is rational, then we can set $a = b = \sqrt 2$, which we already showed to be irrational in class,
and $a^b$ is rational by the assumption in this case.

If it is irrational, then we can set $a =(\sqrt 2)^{\sqrt 2}$ and $b=\sqrt 2$. We compute
$$a^b = \bigl((\sqrt 2)^{\sqrt 2}\bigr)^{\sqrt 2} = (\sqrt 2)^{\sqrt 2 \cdot \sqrt 2}
= (\sqrt 2)^2 = 2,$$
which is rational.\slug

% \proclaim Problem 7. Let $n\ge 1$ be an integer.
% Prove that the number of subsets of $\{1,2,\ldots,n\}$ is $2^n$.
% 
% \proof By induction. The number of subsets of $\{1\}$ is $2 = 2^1$, since the two subsets are $\emptyset$
% and $\{1\}$. Now assume the statement is true for $n$, and note that every subset $A$ of $\{1,2,\ldots,n,n+1\}$
% either contains $n+1$ or it doesn't. If $A$ doesn't contain $n+1$, then $A\subseteq \{1,2,\ldots,n\}$,
% and there are $2^n$ possibilities for $A$, by the induction hypothesis. If $A$ does contain $n+1$,
% then $A = B\cup\{n+1\}$ for some $B\subseteq \{1,2,\ldots,n\}$, and again there are $2^n$ possibilities
% for $A$ in this case. We conclude that
% there are $2\cdot 2^n = 2^{n+1}$ subsets of $\{1,2,\ldots,n+1\}$ in total.\slug

\bye

### 2025-09-20 teaching/math242f25/tut2.tex
new paper and updated 242 tut2

\input fontmac
\input mathmac

\def\To{\Rightarrow}

\maketitle{Math 242 Tutorial 2}{prepared by}{Marcel Goh}{18 September 2025}

\bigskip

\proclaim Problem \advthm. Let $f:D\to E$ be a function and let $B\subseteq E$. Prove that
\medskip
\item{a)} $f\bigl( f^{-1}(B)\bigr) \subseteq B$; and that
\smallskip
\item{b)} if $f$ is surjective, then $f\bigl( f^{-1}(B)\bigr) = B$.
\medskip

\proof Let $y\in f\bigl( f^{-1}(B)\bigr)$. There is some $x\in f^{-1}(B)$ such that $f(x) = y$.
Since $x\in f^{-1}(B)$, we must have $f(x)\in B$. But we have $y = f(x)$, so in fact
$y\in B$. This settles part (a).

For part (b), assume that $f$ is surjective.
In view of part (a) we only have to prove that $B\subseteq f\bigl( f^{-1}(B)\bigr)$. Let $y\in B$;
since $f$ is surjective, there is some $x\in D$ such that $f(x) = y$. This $x$, then, is a member
of $f^{-1}(B)$ (being an element that is mapped to $y$, which we assumed to be in $B$). But then
we see that $y\in f\bigl(f^{-1}(B)\bigr)$, since $y = f(x)$.\slug

\proclaim Problem \advthm. Find the supremum and infimum of each of the following subsets of $\RR$ (and
prove that each is a supremum/infinum).
\medskip
\item{a)} $A = \bigl\{ 1/2^n : n\in \NN\bigr\}$.
\smallskip
\item{b)} $B = \bigl\{ (-1)^n + 1/n : n\in \NN \bigr\}$.
\medskip

\proof For part (a), note that if $m > n$, then $1/2^m < 1/2^n$, so the greatest element of $A$,
namely $1/2$, is also its supremum. The infimum, on the other hand, is $0$, which we now prove.
First we note that $0\le 1/2^n$ for all $n\in \NN$. Then, let any $\eps > 0$ be given. We need
to find some $n$ such that $1/2^n < \eps$. But this can be done, but the Archimedean property of
$\RR$. Let $n$ be a natural number such that $n > \log_2(1/\eps)$. Then
$${1\over 2^n} < {1\over 2^{\log_2(1/\eps)}} = {1\over 1/\eps} = \eps,$$
and we are done.

For part (b), it helps to write out some elements of $B$ for small $n\in \NN$. We have
$$-1 + 1/1 = 0,\qquad 1 + 1/2 = 3/2,\qquad -1 + 1/3 = -2/3, \qquad 1 + 1/4 = 5/4,\qquad
-1 + 1/5 = -4/5,$$
and so on. We see that each even $n$ gives an element of $B$ that is larger than $1$, but
as $n$ gets larger, these elements will get closer and closer to $1$. For odd $n$,
we get an element in the interval $(-1, 0]$, but again, as $n$ gets larger, the $1/n$ term
gets smaller so it these elements get closer and closer to $-1$. We claim, then, that $\sup B=3/2$,
and $\inf B = -1$.

First we prove the claim about the supremum. For all odd $n$,
the number $(-1)^n + 1/n$ equals $-1 + 1/n$, which is nonpositive, so $3/2 > (-1)^n + 1/n$. For all
even $n$, we have $(-1)^n + 1/n = 1+1/n$, which is at most $3/2$. So $3/2$ is an upper bound for the set,
The element $3/2$ is in the set, so no number that is smaller than $3/2$ can be an upper bound on $B$.
Now we prove the claim about the infimum. It is easy to check that $-1$ is a lower bound on $B$, since
all even $n$ give a positive number, and all odd $n$ give $-1+1/n \ge -1$. We just have to show
now that it is the {\it greatest} lower bound. Let $\eps > 0$. We need to find some odd $n$
such that $-1 + 1/n < -1 + \eps$. To this end, we use the Archimedean property of $\RR$ again,
picking some $n$ such that $n > 1/\eps$. Without loss of generality, we can choose $n$ odd (since
if $n$ is even, then $n+1$ is odd and is also greater than $1/\eps$). Then we see that
$$-1 + {1\over n} < -1 + {1\over 1/\eps} = -1+\eps,$$
and we are done.\slug

\proclaim Problem \advthm. Prove that for all $x\in \RR$,
\medskip
\item{a)} $|x| = |-x|$;
\smallskip
\item{b)} $|x| = \sqrt{x^2}$;
\smallskip
\item{c)} $-|x| \le x \le |x|$;
\medskip
\noindent and for all $x,y\in \RR$,
\medskip
\item{d)} $|xy| = |x|\cdot |y|$.
\medskip

\proof If $x$ is nonnegative, then $-x$ is nonpositive, so $|x| = x = |-x|$.
If $x$ is negative, then $-x$ is positive, so $|x| = -x = |-x|$. Either way, $|x| = |-x|$,
and part (a) is done.

For any real number $a$, the square root $\sqrt a$ is defined to be the unique nonnegative real number
$b$ with $b^2 = a$. Note that $|x|^2 = x^2$; this is because if $x$ is nonnegative, then $|x| = x$
and the identity is clear, and if $x$ is negative, then $|x| = -x$ and we have $|x|^2 = (-x)^2
= (-1)^2 x^2 = x^2$. But $|x|$ is nonnegative by construction, and satisfies $|x|^2 = x^2$,
so we conclude that $|x| = \sqrt{x^2}$.

For part (c), note that if $x$ is negative, then $|x| = -x$, so $-|x|\le x < -x = |x|$, and if
$x$ is nonnegative, then $|x| = x$, so $-|x| = -x \le x = |x|$.

Equipped with the identity from part (b), we have
$$|xy| = \sqrt{(xy)^2} = \sqrt{x^2\cdot y^2} = \sqrt{x^2}\sqrt{y^2} = |x|\cdot |y|,$$
since multiplication of real numbers is commutative. This proves part (d).\slug

\proclaim Problem \advthm. Prove the {\it reverse triangle inequality}
$$ |x-y| \ge \bigl| |x|-|y|\bigr|,$$
which holds for all $x,y\in \RR$.

\proof Note first that
$$(x-y)^2 = x^2 - 2xy + y^2.$$
The terms $x^2$ and $y^2$ are both nonnegative, but the middle term, $-2xy$, could be positive or negative,
depending on the relative signs of $x$ and $y$. If we change this term to $-2|x|\cdot |y|$, then
the term must be negative, and the whole right-hand side either stays the same or goes down; in other words,
$$(x-y)^2 \ge x^2 - 2|x|\cdot |y| + y^2.$$
But by part (c) of the previous problem, we see that
$$(x-y)^2 \ge |x|^2 - 2|x|\cdot |y| + |y|^2 = \bigl( |x|- |y|\bigr)^2.$$
Taking square roots of both sides (and applying part (c) of the previous problem once again), we have
$$|x-y| \ge \bigl| |x| - |y| \bigr|,$$
which is what we wanted.\slug

Here's another proof using the ordinary triangle inequality that was proved in class.

\medskip\noindent{\it Alternative proof.}\enspace We add the ``clever'' zero $-y+y$ to $x$, obtaining
$$|x| = \bigl| (x-y) + y\bigr| \le |x-y| + |y|$$
from the triangle inequality. This rearranges to $|x-y| \ge |x|-|y|$. Likewise, we have the similar bound
$$|y| = \bigl| (y-x) + x\bigr| \le |y-x| + |x| = |x-y| + |x|,$$
where after the triangle inequality we used the identity from part (a) of the previous question.
This rearranges to $|x-y| \ge |y|-|x|$, so we have shown that $|x-y| \ge \bigl| |x|-|y|\bigr|$.\slug

\proclaim Problem \advthm. Let $a,b\in \RR$. Show that $a=b$ if and only if for all $\eps>0$, we have
$|a-b|\le \eps$.

\proof The ``only if'' direction is easy. Assume that $a=b$ and let $\eps > 0$ be arbitrary. We have
$$|a-b| = |a-a| = 0 \le \eps.$$
Now we show the ``if'' direction, by contrapositive. (We must show that $a\ne b$ implies
the existence of some $\eps > 0$ such that $|a-b| > \eps$.) Suppose that $a\ne b$, so that $a-b\ne 0$.
This means that $|a-b| > 0$, so we can set $\eps = |a-b| / 2$, which is also positive. Then we have
$$|a-b| > {|a-b|\over 2} = \eps,$$
which settles the proof.\slug

\bye

### 2025-09-30 misc/expo/idempotent.tex
posted tutorial 3

@@
 
-\maketitle{Idempotent theorems and matrices}{}{Marcel K. Goh}{\sl Department of Mathematics and Statistics,
-McGill University}
+\maketitle{Idempotent theorems and matrices}{}{Marcel K. Goh}{23 September 2025}
 
@@
 
-\edef\propopencriterion{\the\thmcount}
+\edef\propopencriterion{\the\sectcount.\the\thmcount}
 \proclaim Proposition \advthm. A set $A$ is open if and only if for every $x\in A$, there is an open
@@
 open neighbourhood $U$ of $X$ and open neighbourhood $V$ of $y$ such that $U\cap V=\emptyset$.
-A set $A$ is {\it compact} if every open cover of $A$ has a finite subcover. If every point of $S$
-has some compact neighbourhood, then $S$ is called {\it locally compact}.
 
 If the singleton set $\{p\}$ is open, then $p$ is called an {\it isolated point} of $S$, and if
-$p$ is isolated for all points $p\in S$, then $S$ is a {\it discrete space}.
+$p$ is isolated for all points $p\in S$, then $S$ is a {\it discrete space}. It is easy to see
+that every subset of a discrete space is open.
 
@@
 
-\edef\propnbhdnet{\the\thmcount}
+\edef\propnbhdnet{\the\sectcount.\the\thmcount}
 \proclaim Proposition \advthm.
 Let $x$ be a point in a topological space $S$ and let $N(x)$ be the set of all open neighbourhoods
-of $x$. Define an order on $N(x)$ by declaring $U\preceq V$ if and only if $U\supset V$. Then the order
+of $x$. Define an order on $N(x)$ by declaring $U\preceq V$ if and only if $U\supseteq V$. Then the order
 $\preceq$ is directed, and $x$ is a limit point of any
@@
 \proof The set $N(x)$ is clearly directed, since for every $U,V\in N(x)$,
-$U \supset U\cap V$ and $V\supset U\cap V$. Now let $(x_U)_{U\in N(x)}$ be any net with
+$U \supseteq U\cap V$ and $V\supseteq U\cap V$. Now let $(x_U)_{U\in N(x)}$ be any net with
 $x_U\in U$ for all $U\in N(x)$. Pick any open set $V$ containing $x$. This set $V$ is a member of
@@
 
-\proclaim Proposition \advthm. A set $A$ is closed if and only if whenever a
-net with $(x_i)_{i\in I}$ satisfying $x_i\in A$ for all $i\in I$ possesses a limit point $x$,
-we must have $x\in A$.
+\proclaim Proposition \advthm. A set $A$ is closed if and only if for every net
+$(x_i)_{i\in I}$ satisfying $x_i\in A$ for all $i\in I$, every limit point $x$ of
+$(x_i)_{i\in I}$
+must be in $A$.
 
@@
 Proposition~{\propopencriterion}, $A$ is closed.\slug
+
+\medskip\boldlabel Cluster points and subnets.
+A point $x$ is a {\it cluster point} of the net $(x_i)_{i\in I}$ if for every open neighbourhood $U$
+of $x$ and any $i\in I$, there is some $j\succeq i$ with $x_i\in U$. Next, consider two directed sets
+$I$ and $J$. A map $\phi:J\to I$ is said to be {\it cofinal} if for every $i\in I$ there is $j\in J$
+such that for all $j'\succeq j$, $\phi(j') \succeq i$. Now if $(x_i)_{i\in I}$ is a net (with elements
+in some topological space $X$) and $\phi:B\to A$ is
+cofinal, then the composition $x\circ \phi : B\to X$ is a net; we say that $x\circ \phi$ is
+a {\it subnet} of the original net.
+The following proposition relates subnets to cluster points.
+
+\proclaim Proposition \advthm.
+A point $x$ is a cluster point of the net $(x_i)_{i\in I}$ if and only if there is a subnet
+of $(x_i)_{i\in I}$ converging to $x$.
+
+\proof Suppose there is a subnet of $(x_i)_{i\in I}$ converging to $x$. That is, there is a directed
+set $J$ and a cofinal map $\phi : J\to I$ such that, setting $y_j = x_{\phi(j)}$ for all $j\in J$,
+the subnet $(y_j)_{j\in J}$ of $(x_i)_{i\in I}$ converges to $x$. Let $U$ be an open neighbourhood
+of $x$ and let $i\in I$. There is some $j_0\in J$ such that $j\succeq j_0$ implies that $y_j\in U$.
+Since $\phi$ is cofinal, there is $j_1\in J$ such that for all $j\succeq j_1$, $\phi(j)\succeq i$.
+Now since $J$ is directed, there is some $j'\in J$ with $j'\succeq j_0$ and $j'\succeq j_1$.
+For this choice of $j'$ we have $\phi(j')\succeq i$ and $x_{\phi(j')} = y_{j'} \in U$. Hence $x$
+is a cluster point of $(x_i)_{i\in I}$.
+
+Conversely, suppose that $x$ is a cluster point of the net $(x_i)_{i\in I}$. Let $N(x)$
+be the set of all neighbourhoods of $x$ as in Proposition~{\propnbhdnet}. We define
+$$J = \bigl\{ (U, i) : U\in N(x), x_i\in U\bigr\}$$
+and impose the preorder $(U,i) \preceq (U', i')$ if and only if $U\supseteq U'$ and $i\preceq i'$.
+For any $(U_0, i_0)$ and $(U_1, i_1)$ in $J$, we can pick $i'\in I$ such that $i\succeq i_0$
+and $i\succeq i_1$. Since $x$ is a cluster point, there is some $i^*\ge i$ such that
+$x_{i^*}\in U_0\cap U_1$, and we see that $(U_0, i_0) \preceq (U_0\cap U_1, i^*)$
+and $(U_1, i_1)\preceq (U_0 \cap U_i, i^*)$. This shows that $J$ is a directed set. We define
+the map $\phi : J\to I$ by sending $(U,i)\mapsto i$. This is certainly cofinal, by our choice of preorder
+on $J$. To show that $\bigl(x_{\phi(j)}\bigr)_{j\in J}$ converges to $x$, let $V$ be any neighbourhood
+of $x$. Since $x$ is a cluster point, there is some $i\in I$ with $x_i\in V$; consider the
+element $j = (V, i)\in J$. For any $j' = (V', i') \in J$ with $(V', i')\succeq (V,i)$, we have
+$V' \subseteq V$ and $x_{i'}\in V'$, so $x_{\phi(j')} = x_{i'}\in V$.\slug
+
+\medskip\boldlabel Compactness.
+A set $A$ is {\it compact} if every open cover of $A$ has a finite subcover. If every point of $S$
+has some compact neighbourhood, then $S$ is called {\it locally compact}.
 

### 2025-09-30 teaching/math242f25/tut3.tex
posted tutorial 3

\input fontmac
\input mathmac

\def\To{\Rightarrow}

\maketitle{Math 242 Tutorial 3}{prepared by}{Marcel Goh}{25 September 2025}

\bigskip

\proclaim Problem 1. Show that the supremum and infimum of a set are both unique.

\noindent[Only after we prove this statement does it makes sense to say, e.g.,
{\it the} supremum of a set, rather than just {\it a} supremum. Hence in the proof we can only refer
to something as being {\it a} supremum.]

\proof We present the proof for suprema. The proof for infima is the same, {\it mutatis mutandis}
(do it as an exercise).

Let $S\subseteq \RR$ be a set and let $s$ and $s'$ both be suprema of $S$.
We want to show that $s = s'$.
Suppose, for a contradiction, that $s\ne s'$. Without loss of generality, assume that
$s<s'$ (by switching the roles of $s$ and $s'$ if necessary).
Since $s'$ is a supremum of $S$, for all $y < s'$ there exists some $x\in S$ such that
$x > y$. Well since $s<s'$ we can apply this with $y=s$ to find some
$x\in S$ with $x > s$. This means that $s$ is not an upper bound of the set $S$,
contradicting our assumption that $s$ was a supremum of $S$. We conclude that $s = s'$.\slug

\proclaim Problem 2. Let $S$ be a nonempty subset of $\RR$ that is bounded from below.
Prove that a real number $t$ is the infimum of $S$ if and only if
\medskip
\item{i)} $t$ is a lower bound of $S$; and
\smallskip
\item{ii)} for all $\eps > 0$, the number $t + \eps$ is {\it not} a lower bound of $S$.
\medskip

\noindent[An analogue of this statement holds for suprema as well (see Question 1, part (a), on Assignment 3).
The proof is the same, so we'll just do the infimum version. You should be able to do the other
statement as an easy exercise by making suitable modifications in this proof.]

\proof First suppose that $t$ is the infimum of $S$. Then it is a lower bound of $S$, so (i) is
immediately satisfied. To show part (ii), let $\eps > 0$ be given. The number $t' = t+\eps$
satisfies $t' > t$, so since $t$ is the greatest lower bound of $S$,
we see that $t+\eps = t'$ is not a lower bound of $S$.

Now suppose that $t$ satisfies conditions (i) and (ii) above. We need to show that it is the
greatest lower bound of $S$. It is certainly a lower bound, by condition (i). To show that it is
the greatest lower bound, let $t'\in \RR$ be a number with $t' > t$. We can let $\eps = t'-t > 0$, so
that $t' = t + \eps$. Then condition (ii) tells us $t' = t+ \eps$ that $t'$ is not a lower bound
of $S$. Hence $t$ is the infimum of $S$.\slug

\proclaim Problem 3. Let $S = \bigl\{ 1 - 1/n^2 : n\in \NN\bigr\}$. Prove that $\sup S = 1$
and $\inf S = 0$.

\proof For $n = 1$ we have $1 - 1/n^2 = 1-1 = 0$. So $0\in S$. On the other hand, $0$
is also a lower bound of $S$, since
$$ 1 - {1\over n^2} \ge 1-{1\over 1^2} = 0 $$
for all $n\in \NN$. This shows that $0=\min S$ and hence $0 = \inf S$.

To show that $\sup S = 1$, we need to show that $1$ is an upper bound of $S$ and that $1-\eps$
is not an upper bound of $S$ for any $\eps > 0$. (This is in light of the supremum version
of the previous problem, or Question 1, part (a) of Assignment 3.) To see that $1$ is an upper bound,
note that for any $n\in \NN$, $1/n^2 > 0$, so $1-1/n^2 < 1$. For the second part, let $\eps > 0$ be
given and pick $n>1/\sqrt\eps$. Then
$$ 1- {1\over n^2} > 1 - {1\over 1/(\sqrt\eps)^2} = 1-{1\over 1/\eps} = 1-\eps,$$
so $1-\eps$ is not an upper bound of $S$. Since $\eps$ was arbitrary, we are done.\slug

\proclaim Problem 4. For any nonempty set $A\subseteq \RR$ define the {\it reflection}
$$ -A = \{ - a : a\in A \}.$$
Prove the following.
\medskip
\item{a)} The number $u$ is an upper bound of $A$ if and only if $-u$ is a lower bound of $-A$.
\smallskip
\item{b)} The set $A$ is bounded from above if and only if $-A$ is bounded from below.
\smallskip
\item{c)} If $A$ has a supremum, then $\inf(-A) = -\sup A$.
\smallskip
\item{d)} Any nonempty set $B\subseteq \RR$ that is bounded from below has an infimum.
\medskip

\proof Let $u$ be an upper bound of $A$. This means that $a\le u$ for all $a\in A$. But then
we have $-a \ge -u$ for all $a\in A$, and we see that $-u$ is a lower bound of $-A$. Conversely,
assume that $-u$ is a lower bound of $-A$. Then $-a \ge -u$ for all $a\in A$, so $a\le u$
for all $a\in A$, which implies that $u$ is an upper bound of $A$. [You could also just have
noticed straightaway that each of these implications is an equivalence, and proved the ``if and only if''
all at once.]

By definition, $A$ is bounded from above if and only if there is some $u\in \RR$
that is an upper bound of $A$. But by part (a),
this is true if and only if there is some $u\in \RR$ such that $-u$ is a lower bound of $-A$,
and this is the definition of $-A$ being bounded from below.

Now we prove part (c). To do so, we must show that $-\sup A$ is a lower bound of $-A$ and that
$-\sup A + \eps$ is not a lower bound of $-A$ for any $\eps > 0$. Well, $\sup A$ is an upper bound
of $A$, so by part (a), $-\sup A$ is a lower bound of $-A$. That settles one of the requirements. For
the other one, let $\eps > 0$ be given. Since $\sup A$ is the least upper bound of $A$,
$\sup A - \eps$ is not an upper bound of $A$; that is, there is some $a\in A$ with $a > \sup A - \eps$.
But this implies that $-a < -\sup A + \eps$, so $-\sup A + \eps$ is not a lower bound of $-A$.
Hence $-\sup A$ is the infimum of $-A$.

Let $B$ be nonempty and bounded from below. Let $A = -B$, so that $A$ is also nonempty
and $-A = B$. By part (b), $A$ is bounded from
above, so the least-upper-bound property of $\RR$ implies that $A$ has a supremum. Now by part (c),
$\inf B = \inf(-A) = -\sup A$, and {\it a fortiori} $\inf B$ exists.\slug

\proclaim Problem 5. Prove the Cauchy--Schwarz inequality, the statement that for all
$u_1,\ldots,u_n, v_1,\ldots,v_n\in \RR$,
$$\biggl(\sum_{i=1}^n u_i v_i\biggr)^2
\le \biggl( \sum_{i=1}^n {u_i}^2\biggr) \biggl( \sum_{i=1}^n {v_i}^2\biggr).$$
[{\it Hint:} Recall from your high school days that
for a quadratic polynomial $ax^2 + bx + c$, the {\it discriminant} is
defined to be the quantity $b^2 - 4ac$. (It appears under the square-root sign
in the quadratic formula.) If it is negative, the polynomial has no real roots,
if it is positive, the polynomial has two real roots, and if it is zero, the polynomial
has one real root.]

\proof If $u_i = 0$ for all $1\le i\le n$, then both sides are $0$ and the inequality
is trivially true. So suppose that at least one of the $u_i$ is nonzero. Then
$$p(x) = (u_1 x + v_1)^2 + (u_2 x + v_2)^2 + \cdots + (u_n x+v_n)^2$$
is a quadratic polynomial in the variable $x$. Rewriting $p(x)$ in summation notation,
we have
$$\eqalign{
p(x) &= \sum_{i=1}^n (u_ix + v_i)^2\cr
&= \sum_{i=1}^n \bigl( u_i^2 x^2 + 2u_i v_i + v_i^2\bigr)\cr
 &= \biggl(\sum_{i=1}^n u_i^2 \biggr) x^2 +
2\biggl(\sum_{i=1}^n u_i v_i\biggr) x + \biggl( \sum_{i=1}^n {v_i}^2\biggr).\cr
}$$
Since it is defined as a sum of squares, $p(x)\ge 0$ for all $x\in \RR$, so it has
at most one real root.
This means its discriminant is nonpositive; in other words,
$$4\biggl(\sum_{i=1}^n u_i v_i\biggr)^2
- 4\biggl( \sum_{i=1}^n {u_i}^2\biggr) \biggl( \sum_{i=1}^n {v_i}^2\biggr) \le 0.$$
Dividing through by $4$ and then rearranging gives exactly the inequality we sought.\slug

\bye

### 2025-10-03 misc/expo/additive.tex
Tutorial 4

@@
 how useful this will be to anyone else. Currently, the bulk of these notes come from
-a series of YouTube lectures of Timothy Gowers (given in early 2022), though I intend
-in future to add more relevant topics as I learn about them.
-
-\bigskip
+a series of YouTube lectures of Timothy Gowers (given in early 2022), though I'm slowly adding more
+relevant topics as I learn about them.
+\vskip36pt
+\noindent{\ninessbx I. Sumsets and the structure of small doubling}
+\smallskip
+\contline{1}{Pl\"unnecke's theorem}{1}
+\smallskip
+\contline{2}{The Balog--Szemer\'edi--Gowers theorem}{5}
+\smallskip
+\contline{3}{Freiman homomorphisms and Ruzsa embedding lemmas}{9}
+\smallskip
+\contline{4}{Rudiments of discrete Fourier analysis}{12}
+\smallskip
+\contline{5}{Bohr sets and Bogolyubov's lemma}{17}
+\smallskip
+\contline{6}{Freiman's theorem in vector spaces over fields of prime order}{19}
+\medskip
+\noindent{\ninessbx II. Arithmetic progressions}
+\smallskip
+\contline{7}{Roth's theorem}{22}
+\medskip
+\noindent{\ninessbx III. Miscellaneous theorems}
+\smallskip
+\contline{8}{Khovanskii's theorem}{27}
+\smallskip
+\contline{9}{The Furstenberg--S\'ark\"ozy theorem}{29}
+\vskip36pt
 
@@
 we like.\slug
-
-\advsect Khovanskii's theorem
-
-The goal of this section is to prove the following wonderful theorem of A.~Khovanskii.
-
-\edef\thmkhovanskii{\the\thmcount}
-\parenproclaim Theorem {\advthm} (Khovanskii, {\rm 1992}). Let $A$ be a finite
-subset of an abelian group and let $f_A(n) = |nA|$ for each $n\in \NN$. Then there exists
-$n_A$ and a polynomial $p_A$ such that $f_A(n) = p_A(n)$ for all $n\ge n_A$.
-
-Consider the example when $A = \{0,1,b-1,b\} \subseteq \ZZ$. Then letting $[a,b] = \{x\in \ZZ : a\le x\le b\}$,
-we have
-$$nA = [0, n] \cup [b-1,\ldots, b+n-1] \cup [2b-2, 2b+n-2] \cup\cdots\cup [n(b-1),nb].$$
-This has size $(n+1)^2$ as long as $n< b-1$, but when $n\ge b-1$, $nA$ is simply the set
-of integers in the range $[0,bn]$, which has size $bn+1$.
-
-We shall present a combinatorial proof of Khovanskii's theorem,
-given by M.~B.~Nathanson and I.~Z.~Ruzsa in 2002.
-Let $A = \{a_1,\ldots, a_r\}$ and
-for $x = (x_1,\ldots, x_r)\in {\NN_0}^r$, write
-$$\sigma(x) = \sum_{i=1}^r x_i\qquad\hbox{and}\qquad
-\alpha(x) = \sum_{i=1}^r a_ix_i.$$
-Then
-$$nA = \{ \alpha(x) : x\in {\NN_0}^r,\, \sigma(x) = n\}.$$
-Let $\prec$ be the lexicographic order on ${\NN_0}^r$; that is, $(x_1,\ldots,x_r)\prec (y_1,\ldots,y_r)$
-if there is some $i\in [r]$ such that $x_i<y_i$ and $x_j = y_j$ for all $1\le j<i$.
-Say that $x$ is {\it useless} if there is $x'\prec x$ such that $\sigma(x') = \sigma(x)$ and
-$\alpha(x') = \alpha(x)$; otherwise, say that $x$ is {\it useful}. It is easy to see that
-$|nA|$ is exactly the number of useful $x$ with $\sigma(x) = n$. Now let $\le$ denote
-the usual partial order on ${\NN_0}^r$, where $(x_1,\ldots,x_r)\le (y_1,\ldots,y_r)$
-if $x_i\le y_i$ for all $1\le i\le r$.
-
-\proclaim Lemma {\advthm}. If $x\le y$ and $x$ is useless, then $y$ is useless.
-
-\proof Since $x$ is useless, we can find $x'\prec x$ with $\sigma(x') = \sigma(x)$ and $\alpha(x') = \alpha(x)$.
-Let $y' = y+x'-x$. Since $x\le y$, $y-x$ is in ${\NN_0}^r$ and therefore so is $y'$. Then since
-$x_i' < x_i$ for the first $i$ at which the two tuples differ, the first coordinate at which $y'$ and
-$y$ differ is also $i$ and we have $y_i'<y_i$. Hence $y'\prec y$. Lastly,
-$$\sigma(y') = \sigma(y+x'-x) = \sigma(y) + \sigma(x') - \sigma(x) = \sigma(y)$$
-and the same is true with $\sigma$ replaced by $\alpha$, so $y$ is useless.\slug
-
-For each useless $x$ we can find a useless $x'$, minimal in the $\le$ order, such that $x'\le x$.
-This lemma gives a converse; it says that if $x'$ is a minimal useless element and $x'\le x$, then
-$x$ is also useless.
-
-\edef\lemminimalfinite{\the\thmcount}
-\proclaim Lemma {\advthm}. For any nonempty $U\subseteq {\NN_0}^r$, the set of minimal elements of
-$U$ (with respect to the $\le$ order) is finite.
-
-\proof We perform induction on $r$. If $r=1$, then it is clear that $U$ has exactly one minimal element.
-Now let $r>1$ and let $u$ be a minimal element of $u$. Then for every $v\in U$ with $u\not\le v$,
-there exists some $i$ and $t< u_i$ such that $v_i = t$. Let
-$$V(i,t) = \{v\in U : v_i = t\}$$
-for all $1\le i\le r$ and $0\le t<u_i$. Each set $V(i,t)$ is order-isomorphic to a subset of ${\NN_0}^{r-1}$,
-so by induction it has finitely many minimal elements. The number of $V(i,t)$ is $r\prod_{i=1}^r u_i$,
-and every minimal element of $u$ is either $u$ or a minimal element of some $V(i,t)$, so there are
-finitely many minimal elements in $U$.\slug
-
-We will need just one more lemma before we are able to prove Khovanskii's theorem. It shows that
-for any fixed $r$-tuple $x'$, the number of $x$ with $x'\le x$ and $\sigma(x) = n$
-is given by a polynomial in the variable $n$, with coefficients depending on $x'$ and $r$.
-
-\proclaim Lemma {\advthm}. Let $x'\in {\NN_0}^r$ and for $n\ge \sigma(x')$, let
-$$C(x', n) = \{n\in {\NN_0}^r : \sigma(x) = n,\,x'\le x\}.$$
-Then
-$$\bigl| C(x',n)\bigr| = {n-\sigma(x') + r-1 \choose r-1}.$$
-
-\proof Given $x\in C(x',n)$, map it to $x-x'$. This is a bijection from $C(x',n)$ to $C(0, n-\sigma(x')$.
-Now let $m = n-\sigma(x')$. The map $(x_1,\ldots,x_r) \mapsto (x_1,+1,\ldots, x_r+1)$ is a bijection
-from $C(0,m)$ to the set of $y\in \NN^r$ such that $y_1+\cdots+y_r = m+r$. Now the map
-$$(y_1,\ldots,y_r) \mapsto \{y_1, y_1+y_2, \ldots, y_1+\cdots+y_r\}$$
-is a bijection from the previous set to the set of subsets of $[m+r]$ that contain $m+r$. (It is
-inverted by the map $\{a_1, \ldots, a_r\} \mapsto (a_1, a_2-a_1, \ldots, a_r-a_{r-1})$.) But this final
-set has size ${m+r-1\choose r-1}$, so we are done.\slug
-
-Without further ado, let us now prove Khovanskii's theorem.
-
-\medskip\noindent {\it Proof of Theorem~{\thmkhovanskii}.}\enspace Let $U$ be the set of all
-useless $r$-tuples. By Lemma~{\lemminimalfinite}, the set $X\subseteq U$ of all minimal useless
-$r$-tuples is finite. Recall that $|nA|$ is the number of useful $x$ such that $\sigma(x) = n$,
-so using the notation of the previous lemma, we have
-$$|nA| = \bigl| C(0,n)\bigr| - \Bigl| \bigcup_{x'\in X} C(x',n)\Bigr|
-=\bigl| C(0,n)\bigr| - \sum_{Y\subseteq X} (-1)^{|Y|} \Bigl| \bigcap_{x'\in Y} C(x',n)\Bigr| $$
-But for any $Y\subseteq X$, $\bigcap_{x'\in Y} C(x',n) = C(x_Y, n)$, where $x_Y$ has for its
-$i$th coordinate the integer $\max_{y\in Y} y_i$. Hence every term in the inclusion-exclusion
-formula depends polynomially on $n$ once
-$$n\ge \sum_{1\le i\le r} \max_{x\in X} x_i.\noskipslug$$
 
@@
 This statement is called Freiman's theorem, as it was first proved by Freiman over the integers.
+A set $A$ for which the ratio
+$|A+A|/|A|$ is small is said to have {\it small doubling}, so Freiman's theorem imposes structure on sets
+with small doubling.
 In these notes we will prove an analogue of it for subsets of $\FF_p^N$, in which ``$d$-dimensional arithmetic
@@
 equivalent to what we just proved: {\sl Any subset of $[N]$
-of size at least $300/\log\log N$ contains a nontrivial arithmetic progression of length $3$.}
+of size at least $300N/\log\log N$ contains a nontrivial arithmetic progression of length $3$.}
+
+\advsect Khovanskii's theorem
+
+The goal of this section is to prove the following wonderful theorem of A.~Khovanskii.
+
+\edef\thmkhovanskii{\the\thmcount}
+\parenproclaim Theorem {\advthm} (Khovanskii, {\rm 1992}). Let $A$ be a finite
+subset of an abelian group and let $f_A(n) = |nA|$ for each $n\in \NN$. Then there exists
+$n_A$ and a polynomial $p_A$ such that $f_A(n) = p_A(n)$ for all $n\ge n_A$.
+
+Consider the example when $A = \{0,1,b-1,b\} \subseteq \ZZ$. Then letting $[a,b] = \{x\in \ZZ : a\le x\le b\}$,
+we have
+$$nA = [0, n] \cup [b-1,\ldots, b+n-1] \cup [2b-2, 2b+n-2] \cup\cdots\cup [n(b-1),nb].$$
+This has size $(n+1)^2$ as long as $n< b-1$, but when $n\ge b-1$, $nA$ is simply the set
+of integers in the range $[0,bn]$, which has size $bn+1$.
+
+We shall present a combinatorial proof of Khovanskii's theorem,
+given by M.~B.~Nathanson and I.~Z.~Ruzsa in 2002.
+Let $A = \{a_1,\ldots, a_r\}$ and
+for $x = (x_1,\ldots, x_r)\in {\NN_0}^r$, write
+$$\sigma(x) = \sum_{i=1}^r x_i\qquad\hbox{and}\qquad
+\alpha(x) = \sum_{i=1}^r a_ix_i.$$
+Then
+$$nA = \{ \alpha(x) : x\in {\NN_0}^r,\, \sigma(x) = n\}.$$
+Let $\prec$ be the lexicographic order on ${\NN_0}^r$; that is, $(x_1,\ldots,x_r)\prec (y_1,\ldots,y_r)$
+if there is some $i\in [r]$ such that $x_i<y_i$ and $x_j = y_j$ for all $1\le j<i$.
+Say that $x$ is {\it useless} if there is $x'\prec x$ such that $\sigma(x') = \sigma(x)$ and
+$\alpha(x') = \alpha(x)$; otherwise, say that $x$ is {\it useful}. It is easy to see that
+$|nA|$ is exactly the number of useful $x$ with $\sigma(x) = n$. Now let $\le$ denote
+the usual partial order on ${\NN_0}^r$, where $(x_1,\ldots,x_r)\le (y_1,\ldots,y_r)$
+if $x_i\le y_i$ for all $1\le i\le r$.
+
+\proclaim Lemma {\advthm}. If $x\le y$ and $x$ is useless, then $y$ is useless.
+
+\proof Since $x$ is useless, we can find $x'\prec x$ with $\sigma(x') = \sigma(x)$ and $\alpha(x') = \alpha(x)$.
+Let $y' = y+x'-x$. Since $x\le y$, $y-x$ is in ${\NN_0}^r$ and therefore so is $y'$. Then since
+$x_i' < x_i$ for the first $i$ at which the two tuples differ, the first coordinate at which $y'$ and
+$y$ differ is also $i$ and we have $y_i'<y_i$. Hence $y'\prec y$. Lastly,
+$$\sigma(y') = \sigma(y+x'-x) = \sigma(y) + \sigma(x') - \sigma(x) = \sigma(y)$$
+and the same is true with $\sigma$ replaced by $\alpha$, so $y$ is useless.\slug
+
+For each useless $x$ we can find a useless $x'$, minimal in the $\le$ order, such that $x'\le x$.
+This lemma gives a converse; it says that if $x'$ is a minimal useless element and $x'\le x$, then
+$x$ is also useless.
+
+\edef\lemminimalfinite{\the\thmcount}
+\proclaim Lemma {\advthm}. For any nonempty $U\subseteq {\NN_0}^r$, the set of minimal elements of
+$U$ (with respect to the $\le$ order) is finite.
+
+\proof We perform induction on $r$. If $r=1$, then it is clear that $U$ has exactly one minimal element.
+Now let $r>1$ and let $u$ be a minimal element of $u$. Then for every $v\in U$ with $u\not\le v$,
+there exists some $i$ and $t< u_i$ such that $v_i = t$. Let
+$$V(i,t) = \{v\in U : v_i = t\}$$
+for all $1\le i\le r$ and $0\le t<u_i$. Each set $V(i,t)$ is order-isomorphic to a subset of ${\NN_0}^{r-1}$,
+so by induction it has finitely many minimal elements. The number of $V(i,t)$ is $r\prod_{i=1}^r u_i$,
+and every minimal element of $u$ is either $u$ or a minimal element of some $V(i,t)$, so there are
+finitely many minimal elements in $U$.\slug
+
+We will need just one more lemma before we are able to prove Khovanskii's theorem. It shows that
+for any fixed $r$-tuple $x'$, the number of $x$ with $x'\le x$ and $\sigma(x) = n$
+is given by a polynomial in the variable $n$, with coefficients depending on $x'$ and $r$.
+
+\proclaim Lemma {\advthm}. Let $x'\in {\NN_0}^r$ and for $n\ge \sigma(x')$, let
+$$C(x', n) = \{n\in {\NN_0}^r : \sigma(x) = n,\,x'\le x\}.$$
+Then
+$$\bigl| C(x',n)\bigr| = {n-\sigma(x') + r-1 \choose r-1}.$$
+
+\proof Given $x\in C(x',n)$, map it to $x-x'$. This is a bijection from $C(x',n)$ to $C(0, n-\sigma(x')$.
+Now let $m = n-\sigma(x')$. The map $(x_1,\ldots,x_r) \mapsto (x_1,+1,\ldots, x_r+1)$ is a bijection
+from $C(0,m)$ to the set of $y\in \NN^r$ such that $y_1+\cdots+y_r = m+r$. Now the map
+$$(y_1,\ldots,y_r) \mapsto \{y_1, y_1+y_2, \ldots, y_1+\cdots+y_r\}$$
+is a bijection from the previous set to the set of subsets of $[m+r]$ that contain $m+r$. (It is
+inverted by the map $\{a_1, \ldots, a_r\} \mapsto (a_1, a_2-a_1, \ldots, a_r-a_{r-1})$.) But this final
+set has size ${m+r-1\choose r-1}$, so we are done.\slug
+
+Without further ado, let us now prove Khovanskii's theorem.
+
+\medskip\noindent {\it Proof of Theorem~{\thmkhovanskii}.}\enspace Let $U$ be the set of all
+useless $r$-tuples. By Lemma~{\lemminimalfinite}, the set $X\subseteq U$ of all minimal useless
+$r$-tuples is finite. Recall that $|nA|$ is the number of useful $x$ such that $\sigma(x) = n$,
+so using the notation of the previous lemma, we have
+$$|nA| = \bigl| C(0,n)\bigr| - \Bigl| \bigcup_{x'\in X} C(x',n)\Bigr|
+=\bigl| C(0,n)\bigr| - \sum_{Y\subseteq X} (-1)^{|Y|} \Bigl| \bigcap_{x'\in Y} C(x',n)\Bigr| $$
+But for any $Y\subseteq X$, $\bigcap_{x'\in Y} C(x',n) = C(x_Y, n)$, where $x_Y$ has for its
+$i$th coordinate the integer $\max_{y\in Y} y_i$. Hence every term in the inclusion-exclusion
+formula depends polynomially on $n$ once
+$$n\ge \sum_{1\le i\le r} \max_{x\in X} x_i.\noskipslug$$
 

### 2025-10-03 teaching/math242f25/tut4.tex
Tutorial 4

\input fontmac
\input mathmac

\def\To{\Rightarrow}

\maketitle{Math 242 Tutorial 4}{prepared by}{Marcel Goh}{2 October 2025}

\bigskip

\proclaim Problem 1. For any sets $A, B\subseteq \RR$, we define the {\it difference set}
to be the set
$$A - B = \{ a- b : a\in A, b\in B\}.$$
Prove that for all sets $U,V,W\subseteq \RR$, we have the inequality
$$|U-W| \le {|U-V| \cdot |V-W|\over |V|}.$$

This
is a very fundamental inequality in the field of additive combinatorics. It is called the
{\it Ruzsa triangle inequality}, for its superficial resemblance to the ordinary triangle inequality.
(When you learn about metric spaces, you'll learn that the resemblance is actually more
than superficial.)

\proof What we have to show can be rearranged to
$$ |U-W| \cdot |V| \le |U-V|\cdot |V-W|.$$
We are done then, if we can furnish an injective function from $(U-W)\times V$
to $(U-V)\times (V-W)$.

First we need to do some bookkeeping. For each $x\in U-W$, let
$u(x)\in U$ and $w(x)\in W$ be such that $u(x) - w(x) = x$. (If there are multiple choices for
this pair $\bigl( u(x), w(x)\bigr)$, just pick one.) Now we define a function.
For $(x,v)\in (U-W)\times V$, we let
$$f(x,v) = \bigl( u(x) - v, w(x) - v\bigr).$$
By construction, the output of this function will be an element of $(U-V)\times(V-W)$, so we have
$f : (U-W)\times V \to (U-V)\times (V-W)$.

We now show that $f$ is injective. Suppose that $(x_1, v_1)\in (U-W)\times V$ and
$(x_2, v_2)\in (U-W)\times V$ are such that $f(x_1,v_1) = f(x_2,v_2)$. Then
we have
$$\bigl( u(x_1) - v_1, w(x_1) - v_1 \bigr) = \bigr( u(x_2) - v_2, w(x_2) - v_2\bigr),$$
so subtracting the second element of each pair from the first element, we have
$$ u(x_1) - w(x_1) = u(x_2) - w(x_2).$$
But by definition of the functions $u(x)$ and $w(x)$, we have $u(x_1) - w(x_1) = x_1$
and $u(x_2) - w(x_2) = x_2$. So we have found that $x_1 = x_2$. Lastly, note that
$$u(x_1) - v_1 = u(x_2) - v_2,$$
so since we have found out that $x_1 = x_2$, we must also have $v_1 = v_2$.\slug

Students of algebra will be delighted to realise that this proof did not use anything particular
about $\RR$. The Ruzsa triangle inequality is valid in any group.

\proclaim Problem 2. In class we took the least upper bound property as an axiom (the completeness axiom)
and proved the nested intervals property as well as the Archimedean property.
Assuming the nested intervals property and the Archimedean property as axioms, prove the
least upper bound property.

\proof Let $S\subseteq \RR$ be a nonempty set that is bounded from above. Let $a\in S$ and $b\in \RR$
be an upper bound of $S$. We have $a\le b$, so we can let $a_1 = a$, $b_1 = b$, and call $I_1 = [a_1, b_1]$.
Note that the length of this interval is $|I_1| = a_1 - b_1 = a-b$.

We now define a sequence of nested intervals, starting with $I_1$. For $n\ge 1$, we assume that $a_n$
and $b_n$ are already defined, and let $x_n$ be the midpoint $(b_n - a_n)/2$. There are two possibilities.
\medskip
\item{i)} If $x_n$ is an upper bound of $S$, then we let $a_{n+1} = a_n$ and $b_{n+1} = x_n$.
\smallskip
\item{ii)} If not, then let $a_{n+1} = x_n$ and $b_{n+1} = b_n$.
\medskip
Since $I_{n+1}$ is half the length of $I_n$ for all $n\in \NN$ and we had $|I_1| = a-b$, we have
$|I_n| = (a-b)/2^{n-1}$. Next, since we only set $b_{n+1} = x_n$ if $x_n$ is an upper bound of $S$
and otherwise we have $b_{n+1} = b_n$, we see that $b_n$ is an upper bound of $S$ for all $n\in \NN$.
Lastly, note that $I_n$ always contains some element of $S$. This we show by induction. It is true for $n=1$,
so for the induction step, suppose it's true for some $n\in \NN$. We want to show that $I_{n+1}$ contains
some element of $S$. If $x_n$ is an upper bound of $S$, then $(x_n, b_n]$ does not contain any element
of $S$, so $I_n = [a_n, x_n]$ must. If $x_n$ not an upper bound of $S$, then there must be some $y\in S$
with $y>x$. But $b_n$ is an upper bound of $S$, so then in this case $I_n = [x_n, b_n]$ and $y$ is in
this range.

By the nested interval property, there is an element $s\in \bigcap_{n\in \NN} I_n$. The claim is that
$s$ is the supremum of $S$. There are two things to show: that it is an upper bound of $S$, and that
no smaller element can be an upper bound of $S$.

Suppose, towards a contradiction, that there is some $y\in S$ with $y > s$. Let $\eps = y-s > 0$. By the
Archimedean property, there is some $N\in \NN$ with $N > \log_2\bigl( (b-a)/\eps\bigr) + 1$. This can
be rearranged to get
$$\eps > {b-a \over 2^{N-1}}.$$
Consider $I_N = [a_N , b_N]$, which has $b_N - a_N = (b-a) / 2^{N-1}$. Since $s\in I_N$, we have
$s \ge a_N$, and
$$ y = s+\eps \ge a_N + \eps > a_N + {b-a\over 2^{N-1}} = a_N + (b_N - a_N) = b_N.$$
This contradicts the fact that $b_N$ is an upper bound of $S$. So we see that $s$ is an upper bound of $S$.

Now let $t < s$. We must show that $t$ is not an upper bound of $S$.
This time, set $\eps = s - t > 0$.
By the Archimedean property again, we can pick $N\in \NN$ large enough so that
$$\eps > {b-a \over 2^{N-1}} = b_N - a_N.$$
Again we note that $s\in I_n = [a_N, b_N]$. Hence $s\le b_N$ and we have
$$ t = s-\eps \le b_N - \eps < b_N - (b_N - a_N) = a_N.$$
But since $I_N$ contains an element of $S$, there is some element of $S$ that is at least $a_N$, which
implies that it is greater than $t$. We conclude that $t$ is not an upper bound of $S$, which
settles the proof that $s = \sup S$.\slug

\proclaim Problem 3. Show that the set $[0,1)$ is uncountable.

\proof If we disallow infinite trailing sequences of $9$s, then every element $x\in [0,1)$ has a unique
decimal representation as
$$x = 0.d_1d_2d_3 \cdots,$$
where, for all $n\in\NN$, $d_n$ is a digit from the set $\{0,1,\ldots,9\}$.

Now we show that there is no surjective function $f : \NN\to [0,1)$. Let $f$ be any function
from $f : \NN = [0,1)$. For all $n\in \NN$, let
$$f(n) = 0.d_{n,1} d_{n,2} d_{n,3} \cdots,$$
where we always use a representation of $f(n)$ that doesn't have an infinite trailing sequence of $9$s.
The claim is that there is some $x\in [0,1)$ that differs from every $f(n)$ by at least one digit.

Consider the diagonal sequence $d_{1,1}, d_{2,2}, d_{3,3}, \ldots$. For all $n$, define
$$ c_n = \cases{d_{n,n} - 1,& if $d_{n,n} > 0$;\cr 1,& if $d_{n,n} = 0$.\cr}.$$
Let $x$ be the element of $[0,1)$ with decimal expansion
$$ x = 0.c_1 c_2 c_3 \cdots.$$
Note that none of these digits are equal to $9$, so there is no infinite trailing sequence of $9$s
in the decimal expansion of $x$. Furthermore, $f(n) \ne x$ for all $n\in \NN$, since $f(n)$ has
the digit $d_{n,n}$ in the $n$th decimal place, whereas $x$ has the digit $c_n \ne d_{n,n}$. So
$f$ is not surjective.

We have proved that there is no surjection from $\NN$ to $[0,1)$, so {\it a fortiori} there is no bijection
from $\NN$ to $[0,1)$. We conclude that $[0,1)$ is uncountable.\slug

\proclaim Problem 4. Let $x$ be a real number. Prove that for every $\eps>0$
there exist two rational numbers $q$ and $q'$ such that $q < x < q'$
and $|q-q'|<\eps$.

\proof We know from class that $\QQ$ is dense in $\RR$. We can apply this with
$x-\eps/2$ and $x$ to get $q\in \QQ$ with
$$ x - {\eps \over 2} < q < x.$$
Then we can apply it with $x$ and $x + \eps/2$ to get $q'\in \QQ$ with
$$ x< q' < x + {\eps \over 2}.$$
By construction, we have $q < x < q'$, as desired. It remains to show that
$ |q - q'| < \eps$. Well, we know that $q' < x+\eps/2$ and $q > x-\eps/2$.
Negating the latter, we get $-q < -x + \eps/2$, and adding this to the former,
we obtain
$$|q-q'| = q' - q < {\eps\over 2} + {\eps \over 2} = \eps.$$
So we are done.\slug

\bye

### 2025-10-03 teaching/math242f25/tut4.tex
typo on page

@@
 
-\proclaim Problem 4. Let $x$ be a real number. Prove that for every $\eps>0$
-there exist two rational numbers $q$ and $q'$ such that $q < x < q'$
-and $|q-q'|<\eps$.
-
-\proof We know from class that $\QQ$ is dense in $\RR$. We can apply this with
-$x-\eps/2$ and $x$ to get $q\in \QQ$ with
-$$ x - {\eps \over 2} < q < x.$$
-Then we can apply it with $x$ and $x + \eps/2$ to get $q'\in \QQ$ with
-$$ x< q' < x + {\eps \over 2}.$$
-By construction, we have $q < x < q'$, as desired. It remains to show that
-$ |q - q'| < \eps$. Well, we know that $q' < x+\eps/2$ and $q > x-\eps/2$.
-Negating the latter, we get $-q < -x + \eps/2$, and adding this to the former,
-we obtain
-$$|q-q'| = q' - q < {\eps\over 2} + {\eps \over 2} = \eps.$$
-So we are done.\slug
+% \proclaim Problem 4. Let $x$ be a real number. Prove that for every $\eps>0$
+% there exist two rational numbers $q$ and $q'$ such that $q < x < q'$
+% and $|q-q'|<\eps$.
+% 
+% \proof We know from class that $\QQ$ is dense in $\RR$. We can apply this with
+% $x-\eps/2$ and $x$ to get $q\in \QQ$ with
+% $$ x - {\eps \over 2} < q < x.$$
+% Then we can apply it with $x$ and $x + \eps/2$ to get $q'\in \QQ$ with
+% $$ x< q' < x + {\eps \over 2}.$$
+% By construction, we have $q < x < q'$, as desired. It remains to show that
+% $ |q - q'| < \eps$. Well, we know that $q' < x+\eps/2$ and $q > x-\eps/2$.
+% Negating the latter, we get $-q < -x + \eps/2$, and adding this to the former,
+% we obtain
+% $$|q-q'| = q' - q < {\eps\over 2} + {\eps \over 2} = \eps.$$
+% So we are done.\slug
 

### 2025-10-10 teaching/math242f25/tut5.tex
Committing in a McDonalds

\input fontmac
\input mathmac

\def\To{\Rightarrow}

\maketitle{Math 242 Tutorial 5}{prepared by}{Marcel Goh}{9 October 2025}

\bigskip

\proclaim Problem \advthm. Prove that
\medskip
\item{a)} $1/a^n\to 0$ as $n\to \infty$, if $a>1$;
\smallskip
\item{b)} $a^n\to 0$ as $n\to \infty$, if $0<a<1$; and
\smallskip
\item{c)} $a^n \to 0$ as $n\to \infty$, if $-1<a<0$.
\medskip

\proof Let $b = a-1$, so that $b>0$. Now for $\eps > 0$ arbitrary, let $N > {1/(b\eps)}$.
Then for all $n\ge N$,
$$ \biggl| {1\over a^n} - 0\biggr| = {1\over a^n} = {1\over (1+b)^n}
\le {1\over 1+nb} < {1\over nb} < {1\over Nb} < \eps$$
by Bernoulli's inequality. This shows that $1/a^n\to 0$.

For part (b), let $c = 1/a$. Then $c>1$ and $a^n = 1/c^n$. It follows from part (a) that
$$\lim_{n\to\infty} a^n = \lim_{n\to\infty} {1\over c^n} = 0.$$

For part (c), note that $0<|a|<1$, so we know that $|a|^n\to 0$ by part (b). But
$$|a^n - 0| = \bigl| |a|^n - 0 \bigr|$$
for all $n\in \NN$, so we see that $a^n\to 0$ as well.\slug

\proclaim Problem \advthm. Prove that
\medskip
\item{a)} $2^n / n! \to 0$ as $n\to \infty$; and that
\smallskip
\item{b)} $n!/n^n \to 0$.
\medskip

\proof For part (a), note that $2/k \le 2/3$ for all $k\ge 3$. Hence
$$ {2^n\over n!} = {2\over 1}\cdot {2\over 2}\cdot {2\over 3} \cdots {2\over n}
\le {2\over 1}\cdot {2\over 2} \cdot \Bigl( {2\over 3}\Bigr)^{n-2}
= 2\cdot \Bigl({3\over 2}\Bigr)^2 \Bigl( {2\over 3}\Bigr)^n
= {9\over 2} \cdot \Bigl( {2\over 3}\Bigr)^n$$
for all $n\ge 2$. We showed in class that $a^n \to 0$ as $n\to \infty$ whenever
$a < 1$. We can apply this with $a = 2/3$ to see that
$$\lim_{n\to \infty} {9\over 2} \cdot \Bigl( {2\over 3}\Bigr)^n = 0.$$
In other words, for every $\eps > 0$ there exists $N\in \NN$ such that for all $n\ge N$,
$${9\over 2} \cdot \Bigl( {2\over 3}\Bigr)^n < \eps.$$
But then we see that
$$0\le {2^n\over n!} \le {9\over 2} \cdot \Bigl( {2\over 3}\Bigr)^n < \eps,$$
so
$$\biggl| {2^n\over n!} - 0\biggr| <\eps$$
for all $n\ge N$. Since $\eps > 0$ was arbitrary, we are done.

For part (b), we expand
$${n!\over n^n} = {1\over n}\cdot {2\over n} \cdots {n\over n} \le {1\over n}.$$
Let $\eps > 0$ and pick $N > 1/\eps$. Then for all $n\ge N$, we have
$$ \biggl| {n!\over n^n} - 0\biggr| = {n!\over n^n} \le {1\over n} < \eps,$$
so we see that $n!/n^n \to 0$.\slug

\proclaim Problem \advthm. Let $a>1$. Prove that $n/a^n\to 0$ as $n\to \infty$.

\proof Let's start by seeing why we can't just do what we did for part (a) of Problem~1 above.
Say we set $b = a-1$, so that $b>0$. Then for $n\in \NN$, we have
$$ \biggl| {n\over a^n} - 0\biggr| = {n\over a^n} = {n\over (1+b)^n}
\le {n\over 1+nb} < {n\over nb} = {1\over b}$$
by Bernoulli's inequality, but now we're stuck, because $1/b > 0$.
Note, however, that we were free to pick $b$ for the purposes
of this proof, so we should try to set it to something that creates a larger power of $n$ in the
denominator. This will cancel the $n$ in the numerator that is giving us problems.

So let us set $b = \sqrt a - 1$. Since $\sqrt a > 1$ whenever $a > 1$, we still have $b>0$.
For any $n\in \NN$, we have
$$ a^n = \bigl((\sqrt a)^2\bigr)^n = (1+b)^{2n} = \bigl((1+b)^n\bigr)^2 \ge (1+nb)^2
> n^2 b^2.$$
Now let $\eps > 0$ be given and choose $N\in\NN$ with $N > 1/(\eps b^2)$. Then
$$ \biggl| {n\over a^n} - 0\biggr| = {n\over a^n} < {n\over n^2 b^2} \le {1\over nb^2}
\le {1\over Nb^2} <\eps$$
for all $n\ge N$. Hence $n/a^n\to 0$.\slug

\proclaim Problem \advthm. Let $x_n$ be a sequence that converges to some limit $x$. For each
$n\in \NN$, let $a_n$ be the average
$$a_n = {1\over n} \sum_{i=1}^n x_i.$$
Show that $a_n$ converges to $x$ as well.

\proof Note that
$$\eqalign{
|a_n - x| &= \biggl| {1\over n}\sum_{i=1}^n x_i  - x\biggr|
= \biggl| {1\over n} \Bigl(\sum_{i=1}^n x_i - nx\Bigr)\biggr|
= \biggl| {1\over n} \Bigl(\sum_{i=1}^n (x_i - x)\Bigr)\biggr|
\le {1\over n}\sum_{i=1}^n |x_i - x|,
}$$
where in the last line we used the triangle inequality.

Let $\eps > 0$ be given. Since $x_n$ converges to $x$, there exists $N_1 \in \NN$ such that
for all $n\ge N_1$, $|x_n - x| <\eps/2$. This is how we will deal with the terms $|x_i - x|$ for
$i\ge N_1$. But we cannot individually bound the terms $|x_i - x|$ when $i < N_1$. But the sum
$$S = \sum_{i=1}^{N_1-1} |x_i - x|.$$
is just a (possibly very large) finite number. Let $N_2 \in \NN$ be large enough to satisfy
$N_2 \ge 2S/\eps$. Finally, let $N = \max\{N_1, N_2\}$. Then for any $n\ge N$ we see that
$$\eqalign{
|a_n - x| &\le {1\over n}\sum_{i=1}^{N_1 - 1} |x_i - x| + {1\over n} \sum_{i=N_1}^n |x_i - x| \cr
&\le {S\over n} + {1\over n} \sum_{i=N_1}^n {\eps\over 2} \cr
&\le {S\over N_2} + {n-N_1 + 1\over n}\cdot {\eps\over 2} \cr
&\le {\eps \over 2} + {\eps\over 2} \cr
&= \eps. \cr
}$$
This shows that $a_n \to x$ as $n\to\infty$.\slug

\bye

### 2025-10-23 teaching/math242f25/tut6.tex
tut6

\input fontmac
\input mathmac

\def\To{\Rightarrow}

\maketitle{Math 242 Tutorial 6}{prepared by}{Marcel Goh}{23 October 2025}

\bigskip

\proclaim Problem \advthm. Determine (with proof) the limits of
\medskip
\item{a)} $n^{1/n^2}$; and
\smallskip
\item{b)} $(n!)^{1/n^2}$
\medskip
\noindent as $n\to \infty$. You may use the fact that $\root n\of n \to 1$ as $n\to \infty$; this you
shall prove on your next assignment.

\proof Note that for all $n\in \NN$, $n\ge 1 = 1^{n^2}$, so we have $n^{1/n^2} \ge 1$.
On the other hand, since $n\le n^n$, taking $n^2$th roots we have $n^{1/n^2} \le \root n \of n$. But
$\root n\of n\to 1$, so by the squeeze theorem $n^{1/n^2}\to 1$ too, as $n\to\infty$.

For part (b), observe first that for all $n\in \NN$, $n! \ge 1\ge 1^{n^2}$, so $(n!)^{1/n^2}\ge 1$.
On the other hand, $n! = 1\cdot 2\cdot 3\cdots n \le n^n$, so taking $n^2$th roots again we have
$(n!)^{1/n^2} \le \root n \of n$. Hence
by the squeeze theorem, we see that $(n!)^{1/n^2} \to 1$.\slug

\proclaim Problem \advthm. Let $(a_n)$ be a sequence with $\lim_{n\to\infty} a_n = a$,
and let $(b_n)$ be a sequence with $b_n\ne 0$ for all $n\in \NN$ and $\lim_{n\to\infty} b_n = b\ne 0$.
\medskip
\item{a)} Prove that $\lim_{n\to\infty} 1/b^n = 1/b$.
\smallskip
\item{b)} Conclude that $\lim_{n\to\infty} (a_n/b_n) = a/b$.
\medskip

\proof Let $\eps > 0$. Since $b_n\to b\ne 0$, there exists $N_1\in \NN$ such that
$$|b-b_n|<{|b|^2\eps\over 2}$$
for all $n\ge N_1$. For the same reason, there exists $N_2\in \NN$ such that
$$|b-b_n| < {|b|\over 2}$$
for all $n\ge N_2$. This second bound implies that for all $n\ge N_2$
$$|b_n| > {|b|\over 2}.$$
So, letting $N = \max\{N_1, N_2\}$, we have
$$\biggl| {1\over b_n} - {1\over b}\biggr|
= {|b-b_n|\over |b_n|\cdot |b|} < {2\over |b|^2}\cdot {|b|^2\eps\over 2} = \eps.$$
We see, then, that $1/b_n\to 1/b$ as $n\to \infty$, since $\eps$ was arbitrary.

For part (b), simply let $c_n = 1/b_n$, with limit $c = 1/b$. By the product law for limits (shown in class),
$$\lim_{n\to\infty} {a_n\over b_n} = \lim_{n\to\infty} a_nc_n = ac = {a\over b},$$
and we are done.
\slug

\proclaim Problem \advthm. Let $(a_n)$ be a sequence, let $(b_n)$ be a null sequence, and let $L\in\RR$.
Show that if there exists $K\in \NN$ such that $|a_n - L|\le |b_n|$ for all $n\ge K$,
then $(a_n)$ converges to $L$.

\proof Let $\eps > 0$. Since $(b_n)$ is a null sequence, there is some $N_1\in \NN$
such that for all $n\ge N_1$ we have $|b_n| = |b_n - 0| < \eps$. Let $N = \max\{N_1, K\}$.
Then for all $n\ge N$, we have
$$|a_n - L|\le |b_n| <\eps,$$
so $a_n\to L$ as $n\to\infty$.\slug
\goodbreak

\proclaim Problem \advthm. Let $(x_n)$ be a sequence of nonnegative real numbers that is
{\it subadditive} in the sense that for all $m,n\in \NN$,
$$x_{m+n} \le x_m + x_n.$$
\medskip
\item{a)} Prove that $x_{qk} \le qx_k$ for all integers $k,q\in \NN$.
\smallskip
\item{b)} Explain why $\inf\{ x_n / n : n\in \NN\}$ exists.
\smallskip
\item{c)} Letting $L = \inf\{x_n/ n : n\in \NN\}$, show that $x_n/n \to L$ as $n\to\infty$.
\medskip

\proof For part (a), we perform induction on $q$. The base case $q=1$ is trivial. For $q\ge 1$,
suppose that $n=(q+1)k+r$. Then
$$x_{(q+1)k} = x_{qk + k} \le x_{qk} + x_k \le qx_k + x_k = (q+1)x_k,$$
where the first inequality is a consequence of subadditivity, and the second inequality follows
from the induction hypothesis.

For part (b), we simply note that $x_n\ge 0$ for all $n\in\NN$. Hence $x_n/n\ge 0$ for all $n\in \NN$,
and the infimum exists by completeness of the reals.

Part (c) is where the real work comes in. Let $\eps > 0$.
Since $L$ is the infimum of the set $\{x_n/n : n\in \NN\}$, there exists some $k\in \NN$ such that
$${x_k\over k} \le L+{\eps\over 2}.$$
Let $M = \max\{x_1,x_2,\ldots,x_{k-1}\}$, and pick $N= \max\{k, 2M/\eps\}$. Now let $n\ge N$ be given.
Since $n\ge k$, by the division algorithm we can write $n = qk+r$ for some integers $q\ge 1$ and
$r\in \{0,\ldots,k-1\}$. Then we have
$$x_n = x_{qk+r} \le x_{qk} + x_r \le qx_k + x_r,$$
by subadditivity and part (a). (Technically we didn't define $x_0$, but we can just say $x_0 = 0$,
and the above holds.) This implies that
$${x_n \over n} - {x_k\over k}
\le {qx_k + x_r\over n} - {x_k\over k}
= x_k \biggl({q\over kq+r}-{1\over k}\biggr) + {x_r\over n}.
$$
But since $r\in \{0,\ldots,k-1\}$, we have $x_r \le M$, by definition of $M$. This,
along with our assumption that $n\ge 2M/\eps$, implies that
$${x_r\over n} \le {M\over n} < {M\over 2M/\eps} = {\eps\over 2}.$$
Furthermore, we have
$${q\over qk+r} \le {q\over qk} = {1\over k},$$
so plugging these inequalities in above, we see that
$${x_n\over n} - {x_k\over k} < {\eps\over 2}.$$
Lastly, we bound
$$
\biggl| {x_n\over n} - L\biggr| = {x_n\over n} - L
= {x_n\over n} - {x_k\over k} + {x_k\over k} - L
< {\eps\over 2} + {\eps\over 2}
= \eps,$$
which completes the proof that $x_n/n\to L$ as $n\to\infty$.

\bye

### 2025-10-31 teaching/math242f25/tut7.tex
tut7

\input fontmac
\input mathmac

\def\To{\Rightarrow}

\maketitle{Math 242 Tutorial 7}{prepared by}{Marcel Goh}{30 October 2025}

\bigskip

\proclaim Problem \advthm. Let $x_1 > 2$ and $x_{n+1} = 1+\sqrt{x_n - 1}$ for all $n\in \NN$.
\medskip
\item{a)} Show that $(x_n)$ is decreasing and bounded from below.
\smallskip
\item{b)} Find the limit of $(x_n)$.
\medskip

\proof First we show that $x_n\ge 2$ for all $n\in\NN$. The case $n=1$ is
given as a hypothesis. Moreover, if we assume that $x_n\ge 2$ for some $n\in \NN$, then $x_n-1 \ge 1$
and $\sqrt{x_n-1}\ge \sqrt 1 = 1$ as well. Hence
$$x_{n+1} = 1 + \sqrt{x_n - 1} \ge 1 + 1 = 2,$$
and the result holds for all $n\in \NN$ by induction.

Recall that $\sqrt u \le u$ for all $u\ge 1$ (this is immediate from the inequality $u\le u\cdot u$).
Let $n\in \NN$. We showed in the previous paragraph that $x_n - 1 \ge 1$, so we have
$\sqrt{x_n - 1} \le x_n - 1$. But then
$$ x_{n+1} = 1+ \sqrt{x_n -1} \le x_n,$$
so we see that the sequence $(x_n)$ is decreasing. We have settled part (a).

For part (b), we claim that for all $n\ge \NN$,
$$0\le x_n - 2 \le {x_1 - 2\over 2^{n-1}}.$$
The first inequality was already shown in part (a). The second inequality we shall
prove by induction on $n$. It clearly holds (with equality) for $n=1$. Then for $n\ge 1$,
we have
$$
x_{n+1} - 2 = \sqrt{x_n -1} - 1 
= {(x_n - 1) - 1\over (x_n-1)+1} 
= {x_n - 2\over x_n}
\le {x_n - 2\over 2}
\le {x_1 - 2\over 2\cdot 2^{n-1}}
= {x_1 - 2\over 2^n},
$$
where the first inequality we use is $x_n\ge 2$ and the second inequality is the induction hypothesis.
Hence the claim holds for all $n\in \NN$.

From here it's easy to show that $x_n \to 2$. Let $\eps > 0$ and let $N\in \NN$
satisfy $N>\log_2\bigl((x_1 - 2)/\eps\bigr)$, which we can do
by the Archimedean property. Then for all $n\ge N$,
we have
$$ |x_n - 2| \le {x_1 - 2\over 2^n} \le {x_1 - 2\over 2^{\log_2((x_1-2)/\eps)}}
= {x_1 - 2\over (x_1 - 2)/\eps} = \eps,$$
which is what we wanted to show.\slug

Consider the sequence $y_n = x_n - 1$, so that we have $y_n > 1$ and $y_{n+1} = \sqrt{y_n}$ for all
$n$. The exercise above showed that $\lim y_n = 1$, which explains why if you start with a number greater
than $1$ in your calculator and then hit the square root button over and over, you end up getting closer
and closer to $1$.

\proclaim Problem \advthm. Determine the limits of the following sequences (the monotone convergence
can be used to prove that each of these series converges, but we will gloss over this in this
exercise).
\medskip
\item{a)} $\bigl((1+1/n)^{n+1}\bigr)$
\smallskip
\item{b)} $\bigl((1+1/(n+1))^n\bigr)$
\smallskip
\item{c)} $\bigl((1+1/n)^{2n}\bigr)$
\smallskip
\item{d)} $\bigl((1-1/n)^{n}\bigr)$
\medskip

\proof For part (a), note that
$$ \lim_{n\to\infty} \biggl(1+{1\over n}\biggr)^{n+1}
= \lim_{n\to\infty} \biggl(1+{1\over n}\biggr)^n \cdot \biggl(1+{1\over n}\biggr)
= \lim_{n\to\infty} \biggl(1+{1\over n}\biggr)^n \cdot \lim_{n\to\infty} \biggl(1+{1\over n}\biggr)
= e\cdot 1 = e,$$
by the product law for limits. Likewise, for part (b) we have
$$ \lim_{n\to\infty} \biggl(1+{1\over n+1}\biggr)^n
= \lim_{n\to\infty} \biggl(1+{1\over n+1}\biggr)^{n+1} \bigg/ \biggl(1+{1\over n+1}\biggr)
= \lim_{n\to\infty} \biggl(1+{1\over n+1}\biggr)^{n+1} \bigg/ \lim_{n\to\infty} \biggl(1+{1\over n+1}\biggr)
= e.$$
For part (c) we use the product law again to find that
$$\lim_{n\to\infty} \biggl(1+{1\over n}\biggr)^{2n}
= \lim_{n\to\infty} \biggl(1+{1\over n}\biggr)\cdot \biggl(1+{1\over n}\biggr)
= \lim_{n\to\infty} \biggl(1+{1\over n}\biggr)\cdot \lim_{n\to\infty} \biggl(1+{1\over n}\biggr)
= e\cdot e = e^2.$$
Lastly, note that for all $n\in \NN$,
$$\biggl(1-{1\over n}\biggr)^{-n} = \biggl({n-1\over n}\biggr)^{-n}
= \biggl({n\over n-1}\biggr)^n = \biggl(1 + {1\over n-1}\biggr)^n.$$
So
$$\lim_{n\to\infty} \biggl(1-{1\over n}\biggr)^{-n}
= \lim_{n\to\infty} \biggl(1 + {1\over n-1}\biggr)
= \lim_{n\to\infty} \biggl(1 + {1\over n}\biggr)^{n+1} = e,$$
by part (a). Then by the reciprocal law (proved last tutorial), we must have
$$\lim_{n\to\infty} \biggl(1-{1\over n}\biggr)^{-n} = 1/e,$$
settling part (d).\slug

\proclaim Problem \advthm. Using the monotone convergence theorem (and without using the least
upper bound property), prove the Archimedean property as well as the nested interval property.

\proof First we prove the Archimedean property. Suppose, towards a contradiction, that there exists
some $M\in \RR$ such that $n \le M$ for all $n\in \NN$. Consider $\NN$ as a sequence $(x_n)$, where
$x_n = n$ for all $n\in \NN$. Then we have $x_{n+1} = 1+x_n$ for all $n\in \NN$, so $(x_n)$
is monotone increasing, and $x_n$ is bounded above by $M$. By the monotone convergence theorem,
$x_n$ converges to some limit, call it $L$. This $L$ satisfies
$$L = \lim_{n\to\infty} x_{n+1} = \lim_{n\to\infty} (x_n + 1) = 1+ \lim_{n\to\infty} x_n = 1+L,$$
so subtracting $L$ from both sides, we arrive at the nonsense statement $0=1$. We conclude that the
Archimedean property is true.

Now we prove the nested interval property. Let $I_n = [a_n, b_n]$ (where $a_n\le b_n$ for all $n\in \NN$)
be a sequence of intervals with
$$I_1\supseteq I_2 \supseteq I_3 \supseteq \cdots.$$
Since $[a_n,b_n] \supseteq [a_{n+1}, b_{n+1}]$ for all $n\in \NN$, the sequence $(a_n)$ is monotone
nondecreasing, and for all $n\in \NN$ one has $a_n\le b_n \le b_1$, so this sequence is also
bounded from above. By the monotone convergence theorem, $(a_n)$ converges to a limit, call it $x$.

First we show that $x\ge a_n$ for all $n\in \NN$. Suppose, for a contradiction, that there is
some $k\in \NN$ such that $a_k > x$. Let $\eps = a_k - x> 0$, and by the definition of convergence
there is some $N\in \NN$ such that for all $n\ge N$, $|a_n - x| < \eps$. But now, since $a_n$
is monotone nondecreasing, choosing any $n\ge \max\{k+1,N\}$, we have
$$\eps > |a_n - x| \ge a_n - x \ge a_k - x = \eps,$$
which is a contradiction.

Next we show that $x\le b_n$ for all $n\in \NN$. Suppose, for a contradiction, that there is
some $k\in \NN$ such that $b_k < x$. Let $\eps = (x-b_k)/2$, and by the definition of convergence
there is some $N\in \NN$ such that for all $n\ge N$, $|a_n - x| < \eps$. Now, letting
$n\ge \max\{k, N\}$, we have
$$a_n > x-\eps > x-2\eps = b_k \ge b_n,$$
which contradicts our assumption that $a_n\le b_n$.

We have shown that $x\in [a_n,b_n]$ for all $n\in \NN$, so
$$x\in \bigcup_{n\in \NN} I_n,$$
and our proof of the nested interval theorem is complete.\slug

This, combined with Problem~2 of our tutorial on 2~October, shows that the monotone convergence
theorem is logically equivalent to the least upper bound property. (And the nested interval
property combined with the Archimedean property are together equivalent to either of these
as well.)

\bye

### 2025-11-07 teaching/math242f25/tut8.tex
tut8

\input fontmac
\input mathmac

\def\To{\Rightarrow}

\maketitle{Math 242 Tutorial 8}{prepared by}{Marcel Goh}{6 November 2025}

\bigskip

\proclaim Problem \advthm. Prove that the Bolzano--Weierstrass theorem implies the monotone
convergence theorem (without using the least upper bound property or anything else we have
shown to be equivalent to it).

\proof Let $(x_n)$ be a monotone nondecreasing sequence that is bounded above; that is, there
is $M\in \RR$ such that $x_n \le M$ for all $n\in \NN$. Then $x_n\in [x_1,M]$ for all
$n\in \NN$ and by the Bolzano--Weierstrass theorem, there is a subsequence $(x_{n_k})$ of
$(x_n)$ as well as some $x\in \RR$ such that $x_{n_k}\to x$ as $k\to\infty$.

First we claim that $x_n \le x$ for all $n\in \NN$. If not, then there is some $r\in \NN$
with $x_r > x$, and letting $\eps = x_r - x$, we see there must be some
$K\in \NN$ such that $\bigl|x_{n_k} - x\bigr| \le \eps$ for all $k\ge K$. Now pick $k\ge K$
large enough so that $r\le n_k$ (this we can do because $n_k$ is an infinite sequence).
For this choice of $k$ we have
$$\eqalign{
x_r - x_{n_k} &= x_{n_r} - x - (x_{n_k} - x)
= \eps - (x_{n_k} - x)
\ge \eps - |x_{n_k} - x|
> \eps - \eps 
= 0,
}$$
which means that $x_r > x_{n_k}$, contradicting the monotonicity of $(x_n)$.

Now we show that $(x_n)$ also converges to $x$.
Let $\eps > 0$. Since the subsequence $(x_{n_k})$ converges to x, there is some $K\in \NN$ such that
for all $k\ge K$,
$$\bigl| x_{n_k} - x\bigr| < \eps.$$
We choose $N = n_k$. Then for all $n\ge N$, we have
$$ x - \eps < x_{n_k} = x_N \le x_n \le x,$$
where $x_N \le x_n$ follows from monotonicity and the last inequality was
proved in the previous paragraph.
Hence in particular, $|x_n - x| < \eps$.\slug

\proclaim Problem \advthm. Cauchy's criterion asserts that every Cauchy sequence converges to a limit.
Assume only Cauchy's criterion and the Archimedean property, and prove the Bolzano--Weierstrass
theorem. (Again, do not resort to the least upper bound property or anything else that we have
shown is equivalent to it.)

\proof Let $(x_n)$ be a sequence that is bounded; that is, there are $a,b\in \RR$ with $a\le b$
such that
$x_n\in [a, n]$ for all $n\in \NN$. We shall build a sequence $(x_{n_k})$ recursively.
For the base case, let
$a_1 = a$ and $b_1 = b$. The interval $[a_1, b_1]$ has infinitely many $x_n$ residing in it.
Let $n_1$ be any integer with $x_{n_1}\in [a_1, b_1]$.

Now for $k\ge 1$, assume that the interval $[a_k, b_k]$ has infinitely many $x_n$ in it. Let
$c_k = (a_k + b_k)/2$ be the midpoint of this interval, and note that (at least) one of
the two intervals $[a_k, c_k]$ and $[c_k, b_k]$ contain infinitely many elements of $(x_n)$.
If it is the first, then let $a_{k+1} = a_k$ and $b_{k+1} = c_k$; otherwise,
let $a_{k+1} = c_k$ and let $b_{k+1} = b_k$. Thus $[a_{k+1}, b_{k+1}]$ contains infinitely
many members of $(x_n)$, and because of this, we can pick $n_{k+1} > n_k$ such that
$x_{n_{k+1}}\in [a_{k+1}, b_{k+1}]$.

It is easy to show by induction that $b_k - a_k \le (b-a)/2^{k-1}$ for all $k\in \NN$.
From this, we can prove
that the sequence $(x_{n_k})$ is Cauchy. Let $\eps > 0$ and pick $K > \log_2 \bigl( (b-a)/\eps\bigr) +1$
by the Archimedean property. Then for any $i,j\ge K$, both $x_{n_i}$
and $x_{n_j}$ are in the interval $[a_K, b_K]$, and consequently
$$\bigl| x_{n_i} - x_{n_j}\bigr| \le b_K - a_K \le {b-a\over 2^{K-1}} < \eps,$$
by our choice of $K$.

But by Cauchy's criterion, every Cauchy sequence is convergent. Hence the subsequence
$(x_{n_k})$ is convergent. This proves the Bolzano--Weierstrass theorem.\slug

\medskip\boldlabel Divergence to $\mathbold\infty$
or $\mathbold-\infty$. We now set up some definitions for the remainder of the tutorial.
We shall say that a sequence $(x_n)$ {\it diverges to $\infty$} if for every
$M\in \RR$ there exists some $N\in \NN$ such that $x_n > M$ for all $n\ge N$.
We say that a sequence $(x_n)$ {\it diverges to $-\infty$} if for all $M\in \RR$ there exists
some $N\in \NN$ such that $x_n < M$ for all $n\ge N$.A

Note that ``diverging to $\infty$'' or ``diverging to $-\infty$'' is not the same as just
``not converging to any $x\in \RR$''. For example, the sequence
$$0,-1,1,0,-2,2,0,-3,3,\ldots$$
does not converge to any $x\in \RR$, nor does it converge to either $-\infty$ or $\infty$. However,
the following result about subsequences is true.

\proclaim Problem \advthm.
Show that for any sequence $(x_n)$ of real numbers, at least one of the following statements holds.
\medskip
\item{a)} There is a subsequence $(x_{n_k})$ of $(x_n)$ that diverges to $\infty$.
\smallskip
\item{b)} There is a subsequence $(x_{n_k})$ of $(x_n)$ that diverges to $-\infty$.
\smallskip
\item{c)} There is some $x\in \RR$ and a subsequence $(x_{n_i})$ of $(x_n)$ such that
$(x_{n_i})$ converges to $x$ as $i\to\infty$.
\medskip

\proof Let $(x_n)$ be any sequence of real numbers.
If (a) holds we are done. If (b) holds we are done. So we assume that (a) and (b)
are both false, and our goal is to prove (c).

Since (a) is false, the sequence $(x_n)$ does not diverge to $\infty$. Hence there exists
$b \in \NN$ such that for all $N\in \NN$, there exists $n\ge N$ with $x_n \le b$.
Let $n_1\ge 1$ such that $x_n\le b$, then let $n_2 > n_1$ such that $x_2\le b$, and so on.
In this manner, we obtain a subsequence $(x_{n_k})$ that is bounded from above by $b$.

Since (b) is false, the sequence $(x_{n_k})$ does not diverge to $-\infty$. So there
is $a\in \NN$ such that for all $K\in \NN$ there is always some $k\ge K$ such that
$x_{n_k} \ge a$. By logic similar to that in the previous paragraph, we see there is a subsequence
$(x_{n_j})$ of $(x_{n_k})$ that is bounded from below by $a$, and hence contained in the
interval $[a,b]$. (We have ``flattened'' out the notation here by simply redefining $n_j = n_{k_j}$.
If we want to be extra precise
with notation, we should have written $(x_{n_{k_j}})$,
since this is a subsequence of $(x_{n_k})$, but this kind of pedantry only creates unreadable nonsense
when we take subsequences of subsequences of subsequences, as in this proof.)

We have found that there is a subsequence $(x_{n_j})$ of $(x_n)$ that is contained in the interval
$[a,b]$. Hence by the Bolzano--Weierstrass theorem, there is a subsequence $(x_{n_i})$
of $(x_{n_j})$ that converges to some limit $x\in \RR$. Of course, $(x_{n_i})$ is a subsequence
of our original sequence $(x_n)$, so we are done.\slug

This proof shows that $\RR\cup \{\infty, -\infty\}$ is a valid {\it compactification} of $\RR$.
This terminology will make more sense once you learn more about topology.

\proclaim Problem 4. Show that $(a^n)$ diverges to $\infty$ for any $a>1$.

\proof Let $M\in \RR$ be given. If $M\le 1$, then since $a^n\ge a>1$ for all $n\in \NN$,
we can simply pick $N=1$ and $a_n> M$ for all $n\ge N$. If $M>1$, then
we can pick $N\in \NN$ satisfying $N>\log_a M > 0$, by using the Archimedean property.
Then for all $n\ge N$, we have
$$a^n \ge a^N > a^{\log_a M} = M.$$
Since $M$ was arbitrary, we see that $(a^n)$ diverges to $\infty$.\slug

\bye

### 2025-11-14 teaching/math242f25/tut9.tex
tut9

\input fontmac
\input mathmac

\def\To{\Rightarrow}
\def\s{\mathord{*}}

\maketitle{Math 242 Tutorial 9}{prepared by}{Marcel Goh}{13 November 2025}

\bigskip

\proclaim Problem \advthm. Let $A\subseteq \RR$. Show that the boundary of $A$ is equal
to the boundary of the complement of $A$.

\proof Simply note that
$$\eqalign{
\partial A
&= \bigl\{ x\in \RR : \hbox{for all}\ \eps>0,
\ V_\eps(x)\cap A\ne\emptyset\ \hbox{and}\ V_\eps(x)\cap A^c\ne\emptyset\bigr\} \cr
&= \bigl\{ x\in \RR : \hbox{for all}\ \eps>0,
\ V_\eps(x)\cap (A^c)^c \ne\emptyset\ \hbox{and}\ V_\eps(x)\cap A^c\ne\emptyset\bigr\} \cr
&= \partial(A^c). \noskipslug\cr
}$$

\proclaim Problem \advthm. Prove the following:
\medskip
\item{a)} A set $U\subseteq \RR$ is open if and only if $U$ doesn't contain any of
its boundary points.
\smallskip
\item{b)} A set $A\subseteq \RR$ is closed if and only if $A$ contains all of its
boundary points.
\medskip

\proof Suppose that $U$ contains a boundary point $x\in \partial U\cap U$. Then
for this choice of $x\in U$, for any $\eps > 0$, $V_\eps(x)\cap U^c \ne\emptyset$.
This means that $V_\eps(x)\not\subseteq U$ for all $\eps>0$. So $U$ is not open.
Conversely, if $U$ is not open, there is some $x\in U$ such that for all
$\eps > 0$, $V_\eps(x)\not\subseteq U$, meaning that $V_\eps(x)\cap U^c \ne\emptyset$.
We also see that $V_\eps(x)\cap U$ contains $x$, so it is also nonempty.
This means that $x$ is a boundary point of $U$; that is, $x$ contains at least one
of its boundary points.

By the previous paragraph, $A^c$ is open if and only if $A^c$ doesn't
contain any boundary points of $A^c$, and by the previous paragraph
this is true if and only if it doesn't contain
any boundary points of $A$.\slug

\proclaim Problem \advthm. Let $\emptyset \ne S\subseteq \RR$. Prove the following.
\medskip
\item{a)} If $S$ is bounded from above, then $\sup S$ is a boundary point of $S$.
\smallskip
\item{b)} If $S$ is bounded from below, then $\inf S$ is a boundary point of $S$.
\medskip

\proof Let $\eps > 0$. Since $\sup S - \eps$ is not an upper bound of $S$, there must
exist some $s\in S$ with $s>\sup S-\eps$. Since $\sup S$ is an upper bound of $S$,
we find that
$$s\in S\cap (\sup S-\eps, \sup S] \subseteq S \cap V_\eps(\sup S).$$
Also, since
$\sup S$ is an upper bound of $S$, the real number $t = \sup S+\eps/2$ is not a member of $S$.
So
$$t\in S^c \cap (\sup S, \sup S+\eps) \subseteq S^c \cap V_\eps(\sup S).$$

The proof of part (b) is similar.\slug

\medskip\boldlabel The Cantor set.
For any scalar $a\in \RR$ and set $S\subseteq \RR$, define the {\it dilate}
$$a\cdot S = \{ a\cdot s : s\in S\}$$
as well as the {\it translate}
$$a+\cdot S = \{a + s : s\in S\}.$$
For integers $n\ge 0$, define $C_n$ recursively by
$C_0 = [0,1]$ and
$$ C_{n+1} = {1\over 3}\cdot C_n \cup \biggl( {2\over 3} + {1\over 3}\cdot C_n\biggr)$$
for $n\ge 0$.
(In other words, $C_{n+1}$ is obtained by removing the ``open middle third'' from $C_n$.)
The {\it Cantor set} $C$ is defined to be $C = \bigcap_{n=0}^\infty C_n$.

\proclaim Problem \advthm. Prove the following.
\medskip
\item{a)} The Cantor set $C$ is closed.
\smallskip
\item{b)} If $a,b\in \RR$ with $a\le b$ are such that $[a,b]\subseteq C$, then $a=b$.
(Hence the only closed intervals in $C$ are singleton sets $\{a\}$.)
\smallskip
\item{c)} The cardinality of $C$ is equal to that of $[0,1]$. (That is, $C$ is uncountable.)
\medskip

\proof Note that $C_0$ is just a closed interval, so {\it a fortiori} it is a finite union
of closed intervals. Now assume that $C_n$ is a finite union of closed intervals. Then
$(1/3)\cdot C_n$ is also a finite union of (the same number of) closed intervals,
and so is $(2/3)+(1/3) C_n$. Hence $C_{n+1}$ is a finite union of (twice as many) closed intervals.
We see by induction that $C_n$ is closed for all $n\ge 0$, and hence $C = \bigcap_{n=0}^\infty$
is closed, being an intersection of closed sets.

For part (b), first observe that $C_0$ is an interval of length $1 = 1/3^0$. Next, assume
that $C_n$ is composed of (a finite number of) pairwise disjoint
intervals, each of length $1/3^n$. Then
$C_{n+1}$ is obtained by taking the disjoint union of two copies of $C_n$, each dilated by $1/3$, so
we conclude that $C_{n+1}$ consists of pairwise disjoint
intervals of length $1/3^{n+1}$. By induction, for all
integers $n\ge 0$, $C_n$ is composed of disjoint intervals of length $1/3^n$. Now, given $a,b\in \RR$
with $a\le b$. If $a<b$, then $b-a > 0$, so we can pick $N > 1/(b-a)$ by the Archimedean property.
Then $C_N$ is composed of disjoint intervals of length $1/3^N < b-a$. We conclude that
$[a,b]$ is not a subset of $C^N$, hence it cannot be contained in $C$.

Lastly, we show that $C$ is uncountable. Note that every real number $x\in [0,1]$ can
be written as
$$x = \sum_{n=1}^\infty {t_n\over 3^n},$$
where $t_n\in \{0,1,2\}$ for all $n\in \NN$.
This is the representation of
$$x = 0.t_1t_2t_3\ldots$$
in base-$3$ (ternary) notation. Ternary representations aren't unique. For instance,
$0.1 = 0.22222\ldots$, just like $0.1 = 0.999999\ldots$ in decimal notation.
In ternary notation, $0.022222\ldots = 1/3$ and $0.2 = 2/3$, so the
first middle third we removed (in going from $C_0$ to $C_1$) contains all numbers of the form
$0.1\s\s\s\ldots$, where the $\s\s\s\ldots$ part can be anything strictly between $000\ldots$
and $222\ldots$. That is, $C_1$ contains all numbers of the form $0.0\s\s\s\ldots$ and
$0.2\s\s\s\ldots$, with no restriction whatsoever on the digits $\s\s\s\ldots$.

In going from $C_1$ to $C_2$, we remove the middle third of $[0,1/3]$ as well as the middle third
of $[2/3,1]$. By the same logic we applied above, we see that $C_2$ contains all numbers
of the form $0.00\s\s\s\ldots$, $0.02\s\s\s\ldots$, $0.20\s\s\s\ldots$, or $0.22\s\s\s\ldots$, with no
restrictions on the digits $\s\s\s\ldots$. By induction, we find every $x\in C$ can be expressed
as
$$x = \sum_{n=1}^\infty {t_n\over 3^n}$$
where $t_n\in \{0,2\}$ for all $n\in \NN$.

Let $f : C\to [0,1]$ map
$$ x = \sum_{n=1}^\infty {t_n \over 3^n},$$
where $t_n\in \{0,2\}$ for all $n\in \NN$, to
$$ f(x) = \sum_{n=1}^\infty {b_n\over 2^n},$$
where we have set
$$b_n = \cases{0, & if $t_n = 0$;\cr 1, & if $t_n = 2$.}$$
We claim that $f$ is surjective. Every $y\in [0,1]$ has at least one binary representation
$$ y = \sum_{n=1}^\infty {b_n\over 2^n},$$
so letting
$$t_n = \cases{0, & if $b_n = 0$;\cr 2, & if $b_n = 1$}$$
and
$$ x = \sum_{n=1}^\infty {t_n \over 3^n},$$
we find that $f(x) = y$. The fact that there is a surjective function $f:C\to [0,1]$ means that
the cardinality of $C$ is at least that of $[0,1]$. But $C\subseteq [0,1]$, so its cardinality is
at most that of $[0,1]$. Hence $|C| = \bigl|[0,1]\bigr|$, and we are done.\slug

\bye

### 2025-11-21 about/vita.tex
new paper and updated 242 tut2
tut10 and other updates

@@
 
-\sectheader{Papers}
+\sectheader{Papers {\rm(in reverse order of acceptance, not necessarily appearance)}}
 
 \begingroup\frenchspacing
-\papercount=8
+\papercount=9
 \pubbegin P
+\papitem (with Hamed Hatami)
+Block complexity and idempotent Schur multipliers.
+To appear in {\sl International Mathematics Research Notices}.
+[arXiv:2506.21752]
 \papitem (with Tsun-Ming Cheung and Luc Devroye)
 A note on plane trees with decreasing labels.
-To appear in {\sl Online Journal of Analytic Combinatorics}.
+{\sl Online Journal of Analytic Combinatorics}\/ {\bf 20} (2025), \#2, 9~pp.
 [arXiv:2502.14596]
@@
 {\sl arXiv:2507.00872} (2025), 14~pp.
-\smallskip
-\thing (with Hamed Hatami)
-Block complexity and idempotent Schur multipliers.
-{\sl arXiv:2506.21752} (2025), 17~pp.
 \smallskip

### 2025-11-21 teaching/math242f25/tut10.tex
tut10 and other updates

\input fontmac
\input mathmac

\def\To{\Rightarrow}

\maketitle{Math 242 Tutorial 10}{prepared by}{Marcel Goh}{20 November 2025}

\bigskip

\proclaim Problem \advthm. Let $p\in \RR$ be a point contained in some interval $I$.
Suppose that $f$ and $g$ are functions defined on $I\setminus\{p\}$ with
$\lim_{x\to p} f(x) = L$ and $\lim_{x\to p} g(x) = M$ for some $L,M\in \RR$. Prove that
\medskip
\item{a)} $\lim_{x\to p} C\cdot f(x) = CL$ for all $C\in \RR$;
\smallskip
\item{b)} $\bigl( \lim_{x\to p} f(x) \bigr) \bigl(\lim_{x\to p} g(x)\bigr) = LM$; and
\smallskip
\item{c)} $\bigl( \lim_{x\to p} f(x) \bigr)\big/\bigl(\lim_{x\to p} g(x)\bigr) = L/M$,
so long as $M\ne 0$ and $g(x)\ne 0$ for all $x\in I\setminus \{p\}$.

\proof Let $x_n$ be a sequence in $I\setminus \{p\}$ that converges to $p$. Then
$f(x_n)\to L$ as $n\to\infty$, and $g(x_n)\to M$ as $n\to \infty$.
This whole question can now be solved using the laws that govern limits of sequences.

For part (a), we use the fact that
$$\lim_{n\to\infty} c\cdot f(x_n) = cL.$$
For part (b), observe that
$$\Bigl(\lim_{n\to\infty} f(x_n)\Bigr)\Bigl(\lim_{n\to\infty} g(x_n)\Bigr)
= \lim_{n\to\infty} f(x_n) g(x_n) = LM.$$
Lastly, for part (c) we have
$$\lim_{n\to\infty} f(x_n)\Big/\lim_{n\to\infty} g(x_n)
= \lim_{n\to\infty} {f(x_n) \over g(x_n)} = {L\over M}.$$

All three parts now follow from the sequential definition of the limit, as well as
the fact that $x_n$ was taken to be an arbitrary sequence.\slug

\proclaim Problem \advthm. Let $f:\RR\to\RR$ be defined by
$$f(x) = \cases{1,& {\rm if $x\in \QQ$;}\cr 0,& {\rm if $x\in \RR\setminus\QQ$.}}$$
(This function is called the {\it Dirichlet function}.) Show that for any $c\in \RR$,
the limit $\lim_{x\to c} f(x)$ does not exist.

\proof Let $c\in \RR$ be given. By the density of $\QQ$ in $\RR$, for any $n\in \NN$
there are infinitely many rational numbers in the ball $V_{1/n}(c)$. Hence there
are also infinitely many rational numbers in the punctured ball $V^*_{1/n}(c)$. Hence
we may construct a sequence $(r_n)$ in $\QQ\setminus\{c\}$ with $|r_n - c| < 1/n$
for all $n\in \NN$. This sequence converges to $c$.

But $\RR\setminus \QQ$ is also dense in $\RR$, so by the same logic as in the previous
paragraph there is a sequence $(s_n)$ in $(\RR\setminus \QQ)\setminus\{c\}$ that converges
to $c$. Now observe that $\lim_{n\to\infty} f(r_n) = 1$ and $\lim_{n\to\infty} f(s_n) = 0$,
so $f(x)$ cannot have any limit at $x=c$.\slug

\proclaim Problem \advthm. Let $f:\RR\to\RR$ be defined by
$$f(x) = \cases{1/q,& {\rm if $x = p/q\in \QQ$ with $p\in \ZZ$, $q\in \NN$,
and $\gcd(p,q) = 1$;}\cr
0,& {\rm if $x\in \RR\setminus\QQ$.}}$$
(This function is called {\it Thomae's function}, or the {\it modified Dirichlet function},
or the {\it stars over Babylon}.)
\medskip
\item{a)} Show that $f$ is periodic with period $1$; that is, $f(x+n) = f(x)$ for all
integers $n$ and all $x\in \RR$.
\smallskip
\item{b)} Show that for any $c\in \QQ$, we have $\lim_{x\to c} f(x)\ne f(c)$.
\smallskip
\item{c)} Show that for any $s\in \RR\setminus\QQ$ one has $\lim_{x\to s} f(x) = f(s)$.
\medskip

\proof
Recall that the sum of an irrational number and a rational number is irrational,
and that the quotient of an irrational number by a rational number is also irrational.

Let $n\in \ZZ$ and $x\in \RR$. If $x$ is irrational, then $x+n$ is also irrational,
and $f(x+n) = 0 = f(x)$. If $x$ is rational, we may express $x = p/q$ with $p\in \ZZ$,
$q\in \NN$, and $\gcd(p,q) = 1$, so that $f(x) = 1/q$. Then
$$ x+n = {p\over q} + n = {p+nq\over q}.$$
If we can show that $\gcd(p+nq,q) = 1$, then $f(x+n) = q$ and we are done.
Suppose that $d$ divides both $p$ and $q$; say, $p = rd$ and $q = sd$ for some integers
$r$ and $s$. Then
$$p + nq = rd + nsd = (r+ns)d,$$
so $d$ divides $p+nq$ as well. On the other
hand, if $d$ divides both $p+nq$ and $q$; say $p+nq = rd$ and $q = sd$ for some $r,s\in \ZZ$.
Then
$$p = (p+nq) - nq = rd - nsd = (r-ns)d,$$
so $d$ divides $p$ as well. We have shown that
the common divisors of $p$ and $q$ are exactly the common divisors of
$p+nq$ and $q$. So $\gcd(p+nq,q) = \gcd(p,q) = 1$.

Let $c\in \QQ$ be arbitrary and express $c = p/q$, where $p$ and $q$ are integers
with $q> 0$ and $\gcd(p,q) = 1$. We have $f(c) = 1/q$ by the definition of $f$.
The claim is that $\lim_{x\to c} f(x) \ne 1/q$.
So we must show that there exists an $\eps$ such that
for all $\delta > 0$, there exists $x\in \RR\setminus \{c\}$ with $|x-c| < \delta$
and $\bigl| f(x) - 1/q \bigr| \ge \eps$.
Fix any positive irrational number $\alpha$. We pick $\eps = 1/q$ and let
$\delta > 0$ be arbitrary. Using the Archimedean property, choose $n\in \NN$ with
$n>\alpha/\delta$ (so that $\alpha/n<\delta$), and set
$$x = c + {\alpha\over n}.$$
From the observation in the first paragraph of this proof,
we see that $x$ is irrational, so $f(x) = 0$ and
$\bigl| f(x) - 1/q\bigl| = 1/q \ge \eps$. On the other hand, we have
$$|x-c| = \biggl| {\alpha\over n}\biggr| < \delta,$$
so we conclude that the limit $\lim_{x\to c} f(x)$ does not exist.

Now let $s$ be irrational, so that $f(s) = 0$. The claim
is that $\lim_{x\to s} f(x)= 0$.
Observe that $s = t+n$ for some integer $n$ and some $t\in (0,1)$,
and if $\lim_{x\to t} f(x) = 0$ then
$$\lim_{x\to s} f(x) = \lim_{x\to t} f(x+n) = \lim_{x\to t} f(x) = L$$
as well.
So to show that $\lim_{x\to s} f(x)= 0$,
it suffices to show that $\lim_{x\to t} f(x)$.
Let $\eps>0$ and pick $m\in \NN$ with $1/m <\eps$ using the Archimedean property.
For each $i\in \{1,\ldots,m\}$, let $k_i$ be the integer with
$$0 < {k_i\over i} < t < {k_i+1\over i}.$$
For all $1\le i\le m$, let
$d_i$ be the minimal distance between $s$ and either $k_i/i$ or $(k_i+1)/i$; that is,
$$ d_i = \min\biggl\{ \biggl| t - {k_i\over i} \biggr|, \biggl| t - {k_i + 1\over i}\biggr|
\biggr\}.$$
Note that $d_i > 0$ for all $1\le i\le m$, so if we set
$$\delta = \{d_1, \ldots, d_m\},$$
then $\delta > 0$, and for all $1\le i\le m$, $|t - k_i/i|\ge \delta$ and
$\bigl| t- (k_i + 1)/i\bigr| \ge \delta$. In other words, for all $1\le i\le m$,
both of the rational
numbers $k_i/i$ and $(k_i+1)/i$ are outside the ball $V_\delta(s)$. What this means
is that any rational number in $V_\delta(t)$ must have a denominator
greater than $m$. Hence for any $x\in \RR$ with $|x-t| < \delta$, either $x$ is
irrational, in which case
$$\bigl| f(x) - 0 \bigr| = |0-0| = 0 < \eps,$$
or $x$ is rational, in which case $x$ can be written as $x = p/q$ with $p\in \ZZ$,
$q > m$, and $\gcd(p,q) = 1$. In this second case,
$$\bigl|f(x) - 0\bigr| = \biggl| {1\over q} - 0\biggr| = {1\over q} < {1\over m}
\le \eps,$$
and we have shown that $\lim_{x\to t} f(x) = 0$, as desired.\slug

\bye

### 2025-11-28 teaching/math242f25/tut11.tex

\input fontmac
\input mathmac

\def\To{\Rightarrow}

\maketitle{Math 242 Tutorial 11}{prepared by}{Marcel Goh}{27 November 2025}

\bigskip

\proclaim Problem \advthm. Let $D\subseteq \RR$ and let $f,g:D\to \RR$ be functions
that are continuous on all of $D$.
\medskip
\item{a)} Show that the function $f-g$ is continuous on all of $D$.
\smallskip
\item{b)} Suppose further that for all $x\in D$ we also have $-x\in D$. (So $D$ is symmetric
about the origin.) Let $h:D\to \RR$ be given by $h(x) = f(-x)$. Show that $h$ is continuous
on all of $D$.
\medskip

\proof Let $p\in D$. Since $f$ and $g$ are both continuous at $p$, we have
$\lim_{x\to p} f(x) = f(p)$ and $\lim_{x\to p} g(x) = g(p)$. Hence
$$\lim_{x\to p} (f-g)(x) = \lim_{x\to p} f(x) - \lim_{x\to p} g(x) = f(p) - g(p),$$
by the algebraic laws for limits of functions.

Now we prove part (b).
Let $\eps > 0$ and let $p\in D$. Since $-p\in D$ and $f$ is continuous on all of $D$,
$f$ is continuous at $-p$. Hence there exists
some $\delta > 0$ such that for all $y\in D$ with $|y-p|< \delta$, one has
$\bigl| f(y) - f(-p)\bigr| < \eps$. Let $x\in D$ be such that $|x-p|<\delta$. Then
$$\bigl|-x-(-p)\bigr| = \bigl|-(x-p)\bigr| = |x-p| < \delta,$$
so letting $y=-x$ above gives tells us that
$$\bigl| h(x) - h(p)\bigr| = \bigl| f(-x) - f(-p)\bigr| < \eps.$$
This shows that $h$ is continuous at $p$, and since $p$ was selected arbitrarily, we conclude that
$h$ is continuous on all of $D$.\slug

These statements generalise to $D\subseteq \RR^n$, a fact you may use in the following problem.

\proclaim Problem \advthm. Let $S^2$ denote a two-dimensional sphere in $\RR^3$,
which can be shifted and scaled so that
$$S^2 = \bigl\{ (x,y,z)\in \RR^3 : x^2 + y^2 + z^2 = 1\bigr\}.$$
Let $f:S^2 \to \RR$ be continuous.
Using the location of roots theorem, show that there exists a point $p\in S^2$
such that $f(-p) = f(p)$.

\proof The set $S^2$, as we have defined it above, is symmetric about the origin of $\RR^3$,
so by the previous problem, the function $g : S^2 \to \RR$ given by $g(v) = f(v)  - f(-v)$ is
continuous. Now fix any point $v\in S^2$ and consider any semicircular arc in $S^2$ from $v$ to $-v$.
Let $\gamma : [0,1]\to \RR^3$ parameterise this arc, so that $\gamma(0) = v$ and $\gamma(1) = -v$ (and
say, $\gamma(1/3)$ is one-third of the way from $v$ to $-v$ along this arc). This is a continuous
function (this technically needs to be proved, but for the purposes of this class, let's just
take it as fact).

Let $h:[0,1]\to \RR$ be given by $h(t) = g\bigl(\gamma(t)\bigr)$. This is continuous, since
it is the composition of continuous functions. First, note that
$$h(0) = g\bigl(\gamma(0)\bigr) = f(v) - f(-v) = -\bigl(f(-v) - f(v)\bigr)
= -g\bigl(\gamma(1)\bigr).$$
If $h(0) = 0$, then $g\bigl(\gamma(0)\bigr) = 0$,
meaning that $g(v) = 0$; that is, $f(v) = f(-v)$, so we can set $p=v$ and we are done.
If $h(0)\ne 0$, then $h(1)\ne 0$ also, and they have opposite signs. Hence by the location
of roots theorem, there is some $t\in (0,1)$ with $h(p) = 0$. Let $p = \gamma(t)$; we have
$$0 = g\bigl(\gamma(t)\bigr) = f(p) - f(-p),$$
which means that $f(p) = f(-p)$, which is what we wanted.\slug

The previous problem shows that if we assume air temperature to be a continuous function on the globe,
then there is some point on Earth that has the same temperature as its antipode (the opposite point
on the globe). In fact, using
a stronger topological
theorem called the {\it Borsuk--Ulam theorem}, one can show that the same thing holds for
{\it pairs} of continuous functions on the sphere. For instance, if we assume barometric pressure
to be continuous on the sphere as well, then there is a point on Earth with the same
air temperature {\it and} the same barometric pressure as its antipode.

\proclaim Problem \advthm. Prove that $\RR$ has the least upper bound property using the
localization of roots theorem (and without using anything else that we have shown to be equivalent
to the completeness axiom).

\proof Suppose, for a contradiction, that there is some $S\subseteq\RR$ that is nonempty and
does not have a least upper bound. We define a function $f:\RR\to \RR$ by
$$f(x) = \cases{1, & if $f$ is an upper bound of $S$;\cr -1, & if $f$ is not an upper bound of $S$.}$$
First we show that $f$ is continuous at every point $c\in \RR$. To do this, let $c\in \RR$
and let $\eps > 0$. There are two cases.

If $f(c) = 1$, then $c$ is an upper bound of $S$. But $S$
does not have a least upper bound, so there is some $b<c$ such that $b$ is also an upper bound of
$c$. In this case, set $\delta = c-b > 0$. For all $x$ with $|x-c|<\delta$, we have $x > b$, so $x$
is an upper bound of $S$ and $f(x) = 1$. We see that
$$\bigl| f(x) - f(c)\bigr| = |1-1| = 0<\eps.$$

If, on the other hand, $f(c) = -1$, then $c$ is not an upper bound of $S$, and there is some $s\in S$
with $s>c$. In this case, set $\delta = s-c > 0$. Then for all $x$ with $|x-c| <\delta$, we must
have $x < s$, so $x$ is not an upper bound of $S$ and $f(x) = 0$. Hence
$$\bigl| f(x) - f(c)\bigr| = \bigl|-1 - (-1)\bigr| = 0 < \eps.$$
We have shown that $f$ is continuous on all of $\RR$.

Now let $s\in S$ and let $t$ be any upper bound of $S$ (so we must have $s\le t)$.
The real number $s-1$ is not an upper bound of $S$, so $f(s-1) = -1$. On the other hand,
$f(t) = 1$. So by the location of roots theorem, there must be some $x\in [s-1, t]$ with
$f(x) = 0$. But this is a contradiction, since $f(x) \in \{-1,1\}$ for all $x\in \RR$.\slug

\proclaim Problem \advthm. Show that a set $f:\RR\to\RR$ is continuous on all of $\RR$
if and only if for every open set $U\subseteq \RR$, the inverse image $f^{-1}(U)$ is open.

\proof Suppose that $f$ is continuous and let $U$ be an open set. We want to show that
$f^{-1}(U)$ is open. Let $x\in f^{-1}(U)$, so that $f(x) \in U$. Since $U$ is open,
there exists some $\eps > 0$ such that $V_\eps\bigl( f(x)\bigr) \subseteq U$. But using this
$\eps$ in the definition of continuity of $f$, there must be some $\delta > 0$
such that for all $y\in V_\delta(x)$, $f(y)$ is in $V_\eps\bigl(f(x)\bigr)$, which is a subset of $U$.
But this implies that $V_\delta(x)\subseteq f^{-1} U$. Since $x$ was arbitrary, we have shown
that $f^{-1}(U)$ is open.

Now suppose that $f:\RR\to\RR$ is a function such that for every open set $U\subseteq \RR$ the set
$f^{-1}(U)$ is also open. Let $x\in \RR$ and $\eps > 0$. The set $U = V_\eps\bigl( f(x)\bigr)$ is open,
so by hypothesis its inverse image $f^{-1}(U)$ is open. Of course, $x$ is in $f^{-1} (U)$,
so there is some $\delta > 0$ such that $V_\delta(x) \subseteq f^{-1}(U)$. Unpacking the definitions
of $V_\delta(x)$ and $U = V_\eps\bigl(f(x)\bigr)$, we see that for this choice of $\delta$, we have
$\bigl| f(x) - f(z)\bigr| < \eps$ whenever $|x-z|<\delta$. Hence $f$ is continuous.\slug

This problem justifies the definition of continuity over $\RR$ that is used in the class, since
over general topological spaces a function is {\it defined} to be continuous if the inverse
image of every open set is open.

\proclaim Problem \advthm. Let $f:\RR\to \RR$ be a continuous function and let $p\in \RR$.
Show that the set
$$S = \bigl\{ x\in \RR : f(x) = p\bigr\}$$
is closed.

\proof The set $\{p\} = [p,p]$ is closed, so its complement $U = \RR\setminus\{p\}$ is open.
Note that $\RR\setminus S = f^{-1}(U)$, so by the previous problem, $\RR\setminus S$ is open.
Hence $S$ is closed.\slug

\bye

### 2025-12-08 about/vita.tex
Updated various things

@@
 \goodbreak
+\sectheader{Invited Talks}
 
+\begingroup\frenchspacing\parindent=10pt
+\thing Block complexity and idempotent Schur multipliers. 7 December 2025, CMS Winter Meeting,
+Toronto.
+\endgroup
+
+\goodbreak
 \sectheader{Published Creative Writing {\rm($*$ indicates nonfiction)}}
 
-\fictioncount=4
+\fictioncount=5
 \pubbegin F
+\ficitem ``Syazwani,'' {\sl The Ex-Puritan} {\bf 71} (2025), online.
 \starficitem ``Lanterns,'' {\sl The Malahat Review} {\bf 229} (2025), 7--13.
@@
 \smallskip
-Founding member and one of two editors-in-chief of
-a semiannual print magazine that publishes poetry, prose, and art.
-In charge of English short fiction submissions and the overall typesetting. Share duties with the other
-editors related to emails, sales, and magazine launches.
+Founding member and one of two directors of a non-profit organization
+that publishes poetry, prose, and art in a semiannual print magazine.
+In charge of English short fiction submissions as well as the overall typesetting.
+Share duties with other editors related to emails, sales, and magazine launches.
 \medbreak
